### 2016-06-02 abstract.tex
Abstract.

\begin{abstract}
ACID model of database transactions advocates serializability as the
correctness criterion for concurrent transactions. 
% Serializability is an easily-stated property,
% providing database programmers with a simple
% implementation-independent model to reason about program invariants in
% presence of concurrent transactions. Unfortunately, 
While serializability is an easily-stated high-level property,
enforcing serializability in a database severely restricts
concurrency.
% , thus limiting the availability and performance
% (latency/throughput) of the data store.  In the database parlance,
% serializibility is \emph{not highly available}~\cite{bailishotos13},
% making it particularly unsuitable for modern-day distributed web
% applications that aspire to be \emph{always on}.
Pragmatic concerns has led to the development of various weaker
alternatives to serializability, which are overwhelmingly
preferred in practice.
% some of which have been
% standardized by the community as \emph{ANSI SQL isolation
% levels}~\cite{berenson}. Indeed, weakly isolated transactions are
% overwhelmingly preferred to serializability in practice, to the extent
% that the default isolation level in popular industrial-strength
% databases is often weaker than serializability.
Unlike serializability, the semantics of weaker isolation levels are
often described informally in terms of an implementation. While there
have been some attempts at formalizing the semantics of weak
isolation, there nonetheless exists a wide chasm between formal models
of weak isolation and the operational models of programs, hindering 
uniform reasoning about program invariants in presence of weakly
isolated transactions.

In this paper, we propose an operational model and a compositional
reasoning framework for concurrent imperative programs with weakly
isolated transactions (\txnimp programs, as we call them).  Our
operational model captures all the behaviors exhibited by a program
under weak isolation without referring to any specific implementation
of isolation. Our reasoning framework enables formal reasoning about
program invariants while taking into account the fine-grained
isolation semantics of transactions. We use a combination of simple
static analysis and manual proofs in our reasoning framework to
formally establish the correctness for several examples, which were
hitherto known to be correct only through experiments.
\end{abstract}

### 2016-06-04 abstract.tex
Examples added.

@@
 While serializability is an easily-stated high-level property,
-enforcing serializability in a database severely restricts
+enforcing it requires the database to severely restrict
 concurrency.
@@
 % applications that aspire to be \emph{always on}.
-Pragmatic concerns has led to the development of various weaker
-alternatives to serializability, which are overwhelmingly
-preferred in practice.
+Pragmatic concerns led to the development of various weaker
+alternatives to serializability, which are overwhelmingly preferred in
+practice.
 % some of which have been
@@
 % databases is often weaker than serializability.
-Unlike serializability, the semantics of weaker isolation levels are
-often described informally in terms of an implementation. While there
-have been some attempts at formalizing the semantics of weak
-isolation, there nonetheless exists a wide chasm between formal models
-of weak isolation and the operational models of programs, hindering 
-uniform reasoning about program invariants in presence of weakly
-isolated transactions.
+However, unlike serializability, the semantics of weaker isolation
+levels are often described informally in terms of an implementation.
+While there have been some attempts at formalization, there
+nonetheless exists a wide chasm between formal models of weak
+isolation and the operational models of programs, hindering uniform
+reasoning about program invariants in presence of weakly isolated
+transactions.
 
@@
 static analysis and manual proofs in our reasoning framework to
-formally establish the correctness for several examples, which were
+formally establish the correctness for several benchmarks, which were
 hitherto known to be correct only through experiments.

### 2016-06-04 motivation.tex
Examples added.

\section{Motivation}

\begin{figure}
\centering
$\{\{\texttt{C}=\texttt{k} \conj \texttt{k}\ge\texttt{a1+a2}\}\}$
\begin{tabular}{l||l}
\begin{txnimpcode}
  txn$\langle$'Wd(C,a1)'$\rangle${
    if (C $\ge$ a1) {
      C := C - a1
    }
  }
\end{txnimpcode}
&
\begin{txnimpcode}
  txn$\langle$'Wd(C,a2)'$\rangle${
    if (C $\ge$ a2) {
      C := C - a2
    }
  }
\end{txnimpcode}
\\
\end{tabular}
$\{\{\texttt{C}=\texttt{k-a1-a2}\}\}$

\caption{Concurrent withdraw transactions}
\label{fig:motiv-eg-1}
\end{figure}

Consider an implementation of a banking application that admits
concurrent withdraw transactions on a checking account (\C{C}), as
shown in Fig.~\ref{fig:motiv-eg-1}. If the initial balance (\C{k}) in
the account is enough to perform both withdraws, then the final
balance, after both transactions commit, is expected to reflect the
effects of both withdraws. The pre and post conditions in
Fig.~\ref{fig:motiv-eg-1} (written in Hoare style) reflect our
expectations. Indeed, invariants are guaranteed to hold if both
withdraw transactions are serialized, making \iso{Serializable}
isolation level a sufficent condition to preserve invaraints. But, is
\iso{Serializable} (SER) isolation necessary?

% Most weak isolation levels allow executions to interleave operations
% (Reads, Writes and Commits) of one transaction with other's, while
% enforcing some additional constraints. Arbitrary interleavings may
% lead to the violation of invaraints. 
Let us consider \iso{Read Committed} (RC), an isolation level weaker
than SER, but is nonetheless the default level in Postgres
9.5~\cite{postgres95} and Oracle 11g~\cite{oracle11g} databases. An RC
transaction is shielded from reading the writes of an uncommitted
transactions, thus preventing the so called \emph{dirty reads}
phenomenon~\cite{berenson}. In the current example, RC isolation
admits following execution traces:\GK{My aim here is to lead the
reader into thinking about executions in terms of implementation
artifacts so that she appreciates the full complexity of the problem.}
% the default isolation level on ,

\begin{tabular}{|p{1.2in}|p{1.2in}|}
\hline
RC Execution 1 & RC Execution 2 \\
\hline
\readf('Wd1',C,k)
\newline
\readf('Wd1',C,k)
\newline
\writef('Wd1',C,k-a1)
\newline
\readf('Wd2',C,k)
\newline
\readf('Wd2',C,k)
\newline
\writef('Wd2',C,k-a2)
\newline
\commitf('Wd2')
\newline
\commitf('Wd1')
&
\readf('Wd2',C,k)
\newline
\readf('Wd2',C,k)
\newline
\writef('Wd2',C,k-a2)
\newline
\readf('Wd1',C,k)
\newline
\readf('Wd1',C,k)
\newline
\writef('Wd1',C,k-a1)
\newline
\commitf('Wd1')
\newline
\commitf('Wd2')\\
\hline
\end{tabular}

Each execution trace is a series of \emph{effects}, where an effect
documents the completion of either a Read, Write or a Commit operation
of a transaction. Effects are tagged by their respective transaction
identifiers (\eg 'Wd1'), associated data objects (\eg \C{C}), and
values read or written (\eg k-a1), if any. In the first trace,
transaction 'Wd1' reads the current balance (\C{k}) and writes the new
balance (\C{k-a1}), but before it commits transaction 'Wd2' executes
and commits, writing the new balance (\C{k-a2}). RC isolation has
prevented 'Wd2' from witnessing the writes of uncommitted transaction
'Wd1'. Committing 'Wd1' now leads to the loss of updates from 'Wd2'
(the so called \emph{lost update anomaly}~\cite{berenson}), resulting
in a balance of \C{k-a1}. The second execution trace describes a
similar scenario with 'Wd1' and 'Wd2' exchanging their roles.

Clearly, \iso{Read Committed} is not sufficient because it loses the
updates of one transaction and violates the invariant in
Fig.~\ref{fig:motiv-eg-1}. Now, consider an alternative unfolding of
Execution 1, where, rather than committing 'Wd1' and overwriting the
updates of 'Wd2', we rollback 'Wd1' to reexecute it starting from the
state that contains the updates of 'Wd2'. No updates are lost in this
alternative execution, leading to the satisfaction of
Fig.~\ref{fig:motiv-eg-1} invariant. Rolling back and reexecuting an
uncommitted transaction to prevent write-write conflicts with a
committed transaction is a strategy often adopted by databases to
implement \iso{Snapshot Isolation} (SI) level. Indeed, SI is
sufficient to enforce the required invaraints in this
example\footnote{\iso{Parallel Snapshot Isolation} (PSI), a weaker
version of SI tailor-made for replicated stores is also sufficient.}.
Moreover, relying on optimistic speculation instead of pessimistic
locking for concurrency control makes SI more efficient than SER,
making it appropriate for transactions in Fig.~\ref{fig:motiv-eg-1}.
Note that we have arrived at his judgment via adhoc reasoning based on
speculations about database's implementation strategies. Such a
reasoning process is necessarily fragile and error-prone. Is there a
better way?

\begin{figure}
\centering
$\{\{\texttt{C+S}\ge\texttt{0}\}\}$
\begin{tabular}{l||l}
\begin{txnimpcode}
  txn$\langle$'WdC'$\rangle${
    if (C+S $\ge$ a1) {
      C := C - a1
    }
  }
\end{txnimpcode}
&
\begin{txnimpcode}
  txn$\langle$'WdS'$\rangle${
    if (C+S $\ge$ a2) {
      S := S - a2
    }
  }
\end{txnimpcode}
\\
\end{tabular}
$\{\{\texttt{C+S}\ge\texttt{0}\}\}$

\caption{Withdraws from current (\C{C}) and savings (\C{S}) accounts.}
\label{fig:motiv-eg-2}
\end{figure}

\iso{Snapshot Isolation} effectively serializes updates to a data
object without necessarily relying on expensive lock-based concurrency
control. While this makes it a better alternative to
\iso{Serializable} isolation, it is however not a replacement for the
latter. Consider a different banking application that maintains a
current (\C{C}) and savings account (\C{S}) for each user. A user can
withdraw either from current or from savings account as long as the
combined balance is at least as much as the amount being withdrawn.
Fig.~\ref{fig:motiv-eg-2} shows a pair of concurrent transactions,
'WdC' and 'WdS', performing withdraws from current and savings
accounts, respectively. Pre and post conditions assert non-negative
combined balance invariant. \iso{Serializable} isolation is sufficient
to preserve the invariant, but \iso{Snapshot Isolation} is not.
Intuitively, this is because there are no \emph{lost updates} in this
example even when both transactions commit concurrently, hence there
is no reason for SI to effectively serialize them. A reasoning
framework for weak isolation should let us formally deduce this fact
by making it impossible to construct a proof of correctness for the
program under SI, but facilitating a proof under SER.

\begin{figure}[t]
\centering
$\{\{\texttt{C+S}\ge\texttt{0}\}\}$
\begin{tabular}{l||l||l}
\begin{txnimpcode}
  txn$\langle$'Xfer'$\rangle${
    C := C - a;
    S := S + a;
  }
\end{txnimpcode}
&
\begin{txnimpcode}
  txn$\langle$'TotBal'$\rangle${
    T := S + C
  }
\end{txnimpcode}
&
\begin{txnimpcode}
  txn$\langle$'AddInt'$\rangle${
    S := S + (T*0.1)
  }
\end{txnimpcode}
\\
\end{tabular}
$\{\{\texttt{C+S}\ge\texttt{0}\}\}$

\caption{Concurrent transfer (\C{Xf}) and interest accumulation (\C{Int}).}
\label{fig:motiv-eg-3}
\end{figure}

The previous two examples preserve invariants on data objects by
effectively serializing transactions that write to those data objects.
However, in general, this is neither sufficient nor necessary. Consider
the banking application in Fig.~\ref{fig:motiv-eg-3} with three
transactions. Transaction 'Xfer' transfers some amount (\C{a}) from
current (\C{C}) to savings (\C{S}) account.  Transaction 'TotBal'
records the total balance in both the accounts in a data object \C{T},
and transaction 'AddIntr' calculates interest on total balance (10\%
of \C{T}) and adds it to the savings account. Using a data object to
record some function of the current state of other data objects is a
standard practice in relational databases, where Materialized
Views~\cite{oraclematview} are created to maintain the results of a
SQL query. Like the previous example, the invariant of interest in
this example is non-negative total balance; if we start with a
non-negative total balance, and we add interest based on the total
balance, we must end up with a non-negative total balance.  However,
unlike the previous example, serializing transactions that write to
\C{C} and \C{S} is not sufficient to guarantee the invariant.  The
following execution demonstrates this anomaly:

Assume that \C{C} and \C{S} are initially 10 and 0, respectively. Let
\C{a} be 200, i.e., we subtract 200 from \C{C} and add it to \C{S}.
This operation must be allowed because it doesn't effect the total
balance, hence the interest. The execution begins with 'TotBal'
transaction reading the savings (\C{S}) balance as 0 (we assume
left-to-right evaluation order for \C{S + C}). However, before it
reads the current (\C{C}) balance, 'Xfer' transaction executes and
commits, changing \C{C} to -190. Subsequently, 'TotBal' reads -190
from \C{C}, and updates \C{T} to -190. 'AddInt', which begins
execution after 'TotBal', reads -190 from \C{T}, calculates interest
as -19, and updates \C{S} to 181. The result is a state with total
balance as -190 + 181 = -9, thus violating the invariant.

Note that transactions 'Xfer' and 'AddInt', which write to the data
objects (\C{C} and \C{S}) relevant to the invariant, are serialized in
the above execution. Transaction 'TotBal' however is not. While
serializing it fixes the problem, it is not necessary. A simpler fix
is to isolate 'TotBal' transaction from the effects of concurrent
'Xfer' transaction, so that it reads \C{C} as 10 despite the
concurrent 'Xfer' transaction committing -190 to \C{C}. This can be
achieved by executing 'TotBal' under \iso{Repeatable Read} (RR)
isolation level, which guarantees that any two reads of a data object
in a transaction return the same value (the \emph{repeatable reads}
property~\cite{berenson}). Allowing 'Xfer' to interfere with 'TotBal'
violates the repeatable read property since a hypothetical read from
\C{S} in 'TotBal' after the interference returns a different value
(200) than the first read (0). Executing 'TotBal' under RR restores
repeatable read by \emph{effectively} prohibiting such interference.

\begin{figure}[t]
\centering
$\{\{\texttt{V=0}\}\}$
\begin{tabular}{l||l||l}
\begin{txnimpcode}
  txn$\langle$'Rd'$\rangle${
    X := V
    Y := V
  }
\end{txnimpcode}
&
\begin{txnimpcode}
  txn$\langle$'Inc1'$\rangle${
    V := V + 1
  }
\end{txnimpcode}
&
\begin{txnimpcode}
  txn$\langle$'Inc2'$\rangle${
    V := V + 1
  }
\end{txnimpcode}
\\
\end{tabular}
$\{\{\texttt{Y}\ge\texttt{X}\}\}$

\caption{View counter (\C{V}) increments (\C{Inc}) and reads (\C{Rd})}
\label{fig:motiv-eg-4}
\end{figure}

Our last example demonstrates that it is not necessary to serialize
transactions, even those that write to the same data objects, to
maintain essential invariants in practical applications. Consider a
counter used to count video views on a YouTube-like application. The
counter need not accurately track the view count, but it must at least
ensure that the count does not decrease (nor does it appear to be
decreasing). Fig.~\ref{fig:motiv-eg-4} shows a simple example
involving a view counter (\C{V}). The three 'Inc' transactions
('Inc[1]' to 'Inc[3]') increment the counter, while the transaction
'Rd' reads the counter twice and writes the values read to \C{X} and
\C{Y}, respectively. The count is 0 initially. Since we expect the
view count to monotonically increase, we require $\C{Y}\ge\C{X}$ at
the end. Under the standard setting, where new writes to data objects
always overwrite previous writes, the postcondition may not hold. For
instance, consider an execution where 'Inc[1]' reads \C{V} as 0 (but
doesn't commit), then 'Inc[2]' and 'Inc[3]' commit serially changing
\C{V} to 2, followed by 'Rd' reading 2 and writing it to \C{X}. Now
'Inc[1]', which has read \C{V} as 0 writes 1 to \C{V} and commits. The
subsequent read of \C{V} by 'Rd' returns 1, which it writes to \C{Y}
and commits. In the final state, \C{X} is 2 and \C{Y} is 1, thus
violating the monotonicity invariant.

However, in an unconventional setting, such as a modern day replicated
data store, writes arriving at a replica need not necessarily
overwrite the existing writes. A \emph{conflict resolution strategy}
may be employed to pick, for example, a write with the largest value
among competing concurrent writes. Under this setting, the write to
\C{V} by 'Inc[1]' with a value (1) less than the current value (2) is
discarded. Transaction 'Rd' therefore continues to read 2, allowing it
to satisfy the postcondition. Note that we did not have to serialize
'Inc' transactions to uphold the monotonicity invariant, although they
are involved in Write-Write conflicts.  The default \iso{Read
Committed} isolation level suffices\footnote{We have implicitly assumed that
'Rd' transaction witnesses monotonically growing state, i.e., the
transactions that appear committed to the first read of \C{V} in 'Rd'
also appear committed to the second read.  Most implementations of RC
enforce this \emph{Monotonic View} invariant. However,
some~\cite{pldi15,bailishat} categorize RC with Monotonic View as a
separate isolation level called \iso{Monotonic Atomic View} (MAV). On
those stores, we need MAV for 'Rd'. }, provided that the conflict
resolution strategy consistently picks writes with larger values. A
reasoning framework for weak isolation must be sensitive to the
semantics of conflict resolution in order to formally prove the
monotonicity invariant in this example.

### 2016-06-04 motivation.tex
Figures added.

@@
 \section{Motivation}
+
+In this section, we motivate the need for a formal reasoning framework
+for weakly isolated transactions through a series of examples. All our
+examples are written in C-like imperative language equipped with a
+\C{txn} lexical block for transactions. Each \C{txn} block is
+associated with a single-quoted string in angle braces that uniquely
+identifies the transaction. We use Hoare triple notation to annotate
+programs with pre and post conditions.
 
@@
 \begin{txnimpcode}
-  txn$\langle$'Wd(C,a1)'$\rangle${
+  txn$\langle$'Wd1'$\rangle${
     if (C $\ge$ a1) {
@@
 \begin{txnimpcode}
-  txn$\langle$'Wd(C,a2)'$\rangle${
+  txn$\langle$'Wd2'$\rangle${
     if (C $\ge$ a2) {
@@
 effects of both withdraws. The pre and post conditions in
-Fig.~\ref{fig:motiv-eg-1} (written in Hoare style) reflect our
-expectations. Indeed, invariants are guaranteed to hold if both
-withdraw transactions are serialized, making \iso{Serializable}
-isolation level a sufficent condition to preserve invaraints. But, is
-\iso{Serializable} (SER) isolation necessary?
+Fig.~\ref{fig:motiv-eg-1} reflect our expectations. Indeed, invariants
+are guaranteed to hold if both withdraw transactions are serialized,
+making \iso{Serializable} isolation (SER) level a sufficent condition
+to preserve invaraints. But, is SER necessary?
 
@@
 Let us consider \iso{Read Committed} (RC), an isolation level weaker
-than SER, but is nonetheless the default level in Postgres
-9.5~\cite{postgres95} and Oracle 11g~\cite{oracle11g} databases. An RC
-transaction is shielded from reading the writes of an uncommitted
-transactions, thus preventing the so called \emph{dirty reads}
-phenomenon~\cite{berenson}. In the current example, RC isolation
-admits following execution traces:\GK{My aim here is to lead the
-reader into thinking about executions in terms of implementation
-artifacts so that she appreciates the full complexity of the problem.}
-% the default isolation level on ,
+than SER that is default in Postgres 9.5~\cite{postgres95} and Oracle
+11g~\cite{oracle11g} databases. An RC transaction is isolated from the
+writes of an uncommitted transactions, thus preventing the so called
+\emph{dirty reads} phenomenon~\cite{berenson}. In the current example,
+RC isolation admits following executions:
 
-\begin{tabular}{|p{1.2in}|p{1.2in}|}
-\hline
-RC Execution 1 & RC Execution 2 \\
-\hline
-\readf('Wd1',C,k)
-\newline
-\readf('Wd1',C,k)
-\newline
-\writef('Wd1',C,k-a1)
-\newline
-\readf('Wd2',C,k)
-\newline
-\readf('Wd2',C,k)
-\newline
-\writef('Wd2',C,k-a2)
-\newline
-\commitf('Wd2')
-\newline
-\commitf('Wd1')
-&
-\readf('Wd2',C,k)
-\newline
-\readf('Wd2',C,k)
-\newline
-\writef('Wd2',C,k-a2)
-\newline
-\readf('Wd1',C,k)
-\newline
-\readf('Wd1',C,k)
-\newline
-\writef('Wd1',C,k-a1)
-\newline
-\commitf('Wd1')
-\newline
-\commitf('Wd2')\\
-\hline
-\end{tabular}
+\begin{figure}[!h]
+\centering
+\subcaptionbox {
+  RC Execution 1
+  \label{fig:motiv-eg-1-a}
+} [
+  0.55\columnwidth
+] {
+  \includegraphics[scale=0.5]{Figures/motiv-eg-1-a}
+}
+%\hspace*{0.5in}
+\subcaptionbox {
+  RC Execution 2
+  \label{fig:motiv-eg-1-b}
+}{
+  \includegraphics[scale=0.5]{Figures/motiv-eg-1-b}
+}
+\end{figure}
 
-Each execution trace is a series of \emph{effects}, where an effect
-documents the completion of either a Read, Write or a Commit operation
-of a transaction. Effects are tagged by their respective transaction
-identifiers (\eg 'Wd1'), associated data objects (\eg \C{C}), and
-values read or written (\eg k-a1), if any. In the first trace,
-transaction 'Wd1' reads the current balance (\C{k}) and writes the new
-balance (\C{k-a1}), but before it commits transaction 'Wd2' executes
-and commits, writing the new balance (\C{k-a2}). RC isolation has
-prevented 'Wd2' from witnessing the writes of uncommitted transaction
-'Wd1'. Committing 'Wd1' now leads to the loss of updates from 'Wd2'
+An execution is a series of Read, Write and Commit operations.
+Subscript of an operation indicates its transaction. For clarity,
+effects of different transactions are shown against differently
+colored backgrounds. In the first execution, the first transaction
+(`Wd1') reads the current balance (\C{k}) and writes the new balance
+(\C{k-a1}), but before it commits transaction `Wd2' executes and
+commits, writing the new balance (\C{k-a2}). RC isolation has
+prevented `Wd2' from witnessing the writes of uncommitted transaction
+`Wd1'. Committing `Wd1' now leads to the loss of updates from `Wd2'
 (the so called \emph{lost update anomaly}~\cite{berenson}), resulting
-in a balance of \C{k-a1}. The second execution trace describes a
-similar scenario with 'Wd1' and 'Wd2' exchanging their roles.
+in a balance of \C{k-a1}. The second execution describes a similar
+scenario with `Wd1' and `Wd2' exchanging their roles.
 
@@
 Fig.~\ref{fig:motiv-eg-1}. Now, consider an alternative unfolding of
-Execution 1, where, rather than committing 'Wd1' and overwriting the
-updates of 'Wd2', we rollback 'Wd1' to reexecute it starting from the
-state that contains the updates of 'Wd2'. No updates are lost in this
+Execution 1, where, rather than committing `Wd1' and overwriting the
+updates of `Wd2', we rollback `Wd1' to reexecute it starting from the
+state that contains the updates of `Wd2'. No updates are lost in this
 alternative execution, leading to the satisfaction of
@@
 Fig.~\ref{fig:motiv-eg-2} shows a pair of concurrent transactions,
-'WdC' and 'WdS', performing withdraws from current and savings
+`WdC' and `WdS', performing withdraws from current and savings
 accounts, respectively. Pre and post conditions assert non-negative
@@
 $\{\{\texttt{C+S}\ge\texttt{0}\}\}$
-\begin{tabular}{l||l||l}
+\begin{tabular}{l||l||}
 \begin{txnimpcode}
@@
 \end{txnimpcode}
-&
+% \\
+% \multicolumn{2}{c} {
+% }
+\end{tabular}
+\begin{tabular}{c}
   \begin{txnimpcode}
@@
   \end{txnimpcode}
-\\
 \end{tabular}
+
 $\{\{\texttt{C+S}\ge\texttt{0}\}\}$
@@
 
-The previous two examples preserve invariants on data objects by
-effectively serializing transactions that write to those data objects.
-However, in general, this is neither sufficient nor necessary. Consider
-the banking application in Fig.~\ref{fig:motiv-eg-3} with three
-transactions. Transaction 'Xfer' transfers some amount (\C{a}) from
-current (\C{C}) to savings (\C{S}) account.  Transaction 'TotBal'
-records the total balance in both the accounts in a data object \C{T},
-and transaction 'AddIntr' calculates interest on total balance (10\%
-of \C{T}) and adds it to the savings account. Using a data object to
-record some function of the current state of other data objects is a
-standard practice in relational databases, where Materialized
-Views~\cite{oraclematview} are created to maintain the results of a
-SQL query. Like the previous example, the invariant of interest in
-this example is non-negative total balance; if we start with a
-non-negative total balance, and we add interest based on the total
-balance, we must end up with a non-negative total balance.  However,
-unlike the previous example, serializing transactions that write to
-\C{C} and \C{S} is not sufficient to guarantee the invariant.  The
-following execution demonstrates this anomaly:
+The previous two examples preserve invariants on \C{C} and \C{S} by
+effectively serializing transactions that write to \C{C} and \C{S}.
+However, in general, serializing a subset of transactions that write
+to the data objects referred by invariants is neither sufficient nor
+necessary to uphold the invariants. Consider the banking application
+in Fig.~\ref{fig:motiv-eg-3} with three transactions. Transaction
+`Xfer' transfers some amount (\C{a}) from current (\C{C}) to savings
+(\C{S}) account.  Transaction `TotBal' records the total balance in
+both the accounts in a data object \C{T}, and transaction `AddIntr'
+calculates interest on total balance (10\% of \C{T}) and adds it to
+the savings account. Using a data object to record some function of
+the current state of other data objects is a standard practice in
+relational databases, where Materialized Views~\cite{oraclematview}
+are created to maintain the results of a SQL query. Like the previous
+example, the invariant of interest in this example is non-negative
+total balance; if we start with a non-negative total balance, and we
+add interest based on the total balance, we must end up with a
+non-negative total balance.  However, unlike the previous example,
+serializing transactions that write to \C{C} and \C{S} is not
+sufficient to guarantee the invariant.  The following execution
+demonstrates this anomaly:
 
-Assume that \C{C} and \C{S} are initially 10 and 0, respectively. Let
-\C{a} be 200, i.e., we subtract 200 from \C{C} and add it to \C{S}.
-This operation must be allowed because it doesn't effect the total
-balance, hence the interest. The execution begins with 'TotBal'
-transaction reading the savings (\C{S}) balance as 0 (we assume
-left-to-right evaluation order for \C{S + C}). However, before it
-reads the current (\C{C}) balance, 'Xfer' transaction executes and
-commits, changing \C{C} to -190. Subsequently, 'TotBal' reads -190
-from \C{C}, and updates \C{T} to -190. 'AddInt', which begins
-execution after 'TotBal', reads -190 from \C{T}, calculates interest
-as -19, and updates \C{S} to 181. The result is a state with total
-balance as -190 + 181 = -9, thus violating the invariant.
+\begin{figure}[!h]
+\centering
+\includegraphics[scale=0.5]{Figures/motiv-eg-3-a}
 
-Note that transactions 'Xfer' and 'AddInt', which write to the data
+\label{fig:motiv-eg-3-a}
+\end{figure}
+
+The subscripts X, TB and AI denote `Xfer', `TotBal' and `AddInt'
+transactions, respectively. Assume that \C{C} and \C{S} are initially
+10 and 0, respectively. Let \C{a} be 200, i.e., we subtract 200 from
+\C{C} and add it to \C{S}.  This operation must be allowed because it
+doesn't effect the total balance, hence the interest. The execution
+begins with `TotBal' transaction reading the savings (\C{S}) balance
+as 0 (we assume left-to-right evaluation order for \C{S + C}).
+However, before it reads the current (\C{C}) balance, `Xfer'
+transaction executes and commits, changing \C{C} to -190.
+Subsequently, `TotBal' reads -190 from \C{C}, and updates \C{T} to
+-190. `AddInt', which begins execution after `TotBal', reads -190 from
+\C{T}, calculates interest as -19, and updates \C{S} to 181. The
+result is a state with total balance as -190 + 181 = -9, thus
+violating the invariant.
+
+Note that transactions `Xfer' and `AddInt', which write to the data
 objects (\C{C} and \C{S}) relevant to the invariant, are serialized in
-the above execution. Transaction 'TotBal' however is not. While
+the above execution. Transaction `TotBal' however is not. While
 serializing it fixes the problem, it is not necessary. A simpler fix
-is to isolate 'TotBal' transaction from the effects of concurrent
-'Xfer' transaction, so that it reads \C{C} as 10 despite the
-concurrent 'Xfer' transaction committing -190 to \C{C}. This can be
-achieved by executing 'TotBal' under \iso{Repeatable Read} (RR)
+is to isolate `TotBal' transaction from the effects of concurrent
+`Xfer' transaction, so that it reads \C{C} as 10 despite the
+concurrent `Xfer' transaction committing -190 to \C{C}. This can be
+achieved by executing `TotBal' under \iso{Repeatable Read} (RR)
 isolation level, which guarantees that any two reads of a data object
 in a transaction return the same value (the \emph{repeatable reads}
-property~\cite{berenson}). Allowing 'Xfer' to interfere with 'TotBal'
+property~\cite{berenson}). Allowing `Xfer' to interfere with `TotBal'
 violates the repeatable read property since a hypothetical read from
-\C{S} in 'TotBal' after the interference returns a different value
-(200) than the first read (0). Executing 'TotBal' under RR restores
+\C{S} in `TotBal' after the interference returns a different value
+(200) than the first read (0). Executing `TotBal' under RR restores
 repeatable read by \emph{effectively} prohibiting such interference.
@@
 \centering
-$\{\{\texttt{V=0}\}\}$
-\begin{tabular}{l||l||l}
+\hspace*{-0.3in}$\{\{\texttt{V=0}\}\}$
+\begin{tabular}{l||l}
 \begin{txnimpcode}
@@
 \begin{txnimpcode}
-  txn$\langle$'Inc1'$\rangle${
+  cobegin (i $\leftarrow$ [1$\ldots$3]) {
+    txn$\langle$'Inc['+i+']'$\rangle${
       V := V + 1
     }
-\end{txnimpcode}
-&
-\begin{txnimpcode}
-  txn$\langle$'Inc2'$\rangle${
-    V := V + 1
   }
 \end{txnimpcode}
+% &
+% \begin{txnimpcode}
+%   txn$\langle$'Inc2'$\rangle${
+%     V := V + 1
+%   }
+% \end{txnimpcode}
 \\
 \end{tabular}
-$\{\{\texttt{Y}\ge\texttt{X}\}\}$
+\hspace*{-0.3in}$\{\{\texttt{Y}\ge\texttt{X}\}\}$
 
@@
 decreasing). Fig.~\ref{fig:motiv-eg-4} shows a simple example
-involving a view counter (\C{V}). The three 'Inc' transactions
-('Inc[1]' to 'Inc[3]') increment the counter, while the transaction
-'Rd' reads the counter twice and writes the values read to \C{X} and
+involving a view counter (\C{V}). The three `Inc' transactions
+(`Inc[1]' to `Inc[3]') increment the counter, while the transaction
+`Rd' reads the counter twice and writes the values read to \C{X} and
 \C{Y}, respectively. The count is 0 initially. Since we expect the
@@
 always overwrite previous writes, the postcondition may not hold. For
-instance, consider an execution where 'Inc[1]' reads \C{V} as 0 (but
-doesn't commit), then 'Inc[2]' and 'Inc[3]' commit serially changing
-\C{V} to 2, followed by 'Rd' reading 2 and writing it to \C{X}. Now
-'Inc[1]', which has read \C{V} as 0 writes 1 to \C{V} and commits. The
-subsequent read of \C{V} by 'Rd' returns 1, which it writes to \C{Y}
+instance, consider an execution where `Inc[1]' reads \C{V} as 0 (but
+doesn't commit), then `Inc[2]' and `Inc[3]' commit serially changing
+\C{V} to 2, followed by `Rd' reading 2 and writing it to \C{X}. Now
+`Inc[1]', which has read \C{V} as 0 writes 1 to \C{V} and commits. The
+subsequent read of \C{V} by `Rd' returns 1, which it writes to \C{Y}
 and commits. In the final state, \C{X} is 2 and \C{Y} is 1, thus
@@
 among competing concurrent writes. Under this setting, the write to
-\C{V} by 'Inc[1]' with a value (1) less than the current value (2) is
-discarded. Transaction 'Rd' therefore continues to read 2, allowing it
+\C{V} by `Inc[1]' with a value (1) less than the current value (2) is
+discarded. Transaction `Rd' therefore continues to read 2, allowing it
 to satisfy the postcondition. Note that we did not have to serialize
-'Inc' transactions to uphold the monotonicity invariant, although they
+`Inc' transactions to uphold the monotonicity invariant, although they
 are involved in Write-Write conflicts.  The default \iso{Read
 Committed} isolation level suffices\footnote{We have implicitly assumed that
-'Rd' transaction witnesses monotonically growing state, i.e., the
-transactions that appear committed to the first read of \C{V} in 'Rd'
+`Rd' transaction witnesses monotonically growing state, i.e., the
+transactions that appear committed to the first read of \C{V} in `Rd'
 also appear committed to the second read.  Most implementations of RC
@@
 separate isolation level called \iso{Monotonic Atomic View} (MAV). On
-those stores, we need MAV for 'Rd'. }, provided that the conflict
+those stores, we need MAV for `Rd'. }, provided that the conflict
 resolution strategy consistently picks writes with larger values. A

### 2016-06-05 introduction.tex
Edits.

\section{Introduction}

Database transactions allow users to group operations on multiple
objects into a coherent unit with useful properties including
atomicity, consistency, (serializable) isolation and durability
(ACID). ACID has been the \emph{de facto} standard for database
transactions for many decades, providing programmers with a simple
implementation-independent model to reason about concurrent
transactions. However, achieving serializability, the ACID-prescribed
standard for isolation, requires databases to severely restrict
concurrency, which brings down the overall throughput of the system.

### 2016-06-06 introduction.tex
Introduction and bib added.

@@
 atomicity, consistency, (serializable) isolation and durability
-(ACID). ACID has been the \emph{de facto} standard for database
-transactions for many decades, providing programmers with a simple
+(ACID). ACID has been the gold standard for database transactions for
+many decades, providing programmers with a simple
 implementation-independent model to reason about concurrent
-transactions. However, achieving serializability, the ACID-prescribed
-standard for isolation, requires databases to severely restrict
-concurrency, which brings down the overall throughput of the system.
+transactions. However, serializability, the ACID-prescribed standard for
+isolation, does not come for free; it requires databases to rely on
+expensive techniques such as two-phase locking~\cite{twopl,ullmanbook}
+that severely restrict concurrency while incurring an additional
+overhead of deadlock detection, rollback and re-execution. Enforcing
+serializability in a replicated data store requires global
+coordination among geo-distributed replicas, which makes the store
+\emph{unavailable} in presence of the inevitable network
+partitions~\cite{cap,sernotavlbl,bailishat,bernsigmod13}. The
+conflict between serializability and pragmatic goals of high
+peformance and availability has motivated the development of multiple
+weaker forms of transaction isolation beginning as early as
+1976~\cite{gray1976}. ANSI SQL 92 standard defines three such weak
+isolation \emph{levels} which are now implemented in
+many relational and NoSQL databases. Weakly isolated transactions have
+been found to significantly outperform serializable transactions on
+benchmark suites, both on single-node~\cite{dbtuningbook} and
+multi-node replicated stores~\cite{bailishat,bailisvldb}, leading to
+their overwhelming adoption. A 2013 study~\cite{bailishotos} of 18
+popular ACID and ``NewSQL'' databases found that only 3 of them offer
+serializability by default, and half, including Oracle 11g, do not
+offer it at all.  A 2015 study~\cite{bailisferal} of a large corpus of
+database applications finds no evidence of any of them changing the
+default isolation level offered by the database. Together, the studies
+establish the prevalance of weak isolation in practice.
+
+Unfortunately, weak isolation introduces new set of behaviors into
+programs that often give rise to anomalous executions and
+incomprehensible results~\cite{pldi15}. To quantify weak isolation
+anomalies, Fekete et al~\cite{feketevldb09} have devised and
+experimented with a microbenchmark suite that executes transactions
+under \iso{Read Committed} isolation level - the default level for 8
+of the 18 databases studied in~\cite{bailishotos}, and found that 25
+out of every 1000 rows in the database violate at least one integrity
+constraint. Bailis et al~\cite{bailisferal} rely on Rails's
+\emph{uniqueness validation} to maintain uniqueness of records while
+serving Linkbench's~\cite{linkbench} insertion workload (6400 records
+distributed over 1000 keys; 64 concurrent clients), and end up with
+more than 10 duplicate records. Rails relies on database transactions
+to validate uniqueness during insertions, which is ok if transactions
+are serializable, but incorrect under \iso{Read Committed} isolation
+used in the experiments. The same study has found that 13\% of all
+invariants among 67 open source Ruby-on-Rails applications are the
+risk of being violated due to weak isolation. The ill effects of weak
+isolation are therefore as prevalent as weak isolation itself.
+
+The root cause of problems with weak isolation is that its semantics
+in context of user programs is not well-understood. The original
+proposal~\cite{gray1976} defines multiple ``degrees'' of weak
+isolation in terms of implementation details, such as the nature and
+duration of locks held in each case. ANSI SQL 92 standard four levels
+of isolation (including serializability) in terms of sevaral
+undesirable \emph{phenomena} (\eg \emph{dirty reads}) each is required
+to prevent. While this is an improvement, it is hard to reason about
+application semantics in terms of the undesirable phenomena. Moreover,
+a later critique~\cite{berenson} uncovers many ambiguities and
+shortcomings in the standard. Adya's thesis~\cite{adyaphd} includes
+first formal definitions of some of the well-known isolation levels in
+the context of a system model of sequentially consistent database
+complete with timestamps and version histories. However, Adya's model,
+which predates the proliferation of weakly consistent replicated data
+stores, is not enough to express the semantics of a host of new
+isolation levels~\cite{psi,nmsi} tailor-made for such stores, or the
+interplay between weak consistency and isolation. Nonetheless,
+regardless of the specification methodology employed, the major
+missing piece in all the aforementioned approaches is the bridge
+between the system model of specifications and the operational model
+of programs that enables formal reasoning about program invariants in
+presence of weak isolation.
+
+There have been encouraging developments in Programming Languages
+research towards addressing the problem of reasoning about application
+invariants on weakly consistent replicated stores~\cite{burckhardt14,
+redblueosdi, redblueatc, ecinec, gotsmanpopl16}. Weak isolation can
+make use of some of the reasoning inventory built for weak consistency
+(as we demonstrate in later sections), but the latter does not subsume
+the former. While weak consistency (as considered by the
+aforementioned research) deals with lazily propagating effects of
+atomic and fully isolated operations, reasoning about weakly isolated
+transactions involves reasoning about the effects of groups of
+operations that are not necessarily atomic or isolated. Indeed, weakly
+isolated transactions have existed on sequentially consistent
+relational databases long before the advent of weakly consistent
+replicated stores, indicating the subtle interplay between consistency
+and isolation. Moreover, a common theme among the aforementioned line
+of work is the assumption of a system with a predefined set of
+consistency levels, and a minimum consistency level that is often
+stronger than eventual consistency. For instance, the reasoning
+framework proposed in~\cite{gotsmanpopl16} crucially relies on the
+assumption of causal consistency with only admissable strengthening
+being strong consistency. None of the ANSI SQL weak isolation levels
+guarantee causal consistency. Moreover, there exist real-world
+distributed stores~\cite{spanner} that offer serializable transactions
+without causal consistency. Evidently, the reality is significantly
+more diverse~\cite{zoo}, with more forms of weak consistency and
+isolation being invented to meet the changing needs of applications.
+What is needed in this context is a reasoning framework parameterized
+over the semantics of weak isolation, that can answer the following
+question:
+\begin{center}
+\emph{Given a program annotated with desired invaraints, and a
+selection of isolation levels for transactions in the program, are the
+invariants guaranteed to hold?}
+\end{center}
+
+In this paper, we propose a reasoning framework that attempts to
+answer this question.

### 2016-06-07 introduction.tex
Intro edited.

@@
 been found to significantly outperform serializable transactions on
-benchmark suites, both on single-node~\cite{dbtuningbook} and
-multi-node replicated stores~\cite{bailishat,bailisvldb}, leading to
+benchmark suites, both on single-node databases and multi-node
+replicated stores~\cite{dbtuningbook,bailishat,bailisvldb}, leading to
 their overwhelming adoption. A 2013 study~\cite{bailishotos} of 18
@@
 invariants among 67 open source Ruby-on-Rails applications are the
-risk of being violated due to weak isolation. The ill effects of weak
-isolation are therefore as prevalent as weak isolation itself.
+risk of being violated due to weak isolation. Indeed, incidents of
+safety violations due to weak isolation have been reported on web
+services in production~\cite{starbucksbug, scimedbug}, including in
+safety-critical applications such as bitcoin
+exchanges~\cite{poloniexbug, bitcoinbug}. Evidently, the ill effects
+of weak isolation are as prevalent as the weak isolation itself.
 
@@
 undesirable \emph{phenomena} (\eg \emph{dirty reads}) each is required
-to prevent. While this is an improvement, it is hard to reason about
-application semantics in terms of the undesirable phenomena. Moreover,
-a later critique~\cite{berenson} uncovers many ambiguities and
-shortcomings in the standard. Adya's thesis~\cite{adyaphd} includes
-first formal definitions of some of the well-known isolation levels in
-the context of a system model of sequentially consistent database
-complete with timestamps and version histories. However, Adya's model,
-which predates the proliferation of weakly consistent replicated data
-stores, is not enough to express the semantics of a host of new
-isolation levels~\cite{psi,nmsi} tailor-made for such stores, or the
-interplay between weak consistency and isolation. Nonetheless,
-regardless of the specification methodology employed, the major
-missing piece in all the aforementioned approaches is the bridge
-between the system model of specifications and the operational model
-of programs that enables formal reasoning about program invariants in
-presence of weak isolation.
+to prevent. While this is an improvement, it requires programmers to
+be clairvoyant about the possible ways various undesirable phenomena
+might manifest in their applications, and determine, in each case, if
+the phenomenon can be allowed without violating application
+invaraints. Adya's thesis~\cite{adyaphd} includes first formal
+definitions of some of the well-known isolation levels in the context
+of a sequentially consistent (SC) database. However, there have been
+no attempts to relate Adya's system model to an operational model of
+programs in a reasoning framework that facilitates formal proofs of
+correctness. For this reason, the unsystematic way of reasoning about
+weak isolation in terms of undesirable phenomena remains popular to
+date, with major database vendors continuing to document their
+isolation levels in terms of the such phenomena~\cite{postgresiso,
+mysqliso, oracleiso}.
 
-There have been encouraging developments in Programming Languages
-research towards addressing the problem of reasoning about application
-invariants on weakly consistent replicated stores~\cite{burckhardt14,
-redblueosdi, redblueatc, ecinec, gotsmanpopl16}. Weak isolation can
-make use of some of the reasoning inventory built for weak consistency
-(as we demonstrate in later sections), but the latter does not subsume
-the former. While weak consistency (as considered by the
-aforementioned research) deals with lazily propagating effects of
-atomic and fully isolated operations, reasoning about weakly isolated
-transactions involves reasoning about the effects of groups of
-operations that are not necessarily atomic or isolated. Indeed, weakly
-isolated transactions have existed on sequentially consistent
-relational databases long before the advent of weakly consistent
-replicated stores, indicating the subtle interplay between consistency
-and isolation. Moreover, a common theme among the aforementioned line
-of work is the assumption of a system with a predefined set of
-consistency levels, and a minimum consistency level that is often
-stronger than eventual consistency. For instance, the reasoning
-framework proposed in~\cite{gotsmanpopl16} crucially relies on the
-assumption of causal consistency with only admissable strengthening
-being strong consistency. None of the ANSI SQL weak isolation levels
-guarantee causal consistency. Moreover, there exist real-world
-distributed stores~\cite{spanner} that offer serializable transactions
-without causal consistency. Evidently, the reality is significantly
-more diverse~\cite{zoo}, with more forms of weak consistency and
-isolation being invented to meet the changing needs of applications.
-What is needed in this context is a reasoning framework parameterized
-over the semantics of weak isolation, that can answer the following
-question:
+There have been encouraging developments in Programming Languages and
+Systems research towards reasoning about application invariants in
+presence of weak consistency~\cite{burckhardt14, redblueosdi,
+redblueatc, ecinec, gotsmanpopl16}. Weak isolation can make use of
+some of the reasoning inventory built for weak consistency (as we
+demonstrate in later sections), but the former problem is strictly
+more general than the latter. At a high level, reasoning about weak
+consistency involves reasoning about concurrent execution of atomic
+and fully isolated operations against different but consistent states
+of the database, where each operation locally preserves invariants. In
+contrast, reasoning about weak isolation involves reasoning about
+concurrent execution of transactions, whose operations witness
+different states which, although consistent individually, may not be
+reconciled into a consistent view. Moreover, operations may violate
+invariants as long as the transaction as a whole preserves them. It is
+easy to see that the latter problem reduces to the former if all
+transactions are singletons. The generality of weak isolation is also
+evident from the fact that weakly isolated transactions have existed
+on sequentially consistent relational databases long before the advent
+of weakly consistent replicated stores. Thus, there is a need for
+reasoning frameworks for weak isolation that generalize those for weak
+consistency in some important ways. The framework should be general
+enough to reason about the semantics of various isolation levels,
+including ANSI SQL isolation levels, in the context of data stores
+with various consistency semantics (\eg sequentially consistent data
+store of~\cite{adyaphd}, causally consistent store assumed
+in~\cite{gotsmanpopl16} etc). Considering the increasing diversity in
+the semantics of weak consistency~\cite{zoo} and
+isolation~\cite{gotsmanconcur}, it is ideal if the reasoning framework
+is parameterized on the semantics of consistency and isolation.  In
+short, what is neededis a reasoning framework that can answer the
+following question:
 \begin{center}
-\emph{Given a program annotated with desired invaraints, and a
-selection of isolation levels for transactions in the program, are the
-invariants guaranteed to hold?}
+\emph{Given a program annotated with desired invaraints, and isolation
+semantics for transactions in the program, assuming that the program
+executes against a data store with a given consistency semantics, are
+the invariants guaranteed to hold?}
 \end{center}

### 2016-06-07 introduction.tex
Intro done.

@@
 serializability by default, and half, including Oracle 11g, do not
-offer it at all.  A 2015 study~\cite{bailisferal} of a large corpus of
-database applications finds no evidence of any of them changing the
-default isolation level offered by the database. Together, the studies
-establish the prevalance of weak isolation in practice.
+offer it at all.  
+% A 2015 study~\cite{bailisferal} of a large corpus of database
+% applications finds no evidence of any of them changing the default
+% isolation level offered by the database. Together, the studies
+% establish the prevalance of weak isolation in practice.
+Weakly isolated transactions are therefore quite prevalent in
+practice.
 
@@
 safety-critical applications such as bitcoin
-exchanges~\cite{poloniexbug, bitcoinbug}. Evidently, the ill effects
-of weak isolation are as prevalent as the weak isolation itself.
+exchanges~\cite{poloniexbug, bitcoinbug}. While enforcing
+serializability for all transactions is sufficient to restore safety,
+it is an overkill considering that 75\% of invariants studied
+in~\cite{bailisferal} can be maintained under one or the other form of
+weak isolation. When to use weak isolation, and in what form, is
+therefore a question facing all database programmers.
 
-The root cause of problems with weak isolation is that its semantics
-in context of user programs is not well-understood. The original
+A major problem with weak isolation is that its semantics in context
+of user programs is not well-understood. The original
 proposal~\cite{gray1976} defines multiple ``degrees'' of weak
@@
 the phenomenon can be allowed without violating application
-invaraints. Adya's thesis~\cite{adyaphd} includes first formal
-definitions of some of the well-known isolation levels in the context
-of a sequentially consistent (SC) database. However, there have been
-no attempts to relate Adya's system model to an operational model of
-programs in a reasoning framework that facilitates formal proofs of
-correctness. For this reason, the unsystematic way of reasoning about
-weak isolation in terms of undesirable phenomena remains popular to
-date, with major database vendors continuing to document their
-isolation levels in terms of the such phenomena~\cite{postgresiso,
-mysqliso, oracleiso}.
+invaraints. This is understandably hard. Adya's thesis~\cite{adyaphd}
+includes first formal definitions of some of the well-known isolation
+levels in the context of a sequentially consistent (SC) database.
+However, there have been no attempts to relate Adya's system model to
+an operational model of programs in a proof system that facilitates
+formal proofs of correctness. For this reason, the unsystematic way of
+reasoning about weak isolation in terms of undesirable phenomena
+remains popular to date, with major database vendors continuing to
+document their isolation levels in terms of the such
+phenomena~\cite{postgresiso, mysqliso, oracleiso}.
 
@@
 concurrent execution of transactions, whose operations witness
-different states which, although consistent individually, may not be
+different states, which, although consistent individually, may not be
 reconciled into a consistent view. Moreover, operations may violate
-invariants as long as the transaction as a whole preserves them. It is
-easy to see that the latter problem reduces to the former if all
-transactions are singletons. The generality of weak isolation is also
-evident from the fact that weakly isolated transactions have existed
-on sequentially consistent relational databases long before the advent
-of weakly consistent replicated stores. Thus, there is a need for
-reasoning frameworks for weak isolation that generalize those for weak
-consistency in some important ways. The framework should be general
-enough to reason about the semantics of various isolation levels,
-including ANSI SQL isolation levels, in the context of data stores
-with various consistency semantics (\eg sequentially consistent data
-store of~\cite{adyaphd}, causally consistent store assumed
-in~\cite{gotsmanpopl16} etc). Considering the increasing diversity in
-the semantics of weak consistency~\cite{zoo} and
-isolation~\cite{gotsmanconcur}, it is ideal if the reasoning framework
-is parameterized on the semantics of consistency and isolation.  In
-short, what is neededis a reasoning framework that can answer the
-following question:
-\begin{center}
-\emph{Given a program annotated with desired invaraints, and isolation
-semantics for transactions in the program, assuming that the program
-executes against a data store with a given consistency semantics, are
-the invariants guaranteed to hold?}
-\end{center}
+invariants as long as the transaction as a whole preserves them.
+Intuitively, the latter problem reduces to the former if all
+transactions are singletons. Furthermore, weak isolation can exist
+independent of weak consistency, as evident from the presence of
+weakly isolated transactions on sequentially consistent relational
+databases long before the advent of weakly consistent NoSQL stores.
+Thus, there is a need for weak isolation reasoning frameworks that
+generalize reasoning under weak consistency in some important ways.
 
-In this paper, we propose a reasoning framework that attempts to
-answer this question.
+% The framework should be general enough to reason about the semantics
+% of multiple isolation levels, including those proposed after the SQL
+% 92 standard, in the context of various stores (\eg sequentially
+% consistent store of~\cite{adyaphd}, causally consistent store
+% of~\cite{gotsmanpopl16} etc). 
+In this paper, we propose a program logic for weakly isolated
+transations. In particular, we propose a set of syntax-directed proof
+rules that lets one build compositional proofs of correctness for
+transactional programs in presence of weak isolation. The key novelty
+of our approach is that it is parametric over the \emph{isolation
+semantics} of transactions in the program, as well as the
+\emph{consistency semantics} of the store. In concrete terms, this
+means that, unlike~\cite{gotsmanpopl16, redblueatc, ecinec}, our
+system model does not assume a minimum consistency/isolation level,
+nor does it assume a predefined set of such levels. Instead, our
+operational semantics admits declarative specifications of transaction
+isolation levels (\eg~\cite{pldi15,gotsmanconcur15}) and store
+consistency level, and generates executions that are guaranteed to
+satisfy the specifications. The result is a flexible system model that
+is general enough to incorporate the semantics of a range of isolation
+levels on variety of stores (\eg sequentially consistent store
+of~\cite{adyaphd}, causally consistent store of~\cite{gotsmanpopl16}
+etc). The parametricity in operational semantics naturally lends
+itself to parametricity in static semantics. Our key technical
+contribution is thus a set of proof rules, for demonstrating that a
+program with a given selection of isolation levels for its
+transactions preserves its invariants when executed on a store
+satisfying a certain consistency gurantees. The paper makes the
+following contributions:
+\begin{itemize}
+  \item We describe \txnimp - a simple imperative language equipped
+  with concurrency and weakly isolated transactions. We describe an
+  operational semantics for \txnimp that generates execution traces
+  satisfying a given set of well-formedness constraints. Building on
+  some ideas expoused in previous work, we show that semantics of
+  isolation levels can be captured as well-formedness constraints on
+  excution traces.
+  \item We propose a compositional proof system for \txnimp programs
+  to show that the program satisfies its invariants in all the
+  executions whose traces satisfy well-formedness constraints
+  corresponding to the chosen isolation levels. Our reasoning
+  framework comes with a rigorous proof of soundness that establishes
+  its correctness with respect to the operational semantics.
+  \item We define a \emph{maximum visibility principle} to reconcile
+  store consistency guarantees with transaction isolation requirements,
+  and use it to instantiate our operational model and the proof system
+  for various kinds of stores.
+  \item We describe case studies modeled after real-world scenarios
+  to demonstrate the applicability and utility of our proof
+  methodology.
+\end{itemize}

### 2016-06-07 introduction.tex
Minor edit.

@@
 programs that often give rise to anomalous executions and
-incomprehensible results~\cite{pldi15}. To quantify weak isolation
+incomprehensible results~\cite{berenson}. To quantify weak isolation
 anomalies, Fekete et al~\cite{feketevldb09} have devised and

### 2016-06-07 abstract.tex
Some edits.

@@
 In this paper, we propose an operational model and a compositional
-reasoning framework for concurrent imperative programs with weakly
-isolated transactions (\txnimp programs, as we call them).  Our
-operational model captures all the behaviors exhibited by a program
-under weak isolation without referring to any specific implementation
-of isolation. Our reasoning framework enables formal reasoning about
-program invariants while taking into account the fine-grained
-isolation semantics of transactions. We use a combination of simple
-static analysis and manual proofs in our reasoning framework to
-formally establish the correctness for several benchmarks, which were
-hitherto known to be correct only through experiments.
+proof system for concurrent imperative programs with weakly isolated
+transactions (\txnimp programs, as we call them).  Our operational
+model captures all the behaviors exhibited by a program under weak
+isolation without referring to any specific implementation of
+isolation. Our proof system enables formal reasoning about program
+invariants while taking into account the fine-grained isolation
+semantics of transactions. Case studies demonstrate the utility of our
+approach.
 \end{abstract}

### 2016-06-11 abstract.tex
intro and abstract

@@
 \begin{abstract}
-ACID model of database transactions advocates serializability as the
-correctness criterion for concurrent transactions. 
-% Serializability is an easily-stated property,
-% providing database programmers with a simple
-% implementation-independent model to reason about program invariants in
-% presence of concurrent transactions. Unfortunately, 
-While serializability is an easily-stated high-level property,
-enforcing it requires the database to severely restrict
-concurrency.
-% , thus limiting the availability and performance
-% (latency/throughput) of the data store.  In the database parlance,
-% serializibility is \emph{not highly available}~\cite{bailishotos13},
-% making it particularly unsuitable for modern-day distributed web
-% applications that aspire to be \emph{always on}.
-Pragmatic concerns led to the development of various weaker
-alternatives to serializability, which are overwhelmingly preferred in
-practice.
-% some of which have been
-% standardized by the community as \emph{ANSI SQL isolation
-% levels}~\cite{berenson}. Indeed, weakly isolated transactions are
-% overwhelmingly preferred to serializability in practice, to the extent
-% that the default isolation level in popular industrial-strength
-% databases is often weaker than serializability.
-However, unlike serializability, the semantics of weaker isolation
-levels are often described informally in terms of an implementation.
-While there have been some attempts at formalization, there
-nonetheless exists a wide chasm between formal models of weak
-isolation and the operational models of programs, hindering uniform
-reasoning about program invariants in presence of weakly isolated
-transactions.
 
-In this paper, we propose an operational model and a compositional
-proof system for concurrent imperative programs with weakly isolated
-transactions (\txnimp programs, as we call them).  Our operational
-model captures all the behaviors exhibited by a program under weak
-isolation without referring to any specific implementation of
-isolation. Our proof system enables formal reasoning about program
-invariants while taking into account the fine-grained isolation
-semantics of transactions. Case studies demonstrate the utility of our
-approach.
+  Serializability is a well-understood mechanism that simplifies
+  reasoning about the behavior of concurrent transactions.  However,
+  its two key properties, \emph{atomicity} - a transaction's effects
+  are seen only when it commits, and \emph{isolation} - a transaction
+  never witnesses the effects of other committed transactions while it
+  is executing, impose significant constraints on implementations, and
+  can negatively impact database performance by restricting
+  opportunities to exploit concurrency.
+
+  As a result, database systems (in practice) support, and often
+  encourage, developers to use weaker alternatives.  Generally, these
+  alternatives break the strong isolation guarantees provided by
+  serializable transactions to permit greater concurrency.
+  Unfortunately, the semantics of weak isolation is often poorly
+  understood, and usually explained informally in terms of low-level
+  implementation artifacts.  Consequently, verifying high-level
+  correctness properties in such environments can be problematic.
+
+  To address this issue, this paper presents a novel program logic
+  that enables compositional reasoning about the behavior of
+  concurrently executing weakly-isolated transactions.  Notably, our
+  development is parametric over a transaction's specific isolation
+  semantics, and the consistency guarantees provided by the underlying
+  data store, allowing it to be applicable over a large range of
+  different isolation and consistency models.  Case studies and
+  experiments on real-world applications demonstrate the utility of
+  our approach, and provide strong evidence that weakly-isolated
+  transactions can be placed on the same formal footing as their
+  strongly-isolated serializable counterparts.
+
 \end{abstract}

### 2016-06-11 introduction.tex
intro and abstract

--- 2016-06-07 introduction.tex
@@
 Database transactions allow users to group operations on multiple
-objects into a coherent unit with useful properties including
-atomicity, consistency, (serializable) isolation and durability
+objects into a coherent unit, equipped with a set of four key
+properties - atomicity, consistency, isolation, and durability
 (ACID). ACID has been the gold standard for database transactions for
-many decades, providing programmers with a simple
-implementation-independent model to reason about concurrent
-transactions. However, serializability, the ACID-prescribed standard for
-isolation, does not come for free; it requires databases to rely on
-expensive techniques such as two-phase locking~\cite{twopl,ullmanbook}
-that severely restrict concurrency while incurring an additional
-overhead of deadlock detection, rollback and re-execution. Enforcing
-serializability in a replicated data store requires global
-coordination among geo-distributed replicas, which makes the store
-\emph{unavailable} in presence of the inevitable network
-partitions~\cite{cap,sernotavlbl,bailishat,bernsigmod13}. The
-conflict between serializability and pragmatic goals of high
-peformance and availability has motivated the development of multiple
-weaker forms of transaction isolation beginning as early as
-1976~\cite{gray1976}. ANSI SQL 92 standard defines three such weak
-isolation \emph{levels} which are now implemented in
-many relational and NoSQL databases. Weakly isolated transactions have
-been found to significantly outperform serializable transactions on
-benchmark suites, both on single-node databases and multi-node
-replicated stores~\cite{dbtuningbook,bailishat,bailisvldb}, leading to
-their overwhelming adoption. A 2013 study~\cite{bailishotos} of 18
-popular ACID and ``NewSQL'' databases found that only 3 of them offer
+many decades, providing programmers with a simple implementation
+independent model to reason about concurrent transactions.  For
+example, concurrency control mechanisms typically enforce isolation by
+making all transactions \emph{serializable}, to ensure that any
+allowed schedule is equivalent to one in which there is no
+interleaving of actions from different concurrently executing
+transactions.
+
+However the strong guarantees provided by serializability does not
+come for free - pessimistic concurrency control methods, for example,
+require databases to provide expensive techniques such as two-phase
+locking that incur overhead to deal with deadlocks, rollbacks, and
+re-execution~\cite{twopl,ullmanbook}.  Similar criticisms apply to
+optimistic multi-version concurrency control methods that must deal
+with timestamp and version management~\cite{BG81}.  Regardless of the
+method employed, enforcing serializability in a replicated data store
+additionally requires global coordination among geo-distributed
+replicas, which makes the store \emph{unavailable} in the presence of
+inevitable network
+partitions~\cite{cap,sernotavlbl,bailishat,bernsigmod13}.
+
+The tension between serializable transactions, which are easy to
+reason about but difficult to implement, and more pragmatic goals
+driven by performance and availability considerations, has motivated
+the development of multiple weaker forms of transaction isolation,
+beginning as early as 1976~\cite{gray1976}. The ANSI SQL 92 standard
+defines three such weak isolation \emph{levels} which are now
+implemented in many relational and NoSQL databases. Not surprisingly,
+weakly isolated transactions have been found to significantly
+outperform serializable transactions on benchmark suites, both on
+single-node databases and multi-node replicated
+stores~\cite{dbtuningbook,bailishat,bailisvldb}, leading to their
+overwhelming adoption. A 2013 study~\cite{bailishotos} of 18 popular
+ACID and ``NewSQL'' databases found that only three of them offer
 serializability by default, and half, including Oracle 11g, do not
 offer it at all.  A 2015 study~\cite{bailisferal} of a large corpus of
-database applications finds no evidence of any of them changing the
-default isolation level offered by the database. Together, the studies
-establish the prevalance of weak isolation in practice.
+database applications finds no evidence that applications manifestly
+change the default isolation level offered by the database. Taken
+together, these studies make clear that weakly isolated transactions
+are quite prevalent in practice.
 
-Unfortunately, weak isolation introduces new set of behaviors into
-programs that often give rise to anomalous executions and
-incomprehensible results~\cite{pldi15}. To quantify weak isolation
-anomalies, Fekete et al~\cite{feketevldb09} have devised and
-experimented with a microbenchmark suite that executes transactions
-under \iso{Read Committed} isolation level - the default level for 8
-of the 18 databases studied in~\cite{bailishotos}, and found that 25
+Unfortunately, weak isolation admits behaviors that are difficult to
+comprehend~\cite{berenson}. To quantify weak isolation anomalies,
+Fekete \emph{et al.}~\cite{feketevldb09} have devised and experimented
+with a microbenchmark suite that executes transactions under a weakly
+isolated \iso{read committed} isolation level - the default level for
+8 of the 18 databases studied in~\cite{bailishotos}, and found that 25
 out of every 1000 rows in the database violate at least one integrity
-constraint. Bailis et al~\cite{bailisferal} rely on Rails's
+constraint. Bailis \empH{et al.}~\cite{bailisferal} rely on Rails's
 \emph{uniqueness validation} to maintain uniqueness of records while
@@
 used in the experiments. The same study has found that 13\% of all
-invariants among 67 open source Ruby-on-Rails applications are the
-risk of being violated due to weak isolation. Indeed, incidents of
-safety violations due to weak isolation have been reported on web
-services in production~\cite{starbucksbug, scimedbug}, including in
+invariants among 67 open source Ruby-on-Rails applications are at risk
+of being violated due to weak isolation. Indeed, incidents of safety
+violations due to weak isolation have been reported on web services in
+production~\cite{starbucksbug, scimedbug}, including in
 safety-critical applications such as bitcoin
-exchanges~\cite{poloniexbug, bitcoinbug}. Evidently, the ill effects
-of weak isolation are as prevalent as the weak isolation itself.
+exchanges~\cite{poloniexbug, bitcoinbug}. While enforcing
+serializability for all transactions would be sufficient to avoid
+these errors and anomalies, it would likely be an overly conservative
+strategy; indeed, 75\% of the invariants studied in~\cite{bailisferal}
+were shown to be preserved under some form of weak isolation.  When to
+use weak isolation, and in what form, is therefore a prominent
+question facing all database programmers.
 
-The root cause of problems with weak isolation is that its semantics
-in context of user programs is not well-understood. The original
+A major problem with weak isolation is that its semantics in the
+context of user programs is not well-understood. The original
 proposal~\cite{gray1976} defines multiple ``degrees'' of weak
 isolation in terms of implementation details, such as the nature and
-duration of locks held in each case. ANSI SQL 92 standard four levels
-of isolation (including serializability) in terms of sevaral
+duration of locks held in each case. The ANSI SQL 92 standard four
+levels of isolation (including serializability) in terms of sevaral
 undesirable \emph{phenomena} (\eg \emph{dirty reads}) each is required
 to prevent. While this is an improvement, it requires programmers to
-be clairvoyant about the possible ways various undesirable phenomena
+be prescient about the possible ways various undesirable phenomena
 might manifest in their applications, and determine, in each case, if
 the phenomenon can be allowed without violating application
-invaraints. Adya's thesis~\cite{adyaphd} includes first formal
-definitions of some of the well-known isolation levels in the context
-of a sequentially consistent (SC) database. However, there have been
-no attempts to relate Adya's system model to an operational model of
-programs in a reasoning framework that facilitates formal proofs of
-correctness. For this reason, the unsystematic way of reasoning about
-weak isolation in terms of undesirable phenomena remains popular to
-date, with major database vendors continuing to document their
-isolation levels in terms of the such phenomena~\cite{postgresiso,
-mysqliso, oracleiso}.
+invariants. This is understandably hard. Adya~\cite{adyaphd} includes
+the first formal definitions of some of well-known isolation levels in
+the context of a sequentially consistent (SC) database.  However,
+there has been little progress relating Adya's system model to a
+formal operational semantics or a proof system that can facilitate
+rigorous correctness arguments.  Consequently, reasoning about weak
+isolation remains an error prone endeavor, with major database vendors
+continuing to document their isolation levels primarily in terms of
+the undesirable phenomena~\cite{postgresiso, mysqliso, oracleiso} a
+particular isolation level may induce; it thus squarely becomes the
+programmer's responsibility to determine if their application can
+yield an execution which can trigger such phenomena.
 
-There have been encouraging developments in Programming Languages and
-Systems research towards reasoning about application invariants in
+Recent results on reasoning about application invariants in the
 presence of weak consistency~\cite{burckhardt14, redblueosdi,
-redblueatc, ecinec, gotsmanpopl16}. Weak isolation can make use of
-some of the reasoning inventory built for weak consistency (as we
-demonstrate in later sections), but the former problem is strictly
-more general than the latter. At a high level, reasoning about weak
-consistency involves reasoning about concurrent execution of atomic
-and fully isolated operations against different but consistent states
-of the database, where each operation locally preserves invariants. In
-contrast, reasoning about weak isolation involves reasoning about
-concurrent execution of transactions, whose operations witness
-different states which, although consistent individually, may not be
-reconciled into a consistent view. Moreover, operations may violate
-invariants as long as the transaction as a whole preserves them. It is
-easy to see that the latter problem reduces to the former if all
-transactions are singletons. The generality of weak isolation is also
-evident from the fact that weakly isolated transactions have existed
-on sequentially consistent relational databases long before the advent
-of weakly consistent replicated stores. Thus, there is a need for
-reasoning frameworks for weak isolation that generalize those for weak
-consistency in some important ways. The framework should be general
-enough to reason about the semantics of various isolation levels,
-including ANSI SQL isolation levels, in the context of data stores
-with various consistency semantics (\eg sequentially consistent data
-store of~\cite{adyaphd}, causally consistent store assumed
-in~\cite{gotsmanpopl16} etc). Considering the increasing diversity in
-the semantics of weak consistency~\cite{zoo} and
-isolation~\cite{gotsmanconcur}, it is ideal if the reasoning framework
-is parameterized on the semantics of consistency and isolation.  In
-short, what is neededis a reasoning framework that can answer the
-following question:
-\begin{center}
-\emph{Given a program annotated with desired invaraints, and isolation
-semantics for transactions in the program, assuming that the program
-executes against a data store with a given consistency semantics, are
-the invariants guaranteed to hold?}
-\end{center}
+  redblueatc, ecinec, gotsmanpopl16} provide insights on possible ways
+we might overcome some of the difficulties in understanding weak
+isolation, even though the latter is strictly more general than the
+former.  At a high level, reasoning about weak consistency involves
+reasoning about the concurrent execution of atomic and strongly
+isolated operations against different but consistent states of the
+database, where each operation locally preserves application
+invariants. In contrast, reasoning about weak isolation involves
+reasoning about concurrent execution of transactions, whose operations
+witness different states, which, although consistent individually, may
+not be obviously reconciled into a consistent view. Moreover,
+operations may violate invariants as long as the transaction as a
+whole preserves them.  Intuitively, the latter problem reduces to the
+former if all transactions contain a single operation. Furthermore,
+weak isolation can exist independent of weak consistency, as evident
+from the presence of weakly isolated transactions on sequentially
+consistent relational databases; such systems existed long before the
+advent of weakly consistent NoSQL stores.  Thus, frameworks for
+reasoning about weak isolation will necessarily have to generalize the
+reasoning frameworks developed for weak consistency in important ways.
 
-In this paper, we propose a reasoning framework that attempts to
-answer this question.
+% The framework should be general enough to reason about the semantics
+% of multiple isolation levels, including those proposed after the SQL
+% 92 standard, in the context of various stores (\eg sequentially
+% consistent store of~\cite{adyaphd}, causally consistent store
+% of~\cite{gotsmanpopl16} etc).
+
+In this paper, we propose a program logic for weakly isolated
+transactions that realizes this goal.  In particular, we develop a set
+of syntax-directed compositional proof rules that allow developers to
+build correctness proofs for transactional programs in the presence of
+a weakly isolated concurrency control mechanism.  A key novelty of our
+approach is that it is parametric over the \emph{isolation semantics}
+of transactions in the program, as well as the \emph{consistency
+  semantics} of the store. In concrete terms, this means that, unlike
+recent work focussed on formalizing the semantics of weak
+consistency~\cite{gotsmanpopl16, redblueatc, ecinec}, our system model
+does not assume a minimum consistency/isolation level, nor does it
+assume a predefined set of such levels.  Instead, our operational
+semantics admits declarative specifications of transaction isolation
+and store consistency levels (\eg~\cite{pldi15,gotsmanconcur15}), and
+generates executions that are guaranteed to satisfy these
+specifications. The result is a flexible system model that is general
+enough to incorporate the semantics of a range of isolation levels on
+a variety of stores (\eg the sequentially consistent store
+of~\cite{adyaphd}, or causally consistent store
+of~\cite{gotsmanpopl16}, etc).  Our key technical contribution is thus
+a set of proof rules that demonstrate that a program with a given
+selection of isolation levels for its transactions preserves its
+invariants when executed on a store satisfying certain consistency
+guarantees.
+
+The paper makes the following contributions:
+\begin{itemize}
+  \item We develop a semantics for a core language equipped with
+    weakly isolated transactions.  Building on some ideas espoused in
+    previous work, we show that isolation semantics can be captured as
+    well-formedness constraints on a program's execution.
+  \item We propose a compositional proof system for this language
+    capable of relating high-level application invariants to the structure
+    of traces induced by the operational semantics, in which transactions
+    are parameterized with specific weak isolation levels.
+  \item We define a \emph{maximum visibility principle} to reconcile
+    store consistency guarantees with transaction isolation
+    requirements, and use it to instantiate our operational model and
+    the proof system for various kinds of stores.
+  \item Case studies modeled after real-world scenarios demonstrate
+    the applicability and utility of our proof methodology.
+\end{itemize}
+
+Our semantic framework and proof methodology thus provide the first
+(to the best of our knowledge) mechanism that precisely relates
+high-level program invariants to low-level store consistency and weak
+isolation guarantees, thereby allowing weakly isolated transactions to
+enjoy the same rigorous reasoning capabilities as their strongly
+isolated (serializable) counterparts.

### 2016-06-12 motivation.tex
edits

@@
 
-In this section, we motivate the need for a formal reasoning framework
-for weakly isolated transactions through a series of examples. All our
-examples are written in C-like imperative language equipped with a
-\C{txn} lexical block for transactions. Each \C{txn} block is
-associated with a single-quoted string in angle braces that uniquely
-identifies the transaction. We use Hoare triple notation to annotate
-programs with pre and post conditions.
+We motivate the need to have a formal reasoning framework for weakly
+isolated transactions through a series of simple examples that
+abstract real-world scenarios.  All of our examples are written in a
+C-like imperative language equipped with a \C{txn} lexical block
+defining a transaction scope.  Each \C{txn} block is associated with
+an identifier that uniquely identifies the transaction. We use Hoare
+triple notation to annotate programs with pre- and post-conditions.
 
@@
 \begin{txnimpcode}
-  txn$\langle$'Wd1'$\rangle${
+  txn '\B{Wd1}' {
     if (C $\ge$ a1) {
@@
 \begin{txnimpcode}
-  txn$\langle$'Wd2'$\rangle${
+  txn '\B{Wd2}'  {
     if (C $\ge$ a2) {
@@
 shown in Fig.~\ref{fig:motiv-eg-1}. If the initial balance (\C{k}) in
-the account is enough to perform both withdraws, then the final
-balance, after both transactions commit, is expected to reflect the
-effects of both withdraws. The pre and post conditions in
-Fig.~\ref{fig:motiv-eg-1} reflect our expectations. Indeed, invariants
-are guaranteed to hold if both withdraw transactions are serialized,
-making \iso{Serializable} isolation (SER) level a sufficent condition
-to preserve invaraints. But, is SER necessary?
+the account is enough to perform both withdraw operations, then the
+final balance, after both transactions commit, is expected to reflect
+the effects of performing these withdraws. The pre- and
+post-conditions in Fig.~\ref{fig:motiv-eg-1} reflect our
+expectations. Indeed, invariants are guaranteed to hold if both
+withdraw transactions are serialized, making a \emph{serializable}
+isolation ({\sc ser}) level a sufficent condition to preserve the
+desired invaraint. But, is {\sc SER} necessary?
 
@@
 % lead to the violation of invaraints. 
-Let us consider \iso{Read Committed} (RC), an isolation level weaker
-than SER that is default in Postgres 9.5~\cite{postgres95} and Oracle
-11g~\cite{oracle11g} databases. An RC transaction is isolated from the
-writes of an uncommitted transactions, thus preventing the so called
-\emph{dirty reads} phenomenon~\cite{berenson}. In the current example,
-RC isolation admits following executions:
+As an alternative, consider the execution of this transaction under a
+\emph{read committed} ({\sc rc}) isolation level, which is weaker than
+     {\sc ser}\footnote{{\sc rc} is in fact the default isolation
+       level in Postgres 9.5~\cite{postgres95} and Oracle
+       11g~\cite{oracle11g} databases.} An {\sc rc} transaction is
+     isolated from the writes of uncommitted transactions, thus
+     preventing the transaction from witnessing \emph{dirty
+       reads}~\cite{berenson}, reads that observe the effects of
+     non-committed transactions.  In the current example, {\sc rc}
+     isolation admits the two executions shown in Fig.~\ref{fig:rc-ex}.
 
@@
 }
+\caption{Two possible executions under an {\sc rc} isolation level.}\label{fig:rc-ex}
 \end{figure}
 
-An execution is a series of Read, Write and Commit operations.
-Subscript of an operation indicates its transaction. For clarity,
-effects of different transactions are shown against differently
-colored backgrounds. In the first execution, the first transaction
-(`Wd1') reads the current balance (\C{k}) and writes the new balance
-(\C{k-a1}), but before it commits transaction `Wd2' executes and
-commits, writing the new balance (\C{k-a2}). RC isolation has
-prevented `Wd2' from witnessing the writes of uncommitted transaction
-`Wd1'. Committing `Wd1' now leads to the loss of updates from `Wd2'
-(the so called \emph{lost update anomaly}~\cite{berenson}), resulting
-in a balance of \C{k-a1}. The second execution describes a similar
-scenario with `Wd1' and `Wd2' exchanging their roles.
+The figure depicts an execution as a series of read, write, and commit
+operations.  The subscript of an operation indicates the transaction
+that executes it\footnote{For clarity, the effects of different
+  transactions are shown in different colored backgrounds.} In the
+first execution, transaction {\bf Wd1} reads the current balance
+(\C{k}) and writes the new balance (\C{k-a1}), but before it commits
+transaction {\bf Wd2} executes and commits, writing the new balance
+(\C{k-a2}). RC isolation prevent {\bf Wd2} from witnessing the
+uncommitted writes of transaction {\bf Wd1}.  Subsequently committing
+{\bf Wd1} leads to the loss of {\bf Wd2}'s updates (the so called
+\emph{lost update anomaly}~\cite{berenson}), resulting in an incorrect
+balance of \C{k-a1}. The second execution describes a similar scenario
+with {\bf Wd1} and {\bf Wd2} exchanging their roles.
 
-Clearly, \iso{Read Committed} is not sufficient because it loses the
-updates of one transaction and violates the invariant in
+Clearly, using a \emph{read committed} isolation level is not
+sufficient because it may result in the loss of updates from one
+transaction and can thus violate the invariant given in
 Fig.~\ref{fig:motiv-eg-1}. Now, consider an alternative unfolding of
-Execution 1, where, rather than committing `Wd1' and overwriting the
-updates of `Wd2', we rollback `Wd1' to reexecute it starting from the
-state that contains the updates of `Wd2'. No updates are lost in this
-alternative execution, leading to the satisfaction of
-Fig.~\ref{fig:motiv-eg-1} invariant. Rolling back and reexecuting an
-uncommitted transaction to prevent write-write conflicts with a
-committed transaction is a strategy often adopted by databases to
-implement \iso{Snapshot Isolation} (SI) level. Indeed, SI is
-sufficient to enforce the required invaraints in this
-example\footnote{\iso{Parallel Snapshot Isolation} (PSI), a weaker
-version of SI tailor-made for replicated stores is also sufficient.}.
-Moreover, relying on optimistic speculation instead of pessimistic
-locking for concurrency control makes SI more efficient than SER,
-making it appropriate for transactions in Fig.~\ref{fig:motiv-eg-1}.
-Note that we have arrived at his judgment via adhoc reasoning based on
-speculations about database's implementation strategies. Such a
-reasoning process is necessarily fragile and error-prone. Is there a
-better way?
+Execution 1, where, rather than committing {\bf Wd1} and overwriting
+the updates of {\bf Wd2}, we rollback {\bf Wd1} to reexecute it
+starting from the state that contains the updates of {\bf Wd2}. No
+updates are lost in this alternative execution, leading to the
+satisfaction of Fig.~\ref{fig:motiv-eg-1}'s invariant. Rolling back
+and reexecuting an uncommitted transaction to prevent write-write
+conflicts with a committed transaction is a strategy often adopted by
+databases that implement \emph{snapshot isolation} ({\sc si}). Indeed,
+{\sc si} is sufficient to enforce the required invariants in this
+example\footnote{\emph{Parallel snapshot isolation} ({\sc psi}), a
+  weaker version of {\sc si} tailor-made for replicated stores is also
+  sufficient.}.  Moreover, relying on optimistic speculation instead
+of pessimistic locking for concurrency control enables {\sc si} to
+often be more efficient than {\sc ser}, making it appropriate for the
+transactions shown in Fig.~\ref{fig:motiv-eg-1}.  Note that we have
+arrived at the applicability of {\sc si} (and the inapplicability of {\sc
+  rc}) via a number of \emph{ad hoc} reasoning steps, several of which
+are based on speculation about a the underlying data store's
+implementation strategies. Such a reasoning process is necessarily
+fragile and error-prone.
 
@@
 \begin{txnimpcode}
-  txn$\langle$'WdC'$\rangle${
+  txn '\B{WdC}' {
     if (C+S $\ge$ a1) {
@@
 \begin{txnimpcode}
-  txn$\langle$'WdS'$\rangle${
+  txn '\B{WdS}' {
     if (C+S $\ge$ a2) {
@@
 
-\iso{Snapshot Isolation} effectively serializes updates to a data
-object without necessarily relying on expensive lock-based concurrency
-control. While this makes it a better alternative to
-\iso{Serializable} isolation, it is however not a replacement for the
-latter. Consider a different banking application that maintains a
-current (\C{C}) and savings account (\C{S}) for each user. A user can
-withdraw either from current or from savings account as long as the
-combined balance is at least as much as the amount being withdrawn.
-Fig.~\ref{fig:motiv-eg-2} shows a pair of concurrent transactions,
-`WdC' and `WdS', performing withdraws from current and savings
-accounts, respectively. Pre and post conditions assert non-negative
-combined balance invariant. \iso{Serializable} isolation is sufficient
-to preserve the invariant, but \iso{Snapshot Isolation} is not.
+{\sc si} effectively serializes updates to a data object without
+relying on expensive lock-based concurrency control. While this makes
+it a better alternative to {\sc ser} in some cases, it cannot however
+serve as a replacement for the latter. Consider a different banking
+application that maintains a current (\C{C}) and savings account
+(\C{S}) for each user. A user can withdraw either from current or from
+savings as long as the combined balance is at least as much as the
+amount being withdrawn.  Fig.~\ref{fig:motiv-eg-2} shows a pair of
+concurrent transactions, {\bf WdC} and {\bf WdS}, performing
+withdrawls from current and savings accounts, resp. The pre-
+and post-conditions for this program assert that executing its
+transactions will not result in a non-negative combined balance.  {\sc
+  ser} is sufficient to preserve the invariant, but {\sc si} is not.
 Intuitively, this is because there are no \emph{lost updates} in this
 example even when both transactions commit concurrently, hence there
-is no reason for SI to effectively serialize them. A reasoning
-framework for weak isolation should let us formally deduce this fact
-by making it impossible to construct a proof of correctness for the
-program under SI, but facilitating a proof under SER.
+is no reason for {\sc si} to serialize them. A reasoning framework for
+weak isolation should let us formally deduce this fact by making it
+impossible to construct a proof of correctness for the program under
+{\sc si}, but facilitating a proof under {\sc ser}.
 
@@
 \begin{txnimpcode}
-  txn$\langle$'Xfer'$\rangle${
+  txn '\B{Xfer}' {
     C := C - a;
@@
 \begin{txnimpcode}
-  txn$\langle$'TotBal'$\rangle${
+  txn '\B{TotBal}' {
     T := S + C
@@
   \begin{txnimpcode}
-    txn$\langle$'AddInt'$\rangle${
+    txn '\B{AddInt}' {
       S := S + (T*0.1)
@@
 
-The previous two examples preserve invariants on \C{C} and \C{S} by
-effectively serializing transactions that write to \C{C} and \C{S}.
-However, in general, serializing a subset of transactions that write
-to the data objects referred by invariants is neither sufficient nor
-necessary to uphold the invariants. Consider the banking application
-in Fig.~\ref{fig:motiv-eg-3} with three transactions. Transaction
-`Xfer' transfers some amount (\C{a}) from current (\C{C}) to savings
-(\C{S}) account.  Transaction `TotBal' records the total balance in
-both the accounts in a data object \C{T}, and transaction `AddIntr'
-calculates interest on total balance (10\% of \C{T}) and adds it to
-the savings account. Using a data object to record some function of
-the current state of other data objects is a standard practice in
-relational databases, where Materialized Views~\cite{oraclematview}
-are created to maintain the results of a SQL query. Like the previous
-example, the invariant of interest in this example is non-negative
-total balance; if we start with a non-negative total balance, and we
-add interest based on the total balance, we must end up with a
-non-negative total balance.  However, unlike the previous example,
-serializing transactions that write to \C{C} and \C{S} is not
-sufficient to guarantee the invariant.  The following execution
-demonstrates this anomaly:
+The previous example preserves invariants on \C{C} and \C{S} by
+effectively serializing those transactions that write to \C{C} and
+\C{S}.  However, in general, serializing a subset of transactions that
+write to the data objects referenced by invariants is neither
+necessary nor sufficient to ensure desired invariants hold. Consider
+the banking application in Fig.~\ref{fig:motiv-eg-3} with three
+transactions. Transaction {\bf Xfer} transfers some amount (\C{a})
+from the current (\C{C}) to the savings (\C{S}) account.  Transaction
+{\bf TotBal} records the total balance in both accounts in a data
+object \C{T}, and transaction {\bf AddIntr} calculates interest on the
+total balance (10\% of \C{T}) and adds it to the savings
+account. Using a data object to record some function of the current
+state of other data objects is a standard practice in relational
+databases, where materialized views~\cite{oraclematview} are created
+to maintain the results of a SQL query. Like the previous example, the
+invariant of interest in this example is a non-negative total balance.
+However, unlike the previous example, serializing transactions that
+write to \C{C} and \C{S} is not sufficient to guarantee the invariant.
+The execution shown in Fig.~\ref{fig:motiv-eg-3-a} demonstrates this anomaly:
 
@@
 \includegraphics[scale=0.5]{Figures/motiv-eg-3-a}
-
+\caption{Serializing transactions that update objects referenced by
+  program invariants may introduce unwanted anomalies.}
 \label{fig:motiv-eg-3-a}
@@
 
-The subscripts X, TB and AI denote `Xfer', `TotBal' and `AddInt'
-transactions, respectively. Assume that \C{C} and \C{S} are initially
-10 and 0, respectively. Let \C{a} be 200, i.e., we subtract 200 from
-\C{C} and add it to \C{S}.  This operation must be allowed because it
-doesn't effect the total balance, hence the interest. The execution
-begins with `TotBal' transaction reading the savings (\C{S}) balance
-as 0 (we assume left-to-right evaluation order for \C{S + C}).
-However, before it reads the current (\C{C}) balance, `Xfer'
-transaction executes and commits, changing \C{C} to -190.
-Subsequently, `TotBal' reads -190 from \C{C}, and updates \C{T} to
--190. `AddInt', which begins execution after `TotBal', reads -190 from
-\C{T}, calculates interest as -19, and updates \C{S} to 181. The
-result is a state with total balance as -190 + 181 = -9, thus
-violating the invariant.
+The subscripts {\bf X}, {\bf TB} and {\bf AI} denote {\bf Xfer}, {\bf
+  TotBal} and {\bf AddInt} transactions, resp. Assume that \C{C} and
+\C{S} are initially 10 and 0, resp.  Assume \C{a} is 200, i.e., we
+subtract 200 from \C{C} and add it to \C{S}.  Transferring amounts
+from \C{C} to \C{S} is allowed because the total balance invariant is
+not affected.  The execution begins with {\bf TotBal} transaction
+reading the savings (\C{S}) balance as 0 (we assume left-to-right
+evaluation order for \C{S + C}).  However, before it reads the current
+(\C{C}) balance, the {\bf Xfer} transaction executes and commits,
+changing \C{C} to -190.  Subsequently, {\bf TotBal} reads -190 from
+\C{C}, and updates \C{T} to -190. {\bf AddInt}, which begins execution
+after {\bf TotBal}, reads -190 from \C{T}, calculates the interest as
+-19, and updates \C{S} to 181. The result is a state with a total
+balance of -9, a value that violates the invariant.
 
-Note that transactions `Xfer' and `AddInt', which write to the data
-objects (\C{C} and \C{S}) relevant to the invariant, are serialized in
-the above execution. Transaction `TotBal' however is not. While
-serializing it fixes the problem, it is not necessary. A simpler fix
-is to isolate `TotBal' transaction from the effects of concurrent
-`Xfer' transaction, so that it reads \C{C} as 10 despite the
-concurrent `Xfer' transaction committing -190 to \C{C}. This can be
-achieved by executing `TotBal' under \iso{Repeatable Read} (RR)
-isolation level, which guarantees that any two reads of a data object
-in a transaction return the same value (the \emph{repeatable reads}
-property~\cite{berenson}). Allowing `Xfer' to interfere with `TotBal'
-violates the repeatable read property since a hypothetical read from
-\C{S} in `TotBal' after the interference returns a different value
-(200) than the first read (0). Executing `TotBal' under RR restores
-repeatable read by \emph{effectively} prohibiting such interference.
+Note that transactions {\bf Xfer} and {\bf AddInt}, which write to the
+data objects (\C{C} and \C{S}) relevant to the invariant, are
+serialized in the above execution. Transaction {\bf TotBal}, however,
+is not.  Although serializing it would fix the problem, it is not
+necessary. A simpler fix would be to isolate {\bf TotBal} transaction
+from the effects of a concurrent executing {\bf Xfer} transaction, so
+that it reads \C{C} as 10 despite {\bf Xfer} committing -190 to
+\C{C}. This can be achieved by executing {\bf TotBal} under a
+\emph{repeatable read} ({\sc rr}) isolation level, which guarantees
+that any two reads of a data object in a transaction return the same
+value.  Allowing {\bf Xfer} to interfere with {\bf TotBal} violates
+the repeatable read property since a hypothetical read from \C{S} in
+{\bf TotBal} after the interference returns a different value (200)
+than the first read (0). Executing {\bf TotBal} under {\sc rr}
+effectively prohibits such interference.
 
@@
 \begin{txnimpcode}
-  txn$\langle$'Rd'$\rangle${
+  txn '\B{Rd}' {
     X := V
@@
   cobegin (i $\leftarrow$ [1$\ldots$3]) {
-    txn$\langle$'Inc['+i+']'$\rangle${
+    txn '\B{Inc}$_{i}$' {
       V := V + 1
@@
 
+%% SJ: not sure this example is needed - seems very complex, and introduces
+%% details about the weak consistency.
 Our last example demonstrates that it is not necessary to serialize
-transactions, even those that write to the same data objects, to
-maintain essential invariants in practical applications. Consider a
-counter used to count video views on a YouTube-like application. The
-counter need not accurately track the view count, but it must at least
-ensure that the count does not decrease (nor does it appear to be
+transactions, even if they write to the same data objects, to maintain
+essential invariants in practical applications. Consider a counter
+used to count video views on a YouTube-like application. The counter
+need not accurately track the view count, but it must at least ensure
+that the count does not decrease (nor does it appear to be
 decreasing). Fig.~\ref{fig:motiv-eg-4} shows a simple example
-involving a view counter (\C{V}). The three `Inc' transactions
-(`Inc[1]' to `Inc[3]') increment the counter, while the transaction
-`Rd' reads the counter twice and writes the values read to \C{X} and
-\C{Y}, respectively. The count is 0 initially. Since we expect the
-view count to monotonically increase, we require $\C{Y}\ge\C{X}$ at
-the end. Under the standard setting, where new writes to data objects
-always overwrite previous writes, the postcondition may not hold. For
-instance, consider an execution where `Inc[1]' reads \C{V} as 0 (but
-doesn't commit), then `Inc[2]' and `Inc[3]' commit serially changing
-\C{V} to 2, followed by `Rd' reading 2 and writing it to \C{X}. Now
-`Inc[1]', which has read \C{V} as 0 writes 1 to \C{V} and commits. The
-subsequent read of \C{V} by `Rd' returns 1, which it writes to \C{Y}
-and commits. In the final state, \C{X} is 2 and \C{Y} is 1, thus
-violating the monotonicity invariant.
+involving a view counter (\C{V}). The three {\bf Inc} transactions
+({\bf Inc$_{1}$}, {\bf Inc$_{2}$}, and {\bf Inc$_{3}$}) increment the counter,
+while {\bf Rd} transaction reads the counter twice and writes the
+values read to \C{X} and \C{Y}, resp. The count is 0 initially. Since
+we expect the view count to monotonically increase, we require
+$\C{Y}\ge\C{X}$ after all transactions commit.  Under a sequentially
+consistent setting, where new writes to data objects always overwrite
+previous writes, executing the increment transactions under a weak
+isolation level can violate the monotonicity invariant, by allowing a
+transaction to read a stale value for \C{V}, a value that has already
+been updated by another.
 
-However, in an unconventional setting, such as a modern day replicated
-data store, writes arriving at a replica need not necessarily
-overwrite the existing writes. A \emph{conflict resolution strategy}
-may be employed to pick, for example, a write with the largest value
-among competing concurrent writes. Under this setting, the write to
-\C{V} by `Inc[1]' with a value (1) less than the current value (2) is
-discarded. Transaction `Rd' therefore continues to read 2, allowing it
-to satisfy the postcondition. Note that we did not have to serialize
-`Inc' transactions to uphold the monotonicity invariant, although they
-are involved in Write-Write conflicts.  The default \iso{Read
-Committed} isolation level suffices\footnote{We have implicitly assumed that
-`Rd' transaction witnesses monotonically growing state, i.e., the
-transactions that appear committed to the first read of \C{V} in `Rd'
-also appear committed to the second read.  Most implementations of RC
-enforce this \emph{Monotonic View} invariant. However,
-some~\cite{pldi15,bailishat} categorize RC with Monotonic View as a
-separate isolation level called \iso{Monotonic Atomic View} (MAV). On
-those stores, we need MAV for `Rd'. }, provided that the conflict
-resolution strategy consistently picks writes with larger values. A
-reasoning framework for weak isolation must be sensitive to the
-semantics of conflict resolution in order to formally prove the
+%% the post-condition may not hold. For instance,
+%% consider an execution where `{\bf Inc[1]}' reads \C{V} as 0 (but doesn't
+%% commit), then `{\bf Inc[2]}' and `Inc[3]' commit serially changing \C{V} to
+%% 2, followed by `Rd' reading 2 and writing it to \C{X}. Now `Inc[1]',
+%% which has read \C{V} as 0 writes 1 to \C{V} and commits. The
+%% subsequent read of \C{V} by `Rd' returns 1, which it writes to \C{Y}
+%% and commits. In the final state, \C{X} is 2 and \C{Y} is 1, thus
+%% violating the monotonicity invariant.
+
+However, when executed in a weakly consistent environment, such as
+those found in modern day replicated data stores, writes arriving at a
+replica need not necessarily overwrite existing writes. A
+\emph{conflict resolution strategy} may be employed to pick, for
+example, a write with the largest value among competing concurrent
+writes. Under this setting, the write to \C{V} by {\bf Inc$_{1}$} with a
+value less than the current value is discarded. Transaction
+{\bf Rd} therefore continues to read a value consistent with the invariant, allowing it to satisfy the
+post-condition. Note that we did not have to serialize any of the {\bf Inc}
+transactions to uphold the invariant, although they are
+all involved in write-write conflicts.  A \emph{read committed}
+isolation level suffices\footnote{We have implicitly assumed that `Rd'
+  transaction witnesses monotonically growing state, i.e., the
+  transactions that appear committed to the first read of \C{V} in
+  `Rd' also appear committed to the second read.  Most implementations
+  of RC enforce this \emph{Monotonic View} invariant. However,
+  some~\cite{pldi15,bailishat} categorize {\sc rc} with \emph{monotonic view} as a
+  separate isolation level called \emph{monotonic atomic view}
+  ({\sc mav}). On those stores, we need {\bf Rd} to execute under {\sc mav} isolation.}, provided that the
+conflict resolution strategy consistently picks writes with larger
+values. A reasoning framework for weak isolation must be sensitive to
+the semantics of conflict resolution in order to formally prove the
 monotonicity invariant in this example.

### 2016-06-15 motivation.tex
Motivation edits.

--- 2016-06-04 motivation.tex
@@
 
-In this section, we motivate the need for a formal reasoning framework
-for weakly isolated transactions through a series of examples. All our
-examples are written in C-like imperative language equipped with a
-\C{txn} lexical block for transactions. Each \C{txn} block is
-associated with a single-quoted string in angle braces that uniquely
-identifies the transaction. We use Hoare triple notation to annotate
-programs with pre and post conditions.
+In this section, we motivate our proof system for weak isolation
+through a series of examples written in \txnimp - a C-like imperative
+language equipped with a \C{txn} lexical block for transactions. Each
+\C{txn} block is associated with a single-quoted string in angle
+braces that uniquely identifies the transaction. We use Hoare triple
+notation to annotate programs with pre and post conditions.
 
@@
 % lead to the violation of invaraints. 
-Let us consider \iso{Read Committed} (RC), an isolation level weaker
-than SER that is default in Postgres 9.5~\cite{postgres95} and Oracle
-11g~\cite{oracle11g} databases. An RC transaction is isolated from the
-writes of an uncommitted transactions, thus preventing the so called
-\emph{dirty reads} phenomenon~\cite{berenson}. In the current example,
-RC isolation admits following executions:
+Let us consider \iso{Read Committed} (RC), a weak isolation level
+default in Postgres 9.5 and Oracle 11g databases. An RC transaction is
+isolated from the writes of an uncommitted transactions, thus
+preventing the so called \emph{dirty reads}
+phenomenon~\cite{berenson}. In the current example, RC isolation
+admits following executions on an SC store, such as a relational
+database:
 
@@
 commits, writing the new balance (\C{k-a2}). RC isolation has
-prevented `Wd2' from witnessing the writes of uncommitted transaction
-`Wd1'. Committing `Wd1' now leads to the loss of updates from `Wd2'
-(the so called \emph{lost update anomaly}~\cite{berenson}), resulting
-in a balance of \C{k-a1}. The second execution describes a similar
-scenario with `Wd1' and `Wd2' exchanging their roles.
+prevented the writes of uncommitted transaction `Wd1' from being
+\emph{visible} in `Wd2'. Committing `Wd1' now leads to the loss of
+updates from `Wd2' (the so called \emph{lost update
+anomaly}~\cite{berenson}), resulting in a balance of \C{k-a1}. The
+second execution describes a similar scenario with `Wd1' and `Wd2'
+exchanging their roles. Clearly, RC is not sufficient because it loses
+the updates of one transaction resulting in the violation of post
+condition.  We need an isolation level that prevents lost updates.
+\iso{Snapshot Isolation} (SI)~\cite{berenson} fits this requirement by
+its definition. SI implementations effectively serialize transactions
+that update a shared data object mostly by aborting and re-executing a
+transaction if write-write conflicts are detected during its commit.
+Since SI, unlike SER, need not necessarily rely on expensive
+lock-based concurrency control, it is also more efficient, making it
+appropriate appropriate for both `Wd' transactions.
 
-Clearly, \iso{Read Committed} is not sufficient because it loses the
-updates of one transaction and violates the invariant in
-Fig.~\ref{fig:motiv-eg-1}. Now, consider an alternative unfolding of
-Execution 1, where, rather than committing `Wd1' and overwriting the
-updates of `Wd2', we rollback `Wd1' to reexecute it starting from the
-state that contains the updates of `Wd2'. No updates are lost in this
-alternative execution, leading to the satisfaction of
-Fig.~\ref{fig:motiv-eg-1} invariant. Rolling back and reexecuting an
-uncommitted transaction to prevent write-write conflicts with a
-committed transaction is a strategy often adopted by databases to
-implement \iso{Snapshot Isolation} (SI) level. Indeed, SI is
-sufficient to enforce the required invaraints in this
-example\footnote{\iso{Parallel Snapshot Isolation} (PSI), a weaker
-version of SI tailor-made for replicated stores is also sufficient.}.
-Moreover, relying on optimistic speculation instead of pessimistic
-locking for concurrency control makes SI more efficient than SER,
-making it appropriate for transactions in Fig.~\ref{fig:motiv-eg-1}.
-Note that we have arrived at his judgment via adhoc reasoning based on
-speculations about database's implementation strategies. Such a
-reasoning process is necessarily fragile and error-prone. Is there a
-better way?
+Thinking in terms of anomalies, as described above, is how database
+programmers are often encouraged to reason about weak isolation.
+Unfortunately, such reasoning does not rest on any sound foundation,
+thus highly error-prone. Reasoning in terms of the implementations of
+weak isolation is not any better since it requires application
+programmers to understand and reason about low-level implementation
+details of the database (or, databases) that are far removed from the
+application semantics. An attractive alternative in this context is a
+formal proof system that combines declarative reasoning about
+isolation guarantees with operational reasoning about programs. We
+demonstrate how our proof system makes this possible in the context of
+the current example.
+
+Firstly, we note that the example in Fig.~\ref{fig:motiv-eg-1} is a
+concurrent program, hence admits \emph{rely-guarantee} style
+reasoning~\cite{rgjones}. Rely-guarantee is compositional proof
+technique that allows us to reason about one thread at a time by
+abstracting away the interference due to remaining threads
+(collectively called \emph{the environment}) into a \emph{rely}
+relation. In an ordinary concurrent program, every environment step is
+a valid interference in the current thread. However, in presence of
+transactions and variable isolation, determining what constitutes an
+interference and what does not is a non-trivial problem. For example,
+inside a serializable transaction no interference is a valid
+interference, whereas inside a transaction executing under \iso{Read
+Uncommitted}, the weakest isolation level, all interferences are
+valid. Between these two extremes are various levels of isolation
+that admit some interferences while prohibiting others. 
+% For example, \iso{Read Committed} isolation admits interference of a
+% committed transaction, but not that of an uncommitted transaction.
+% \iso{Snapshot Isolation} admits interference from committed
+% non-conflicting transactions, and so on. 
+The key aspect of our proof system is that it facilitates
+rely-guarantee reasoning while allowing the programmer to ignore any
+interference that is invalid as per the chosen isolation level or the
+store consistency level. For instance, in the current example, `Wd2'
+executes concurrently with `Wd1', hence latter's rely relation is
+required to model all possible interferences from `Wd1'. However,
+since RC prevents the effects of an uncommitted transaction from being
+visible, our proof system allows the programmer to effectively ignore
+the interference from `Wd2' unless the interference includes its
+commit. RC execution 1 shown above is an example of a case where
+interference from `Wd2' includes its commit. Such interference is not
+prohibited by RC, and the SC store makes \emph{all} effects of `Wd2'
+immediately visible. Our proof system insists that the
+programmer consider all effects of `Wd2' and conclude the value of
+\C{C} as \C{k-a2}, thus leading (rightfully) to the failure of proof.
+On the other hand, \iso{Snapshot Isolation} proscribes any
+interference from a concurrent transaction performing conflicting
+writes. The RC executions shown above are therefore invalid under SI.
+Accordingly, programmer is allowed to ignore any interference from
+`Wd2', thus facilitating the proof of postcondition.
+
+As evident from the above example, a proof system for weak isolation
+has to be flexible enough to admit different kinds of reasoning for
+different isolation and consistency levels. One way this can be
+achieved is by having a separate proof rule for each combination of
+transaction isolation level and store consistency level, such that the
+rule reflects the operational semantics of the isolation level on the
+store. Given the large and ever-increasing diversity among isolation
+and consistency semantics, this approach is unlikely to scale. Our
+proof system avoids this problem by being parametric over the
+semantics of isolation and consistency levels. This approach is based
+on a couple of observations. First, as noted by some
+authors~\cite{pldi15,gotsmanconcur15}, semantics of various isolation
+and consistency levels can be captured as well-formedness conditions
+over execution traces. If an execution trace satisfies specified
+well-formedness conditions, then the program is considered to have
+experienced the required levels of consistency and isolation in that
+execution. Second, well-formedness conditions are often
+\emph{prefix-complete}, meaning that every prefix of a well-formed
+trace is also well-formed, or conversely, an ill-formed trace cannot
+be extended to a well-formed one. Prefix-completeness allows us to
+reason about programs operationally, in terms of an imaginary machine
+whose primary artifact is an execution trace. The machine starts from
+a well-formed trace, executes a read/write/commit operation, and
+extends the trace while preserving well-formedness. If well-formedness
+cannot be preserved by taking a step, the machine gets ``stuck''. In
+executions that run to completion, the machine models a store whose
+consistency and isolation guarantees are captured by the trace
+well-formedness conditions. Thus, to show that program's invariants
+are preserved under weak isolation, it suffices to show that they are
+preserved in all executions of the imaginary machine that do not get
+stuck. This is indeed how our proof system facilitates program
+verification. The ``invalid interference'' described informally before
+corresponds to a case of machine getting stuck due to the violation of
+trace well-formedness conditions. The proof system inherits the
+parametricity of the machine over trace well-formedness conditions,
+leading to a generic set of parameterized proof rules that can be
+instantiated for various store consistency and transaction isolation
+levels.
+
+We now demonstrate how rely-guarantee reasoning, parametricity ,and
+trace well-formedness conditions facilitate the proof of the example
+in Fig.~\ref{fig:motiv-eg-1}. The trace-wellformedness condition
+corresponding to \iso{Snapshot Isolation} is shown below\footnote{This
+is a simplified version of the actual specification of SI
+(\S~\ref{sec:opsem}), which is a little more involved.}:
+\begin{smathpar}
+\begin{array}{lcl}
+\underE{\C{SI}(T_j)} & \Rightarrow & \forall T_i. T_i \neq T_j \conj \\
+  & & (\exists \C{X}.~\underE{T_i \wrstoar \C{X}} \conj 
+                  \underE{T_j \wrstoar \C{X}}) \Rightarrow \\
+  &  & \underE{T_i \hboar T_j} \disj \underE{T_j \hboar \C{COMMIT}(T_i)} \\
+\end{array}
+\end{smathpar}
+A transaction $T_j$ that writes to a variable $\C{X}$ has experienced
+snapshot isolation in an execution $\E$ only if every other
+transaction $T_i$ that also wrote to \C{X} in $\E$ either happened
+before (i.e., committed before) $T_j$, or executed concurrently but
+commited after $T_j$. Here, $\E$ denotes an execution trace, and
+$\underE{\phi}$ asserts that proposition $\phi$ about $\E$ is valid.
+Since we require both `Wd1' and `Wd2' to be executed under SI, the
+trace invariant ($\I$) for the example is $\C{SI(Wd1)} \wedge
+\C{SI(Wd2)}$. That is, if the machine extends an execution $\E$ that
+satisfies $\I$ (i.e.,$\underE{\I}$) to $\E'$, then $\E'$ also satisfies
+$\I$ (i.e., $\E' \Vdash \I$). Since we reason in terms of traces, pre
+and post conditions need to be written with respect to a trace.
 
@@
 \centering
-$\{\{\texttt{C+S}\ge\texttt{0}\}\}$
-\begin{tabular}{l||l}
 \begin{txnimpcode}
-  txn$\langle$'WdC'$\rangle${
-    if (C+S $\ge$ a1) {
-      C := C - a1
+ $\begin{decoration}
+ P_1:\{ {\neg\committed(\C{Wd2})} \Rightarrow \C{C = k} \conj\\
+        \hspace*{0.3in}{\committed(\C{Wd2})} \Rightarrow \C{Wd2}
+        \hboar \C{Wd1} \wedge \C{Wd2} \wrstoar \C{C} \wedge \C{C = k-a2} \}
+ \end{decoration}$
+  txn$\langle$'Wd1'$\rangle${
+   $\begin{decoration}
+    \phi_1 : \{{\neg\committed(\C{Wd2})} \Rightarrow \C{C = k} \conj
+      {\committed(\C{Wd2})} \Rightarrow \C{Wd2} \wrstoar \C{C} \conj\\
+       \hspace*{0.3in}{\committed(\C{Wd2})} \wedge
+        {\C{Wd2} \hboar \C{Wd1}} 
+       \Rightarrow \C{C = k-a2} \}
+    \end{decoration}$ 
+    v1 = C
+   $\begin{decoration}
+    \phi_2 : \{{\neg\committed(\C{Wd2})} \Rightarrow \C{C = k} \wedge \C{v1 = k} \conj\\
+       \hspace*{0.3in}{\committed(\C{Wd2})} \Rightarrow \C{Wd2} \wrstoar \C{C} \conj \\
+       \hspace*{0.3in}{\committed(\C{Wd2})} \wedge
+        {\C{Wd2} \hboar \C{Wd1}} 
+       \Rightarrow \C{C = k-a2} \wedge \C{v1 = k-a2}\}
+    \end{decoration}$ 
+    if (v1 $\ge$ a1) {
+      v2 = C;
+     $\begin{decoration}
+      \phi_3 : \{{\neg\committed(\C{Wd2})} \Rightarrow \C{C = k} \wedge \C{v2 = k} \conj\\
+       \hspace*{0.3in}{\committed(\C{Wd2})} \Rightarrow \C{Wd2} \wrstoar \C{C} \conj \\
+         \hspace*{0.3in}{\committed(\C{Wd2})} \wedge
+          {\C{Wd2} \hboar \C{Wd1}} \Rightarrow \C{C = k-a2} \wedge \\
+          \hspace*{1.9in}\C{v2 = k-a2}\}
+      \end{decoration}$ 
+      v3 = v2 - a1;
+     $\begin{decoration}
+      \phi_4 : \{{\neg\committed(\C{Wd2})} \Rightarrow \C{C = k} \wedge \C{v3 = k-a1} \conj\\
+         \hspace*{0.3in}{\committed(\C{Wd2})} \Rightarrow \C{Wd2} \wrstoar \C{C} \conj \\
+         \hspace*{0.3in}{\committed(\C{Wd2})} \wedge
+          {\C{Wd2} \hboar \C{Wd1}} 
+         \Rightarrow \C{C = k-a2} \wedge \\
+         \hspace*{1.9in}\C{v3 = k-a2-a1}\}
+      \end{decoration}$ 
+      C := v3
+     $\begin{decoration}
+      \phi_5 : \{{\neg\committed(\C{Wd2})} \Rightarrow \C{C = k-a1}) 
+                \conj\\
+         \hspace*{0.3in}{\committed(\C{Wd2})} 
+                \Rightarrow \C{C = k-a1-a2}\}
+      \end{decoration}$ 
     }
   }
+ $\begin{decoration}
+  Q_1 : \{{\neg\committed(\C{Wd2})} \Rightarrow \C{C = k-a1}
+            \conj \committed(\C{Wd1}) \conj\\
+      \hspace*{0.12in}{\committed(\C{Wd2})} 
+          \Rightarrow \C{C = k-a1-a2} \}
+  \end{decoration}$ 
 \end{txnimpcode}
-&
-\begin{txnimpcode}
-  txn$\langle$'WdS'$\rangle${
-    if (C+S $\ge$ a2) {
-      S := S - a2
-    }
-  }
-\end{txnimpcode}
-\\
-\end{tabular}
-$\{\{\texttt{C+S}\ge\texttt{0}\}\}$
 
-\caption{Withdraws from current (\C{C}) and savings (\C{S}) accounts.}
-\label{fig:motiv-eg-2}
+\caption{`Wd1' transaction decorated with assertions}
+\label{fig:wd1-decorated}
 \end{figure}
 
-\iso{Snapshot Isolation} effectively serializes updates to a data
-object without necessarily relying on expensive lock-based concurrency
-control. While this makes it a better alternative to
-\iso{Serializable} isolation, it is however not a replacement for the
-latter. Consider a different banking application that maintains a
-current (\C{C}) and savings account (\C{S}) for each user. A user can
-withdraw either from current or from savings account as long as the
-combined balance is at least as much as the amount being withdrawn.
-Fig.~\ref{fig:motiv-eg-2} shows a pair of concurrent transactions,
-`WdC' and `WdS', performing withdraws from current and savings
-accounts, respectively. Pre and post conditions assert non-negative
-combined balance invariant. \iso{Serializable} isolation is sufficient
-to preserve the invariant, but \iso{Snapshot Isolation} is not.
-Intuitively, this is because there are no \emph{lost updates} in this
-example even when both transactions commit concurrently, hence there
-is no reason for SI to effectively serialize them. A reasoning
-framework for weak isolation should let us formally deduce this fact
-by making it impossible to construct a proof of correctness for the
-program under SI, but facilitating a proof under SER.
+The fully decorated implementation of transaction `Wd1' is shown in
+Fig.~\ref{fig:wd1-decorated}. Decorated `Wd2' is similar. Assignment
+statements are broken down and temporary local variables (\C{v1},
+\C{v2} and \C{v3}) are introduced so as to separate shared variable
+reads and writes.  Since we reason in terms of executions, all
+assertions implicitly refer to the current execution (just as hoare
+triples implicitly refer to the current state). Also implicit is the
+invariant \C{k $\ge$ a1+a2}. Proposition $\committed(T)$ indicates
+that the transaction $T$ is already committed. Precondition ($P_1$) of
+`Wd1' accounts for the possibility of `Wd2' committing before `Wd1',
+writing \C{k-a2} to \C{C}.  Precondition of `Wd2' is similar.  Since
+neither `Wd1' nor `Wd2' are committed at the beginning, the
+precondition in Fig.~\ref{fig:motiv-eg-1} is extended to include
+$\neg\committed(\C{Wd1}) \wedge \neg\committed(\C{Wd2})$, from which
+$P_1$ follows. Once the execution is inside `Wd1', commit of `Wd2'
+($\committed(\C{Wd2})$) may mean that either `Wd2' happened before
+`Wd1', or that it committed concurrently with `Wd1'. Since SI
+proscribes the latter possibility, we only consider the case when
+$\C{Wd2} \hboar \C{Wd1}$. A proof for $\C{Wd2} \hboar \C{Wd1}$ is
+obtained subsequently, allowing us to get rid of the special case.
 
-\begin{figure}[t]
-\centering
-$\{\{\texttt{C+S}\ge\texttt{0}\}\}$
-\begin{tabular}{l||l||}
-\begin{txnimpcode}
-  txn$\langle$'Xfer'$\rangle${
-    C := C - a;
-    S := S + a;
-  }
-\end{txnimpcode}
-&
-\begin{txnimpcode}
-  txn$\langle$'TotBal'$\rangle${
-    T := S + C
-  }
-\end{txnimpcode}
-% \\
-% \multicolumn{2}{c} {
-% }
-\end{tabular}
-\begin{tabular}{c}
-  \begin{txnimpcode}
-    txn$\langle$'AddInt'$\rangle${
-      S := S + (T*0.1)
-    }
-  \end{txnimpcode}
-\end{tabular}
+The first obligation is to show that the assertions that decorate
+`Wd1' are indeed valid. This entails proving that, for each statement
+of `Wd1', if evaluating the statement and extending an execution that
+satisfies its precondition results in a well-formed execution, then
+the extended execution must satisfy the postcondition. The
+well-formedness assumption lets us ignore the executions that violate
+the trace invariant lead the machine to a stuck state. This is useful
+in case of the assignment statement that writes \C{v3} to \C{C}. Since
+`Wd2' also writes to \C{C}, the trace invariant ($\I$) requires
+$\C{Wd2} \hboar \C{Wd1}$ if $\committed(\C{Wd2})$ (the alternative
+$\C{Wd1} \hboar \C{COMMIT}(\C{Wd2})$ allowed by \C{SI(Wd1)} is
+impossible in this case because \C{Wd2} has already committed, while
+\C{Wd1} is still in progress). This lets us treat the special case
+$\committed(\C{Wd2})\conj \C{Wd2} \hboar \C{Wd1}$ as being equivalent
+to the general case $\committed(\C{Wd2})$, which is required to prove
+the postcondition of `Wd1'. Observe that \C{RC(Wd1)} does not justify
+this generalization, hence replacing \C{SI(Wd1)} by \C{RC(Wd1)} is not
+sufficient to prove the postcondition of `Wd1'.
 
-$\{\{\texttt{C+S}\ge\texttt{0}\}\}$
-
-\caption{Concurrent transfer (\C{Xf}) and interest accumulation (\C{Int}).}
-\label{fig:motiv-eg-3}
-\end{figure}
-
-The previous two examples preserve invariants on \C{C} and \C{S} by
-effectively serializing transactions that write to \C{C} and \C{S}.
-However, in general, serializing a subset of transactions that write
-to the data objects referred by invariants is neither sufficient nor
-necessary to uphold the invariants. Consider the banking application
-in Fig.~\ref{fig:motiv-eg-3} with three transactions. Transaction
-`Xfer' transfers some amount (\C{a}) from current (\C{C}) to savings
-(\C{S}) account.  Transaction `TotBal' records the total balance in
-both the accounts in a data object \C{T}, and transaction `AddIntr'
-calculates interest on total balance (10\% of \C{T}) and adds it to
-the savings account. Using a data object to record some function of
-the current state of other data objects is a standard practice in
-relational databases, where Materialized Views~\cite{oraclematview}
-are created to maintain the results of a SQL query. Like the previous
-example, the invariant of interest in this example is non-negative
-total balance; if we start with a non-negative total balance, and we
-add interest based on the total balance, we must end up with a
-non-negative total balance.  However, unlike the previous example,
-serializing transactions that write to \C{C} and \C{S} is not
-sufficient to guarantee the invariant.  The following execution
-demonstrates this anomaly:
-
-\begin{figure}[!h]
-\centering
-\includegraphics[scale=0.5]{Figures/motiv-eg-3-a}
-
-\label{fig:motiv-eg-3-a}
-\end{figure}
-
-The subscripts X, TB and AI denote `Xfer', `TotBal' and `AddInt'
-transactions, respectively. Assume that \C{C} and \C{S} are initially
-10 and 0, respectively. Let \C{a} be 200, i.e., we subtract 200 from
-\C{C} and add it to \C{S}.  This operation must be allowed because it
-doesn't effect the total balance, hence the interest. The execution
-begins with `TotBal' transaction reading the savings (\C{S}) balance
-as 0 (we assume left-to-right evaluation order for \C{S + C}).
-However, before it reads the current (\C{C}) balance, `Xfer'
-transaction executes and commits, changing \C{C} to -190.
-Subsequently, `TotBal' reads -190 from \C{C}, and updates \C{T} to
--190. `AddInt', which begins execution after `TotBal', reads -190 from
-\C{T}, calculates interest as -19, and updates \C{S} to 181. The
-result is a state with total balance as -190 + 181 = -9, thus
-violating the invariant.
-
-Note that transactions `Xfer' and `AddInt', which write to the data
-objects (\C{C} and \C{S}) relevant to the invariant, are serialized in
-the above execution. Transaction `TotBal' however is not. While
-serializing it fixes the problem, it is not necessary. A simpler fix
-is to isolate `TotBal' transaction from the effects of concurrent
-`Xfer' transaction, so that it reads \C{C} as 10 despite the
-concurrent `Xfer' transaction committing -190 to \C{C}. This can be
-achieved by executing `TotBal' under \iso{Repeatable Read} (RR)
-isolation level, which guarantees that any two reads of a data object
-in a transaction return the same value (the \emph{repeatable reads}
-property~\cite{berenson}). Allowing `Xfer' to interfere with `TotBal'
-violates the repeatable read property since a hypothetical read from
-\C{S} in `TotBal' after the interference returns a different value
-(200) than the first read (0). Executing `TotBal' under RR restores
-repeatable read by \emph{effectively} prohibiting such interference.
-
-\begin{figure}[t]
-\centering
-\hspace*{-0.3in}$\{\{\texttt{V=0}\}\}$
-\begin{tabular}{l||l}
-\begin{txnimpcode}
-  txn$\langle$'Rd'$\rangle${
-    X := V
-    Y := V
-  }
-\end{txnimpcode}
-&
-\begin{txnimpcode}
-  cobegin (i $\leftarrow$ [1$\ldots$3]) {
-    txn$\langle$'Inc['+i+']'$\rangle${
-      V := V + 1
-    }
-  }
-\end{txnimpcode}
-% &
-% \begin{txnimpcode}
-%   txn$\langle$'Inc2'$\rangle${
-%     V := V + 1
-%   }
-% \end{txnimpcode}
-\\
-\end{tabular}
-\hspace*{-0.3in}$\{\{\texttt{Y}\ge\texttt{X}\}\}$
-
-\caption{View counter (\C{V}) increments (\C{Inc}) and reads (\C{Rd})}
-\label{fig:motiv-eg-4}
-\end{figure}
-
-Our last example demonstrates that it is not necessary to serialize
-transactions, even those that write to the same data objects, to
-maintain essential invariants in practical applications. Consider a
-counter used to count video views on a YouTube-like application. The
-counter need not accurately track the view count, but it must at least
-ensure that the count does not decrease (nor does it appear to be
-decreasing). Fig.~\ref{fig:motiv-eg-4} shows a simple example
-involving a view counter (\C{V}). The three `Inc' transactions
-(`Inc[1]' to `Inc[3]') increment the counter, while the transaction
-`Rd' reads the counter twice and writes the values read to \C{X} and
-\C{Y}, respectively. The count is 0 initially. Since we expect the
-view count to monotonically increase, we require $\C{Y}\ge\C{X}$ at
-the end. Under the standard setting, where new writes to data objects
-always overwrite previous writes, the postcondition may not hold. For
-instance, consider an execution where `Inc[1]' reads \C{V} as 0 (but
-doesn't commit), then `Inc[2]' and `Inc[3]' commit serially changing
-\C{V} to 2, followed by `Rd' reading 2 and writing it to \C{X}. Now
-`Inc[1]', which has read \C{V} as 0 writes 1 to \C{V} and commits. The
-subsequent read of \C{V} by `Rd' returns 1, which it writes to \C{Y}
-and commits. In the final state, \C{X} is 2 and \C{Y} is 1, thus
-violating the monotonicity invariant.
-
-However, in an unconventional setting, such as a modern day replicated
-data store, writes arriving at a replica need not necessarily
-overwrite the existing writes. A \emph{conflict resolution strategy}
-may be employed to pick, for example, a write with the largest value
-among competing concurrent writes. Under this setting, the write to
-\C{V} by `Inc[1]' with a value (1) less than the current value (2) is
-discarded. Transaction `Rd' therefore continues to read 2, allowing it
-to satisfy the postcondition. Note that we did not have to serialize
-`Inc' transactions to uphold the monotonicity invariant, although they
-are involved in Write-Write conflicts.  The default \iso{Read
-Committed} isolation level suffices\footnote{We have implicitly assumed that
-`Rd' transaction witnesses monotonically growing state, i.e., the
-transactions that appear committed to the first read of \C{V} in `Rd'
-also appear committed to the second read.  Most implementations of RC
-enforce this \emph{Monotonic View} invariant. However,
-some~\cite{pldi15,bailishat} categorize RC with Monotonic View as a
-separate isolation level called \iso{Monotonic Atomic View} (MAV). On
-those stores, we need MAV for `Rd'. }, provided that the conflict
-resolution strategy consistently picks writes with larger values. A
-reasoning framework for weak isolation must be sensitive to the
-semantics of conflict resolution in order to formally prove the
-monotonicity invariant in this example.
+The second obligation is to prove that assertions are valid despite
+the interference from the the concurrent thread executing `Wd2'. The
+interference is given by the following rely relation ($R_1$):
+\begin{smathpar}
+\begin{array}{lcl}
+  R_1 & = & \{ (\E,\E') \;|\; \neg\underE{\committed(\C{Wd2})} \conj 
+        \underE{\I} \conj \E'\Vdash \I \conj\\
+%       \conj (\E'-\E) \subseteq \C{Wd2} \conj \\
+%   & & \hspace*{0.5in} \E' \Vdash \committed(\C{Wd2}) ~\Rightarrow~ \E'
+%       \Vdash \C{Wd2} \wrstoar \C{C} \conj \\ 
+%   & & \hspace*{0.5in} \underE{\committed(\C{Wd1})} \conj \E' \Vdash
+%   \committed(\C{Wd2}) \Rightarrow \C{C=k-a1-a2} \\
+    & & \hspace*{0.5in} \neg\underE{\committed(\C{Wd1})} \conj
+        \C{COMMIT(Wd2)} \in (\E'-\E) \\
+    & & \hspace*{0.8in}\Rightarrow \C{C=k-a2} \conj 
+        \E' \Vdash \C{Wd2} \wrstoar \C{C} \conj \\
+    & & \hspace*{0.5in} \underE{\committed(\C{Wd1})} \conj
+        \C{COMMIT(Wd2)} \in (\E'-\E) \\
+    & & \hspace*{0.6in}\Rightarrow \C{C=k-a1-a2} \conj
+        \E' \Vdash \C{Wd2} \wrstoar \C{C} \}\\
+\end{array}
+\end{smathpar}
+The rely relation says the following about the concurrent thread: (1).
+It may interfere only if `Wd2' is not already committed, (2). Any
+interference takes well-formed executions to well-formed executions,
+(3). If the interference commits `Wd2', then `Wd2' should have already
+written to \C{C}, and (4). The value written is either $\C{k-a1-a2}$
+or $\C{k-a1}$ depending on whether or not `Wd1' has already committed.
+Interference from `Wd2' is allowed between any two operations of
+`Wd1', and assertions are required to be valid despite the
+interference. For instance, if the execution ($\E$) immediately after
+evaluating $\C{v1 = C}$ satisfies $\phi_2$ (i.e., $\underE{\phi_2}$),
+and an interference from `Wd2' extends $\E$ to $\E'$ (i.e.,
+$R_1(\E,\E')$), then the execution $\E'$ just before evaluating the
+condition must satisfy $\phi_2$ (i.e., $\E' \Vdash \phi_2$). Indeed,
+all $\phi_i$'s do satisfy this \emph{stability} condition, thus
+lending the validity to the proof.

### 2016-06-15 abstract.tex

@@
   Serializability is a well-understood mechanism that simplifies
-  reasoning about the behavior of concurrent transactions.  However,
-  its two key properties, \emph{atomicity} - a transaction's effects
-  are seen only when it commits, and \emph{isolation} - a transaction
-  never witnesses the effects of other committed transactions while it
-  is executing, impose significant constraints on implementations, and
-  can negatively impact database performance by restricting
-  opportunities to exploit concurrency.
+  reasoning about the behavior of concurrent transactions by ensuring
+  an executing transaction is \emph{isolated} from the effects of
+  other concurrenty executing transactions once it begins executing.
+  However, enforcing serializabity comes at the cost in performance
+  because it necessarily restricts opportunities to exploit
+  concurrency even when such opportunities would not violate
+  application-specific invariants.
 
@@
   Unfortunately, the semantics of weak isolation is often poorly
-  understood, and usually explained informally in terms of low-level
-  implementation artifacts.  Consequently, verifying high-level
-  correctness properties in such environments can be problematic.
+  understood, and usually explained only informally in terms of
+  low-level implementation artifacts.  Consequently, verifying
+  high-level correctness properties in such environments is
+  challenging.
 
-  To address this issue, this paper presents a novel program logic
-  that enables compositional reasoning about the behavior of
-  concurrently executing weakly-isolated transactions.  Notably, our
-  development is parametric over a transaction's specific isolation
-  semantics, and the consistency guarantees provided by the underlying
-  data store, allowing it to be applicable over a large range of
-  different isolation and consistency models.  Case studies and
-  experiments on real-world applications demonstrate the utility of
-  our approach, and provide strong evidence that weakly-isolated
-  transactions can be placed on the same formal footing as their
-  strongly-isolated serializable counterparts.
+  To address this issue, this paper presents a program logic that
+  enables compositional reasoning about the behavior of concurrently
+  executing weakly isolated transactions.  Notably, our development is
+  parametric over a transaction's specific isolation semantics, and
+  the consistency guarantees provided by the underlying data store,
+  allowing it to be applicable over a large range of concurrency
+  control mechanisms.  Case studies and experiments on real-world
+  applications demonstrate the utility of our approach, and provide
+  strong evidence that weakly isolated transactions can be placed on
+  the same formal footing as their strongly-isolated serializable
+  counterparts.
 

### 2016-06-15 introduction.tex

@@
 objects into a coherent unit, equipped with a set of four key
-properties - atomicity, consistency, isolation, and durability
-(ACID). ACID has been the gold standard for database transactions for
-many decades, providing programmers with a simple implementation
-independent model to reason about concurrent transactions.  For
-example, concurrency control mechanisms typically enforce isolation by
-making all transactions \emph{serializable}, to ensure that any
-allowed schedule is equivalent to one in which there is no
-interleaving of actions from different concurrently executing
-transactions.
+properties - atomicity, consistency, isolation, and durability (ACID).
+Concurrency control mechanisms provide specific instantiations of
+these properties to yield different ACID variants that characterize
+how and when the effects of concurrently executing transactions become
+visible to one another.  \emph{Serializability} is a particularly
+well-studied instantiation that imposes strong atomicity and isolation
+constraints on transaction execution, ensuring that any permissible
+concurrent schedule is equivalent to a serial one in which in which no
+interleaving of actions from different transactions are allowed.
 
-However the strong guarantees provided by serializability does not
-come for free - pessimistic concurrency control methods, for example,
-require databases to provide expensive techniques such as two-phase
-locking that incur overhead to deal with deadlocks, rollbacks, and
-re-execution~\cite{twopl,ullmanbook}.  Similar criticisms apply to
-optimistic multi-version concurrency control methods that must deal
-with timestamp and version management~\cite{BG81}.  Regardless of the
-method employed, enforcing serializability in a replicated data store
-additionally requires global coordination among geo-distributed
-replicas, which makes the store \emph{unavailable} in the presence of
-inevitable network
-partitions~\cite{cap,sernotavlbl,bailishat,bernsigmod13}.
+The guarantees provided by serializability, and strong isolation in
+particular, do not come for free, however - pessimistic concurrency
+control methods, for example, require databases to provide expensive
+techniques such as two-phase locking that incur overhead to deal with
+deadlocks, rollbacks, and re-execution~\cite{twopl,ullmanbook}.
+Similar criticisms apply to optimistic multi-version concurrency
+control methods that must deal with timestamp and version
+management~\cite{BG81}.  Regardless of the particular method employed,
+enforcing serializability in a replicated data store additionally
+requires global coordination among geo-distributed replicas, which
+makes the store \emph{unavailable} in the presence of inevitable
+network partitions~\cite{cap,sernotavlbl,bailishat,bernsigmod13}.
 
@@
 beginning as early as 1976~\cite{gray1976}. The ANSI SQL 92 standard
-defines three such weak isolation \emph{levels} which are now
+defines three such weak isolation levels which are now
 implemented in many relational and NoSQL databases. Not surprisingly,
@@
 together, these studies make clear that weakly isolated transactions
-are quite prevalent in practice.
+are quite prevalent in practice, and strongly isolated (serializable)
+transactions are often eschewed.
 
@@
 with a microbenchmark suite that executes transactions under a weakly
-isolated \iso{read committed} isolation level - the default level for
+isolated \emph{read committed} isolation level - the default level for
 8 of the 18 databases studied in~\cite{bailishotos}, and found that 25
 out of every 1000 rows in the database violate at least one integrity
-constraint. Bailis \empH{et al.}~\cite{bailisferal} rely on Rails's
+constraint. Bailis \emph{et al.}~\cite{bailisferal} rely on Rails'
 \emph{uniqueness validation} to maintain uniqueness of records while
@@
 more than 10 duplicate records. Rails relies on database transactions
-to validate uniqueness during insertions, which is ok if transactions
-are serializable, but incorrect under \iso{Read Committed} isolation
-used in the experiments. The same study has found that 13\% of all
-invariants among 67 open source Ruby-on-Rails applications are at risk
-of being violated due to weak isolation. Indeed, incidents of safety
-violations due to weak isolation have been reported on web services in
-production~\cite{starbucksbug, scimedbug}, including in
-safety-critical applications such as bitcoin
+to validate uniqueness during insertions, which is sensible if
+transactions are serializable, but incorrect under the \emph{read
+  committed} isolation level used in the experiments. The same study
+has found that 13\% of all invariants among 67 open source
+Ruby-on-Rails applications are at risk of being violated due to weak
+isolation. Indeed, incidents of safety violations due to executing
+applications in a weakly isolated environment have been reported on
+web services in production~\cite{starbucksbug, scimedbug}, including
+in safety-critical applications such as bitcoin
 exchanges~\cite{poloniexbug, bitcoinbug}. While enforcing
@@
 proposal~\cite{gray1976} defines multiple ``degrees'' of weak
-isolation in terms of implementation details, such as the nature and
+isolation in terms of implementation details such as the nature and
 duration of locks held in each case. The ANSI SQL 92 standard four
-levels of isolation (including serializability) in terms of sevaral
+levels of isolation (including serializability) in terms of various
 undesirable \emph{phenomena} (\eg \emph{dirty reads}) each is required
@@
 the phenomenon can be allowed without violating application
-invariants. This is understandably hard. Adya~\cite{adyaphd} includes
+invariants. This is understandably hard. Adya~\cite{adyaphd} presents
 the first formal definitions of some of well-known isolation levels in
@@
 Recent results on reasoning about application invariants in the
-presence of weak consistency~\cite{burckhardt14, redblueosdi,
+presence of weak \emph{consistency}~\cite{burckhardt14, redblueosdi,
   redblueatc, ecinec, gotsmanpopl16} provide insights on possible ways
@@
 former.  At a high level, reasoning about weak consistency involves
-reasoning about the concurrent execution of atomic and strongly
+reasoning about the concurrent execution of strongly atomic and
 isolated operations against different but consistent states of the
-database, where each operation locally preserves application
-invariants. In contrast, reasoning about weak isolation involves
-reasoning about concurrent execution of transactions, whose operations
-witness different states, which, although consistent individually, may
-not be obviously reconciled into a consistent view. Moreover,
-operations may violate invariants as long as the transaction as a
-whole preserves them.  Intuitively, the latter problem reduces to the
-former if all transactions contain a single operation. Furthermore,
-weak isolation can exist independent of weak consistency, as evident
-from the presence of weakly isolated transactions on sequentially
-consistent relational databases; such systems existed long before the
-advent of weakly consistent NoSQL stores.  Thus, frameworks for
-reasoning about weak isolation will necessarily have to generalize the
-reasoning frameworks developed for weak consistency in important ways.
+database or data store, where each operation locally preserves
+application invariants. In contrast, reasoning about weak isolation
+involves reasoning about the concurrent execution of transactions,
+whose operations witness different states which, although consistent
+individually, may not be obviously reconciled into a consistent global
+view.  Moreover, operations may violate invariants as long as the
+transaction as a whole preserves them.  Intuitively, the latter
+problem reduces to the former if all transactions contain a single
+operation. Furthermore, weak isolation can exist independent of weak
+consistency, as evident from the presence of weakly isolated
+transactions on sequentially consistent relational databases; such
+systems existed long before the advent of weakly consistent NoSQL
+stores.  Thus, frameworks for reasoning about weak isolation will
+necessarily have to generalize the reasoning frameworks developed for
+weak consistency in new and important ways.
 
@@
 build correctness proofs for transactional programs in the presence of
-a weakly isolated concurrency control mechanism.  A key novelty of our
+a weakly isolated concurrency control mechanism  A key novelty of our
 approach is that it is parametric over the \emph{isolation semantics}
 of transactions in the program, as well as the \emph{consistency
-  semantics} of the store. In concrete terms, this means that, unlike
-recent work focussed on formalizing the semantics of weak
+  semantics} of the underlying store. In concrete terms, this means
+that, unlike recent work focussed on formalizing the semantics of weak
 consistency~\cite{gotsmanpopl16, redblueatc, ecinec}, our system model
-does not assume a minimum consistency/isolation level, nor does it
+does not assume a minimum consistency or isolation level, nor does it
 assume a predefined set of such levels.  Instead, our operational
@@
 a variety of stores (\eg the sequentially consistent store
-of~\cite{adyaphd}, or causally consistent store
+of~\cite{adyaphd}, or the causally consistent store
 of~\cite{gotsmanpopl16}, etc).  Our key technical contribution is thus
@@
 invariants when executed on a store satisfying certain consistency
-guarantees.
+guarantees (e.g., causal visibility).
 
@@
   \item We develop a semantics for a core language equipped with
-    weakly isolated transactions.  Building on some ideas espoused in
-    previous work, we show that isolation semantics can be captured as
-    well-formedness constraints on a program's execution.
+    weakly isolated transactions, demonstrating that that a general
+    isolation semantics can be expressed as well-formedness constraints
+    on a program's execution.
   \item We propose a compositional proof system for this language
@@
     requirements, and use it to instantiate our operational model and
-    the proof system for various kinds of stores.
+    the proof system for various kinds of (weakly consistent) data
+    stores.
   \item Case studies modeled after real-world scenarios demonstrate
@@
 
-Our semantic framework and proof methodology thus provide the first
-(to the best of our knowledge) mechanism that precisely relates
-high-level program invariants to low-level store consistency and weak
-isolation guarantees, thereby allowing weakly isolated transactions to
-enjoy the same rigorous reasoning capabilities as their strongly
-isolated (serializable) counterparts.
+Our results thus provide the first (to the best of our knowledge)
+mechanism that precisely relates high-level program invariants to
+low-level store consistency and weak isolation guarantees, thereby
+allowing weakly isolated transactions to enjoy the same rigorous
+reasoning capabilities as their strongly isolated (serializable)
+counterparts.

### 2016-06-15 motivation.tex

@@
 
-In this section, we motivate our proof system for weak isolation
-through a series of examples written in \txnimp - a C-like imperative
-language equipped with a \C{txn} lexical block for transactions. Each
+In this section, we motivate our proof system for weak isolation via
+an example written in \txnimp - a C-like imperative language equipped
+with a \C{txn} lexical block defining a transaction scope. Each
 \C{txn} block is associated with a single-quoted string in angle
 braces that uniquely identifies the transaction. We use Hoare triple
-notation to annotate programs with pre and post conditions.
+notation to annotate programs with pre- and post- conditions.
 
@@
 
-% Most weak isolation levels allow executions to interleave operations
-% (Reads, Writes and Commits) of one transaction with other's, while
-% enforcing some additional constraints. Arbitrary interleavings may
-% lead to the violation of invaraints. 
-Let us consider \iso{Read Committed} (RC), a weak isolation level
-default in Postgres 9.5 and Oracle 11g databases. An RC transaction is
-isolated from the writes of an uncommitted transactions, thus
-preventing the so called \emph{dirty reads}
-phenomenon~\cite{berenson}. In the current example, RC isolation
-admits following executions on an SC store, such as a relational
-database:
+As an alternative, consider the execution of this transaction under a
+\emph{read committed} ({\sc rc}) isolation level, which is weaker than
+{\sc ser}\footnote{{\sc rc} is in fact the default isolation level in
+Postgres 9.5 and Oracle 11g databases.} An {\sc rc} transaction is
+isolated from the writes of uncommitted transactions, thus preventing
+the transaction from witnessing \emph{dirty reads}~\cite{berenson},
+reads that observe the effects of non-committed transactions. In the
+current example, {\sc rc} isolation admits the two executions shown in
+Fig.~\ref{fig:rc-ex} on an SC store, such as a relational database:
 
@@
 \end{figure}
-
-An execution is a series of Read, Write and Commit operations.
-Subscript of an operation indicates its transaction. For clarity,
-effects of different transactions are shown against differently
-colored backgrounds. In the first execution, the first transaction
-(`Wd1') reads the current balance (\C{k}) and writes the new balance
-(\C{k-a1}), but before it commits transaction `Wd2' executes and
-commits, writing the new balance (\C{k-a2}). RC isolation has
-prevented the writes of uncommitted transaction `Wd1' from being
-\emph{visible} in `Wd2'. Committing `Wd1' now leads to the loss of
-updates from `Wd2' (the so called \emph{lost update
-anomaly}~\cite{berenson}), resulting in a balance of \C{k-a1}. The
-second execution describes a similar scenario with `Wd1' and `Wd2'
-exchanging their roles. Clearly, RC is not sufficient because it loses
-the updates of one transaction resulting in the violation of post
-condition.  We need an isolation level that prevents lost updates.
-\iso{Snapshot Isolation} (SI)~\cite{berenson} fits this requirement by
-its definition. SI implementations effectively serialize transactions
-that update a shared data object mostly by aborting and re-executing a
-transaction if write-write conflicts are detected during its commit.
-Since SI, unlike SER, need not necessarily rely on expensive
-lock-based concurrency control, it is also more efficient, making it
-appropriate appropriate for both `Wd' transactions.
+The figure depicts an execution as a series of read, write, and commit
+operations.  The subscript of an operation indicates the transaction
+that executes it\footnote{For clarity, the effects of different
+transactions are shown in different colored backgrounds.} In the first
+execution, transaction {\bf Wd1} reads the current balance (\C{k}) and
+writes the new balance (\C{k-a1}), but before it commits transaction
+{\bf Wd2} executes and commits, writing the new balance (\C{k-a2}). RC
+isolation prevent {\bf Wd2} from witnessing the uncommitted writes of
+transaction {\bf Wd1}.  Subsequently committing {\bf Wd1} leads to the
+loss of {\bf Wd2}'s updates (the so called \emph{lost update
+anomaly}~\cite{berenson}), resulting in an incorrect balance of
+\C{k-a1}. The second execution describes a similar scenario with {\bf
+Wd1} and {\bf Wd2} exchanging their roles.  Clearly, RC is not
+sufficient because it loses the updates of one transaction resulting
+in the violation of post condition.  We need an isolation level that
+prevents lost updates.  \iso{Snapshot Isolation} (SI)~\cite{berenson}
+fits this requirement by its definition. SI implementations
+effectively serialize transactions that update a shared data object
+mostly by aborting and re-executing a transaction if write-write
+conflicts are detected during its commit.  Since SI, unlike SER, need
+not necessarily rely on expensive lock-based concurrency control, it
+is also more efficient, making it appropriate appropriate for both
+`Wd' transactions.
 

### 2016-06-16 motivation.tex
Finally motivated!

@@
 
+\begin{figure}
+\centering
+  \includegraphics[scale=0.4]{Figures/motiv-eg-1-hb}
+
+\label{fig:motiv-eg-1-hb}
+\caption{$\hbZ$ arrows allowed by SI (solid) and RC (solid and dashed)
+isolation levels for the example in Fig.~\ref{fig:motiv-eg-1}. Each
+arrow represents $\hbZ$ relationship between all operations of `Wd2'
+and operations in `Wd1' that follow the arrow head. Letters R, W and C
+stand for read, write and commit, respectively.}
+\end{figure}
+
 Firstly, we note that the example in Fig.~\ref{fig:motiv-eg-1} is a
@@
 % non-conflicting transactions, and so on. 
-The key aspect of our proof system is that it facilitates
-rely-guarantee reasoning while allowing the programmer to ignore any
-interference that is invalid as per the chosen isolation level or the
-store consistency level. For instance, in the current example, `Wd2'
-executes concurrently with `Wd1', hence latter's rely relation is
-required to model all possible interferences from `Wd1'. However,
-since RC prevents the effects of an uncommitted transaction from being
-visible, our proof system allows the programmer to effectively ignore
-the interference from `Wd2' unless the interference includes its
-commit. RC execution 1 shown above is an example of a case where
-interference from `Wd2' includes its commit. Such interference is not
-prohibited by RC, and the SC store makes \emph{all} effects of `Wd2'
-immediately visible. Our proof system insists that the
-programmer consider all effects of `Wd2' and conclude the value of
-\C{C} as \C{k-a2}, thus leading (rightfully) to the failure of proof.
-On the other hand, \iso{Snapshot Isolation} proscribes any
-interference from a concurrent transaction performing conflicting
-writes. The RC executions shown above are therefore invalid under SI.
-Accordingly, programmer is allowed to ignore any interference from
-`Wd2', thus facilitating the proof of postcondition.
-
-As evident from the above example, a proof system for weak isolation
-has to be flexible enough to admit different kinds of reasoning for
-different isolation and consistency levels. One way this can be
-achieved is by having a separate proof rule for each combination of
-transaction isolation level and store consistency level, such that the
-rule reflects the operational semantics of the isolation level on the
-store. Given the large and ever-increasing diversity among isolation
-and consistency semantics, this approach is unlikely to scale. Our
-proof system avoids this problem by being parametric over the
-semantics of isolation and consistency levels. This approach is based
-on a couple of observations. First, as noted by some
-authors~\cite{pldi15,gotsmanconcur15}, semantics of various isolation
-and consistency levels can be captured as well-formedness conditions
-over execution traces. If an execution trace satisfies specified
-well-formedness conditions, then the program is considered to have
-experienced the required levels of consistency and isolation in that
-execution. Second, well-formedness conditions are often
-\emph{prefix-complete}, meaning that every prefix of a well-formed
-trace is also well-formed, or conversely, an ill-formed trace cannot
-be extended to a well-formed one. Prefix-completeness allows us to
-reason about programs operationally, in terms of an imaginary machine
-whose primary artifact is an execution trace. The machine starts from
-a well-formed trace, executes a read/write/commit operation, and
-extends the trace while preserving well-formedness. If well-formedness
-cannot be preserved by taking a step, the machine gets ``stuck''. In
-executions that run to completion, the machine models a store whose
-consistency and isolation guarantees are captured by the trace
-well-formedness conditions. Thus, to show that program's invariants
-are preserved under weak isolation, it suffices to show that they are
-preserved in all executions of the imaginary machine that do not get
-stuck. This is indeed how our proof system facilitates program
-verification. The ``invalid interference'' described informally before
-corresponds to a case of machine getting stuck due to the violation of
-trace well-formedness conditions. The proof system inherits the
-parametricity of the machine over trace well-formedness conditions,
-leading to a generic set of parameterized proof rules that can be
-instantiated for various store consistency and transaction isolation
-levels.
-
-We now demonstrate how rely-guarantee reasoning, parametricity ,and
-trace well-formedness conditions facilitate the proof of the example
-in Fig.~\ref{fig:motiv-eg-1}. The trace-wellformedness condition
-corresponding to \iso{Snapshot Isolation} is shown below\footnote{This
-is a simplified version of the actual specification of SI
-(\S~\ref{sec:opsem}), which is a little more involved.}:
+For the rely guarantee approach to be useful it has to constrain the
+interference in accordance with the chosen level of isolation. The
+first key observation we make is that we can constrain interference by
+constraining the nature of \emph{happens-before} ($\hbZ$) relation
+between transactions or their respective constituents. We make use of
+the $\hbZ$ relation to axiomatize the interference characteristic of
+isolation levels as their specifications ($\psi$). For example,
+\iso{Read Committed} specification allows an operation $\eta_1$ of
+$T_1$ to happen before an operation $\eta_2$ of $T_2$ only if every
+operation in $T_1$ (including its commit) happens before $\eta_2$ (we
+write $T_1 \hboar \eta_2$ to denote this). RC spec does not require
+$T_1$ to happen before all operations of $T_2$, thus allowing it to
+interfere in $T_2$, as possible under RC isolation. \iso{Snapshot
+Isolation} spec requires two transactions that write to the same
+variable to be related by happens-before.  This effectively proscribes
+interference due to $T_1$ in $T_2$ (or, $T_2$ in $T_1$).
+Formally\footnote{Note that specs are simplified for illustration. The
+actual specs (\S~\ref{sec:opsem}) are more nuanced.  }:
 \begin{smathpar}
 \begin{array}{lcl}
-\underE{\C{SI}(T_j)} & \Rightarrow & \forall T_i. T_i \neq T_j \conj \\
-  & & (\exists \C{X}.~\underE{T_i \wrstoar \C{X}} \conj 
-                  \underE{T_j \wrstoar \C{X}}) \Rightarrow \\
-  &  & \underE{T_i \hboar T_j} \disj \underE{T_j \hboar \C{COMMIT}(T_i)} \\
+\psi_{RC} & \defeq & \forall \eta_1,\eta_2,T_1,T_2.\; \txn(\eta_1) = T_1 
+  \conj \txn(\eta_2) = T_2 \\
+  & & \hspace*{0.6in}\conj T_1 \neq T_2 \conj \eta_1 \hboar
+  \eta_2 \Rightarrow T_1 \hboar \eta_2 \\
+\psi_{SI} & \defeq & \forall T_1,T_2.\; T_1 \neq T_2 \conj
+  (\exists \C{X}.~{T_1 \wrstoar \C{X}} \conj 
+                      {T_2 \wrstoar \C{X}})\\
+  &  & \hspace*{0.6in}\Rightarrow{T_1 \hboar T_2} \disj {T_2 \hboar T_1} \\
 \end{array}
 \end{smathpar}
-A transaction $T_j$ that writes to a variable $\C{X}$ has experienced
-snapshot isolation in an execution $\E$ only if every other
-transaction $T_i$ that also wrote to \C{X} in $\E$ either happened
-before (i.e., committed before) $T_j$, or executed concurrently but
-commited after $T_j$. Here, $\E$ denotes an execution trace, and
-$\underE{\phi}$ asserts that proposition $\phi$ about $\E$ is valid.
-Since we require both `Wd1' and `Wd2' to be executed under SI, the
-trace invariant ($\I$) for the example is $\C{SI(Wd1)} \wedge
-\C{SI(Wd2)}$. That is, if the machine extends an execution $\E$ that
-satisfies $\I$ (i.e.,$\underE{\I}$) to $\E'$, then $\E'$ also satisfies
-$\I$ (i.e., $\E' \Vdash \I$). Since we reason in terms of traces, pre
-and post conditions need to be written with respect to a trace.
+Fig.~\ref{fig:motiv-eg-1-hb} is a visualization of $\hbZ$ arrows
+from `Wd2' to `Wd1' allowed by $\psi_{RC}$and $\psi_{SI}$. All
+arrows re legal under $\psi_{RC}$ because in no case does an operation
+$\eta_1$ from `Wd1' happen before $\eta_2$ of `Wd2' without the commit
+of `Wd1' also happening before $\eta_2$. $\psi_{SI}$ however disallows
+all dashed arrows, becase dashed arrows establish $\hbZ$ between `Wd2'
+and only a subset of operations in `Wd1'. Ligher (blue) arrows with
+hollow heads denote $\hbZ$ relationships that do not effect the value
+of \C{C} in a way that causes the program to violate its
+postcondition.  Darker (black) arrows with solid heads denote $\hbZ$
+relationships that lead to the violation of postcondition. The task of
+the reasoning framework is to determine if all $\hbZ$ relationships
+allowed by an isolation level lead to the satisfaction of the
+postcondition.
 
-\begin{figure}
-\centering
-\begin{txnimpcode}
- $\begin{decoration}
- P_1:\{ {\neg\committed(\C{Wd2})} \Rightarrow \C{C = k} \conj\\
-        \hspace*{0.3in}{\committed(\C{Wd2})} \Rightarrow \C{Wd2}
-        \hboar \C{Wd1} \wedge \C{Wd2} \wrstoar \C{C} \wedge \C{C = k-a2} \}
- \end{decoration}$
-  txn$\langle$'Wd1'$\rangle${
-   $\begin{decoration}
-    \phi_1 : \{{\neg\committed(\C{Wd2})} \Rightarrow \C{C = k} \conj
-      {\committed(\C{Wd2})} \Rightarrow \C{Wd2} \wrstoar \C{C} \conj\\
-       \hspace*{0.3in}{\committed(\C{Wd2})} \wedge
-        {\C{Wd2} \hboar \C{Wd1}} 
-       \Rightarrow \C{C = k-a2} \}
-    \end{decoration}$ 
-    v1 = C
-   $\begin{decoration}
-    \phi_2 : \{{\neg\committed(\C{Wd2})} \Rightarrow \C{C = k} \wedge \C{v1 = k} \conj\\
-       \hspace*{0.3in}{\committed(\C{Wd2})} \Rightarrow \C{Wd2} \wrstoar \C{C} \conj \\
-       \hspace*{0.3in}{\committed(\C{Wd2})} \wedge
-        {\C{Wd2} \hboar \C{Wd1}} 
-       \Rightarrow \C{C = k-a2} \wedge \C{v1 = k-a2}\}
-    \end{decoration}$ 
-    if (v1 $\ge$ a1) {
-      v2 = C;
-     $\begin{decoration}
-      \phi_3 : \{{\neg\committed(\C{Wd2})} \Rightarrow \C{C = k} \wedge \C{v2 = k} \conj\\
-       \hspace*{0.3in}{\committed(\C{Wd2})} \Rightarrow \C{Wd2} \wrstoar \C{C} \conj \\
-         \hspace*{0.3in}{\committed(\C{Wd2})} \wedge
-          {\C{Wd2} \hboar \C{Wd1}} \Rightarrow \C{C = k-a2} \wedge \\
-          \hspace*{1.9in}\C{v2 = k-a2}\}
-      \end{decoration}$ 
-      v3 = v2 - a1;
-     $\begin{decoration}
-      \phi_4 : \{{\neg\committed(\C{Wd2})} \Rightarrow \C{C = k} \wedge \C{v3 = k-a1} \conj\\
-         \hspace*{0.3in}{\committed(\C{Wd2})} \Rightarrow \C{Wd2} \wrstoar \C{C} \conj \\
-         \hspace*{0.3in}{\committed(\C{Wd2})} \wedge
-          {\C{Wd2} \hboar \C{Wd1}} 
-         \Rightarrow \C{C = k-a2} \wedge \\
-         \hspace*{1.9in}\C{v3 = k-a2-a1}\}
-      \end{decoration}$ 
-      C := v3
-     $\begin{decoration}
-      \phi_5 : \{{\neg\committed(\C{Wd2})} \Rightarrow \C{C = k-a1}) 
-                \conj\\
-         \hspace*{0.3in}{\committed(\C{Wd2})} 
-                \Rightarrow \C{C = k-a1-a2}\}
-      \end{decoration}$ 
-    }
-  }
- $\begin{decoration}
-  Q_1 : \{{\neg\committed(\C{Wd2})} \Rightarrow \C{C = k-a1}
-            \conj \committed(\C{Wd1}) \conj\\
-      \hspace*{0.12in}{\committed(\C{Wd2})} 
-          \Rightarrow \C{C = k-a1-a2} \}
-  \end{decoration}$ 
-\end{txnimpcode}
+% There is a $\hb$ edge (first dashed) that is allowed by RC but
+% prohibited by SI, which nonetheless does not lead to invariant
+% violation.
 
-\caption{`Wd1' transaction decorated with assertions}
-\label{fig:wd1-decorated}
-\end{figure}
-
-The fully decorated implementation of transaction `Wd1' is shown in
-Fig.~\ref{fig:wd1-decorated}. Decorated `Wd2' is similar. Assignment
-statements are broken down and temporary local variables (\C{v1},
-\C{v2} and \C{v3}) are introduced so as to separate shared variable
-reads and writes.  Since we reason in terms of executions, all
-assertions implicitly refer to the current execution (just as hoare
-triples implicitly refer to the current state). Also implicit is the
-invariant \C{k $\ge$ a1+a2}. Proposition $\committed(T)$ indicates
-that the transaction $T$ is already committed. Precondition ($P_1$) of
-`Wd1' accounts for the possibility of `Wd2' committing before `Wd1',
-writing \C{k-a2} to \C{C}.  Precondition of `Wd2' is similar.  Since
-neither `Wd1' nor `Wd2' are committed at the beginning, the
-precondition in Fig.~\ref{fig:motiv-eg-1} is extended to include
-$\neg\committed(\C{Wd1}) \wedge \neg\committed(\C{Wd2})$, from which
-$P_1$ follows. Once the execution is inside `Wd1', commit of `Wd2'
-($\committed(\C{Wd2})$) may mean that either `Wd2' happened before
-`Wd1', or that it committed concurrently with `Wd1'. Since SI
-proscribes the latter possibility, we only consider the case when
-$\C{Wd2} \hboar \C{Wd1}$. A proof for $\C{Wd2} \hboar \C{Wd1}$ is
-obtained subsequently, allowing us to get rid of the special case.
-
-The first obligation is to show that the assertions that decorate
-`Wd1' are indeed valid. This entails proving that, for each statement
-of `Wd1', if evaluating the statement and extending an execution that
-satisfies its precondition results in a well-formed execution, then
-the extended execution must satisfy the postcondition. The
-well-formedness assumption lets us ignore the executions that violate
-the trace invariant lead the machine to a stuck state. This is useful
-in case of the assignment statement that writes \C{v3} to \C{C}. Since
-`Wd2' also writes to \C{C}, the trace invariant ($\I$) requires
-$\C{Wd2} \hboar \C{Wd1}$ if $\committed(\C{Wd2})$ (the alternative
-$\C{Wd1} \hboar \C{COMMIT}(\C{Wd2})$ allowed by \C{SI(Wd1)} is
-impossible in this case because \C{Wd2} has already committed, while
-\C{Wd1} is still in progress). This lets us treat the special case
-$\committed(\C{Wd2})\conj \C{Wd2} \hboar \C{Wd1}$ as being equivalent
-to the general case $\committed(\C{Wd2})$, which is required to prove
-the postcondition of `Wd1'. Observe that \C{RC(Wd1)} does not justify
-this generalization, hence replacing \C{SI(Wd1)} by \C{RC(Wd1)} is not
-sufficient to prove the postcondition of `Wd1'.
-
-The second obligation is to prove that assertions are valid despite
-the interference from the the concurrent thread executing `Wd2'. The
-interference is given by the following rely relation ($R_1$):
+Specification of an isolation level encodes its interference
+characteristic as constraints over $\hbZ$ relation. However, for this
+to be useful in reasoning about programs, rely guarantee framework
+should be able to use the specification to determine if an
+interference is valid or invalid, allowing the programmer to only
+focus on the former. Our second key observation is that this is
+possible if the reasoning framework adequately tracks $\hbZ$ at each
+program point, while preserving $\hbZ$ constraints as invariants
+between program points. An interference that leads to the violation of
+$\hbZ$ constraints (i.e., an invalid interference) is thus
+automatically prohibited. For instance, consider the program point
+after the write to \C{C} in `Wd1'. The expected invariant ($\phi$) at
+that program point is shown below ($\committed$ stands for
+``committed''):
 \begin{smathpar}
-\begin{array}{lcl}
-  R_1 & = & \{ (\E,\E') \;|\; \neg\underE{\committed(\C{Wd2})} \conj 
-        \underE{\I} \conj \E'\Vdash \I \conj\\
-%       \conj (\E'-\E) \subseteq \C{Wd2} \conj \\
-%   & & \hspace*{0.5in} \E' \Vdash \committed(\C{Wd2}) ~\Rightarrow~ \E'
-%       \Vdash \C{Wd2} \wrstoar \C{C} \conj \\ 
-%   & & \hspace*{0.5in} \underE{\committed(\C{Wd1})} \conj \E' \Vdash
-%   \committed(\C{Wd2}) \Rightarrow \C{C=k-a1-a2} \\
-    & & \hspace*{0.5in} \neg\underE{\committed(\C{Wd1})} \conj
-        \C{COMMIT(Wd2)} \in (\E'-\E) \\
-    & & \hspace*{0.8in}\Rightarrow \C{C=k-a2} \conj 
-        \E' \Vdash \C{Wd2} \wrstoar \C{C} \conj \\
-    & & \hspace*{0.5in} \underE{\committed(\C{Wd1})} \conj
-        \C{COMMIT(Wd2)} \in (\E'-\E) \\
-    & & \hspace*{0.6in}\Rightarrow \C{C=k-a1-a2} \conj
-        \E' \Vdash \C{Wd2} \wrstoar \C{C} \}\\
+\begin{array}{l}
+  \neg\committed(\C{Wd1}) \conj (\neg\committed(\C{Wd2}) \Rightarrow
+  \C{C = k-a1}) 
+                \\
+      \C{Wd1} \wrstoar \C{C} \conj (\committed(\C{Wd2})
+                \Rightarrow \C{C = k-a1-a2})\}
 \end{array}
 \end{smathpar}
-The rely relation says the following about the concurrent thread: (1).
-It may interfere only if `Wd2' is not already committed, (2). Any
-interference takes well-formed executions to well-formed executions,
-(3). If the interference commits `Wd2', then `Wd2' should have already
-written to \C{C}, and (4). The value written is either $\C{k-a1-a2}$
-or $\C{k-a1}$ depending on whether or not `Wd1' has already committed.
-Interference from `Wd2' is allowed between any two operations of
-`Wd1', and assertions are required to be valid despite the
-interference. For instance, if the execution ($\E$) immediately after
-evaluating $\C{v1 = C}$ satisfies $\phi_2$ (i.e., $\underE{\phi_2}$),
-and an interference from `Wd2' extends $\E$ to $\E'$ (i.e.,
-$R_1(\E,\E')$), then the execution $\E'$ just before evaluating the
-condition must satisfy $\phi_2$ (i.e., $\E' \Vdash \phi_2$). Indeed,
-all $\phi_i$'s do satisfy this \emph{stability} condition, thus
-lending the validity to the proof.
+$\phi$ asserts that `Wd1' is not yet committed, and that it wrote to
+\C{C}, and the value of \C{C} is either \C{k-a1-a2} or \C{k-a1}
+depending on whether or not `Wd2' is committed. If $\phi$ stays
+invariant until `Wd1' commits, then the postcondition (\C{C =
+k-a1-a2}) can be established easily. However, an interference from
+`Wd2' at this stage (captured by the last dashed arrow in
+Fig.~\ref{fig:motiv-eg-1-hb}) may violate the invariant by writing
+$\C{k-a1}$ to \C{C} and committing `Wd2', thus leading to
+$\committed(\C{Wd2}) \Rightarrow \C{C=k-a2}$. Fortunately,
+\iso{Snapshot Isolation} prevents this interference, and this can be
+proved by showing that an interference from `Wd2' starting from a
+execution state that satisfies $\psi_{SI} \wedge \phi$ leads to an
+execution state where neither $\C{Wd2} \hboar \C{Wd1}$ nor $\C{Wd1}
+\hboar \C{Wd2}$ holds (former does not hold because at least one write
+from `Wd1' has happened before at least one operation of `Wd2', and
+the latter doesn't hold because `Wd1' hasn't yet committed and `Wd2'
+has already begun). Since `Wd2' also writes to \C{C}, this violates
+the $\psi_{SI}$ constraint which we assume to be an invariant. A proof
+for postcondition now follows from the contradiction. It is
+informative to note that if the invariant is $\psi_{RC}$ instead of
+$\psi_{SI}$, we cannot derive a contradiction and we cannot rule out
+the interference, which causes (rightfully) the proof to fail.
+
+We have thus far assumed an SC store that, in the absence of
+transactions with special isolation requirements, makes effects of an
+operation to be immediately visible to subsequent operations. Thus,
+the natural behavior of SC store to totally order all operations w.r.t
+$\hbZ$ relation could be in conflict with the constraints imposed by
+weak isolation. To meet the requirements of weaker isolation levels,
+say RC, the store has to adapt itself to \emph{hide} the effects of
+concurrent transactions until they are committed. Once committed
+though, the effects need to be immediately and perennially visible to
+subsequent operations. This semantics of an SC store can be built into
+the reasoning framework leading to a proof system tailor-made for such
+stores. However, for the reasoning framework to be useful across the
+stores, it has to be parametric over store consistency semantics, and
+should be able to reconcile conflicts betweenconsistency and isolation
+constraints. Later sections demonstrate how our reasoning framework
+makes this possible.

### 2016-06-16 opsem.tex
Finally motivated!

\section{\txnimp}

As evident from the above example, a proof system for weak isolation
has to be flexible enough to admit different kinds of reasoning for
different isolation and consistency levels. One way this can be
achieved is by having a separate proof rule for each combination of
transaction isolation level and store consistency level, such that the
rule reflects the operational semantics of the isolation level on the
store. Given the large and ever-increasing diversity among isolation
and consistency semantics, this approach is unlikely to scale. Our
proof system avoids this problem by being parametric over the
semantics of isolation and consistency levels. This approach is based
on a couple of observations. First, as noted by some
authors~\cite{pldi15,gotsmanconcur15}, semantics of various isolation
and consistency levels can be captured as well-formedness conditions
over execution traces. If an execution trace satisfies specified
well-formedness conditions, then the program is considered to have
experienced the required levels of consistency and isolation in that
execution. Second, well-formedness conditions are often
\emph{prefix-complete}, meaning that every prefix of a well-formed
trace is also well-formed, or conversely, an ill-formed trace cannot
be extended to a well-formed one. Prefix-completeness allows us to
reason about programs operationally, in terms of an imaginary machine
whose primary artifact is an execution trace. The machine starts from
a well-formed trace, executes a read/write/commit operation, and
extends the trace while preserving well-formedness. If well-formedness
cannot be preserved by taking a step, the machine gets ``stuck''. In
executions that run to completion, the machine models a store whose
consistency and isolation guarantees are captured by the trace
well-formedness conditions. Thus, to show that program's invariants
are preserved under weak isolation, it suffices to show that they are
preserved in all executions of the imaginary machine that do not get
stuck. This is indeed how our proof system facilitates program
verification. The ``invalid interference'' described informally before
corresponds to a case of machine getting stuck due to the violation of
trace well-formedness conditions. The proof system inherits the
parametricity of the machine over trace well-formedness conditions,
leading to a generic set of parameterized proof rules that can be
instantiated for various store consistency and transaction isolation
levels.

### 2016-06-16 motivation.tex
Edits.

@@
 
-\label{fig:motiv-eg-1-hb}
 \caption{$\hbZ$ arrows allowed by SI (solid) and RC (solid and dashed)
@@
 stand for read, write and commit, respectively.}
+\label{fig:motiv-eg-1-hb}
 \end{figure}
@@
 transactions with special isolation requirements, makes effects of an
-operation to be immediately visible to subsequent operations. Thus,
-the natural behavior of SC store to totally order all operations w.r.t
+operation to be immediately visible to subsequent operations. The
+natural behavior of SC store to totally order all operations w.r.t
 $\hbZ$ relation could be in conflict with the constraints imposed by

### 2016-06-18 opsem.tex
Opsem with ECtx added.
section 3 part 1.

@@
 \section{\txnimp}
+\label{sec:opsem}
 
-As evident from the above example, a proof system for weak isolation
-has to be flexible enough to admit different kinds of reasoning for
-different isolation and consistency levels. One way this can be
-achieved is by having a separate proof rule for each combination of
-transaction isolation level and store consistency level, such that the
-rule reflects the operational semantics of the isolation level on the
-store. Given the large and ever-increasing diversity among isolation
-and consistency semantics, this approach is unlikely to scale. Our
-proof system avoids this problem by being parametric over the
-semantics of isolation and consistency levels. This approach is based
-on a couple of observations. First, as noted by some
-authors~\cite{pldi15,gotsmanconcur15}, semantics of various isolation
-and consistency levels can be captured as well-formedness conditions
-over execution traces. If an execution trace satisfies specified
-well-formedness conditions, then the program is considered to have
-experienced the required levels of consistency and isolation in that
-execution. Second, well-formedness conditions are often
-\emph{prefix-complete}, meaning that every prefix of a well-formed
-trace is also well-formed, or conversely, an ill-formed trace cannot
-be extended to a well-formed one. Prefix-completeness allows us to
-reason about programs operationally, in terms of an imaginary machine
-whose primary artifact is an execution trace. The machine starts from
-a well-formed trace, executes a read/write/commit operation, and
-extends the trace while preserving well-formedness. If well-formedness
-cannot be preserved by taking a step, the machine gets ``stuck''. In
-executions that run to completion, the machine models a store whose
-consistency and isolation guarantees are captured by the trace
-well-formedness conditions. Thus, to show that program's invariants
-are preserved under weak isolation, it suffices to show that they are
-preserved in all executions of the imaginary machine that do not get
-stuck. This is indeed how our proof system facilitates program
-verification. The ``invalid interference'' described informally before
-corresponds to a case of machine getting stuck due to the violation of
-trace well-formedness conditions. The proof system inherits the
-parametricity of the machine over trace well-formedness conditions,
-leading to a generic set of parameterized proof rules that can be
-instantiated for various store consistency and transaction isolation
-levels.
+\input{txnimp}
+
+\subsection{Syntax and Semantics}
+\label{sec:syntax}
+
+Fig.~\ref{fig:txnimp} shows the syntax and small-step semantics of
+\txnimp. Natural numbers, (shared) variables and arithmetic
+expressions constitute the syntactic class of expressions ($e$).
+Commands ($c$) include $\cskip$, assignment statements, transaction
+(\C{txn}) lexical blocks, and their sequential and parallel
+composition. We let $T_i$ for $i \in \mathbb{N}$ range over
+transaction identifiers. When it is evident we are referring to a
+transaction, we use the number $i$ instead of $T_i$ for identification
+(\eg, in $\C{txn}\langle i \rangle$). For notational convenience we let
+$t$ range over both expressions and commands.
+
+Small-step semantics of \txnimp are defined in terms of an abstract
+machine that generates an execution trace, also called an execution
+($\E$). First artifact of the trace is a set ($\A$) of \emph{effects},
+where an effect ($\eta$) documents a read (\C{RD(X)}), or a write
+(\C{WR(X)}), or a transaction commit (\C{COMMIT}) operation performed
+during the execution. Any value associated with the operation (\eg,
+value read or value written) is also documented, and can be accessed
+via $\rval$. Every effect has a unique identifier accessible via
+$\id$, and a transction idenfier accessible via $\txn$.  The latter
+identifies the transaction that generated the effect. In every step of
+the evaluation, abstract machine reduces a \txnimp term by executing a
+read, write or commit operation, generating an effect, and extending
+the trace. Since effects include transaction identifiers, semantics
+distinguishes between terms ($t$) of different transactions. For
+example, $\txnbox{t}_i$ denotes a term $t$ inside a transaction $T_i$.
+Evaluation contexts are also appropriately marked. For example,
+$\ectx_i$ denotes evaluation context for a term inside $T_i$. The
+other component of the execution trace is the visibility relation
+($\visZ$) between the effects. The intent and mechanics of $\visZ$
+become clear as we describe the evaluation rules in the following.
+
+Fundamental to our operational semantics is the notation of a trace
+invariant ($\I$), which is a function from traces ($\E$) to
+propositions ($\texttt{Prop}$) that declare well-formedness
+constraints over traces. Abstract machine takes a step only if the
+resulting trace satisfies the well-formedness constraints imposed by
+$\I$. This behaviour is captured by the auxiliary reduction rule
+\rulelabel{E-Aux} that factors out the trace extension aspect of the
+evaluation by abstracting away operation-specific behaviour as a
+function ($\mathcal{F}$) that generates appropriate effect.
+\rulelabel{E-Aux} uses $\mathcal{F}$ to generate a new effect and
+extend the trace ($\E = (\A,\visZ)$) \emph{only if} the
+well-formedness constraints imposed by $\I$ on $\E$ (i.e., $\I(\E')$)
+are satisfied. Otherwise, it gets stuck. In an execution that runs to
+completion, every small-step preserves well-formedness of trace, thus
+ensuring the invariance of $\I$. Note that the operational semantics
+makes no assumptions about $\I$ other than its type. As such it is
+parametric over $\I$, and can be instantiated with any
+trace-parametric proposition that expresses constraints over the given
+trace. While the machine takes a step only if the constraints are
+satisfied, it neither defines nor explicitly assumes an oracle to
+check the satisfaction; there is no motivation for doing so in the
+semantics.
+
+As explained in \S\ref{sec:motivation}, semantics of various isolation
+levels can be captured as constraints over the happens-before ($\hbZ$)
+relation. $\hbZ$ is however a derived relation in our model, composed
+of more fundamental \emph{session order} ($\soZ$) and
+\emph{visibility} ($\visZ$) relations. Session order captures the
+sequential order of operations within a transaction. In particular, it
+relates two effects, $\eta_1$ and $\eta_2$, such that $\txn(\eta_1) =
+\txn(\eta_2)$ and $\id(\eta_1) < \id(\eta_2)$. Small-step semantics
+assigns monotonically increasing identifiers to effects, as captured
+by the $\id(\eta) > \maxId(\A)$ condition of \rulelabel{E-Aux}.
+Evaluation contexts ($\ectx_i$) for transaction-bound terms are
+defined so as to enforce a deterministic sequential order of execution
+within a transaction, leading to a deterministic total order among
+effect ids - the session order. Visibility ($\visZ$) on the other
+hand relates effects across concurrent transactions, thus reflects the
+non-determinism inherent in concurrency. Intuitively, $\visZ$ relates
+$\eta_1$ to $\eta_2$ if and only if $\eta_1$ was \emph{visible} to the
+operation that generated $\eta_2$ during its execution, thus 
+effecting its return value ($\rval(\eta_2)$). For example, a read
+operation over \C{X} may pick the value ($\rval$) of the write effect
+with highest id among the visible effects (this is made possible by
+appropriately defining $\interp{\cdot}$ in \rulelabel{E-Var},
+as we show later). Thus, the value of a read depends on what write
+effects it can witness. An operation can only witness the effects of
+already concluded operations, which varies between executions due to
+the non-deterministic order of evaluating the parallel composition
+(cf. evaluation contexts ($\ectx$) for parallel composition). A more
+notable source of non-determinism, however, is the \rulelabel{E-Aux}
+rule, which allows the machine to expose an arbitrary subset ($S$) of
+existing effects ($\A$) to the incoming operation. In other words, the
+machine is not obligated to reveal the effects of all previous
+operations to an incoming operation. This relaxation allows the
+abstract machine to model the semantics of weakly consistent data
+stores. For instance, operations issued to an eventually consistent
+(EC) replicated store could be dispatched to different replicas whose
+states may not be in any well-defined relationship. By allowing
+operations to witness arbitrary subsets of the global state, the
+abstract machine effectively models an EC store.
+
+However, a machine that lets operations witness arbitrary subset of
+the global state offers no isolation whatsoever. For example, it lets
+a read operation witness writes of an uncommitted transaction,
+violating RC isolation. Fortunately, our ability to express isolation
+semantics as constraints over happens-before order through $\visZ$ and
+$\soZ$ relations, and the property of the abstract machine to be
+parametric over the trace invariant ($\I$) lets us solve this problem.
+In particular, we continue defining isolation semantics as constraints
+over $\visZ$ and $\soZ$, but confine their domain of interpretation to
+the given trace so that they now become trace well-formedness
+constraints. Well-formedness constraints can be combined into a trace
+invariant ($\I$). Since the abstract machine is guaranteed to preserve
+$\I$ at every step of the execution, operations are guaranteed to
+experience the level of isolation specified by $\I$.  Thus, in
+executions that run to completion abstract machine models a store that
+provides the required levels of isolation. Notably, the machine
+achieves this without confining to any particular implementation of
+the isolation, by relying on a declarative specification of isolation
+as trace well-formedness constraints.  \S\ref{sec:ansi-isolation}
+shows specifications of various ANSI SQL isolation levels stated as
+trace well-formedness constraints.
+
+As described previously, \rulelabel{E-Aux} abstracts away the
+operation-specific behaviour of a machine step as a function ($\F$)
+that accepts a set ($S$) of effects chose by the machine to make
+visible to the operation, interprets the operation w.r.t $S$, and
+returns an appropriate effect that encodes its return value. Rules
+\rulelabel{E-Var}, \rulelabel{E-Asgn} and \rulelabel{E-Commit} define
+such functions for read, write and commit operations, respectively.
+The effect returned by the function in each case includes its
+transaction id ($T_i$) along with an arbitrarily chosen effect id
+($j$) that is later verified to be unique in \rulelabel{E-Aux}. The
+$\rval$ for a write is the value being written, and for commit it is
+$\bot$. In case of a read, the value read depends on the how the read
+operation chooses to interpret the given set ($S$) of visible effects.
+The interpretation may depend on the application semantics. For
+example, a monotonically increasing counter application may choose to
+let a write with largest value determine the value of a
+read. A more straightforward interpretation has already been described
+before. To accommodate multiple interpretations, operational semantics
+is made parametric over an interpretation function ($\interp{\cdot}$)
+that accepts a set of effects and a variable name, and returns the
+value associated with the variable. A straightforward interpretation
+function that chooses the last write (i.e., write with largest id) is
+shown below:
+\begin{smathpar}
+\begin{array}{lcl}
+  \isMax(S,\eta) & \Leftrightarrow &  \forall (\eta'\in S).  
+  \kind(\eta') = \kind(\eta) \\
+  & & \hspace*{0.4in}\Rightarrow \eta' = \eta \disj \id(\eta') < \id(\eta)\\
+
+\interp{S}(X) & = & \C{if}\;(\exists (\eta \in S). \kind(\eta) = \C{WR}(X) 
+  \wedge \isMax(S,\eta)) \\
+  & & \C{then}\;\rval(\eta)\;\C{else}\;0\\
+\end{array}
+\end{smathpar}
+Rules \rulelabel{E-Top-Ctx} and \rulelabel{E-Txn-Ctx} rules define
+congruence rules for top-level terms and transaction-bound terms,
+respectively. The rules and evaluation contexts ($\ectx$ and
+$\ectx_i$) are defined such that only a certain kinds of terms are
+allowed at the top-level and inside a transaction. In particular, a
+\txnimp program at the top-level can either be a transaction, or a
+parallel composition of transactions. A command inside a \C{txn} block
+can either be an assignment, or a sequential composition of
+assignments. Nested transactions and transactions that fork multiple
+threads are forbidden to focus on the problems relevant to this paper.
+
+\subsection{Isolation Specifications}
+\label{sec:ansi-isolation}
+
+\subsection{Store Consistency}
+\label{sec:store-consistency}

### 2016-06-18 opsem.tex
Typos.

@@
 Small-step semantics of \txnimp are defined in terms of an abstract
-machine that generates an execution trace, also called an execution
-($\E$). First artifact of the trace is a set ($\A$) of \emph{effects},
+machine that generates an execution trace ($\E$). First artifact of the trace is a set ($\A$) of \emph{effects},
 where an effect ($\eta$) documents a read (\C{RD(X)}), or a write
@@
 via $\rval$. Every effect has a unique identifier accessible via
-$\id$, and a transction idenfier accessible via $\txn$.  The latter
+$\id$, and a transction identifier accessible via $\txn$.  The latter
 identifies the transaction that generated the effect. In every step of
@@
 
-Fundamental to our operational semantics is the notation of a trace
+Fundamental to our operational semantics is the notion of a trace
 invariant ($\I$), which is a function from traces ($\E$) to
@@
 satisfied, it neither defines nor explicitly assumes an oracle to
-check the satisfaction; there is no motivation for doing so in the
+check the satisfaction; there is no need for doing so in the
 semantics.

### 2016-06-18 abstract.tex
edits to Secs. 1 and 2

@@
   other concurrenty executing transactions once it begins executing.
-  However, enforcing serializabity comes at the cost in performance
+  However, enforcing serializability comes at the cost in performance
   because it necessarily restricts opportunities to exploit
@@
   As a result, database systems (in practice) support, and often
-  encourage, developers to use weaker alternatives.  Generally, these
-  alternatives break the strong isolation guarantees provided by
+  encourage, developers to use weaker alternatives.  These
+  alternatives break the strong isolation guarantees offered by
   serializable transactions to permit greater concurrency.
@@
   high-level correctness properties in such environments is
-  challenging.
+  problematic.
 
-  To address this issue, this paper presents a program logic that
-  enables compositional reasoning about the behavior of concurrently
-  executing weakly isolated transactions.  Notably, our development is
+  To address this issue, we present a program logic that enables
+  compositional reasoning about the behavior of concurrently executing
+  weakly-isolated transactions.  Notably, our development is
   parametric over a transaction's specific isolation semantics, and
@@
   applications demonstrate the utility of our approach, and provide
-  strong evidence that weakly isolated transactions can be placed on
+  strong evidence that weakly-isolated transactions can be placed on
   the same formal footing as their strongly-isolated serializable

### 2016-06-18 introduction.tex
edits to Secs. 1 and 2

@@
 constraints on transaction execution, ensuring that any permissible
-concurrent schedule is equivalent to a serial one in which in which no
-interleaving of actions from different transactions are allowed.
+concurrent schedule yields results equivalent to a serial one in which
+there is no interleaving of actions from different transactions.
 
@@
 control methods that must deal with timestamp and version
-management~\cite{BG81}.  Regardless of the particular method employed,
-enforcing serializability in a replicated data store additionally
-requires global coordination among geo-distributed replicas, which
-makes the store \emph{unavailable} in the presence of inevitable
-network partitions~\cite{cap,sernotavlbl,bailishat,bernsigmod13}.
+management~\cite{BG81}.  Regardless of the particular concurrency
+control method employed, enforcing serializability in a replicated
+data store additionally requires global coordination among
+geo-distributed replicas, which makes the store \emph{unavailable} in
+the presence of network
+partitions~\cite{cap,sernotavlbl,bailishat,bernsigmod13}.
 
@@
 driven by performance and availability considerations, has motivated
-the development of multiple weaker forms of transaction isolation,
+the development of different weaker forms of transaction isolation,
 beginning as early as 1976~\cite{gray1976}. The ANSI SQL 92 standard
-defines three such weak isolation levels which are now
-implemented in many relational and NoSQL databases. Not surprisingly,
-weakly isolated transactions have been found to significantly
-outperform serializable transactions on benchmark suites, both on
-single-node databases and multi-node replicated
-stores~\cite{dbtuningbook,bailishat,bailisvldb}, leading to their
-overwhelming adoption. A 2013 study~\cite{bailishotos} of 18 popular
-ACID and ``NewSQL'' databases found that only three of them offer
-serializability by default, and half, including Oracle 11g, do not
-offer it at all.  A 2015 study~\cite{bailisferal} of a large corpus of
-database applications finds no evidence that applications manifestly
-change the default isolation level offered by the database. Taken
-together, these studies make clear that weakly isolated transactions
-are quite prevalent in practice, and strongly isolated (serializable)
-transactions are often eschewed.
+defines three such weak isolation levels which are now implemented in
+many relational and NoSQL databases. Not surprisingly, weakly-isolated
+transactions have been found to significantly outperform serializable
+transactions on benchmark suites, both on single-node databases and
+multi-node replicated stores~\cite{dbtuningbook,bailishat,bailisvldb},
+leading to their overwhelming adoption. A 2013
+study~\cite{bailishotos} of 18 popular ACID and ``NewSQL'' databases
+found that only three of them offer serializability by default, and
+half, including Oracle 11g, do not offer it at all.  A 2015
+study~\cite{bailisferal} of a large corpus of database applications
+finds no evidence that applications manifestly change the default
+isolation level offered by the database. Taken together, these studies
+make clear that weakly-isolated transactions are quite prevalent in
+practice, and strongly-isolated (serializable) transactions are often
+eschewed.
 
@@
 isolation. Indeed, incidents of safety violations due to executing
-applications in a weakly isolated environment have been reported on
+applications in a weakly-isolated environment have been reported on
 web services in production~\cite{starbucksbug, scimedbug}, including
@@
 isolation in terms of implementation details such as the nature and
-duration of locks held in each case. The ANSI SQL 92 standard four
-levels of isolation (including serializability) in terms of various
-undesirable \emph{phenomena} (\eg \emph{dirty reads}) each is required
-to prevent. While this is an improvement, it requires programmers to
-be prescient about the possible ways various undesirable phenomena
-might manifest in their applications, and determine, in each case, if
-the phenomenon can be allowed without violating application
-invariants. This is understandably hard. Adya~\cite{adyaphd} presents
-the first formal definitions of some of well-known isolation levels in
-the context of a sequentially consistent (SC) database.  However,
-there has been little progress relating Adya's system model to a
-formal operational semantics or a proof system that can facilitate
-rigorous correctness arguments.  Consequently, reasoning about weak
-isolation remains an error prone endeavor, with major database vendors
-continuing to document their isolation levels primarily in terms of
-the undesirable phenomena~\cite{postgresiso, mysqliso, oracleiso} a
-particular isolation level may induce; it thus squarely becomes the
-programmer's responsibility to determine if their application can
-yield an execution which can trigger such phenomena.
+duration of locks held in each case. The ANSI SQL 92 standard defines
+four levels of isolation (including serializability) in terms of
+various undesirable \emph{phenomena} (\eg \emph{dirty reads}) each is
+required to prevent. While this is an improvement, it requires
+programmers to be prescient about the possible ways various
+undesirable phenomena might manifest in their applications, and
+determine, in each case, if the phenomenon can be allowed without
+violating application invariants. This is understandably
+hard. Adya~\cite{adyaphd} presents the first formal definitions of
+some of well-known isolation levels in the context of a sequentially
+consistent (SC) database.  However, there has been little progress
+relating Adya's system model to a formal operational semantics or a
+proof system that can facilitate rigorous correctness arguments.
+Consequently, reasoning about weak isolation remains an error prone
+endeavor, with major database vendors continuing to document their
+isolation levels primarily in terms of the undesirable
+phenomena~\cite{postgresiso, mysqliso, oracleiso} a particular
+isolation level may induce; it thus squarely becomes the programmer's
+responsibility to determine if their application can yield an
+execution which can trigger such phenomena.
 
@@
 presence of weak \emph{consistency}~\cite{burckhardt14, redblueosdi,
-  redblueatc, ecinec, gotsmanpopl16} provide insights on possible ways
-we might overcome some of the difficulties in understanding weak
-isolation, even though the latter is strictly more general than the
-former.  At a high level, reasoning about weak consistency involves
-reasoning about the concurrent execution of strongly atomic and
-isolated operations against different but consistent states of the
-database or data store, where each operation locally preserves
-application invariants. In contrast, reasoning about weak isolation
-involves reasoning about the concurrent execution of transactions,
-whose operations witness different states which, although consistent
+  redblueatc, ecinec, gotsmanpopl16} address broadly related concerns.
+At a high level, reasoning about weak consistency involves reasoning
+about the concurrent execution of strongly atomic and isolated
+operations against different but consistent states of the database or
+data store, where each operation locally preserves application
+invariants. In contrast, reasoning about weak isolation involves
+reasoning about the concurrent execution of transactions, whose
+operations witness different states which, although consistent
 individually, may not be obviously reconciled into a consistent global
@@
 
-In this paper, we propose a program logic for weakly isolated
+In this paper, we propose a program logic for weakly-isolated
 transactions that realizes this goal.  In particular, we develop a set
@@
 consistency~\cite{gotsmanpopl16, redblueatc, ecinec}, our system model
-does not assume a minimum consistency or isolation level, nor does it
-assume a predefined set of such levels.  Instead, our operational
-semantics admits declarative specifications of transaction isolation
-and store consistency levels (\eg~\cite{pldi15,gotsmanconcur15}), and
-generates executions that are guaranteed to satisfy these
-specifications. The result is a flexible system model that is general
-enough to incorporate the semantics of a range of isolation levels on
-a variety of stores (\eg the sequentially consistent store
-of~\cite{adyaphd}, or the causally consistent store
+does not assume a minimum or predefined set of consistency or
+isolation levels. Instead, our operational semantics admits declarative
+specifications of transaction isolation and store consistency levels
+(\eg~\cite{pldi15,gotsmanconcur15}), and generates executions that are
+guaranteed to satisfy these specifications. The result is a flexible
+system model that is general enough to incorporate the semantics of a
+range of isolation levels on a variety of stores (\eg the sequentially
+consistent store of~\cite{adyaphd}, or the causally consistent store
 of~\cite{gotsmanpopl16}, etc).  Our key technical contribution is thus
@@
   \item We develop a semantics for a core language equipped with
-    weakly isolated transactions, demonstrating that that a general
+    weakly isolated transactions, demonstrating that a general
     isolation semantics can be expressed as well-formedness constraints
@@
 mechanism that precisely relates high-level program invariants to
-low-level store consistency and weak isolation guarantees, thereby
-allowing weakly isolated transactions to enjoy the same rigorous
+low-level weak isolation and store consistency guarantees, thereby
+allowing weakly-isolated transactions to enjoy the same rigorous
 reasoning capabilities as their strongly isolated (serializable)

### 2016-06-18 motivation.tex
edits to Secs. 1 and 2

@@
 balance, after both transactions commit, is expected to reflect the
-effects of both withdraws. The pre and post conditions in
+effects of both withdraws. The pre- and post-conditions in
 Fig.~\ref{fig:motiv-eg-1} reflect our expectations. Indeed, invariants
 are guaranteed to hold if both withdraw transactions are serialized,
-making \iso{Serializable} isolation (SER) level a sufficent condition
-to preserve invaraints. But, is SER necessary?
+making \iso{Serializable} isolation ({\sc ser}) level a sufficent
+condition to preserve invaraints. But, is {\sc ser} necessary?
 
@@
 \emph{read committed} ({\sc rc}) isolation level, which is weaker than
-{\sc ser}\footnote{{\sc rc} is in fact the default isolation level in
-Postgres 9.5 and Oracle 11g databases.} An {\sc rc} transaction is
-isolated from the writes of uncommitted transactions, thus preventing
-the transaction from witnessing \emph{dirty reads}~\cite{berenson},
-reads that observe the effects of non-committed transactions. In the
-current example, {\sc rc} isolation admits the two executions shown in
-Fig.~\ref{fig:rc-ex} on an SC store, such as a relational database:
+     {\sc ser}\footnote{{\sc rc} is in fact the default isolation
+       level in Postgres 9.5 and Oracle 11g databases.} An {\sc rc}
+     transaction is isolated from the writes of uncommitted
+     transactions, thus preventing the transaction from witnessing
+     \emph{dirty reads}~\cite{berenson}, reads that observe the
+     effects of non-committed transactions. In the current example,
+     {\sc rc} isolation admits the two executions shown in
+     Fig.~\ref{fig:rc-ex} on a strongly consistent ({\sc sc}) store,
+     such as a relational database:
 
@@
 \subcaptionbox {
-  RC Execution 1
+  {\sc rc} Execution 1
   \label{fig:motiv-eg-1-a}
@@
 \subcaptionbox {
-  RC Execution 2
+  {\sc rc} Execution 2
   \label{fig:motiv-eg-1-b}
@@
 that executes it\footnote{For clarity, the effects of different
-transactions are shown in different colored backgrounds.} In the first
-execution, transaction {\bf Wd1} reads the current balance (\C{k}) and
-writes the new balance (\C{k-a1}), but before it commits transaction
-{\bf Wd2} executes and commits, writing the new balance (\C{k-a2}). RC
-isolation prevent {\bf Wd2} from witnessing the uncommitted writes of
-transaction {\bf Wd1}.  Subsequently committing {\bf Wd1} leads to the
-loss of {\bf Wd2}'s updates (the so called \emph{lost update
-anomaly}~\cite{berenson}), resulting in an incorrect balance of
-\C{k-a1}. The second execution describes a similar scenario with {\bf
-Wd1} and {\bf Wd2} exchanging their roles.  Clearly, RC is not
-sufficient because it loses the updates of one transaction resulting
-in the violation of post condition.  We need an isolation level that
-prevents lost updates.  \iso{Snapshot Isolation} (SI)~\cite{berenson}
-fits this requirement by its definition. SI implementations
-effectively serialize transactions that update a shared data object
-mostly by aborting and re-executing a transaction if write-write
-conflicts are detected during its commit.  Since SI, unlike SER, need
-not necessarily rely on expensive lock-based concurrency control, it
-is also more efficient, making it appropriate appropriate for both
-`Wd' transactions.
+  transactions are shown in different colored backgrounds.} In the
+first execution, transaction {\bf Wd1} reads the current balance
+(\C{k}) and writes the new balance (\C{k-a1}), but before it commits
+transaction {\bf Wd2} executes and commits, writing the new balance
+(\C{k-a2}). RC isolation prevent {\bf Wd2} from witnessing the
+uncommitted writes of transaction {\bf Wd1}.  Subsequently committing
+{\bf Wd1} leads to the loss of {\bf Wd2}'s updates (the so called
+\emph{lost update anomaly}~\cite{berenson}), resulting in an incorrect
+balance of \C{k-a1}. The second execution describes a similar scenario
+with {\bf Wd1} and {\bf Wd2} exchanging their roles.  Clearly, {\sc
+  rc} is an excessively weak isolation level for this program because
+it loses the updates of one transaction, resulting in the violation of
+the post-condition.  A stronger isolation level that prevents lost
+updates is required.  \iso{Snapshot Isolation} ({\sc
+  si})~\cite{berenson} fits this requirement; {\sc si} effectively
+serializes transactions that update a shared data object by aborting
+and re-executing a transaction if write-write conflicts are detected
+during its commit.\footnote{{\sc si} is not functionally the same as
+  {\sc ser}, however; the differences in the two manifest in the \emph{absence}
+  of any write-write conflicts.}   Since {\sc si},
+unlike {\sc ser}, does not rely on expensive lock-based
+concurrency control, it is also more efficient, making it appropriate
+appropriate for both {\bf Wd} transactions.
 
@@
 Unfortunately, such reasoning does not rest on any sound foundation,
-thus highly error-prone. Reasoning in terms of the implementations of
-weak isolation is not any better since it requires application
-programmers to understand and reason about low-level implementation
-details of the database (or, databases) that are far removed from the
-application semantics. An attractive alternative in this context is a
-formal proof system that combines declarative reasoning about
+and is thus highly error-prone. Reasoning in terms of how weak
+isolation variants are implementated is no better since it requires
+application programmers to understand low-level implementation details
+of the database that are naturally far removed from the application
+semantics. An attractive (and principled) alternative in this context
+is a formal proof system that combines declarative reasoning about
 isolation guarantees with operational reasoning about programs. We
@@
 
-\caption{$\hbZ$ arrows allowed by SI (solid) and RC (solid and dashed)
-isolation levels for the example in Fig.~\ref{fig:motiv-eg-1}. Each
-arrow represents $\hbZ$ relationship between all operations of `Wd2'
-and operations in `Wd1' that follow the arrow head. Letters R, W and C
-stand for read, write and commit, respectively.}
+\caption{$\hbZ$ arrows allowed by {\sc si} (solid) and {\sc rc} (solid
+  and dashed) isolation levels for the example in
+  Fig.~\ref{fig:motiv-eg-1}. Each arrow represents an $\hbZ$ relationship
+  among all operations of `Wd2' and operations in `Wd1' that follow
+  the arrow head. Letters R, W and C stand for read, write and commit,
+  respectively.}
 \label{fig:motiv-eg-1-hb}
@@
 
-Firstly, we note that the example in Fig.~\ref{fig:motiv-eg-1} is a
-concurrent program, hence admits \emph{rely-guarantee} style
-reasoning~\cite{rgjones}. Rely-guarantee is compositional proof
-technique that allows us to reason about one thread at a time by
-abstracting away the interference due to remaining threads
-(collectively called \emph{the environment}) into a \emph{rely}
-relation. In an ordinary concurrent program, every environment step is
-a valid interference in the current thread. However, in presence of
-transactions and variable isolation, determining what constitutes an
-interference and what does not is a non-trivial problem. For example,
-inside a serializable transaction no interference is a valid
-interference, whereas inside a transaction executing under \iso{Read
+First, we note that the example in Fig.~\ref{fig:motiv-eg-1} is a
+concurrent program, and hence is amenable to \emph{rely-guarantee}
+style reasoning~\cite{rgjones}. Rely-guarantee is a compositional
+proof technique that allows us to reason about one thread at a time by
+abstracting away interference induced by other threads (collectively
+called \emph{the environment}) into a \emph{rely} relation. In an
+ordinary concurrent program, every environment step is a valid
+interference in the current thread. However, in the presence of
+transactions executing under weak isolation, determining what
+constitutes an interference and what does not is a non-trivial
+problem. For example, inside a serializable transaction there is no
+interference, wherase inside a transaction executing under \iso{Read
   Uncommitted}, the weakest isolation level, all interferences are
-valid. Between these two extremes are various levels of isolation
-that admit some interferences while prohibiting others. 
+valid. Between these two extremes are various levels of isolation that
+admit some interferences while prohibiting others.
 % For example, \iso{Read Committed} isolation admits interference of a
@@
 % non-conflicting transactions, and so on. 
-For the rely guarantee approach to be useful it has to constrain the
+For the rely-guarantee approach to be useful, it has to thus constrain the
 interference in accordance with the chosen level of isolation. The
 first key observation we make is that we can constrain interference by
-constraining the nature of \emph{happens-before} ($\hbZ$) relation
-between transactions or their respective constituents. We make use of
-the $\hbZ$ relation to axiomatize the interference characteristic of
-isolation levels as their specifications ($\psi$). For example,
-\iso{Read Committed} specification allows an operation $\eta_1$ of
+constraining the definition of the \emph{happens-before} ($\hbZ$) relation
+between transactions, or their respective constituents.  To do so, we
+axiomatize  the $\hbZ$ relation to capture the interference characteristics of
+different isolation levels as their specifications ($\psi$). For example,
+a \iso{Read Committed} specification allows an operation $\eta_1$ of
 $T_1$ to happen before an operation $\eta_2$ of $T_2$ only if every
 operation in $T_1$ (including its commit) happens before $\eta_2$ (we
-write $T_1 \hboar \eta_2$ to denote this). RC spec does not require
-$T_1$ to happen before all operations of $T_2$, thus allowing it to
-interfere in $T_2$, as possible under RC isolation. \iso{Snapshot
-Isolation} spec requires two transactions that write to the same
-variable to be related by happens-before.  This effectively proscribes
-interference due to $T_1$ in $T_2$ (or, $T_2$ in $T_1$).
-Formally\footnote{Note that specs are simplified for illustration. The
-actual specs (\S~\ref{sec:opsem}) are more nuanced.  }:
+write $T_1 \hboar \eta_2$ to denote this). The {\sc rc} specification does not require
+$T_1$ to execute and commit before all operations of $T_2$, thus allowing it to
+interfere in $T_2$. The  \iso{Snapshot
+Isolation} specification requires two transactions that write to the same
+variable to be related by happens-before.  This effectively prohibits
+interference due to actions in $T_1$ being interleaved in $T_2$ (or, vise versa).
+Formally\footnote{Note that specifications are simplified for illustration. The
+actual specifications described in (\S~\ref{sec:opsem}) are more nuanced.  }:
 \begin{smathpar}
@@
 \end{smathpar}
-Fig.~\ref{fig:motiv-eg-1-hb} is a visualization of $\hbZ$ arrows
-from `Wd2' to `Wd1' allowed by $\psi_{RC}$and $\psi_{SI}$. All
-arrows re legal under $\psi_{RC}$ because in no case does an operation
-$\eta_1$ from `Wd1' happen before $\eta_2$ of `Wd2' without the commit
-of `Wd1' also happening before $\eta_2$. $\psi_{SI}$ however disallows
-all dashed arrows, becase dashed arrows establish $\hbZ$ between `Wd2'
-and only a subset of operations in `Wd1'. Ligher (blue) arrows with
-hollow heads denote $\hbZ$ relationships that do not effect the value
+Fig.~\ref{fig:motiv-eg-1-hb} is a visualization of $\hbZ$ 
+from {\bf Wd2} to {\bf Wd1} allowed by $\psi_{RC}$ and $\psi_{SI}$. All
+arrows are legal under $\psi_{RC}$ because in no case does an operation
+$\eta_1$ from {\bf Wd1} happen-before $\eta_2$ of {\bf Wd2} without the commit
+of {\bf Wd1} also happening before $\eta_2$. $\psi_{SI}$ however disallows
+all dashed arrows, because dashed arrows establish $\hbZ$ between actions in {\bf Wd2}
+and only a subset of the operations in {\bf Wd1}. Ligher (blue) arrows with
+hollow heads denote $\hbZ$ relationships that do not affect the value
 of \C{C} in a way that causes the program to violate its
-postcondition.  Darker (black) arrows with solid heads denote $\hbZ$
-relationships that lead to the violation of postcondition. The task of
+post-condition.  Darker (black) arrows with solid heads denote $\hbZ$
+relationships that lead to the violation of the post-condition. The task of
 the reasoning framework is to determine if all $\hbZ$ relationships
 allowed by an isolation level lead to the satisfaction of the
-postcondition.
+post-condition.
 
@@
 Specification of an isolation level encodes its interference
-characteristic as constraints over $\hbZ$ relation. However, for this
-to be useful in reasoning about programs, rely guarantee framework
+characteristic as constraints over the $\hbZ$ relation. However, for this
+to be useful in reasoning about programs, the rely-guarantee framework
 should be able to use the specification to determine if an
@@
 automatically prohibited. For instance, consider the program point
-after the write to \C{C} in `Wd1'. The expected invariant ($\phi$) at
+after the write to \C{C} in {\bf Wd1}. The expected invariant ($\phi$) at
 that program point is shown below ($\committed$ stands for
@@
 \end{smathpar}
-$\phi$ asserts that `Wd1' is not yet committed, and that it wrote to
+$\phi$ asserts that {\bf Wd1} is not yet committed, and that it wrote to
 \C{C}, and the value of \C{C} is either \C{k-a1-a2} or \C{k-a1}
-depending on whether or not `Wd2' is committed. If $\phi$ stays
-invariant until `Wd1' commits, then the postcondition (\C{C =
+depending on whether or not {\bf Wd2} is committed. If $\phi$ remains
+invariant until {\bf Wd1} commits, then the post-condition (\C{C =
 k-a1-a2}) can be established easily. However, an interference from
-`Wd2' at this stage (captured by the last dashed arrow in
+{\bf Wd2} at this stage (captured by the last dashed arrow in
 Fig.~\ref{fig:motiv-eg-1-hb}) may violate the invariant by writing
-$\C{k-a1}$ to \C{C} and committing `Wd2', thus leading to
+$\C{k-a1}$ to \C{C} and committing {\bf Wd2}, thus leading to
 $\committed(\C{Wd2}) \Rightarrow \C{C=k-a2}$. Fortunately,
 \iso{Snapshot Isolation} prevents this interference, and this can be
-proved by showing that an interference from `Wd2' starting from a
+shown by demonstrating that an interference from {\bf Wd2} starting from a
 execution state that satisfies $\psi_{SI} \wedge \phi$ leads to an
@@
 \hboar \C{Wd2}$ holds (former does not hold because at least one write
-from `Wd1' has happened before at least one operation of `Wd2', and
-the latter doesn't hold because `Wd1' hasn't yet committed and `Wd2'
-has already begun). Since `Wd2' also writes to \C{C}, this violates
+from {\bf Wd1} has happened before at least one operation of {\bf Wd2}, and
+the latter does not hold because {\bf Wd1} has not yet committed, and {\bf Wd2}
+has already begun). Since {\bf Wd2} also writes to \C{C}, this violates
 the $\psi_{SI}$ constraint which we assume to be an invariant. A proof
-for postcondition now follows from the contradiction. It is
+that the post-condition holds now follows from the contradiction. It is
 informative to note that if the invariant is $\psi_{RC}$ instead of
 $\psi_{SI}$, we cannot derive a contradiction and we cannot rule out
-the interference, which causes (rightfully) the proof to fail.
+the interference, which (rightfully) causes  the proof to fail.
 
-We have thus far assumed an SC store that, in the absence of
-transactions with special isolation requirements, makes effects of an
-operation to be immediately visible to subsequent operations. The
-natural behavior of SC store to totally order all operations w.r.t
-$\hbZ$ relation could be in conflict with the constraints imposed by
-weak isolation. To meet the requirements of weaker isolation levels,
-say RC, the store has to adapt itself to \emph{hide} the effects of
-concurrent transactions until they are committed. Once committed
-though, the effects need to be immediately and perennially visible to
-subsequent operations. This semantics of an SC store can be built into
-the reasoning framework leading to a proof system tailor-made for such
-stores. However, for the reasoning framework to be useful across the
-stores, it has to be parametric over store consistency semantics, and
-should be able to reconcile conflicts betweenconsistency and isolation
-constraints. Later sections demonstrate how our reasoning framework
-makes this possible.
+We have thus far assumed a strongly consistent ({\sc SC}) store that,
+in the absence of transactions with special isolation requirements,
+makes the effects of any operation immediately visible to all
+subsequent operations. The natural behavior of {\sc SC} to totally
+order all operations w.r.t the $\hbZ$ relation could be in conflict
+with the constraints imposed by weak isolation. To meet the
+requirements of weaker isolation levels, say {\sc rc}, the store has
+to adapt itself to \emph{hide} the effects of concurrent transactions
+until they are committed. Once committed though, effects need to be
+immediately and perennially visible to subsequent operations. This
+semantics can be built into the reasoning framework leading to a proof
+system tailor-made for such stores. However, for the reasoning
+framework to be truly useful, it has to be parametric over different
+store consistency semantics, supporting stores that are weaker than
+{\sc sc} (e.g., eventually consistent), and should be able to
+reconcile conflicts between consistency and isolation
+constraints.  We demonstrate how our reasoning framework
+makes this possible in the following sections.

### 2016-06-19 ansi-isolation.tex
Isolation Level Specs.

\subsection{Isolation Specifications}
\label{sec:ansi-isolation}

\begin{figure*}[t]
\begin{smathpar}
\begin{array}{lcl}
(\A,\visZ) \Vdash \eta \in T_i & \Leftrightarrow & \eta \in \A \conj \txn(\eta) = T_i\\
(\A,\visZ) \Vdash \eta_1 \visar \eta_2 & \Leftrightarrow & \{\eta_1,\eta_2\}
        \subseteq \A \conj (\eta_1,\eta_2) \in \visZ\\
(\A,\visZ) \Vdash \eta_1 \soar \eta_2 & \Leftrightarrow & \{\eta_1,\eta_2\}
        \subseteq \A \conj \txn(\eta_1)=\txn(\eta_2) \conj \id(\eta_1)
        < \id(\eta_2)\\
\E \Vdash S \subseteq T_i & \Leftrightarrow & \forall \eta.~ \eta
        \in S \Rightarrow \underE{\eta \in T_i} \\
\E \Vdash T_i \visar \eta & \Leftrightarrow &\forall\eta_1
        .\,(\E \Vdash \eta_1 \in T_i) \Rightarrow \E \Vdash \eta_1 \visar \eta \\
\underE{T_i \visar T_j} & \Leftrightarrow &  \forall\eta_1,\eta_2.\,
    %\sameobj{\eta_1}{\eta_2}  \Rightarrow 
    \underE{\eta_1\in T_i} \conj \underE{\eta_2 \in T_j} \Rightarrow 
    \underE{\eta_1 \visar \eta_2} \\
\underE{T_i \invisar T_j} & \Leftrightarrow &  \forall\eta_1,\eta_2.\,
        %\sameobj{\eta_1}{\eta_2}\Rightarrow 
        \underE{\eta_1\in T_i} \conj \underE{\eta_2 \in T_j} \Rightarrow 
        \neg (\underE{\eta_1 \visar \eta_2})\\
\underE{T_i \wrstoar X} & \Leftrightarrow & \exists\eta.~
        \underE{\eta \in T_i} \conj \kind(\eta) = \C{WR}(X)\\
\underE{\C{RMWVis}(T_j)} & \Leftrightarrow & \forall\eta_1,\eta_2.\,
       \underE{\{\eta_1,\eta_2\} \subseteq T_j} \conj
       \underE{\eta_1 \soar \eta_2} \Rightarrow \underE{\eta_1 \visar
       \eta_2}\\
\underE{\C{MonotonicVis}(T_j)} & \Leftrightarrow & 
       \underE{\C{RMWVis}(T_j)} \conj 
       \forall\eta,\eta_1,\eta_2.\, \underE{\{\eta_1,\eta_2\} \in T_j} 
          \conj \\
  &   & \hspace*{1.2in}\underE{\eta \visar \eta_1} \conj
        \underE{\eta_1 \soar \eta_2} \Rightarrow \underE{\eta \visar
        \eta_2} \\
%  \C{CausalVis}(T_i) & \Leftrightarrow & 
%         \C{MonotonicVis}(T_i) \conj \C{RMWVis}(T_i)\\
\underE{\C{AtomicVis}(T_j)} & \Leftrightarrow & 
       \forall\eta_1,\eta_2.\, \neg(\underE{\eta_1 \in T_j}) \conj
       \underE{\eta_2 \in T_j} \conj
       \underE{\eta_1 \visar \eta_2} \Rightarrow \underE{\txn(\eta_1)
       \visar \eta_2}\\
\underE{\C{CommitVis}(T_j)} & \Leftrightarrow & 
       \forall\eta_1,\eta_2.~ \neg(\underE{\eta_1 \in T_j}) \conj 
          \underE{\eta_2 \in T_j} \conj
       \underE{\eta_1 \visar \eta_2} \Rightarrow\\
  &   & \hspace*{1.2in}\exists\eta.\, \underE{\eta \in \txn(\eta_1)} 
        \conj \kind(\eta) = \C{COMMIT} 
        \conj \underE{\eta \visar \eta_2}\\
% \underE{\C{TransVis}(T_j)} & \Leftrightarrow &  \forall
%        \eta_1,\eta_2,\eta_3.\, \underE{\eta_3 \in T_j} \conj
%        \underE{\eta_1 \visar \eta_2} \conj
%        \underE{\eta_2 \visar \eta_3} \Rightarrow \underE{\eta_1 \visar
%        \eta_3} \\
\underE{\C{SnapshotVis}(T_i,T_j)} & \Leftrightarrow &  \underE{T_i
       \visar T_j} \disj \underE{T_i \invisar T_j}\\
\underE{\C{OneWaySER}(T_i,T_j)} & \Leftrightarrow &  \underE{T_i
       \visar T_j} \disj (\underE{T_i \invisar T_j} \conj \\
  &   &\hspace*{1.2in}\exists \eta.~\underE{\eta\in T_i} \conj \kind(\eta) =
       \C{COMMIT} \Rightarrow \underE{T_j \visar \eta})\\
\underE{\C{RC}(T_j)} & \Leftrightarrow & \underE{\C{AtomicVis}(T_j)} 
        \conj \underE{\C{CommitVis}(T_j)}\\
\underE{\C{MAV}(T_j)} & \Leftrightarrow & \underE{\C{RC}(T_j)} \conj
      \underE{\C{MonotonicVis}(T_j)} \\
\underE{\C{RR}(T_j)} & \Leftrightarrow & \underE{\C{MAV}(T_j)}
       \conj \forall T_i.\,T_i \neq T_j \Rightarrow 
        \underE{\C{SnapshotVis}(T_i,T_j)} \\
% &   & \hspace*{2in}\conj  \C{SnapshotVis}(T_i,T_j)\\
\underE{\C{SI}(T_j)} & \Leftrightarrow &  \underE{\C{RR}(T_j)}
       \conj \forall T_i.\,(T_i \neq T_j \conj \exists X.\, \underE{T_i \wrstoar X} \conj 
        \underE{T_j \wrstoar X})\\
%      (\exists X.\, T_i \wrstoar X \conj T_j \wrstoar X)
%      \Rightarrow  \C{TotalVis}(T_i,T_j)\\
  &   & \hspace*{2in} \Rightarrow \underE{\C{OneWaySER}(T_i,T_j)}\\
\underE{\C{SER}(T_j)} & \Leftrightarrow & \underE{\C{MAV}(T_j)}
       \conj \forall T_i.\,T_i \neq T_j 
       \Rightarrow \underE{\C{OneWaySER}(T_i,T_j)}\\
% &   & \hspace*{2in}\conj  \C{TotalVis}(T_i,T_j)\\
\end{array}
\end{smathpar}

\caption{Standard isolation guarantees expressed as trace
well-formedness constraints}
\label{fig:ansi-isolation}
\end{figure*}

Fig.~\ref{fig:ansi-isolation} shows specifications of standard
isolation guarantees expressed as constraints over trace
well-formedness. For brevity and convenience, we adopt some notations.
An execution trace is destructed into $\A$ and $\visZ$ whenever
individual components of the pair are needed.  Otherwise it is written
$\E$. When $\psi$ is a proposition, we write $\underE{\psi}$ to denote
an interpretation of $\psi$ in the context of the trace $\E$. Such
interpretations are defined on a case-by-case basis in
Fig.~\ref{fig:ansi-isolation}. In the following, we give informal
explanations for each definition.

In the context of a trace $(\A,\visZ)$, An effect $\eta$ is said to
belong to a transaction $T_i$ if $\eta$ belongs to the effect set $A$
and its transaction identifier is $T_i$. The containment relation is
trivially lifted to the set of effects to define $\underE{S \subseteq
T_i}$.  The interpretations of $\eta_1 \visar \eta_2$ and $\eta_1
\soar \eta_2$ are straightforward and explained in \S\ref{sec:syntax}.
A transaction $T_i$ is said to be visible to an effect $\eta$ if every
effect $\eta_1$ of $T_i$ recorded by the trace is visible to $\eta$.
$T_i$ may be visible to $\eta$ but may not be visible to every other
effect in the transaction. For a transaction $T_i$ to be considered to
be visible to a transaction $T_j$ in the context of a trace $\E$
(written $\underE{T_i \visar T_j}$), every effect ($\eta_1$) of $T_i$
present in $\E$ must be visible to every effect ($\eta_2$) of $T_j$ in
$\E$. Conversely, if none of the effects of $T_i$ present in $\E$ are
visible to any effect of $T_j$, then $T_i$ is considered invisible to
$T_j$ under $\E$ (written $\underE{T_i \invisar T_j}$). Transaction
$T_i$ is said to have written to a variable $X$ under $\E$ (i.e.,
$\underE{T_i \wrstoar X}$) if there exists a write-to-$X$ effect from
$T_i$ in $\E$.

Various isolation guarantees are defined as propositions indexed by
a transaction identifier $T_j$. Transaction $T_j$ is said to have
experienced \emph{read-my-writes} visibility under $\E$ if every
effect ($\eta_1$) of $T_j$ is visible to every subsequent effect
($\eta_2$) of the same transaction in $\E$. This lets $T_j$ to never
lose its own updates. Monotonic visibility adds one more constraint to
read-my-writes; besides requiring $\eta_1$ to be visible to $\eta_2$,
it also requires every effect ($\eta$) visible to $\eta_1$ in $\E$ to
be visible to $\eta_2$ as well. Thus $T_j$ witnesses monotonically
increasing state as time progresses. Atomic visibility allows an
effect $\eta_2$ of $T_j$ to witness an effect $\eta_1$ of $T_i$ only
if all effects of $T_i$ in $\E$ are also visible to $\eta_2$. Atomic
visibility thus prevents a transaction from being partially visible.
However, atomic visiblity does not prevent an uncommitted transaction
from being visible. To confine visibility to only committed
transactions, we need an additional constraint that requires an effect
($\eta_2$) of $T_j$ to also witness the commit effect of $T_i$
whenever it witnesses some effect ($\eta_1$) of $T_i$. This additional
constraint is captured by the $\mathtt{CommitVis}$ specification. A
transaction $T_i$ is said to be snapshot-visible to a transaction
$T_j$ if either it is visible to $T_j$, or it is invisible; it is
forbidden for only a suffix (more generally, a subset) of $T_j$ to
witness $T_i$. Snapshot visibility does not prevent $T_i$ and $T_j$ to
execute and commit oblivious of each other. One-way
serializability\footnote{\GK{This is a name I invented. Please feel
free to rename it appropriately.}} proscribes this possiblity. $T_i$
is said to be one-way-serializable w.r.t $T_j$ if either it is visible
to $T_j$ or $T_j$ is visible to the commit effect of $T_i$.  Note that
both $\mathtt{SnapshotVis}$ and $\mathtt{OneWaySER}$ are asymmetric
definitions that guarantee complete visiblity or invisibility of $T_i$
to $T_j$, but not the converse. While $\mathtt{SnapshotVis}$ provides
no guarantees to $T_i$, $\mathtt{OneWaySER}$ guarantees that at least
the commit effect of $T_i$ witnesses $T_j$.

ANSI SQL 92 standard requires \iso{Read Committted} isolation to avoid
dirty reads phenomenon, which is achieved by enforcing
$\mathtt{AtomicVis}$ and $\mathtt{CommitVis}$ guarantees. \C{RC}
specification is therefore a combination of these two guarantees. The
specification also agrees with the description and
implementation~\cite{bailishat,pldi15} of RC for highly available
replicated stores. On relational databases, however, RC has also come
to be associated with the $\mathtt{MonotonicVis}$ guarantee.
Nonetheless, $\mathtt{AtomicVis}$ and $\mathtt{CommitVis}$ are
sufficient to reason about RC isolation on relational stores too. As
we demonstrate in \S\ref{sec:store-consistency}, the combination of
these guarantees with the SC property of relational stores
automatically leads to the monotonicity guarantee, which probably
explains why RC comes with $\mathtt{MonotonicVis}$ on such stores. On
weakly consistent stores however, $\mathtt{AtomicVis}$ and
$\mathtt{CommitVis}$ do not imply $\mathtt{MonotonicVis}$. A stronger
isolation level called \emph{Monotonic Atomic View} (\C{MAV} of
Fig.~\ref{fig:ansi-isolation})~\cite{bailishat,pldi15} was proposed to
explicitly extend RC with monotonicity on such stores. \iso{Repeatable
Read} spec (\C{RR}) extends \C{MAV} with snapshot visibility guarantee
described before. \iso{Snapshot Isolation} spec (\C{SI}) extends
\C{RR} with one-way serializability guarantee w.r.t the transactions
that perform conflicting writes (i.e., writes to the same shared
variable). Finally, \iso{Serializable} isolation extends one-way
$\mathtt{SnapshotVis}$ to all transactions, including those that do
not conflict.

It is worth noting that, unlike some previous proposals
~\cite{gotsmanconcur15}, our specifications for SI and SER do not
necessarily impose a total order among transactions (conflicting or
otherwise). In reality, total order under SER (resp. SI) is guaranteed
only if the store executes all transactions under SER (resp. SI)
isolation. Our specifications admit this possibility, and derive total
order under the assumption of homogenity. However, databases almost
always allow isolation levels to be configured on a per-transaction
basis, allowing transactions at various isolation levels to coexist.
Specifications that assume homogenity are incorrect under this
setting\footnote{On the other hand, Fig.~\ref{fig:ansi-isolation}
presents slightly weaker versions of \C{SI} and \C{SER} specs in the
interest of clarity.}.

### 2016-06-20 ansi-isolation.tex
Example added.

@@
 $\mathtt{CommitVis}$ do not imply $\mathtt{MonotonicVis}$. A stronger
-isolation level called \emph{Monotonic Atomic View} (\C{MAV} of
+isolation level called \iso{Monotonic Atomic View} (\C{MAV} of
 Fig.~\ref{fig:ansi-isolation})~\cite{bailishat,pldi15} was proposed to
@@
 $\mathtt{SnapshotVis}$ to all transactions, including those that do
-not conflict.
+not conflict\footnote{Fig.~\ref{fig:ansi-isolation} presents slightly
+weaker versions of \C{SI} and \C{SER} specs in the interest of
+clarity.}.
 
@@
 Specifications that assume homogenity are incorrect under this
-setting\footnote{On the other hand, Fig.~\ref{fig:ansi-isolation}
-presents slightly weaker versions of \C{SI} and \C{SER} specs in the
-interest of clarity.}.
+setting.
+
+Having specified isolation levels as trace well-formedness
+constraints, we can now construct trace invariants ($\I$) for \txnimp
+programs by composing isolation specifications for various
+transactions. For instance, the following trace invariant enforces SI
+for both transactions of the program in Fig.~\ref{fig:motiv-eg-1},
+allowing it to satisfy its postcondition:
+\begin{smathpar}
+\I \;=\; \lambda\E.~ \underE{\C{SI(Wd1)}} \conj \underE{\C{SI(Wd2)}}
+\end{smathpar}
+In contrast, the following trace invariant enforces RC for one and SI
+for another, leading to a possible violation of the postcondition:
+\begin{smathpar}
+\I \;=\; \lambda\E.~ \underE{\C{RC(Wd1)}} \conj \underE{\C{SI(Wd2)}}
+\end{smathpar}

### 2016-06-20 store-consistency.tex
Store Consistency first commit.

\subsection{Store Consistency}
\label{sec:store-consistency}

Under a trivial trace invariant (i.e., $\I(\E) = true$), the abstract
machine of Fig.~\ref{fig:txnimp} assumes the semantics of an EC store;
it allows operations of a transaction witness arbitrary subsets of the
global state. A non-trivial $\I$ composed of isolation specifications
from Fig.~\ref{fig:ansi-isolation} induces the machine to provide
non-trivial isolation guarantees for transactions. However, weak
isolation levels often only constrain the visibility sets of
operations by dictating what \emph{not} to see; not what to see.  For
instance, \iso{Repeatable Read} isolation prohibits operations of a
transaction from witnessing different states. It however does not
prohibit all operations of a transaction from witnessing an aribitrary
subset of the global state. Consequently, the machine can remain an EC
store even while providing non-trivial isolation. How then to model
the semantics of an SC store, such as a relational database, with
variable isolation?

The answer lies in enforcing store-specific consistency constraints,
along with transaction-specific isolation constraints, via the trace
invariant $\I$. In particular, we split $\I$ into two components: (1).
$\I_s$, the store-specific invariant, and (2). $\I_c$, the
program-specific (or, client-specific) invariant, to capture
consistency and isolation constraints, respectively. $\I$ is now a
conjunction:
\begin{smathpar}
  \I \;=\; \lambda\E.~\I_s(\E) \wedge \I_c(\E)
\end{smathpar}
The trace invariant we wrote in \S\ref{sec:ansi-isolation} for the
withdraw program of Fig.~\ref{fig:motiv-eg-1} now becomes its $\I_c$
and remains an invariant regardless of the store. $\I_s$ however
changes from store to store. We now consider various stores and
describe their corresponding $\I_s$.

\paragraph{An EC store} An EC store provides no additional consistency
guarantees besides those provided by the machine. Hence, its
$\I_s(\E)$ is $true$ for any $\E$.

\paragraph{An SC store} An SC store guarantees total order of all
operations w.r.t $\visZ$ that is consistent with their chronological
order. A straightforward $\I_s$ for this store formalizes this
guarantee:
\begin{smathpar}
\begin{array}{l}
  \I_s \;=\; \lambda\E.~\forall\eta_1,\eta_2.~\{\eta_1,\eta_2\}
  \subseteq \E.\A \conj \id(\eta_1) < \id(\eta_2) \\
  \hspace*{2in}\Rightarrow \underE{\eta_1 \visar \eta_2}
\end{array}
\end{smathpar}
Unfortunately, the above $\I_s$ is inconsistent with all isolation
specifications of Fig.~\ref{fig:ansi-isolation}. For instance, let
$\eta_1$ be an effect of an uncommitted transaction and let $\eta_2$
be an effect of a \iso{Read Committed} transaction such that
$\id(\eta_1) < \id(\eta_2)$. While \C{RC} prohibits $\eta_1$ from
being visible $\eta_2$, $\I_s$ requires $\eta_1$ to be visible to
$\eta_2$. The machine, which cannot satisfy both constraints
simultaneously, gets stuck.

The way out of this impasse is to recognize that SC property of the
store needs to be weakened to admit weak isolation.

### 2016-06-20 store-consistency.tex
Store Consistency.

@@
 operations w.r.t $\visZ$ that is consistent with their chronological
-order. A straightforward $\I_s$ for this store formalizes this
-guarantee:
+order. A straightforward $\I_s$ for this store is the SC guarantee
+defined below:
 \begin{smathpar}
 \begin{array}{l}
-  \I_s \;=\; \lambda\E.~\forall\eta_1,\eta_2.~\{\eta_1,\eta_2\}
+  \C{SC}(\E) \;=\; \forall\eta_1,\eta_2.~\{\eta_1,\eta_2\}
   \subseteq \E.\A \conj \id(\eta_1) < \id(\eta_2) \\
@@
 \end{smathpar}
-Unfortunately, the above $\I_s$ is inconsistent with all isolation
-specifications of Fig.~\ref{fig:ansi-isolation}. For instance, let
-$\eta_1$ be an effect of an uncommitted transaction and let $\eta_2$
-be an effect of a \iso{Read Committed} transaction such that
-$\id(\eta_1) < \id(\eta_2)$. While \C{RC} prohibits $\eta_1$ from
-being visible $\eta_2$, $\I_s$ requires $\eta_1$ to be visible to
-$\eta_2$. The machine, which cannot satisfy both constraints
-simultaneously, gets stuck.
+Unfortunately, $\I_s=\C{SC}$ conflicts with all isolation
+specifications of Fig.~\ref{fig:ansi-isolation}. For instance,
+consider a case where $\I_c(\E) \;=\; \forall
+T_i.~\underE{\C{RC}(T_i)}$. That is, all transactions execute under
+\iso{Read Committed} isolation. Imagine a sample execution where
+$\eta_1$'s transaction is not yet committed when $\eta_2$ is
+generated. Letting $\eta_1$ be visible to $\eta_2$ violates $\I_c$,
+whereas not letting it be visible violates $\I_{s}$. The only way to
+satisfy both the invariants is to rule out all the executions that
+interleave the operations of one transaction with the other, thereby
+enforcing serializability and the ACID model of
+transactions\footnote{Thus, serializability is a natural
+generalization of SC to transactions.}. In general, when $\I_s$ is in
+conflict (but not inconsistent) with $\I_c$, the only way to enforce
+both is to restrict concurrency. Clearly, this is unacceptable since
+it defeats the very purpose of weak isolation. 
+% How then do we enforce weak isolation on a strongly consistent
+% machine?
 
-The way out of this impasse is to recognize that SC property of the
-store needs to be weakened to admit weak isolation.
+Relational databases demonstrate a way of out this impasse.
+Implementations of weak isolation levels on relational databases
+implicitly relax the SC requirement so as to maximize concurrency and
+improve performance. Consequently, execution traces do not necessarily
+satisfy \C{SC}. This approach hints at a general solution reconciling
+$\I_s$ with $\I_c$, which is to weaken the former. However, there are
+many ways $\I_s$ can be weakened; for instance, by setting it to
+$true$, which makes the store EC. In reality though, weak isolation on
+databases does not cause so drastic a reduction in consistency. It is
+therefore possible that there exists a a principled approach to weaken
+$\I_s$ that adequately captures the underlying reality. Such a
+principle can be gleaned by observing that data stores often make
+\emph{recency} commitments~\cite{bailishat}, by which they aim to make
+the most recent data available to clients. SC is infact an extreme
+form of a recency commitment, which guarantees that reads witness most
+recent writes to a data item. In case this conflicts with weak
+isolation guarantees, such as RR which requires all reads in a
+transaction to witness the same state, the natural way to weaken
+consistency while prioritizing recency is to let the reads witness
+most recent state that \emph{does not} violate the isolation
+constraints. In the context of the operational model of
+Fig.~\ref{fig:txnimp}, this translates to making visible to an
+operation the largest subset ($S$) of the global state ($\A$) that
+does not lead to the violation of isolation constraints. Generalizing
+this approach to any $\I_s$ and $\I_c$ gives the \emph{maximum
+visiblity principle}, which requires the weakened consistency
+guarantee ($\I_s'$) of a store to enforce all visibility relationships
+imposed by the actual consistency guarantee ($\I_s$) on a given trace,
+unless enforcing such a relationship violates $\I_c$.  Formally,
+maximum visibility weakening of $\I_s$ is an $\I_s'$, such that:
+\begin{itemize}
+  \item $\I_s'$ is weaker than $\I_s$: 
+      $\forall\E.~ \I_s(\E) \Rightarrow \I_s'(\E)$, and
+  \item In every trace $\E$ that satisfies $\I_c$, and for every pair
+  of effects $\eta_1$ and $\eta_2$ in $\E$, if $\I_s(\E)$ requires
+  $\eta_1$ to be visible to $\eta_2$, then so does $\I_s'(\E)$ unless
+  extending $\E$ with $\visZ(\eta_1,\eta_2)$ violates
+  $\I_c$\footnote{\GK{ToDo: consider other well-formedness conditions
+  on trace, such as acyclicity of $\visZ$ and $\soZ$. Are they needed
+  (considering that the machine never violates them)? Encode the
+  specifications in Z3 and make sure they are consistent.}}:
+  \begin{smathpar}
+  \begin{array}{l}
+  \forall\E,\eta_1,\eta_2.~ \I_c(\E) \Rightarrow (\I_s(\E)
+    \Rightarrow \underE{\eta_1 \visar \eta_2}) \Rightarrow \\
+    \hspace*{0.5in}(\I_s'(\E) \Rightarrow \underE{\eta_1 \visar
+    \eta_2} \disj \neg\I_c(\E\,\cup\,(\emptyset,\{(\eta_1,\eta_2)\})))
+  \end{array}
+  \end{smathpar}
+\end{itemize}
+
+Applying maximum visibility principle, we weaken \C{SC} to obtain the
+following store trace invariant ($\I_s$) for an SC store whose
+isolation constraints are captured by $\I_c$:
+\begin{smathpar}
+\begin{array}{lcl}
+\I_s(\E) & = & \forall \eta_1,\eta_2.\, \{\eta_1,\eta_2\},
+    \subseteq \E.\A \conj \id(\eta_1) <
+    \id(\eta_2) \\
+    & & \hspace*{0.5in} \Rightarrow 
+      \underE{\eta_1 \visar \eta_2} \disj \neg\I_c(\E
+    \cup (\emptyset,\{(\eta_1,\eta_2)\}))\\
+\end{array}
+\end{smathpar}
+As usual $\I = \I_s \wedge \I_c$. The resultant operational model
+hides an existing effect ($\eta\in\A$) from being visible to the
+current operation if and only if showing it results in the violation
+of $\I_c$.
+
+\paragraph{A CC store} A causally consistent data store, such as the
+one underlying the reasoning framework of~\cite{gotsmanpopl16}, allows
+operations to only witness a causally consistent snapshot of the
+global state. A straightforward trace invariant ($\I_s$) for this
+store is the \C{CC} guarantee that requires an effect to witness all
+causally preceding effects regardless of the isolation requirements of
+its transaction:
+\begin{smathpar}
+\begin{array}{lccl}
+\C{CC}(\E) & \;=\; &  & \forall \eta_1,\eta_2.\, 
+      \E \Vdash \eta_1 \soar \eta_2 \Rightarrow  \underE{\eta_1 \visar
+      \eta_2}\\
+    &   & \wedge & \forall\eta_1,\eta_2,\eta_3.\,\underE{\eta_1 \visar
+      \eta_2} \conj \underE{\eta_2 \visoar \eta_3} \\
+    &   & &\hspace*{0.5in} \Rightarrow \underE{\eta_1 \visar \eta_3}\\
+\end{array}
+\end{smathpar}
+However, like $\I_s = \C{SC}$, $\I_s = \C{CC}$ may induce the store to
+meet isolation requirements by restricting concurrency, thus failing
+to abide by the spirit of weak isolation. In order to admit weak
+isolation without restricting concurrency, we weaken $\I_s$ while
+being guided by the maximum visibility principle. The weakened $\I_s$
+is shown below:
+\begin{smathpar}
+\begin{array}{lccl}
+\C{CC}(\E) & \;=\; &  & \forall \eta_1,\eta_2.\, 
+      \E \Vdash \eta_1 \soar \eta_2 \Rightarrow  \underE{\eta_1 \visar
+      \eta_2} \\
+    & & & \hspace*{0.6in}\disj \neg\I_c(\E \cup 
+                (\emptyset,\{(\eta_1,\eta_2)\}))\\
+    &   & \wedge & \forall\eta_1,\eta_2,\eta_3.\,\underE{\eta_1 \visar
+      \eta_2} \conj \underE{\eta_2 \visoar \eta_3} \\
+    &   & &\hspace*{0.3in} \Rightarrow \underE{\eta_1 \visar \eta_3}
+      \disj \neg\I_c(\E \cup (\emptyset,\{(\eta_1,\eta_2)\}))\\
+\end{array}
+\end{smathpar}
+Instantiating the parameter $\I$  with $\I_s \wedge \I_c$ results in
+an operational semantics that admits the violation of causal
+consistency if and only if the violation is inevitable to enforce
+$\I_c$.

### 2016-06-20 store-consistency.tex
Edits.

@@
 \end{smathpar}
-Instantiating the parameter $\I$  with $\I_s \wedge \I_c$ results in
-an operational semantics that admits the violation of causal
-consistency if and only if the violation is inevitable to enforce
+Instantiating the parameter $\I$  with $\I_s \wedge \I_c$ in
+Fig.~\ref{fig:txnimp} results in an operational semantics that admits
+the violation of causal consistency if and only if the violation is
+inevitable to enforce
 $\I_c$.

### 2016-06-20 motivation.tex
edits to op sem

@@
 }
+\caption{A possible execution of the program shown in Fig.~\ref{fig:motiv-eg-1} under
+  a read committed isolation level.}
 \end{figure}
@@
 concurrent program, and hence is amenable to \emph{rely-guarantee}
-style reasoning~\cite{rgjones}. Rely-guarantee is a compositional
-proof technique that allows us to reason about one thread at a time by
-abstracting away interference induced by other threads (collectively
-called \emph{the environment}) into a \emph{rely} relation. In an
-ordinary concurrent program, every environment step is a valid
-interference in the current thread. However, in the presence of
-transactions executing under weak isolation, determining what
-constitutes an interference and what does not is a non-trivial
-problem. For example, inside a serializable transaction there is no
-interference, wherase inside a transaction executing under \iso{Read
-  Uncommitted}, the weakest isolation level, all interferences are
-valid. Between these two extremes are various levels of isolation that
-admit some interferences while prohibiting others.
-% For example, \iso{Read Committed} isolation admits interference of a
-% committed transaction, but not that of an uncommitted transaction.
-% \iso{Snapshot Isolation} admits interference from committed
-% non-conflicting transactions, and so on. 
-For the rely-guarantee approach to be useful, it has to thus constrain the
-interference in accordance with the chosen level of isolation. The
-first key observation we make is that we can constrain interference by
-constraining the definition of the \emph{happens-before} ($\hbZ$) relation
-between transactions, or their respective constituents.  To do so, we
-axiomatize  the $\hbZ$ relation to capture the interference characteristics of
-different isolation levels as their specifications ($\psi$). For example,
-a \iso{Read Committed} specification allows an operation $\eta_1$ of
+style reasoning~\cite{rgjones}, a compositional proof technique that
+allows us to reason about one thread at a time by abstracting away
+interference induced by other threads (collectively called \emph{the
+  environment}) into a \emph{rely} relation. In an ordinary concurrent
+program, every environment step is a valid interference in the current
+thread. However, in the presence of transactions executing under weak
+isolation, determining what constitutes an interference is a
+non-trivial problem. For example, inside a serializable transaction
+there is no interference, whereas within a transaction executing under
+\iso{Read Uncommitted}, the weakest isolation level, all interferences
+are valid. Between these two extremes are various levels of isolation
+that admit some interferences while prohibiting others.  For example,
+\iso{Read Committed} isolation admits interference of a committed
+transaction, but not that of an uncommitted transaction.
+\iso{Snapshot Isolation} admits interference from committed
+non-conflicting transactions, and so on.
+
+For the rely-guarantee approach to be useful, it has to thus constrain
+allowed interference in accordance with the chosen level of isolation
+associated with a transaction. The first key observation we make is
+that we can constrain interference by constraining the definition of
+the \emph{happens-before} ($\hbZ$) relation between transactions, or
+their respective constituents.  To do so, we axiomatize the $\hbZ$
+relation to capture the interference characteristics of different
+isolation levels as their specifications ($\psi$). For example, a
+\iso{Read Committed} specification allows an operation $\eta_1$ of
 $T_1$ to happen before an operation $\eta_2$ of $T_2$ only if every
 operation in $T_1$ (including its commit) happens before $\eta_2$ (we
-write $T_1 \hboar \eta_2$ to denote this). The {\sc rc} specification does not require
-$T_1$ to execute and commit before all operations of $T_2$, thus allowing it to
-interfere in $T_2$. The  \iso{Snapshot
-Isolation} specification requires two transactions that write to the same
-variable to be related by happens-before.  This effectively prohibits
-interference due to actions in $T_1$ being interleaved in $T_2$ (or, vise versa).
-Formally\footnote{Note that specifications are simplified for illustration. The
-actual specifications described in (\S~\ref{sec:opsem}) are more nuanced.  }:
+write $T_1 \hboar \eta_2$ to denote this). The {\sc rc} specification
+does not require $T_1$ to execute and commit before all operations of
+$T_2$, thus allowing it to interfere in $T_2$. The \iso{Snapshot
+  Isolation} specification requires two transactions that write to the
+same variable to be related by happens-before.  This effectively
+prohibits interference due to actions in $T_1$ being interleaved in
+$T_2$ (or, vise versa).  Formally\footnote{Note that specifications
+  are simplified for illustration. The actual specifications described
+  in (\S~\ref{sec:opsem}) are more nuanced.  }:
 \begin{smathpar}
@@
 
-Specification of an isolation level encodes its interference
+The specification of an isolation level encodes its interference
 characteristic as constraints over the $\hbZ$ relation. However, for this
@@
 \iso{Snapshot Isolation} prevents this interference, and this can be
-shown by demonstrating that an interference from {\bf Wd2} starting from a
+shown by demonstrating that an interference from {\bf Wd2} starting from an
 execution state that satisfies $\psi_{SI} \wedge \phi$ leads to an
 execution state where neither $\C{Wd2} \hboar \C{Wd1}$ nor $\C{Wd1}
-\hboar \C{Wd2}$ holds (former does not hold because at least one write
+\hboar \C{Wd2}$ holds (the former does not hold because at least one write
 from {\bf Wd1} has happened before at least one operation of {\bf Wd2}, and
@@
 immediately and perennially visible to subsequent operations. This
-semantics can be built into the reasoning framework leading to a proof
+semantics can be built into the reasoning framework, leading to a proof
 system tailor-made for such stores. However, for the reasoning

### 2016-06-20 opsem.tex
Isolation Level Specs.
Store Consistency first commit.
edits to op sem

@@
 Fig.~\ref{fig:txnimp} shows the syntax and small-step semantics of
-\txnimp. Natural numbers, (shared) variables and arithmetic
-expressions constitute the syntactic class of expressions ($e$).
-Commands ($c$) include $\cskip$, assignment statements, transaction
-(\C{txn}) lexical blocks, and their sequential and parallel
-composition. We let $T_i$ for $i \in \mathbb{N}$ range over
-transaction identifiers. When it is evident we are referring to a
+\txnimp, a core language that we will use to formalize the intuitions
+presented in the previous section. Natural numbers, (shared) variables
+and arithmetic expressions constitute the syntactic class of
+expressions ($e$).  Commands ($c$) include $\cskip$, assignment
+statements, transaction (\C{txn}) lexical blocks, and their sequential
+and parallel composition. We let $T_i$ for $i \in \mathbb{N}$ range
+over transaction identifiers. When it is evident we are referring to a
 transaction, we use the number $i$ instead of $T_i$ for identification
-(\eg, in $\C{txn}\langle i \rangle$). For notational convenience we let
-$t$ range over both expressions and commands.
+(\eg, in $\C{txn}\langle i \rangle$). For notational convenience, we
+let $t$ range over both expressions and commands.
 
-Small-step semantics of \txnimp are defined in terms of an abstract
-machine that generates an execution trace ($\E$). First artifact of the trace is a set ($\A$) of \emph{effects},
-where an effect ($\eta$) documents a read (\C{RD(X)}), or a write
-(\C{WR(X)}), or a transaction commit (\C{COMMIT}) operation performed
-during the execution. Any value associated with the operation (\eg,
-value read or value written) is also documented, and can be accessed
-via $\rval$. Every effect has a unique identifier accessible via
-$\id$, and a transction identifier accessible via $\txn$.  The latter
-identifies the transaction that generated the effect. In every step of
-the evaluation, abstract machine reduces a \txnimp term by executing a
-read, write or commit operation, generating an effect, and extending
-the trace. Since effects include transaction identifiers, semantics
-distinguishes between terms ($t$) of different transactions. For
-example, $\txnbox{t}_i$ denotes a term $t$ inside a transaction $T_i$.
-Evaluation contexts are also appropriately marked. For example,
-$\ectx_i$ denotes evaluation context for a term inside $T_i$. The
-other component of the execution trace is the visibility relation
-($\visZ$) between the effects. The intent and mechanics of $\visZ$
-become clear as we describe the evaluation rules in the following.
+We define a small-step operational semantics for this language in
+terms of an abstract machine that generates an execution trace
+($\E$). The first component of the trace is a set ($\A$) of
+\emph{effects}, where an effect ($\eta$) documents a read (\C{RD(X)}),
+a write (\C{WR(X)}), or a transaction commit (\C{COMMIT}) operation
+performed during the execution. Any value associated with the
+operation (\eg, value read or value written) is also documented, and
+can be accessed via $\rval$. Every effect has a unique identifier
+accessible via $\id$, and a transction identifier accessible via
+$\txn$.  The latter identifies the transaction that generated the
+effect. In every step of the evaluation, the machine reduces a \txnimp
+term by executing a read, write or commit operation, generating an
+effect, and extending the trace. Since effects include transaction
+identifiers, the semantics distinguishes between terms ($t$) of
+different transactions. For example, $\txnbox{t}_i$ denotes a term $t$
+inside a transaction $T_i$.  Evaluation contexts are also
+appropriately marked. For example, $\ectx_i$ denotes the evaluation
+context for a term inside $T_i$. The other component of an execution
+trace is a visibility relation ($\visZ$) that establishes a visibility
+property between effects among different transactions. The intent and
+mechanics of $\visZ$ is described below.
 
-Fundamental to our operational semantics is the notion of a trace
-invariant ($\I$), which is a function from traces ($\E$) to
-propositions ($\texttt{Prop}$) that declare well-formedness
-constraints over traces. Abstract machine takes a step only if the
-resulting trace satisfies the well-formedness constraints imposed by
-$\I$. This behaviour is captured by the auxiliary reduction rule
-\rulelabel{E-Aux} that factors out the trace extension aspect of the
-evaluation by abstracting away operation-specific behaviour as a
-function ($\mathcal{F}$) that generates appropriate effect.
-\rulelabel{E-Aux} uses $\mathcal{F}$ to generate a new effect and
-extend the trace ($\E = (\A,\visZ)$) \emph{only if} the
-well-formedness constraints imposed by $\I$ on $\E$ (i.e., $\I(\E')$)
-are satisfied. Otherwise, it gets stuck. In an execution that runs to
-completion, every small-step preserves well-formedness of trace, thus
-ensuring the invariance of $\I$. Note that the operational semantics
-makes no assumptions about $\I$ other than its type. As such it is
-parametric over $\I$, and can be instantiated with any
-trace-parametric proposition that expresses constraints over the given
-trace. While the machine takes a step only if the constraints are
-satisfied, it neither defines nor explicitly assumes an oracle to
-check the satisfaction; there is no need for doing so in the
-semantics.
+Fundamental to our development is the notion of a trace invariant
+($\I$), is a function from traces ($\E$) to propositions
+($\texttt{Prop}$) that defines well-formedness constraints over
+traces. The machine takes a step only if the resulting trace satisfies
+the constraints imposed by $\I$. This behaviour is captured by the
+auxiliary reduction rule \rulelabel{E-Aux} that factors out the trace
+extension aspect of the evaluation by abstracting away
+operation-specific behaviour as a function ($\mathcal{F}$) that
+generates an appropriate effect.  \rulelabel{E-Aux} uses $\mathcal{F}$
+to generate a new effect and extend the trace ($\E = (\A,\visZ)$)
+\emph{only if} the well-formedness constraints imposed by $\I$ on $\E$
+(i.e., $\I(\E')$) are satisfied. Otherwise, it gets stuck. In an
+execution that runs to completion, every small-step preserves the
+well-formedness of a trace, thus ensuring the invariance of $\I$.
+Note that the semantics makes no assumptions about $\I$ other than its
+type. As such, it can be instantiated with any trace-parametric
+proposition that expresses constraints over the given trace. While the
+machine takes a step only if the constraints are satisfied, it neither
+defines nor explicitly assumes an oracle to check satisfaction.
 
-As explained in \S\ref{sec:motivation}, semantics of various isolation
-levels can be captured as constraints over the happens-before ($\hbZ$)
-relation. $\hbZ$ is however a derived relation in our model, composed
-of more fundamental \emph{session order} ($\soZ$) and
-\emph{visibility} ($\visZ$) relations. Session order captures the
-sequential order of operations within a transaction. In particular, it
-relates two effects, $\eta_1$ and $\eta_2$, such that $\txn(\eta_1) =
-\txn(\eta_2)$ and $\id(\eta_1) < \id(\eta_2)$. Small-step semantics
-assigns monotonically increasing identifiers to effects, as captured
-by the $\id(\eta) > \maxId(\A)$ condition of \rulelabel{E-Aux}.
-Evaluation contexts ($\ectx_i$) for transaction-bound terms are
-defined so as to enforce a deterministic sequential order of execution
-within a transaction, leading to a deterministic total order among
-effect ids - the session order. Visibility ($\visZ$) on the other
-hand relates effects across concurrent transactions, thus reflects the
-non-determinism inherent in concurrency. Intuitively, $\visZ$ relates
-$\eta_1$ to $\eta_2$ if and only if $\eta_1$ was \emph{visible} to the
-operation that generated $\eta_2$ during its execution, thus 
-effecting its return value ($\rval(\eta_2)$). For example, a read
-operation over \C{X} may pick the value ($\rval$) of the write effect
-with highest id among the visible effects (this is made possible by
-appropriately defining $\interp{\cdot}$ in \rulelabel{E-Var},
-as we show later). Thus, the value of a read depends on what write
-effects it can witness. An operation can only witness the effects of
-already concluded operations, which varies between executions due to
-the non-deterministic order of evaluating the parallel composition
-(cf. evaluation contexts ($\ectx$) for parallel composition). A more
-notable source of non-determinism, however, is the \rulelabel{E-Aux}
-rule, which allows the machine to expose an arbitrary subset ($S$) of
-existing effects ($\A$) to the incoming operation. In other words, the
-machine is not obligated to reveal the effects of all previous
-operations to an incoming operation. This relaxation allows the
-abstract machine to model the semantics of weakly consistent data
-stores. For instance, operations issued to an eventually consistent
-(EC) replicated store could be dispatched to different replicas whose
-states may not be in any well-defined relationship. By allowing
-operations to witness arbitrary subsets of the global state, the
-abstract machine effectively models an EC store.
+As described in \S\ref{sec:motivation}, the semantics of various
+isolation levels can be captured as constraints over the
+happens-before ($\hbZ$) relation. $\hbZ$ is however a derived relation
+in our model, composed of more fundamental \emph{session order}
+($\soZ$) and \emph{visibility} ($\visZ$) relations.  A session order
+relation captures the sequential order of operations within a
+transaction. In particular, it relates two effects, $\eta_1$ and
+$\eta_2$, such that $\txn(\eta_1) = \txn(\eta_2)$ and $\id(\eta_1) <
+\id(\eta_2)$.  The semantics assigns monotonically increasing
+identifiers to effects, as defined by the $\id(\eta) > \maxId(\A)$
+condition of \rulelabel{E-Aux}.  Evaluation contexts ($\ectx_i$) for
+transaction-bound terms are defined so as to enforce a deterministic
+sequential order of execution within a transaction, leading to a
+deterministic total order among effect ids, which defines the session
+order relation. Visibility ($\visZ$) on the other hand relates effects
+across concurrent transactions.  Intuitively, $\visZ$ relates $\eta_1$
+to $\eta_2$ if and only if $\eta_1$ was \emph{visible} to the
+operation that generated $\eta_2$ during its execution, thus effecting
+its return value ($\rval(\eta_2)$). For example, a read operation over
+\C{X} may pick the value ($\rval$) of the write effect with highest id
+among the visible effects (this is made possible by appropriately
+defining $\interp{\cdot}$ in \rulelabel{E-Var}, as we show
+later). Thus, the value of a read depends on what write effects it can
+witness. An operation can only witness the effects of already
+concluded operations, which varies between executions due to the
+non-deterministic order of evaluating the parallel composition
+(cf. evaluation contexts ($\ectx$) for parallel composition) of
+transactions. A more notable source of non-determinism, however, is
+the \rulelabel{E-Aux} rule, which allows the machine to expose an
+arbitrary subset ($S$) of existing effects ($\A$) to the incoming
+operation. In other words, the machine is not obligated to reveal the
+effects of all previous operations to an incoming operation. This
+relaxation allows the abstract machine to model the semantics of
+weakly consistent data stores. For instance, operations issued to an
+eventually consistent (EC) replicated store could be dispatched to
+different replicas whose states may not be in any well-defined
+relationship. By allowing operations to witness arbitrary subsets of
+the global state, the semantics models the weak visibility properties
+of such stores.
 
 However, a machine that lets operations witness arbitrary subset of
-the global state offers no isolation whatsoever. For example, it lets
-a read operation witness writes of an uncommitted transaction,
-violating RC isolation. Fortunately, our ability to express isolation
-semantics as constraints over happens-before order through $\visZ$ and
-$\soZ$ relations, and the property of the abstract machine to be
-parametric over the trace invariant ($\I$) lets us solve this problem.
-In particular, we continue defining isolation semantics as constraints
-over $\visZ$ and $\soZ$, but confine their domain of interpretation to
-the given trace so that they now become trace well-formedness
-constraints. Well-formedness constraints can be combined into a trace
-invariant ($\I$). Since the abstract machine is guaranteed to preserve
-$\I$ at every step of the execution, operations are guaranteed to
-experience the level of isolation specified by $\I$.  Thus, in
-executions that run to completion abstract machine models a store that
-provides the required levels of isolation. Notably, the machine
-achieves this without confining to any particular implementation of
-the isolation, by relying on a declarative specification of isolation
-as trace well-formedness constraints.  \S\ref{sec:ansi-isolation}
-shows specifications of various ANSI SQL isolation levels stated as
-trace well-formedness constraints.
+the global state offers no isolation whatsoever. For example, it may
+allow a read operation to witness writes of an uncommitted
+transaction, violating RC isolation. Fortunately, our ability to
+express an isolation semantics as constraints over happens-before
+order through $\visZ$ and $\soZ$ relations, and the property of the
+abstract machine to be parametric over the trace invariant ($\I$),
+lets us solve this problem.  In particular, we continue to define
+isolation semantics as constraints over $\visZ$ and $\soZ$, but
+confine their domain of interpretation to the given trace so that they
+now become trace well-formedness constraints. Well-formedness
+constraints can be combined into a trace invariant ($\I$). Since the
+abstract machine is guaranteed to preserve $\I$ at every step of the
+execution, operations are guaranteed to experience the level of
+isolation specified by $\I$.  Thus, in executions that run to
+completion, the abstract machine models a store that provides the
+required levels of isolation. Notably, the machine achieves this
+without being restricted to any particular isolation level by relying
+on a declarative specification of isolation as trace well-formedness
+constraints.  \S\ref{sec:ansi-isolation} shows the specification of
+various ANSI SQL isolation levels stated as trace well-formedness
+constraints.
 
@@
 operation-specific behaviour of a machine step as a function ($\F$)
-that accepts a set ($S$) of effects chose by the machine to make
+that accepts a set ($S$) of effects chosen by the machine to make
 visible to the operation, interprets the operation w.r.t $S$, and
@@
 example, a monotonically increasing counter application may choose to
-let a write with largest value determine the value of a
+let a write with the largest value determine the value of a
 read. A more straightforward interpretation has already been described
-before. To accommodate multiple interpretations, operational semantics
+before. To accommodate multiple interpretations, the semantics
 is made parametric over an interpretation function ($\interp{\cdot}$)
@@
 \end{smathpar}
-Rules \rulelabel{E-Top-Ctx} and \rulelabel{E-Txn-Ctx} rules define
-congruence rules for top-level terms and transaction-bound terms,
+Rules \rulelabel{E-Top-Ctx} and \rulelabel{E-Txn-Ctx} define
+congruence properties for top-level terms and transaction-bound terms,
 respectively. The rules and evaluation contexts ($\ectx$ and
-$\ectx_i$) are defined such that only a certain kinds of terms are
+$\ectx_i$) are defined such that only certain kinds of terms are
 allowed at the top-level and inside a transaction. In particular, a
@@
 can either be an assignment, or a sequential composition of
-assignments. Nested transactions and transactions that fork multiple
-threads are forbidden to focus on the problems relevant to this paper.
+assignments.
 
-\subsection{Isolation Specifications}
-\label{sec:ansi-isolation}
+\input{ansi-isolation}
 
-\subsection{Store Consistency}
-\label{sec:store-consistency}
+\input{store-consistency}

### 2016-06-20 ansi-isolation.tex
edits

@@
 
-Fig.~\ref{fig:ansi-isolation} shows specifications of standard
+Fig.~\ref{fig:ansi-isolation} shows the specification of standard
 isolation guarantees expressed as constraints over trace
-well-formedness. For brevity and convenience, we adopt some notations.
+well-formedness. For brevity and convenience, we adopt some notation.
 An execution trace is destructed into $\A$ and $\visZ$ whenever
-individual components of the pair are needed.  Otherwise it is written
-$\E$. When $\psi$ is a proposition, we write $\underE{\psi}$ to denote
+individual components of the pair are needed.  Otherwise, it is written
+as $\E$. When $\psi$ is a proposition, we write $\underE{\psi}$ to denote
 an interpretation of $\psi$ in the context of the trace $\E$. Such
@@
 
-In the context of a trace $(\A,\visZ)$, An effect $\eta$ is said to
+In the context of a trace $(\A,\visZ)$, an effect $\eta$ is said to
 belong to a transaction $T_i$ if $\eta$ belongs to the effect set $A$
@@
 trivially lifted to the set of effects to define $\underE{S \subseteq
-T_i}$.  The interpretations of $\eta_1 \visar \eta_2$ and $\eta_1
+T_i}$.  The interpretation of $\eta_1 \visar \eta_2$ and $\eta_1
 \soar \eta_2$ are straightforward and explained in \S\ref{sec:syntax}.
@@
 forbidden for only a suffix (more generally, a subset) of $T_j$ to
-witness $T_i$. Snapshot visibility does not prevent $T_i$ and $T_j$ to
-execute and commit oblivious of each other. One-way
-serializability\footnote{\GK{This is a name I invented. Please feel
-free to rename it appropriately.}} proscribes this possiblity. $T_i$
+witness $T_i$. Observe that snapshot visibility permits $T_i$ and $T_j$ to
+execute and commit while being  oblivious of each other. One-way
+serializability proscribes this possiblity. $T_i$
 is said to be one-way-serializable w.r.t $T_j$ if either it is visible
@@
 
-ANSI SQL 92 standard requires \iso{Read Committted} isolation to avoid
-dirty reads phenomenon, which is achieved by enforcing
-$\mathtt{AtomicVis}$ and $\mathtt{CommitVis}$ guarantees. \C{RC}
+The ANSI SQL 92 standard requires \iso{Read Committted} isolation to avoid
+the dirty reads phenomenon, which is achieved by enforcing
+$\mathtt{AtomicVis}$ and $\mathtt{CommitVis}$ guarantees. The {\sc rc}
 specification is therefore a combination of these two guarantees. The
 specification also agrees with the description and
-implementation~\cite{bailishat,pldi15} of RC for highly available
-replicated stores. On relational databases, however, RC has also come
+implementation~\cite{bailishat,pldi15} of {\sc rc} for highly available
+replicated stores. On relational databases, however, {\sc rc} has also come
 to be associated with the $\mathtt{MonotonicVis}$ guarantee.
 Nonetheless, $\mathtt{AtomicVis}$ and $\mathtt{CommitVis}$ are
-sufficient to reason about RC isolation on relational stores too. As
+sufficient to reason about {\sc rc} isolation on relational stores too. As
 we demonstrate in \S\ref{sec:store-consistency}, the combination of
-these guarantees with the SC property of relational stores
+these guarantees with the {\sc sc} property of relational stores
 automatically leads to the monotonicity guarantee, which probably
-explains why RC comes with $\mathtt{MonotonicVis}$ on such stores. On
+explains why {\sc rc} comes with $\mathtt{MonotonicVis}$ on such stores. On
 weakly consistent stores however, $\mathtt{AtomicVis}$ and
 $\mathtt{CommitVis}$ do not imply $\mathtt{MonotonicVis}$. A stronger
-isolation level called \iso{Monotonic Atomic View} (\C{MAV} of
+isolation level called \iso{Monotonic Atomic View} ({\sc mav} of
 Fig.~\ref{fig:ansi-isolation})~\cite{bailishat,pldi15} was proposed to
-explicitly extend RC with monotonicity on such stores. \iso{Repeatable
-Read} spec (\C{RR}) extends \C{MAV} with snapshot visibility guarantee
-described before. \iso{Snapshot Isolation} spec (\C{SI}) extends
-\C{RR} with one-way serializability guarantee w.r.t the transactions
+explicitly extend {\sc rc} with monotonicity on such stores. \iso{Repeatable
+Read} spec ({\sc rr}) extends {\sc mav} with the snapshot visibility guarantee
+as described before. \iso{Snapshot Isolation} spec ({\sc si}) extends
+{\sc rr} with a one-way serializability guarantee w.r.t the transactions
 that perform conflicting writes (i.e., writes to the same shared
@@
 not conflict\footnote{Fig.~\ref{fig:ansi-isolation} presents slightly
-weaker versions of \C{SI} and \C{SER} specs in the interest of
+weaker versions of {\sc si} and {\sc ser} specs in the interest of
 clarity.}.
 
-It is worth noting that, unlike some previous proposals
-~\cite{gotsmanconcur15}, our specifications for SI and SER do not
+Note that, unlike recent proposals (\emph{e.g.},
+~\cite{gotsmanconcur15}, our specifications for {\sc si} and {\sc ser} do not
 necessarily impose a total order among transactions (conflicting or
-otherwise). In reality, total order under SER (resp. SI) is guaranteed
-only if the store executes all transactions under SER (resp. SI)
-isolation. Our specifications admit this possibility, and derive total
+otherwise). In reality, a total order under {\sc ser} (resp. {\sc si}) is guaranteed
+only if the store executes all transactions under {\sc ser} (resp. {\sc SI})
+isolation. Our specifications admit this possibility, and derive a total
 order under the assumption of homogenity. However, databases almost
@@
 programs by composing isolation specifications for various
-transactions. For instance, the following trace invariant enforces SI
+transactions. For instance, the following trace invariant enforces {\sc si}
 for both transactions of the program in Fig.~\ref{fig:motiv-eg-1},
@@
 \end{smathpar}
-In contrast, the following trace invariant enforces RC for one and SI
+In contrast, the following trace invariant enforces {\sc rc} for one and {\sc si}
 for another, leading to a possible violation of the postcondition:

### 2016-06-21 ansi-isolation.tex
edits

@@
 
-Note that, unlike recent proposals (\emph{e.g.},
-~\cite{gotsmanconcur15}, our specifications for {\sc si} and {\sc ser} do not
-necessarily impose a total order among transactions (conflicting or
-otherwise). In reality, a total order under {\sc ser} (resp. {\sc si}) is guaranteed
-only if the store executes all transactions under {\sc ser} (resp. {\sc SI})
-isolation. Our specifications admit this possibility, and derive a total
-order under the assumption of homogenity. However, databases almost
-always allow isolation levels to be configured on a per-transaction
-basis, allowing transactions at various isolation levels to coexist.
-Specifications that assume homogenity are incorrect under this
-setting.
+In contrast to recent proposals (\emph{e.g.},
+~\cite{gotsmanconcur15}), our specifications for {\sc si} and {\sc
+  ser} do not necessarily impose a total order among transactions
+(conflicting or otherwise). In reality, a total order under {\sc ser}
+(resp. {\sc si}) is guaranteed only if the store executes all
+transactions under {\sc ser} (resp. {\sc SI}) isolation. Our
+specifications admit this possibility, and derive a total order under
+the assumption of homogenity. However, databases almost always allow
+isolation levels to be configured on a per-transaction basis, allowing
+transactions at various isolation levels to coexist.  Specifications
+that assume homogenity are incorrect under this setting.
 

### 2016-06-21 store-consistency.tex
edits

@@
 Under a trivial trace invariant (i.e., $\I(\E) = true$), the abstract
-machine of Fig.~\ref{fig:txnimp} assumes the semantics of an EC store;
-it allows operations of a transaction witness arbitrary subsets of the
+machine of Fig.~\ref{fig:txnimp} assumes the semantics of an {\sc ec} store;
+it allows operations of a transaction to witness arbitrary subsets of the
 global state. A non-trivial $\I$ composed of isolation specifications
@@
 instance, \iso{Repeatable Read} isolation prohibits operations of a
-transaction from witnessing different states. It however does not
+transaction from witnessing different states. It, however, does not
 prohibit all operations of a transaction from witnessing an aribitrary
-subset of the global state. Consequently, the machine can remain an EC
+subset of the global state. Consequently, the machine can remain an {\sc ec}
 store even while providing non-trivial isolation. How then to model
-the semantics of an SC store, such as a relational database, with
-variable isolation?
+the semantics of an {\sc sc} store, such as a relational database, with
+variable (weak) isolation?
 
@@
 \end{smathpar}
-The trace invariant we wrote in \S\ref{sec:ansi-isolation} for the
+The trace invariant defined in \S\ref{sec:ansi-isolation} for the
 withdraw program of Fig.~\ref{fig:motiv-eg-1} now becomes its $\I_c$
@@
 changes from store to store. We now consider various stores and
-describe their corresponding $\I_s$.
+describe their corresponding store invariants.
 
-\paragraph{An EC store} An EC store provides no additional consistency
+\paragraph{An EC store} An {\sc ec} store provides no additional consistency
 guarantees besides those provided by the machine. Hence, its
@@
 
-\paragraph{An SC store} An SC store guarantees total order of all
+\paragraph{An SC store} An {\sc sc} store guarantees total order of all
 operations w.r.t $\visZ$ that is consistent with their chronological
-order. A straightforward $\I_s$ for this store is the SC guarantee
+order. A straightforward $\I_s$ for this store is the {\sc sc} guarantee
 defined below:
@@
 transactions\footnote{Thus, serializability is a natural
-generalization of SC to transactions.}. In general, when $\I_s$ is in
+generalization of {\sc sc} to transactions.}. In general, when $\I_s$ is in
 conflict (but not inconsistent) with $\I_c$, the only way to enforce
-both is to restrict concurrency. Clearly, this is unacceptable since
+both invariant sets is to restrict concurrency. Clearly, this is unacceptable since
 it defeats the very purpose of weak isolation. 
@@
 Implementations of weak isolation levels on relational databases
-implicitly relax the SC requirement so as to maximize concurrency and
-improve performance. Consequently, execution traces do not necessarily
-satisfy \C{SC}. This approach hints at a general solution reconciling
-$\I_s$ with $\I_c$, which is to weaken the former. However, there are
-many ways $\I_s$ can be weakened; for instance, by setting it to
-$true$, which makes the store EC. In reality though, weak isolation on
-databases does not cause so drastic a reduction in consistency. It is
-therefore possible that there exists a a principled approach to weaken
-$\I_s$ that adequately captures the underlying reality. Such a
-principle can be gleaned by observing that data stores often make
-\emph{recency} commitments~\cite{bailishat}, by which they aim to make
-the most recent data available to clients. SC is infact an extreme
-form of a recency commitment, which guarantees that reads witness most
-recent writes to a data item. In case this conflicts with weak
-isolation guarantees, such as RR which requires all reads in a
-transaction to witness the same state, the natural way to weaken
-consistency while prioritizing recency is to let the reads witness
-most recent state that \emph{does not} violate the isolation
-constraints. In the context of the operational model of
-Fig.~\ref{fig:txnimp}, this translates to making visible to an
-operation the largest subset ($S$) of the global state ($\A$) that
-does not lead to the violation of isolation constraints. Generalizing
-this approach to any $\I_s$ and $\I_c$ gives the \emph{maximum
-visiblity principle}, which requires the weakened consistency
-guarantee ($\I_s'$) of a store to enforce all visibility relationships
-imposed by the actual consistency guarantee ($\I_s$) on a given trace,
-unless enforcing such a relationship violates $\I_c$.  Formally,
-maximum visibility weakening of $\I_s$ is an $\I_s'$, such that:
+implicitly relax the {\sc sc} requirement so as to maximize
+concurrency and improve performance. Consequently, execution traces do
+not necessarily satisfy {\sc sc}. This approach hints at a general
+solution reconciling $\I_s$ with $\I_c$, that entails weakening the
+former. However, there are many ways $\I_s$ can be weakened; for
+instance, by setting it to $true$, which makes the store EC. In
+reality though, weak isolation on databases does not cause so drastic
+a reduction in consistency. It is therefore possible that there exists
+a principled approach to weaken $\I_s$ that adequately captures the
+underlying reality. Such a principle can be gleaned by observing that
+data stores often make \emph{recency} commitments~\cite{bailishat}, by
+which they aim to make the most recent data available to clients; {\sc
+  sc} is, in fact, an extreme form of a recency commitment, which
+guarantees that reads witness the most recent writes to a data
+item. In case this conflicts with weak isolation guarantees, such as
+{\sc rr} which requires all reads in a transaction to witness the same
+state, the natural way to weaken consistency while prioritizing
+recency is to let reads witness the most recent state that \emph{does
+  not} violate the isolation constraints. In the context of the
+operational model of Fig.~\ref{fig:txnimp}, this translates to making
+the largest subset ($S$) of the global state ($\A$) that does not lead
+to the violation of isolation constraints visible to an
+operation. Generalizing this approach to any $\I_s$ and $\I_c$ yields
+a \emph{maximum visiblity principle}, which requires the weakened
+consistency guarantee ($\I_s'$) of a store to enforce all visibility
+relationships imposed by the actual consistency guarantee ($\I_s$) on
+a given trace, unless enforcing such a relationship violates $\I_c$.
+Formally, the maximum visibility weakening of $\I_s$ is a new store
+consistency relation $\I_s'$, such that:
 \begin{itemize}
@@
 
-Applying maximum visibility principle, we weaken \C{SC} to obtain the
-following store trace invariant ($\I_s$) for an SC store whose
+Applying this principle, we can weaken {\sc sc} to obtain the
+following store trace invariant ($\I_s$) for an {\sc sc} store whose
 isolation constraints are captured by $\I_c$:
@@
 global state. A straightforward trace invariant ($\I_s$) for this
-store is the \C{CC} guarantee that requires an effect to witness all
+store is the causally consistent guarantee that requires an effect to witness all
 causally preceding effects regardless of the isolation requirements of
@@
 \end{smathpar}
-However, like $\I_s = \C{SC}$, $\I_s = \C{CC}$ may induce the store to
+However, like $\I_s = \C{SC}$, $\I_s = \C{CC}$ we may induce the store to
 meet isolation requirements by restricting concurrency, thus failing

### 2016-06-21 ansi-isolation.tex
Example added in Sec. 3.1.

@@
 
-Having specified isolation levels as trace well-formedness
-constraints, we can now construct trace invariants ($\I$) for \txnimp
-programs by composing isolation specifications for various
-transactions. For instance, the following trace invariant enforces {\sc si}
-for both transactions of the program in Fig.~\ref{fig:motiv-eg-1},
-allowing it to satisfy its postcondition:
-\begin{smathpar}
-\I \;=\; \lambda\E.~ \underE{\C{SI(Wd1)}} \conj \underE{\C{SI(Wd2)}}
-\end{smathpar}
-In contrast, the following trace invariant enforces {\sc rc} for one and {\sc si}
-for another, leading to a possible violation of the postcondition:
-\begin{smathpar}
-\I \;=\; \lambda\E.~ \underE{\C{RC(Wd1)}} \conj \underE{\C{SI(Wd2)}}
-\end{smathpar}
+% Having specified isolation levels as trace well-formedness
+% constraints, we can now construct trace invariants ($\I$) for
+% \txnimp programs by composing isolation specifications for various
+% transactions. For instance, the following trace invariant enforces
+% {\sc si} for both transactions of the program in
+% Fig.~\ref{fig:motiv-eg-1}, allowing it to satisfy its postcondition:
+% \begin{smathpar}
+% \I \;=\; \lambda\E.~ \underE{\C{SI(Wd1)}} \conj \underE{\C{SI(Wd2)}}
+% \end{smathpar}
+% In contrast, the following trace invariant enforces {\sc rc} for one and {\sc si}
+% for another, leading to a possible violation of the postcondition:
+% \begin{smathpar}
+% \I \;=\; \lambda\E.~ \underE{\C{RC(Wd1)}} \conj \underE{\C{SI(Wd2)}}
+% \end{smathpar}

### 2016-06-21 motivation.tex
Example added in Sec. 3.1.

@@
 \section{Motivation}
+\label{sec:motivation}
 
@@
 operation in $T_1$ (including its commit) happens before $\eta_2$ (we
-write $T_1 \hboar \eta_2$ to denote this). The {\sc rc} specification
+let $T_1 \hboar \eta_2$ denote this). The {\sc rc} specification
 does not require $T_1$ to execute and commit before all operations of
@@
 $T_2$ (or, vise versa).  Formally\footnote{Note that specifications
-  are simplified for illustration. The actual specifications described
-  in (\S~\ref{sec:opsem}) are more nuanced.  }:
+    presented here are solely for illustration. The actual
+    specifications described in \S\ref{sec:ansi-isolation} are more nuanced.}:
 \begin{smathpar}
 \begin{array}{lcl}
-\psi_{RC} & \defeq & \forall \eta_1,\eta_2,T_1,T_2.\; \txn(\eta_1) = T_1 
+\psi_{RC} & \defeq & \forall T_1,T_2,\eta_1,\eta_2,.\; \txn(\eta_1) = T_1 
   \conj \txn(\eta_2) = T_2 \\

### 2016-06-21 opsem.tex
Example added in Sec. 3.1.

@@
 property between effects among different transactions. The intent and
-mechanics of $\visZ$ is described below.
+mechanics of $\visZ$ is described in the sequel.
 
 Fundamental to our development is the notion of a trace invariant
-($\I$), is a function from traces ($\E$) to propositions
-($\texttt{Prop}$) that defines well-formedness constraints over
-traces. The machine takes a step only if the resulting trace satisfies
-the constraints imposed by $\I$. This behaviour is captured by the
+($\I$). $\I$ is a function from traces ($\E$) to propositions
+($\texttt{Prop}$) that define well-formedness constraints over traces.
+The machine takes a step only if the resulting trace satisfies the
+constraints imposed by $\I$. This behaviour is captured by the
 auxiliary reduction rule \rulelabel{E-Aux} that factors out the trace
-extension aspect of the evaluation by abstracting away
-operation-specific behaviour as a function ($\mathcal{F}$) that
-generates an appropriate effect.  \rulelabel{E-Aux} uses $\mathcal{F}$
-to generate a new effect and extend the trace ($\E = (\A,\visZ)$)
-\emph{only if} the well-formedness constraints imposed by $\I$ on $\E$
-(i.e., $\I(\E')$) are satisfied. Otherwise, it gets stuck. In an
-execution that runs to completion, every small-step preserves the
-well-formedness of a trace, thus ensuring the invariance of $\I$.
-Note that the semantics makes no assumptions about $\I$ other than its
-type. As such, it can be instantiated with any trace-parametric
-proposition that expresses constraints over the given trace. While the
-machine takes a step only if the constraints are satisfied, it neither
-defines nor explicitly assumes an oracle to check satisfaction.
+extension aspect of the evaluation by abstracting away the
+operation-specific behaviour as a function that generates an
+appropriate effect. We let $\mathcal{F}$ denote this function.
+\rulelabel{E-Aux} uses $\mathcal{F}$ to generate a new effect and
+extend the trace ($\E = (\A,\visZ)$) \emph{only if} the
+well-formedness constraints imposed by $\I$ on $\E$ (i.e., $\I(\E')$)
+are satisfied. Otherwise, it gets stuck. In an execution that runs to
+completion, every small-step preserves the well-formedness of a trace,
+thus ensuring the invariance of $\I$.  Note that the semantics makes
+no assumptions about $\I$ other than its type. As such, it can be
+instantiated with any trace-parametric proposition that expresses
+constraints over the given trace. For instance, consider the
+$\psi_{RC}$ specification from \S\ref{sec:motivation}, but with
+bounded $T_1$ and $T_2$ instantiated with \C{Wd1} and \C{Wd2},
+respectively. The instantiated specification is the following
+proposition:
+\begin{smathpar}
+\begin{array}{l}
+  \forall \eta_1,\eta_2,.\; \txn(\eta_1) = \C{Wd1}
+  \conj \txn(\eta_2) = \C{Wd2} \\
+  \hspace*{0.6in}\conj \C{Wd1} \neq \C{Wd2} \conj \eta_1 \hboar
+  \eta_2 \Rightarrow \C{Wd1} \hboar \eta_2 \\
+\end{array}
+\end{smathpar}
+It is easy to intepret the above proposition in the context of a trace
+$\E$ that captures an execution of the program in
+Fig.~\ref{fig:motiv-eg-1}. Such a trace-parametric proposition can be
+used to instantiate the trace invariant $\I$ in Fig.~\ref{fig:txnimp}.
+The resultant operational semantics describes an abstract machine that
+gets stuck if an operation of \C{Wd2} is executed in a state that
+incorporates some, but not all the effects (including the \C{COMMIT})
+of \C{Wd1}. 
+% While the machine takes a step only if the constraints are
+% satisfied, it neither defines nor explicitly assumes an oracle to
+% check satisfaction.
 
@@
 concluded operations, which varies between executions due to the
-non-deterministic order of evaluating the parallel composition
-(cf. evaluation contexts ($\ectx$) for parallel composition) of
+non-deterministic order of evaluating the parallel composition of
 transactions. A more notable source of non-determinism, however, is
@@
 required levels of isolation. Notably, the machine achieves this
-without being restricted to any particular isolation level by relying
-on a declarative specification of isolation as trace well-formedness
-constraints.  \S\ref{sec:ansi-isolation} shows the specification of
-various ANSI SQL isolation levels stated as trace well-formedness
-constraints.
+without defining an operation semantics for isolation levels, solely
+by relying on their declarative characterization as trace
+well-formedness constraints.  \S\ref{sec:ansi-isolation} shows the
+specification of various ANSI SQL isolation levels stated as trace
+well-formedness constraints.
 

### 2016-06-22 ansi-isolation.tex
Edits.

@@
 isolation guarantees expressed as constraints over trace
-well-formedness. For brevity and convenience, we adopt some notation.
-An execution trace is destructed into $\A$ and $\visZ$ whenever
-individual components of the pair are needed.  Otherwise, it is written
-as $\E$. When $\psi$ is a proposition, we write $\underE{\psi}$ to denote
-an interpretation of $\psi$ in the context of the trace $\E$. Such
+well-formedness. For brevity and convenience, we adopt few notations
+in this and subsequent sections.  An execution trace is destructed
+into $\A$ and $\visZ$ whenever individual components of the pair are
+needed. Otherwise, it is written as $\E$. Sometimes, the dot notation
+(\eg~$\E.\A$) is also used. Since $\A$ and $\visZ$ are both sets, we
+lift the operations on sets to pairs of sets when updating $\E$. For
+example, $\E' = \E \cup (\{\eta_2\},\{(\eta_1,\eta_2)\})$ expands to
+$\E' = (\E.\A \cup \{\eta_2\},\,\E.\visZ \cup \{(\eta_1,\eta_2)\})$.
+When $\psi$ is a proposition, $\underE{\psi}$ denotes the
+interpretation of $\psi$ in the context of the trace $\E$. Such
 interpretations are defined on a case-by-case basis in
@@
 
-The ANSI SQL 92 standard requires \iso{Read Committted} isolation to avoid
-the dirty reads phenomenon, which is achieved by enforcing
+The ANSI SQL 92 standard requires \iso{Read Committted} isolation to
+avoid the dirty reads phenomenon, which is achieved by enforcing
 $\mathtt{AtomicVis}$ and $\mathtt{CommitVis}$ guarantees. The {\sc rc}
@@
 specification also agrees with the description and
-implementation~\cite{bailishat,pldi15} of {\sc rc} for highly available
-replicated stores. On relational databases, however, {\sc rc} has also come
-to be associated with the $\mathtt{MonotonicVis}$ guarantee.
-Nonetheless, $\mathtt{AtomicVis}$ and $\mathtt{CommitVis}$ are
-sufficient to reason about {\sc rc} isolation on relational stores too. As
-we demonstrate in \S\ref{sec:store-consistency}, the combination of
-these guarantees with the {\sc sc} property of relational stores
+implementation~\cite{bailishat,pldi15} of {\sc rc} for highly
+available replicated stores. On relational databases, however, {\sc
+rc} has also come to be associated with the $\mathtt{MonotonicVis}$
+guarantee.  Nonetheless, $\mathtt{AtomicVis}$ and $\mathtt{CommitVis}$
+are sufficient to reason about {\sc rc} isolation on relational stores
+too. The combination of these guarantees with the {\sc sc} property of
+relational stores (formalized in \S\ref{sec:store-consistency})
 automatically leads to the monotonicity guarantee, which probably
-explains why {\sc rc} comes with $\mathtt{MonotonicVis}$ on such stores. On
-weakly consistent stores however, $\mathtt{AtomicVis}$ and
+explains why {\sc rc} comes with $\mathtt{MonotonicVis}$ on such
+stores. On weakly consistent stores however, $\mathtt{AtomicVis}$ and
 $\mathtt{CommitVis}$ do not imply $\mathtt{MonotonicVis}$. A stronger
@@
 Fig.~\ref{fig:ansi-isolation})~\cite{bailishat,pldi15} was proposed to
-explicitly extend {\sc rc} with monotonicity on such stores. \iso{Repeatable
-Read} spec ({\sc rr}) extends {\sc mav} with the snapshot visibility guarantee
-as described before. \iso{Snapshot Isolation} spec ({\sc si}) extends
-{\sc rr} with a one-way serializability guarantee w.r.t the transactions
-that perform conflicting writes (i.e., writes to the same shared
-variable). Finally, \iso{Serializable} isolation extends one-way
+explicitly extend {\sc rc} with monotonicity on such stores. Sample
+{\sc rc} executions shown in Fig.~\ref{fig:rc-executions} are also
+{\sc mav}. \iso{Repeatable Read} spec ({\sc rr}) extends {\sc mav}
+with the snapshot visibility guarantee as described before.
+\iso{Snapshot Isolation} spec ({\sc si}) extends {\sc rr} with a
+one-way serializability guarantee w.r.t the transactions that perform
+conflicting writes (i.e., writes to the same shared variable).
+Finally, \iso{Serializable} isolation extends one-way
 $\mathtt{SnapshotVis}$ to all transactions, including those that do
@@
 weaker versions of {\sc si} and {\sc ser} specs in the interest of
-clarity.}.
+clarity.}. \GK{ToDo: Figures. Need to figure out a convention to
+uniformly represent various isolation levels. Current idea: concurrent
+timelines with arrows in both directions.}
 

### 2016-06-22 store-consistency.tex
Edits.

@@
 
+It is worthwhile to note that, although conflict-free $\I_s$ shown
+above captures the behaviour of an {\sc SC} database in presence of
+isolation constraints, it does not necessarily describe how the
+database achieves this behaviour (i.e., its implementation). It is
+unusual for databases to check for satisfaction of isolation
+constraints everytime an operation is executed. Efficient
+implementation strategies (\eg multi-versioning) are employed to
+achieve the twin goals of isolation and consistency/recency.
+
 \paragraph{A CC store} A causally consistent data store, such as the

### 2016-06-23 introduction.tex
edits

@@
 particular, do not come for free, however - pessimistic concurrency
-control methods, for example, require databases to provide expensive
-techniques such as two-phase locking that incur overhead to deal with
-deadlocks, rollbacks, and re-execution~\cite{twopl,ullmanbook}.
-Similar criticisms apply to optimistic multi-version concurrency
-control methods that must deal with timestamp and version
-management~\cite{BG81}.  Regardless of the particular concurrency
-control method employed, enforcing serializability in a replicated
-data store additionally requires global coordination among
-geo-distributed replicas, which makes the store \emph{unavailable} in
-the presence of network
+control methods that enforce serializability, for example, require
+databases to provide expensive techniques such as two-phase locking
+that incur overhead to deal with deadlocks, rollbacks, and
+re-execution~\cite{twopl,ullmanbook}.  Similar criticisms apply to
+optimistic multi-version concurrency control methods that must deal
+with timestamp and version management~\cite{BG81}.  Regardless of the
+particular concurrency control method employed, enforcing
+serializability in a replicated data store additionally requires
+global coordination among geo-distributed replicas, which makes the
+store \emph{unavailable} in the presence of network
 partitions~\cite{cap,sernotavlbl,bailishat,bernsigmod13}.
@@
 The tension between serializable transactions, which are easy to
-reason about but difficult to implement, and more pragmatic goals
-driven by performance and availability considerations, has motivated
-the development of different weaker forms of transaction isolation,
-beginning as early as 1976~\cite{gray1976}. The ANSI SQL 92 standard
-defines three such weak isolation levels which are now implemented in
-many relational and NoSQL databases. Not surprisingly, weakly-isolated
-transactions have been found to significantly outperform serializable
-transactions on benchmark suites, both on single-node databases and
-multi-node replicated stores~\cite{dbtuningbook,bailishat,bailisvldb},
-leading to their overwhelming adoption. A 2013
-study~\cite{bailishotos} of 18 popular ACID and ``NewSQL'' databases
-found that only three of them offer serializability by default, and
-half, including Oracle 11g, do not offer it at all.  A 2015
-study~\cite{bailisferal} of a large corpus of database applications
-finds no evidence that applications manifestly change the default
-isolation level offered by the database. Taken together, these studies
-make clear that weakly-isolated transactions are quite prevalent in
-practice, and strongly-isolated (serializable) transactions are often
-eschewed.
+reason about but difficult to implement, and more pragmatic goals but
+weaker variants that are driven by performance and availability
+considerations, has motivated the development of weaker forms of
+transaction isolation, beginning as early as 1976~\cite{gray1976}. The
+ANSI SQL 92 standard defines three such weak isolation levels which
+are now implemented in many relational and NoSQL databases. Not
+surprisingly, weakly-isolated transactions have been found to
+significantly outperform serializable transactions on benchmark
+suites, both on single-node databases and multi-node replicated
+stores~\cite{dbtuningbook,bailishat,bailisvldb}, leading to their
+overwhelming adoption. A 2013 study~\cite{bailishotos} of 18 popular
+ACID and ``NewSQL'' databases found that only three of them offer
+serializability by default, and half, including Oracle 11g, do not
+offer it at all.  A 2015 study~\cite{bailisferal} of a large corpus of
+database applications finds no evidence that applications manifestly
+change the default isolation level offered by the database. Taken
+together, these studies make clear that weakly-isolated transactions
+are quite prevalent in practice, and strongly-isolated (serializable)
+transactions are often eschewed.
 
@@
 
-Our results thus provide the first (to the best of our knowledge)
+\noindent Our results thus provide the first (to the best of our knowledge)
 mechanism that precisely relates high-level program invariants to
@@
 counterparts.
+
+The remainder of the paper is organized as follows.  The next section
+provides motivation and background on serializable and weakly-isolated
+transactions. \S\ref{sec:opsem} presents an operational semantics for
+a core language that supports weakly-isolated transactions, and which
+is parameterized over different isolation and store-consistency
+notions. \S\ref{sec:proofrules} formalizes the proof systemt that we
+use to reason about program invariants, and establishes the soundness
+of these rules with respect to the semantics.  We describe the impact
+of our reasoning framework in the context of several real-world case
+studies in \S\ref{sec:casestudies}.  Related work and conclusions are
+given in \S\ref{sec:relatedwork}.

### 2016-06-23 reasoning.tex

\section{The Reasoning Framework}
\label{sec:reasoning}

We now describe a proof system that lets us prove that a \txnimp
program satisfies its high-level invariants in all small-step
executions that satisfy a trace invariant\footnote{It is important to
mind the distinction between high-level invariants (\eg~\C{X$\ge$0})
and trace invariants (\eg~\C{RC}). The former constitute \emph{proof}
\emph{obligations} for the programmer, whereas the latter capture
\emph{assumptions} about the machine. } ($\I$).  Since the trace
invariant is a conjunction of store consistency ($\I_s$) and
transaction isolation ($\I_c$) constraints, the proof demonstrates
that program preserves its invariants when executed on a given store
under a given selection of isolation levels for its transactions.

Our proof system generalizes rely-guarantee reasoning to a setting
that includes weakly consistent stores and weakly isolated
transactions.

\input{rg-rules}

### 2016-06-23 motivation.tex
Edits.
edits

@@
   a read committed isolation level.}
+\label{fig:rc-executions}
 \end{figure}
@@
 appropriate for both {\bf Wd} transactions.
+
+\begin{figure}
+\begin{smathpar}
+\begin{array}{lcl}
+\psi_{RC} & \defeq & \forall T_1,T_2,\eta_1,\eta_2,.\; \txn(\eta_1) = T_1 
+  \conj \txn(\eta_2) = T_2 \\
+  & & \hspace*{0.6in}\conj T_1 \neq T_2 \conj \eta_1 \hboar
+  \eta_2 \Rightarrow T_1 \hboar \eta_2 \\
+\psi_{SI} & \defeq & \forall T_1,T_2.\; T_1 \neq T_2 \conj
+  (\exists \C{X}.~{T_1 \wrstoar \C{X}} \conj 
+                      {T_2 \wrstoar \C{X}})\\
+  &  & \hspace*{0.6in}\Rightarrow{T_1 \hboar T_2} \disj {T_2 \hboar T_1} \\
+\end{array}
+\end{smathpar}
+\caption{Visibility and interference properties for different weak isolation levels can be
+  captured by different instantiations of their happens-before relation.}
+\label{fig:interference-ex}
+\end{figure}
 
@@
 prohibits interference due to actions in $T_1$ being interleaved in
-$T_2$ (or, vise versa).  Formally\footnote{Note that specifications
+$T_2$ (or, vise versa).  We formalize these properties in Fig.~\ref{fig:interference-ex}\footnote{Note that specifications
     presented here are solely for illustration. The actual
-    specifications described in \S\ref{sec:ansi-isolation} are more nuanced.}:
-\begin{smathpar}
-\begin{array}{lcl}
-\psi_{RC} & \defeq & \forall T_1,T_2,\eta_1,\eta_2,.\; \txn(\eta_1) = T_1 
-  \conj \txn(\eta_2) = T_2 \\
-  & & \hspace*{0.6in}\conj T_1 \neq T_2 \conj \eta_1 \hboar
-  \eta_2 \Rightarrow T_1 \hboar \eta_2 \\
-\psi_{SI} & \defeq & \forall T_1,T_2.\; T_1 \neq T_2 \conj
-  (\exists \C{X}.~{T_1 \wrstoar \C{X}} \conj 
-                      {T_2 \wrstoar \C{X}})\\
-  &  & \hspace*{0.6in}\Rightarrow{T_1 \hboar T_2} \disj {T_2 \hboar T_1} \\
-\end{array}
-\end{smathpar}
+    specifications described in \S\ref{sec:ansi-isolation} are more nuanced.}.
+
 Fig.~\ref{fig:motiv-eg-1-hb} is a visualization of $\hbZ$ 
@@
 ``committed''):
+
 \begin{smathpar}
@@
 \end{smathpar}
-$\phi$ asserts that {\bf Wd1} is not yet committed, and that it wrote to
+
+\noindent $\phi$ asserts that {\bf Wd1} is not yet committed, and that it wrote to
 \C{C}, and the value of \C{C} is either \C{k-a1-a2} or \C{k-a1}

### 2016-06-23 opsem.tex
edits

@@
+\subsection{\txnimp: Syntax and Semantics}
 \section{\txnimp}
@@
 
+\label{sec:syntax}
+
 \input{txnimp}
-
-\subsection{Syntax and Semantics}
-\label{sec:syntax}
 
@@
 proposition:
+
 \begin{smathpar}
@@
 \end{smathpar}
-It is easy to intepret the above proposition in the context of a trace
+
+\noindent It is easy to intepret the above proposition in the context of a trace
 $\E$ that captures an execution of the program in
@@
 shown below:
+
 \begin{smathpar}
@@
 \end{smathpar}
-Rules \rulelabel{E-Top-Ctx} and \rulelabel{E-Txn-Ctx} define
+
+\noindent Rules \rulelabel{E-Top-Ctx} and \rulelabel{E-Txn-Ctx} define
 congruence properties for top-level terms and transaction-bound terms,

### 2016-06-23 reasoning.tex
Secs. 4.1 and 4.2 done.

@@
 
-We now describe a proof system that lets us prove that a \txnimp
+We now describe a proof system that lets us demonstrate that a \txnimp
 program satisfies its high-level invariants in all small-step
-executions that satisfy a trace invariant\footnote{It is important to
-mind the distinction between high-level invariants (\eg~\C{X$\ge$0})
-and trace invariants (\eg~\C{RC}). The former constitute \emph{proof}
-\emph{obligations} for the programmer, whereas the latter capture
-\emph{assumptions} about the machine. } ($\I$).  Since the trace
-invariant is a conjunction of store consistency ($\I_s$) and
-transaction isolation ($\I_c$) constraints, the proof demonstrates
-that program preserves its invariants when executed on a given store
-under a given selection of isolation levels for its transactions.
+executions that satisfy a chosen trace invariant\footnote{It is
+important to mind the distinction between high-level invariants
+(\eg~\C{X$\ge$0}) and trace invariants (\eg~\C{RC}). The former
+constitute \emph{proof} \emph{obligations} for the programmer, whereas
+the latter capture \emph{assumptions} about the machine. } ($\I$).
+Since the trace invariant is a conjunction of store consistency
+($\I_s$) and transaction isolation ($\I_c$) constraints, the
+demonstration is a proof that the program preserves its invariants
+when executed on a chosen store under a selection of isolation levels
+for its transactions.
 
-Our proof system generalizes rely-guarantee reasoning to a setting
-that includes weakly consistent stores and weakly isolated
-transactions.
+As mentioned earlier, our proof system is a rely-guarantee framework
+that admits compositional reasoning by abstracting away the
+environmental interference as a rely relation. A conspicuous
+difference between a standard development of rely-guarantee and ours
+is that, while the former reasons in terms of program states (variable
+to value bindings), we reason in terms of executions as captured by
+their traces ($\E$). In particular, our rely ($R$) and guarantee ($G$)
+relations relate executions (i.e., $R,\,Q\,\subseteq\,\E\times\E$),
+and pre ($P$) and post ($Q$) conditions are assertions over executions
+(i.e., $P,\,Q\, : \E \rightarrow \Prop$). Our development nonetheless
+facilitates state-based reasoning via the interpretation function
+($\interp{\cdot}$) introduced in \S\ref{sec:syntax}. For example, if a
+\emph{bi-state} rely relation relates every pair of states $\sigma$
+and $\sigma'$ such that $\sigma'(X) \ge \sigma(X) \ge 0$, the
+corresponding \emph{bi-execution} rely relation relates every pair of
+executions $\E$ and $\E'$ such that $\interp{\E}(X) \ge
+\interp{\E'}(X) \ge 0$.  Assertions on state are also written
+similarly. For instance, consider the postcondition
+$\C{C\,=\,k-a1-a2}$ of the program in Fig.~\ref{fig:motiv-eg-1}. The
+corresponding assertion on the post-state
+($\lambda\sigma.~\sigma(\C{C}) \,=\, \C{k-a1-a2}$) asserts that in all
+states resulting from executing the program, the value of \C{C} is
+\C{k-a1-a2}. The equivalent execution-based assertion
+($\lambda\E.~\interp{\E}(\C{C})\,=\,\C{k-a1-a2}$) asserts that in all
+executions of the program, the value of \C{C} is \C{k-a1-a2}.
+However, having access to an execution facilitates assertions that go
+beyond the state. An example is the invariant at a program point
+inside \C{Wd1} described abstractly in \S\ref{sec:motivation}, and
+reified as an execution-based assertion below:
+\begin{smathpar}
+\begin{array}{l}
+  \lambda\E.~ \neg(\underE{\committed(\C{Wd1}}) \conj 
+      \underE{\C{Wd1} \wrstoar \C{C}}\\
+  \hspace*{0.2in}\conj(\neg(\underE{\committed(\C{Wd2})}) \Rightarrow 
+      \C{\interp{\E}(C) = k-a1}) \\
+   \hspace*{0.2in}\conj (\underE{\committed(\C{Wd2})} \Rightarrow 
+      \C{\interp{\E}(C) = k-a1-a2})
+\end{array}
+\end{smathpar}
+% Where, $\committed$ is defined as:
+% \begin{smathpar}
+% \begin{array}{lcl}
+% \underE{\committed(T_i)} & \Leftrightarrow &
+%   \exists\eta.~\underE{\eta\in T_i} \conj \kind(\eta) = \C{COMMIT}\\
+% \end{array}
+% \end{smathpar}
+One of the conjuncts asserts that \C{Wd1} has written to \C{C} - a
+fact which cannot be deduced solely from the value of \C{C} (esp. in
+presence of interference), but can be expressed as a proposition over
+the execution trace.
+
+\subsection{The Rely-Guarantee Judgment}
 
 \input{rg-rules}
+
+The standard rely guarantee judgment is the quintuple $\rg{P,R}{c}{G,Q}$,
+which intuitively means the following: if a command $c$ is executed in
+a state that satisfies its precondition $P$, provided that every
+interference step during the execution is contained inside the rely
+relation $R$, then the effect that each step of executing $c$ has on
+the state is captured by $G$, and the final state of execution
+satisfies the postcondition $Q$. The terms ``execution'' and ``step''
+are defined with respect to an operational semantics, which in our
+case, is parameterized on the trace invariant $\I$. Consequently, our
+rely-guarantee judgment for commands is the sextuple\footnote{$\I$ is
+written in the same group as $P$ to suggest that it is an assumption;
+not an obligation.} $\rg{\I,P,R}{c}{G,Q}$, whose semantics differs
+from the quintuple in that it requires every step of the execution to
+preserve the trace invariant $\I$ for the postcondition to be valid.
+As usual, we write $\txnbox{t}_i$ to denote a term $t$ inside a
+transaction $T_i$.  \txnimp expressions are side-effecting (they
+generate \C{RD} effects), and admit interference during their
+evaluation. Therefore, like the commands, expressions need to be
+reasoned about explicitly. But, unlike commands, expressions evaluate
+to values, and reasoning framework should also admit assertions on such
+values. We therefore define a separate judgment for expressions - the
+septuple $\rg{\I,P,R}{e}{G,C,Q}$. The new entrant $C : \mathbb{N}
+\rightarrow \Prop$ is an assertion on the return value of $e$.
+
+The rules that define rely-guarantee judgment are shown in
+Fig.~\ref{fig:rg-rules}. Like the standard definitions of rely
+guarantee, our definition also relies on the \emph{stability}
+condition, which requires pre and post conditions to hold despite the
+interference from concurrent threads (captured by $R$). The stability
+of pre and post conditions can be predicated on the assumption that
+the interference preserves the trace invariant $\I$. Formally:
+\begin{smathpar}
+\begin{array}{lcl}
+  \underI{\stable(R,P)} & \defeq & \forall \E, \E'.\, 
+  \I(\E) \conj P(\E) \conj R(\E,\E') \\
+  &   & \hspace*{1in}\conj \I(\E') \Rightarrow P(\E')\\
+\end{array}
+\end{smathpar}
+However, the assumption that the interference preserves $\I$ (or,
+equivalently, $\I$ withstands inteferference) needs to
+be justified separately. We call this the stability requirement on
+$\I$:
+\begin{smathpar}
+\begin{array}{lcl}
+\stable(R,\I)& \Leftrightarrow & \forall \E, \E'.\, 
+  \I(\E) \conj R(\E,\E')\Rightarrow \I(\E')\\
+\end{array}
+\end{smathpar}
+
+The rule \rulelabel{RG-Var} defines the rely-guarantee judgment for
+shared variable reads inside a transaction $T_i$. It requires $\I$ to
+be stable, and pre and post conditions to be stable relative to $\I$.
+The quantified premise effectively requires a proof that if the
+abstract machine of Fig.~\ref{fig:txnimp} takes a step starting from
+an execution $\E$ that satisfies the precondition $P$, then the
+resultant execution $\E'$ satisfies the postcondition $Q$, and that
+the guarantee $G$ faithfully captures the transition from $\E$ to
+$\E'$. It is informative to compare this premise with the premise of
+the \rulelabel{E-Aux} reduction rule of Fig.~\ref{fig:txnimp}. Similar
+premises also appear in \rulelabel{RG-Asgn} and \rulelabel{RG-Txn}
+rules, which define rely-guarantee judgments for assignments and
+transactions, respectively. \rulelabel{RG-Var} however also requires
+the return value ($\interp{S}(X)$) of the read to satisfy the
+assertion $C$ meant for the value. \rulelabel{RG-Arith} defines the
+RG judgment for an arithmetic expression $e_1\pm e_2$ in terms of the
+corresponding judgments for the constitutent expressions $e_1$ and
+$e_2$. Note that the pre-condition of $e_2$ ($Q'$) is also the
+postcondition of $e_1$, reflecting the order of evaluation established
+in Fig.~\ref{fig:txnimp}. The quantified premise requires any value
+resulting from evaluating $e_1 \pm e_2$ to satisfy the assertion $C$,
+provided that $e_1$ and $e_2$ always evaluate to values that satisfy
+$C_1$ and $C_2$, respectively.
+
+RG judgment for the assignment $X:=e$ makes use of the corresponding
+judgment for the RHS expression $e$. The quantified premise asserts
+that evaluating the assigment after evaluating $e$ to a value $v$ and
+an execution $\E$, should result in an execution $\E'$ that satisfies
+the postcondition $Q$, while being related to $\E$ via the guarantee
+$G$. RG judgment for the \C{txn} lexical block is similar.  It uses
+the judgment for the transaction-bound command $c$ (i.e.,
+$\txnbox{c}_i$) to obtain an invariant $Q'$ for the execution $\E$
+before the commit, and verifies that committing the transaction under
+$\E$ results in an execution $\E'$ that satisfies transaction's
+postcondition $Q$. As usual, $\E$ and $\E'$ need to be related by $G$.
+The rules for sequential composition of transaction-bound comands
+(\rulelabel{RG-Seq}) and parallel composition of top-level commands
+(\rulelabel{RG-Par}) are straightforward, and more-or-less same as the
+corresponding rules in classical rely-guarantee.
+
+The three \rulelabel{RG-Conseq} rules define ways to strengthen or
+weaken relations and assertions associated with RG judgments of
+transaction-bound expressions, commands and top-level commands,
+respectively. In particular, they allow the precondition $P$ and the
+rely relation $R$ to be strengthened\footnote{A stronger relation is a
+smaller set than a weaker relation, hence a stronger $R$ admits lesser
+interference than a weaker $R$}, and the postcondition $Q$ (also, $C$
+in the case of expressions) and the guarantee relation $G$ to be
+weakened. This kind of strengthening/weakening is also admitted in the
+classical RG framework. The most notable aspect of
+\rulelabel{RG-Conseq} rules however is that they allow the trace
+invariant $\I$ to be strengthened. Considering that $\I$ captures
+consistency and isolation properties, this means that a program proved
+correct under weaker consistency and isolation levels is also correct
+under stronger levels.  Parametricity over the trace invariant $\I$,
+combined with the ability to strengthen $\I$ as needed, allows our
+proof system to admit a highly flexible proof strategy to prove
+programs correct on various stores. Unlike the existing approaches,
+which require programs to be proved correct with respect to a
+pre-defined store (\eg~A CC store in~\cite{gotsmanpopl16}), our
+reasoning framework allows programmers to \emph{define} a store of
+their choice (by defining $\I$ appropriately) and then prove programs
+correct with respect to this store. Thus, programmers can prove their
+programs correct with respect to an imaginary store that provides only
+the required consistency and isolation guarantees. Soundness of
+strengthening $\I$ guarantees that the program can now be executed
+safely on any real store (\eg~a relational database, or a NoSQL store)
+that offers same or stronger consistency and isolation guarantees as
+the imaginary store. The following subsection proves the soundness of
+our proof system.
+
+\subsection{Semantics and Soundness}
+
+We now formalize the semantics of the rely-guarantee judgment of
+Fig.~\ref{fig:rg-rules} through a series of definitions.
+
+\begin{definition}[Step-indexed reflexive transitive closure]
+Forall $A$, $R: A \rightarrow A \rightarrow \mathbb{P}$, and $n :
+\mathbb{N}$, step-indexed reflexive transitive closure $R^n$ of $R$ is
+the smallest relation satisfying the following
+properties:
+\begin{itemize}
+\item $\forall (x:A).\, R^0 (x,x)$
+\item $\forall (n:\mathbb{N})(x,y,z : A).\, R(x,y) \conj R^n(y,z) \Rightarrow
+R^{n+1}(x,z)$
+\end{itemize}
+\end{definition}
+
+\begin{definition}[Interleaved step and multi-step relations]
+Interleaved step relation interleaves the thread-local reduction with
+the interference from concurrent threads captured as the rely relation
+($R$).  It is defined as following:
+\begin{smathpar}
+\begin{array}{lcl}
+\I \vdash (c,\E) \rstepsto (c',\E') & \defeq & \I \vdash 
+  (c,\E) \stepsto (c',\E') \\
+  &   & \disj (c' = c \conj R(\E, \E') \conj \I(\E'))\\
+\end{array}
+\end{smathpar}
+The interleaved step relation for transaction bound expressions
+($\txnbox{e}_i$) and commands ($\txnbox{c}_i$) is defined similarly.
+Interleaved multi-step relation ($\stepssto{n}$) is the step-indexed
+reflexive transitive closure of the interleaved step relation.
+\end{definition}
+
+\begin{definition}[Semantics of the RG judgment]
+Semantics of the RG sextuple $\rg{\mathbb{I},P,R}{c}{G,Q}$ is defined
+in terms of the interleaved step relation as following:
+\begin{smathpar}
+\begin{array}{l}
+\hspace*{-0.3in}
+\rg{\mathbb{I},P,R}{c}{G,Q} \;\defeq\; \forall \E.\, P(\E)
+  \wedge \mathbb{I}(\E) \\
+\hspace*{0.4in}\Rightarrow (\forall n,\E'.\; \I \vdash (c,\E) 
+    \rstepssto{n} (\cskip,\E') \Rightarrow Q(\E')) \\
+\hspace*{0.5in}\conj \texttt{step-guaranteed}(\I,R,G,c,\E)\\
+\end{array}
+\end{smathpar}
+The first conjunct in the consequent is called the \emph{Hoare
+consequent} since it ascribes Hoare triple semantics to RG sextuple.
+The second conjunct, called the \emph{guarantee consequent}, uses the
+$\texttt{step-guaranteed}$ predicate defined below:
+\begin{smathpar}
+\begin{array}{l}
+\texttt{step-guaranteed}(\I,R,G,c,\E) \;\defeq\; \forall n,\E',c'',\E''.\\
+\hspace*{0.2in}\I \vdash (c,\E) \rstepssto{n} (c',\E') \conj \I \vdash (c',\E') \stepsto
+  (c'',\E'') \Rightarrow G(\E',\E'')\\
+\end{array}
+\end{smathpar}
+The guarantee consequent requires $G$ to capture the trace effect of
+every small-step reduction of $c$, where reduction can be interleaved
+by the interference ($R$) from concurrent threads. Semantics of the RG
+sextuple for transaction-bound commands ($\txnbox{c}_i$) is defined
+similarly. Expressions, unlike commands, evaluate to a value $v$, and
+the semantics of their RG septuple
+($\rg{\I,P,R}{\txnbox{e}_i}{G,C,Q}$) differs slightly in that its
+Hoare consequent requires the value $v$ to satisfy the assertion $C$. 
+\end{definition}
+
+It is worthwhile to note that the semantics of all RG judgments,
+including the judgments for transaction-bound terms,  make similar
+demands of the guarantee relation. Considering that transactions are
+atomic (though not isolated), at a first glance, it might seem odd
+that a transaction's guarantee is required to make explicit every step
+of its reduction. This requirement is however justifed because, in
+reality, a transaction's atomicity is predicated on the isolation
+settings of the observer. A \iso{Read Uncommitted} transaction, for
+example, can observe the internal state of a transaction, regardless
+of the latter's preferences\footnote{Thus, there is no such thing as
+atomicity; it's all about isolation.}. In the interest of modular
+verification, the transaction must therefore make its internal state
+available via its guarantee relation.
+
+Soundness of rely-guarantee judgments can now be stated
+thus\footnote{Supplemntary contains a rigorous proof of the soundness
+theorem, and all relevant lemmas.}:
+\begin{theorem}[Soundness] 
+Rely-guarantee judgments defined by the rules in
+Fig.~\ref{fig:rg-rules} are sound with respect to their semantics
+defined above. 
+\end{theorem}
+\noindent In particular, if $\rg{\I,P,R}{c}{G,Q}$ can be derived using
+the rules of Fig.~\ref{fig:rg-rules}, then (a). every interleaved
+multi-step reduction of $c$ starting from a trace that satisfies $P$
+and $\I$, results in a trace that satisfies $Q$, and (b). the effect
+that every small-step reduction of $c$ along the way has on the trace
+is contained in $G$. Soundness of RG judgment for transaction-bound
+commands ($\txnbox{c}_i$) is stated similarly.  For expressions,
+soundness of the judgment $\rg{\I,P,R}{\txnbox{e}_i}{G,C,Q}$ also
+proves that $e$ is always evaluated to a value that satisfies $C$.
+
+\subsection{Soundness under Eventually Consistent Replication}
+
+\GK{Here, I give intuitive explanation of soundness in the context of
+EC replication. This answers Gustavo's question of why Gotsman-like
+special case reasoning about replicas is not needed in our setting.}
+
+\subsection{Examples}
+
+\GK{Here, I plan to include the full example that we dropped from
+\S\ref{sec:motivation}.}

### 2016-06-23 store-consistency.tex

--- 2016-06-21 store-consistency.tex
@@
 
-\paragraph{An EC store} An {\sc ec} store provides no additional consistency
-guarantees besides those provided by the machine. Hence, its
-$\I_s(\E)$ is $true$ for any $\E$.
+\paragraph{An EC store} An eventually-consistent {\sc ec} store provides no
+additional consistency guarantees besides those provided by the operational
+semantics.  Hence, its trace invariant, $\I_s(\E)$, for any execution
+$\E$ is always \emph{true}.
 
-\paragraph{An SC store} An {\sc sc} store guarantees total order of all
-operations w.r.t $\visZ$ that is consistent with their chronological
-order. A straightforward $\I_s$ for this store is the {\sc sc} guarantee
-defined below:
+\paragraph{An SC store} A strongly-consistent {\sc sc} store guarantees a total order of
+all operations w.r.t $\visZ$ consistent with their chronological
+order. A straightforward definition of $\I_s$ for this store is:
 \begin{smathpar}
@@
 consider a case where $\I_c(\E) \;=\; \forall
-T_i.~\underE{\C{RC}(T_i)}$. That is, all transactions execute under
+T_i.~\underE{\C{RC}(T_i)}$, a constraint that dictates all transactions execute under
 \iso{Read Committed} isolation. Imagine a sample execution where
@@
 both invariant sets is to restrict concurrency. Clearly, this is unacceptable since
-it defeats the very purpose of weak isolation. 
+it defeats the very purpose of supporting weak isolation. 
 % How then do we enforce weak isolation on a strongly consistent
@@
 not necessarily satisfy {\sc sc}. This approach hints at a general
-solution reconciling $\I_s$ with $\I_c$, that entails weakening the
-former. However, there are many ways $\I_s$ can be weakened; for
-instance, by setting it to $true$, which makes the store EC. In
-reality though, weak isolation on databases does not cause so drastic
-a reduction in consistency. It is therefore possible that there exists
-a principled approach to weaken $\I_s$ that adequately captures the
-underlying reality. Such a principle can be gleaned by observing that
-data stores often make \emph{recency} commitments~\cite{bailishat}, by
-which they aim to make the most recent data available to clients; {\sc
-  sc} is, in fact, an extreme form of a recency commitment, which
-guarantees that reads witness the most recent writes to a data
-item. In case this conflicts with weak isolation guarantees, such as
-{\sc rr} which requires all reads in a transaction to witness the same
-state, the natural way to weaken consistency while prioritizing
-recency is to let reads witness the most recent state that \emph{does
-  not} violate the isolation constraints. In the context of the
-operational model of Fig.~\ref{fig:txnimp}, this translates to making
-the largest subset ($S$) of the global state ($\A$) that does not lead
-to the violation of isolation constraints visible to an
-operation. Generalizing this approach to any $\I_s$ and $\I_c$ yields
-a \emph{maximum visiblity principle}, which requires the weakened
+solution reconciling $\I_s$ with $\I_c$ that entails weakening the
+former sufficiently to satisfy the latter. However, there are many
+ways $\I_s$ can be weakened; for instance, by setting it to \emph{true}
+which makes the store EC. In reality though, weak isolation on
+databases does not cause so drastic a reduction in consistency. It is
+therefore possible that there exists a principled approach to weaken
+$\I_s$ that adequately captures the underlying properties provided by the store. Such
+a principle can be gleaned by observing that data stores often make
+\emph{recency} commitments~\cite{bailishat}, by which they aim to make
+the most recent data available to clients; {\sc sc} is, in fact, an
+extreme form of a recency commitment, which guarantees that reads
+\emph{always} witness the most recent writes to a data item. In case this conflicts
+with weak isolation guarantees, such as {\sc rr} which only requires that
+reads in a transaction witness the same state, the natural way to
+weaken consistency while prioritizing recency is to let reads witness
+the most recent state that \emph{does not} violate isolation
+constraints. In the context of the operational model of
+Fig.~\ref{fig:txnimp}, this translates to making the largest subset
+($S$) of the global state ($\A$) that does not lead to the violation
+of {\sc rr} isolation constraints visible to an operation.
+
+Generalizing this intuition to any $\I_s$ and $\I_c$ yields a
+\emph{maximum visiblity principle}, which requires the weakened
 consistency guarantee ($\I_s'$) of a store to enforce all visibility
@@
 isolation constraints are captured by $\I_c$:
+
 \begin{smathpar}
@@
 \end{smathpar}
-As usual $\I = \I_s \wedge \I_c$. The resultant operational model
+
+\noindent As usual $\I = \I_s \wedge \I_c$. The resultant operational model
 hides an existing effect ($\eta\in\A$) from being visible to the
@@
 
-\paragraph{A CC store} A causally consistent data store, such as the
-one underlying the reasoning framework of~\cite{gotsmanpopl16}, allows
-operations to only witness a causally consistent snapshot of the
+It is worthwhile to note that although the conflict-free definition of
+$\I_s$ shown above captures the behaviour of an {\sc sc} database in
+the presence of isolation constraints, it does not necessarily describe
+how the database achieves this behaviour (i.e., its
+implementation). It is unusual for databases to check for satisfaction
+of isolation constraints every time an operation is executed. Efficient
+implementation strategies (\eg multi-versioning) are employed to
+achieve the twin goals of isolation and consistency/recency. \footnote{\SJ{This
+  paragraph reads poorly - not quite sure what the reader should conclude
+  from this.  A negative interpretation is that we can formalize the
+  relationship between $I_s$ and $I_c$ but the formalization may be
+  far removed from reality and practicality.}}
+
+\paragraph{A CC store} A causally consistent data store~\cite{gotsmanpopl16,LBC16}
+allows operations to only witness a causally consistent snapshot of the
 global state. A straightforward trace invariant ($\I_s$) for this
@@
 \end{smathpar}
-However, like $\I_s = \C{SC}$, $\I_s = \C{CC}$ we may induce the store to
-meet isolation requirements by restricting concurrency, thus failing
-to abide by the spirit of weak isolation. In order to admit weak
-isolation without restricting concurrency, we weaken $\I_s$ while
+In order to admit weak isolation behavior without restricting concurrency, we weaken $\I_s$ while
 being guided by the maximum visibility principle. The weakened $\I_s$
 is shown below:
+
 \begin{smathpar}
@@
 \end{smathpar}
-Instantiating the parameter $\I$  with $\I_s \wedge \I_c$ in
+
+\noindent Instantiating the parameter $\I$  with $\I_s \wedge \I_c$ in
 Fig.~\ref{fig:txnimp} results in an operational semantics that admits
-the violation of causal consistency if and only if the violation is
-inevitable to enforce
-$\I_c$.
+violation of causal consistency if and only if the violation is
+inevitable to enforce $\I_c$.

### 2016-06-24 reasoning.tex
No more bashing [17].

@@
 proof system to admit a highly flexible proof strategy to prove
-programs correct on various stores. Unlike the existing approaches,
-which require programs to be proved correct with respect to a
-pre-defined store (\eg~A CC store in~\cite{gotsmanpopl16}), our
-reasoning framework allows programmers to \emph{define} a store of
-their choice (by defining $\I$ appropriately) and then prove programs
-correct with respect to this store. Thus, programmers can prove their
-programs correct with respect to an imaginary store that provides only
-the required consistency and isolation guarantees. Soundness of
-strengthening $\I$ guarantees that the program can now be executed
-safely on any real store (\eg~a relational database, or a NoSQL store)
-that offers same or stronger consistency and isolation guarantees as
-the imaginary store. The following subsection proves the soundness of
-our proof system.
+programs correct on various stores. Instead of requiring programs to
+be proved correct with respect to a pre-defined store (\eg~A CC store
+like in~\cite{gotsmanpopl16}), our reasoning framework allows
+programmers to \emph{define} a store of their choice (by defining $\I$
+appropriately) and then prove programs correct with respect to this
+store. Thus, programmers can prove their programs correct with respect
+to an imaginary store that provides only the required consistency and
+isolation guarantees. Soundness of strengthening $\I$ guarantees that
+the program can now be executed safely on any real store (\eg~a
+relational database, or a NoSQL store) that offers same or stronger
+consistency and isolation guarantees as the imaginary store. The
+following subsection proves the soundness of our proof system.
 
@@
 
-\GK{Here, I give intuitive explanation of soundness in the context of
-EC replication. This answers Gustavo's question of why Gotsman-like
-special case reasoning about replicas is not needed in our setting.}
-
 \subsection{Examples}

### 2016-06-24 reasoning.tex
RG section made succinct.

@@
 However, having access to an execution facilitates assertions that go
-beyond the state. An example is the invariant at a program point
-inside \C{Wd1} described abstractly in \S\ref{sec:motivation}, and
-reified as an execution-based assertion below:
+beyond the state. An example is an invariant of \C{Wd1} described
+abstractly in \S\ref{sec:motivation}, and reified as an
+execution-based assertion below:
 \begin{smathpar}
@@
 
-The standard rely guarantee judgment is the quintuple $\rg{P,R}{c}{G,Q}$,
-which intuitively means the following: if a command $c$ is executed in
-a state that satisfies its precondition $P$, provided that every
-interference step during the execution is contained inside the rely
-relation $R$, then the effect that each step of executing $c$ has on
-the state is captured by $G$, and the final state of execution
-satisfies the postcondition $Q$. The terms ``execution'' and ``step''
-are defined with respect to an operational semantics, which in our
-case, is parameterized on the trace invariant $\I$. Consequently, our
-rely-guarantee judgment for commands is the sextuple\footnote{$\I$ is
-written in the same group as $P$ to suggest that it is an assumption;
-not an obligation.} $\rg{\I,P,R}{c}{G,Q}$, whose semantics differs
-from the quintuple in that it requires every step of the execution to
-preserve the trace invariant $\I$ for the postcondition to be valid.
-As usual, we write $\txnbox{t}_i$ to denote a term $t$ inside a
-transaction $T_i$.  \txnimp expressions are side-effecting (they
-generate \C{RD} effects), and admit interference during their
-evaluation. Therefore, like the commands, expressions need to be
-reasoned about explicitly. But, unlike commands, expressions evaluate
-to values, and reasoning framework should also admit assertions on such
-values. We therefore define a separate judgment for expressions - the
-septuple $\rg{\I,P,R}{e}{G,C,Q}$. The new entrant $C : \mathbb{N}
-\rightarrow \Prop$ is an assertion on the return value of $e$.
+The standard rely guarantee judgment is the quintuple
+$\rg{P,R}{c}{G,Q}$, which intuitively means the following: if a
+command $c$ is executed in a state that satisfies its precondition
+$P$, provided that every interference step during the execution is
+contained inside the rely relation $R$, then the effect that each step
+of executing $c$ has on the state is captured by $G$, and the final
+state of execution satisfies the postcondition $Q$. The terms
+``execution'' and ``step'' are defined with respect to an operational
+semantics, which, in our case, is parameterized on the trace invariant
+$\I$. Consequently, our rely-guarantee judgment for commands is the
+sextuple $\rg{\I,P,R}{c}{G,Q}$, whose semantics differs from the
+quintuple in that it requires every step of the execution to preserve
+the trace invariant $\I$ for the postcondition to be valid.  As usual,
+if $c$ is inside a transaction $T_i$, we write $\txnbox{c}_i$. \txnimp
+expressions are side-effecting (they generate \C{RD} effects), and
+admit interference during their evaluation.  Therefore, like the
+commands, expressions need to be reasoned about explicitly. But,
+unlike commands, expressions evaluate to values, and the reasoning
+framework should also admit assertions on such values.  We therefore
+define a separate judgment for expressions - the septuple
+$\rg{\I,P,R}{e}{G,C,Q}$. The new entrant $C : \mathbb{N} \rightarrow
+\Prop$ is an assertion on the return value of $e$.
 
@@
 However, the assumption that the interference preserves $\I$ (or,
-equivalently, $\I$ withstands inteferference) needs to
-be justified separately. We call this the stability requirement on
-$\I$:
+dually, $\I$ withstands inteferference) needs to be justified
+separately. We call this the stability requirement on $\I$:
 \begin{smathpar}
@@
 corresponding judgments for the constitutent expressions $e_1$ and
-$e_2$. Note that the pre-condition of $e_2$ ($Q'$) is also the
-postcondition of $e_1$, reflecting the order of evaluation established
-in Fig.~\ref{fig:txnimp}. The quantified premise requires any value
-resulting from evaluating $e_1 \pm e_2$ to satisfy the assertion $C$,
-provided that $e_1$ and $e_2$ always evaluate to values that satisfy
-$C_1$ and $C_2$, respectively.
+$e_2$. The quantified premise requires any value resulting from
+evaluating $e_1 \pm e_2$ to satisfy the assertion $C$, provided that
+$e_1$ and $e_2$ always evaluate to values that satisfy $C_1$ and
+$C_2$, respectively. The rules for sequential and parallel composition
+of comands are more-or-less same as the corresponding rules in
+classical rely-guarantee, hence elided.
 
-RG judgment for the assignment $X:=e$ makes use of the corresponding
-judgment for the RHS expression $e$. The quantified premise asserts
-that evaluating the assigment after evaluating $e$ to a value $v$ and
-an execution $\E$, should result in an execution $\E'$ that satisfies
-the postcondition $Q$, while being related to $\E$ via the guarantee
-$G$. RG judgment for the \C{txn} lexical block is similar.  It uses
-the judgment for the transaction-bound command $c$ (i.e.,
-$\txnbox{c}_i$) to obtain an invariant $Q'$ for the execution $\E$
-before the commit, and verifies that committing the transaction under
-$\E$ results in an execution $\E'$ that satisfies transaction's
-postcondition $Q$. As usual, $\E$ and $\E'$ need to be related by $G$.
-The rules for sequential composition of transaction-bound comands
-(\rulelabel{RG-Seq}) and parallel composition of top-level commands
-(\rulelabel{RG-Par}) are straightforward, and more-or-less same as the
-corresponding rules in classical rely-guarantee.
+% RG judgment for the assignment $X:=e$ makes use of the corresponding
+% judgment for the RHS expression $e$. The quantified premise asserts
+% that evaluating the assigment after evaluating $e$ to a value $v$ and
+% an execution $\E$, should result in an execution $\E'$ that satisfies
+% the postcondition $Q$, while being related to $\E$ via the guarantee
+% $G$. RG judgment for the \C{txn} lexical block is similar.  It uses
+% the judgment for the transaction-bound command $c$ (i.e.,
+% $\txnbox{c}_i$) to obtain an invariant $Q'$ for the execution $\E$
+% before the commit, and verifies that committing the transaction under
+% $\E$ results in an execution $\E'$ that satisfies transaction's
+% postcondition $Q$. As usual, $\E$ and $\E'$ need to be related by $G$.
+% The rules for sequential composition of transaction-bound comands
+% (\rulelabel{RG-Seq}) and parallel composition of top-level commands
+% (\rulelabel{RG-Par}) are straightforward, and more-or-less same as the
+% corresponding rules in classical rely-guarantee.
 
-The three \rulelabel{RG-Conseq} rules define ways to strengthen or
-weaken relations and assertions associated with RG judgments of
-transaction-bound expressions, commands and top-level commands,
-respectively. In particular, they allow the precondition $P$ and the
-rely relation $R$ to be strengthened\footnote{A stronger relation is a
-smaller set than a weaker relation, hence a stronger $R$ admits lesser
+The \rulelabel{RG-Conseq} rule defines ways to strengthen or weaken
+relations and assertions associated with the RG judgment of
+transaction-bound expressions. Similar rules exist for
+transaction-bound and top-level commands, but are not shown. In
+particular, the rules allow the precondition $P$ and the rely relation
+$R$ to be strengthened\footnote{A stronger relation is a smaller set
+than a weaker relation, hence a stronger $R$ admits lesser
 interference than a weaker $R$}, and the postcondition $Q$ (also, $C$
@@
 invariant $\I$ to be strengthened. Considering that $\I$ captures
-consistency and isolation properties, this means that a program proved
-correct under weaker consistency and isolation levels is also correct
-under stronger levels.  Parametricity over the trace invariant $\I$,
-combined with the ability to strengthen $\I$ as needed, allows our
-proof system to admit a highly flexible proof strategy to prove
-programs correct on various stores. Instead of requiring programs to
-be proved correct with respect to a pre-defined store (\eg~A CC store
-like in~\cite{gotsmanpopl16}), our reasoning framework allows
-programmers to \emph{define} a store of their choice (by defining $\I$
-appropriately) and then prove programs correct with respect to this
-store. Thus, programmers can prove their programs correct with respect
-to an imaginary store that provides only the required consistency and
-isolation guarantees. Soundness of strengthening $\I$ guarantees that
-the program can now be executed safely on any real store (\eg~a
-relational database, or a NoSQL store) that offers same or stronger
-consistency and isolation guarantees as the imaginary store. The
-following subsection proves the soundness of our proof system.
+isolation properties, this means that a program proved correct under
+weaker isolation levels is also correct under stronger levels.
+Parametricity over the trace invariant $\I$, combined with the ability
+to strengthen $\I$ as needed, allows our proof system to admit a
+highly flexible proof strategy to prove programs correct on various
+stores. For example, programmers can \emph{define} isolation
+guarantees of their choice (by defining $\I$ appropriately) and then
+prove programs correct assuming the guarantees hold. Soundness of
+strengthening $\I$ assures that the program can now be executed safely
+on any real store that offers same or stronger isolation guarantees
+than those assumed. The following subsection proves the soundness of
+our proof system.
 
@@
 Soundness of rely-guarantee judgments can now be stated
-thus\footnote{Supplemntary contains a rigorous proof of the soundness
+thus\footnote{Supplementary contains a rigorous proof of the soundness
 theorem, and all relevant lemmas.}:

### 2016-06-25 reasoning.tex
Typo.
4.3 Done.

@@
 \subsection{The Rely-Guarantee Judgment}
+\label{sec:rely-guarantee}
 
@@
 \begin{array}{lcl}
-\stable(R,\I)& \Leftrightarrow & \forall \E, \E'.\, 
+\stable(R,\I)& \defeq & \forall \E, \E'.\, 
   \I(\E) \conj R(\E,\E')\Rightarrow \I(\E')\\
@@
 The guarantee consequent requires $G$ to capture the trace effect of
-every small-step reduction of $c$, where reduction can be interleaved
-by the interference ($R$) from concurrent threads. Semantics of the RG
+every small-step of $c$, where the reduction can be interleaved by the
+interference ($R$) from concurrent threads. Semantics of the RG
 sextuple for transaction-bound commands ($\txnbox{c}_i$) is defined
@@
 and $\I$, results in a trace that satisfies $Q$, and (b). the effect
-that every small-step reduction of $c$ along the way has on the trace
-is contained in $G$. Soundness of RG judgment for transaction-bound
-commands ($\txnbox{c}_i$) is stated similarly.  For expressions,
-soundness of the judgment $\rg{\I,P,R}{\txnbox{e}_i}{G,C,Q}$ also
-proves that $e$ is always evaluated to a value that satisfies $C$.
+that every small-step of $c$ has on the trace is contained in $G$.
+Soundness of RG judgment for transaction-bound commands
+($\txnbox{c}_i$) is stated similarly.  For expressions, soundness of
+the judgment $\rg{\I,P,R}{\txnbox{e}_i}{G,C,Q}$ also proves that $e$
+is always evaluated to a value that satisfies $C$.
 
 \subsection{Soundness under Eventually Consistent Replication}
+
+\begin{figure}
+\centering
+\subcaptionbox {
+  Replica model
+  \label{fig:ec-theirs}
+} [
+  0.4\columnwidth
+] {
+  \includegraphics[scale=0.7]{Figures/ec-theirs}
+
+}
+\hspace*{0.1in}
+\subcaptionbox {
+  Subset model
+  \label{fig:ec-ours}
+}{
+  \includegraphics[scale=0.45]{Figures/ec-ours}
+} \caption{In the replica model, operation $\op$ generates effect
+$\eta$ at replica $R_1$, which is then merged to $R_2$. If the
+\emph{store is {\sc cc}}, then $R_2$'s state at merge event is same or
+larger than $R_1$'s state at generation event (the difference is
+highlighted). In our subset model, $\op$ witnesses $S_1 \subseteq
+\E.\A$ and generates $\eta$, which is immediately added to $\E.\A$. A
+later operation may witness $S_2 \subseteq \E.\A$, and if the
+\emph{operation is} {\sc cc} and $\eta \in S_2$, then it also
+witnesses $S_1$ (i.e., $S_1 \subseteq S_2$). } 
+% Moreover, Like $R_2 - R_1$, if all effects in $S_2 - S_1$ are
+% concurrent with $\eta$, i.e., $\not\exists\eta'.~\eta' \in S_2 - S_1
+% \conj % visZ(\eta,\eta')$, then any precondition $P$ that is valid
+% when $\op$ executed is also valid when $\eta$ is witnessed because
+% of the stability condition.
+\label{fig:ec-theirs-vs-ours}
+\end{figure}
+
+In this section, we present an intuitive explanation of the soundness
+result in context of eventually consistent replication. 
+% In other words, we explain why our approach is a sound way of
+% reasoning about transactions under eventually consistent replication
+% despite our system model not including the standard artifacts
+% associated with a replicated store.
+
+Existing approaches to reason about eventually consistent replication
+necessarily involve reasoning in terms of \emph{replicas} of data. The
+primary challenge in this setting is to ensure that the assumptions
+made and guarantees enforced by an operation at one replica carry over
+to other replicas that merge its effects, thus preserving the overall
+integrity of the system. Reasoning frameworks, such
+as~\cite{gotsmanpopl16}, address this challenge by imposing
+restrictions on how various replica states differ, i.e., by
+strengthening the consistency of the store. Our view of eventually
+consistent replication however does not explicitly involve replicas.
+Fig.~\ref{fig:ec-theirs-vs-ours} contrasts our model of EC replication
+with the conventional replica-based model.  Under our model, the
+notion of a replica is subsumed by the concept of visibility; a
+replica is defined by the subset ($S$) of global state ($\E.\A$) that
+an operation witnesses.  Constraints over replica states therefore
+manifest as constraints over the visibility relation. For example,
+instead of requiring the store to be causally consistent, an operation
+can request to witness a causally consistent subset of the state; such
+demands can be made via the trace invariant $\I$. For a precondition
+($P$) of the operation to be useful, it has to be an assertion over
+every causally consistent subset of the global state. Since any
+replica that eventually executes the operation has to expose one such
+subset ($S$), the precondition is guaranteed to hold regardless of the
+replica. There is however one problem with this explanation; by
+considering subsets of just one global state, it ignores the fact that
+the global state (hence, the replica states) change during the
+execution of the operation. Existing approaches account for this
+change by distinguishing between \emph{effect generation} event at one
+replica and \emph{effect merge} event at another replica, and
+requiring that certain invariants be preserved between these two
+events at different replicas. Our framework folds all of this into a
+stability condition (\S\ref{sec:rely-guarantee}). Since any change to
+the global state during the execution of the operation is an
+interference, and $P$ is required to be stable with respect to any
+such interference, it follows that $P$ is valid on every replica, thus
+ensuring that assumptions made at the generation event are also valid
+at the merge event.
 

### 2016-06-25 opsem.tex
RG section made succinct.
edits
Sec. 5 created.

@@
+\section{\txnimp}
+\label{sec:syntax}
+
 \subsection{\txnimp: Syntax and Semantics}
-\section{\txnimp}
 \label{sec:opsem}
-
-\label{sec:syntax}
 
@@
 in our model, composed of more fundamental \emph{session order}
-($\soZ$) and \emph{visibility} ($\visZ$) relations.  A session order
-relation captures the sequential order of operations within a
-transaction. In particular, it relates two effects, $\eta_1$ and
-$\eta_2$, such that $\txn(\eta_1) = \txn(\eta_2)$ and $\id(\eta_1) <
-\id(\eta_2)$.  The semantics assigns monotonically increasing
-identifiers to effects, as defined by the $\id(\eta) > \maxId(\A)$
-condition of \rulelabel{E-Aux}.  Evaluation contexts ($\ectx_i$) for
+($\soZ$) and \emph{visibility} ($\visZ$) relations, which shall
+henceforth replace $\hbZ$ in the discussion.  A session order relation
+captures the sequential order of operations within a transaction. In
+particular, it relates two effects, $\eta_1$ and $\eta_2$, such that
+$\txn(\eta_1) = \txn(\eta_2)$ and $\id(\eta_1) < \id(\eta_2)$.  The
+semantics assigns monotonically increasing identifiers to effects, as
+defined by the $\id(\eta) > \maxId(\A)$ condition of
+\rulelabel{E-Aux}.  Evaluation contexts ($\ectx_i$) for
 transaction-bound terms are defined so as to enforce a deterministic
@@
 among the visible effects (this is made possible by appropriately
-defining $\interp{\cdot}$ in \rulelabel{E-Var}, as we show
-later). Thus, the value of a read depends on what write effects it can
+defining $\interp{\cdot}$ in \rulelabel{E-Var}, as we show later).
+Thus, the value of a read depends on what write effects it can
 witness. An operation can only witness the effects of already
@@
 the global state, the semantics models the weak visibility properties
-of such stores.
-
-However, a machine that lets operations witness arbitrary subset of
-the global state offers no isolation whatsoever. For example, it may
-allow a read operation to witness writes of an uncommitted
-transaction, violating RC isolation. Fortunately, our ability to
-express an isolation semantics as constraints over happens-before
-order through $\visZ$ and $\soZ$ relations, and the property of the
-abstract machine to be parametric over the trace invariant ($\I$),
-lets us solve this problem.  In particular, we continue to define
-isolation semantics as constraints over $\visZ$ and $\soZ$, but
-confine their domain of interpretation to the given trace so that they
-now become trace well-formedness constraints. Well-formedness
-constraints can be combined into a trace invariant ($\I$). Since the
-abstract machine is guaranteed to preserve $\I$ at every step of the
+of such stores. Stronger visiblity properties needed for stronger
+isolation can however be recovered by imposing well-formedness
+constraints over $\visZ$ via the trace invariant ($\I$). Since the
+abstract machine is obligated to satisfy $\I$ at every step of the
 execution, operations are guaranteed to experience the level of
@@
 well-formedness constraints.
+
+% However, a machine that lets operations witness arbitrary subset of
+% the global state offers no isolation whatsoever. For example, it may
+% allow a read operation to witness writes of an uncommitted
+% transaction, violating RC isolation. Fortunately, our ability to
+% express an isolation semantics as constraints over happens-before
+% order through $\visZ$ and $\soZ$ relations, and the property of the
+% abstract machine to be parametric over the trace invariant ($\I$),
+% lets us solve this problem.
+% In particular, we continue to define isolation semantics as
+% constraints over $\visZ$ and $\soZ$, but confine their domain of
+% interpretation to the given trace so that they now become trace
+% well-formedness constraints. Well-formedness constraints can be
+% combined into a trace invariant ($\I$).
 
@@
 \input{ansi-isolation}
-
-\input{store-consistency}

### 2016-06-25 reasoning.tex
Sec. 5 created.

--- 2016-06-24 reasoning.tex
@@
 the latter capture \emph{assumptions} about the machine. } ($\I$).
-Since the trace invariant is a conjunction of store consistency
-($\I_s$) and transaction isolation ($\I_c$) constraints, the
-demonstration is a proof that the program preserves its invariants
-when executed on a chosen store under a selection of isolation levels
-for its transactions.
+Since the trace invariant captures transaction isolation constraints,
+the demonstration is a proof that the given selection of isolation
+levels for transactions is sufficient to enforce program's invariants.
 
@@
 and $\I$, results in a trace that satisfies $Q$, and (b). the effect
-that every small-step reduction of $c$ along the way has on the trace
-is contained in $G$. Soundness of RG judgment for transaction-bound
-commands ($\txnbox{c}_i$) is stated similarly.  For expressions,
-soundness of the judgment $\rg{\I,P,R}{\txnbox{e}_i}{G,C,Q}$ also
-proves that $e$ is always evaluated to a value that satisfies $C$.
-
-\subsection{Soundness under Eventually Consistent Replication}
-
-\subsection{Examples}
-
-\GK{Here, I plan to include the full example that we dropped from
-\S\ref{sec:motivation}.}
+that every small-step of $c$ has on the trace is contained in $G$.
+Soundness of RG judgment for transaction-bound commands
+($\txnbox{c}_i$) is stated similarly.  For expressions, soundness of
+the judgment $\rg{\I,P,R}{\txnbox{e}_i}{G,C,Q}$ also proves that $e$
+is always evaluated to a value that satisfies $C$.

### 2016-06-25 store-consistency.tex

@@
-\subsection{Store Consistency}
+\section{Store Consistency}
 \label{sec:store-consistency}
@@
 inevitable to enforce $\I_c$.
+
+\subsection{Eventually Consistent Replication}
+
+\begin{figure}
+\centering
+\subcaptionbox {
+  Replica model
+  \label{fig:ec-theirs}
+} [
+  0.4\columnwidth
+] {
+  \includegraphics[scale=0.7]{Figures/ec-theirs}
+
+}
+\hspace*{0.1in}
+\subcaptionbox {
+  Subset model
+  \label{fig:ec-ours}
+}{
+  \includegraphics[scale=0.45]{Figures/ec-ours}
+} \caption{In the replica model, operation $\op$ generates effect
+$\eta$ at replica $R_1$, which is then merged to $R_2$. If the
+\emph{store is {\sc cc}}, then $R_2$'s state at merge event is same or
+larger than $R_1$'s state at generation event (the difference is
+highlighted). In our subset model, $\op$ witnesses $S_1 \subseteq
+\E.\A$ and generates $\eta$, which is immediately added to $\E.\A$. A
+later operation may witness $S_2 \subseteq \E.\A$, and if the
+\emph{operation is} {\sc cc} and $\eta \in S_2$, then it also
+witnesses $S_1$ (i.e., $S_1 \subseteq S_2$). } 
+% Moreover, Like $R_2 - R_1$, if all effects in $S_2 - S_1$ are
+% concurrent with $\eta$, i.e., $\not\exists\eta'.~\eta' \in S_2 - S_1
+% \conj % visZ(\eta,\eta')$, then any precondition $P$ that is valid
+% when $\op$ executed is also valid when $\eta$ is witnessed because
+% of the stability condition.
+\label{fig:ec-theirs-vs-ours}
+\end{figure}
+
+In this section, we present an intuitive explanation of the soundness
+result in context of eventually consistent replication. 
+% In other words, we explain why our approach is a sound way of
+% reasoning about transactions under eventually consistent replication
+% despite our system model not including the standard artifacts
+% associated with a replicated store.
+
+Existing approaches to reason about eventually consistent replication
+necessarily involve reasoning in terms of \emph{replicas} of data. The
+primary challenge in this setting is to ensure that the assumptions
+made and guarantees enforced by an operation at one replica carry over
+to other replicas that merge its effects, thus preserving the overall
+integrity of the system. Reasoning frameworks, such
+as~\cite{gotsmanpopl16}, address this challenge by imposing
+restrictions on how various replica states differ, i.e., by
+strengthening the consistency of the store. Our view of eventually
+consistent replication however does not explicitly involve replicas.
+Fig.~\ref{fig:ec-theirs-vs-ours} contrasts our model of EC replication
+with the conventional replica-based model.  Under our model, the
+notion of a replica is subsumed by the concept of visibility; a
+replica is defined by the subset ($S$) of global state ($\E.\A$) that
+an operation witnesses.  Constraints over replica states therefore
+manifest as constraints over the visibility relation. For example,
+instead of requiring the store to be causally consistent, an operation
+can request to witness a causally consistent subset of the state; such
+demands can be made via the trace invariant $\I$. For a precondition
+($P$) of the operation to be useful, it has to be an assertion over
+every causally consistent subset of the global state. Since any
+replica that eventually executes the operation has to expose one such
+subset ($S$), the precondition is guaranteed to hold regardless of the
+replica. There is however one problem with this explanation; by
+considering subsets of just one global state, it ignores the fact that
+the global state (hence, the replica states) change during the
+execution of the operation. Existing approaches account for this
+change by distinguishing between \emph{effect generation} event at one
+replica and \emph{effect merge} event at another replica, and
+requiring that certain invariants be preserved between these two
+events at different replicas. Our framework folds all of this into a
+stability condition (\S\ref{sec:rely-guarantee}). Since any change to
+the global state during the execution of the operation is an
+interference, and $P$ is required to be stable with respect to any
+such interference, it follows that $P$ is valid on every replica, thus
+ensuring that assumptions made at the generation event are also valid
+at the merge event.
+
+\subsection{Example}
+
+\GK{Here, I plan to include the full example that we dropped from
+\S\ref{sec:motivation}.}

### 2016-06-25 reasoning.tex
Definitions are now numbered

@@
 
-\begin{definition}[Step-indexed reflexive transitive closure]
+\begin{definition}[\bfseries Step-indexed reflexive transitive closure]
 Forall $A$, $R: A \rightarrow A \rightarrow \mathbb{P}$, and $n :
@@
 
-\begin{definition}[Interleaved step and multi-step relations]
+\begin{definition}[\bfseries Interleaved step and multi-step relations]
 Interleaved step relation interleaves the thread-local reduction with
@@
 
-\begin{definition}[Semantics of the RG judgment]
+\begin{definition}[\bfseries Semantics of the RG judgment]
+\label{def:rg-semantics}
 Semantics of the RG sextuple $\rg{\mathbb{I},P,R}{c}{G,Q}$ is defined
@@
 theorem, and all relevant lemmas.}:
-\begin{theorem}[Soundness] 
+\begin{theorem}[\bfseries Soundness] 
 Rely-guarantee judgments defined by the rules in
-Fig.~\ref{fig:rg-rules} are sound with respect to their semantics
-defined above. 
+Fig.~\ref{fig:rg-rules} are sound with respect to
+Definition~\ref{def:rg-semantics}. 
 \end{theorem}

### 2016-06-25 store-consistency.tex
store consistency edited.

@@
-\section{Store Consistency}
+\section{Data Stores and Consistency}
 \label{sec:store-consistency}
 
-Under a trivial trace invariant (i.e., $\I(\E) = true$), the abstract
-machine of Fig.~\ref{fig:txnimp} assumes the semantics of an {\sc ec} store;
-it allows operations of a transaction to witness arbitrary subsets of the
-global state. A non-trivial $\I$ composed of isolation specifications
-from Fig.~\ref{fig:ansi-isolation} induces the machine to provide
-non-trivial isolation guarantees for transactions. However, weak
-isolation levels often only constrain the visibility sets of
-operations by dictating what \emph{not} to see; not what to see.  For
-instance, \iso{Repeatable Read} isolation prohibits operations of a
-transaction from witnessing different states. It, however, does not
-prohibit all operations of a transaction from witnessing an aribitrary
-subset of the global state. Consequently, the machine can remain an {\sc ec}
-store even while providing non-trivial isolation. How then to model
-the semantics of an {\sc sc} store, such as a relational database, with
-variable (weak) isolation?
+The abstract machine
+of Fig.~\ref{fig:txnimp} allows operations to witness arbitrary
+subsets of the global state, thus assuming the semantics of an
+eventually consistent ({\sc ec}) data store\footnote{{\sc ec}
+guarantees that in the absence of further updates, all reads witness
+the same global state \emph{eventually}. In any finite trace, however,
+there are no guarantees on what a read witnesses.}. There are however
+stores, such as relational databases, that provide stronger
+consistency guarantees than {\sc ec}. Like the isolation levels of
+transactions, the consistency level of the underlying store also
+affects the semantics of a program in non-trivial ways. In this
+section, we demonstrate how the semantics of stronger stores with
+on-demand weak isolation can be captured in our operational model
+and reasoning framework.
 
-The answer lies in enforcing store-specific consistency constraints,
-along with transaction-specific isolation constraints, via the trace
-invariant $\I$. In particular, we split $\I$ into two components: (1).
-$\I_s$, the store-specific invariant, and (2). $\I_c$, the
-program-specific (or, client-specific) invariant, to capture
-consistency and isolation constraints, respectively. $\I$ is now a
-conjunction:
-\begin{smathpar}
-  \I \;=\; \lambda\E.~\I_s(\E) \wedge \I_c(\E)
-\end{smathpar}
-The trace invariant defined in \S\ref{sec:ansi-isolation} for the
-withdraw program of Fig.~\ref{fig:motiv-eg-1} now becomes its $\I_c$
-and remains an invariant regardless of the store. $\I_s$ however
-changes from store to store. We now consider various stores and
-describe their corresponding store invariants.
+% A non-trivial $\I$ composed of isolation specifications from
+% Fig.~\ref{fig:ansi-isolation} induces the machine to provide
+% non-trivial isolation guarantees for transactions. However, weak
+% isolation levels often only constrain the visibility sets of
+% operations by dictating what \emph{not} to see; not what to see.  For
+% instance, \iso{Repeatable Read} isolation prohibits operations of a
+% transaction from witnessing different states. It, however, does not
+% prohibit all operations of a transaction from witnessing an aribitrary
+% subset of the global state. Consequently, the machine can remain an
+% {\sc ec} store even while providing non-trivial isolation. How then to
+% model the semantics of an {\sc sc} store, such as a relational
+% database, with variable (weak) isolation?
 
-\paragraph{An EC store} An eventually-consistent {\sc ec} store provides no
-additional consistency guarantees besides those provided by the operational
-semantics.  Hence, its trace invariant, $\I_s(\E)$, for any execution
-$\E$ is always \emph{true}.
+Firstly, we observe that the semantics of stronger stores can be
+captured by store-specific consistency constraints, along with
+transaction-specific isolation constraints, via the trace invariant
+$\I$. In particular, we split $\I$ into two components: (1).  $\I_s$,
+the store-specific invariant, and (2). $\I_c$, the program-specific
+(or, client-specific) invariant, to capture consistency and isolation
+constraints, respectively. $\I$ is now a conjunction:
+  $\I \;=\; \lambda\E.~\I_s(\E) \wedge \I_c(\E)$.
+While the program-specific trace invariant ($\I_c$) and remains an
+invariant regardless of the store, the store specific invariant
+($\I_s$) changes from store to store depending on the consisteny
+level. $\I_s$ for an {\sc ec} store is $true$. Stronger stores, such
+as an {\sc sc} store, have non-trivial $\I_s$.
 
-\paragraph{An SC store} A strongly-consistent {\sc sc} store guarantees a total order of
-all operations w.r.t $\visZ$ consistent with their chronological
-order. A straightforward definition of $\I_s$ for this store is:
+A strongly-consistent {\sc sc} store guarantees a total order of all
+operations w.r.t $\visZ$ consistent with their chronological order. A
+straightforward $\I_s$ for this store is the \C{SC} property
+formalized below:
 \begin{smathpar}
@@
 consider a case where $\I_c(\E) \;=\; \forall
-T_i.~\underE{\C{RC}(T_i)}$, a constraint that dictates all transactions execute under
-\iso{Read Committed} isolation. Imagine a sample execution where
-$\eta_1$'s transaction is not yet committed when $\eta_2$ is
-generated. Letting $\eta_1$ be visible to $\eta_2$ violates $\I_c$,
-whereas not letting it be visible violates $\I_{s}$. The only way to
-satisfy both the invariants is to rule out all the executions that
-interleave the operations of one transaction with the other, thereby
-enforcing serializability and the ACID model of
-transactions\footnote{Thus, serializability is a natural
-generalization of {\sc sc} to transactions.}. In general, when $\I_s$ is in
-conflict (but not inconsistent) with $\I_c$, the only way to enforce
-both invariant sets is to restrict concurrency. Clearly, this is unacceptable since
-it defeats the very purpose of supporting weak isolation. 
+T_i.~\underE{\C{RC}(T_i)}$, a constraint that dictates all
+transactions execute under \iso{Read Committed} isolation. Imagine a
+sample execution where $\eta_1$'s transaction is not yet committed
+when $\eta_2$ is generated. Letting $\eta_1$ be visible to $\eta_2$
+violates $\I_c$, whereas not letting it be visible violates $\I_{s}$.
+The only way to satisfy both the invariants is to rule out all the
+executions that interleave the operations of one transaction with the
+other, thereby enforcing serializability \footnote{Thus,
+serializability is a natural generalization of {\sc sc} to
+transactions.}. In general, when $\I_s$ is in conflict (but not
+inconsistent) with $\I_c$, the only way to enforce both invariant sets
+is to restrict concurrency. Clearly, this is unacceptable since it
+defeats the very purpose of supporting weak isolation. 
 % How then do we enforce weak isolation on a strongly consistent
@@
 
-Relational databases demonstrate a way of out this impasse.
-Implementations of weak isolation levels on relational databases
-implicitly relax the {\sc sc} requirement so as to maximize
-concurrency and improve performance. Consequently, execution traces do
-not necessarily satisfy {\sc sc}. This approach hints at a general
-solution reconciling $\I_s$ with $\I_c$ that entails weakening the
-former sufficiently to satisfy the latter. However, there are many
-ways $\I_s$ can be weakened; for instance, by setting it to \emph{true}
-which makes the store EC. In reality though, weak isolation on
-databases does not cause so drastic a reduction in consistency. It is
-therefore possible that there exists a principled approach to weaken
-$\I_s$ that adequately captures the underlying properties provided by the store. Such
-a principle can be gleaned by observing that data stores often make
-\emph{recency} commitments~\cite{bailishat}, by which they aim to make
-the most recent data available to clients; {\sc sc} is, in fact, an
-extreme form of a recency commitment, which guarantees that reads
-\emph{always} witness the most recent writes to a data item. In case this conflicts
-with weak isolation guarantees, such as {\sc rr} which only requires that
-reads in a transaction witness the same state, the natural way to
-weaken consistency while prioritizing recency is to let reads witness
-the most recent state that \emph{does not} violate isolation
-constraints. In the context of the operational model of
-Fig.~\ref{fig:txnimp}, this translates to making the largest subset
-($S$) of the global state ($\A$) that does not lead to the violation
-of {\sc rr} isolation constraints visible to an operation.
-
-Generalizing this intuition to any $\I_s$ and $\I_c$ yields a
-\emph{maximum visiblity principle}, which requires the weakened
-consistency guarantee ($\I_s'$) of a store to enforce all visibility
-relationships imposed by the actual consistency guarantee ($\I_s$) on
-a given trace, unless enforcing such a relationship violates $\I_c$.
-Formally, the maximum visibility weakening of $\I_s$ is a new store
-consistency relation $\I_s'$, such that:
+In practice, relational databases resolve the conflict by prioritizing
+weak isolation (thus, concurrency and performance) over strong
+consistency, so the execution traces do not necessarily satisfy {\sc
+sc}. In particular, visibility constraints imposed by {\sc sc} are
+violated if (and only if) they are found to be in conflict with the
+constraints imposed by the isolation level. In the context of
+aforementioned example, $\eta_1$ is not made visible to $\eta_2$
+because doing so would violate $\I_c$. However, if $\I_c$ is true,
+then the store makes $\eta_1$ visible to $\eta_2$ to honor its
+consistency commitment\footnote{The term
+\emph{recency}~\cite{bailisvldb} is often used in place of consistency
+to capture the best-effort nature of the {\sc sc} guarantee on
+databases.}. We generalize this approach to any $\I_s$ and $\I_c$ by
+defining a \emph{maximum visiblity principle} to determine an
+acceptable weakening of $\I_s$ in case of a conflict with $\I_c$.  The
+principle requires the weakened consistency guarantee ($\I_s'$) of the
+store to enforce all visibility relationships imposed by the actual
+consistency guarantee ($\I_s$), unless enforcing such a relationship
+violates $\I_c$. Formally:
+\begin{definition}
+$\I_s' : \E \rightarrow \Prop$ is said to be a maximum visibility
+weakening of $\I_s : \E \rightarrow \Prop$ if and only if:
 \begin{itemize}
@@
 \end{itemize}
+\end{definition}
 
@@
 isolation constraints are captured by $\I_c$:
-
 \begin{smathpar}
@@
 \end{smathpar}
+Instantiating the parameter $\I$  with $\I_s \wedge \I_c$ in
+Fig.~\ref{fig:txnimp} results in an operational semantics that
+describes a practical {\sc sc} store, such as a relational database.
 
-\noindent As usual $\I = \I_s \wedge \I_c$. The resultant operational model
-hides an existing effect ($\eta\in\A$) from being visible to the
-current operation if and only if showing it results in the violation
-of $\I_c$.
+Semantics of a causally consistent data
+store~\cite{gotsmanpopl16,LBC16} can also be obtained: by defining
+$\I_s$ to be the causal consistency ({\sc cc}) invariant that is
+appropriately weakened to resolve conflicts with $\I_c$. Details are
+relegated to the supplementary in the interest of space.
 
-It is worthwhile to note that although the conflict-free definition of
-$\I_s$ shown above captures the behaviour of an {\sc sc} database in
-the presence of isolation constraints, it does not necessarily describe
-how the database achieves this behaviour (i.e., its
-implementation). It is unusual for databases to check for satisfaction
-of isolation constraints every time an operation is executed. Efficient
-implementation strategies (\eg multi-versioning) are employed to
-achieve the twin goals of isolation and consistency/recency. \footnote{\SJ{This
-  paragraph reads poorly - not quite sure what the reader should conclude
-  from this.  A negative interpretation is that we can formalize the
-  relationship between $I_s$ and $I_c$ but the formalization may be
-  far removed from reality and practicality.}}
+\begin{remark}
+Note that the purpose of maximum visibility principle is to
+rationalize the observed behaviour of databases in practice. In
+particular, it is not intended to be a guiding principle to engineer
+data stores, and need not be objectively better than similar
+principles.
+\end{remark}
 
-\paragraph{A CC store} A causally consistent data store~\cite{gotsmanpopl16,LBC16}
-allows operations to only witness a causally consistent snapshot of the
-global state. A straightforward trace invariant ($\I_s$) for this
-store is the causally consistent guarantee that requires an effect to witness all
-causally preceding effects regardless of the isolation requirements of
-its transaction:
-\begin{smathpar}
-\begin{array}{lccl}
-\C{CC}(\E) & \;=\; &  & \forall \eta_1,\eta_2.\, 
-      \E \Vdash \eta_1 \soar \eta_2 \Rightarrow  \underE{\eta_1 \visar
-      \eta_2}\\
-    &   & \wedge & \forall\eta_1,\eta_2,\eta_3.\,\underE{\eta_1 \visar
-      \eta_2} \conj \underE{\eta_2 \visoar \eta_3} \\
-    &   & &\hspace*{0.5in} \Rightarrow \underE{\eta_1 \visar \eta_3}\\
-\end{array}
-\end{smathpar}
-In order to admit weak isolation behavior without restricting concurrency, we weaken $\I_s$ while
-being guided by the maximum visibility principle. The weakened $\I_s$
-is shown below:
+% \paragraph{A CC store} A causally consistent data store~\cite{gotsmanpopl16,LBC16}
+% allows operations to only witness a causally consistent snapshot of the
+% global state. The store-specific invariant obtained by weakening the
+% {\sc cc} guarantee to account for conflicts with $\I_c$ is shown
+% below (the original {\sc cc} does not contain the $\neg\I_c(\dots)$
+% disjuncts): 
+% \begin{smathpar}
+% \begin{array}{lccl}
+% \C{CC}(\E) & \;=\; &  & \forall \eta_1,\eta_2.\, 
+%       \E \Vdash \eta_1 \soar \eta_2 \Rightarrow  \underE{\eta_1 \visar
+%       \eta_2} \\
+%     & & & \hspace*{0.6in}\disj \neg\I_c(\E \cup 
+%                 (\emptyset,\{(\eta_1,\eta_2)\}))\\
+%     &   & \wedge & \forall\eta_1,\eta_2,\eta_3.\,\underE{\eta_1 \visar
+%       \eta_2} \conj \underE{\eta_2 \visoar \eta_3} \\
+%     &   & &\hspace*{0.3in} \Rightarrow \underE{\eta_1 \visar \eta_3}
+%       \disj \neg\I_c(\E \cup (\emptyset,\{(\eta_1,\eta_2)\}))\\
+% \end{array}
+% \end{smathpar}
 
-\begin{smathpar}
-\begin{array}{lccl}
-\C{CC}(\E) & \;=\; &  & \forall \eta_1,\eta_2.\, 
-      \E \Vdash \eta_1 \soar \eta_2 \Rightarrow  \underE{\eta_1 \visar
-      \eta_2} \\
-    & & & \hspace*{0.6in}\disj \neg\I_c(\E \cup 
-                (\emptyset,\{(\eta_1,\eta_2)\}))\\
-    &   & \wedge & \forall\eta_1,\eta_2,\eta_3.\,\underE{\eta_1 \visar
-      \eta_2} \conj \underE{\eta_2 \visoar \eta_3} \\
-    &   & &\hspace*{0.3in} \Rightarrow \underE{\eta_1 \visar \eta_3}
-      \disj \neg\I_c(\E \cup (\emptyset,\{(\eta_1,\eta_2)\}))\\
-\end{array}
-\end{smathpar}
-
-\noindent Instantiating the parameter $\I$  with $\I_s \wedge \I_c$ in
-Fig.~\ref{fig:txnimp} results in an operational semantics that admits
-violation of causal consistency if and only if the violation is
-inevitable to enforce $\I_c$.
+% \noindent Instantiating the parameter $\I$  with $\I_s \wedge \I_c$ in
+% Fig.~\ref{fig:txnimp} results in an operational semantics that admits
+% violation of causal consistency if and only if the violation is
+% inevitable to enforce $\I_c$.
 

### 2016-06-25 store-consistency.tex
edits to sec. 5

@@
 
-The abstract machine
-of Fig.~\ref{fig:txnimp} allows operations to witness arbitrary
-subsets of the global state, thus assuming the semantics of an
-eventually consistent ({\sc ec}) data store\footnote{{\sc ec}
-guarantees that in the absence of further updates, all reads witness
-the same global state \emph{eventually}. In any finite trace, however,
-there are no guarantees on what a read witnesses.}. There are however
-stores, such as relational databases, that provide stronger
-consistency guarantees than {\sc ec}. Like the isolation levels of
-transactions, the consistency level of the underlying store also
-affects the semantics of a program in non-trivial ways. In this
-section, we demonstrate how the semantics of stronger stores with
-on-demand weak isolation can be captured in our operational model
-and reasoning framework.
+The abstract machine of Fig.~\ref{fig:txnimp} allows operations to
+witness arbitrary subsets of the global state, thus assuming the
+semantics of an eventually consistent ({\sc ec}) data
+store\footnote{{\sc ec} guarantees that in the absence of further
+updates, all reads witness the same global state \emph{eventually}. In
+any finite trace, however, there are no guarantees on what a read
+witnesses.}. There are however stores, such as relational databases,
+that provide stronger consistency guarantees than {\sc ec}. Like the
+isolation levels of transactions, the consistency level of the
+underlying store also affects the semantics of a program in
+non-trivial ways. In this section, we demonstrate how the semantics of
+stronger stores with on-demand weak isolation can be captured in our
+operational model so that we can use the reasoning framework developed
+in \S\ref{sec:reasoning} to reason about such stores.
 
@@
 constraints, respectively. $\I$ is now a conjunction:
-  $\I \;=\; \lambda\E.~\I_s(\E) \wedge \I_c(\E)$.
+  $\I \,=\, \lambda\E.~\I_s(\E) \wedge \I_c(\E)$ (often simplified to
+  $\I \,=\, \I_s\wedge\I_c$).
 While the program-specific trace invariant ($\I_c$) and remains an
@@
 because doing so would violate $\I_c$. However, if $\I_c$ is true,
-then the store makes $\eta_1$ visible to $\eta_2$ to honor its
+then the store needs to make $\eta_1$ visible to $\eta_2$ to honor its
 consistency commitment\footnote{The term
@@
 consistency guarantee ($\I_s$), unless enforcing such a relationship
-violates $\I_c$. Formally:
-\begin{definition}
-$\I_s' : \E \rightarrow \Prop$ is said to be a maximum visibility
-weakening of $\I_s : \E \rightarrow \Prop$ if and only if:
-\begin{itemize}
-  \item $\I_s'$ is weaker than $\I_s$: 
-      $\forall\E.~ \I_s(\E) \Rightarrow \I_s'(\E)$, and
-  \item In every trace $\E$ that satisfies $\I_c$, and for every pair
-  of effects $\eta_1$ and $\eta_2$ in $\E$, if $\I_s(\E)$ requires
-  $\eta_1$ to be visible to $\eta_2$, then so does $\I_s'(\E)$ unless
-  extending $\E$ with $\visZ(\eta_1,\eta_2)$ violates
-  $\I_c$\footnote{\GK{ToDo: consider other well-formedness conditions
-  on trace, such as acyclicity of $\visZ$ and $\soZ$. Are they needed
-  (considering that the machine never violates them)? Encode the
-  specifications in Z3 and make sure they are consistent.}}:
-  \begin{smathpar}
-  \begin{array}{l}
-  \forall\E,\eta_1,\eta_2.~ \I_c(\E) \Rightarrow (\I_s(\E)
-    \Rightarrow \underE{\eta_1 \visar \eta_2}) \Rightarrow \\
-    \hspace*{0.5in}(\I_s'(\E) \Rightarrow \underE{\eta_1 \visar
-    \eta_2} \disj \neg\I_c(\E\,\cup\,(\emptyset,\{(\eta_1,\eta_2)\})))
-  \end{array}
-  \end{smathpar}
-\end{itemize}
-\end{definition}
+violates $\I_c$. The formal definition of the principle is relegated
+to the supplementary in the interest of space, but it can be
+understood in the context of an {\sc sc} store, where it weakens the
+store invariant to the following:
+% Formally:
+% \begin{definition}
+% $\I_s' : \E \rightarrow \Prop$ is said to be a maximum visibility
+% weakening of $\I_s : \E \rightarrow \Prop$ if and only if:
+% \begin{itemize}
+%   \item $\I_s'$ is weaker than $\I_s$: 
+%       $\forall\E.~ \I_s(\E) \Rightarrow \I_s'(\E)$, and
+%   \item In every trace $\E$ that satisfies $\I_c$, and for every pair
+%   of effects $\eta_1$ and $\eta_2$ in $\E$, if $\I_s(\E)$ requires
+%   $\eta_1$ to be visible to $\eta_2$, then so does $\I_s'(\E)$ unless
+%   extending $\E$ with $\visZ(\eta_1,\eta_2)$ violates
+%   $\I_c$\footnote{\GK{ToDo: consider other well-formedness conditions
+%   on trace, such as acyclicity of $\visZ$ and $\soZ$. Are they needed
+%   (considering that the machine never violates them)? Encode the
+%   specifications in Z3 and make sure they are consistent.}}:
+%   \begin{smathpar}
+%   \begin{array}{l}
+%   \forall\E,\eta_1,\eta_2.~ \I_c(\E) \Rightarrow (\I_s(\E)
+%     \Rightarrow \underE{\eta_1 \visar \eta_2}) \Rightarrow \\
+%     \hspace*{0.5in}(\I_s'(\E) \Rightarrow \underE{\eta_1 \visar
+%     \eta_2} \disj \neg\I_c(\E\,\cup\,(\emptyset,\{(\eta_1,\eta_2)\})))
+%   \end{array}
+%   \end{smathpar}
+% \end{itemize}
+% \end{definition}
 
-Applying this principle, we can weaken {\sc sc} to obtain the
-following store trace invariant ($\I_s$) for an {\sc sc} store whose
-isolation constraints are captured by $\I_c$:
+% Applying this principle, we can weaken {\sc sc} to obtain the
+% following store trace invariant ($\I_s$) for an {\sc sc} store whose
+% isolation constraints are captured by $\I_c$:
 \begin{smathpar}
@@
 \end{smathpar}
+$\I_s$ requires a trace $\E$ to satisfy the visibility constraints of
+{\sc sc} except in cases where they are in conflict with $\I_c$.
 Instantiating the parameter $\I$  with $\I_s \wedge \I_c$ in
 Fig.~\ref{fig:txnimp} results in an operational semantics that
-describes a practical {\sc sc} store, such as a relational database.
+describes relational database-like stores.
 
@@
 result in context of eventually consistent replication.
-% In other words, we explain why our approach is a sound way of
-% reasoning about transactions under eventually consistent replication
-% despite our system model not including the standard artifacts
-% associated with a replicated store.
 

### 2016-06-25 store-consistency.tex

--- 2016-06-25 store-consistency.tex
@@
 consistency guarantee ($\I_s$), unless enforcing such a relationship
-violates $\I_c$. Formally:
+violates $\I_c$. 
+% The formal definition of the principle is relegated
+% to the supplementary in the interest of space, but it can be
+% understood in the context of an {\sc sc} store, where it weakens the
+% store invariant to the following:
+Formally:
 \begin{definition}
@@
 \end{definition}
-
 Applying this principle, we can weaken {\sc sc} to obtain the
@@
 
-\subsection{Eventually Consistent Replication}
+\subsection{Weakly Consistent Replication}
 
@@
 
-In this section, we present an intuitive explanation of the soundness
-result in context of eventually consistent replication. 
-% In other words, we explain why our approach is a sound way of
-% reasoning about transactions under eventually consistent replication
-% despite our system model not including the standard artifacts
-% associated with a replicated store.
+Reasoning under weakly consistent replication has received special
+attention in the recent research~\cite{gotsmanpopl16}. Our operational
+semantics and proof system are however more general and admit
+replication as a special case. In this section, we explain how our
+reasoning approach works under weakly consistent replication by
+comparing our model with the standard model of replication.
 
-Existing approaches to reason about eventually consistent replication
-necessarily involve reasoning in terms of \emph{replicas} of data. The
-primary challenge in this setting is to ensure that the assumptions
-made and guarantees enforced by an operation at one replica carry over
-to other replicas that merge its effects, thus preserving the overall
-integrity of the system. Reasoning frameworks, such
-as~\cite{gotsmanpopl16}, address this challenge by imposing
-restrictions on how various replica states differ, i.e., by
-strengthening the consistency of the store. Our view of eventually
-consistent replication however does not explicitly involve replicas.
-Fig.~\ref{fig:ec-theirs-vs-ours} contrasts our model of EC replication
-with the conventional replica-based model.  Under our model, the
-notion of a replica is subsumed by the concept of visibility; a
-replica is defined by the subset ($S$) of global state ($\E.\A$) that
-an operation witnesses.  Constraints over replica states therefore
-manifest as constraints over the visibility relation. For example,
-instead of requiring the store to be causally consistent, an operation
-can request to witness a causally consistent subset of the state; such
-demands can be made via the trace invariant $\I$. For a precondition
-($P$) of the operation to be useful, it has to be an assertion over
-every causally consistent subset of the global state. Since any
-replica that eventually executes the operation has to expose one such
-subset ($S$), the precondition is guaranteed to hold regardless of the
-replica. There is however one problem with this explanation; by
-considering subsets of just one global state, it ignores the fact that
-the global state (hence, the replica states) change during the
-execution of the operation. Existing approaches account for this
-change by distinguishing between \emph{effect generation} event at one
-replica and \emph{effect merge} event at another replica, and
-requiring that certain invariants be preserved between these two
-events at different replicas. Our framework folds all of this into a
-stability condition (\S\ref{sec:rely-guarantee}). Since any change to
-the global state during the execution of the operation is an
-interference, and $P$ is required to be stable with respect to any
-such interference, it follows that $P$ is valid on every replica, thus
-ensuring that assumptions made at the generation event are also valid
-at the merge event.
+Reasoning about eventually consistent replication often involves
+reasoning in terms of \emph{replicas} of data. The primary challenge
+in this setting is to ensure that the assumptions made and guarantees
+enforced by an operation at one replica carry over to other replicas
+that merge its effects, thus preserving the overall integrity of the
+system. In~\cite{gotsmanpopl16}, this challenge is partly addressed by
+imposing restrictions on how various replica states differ, i.e., by
+fixing a system model with a stronger baseline consistency ({\sc cc})
+than {\sc ec}. This unfortunately restricts the applicability of
+reasoning approach from being extended to weaker stores such as
+Bayou~\cite{bayou} or Quelea~\cite{pldi15}. Our view of replication
+however does not explicitly involve replicas.
+Fig.~\ref{fig:ec-theirs-vs-ours} contrasts our model of weakly
+consistent replication with the conventional replica-based model.
+Under our model, the notion of a replica is subsumed by the concept of
+visibility; a replica is defined by the subset ($S$) of global state
+($\E.\A$) that an operation witnesses. Constraints over replica states
+therefore manifest as constraints over the visibility relation.  For
+example, instead of requiring the store to be causally consistent, an
+operation can request to witness a causally consistent subset of the
+state; such demands can be made via the trace invariant $\I$. For a
+precondition ($P$) of the operation to be useful, it has to be an
+assertion over every causally consistent subset of the global state.
+Since any replica that eventually executes the operation has to expose
+one such subset ($S$), the precondition is guaranteed to hold
+regardless of the replica. There is however one problem with this
+explanation; by considering subsets of just one global state, it
+ignores the fact that the global state (hence, the replica states)
+change during the execution of the operation. To account for this
+change,~\cite{gotsmanpopl16} distinguishes between \emph{effect
+generation} event at one replica and \emph{effect merge} event at
+another replica, and requires that certain invariants be preserved
+between these two events at different replicas. Our framework folds
+all of this into a stability condition (\S\ref{sec:rely-guarantee}).
+Since any change to the global state during the execution of the
+operation is an interference, and $P$ is required to be stable with
+respect to any such interference, it follows that $P$ is valid on
+every replica, thus ensuring that assumptions made at the generation
+event are also valid at the merge event.
 

### 2016-06-26 abstract.tex
edits to secs. 1 - 4

@@
   reasoning about the behavior of concurrent transactions by ensuring
-  an executing transaction is \emph{isolated} from the effects of
-  other concurrenty executing transactions once it begins executing.
-  However, enforcing serializability comes at the cost in performance
-  because it necessarily restricts opportunities to exploit
-  concurrency even when such opportunities would not violate
-  application-specific invariants.
+  transactions are \emph{isolated} from each other's effects once they
+  begin executing.  However, enforcing serializability comes at a
+  steep cost in performance because it necessarily restricts
+  opportunities to exploit concurrency even when such opportunities
+  would not violate application-specific invariants.
 

### 2016-06-26 ansi-isolation.tex
edits to secs. 1 - 4

@@
 
-\begin{figure*}[t]
+\begin{figure*}[t!]
 \begin{smathpar}
@@
 $\E' = (\E.\A \cup \{\eta_2\},\,\E.\visZ \cup \{(\eta_1,\eta_2)\})$.
-When $\psi$ is a proposition, $\underE{\psi}$ denotes the
+When $\psi$ is a formula, $\underE{\psi}$ denotes the
 interpretation of $\psi$ in the context of the trace $\E$. Such
@@
   T_i}$.  The interpretation of $\eta_1 \visar \eta_2$ and $\eta_1
-\soar \eta_2$ are straightforward and explained in \S\ref{sec:syntax}.
-A transaction $T_i$ is said to be visible to an effect $\eta$ if every
-effect $\eta_1$ of $T_i$ recorded by the trace is visible to $\eta$.
-$T_i$ may be visible to $\eta$ but may not be visible to every other
-effect in the transaction. For a transaction $T_i$ to be considered to
-be visible to a transaction $T_j$ in the context of a trace $\E$
-(written $\underE{T_i \visar T_j}$), every effect ($\eta_1$) of $T_i$
-present in $\E$ must be visible to every effect ($\eta_2$) of $T_j$ in
+\soar \eta_2$ are as explained earlier.  A transaction $T_i$ is said
+to be visible to an effect $\eta$ if every effect $\eta_1$ of $T_i$
+recorded by the trace is visible to $\eta$.  $T_i$ may be visible to
+$\eta$ but may not be visible to every other effect in the
+transaction. For a transaction $T_i$ to be considered to be visible to
+a transaction $T_j$ in the context of a trace $\E$ (written
+$\underE{T_i \visar T_j}$), every effect ($\eta_1$) of $T_i$ present
+in $\E$ must be visible to every effect ($\eta_2$) of $T_j$ in
 $\E$. Conversely, if none of the effects of $T_i$ present in $\E$ are
@@
 relational stores (formalized in \S\ref{sec:store-consistency})
-automatically leads to the monotonicity guarantee, which probably
+automatically leads to the monotonicity guarantee, which 
 explains why {\sc rc} comes with $\mathtt{MonotonicVis}$ on such
@@
 isolation level called \iso{Monotonic Atomic View} ({\sc mav} of
-Fig.~\ref{fig:ansi-isolation})~\cite{bailishat,pldi15} was proposed to
+Fig.~\ref{fig:ansi-isolation})~\cite{bailishat,pldi15} has been proposed to
 explicitly extend {\sc rc} with monotonicity on such stores. Sample
 {\sc rc} executions shown in Fig.~\ref{fig:rc-executions} are also
-{\sc mav}. \iso{Repeatable Read} spec ({\sc rr}) extends {\sc mav}
+{\sc mav}. \iso{Repeatable Read} ({\sc rr}) isolation  extends {\sc mav}
 with the snapshot visibility guarantee as described before.
@@
 
-In contrast to recent proposals (\emph{e.g.},
-~\cite{gotsmanconcur15}), our specifications for {\sc si} and {\sc
-  ser} do not necessarily impose a total order among transactions
-(conflicting or otherwise). In reality, a total order under {\sc ser}
-(resp. {\sc si}) is guaranteed only if the store executes all
-transactions under {\sc ser} (resp. {\sc SI}) isolation. Our
-specifications admit this possibility, and derive a total order under
-the assumption of homogenity. However, databases almost always allow
-isolation levels to be configured on a per-transaction basis, allowing
-transactions at various isolation levels to coexist.  Specifications
-that assume homogenity are incorrect under this setting.
+%% SJ: Not sure this paragraph is necessary here.
+%% In contrast to recent proposals (\emph{e.g.},
+%% ~\cite{gotsmanconcur15}), our specifications for {\sc si} and {\sc
+%%   ser} do not necessarily impose a total order among transactions
+%% (conflicting or otherwise). In reality, a total order under {\sc ser}
+%% (resp. {\sc si}) is guaranteed only if the store executes all
+%% transactions under {\sc ser} (resp. {\sc si}) isolation. Our
+%% specifications admit this possibility, and derive a total order under
+%% the assumption of homogenity. However, databases almost always allow
+%% isolation levels to be configured on a per-transaction basis, allowing
+%% transactions at various isolation levels to coexist.  Specifications
+%% that assume homogenity are incorrect under this setting.
 

### 2016-06-26 introduction.tex
edits to secs. 1 - 4

@@
 The tension between serializable transactions, which are easy to
-reason about but difficult to implement, and more pragmatic goals but
-weaker variants that are driven by performance and availability
-considerations, has motivated the development of weaker forms of
-transaction isolation, beginning as early as 1976~\cite{gray1976}. The
-ANSI SQL 92 standard defines three such weak isolation levels which
-are now implemented in many relational and NoSQL databases. Not
-surprisingly, weakly-isolated transactions have been found to
-significantly outperform serializable transactions on benchmark
-suites, both on single-node databases and multi-node replicated
-stores~\cite{dbtuningbook,bailishat,bailisvldb}, leading to their
-overwhelming adoption. A 2013 study~\cite{bailishotos} of 18 popular
-ACID and ``NewSQL'' databases found that only three of them offer
-serializability by default, and half, including Oracle 11g, do not
-offer it at all.  A 2015 study~\cite{bailisferal} of a large corpus of
-database applications finds no evidence that applications manifestly
-change the default isolation level offered by the database. Taken
-together, these studies make clear that weakly-isolated transactions
-are quite prevalent in practice, and strongly-isolated (serializable)
-transactions are often eschewed.
+reason about but difficult to implement, and more pragmatic variants
+that are driven by performance and availability considerations, has
+motivated the development of weaker forms of transaction isolation,
+beginning as early as 1976~\cite{gray1976}. The ANSI SQL 92 standard
+defines three such weak isolation levels which are now implemented in
+many relational and NoSQL databases. Not surprisingly, weakly-isolated
+transactions have been found to significantly outperform serializable
+transactions on benchmark suites, both on single-node databases and
+multi-node replicated stores~\cite{dbtuningbook,bailishat,bailisvldb},
+leading to their overwhelming adoption. A 2013
+study~\cite{bailishotos} of 18 popular ACID and ``NewSQL'' databases
+found that only three of them offer serializability by default, and
+half, including Oracle 11g, do not offer it at all.  A 2015
+study~\cite{bailisferal} of a large corpus of database applications
+finds no evidence that applications manifestly change the default
+isolation level offered by the database. Taken together, these studies
+make clear that weakly-isolated transactions are quite prevalent in
+practice, and strongly-isolated (serializable) transactions are often
+eschewed altogether.
 
@@
 Fekete \emph{et al.}~\cite{feketevldb09} have devised and experimented
-with a microbenchmark suite that executes transactions under a weakly
-isolated \emph{read committed} isolation level - the default level for
-8 of the 18 databases studied in~\cite{bailishotos}, and found that 25
-out of every 1000 rows in the database violate at least one integrity
-constraint. Bailis \emph{et al.}~\cite{bailisferal} rely on Rails'
-\emph{uniqueness validation} to maintain uniqueness of records while
-serving Linkbench's~\cite{linkbench} insertion workload (6400 records
-distributed over 1000 keys; 64 concurrent clients), and end up with
-more than 10 duplicate records. Rails relies on database transactions
-to validate uniqueness during insertions, which is sensible if
-transactions are serializable, but incorrect under the \emph{read
-  committed} isolation level used in the experiments. The same study
-has found that 13\% of all invariants among 67 open source
+with a microbenchmark suite that executes transactions under a
+weakly-isolated \emph{read committed} isolation level - the default
+level for 8 of the 18 databases studied in~\cite{bailishotos}, and
+found that 25 out of every 1000 rows in the database violate at least
+one integrity constraint. Bailis \emph{et al.}~\cite{bailisferal} rely
+on Rails' \emph{uniqueness validation} to maintain uniqueness of
+records while serving Linkbench's~\cite{linkbench} insertion workload
+(6400 records distributed over 1000 keys; 64 concurrent clients), and
+report discovering more than 10 duplicate records.  Rails relies on
+database transactions to validate uniqueness during insertions, which
+is sensible if transactions are serializable, but incorrect under the
+\emph{read committed} isolation level used in the experiments. The
+same study has found that 13\% of all invariants among 67 open source
 Ruby-on-Rails applications are at risk of being violated due to weak
@@
 determine, in each case, if the phenomenon can be allowed without
-violating application invariants. This is understandably
-hard. Adya~\cite{adyaphd} presents the first formal definitions of
-some of well-known isolation levels in the context of a sequentially
-consistent (SC) database.  However, there has been little progress
-relating Adya's system model to a formal operational semantics or a
-proof system that can facilitate rigorous correctness arguments.
-Consequently, reasoning about weak isolation remains an error prone
-endeavor, with major database vendors continuing to document their
-isolation levels primarily in terms of the undesirable
+violating application invariants. This is understandably hard,
+especially in the absence of any formal underpinning to define weak
+isolation semantics.  Adya~\cite{adyaphd} presents the first formal
+definitions of some of well-known isolation levels in the context of a
+sequentially consistent (SC) database.  However, there has been little
+progress relating Adya's system model to a formal operational
+semantics or a proof system that can facilitate rigorous correctness
+arguments.  Consequently, reasoning about weak isolation remains an
+error prone endeavor, with major database vendors continuing to
+document their isolation levels primarily in terms of the undesirable
 phenomena~\cite{postgresiso, mysqliso, oracleiso} a particular
-isolation level may induce; it thus squarely becomes the programmer's
-responsibility to determine if their application can yield an
-execution which can trigger such phenomena.
+isolation level may induce, placing the burden squarely on the
+programmer to determine application correctness.
 
@@
 At a high level, reasoning about weak consistency involves reasoning
-about the concurrent execution of strongly atomic and isolated
-operations against different but consistent states of the database or
-data store, where each operation locally preserves application
-invariants. In contrast, reasoning about weak isolation involves
-reasoning about the concurrent execution of transactions, whose
-operations witness different states which, although consistent
-individually, may not be obviously reconciled into a consistent global
-view.  Moreover, operations may violate invariants as long as the
-transaction as a whole preserves them.  Intuitively, the latter
-problem reduces to the former if all transactions contain a single
+about the concurrent execution of strongly atomic and fully isolated
+operations against different (potentially replicated), but consistent
+states of the database or data store, where each operation locally
+preserves application invariants. In contrast, reasoning about weak
+isolation involves reasoning about the concurrent execution of
+transactions, whose operations witness different states which,
+although consistent individually, may not be obviously reconciled into
+a consistent global view.  Intuitively, the latter problem reduces to
+the former if all transactions contain a single
 operation. Furthermore, weak isolation can exist independent of weak
@@
 build correctness proofs for transactional programs in the presence of
-a weakly isolated concurrency control mechanism A key novelty of our
+a weakly isolated concurrency control mechanism.  A key novelty of our
 approach is that it is parametric over the \emph{isolation semantics}
@@
 does not assume a minimum or predefined set of consistency or
-isolation levels. Instead, our operational semantics admits declarative
-specifications of transaction isolation and store consistency levels
-(\eg~\cite{pldi15,gotsmanconcur15}), and generates executions that are
-guaranteed to satisfy these specifications. The result is a flexible
-system model that is general enough to incorporate the semantics of a
-range of isolation levels on a variety of stores (\eg the sequentially
-consistent store of~\cite{adyaphd}, or the causally consistent store
-of~\cite{gotsmanpopl16}, etc).  Our key technical contribution is thus
-a set of proof rules that demonstrate that a program with a given
-selection of isolation levels for its transactions preserves its 
-invariants when executed on a store satisfying certain consistency
-guarantees (e.g., causal visibility).
+isolation levels. Instead, our operational semantics admits
+declarative specifications of transaction isolation and store
+consistency levels (\eg~\cite{pldi15,gotsmanconcur15}), and generates
+executions that are guaranteed to satisfy these specifications. The
+result is a flexible system model that is general enough to
+incorporate the semantics of a range of isolation levels on a variety
+of stores (\eg the sequentially consistent store of~\cite{adyaphd}, or
+the causally consistent store of~\cite{gotsmanpopl16}, etc).  Our key
+technical contribution is thus a set of proof rules that demonstrate
+that a program with a given selection of isolation levels for its
+transactions preserves its invariants when executed on a store
+equipped with certain consistency guarantees (e.g., causal
+visibility).
 
@@
 
-\noindent Our results thus provide the first (to the best of our knowledge)
-mechanism that precisely relates high-level program invariants to
-low-level weak isolation and store consistency guarantees, thereby
-allowing weakly-isolated transactions to enjoy the same rigorous
-reasoning capabilities as their strongly isolated (serializable)
-counterparts.
+\noindent Our results thus provide the first (to the best of our
+knowledge) mechanism that precisely relates high-level program
+invariants to low-level weak isolation and store consistency
+guarantees, thereby allowing weakly-isolated transactions to enjoy the
+same rigorous reasoning capabilities as their strongly isolated
+(serializable) counterparts.
 
@@
 transactions. \S\ref{sec:opsem} presents an operational semantics for
-a core language that supports weakly-isolated transactions, and which
-is parameterized over different isolation and store-consistency
-notions. \S\ref{sec:proofrules} formalizes the proof systemt that we
-use to reason about program invariants, and establishes the soundness
-of these rules with respect to the semantics.  We describe the impact
-of our reasoning framework in the context of several real-world case
-studies in \S\ref{sec:casestudies}.  Related work and conclusions are
-given in \S\ref{sec:relatedwork}.
+a core language that supports weakly-isolated transactions,
+parameterized over different isolation notions. \S\ref{sec:proofrules}
+formalizes the proof systemt that we use to reason about program
+invariants, and establishes the soundness of these rules with respect
+to the semantics. \S\ref{sec:store-consistency} generalizes the
+framework to integrate support for weakly consistent data stores.  We
+describe the impact of our reasoning framework in the context of
+several real-world case studies in \S\ref{sec:casestudies}.  Related
+work and conclusions are given in \S\ref{sec:relatedwork}.

### 2016-06-26 motivation.tex
Secs. 4.1 and 4.2 done.
edits to secs. 1 - 4

@@
 operations.  The subscript of an operation indicates the transaction
-that executes it\footnote{For clarity, the effects of different
+that executes it.\footnote{For clarity, the effects of different
   transactions are shown in different colored backgrounds.} In the
@@
 concurrency control, it is also more efficient, making it appropriate
-appropriate for both {\bf Wd} transactions.
+for both {\bf Wd} transactions.
 
@@
 isolation variants are implementated is no better since it requires
-application programmers to understand low-level implementation details
+programmers to understand low-level implementation details
 of the database that are naturally far removed from the application
@@
 style reasoning~\cite{rgjones}, a compositional proof technique that
-allows us to reason about one thread at a time by abstracting away
-interference induced by other threads (collectively called \emph{the
-  environment}) into a \emph{rely} relation. In an ordinary concurrent
-program, every environment step is a valid interference in the current
-thread. However, in the presence of transactions executing under weak
-isolation, determining what constitutes an interference is a
-non-trivial problem. For example, inside a serializable transaction
-there is no interference, whereas within a transaction executing under
-\iso{Read Uncommitted}, the weakest isolation level, all interferences
-are valid. Between these two extremes are various levels of isolation
-that admit some interferences while prohibiting others.  For example,
+allows us to reason about the behaviour of individual threads by
+abstracting away interference induced by other threads (collectively
+called \emph{the environment}) into a \emph{rely} relation. In an
+ordinary concurrent program, every environment step is a valid
+interference in the current thread. However, in the presence of
+transactions executing under weak isolation, determining what
+constitutes an interference is a non-trivial problem. For example,
+inside a serializable transaction there is no interference, whereas
+within a transaction executing under a \iso{Read Uncommitted}
+isolation level, the weakest isolation level, all interferences are
+valid. Between these two extremes are various levels of isolation that
+admit some interferences while prohibiting others.  For example,
 \iso{Read Committed} isolation admits interference of a committed
@@
 allowed interference in accordance with the chosen level of isolation
-associated with a transaction. The first key observation we make is
+associated with a transaction. The first observation we make is
 that we can constrain interference by constraining the definition of
-the \emph{happens-before} ($\hbZ$) relation between transactions, or
-their respective constituents.  To do so, we axiomatize the $\hbZ$
-relation to capture the interference characteristics of different
-isolation levels as their specifications ($\psi$). For example, a
-\iso{Read Committed} specification allows an operation $\eta_1$ of
-$T_1$ to happen before an operation $\eta_2$ of $T_2$ only if every
-operation in $T_1$ (including its commit) happens before $\eta_2$ (we
-let $T_1 \hboar \eta_2$ denote this). The {\sc rc} specification
-does not require $T_1$ to execute and commit before all operations of
-$T_2$, thus allowing it to interfere in $T_2$. The \iso{Snapshot
-  Isolation} specification requires two transactions that write to the
-same variable to be related by happens-before.  This effectively
-prohibits interference due to actions in $T_1$ being interleaved in
-$T_2$ (or, vise versa).  We formalize these properties in Fig.~\ref{fig:interference-ex}\footnote{Note that specifications
+the \emph{happens-before} ($\hbZ$) relation that dictates visibility
+actions between transactions, or their respective constituents.  To do
+so, we axiomatize the $\hbZ$ relation to capture the interference
+characteristics of different isolation levels as their specifications
+($\psi$). For example, a \iso{Read Committed} specification allows an
+operation $\eta_1$ of $T_1$ to happen before an operation $\eta_2$ of
+$T_2$ only if every operation in $T_1$ (including its commit) happens
+before $\eta_2$ (we let $T_1 \hboar \eta_2$ denote this). The 
+specification does not require $T_1$ to execute and commit before all
+operations of $T_2$, thus allowing it to interfere in $T_2$. The
+\iso{Snapshot Isolation} specification requires two transactions that
+write to the same variable to be related by happens-before.  This
+effectively prohibits interference due to actions in $T_1$ being
+interleaved in $T_2$ (or, vise versa).  We formalize these properties
+in Fig.~\ref{fig:interference-ex}.\footnote{Note that specifications
   presented here are solely for illustration. The actual
-    specifications described in \S\ref{sec:ansi-isolation} are more nuanced.}.
+  specifications described in \S\ref{sec:ansi-isolation} are more
+  nuanced.}.
 
@@
 The specification of an isolation level encodes its interference
-characteristic as constraints over the $\hbZ$ relation. However, for this
-to be useful in reasoning about programs, the rely-guarantee framework
-should be able to use the specification to determine if an
+characteristic as constraints over the $\hbZ$ relation. However, for
+this to be useful in reasoning about programs, the rely-guarantee
+framework should be able to use the specification to determine if an
 interference is valid or invalid, allowing the programmer to only
-focus on the former. Our second key observation is that this is
-possible if the reasoning framework adequately tracks $\hbZ$ at each
-program point, while preserving $\hbZ$ constraints as invariants
-between program points. An interference that leads to the violation of
-$\hbZ$ constraints (i.e., an invalid interference) is thus
-automatically prohibited. For instance, consider the program point
-after the write to \C{C} in {\bf Wd1}. The expected invariant ($\phi$) at
-that program point is shown below ($\committed$ stands for
-``committed''):
+focus on the former. Our second observation is that this is possible
+if the reasoning framework adequately tracks $\hbZ$ at each program
+point, while preserving $\hbZ$ constraints as invariants between
+program points. An interference that leads to the violation of $\hbZ$
+constraints (i.e., an invalid interference) is thus automatically
+prohibited. For instance, consider the program point after the write
+to \C{C} in {\bf Wd1}. The expected invariant ($\phi$) at that program
+point is shown below ($\committed$ stands for ``committed''):
 
@@
       \C{Wd1} \wrstoar \C{C} \conj (\committed(\C{Wd2})
-                \Rightarrow \C{C = k-a1-a2})\}
+                \Rightarrow \C{C = k-a1-a2})
 \end{array}
@@
 
-We have thus far assumed a strongly consistent ({\sc SC}) store that,
+We have thus far assumed a strongly consistent ({\sc sc}) store that,
 in the absence of transactions with special isolation requirements,
 makes the effects of any operation immediately visible to all
-subsequent operations. The natural behavior of {\sc SC} to totally
+subsequent operations. The natural behavior of {\sc sc} to totally
 order all operations w.r.t the $\hbZ$ relation could be in conflict
@@
 until they are committed. Once committed though, effects need to be
-immediately and perennially visible to subsequent operations. This
-semantics can be built into the reasoning framework, leading to a proof
-system tailor-made for such stores. However, for the reasoning
-framework to be truly useful, it has to be parametric over different
-store consistency semantics, supporting stores that are weaker than
-{\sc sc} (e.g., eventually consistent), and should be able to
-reconcile conflicts between consistency and isolation
-constraints.  We demonstrate how our reasoning framework
-makes this possible in the following sections.
+immediately visible to subsequent operations. This semantics can be
+built into the reasoning framework, leading to a proof system
+tailor-made for such stores. However, for the reasoning framework to
+be truly useful, it has to be parametric over different store
+consistency semantics, supporting stores that are weaker than {\sc sc}
+(e.g., eventually consistent), and should be able to reconcile
+conflicts between consistency and isolation constraints.  We
+demonstrate how our reasoning framework makes this possible in the
+following sections.

### 2016-06-26 opsem.tex
edits to secs. 1 - 4

@@
-\section{\txnimp}
+\section{\txnimp: Syntax and Semantics}
+\label{sec:opsem}
+
 \label{sec:syntax}
-
-\subsection{\txnimp: Syntax and Semantics}
-\label{sec:opsem}
 
@@
 Fundamental to our development is the notion of a trace invariant
-($\I$). $\I$ is a function from traces ($\E$) to propositions
-($\texttt{Prop}$) that define well-formedness constraints over traces.
-The machine takes a step only if the resulting trace satisfies the
-constraints imposed by $\I$. This behaviour is captured by the
-auxiliary reduction rule \rulelabel{E-Aux} that factors out the trace
-extension aspect of the evaluation by abstracting away the
-operation-specific behaviour as a function that generates an
-appropriate effect. We let $\mathcal{F}$ denote this function.
-\rulelabel{E-Aux} uses $\mathcal{F}$ to generate a new effect and
-extend the trace ($\E = (\A,\visZ)$) \emph{only if} the
-well-formedness constraints imposed by $\I$ on $\E$ (i.e., $\I(\E')$)
-are satisfied. Otherwise, it gets stuck. In an execution that runs to
-completion, every small-step preserves the well-formedness of a trace,
-thus ensuring the invariance of $\I$.  Note that the semantics makes
-no assumptions about $\I$ other than its type. As such, it can be
-instantiated with any trace-parametric proposition that expresses
-constraints over the given trace. For instance, consider the
-$\psi_{RC}$ specification from \S\ref{sec:motivation}, but with
-bounded $T_1$ and $T_2$ instantiated with \C{Wd1} and \C{Wd2},
-respectively. The instantiated specification is the following
-proposition:
+($\I$). $\I$ is a function from traces ($\E$) to first-order logical
+formula ($\texttt{Prop}$) that define well-formedness
+constraints over traces.  The machine takes a step only if the
+resulting trace satisfies the constraints imposed by $\I$. This
+behaviour is captured by the auxiliary reduction rule
+\rulelabel{E-Aux} that factors out the trace extension aspect of the
+evaluation by abstracting away the operation-specific behaviour as a
+function that generates an appropriate effect. We let $\mathcal{F}$
+denote this function.  \rulelabel{E-Aux} uses $\mathcal{F}$ to
+generate a new effect and extend the trace ($\E = (\A,\visZ)$)
+\emph{only if} the well-formedness constraints imposed by $\I$ on $\E$
+(i.e., $\I(\E')$) are satisfied. Otherwise, it gets stuck. In an
+execution that runs to completion, every small-step preserves the
+well-formedness of a trace, thus ensuring the invariance of $\I$.
+Note that the semantics makes no assumptions about $\I$ other than its
+type. As such, it can be instantiated with any trace-parametric
+proposition that expresses constraints over the given trace. For
+instance, consider the $\psi_{RC}$ specification from
+\S\ref{sec:motivation}, but with bounded $T_1$ and $T_2$ instantiated
+with \C{Wd1} and \C{Wd2}, respectively. The instantiated specification
+is the following term:
 
@@
 
-\noindent It is easy to intepret the above proposition in the context of a trace
-$\E$ that captures an execution of the program in
-Fig.~\ref{fig:motiv-eg-1}. Such a trace-parametric proposition can be
+\noindent It is easy to interpret the above specification in the
+context of a trace $\E$ that captures an execution of the program in
+Fig.~\ref{fig:motiv-eg-1}. Such a trace-parametric formula can be
 used to instantiate the trace invariant $\I$ in Fig.~\ref{fig:txnimp}.
@@
 its return value ($\rval(\eta_2)$). For example, a read operation over
-\C{X} may pick the value ($\rval$) of the write effect with highest id
-among the visible effects (this is made possible by appropriately
+\C{X} may pick the value ($\rval$) of the write effect with the highest id
+among its visible effects (this is made possible by appropriately
 defining $\interp{\cdot}$ in \rulelabel{E-Var}, as we show later).
@@
 non-deterministic order of evaluating the parallel composition of
-transactions. A more notable source of non-determinism, however, is
-the \rulelabel{E-Aux} rule, which allows the machine to expose an
+transactions.
+
+A more notable source of non-determinism, however, is the
+\rulelabel{E-Aux} rule, which allows the machine to expose an
 arbitrary subset ($S$) of existing effects ($\A$) to the incoming
@@
 relaxation allows the abstract machine to model the semantics of
-weakly consistent data stores. For instance, operations issued to an
-eventually consistent (EC) replicated store could be dispatched to
-different replicas whose states may not be in any well-defined
+weakly-consistent data stores. For instance, operations issued to an
+eventually consistent ({\sc ec}) replicated store could be dispatched
+to different replicas whose states may not be in any well-defined
 relationship. By allowing operations to witness arbitrary subsets of
 the global state, the semantics models the weak visibility properties
-of such stores. Stronger visiblity properties needed for stronger
-isolation can however be recovered by imposing well-formedness
-constraints over $\visZ$ via the trace invariant ($\I$). Since the
-abstract machine is obligated to satisfy $\I$ at every step of the
-execution, operations are guaranteed to experience the level of
-isolation specified by $\I$.  Thus, in executions that run to
-completion, the abstract machine models a store that provides the
-required levels of isolation.  Notably, the machine achieves this
-without defining an operation semantics for isolation levels, solely
-by relying on their declarative characterization as trace
-well-formedness constraints.  \S\ref{sec:ansi-isolation} shows the
-specification of various ANSI SQL isolation levels stated as trace
-well-formedness constraints.
+of such stores; we elaborate on the implication of this style of
+definition in \S\ref{sec:store-consistency}.  Stronger visiblity
+properties can be expressed by imposing well-formedness constraints
+over $\visZ$ via the trace invariant ($\I$). Since the abstract
+machine is obligated to satisfy $\I$ at every step of the execution,
+operations are guaranteed to experience the level of isolation
+specified by $\I$.  Thus, in executions that run to completion, the
+abstract machine models a store that provides the required levels of
+isolation.  Notably, the machine achieves this without defining an
+operation semantics for isolation levels, solely by relying on their
+declarative characterization as trace well-formedness constraints.
+\S\ref{sec:ansi-isolation} shows the specification of various ANSI SQL
+isolation levels stated as trace well-formedness constraints.
 
@@
 $\rval$ for a write is the value being written, and for commit it is
-$\bot$. In case of a read, the value read depends on the how the read
+$\bot$. In case of a read, the value read depends on how the read
 operation chooses to interpret the given set ($S$) of visible effects.
-The interpretation may depend on the application semantics. For
-example, a monotonically increasing counter application may choose to
-let a write with the largest value determine the value of a
-read. A more straightforward interpretation has already been described
-before. To accommodate multiple interpretations, the semantics
-is made parametric over an interpretation function ($\interp{\cdot}$)
-that accepts a set of effects and a variable name, and returns the
-value associated with the variable. A straightforward interpretation
-function that chooses the last write (i.e., write with largest id) is
-shown below:
+The interpretation may depend on application semantics. For example, a
+monotonically increasing counter application may choose to let a write
+with the largest value determine the value of a read. A more
+straightforward interpretation has already been described before. To
+accommodate multiple interpretations, the semantics is made parametric
+over an interpretation function ($\interp{\cdot}$) that accepts a set
+of effects and a variable name, and returns the value associated with
+the variable. A straightforward interpretation function that chooses
+the last write (i.e., write with largest id) is shown below:
 

### 2016-06-26 reasoning.tex
edits to secs. 1 - 4

@@
 program satisfies its high-level invariants in all small-step
-executions that satisfy a chosen trace invariant\footnote{It is
-important to mind the distinction between high-level invariants
+executions that satisfy a chosen trace invariant ($\I$).\footnote{It is
+important to keep in mind the distinction between high-level invariants
 (\eg~\C{X$\ge$0}) and trace invariants (\eg~\C{RC}). The former
 constitute \emph{proof} \emph{obligations} for the programmer, whereas
-the latter capture \emph{assumptions} about the machine. } ($\I$).
-Since the trace invariant captures transaction isolation constraints,
+the latter captures \emph{assumptions} about the operational characteristics
+of transactions and the underlying data store. } 
+Since the trace invariant captures transaction-isolation constraints,
 the demonstration is a proof that the given selection of isolation
-levels for transactions is sufficient to enforce program's invariants.
+levels for transactions is sufficient to enforce a program's invariants.
 
 As mentioned earlier, our proof system is a rely-guarantee framework
-that admits compositional reasoning by abstracting away the
+that admits compositional reasoning by abstracting away 
 environmental interference as a rely relation. A conspicuous
@@
 relations relate executions (i.e., $R,\,Q\,\subseteq\,\E\times\E$),
-and pre ($P$) and post ($Q$) conditions are assertions over executions
+and pre- ($P$) and post- ($Q$) conditions as assertions over executions
 (i.e., $P,\,Q\, : \E \rightarrow \Prop$). Our development nonetheless
@@
 executions $\E$ and $\E'$ such that $\interp{\E}(X) \ge
-\interp{\E'}(X) \ge 0$.  Assertions on state are also written
+\interp{\E'}(X) \ge 0$.  Assertions on states are also written
 similarly. For instance, consider the postcondition
@@
 execution-based assertion below:
+
 \begin{smathpar}
@@
 % \end{smathpar}
-One of the conjuncts asserts that \C{Wd1} has written to \C{C} - a
-fact which cannot be deduced solely from the value of \C{C} (esp. in
+
+\noindent One of the conjuncts asserts that \C{Wd1} has written to \C{C} - a
+fact which cannot be deduced solely from the value of \C{C} (esp. in the
 presence of interference), but can be expressed as a proposition over
@@
 
-The standard rely guarantee judgment is the quintuple
-$\rg{P,R}{c}{G,Q}$, which intuitively means the following: if a
+The standard rely-guarantee judgment is a quintuple
+$\rg{P,R}{c}{G,Q}$, which informally asserts that if a
 command $c$ is executed in a state that satisfies its precondition
@@
 sextuple $\rg{\I,P,R}{c}{G,Q}$, whose semantics differs from the
-quintuple in that it requires every step of the execution to preserve
+quintuple in that it requires every step of the execution to additionally preserve
 the trace invariant $\I$ for the postcondition to be valid.  As usual,
@@
 expressions are side-effecting (they generate \C{RD} effects), and
-admit interference during their evaluation.  Therefore, like the
-commands, expressions need to be reasoned about explicitly. But,
-unlike commands, expressions evaluate to values, and the reasoning
+admit interference during their evaluation.  Therefore, like 
+commands, expressions need to be reasoned about explicitly.  In contrast to
+commands, however, expressions evaluate to values, and the reasoning
 framework should also admit assertions on such values.  We therefore
@@
 The rules that define rely-guarantee judgment are shown in
-Fig.~\ref{fig:rg-rules}. Like the standard definitions of rely
-guarantee, our definition also relies on the \emph{stability}
-condition, which requires pre and post conditions to hold despite the
-interference from concurrent threads (captured by $R$). The stability
-of pre and post conditions can be predicated on the assumption that
-the interference preserves the trace invariant $\I$. Formally:
+Fig.~\ref{fig:rg-rules}. Like standard rely-guarantee definitions,
+these definitions also require a \emph{stability} condition, which
+requires pre- and post- conditions to hold despite any interference
+from concurrent threads (captured by $R$). Stability can be predicated
+on the assumption that interference preserves the trace invariant
+$\I$. Formally:
+
 \begin{smathpar}
@@
 \end{smathpar}
-However, the assumption that the interference preserves $\I$ (or,
-dually, $\I$ withstands inteferference) needs to be justified
+
+\noindent However, the assumption that interference preserves $\I$ (or,
+dually, $\I$ withstands inteference) needs to be justified
 separately. We call this the stability requirement on $\I$:
+
 \begin{smathpar}
@@
 shared variable reads inside a transaction $T_i$. It requires $\I$ to
-be stable, and pre and post conditions to be stable relative to $\I$.
-The quantified premise effectively requires a proof that if the
+be stable, and pre- and post- conditions to be stable relative to
+$\I$.  The quantified premise effectively requires a proof that if the
 abstract machine of Fig.~\ref{fig:txnimp} takes a step starting from
@@
 the return value ($\interp{S}(X)$) of the read to satisfy the
-assertion $C$ meant for the value. \rulelabel{RG-Arith} defines the
-RG judgment for an arithmetic expression $e_1\pm e_2$ in terms of the
+assertion $C$ meant for the value. \rulelabel{RG-Arith} defines the RG
+judgment for an arithmetic expression $e_1\pm e_2$ in terms of the
 corresponding judgments for the constitutent expressions $e_1$ and
@@
 $C_2$, respectively. The rules for sequential and parallel composition
-of comands are more-or-less same as the corresponding rules in
-classical rely-guarantee, hence elided.
+of commands are essentially the same as their counterparts in a
+standard rely-guarantee formulation and hence elided.
 
@@
 
-The \rulelabel{RG-Conseq} rule defines ways to strengthen or weaken
+The \rulelabel{RG-Conseq} rule define ways to strengthen or weaken
 relations and assertions associated with the RG judgment of
 transaction-bound expressions. Similar rules exist for
-transaction-bound and top-level commands, but are not shown. In
-particular, the rules allow the precondition $P$ and the rely relation
+transaction-bound and top-level commands, but are not
+shown.\footnote{The supplementary materials provide the complete set
+  of rules.} As is the case with a standard rely-guarantee
+formulation,the rules allow the precondition $P$ and the rely relation
 $R$ to be strengthened\footnote{A stronger relation is a smaller set
-than a weaker relation, hence a stronger $R$ admits lesser
-interference than a weaker $R$}, and the postcondition $Q$ (also, $C$
-in the case of expressions) and the guarantee relation $G$ to be
-weakened. This kind of strengthening/weakening is also admitted in the
-classical RG framework. The most notable aspect of
-\rulelabel{RG-Conseq} rules however is that they allow the trace
-invariant $\I$ to be strengthened. Considering that $\I$ captures
-isolation properties, this means that a program proved correct under
-weaker isolation levels is also correct under stronger levels.
-Parametricity over the trace invariant $\I$, combined with the ability
-to strengthen $\I$ as needed, allows our proof system to admit a
-highly flexible proof strategy to prove programs correct on various
-stores. For example, programmers can \emph{define} isolation
+  than a weaker relation, hence a stronger $R$ admits fewer
+  interferences than a weaker one}, and the postcondition $Q$ (also,
+$C$ in the case of expressions) and the guarantee relation $G$ to be
+weakened.  The most notable aspect of the \rulelabel{RG-Conseq} rules
+is that they allow the trace invariant $\I$ to be
+strengthened. Considering that $\I$ captures isolation properties,
+this means that a program proven correct under weaker isolation levels
+is also correct under stronger ones.  Parametricity over the trace
+invariant $\I$, combined with the ability to strengthen $\I$ as
+needed, allows our proof system to support a highly flexible proof
+strategy to prove programs correct over various isolation
+variants. For example, programmers can \emph{define} isolation
 guarantees of their choice (by defining $\I$ appropriately) and then
-prove programs correct assuming the guarantees hold. Soundness of
+prove programs correct assuming the guarantees hold.  The soundness of
 strengthening $\I$ assures that the program can now be executed safely
-on any real store that offers same or stronger isolation guarantees
-than those assumed. The following subsection proves the soundness of
-our proof system.
+on any data store that offers isolation guarantees at least as strong
+as those assumed.
 
@@
 
-We now formalize the semantics of the rely-guarantee judgment of
-Fig.~\ref{fig:rg-rules} through a series of definitions.
-
 \begin{definition}[\bfseries Step-indexed reflexive transitive closure]
-Forall $A$, $R: A \rightarrow A \rightarrow \mathbb{P}$, and $n :
+Forall $A:\text{Type}$, $R: A \rightarrow A \rightarrow \mathbb{P}$, and $n :
 \mathbb{N}$, step-indexed reflexive transitive closure $R^n$ of $R$ is
@@
 \begin{definition}[\bfseries Interleaved step and multi-step relations]
-Interleaved step relation interleaves the thread-local reduction with
-the interference from concurrent threads captured as the rely relation
-($R$).  It is defined as following:
+An Interleaved step relation interleaves thread-local reductions with
+interference from concurrent threads captured as the rely relation
+($R$).  It is defined thus:
 \begin{smathpar}
@@
 ($\txnbox{e}_i$) and commands ($\txnbox{c}_i$) is defined similarly.
-Interleaved multi-step relation ($\stepssto{n}$) is the step-indexed
+An Interleaved multi-step relation ($\stepssto{n}$) is the step-indexed
 reflexive transitive closure of the interleaved step relation.
@@
 \label{def:rg-semantics}
-Semantics of the RG sextuple $\rg{\mathbb{I},P,R}{c}{G,Q}$ is defined
-in terms of the interleaved step relation as following:
+The semantics of the RG sextuple $\rg{\mathbb{I},P,R}{c}{G,Q}$ is defined
+in terms of the interleaved step relation thus:
 \begin{smathpar}
@@
 The first conjunct in the consequent is called the \emph{Hoare
-consequent} since it ascribes Hoare triple semantics to RG sextuple.
+consequent} since it ascribes Hoare triple semantics to an RG sextuple.
 The second conjunct, called the \emph{guarantee consequent}, uses the
@@
 every small-step of $c$, where the reduction can be interleaved by the
-interference ($R$) from concurrent threads. Semantics of the RG
+interference ($R$) from concurrent threads. The semantics of the RG
 sextuple for transaction-bound commands ($\txnbox{c}_i$) is defined
 similarly. Expressions, unlike commands, evaluate to a value $v$, and
-the semantics of their RG septuple
-($\rg{\I,P,R}{\txnbox{e}_i}{G,C,Q}$) differs slightly in that its
+the semantics of their RG septuple ($\rg{\I,P,R}{\txnbox{e}_i}{G,C,Q}$) differs slightly in that its
 Hoare consequent requires the value $v$ to satisfy the assertion $C$. 
@@
 
-It is worthwhile to note that the semantics of all RG judgments,
-including the judgments for transaction-bound terms,  make similar
-demands of the guarantee relation. Considering that transactions are
-atomic (though not isolated), at a first glance, it might seem odd
-that a transaction's guarantee is required to make explicit every step
-of its reduction. This requirement is however justifed because, in
-reality, a transaction's atomicity is predicated on the isolation
-settings of the observer. A \iso{Read Uncommitted} transaction, for
-example, can observe the internal state of a transaction, regardless
-of the latter's preferences\footnote{Thus, there is no such thing as
-atomicity; it's all about isolation.}. In the interest of modular
-verification, the transaction must therefore make its internal state
-available via its guarantee relation.
+Note that the semantics of all RG judgments, including the judgments
+for transaction-bound terms, make similar demands of the guarantee
+relation. Given that transactions are atomic (though not isolated), it
+is not immediately apparent why a transaction's guarantee is required
+to make explicit every step of its reduction. This requirement is
+justifed however because, in reality, a transaction's atomicity is
+predicated on the isolation settings of the observer. A \iso{Read
+  Uncommitted} transaction, for example, is permitted to observe the
+internal state of a transaction $T$ even if $T$ is claimed to execute
+atomically.  In the interest of modular verification, the transaction
+must therefore make its internal state available via its guarantee
+relation.
 
-Soundness of rely-guarantee judgments can now be stated
-thus\footnote{Supplementary contains a rigorous proof of the soundness
-theorem, and all relevant lemmas.}:
+\noindent The soundness of rely-guarantee judgments can be stated
+thus\footnote{Full proofs are provided in the supplementary material.}:
 \begin{theorem}[\bfseries Soundness] 
-Rely-guarantee judgments defined by the rules in
+The rely-guarantee judgments defined by the rules in
 Fig.~\ref{fig:rg-rules} are sound with respect to
@@
 \noindent In particular, if $\rg{\I,P,R}{c}{G,Q}$ can be derived using
-the rules of Fig.~\ref{fig:rg-rules}, then (a). every interleaved
+the rules of Fig.~\ref{fig:rg-rules}, then (a) every interleaved
 multi-step reduction of $c$ starting from a trace that satisfies $P$
-and $\I$, results in a trace that satisfies $Q$, and (b). the effect
+and $\I$, results in a trace that satisfies $Q$, and (b) the effect
 that every small-step of $c$ has on the trace is contained in $G$.
-Soundness of RG judgment for transaction-bound commands
+Soundness of the RG judgment for transaction-bound commands
 ($\txnbox{c}_i$) is stated similarly.  For expressions, soundness of

### 2016-06-26 ansi-isolation.tex
Figures added to 3.2. Edits done.

--- 2016-06-22 ansi-isolation.tex
@@
 
-\begin{figure*}[t]
+\begin{figure*}[t!]
 \begin{smathpar}
 \begin{array}{lcl}
-(\A,\visZ) \Vdash \eta \in T_i & \Leftrightarrow & \eta \in \A \conj \txn(\eta) = T_i\\
-(\A,\visZ) \Vdash \eta_1 \visar \eta_2 & \Leftrightarrow & \{\eta_1,\eta_2\}
+(\A,\visZ) \Vdash \eta \in T_i & \defeq & \eta \in \A \conj \txn(\eta) = T_i\\
+(\A,\visZ) \Vdash \eta_1 \visar \eta_2 & \defeq & \{\eta_1,\eta_2\}
         \subseteq \A \conj (\eta_1,\eta_2) \in \visZ\\
-(\A,\visZ) \Vdash \eta_1 \soar \eta_2 & \Leftrightarrow & \{\eta_1,\eta_2\}
+(\A,\visZ) \Vdash \eta_1 \soar \eta_2 & \defeq & \{\eta_1,\eta_2\}
         \subseteq \A \conj \txn(\eta_1)=\txn(\eta_2) \conj \id(\eta_1)
         < \id(\eta_2)\\
-\E \Vdash S \subseteq T_i & \Leftrightarrow & \forall \eta.~ \eta
+\E \Vdash S \subseteq T_i & \defeq & \forall \eta.~ \eta
         \in S \Rightarrow \underE{\eta \in T_i} \\
-\E \Vdash T_i \visar \eta & \Leftrightarrow &\forall\eta_1
+\E \Vdash T_i \visar \eta & \defeq &\forall\eta_1
         .\,(\E \Vdash \eta_1 \in T_i) \Rightarrow \E \Vdash \eta_1 \visar \eta \\
-\underE{T_i \visar T_j} & \Leftrightarrow &  \forall\eta_1,\eta_2.\,
+\underE{T_i \visar T_j} & \defeq &  \forall\eta_1,\eta_2.\,
     %\sameobj{\eta_1}{\eta_2}  \Rightarrow 
@@
     \underE{\eta_1 \visar \eta_2} \\
-\underE{T_i \invisar T_j} & \Leftrightarrow &  \forall\eta_1,\eta_2.\,
+\underE{T_i \invisar T_j} & \defeq &  \forall\eta_1,\eta_2.\,
         %\sameobj{\eta_1}{\eta_2}\Rightarrow 
@@
         \neg (\underE{\eta_1 \visar \eta_2})\\
-\underE{T_i \wrstoar X} & \Leftrightarrow & \exists\eta.~
+\underE{T_i \wrstoar X} & \defeq & \exists\eta.~
         \underE{\eta \in T_i} \conj \kind(\eta) = \C{WR}(X)\\
-\underE{\C{RMWVis}(T_j)} & \Leftrightarrow & \forall\eta_1,\eta_2.\,
+\underE{T_i \rdsfmar X} & \defeq & \exists\eta.~
+        \underE{\eta \in T_i} \conj \kind(\eta) = \C{RD}(X)\\
+\underE{T_i \usesar X} & \defeq & \underE{T_i \wrstoar X} \disj
+      \underE{T_i \rdsfmar X}\\
+\underE{\C{RMWVis}(T_j)} & \defeq & \forall\eta_1,\eta_2.\,
        \underE{\{\eta_1,\eta_2\} \subseteq T_j} \conj
@@
        \eta_2}\\
-\underE{\C{MonotonicVis}(T_j)} & \Leftrightarrow & 
+\underE{\C{MonotonicVis}(T_j)} & \defeq & 
        \underE{\C{RMWVis}(T_j)} \conj 
@@
         \eta_2} \\
-%  \C{CausalVis}(T_i) & \Leftrightarrow & 
+%  \C{CausalVis}(T_i) & \defeq & 
 %         \C{MonotonicVis}(T_i) \conj \C{RMWVis}(T_i)\\
-\underE{\C{AtomicVis}(T_j)} & \Leftrightarrow & 
+\underE{\C{AtomicVis}(T_j)} & \defeq & 
        \forall\eta_1,\eta_2.\, \neg(\underE{\eta_1 \in T_j}) \conj
@@
        \visar \eta_2}\\
-\underE{\C{CommitVis}(T_j)} & \Leftrightarrow & 
+\underE{\C{CommitVis}(T_j)} & \defeq & 
        \forall\eta_1,\eta_2.~ \neg(\underE{\eta_1 \in T_j}) \conj 
@@
         \conj \underE{\eta \visar \eta_2}\\
-% \underE{\C{TransVis}(T_j)} & \Leftrightarrow &  \forall
+% \underE{\C{TransVis}(T_j)} & \defeq &  \forall
 %        \eta_1,\eta_2,\eta_3.\, \underE{\eta_3 \in T_j} \conj
@@
 %        \eta_3} \\
-\underE{\C{SnapshotVis}(T_i,T_j)} & \Leftrightarrow &  \underE{T_i
+\underE{\C{RC}(T_j)} & \defeq & \underE{\C{AtomicVis}(T_j)} 
+        \conj \underE{\C{CommitVis}(T_j)}\\
+\underE{\C{MAV}(T_j)} & \defeq & \underE{\C{RC}(T_j)} \conj
+      \underE{\C{MonotonicVis}(T_j)} \\
+\underE{\C{SnapshotVis}(T_i,T_j)} & \defeq &  \underE{T_i
        \visar T_j} \disj \underE{T_i \invisar T_j}\\
-\underE{\C{OneWaySER}(T_i,T_j)} & \Leftrightarrow &  \underE{T_i
-       \visar T_j} \disj (\underE{T_i \invisar T_j} \conj \\
-  &   &\hspace*{1.2in}\exists \eta.~\underE{\eta\in T_i} \conj \kind(\eta) =
-       \C{COMMIT} \Rightarrow \underE{T_j \visar \eta})\\
-\underE{\C{RC}(T_j)} & \Leftrightarrow & \underE{\C{AtomicVis}(T_j)} 
-        \conj \underE{\C{CommitVis}(T_j)}\\
-\underE{\C{MAV}(T_j)} & \Leftrightarrow & \underE{\C{RC}(T_j)} \conj
-      \underE{\C{MonotonicVis}(T_j)} \\
-\underE{\C{RR}(T_j)} & \Leftrightarrow & \underE{\C{MAV}(T_j)}
+\underE{\C{RR}(T_j)} & \defeq & \underE{\C{MAV}(T_j)}
        \conj \forall T_i.\,T_i \neq T_j \Rightarrow 
@@
 % &   & \hspace*{2in}\conj  \C{SnapshotVis}(T_i,T_j)\\
-\underE{\C{SI}(T_j)} & \Leftrightarrow &  \underE{\C{RR}(T_j)}
+\underE{\C{SnapshotSER}(T_i,T_j,X)} & \defeq &  \underE{T_i
+       \visar T_j} \disj (\underE{T_i \invisar T_j} \conj \\
+  &   &\hspace*{1.2in}\exists \eta.~\underE{\eta\in T_i} \conj \kind(\eta) =
+       \C{WR(X)} \Rightarrow \underE{T_j \visar \eta})\\
+\underE{\C{SI}(T_j)} & \defeq &  \underE{\C{RR}(T_j)}
        \conj \forall T_i.\,(T_i \neq T_j \conj \exists X.\, \underE{T_i \wrstoar X} \conj 
@@
 %      \Rightarrow  \C{TotalVis}(T_i,T_j)\\
-  &   & \hspace*{2in} \Rightarrow \underE{\C{OneWaySER}(T_i,T_j)}\\
-\underE{\C{SER}(T_j)} & \Leftrightarrow & \underE{\C{MAV}(T_j)}
-       \conj \forall T_i.\,T_i \neq T_j 
-       \Rightarrow \underE{\C{OneWaySER}(T_i,T_j)}\\
-% &   & \hspace*{2in}\conj  \C{TotalVis}(T_i,T_j)\\
+  &   & \hspace*{2in} \Rightarrow \underE{\C{SnapshotSER}(T_i,T_j,X)}\\
+\underE{\C{SER}(T_j)} & \defeq & \underE{\C{RR}(T_j)}
+        \conj \forall T_i.\,(T_i \neq T_j \conj \exists X.\, 
+        \underE{T_i \wrstoar X} \conj \underE{T_j \usesar X})\\
+  &   & \hspace*{2in} \Rightarrow \underE{\C{SnapshotSER}(T_i,T_j,X)}\\
 \end{array}
@@
 $\underE{T_i \wrstoar X}$) if there exists a write-to-$X$ effect from
-$T_i$ in $\E$.
+$T_i$ in $\E$. \emph{Reads-from} is similar. $T_i$ \emph{uses} $X$ if
+it reads or writes $X$.
 
@@
 visibility thus prevents a transaction from being partially visible.
-However, atomic visiblity does not prevent an uncommitted transaction
-from being visible. To confine visibility to only committed
-transactions, we need an additional constraint that requires an effect
-($\eta_2$) of $T_j$ to also witness the commit effect of $T_i$
-whenever it witnesses some effect ($\eta_1$) of $T_i$. This additional
-constraint is captured by the $\mathtt{CommitVis}$ specification. A
-transaction $T_i$ is said to be snapshot-visible to a transaction
-$T_j$ if either it is visible to $T_j$, or it is invisible; it is
-forbidden for only a suffix (more generally, a subset) of $T_j$ to
-witness $T_i$. Observe that snapshot visibility permits $T_i$ and $T_j$ to
-execute and commit while being  oblivious of each other. One-way
-serializability proscribes this possiblity. $T_i$
-is said to be one-way-serializable w.r.t $T_j$ if either it is visible
-to $T_j$ or $T_j$ is visible to the commit effect of $T_i$.  Note that
-both $\mathtt{SnapshotVis}$ and $\mathtt{OneWaySER}$ are asymmetric
-definitions that guarantee complete visiblity or invisibility of $T_i$
-to $T_j$, but not the converse. While $\mathtt{SnapshotVis}$ provides
-no guarantees to $T_i$, $\mathtt{OneWaySER}$ guarantees that at least
-the commit effect of $T_i$ witnesses $T_j$.
+However, atomic visiblity does not prevent an uncommitted
+transaction from being visible. This is addressed by \C{CommitVis},
+which requires the commit effect of $T_i$ to be visible whenever any
+effect of $T_i$ is visible.
 
-The ANSI SQL 92 standard requires \iso{Read Committted} isolation to
-avoid the dirty reads phenomenon, which is achieved by enforcing
-$\mathtt{AtomicVis}$ and $\mathtt{CommitVis}$ guarantees. The {\sc rc}
+ANSI SQL 92 standard requires \iso{Read Committted} isolation to avoid
+the dirty reads phenomenon, which is achieved by enforcing
+$\mathtt{AtomicVis}$ and $\mathtt{CommitVis}$ guarantees. {\sc rc}
 specification is therefore a combination of these two guarantees. The
@@
 relational stores (formalized in \S\ref{sec:store-consistency})
-automatically leads to the monotonicity guarantee, which probably
-explains why {\sc rc} comes with $\mathtt{MonotonicVis}$ on such
-stores. On weakly consistent stores however, $\mathtt{AtomicVis}$ and
-$\mathtt{CommitVis}$ do not imply $\mathtt{MonotonicVis}$. A stronger
-isolation level called \iso{Monotonic Atomic View} ({\sc mav} of
+automatically leads to the monotonicity guarantee, which explains why
+{\sc rc} comes with $\mathtt{MonotonicVis}$ on such stores regardless
+of the implementation. On weakly consistent stores however,
+$\mathtt{AtomicVis}$ and $\mathtt{CommitVis}$ do not imply
+$\mathtt{MonotonicVis}$. A stronger isolation level called
+\iso{Monotonic Atomic View} ({\sc mav} of
 Fig.~\ref{fig:ansi-isolation})~\cite{bailishat,pldi15} was proposed to
@@
 {\sc rc} executions shown in Fig.~\ref{fig:rc-executions} are also
-{\sc mav}. \iso{Repeatable Read} spec ({\sc rr}) extends {\sc mav}
-with the snapshot visibility guarantee as described before.
-\iso{Snapshot Isolation} spec ({\sc si}) extends {\sc rr} with a
-one-way serializability guarantee w.r.t the transactions that perform
-conflicting writes (i.e., writes to the same shared variable).
-Finally, \iso{Serializable} isolation extends one-way
-$\mathtt{SnapshotVis}$ to all transactions, including those that do
-not conflict\footnote{Fig.~\ref{fig:ansi-isolation} presents slightly
-weaker versions of {\sc si} and {\sc ser} specs in the interest of
-clarity.}. \GK{ToDo: Figures. Need to figure out a convention to
-uniformly represent various isolation levels. Current idea: concurrent
-timelines with arrows in both directions.}
+{\sc mav}.
+
+\begin{figure}
+\centering
+\subcaptionbox {
+  {\sc rr}($T_1$), {\sc mav}($T_2$).
+  \label{fig:ansi-iso-eg-rr}
+} [
+  0.33\columnwidth
+] {
+  \includegraphics[scale=0.36]{Figures/rr-eg}
+}
+\subcaptionbox {
+  {\sc si}($T_1$), {\sc mav}($T_2$)
+  \label{fig:ansi-iso-eg-si}
+} [
+  0.33\columnwidth
+] {
+  \includegraphics[scale=0.36]{Figures/si-eg-2}
+}
+\subcaptionbox {
+  {\sc ser}($T_3$), {\sc mav}($T_4$)
+  \label{fig:ansi-iso-eg-ser}
+}{
+  \includegraphics[scale=0.32]{Figures/ser-eg}
+}
+
+\caption{Sample executions of concurrent transactions at different
+isolation levels. Solid (black) arrows indicate the timeline. Points
+on the timeline mark the time when an operation is executed. Dotted
+(blue) arrows denote $\visZ$. }
+\label{fig:ansi-iso-eg}
+\end{figure}
+
+A transaction $T_i$ is said to be snapshot-visible to a transaction
+$T_j$ if either it is visible to $T_j$, or it is invisible; it is
+forbidden for only a suffix (more generally, a subset) of $T_j$ to
+witness $T_i$. The specification of\iso{Repeatable Read} ({\sc rr})
+isolation extends {\sc mav} with the snapshot visibility guarantee.
+Observe that snapshot visibility permits $T_i$ and $T_j$ to execute
+and commit while being oblivious of each other. This scenario is
+captured in Fig.~\ref{fig:ansi-iso-eg-rr}, where transactions $T_1$
+and $T_2$, which perform conflicting writes, execute against a
+snapshot of the database and commit concurrently.
+
+\iso{Snapshot Isolation} ({\sc SI}) proscribes this possibility. If
+$T_i$ and $T_j$ both write to the same shared variable, then {\sc si}
+insists that either $T_i$ be visible to $T_j$, or $T_j$ be visible to
+the conflicting write of $T_i$ (this is captured by the auxiliary
+definition \C{SnapshotSER}). Fig.~\ref{fig:ansi-iso-eg-si} shows an
+execution where {\sc si} transaction $T_1$ witnesses a snapshot of the
+database that doesn't include $T_2$, but the conflicting write to $X$
+in $T_2$, and the subsequent operations (due to the monotonicity
+requirement of {\sc mav}) witness $T_1$. Thus, $T_2$ is aware of $T_1$
+when it committed. Lastly, \iso{Serializable} isolation extends
+$\mathtt{SnapshotSER}$ to also cover non-conflicting transactions that
+write to variables read in the current transaction.
+Fig.~\ref{fig:ansi-iso-eg-ser} shows an execution of an {\sc ser}
+transacton $T_3$ and a {\sc mav transaction} $T_4$, which write to $X$
+and $Y$, respectively. The snapshot witnessed by the {\sc ser}
+transaction $T_3$ does not include $T_4$, but the write to $Y$ in
+$T_4$, although non-conflicting, witnesses $T_1$ because $Y$ is read
+by $T_1$. Visibility includes other operations of $T_4$ because of
+{\sc mav}.
+
+% Note that both $\mathtt{SnapshotVis}$ and $\mathtt{OneWaySER}$
+% are asymmetric definitions that guarantee complete visiblity or
+% invisibility of $T_i$ to $T_j$, but not the converse. While
+% $\mathtt{SnapshotVis}$ provides no guarantees to $T_i$,
+% $\mathtt{OneWaySER}$ guarantees that at least the commit effect of
+% $T_i$ witnesses $T_j$.
+
+% \iso{Snapshot Isolation} spec ({\sc si}) extends {\sc rr} with a
+% one-way serializability guarantee w.r.t the transactions that perform
+% conflicting writes (i.e., writes to the same shared variable).
+% \footnote{Fig.~\ref{fig:ansi-isolation} presents slightly
+% weaker versions of {\sc si} and {\sc ser} specs in the interest of
+% clarity.}. Fig.~\ref{fig:ansi-iso-eg} shows sample executions of
+% transactions $T_1$ and $T_2$. Both transactions read and write to
+% shared variables $X$ and $Y$. In the first execution, $T_2$ commits
+% while $T_1$ is still in progress, but {\sc rr} isolation prevents
+% $T_1$ from witnessing the writes of $T_2$. In the second execution,
 

### 2016-06-26 store-consistency.tex
Figures added to 3.2. Edits done.

@@
 because doing so would violate $\I_c$. However, if $\I_c$ is true,
-then the store needs to make $\eta_1$ visible to $\eta_2$ to honor its
-consistency commitment\footnote{The term
-\emph{recency}~\cite{bailisvldb} is often used in place of consistency
-to capture the best-effort nature of the {\sc sc} guarantee on
-databases.}. We generalize this approach to any $\I_s$ and $\I_c$ by
-defining a \emph{maximum visiblity principle} to determine an
-acceptable weakening of $\I_s$ in case of a conflict with $\I_c$.  The
-principle requires the weakened consistency guarantee ($\I_s'$) of the
-store to enforce all visibility relationships imposed by the actual
-consistency guarantee ($\I_s$), unless enforcing such a relationship
-violates $\I_c$. 
+then the store makes $\eta_1$ visible to $\eta_2$ to honor its
+consistency commitment\footnote{The term \emph{recency
+commitment}~\cite{bailisvldb} is often used in practice, perhaps to
+capture the best-effort nature of {\sc sc}.}. We generalize this
+approach to any $\I_s$ and $\I_c$ by defining a \emph{maximum
+visiblity principle} to determine an acceptable weakening of $\I_s$ in
+case of a conflict with $\I_c$.  The principle requires the weakened
+consistency guarantee ($\I_s'$) of the store to enforce all visibility
+relationships imposed by the actual consistency guarantee ($\I_s$),
+unless enforcing such a relationship violates $\I_c$. 
 % The formal definition of the principle is relegated
@@
 particular, it is not intended to be a guiding principle to engineer
-data stores, and need not be objectively better than similar
-principles.
+data stores.
 \end{remark}
@@
 Reasoning under weakly consistent replication has received special
-attention in the recent research~\cite{gotsmanpopl16}. Our operational
-semantics and proof system are however more general and admit
-replication as a special case. In this section, we explain how our
-reasoning approach works under weakly consistent replication by
-comparing our model with the standard model of replication.
+attention in the recent work~\cite{gotsmanpopl16}. Our operational
+semantics and proof system are general enough to admit replication as
+a special case. In this section, we explain how our reasoning approach
+works under weakly consistent replication by comparing our model with
+the standard model of replication.
 
@@
 
-\GK{Here, I plan to include the full example that we dropped from
-\S\ref{sec:motivation}.}
+\begin{figure}
+\centering
+\begin{txnimpcode}
+ $\begin{decoration}
+ P_1:\{ {\neg\committed(\C{Wd2})} \Rightarrow \C{C = k} \conj\\
+        \hspace*{0.3in}{\committed(\C{Wd2})} \Rightarrow \C{Wd2}
+        \hboar \C{Wd1} \wedge \C{Wd2} \wrstoar \C{C} \wedge \C{C = k-a2} \}
+ \end{decoration}$
+  txn$\langle$'Wd1'$\rangle${
+   $\begin{decoration}
+    \phi_1 : \{{\neg\committed(\C{Wd2})} \Rightarrow \C{C = k} \conj
+      {\committed(\C{Wd2})} \Rightarrow \C{Wd2} \wrstoar \C{C} \conj\\
+       \hspace*{0.3in}{\committed(\C{Wd2})} \wedge
+        {\C{Wd2} \hboar \C{Wd1}} 
+       \Rightarrow \C{C = k-a2} \}
+    \end{decoration}$ 
+    v1 = C
+   $\begin{decoration}
+    \phi_2 : \{{\neg\committed(\C{Wd2})} \Rightarrow \C{C = k} \wedge \C{v1 = k} \conj\\
+       \hspace*{0.3in}{\committed(\C{Wd2})} \Rightarrow \C{Wd2} \wrstoar \C{C} \conj \\
+       \hspace*{0.3in}{\committed(\C{Wd2})} \wedge
+        {\C{Wd2} \hboar \C{Wd1}} 
+       \Rightarrow \C{C = k-a2} \wedge \C{v1 = k-a2}\}
+    \end{decoration}$ 
+    if (v1 $\ge$ a1) {
+      v2 = C;
+     $\begin{decoration}
+      \phi_3 : \{{\neg\committed(\C{Wd2})} \Rightarrow \C{C = k} \wedge \C{v2 = k} \conj\\
+       \hspace*{0.3in}{\committed(\C{Wd2})} \Rightarrow \C{Wd2} \wrstoar \C{C} \conj \\
+         \hspace*{0.3in}{\committed(\C{Wd2})} \wedge
+          {\C{Wd2} \hboar \C{Wd1}} \Rightarrow \C{C = k-a2} \wedge \\
+          \hspace*{1.9in}\C{v2 = k-a2}\}
+      \end{decoration}$ 
+      v3 = v2 - a1;
+     $\begin{decoration}
+      \phi_4 : \{{\neg\committed(\C{Wd2})} \Rightarrow \C{C = k} \wedge \C{v3 = k-a1} \conj\\
+         \hspace*{0.3in}{\committed(\C{Wd2})} \Rightarrow \C{Wd2} \wrstoar \C{C} \conj \\
+         \hspace*{0.3in}{\committed(\C{Wd2})} \wedge
+          {\C{Wd2} \hboar \C{Wd1}} 
+         \Rightarrow \C{C = k-a2} \wedge \\
+         \hspace*{1.9in}\C{v3 = k-a2-a1}\}
+      \end{decoration}$ 
+      C := v3
+     $\begin{decoration}
+      \phi_5 : \{{\neg\committed(\C{Wd2})} \Rightarrow \C{C = k-a1}) 
+                \conj\\
+         \hspace*{0.3in}{\committed(\C{Wd2})} 
+                \Rightarrow \C{C = k-a1-a2}\}
+      \end{decoration}$ 
+    }
+  }
+ $\begin{decoration}
+  Q_1 : \{{\neg\committed(\C{Wd2})} \Rightarrow \C{C = k-a1}
+            \conj \committed(\C{Wd1}) \conj\\
+      \hspace*{0.12in}{\committed(\C{Wd2})} 
+          \Rightarrow \C{C = k-a1-a2} \}
+  \end{decoration}$ 
+\end{txnimpcode}
+
+\caption{`Wd1' transaction decorated with assertions}
+\label{fig:wd1-decorated}
+\end{figure}
+
+The fully decorated implementation of transaction `Wd1' is shown in
+Fig.~\ref{fig:wd1-decorated}. Decorated `Wd2' is similar. Assignment
+statements are broken down and temporary local variables (\C{v1},
+\C{v2} and \C{v3}) are introduced so as to separate shared variable
+reads and writes.  Since we reason in terms of executions, all
+assertions implicitly refer to the current execution (just as hoare
+triples implicitly refer to the current state). Also implicit is the
+invariant \C{k $\ge$ a1+a2}. Proposition $\committed(T)$ indicates
+that the transaction $T$ is already committed. Precondition ($P_1$) of
+`Wd1' accounts for the possibility of `Wd2' committing before `Wd1',
+writing \C{k-a2} to \C{C}.  Precondition of `Wd2' is similar.  Since
+neither `Wd1' nor `Wd2' are committed at the beginning, the
+precondition in Fig.~\ref{fig:motiv-eg-1} is extended to include
+$\neg\committed(\C{Wd1}) \wedge \neg\committed(\C{Wd2})$, from which
+$P_1$ follows. Once the execution is inside `Wd1', commit of `Wd2'
+($\committed(\C{Wd2})$) may mean that either `Wd2' happened before
+`Wd1', or that it committed concurrently with `Wd1'. Since SI
+proscribes the latter possibility, we only consider the case when
+$\C{Wd2} \hboar \C{Wd1}$. A proof for $\C{Wd2} \hboar \C{Wd1}$ is
+obtained subsequently, allowing us to get rid of the special case.
+
+The first obligation is to show that the assertions that decorate
+`Wd1' are indeed valid. This entails proving that, for each statement
+of `Wd1', if evaluating the statement and extending an execution that
+satisfies its precondition results in a well-formed execution, then
+the extended execution must satisfy the postcondition. The
+well-formedness assumption lets us ignore the executions that violate
+the trace invariant lead the machine to a stuck state. This is useful
+in case of the assignment statement that writes \C{v3} to \C{C}. Since
+`Wd2' also writes to \C{C}, the trace invariant ($\I$) requires
+$\C{Wd2} \hboar \C{Wd1}$ if $\committed(\C{Wd2})$ (the alternative
+$\C{Wd1} \hboar \C{COMMIT}(\C{Wd2})$ allowed by \C{SI(Wd1)} is
+impossible in this case because \C{Wd2} has already committed, while
+\C{Wd1} is still in progress). This lets us treat the special case
+$\committed(\C{Wd2})\conj \C{Wd2} \hboar \C{Wd1}$ as being equivalent
+to the general case $\committed(\C{Wd2})$, which is required to prove
+the postcondition of `Wd1'. Observe that \C{RC(Wd1)} does not justify
+this generalization, hence replacing \C{SI(Wd1)} by \C{RC(Wd1)} is not
+sufficient to prove the postcondition of `Wd1'.
+
+The second obligation is to prove that assertions are valid despite
+the interference from the the concurrent thread executing `Wd2'. The
+interference is given by the following rely relation ($R_1$):
+\begin{smathpar}
+\begin{array}{lcl}
+  R_1 & = & \{ (\E,\E') \;|\; \neg\underE{\committed(\C{Wd2})} \conj 
+        \underE{\I} \conj \E'\Vdash \I \conj\\
+%       \conj (\E'-\E) \subseteq \C{Wd2} \conj \\
+%   & & \hspace*{0.5in} \E' \Vdash \committed(\C{Wd2}) ~\Rightarrow~ \E'
+%       \Vdash \C{Wd2} \wrstoar \C{C} \conj \\ 
+%   & & \hspace*{0.5in} \underE{\committed(\C{Wd1})} \conj \E' \Vdash
+%   \committed(\C{Wd2}) \Rightarrow \C{C=k-a1-a2} \\
+    & & \hspace*{0.5in} \neg\underE{\committed(\C{Wd1})} \conj
+        \C{COMMIT(Wd2)} \in (\E'-\E) \\
+    & & \hspace*{0.8in}\Rightarrow \C{C=k-a2} \conj 
+        \E' \Vdash \C{Wd2} \wrstoar \C{C} \conj \\
+    & & \hspace*{0.5in} \underE{\committed(\C{Wd1})} \conj
+        \C{COMMIT(Wd2)} \in (\E'-\E) \\
+    & & \hspace*{0.6in}\Rightarrow \C{C=k-a1-a2} \conj
+        \E' \Vdash \C{Wd2} \wrstoar \C{C} \}\\
+\end{array}
+\end{smathpar}
+The rely relation says the following about the concurrent thread: (1).
+It may interfere only if `Wd2' is not already committed, (2). Any
+interference takes well-formed executions to well-formed executions,
+(3). If the interference commits `Wd2', then `Wd2' should have already
+written to \C{C}, and (4). The value written is either $\C{k-a1-a2}$
+or $\C{k-a1}$ depending on whether or not `Wd1' has already committed.
+Interference from `Wd2' is allowed between any two operations of
+`Wd1', and assertions are required to be valid despite the
+interference. For instance, if the execution ($\E$) immediately after
+evaluating $\C{v1 = C}$ satisfies $\phi_2$ (i.e., $\underE{\phi_2}$),
+and an interference from `Wd2' extends $\E$ to $\E'$ (i.e.,
+$R_1(\E,\E')$), then the execution $\E'$ just before evaluating the
+condition must satisfy $\phi_2$ (i.e., $\E' \Vdash \phi_2$). Indeed,
+all $\phi_i$'s do satisfy this \emph{stability} condition, thus
+lending the validity to the proof.

### 2016-06-26 store-consistency.tex
Edits.

@@
 semantics and proof system are general enough to admit replication as
-a special case. In this section, we explain how our reasoning approach
-works under weakly consistent replication by comparing our model with
-the standard model of replication.
+a special case. In this section, we explain how the standard artifacts
+of weakly consistent replication manifest in our reasoning framework.
 
-Reasoning about eventually consistent replication often involves
-reasoning in terms of \emph{replicas} of data. The primary challenge
-in this setting is to ensure that the assumptions made and guarantees
-enforced by an operation at one replica carry over to other replicas
-that merge its effects, thus preserving the overall integrity of the
-system. In~\cite{gotsmanpopl16}, this challenge is partly addressed by
+Reasoning about weakly consistent replication often involves reasoning
+in terms of \emph{replicas} of data. The primary challenge in this
+setting is to ensure that the assumptions made and guarantees enforced
+by an operation at one replica carry over to other replicas that merge
+its effects, thus preserving the overall integrity of the system.
+In~\cite{gotsmanpopl16}, this challenge is partly addressed by
 imposing restrictions on how various replica states differ, i.e., by
 fixing a system model with a stronger baseline consistency ({\sc cc})
-than {\sc ec}. This unfortunately restricts the applicability of
-reasoning approach from being extended to weaker stores such as
-Bayou~\cite{bayou} or Quelea~\cite{pldi15}. Our view of replication
-however does not explicitly involve replicas.
-Fig.~\ref{fig:ec-theirs-vs-ours} contrasts our model of weakly
-consistent replication with the conventional replica-based model.
-Under our model, the notion of a replica is subsumed by the concept of
-visibility; a replica is defined by the subset ($S$) of global state
-($\E.\A$) that an operation witnesses. Constraints over replica states
-therefore manifest as constraints over the visibility relation.  For
-example, instead of requiring the store to be causally consistent, an
-operation can request to witness a causally consistent subset of the
-state; such demands can be made via the trace invariant $\I$. For a
-precondition ($P$) of the operation to be useful, it has to be an
-assertion over every causally consistent subset of the global state.
-Since any replica that eventually executes the operation has to expose
-one such subset ($S$), the precondition is guaranteed to hold
-regardless of the replica. There is however one problem with this
-explanation; by considering subsets of just one global state, it
-ignores the fact that the global state (hence, the replica states)
-change during the execution of the operation. To account for this
+than {\sc ec}. This unfortunately restricts the reasoning approach
+from being applied to weaker stores such as Bayou~\cite{bayou} or
+Quelea~\cite{pldi15}. Our view of replication however does not
+explicitly involve replicas.  Fig.~\ref{fig:ec-theirs-vs-ours}
+contrasts our model of weakly consistent replication with the
+conventional replica-based model.  Under our model, the notion of a
+replica is subsumed by the concept of visibility; a replica is defined
+by the subset ($S$) of global state ($\E.\A$) that an operation
+witnesses. Constraints over replica states therefore manifest as
+constraints over the visibility relation.  For example, instead of
+requiring the store to be causally consistent, an operation can
+request to witness a causally consistent subset of the state; such
+demands can be made via the trace invariant $\I$. For a precondition
+($P$) of the operation to be useful, it has to be an assertion over
+every causally consistent subset of the global state.  Since any
+replica that eventually executes the operation has to expose one such
+subset ($S$), the precondition is guaranteed to hold regardless of the
+replica. There is however one problem with this explanation; by
+considering subsets of just one global state, it ignores the fact that
+the global state (hence, the replica states) change during the
+execution of the operation. To account for this
 change,~\cite{gotsmanpopl16} distinguishes between \emph{effect

### 2016-06-27 store-consistency.tex
Example added.

@@
 
+We shall now consider the proof of the example in
+Fig.~\ref{fig:motiv-eg-1} in greater detail. We assume an {\sc sc}
+store, such as a relational database, whose store-specific trace
+invaraint ($\I_s$) was shown previously. Both transactions are run at
+{\sc si} isolation, hence $\I_c$ is $\lambda\E.~\underE{\C{SI(Wd1)}}
+\conj \underE{\C{SI(Wd1)}}$. As usual, $\I$ is $\I_s \conj \I_c$.
+
 \begin{figure}
@@
         \hspace*{0.3in}{\committed(\C{Wd2})} \Rightarrow \C{Wd2}
-        \hboar \C{Wd1} \wedge \C{Wd2} \wrstoar \C{C} \wedge \C{C = k-a2} \}
+        \visar \C{Wd1} \wedge \C{Wd2} \wrstoar \C{C} \wedge \C{C = k-a2} \}
  \end{decoration}$
@@
        \hspace*{0.3in}{\committed(\C{Wd2})} \wedge
-        {\C{Wd2} \hboar \C{Wd1}} 
+        {\C{Wd2} \visar \C{Wd1}} 
        \Rightarrow \C{C = k-a2} \}
@@
        \hspace*{0.3in}{\committed(\C{Wd2})} \wedge
-        {\C{Wd2} \hboar \C{Wd1}} 
+        {\C{Wd2} \visar \C{Wd1}} 
        \Rightarrow \C{C = k-a2} \wedge \C{v1 = k-a2}\}
@@
          \hspace*{0.3in}{\committed(\C{Wd2})} \wedge
-          {\C{Wd2} \hboar \C{Wd1}} \Rightarrow \C{C = k-a2} \wedge \\
+          {\C{Wd2} \visar \C{Wd1}} \Rightarrow \C{C = k-a2} \wedge \\
           \hspace*{1.9in}\C{v2 = k-a2}\}
@@
          \hspace*{0.3in}{\committed(\C{Wd2})} \wedge
-          {\C{Wd2} \hboar \C{Wd1}} 
+          {\C{Wd2} \visar \C{Wd1}} 
          \Rightarrow \C{C = k-a2} \wedge \\
@@
 
-\caption{`Wd1' transaction decorated with assertions}
+\caption{\C{Wd1} transaction decorated with assertions}
 \label{fig:wd1-decorated}
@@
 
-The fully decorated implementation of transaction `Wd1' is shown in
-Fig.~\ref{fig:wd1-decorated}. Decorated `Wd2' is similar. Assignment
-statements are broken down and temporary local variables (\C{v1},
-\C{v2} and \C{v3}) are introduced so as to separate shared variable
-reads and writes.  Since we reason in terms of executions, all
-assertions implicitly refer to the current execution (just as hoare
-triples implicitly refer to the current state). Also implicit is the
-invariant \C{k $\ge$ a1+a2}. Proposition $\committed(T)$ indicates
-that the transaction $T$ is already committed. Precondition ($P_1$) of
-`Wd1' accounts for the possibility of `Wd2' committing before `Wd1',
-writing \C{k-a2} to \C{C}.  Precondition of `Wd2' is similar.  Since
-neither `Wd1' nor `Wd2' are committed at the beginning, the
-precondition in Fig.~\ref{fig:motiv-eg-1} is extended to include
-$\neg\committed(\C{Wd1}) \wedge \neg\committed(\C{Wd2})$, from which
-$P_1$ follows. Once the execution is inside `Wd1', commit of `Wd2'
-($\committed(\C{Wd2})$) may mean that either `Wd2' happened before
-`Wd1', or that it committed concurrently with `Wd1'. Since SI
-proscribes the latter possibility, we only consider the case when
-$\C{Wd2} \hboar \C{Wd1}$. A proof for $\C{Wd2} \hboar \C{Wd1}$ is
-obtained subsequently, allowing us to get rid of the special case.
+The fully decorated implementation of transaction \C{Wd1} is shown in
+Fig.~\ref{fig:wd1-decorated}. Decorated \C{Wd2} is similar and not
+shown. Assignment statements are broken down and temporary local
+variables (\C{v1}, \C{v2} and \C{v3}) are introduced so as to separate
+shared variable reads and writes. All assertions implicitly refer to
+the current execution ($\E$), just as hoare triples implicitly refer
+to the current state. The context for propositions is also the
+implicit, i.e., we write $\psi$ instead of $\underE{\psi}$. The
+proposition \C{k $\ge$ a1+a2} remains an invariant, hence elided. The
+precondition ($P_1$) of \C{Wd1} accounts for the possibility of
+\C{Wd2} committing before \C{Wd1}, writing \C{k-a2} to \C{C}.
+Precondition of \C{Wd2} is similar.  Since neither \C{Wd1} nor \C{Wd2}
+are committed at the beginning, the execution-based assertion
+corresponding to the precondition ($P$) of Fig.~\ref{fig:motiv-eg-1}
+must include $\neg\committed(\C{Wd1}) \wedge \neg\committed(\C{Wd2})$,
+from which $P_1$ follows. Once the execution is inside \C{Wd1}, commit
+of \C{Wd2} ($\committed(\C{Wd2})$) may mean that either \C{Wd2}
+happened before \C{Wd1} (hence can be visible to all of \C{Wd1}), or
+that it committed concurrently with \C{Wd1} (hence cannot be visible
+to all of \C{Wd2}).  Since SI proscribes the latter possibility, we
+only consider the case when $\C{Wd2} \visar \C{Wd1}$. A proof for
+$\C{Wd2} \visar \C{Wd1}$ is obtained subsequently, allowing us to get
+rid of the special case.
 
-The first obligation is to show that the assertions that decorate
-`Wd1' are indeed valid. This entails proving that, for each statement
-of `Wd1', if evaluating the statement and extending an execution that
-satisfies its precondition results in a well-formed execution, then
-the extended execution must satisfy the postcondition. The
-well-formedness assumption lets us ignore the executions that violate
-the trace invariant lead the machine to a stuck state. This is useful
-in case of the assignment statement that writes \C{v3} to \C{C}. Since
-`Wd2' also writes to \C{C}, the trace invariant ($\I$) requires
-$\C{Wd2} \hboar \C{Wd1}$ if $\committed(\C{Wd2})$ (the alternative
-$\C{Wd1} \hboar \C{COMMIT}(\C{Wd2})$ allowed by \C{SI(Wd1)} is
-impossible in this case because \C{Wd2} has already committed, while
-\C{Wd1} is still in progress). This lets us treat the special case
-$\committed(\C{Wd2})\conj \C{Wd2} \hboar \C{Wd1}$ as being equivalent
-to the general case $\committed(\C{Wd2})$, which is required to prove
-the postcondition of `Wd1'. Observe that \C{RC(Wd1)} does not justify
-this generalization, hence replacing \C{SI(Wd1)} by \C{RC(Wd1)} is not
-sufficient to prove the postcondition of `Wd1'.
+First, we focus on the sequential aspect of the proof and show that
+the assertions that decorate \C{Wd1} are indeed valid. The proof for
+each triple follows from the rule for the corresponding term in
+Fig.~\ref{fig:rg-rules}. For illustration, we consider the triple
+corresponding to the assignment statement (line ??). The precondition
+($\phi_4$) asserts that \C{Wd2} writes to \C{C}, and the value of the
+temporary variable \C{v3} is \C{k-a2-a1}. The RG rule
+(\rulelabel{RG-Asgn}) for assignment statement allows us to easiliy
+conclude the following about the execution state after the assignment:
+\begin{smathpar}
+\begin{array}{l}
+    {\neg\committed(\C{Wd1}) \conj \C{Wd1} \wrstoar \C{C} 
+    \conj (\neg\committed(\C{Wd2})} \Rightarrow \C{C = k}) \conj\\
+    ({\committed(\C{Wd2})} \Rightarrow \C{Wd2} \wrstoar
+    \C{C}) \conj \\
+    ({\committed(\C{Wd2})} \wedge {\C{Wd2} \visar \C{Wd1}} 
+    \Rightarrow \C{C = k-a2} \wedge \C{C = k-a2-a1})
+\end{array}
+\end{smathpar}
+The rule also lets us assume that the execution satisfies trace
+invariant ($\I$), which asserts {\sc si} for both transactions.  Since
+both transactions write to \C{C}, {\sc si} requires either $\C{Wd1}
+\visar \C{Wd2}$, or $\C{Wd2} \visar \C{Wd1}$. Since \C{Wd1} is not yet
+committed ($\neg\committed(\C{Wd1})$), {\sc si} prohibits the former
+possibility, allowing us the deduce the $\C{Wd2} \visar \C{Wd1}$. This
+lets us derive ${\committed(\C{Wd2})}  \Rightarrow \C{C = k-a2} \wedge
+\C{C = k-a2-a1}$, allowing us to prove the postcondition ($\phi_5$).
 
-The second obligation is to prove that assertions are valid despite
-the interference from the the concurrent thread executing `Wd2'. The
-interference is given by the following rely relation ($R_1$):
+The second part of the proof is to show that assertions are stable
+despite the interference from the the concurrent thread executing
+\C{Wd2}. The interference is given by the following rely relation
+($R_1$):
 \begin{smathpar}
@@
 The rely relation says the following about the concurrent thread: (1).
-It may interfere only if `Wd2' is not already committed, (2). Any
+It may interfere only if \C{Wd2} is not already committed, (2). Any
 interference takes well-formed executions to well-formed executions,
-(3). If the interference commits `Wd2', then `Wd2' should have already
+(3). If the interference commits \C{Wd2}, then \C{Wd2} should have already
 written to \C{C}, and (4). The value written is either $\C{k-a1-a2}$
-or $\C{k-a1}$ depending on whether or not `Wd1' has already committed.
-Interference from `Wd2' is allowed between any two operations of
-`Wd1', and assertions are required to be valid despite the
-interference. For instance, if the execution ($\E$) immediately after
-evaluating $\C{v1 = C}$ satisfies $\phi_2$ (i.e., $\underE{\phi_2}$),
-and an interference from `Wd2' extends $\E$ to $\E'$ (i.e.,
-$R_1(\E,\E')$), then the execution $\E'$ just before evaluating the
-condition must satisfy $\phi_2$ (i.e., $\E' \Vdash \phi_2$). Indeed,
-all $\phi_i$'s do satisfy this \emph{stability} condition, thus
-lending the validity to the proof.
+or $\C{k-a1}$ depending on whether or not \C{Wd1} has already committed.
+Stability for most assertions is straightforward. The only non-trivial
+proof is for $\phi_5$, where {\sc si} condition should be used to
+show that any interference from \C{Wd2} is invalid. The proof proceeds
+on the same lines as the Hoare proof discussed above; it is also
+discussed in \S\ref{sec:motivation}.
+
+The final aspect of the proof is to show that any interference from
+\C{Wd1} is contained in its guarantee relation $G_1$, which becomes
+the rely relation ($R_2$) for \C{Wd2}. $R_2$ (hence $G_1$) is same as
+$R_1$ shown above, but with \C{Wd2} and \C{Wd1} interchanged. $\ldots$

### 2016-06-27 reasoning.tex
edits

@@
 
-\noindent The soundness of rely-guarantee judgments can be stated
-thus\footnote{Full proofs are provided in the supplementary material.}:
 \begin{theorem}[\bfseries Soundness] 
@@
 Fig.~\ref{fig:rg-rules} are sound with respect to
-Definition~\ref{def:rg-semantics}. 
+the semantics of Definition~\ref{def:rg-semantics}.\footnote{Full proofs are provided in the supplementary material.}
 \end{theorem}

### 2016-06-27 store-consistency.tex
edits

@@
 
-The abstract machine of Fig.~\ref{fig:txnimp} allows operations to
-witness arbitrary subsets of the global state, thus assuming the
-semantics of an eventually consistent ({\sc ec}) data
-store\footnote{{\sc ec} guarantees that in the absence of further
-updates, all reads witness the same global state \emph{eventually}. In
-any finite trace, however, there are no guarantees on what a read
-witnesses.}. There are however stores, such as relational databases,
-that provide stronger consistency guarantees than {\sc ec}. Like the
-isolation levels of transactions, the consistency level of the
-underlying store also affects the semantics of a program in
-non-trivial ways. In this section, we demonstrate how the semantics of
-stronger stores with on-demand weak isolation can be captured in our
-operational model so that we can use the reasoning framework developed
-in \S\ref{sec:reasoning} to reason about such stores.
+The operational semantics of Fig.~\ref{fig:txnimp} allows operations
+to witness arbitrary subsets of the global state, effectively mimicing the
+behaviour of an eventually consistent ({\sc ec}) data
+store\footnote{Eventual consistency guarantees that in the absence of further
+  updates, all reads witness the same global state
+  \emph{eventually}. In any finite trace, however, there are no
+  guarantees on what a read may witness.}. There are, however, data stores,
+such as relational databases, that provide stronger consistency
+guarantees than {\sc ec}. Like the isolation levels of transactions,
+the consistency level of the underlying store also affects the
+semantics of a program in non-trivial ways. In this section, we
+demonstrate how the semantics of stronger stores with on-demand weak
+isolation can be captured in our operational model.
 
@@
 
-Firstly, we observe that the semantics of stronger stores can be
-captured by store-specific consistency constraints, along with
+First, we observe that the semantics of a weakly-consistent data store
+can be captured by store-specific consistency constraints, along with
 transaction-specific isolation constraints, via the trace invariant
-$\I$. In particular, we split $\I$ into two components: (1).  $\I_s$,
-the store-specific invariant, and (2). $\I_c$, the program-specific
-(or, client-specific) invariant, to capture consistency and isolation
-constraints, respectively. $\I$ is now a conjunction:
-  $\I \,=\, \lambda\E.~\I_s(\E) \wedge \I_c(\E)$ (often simplified to
-  $\I \,=\, \I_s\wedge\I_c$).
-While the program-specific trace invariant ($\I_c$) and remains an
-invariant regardless of the store, the store specific invariant
-($\I_s$) changes from store to store depending on the consisteny
-level. $\I_s$ for an {\sc ec} store is $true$. Stronger stores, such
-as an {\sc sc} store, have non-trivial $\I_s$.
+$\I$. In particular, we can split $\I$ into two components: (1).
+$\I_s$, the store-specific invariant, and (2). $\I_c$, the
+program-specific (or, client-specific) invariant, to capture
+consistency and isolation constraints, resp.  $\I$ is now a
+conjunction: $\I \,=\, \lambda\E.~\I_s(\E) \wedge \I_c(\E)$ (often
+simplified to $\I \,=\, \I_s\wedge\I_c$).  While the program-specific
+trace invariant ($\I_c$) remains an invariant regardless of the
+specific consistency and visibility features exposed by the underlying
+data store, the store specific invariant ($\I_s$) changes from store
+to store depending on the consisteny level.  For an eventually
+consistent store, $\I_s$ is simply \emph{true}. Stronger stores, such
+as those that support strong consistenty, have non-trivial definitions
+for $\I_s$.
 
@@
 \begin{array}{l}
-  \C{SC}(\E) \;=\; \forall\eta_1,\eta_2.~\{\eta_1,\eta_2\}
+  \C{sc}(\E) \;=\; \forall\eta_1,\eta_2.~\{\eta_1,\eta_2\}
   \subseteq \E.\A \conj \id(\eta_1) < \id(\eta_2) \\
@@
 violates $\I_c$, whereas not letting it be visible violates $\I_{s}$.
-The only way to satisfy both the invariants is to rule out all the
+The only way to satisfy both invariants is to rule out all 
 executions that interleave the operations of one transaction with the
-other, thereby enforcing serializability \footnote{Thus,
-serializability is a natural generalization of {\sc sc} to
-transactions.}. In general, when $\I_s$ is in conflict (but not
+other, thereby enforcing serializability.  In general, when $\I_s$ conflicts (but is not
 inconsistent) with $\I_c$, the only way to enforce both invariant sets
@@
 
-In practice, relational databases resolve the conflict by prioritizing
-weak isolation (thus, concurrency and performance) over strong
-consistency, so the execution traces do not necessarily satisfy {\sc
-sc}. In particular, visibility constraints imposed by {\sc sc} are
-violated if (and only if) they are found to be in conflict with the
-constraints imposed by the isolation level. In the context of
-aforementioned example, $\eta_1$ is not made visible to $\eta_2$
+In practice, relational databases resolve such conflicts by
+prioritizing weak isolation (thus, concurrency and performance) over
+strong consistency, so the execution traces do not necessarily satisfy
+{\sc sc}. In particular, visibility constraints imposed by {\sc sc}
+are violated iff they are found to be in conflict with the constraints
+imposed by a transaction isolation level. In the context of
+the aforementioned example, $\eta_1$ is not made visible to $\eta_2$
 because doing so would violate $\I_c$. However, if $\I_c$ is true,
@@
 consistency commitment\footnote{The term \emph{recency
-commitment}~\cite{bailisvldb} is often used in practice, perhaps to
-capture the best-effort nature of {\sc sc}.}. We generalize this
+    commitment}~\cite{bailisvldb} is often used in practice
+  to capture the best-effort nature of {\sc sc}.}. We generalize this
 approach to any $\I_s$ and $\I_c$ by defining a \emph{maximum
-visiblity principle} to determine an acceptable weakening of $\I_s$ in
-case of a conflict with $\I_c$.  The principle requires the weakened
-consistency guarantee ($\I_s'$) of the store to enforce all visibility
-relationships imposed by the actual consistency guarantee ($\I_s$),
-unless enforcing such a relationship violates $\I_c$. 
+  visiblity principle} to determine an acceptable weakening of $\I_s$
+in case of a conflict with $\I_c$.  The principle requires the
+weakened consistency guarantee ($\I_s'$) of the store to enforce all
+visibility relationships imposed by the actual consistency guarantee
+($\I_s$), unless enforcing such a relationship violates $\I_c$.
 % The formal definition of the principle is relegated
@@
 Fig.~\ref{fig:txnimp} results in an operational semantics that
-describes relational database-like stores.
+describes many kinds of relational database-like stores.
 
-Semantics of a causally consistent data
-store~\cite{gotsmanpopl16,LBC16} can also be obtained: by defining
+For example, the semantics of a causally consistent data
+store~\cite{gotsmanpopl16,LBC16} can also be obtained by defining
 $\I_s$ to be the causal consistency ({\sc cc}) invariant that is
-appropriately weakened to resolve conflicts with $\I_c$. Details are
-relegated to the supplementary in the interest of space.
+appropriately weakened to resolve conflicts with $\I_c$.\footnote{Details
+of this instantiation are relegated to the supplementary in the interest of space.}
 
-\begin{remark}
-Note that the purpose of maximum visibility principle is to
-rationalize the observed behaviour of databases in practice. In
-particular, it is not intended to be a guiding principle to engineer
-data stores.
-\end{remark}
+%% \begin{remark}
+%% Note that the purpose of maximum visibility principle is to
+%% rationalize the observed behaviour of databases in practice. In
+%% particular, it is not intended to be a guiding principle to engineer
+%% data stores.
+%% \end{remark}
 
@@
 Reasoning under weakly consistent replication has received special
-attention in the recent work~\cite{gotsmanpopl16}. Our operational
+attention in recent work~\cite{gotsmanpopl16}. Our operational
 semantics and proof system are general enough to admit replication as
 a special case. In this section, we explain how the standard artifacts
-of weakly consistent replication manifest in our reasoning framework.
+of weakly-consistent replication manifest in our reasoning framework.
 
-Reasoning about weakly consistent replication often involves reasoning
-in terms of \emph{replicas} of data. The primary challenge in this
+Reasoning about weakly-consistent replication involves reasoning in
+terms of \emph{replicas} of data. The primary challenge in this
 setting is to ensure that the assumptions made and guarantees enforced
 by an operation at one replica carry over to other replicas that merge
-its effects, thus preserving the overall integrity of the system.
+their effects, thus preserving the overall integrity of the system.
 In~\cite{gotsmanpopl16}, this challenge is partly addressed by
@@
 than {\sc ec}. This unfortunately restricts the reasoning approach
-from being applied to weaker stores such as Bayou~\cite{bayou} or
-Quelea~\cite{pldi15}. Our view of replication however does not
-explicitly involve replicas.  Fig.~\ref{fig:ec-theirs-vs-ours}
-contrasts our model of weakly consistent replication with the
-conventional replica-based model.  Under our model, the notion of a
-replica is subsumed by the concept of visibility; a replica is defined
-by the subset ($S$) of global state ($\E.\A$) that an operation
-witnesses. Constraints over replica states therefore manifest as
-constraints over the visibility relation.  For example, instead of
-requiring the store to be causally consistent, an operation can
-request to witness a causally consistent subset of the state; such
-demands can be made via the trace invariant $\I$. For a precondition
-($P$) of the operation to be useful, it has to be an assertion over
-every causally consistent subset of the global state.  Since any
-replica that eventually executes the operation has to expose one such
-subset ($S$), the precondition is guaranteed to hold regardless of the
-replica. There is however one problem with this explanation; by
-considering subsets of just one global state, it ignores the fact that
-the global state (hence, the replica states) change during the
-execution of the operation. To account for this
-change,~\cite{gotsmanpopl16} distinguishes between \emph{effect
-generation} event at one replica and \emph{effect merge} event at
+from being applied to stores that provide guarantees weaker than
+causal consistency (e.g., \cite{bayou,pldi15}).
+
+Notably, our view of replication however does not explicitly involve
+replicas.  Fig.~\ref{fig:ec-theirs-vs-ours} contrasts our model of
+weakly-consistent replication with a conventional replica-based
+model.  Under our model, the notion of a replica is subsumed by the
+concept of visibility; a replica is defined by the subset ($S$) of
+global state ($\E.\A$) that an operation witnesses. Constraints over
+replica states therefore manifest as constraints over a specific visibility
+relation.  For example, instead of requiring the store to be causally
+consistent, an operation can request to witness a causally consistent
+subset of the state; such demands can be made via the trace invariant
+$\I$. For a precondition ($P$) of the operation to be useful, it has
+to be an assertion over every causally consistent subset of the global
+state.  Since any replica that eventually executes the operation has
+to expose one such subset ($S$), the precondition is guaranteed to
+hold regardless of the replica. There is however one problem with this
+explanation - by considering subsets of just one global state, it
+ignores the fact that the global state (hence, the replica states)
+change during the execution of the operation. To account for this
+change,~\cite{gotsmanpopl16} distinguishes between an \emph{effect
+  generation} event at one replica and an \emph{effect merge} event at
 another replica, and requires that certain invariants be preserved
 between these two events at different replicas. Our framework folds
-all of this into a stability condition (\S\ref{sec:rely-guarantee}).
+all such machinery into a stability condition (\S\ref{sec:rely-guarantee}).
 Since any change to the global state during the execution of the
@@
 ($R_1$):
+
 \begin{smathpar}
@@
 \end{smathpar}
-The rely relation says the following about the concurrent thread: (1).
+
+\noindent The rely relation says the following about the concurrent thread: (1).
 It may interfere only if \C{Wd2} is not already committed, (2). Any

### 2016-06-28 related-work.tex
Related work done.

\section{Related Work}

\paragraph{Specifications of weak isolation}
Adya~\cite{adyaphd} specified several isolation levels in terms of
\emph{dependency graphs} between transactions, by specifying what
dependency cycles are forbidden in each case. The operational nature
of Adya's specifications make them suitable for runtime monitoring and
anamoly detection~\cite{kemmevldb,feketesigmod08,pssi2011}, whereas
declarative nature of our specifications make them suitable for
reasoning about programs. Sivaramakrishnan et al.~\cite{pldi15} specify
isolation levels declaratively as trace well-formedness conditions,
but their specifications implicitly assume a complete trace with only
committed transactions, hence not suitable to operationally reason
about a program as it builds the trace. For instance, their {\sc rc}
specification allows the effects of a not-yet-committed transaction to
be visible to an {\sc rc} transaction.  Nonetheless, our
specifications belong to the same fragment of first-order logic as
theirs, which is known to be decidable. Cerone et
al.~\cite{gotsmanconcur15} present a framework to specify isolation
levels with atomic visibility, but their specifications are also for
the complete traces.  Moreover, their specifications for {\sc si} and
{\sc ser} implicitly assume homogenity, i.e., they require all
transactions to execute at {\sc si} (resp.  {\sc ser}), which is often
not the case in practice. Like ours, all the aforementioned
specification frameworks, use the vocabulary introduced
in~\cite{burckhardt14} to specify replicated data types and forms of
eventual consistency. However, none of them come with an associated
reasoning framework that can use such specifications productively
towards verifying programs under weak isolation.

\paragraph{Reasoning under weak isolation} In~\cite{feketessi}, Fekete
et al. propose a theory to characterize non-serializable executions
that arise under {\sc si}. Fekete~\cite{fekete2005} also proposes an
algorithm that allocates either {\sc si} or {\sc ser} isolation levels
to transactions while guaranteeing the serializability of the
execution. In~\cite{gotsmanpodc16}, Cerone et al. improve on Adya's
{\sc si} specification and use it to derive a static analysis that
determines the safety of substituting {\sc si} with a weaker variant
called \iso{Parallel Snapshot Isolation}~\cite{psi}. All of them focus
on establishing the equivalence of executions between a pair of
isolation levels, without taking application invariants into account.
Bernstein et al.~\cite{bern2000} propose informal semantic conditions
to ensure the satisfaction of application invariants under weaker
isolation levels.  However, no formal proof system has been proposed.
Furthermore, all the aforementioned approaches are tailor-made for a
finite set of well-understood isolation levels (rooted
in~\cite{berenson}) under a pre-defined consistency model (usually
{\sc sc}).

\paragraph{Reasoning under weak consistency} There have been many
recent proposals focusing on reasoning about programs under weak
consistency~\cite{bailisvldb, alvarocalm, gotsmanpopl16,redblueatc,
redblueosdi, ecinec}. All of them assume a system model that offers a
choice between a \emph{coordination-free} weak consistency level ({\sc
ec} in~\cite{redblueosdi, redblueatc, ecinec, alvarocalm, bailisvldb};
{\sc cc}\footnote{It is worth noting that, although causal consistency
does not require global coordination, it is nonetheless not
\emph{highly available}~\cite{bailishat}.} in~\cite{gotsmanpopl16})
and strong consistency ({\sc sc}), which requires coordination.  They
propose various conditions (\eg \emph{invariant
confluence}~\cite{bailisvldb}, \emph{logical
monotonicity}~\cite{alvarocalm}, \emph{invariant
safety}~\cite{redblueatc}), or proof systems (\eg~rely-guarantee
framework of~\cite{gotsmanpopl16}) to show that an allotment of atomic
and fully isolated operations to these levels preserves application
invariants.  In contrast, we admit weakly isolated transactions (sets
of atomic operations), and our system model accepts
\emph{specifications} of consistency and isolation levels drawn from
an expressive logic. Gotsman et al.~\cite{gotsmanpopl16} adapt
\iso{Parallel Snapshot Isolation} to the aforementioned setting by
interpreting it as a consistency level that serializes writes to
objects. They subsequently propose a dedicated proof rule to prove
program invaraints under this model. By parameterizing our proof
system on isolation specifications, we avoid the need to define a
separate proof rule for each new isolation level.

\paragraph{Reasoning under relaxed memory} Considering the
correspondences between axiomatic models of relaxed
memory~\cite{battycpp} and weakly consistent
replication~\cite{burckhardt14}, it is reasonable to expect similar
reasoning approaches to work in both the cases. Ridge~\cite{rgtso}
generalizes rely-guarantee reasoning to x86-TSO relaxed memory model.
Likewise, Vafeadis et al.~\cite{rsl13} generalize concurrent
separation logic (CSL)~\cite{csl} to C11 relaxed memory model.
Ferreira et al.~\cite{ferreira10} propose parameterized operational
semantics for relaxed memory models, but the parameterization is over
a relation between relaxed memory programs and related {\sc sc}
programs. They also prove the soundness of CSL for the weakest memory
model expressible in their semantics, but neither a parameterized (in
the aforementioned sense) CSL, nor a CSL specialized for any relaxed
memory model is proposed. In fact, to the best of our knowledge, there
does not exist a proof system that combines axiomatic semantics of
memory models and operational semantics of programs in a unified
reasoning framework.

### 2016-06-28 ansi-isolation.tex

@@
 
-ANSI SQL 92 standard requires \iso{Read Committted} isolation to avoid
+The ANSI SQL 92 standard requires \iso{Read Committted} isolation to avoid
 the dirty reads phenomenon, which is achieved by enforcing
-$\mathtt{AtomicVis}$ and $\mathtt{CommitVis}$ guarantees. {\sc rc}
+$\mathtt{AtomicVis}$ and $\mathtt{CommitVis}$ guarantees. The {\sc rc}
 specification is therefore a combination of these two guarantees. The
@@
 Fig.~\ref{fig:ansi-isolation})~\cite{bailishat,pldi15} was proposed to
-explicitly extend {\sc rc} with monotonicity on such stores. Sample
+explicitly extend {\sc rc} with monotonicity on such stores. The sample
 {\sc rc} executions shown in Fig.~\ref{fig:rc-executions} are also
@@
 forbidden for only a suffix (more generally, a subset) of $T_j$ to
-witness $T_i$. The specification of\iso{Repeatable Read} ({\sc rr})
+witness $T_i$.  \iso{Repeatable Read} ({\sc rr})
 isolation extends {\sc mav} with the snapshot visibility guarantee.
@@
 requirement of {\sc mav}) witness $T_1$. Thus, $T_2$ is aware of $T_1$
-when it committed. Lastly, \iso{Serializable} isolation extends
+when it commits. Lastly, \iso{Serializable} isolation extends
 $\mathtt{SnapshotSER}$ to also cover non-conflicting transactions that

### 2016-06-28 related-work.tex
Edits.

@@
 ec} in~\cite{redblueosdi, redblueatc, ecinec, alvarocalm, bailisvldb};
-{\sc cc}\footnote{It is worth noting that, although causal consistency
-does not require global coordination, it is nonetheless not
-\emph{highly available}~\cite{bailishat}.} in~\cite{gotsmanpopl16})
-and strong consistency ({\sc sc}), which requires coordination.  They
-propose various conditions (\eg \emph{invariant
-confluence}~\cite{bailisvldb}, \emph{logical
+{\sc cc} in~\cite{gotsmanpopl16}) and strong consistency ({\sc sc}),
+which requires coordination.  They propose various conditions (\eg
+\emph{invariant confluence}~\cite{bailisvldb}, \emph{logical
 monotonicity}~\cite{alvarocalm}, \emph{invariant

### 2016-06-28 store-consistency.tex
Related work done.
Edits.

@@
 consistency commitment\footnote{The term \emph{recency
-    commitment}~\cite{bailisvldb} is often used in practice
+    commitment}~\cite{bailishat} is often used in practice
   to capture the best-effort nature of {\sc sc}.}. We generalize this
@@
 \subsection{Weakly Consistent Replication}
+\label{sec:replication}
 
@@
 than {\sc ec}. This unfortunately restricts the reasoning approach
-from being applied to stores that provide guarantees weaker than
-causal consistency (e.g., \cite{bayou,pldi15}).
+from being applied to stores (e.g., \cite{bayou,pldi15}) that provide
+guarantees weaker than causal consistency, such as causal visibility
+or read-my-writes~\cite{zoo}. Causal consistency is not baseline
+consistency in these stores because it is not \emph{highly
+available}~\cite{bailishat}.
 
@@
 change during the execution of the operation. To account for this
-change,~\cite{gotsmanpopl16} distinguishes between an \emph{effect
-  generation} event at one replica and an \emph{effect merge} event at
-another replica, and requires that certain invariants be preserved
-between these two events at different replicas. Our framework folds
-all such machinery into a stability condition (\S\ref{sec:rely-guarantee}).
-Since any change to the global state during the execution of the
-operation is an interference, and $P$ is required to be stable with
-respect to any such interference, it follows that $P$ is valid on
-every replica, thus ensuring that assumptions made at the generation
-event are also valid at the merge event.
+change,~\cite{gotsmanpopl16} distinguishes between effect generation
+event at one replica $r_1$ and effect merge event at replica $r_2$,
+and requires that only \emph{non-conflicting} operations execute
+between these two events at $r_2$, and that they preserve certain
+invariants. Our framework folds all such machinery into a stability
+condition predicated on $\I$ (\S\ref{sec:rely-guarantee}).  Since any
+change to the global state during the execution of the operation is an
+interference, and $P$ is required to be stable with respect to any
+such interference, it follows that $P$ is valid on every replica, thus
+ensuring that assumptions made at the generation event are also valid
+at the merge event.
 

### 2016-06-28 store-consistency.tex
Example done.

@@
 $\begin{decoration}
- P_1:\{ {\neg\committed(\C{Wd2})} \Rightarrow \C{C = k} \conj\\
-        \hspace*{0.3in}{\committed(\C{Wd2})} \Rightarrow \C{Wd2}
-        \visar \C{Wd1} \wedge \C{Wd2} \wrstoar \C{C} \wedge \C{C = k-a2} \}
+  \hspace*{0.3in}P : \{\neg\committed(\C{Wd1}) \conj \neg\committed(\C{Wd2}) \conj
+                          \C{C} = \C{k}\}
+\end{decoration}$
+
+ $\begin{decoration}
+ P_1:\{ \neg\committed(\C{Wd1}) \conj
+        (\neg\committed(\C{Wd2}) \Rightarrow \C{C = k}) \conj\\
+        \hspace*{0.3in}{\committed(\C{Wd2})} \Rightarrow 
+            \C{Wd2} \wrstoar \C{C} \wedge \C{C = k-a2} \}
  \end{decoration}$
   txn$\langle$'Wd1'$\rangle${
-   $\begin{decoration}
-    \phi_1 : \{{\neg\committed(\C{Wd2})} \Rightarrow \C{C = k} \conj
-      {\committed(\C{Wd2})} \Rightarrow \C{Wd2} \wrstoar \C{C} \conj\\
-       \hspace*{0.3in}{\committed(\C{Wd2})} \wedge
-        {\C{Wd2} \visar \C{Wd1}} 
-       \Rightarrow \C{C = k-a2} \}
-    \end{decoration}$ 
-    v1 = C
-   $\begin{decoration}
-    \phi_2 : \{{\neg\committed(\C{Wd2})} \Rightarrow \C{C = k} \wedge \C{v1 = k} \conj\\
-       \hspace*{0.3in}{\committed(\C{Wd2})} \Rightarrow \C{Wd2} \wrstoar \C{C} \conj \\
-       \hspace*{0.3in}{\committed(\C{Wd2})} \wedge
-        {\C{Wd2} \visar \C{Wd1}} 
-       \Rightarrow \C{C = k-a2} \wedge \C{v1 = k-a2}\}
-    \end{decoration}$ 
-    if (v1 $\ge$ a1) {
-      v2 = C;
-     $\begin{decoration}
-      \phi_3 : \{{\neg\committed(\C{Wd2})} \Rightarrow \C{C = k} \wedge \C{v2 = k} \conj\\
-       \hspace*{0.3in}{\committed(\C{Wd2})} \Rightarrow \C{Wd2} \wrstoar \C{C} \conj \\
-         \hspace*{0.3in}{\committed(\C{Wd2})} \wedge
-          {\C{Wd2} \visar \C{Wd1}} \Rightarrow \C{C = k-a2} \wedge \\
-          \hspace*{1.9in}\C{v2 = k-a2}\}
-      \end{decoration}$ 
-      v3 = v2 - a1;
+              $\vdots$ 
      $\begin{decoration}
      \phi_4 : \{{\neg\committed(\C{Wd2})} \Rightarrow \C{C = k} \wedge \C{v3 = k-a1} \conj\\
-         \hspace*{0.3in}{\committed(\C{Wd2})} \Rightarrow \C{Wd2} \wrstoar \C{C} \conj \\
-         \hspace*{0.3in}{\committed(\C{Wd2})} \wedge
+        \hspace*{0.3in}({\committed(\C{Wd2})} \Rightarrow \C{Wd2}
+        \wrstoar \C{C}) \conj \neg\committed(\C{Wd1}) \conj \\
+        \hspace*{0.32in}{\committed(\C{Wd2})} \wedge
          {\C{Wd2} \visar \C{Wd1}} 
@@
      $\begin{decoration}
-      \phi_5 : \{{\neg\committed(\C{Wd2})} \Rightarrow \C{C = k-a1}) 
-                \conj\\
-         \hspace*{0.3in}{\committed(\C{Wd2})} 
-                \Rightarrow \C{C = k-a1-a2}\}
+      \phi_5 : \{({\neg\committed(\C{Wd2})} \Rightarrow \C{C = k-a1}) 
+          \conj \neg\committed(\C{Wd1}) \conj\\
+         \hspace*{0.4in}{\committed(\C{Wd2})} 
+            \Rightarrow \C{Wd2} \wrstoar \C{C} \wedge \C{C = k-a1-a2}\}
       \end{decoration}$ 
-    }
+              $\vdots$ 
   }
  $\begin{decoration}
-  Q_1 : \{{\neg\committed(\C{Wd2})} \Rightarrow \C{C = k-a1}
+  Q_1 : \{{(\neg\committed(\C{Wd2})} \Rightarrow \C{C = k-a1})
             \conj \committed(\C{Wd1}) \conj\\
-      \hspace*{0.12in}{\committed(\C{Wd2})} 
+      \hspace*{0.4in}{\committed(\C{Wd2})} 
           \Rightarrow \C{C = k-a1-a2} \}
+  \end{decoration}$
+
+$\begin{decoration}
+  \hspace*{0.3in}Q : \{\committed(\C{Wd1}) \conj \committed(\C{Wd2}) \conj
+                          \C{C} = \C{k-a1-a2}\}
 \end{decoration}$
@@
 
-The fully decorated implementation of transaction \C{Wd1} is shown in
-Fig.~\ref{fig:wd1-decorated}. Decorated \C{Wd2} is similar and not
-shown. Assignment statements are broken down and temporary local
-variables (\C{v1}, \C{v2} and \C{v3}) are introduced so as to separate
-shared variable reads and writes. All assertions implicitly refer to
-the current execution ($\E$), just as hoare triples implicitly refer
-to the current state. The context for propositions is also the
-implicit, i.e., we write $\psi$ instead of $\underE{\psi}$. The
-proposition \C{k $\ge$ a1+a2} remains an invariant, hence elided. The
-precondition ($P_1$) of \C{Wd1} accounts for the possibility of
-\C{Wd2} committing before \C{Wd1}, writing \C{k-a2} to \C{C}.
-Precondition of \C{Wd2} is similar.  Since neither \C{Wd1} nor \C{Wd2}
-are committed at the beginning, the execution-based assertion
-corresponding to the precondition ($P$) of Fig.~\ref{fig:motiv-eg-1}
-must include $\neg\committed(\C{Wd1}) \wedge \neg\committed(\C{Wd2})$,
-from which $P_1$ follows. Once the execution is inside \C{Wd1}, commit
-of \C{Wd2} ($\committed(\C{Wd2})$) may mean that either \C{Wd2}
-happened before \C{Wd1} (hence can be visible to all of \C{Wd1}), or
-that it committed concurrently with \C{Wd1} (hence cannot be visible
-to all of \C{Wd2}).  Since SI proscribes the latter possibility, we
-only consider the case when $\C{Wd2} \visar \C{Wd1}$. A proof for
-$\C{Wd2} \visar \C{Wd1}$ is obtained subsequently, allowing us to get
-rid of the special case.
+Decorated parts of \C{Wd1} relevant to the discussion are shown in
+Fig.~\ref{fig:wd1-decorated}.  \C{Wd2} is similar and not
+shown\footnote{Supplementary contains fully decorated program.}.
+Temporary name \C{v3} has been introduced to capture the result of the
+RHS expression \C{C-a2}.  Assertions implicitly refer to the current
+execution ($\E$), just as hoare triples implicitly refer to the
+current state.  The context for propositions is also the implicit,
+i.e., we write $\psi$ instead of $\underE{\psi}$. The proposition \C{k
+$\ge$ a1+a2} remains an invariant, hence elided. The precondition
+$P_1$ of \C{Wd1} accounts for the possibility of \C{Wd2} committing
+before \C{Wd1}, writing \C{k-a2} to \C{C}. Since neither \C{Wd1} nor
+\C{Wd2} are committed at the beginning, the precondition $P$ of the
+program is extended with $\neg\committed(\C{Wd1}) \wedge
+\neg\committed(\C{Wd2})$, from which $P_1$ follows. The postcondition
+$Q_1$ of \C{Wd1} asserts different values for \C{C} depending on
+whether or not \C{Wd2} has already committed.  It also asserts that
+\C{Wd1} has committed. The postcondition $Q_2$ of \C{Wd2} (not shown)
+is similar, but with the roles of \C{Wd1} and \C{Wd2} exchanged. The
+postcondition $Q$ follows from $Q_1 \wedge Q_2$, hence valid. It
+remains to show that \C{Wd1} satisfies its specification (\C{Wd2} must
+also satisfy its spec, but the proof is similar).
 
 First, we focus on the sequential aspect of the proof and show that
-the assertions that decorate \C{Wd1} are indeed valid. The proof for
-each triple follows from the rule for the corresponding term in
-Fig.~\ref{fig:rg-rules}. For illustration, we consider the triple
-corresponding to the assignment statement (line ??). The precondition
-($\phi_4$) asserts that \C{Wd2} writes to \C{C}, and the value of the
-temporary variable \C{v3} is \C{k-a2-a1}. The RG rule
-(\rulelabel{RG-Asgn}) for assignment statement allows us to easiliy
-conclude the following about the execution state after the assignment:
+the assertions that decorate \C{Wd1} are indeed valid. We consider the
+triple associated with the assignment to \C{C} for illustration. The
+precondition $\phi_4$ asserts different values for \C{C} and \C{v3}
+(temporary alias of RHS expression) depending on whether or not
+\C{Wd2} is committed. Inside \C{Wd1}, $\committed(\C{Wd2})$ may mean
+that either \C{Wd2} happened before \C{Wd1} (hence can be visible to
+all of \C{Wd1}), or that it committed concurrently with \C{Wd1} (hence
+cannot be visible to all of \C{Wd2}).  Nonetheless, $\phi_4$ only
+considers the case when $\C{Wd2} \visar \C{Wd1}$. As we show below,
+this is sufficient to prove the postcondition $\phi_5$, which appears
+to make a stronger claim about the value of \C{C}. The proof follows
+from the RG rule \rulelabel{RG-Asgn} for assignments, which allows us
+to easiliy conclude the following about the execution state after the
+assignment:
 \begin{smathpar}
@@
     {\neg\committed(\C{Wd1}) \conj \C{Wd1} \wrstoar \C{C} 
-    \conj (\neg\committed(\C{Wd2})} \Rightarrow \C{C = k}) \conj\\
+    \conj (\neg\committed(\C{Wd2})} \Rightarrow \C{C = k-a1}) \conj\\
     ({\committed(\C{Wd2})} \Rightarrow \C{Wd2} \wrstoar
@@
 invariant ($\I$), which asserts {\sc si} for both transactions.  Since
-both transactions write to \C{C}, {\sc si} requires either $\C{Wd1}
-\visar \C{Wd2}$, or $\C{Wd2} \visar \C{Wd1}$. Since \C{Wd1} is not yet
-committed ($\neg\committed(\C{Wd1})$), {\sc si} prohibits the former
-possibility, allowing us the deduce the $\C{Wd2} \visar \C{Wd1}$. This
-lets us derive ${\committed(\C{Wd2})}  \Rightarrow \C{C = k-a2} \wedge
-\C{C = k-a2-a1}$, allowing us to prove the postcondition ($\phi_5$).
+both transactions now write to \C{C}, {\sc si} requires either
+$\C{Wd1} \visar \C{Wd2}$, or $\C{Wd2} \visar \C{Wd1}$. Since \C{Wd1}
+is not yet committed ($\neg\committed(\C{Wd1})$), {\sc si} (which
+subsumes {\sc rc}) prohibits the former possibility, allowing us the
+deduce $\C{Wd2} \visar \C{Wd1}$. This lets us derive
+${\committed(\C{Wd2})}  \Rightarrow \C{C = k-a2} \wedge \C{C =
+k-a2-a1}$, allowing us to prove the postcondition $\phi_5$.
 
@@
 despite the interference from the the concurrent thread executing
-\C{Wd2}. The interference is given by the following rely relation
-($R_1$):
+\C{Wd2}. We formalize the interference as a rely relation that
+describes the semantics of \C{Wd2} at just the right level of
+abstraction. In particular, we would like to capture two important
+facts about \C{Wd2}: 
+\begin{itemize}
+  \item  If \C{Wd2} commits, then it should have already written to
+  \C{C}. This fact is needed to show that \C{Wd2} conflicts with
+  \C{Wd1}, so that \C{SI(Wd1)} property can be exploited to deduce
+  $\C{Wd2} \visar \C{Wd1}$ from $\committed(\C{Wd2})$ in the
+  sequential proof.
+
+  \item The value written to \C{C} is either $\C{k-a1-a2}$ or
+  $\C{k-a1}$ depending on whether or not \C{Wd1} has already
+  committed. This is clearly necessary to deduce the value of \C{C}
+  before and after \C{Wd1} commits.
+\end{itemize}
+ The resultant rely relation ($R_1$) is shown below (note: we
+write $(\E'-\E) \Vdash {\committed(\C{Wd2})}$ to express that \C{Wd2}
+is not committed in $\E$, but committed in $\E'$.):
 
@@
 \begin{array}{lcl}
-  R_1 & = & \{ (\E,\E') \;|\; \neg\underE{\committed(\C{Wd2})} \conj 
-        \underE{\I} \conj \E'\Vdash \I \conj\\
+  R_1 & = & \{ (\E,\E') \;|\; \neg\underE{\committed(\C{Wd1})} \conj
+        (\E'-\E) \Vdash {\committed(\C{Wd2})}\\
+%       \neg\underE{\committed(\C{Wd2})} \conj \I(\E) \conj \I(\E')\\
 %       \conj (\E'-\E) \subseteq \C{Wd2} \conj \\
@@
 %   \committed(\C{Wd2}) \Rightarrow \C{C=k-a1-a2} \\
-    & & \hspace*{0.5in} \neg\underE{\committed(\C{Wd1})} \conj
-        \C{COMMIT(Wd2)} \in (\E'-\E) \\
+%   & & \hspace*{0.5in} \neg\underE{\committed(\C{Wd1})} \conj
+%       (\E'-\E) \Vdash {\committed(\C{Wd2})}\\
     & & \hspace*{0.8in}\Rightarrow \C{C=k-a2} \conj 
-        \E' \Vdash \C{Wd2} \wrstoar \C{C} \conj \\
-    & & \hspace*{0.5in} \underE{\committed(\C{Wd1})} \conj
-        \C{COMMIT(Wd2)} \in (\E'-\E) \\
+        \E' \Vdash \C{Wd2} \wrstoar \C{C} \\
+    & & \hspace*{0.37in} \conj \underE{\committed(\C{Wd1})} \conj
+        (\E'-\E) \Vdash {\committed(\C{Wd2})}\\
+%       \C{COMMIT(Wd2)} \in (\E'.\A-\E.\A) \\
     & & \hspace*{0.6in}\Rightarrow \C{C=k-a1-a2} \conj
-        \E' \Vdash \C{Wd2} \wrstoar \C{C} \}\\
+        \E' \Vdash \C{Wd2} \wrstoar \C{C} \\
+    & & \hspace*{0.37in}\conj \forall(\eta \in \E'.\A - 
+        \E.\A).~\txn(\eta) = \C{Wd2}\}\\
 \end{array}
 \end{smathpar}
+For the sake of completeness, we also explicitly represent the fact
+that all the effects added by $R_1$ are of \C{Wd2}. This is
+needed to prove that an interference from \C{Wd2} does effect the
+local state of \C{Wd1}.
 
-\noindent The rely relation says the following about the concurrent thread: (1).
-It may interfere only if \C{Wd2} is not already committed, (2). Any
-interference takes well-formed executions to well-formed executions,
-(3). If the interference commits \C{Wd2}, then \C{Wd2} should have already
-written to \C{C}, and (4). The value written is either $\C{k-a1-a2}$
-or $\C{k-a1}$ depending on whether or not \C{Wd1} has already committed.
-Stability for most assertions is straightforward. The only non-trivial
-proof is for $\phi_5$, where {\sc si} condition should be used to
-show that any interference from \C{Wd2} is invalid. The proof proceeds
-on the same lines as the Hoare proof discussed above; it is also
-discussed in \S\ref{sec:motivation}.
+The stability for $\phi_4$ w.r.t $R_1$ is trivial to establish: if
+$R_1$ does not commit \C{Wd2}, then the value of \C{C} is unaffected,
+and if it commits \C{Wd2}, then $\C{Wd2}\not\visar\C{Wd1}$, hence
+$\phi_4$ doesn't care. For $\phi_5$, {\sc si} condition should be used
+to show that any interference from \C{Wd2} is invalid. The proof
+proceeds on the same lines as the Hoare proof discussed above; it is
+also discussed in detail in \S\ref{sec:motivation}.
 
 The final aspect of the proof is to show that any interference from
-\C{Wd1} is contained in its guarantee relation $G_1$, which becomes
-the rely relation ($R_2$) for \C{Wd2}. $R_2$ (hence $G_1$) is same as
-$R_1$ shown above, but with \C{Wd2} and \C{Wd1} interchanged. $\ldots$
+\C{Wd1} is contained in its guarantee relation $G_1$, which is also
+the rely relation $R_2$ for \C{Wd2}. Since the implementation of
+\C{Wd2} is same as \C{Wd1}, $R_2$ (hence $G_1$) is same as $R_1$ shown
+above, but with \C{Wd2} and \C{Wd1} interchanged. Continuing our focus
+on the assigment statement, the guarantee proof requires us to show
+that if executing the assignment takes the execution from $\E$ to
+$\E'$, such that $\E'$ satisfies $\I$, then $(\E,\E') \in G_1$.  The
+proof follows from the \rulelabel{RG-Asgn} rule, which establishes
+that (a). the assignment statement does not commit \C{Wd2}, and (b).
+it adds a write effect $\eta$ such that $\txn(\eta)=\C{Wd1}$. This
+finishes the proof for the triple in focus. The proofs for other
+triples of \C{Wd1} can also be constructed similarly. The combination
+of correctness proofs for \C{Wd1} and \C{Wd2} yeilds a correctness
+proof for the program.

### 2016-06-29 store-consistency.tex
Edit.
edits to store-consistency

--- 2016-06-28 store-consistency.tex
@@
 The operational semantics of Fig.~\ref{fig:txnimp} allows operations
-to witness arbitrary subsets of the global state, effectively mimicing the
+to witness arbitrary subsets of the global state, effectively mimicking the
 behaviour of an eventually consistent ({\sc ec}) data
@@
 
-A strongly-consistent {\sc sc} store guarantees a total order of all
+A strongly-consistent {\sc sc} store guarantees a total order on all
 operations w.r.t $\visZ$ consistent with their chronological order. A
@@
 are violated iff they are found to be in conflict with the constraints
-imposed by a transaction isolation level. In the context of
+imposed by a transaction's isolation level. In the context of
 the aforementioned example, $\eta_1$ is not made visible to $\eta_2$
@@
     Fig.~\ref{fig:txnimp} results in an operational semantics that
-describes many kinds of relational database-like stores.
-
-For example, the semantics of a causally consistent data
-store~\cite{gotsmanpopl16,LBC16} can also be obtained by defining
-$\I_s$ to be the causal consistency ({\sc cc}) invariant that is
-appropriately weakened to resolve conflicts with $\I_c$.\footnote{Details
-of this instantiation are relegated to the supplementary in the interest of space.}
+    describes many kinds of relational database-like data stores,
+    including causally-consistent data stores, among
+    others.\footnote{Details of this instantiation are relegated to
+      the supplementary in the interest of space.}
 
@@
 
-Reasoning under weakly consistent replication has received special
+Reasoning under weakly-consistent replication has received special
 attention in recent work~\cite{gotsmanpopl16}. Our operational
 semantics and proof system are general enough to admit replication as
-a special case. In this section, we explain how the standard artifacts
-of weakly-consistent replication manifest in our reasoning framework.
+a special case of our formulation. In this section, we explain how the
+standard artifacts of weakly-consistent replication manifest in our
+reasoning framework.
 
-Reasoning about weakly-consistent replication involves reasoning in
-terms of \emph{replicas} of data. The primary challenge in this
-setting is to ensure that the assumptions made and guarantees enforced
-by an operation at one replica carry over to other replicas that merge
-their effects, thus preserving the overall integrity of the system.
-In~\cite{gotsmanpopl16}, this challenge is partly addressed by
+The primary challenge in this setting is to ensure that the
+assumptions made and guarantees enforced by an operation at one
+replica carry over to other replicas that merge their effects, thus
+preserving the overall integrity of the system.  In prior
+work~\cite{lbc16,gotsmanpopl16}, this challenge is partly addressed by
 imposing restrictions on how various replica states differ, i.e., by
@@
 than {\sc ec}. This unfortunately restricts the reasoning approach
-from being applied to stores (e.g., \cite{bayou,pldi15}) that provide
+from being applied to data stores (e.g., \cite{bayou,pldi15}) that provide
 guarantees weaker than causal consistency, such as causal visibility
@@
 
-Notably, our view of replication however does not explicitly involve
-replicas.  Fig.~\ref{fig:ec-theirs-vs-ours} contrasts our model of
-weakly-consistent replication with a conventional replica-based
-model.  Under our model, the notion of a replica is subsumed by the
-concept of visibility; a replica is defined by the subset ($S$) of
-global state ($\E.\A$) that an operation witnesses. Constraints over
-replica states therefore manifest as constraints over a specific visibility
-relation.  For example, instead of requiring the store to be causally
-consistent, an operation can request to witness a causally consistent
-subset of the state; such demands can be made via the trace invariant
-$\I$. For a precondition ($P$) of the operation to be useful, it has
-to be an assertion over every causally consistent subset of the global
-state.  Since any replica that eventually executes the operation has
-to expose one such subset ($S$), the precondition is guaranteed to
-hold regardless of the replica. There is however one problem with this
+Notably, our view of replication does not explicitly involve replicas.
+Fig.~\ref{fig:ec-theirs-vs-ours} contrasts our model of
+weakly-consistent replication with a conventional replica-based model.
+Under our model, the notion of a replica is subsumed by the concept of
+visibility; a replica is defined by the subset ($S$) of global state
+($\E.\A$) that an operation witnesses. Constraints over replica states
+therefore manifest as constraints over a specific visibility relation.
+For example, instead of requiring the store to be causally consistent,
+an operation can witness a causally consistent subset of the state;
+such demands can be made via the trace invariant $\I$. For a
+precondition ($P$) of the operation to be useful, it has to be an
+assertion over every causally consistent subset of the global state.
+Since any replica that eventually executes the operation has to expose
+one such subset ($S$), the precondition is guaranteed to hold
+regardless of the replica. There is however one problem with this
 explanation - by considering subsets of just one global state, it
 ignores the fact that the global state (hence, the replica states)
-change during the execution of the operation. To account for this
-change,~\cite{gotsmanpopl16} distinguishes between effect generation
+change during the execution of the operation. To account for such
+changes, we might choose to distinguish between effect generation
 event at one replica $r_1$ and effect merge event at replica $r_2$,
-and requires that only \emph{non-conflicting} operations execute
-between these two events at $r_2$, and that they preserve certain
-invariants. Our framework folds all such machinery into a stability
-condition predicated on $\I$ (\S\ref{sec:rely-guarantee}).  Since any
-change to the global state during the execution of the operation is an
-interference, and $P$ is required to be stable with respect to any
-such interference, it follows that $P$ is valid on every replica, thus
-ensuring that assumptions made at the generation event are also valid
-at the merge event.
+requiring that \emph{non-conflicting} operations execute between these
+two events at $r_2$, and that they preserve certain
+invariants~\cite{gotsmanpopl16}.  Instead, our framework folds all such
+machinery into a stability condition predicated on $\I$
+(\S\ref{sec:rely-guarantee}).  Since any change to the global state
+during the execution of the operation is an interference, and $P$ is
+required to be stable with respect to any such interference, it
+follows that $P$ is valid on every replica, thus ensuring that
+assumptions made at a generation event is also valid at the merge
+event.
 
@@
 conclude the following about the execution state after the assignment:
+
 \begin{smathpar}
@@
 \end{smathpar}
-The rule also lets us assume that the execution satisfies trace
+
+\noindent The rule also lets us assume that the execution satisfies trace
 invariant ($\I$), which asserts {\sc si} for both transactions.  Since

### 2016-06-29 store-consistency.tex

--- 2016-06-28 store-consistency.tex
@@
 assignment:
+>>>>>>> 70ae6b11c8f648c323413d6fe8f8fb1c2cba75ae
 \begin{smathpar}
@@
 For the sake of completeness, we also explicitly represent the fact
-that all the effects added by $R_1$ are of \C{Wd2}. This is
-needed to prove that an interference from \C{Wd2} does effect the
-local state of \C{Wd1}.
+that all the effects added by $R_1$ are of \C{Wd2}. This is needed to
+prove that an interference from \C{Wd2} does not affect the local
+state of \C{Wd1}.
 

### 2016-06-29 abstract.tex
edits

@@
   serializable transactions to permit greater concurrency.
-  Unfortunately, the semantics of weak isolation is often poorly
-  understood, and usually explained only informally in terms of
-  low-level implementation artifacts.  Consequently, verifying
-  high-level correctness properties in such environments is
-  problematic.
+  Unfortunately, the semantics of weak isolation is poorly understood,
+  and usually explained only informally in terms of low-level
+  implementation artifacts.  Consequently, verifying high-level
+  correctness properties in such environments is problematic.
 

### 2016-06-29 introduction.tex
edits

@@
 comprehend~\cite{berenson}. To quantify weak isolation anomalies,
-Fekete \emph{et al.}~\cite{feketevldb09} have devised and experimented
+Fekete \emph{et al.}~\cite{feketevldb09} devised and experimented
 with a microbenchmark suite that executes transactions under a
@@
   redblueatc, ecinec, gotsmanpopl16} address broadly related concerns.
-At a high level, reasoning about weak consistency involves reasoning
+At a high-level, reasoning about weak consistency involves reasoning
 about the concurrent execution of strongly atomic and fully isolated
@@
   \item We develop a semantics for a core language equipped with
-    weakly isolated transactions, demonstrating that a general
-    isolation semantics can be expressed as well-formedness constraints
-    on a program's execution.
+    weakly-isolated transactions, demonstrating that a general
+    parameteric isolation semantics can be expressed as
+    well-formedness constraints on a program's execution.
   \item We propose a compositional proof system for this language
@@
     of traces induced by the operational semantics, in which transactions
-    are parameterized with specific weak isolation levels.
+    are associated with specific weak-isolation levels.
   \item We define a \emph{maximum visibility principle} to reconcile
@@
 guarantees, thereby allowing weakly-isolated transactions to enjoy the
-same rigorous reasoning capabilities as their strongly isolated
+same rigorous reasoning capabilities as their strongly-isolated
 (serializable) counterparts.
@@
 a core language that supports weakly-isolated transactions,
-parameterized over different isolation notions. \S\ref{sec:proofrules}
-formalizes the proof systemt that we use to reason about program
+parameterized over different isolation notions. \S\ref{sec:reasoning}
+formalizes the proof system that we use to reason about program
 invariants, and establishes the soundness of these rules with respect

### 2016-06-29 motivation.tex
edits

@@
 
-In this section, we motivate our proof system for weak isolation via
-an example written in \txnimp - a C-like imperative language equipped
-with a \C{txn} lexical block defining a transaction scope. Each
+In this section, we motivate our ideas via an example written in
+\txnimp - a C-like imperative language equipped with a \C{txn} lexical
+block defining a transaction scope.  For exposition convenience, each
 \C{txn} block is associated with a single-quoted string in angle
@@
 effects of both withdraws. The pre- and post-conditions in
-Fig.~\ref{fig:motiv-eg-1} reflect our expectations. Indeed, invariants
+Fig.~\ref{fig:motiv-eg-1} reflect this expectation. Indeed, invariants
 are guaranteed to hold if both withdraw transactions are serialized,
@@
 and re-executing a transaction if write-write conflicts are detected
-during its commit.\footnote{{\sc si} is not functionally the same as
-  {\sc ser}, however; the differences in the two manifest in the \emph{absence}
-  of any write-write conflicts.}   Since {\sc si},
+during its commit.\footnote{{\sc si} exhibits different behaviour from
+  {\sc ser} in the absence of write-write conflicts.}   Since {\sc si},
 unlike {\sc ser}, does not rely on expensive lock-based
@@
 isolation variants are implementated is no better since it requires
-programmers to understand low-level implementation details
-of the database that are naturally far removed from the application
-semantics. An attractive (and principled) alternative in this context
-is a formal proof system that combines declarative reasoning about
+programmers to understand low-level implementation details of the
+database that are naturally far removed from application semantics. An
+attractive (and principled) alternative in this context would be a
+formal proof system that combines declarative reasoning about
 isolation guarantees with operational reasoning about programs. We
@@
 allows us to reason about the behaviour of individual threads by
-abstracting away interference induced by other threads (collectively
+abstracting away interferences induced by other threads (collectively
 called \emph{the environment}) into a \emph{rely} relation. In an
@@
 The specification of an isolation level encodes its interference
-characteristic as constraints over the $\hbZ$ relation. However, for
+characteristics as constraints over the $\hbZ$ relation. However, for
 this to be useful in reasoning about programs, the rely-guarantee
@@
 consistency semantics, supporting stores that are weaker than {\sc sc}
-(e.g., eventually consistent), and should be able to reconcile
-conflicts between consistency and isolation constraints.  We
+(e.g., causally or eventually consistent), and should be able to
+reconcile conflicts between consistency and isolation constraints.  We
 demonstrate how our reasoning framework makes this possible in the

### 2016-06-29 opsem.tex
edits

@@
 $\txn$.  The latter identifies the transaction that generated the
-effect. In every step of the evaluation, the machine reduces a \txnimp
+effect. In every step of the evaluation, the machine reduces a \txnimp\
 term by executing a read, write or commit operation, generating an
@@
 isolation.  Notably, the machine achieves this without defining an
-operation semantics for isolation levels, solely by relying on their
-declarative characterization as trace well-formedness constraints.
+operationalsemantics for isolation levels, instead solely relying on
+their declarative characterization as trace well-formedness
+constraints to enforce isolation guarantees.
 \S\ref{sec:ansi-isolation} shows the specification of various ANSI SQL

### 2016-06-29 reasoning.tex
edits

@@
 
-The rule \rulelabel{RG-Var} defines the rely-guarantee judgment for
+\noindent The rule \rulelabel{RG-Var} defines the rely-guarantee judgment for
 shared variable reads inside a transaction $T_i$. It requires $\I$ to
@@
 
-The \rulelabel{RG-Conseq} rule define ways to strengthen or weaken
-relations and assertions associated with the RG judgment of
+The \rulelabel{RG-Conseq} rule define ways to strengthen or
+weaken relations and assertions associated with the RG judgment of
 transaction-bound expressions. Similar rules exist for
@@
 formulation,the rules allow the precondition $P$ and the rely relation
-$R$ to be strengthened\footnote{A stronger relation is a smaller set
-  than a weaker relation, hence a stronger $R$ admits fewer
-  interferences than a weaker one}, and the postcondition $Q$ (also,
-$C$ in the case of expressions) and the guarantee relation $G$ to be
-weakened.  The most notable aspect of the \rulelabel{RG-Conseq} rules
-is that they allow the trace invariant $\I$ to be
-strengthened. Considering that $\I$ captures isolation properties,
-this means that a program proven correct under weaker isolation levels
-is also correct under stronger ones.  Parametricity over the trace
-invariant $\I$, combined with the ability to strengthen $\I$ as
-needed, allows our proof system to support a highly flexible proof
-strategy to prove programs correct over various isolation
-variants. For example, programmers can \emph{define} isolation
-guarantees of their choice (by defining $\I$ appropriately) and then
-prove programs correct assuming the guarantees hold.  The soundness of
-strengthening $\I$ assures that the program can now be executed safely
-on any data store that offers isolation guarantees at least as strong
-as those assumed.
+$R$ to be strengthened, and the postcondition $Q$ (also, $C$ in the
+case of expressions) and the guarantee relation $G$ to be weakened.
+The most notable aspect of the \rulelabel{RG-Conseq} rules is that
+they allow the trace invariant $\I$ to be strengthened. Considering
+that $\I$ captures isolation properties, this means that a program
+proven correct under weaker isolation levels is also correct under
+stronger ones.  Parametricity over the trace invariant $\I$, combined
+with the ability to strengthen $\I$ as needed, allows our proof system
+to support a highly flexible proof strategy to prove programs correct
+over various isolation variants. For example, programmers can
+\emph{define} isolation guarantees of their choice (by defining $\I$
+appropriately) and then prove programs correct assuming the guarantees
+hold.  The soundness of strengthening $\I$ ensures that a program can
+be executed safely on any data store that offers isolation guarantees
+at least as strong as those assumed.
 
@@
 \begin{definition}[\bfseries Step-indexed reflexive transitive closure]
-Forall $A:\text{Type}$, $R: A \rightarrow A \rightarrow \mathbb{P}$, and $n :
-\mathbb{N}$, step-indexed reflexive transitive closure $R^n$ of $R$ is
+For all $A:\text{Type}$, $R: A \rightarrow A \rightarrow \mathbb{P}$, and $n :
+\mathbb{N}$, the step-indexed reflexive transitive closure $R^n$ of $R$ is
 the smallest relation satisfying the following
@@
 \begin{definition}[\bfseries Interleaved step and multi-step relations]
-An Interleaved step relation interleaves thread-local reductions with
+An interleaved step relation interleaves thread-local reductions with
 interference from concurrent threads captured as the rely relation
 ($R$).  It is defined thus:
+
 \begin{smathpar}
@@
 \end{smathpar}
+
 The interleaved step relation for transaction bound expressions
 ($\txnbox{e}_i$) and commands ($\txnbox{c}_i$) is defined similarly.
-An Interleaved multi-step relation ($\stepssto{n}$) is the step-indexed
+An interleaved multi-step relation ($\stepssto{n}$) is the step-indexed
 reflexive transitive closure of the interleaved step relation.
@@
 in terms of the interleaved step relation thus:
+
 \begin{smathpar}
@@
 \end{smathpar}
-The first conjunct in the consequent is called the \emph{Hoare
+
+\noindent The first conjunct in the consequent is called the \emph{Hoare
 consequent} since it ascribes Hoare triple semantics to an RG sextuple.
@@
 $\texttt{step-guaranteed}$ predicate defined below:
+
 \begin{smathpar}
@@
 \end{smathpar}
-The guarantee consequent requires $G$ to capture the trace effect of
+
+\noindent The guarantee consequent requires $G$ to capture the trace effect of
 every small-step of $c$, where the reduction can be interleaved by the
@@
 \end{theorem}
+
 \noindent In particular, if $\rg{\I,P,R}{c}{G,Q}$ can be derived using

### 2016-06-29 related-work.tex
edits

@@
 \section{Related Work}
+\label{sec:relatedwork}
 
-\paragraph{Specifications of weak isolation}
-Adya~\cite{adyaphd} specified several isolation levels in terms of
-\emph{dependency graphs} between transactions, by specifying what
-dependency cycles are forbidden in each case. The operational nature
-of Adya's specifications make them suitable for runtime monitoring and
-anamoly detection~\cite{kemmevldb,feketesigmod08,pssi2011}, whereas
+\paragraph{Specifying weak isolation.}
+Adya~\cite{adyaphd} specified several weak isolation levels in terms
+of \emph{dependency graphs} between transactions, by defining the
+kinds of dependency cycles that are forbidden in each case. The
+operational nature of Adya's specifications make them suitable for
+runtime monitoring and anamoly
+detection~\cite{kemmevldb,feketesigmod08,pssi2011}, whereas the
 declarative nature of our specifications make them suitable for
-reasoning about programs. Sivaramakrishnan et al.~\cite{pldi15} specify
-isolation levels declaratively as trace well-formedness conditions,
-but their specifications implicitly assume a complete trace with only
-committed transactions, hence not suitable to operationally reason
-about a program as it builds the trace. For instance, their {\sc rc}
-specification allows the effects of a not-yet-committed transaction to
-be visible to an {\sc rc} transaction.  Nonetheless, our
-specifications belong to the same fragment of first-order logic as
-theirs, which is known to be decidable. Cerone et
-al.~\cite{gotsmanconcur15} present a framework to specify isolation
-levels with atomic visibility, but their specifications are also for
-the complete traces.  Moreover, their specifications for {\sc si} and
-{\sc ser} implicitly assume homogenity, i.e., they require all
-transactions to execute at {\sc si} (resp.  {\sc ser}), which is often
-not the case in practice. Like ours, all the aforementioned
-specification frameworks, use the vocabulary introduced
-in~\cite{burckhardt14} to specify replicated data types and forms of
-eventual consistency. However, none of them come with an associated
-reasoning framework that can use such specifications productively
-towards verifying programs under weak isolation.
+reasoning about program behaviour. Sivaramakrishnan \emph{et
+  al.}~\cite{pldi15} specify isolation levels declaratively as trace
+well-formedness conditions, but their specifications implicitly assume
+a complete trace with only committed transactions, and is thus not
+suitable to reason operationally about a program as it builds the
+trace. Cerone \emph{et al.}~\cite{gotsmanconcur15} present a framework
+to specify isolation levels with atomic visibility, but their
+specifications are also for complete traces; Moreover, their
+formulation implicitly assumes homogenity, i.e., they require all
+transactions to execute using either {\sc si} or {\sc ser}, which is
+often not the case in practice. While all the aforementioned
+specification frameworks use the vocabulary introduced
+in~\cite{burckhardt14} to specify replicated data types and different
+forms of weak consistency, in contrast to this work, none of them come
+with an associated reasoning framework that can use such
+specifications productively towards verifying programs under weak
+isolation.
 
 \paragraph{Reasoning under weak isolation} In~\cite{feketessi}, Fekete
-et al. propose a theory to characterize non-serializable executions
+\emph{et al.} propose a theory to characterize non-serializable executions
 that arise under {\sc si}. Fekete~\cite{fekete2005} also proposes an
@@
 
-\paragraph{Reasoning under weak consistency} There have been many
-recent proposals focusing on reasoning about programs under weak
-consistency~\cite{bailisvldb, alvarocalm, gotsmanpopl16,redblueatc,
-redblueosdi, ecinec}. All of them assume a system model that offers a
-choice between a \emph{coordination-free} weak consistency level ({\sc
-ec} in~\cite{redblueosdi, redblueatc, ecinec, alvarocalm, bailisvldb};
-{\sc cc} in~\cite{gotsmanpopl16}) and strong consistency ({\sc sc}),
-which requires coordination.  They propose various conditions (\eg
-\emph{invariant confluence}~\cite{bailisvldb}, \emph{logical
-monotonicity}~\cite{alvarocalm}, \emph{invariant
-safety}~\cite{redblueatc}), or proof systems (\eg~rely-guarantee
-framework of~\cite{gotsmanpopl16}) to show that an allotment of atomic
-and fully isolated operations to these levels preserves application
-invariants.  In contrast, we admit weakly isolated transactions (sets
-of atomic operations), and our system model accepts
+\paragraph{Reasoning under weak consistency} There have been several
+recent proposals on reasoning techniques for programs executing under
+weak consistency~\cite{bailisvldb, alvarocalm,
+  gotsmanpopl16,redblueatc, redblueosdi, ecinec}. All of them assume a
+system model that offers a choice between a \emph{coordination-free}
+weak consistency level (\emph{e.g.}, {\sc ec}~\cite{redblueosdi,
+  redblueatc, ecinec, alvarocalm, bailisvldb} or causal consistency
+({\sc cc})~\cite{gotsmanpopl16}). In all these efforts, the key
+technical issue involves proving that atomic and fully isolated
+operations preserve application invariants when executed on these
+consistency levels.  In contrast, we admit weakly isolated
+transactions (sets of atomic operations), and our system model accepts
 \emph{specifications} of consistency and isolation levels drawn from
-an expressive logic. Gotsman et al.~\cite{gotsmanpopl16} adapt
-\iso{Parallel Snapshot Isolation} to the aforementioned setting by
-interpreting it as a consistency level that serializes writes to
-objects. They subsequently propose a dedicated proof rule to prove
-program invaraints under this model. By parameterizing our proof
-system on isolation specifications, we avoid the need to define a
-separate proof rule for each new isolation level.
+an expressive logic.  In recent work~\cite{gotsmanpopl16},
+\iso{Parallel Snapshot Isolation} is adapted to the aforementioned
+setting by interpreting it as a consistency level that serializes
+writes to objects; a dedicated proof rule is developed to help prove
+prove program invariants hold under this model. By parameterizing our
+proof system over a gamut of weak isolation specifications, we avoid
+the need to define a separate proof rule for each new isolation level.
 
@@
 reasoning approaches to work in both the cases. Ridge~\cite{rgtso}
-generalizes rely-guarantee reasoning to x86-TSO relaxed memory model.
-Likewise, Vafeadis et al.~\cite{rsl13} generalize concurrent
-separation logic (CSL)~\cite{csl} to C11 relaxed memory model.
-Ferreira et al.~\cite{ferreira10} propose parameterized operational
+generalizes rely-guarantee reasoning to the x86-TSO memory model.
+Likewise, Vafeadis \emph{et al.}~\cite{rsl13} generalize concurrent
+separation logic (CSL)~\cite{csl} to the C11 relaxed memory model.
+Ferreira et al.~\cite{ferreira10} propose a parameterized operational
 semantics for relaxed memory models, but the parameterization is over
@@
 memory models and operational semantics of programs in a unified
-reasoning framework.
+reasoning framework.  \SJ{cite plan b: a buffered memory model for Java.}

### 2016-06-29 store-consistency.tex
edits

@@
 \begin{array}{l}
-  \C{sc}(\E) \;=\; \forall\eta_1,\eta_2.~\{\eta_1,\eta_2\}
+  \C{SC}(\E) \;=\; \forall\eta_1,\eta_2.~\{\eta_1,\eta_2\}
   \subseteq \E.\A \conj \id(\eta_1) < \id(\eta_2) \\
@@
 
-We shall now consider the proof of the example in
-Fig.~\ref{fig:motiv-eg-1} in greater detail. We assume an {\sc sc}
-store, such as a relational database, whose store-specific trace
-invaraint ($\I_s$) was shown previously. Both transactions are run at
-{\sc si} isolation, hence $\I_c$ is $\lambda\E.~\underE{\C{SI(Wd1)}}
-\conj \underE{\C{SI(Wd1)}}$. As usual, $\I$ is $\I_s \conj \I_c$.
+We now consider the proof of the example in Fig.~\ref{fig:motiv-eg-1}
+in greater detail. We assume an {\sc sc} store, such as a relational
+database, whose store-specific trace invaraint ($\I_s$) was shown
+previously. Both transactions are run at {\sc si} isolation, hence
+$\I_c$ is $\lambda\E.~\underE{\C{SI(Wd1)}} \conj
+\underE{\C{SI(Wd1)}}$. As usual, $\I$ is $\I_s \conj \I_c$.
 
@@
 Temporary name \C{v3} has been introduced to capture the result of the
-RHS expression \C{C-a2}.  Assertions implicitly refer to the current
+RHS expression \C{C-a1}.  Assertions implicitly refer to the current
 execution ($\E$), just as hoare triples implicitly refer to the
@@
 from the RG rule \rulelabel{RG-Asgn} for assignments, which allows us
-to easiliy conclude the following about the execution state after the
+to easily conclude the following about the execution state after the
 assignment:
->>>>>>> 70ae6b11c8f648c323413d6fe8f8fb1c2cba75ae
+
 \begin{smathpar}

### 2016-06-30 store-consistency.tex
edits to sec. 5

@@
 % inevitable to enforce $\I_c$.
+
+A distinguishing feature of our framework is that it is parameteric
+over different isolation and consistency instantiations, whose
+properties can be encapsulated within the trace invariant $\I$.  While
+it is true that most relational databases implement standard ANSI SQL
+isolation levels similar to the ones shown in
+Fig.~\ref{fig:ansi-isolation}, there is no clear agreement on the
+exact semantics of these isolation levels on various weakly-consistent
+stores.  Thus, what Oracle calls serializable isolation is a slightly
+stronger variant of snapshot isolation.  Having a parameterized
+semantics and proof framework built on this semantics allows us to
+support these variants without the need for semantics re-engineering
+every time an application is retargeted to a new platform.  Moreover,
+the need for high availability on geo-replicated stores is
+incentivizing the development of new isolation levels. \emph{Parallel
+  Snapshot Isolation}, for example, was proposed in 2011, and
+\emph{Non-Monotonic Snapshot Isolation}~\cite{...} in 2013, both for
+replicated stores. Commercial eventually-consistent stores have only
+just started adopting transactions (the first release of Cassandra
+with support for multi-key transactions is slated for later this
+year), and all indications are that we should expect to see new
+isolation instantiations in the future.
+
+Without parameterization, we would need to carefully engineer a
+reasoning framework for each isolation and consistency variant.
+Clearly, each such framework would be incomparable to one another. On
+the other hand, our treatment allows us to accommodate the myriad of
+different isolation and store-consistency instances without the need
+for bespoke specialization.
 

### 2016-06-30 related-work.tex
edits

@@
 to transactions while guaranteeing the serializability of the
-execution. In~\cite{gotsmanpodc16}, Cerone et al. improve on Adya's
+execution. In~\cite{gotsmanpodc16}, Cerone \emph{et al.} improve on Adya's
 {\sc si} specification and use it to derive a static analysis that
@@
 isolation levels, without taking application invariants into account.
-Bernstein et al.~\cite{bern2000} propose informal semantic conditions
+Bernstein \emph{et al.}~\cite{bern2000} propose informal semantic conditions
 to ensure the satisfaction of application invariants under weaker
@@
 separation logic (CSL)~\cite{csl} to the C11 relaxed memory model.
-Ferreira et al.~\cite{ferreira10} propose a parameterized operational
-semantics for relaxed memory models, but the parameterization is over
-a relation between relaxed memory programs and related {\sc sc}
-programs. They also prove the soundness of CSL for the weakest memory
-model expressible in their semantics, but neither a parameterized (in
-the aforementioned sense) CSL, nor a CSL specialized for any relaxed
-memory model is proposed. In fact, to the best of our knowledge, there
-does not exist a proof system that combines axiomatic semantics of
-memory models and operational semantics of programs in a unified
-reasoning framework.  \SJ{cite plan b: a buffered memory model for Java.}
+Ferreira \emph{et al.}~\cite{ferreira10} propose a parameterized
+operational semantics for relaxed memory models, but the
+parameterization is over a relation between relaxed memory programs
+and related {\sc sc} programs. They also prove the soundness of CSL
+for the weakest memory model expressible in their semantics, but
+neither a parameterized (in the aforementioned sense) CSL, nor a CSL
+specialized for any relaxed memory model is proposed.  ~\cite{DLZ+13}
+presents a \emph{buffered memory model} for Java that deines an
+axiomatic definition for the JMM in terms of memory reorderings, and
+an operational instantiation consistent with the total store ordering
+(TSO) memory model found on x86 processors, and demonstrates the
+equivalence of the two definitions.  However, there is no attempt to
+parameterize the model over different reordering or consistency
+definitions as we demonstrate here.

### 2016-06-30 store-consistency.tex
edits

@@
 
-A distinguishing feature of our framework is that it is parameteric
-over different isolation and consistency instantiations, whose
-properties can be encapsulated within the trace invariant $\I$.  While
-it is true that most relational databases implement standard ANSI SQL
-isolation levels similar to the ones shown in
-Fig.~\ref{fig:ansi-isolation}, there is no clear agreement on the
-exact semantics of these isolation levels on various weakly-consistent
-stores.  Thus, what Oracle calls serializable isolation is a slightly
-stronger variant of snapshot isolation.  Having a parameterized
-semantics and proof framework built on this semantics allows us to
-support these variants without the need for semantics re-engineering
-every time an application is retargeted to a new platform.  Moreover,
-the need for high availability on geo-replicated stores is
-incentivizing the development of new isolation levels. \emph{Parallel
-  Snapshot Isolation}, for example, was proposed in 2011, and
-\emph{Non-Monotonic Snapshot Isolation}~\cite{...} in 2013, both for
-replicated stores. Commercial eventually-consistent stores have only
-just started adopting transactions (the first release of Cassandra
-with support for multi-key transactions is slated for later this
-year), and all indications are that we should expect to see new
-isolation instantiations in the future.
+\paragraph{Discussion.} A distinguishing feature of our framework is that it
+    is parameteric over different isolation and consistency
+    instantiations, whose properties can be encapsulated within the
+    trace invariant $\I$.  While it is true that most relational
+    databases implement standard ANSI SQL isolation levels similar to
+    the ones shown in Fig.~\ref{fig:ansi-isolation}, there is no clear
+    agreement on the exact semantics of these isolation levels on
+    various weakly-consistent stores.  Thus, what Oracle calls
+    serializable isolation is a slightly stronger variant of snapshot
+    isolation.  Having a parameterized semantics and proof framework
+    built on this semantics allows us to support these variants
+    without the need for semantics re-engineering every time an
+    application is retargeted to a new platform.
 
-Without parameterization, we would need to carefully engineer a
-reasoning framework for each isolation and consistency variant.
-Clearly, each such framework would be incomparable to one another. On
-the other hand, our treatment allows us to accommodate the myriad of
-different isolation and store-consistency instances without the need
-for bespoke specialization.
+    Moreover, the need for high availability on geo-replicated stores
+    is incentivizing the development of new isolation
+    levels. \emph{Parallel Snapshot Isolation}, for example, was
+    proposed in 2011, and \emph{Non-Monotonic Snapshot
+      Isolation}~\cite{...} in 2013, both for replicated
+    stores. Commercial eventually-consistent stores have only just
+    started adopting transactions (the first release of Cassandra with
+    support for multi-key transactions is slated for later this year),
+    and all indications are that we should expect to see new isolation
+    instantiations in the future.  Without parameterization, we would
+    need to carefully engineer a reasoning framework for each
+    isolation and consistency variant.  Clearly, each such framework
+    would be incomparable to one another. On the other hand, our
+    treatment allows us to accommodate the myriad of different
+    isolation and store-consistency instances without the need for
+    bespoke specialization.
 

### 2016-07-01 store-consistency.tex
Edits to the note about parameterization.

--- 2016-06-29 store-consistency.tex
@@
 proof for the program.
+
+\paragraph{Discussion.} As evident from the example, instantiating the
+proof system for a specfic combination of consistency and isolation
+settings, and then deriving facts specific to that setting
+characterizes our reasoning approach. While intuition suggests that
+some instantiations, such as {\sc sc} with ANSI SQL isolation, are
+general enough to cover many practical cases, it not the case in
+reality owing to the subtle differences in the semantics of isolation
+levels among databases. For instance, Postgres's \iso{Repeatable Read}
+isolation is stronger than ANSI SQL's or MySQL's in non-trivial ways.
+Furthermore, the need for high availability on geo-replicated stores
+has led to the development of of new isolation levels which are yet to
+be standardized. For example, \emph{Parallel Snapshot
+Isolation}~\cite{psi} was proposed in 2011, and \emph{Non-Monotonic
+Snapshot Isolation}~\cite{nmsi} in 2013, both targeted for replicated
+stores. As more commercial weakly consistent stores adopt variants of
+transactions (like the \emph{lightweight transactions} of Cassandra),
+it is reasonable to expect novel forms of isolation to be proposed.
+While it is possible to carefully engineer a reasoning framework for
+each combination of consistency and isolation, it leads to multiple
+proof systems with no obvious to compare proofs.  Having a
+parameterized semantics and proof framework built on this semantics
+allows us to support these variants as distinct, yet comparable
+instantiations, thus avoiding the need for semantics re-engineering
+every time an application is retargeted to a new platform
+
+%   A distinguishing feature of our framework is that it
+%   is parameteric over different isolation and consistency
+%   instantiations, whose properties can be encapsulated within the
+%   trace invariant $\I$.  While it is true that most relational
+%   databases implement standard ANSI SQL isolation levels similar to
+%   the ones shown in Fig.~\ref{fig:ansi-isolation}, there is no clear
+%   agreement on the exact semantics of these isolation levels on
+%   various weakly-consistent stores.  Thus, what Oracle calls
+%   serializable isolation is a slightly stronger variant of snapshot
+%   isolation.  Having a parameterized semantics and proof framework
+%   built on this semantics allows us to support these variants
+%   without the need for semantics re-engineering every time an
+%   application is retargeted to a new platform.
+
+%   Moreover, the need for high availability on geo-replicated stores
+%   is incentivizing the development of new isolation
+%   levels. \emph{Parallel Snapshot Isolation}, for example, was
+%   proposed in 2011, and \emph{Non-Monotonic Snapshot
+%     Isolation}~\cite{...} in 2013, both for replicated
+%   stores. Commercial eventually-consistent stores have only just
+%   started adopting transactions (the first release of Cassandra with
+%   support for multi-key transactions is slated for later this year),
+%   and all indications are that we should expect to see new isolation
+%   instantiations in the future.  Without parameterization, we would
+%   need to carefully engineer a reasoning framework for each
+%   isolation and consistency variant.  Clearly, each such framework
+%   would be incomparable to one another. On the other hand, our
+%   treatment allows us to accommodate the myriad of different
+%   isolation and store-consistency instances without the need for
+%   bespoke specialization.

### 2016-07-01 store-consistency.tex
edits

@@
 general enough to cover many practical cases, it not the case in
-reality owing to the subtle differences in the semantics of isolation
-levels among databases. For instance, Postgres's \iso{Repeatable Read}
-isolation is stronger than ANSI SQL's or MySQL's in non-trivial ways.
-Furthermore, the need for high availability on geo-replicated stores
-has led to the development of of new isolation levels which are yet to
-be standardized. For example, \emph{Parallel Snapshot
-Isolation}~\cite{psi} was proposed in 2011, and \emph{Non-Monotonic
-Snapshot Isolation}~\cite{nmsi} in 2013, both targeted for replicated
-stores. As more commercial weakly consistent stores adopt variants of
-transactions (like the \emph{lightweight transactions} of Cassandra),
-it is reasonable to expect novel forms of isolation to be proposed.
-While it is possible to carefully engineer a reasoning framework for
-each combination of consistency and isolation, it leads to multiple
-proof systems with no obvious to compare proofs.  Having a
-parameterized semantics and proof framework built on this semantics
-allows us to support these variants as distinct, yet comparable
-instantiations, thus avoiding the need for semantics re-engineering
-every time an application is retargeted to a new platform
+reality given the subtle differences in the semantics of isolation
+levels among different database implementations. For instance,
+Postgres' \iso{Repeatable Read} isolation is stronger than ANSI SQL's
+or MySQL's in non-trivial ways.  Furthermore, the need for high
+availability on geo-replicated stores has led to the development of
+new isolation levels which are yet to be standardized. For example,
+\emph{Parallel Snapshot Isolation}~\cite{psi} was proposed in 2011,
+and \emph{Non-Monotonic Snapshot Isolation}~\cite{nmsi} in 2013, both
+targeted for replicated stores. As more commercial weakly consistent
+stores adopt variants of transactions, like the recently introduced
+\emph{lightweight transactions} of Cassandra, it is reasonable to
+expect new isolation definitions to continue to be proposed.  While it
+is possible to carefully engineer a reasoning framework for each
+combination of consistency and isolation, such a strategy would lead
+to multiple semantic definitions and proof systems with no obvious to
+compare and relate them.  Having a parameterized semantics and proof
+framework built on the framework defined above allows us to support
+these variants as distinct, yet comparable instantiations, thus
+avoiding the need for semantics re-engineering every time an
+application is retargeted to a new platform.
 

### 2016-07-01 store-consistency.tex
Fixed typo.

--- 2016-07-01 store-consistency.tex
@@
 proof system for a specfic combination of consistency and isolation
-settings, and then deriving facts specific to that setting
-characterizes our reasoning approach. While intuition suggests that
-some instantiations, such as {\sc sc} with ANSI SQL isolation, are
-general enough to cover many practical cases, it not the case in
+settings, and then deriving facts specific to that setting is one of
+the characteristic our reasoning approach. While intuition suggests
+that some instantiations, such as {\sc sc} with ANSI SQL isolation,
+are general enough to cover many practical cases, it not the case in
 reality owing to the subtle differences in the semantics of isolation
@@
 each combination of consistency and isolation, it leads to multiple
-proof systems with no obvious to compare proofs.  Having a
+proof systems with no obvious way to compare proofs.  Having a
 parameterized semantics and proof framework built on this semantics

### 2016-07-01 related-work.tex

@@
 specialized for any relaxed memory model is proposed.  ~\cite{DLZ+13}
-presents a \emph{buffered memory model} for Java that deines an
+presents a \emph{buffered memory model} for Java that defines an
 axiomatic definition for the JMM in terms of memory reorderings, and
-an operational instantiation consistent with the total store ordering
-(TSO) memory model found on x86 processors, and demonstrates the
-equivalence of the two definitions.  However, there is no attempt to
-parameterize the model over different reordering or consistency
-definitions as we demonstrate here.
+an operational instantiation consistent with the TSO memory model
+found on x86 processors, and demonstrates the equivalence of the two
+definitions.  However, there is no attempt to parameterize the model
+over different reordering or consistency definitions, which is a central
+focus of this paper.

### 2016-07-01 store-consistency.tex

--- 2016-07-01 store-consistency.tex
@@
 combination of consistency and isolation, such a strategy would lead
-to multiple semantic definitions and proof systems with no obvious to
-compare and relate them.  Having a parameterized semantics and proof
-framework built on the framework defined above allows us to support
-these variants as distinct, yet comparable instantiations, thus
-avoiding the need for semantics re-engineering every time an
+to multiple semantic definitions and proof systems with no obvious way
+to compare and relate them.  Having a parameterized semantics and a
+proof framework built on the such semantics allows us to support these
+variants as distinct, yet comparable instantiations, thus avoiding the
+need for semantics re-engineering every time an
 application is retargeted to a new platform.

### 2016-07-02 case-studies.tex
Case studies.

\section{Case Studies}
\label{sec:case-studies}

In this section, we present two case studies that demonstrate the
applicability of our reasoning approach in varied settings, and the
consequent practical utility.

\paragraph{A video view counter} The view count of a video on a
YouTube-like video-sharing website must never appear to be decreasing,
although it may not necessarily reflect the actual number of views.
We call this the monotonicity invariant of the counter. To understand
whether a straightforward implementation of the view counter preserves
the monotonicity invariant, we considered a \txnimp program that is
representative of the counter application. In particular, we
considered four concurrent transactions that operate on a shared
counter: a reader transaction that reads the counter twice obtaining
the values $k_1$ and $k_2$, respectively, and three writer
transactions, each incrementing the counter once. If the reader
witnesses a monotonically increasing view count, then $k_2\ge k_1$. We
then instantiated our reasoning framework for an {\sc sc} store with
{\sc rc} as the default isolation level (Postgres is one such store),
with the aim of verifying the program under this instantiation.  Our
attempts were however unsuccessful owing to our inability to prove
that a writer transaction always commits a value no smaller than the
current value of the counter. This points to the existence of
anamolous executions where writers may commit a value smaller than the
current value of the counter. To understand the prevalence of such
anomalies, we implemented a view counter application in Ruby-on-Rails
using standard templates and techniques\footnote{Source code of the
applications, experiments, and the tools we developed to analyze Ruby
can be found at the following anonymized url: \url{...}}. The
application stores the video count in a Postgres database, and
supports an increment operation and a read operation on the count. In
our experiments with 16 (resp. 32) concurrent writers, readers
observed 7 (17) violations of monotonicity, in 10 rounds of operation,
thus confirming that anomalies are indeed possible.

To determine the weakest ANSI SQL isolation levels needed to restore
the monotonicity invariant, we instantiated our reasoning framework
for an {\sc sc} store with different isolation levels for readers and
writers.  With half-a-day of effort, we were able to prove that
\iso{Read Uncommmitted}, the weakest isolation level, is sufficient
for readers, whereas \iso{Serializable} isolation is needed for
writers. Our experiments with uncommitted reads and serializable
writes resulted in no observed violations of monotonicity, confirming
our judgment. While this exercise let us determine the appropriate
isolation levels for an {\sc sc} store with ANSI SQL isolation levels,
we were however surprised to learn from Postgres's
documentation~\cite{postgres} that its isolation levels are stronger
than ANSI SQL's in non-trivial ways. We subsequently formalized the
guarantees and repeated the proof effort to discover that Postgres's
\iso{Repeatable Read} is in fact sufficient for writers - an
observation we confirmed through experiments. However, the same choice
of isolation levels led to monotonicity violations on MySQL, because
MySQL's {\sc rr} is weaker than Postgres's.

To understand the effect of changing the store consistency, we
reimplemented the counter application for Cassandra, an {\sc ec} store
(We used the \emph{shim layer} of~\cite{pldi15} to equip Cassandra
with {\sc cc} and {\sc sc} consistency levels, and {\sc rc}, {\sc
mav}, and {\sc rr} isolation levels). We then repeated our experiments
with uncommitted reads and serializable writes - a combination which
was proved to preserve monotonicity on an {\sc sc} store, but
nonetheless observed monotonicity violations. The reason for such
violations became evident as we considered an instantiation our proof
framework for an {\sc ec} store. Under this instantiation, it is
impossible to derive the \C{MonotonicVis} (\S\ref{sec:ansi-isolation})
property needed to show that successive reads in a session witness
monotonically increasing state. We were however able to derive the
property assuming either causal consistency or {\sc mav} isolation
(both supported by the shim layer). We experimented with both these
options and were able to recover the monotonicity invariant.

\paragraph{Microblog}

Next, we focus our attention on a Twitter-like microblogging site that
is one the sample applications shipped with a popular textbook on
Ruby-on-Rails~\cite{railsbook}. The application is built to
demonstrate various capabilities of Rails' ORM framework, and, as
such, is a good representative of Ruby-on-Rails applications.  For
instance, instead of relying on database capabilities, it uses Rails's
\emph{has\_many} association to describe the relationship between a
user entity and a micropost entity. Further, it qualifies the
association with \emph{dependent: destroy}, thus directing Rails to
destroy all microposts if the corresponding author is deleted. Thus,
the application entrusts Rails with the responsibility to enforce
referential integrity between microposts and users. Rails in turn
relies on database-backed transactions to enforce such constraints,
thus making application semantics susceptible to the changes in
isolation level.  For instance, deletion of a user is done within a
transaction ($T_1$) that also deletes the microposts authored by
(hence, \emph{dependent} on) the user. While this is effective if all
transactions are {\sc ser}, under the default {\sc rc} isolation, it
lets concurrent transactions witness referential integrity violations.
For instance, consider a transaction ($T_2$) that populates a timeline
by first reading microposts, and then reading the author information
associated with each micropost. It is possible for an author to be
deleted (due to the interference from $T_1$) between the two reads,
leading $T_2$ to witness a referential integrity violation, although
none really exists in the database.  Indeed, we have verified this
hypothesis by instantiating our reasoning framework for {\sc sc}
machine and {\sc rc} isolation, and extracting the aforementioned
anomalous execution from our failure to prove referential
integrity\footnote{We modeled referential integrity using two
variables, $X$ and $Y$: $Y$ is 1 only if $X$ is 1.  Otherwise, both
are 0.}. To measure the likelihood of such anomalies in practice, we
first populated a Postgres database with 1000 user accounts, each user
with 50 microposts and 20 followers, choosen uniformly at random. We
then constructed experiments with 64 concurrent clients performing
deletions and timeline reads of randomly choosen users in 1:7 ratio,
and witnessed a maximum of 250 violations of referential integrity.

Clearly, {\sc rc} is insufficient. In order to determine the weakest
isolation levels needed to enforce referential integrity, we
instantiated our framework for an {\sc sc} store with various Postgres
isolation levels for user deletions and timeline reads, and obtained a
proof that {\sc rc} is sufficient for deletions, whereas reads require
{\sc rr}. Repeating the experiment with these isolation settings
resulted in no violations, as expected.

### 2016-07-02 case-studies.tex
Edits.

@@
 although it may not necessarily reflect the actual number of views.
-We call this the monotonicity invariant of the counter. To understand
-whether a straightforward implementation of the view counter preserves
-the monotonicity invariant, we considered a \txnimp program that is
-representative of the counter application. In particular, we
-considered four concurrent transactions that operate on a shared
-counter: a reader transaction that reads the counter twice obtaining
-the values $k_1$ and $k_2$, respectively, and three writer
-transactions, each incrementing the counter once. If the reader
-witnesses a monotonically increasing view count, then $k_2\ge k_1$. We
-then instantiated our reasoning framework for an {\sc sc} store with
-{\sc rc} as the default isolation level (Postgres is one such store),
-with the aim of verifying the program under this instantiation.  Our
-attempts were however unsuccessful owing to our inability to prove
-that a writer transaction always commits a value no smaller than the
-current value of the counter. This points to the existence of
-anamolous executions where writers may commit a value smaller than the
-current value of the counter. To understand the prevalence of such
-anomalies, we implemented a view counter application in Ruby-on-Rails
-using standard templates and techniques\footnote{Source code of the
-applications, experiments, and the tools we developed to analyze Ruby
-can be found at the following anonymized url: \url{...}}. The
-application stores the video count in a Postgres database, and
-supports an increment operation and a read operation on the count. In
-our experiments with 16 (resp. 32) concurrent writers, readers
-observed 7 (17) violations of monotonicity, in 10 rounds of operation,
-thus confirming that anomalies are indeed possible.
+We call this the monotonicity invariant. To understand whether a
+straightforward implementation of the view counter preserves the
+invariant, we considered a \txnimp program that is representative of
+the counter application. In particular, we considered four concurrent
+transactions that operate on a shared counter: a reader transaction
+that reads the counter twice obtaining the values $k_1$ and $k_2$,
+respectively, and three writer transactions, each incrementing the
+counter once. If the reader witnesses a monotonically increasing view
+count, then $k_2\ge k_1$. We then instantiated our reasoning framework
+for an {\sc sc} store with {\sc rc} as the default isolation level
+(Postgres is one such store), with the aim of verifying the program
+under this instantiation.  Our attempts were however unsuccessful
+owing to our inability to prove that a writer transaction always
+commits a value no smaller than the current value of the counter. This
+points to the existence of anamolous executions where writers may
+commit a value smaller than the current value of the counter. To
+understand the prevalence of such anomalies, we implemented a view
+counter application in Ruby-on-Rails using standard templates and
+techniques\footnote{Source code of the applications, experiments, and
+the tools we developed to analyze Ruby can be found at the following
+anonymized url: \url{...}}. The application stores the video count in
+a Postgres database, and supports an increment operation and a read
+operation on the count. In our experiments with 16 (resp. 32)
+concurrent writers, readers observed 7 (17) violations of
+monotonicity, in 10 rounds of operation, thus confirming that
+anomalies are indeed possible.
 
@@
 (both supported by the shim layer). We experimented with both these
-options and were able to recover the monotonicity invariant.
+options and were able to recover the invariant.
 

### 2016-07-03 case-studies.tex
Bib fixed. Case studies edited.

@@
 
-\paragraph{A video view counter} The view count of a video on a
-YouTube-like video-sharing website must never appear to be decreasing,
-although it may not necessarily reflect the actual number of views.
-We call this the monotonicity invariant. To understand whether a
-straightforward implementation of the view counter preserves the
-invariant, we considered a \txnimp program that is representative of
-the counter application. In particular, we considered four concurrent
-transactions that operate on a shared counter: a reader transaction
-that reads the counter twice obtaining the values $k_1$ and $k_2$,
-respectively, and three writer transactions, each incrementing the
-counter once. If the reader witnesses a monotonically increasing view
-count, then $k_2\ge k_1$. We then instantiated our reasoning framework
-for an {\sc sc} store with {\sc rc} as the default isolation level
-(Postgres is one such store), with the aim of verifying the program
-under this instantiation.  Our attempts were however unsuccessful
-owing to our inability to prove that a writer transaction always
-commits a value no smaller than the current value of the counter. This
-points to the existence of anamolous executions where writers may
-commit a value smaller than the current value of the counter. To
-understand the prevalence of such anomalies, we implemented a view
-counter application in Ruby-on-Rails using standard templates and
-techniques\footnote{Source code of the applications, experiments, and
-the tools we developed to analyze Ruby can be found at the following
+\paragraph{A video view counter} Our first application is a
+Ruby-on-Rails implementation of a counter to count the number of views
+a video has garnered on a video-sharing website like
+YouTube\footnote{Source code of the applications, experiments, and the
+tools we developed to analyze Ruby can be found at the following
 anonymized url: \url{...}}. The application stores the video count in
-a Postgres database, and supports an increment operation and a read
-operation on the count. In our experiments with 16 (resp. 32)
-concurrent writers, readers observed 7 (17) violations of
-monotonicity, in 10 rounds of operation, thus confirming that
-anomalies are indeed possible.
+a Postgres database, and supports a read operation (a database read
+transaction), and an increment operation (a read-increment-write
+sequence wrapped in a transaction). The application is expected to
+preserve the \emph{monotonicity invariant}, namely that the view count
+should never appear to be decreasing. However, the default {\sc rc}
+isolation of Postgres may lead to anomalous executions as confirmed by
+our experiments; with 16 (resp.  32) concurrent writers performing
+single increments, we observed a total of 7 (17) violations of
+monotonicity. We subsequently instantiated our reasoning framework for
+an {\sc sc} store with Postgres isolation levels (which are stronger
+than ANSI SQL's and MySQL's), and spent 5 man hours determining
+appropriate isolation levels needed to enforce the
+invariant\footnote{The representative \txnimp programs we used to
+formally reason about scenarios described in this section can be found
+in the supplementary.} In particular, we were able to formally prove
+that to preserve the invariant, writers need to execute at
+\iso{Repeatable Read} isolation, while readers can continue at {\sc
+rc}. Repeating the experiments with this configuration led to no
+violations, as expected. While executing all transactions at {\sc ser}
+also prevents violations, it led to an additional 47\% increase in
+latency on an average (16 concurrent writers; 5 rounds). We repeated
+the exercise for MySQL, and found that its {\sc rr} is not sufficient
+to enforce monotonicity; {\sc ser} is needed.
 
-To determine the weakest ANSI SQL isolation levels needed to restore
-the monotonicity invariant, we instantiated our reasoning framework
-for an {\sc sc} store with different isolation levels for readers and
-writers.  With half-a-day of effort, we were able to prove that
-\iso{Read Uncommmitted}, the weakest isolation level, is sufficient
-for readers, whereas \iso{Serializable} isolation is needed for
-writers. Our experiments with uncommitted reads and serializable
-writes resulted in no observed violations of monotonicity, confirming
-our judgment. While this exercise let us determine the appropriate
-isolation levels for an {\sc sc} store with ANSI SQL isolation levels,
-we were however surprised to learn from Postgres's
-documentation~\cite{postgres} that its isolation levels are stronger
-than ANSI SQL's in non-trivial ways. We subsequently formalized the
-guarantees and repeated the proof effort to discover that Postgres's
-\iso{Repeatable Read} is in fact sufficient for writers - an
-observation we confirmed through experiments. However, the same choice
-of isolation levels led to monotonicity violations on MySQL, because
-MySQL's {\sc rr} is weaker than Postgres's.
-
+% To understand whether a straightforward implementation of the view
+% counter preserves the invariant, we considered a \txnimp program
+% that is representative of the counter application. In particular, we
+% considered four concurrent transactions that operate on a shared
+% counter: a reader transaction that reads the counter twice obtaining
+% the values $k_1$ and $k_2$, respectively, and three writer
+% transactions, each incrementing the counter once. If the reader
+% witnesses a monotonically increasing view count, then $k_2\ge k_1$.
+% We then instantiated our reasoning framework for an {\sc sc} store
+% with {\sc rc} as the default isolation level (Postgres is one such
+% store), with the aim of verifying the program under this
+% instantiation.  Our attempts were however unsuccessful owing to our
+% inability to prove that a writer transaction always commits a value
+% no smaller than the current value of the counter. This points to the
+% existence of anamolous executions where  To understand the
+% prevalence of such anomalies, we implemented a view counter
+% application in Ruby-on-Rails using standard templates and
+% techniques.   
 To understand the effect of changing the store consistency, we
-reimplemented the counter application for Cassandra, an {\sc ec} store
-(We used the \emph{shim layer} of~\cite{pldi15} to equip Cassandra
-with {\sc cc} and {\sc sc} consistency levels, and {\sc rc}, {\sc
-mav}, and {\sc rr} isolation levels). We then repeated our experiments
-with uncommitted reads and serializable writes - a combination which
-was proved to preserve monotonicity on an {\sc sc} store, but
-nonetheless observed monotonicity violations. The reason for such
-violations became evident as we considered an instantiation our proof
-framework for an {\sc ec} store. Under this instantiation, it is
-impossible to derive the \C{MonotonicVis} (\S\ref{sec:ansi-isolation})
-property needed to show that successive reads in a session witness
-monotonically increasing state. We were however able to derive the
-property assuming either causal consistency or {\sc mav} isolation
-(both supported by the shim layer). We experimented with both these
-options and were able to recover the invariant.
+reimplemented the counter as a \emph{replicated data type}
+(RDT)~\cite{crdt,burckhardt14} on top of Cassandra
+\emph{bolted-on}~\cite{bailisbolton} with {\sc cc} and {\sc sc}
+consistency levels, and ANSI SQL isolation levels. Keeping the
+consistency level at {\sc ec}, we then repeated our experiments with
+committed reads and serializable increments - a combination which was
+proved to preserve monotonicity on an {\sc sc} store, but nonetheless
+observed monotonicity violations. On the other hand, instantiating the
+store consistency with {\sc cc} allowed us to obtain a proof, and
+recover the monotonicity invariant. We in fact obtained a proof after
+relaxing the serializability requirement for writes, by changing the
+definition of the $\interp{\E}(X)$ (\S\ref{sec:opsem}) to return the
+value of the largest write-to-$X$ instead of the latest write-to-$X$.
+We implemented this version of the counter RDT, and obtained a 52\%
+improvement in latency compared to the previous
+implementation\footnote{The tradeoff is that the view count is now no
+longer accurate, although it grows monotonically.}.
 
@@
 is one the sample applications shipped with a popular textbook on
-Ruby-on-Rails~\cite{railsbook}. The application is built to
-demonstrate various capabilities of Rails' ORM framework, and, as
-such, is a good representative of Ruby-on-Rails applications.  For
-instance, instead of relying on database capabilities, it uses Rails's
-\emph{has\_many} association to describe the relationship between a
-user entity and a micropost entity. Further, it qualifies the
-association with \emph{dependent: destroy}, thus directing Rails to
-destroy all microposts if the corresponding author is deleted. Thus,
-the application entrusts Rails with the responsibility to enforce
-referential integrity between microposts and users. Rails in turn
-relies on database-backed transactions to enforce such constraints,
-thus making application semantics susceptible to the changes in
-isolation level.  For instance, deletion of a user is done within a
-transaction ($T_1$) that also deletes the microposts authored by
-(hence, \emph{dependent} on) the user. While this is effective if all
-transactions are {\sc ser}, under the default {\sc rc} isolation, it
-lets concurrent transactions witness referential integrity violations.
-For instance, consider a transaction ($T_2$) that populates a timeline
-by first reading microposts, and then reading the author information
-associated with each micropost. It is possible for an author to be
-deleted (due to the interference from $T_1$) between the two reads,
-leading $T_2$ to witness a referential integrity violation, although
-none really exists in the database.  Indeed, we have verified this
-hypothesis by instantiating our reasoning framework for {\sc sc}
-machine and {\sc rc} isolation, and extracting the aforementioned
-anomalous execution from our failure to prove referential
-integrity\footnote{We modeled referential integrity using two
-variables, $X$ and $Y$: $Y$ is 1 only if $X$ is 1.  Otherwise, both
-are 0.}. To measure the likelihood of such anomalies in practice, we
-first populated a Postgres database with 1000 user accounts, each user
-with 50 microposts and 20 followers, choosen uniformly at random. We
-then constructed experiments with 64 concurrent clients performing
-deletions and timeline reads of randomly choosen users in 1:7 ratio,
-and witnessed a maximum of 250 violations of referential integrity.
+Ruby-on-Rails~\cite{railsbook}. One of the invariants that the
+application chooses to enforce is the referential integrity between
+microposts and authors. The application enforces this invariant
+entirely in the user space by relying on database-backed transactions
+(this behaviour is typical of Rails applications~\cite{bailisferal}).
+For instance, deletion of a user is carried out in a transaction that
+also includes the deletion of microposts authored by the user.
+However, concurrent observers can nonetheless witness violations of
+referential integrity. To measure the likelihood of such anomalies in
+practice, we first populated a Postgres database with 1000 user
+accounts, each user with 50 microposts and 20 followers, choosen
+uniformly at random. We then constructed experiments with 64
+concurrent clients performing deletions and timeline reads of randomly
+choosen users in 1:7 ratio, and witnessed a maximum of 250 violations
+of referential integrity. We then instantiated our framework for an
+{\sc sc} store with various ANSI SQL isolation levels for user
+deletions and timeline reads, and obtained a proof that {\sc rc} is
+sufficient for deletions, whereas reads require {\sc rr}. The proof is
+applicable for MySQL, which implements ANSI isolation.  Since
+Postgres's isolation levels are stronger than corresponding ANSI
+levels, using \rulelabel{RG-Conseq} rule (Fig.~\ref{fig:rg-rules}), we
+also obtain a correctness proof for Postgres.
 
-Clearly, {\sc rc} is insufficient. In order to determine the weakest
-isolation levels needed to enforce referential integrity, we
-instantiated our framework for an {\sc sc} store with various Postgres
-isolation levels for user deletions and timeline reads, and obtained a
-proof that {\sc rc} is sufficient for deletions, whereas reads require
-{\sc rr}. Repeating the experiment with these isolation settings
-resulted in no violations, as expected.
+As evident from our case studies, the there already exists a
+significant diversity among the weak consistency and isolation
+variants implemented on various real-world data stores. More such
+variants have been developed in the recent years to meet the need for
+high availability on geo-replicated stores. For example,
+\emph{Parallel Snapshot Isolation}~\cite{psi} was proposed in 2011,
+and \emph{Non-Monotonic Snapshot Isolation}~\cite{nmsi} in 2013, both
+targeted for replicated stores. As more commercial weakly consistent
+stores adopt various kinds of transactional semantics, like the
+recently introduced \emph{lightweight transactions} of Cassandra, it
+is reasonable to expect new isolation definitions to continue to be
+proposed. It may be possible to carefully engineer a reasoning
+framework for each combination of consistency and isolation, but such
+a strategy would lead to multiple semantic definitions and proof
+systems with no obvious way to compare and relate them. Having a
+parameterized semantics and a proof framework built on the such
+semantics allows us to support these variants as distinct, yet
+comparable instantiations, as we demonstrate in our case studies.
+%  GK: what does the following sentence mean?
+%  As a consequence, we avoid the need for semantics re-engineering
+%  every time an application is retargeted to a new platform.

### 2016-07-03 introduction.tex
Bib fixed. Case studies edited.

@@
 arguments.  Consequently, reasoning about weak isolation remains an
-error prone endeavor, with major database vendors continuing to
-document their isolation levels primarily in terms of the undesirable
-phenomena~\cite{postgresiso, mysqliso, oracleiso} a particular
-isolation level may induce, placing the burden squarely on the
-programmer to determine application correctness.
+error prone endeavor, with major database vendors~\cite{postgresiso,
+mysqliso, oracleiso} continuing to document their isolation levels
+primarily in terms of the undesirable phenomena a particular isolation
+level may induce, placing the burden squarely on the programmer to
+determine application correctness.
 

### 2016-07-03 store-consistency.tex
Bib fixed. Case studies edited.

--- 2016-07-01 store-consistency.tex
@@
 
-\paragraph{Discussion.} As evident from the example, instantiating the
-proof system for a specfic combination of consistency and isolation
-settings, and then deriving facts specific to that setting is one of
-the characteristic our reasoning approach. While intuition suggests
-that some instantiations, such as {\sc sc} with ANSI SQL isolation,
-are general enough to cover many practical cases, it not the case in
-reality owing to the subtle differences in the semantics of isolation
-levels among databases. For instance, Postgres's \iso{Repeatable Read}
-isolation is stronger than ANSI SQL's or MySQL's in non-trivial ways.
-Furthermore, the need for high availability on geo-replicated stores
-has led to the development of of new isolation levels which are yet to
-be standardized. For example, \emph{Parallel Snapshot
-Isolation}~\cite{psi} was proposed in 2011, and \emph{Non-Monotonic
-Snapshot Isolation}~\cite{nmsi} in 2013, both targeted for replicated
-stores. As more commercial weakly consistent stores adopt variants of
-transactions (like the \emph{lightweight transactions} of Cassandra),
-it is reasonable to expect novel forms of isolation to be proposed.
-While it is possible to carefully engineer a reasoning framework for
-each combination of consistency and isolation, it leads to multiple
-proof systems with no obvious way to compare proofs.  Having a
-parameterized semantics and proof framework built on this semantics
-allows us to support these variants as distinct, yet comparable
-instantiations, thus avoiding the need for semantics re-engineering
-every time an application is retargeted to a new platform
+% \paragraph{Discussion.} As evident from the example, instantiating the
+% proof system for a specfic combination of consistency and isolation
+% settings, and then deriving facts specific to that setting
+% characterizes our reasoning approach. While intuition suggests that
+% some instantiations, such as {\sc sc} with ANSI SQL isolation, are
+% general enough to cover many practical cases, it not the case in
+% reality given the subtle differences in the semantics of isolation
+% levels among different database implementations. For instance,
+% Postgres' \iso{Repeatable Read} isolation is stronger than ANSI SQL's
+% or MySQL's in non-trivial ways.  Furthermore, the need for high
+% availability on geo-replicated stores has led to the development of
+% new isolation levels which are yet to be standardized. For example,
+% \emph{Parallel Snapshot Isolation}~\cite{psi} was proposed in 2011,
+% and \emph{Non-Monotonic Snapshot Isolation}~\cite{nmsi} in 2013, both
+% targeted for replicated stores. As more commercial weakly consistent
+% stores adopt variants of transactions, like the recently introduced
+% \emph{lightweight transactions} of Cassandra, it is reasonable to
+% expect new isolation definitions to continue to be proposed.  While it
+% is possible to carefully engineer a reasoning framework for each
+% combination of consistency and isolation, such a strategy would lead
+% to multiple semantic definitions and proof systems with no obvious way
+% to compare and relate them.  Having a parameterized semantics and a
+% proof framework built on the such semantics allows us to support these
+% variants as distinct, yet comparable instantiations, thus avoiding the
+% need for semantics re-engineering every time an
+% application is retargeted to a new platform.
 

### 2016-07-03 case-studies.tex
Edits.

@@
 sufficient for deletions, whereas reads require {\sc rr}. The proof is
-applicable for MySQL, which implements ANSI isolation.  Since
-Postgres's isolation levels are stronger than corresponding ANSI
-levels, using \rulelabel{RG-Conseq} rule (Fig.~\ref{fig:rg-rules}), we
-also obtain a correctness proof for Postgres.
+immediately applicable for MySQL, which implements ANSI SQL isolation
+levels. Since Postgres's isolation levels are stronger than the
+corresponding ANSI levels, \rulelabel{RG-Conseq} rule
+(Fig.~\ref{fig:rg-rules}) lets us deduce that the same choice of
+isolation levels also work on Postgres. Indeed, experiments confirmed
+this judgment.
 

### 2016-07-03 ansi-isolation.tex
Typos fixed.

@@
 visibility thus prevents a transaction from being partially visible.
-However, atomic visiblity does not prevent an uncommitted
+However, atomic visibility does not prevent an uncommitted
 transaction from being visible. This is addressed by \C{CommitVis},
@@
 
-The ANSI SQL 92 standard requires \iso{Read Committted} isolation to avoid
+The ANSI SQL 92 standard requires \iso{Read Committed} isolation to avoid
 the dirty reads phenomenon, which is achieved by enforcing
@@
 Fig.~\ref{fig:ansi-iso-eg-ser} shows an execution of an {\sc ser}
-transacton $T_3$ and a {\sc mav transaction} $T_4$, which write to $X$
+transaction $T_3$ and a {\sc mav transaction} $T_4$, which write to $X$
 and $Y$, respectively. The snapshot witnessed by the {\sc ser}

### 2016-07-03 case-studies.tex
Typos fixed.

@@
 To understand the effect of changing the store consistency, we
-reimplemented the counter as a \emph{replicated data type}
+re-implemented the counter as a \emph{replicated data type}
 (RDT)~\cite{crdt,burckhardt14} on top of Cassandra
@@
 improvement in latency compared to the previous
-implementation\footnote{The tradeoff is that the view count is now no
+implementation\footnote{The trade-off is that the view count is now no
 longer accurate, although it grows monotonically.}.
@@
 practice, we first populated a Postgres database with 1000 user
-accounts, each user with 50 microposts and 20 followers, choosen
+accounts, each user with 50 microposts and 20 followers, chosen
 uniformly at random. We then constructed experiments with 64
 concurrent clients performing deletions and timeline reads of randomly
-choosen users in 1:7 ratio, and witnessed a maximum of 250 violations
+chosen users in 1:7 ratio, and witnessed a maximum of 250 violations
 of referential integrity. We then instantiated our framework for an

### 2016-07-03 introduction.tex
Typos fixed.

@@
   semantics} of the underlying store. In concrete terms, this means
-that, unlike recent work focussed on formalizing the semantics of weak
+that, unlike recent work focused on formalizing the semantics of weak
 consistency~\cite{gotsmanpopl16, redblueatc, ecinec}, our system model
@@
     weakly-isolated transactions, demonstrating that a general
-    parameteric isolation semantics can be expressed as
+    parametric isolation semantics can be expressed as
     well-formedness constraints on a program's execution.

### 2016-07-03 opsem.tex
edits to sec. 5
Typos fixed.

@@
 can be accessed via $\rval$. Every effect has a unique identifier
-accessible via $\id$, and a transction identifier accessible via
+accessible via $\id$, and a transaction identifier accessible via
 $\txn$.  The latter identifies the transaction that generated the
@@
 of such stores; we elaborate on the implication of this style of
-definition in \S\ref{sec:store-consistency}.  Stronger visiblity
+definition in \S\ref{sec:store-consistency}.  Stronger visibility
 properties can be expressed by imposing well-formedness constraints
@@
 isolation.  Notably, the machine achieves this without defining an
-operationalsemantics for isolation levels, instead solely relying on
+operational semantics for isolation levels, instead solely relying on
 their declarative characterization as trace well-formedness

### 2016-07-03 reasoning.tex
Typos fixed.

@@
 \interp{\E'}(X) \ge 0$.  Assertions on states are also written
-similarly. For instance, consider the postcondition
+similarly. For instance, consider the post-condition
 $\C{C\,=\,k-a1-a2}$ of the program in Fig.~\ref{fig:motiv-eg-1}. The
@@
 of executing $c$ has on the state is captured by $G$, and the final
-state of execution satisfies the postcondition $Q$. The terms
+state of execution satisfies the post-condition $Q$. The terms
 ``execution'' and ``step'' are defined with respect to an operational
@@
 quintuple in that it requires every step of the execution to additionally preserve
-the trace invariant $\I$ for the postcondition to be valid.  As usual,
+the trace invariant $\I$ for the post-condition to be valid.  As usual,
 if $c$ is inside a transaction $T_i$, we write $\txnbox{c}_i$. \txnimp
@@
 \noindent However, the assumption that interference preserves $\I$ (or,
-dually, $\I$ withstands inteference) needs to be justified
+dually, $\I$ withstands interference) needs to be justified
 separately. We call this the stability requirement on $\I$:
@@
 an execution $\E$ that satisfies the precondition $P$, then the
-resultant execution $\E'$ satisfies the postcondition $Q$, and that
+resultant execution $\E'$ satisfies the post-condition $Q$, and that
 the guarantee $G$ faithfully captures the transition from $\E$ to
@@
 judgment for an arithmetic expression $e_1\pm e_2$ in terms of the
-corresponding judgments for the constitutent expressions $e_1$ and
+corresponding judgments for the constituent expressions $e_1$ and
 $e_2$. The quantified premise requires any value resulting from
@@
 formulation,the rules allow the precondition $P$ and the rely relation
-$R$ to be strengthened, and the postcondition $Q$ (also, $C$ in the
+$R$ to be strengthened, and the post-condition $Q$ (also, $C$ in the
 case of expressions) and the guarantee relation $G$ to be weakened.
@@
 to make explicit every step of its reduction. This requirement is
-justifed however because, in reality, a transaction's atomicity is
+justified however because, in reality, a transaction's atomicity is
 predicated on the isolation settings of the observer. A \iso{Read

### 2016-07-03 related-work.tex
Case studies.
Bib fixed. Case studies edited.
Typos fixed.

@@
 operational nature of Adya's specifications make them suitable for
-runtime monitoring and anamoly
+runtime monitoring and anomaly
 detection~\cite{kemmevldb,feketesigmod08,pssi2011}, whereas the
@@
 specifications are also for complete traces; Moreover, their
-formulation implicitly assumes homogenity, i.e., they require all
+formulation implicitly assumes homogeneity, i.e., they require all
 transactions to execute using either {\sc si} or {\sc ser}, which is
@@
 generalizes rely-guarantee reasoning to the x86-TSO memory model.
-Likewise, Vafeadis \emph{et al.}~\cite{rsl13} generalize concurrent
+Likewise, Vafeiadis \emph{et al.}~\cite{rsl13} generalize concurrent
 separation logic (CSL)~\cite{csl} to the C11 relaxed memory model.

### 2016-07-03 store-consistency.tex
Typos fixed.

@@
 data store, the store specific invariant ($\I_s$) changes from store
-to store depending on the consisteny level.  For an eventually
+to store depending on the consistency level.  For an eventually
 consistent store, $\I_s$ is simply \emph{true}. Stronger stores, such
-as those that support strong consistenty, have non-trivial definitions
+as those that support strong consistent, have non-trivial definitions
 for $\I_s$.
@@
 approach to any $\I_s$ and $\I_c$ by defining a \emph{maximum
-  visiblity principle} to determine an acceptable weakening of $\I_s$
+  visibility principle} to determine an acceptable weakening of $\I_s$
 in case of a conflict with $\I_c$.  The principle requires the
@@
 in greater detail. We assume an {\sc sc} store, such as a relational
-database, whose store-specific trace invaraint ($\I_s$) was shown
+database, whose store-specific trace invariant ($\I_s$) was shown
 previously. Both transactions are run at {\sc si} isolation, hence
@@
 RHS expression \C{C-a1}.  Assertions implicitly refer to the current
-execution ($\E$), just as hoare triples implicitly refer to the
+execution ($\E$), just as Hoare triples implicitly refer to the
 current state.  The context for propositions is also the implicit,
@@
 program is extended with $\neg\committed(\C{Wd1}) \wedge
-\neg\committed(\C{Wd2})$, from which $P_1$ follows. The postcondition
+\neg\committed(\C{Wd2})$, from which $P_1$ follows. The post-condition
 $Q_1$ of \C{Wd1} asserts different values for \C{C} depending on
 whether or not \C{Wd2} has already committed.  It also asserts that
-\C{Wd1} has committed. The postcondition $Q_2$ of \C{Wd2} (not shown)
+\C{Wd1} has committed. The post-condition $Q_2$ of \C{Wd2} (not shown)
 is similar, but with the roles of \C{Wd1} and \C{Wd2} exchanged. The
-postcondition $Q$ follows from $Q_1 \wedge Q_2$, hence valid. It
+post-condition $Q$ follows from $Q_1 \wedge Q_2$, hence valid. It
 remains to show that \C{Wd1} satisfies its specification (\C{Wd2} must
@@
 considers the case when $\C{Wd2} \visar \C{Wd1}$. As we show below,
-this is sufficient to prove the postcondition $\phi_5$, which appears
+this is sufficient to prove the post-condition $\phi_5$, which appears
 to make a stronger claim about the value of \C{C}. The proof follows
@@
 ${\committed(\C{Wd2})}  \Rightarrow \C{C = k-a2} \wedge \C{C =
-k-a2-a1}$, allowing us to prove the postcondition $\phi_5$.
+k-a2-a1}$, allowing us to prove the post-condition $\phi_5$.
 
@@
 above, but with \C{Wd2} and \C{Wd1} interchanged. Continuing our focus
-on the assigment statement, the guarantee proof requires us to show
+on the assignment statement, the guarantee proof requires us to show
 that if executing the assignment takes the execution from $\E$ to
@@
 triples of \C{Wd1} can also be constructed similarly. The combination
-of correctness proofs for \C{Wd1} and \C{Wd2} yeilds a correctness
+of correctness proofs for \C{Wd1} and \C{Wd2} yields a correctness
 proof for the program.

### 2016-07-03 ansi-isolation.tex
Typos fixed. ANSI ISO figure split.

@@
 
-\begin{figure*}[t!]
-\begin{smathpar}
-\begin{array}{lcl}
-(\A,\visZ) \Vdash \eta \in T_i & \defeq & \eta \in \A \conj \txn(\eta) = T_i\\
-(\A,\visZ) \Vdash \eta_1 \visar \eta_2 & \defeq & \{\eta_1,\eta_2\}
-        \subseteq \A \conj (\eta_1,\eta_2) \in \visZ\\
-(\A,\visZ) \Vdash \eta_1 \soar \eta_2 & \defeq & \{\eta_1,\eta_2\}
-        \subseteq \A \conj \txn(\eta_1)=\txn(\eta_2) \conj \id(\eta_1)
-        < \id(\eta_2)\\
-\E \Vdash S \subseteq T_i & \defeq & \forall \eta.~ \eta
-        \in S \Rightarrow \underE{\eta \in T_i} \\
-\E \Vdash T_i \visar \eta & \defeq &\forall\eta_1
-        .\,(\E \Vdash \eta_1 \in T_i) \Rightarrow \E \Vdash \eta_1 \visar \eta \\
-\underE{T_i \visar T_j} & \defeq &  \forall\eta_1,\eta_2.\,
-    %\sameobj{\eta_1}{\eta_2}  \Rightarrow 
-    \underE{\eta_1\in T_i} \conj \underE{\eta_2 \in T_j} \Rightarrow 
-    \underE{\eta_1 \visar \eta_2} \\
-\underE{T_i \invisar T_j} & \defeq &  \forall\eta_1,\eta_2.\,
-        %\sameobj{\eta_1}{\eta_2}\Rightarrow 
-        \underE{\eta_1\in T_i} \conj \underE{\eta_2 \in T_j} \Rightarrow 
-        \neg (\underE{\eta_1 \visar \eta_2})\\
-\underE{T_i \wrstoar X} & \defeq & \exists\eta.~
-        \underE{\eta \in T_i} \conj \kind(\eta) = \C{WR}(X)\\
-\underE{T_i \rdsfmar X} & \defeq & \exists\eta.~
-        \underE{\eta \in T_i} \conj \kind(\eta) = \C{RD}(X)\\
-\underE{T_i \usesar X} & \defeq & \underE{T_i \wrstoar X} \disj
-      \underE{T_i \rdsfmar X}\\
-\underE{\C{RMWVis}(T_j)} & \defeq & \forall\eta_1,\eta_2.\,
-       \underE{\{\eta_1,\eta_2\} \subseteq T_j} \conj
-       \underE{\eta_1 \soar \eta_2} \Rightarrow \underE{\eta_1 \visar
-       \eta_2}\\
-\underE{\C{MonotonicVis}(T_j)} & \defeq & 
-       \underE{\C{RMWVis}(T_j)} \conj 
-       \forall\eta,\eta_1,\eta_2.\, \underE{\{\eta_1,\eta_2\} \in T_j} 
-          \conj \\
-  &   & \hspace*{1.2in}\underE{\eta \visar \eta_1} \conj
-        \underE{\eta_1 \soar \eta_2} \Rightarrow \underE{\eta \visar
-        \eta_2} \\
-%  \C{CausalVis}(T_i) & \defeq & 
-%         \C{MonotonicVis}(T_i) \conj \C{RMWVis}(T_i)\\
-\underE{\C{AtomicVis}(T_j)} & \defeq & 
-       \forall\eta_1,\eta_2.\, \neg(\underE{\eta_1 \in T_j}) \conj
-       \underE{\eta_2 \in T_j} \conj
-       \underE{\eta_1 \visar \eta_2} \Rightarrow \underE{\txn(\eta_1)
-       \visar \eta_2}\\
-\underE{\C{CommitVis}(T_j)} & \defeq & 
-       \forall\eta_1,\eta_2.~ \neg(\underE{\eta_1 \in T_j}) \conj 
-          \underE{\eta_2 \in T_j} \conj
-       \underE{\eta_1 \visar \eta_2} \Rightarrow\\
-  &   & \hspace*{1.2in}\exists\eta.\, \underE{\eta \in \txn(\eta_1)} 
-        \conj \kind(\eta) = \C{COMMIT} 
-        \conj \underE{\eta \visar \eta_2}\\
-% \underE{\C{TransVis}(T_j)} & \defeq &  \forall
-%        \eta_1,\eta_2,\eta_3.\, \underE{\eta_3 \in T_j} \conj
-%        \underE{\eta_1 \visar \eta_2} \conj
-%        \underE{\eta_2 \visar \eta_3} \Rightarrow \underE{\eta_1 \visar
-%        \eta_3} \\
-\underE{\C{RC}(T_j)} & \defeq & \underE{\C{AtomicVis}(T_j)} 
-        \conj \underE{\C{CommitVis}(T_j)}\\
-\underE{\C{MAV}(T_j)} & \defeq & \underE{\C{RC}(T_j)} \conj
-      \underE{\C{MonotonicVis}(T_j)} \\
-\underE{\C{SnapshotVis}(T_i,T_j)} & \defeq &  \underE{T_i
-       \visar T_j} \disj \underE{T_i \invisar T_j}\\
-\underE{\C{RR}(T_j)} & \defeq & \underE{\C{MAV}(T_j)}
-       \conj \forall T_i.\,T_i \neq T_j \Rightarrow 
-        \underE{\C{SnapshotVis}(T_i,T_j)} \\
-% &   & \hspace*{2in}\conj  \C{SnapshotVis}(T_i,T_j)\\
-\underE{\C{SnapshotSER}(T_i,T_j,X)} & \defeq &  \underE{T_i
-       \visar T_j} \disj (\underE{T_i \invisar T_j} \conj \\
-  &   &\hspace*{1.2in}\exists \eta.~\underE{\eta\in T_i} \conj \kind(\eta) =
-       \C{WR(X)} \Rightarrow \underE{T_j \visar \eta})\\
-\underE{\C{SI}(T_j)} & \defeq &  \underE{\C{RR}(T_j)}
-       \conj \forall T_i.\,(T_i \neq T_j \conj \exists X.\, \underE{T_i \wrstoar X} \conj 
-        \underE{T_j \wrstoar X})\\
-%      (\exists X.\, T_i \wrstoar X \conj T_j \wrstoar X)
-%      \Rightarrow  \C{TotalVis}(T_i,T_j)\\
-  &   & \hspace*{2in} \Rightarrow \underE{\C{SnapshotSER}(T_i,T_j,X)}\\
-\underE{\C{SER}(T_j)} & \defeq & \underE{\C{RR}(T_j)}
-        \conj \forall T_i.\,(T_i \neq T_j \conj \exists X.\, 
-        \underE{T_i \wrstoar X} \conj \underE{T_j \usesar X})\\
-  &   & \hspace*{2in} \Rightarrow \underE{\C{SnapshotSER}(T_i,T_j,X)}\\
-\end{array}
-\end{smathpar}
+% ISOLATION SPECS
+% ---------------
+\input{iso-specs}
 
-\caption{Standard isolation guarantees expressed as trace
-well-formedness constraints}
-\label{fig:ansi-isolation}
-\end{figure*}
+We now describe specifications of standard isolation guarantees
+expressed as constraints over trace well-formedness. For brevity and
+convenience, we adopt few notations in this and subsequent sections.
+An execution trace is destructed as ($\A$,$\visZ$) whenever individual
+components of the pair are needed. Otherwise, it is written as $\E$.
+Sometimes, the dot notation (\eg~$\E.\A$) is also used. Since $\A$ and
+$\visZ$ are both sets, we lift the operations on sets to pairs of sets
+when updating $\E$. For example, $\E' = \E \cup
+(\{\eta_2\},\{(\eta_1,\eta_2)\})$ expands to $\E' = (\E.\A \cup
+\{\eta_2\},\,\E.\visZ \cup \{(\eta_1,\eta_2)\})$.  When $\psi$ is a
+formula, $\underE{\psi}$ denotes the interpretation of $\psi$ in the
+context of the trace $\E$. Such interpretations are defined on a
+case-by-case basis in Figs.~\ref{fig:rel-defs}
+and~\ref{fig:ansi-isolation}.
 
-Fig.~\ref{fig:ansi-isolation} shows the specification of standard
-isolation guarantees expressed as constraints over trace
-well-formedness. For brevity and convenience, we adopt few notations
-in this and subsequent sections.  An execution trace is destructed
-into $\A$ and $\visZ$ whenever individual components of the pair are
-needed. Otherwise, it is written as $\E$. Sometimes, the dot notation
-(\eg~$\E.\A$) is also used. Since $\A$ and $\visZ$ are both sets, we
-lift the operations on sets to pairs of sets when updating $\E$. For
-example, $\E' = \E \cup (\{\eta_2\},\{(\eta_1,\eta_2)\})$ expands to
-$\E' = (\E.\A \cup \{\eta_2\},\,\E.\visZ \cup \{(\eta_1,\eta_2)\})$.
-When $\psi$ is a formula, $\underE{\psi}$ denotes the
-interpretation of $\psi$ in the context of the trace $\E$. Such
-interpretations are defined on a case-by-case basis in
-Fig.~\ref{fig:ansi-isolation}. In the following, we give informal
-explanations for each definition.
+Fig.~\ref{fig:rel-defs} shows various relations defined over elements
+in a trace. In the context of a trace $(\A,\visZ)$, an effect $\eta$
+is said to belong to a transaction $T_i$ if $\eta$ belongs to the
+effect set $A$ and its transaction identifier is $T_i$. The
+containment relation is trivially lifted to the set of effects to
+define $\underE{S \subseteq T_i}$.  Visibility and session order
+relations are denoted by $\eta_1 \visar \eta_2$ and $\eta_1 \soar
+\eta_2$, respectively. A transaction $T_i$ is said to be visible to an
+effect $\eta$ if every effect $\eta_1$ of $T_i$ recorded by the trace
+is visible to $\eta$.  $T_i$ may be visible to $\eta$ but may not be
+visible to every other effect in the $\txn(\eta)$. For a transaction
+$T_i$ to be considered to be visible to a transaction $T_j$ in the
+context of a trace $\E$ (written $\underE{T_i \visar T_j}$), every
+effect ($\eta_1$) of $T_i$ present in $\E$ must be visible to every
+effect ($\eta_2$) of $T_j$ in $\E$.  Conversely, if none of the
+effects of $T_i$ present in $\E$ are visible to any effect of $T_j$,
+then $T_i$ is considered invisible to $T_j$ under $\E$ (written
+$\underE{T_i \invisar T_j}$). Transaction $T_i$ is said to have
+written to a variable $X$ under $\E$ (i.e., $\underE{T_i \wrstoar X}$)
+if there exists a write-to-$X$ effect from $T_i$ in $\E$.
+\emph{Reads-from} ($\underE{T_i \rdsfmar X}$) is defined similarly.
+$T_i \usesar X$ if it reads or writes $X$.
 
-In the context of a trace $(\A,\visZ)$, an effect $\eta$ is said to
-belong to a transaction $T_i$ if $\eta$ belongs to the effect set $A$
-and its transaction identifier is $T_i$. The containment relation is
-trivially lifted to the set of effects to define $\underE{S \subseteq
-  T_i}$.  The interpretation of $\eta_1 \visar \eta_2$ and $\eta_1
-\soar \eta_2$ are as explained earlier.  A transaction $T_i$ is said
-to be visible to an effect $\eta$ if every effect $\eta_1$ of $T_i$
-recorded by the trace is visible to $\eta$.  $T_i$ may be visible to
-$\eta$ but may not be visible to every other effect in the
-transaction. For a transaction $T_i$ to be considered to be visible to
-a transaction $T_j$ in the context of a trace $\E$ (written
-$\underE{T_i \visar T_j}$), every effect ($\eta_1$) of $T_i$ present
-in $\E$ must be visible to every effect ($\eta_2$) of $T_j$ in
-$\E$. Conversely, if none of the effects of $T_i$ present in $\E$ are
-visible to any effect of $T_j$, then $T_i$ is considered invisible to
-$T_j$ under $\E$ (written $\underE{T_i \invisar T_j}$). Transaction
-$T_i$ is said to have written to a variable $X$ under $\E$ (i.e.,
-$\underE{T_i \wrstoar X}$) if there exists a write-to-$X$ effect from
-$T_i$ in $\E$. \emph{Reads-from} is similar. $T_i$ \emph{uses} $X$ if
-it reads or writes $X$.
+Fig.~\ref{fig:ansi-isolation} shows various isolation guarantees
+defined as propositions indexed by transaction identifiers.
+Transaction $T_j$ is said to have experienced \emph{read-my-writes}
+visibility under $\E$ if every effect ($\eta_1$) of $T_j$ is visible
+to every subsequent effect ($\eta_2$) of the same transaction in $\E$.
+This lets $T_j$ to never lose its own updates. Monotonic visibility
+adds one more constraint to read-my-writes; besides requiring $\eta_1$
+to be visible to $\eta_2$, it also requires every effect ($\eta$)
+visible to $\eta_1$ in $\E$ to be visible to $\eta_2$ as well. Thus,
+later operations in a transaction witness at least the same set of
+effects witnessed by the earlier operations, if not more (hence,
+``monotonic'' visibility). Atomic visibility allows an effect
+$\eta_2$ of $T_j$ to witness an effect $\eta_1$ of $T_i$ only if all
+effects of $T_i$ in $\E$ are also visible to $\eta_2$. Atomic
+visibility thus prevents a transaction from being partially visible.
+However, atomic visibility does not prevent an uncommitted transaction
+from being visible. This is addressed by \C{CommitVis}, which requires
+the commit effect of $T_i$ to be visible whenever any effect of $T_i$
+is visible.
 
-Various isolation guarantees are defined as propositions indexed by
-a transaction identifier $T_j$. Transaction $T_j$ is said to have
-experienced \emph{read-my-writes} visibility under $\E$ if every
-effect ($\eta_1$) of $T_j$ is visible to every subsequent effect
-($\eta_2$) of the same transaction in $\E$. This lets $T_j$ to never
-lose its own updates. Monotonic visibility adds one more constraint to
-read-my-writes; besides requiring $\eta_1$ to be visible to $\eta_2$,
-it also requires every effect ($\eta$) visible to $\eta_1$ in $\E$ to
-be visible to $\eta_2$ as well. Thus $T_j$ witnesses monotonically
-increasing state as time progresses. Atomic visibility allows an
-effect $\eta_2$ of $T_j$ to witness an effect $\eta_1$ of $T_i$ only
-if all effects of $T_i$ in $\E$ are also visible to $\eta_2$. Atomic
-visibility thus prevents a transaction from being partially visible.
-However, atomic visibility does not prevent an uncommitted
-transaction from being visible. This is addressed by \C{CommitVis},
-which requires the commit effect of $T_i$ to be visible whenever any
-effect of $T_i$ is visible.
-
-The ANSI SQL 92 standard requires \iso{Read Committed} isolation to avoid
-the dirty reads phenomenon, which is achieved by enforcing
+The ANSI SQL 92 standard requires \iso{Read Committed} isolation to
+avoid the dirty reads phenomenon, which is achieved by enforcing
 $\mathtt{AtomicVis}$ and $\mathtt{CommitVis}$ guarantees. The {\sc rc}
-specification is therefore a combination of these two guarantees. The
-specification also agrees with the description and
+specification\footnote{Note the distinction in fonts: {\sc rc} is an
+abbreviation of \iso{Read Committted}, whereas \C{RC} is its
+specification.} (\C{RC}) is therefore a combination of these two
+guarantees. The specification also agrees with the description and
 implementation~\cite{bailishat,pldi15} of {\sc rc} for highly
@@
 \iso{Monotonic Atomic View} ({\sc mav} of
-Fig.~\ref{fig:ansi-isolation})~\cite{bailishat,pldi15} was proposed to
-explicitly extend {\sc rc} with monotonicity on such stores. The sample
-{\sc rc} executions shown in Fig.~\ref{fig:rc-executions} are also
-{\sc mav}.
+Fig.~\ref{fig:ansi-isolation})~\cite{bailishat} was proposed to
+explicitly extend {\sc rc} with monotonicity on such stores.
 
@@
 \subcaptionbox {
-  {\sc rr}($T_1$), {\sc mav}($T_2$).
+  {\sc rr}($T_1$), {\sc rr}($T_2$).
   \label{fig:ansi-iso-eg-rr}
@@
 
-A transaction $T_i$ is said to be snapshot-visible to a transaction
-$T_j$ if either it is visible to $T_j$, or it is invisible; it is
+Snapshot visibility (\C{SnapshotVis}) captures the scenario where a
+transaction executes against a static snapshot of the database.  A
+transaction $T_i$ is said to be snapshot-visible to a transaction
+$T_j$ if either it is visible to $T_j$ (i.e., $T_i$ is included in the
+snapshot), or it is invisible (i.e., it is not included); it is
 forbidden for only a suffix (more generally, a subset) of $T_j$ to
-witness $T_i$.  \iso{Repeatable Read} ({\sc rr})
-isolation extends {\sc mav} with the snapshot visibility guarantee.
-Observe that snapshot visibility permits $T_i$ and $T_j$ to execute
-and commit while being oblivious of each other. This scenario is
-captured in Fig.~\ref{fig:ansi-iso-eg-rr}, where transactions $T_1$
-and $T_2$, which perform conflicting writes, execute against a
-snapshot of the database and commit concurrently.
+witness $T_i$. \iso{Repeatable Read} ({\sc rr}) isolation extends {\sc
+mav} with the snapshot visibility guarantee.  Observe that snapshot
+visibility permits $T_i$ and $T_j$ to execute and commit while being
+oblivious of each other. This scenario is captured in
+Fig.~\ref{fig:ansi-iso-eg-rr}, where transactions $T_1$ and $T_2$,
+which perform conflicting writes, execute against a snapshot of the
+database and commit concurrently. While the actual values read and
+written by $T_1$ and $T_2$ are unimportant (hence, elided), it is
+important to note the \emph{absence} of visiblity arrows from $T_2$ to
+$T_1$, although $T_2$ commits before $T_1$'s read-from-$Y$.
 
-\iso{Snapshot Isolation} ({\sc SI}) proscribes this possibility. If
+\iso{Snapshot Isolation} ({\sc si}) proscribes this possibility. If
 $T_i$ and $T_j$ both write to the same shared variable, then {\sc si}
@@
 % \iso{Snapshot Isolation} spec ({\sc si}) extends {\sc rr} with a
-% one-way serializability guarantee w.r.t the transactions that perform
+% one-way serializability guarantee w.r.t. the transactions that perform
 % conflicting writes (i.e., writes to the same shared variable).

### 2016-07-03 motivation.tex
Typos fixed. ANSI ISO figure split.

@@
 
-As an alternative, consider the execution of this transaction under a
-\emph{read committed} ({\sc rc}) isolation level, which is weaker than
-     {\sc ser}\footnote{{\sc rc} is in fact the default isolation
-       level in Postgres 9.5 and Oracle 11g databases.} An {\sc rc}
-     transaction is isolated from the writes of uncommitted
-     transactions, thus preventing the transaction from witnessing
-     \emph{dirty reads}~\cite{berenson}, reads that observe the
-     effects of non-committed transactions. In the current example,
-     {\sc rc} isolation admits the two executions shown in
-     Fig.~\ref{fig:rc-ex} on a strongly consistent ({\sc sc}) store,
-     such as a relational database:
+As an alternative, consider the execution of this transaction under
+\emph{Read Committed} ({\sc rc}) isolation level, which is weaker than
+{\sc ser}\footnote{{\sc rc} is in fact the default isolation level in
+Postgres 9.5 and Oracle 11g databases.} An {\sc rc} transaction is
+isolated from the writes of uncommitted transactions, and is therefore
+free from \emph{dirty reads}~\cite{berenson} that read uncommitted
+data. In the current example, {\sc rc} isolation admits the following
+two executions (Fig.~\ref{fig:rc-ex}) on a strongly consistent ({\sc
+sc}) store, such as a relational database:
 
@@
   a read committed isolation level.}
-\label{fig:rc-executions}
+\label{fig:rc-ex}
 \end{figure}
@@
 operations.  The subscript of an operation indicates the transaction
-that executes it.\footnote{For clarity, the effects of different
-  transactions are shown in different colored backgrounds.} In the
-first execution, transaction {\bf Wd1} reads the current balance
-(\C{k}) and writes the new balance (\C{k-a1}), but before it commits
-transaction {\bf Wd2} executes and commits, writing the new balance
-(\C{k-a2}). RC isolation prevent {\bf Wd2} from witnessing the
-uncommitted writes of transaction {\bf Wd1}.  Subsequently committing
-{\bf Wd1} leads to the loss of {\bf Wd2}'s updates (the so called
-\emph{lost update anomaly}~\cite{berenson}), resulting in an incorrect
-balance of \C{k-a1}. The second execution describes a similar scenario
-with {\bf Wd1} and {\bf Wd2} exchanging their roles.  Clearly, {\sc
-  rc} is an excessively weak isolation level for this program because
-it loses the updates of one transaction, resulting in the violation of
-the post-condition.  A stronger isolation level that prevents lost
-updates is required.  \iso{Snapshot Isolation} ({\sc
+that executes it. In the first execution, transaction \C{Wd1} reads
+the current balance (\C{k}) and writes the new balance (\C{k-a1}), but
+before it commits transaction \C{Wd2} executes and commits, writing
+the new balance (\C{k-a2}). {\sc rc} isolation prevent \C{Wd2} from
+witnessing the uncommitted writes of transaction \C{Wd1}.
+Subsequently committing \C{Wd1} leads to the loss of \C{Wd2}'s updates
+(the so called \emph{lost update anomaly}~\cite{berenson}), resulting
+in an incorrect balance of \C{k-a1}. The second execution describes a
+similar scenario with \C{Wd1} and \C{Wd2} exchanging their roles.
+Clearly, {\sc rc} is an excessively weak isolation level for this
+program because it loses the updates of one transaction, resulting in
+the violation of the post-condition.  A stronger isolation level that
+prevents lost updates is required. \iso{Snapshot Isolation} ({\sc
 si})~\cite{berenson} fits this requirement; {\sc si} effectively
@@
 and re-executing a transaction if write-write conflicts are detected
-during its commit.\footnote{{\sc si} exhibits different behaviour from
-  {\sc ser} in the absence of write-write conflicts.}   Since {\sc si},
-unlike {\sc ser}, does not rely on expensive lock-based
-concurrency control, it is also more efficient, making it appropriate
-for both {\bf Wd} transactions.
+during its commit.\footnote{{\sc si} however does not serialize
+transactions in the absence of write-write conflicts.} Since {\sc si},
+unlike {\sc ser}, does not rely on expensive lock-based concurrency
+control, it is also more efficient, making it appropriate for both
+\C{Wd} transactions.
 
@@
 programmers to understand low-level implementation details of the
-database that are naturally far removed from application semantics. An
-attractive (and principled) alternative in this context would be a
-formal proof system that combines declarative reasoning about
-isolation guarantees with operational reasoning about programs. We
-demonstrate how our proof system makes this possible in the context of
-the current example.
+database that are far removed from the application semantics. An
+attractive  alternative in this context would be a principled
+reasoning approach that combines declarative reasoning about isolation
+guarantees with operational reasoning about programs. We demonstrate
+how our proof system makes this possible in the context of the current
+example.
 
@@
 
-\caption{$\hbZ$ arrows allowed by {\sc si} (solid) and {\sc rc} (solid
+\caption{$\visZ$ arrows allowed by {\sc si} (solid) and {\sc rc} (solid
   and dashed) isolation levels for the example in
-  Fig.~\ref{fig:motiv-eg-1}. Each arrow represents an $\hbZ$ relationship
+  Fig.~\ref{fig:motiv-eg-1}. Each arrow represents an $\visZ$ relationship
   among all operations of `Wd2' and operations in `Wd1' that follow
@@
 interference in the current thread. However, in the presence of
-transactions executing under weak isolation, determining what
-constitutes an interference is a non-trivial problem. For example,
-inside a serializable transaction there is no interference, whereas
-within a transaction executing under a \iso{Read Uncommitted}
-isolation level, the weakest isolation level, all interferences are
-valid. Between these two extremes are various levels of isolation that
-admit some interferences while prohibiting others.  For example,
-\iso{Read Committed} isolation admits interference of a committed
-transaction, but not that of an uncommitted transaction.
-\iso{Snapshot Isolation} admits interference from committed
-non-conflicting transactions, and so on.
+transactions executing under various levels of weak isolation,
+determining what constitutes an interference is a non-trivial problem.
+For example, a \iso{Serializable} transaction admits no interference,
+whereas within a transaction executing under \iso{Read Uncommitted}
+isolation all interferences are valid. Between these two extremes are
+various levels of isolation that admit some interferences while
+prohibiting others. For example, \iso{Read Committed} isolation admits
+interference of a committed transaction, but not that of an
+uncommitted transaction. \iso{Snapshot Isolation} admits interference
+from committed non-conflicting transactions, and so on.
 
@@
 allowed interference in accordance with the chosen level of isolation
-associated with a transaction. The first observation we make is
-that we can constrain interference by constraining the definition of
-the \emph{happens-before} ($\hbZ$) relation that dictates visibility
+associated with a transaction. The first observation we make is that
+we can constrain interference by constraining the definition of the
+\emph{happens-before} ($\hbZ$) relation that dictates visibility
 actions between transactions, or their respective constituents.  To do
 so, we axiomatize the $\hbZ$ relation to capture the interference
-characteristics of different isolation levels as their specifications
-($\psi$). For example, a \iso{Read Committed} specification allows an
-operation $\eta_1$ of $T_1$ to happen before an operation $\eta_2$ of
-$T_2$ only if every operation in $T_1$ (including its commit) happens
-before $\eta_2$ (we let $T_1 \hboar \eta_2$ denote this). The 
-specification does not require $T_1$ to execute and commit before all
-operations of $T_2$, thus allowing it to interfere in $T_2$. The
-\iso{Snapshot Isolation} specification requires two transactions that
-write to the same variable to be related by happens-before.  This
-effectively prohibits interference due to actions in $T_1$ being
-interleaved in $T_2$ (or, vise versa).  We formalize these properties
-in Fig.~\ref{fig:interference-ex}.\footnote{Note that specifications
-  presented here are solely for illustration. The actual
-  specifications described in \S\ref{sec:ansi-isolation} are more
-  nuanced.}.
+characteristics of different isolation levels as their specifications.
+Two such specifications are shown\footnote{Note that specifications
+presented here are solely for illustration. The actual specifications
+described in \S\ref{sec:ansi-isolation} are more nuanced.} in
+Fig.~\ref{fig:interference-ex}.  \iso{Read Committed} specification
+($\psi_{RC}$) allows an operation $\eta_1$ of $T_1$ to happen before
+an operation $\eta_2$ of $T_2$ only if every operation in $T_1$
+(including its commit) happens before $\eta_2$ (we let $T_1 \hboar
+\eta_2$ denote this). The specification does not require $T_1$ to
+execute and commit before all operations of $T_2$, thus allowing it to
+interfere in $T_2$. The \iso{Snapshot Isolation} specification
+($\psi_{SI}$) requires two transactions that write to the same
+variable to be related by happens-before.  This effectively prohibits
+interference due to actions in $T_1$ being interleaved in $T_2$ (or,
+vise versa).
 
 Fig.~\ref{fig:motiv-eg-1-hb} is a visualization of $\hbZ$ 
-from {\bf Wd2} to {\bf Wd1} allowed by $\psi_{RC}$ and $\psi_{SI}$. All
+from \C{Wd2} to \C{Wd1} allowed by $\psi_{RC}$ and $\psi_{SI}$. All
 arrows are legal under $\psi_{RC}$ because in no case does an operation
-$\eta_1$ from {\bf Wd1} happen-before $\eta_2$ of {\bf Wd2} without the commit
-of {\bf Wd1} also happening before $\eta_2$. $\psi_{SI}$ however disallows
-all dashed arrows, because dashed arrows establish $\hbZ$ between actions in {\bf Wd2}
-and only a subset of the operations in {\bf Wd1}. Ligher (blue) arrows with
+$\eta_1$ from \C{Wd1} happen-before $\eta_2$ of \C{Wd2} without the commit
+of \C{Wd1} also happening before $\eta_2$. $\psi_{SI}$ however disallows
+all dashed arrows, because dashed arrows establish $\hbZ$ between actions in \C{Wd2}
+and only a subset of the operations in \C{Wd1}. Ligher (blue) arrows with
 hollow heads denote $\hbZ$ relationships that do not affect the value
@@
 prohibited. For instance, consider the program point after the write
-to \C{C} in {\bf Wd1}. The expected invariant ($\phi$) at that program
+to \C{C} in \C{Wd1}. The expected invariant ($\phi$) at that program
 point is shown below ($\committed$ stands for ``committed''):
@@
 
-\noindent $\phi$ asserts that {\bf Wd1} is not yet committed, and that it wrote to
+\noindent $\phi$ asserts that \C{Wd1} is not yet committed, and that it wrote to
 \C{C}, and the value of \C{C} is either \C{k-a1-a2} or \C{k-a1}
-depending on whether or not {\bf Wd2} is committed. If $\phi$ remains
-invariant until {\bf Wd1} commits, then the post-condition (\C{C =
+depending on whether or not \C{Wd2} is committed. If $\phi$ remains
+invariant until \C{Wd1} commits, then the post-condition (\C{C =
 k-a1-a2}) can be established easily. However, an interference from
-{\bf Wd2} at this stage (captured by the last dashed arrow in
+\C{Wd2} at this stage (captured by the last dashed arrow in
 Fig.~\ref{fig:motiv-eg-1-hb}) may violate the invariant by writing
-$\C{k-a1}$ to \C{C} and committing {\bf Wd2}, thus leading to
+$\C{k-a1}$ to \C{C} and committing \C{Wd2}, thus leading to
 $\committed(\C{Wd2}) \Rightarrow \C{C=k-a2}$. Fortunately,
 \iso{Snapshot Isolation} prevents this interference, and this can be
-shown by demonstrating that an interference from {\bf Wd2} starting from an
+shown by demonstrating that an interference from \C{Wd2} starting from an
 execution state that satisfies $\psi_{SI} \wedge \phi$ leads to an
@@
 \hboar \C{Wd2}$ holds (the former does not hold because at least one write
-from {\bf Wd1} has happened before at least one operation of {\bf Wd2}, and
-the latter does not hold because {\bf Wd1} has not yet committed, and {\bf Wd2}
-has already begun). Since {\bf Wd2} also writes to \C{C}, this violates
+from \C{Wd1} has happened before at least one operation of \C{Wd2}, and
+the latter does not hold because \C{Wd1} has not yet committed, and \C{Wd2}
+has already begun). Since \C{Wd2} also writes to \C{C}, this violates
 the $\psi_{SI}$ constraint which we assume to be an invariant. A proof
@@
 subsequent operations. The natural behavior of {\sc sc} to totally
-order all operations w.r.t the $\hbZ$ relation could be in conflict
+order all operations w.r.t. the $\hbZ$ relation could be in conflict
 with the constraints imposed by weak isolation. To meet the

### 2016-07-03 reasoning.tex
Typos fixed. ANSI ISO figure split.

@@
 $\rg{P,R}{c}{G,Q}$, which informally asserts that if a
-command $c$ is executed in a state that satisfies its precondition
+command $c$ is executed in a state that satisfies its pre-condition
 $P$, provided that every interference step during the execution is
@@
 abstract machine of Fig.~\ref{fig:txnimp} takes a step starting from
-an execution $\E$ that satisfies the precondition $P$, then the
+an execution $\E$ that satisfies the pre-condition $P$, then the
 resultant execution $\E'$ satisfies the post-condition $Q$, and that
@@
 % an execution $\E$, should result in an execution $\E'$ that satisfies
-% the postcondition $Q$, while being related to $\E$ via the guarantee
+% the post-condition $Q$, while being related to $\E$ via the guarantee
 % $G$. RG judgment for the \C{txn} lexical block is similar.  It uses
@@
 % $\E$ results in an execution $\E'$ that satisfies transaction's
-% postcondition $Q$. As usual, $\E$ and $\E'$ need to be related by $G$.
+% post-condition $Q$. As usual, $\E$ and $\E'$ need to be related by $G$.
 % The rules for sequential composition of transaction-bound comands
@@
   of rules.} As is the case with a standard rely-guarantee
-formulation,the rules allow the precondition $P$ and the rely relation
+formulation,the rules allow the pre-condition $P$ and the rely relation
 $R$ to be strengthened, and the post-condition $Q$ (also, $C$ in the

### 2016-07-03 store-consistency.tex
Typos fixed. ANSI ISO figure split.

@@
 A strongly-consistent {\sc sc} store guarantees a total order on all
-operations w.r.t $\visZ$ consistent with their chronological order. A
+operations w.r.t. $\visZ$ consistent with their chronological order. A
 straightforward $\I_s$ for this store is the \C{SC} property
@@
   extending $\E$ with $\visZ(\eta_1,\eta_2)$ violates
-  $\I_c$\footnote{\GK{ToDo: consider other well-formedness conditions
-  on trace, such as acyclicity of $\visZ$ and $\soZ$. Are they needed
-  (considering that the machine never violates them)? Encode the
-  specifications in Z3 and make sure they are consistent.}}:
+  $\I_c$:
   \begin{smathpar}
@@
 % concurrent with $\eta$, i.e., $\not\exists\eta'.~\eta' \in S_2 - S_1
-% \conj % visZ(\eta,\eta')$, then any precondition $P$ that is valid
+% \conj % visZ(\eta,\eta')$, then any pre-condition $P$ that is valid
 % when $\op$ executed is also valid when $\eta$ is witnessed because
@@
 such demands can be made via the trace invariant $\I$. For a
-precondition ($P$) of the operation to be useful, it has to be an
+pre-condition ($P$) of the operation to be useful, it has to be an
 assertion over every causally consistent subset of the global state.
 Since any replica that eventually executes the operation has to expose
-one such subset ($S$), the precondition is guaranteed to hold
+one such subset ($S$), the pre-condition is guaranteed to hold
 regardless of the replica. There is however one problem with this
@@
 i.e., we write $\psi$ instead of $\underE{\psi}$. The proposition \C{k
-$\ge$ a1+a2} remains an invariant, hence elided. The precondition
+$\ge$ a1+a2} remains an invariant, hence elided. The pre-condition
 $P_1$ of \C{Wd1} accounts for the possibility of \C{Wd2} committing
 before \C{Wd1}, writing \C{k-a2} to \C{C}. Since neither \C{Wd1} nor
-\C{Wd2} are committed at the beginning, the precondition $P$ of the
+\C{Wd2} are committed at the beginning, the pre-condition $P$ of the
 program is extended with $\neg\committed(\C{Wd1}) \wedge
@@
 triple associated with the assignment to \C{C} for illustration. The
-precondition $\phi_4$ asserts different values for \C{C} and \C{v3}
+pre-condition $\phi_4$ asserts different values for \C{C} and \C{v3}
 (temporary alias of RHS expression) depending on whether or not
@@
 
-The stability for $\phi_4$ w.r.t $R_1$ is trivial to establish: if
+The stability for $\phi_4$ w.r.t. $R_1$ is trivial to establish: if
 $R_1$ does not commit \C{Wd2}, then the value of \C{C} is unaffected,

### 2016-07-03 ansi-isolation.tex
SI and SER figs modified.

@@
 execution where {\sc si} transaction $T_1$ witnesses a snapshot of the
-database that doesn't include $T_2$, but the conflicting write to $X$
-in $T_2$, and the subsequent operations (due to the monotonicity
-requirement of {\sc mav}) witness $T_1$. Thus, $T_2$ is aware of $T_1$
-when it commits. Lastly, \iso{Serializable} isolation extends
-$\mathtt{SnapshotSER}$ to also cover non-conflicting transactions that
-write to variables read in the current transaction.
-Fig.~\ref{fig:ansi-iso-eg-ser} shows an execution of an {\sc ser}
-transaction $T_3$ and a {\sc mav transaction} $T_4$, which write to $X$
-and $Y$, respectively. The snapshot witnessed by the {\sc ser}
-transaction $T_3$ does not include $T_4$, but the write to $Y$ in
-$T_4$, although non-conflicting, witnesses $T_1$ because $Y$ is read
-by $T_1$. Visibility includes other operations of $T_4$ because of
-{\sc mav}.
+database that doesn't include $T_2$. Due to \C{SI($T_1$)}, the
+conflicting write to $X$ in $T_2$ is now required to witness $T_1$, as
+captured by the direction of the $\visZ$ arrow (subsequent operations
+also witness $T_1$ because $T_2$ is executing under {\sc mav}).
+Lastly, \iso{Serializable} isolation extends $\mathtt{SnapshotSER}$ to
+also cover non-conflicting transactions that write to variables read
+in the current transaction.  Fig.~\ref{fig:ansi-iso-eg-ser} shows an
+execution of an {\sc ser} transaction $T_3$ and a {\sc mav
+transaction} $T_4$, which write to $X$ and $Y$, respectively. The
+snapshot witnessed by the {\sc ser} transaction $T_3$ does not include
+$T_4$, but the write to $Y$ in $T_4$, although non-conflicting,
+witnesses $T_1$ because $Y$ is read by $T_1$. Visibility includes
+other operations of $T_4$ because of {\sc mav}.
 

### 2016-07-03 ansi-isolation.tex
Edits.

@@
 in the current transaction.  Fig.~\ref{fig:ansi-iso-eg-ser} shows an
-execution of an {\sc ser} transaction $T_3$ and a {\sc mav
-transaction} $T_4$, which write to $X$ and $Y$, respectively. The
+execution of an {\sc ser} transaction $T_3$ and a {\sc mav}
+transaction $T_4$, which write to $X$ and $Y$, respectively. The
 snapshot witnessed by the {\sc ser} transaction $T_3$ does not include
@@
 witnesses $T_1$ because $Y$ is read by $T_1$. Visibility includes
-other operations of $T_4$ because of {\sc mav}.
+other operations of $T_4$ because of {\sc mav}. Note that {\sc ser}
+does not guarantee total order of transactions w.r.t. $\visZ$ because,
+in reality, databases do not guarantee serializable execution unless
+\emph{all} transactions choose {\sc ser}.
 

### 2016-07-03 ansi-isolation.tex
Edits.

@@
 forbidden for only a suffix (more generally, a subset) of $T_j$ to
-witness $T_i$. \iso{Repeatable Read} ({\sc rr}) isolation extends {\sc
-mav} with the snapshot visibility guarantee.  Observe that snapshot
-visibility permits $T_i$ and $T_j$ to execute and commit while being
-oblivious of each other. This scenario is captured in
-Fig.~\ref{fig:ansi-iso-eg-rr}, where transactions $T_1$ and $T_2$,
-which perform conflicting writes, execute against a snapshot of the
-database and commit concurrently. While the actual values read and
-written by $T_1$ and $T_2$ are unimportant (hence, elided), it is
-important to note the \emph{absence} of visiblity arrows from $T_2$ to
-$T_1$, although $T_2$ commits before $T_1$'s read-from-$Y$.
+witness $T_i$. Specification of ANSI SQL \iso{Repeatable Read}
+isolation (\C{RR}) extends \C{MAV} with the snapshot visibility
+guarantee.  Observe that snapshot visibility permits $T_i$ and $T_j$
+to execute and commit while being oblivious of each other. This
+scenario is captured in Fig.~\ref{fig:ansi-iso-eg-rr}, where
+transactions $T_1$ and $T_2$, which perform conflicting writes,
+execute against a snapshot of the database and commit concurrently.
+While the actual values read and written by $T_1$ and $T_2$ are
+unimportant (hence, elided), it is important to note the
+\emph{absence} of visiblity arrows from $T_2$ to $T_1$, although $T_2$
+commits before $T_1$'s read-from-$Y$.
 

### 2016-07-03 store-consistency.tex
Previously uncommitted parts of ANSI ISO figure.

@@
 
-First, we observe that the semantics of a weakly-consistent data store
-can be captured by store-specific consistency constraints, along with
+First, we observe that the consistency level of a data store can be
+captured by store-specific consistency constraints, along with
 transaction-specific isolation constraints, via the trace invariant
@@
 program-specific (or, client-specific) invariant, to capture
-consistency and isolation constraints, resp.  $\I$ is now a
+consistency and isolation constraints, respectively.  $\I$ is now a
 conjunction: $\I \,=\, \lambda\E.~\I_s(\E) \wedge \I_c(\E)$ (often
 simplified to $\I \,=\, \I_s\wedge\I_c$).  While the program-specific
-trace invariant ($\I_c$) remains an invariant regardless of the
-specific consistency and visibility features exposed by the underlying
-data store, the store specific invariant ($\I_s$) changes from store
-to store depending on the consistency level.  For an eventually
-consistent store, $\I_s$ is simply \emph{true}. Stronger stores, such
-as those that support strong consistent, have non-trivial definitions
-for $\I_s$.
+trace invariant ($\I_c$) is independent of the underlying data store,
+the store specific invariant ($\I_s$) changes from store to store
+depending on the consistency level.  For an eventually consistent
+store, $\I_s$ is simply \emph{true}. Stronger stores, such as those
+that support strong consistency, have non-trivial definitions for
+$\I_s$.
 
@@
 violates $\I_c$, whereas not letting it be visible violates $\I_{s}$.
-The only way to satisfy both invariants is to rule out all 
-executions that interleave the operations of one transaction with the
-other, thereby enforcing serializability.  In general, when $\I_s$ conflicts (but is not
-inconsistent) with $\I_c$, the only way to enforce both invariant sets
-is to restrict concurrency. Clearly, this is unacceptable since it
+The only way to satisfy both invariants is to rule out all executions
+that interleave the operations of one transaction with the other,
+thereby enforcing serializability.  In general, when $\I_s$ conflicts
+(but is not inconsistent) with $\I_c$, the only way to enforce both is
+to restrict concurrency. Clearly, this is unacceptable since it
 defeats the very purpose of supporting weak isolation. 
@@
 are violated iff they are found to be in conflict with the constraints
-imposed by a transaction's isolation level. In the context of
-the aforementioned example, $\eta_1$ is not made visible to $\eta_2$
-because doing so would violate $\I_c$. However, if $\I_c$ is true,
-then the store makes $\eta_1$ visible to $\eta_2$ to honor its
-consistency commitment\footnote{The term \emph{recency
-    commitment}~\cite{bailishat} is often used in practice
-  to capture the best-effort nature of {\sc sc}.}. We generalize this
-approach to any $\I_s$ and $\I_c$ by defining a \emph{maximum
-  visibility principle} to determine an acceptable weakening of $\I_s$
-in case of a conflict with $\I_c$.  The principle requires the
-weakened consistency guarantee ($\I_s'$) of the store to enforce all
-visibility relationships imposed by the actual consistency guarantee
-($\I_s$), unless enforcing such a relationship violates $\I_c$.
+imposed by a transaction's isolation level. In the context of the
+aforementioned example, $\eta_1$ is not made visible to $\eta_2$
+because doing so would violate $\I_c$. However, if $\I_c$ is redefined
+to \emph{true}, then the store makes $\eta_1$ visible to $\eta_2$ to
+honor its consistency commitment\footnote{The term \emph{recency
+commitment}~\cite{bailishat} is often used in practice to capture the
+best-effort nature of {\sc sc}.}. We generalize this approach to any
+$\I_s$ and $\I_c$ by defining a \emph{maximum visibility principle} to
+determine an acceptable weakening of $\I_s$ in case of a conflict with
+$\I_c$.  The principle requires the weakened consistency guarantee
+($\I_s'$) of the store to enforce all visibility relationships imposed
+by the actual consistency guarantee ($\I_s$), unless enforcing such a
+relationship violates $\I_c$.
 % The formal definition of the principle is relegated
@@
 \end{smathpar}
+% GK : Please talk to me before changing the substance of the following paragraph.
 $\I_s$ requires a trace $\E$ to satisfy the visibility constraints of
@@
 Fig.~\ref{fig:txnimp} results in an operational semantics that
-    describes many kinds of relational database-like data stores,
-    including causally-consistent data stores, among
-    others.\footnote{Details of this instantiation are relegated to
-      the supplementary in the interest of space.}
+describes {\sc sc} stores with on-demand weak isolation, such as
+off-the-shelf relational databases. Similar approach can be employed
+to capture the semantics of other kinds of stores, such as a causally
+consistent ({\sc cc}) data store, as specific instantiations of our
+operational semantics\footnote{Details of the {\sc cc} instantiation
+are relegated to the supplementary in the interest of space.}.
 

### 2016-07-03 case-studies.tex
edits

--- 2016-07-03 case-studies.tex
@@
 In this section, we present two case studies that demonstrate the
-applicability of our reasoning approach in varied settings, and the
+applicability of our reasoning approach in varied settings, and its
 consequent practical utility.
@@
   tools we developed to analyze Ruby can be found at the following
-anonymized url: \url{...}}. The application stores the video count in
-a Postgres database, and supports a read operation (a database read
+  anonymized url: \url{...}}. The application stores the video count
+in a Postgres database, and supports a read operation (a database read
 transaction), and an increment operation (a read-increment-write
@@
 invariant\footnote{The representative \txnimp programs we used to
-formally reason about scenarios described in this section can be found
-in the supplementary.} In particular, we were able to formally prove
-that to preserve the invariant, writers need to execute at
-\iso{Repeatable Read} isolation, while readers can continue at {\sc
-rc}. Repeating the experiments with this configuration led to no
-violations, as expected. While executing all transactions at {\sc ser}
-also prevents violations, it led to an additional 47\% increase in
-latency on an average (16 concurrent writers; 5 rounds). We repeated
-the exercise for MySQL, and found that its {\sc rr} is not sufficient
-to enforce monotonicity; {\sc ser} is needed.
+  formally reason about scenarios described in this section can be
+  found in the supplementary material.} In particular, we were able to
+formally prove that to preserve the invariant, writers need to execute
+at a \iso{Repeatable Read} isolation level, while readers can continue
+to operate using \iso{Read Committed} isolation.  While executing all
+transactions at {\sc ser} also prevents violations, it led to an
+additional 47\% increase in latency on an average (16 concurrent
+writers; 5 rounds). We repeated the exercise for MySQL, and found that
+using {\sc rr} was not sufficient to enforce monotonicity, but that
+{\sc ser} was needed, due to differences in how {\sc rr} is
+implemented between the two systems.
 
@@
 % techniques.   
-To understand the effect of changing the store consistency, we
+To understand the effect of changing the store consistency level, we
 reimplemented the counter as a \emph{replicated data type}
@@
 observed monotonicity violations. On the other hand, instantiating the
-store consistency with {\sc cc} allowed us to obtain a proof, and
-recover the monotonicity invariant. We in fact obtained a proof after
-relaxing the serializability requirement for writes, by changing the
-definition of the $\interp{\E}(X)$ (\S\ref{sec:opsem}) to return the
-value of the largest write-to-$X$ instead of the latest write-to-$X$.
-We implemented this version of the counter RDT, and obtained a 52\%
-improvement in latency compared to the previous
+store consistency level with {\sc cc} allowed us to obtain a proof,
+and recover the monotonicity invariant. We in fact obtained a proof
+after relaxing the serializability requirement for writes, by changing
+the definition of the $\interp{\E}(X)$ (\S\ref{sec:opsem}) to return
+the value of the largest write-to-$X$ instead of the latest
+write-to-$X$.  We implemented this version of the counter RDT, and
+obtained a 52\% improvement in latency compared to the previous
 implementation\footnote{The tradeoff is that the view count is now no
-longer accurate, although it grows monotonically.}.
+  longer accurate, although it grows monotonically, thus preserving
+  the application invariant.}.
 

### 2016-07-03 related-work.tex
edits

@@
 \paragraph{Specifying weak isolation.}
-Adya~\cite{adyaphd} specified several weak isolation levels in terms
+Adya~\cite{adyaphd} specifies several weak isolation levels in terms
 of \emph{dependency graphs} between transactions, by defining the
@@
 operational nature of Adya's specifications make them suitable for
-runtime monitoring and anomaly
+runtime monitoring and anamoly
 detection~\cite{kemmevldb,feketesigmod08,pssi2011}, whereas the
-declarative nature of our specifications make them suitable for
+declarative nature of our specifications make them suitable for formal
 reasoning about program behaviour. Sivaramakrishnan \emph{et
@@
 suitable to reason operationally about a program as it builds the
-trace. Cerone \emph{et al.}~\cite{gotsmanconcur15} present a framework
-to specify isolation levels with atomic visibility, but their
-specifications are also for complete traces; Moreover, their
-formulation implicitly assumes homogeneity, i.e., they require all
-transactions to execute using either {\sc si} or {\sc ser}, which is
-often not the case in practice. While all the aforementioned
-specification frameworks use the vocabulary introduced
-in~\cite{burckhardt14} to specify replicated data types and different
-forms of weak consistency, in contrast to this work, none of them come
+trace; their framework also requires application invariants be
+expressed in terms of the visibility relations used to define
+isolation and consistency properties.  Cerone \emph{et
+  al.}~\cite{gotsmanconcur15} present a framework to specify isolation
+levels with atomic visibility, but their specifications are also for
+complete traces; moreover, their formulation implicitly assumes
+homogenity, \emph{i.e.}, they require all transactions to execute
+using either {\sc si} or {\sc ser}, which is often not the case in
+practice. While all the aforementioned specification frameworks use
+the vocabulary introduced in~\cite{burckhardt14} to specify replicated
+data types and different forms of weak consistency, none of them come
 with an associated reasoning framework that can use such
@@
 \paragraph{Reasoning under weak isolation} In~\cite{feketessi}, Fekete
-\emph{et al.} propose a theory to characterize non-serializable executions
-that arise under {\sc si}. Fekete~\cite{fekete2005} also proposes an
-algorithm that allocates either {\sc si} or {\sc ser} isolation levels
-to transactions while guaranteeing the serializability of the
-execution. In~\cite{gotsmanpodc16}, Cerone \emph{et al.} improve on Adya's
-{\sc si} specification and use it to derive a static analysis that
-determines the safety of substituting {\sc si} with a weaker variant
-called \iso{Parallel Snapshot Isolation}~\cite{psi}. All of them focus
-on establishing the equivalence of executions between a pair of
-isolation levels, without taking application invariants into account.
-Bernstein \emph{et al.}~\cite{bern2000} propose informal semantic conditions
-to ensure the satisfaction of application invariants under weaker
-isolation levels.  However, no formal proof system has been proposed.
-Furthermore, all the aforementioned approaches are tailor-made for a
-finite set of well-understood isolation levels (rooted
-in~\cite{berenson}) under a pre-defined consistency model (usually
-{\sc sc}).
+\emph{et al.} propose a theory to characterize non-serializable
+executions that arise under {\sc si}. Fekete~\cite{fekete2005} also
+proposes an algorithm that allocates either {\sc si} or {\sc ser}
+isolation levels to transactions while guaranteeing the
+serializability of the execution. In~\cite{gotsmanpodc16}, Cerone
+\emph{et al.} improve on Adya's {\sc si} specification and use it to
+derive a static analysis that determines the safety of substituting
+{\sc si} with a weaker variant called \iso{Parallel Snapshot
+  Isolation}~\cite{psi}.  These efforts focus on establishing the
+equivalence of executions between a pair of isolation levels, without
+taking application invariants into account.  Bernstein \emph{et
+  al.}~\cite{bern2000} propose informal semantic conditions to ensure
+the satisfaction of application invariants under weaker isolation
+levels.  However, no formal proof system is given.  Furthermore, these
+techniques are tailor-made for a finite set of well-understood
+isolation levels (rooted in~\cite{berenson}) under a pre-defined
+store consistency model (usually {\sc sc}).
 
@@
 weak consistency level (\emph{e.g.}, {\sc ec}~\cite{redblueosdi,
-  redblueatc, ecinec, alvarocalm, bailisvldb} or causal consistency
-({\sc cc})~\cite{gotsmanpopl16}). In all these efforts, the key
-technical issue involves proving that atomic and fully isolated
-operations preserve application invariants when executed on these
-consistency levels.  In contrast, we admit weakly isolated
-transactions (sets of atomic operations), and our system model accepts
-\emph{specifications} of consistency and isolation levels drawn from
-an expressive logic.  In recent work~\cite{gotsmanpopl16},
-\iso{Parallel Snapshot Isolation} is adapted to the aforementioned
-setting by interpreting it as a consistency level that serializes
-writes to objects; a dedicated proof rule is developed to help prove
-prove program invariants hold under this model. By parameterizing our
-proof system over a gamut of weak isolation specifications, we avoid
-the need to define a separate proof rule for each new isolation level.
+  redblueatc, ecinec, alvarocalm, bailisvldb}) or causal
+consistency~\cite{gotsmanpopl16}). The key technical issue in these
+efforts involves proving that atomic and fully isolated operations
+preserve application invariants when executed on these consistency
+levels.  In contrast, we admit weakly isolated transactions, and our
+system model accepts \emph{specifications} of consistency and
+isolation levels drawn from an expressive logic.  In recent
+work~\cite{gotsmanpopl16}, \iso{Parallel Snapshot Isolation} is
+adapted to the aforementioned setting by interpreting it as a
+consistency level that serializes writes to objects; a dedicated proof
+rule is developed to help prove prove program invariants hold under
+this model. By parameterizing our proof system over a gamut of weak
+isolation specifications, we avoid the need to define a separate proof
+rule for each new isolation level we may encounter.
 
-\paragraph{Reasoning under relaxed memory} Considering the
-correspondences between axiomatic models of relaxed
-memory~\cite{battycpp} and weakly consistent
-replication~\cite{burckhardt14}, it is reasonable to expect similar
-reasoning approaches to work in both the cases. Ridge~\cite{rgtso}
-generalizes rely-guarantee reasoning to the x86-TSO memory model.
-Likewise, Vafeiadis \emph{et al.}~\cite{rsl13} generalize concurrent
-separation logic (CSL)~\cite{csl} to the C11 relaxed memory model.
-Ferreira \emph{et al.}~\cite{ferreira10} propose a parameterized
-operational semantics for relaxed memory models, but the
-parameterization is over a relation between relaxed memory programs
-and related {\sc sc} programs. They also prove the soundness of CSL
-for the weakest memory model expressible in their semantics, but
-neither a parameterized (in the aforementioned sense) CSL, nor a CSL
-specialized for any relaxed memory model is proposed.  ~\cite{DLZ+13}
-presents a \emph{buffered memory model} for Java that defines an
-axiomatic definition for the JMM in terms of memory reorderings, and
-an operational instantiation consistent with the TSO memory model
-found on x86 processors, and demonstrates the equivalence of the two
-definitions.  However, there is no attempt to parameterize the model
-over different reordering or consistency definitions, which is a central
-focus of this paper.
+\paragraph{Reasoning under relaxed memory} The reasoning mechanisms
+used to describe and prove properties about weakly-isolated
+transactions bear resemblance to those used to formalize relaxed
+memory behaviour~\cite{battycpp}.  Ridge~\cite{rgtso} generalizes
+rely-guarantee reasoning to the x86-TSO memory model.  Likewise,
+Vafeiadis \emph{et al.}~\cite{rsl13} generalize concurrent separation
+logic (CSL)~\cite{csl} to the C11 relaxed memory model.  Ferreira
+\emph{et al.}~\cite{ferreira10} propose a parameterized operational
+semantics for relaxed memory models, but the parameterization is over
+a relation between relaxed memory programs and related {\sc sc}
+programs. They also prove the soundness of CSL for the weakest memory
+model expressible in their semantics, but neither a parameterized (in
+the aforementioned sense) CSL, nor a CSL specialized for any relaxed
+memory model is proposed.  ~\cite{DLZ+13} presents a \emph{buffered
+  memory model} for Java that defines an axiomatic definition for the
+JMM in terms of memory reorderings, and an operational instantiation
+consistent with the TSO memory model found on x86 processors, and
+demonstrates the equivalence of the two definitions.  However, there
+is no attempt to parameterize the model over different reordering or
+consistency definitions, which is a central focus of this paper.

### 2016-07-03 store-consistency.tex
edits

--- 2016-07-03 store-consistency.tex
@@
 data store, the store specific invariant ($\I_s$) changes from store
-to store depending on the consistency level.  For an eventually
+to store depending on the consisteny level.  For an eventually
 consistent store, $\I_s$ is simply \emph{true}. Stronger stores, such
-as those that support strong consistent, have non-trivial definitions
+as those that support strong consistenty, have non-trivial definitions
 for $\I_s$.
@@
 approach to any $\I_s$ and $\I_c$ by defining a \emph{maximum
-  visibility principle} to determine an acceptable weakening of $\I_s$
+  visiblity principle} to determine an acceptable weakening of $\I_s$
 in case of a conflict with $\I_c$.  The principle requires the
@@
 than {\sc ec}. This unfortunately restricts the reasoning approach
-from being applied to data stores (e.g., \cite{bayou,pldi15}) that provide
-guarantees weaker than causal consistency, such as causal visibility
-or read-my-writes~\cite{zoo}. Causal consistency is not baseline
-consistency in these stores because it is not \emph{highly
+from being applied to data stores (e.g., \cite{bayou,pldi15}) that
+provide guarantees weaker than causal consistency, such as causal
+visibility or read-my-writes~\cite{zoo}. Causal consistency is not
+baseline consistency in these stores because it is not \emph{highly
   available}~\cite{bailishat}.
@@
 change during the execution of the operation. To account for such
-changes, we might choose to distinguish between effect generation
-event at one replica $r_1$ and effect merge event at replica $r_2$,
+changes, we might choose to distinguish between an effect generation
+event at one replica $r_1$ and an effect merge event at replica $r_2$,
 requiring that \emph{non-conflicting} operations execute between these
@@
 in greater detail. We assume an {\sc sc} store, such as a relational
-database, whose store-specific trace invariant ($\I_s$) was shown
+database, whose store-specific trace invaraint ($\I_s$) was shown
 previously. Both transactions are run at {\sc si} isolation, hence
@@
 RHS expression \C{C-a1}.  Assertions implicitly refer to the current
-execution ($\E$), just as Hoare triples implicitly refer to the
+execution ($\E$), just as hoare triples implicitly refer to the
 current state.  The context for propositions is also the implicit,
@@
 program is extended with $\neg\committed(\C{Wd1}) \wedge
-\neg\committed(\C{Wd2})$, from which $P_1$ follows. The post-condition
+\neg\committed(\C{Wd2})$, from which $P_1$ follows. The postcondition
 $Q_1$ of \C{Wd1} asserts different values for \C{C} depending on
 whether or not \C{Wd2} has already committed.  It also asserts that
-\C{Wd1} has committed. The post-condition $Q_2$ of \C{Wd2} (not shown)
+\C{Wd1} has committed. The postcondition $Q_2$ of \C{Wd2} (not shown)
 is similar, but with the roles of \C{Wd1} and \C{Wd2} exchanged. The
-post-condition $Q$ follows from $Q_1 \wedge Q_2$, hence valid. It
+postcondition $Q$ follows from $Q_1 \wedge Q_2$, hence valid. It
 remains to show that \C{Wd1} satisfies its specification (\C{Wd2} must
@@
 considers the case when $\C{Wd2} \visar \C{Wd1}$. As we show below,
-this is sufficient to prove the post-condition $\phi_5$, which appears
+this is sufficient to prove the postcondition $\phi_5$, which appears
 to make a stronger claim about the value of \C{C}. The proof follows
@@
 ${\committed(\C{Wd2})}  \Rightarrow \C{C = k-a2} \wedge \C{C =
-k-a2-a1}$, allowing us to prove the post-condition $\phi_5$.
+k-a2-a1}$, allowing us to prove the postcondition $\phi_5$.
 
@@
 above, but with \C{Wd2} and \C{Wd1} interchanged. Continuing our focus
-on the assignment statement, the guarantee proof requires us to show
+on the assigment statement, the guarantee proof requires us to show
 that if executing the assignment takes the execution from $\E$ to
@@
 triples of \C{Wd1} can also be constructed similarly. The combination
-of correctness proofs for \C{Wd1} and \C{Wd2} yields a correctness
+of correctness proofs for \C{Wd1} and \C{Wd2} yeilds a correctness
 proof for the program.

### 2016-07-04 case-studies.tex

@@
 observed monotonicity violations. On the other hand, instantiating the
-store consistency level with {\sc cc} allowed us to obtain a proof,
-and recover the monotonicity invariant. We in fact obtained a proof
-after relaxing the serializability requirement for writes, by changing
-the definition of the $\interp{\E}(X)$ (\S\ref{sec:opsem}) to return
-the value of the largest write-to-$X$ instead of the latest
-write-to-$X$.  We implemented this version of the counter RDT, and
-obtained a 52\% improvement in latency compared to the previous
-implementation\footnote{The tradeoff is that the view count is now no
-  longer accurate, although it grows monotonically, thus preserving
-  the application invariant.}.
+store consistency level with {\sc cc} allowed us to obtain a proof, and
+recover the monotonicity invariant. We in fact obtained a proof after
+relaxing the serializability requirement for writes, by changing the
+definition of the $\interp{\E}(X)$ (\S\ref{sec:opsem}) to return the
+value of the largest write-to-$X$ instead of the latest write-to-$X$.
+We implemented this version of the counter RDT, and obtained a 52\%
+improvement in latency compared to the previous
+implementation\footnote{The trade-off is that the view count is now no
+longer accurate, although it grows monotonically.}.
 

### 2016-07-04 opsem.tex

@@
 ($\I$). $\I$ is a function from traces ($\E$) to first-order logical
-formula ($\texttt{Prop}$) that define well-formedness
+formula ($\Prop$) that define well-formedness
 constraints over traces.  The machine takes a step only if the
@@
 
+% TRACE RELATIONS
+% ---------------
+\input{trace-relations}
+
 % However, a machine that lets operations witness arbitrary subset of
@@
 that accepts a set ($S$) of effects chosen by the machine to make
-visible to the operation, interprets the operation w.r.t $S$, and
+visible to the operation, interprets the operation w.r.t. $S$, and
 returns an appropriate effect that encodes its return value. Rules
@@
 monotonically increasing counter application may choose to let a write
-with the largest value determine the value of a read. A more
-straightforward interpretation has already been described before. To
-accommodate multiple interpretations, the semantics is made parametric
-over an interpretation function ($\interp{\cdot}$) that accepts a set
-of effects and a variable name, and returns the value associated with
-the variable. A straightforward interpretation function that chooses
-the last write (i.e., write with largest id) is shown below:
+with the largest value determine the value of a read. To accommodate
+multiple interpretations, the semantics is made parametric over an
+interpretation function ($\interp{\cdot}$) that accepts a set of
+effects and a variable name, and returns the value associated with the
+variable. A straightforward interpretation function that chooses the
+last write (i.e., write with largest id) is shown below:
 

### 2016-07-04 reasoning.tex

@@
 As mentioned earlier, our proof system is a rely-guarantee framework
-that admits compositional reasoning by abstracting away 
-environmental interference as a rely relation. A conspicuous
-difference between a standard development of rely-guarantee and ours
-is that, while the former reasons in terms of program states (variable
-to value bindings), we reason in terms of executions as captured by
-their traces ($\E$). In particular, our rely ($R$) and guarantee ($G$)
+that admits compositional reasoning by abstracting away environmental
+interference as a rely relation. A conspicuous difference between a
+standard development of rely-guarantee and ours is that, while the
+former reasons in terms of program states (variable to value
+bindings), we reason in terms of executions as captured by their
+traces ($\E$). In particular, our rely ($R$) and guarantee ($G$)
 relations relate executions (i.e., $R,\,Q\,\subseteq\,\E\times\E$),
-and pre- ($P$) and post- ($Q$) conditions as assertions over executions
-(i.e., $P,\,Q\, : \E \rightarrow \Prop$). Our development nonetheless
-facilitates state-based reasoning via the interpretation function
+and our pre- ($P$) and post- ($Q$) conditions are assertions over
+executions (i.e., $P,\,Q\, : \E \rightarrow \Prop$). Our development
+also facilitates state-based reasoning via the interpretation function
 ($\interp{\cdot}$) introduced in \S\ref{sec:syntax}. For example, if a
@@
 The standard rely-guarantee judgment is a quintuple
-$\rg{P,R}{c}{G,Q}$, which informally asserts that if a
-command $c$ is executed in a state that satisfies its pre-condition
-$P$, provided that every interference step during the execution is
-contained inside the rely relation $R$, then the effect that each step
-of executing $c$ has on the state is captured by $G$, and the final
-state of execution satisfies the post-condition $Q$. The terms
-``execution'' and ``step'' are defined with respect to an operational
-semantics, which, in our case, is parameterized on the trace invariant
-$\I$. Consequently, our rely-guarantee judgment for commands is the
-sextuple $\rg{\I,P,R}{c}{G,Q}$, whose semantics differs from the
-quintuple in that it requires every step of the execution to additionally preserve
-the trace invariant $\I$ for the post-condition to be valid.  As usual,
-if $c$ is inside a transaction $T_i$, we write $\txnbox{c}_i$. \txnimp
-expressions are side-effecting (they generate \C{RD} effects), and
-admit interference during their evaluation.  Therefore, like 
-commands, expressions need to be reasoned about explicitly.  In contrast to
-commands, however, expressions evaluate to values, and the reasoning
+$\rg{P,R}{c}{G,Q}$, which informally asserts that if a command $c$ is
+executed in a state that satisfies its pre-condition $P$, provided
+that every interference step during the execution is contained inside
+the rely relation $R$, then the effect that each step of executing $c$
+has on the state is captured by $G$, and the final state of execution
+satisfies the post-condition $Q$. The terms ``execution'' and ``step''
+are defined with respect to an operational semantics, which, in our
+case, is parameterized on the trace invariant $\I$. Consequently, our
+rely-guarantee judgment for commands is the sextuple
+$\rg{\I,P,R}{c}{G,Q}$, whose semantics differs from the quintuple in
+that it requires every step of the execution to additionally preserve
+the trace invariant $\I$ for the post-condition to be valid.  As
+usual, if $c$ is inside a transaction $T_i$, we write $\txnbox{c}_i$.
+\txnimp expressions are side-effecting (they generate \C{RD} effects),
+and admit interference during their evaluation.  Therefore, like
+commands, expressions need to be reasoned about explicitly. But,
+unlike commands, expressions evaluate to values, and the reasoning
 framework should also admit assertions on such values.  We therefore

### 2016-07-04 case-studies.tex
Case studies changed significantly.

--- 2016-07-03 case-studies.tex
@@
 anonymized url: \url{...}}. The application stores the video count in
-a Postgres database, and supports a read operation (a database read
+a MySQL database, and supports a read operation (a database read
 transaction), and an increment operation (a read-increment-write
@@
 preserve the \emph{monotonicity invariant}, namely that the view count
-should never appear to be decreasing. However, the default {\sc rc}
-isolation of Postgres may lead to anomalous executions as confirmed by
-our experiments; with 16 (resp.  32) concurrent writers performing
-single increments, we observed a total of 7 (17) violations of
-monotonicity. We subsequently instantiated our reasoning framework for
-an {\sc sc} store with Postgres isolation levels (which are stronger
-than ANSI SQL's and MySQL's), and spent 5 man hours determining
-appropriate isolation levels needed to enforce the
+should never appear to be decreasing. However, the default {\sc rr}
+isolation of MySQL may lead to anomalous executions as confirmed by
+our experiments; with 32 concurrent writers performing single
+increments, and one reader issuing a constant stream of reads, we
+observed a maximum of 12 violations of monotonicity in 10 rounds of
+operation. We subsequently instantiated our reasoning framework for an
+{\sc sc} store with ANSI SQL isolation levels, and spent 5 man hours
+determining appropriate isolation levels needed to enforce the
 invariant\footnote{The representative \txnimp programs we used to
@@
 that to preserve the invariant, writers need to execute at
-\iso{Repeatable Read} isolation, while readers can continue at {\sc
-rc}. Repeating the experiments with this configuration led to no
-violations, as expected. While executing all transactions at {\sc ser}
-also prevents violations, it led to an additional 47\% increase in
-latency on an average (16 concurrent writers; 5 rounds). We repeated
-the exercise for MySQL, and found that its {\sc rr} is not sufficient
-to enforce monotonicity; {\sc ser} is needed.
+\iso{Serializable} isolation, while readers can execute at \iso{Read
+Uncommitted}. Repeating the experiments with this configuration led to
+no violations, as expected. While executing all transactions at {\sc
+ser} also prevents violations, it led to an additional 113\% increase
+in the latency of writes on an average (32 concurrent writers; 5
+rounds). We repeated the experiments with the same configuration on
+Postgres, and found no violations. However, Postgres's isolation
+levels~\cite{postgresiso} are stronger than ANSI SQL's in non-trivial
+ways. We therefore formalized Postgres's guarantees and repeated
+relevant parts of the proof to discover that \iso{Repeatable Read} is
+in fact sufficient for writers - an observation we confirmed through
+experiments. As usual, using {\sc ser} for all transactions also
+suffices, but on Postgres it led to an additional 47\% increase in
+write latency.
 
@@
 To understand the effect of changing the store consistency, we
-re-implemented the counter as a \emph{replicated data type}
+re-implemented the counter as a op-based \emph{replicated data type}
 (RDT)~\cite{crdt,burckhardt14} on top of Cassandra
 \emph{bolted-on}~\cite{bailisbolton} with {\sc cc} and {\sc sc}
-consistency levels, and ANSI SQL isolation levels. Keeping the
-consistency level at {\sc ec}, we then repeated our experiments with
-committed reads and serializable increments - a combination which was
-proved to preserve monotonicity on an {\sc sc} store, but nonetheless
-observed monotonicity violations. On the other hand, instantiating the
-store consistency with {\sc cc} allowed us to obtain a proof, and
-recover the monotonicity invariant. We in fact obtained a proof after
-relaxing the serializability requirement for writes, by changing the
+consistency levels (both on-demand; {\sc ec} is default), and ANSI SQL
+isolation levels. Keeping the consistency level at {\sc ec}, we then
+repeated our experiments with committed reads and serializable
+increments - a combination which was proved to preserve monotonicity
+on an {\sc sc} store, but nonetheless observed monotonicity
+violations. The reason for such violations became evident as we
+considered an instantiation our proof framework for an {\sc ec} store.
+Under this instantiation, it is impossible to derive the
+\C{MonotonicVis} (\S\ref{sec:ansi-isolation}) property needed to show
+that successive reads in a session witness monotonically increasing
+state. However, assuming {\sc cc} store consistency for reads allowed
+us to complete the proof, and recover the monotonicity invariant. In
+fact, we were able to complete the proof even after relaxing the
+serializability requirement for increments, just by changing the
 definition of the $\interp{\E}(X)$ (\S\ref{sec:opsem}) to return the
 value of the largest write-to-$X$ instead of the latest write-to-$X$.
-We implemented this version of the counter RDT, and obtained a 52\%
-improvement in latency compared to the previous
-implementation\footnote{The trade-off is that the view count is now no
-longer accurate, although it grows monotonically.}.
+In other words, if the reads chose the largest valued write effect,
+instead of the latest write effect they witness, then counter
+increments need not be serializable for reads to witness a
+monotonically increasing count\footnote{The trade-off is that the view
+count is now no longer accurate, although it grows monotonically.}.
+We implemented this version of the counter as an op-based RDT, which
+uses {\sc cc} reads and {\sc ec} increments, both un-isolated (i.e.,
+\emph{Read Uncommitted}), and obtained a 52\% improvement in overall
+latency compared to the version that uses serializable increments.
 
@@
 entirely in the user space by relying on database-backed transactions
-(this behaviour is typical of Rails applications~\cite{bailisferal}).
+(this behavior is typical of Rails applications~\cite{bailisferal}).
 For instance, deletion of a user is carried out in a transaction that
@@
 chosen users in 1:7 ratio, and witnessed a maximum of 250 violations
-of referential integrity. We then instantiated our framework for an
-{\sc sc} store with various ANSI SQL isolation levels for user
-deletions and timeline reads, and obtained a proof that {\sc rc} is
-sufficient for deletions, whereas reads require {\sc rr}. The proof is
-immediately applicable for MySQL, which implements ANSI SQL isolation
-levels. Since Postgres's isolation levels are stronger than the
-corresponding ANSI levels, \rulelabel{RG-Conseq} rule
-(Fig.~\ref{fig:rg-rules}) lets us deduce that the same choice of
-isolation levels also work on Postgres. Indeed, experiments confirmed
-this judgment.
+of referential integrity. To determine appropriate isolation levels,
+we then instantiated our framework for an {\sc sc} store with various
+ANSI SQL isolation levels for user deletions and timeline reads, and
+obtained a proof that {\sc rc} is sufficient for deletions, whereas
+reads require {\sc rr}. The proof is immediately applicable for MySQL,
+which implements ANSI SQL isolation levels. Repeating our experiments
+for MySQL with this configuration resulted in no violations (as
+expectected), while yeilding a 38\% better overall latency than {\sc
+ser} configuration (96 concurrent clients; 5 rounds).  Since
+Postgres's isolation levels are stronger than the corresponding ANSI
+levels, \rulelabel{RG-Conseq} rule (Fig.~\ref{fig:rg-rules}) lets us
+deduce that the same choice of isolation levels also work on Postgres.
+Indeed, experiments confirm this judgment.
 

### 2016-07-04 opsem.tex
Case studies changed significantly.

@@
 resulting trace satisfies the constraints imposed by $\I$. This
-behaviour is captured by the auxiliary reduction rule
+behavior is captured by the auxiliary reduction rule
 \rulelabel{E-Aux} that factors out the trace extension aspect of the
-evaluation by abstracting away the operation-specific behaviour as a
+evaluation by abstracting away the operation-specific behavior as a
 function that generates an appropriate effect. We let $\mathcal{F}$
@@
 As described previously, \rulelabel{E-Aux} abstracts away the
-operation-specific behaviour of a machine step as a function ($\F$)
+operation-specific behavior of a machine step as a function ($\F$)
 that accepts a set ($S$) of effects chosen by the machine to make

### 2016-07-04 related-work.tex
Case studies changed significantly.

--- 2016-07-03 related-work.tex
@@
 declarative nature of our specifications make them suitable for
-reasoning about program behaviour. Sivaramakrishnan \emph{et
+reasoning about program behavior. Sivaramakrishnan \emph{et
   al.}~\cite{pldi15} specify isolation levels declaratively as trace

### 2016-07-04 store-consistency.tex
Case studies changed significantly.

--- 2016-07-03 store-consistency.tex
@@
 to witness arbitrary subsets of the global state, effectively mimicking the
-behaviour of an eventually consistent ({\sc ec}) data
+behavior of an eventually consistent ({\sc ec}) data
 store\footnote{Eventual consistency guarantees that in the absence of further
@@
 %% Note that the purpose of maximum visibility principle is to
-%% rationalize the observed behaviour of databases in practice. In
+%% rationalize the observed behavior of databases in practice. In
 %% particular, it is not intended to be a guiding principle to engineer

### 2016-07-04 motivation.tex
Edits.

@@
 \centering
-$\{\{\texttt{C}=\texttt{k} \conj \texttt{k}\ge\texttt{a1+a2}\}\}$
+$\{\texttt{C}=\texttt{k} \conj \texttt{k}\ge\texttt{a1+a2}\}$
 \begin{tabular}{l||l}
@@
 \end{tabular}
-$\{\{\texttt{C}=\texttt{k-a1-a2}\}\}$
+$\{\texttt{C}=\texttt{k-a1-a2}\}$
 
@@
 style reasoning~\cite{rgjones}, a compositional proof technique that
-allows us to reason about the behaviour of individual threads by
+allows us to reason about the behavior of individual threads by
 abstracting away interferences induced by other threads (collectively

### 2016-07-04 case-studies.tex
various edits to regain space

@@
 
-\paragraph{A video view counter} Our first application is a
+\paragraph{A video view counter.} Our first application is a
 Ruby-on-Rails implementation of a counter to count the number of views
@@
   tools we developed to analyze Ruby can be found at the following
-anonymized url: \url{...}}. The application stores the video count in
-a MySQL database, and supports a read operation (a database read
+  anonymized url: \url{...}}. The application stores the video count
+in a MySQL database, and supports a read operation (a database read
 transaction), and an increment operation (a read-increment-write
@@
 preserve the \emph{monotonicity invariant}, namely that the view count
-should never appear to be decreasing. However, the default {\sc rr}
-isolation of MySQL may lead to anomalous executions as confirmed by
-our experiments; with 32 concurrent writers performing single
-increments, and one reader issuing a constant stream of reads, we
-observed a maximum of 12 violations of monotonicity in 10 rounds of
-operation. We subsequently instantiated our reasoning framework for an
-{\sc sc} store with ANSI SQL isolation levels, and spent 5 man hours
-determining appropriate isolation levels needed to enforce the
-invariant\footnote{The representative \txnimp programs we used to
-formally reason about scenarios described in this section can be found
-in the supplementary.} In particular, we were able to formally prove
-that to preserve the invariant, writers need to execute at
-\iso{Serializable} isolation, while readers can execute at \iso{Read
-Uncommitted}. Repeating the experiments with this configuration led to
-no violations, as expected. While executing all transactions at {\sc
-ser} also prevents violations, it led to an additional 113\% increase
-in the latency of writes on an average (32 concurrent writers; 5
-rounds). We repeated the experiments with the same configuration on
-Postgres, and found no violations. However, Postgres's isolation
-levels~\cite{postgresiso} are stronger than ANSI SQL's in non-trivial
-ways. We therefore formalized Postgres's guarantees and repeated
-relevant parts of the proof to discover that \iso{Repeatable Read} is
-in fact sufficient for writers - an observation we confirmed through
-experiments. As usual, using {\sc ser} for all transactions also
-suffices, but on Postgres it led to an additional 47\% increase in
-write latency.
+should never appear to be decreasing. However, the default
+\iso{Repeatable Read} isolation level of MySQL leads to anomalous
+executions - with 32 concurrent writers performing single increments,
+and one reader issuing a constant stream of reads, we observed 12
+violations of monotonicity in 10 rounds. We subsequently instantiated
+our reasoning framework for an {\sc sc} store with ANSI SQL isolation
+levels, and spent 5 man hours determining appropriate isolation levels
+needed to enforce the invariant\footnote{The representative \txnimp
+  programs we used to formally reason about scenarios described in
+  this section can be found in the supplementary.} In particular, we
+were able to formally prove that to preserve the invariant, writers
+need to execute at \iso{Serializable} isolation, while readers can
+execute at \iso{Read Uncommitted}. Repeating the experiments with this
+configuration led to no violations, as expected. While executing all
+transactions at {\sc ser} also prevents violations, it led to an
+additional 113\% increase in the latency of writes on an average (32
+concurrent writers; 5 rounds). We repeated the experiments with the
+same configuration on Postgres, and found no violations. However,
+Postgres's isolation levels~\cite{postgresiso} are stronger than ANSI
+SQL's in non-trivial ways. We therefore formalized Postgres'
+guarantees and repeated relevant parts of the proof to discover that
+\iso{Repeatable Read} is in fact sufficient for writers - an
+observation we confirmed through experiments. As usual, using {\sc
+  ser} for all transactions also suffices, but on Postgres results in
+an additional 47\% increase in write latency.
 
@@
 % techniques.   
-To understand the effect of changing the store consistency, we
-re-implemented the counter as a op-based \emph{replicated data type}
+To understand the effect of changing the store consistency levels, we
+re-implemented the counter as an op-based \emph{replicated data type}
 (RDT)~\cite{crdt,burckhardt14} on top of Cassandra
@@
 on an {\sc sc} store, but nonetheless observed monotonicity
-violations. The reason for such violations became evident as we
-considered an instantiation our proof framework for an {\sc ec} store.
+violations. The reason for such violations became evident when
+we instantiated our proof framework for an {\sc ec} store.
 Under this instantiation, it is impossible to derive the
@@
 latency compared to the version that uses serializable increments.
-
-\paragraph{Microblog}
-
-Next, we focus our attention on a Twitter-like microblogging site that
-is one the sample applications shipped with a popular textbook on
-Ruby-on-Rails~\cite{railsbook}. One of the invariants that the
-application chooses to enforce is the referential integrity between
-microposts and authors. The application enforces this invariant
-entirely in the user space by relying on database-backed transactions
-(this behavior is typical of Rails applications~\cite{bailisferal}).
-For instance, deletion of a user is carried out in a transaction that
-also includes the deletion of microposts authored by the user.
-However, concurrent observers can nonetheless witness violations of
-referential integrity. To measure the likelihood of such anomalies in
-practice, we first populated a Postgres database with 1000 user
-accounts, each user with 50 microposts and 20 followers, chosen
-uniformly at random. We then constructed experiments with 64
-concurrent clients performing deletions and timeline reads of randomly
-chosen users in 1:7 ratio, and witnessed a maximum of 250 violations
-of referential integrity. To determine appropriate isolation levels,
-we then instantiated our framework for an {\sc sc} store with various
+\vspace*{-3pt}
+\paragraph{Microblog.} Our second set of experiments focused on a Twitter-like microblogging
+site, also written using Ruby-on-Rails~\cite{railsbook}. One of the
+invariants that the application chooses to enforce is referential
+integrity between microposts and authors. The application enforces
+this invariant entirely in user space by relying on database-backed
+transactions (this behavior is typical of Rails
+applications~\cite{bailisferal}).  For instance, deletion of a user is
+carried out in a transaction that also includes the deletion of
+microposts authored by the user.  However, concurrent observers can
+nonetheless witness violations of referential integrity. To measure
+the likelihood of such anomalies, we first populated a Postgres
+database with 1000 user accounts, each user with 50 microposts and 20
+followers, chosen uniformly at random. We then constructed experiments
+with 64 concurrent clients performing deletions and timeline reads of
+randomly chosen users in a 1:7 ratio, and witnessed 250 violations of
+referential integrity. To determine appropriate isolation levels, we
+then instantiated our framework for an {\sc sc} store with various
 ANSI SQL isolation levels for user deletions and timeline reads, and
@@
 for MySQL with this configuration resulted in no violations (as
-expectected), while yeilding a 38\% better overall latency than {\sc
+expectected), while yielding a 38\% reduction in latency than an {\sc
   ser} configuration (96 concurrent clients; 5 rounds).  Since
@@
 levels, \rulelabel{RG-Conseq} rule (Fig.~\ref{fig:rg-rules}) lets us
-deduce that the same choice of isolation levels also work on Postgres.
-Indeed, experiments confirm this judgment.
+immediately deduce that the same choice of isolation levels also work
+on Postgres.
 
-As evident from our case studies, the there already exists a
-significant diversity among the weak consistency and isolation
-variants implemented on various real-world data stores. More such
-variants have been developed in the recent years to meet the need for
-high availability on geo-replicated stores. For example,
-\emph{Parallel Snapshot Isolation}~\cite{psi} was proposed in 2011,
-and \emph{Non-Monotonic Snapshot Isolation}~\cite{nmsi} in 2013, both
-targeted for replicated stores. As more commercial weakly consistent
-stores adopt various kinds of transactional semantics, like the
-recently introduced \emph{lightweight transactions} of Cassandra, it
-is reasonable to expect new isolation definitions to continue to be
-proposed. It may be possible to carefully engineer a reasoning
-framework for each combination of consistency and isolation, but such
-a strategy would lead to multiple semantic definitions and proof
-systems with no obvious way to compare and relate them. Having a
-parameterized semantics and a proof framework built on the such
-semantics allows us to support these variants as distinct, yet
-comparable instantiations, as we demonstrate in our case studies.
+Clearly, there exists a significant diversity among the weak
+consistency and isolation variants implemented on various real-world
+data stores. More such variants have been developed in recent years to
+meet the need for high availability (\emph{e.g.}, \emph{Parallel
+  Snapshot Isolation}~\cite{psi} (2011) or \emph{Non-Monotonic
+  Snapshot Isolation}~\cite{nmsi} (2013)).  As more commercial
+weakly-consistent stores adopt various kinds of transactional
+semantics, like Cassandra's recently introduced \emph{lightweight
+  transactions}, it is reasonable to expect new isolation
+definitions to continue to be proposed.  While it may be possible to
+carefully engineer a reasoning framework for each combination of
+consistency and isolation, such a strategy would lead to multiple
+semantic definitions and proof systems with no obvious way to compare
+and relate them. As illustrated here, having a parameterized semantics
+and a proof framework built on top of it allows us to support these
+variants as distinct, yet comparable, instantiations.
+
 %  GK: what does the following sentence mean?

### 2016-07-04 introduction.tex
various edits to regain space

@@
 primarily in terms of the undesirable phenomena a particular isolation
-level may induce, placing the burden squarely on the programmer to
-determine application correctness.
+level may induce, placing the burden on the programmer to determine
+application correctness.
 
@@
 
-\noindent Our results thus provide the first (to the best of our
+\noindent Our results provide the first (to the best of our
 knowledge) mechanism that precisely relates high-level program
@@
 describe the impact of our reasoning framework in the context of
-several real-world case studies in \S\ref{sec:casestudies}.  Related
+several real-world case studies in \S\ref{sec:case-studies}.  Related
 work and conclusions are given in \S\ref{sec:relatedwork}.

### 2016-07-04 motivation.tex
various edits to regain space

@@
 $\{\texttt{C}=\texttt{k-a1-a2}\}$
-
-\caption{Concurrent withdraw transactions}
+\caption{\small Concurrent withdraw transactions}
 \label{fig:motiv-eg-1}
+\vspace*{-10pt}
 \end{figure}
@@
 }
-\caption{A possible execution of the program shown in Fig.~\ref{fig:motiv-eg-1} under
+\caption{\small A possible execution of the program shown in Fig.~\ref{fig:motiv-eg-1} under
   a read committed isolation level.}
@@
 \end{smathpar}
-\caption{Visibility and interference properties for different weak isolation levels can be
+\caption{\small Visibility and interference properties for different weak isolation levels can be
   captured by different instantiations of their happens-before relation.}
@@
 
-\caption{$\visZ$ arrows allowed by {\sc si} (solid) and {\sc rc} (solid
+\caption{\small $\hbZ$ arrows allowed by {\sc si} (solid) and {\sc rc} (solid
   and dashed) isolation levels for the example in
-  Fig.~\ref{fig:motiv-eg-1}. Each arrow represents an $\visZ$ relationship
+  Fig.~\ref{fig:motiv-eg-1}. Each arrow represents an $\hbZ$ relationship
   among all operations of `Wd2' and operations in `Wd1' that follow
@@
 \label{fig:motiv-eg-1-hb}
+\vspace*{-8pt}
 \end{figure}

### 2016-07-04 reasoning.tex
various edits to regain space

@@
 on the assumption that interference preserves the trace invariant
-$\I$. Formally:
+$\I$. Formally:\vspace*{-10pt}
 
@@
 dually, $\I$ withstands interference) needs to be justified
-separately. We call this the stability requirement on $\I$:
+separately. We call this the stability requirement on $\I$:\vspace*{-10pt}
 
@@
 interference from concurrent threads captured as the rely relation
-($R$).  It is defined thus:
+($R$).  It is defined thus:\vspace*{-10pt}
 
@@
 The semantics of the RG sextuple $\rg{\mathbb{I},P,R}{c}{G,Q}$ is defined
-in terms of the interleaved step relation thus:
+in terms of the interleaved step relation thus:\vspace*{-10pt}
 
@@
 The second conjunct, called the \emph{guarantee consequent}, uses the
-$\texttt{step-guaranteed}$ predicate defined below:
+$\texttt{step-guaranteed}$ predicate defined below:\vspace*{-10pt}
 

### 2016-07-04 store-consistency.tex
various edits to regain space

@@
 describes {\sc sc} stores with on-demand weak isolation, such as
-off-the-shelf relational databases. Similar approach can be employed
+off-the-shelf relational databases. Similar approaches can be employed
 to capture the semantics of other kinds of stores, such as a causally
@@
 ] {
-  \includegraphics[scale=0.7]{Figures/ec-theirs}
+  \includegraphics[scale=0.5]{Figures/ec-theirs}
 
@@
 }{
-  \includegraphics[scale=0.45]{Figures/ec-ours}
-} \caption{In the replica model, operation $\op$ generates effect
+  \includegraphics[scale=0.35]{Figures/ec-ours}
+} \caption{\small In the replica model, operation $\op$ generates effect
 $\eta$ at replica $R_1$, which is then merged to $R_2$. If the
@@
 \end{txnimpcode}
-
-\caption{\C{Wd1} transaction decorated with assertions}
+\vspace*{-8pt}
+\caption{\small \C{Wd1} transaction decorated with assertions}
 \label{fig:wd1-decorated}
+\vspace*{-12pt}
 \end{figure}
@@
 to easily conclude the following about the execution state after the
-assignment:
+assignment:\vspace*{-12pt}
 

### 2016-07-04 related-work.tex
edits

--- 2016-07-04 related-work.tex
@@
 isolation and consistency properties.  Cerone \emph{et
-  al.}~\cite{gotsmanconcur15} present a framework to specify isolation
-levels with atomic visibility, but their specifications are also for
-complete traces; moreover, their formulation implicitly assumes
-homogenity, \emph{i.e.}, they require all transactions to execute
-using either {\sc si} or {\sc ser}, which is often not the case in
-practice. While all the aforementioned specification frameworks use
-the vocabulary introduced in~\cite{burckhardt14} to specify replicated
-data types and different forms of weak consistency, none of them come
-with an associated reasoning framework that can use such
-specifications productively towards verifying programs under weak
-isolation.
+  al.}~\cite{gotsmanconcur15} specify isolation levels with atomic
+visibility, but their specifications are also for complete traces, and
+assumes all transactions execute using either {\sc si} or {\sc ser},
+which is often not the case in practice. While all the aforementioned
+specification frameworks use the vocabulary introduced
+in~\cite{burckhardt14} to specify replicated data types, none of them
+are equipped with a reasoning framework that can use such
+specifications to verify programs under weak isolation.
 
@@
 isolation levels to transactions while guaranteeing the
-serializability of the execution. In~\cite{gotsmanpodc16}, Cerone
-\emph{et al.} improve on Adya's {\sc si} specification and use it to
-derive a static analysis that determines the safety of substituting
-{\sc si} with a weaker variant called \iso{Parallel Snapshot
-  Isolation}~\cite{psi}.  These efforts focus on establishing the
-equivalence of executions between a pair of isolation levels, without
-taking application invariants into account.  Bernstein \emph{et
-  al.}~\cite{bern2000} propose informal semantic conditions to ensure
-the satisfaction of application invariants under weaker isolation
-levels.  However, no formal proof system is given.  Furthermore, these
+serializability. In~\cite{gotsmanpodc16}, Cerone \emph{et al.} improve
+on Adya's {\sc si} specification and use it to derive a static
+analysis that determines the safety of substituting {\sc si} with a
+weaker variant called \iso{Parallel Snapshot Isolation}~\cite{psi}.
+These efforts focus on establishing the equivalence of executions
+between a pair of isolation levels, without taking application
+invariants into account.  Bernstein \emph{et al.}~\cite{bern2000}
+propose informal semantic conditions to ensure the satisfaction of
+application invariants under weaker isolation levels.  All these
 techniques are tailor-made for a finite set of well-understood
-isolation levels (rooted in~\cite{berenson}) under a pre-defined
-store consistency model (usually {\sc sc}).
+isolation levels (rooted in~\cite{berenson}) under a pre-defined store
+consistency model (usually {\sc sc}).
 
 \paragraph{Reasoning under weak consistency} There have been several
-recent proposals on reasoning techniques for programs executing under
-weak consistency~\cite{bailisvldb, alvarocalm,
-  gotsmanpopl16,redblueatc, redblueosdi, ecinec}. All of them assume a
-system model that offers a choice between a \emph{coordination-free}
-weak consistency level (\emph{e.g.}, {\sc ec}~\cite{redblueosdi,
-  redblueatc, ecinec, alvarocalm, bailisvldb}) or causal
-consistency~\cite{gotsmanpopl16}). The key technical issue in these
-efforts involves proving that atomic and fully isolated operations
-preserve application invariants when executed on these consistency
-levels.  In contrast, we admit weakly isolated transactions, and our
-system model accepts \emph{specifications} of consistency and
-isolation levels drawn from an expressive logic.  In recent
-work~\cite{gotsmanpopl16}, \iso{Parallel Snapshot Isolation} is
-adapted to the aforementioned setting by interpreting it as a
-consistency level that serializes writes to objects; a dedicated proof
-rule is developed to help prove prove program invariants hold under
-this model. By parameterizing our proof system over a gamut of weak
-isolation specifications, we avoid the need to define a separate proof
-rule for each new isolation level we may encounter.
+recent proposals to reason about programs executing under weak
+consistency~\cite{bailisvldb, alvarocalm, gotsmanpopl16,redblueatc,
+  redblueosdi, ecinec}. All of them assume a system model that offers
+a choice between a \emph{coordination-free} weak consistency level
+(\emph{e.g.}, eventual consistency~\cite{redblueosdi, redblueatc,
+  ecinec, alvarocalm, bailisvldb}) or causal
+consistency~\cite{LBC16,gotsmanpopl16}). The key technical issue in
+these efforts involves proving that atomic and fully isolated
+operations preserve application invariants when executed on these
+consistency levels.  In contrast, we admit weakly isolated
+transactions, and our system model accepts \emph{specifications} of
+consistency and isolation levels drawn from an expressive logic.
+Gotsman \emph{et al.}~\cite{gotsmanpopl16} adapts \iso{Parallel
+  Snapshot Isolation} to the aforementioned setting by interpreting it
+as a consistency level that serializes writes to objects; a dedicated
+proof rule is developed to help prove prove program invariants hold
+under this model. By parameterizing our proof system over a gamut of
+weak isolation specifications, we avoid the need to define a separate
+proof rule for each new isolation level we may encounter.
 
@@
 a relation between relaxed memory programs and related {\sc sc}
-programs. They also prove the soundness of CSL for the weakest memory
-model expressible in their semantics, but neither a parameterized (in
-the aforementioned sense) CSL, nor a CSL specialized for any relaxed
-memory model is proposed.  ~\cite{DLZ+13} presents a \emph{buffered
+programs. Demange \emph{et al.}~\cite{DLZ+13} present a \emph{buffered
   memory model} for Java that defines an axiomatic definition for the
 JMM in terms of memory reorderings, and an operational instantiation
-consistent with the TSO memory model found on x86 processors, and
-demonstrates the equivalence of the two definitions.  However, there
-is no attempt to parameterize the model over different reordering or
-consistency definitions, which is a central focus of this paper.
+consistent with the TSO memory model, and demonstrate the equivalence
+of the two definitions.

### 2016-07-04 ansi-isolation.tex
various edits to regain space
typo fixes mostly

@@
 written to a variable $X$ under $\E$ (i.e., $\underE{T_i \wrstoar X}$)
-if there exists a write-to-$X$ effect from $T_i$ in $\E$.
+if there exists a writes-to-$X$ effect from $T_i$ in $\E$.
 \emph{Reads-from} ($\underE{T_i \rdsfmar X}$) is defined similarly.
-$T_i \usesar X$ if it reads or writes $X$.
+$T_i \usesar X$ holds if $T_i$ reads or writes $X$.
 
@@
 This lets $T_j$ to never lose its own updates. Monotonic visibility
-adds one more constraint to read-my-writes; besides requiring $\eta_1$
+adds one more constraint to \emph{read-my-writes}; besides requiring $\eta_1$
 to be visible to $\eta_2$, it also requires every effect ($\eta$)
@@
 $\mathtt{AtomicVis}$ and $\mathtt{CommitVis}$ guarantees. The {\sc rc}
-specification\footnote{Note the distinction in fonts: {\sc rc} is an
-abbreviation of \iso{Read Committted}, whereas \C{RC} is its
-specification.} (\C{RC}) is therefore a combination of these two
+specification (\C{RC})\footnote{In the following, we use small caps to abbreviate
+  isolation levels (\emph{e.g.,} {\sc rc} for \emph{Read Committed}), and
+  typewriter font for abbreviations of the specification of an isolation
+  level (\emph{e.g.,} \C{RC} denotes the specification for {\sc rc} as
+  given in Fig.~\ref{fig:ansi-isolation}).} is therefore a combination of these two
 guarantees. The specification also agrees with the description and
@@
 
-\begin{figure}
-\centering
-\subcaptionbox {
-  {\sc rr}($T_1$), {\sc rr}($T_2$).
-  \label{fig:ansi-iso-eg-rr}
-} [
-  0.33\columnwidth
-] {
-  \includegraphics[scale=0.36]{Figures/rr-eg}
-}
-\subcaptionbox {
-  {\sc si}($T_1$), {\sc mav}($T_2$)
-  \label{fig:ansi-iso-eg-si}
-} [
-  0.33\columnwidth
-] {
-  \includegraphics[scale=0.36]{Figures/si-eg-2}
-}
-\subcaptionbox {
-  {\sc ser}($T_3$), {\sc mav}($T_4$)
-  \label{fig:ansi-iso-eg-ser}
-}{
-  \includegraphics[scale=0.32]{Figures/ser-eg}
-}
-
-\caption{Sample executions of concurrent transactions at different
-isolation levels. Solid (black) arrows indicate the timeline. Points
-on the timeline mark the time when an operation is executed. Dotted
-(blue) arrows denote $\visZ$. }
-\label{fig:ansi-iso-eg}
-\end{figure}
-
 Snapshot visibility (\C{SnapshotVis}) captures the scenario where a
@@
 forbidden for only a suffix (more generally, a subset) of $T_j$ to
-witness $T_i$. Specification of ANSI SQL \iso{Repeatable Read}
+witness $T_i$. The specification of ANSI SQL \iso{Repeatable Read}
 isolation (\C{RR}) extends \C{MAV} with the snapshot visibility
@@
 other operations of $T_4$ because of {\sc mav}. Note that {\sc ser}
-does not guarantee total order of transactions w.r.t. $\visZ$ because,
-in reality, databases do not guarantee serializable execution unless
-\emph{all} transactions choose {\sc ser}.
+does not guarantee a total order on transactions w.r.t. $\visZ$
+because, in practice, databases do not guarantee global serializable
+execution unless \emph{all} transactions choose {\sc ser}.
+
+\begin{figure}
+\centering
+\subcaptionbox {
+  {\sc rr}($T_1$), {\sc rr}($T_2$).
+  \label{fig:ansi-iso-eg-rr}
+} [
+  0.33\columnwidth
+] {
+  \includegraphics[scale=0.36]{Figures/rr-eg}
+}
+\subcaptionbox {
+  {\sc si}($T_1$), {\sc mav}($T_2$)
+  \label{fig:ansi-iso-eg-si}
+} [
+  0.33\columnwidth
+] {
+  \includegraphics[scale=0.36]{Figures/si-eg-2}
+}
+\subcaptionbox {
+  {\sc ser}($T_3$), {\sc mav}($T_4$)
+  \label{fig:ansi-iso-eg-ser}
+}{
+  \includegraphics[scale=0.32]{Figures/ser-eg}
+}
+
+\caption{\small Sample executions of concurrent transactions at different
+isolation levels. Solid (black) arrows indicate the timeline. Points
+on the timeline mark the time when an operation is executed. Dotted
+(blue) arrows denote $\visZ$. }
+\label{fig:ansi-iso-eg}
+\vspace*{-12pt}
+\end{figure}
 

### 2016-07-04 case-studies.tex
typo fixes mostly

@@
 levels, and spent 5 man hours determining appropriate isolation levels
-needed to enforce the invariant\footnote{The representative \txnimp
+needed to enforce the invariant.\footnote{The representative \txnimp
   programs we used to formally reason about scenarios described in
@@
 were able to formally prove that to preserve the invariant, writers
-need to execute at \iso{Serializable} isolation, while readers can
-execute at \iso{Read Uncommitted}. Repeating the experiments with this
-configuration led to no violations, as expected. While executing all
-transactions at {\sc ser} also prevents violations, it led to an
-additional 113\% increase in the latency of writes on an average (32
-concurrent writers; 5 rounds). We repeated the experiments with the
-same configuration on Postgres, and found no violations. However,
-Postgres's isolation levels~\cite{postgresiso} are stronger than ANSI
-SQL's in non-trivial ways. We therefore formalized Postgres'
-guarantees and repeated relevant parts of the proof to discover that
-\iso{Repeatable Read} is in fact sufficient for writers - an
-observation we confirmed through experiments. As usual, using {\sc
+need to execute using \iso{Serializable} isolation, while readers can
+execute using \iso{Read Uncommitted}. Repeating the experiments with
+this configuration led to no violations, as expected. While executing
+all transactions at {\sc ser} also prevents violations, it led to an
+additional 113\% increase in the latency of writes on average, using a
+configuration with 32 concurrent writers, and 5 rounds. We repeated
+the experiments with the same configuration on Postgres, and found no
+violations. However, Postgres' isolation levels~\cite{postgresiso} are
+stronger than ANSI SQL's in non-trivial ways. We therefore formalized
+Postgres' guarantees and repeated relevant parts of the proof to
+discover that \iso{Repeatable Read} is in fact sufficient for writers
+- an observation confirmed through experiments. As usual, using {\sc
   ser} for all transactions also suffices, but on Postgres results in
@@
 % techniques.   
-To understand the effect of changing the store consistency levels, we
+To understand the effect of changing store consistency levels, we
 re-implemented the counter as an op-based \emph{replicated data type}
@@
 increments need not be serializable for reads to witness a
-monotonically increasing count\footnote{The trade-off is that the view
-count is now no longer accurate, although it grows monotonically.}.
+monotonically increasing count.\footnote{The trade-off is that the view
+count is now no longer accurate, although it grows monotonically.}
 We implemented this version of the counter as an op-based RDT, which
@@
 \paragraph{Microblog.} Our second set of experiments focused on a Twitter-like microblogging
-site, also written using Ruby-on-Rails~\cite{railsbook}. One of the
-invariants that the application chooses to enforce is referential
+site, also written using Ruby-on-Rails~\cite{railsbook}.  An important
+invariant that must be preserved is referential
 integrity between microposts and authors. The application enforces
@@
 carried out in a transaction that also includes the deletion of
-microposts authored by the user.  However, concurrent observers can
-nonetheless witness violations of referential integrity. To measure
+microposts authored by the user.  However, it is possible for concurrent observers 
+to nonetheless witness invariant violations. To measure
 the likelihood of such anomalies, we first populated a Postgres
@@
 reads require {\sc rr}. The proof is immediately applicable for MySQL,
-which implements ANSI SQL isolation levels. Repeating our experiments
+which implements ANSI SQL isolation. Repeating our experiments
 for MySQL with this configuration resulted in no violations (as
-expectected), while yielding a 38\% reduction in latency than an {\sc
+expected), while yielding a 38\% reduction in latency than an {\sc
   ser} configuration (96 concurrent clients; 5 rounds).  Since
-Postgres's isolation levels are stronger than the corresponding ANSI
+Postgres isolation levels are stronger than the corresponding ANSI
 levels, \rulelabel{RG-Conseq} rule (Fig.~\ref{fig:rg-rules}) lets us

### 2016-07-04 introduction.tex
typo fixes mostly

@@
     of traces induced by the operational semantics, in which transactions
-    are associated with specific weak-isolation levels.
+    are associated with specific weak isolation levels.
   \item We define a \emph{maximum visibility principle} to reconcile
@@
     requirements, and use it to instantiate our operational model and
-    the proof system for various kinds of (weakly consistent) data
+    the proof system for various kinds of (weakly-consistent) data
     stores.

### 2016-07-04 motivation.tex
typo fixes mostly

@@
 As an alternative, consider the execution of this transaction under
-\emph{Read Committed} ({\sc rc}) isolation level, which is weaker than
-{\sc ser}\footnote{{\sc rc} is in fact the default isolation level in
+a \emph{Read Committed} ({\sc rc}) isolation level, which is weaker than
+{\sc ser}.\footnote{{\sc rc} is in fact the default isolation level in
 Postgres 9.5 and Oracle 11g databases.} An {\sc rc} transaction is
@@
 the current balance (\C{k}) and writes the new balance (\C{k-a1}), but
-before it commits transaction \C{Wd2} executes and commits, writing
+before it commits, transaction \C{Wd2} executes and commits, writing
 the new balance (\C{k-a2}). {\sc rc} isolation prevent \C{Wd2} from
@@
 Subsequently committing \C{Wd1} leads to the loss of \C{Wd2}'s updates
-(the so called \emph{lost update anomaly}~\cite{berenson}), resulting
+(the so-called \emph{lost update anomaly}~\cite{berenson}), resulting
 in an incorrect balance of \C{k-a1}. The second execution describes a
@@
 whereas within a transaction executing under \iso{Read Uncommitted}
-isolation all interferences are valid. Between these two extremes are
+isolation, all interferences are valid. Between these two extremes are
 various levels of isolation that admit some interferences while
 prohibiting others. For example, \iso{Read Committed} isolation admits
-interference of a committed transaction, but not that of an
-uncommitted transaction. \iso{Snapshot Isolation} admits interference
+interference of a committed transaction, but not those from
+uncommitted transactions. \iso{Snapshot Isolation} admits interference
 from committed non-conflicting transactions, and so on.
@@
 described in \S\ref{sec:ansi-isolation} are more nuanced.} in
-Fig.~\ref{fig:interference-ex}.  \iso{Read Committed} specification
-($\psi_{RC}$) allows an operation $\eta_1$ of $T_1$ to happen before
+Fig.~\ref{fig:interference-ex}.  The \iso{Read Committed} specification
+($\psi_{RC}$) allows an operation $\eta_1$ of $T_1$ to happen-before
 an operation $\eta_2$ of $T_2$ only if every operation in $T_1$
-(including its commit) happens before $\eta_2$ (we let $T_1 \hboar
+(including its commit) happens-before $\eta_2$ (we let $T_1 \hboar
 \eta_2$ denote this). The specification does not require $T_1$ to
@@
 all dashed arrows, because dashed arrows establish $\hbZ$ between actions in \C{Wd2}
-and only a subset of the operations in \C{Wd1}. Ligher (blue) arrows with
+and only a subset of the operations in \C{Wd1}. Lighter (blue) arrows with
 hollow heads denote $\hbZ$ relationships that do not affect the value
@@
 framework should be able to use the specification to determine if an
-interference is valid or invalid, allowing the programmer to only
+interference is valid or not, allowing the programmer to only
 focus on the former. Our second observation is that this is possible
@@
 to \C{C} in \C{Wd1}. The expected invariant ($\phi$) at that program
-point is shown below ($\committed$ stands for ``committed''):
+point is shown below ($\committed$ stands for ``committed''):\vspace*{-8pt}
 
 \begin{smathpar}
-\begin{array}{l}
-  \neg\committed(\C{Wd1}) \conj (\neg\committed(\C{Wd2}) \Rightarrow
-  \C{C = k-a1}) 
-                \\
-      \C{Wd1} \wrstoar \C{C} \conj (\committed(\C{Wd2})
+\begin{array}{c}
+ \neg\committed(\C{Wd1}) \conj \C{Wd1} \wrstoar \C{C}  \conj  \\
+    (\neg\committed(\C{Wd2}) \Rightarrow \C{C = k-a1}) \conj (\committed(\C{Wd2})
                 \Rightarrow \C{C = k-a1-a2})
@@
 \end{smathpar}
+\vspace*{-8pt}
 

### 2016-07-04 case-studies.tex
Final set of major edits.

--- 2016-07-04 case-studies.tex
@@
 applicability of our reasoning approach in varied settings, and its
-consequent practical utility.
+consequent practical utility. Both our case studies target
+Ruby-on-Rails applications. To analyze these, and other similar
+open-source applications, we developed a symbolic execution engine
+called {\sc MagLev} that compiles a Rails program to an abstract
+program in an extended version of \txnimp\footnote{The details of {\sc
+MagLev} can be found in the supplementary. Its source code, along with
+that of our applications and experiments, is available at the
+following anonymized url: \url{...}}. The focus of our formal
+verification efforts are such \txnimp programs.
 
-\paragraph{A video view counter.} Our first application is a
-Ruby-on-Rails implementation of a counter to count the number of views
-a video has garnered on a video-sharing website like
-YouTube\footnote{Source code of the applications, experiments, and the
-  tools we developed to analyze Ruby can be found at the following
-  anonymized url: \url{...}}. The application stores the video count
-in a MySQL database, and supports a read operation (a database read
-transaction), and an increment operation (a read-increment-write
-sequence wrapped in a transaction). The application is expected to
-preserve the \emph{monotonicity invariant}, namely that the view count
-should never appear to be decreasing. However, the default
-\iso{Repeatable Read} isolation level of MySQL leads to anomalous
-executions - with 32 concurrent writers performing single increments,
-and one reader issuing a constant stream of reads, we observed 12
-violations of monotonicity in 10 rounds. We subsequently instantiated
-our reasoning framework for an {\sc sc} store with ANSI SQL isolation
-levels, and spent 5 man hours determining appropriate isolation levels
-needed to enforce the invariant\footnote{The representative \txnimp
-  programs we used to formally reason about scenarios described in
-  this section can be found in the supplementary.} In particular, we
-were able to formally prove that to preserve the invariant, writers
-need to execute at \iso{Serializable} isolation, while readers can
-execute at \iso{Read Uncommitted}. Repeating the experiments with this
-configuration led to no violations, as expected. While executing all
-transactions at {\sc ser} also prevents violations, it led to an
-additional 113\% increase in the latency of writes on an average (32
-concurrent writers; 5 rounds). We repeated the experiments with the
-same configuration on Postgres, and found no violations. However,
-Postgres's isolation levels~\cite{postgresiso} are stronger than ANSI
-SQL's in non-trivial ways. We therefore formalized Postgres'
-guarantees and repeated relevant parts of the proof to discover that
-\iso{Repeatable Read} is in fact sufficient for writers - an
-observation we confirmed through experiments. As usual, using {\sc
-  ser} for all transactions also suffices, but on Postgres results in
-an additional 47\% increase in write latency.
+\paragraph{A video view counter.} Our first application is a Rails
+implementation of a counter to count the number of views a video has
+garnered on a video-sharing website like YouTube. The application
+stores the video count in a MySQL database, and supports a read
+operation (a database read transaction), and an increment operation (a
+read-increment-write sequence wrapped in a transaction).  The
+application is expected to preserve the \emph{monotonicity invariant},
+namely that the view count should never appear to be decreasing.
+However, the default \iso{Repeatable Read} isolation level of MySQL
+leads to anomalous executions - with 32 concurrent writers performing
+single increments, and one reader issuing a constant stream of reads,
+we observed 12 violations of monotonicity in 10 rounds. We
+subsequently instantiated our reasoning framework for an {\sc sc}
+store with ANSI SQL isolation levels, and spent 5 man hours
+determining appropriate isolation levels needed to enforce the
+invariant. In particular, we were able to formally prove that to
+preserve the invariant, writers need to execute at \iso{Serializable}
+isolation, while readers can execute at \iso{Read Uncommitted}.
+Repeating the experiments with this configuration led to no
+violations, as expected. While executing all transactions at {\sc ser}
+also prevents violations, it led to an additional 113\% increase in
+the latency of writes on an average (32 concurrent writers; 5 rounds).
+We repeated the experiments with the same configuration on Postgres,
+and found no violations. However, Postgres's isolation
+levels~\cite{postgresiso} are stronger than ANSI SQL's in non-trivial
+ways. We therefore formalized Postgres' guarantees and repeated
+relevant parts of the proof to discover that \iso{Repeatable Read} is
+in fact sufficient for writers - an observation we confirmed through
+experiments. As usual, using {\sc ser} for all transactions also
+suffices, but on Postgres results in an additional 47\% increase in
+write latency.
 
@@
 \vspace*{-3pt}
-\paragraph{Microblog.} Our second set of experiments focused on a Twitter-like microblogging
-site, also written using Ruby-on-Rails~\cite{railsbook}. One of the
-invariants that the application chooses to enforce is referential
-integrity between microposts and authors. The application enforces
-this invariant entirely in user space by relying on database-backed
-transactions (this behavior is typical of Rails
+\paragraph{Microblog.} Our second set of experiments focused on a
+Twitter-like microblogging application taken from~\cite{railsbook}.
+One of the invariants that the application chooses to enforce is
+referential integrity between microposts and authors. The application
+enforces this invariant entirely in user space by relying on
+database-backed transactions (this behavior is typical of Rails
 applications~\cite{bailisferal}).  For instance, deletion of a user is
@@
 randomly chosen users in a 1:7 ratio, and witnessed 250 violations of
-referential integrity. To determine appropriate isolation levels, we
-then instantiated our framework for an {\sc sc} store with various
-ANSI SQL isolation levels for user deletions and timeline reads, and
-obtained a proof that {\sc rc} is sufficient for deletions, whereas
-reads require {\sc rr}. The proof is immediately applicable for MySQL,
-which implements ANSI SQL isolation levels. Repeating our experiments
-for MySQL with this configuration resulted in no violations (as
-expectected), while yielding a 38\% reduction in latency than an {\sc
-  ser} configuration (96 concurrent clients; 5 rounds).  Since
-Postgres's isolation levels are stronger than the corresponding ANSI
-levels, \rulelabel{RG-Conseq} rule (Fig.~\ref{fig:rg-rules}) lets us
+referential integrity (the isolation level was left at the default
+{\sc rc} level for experiments). To determine appropriate isolation
+levels, we then instantiated our framework for an {\sc sc} store with
+various ANSI SQL isolation levels for user deletions and timeline
+reads, and obtained a proof that {\sc rc} is sufficient for deletions,
+whereas reads require {\sc rr}. The proof is immediately applicable
+for MySQL, which implements ANSI SQL isolation levels. Repeating our
+experiments for MySQL with this configuration resulted in no
+violations (as expectected), while yielding a 38\% reduction in
+latency than an {\sc ser} configuration.  Since Postgres's isolation
+levels are stronger than the corresponding ANSI levels,
+\rulelabel{RG-Conseq} rule (Fig.~\ref{fig:rg-rules}) lets us
 immediately deduce that the same choice of isolation levels also work
@@
 variants as distinct, yet comparable, instantiations.
-
-%  GK: what does the following sentence mean?
-%  As a consequence, we avoid the need for semantics re-engineering
-%  every time an application is retargeted to a new platform.

### 2016-07-04 motivation.tex
Final set of major edits.

--- 2016-07-04 motivation.tex
@@
 free from \emph{dirty reads}~\cite{berenson} that read uncommitted
-data. In the current example, {\sc rc} isolation admits the following
-two executions (Fig.~\ref{fig:rc-ex}) on a strongly consistent ({\sc
-sc}) store, such as a relational database:
+data. In the current example, {\sc rc} isolation admits the two
+executions shown in Fig.~\ref{fig:rc-ex} on a strongly consistent
+({\sc sc}) store, such as a relational database.
 
@@
 ] {
-  \includegraphics[scale=0.5]{Figures/motiv-eg-1-a}
+  \includegraphics[scale=0.6]{Figures/motiv-eg-1-a}
 }
@@
 }{
-  \includegraphics[scale=0.5]{Figures/motiv-eg-1-b}
+  \includegraphics[scale=0.6]{Figures/motiv-eg-1-b}
 }
 \caption{\small A possible execution of the program shown in Fig.~\ref{fig:motiv-eg-1} under
-  a read committed isolation level.}
+  \iso{Read Committed} isolation level.}
 \label{fig:rc-ex}
 \end{figure}
+
 The figure depicts an execution as a series of read, write, and commit
@@
 
-\begin{figure}
-\begin{smathpar}
-\begin{array}{lcl}
-\psi_{RC} & \defeq & \forall T_1,T_2,\eta_1,\eta_2,.\; \txn(\eta_1) = T_1 
-  \conj \txn(\eta_2) = T_2 \\
-  & & \hspace*{0.6in}\conj T_1 \neq T_2 \conj \eta_1 \hboar
-  \eta_2 \Rightarrow T_1 \hboar \eta_2 \\
-\psi_{SI} & \defeq & \forall T_1,T_2.\; T_1 \neq T_2 \conj
-  (\exists \C{X}.~{T_1 \wrstoar \C{X}} \conj 
-                      {T_2 \wrstoar \C{X}})\\
-  &  & \hspace*{0.6in}\Rightarrow{T_1 \hboar T_2} \disj {T_2 \hboar T_1} \\
-\end{array}
-\end{smathpar}
-\caption{\small Visibility and interference properties for different weak isolation levels can be
-  captured by different instantiations of their happens-before relation.}
-\label{fig:interference-ex}
-\end{figure}
-
 Thinking in terms of anomalies, as described above, is how database
@@
 example.
-
-\begin{figure}
-\centering
-  \includegraphics[scale=0.4]{Figures/motiv-eg-1-hb}
-
-\caption{\small $\hbZ$ arrows allowed by {\sc si} (solid) and {\sc rc} (solid
-  and dashed) isolation levels for the example in
-  Fig.~\ref{fig:motiv-eg-1}. Each arrow represents an $\hbZ$ relationship
-  among all operations of `Wd2' and operations in `Wd1' that follow
-  the arrow head. Letters R, W and C stand for read, write and commit,
-  respectively.}
-\label{fig:motiv-eg-1-hb}
-\vspace*{-8pt}
-\end{figure}
 
@@
 
+\begin{figure}
+\begin{smathpar}
+\begin{array}{lcl}
+\psi_{RC} & \defeq & \forall T_1,T_2,\eta_1,\eta_2,.\; \txn(\eta_1) = T_1 
+  \conj \txn(\eta_2) = T_2 \\
+  & & \hspace*{0.6in}\conj T_1 \neq T_2 \conj \eta_1 \hboar
+  \eta_2 \Rightarrow T_1 \hboar \eta_2 \\
+\psi_{SI} & \defeq & \forall T_1,T_2.\; T_1 \neq T_2 \conj
+  (\exists \C{X}.~{T_1 \wrstoar \C{X}} \conj 
+                      {T_2 \wrstoar \C{X}})\\
+  &  & \hspace*{0.6in}\Rightarrow{T_1 \hboar T_2} \disj {T_2 \hboar T_1} \\
+\end{array}
+\end{smathpar}
+\caption{\small Visibility and interference properties for different weak isolation levels can be
+  captured by different instantiations of their happens-before relation.}
+\label{fig:interference-ex}
+\end{figure}
+
+\begin{figure}
+\centering
+  \includegraphics[scale=0.4]{Figures/motiv-eg-1-hb}
+
+\caption{\small $\hbZ$ arrows allowed by {\sc si} (solid) and {\sc rc} (solid
+  and dashed) isolation levels for the example in
+  Fig.~\ref{fig:motiv-eg-1}. Each arrow represents an $\hbZ$ relationship
+  among all operations of \C{Wd2} and operations in \C{Wd1} that follow
+  the arrow head. Letters R, W and C stand for read, write and commit,
+  respectively.}
+\label{fig:motiv-eg-1-hb}
+\vspace*{-8pt}
+\end{figure}
+
 For the rely-guarantee approach to be useful, it has to thus constrain
@@
 we can constrain interference by constraining the definition of the
-\emph{happens-before} ($\hbZ$) relation that dictates visibility
-actions between transactions, or their respective constituents.  To do
-so, we axiomatize the $\hbZ$ relation to capture the interference
-characteristics of different isolation levels as their specifications.
-Two such specifications are shown\footnote{Note that specifications
-presented here are solely for illustration. The actual specifications
-described in \S\ref{sec:ansi-isolation} are more nuanced.} in
-Fig.~\ref{fig:interference-ex}.  \iso{Read Committed} specification
-($\psi_{RC}$) allows an operation $\eta_1$ of $T_1$ to happen before
-an operation $\eta_2$ of $T_2$ only if every operation in $T_1$
-(including its commit) happens before $\eta_2$ (we let $T_1 \hboar
-\eta_2$ denote this). The specification does not require $T_1$ to
-execute and commit before all operations of $T_2$, thus allowing it to
-interfere in $T_2$. The \iso{Snapshot Isolation} specification
+\emph{happens-before} ($\hbZ$) relation, which ultimately dictates
+what actions become visible to a transaction, or its constituents, and
+when. To do so, we axiomatize the $\hbZ$ relation to capture the
+interference characteristics of different isolation levels as their
+specifications.  Two such specifications are shown\footnote{Note that
+specifications presented here are solely for illustration. The actual
+specifications described in \S\ref{sec:ansi-isolation} are more
+nuanced.} in Fig.~\ref{fig:interference-ex}.  \iso{Read Committed}
+specification ($\psi_{RC}$) allows an operation $\eta_1$ of $T_1$ to
+happen before an operation $\eta_2$ of $T_2$ only if every operation
+in $T_1$ (including its commit) happens before $\eta_2$ (we let $T_1
+\hboar \eta_2$ denote this). The specification does not require $T_1$
+to execute and commit before all operations of $T_2$, thus allowing it
+to interfere in $T_2$. The \iso{Snapshot Isolation} specification
 ($\psi_{SI}$) requires two transactions that write to the same
@@
 
-Fig.~\ref{fig:motiv-eg-1-hb} is a visualization of $\hbZ$ 
-from \C{Wd2} to \C{Wd1} allowed by $\psi_{RC}$ and $\psi_{SI}$. All
-arrows are legal under $\psi_{RC}$ because in no case does an operation
-$\eta_1$ from \C{Wd1} happen-before $\eta_2$ of \C{Wd2} without the commit
-of \C{Wd1} also happening before $\eta_2$. $\psi_{SI}$ however disallows
-all dashed arrows, because dashed arrows establish $\hbZ$ between actions in \C{Wd2}
-and only a subset of the operations in \C{Wd1}. Ligher (blue) arrows with
-hollow heads denote $\hbZ$ relationships that do not affect the value
-of \C{C} in a way that causes the program to violate its
-post-condition.  Darker (black) arrows with solid heads denote $\hbZ$
-relationships that lead to the violation of the post-condition. The task of
-the reasoning framework is to determine if all $\hbZ$ relationships
-allowed by an isolation level lead to the satisfaction of the
-post-condition.
+Fig.~\ref{fig:motiv-eg-1-hb} is a visualization of $\hbZ$ from \C{Wd2}
+to \C{Wd1} allowed by $\psi_{RC}$ and $\psi_{SI}$. All arrows are
+legal under $\psi_{RC}$ because in no case does an operation $\eta_1$
+from \C{Wd1} happen-before $\eta_2$ of \C{Wd2} without the commit of
+\C{Wd1} also happening before $\eta_2$. $\psi_{SI}$ however disallows
+all dashed arrows, because dashed arrows establish $\hbZ$ between
+actions in \C{Wd2} and only a subset of the operations in \C{Wd1}.
+Since the result of an operation (\eg a read) depends on what happens
+before that operation, $\hbZ$ relation ultimately affects the result
+of the program. The solid arrow and the first (from left) dashed arrow
+in Fig.~\ref{fig:motiv-eg-1-hb} denote $\hbZ$ relationships that do
+not affect the value of \C{C} in a way that causes the program to
+violate its post-condition. The last two dashed arrows denote $\hbZ$
+relationships that lead to the violation of the post-condition. The
+task of the reasoning framework is to determine if all $\hbZ$
+relationships allowed by an isolation level lead to the satisfaction
+of the post-condition.
 
@@
 
-\noindent $\phi$ asserts that \C{Wd1} is not yet committed, and that it wrote to
-\C{C}, and the value of \C{C} is either \C{k-a1-a2} or \C{k-a1}
-depending on whether or not \C{Wd2} is committed. If $\phi$ remains
-invariant until \C{Wd1} commits, then the post-condition (\C{C =
-k-a1-a2}) can be established easily. However, an interference from
+\noindent $\phi$ asserts that \C{Wd1} is not yet committed, and that
+it wrote to \C{C}, and the value of \C{C} is either \C{k-a1-a2} or
+\C{k-a1} depending on whether or not \C{Wd2} is committed. If $\phi$
+remains invariant until \C{Wd1} commits, then the post-condition (\C{C
+= k-a1-a2}) can be established easily. However, an interference from
 \C{Wd2} at this stage (captured by the last dashed arrow in
@@
 \iso{Snapshot Isolation} prevents this interference, and this can be
-shown by demonstrating that an interference from \C{Wd2} starting from an
-execution state that satisfies $\psi_{SI} \wedge \phi$ leads to an
+shown by demonstrating that an interference from \C{Wd2} starting from
+an execution state that satisfies $\psi_{SI} \wedge \phi$ leads to an
 execution state where neither $\C{Wd2} \hboar \C{Wd1}$ nor $\C{Wd1}
-\hboar \C{Wd2}$ holds (the former does not hold because at least one write
-from \C{Wd1} has happened before at least one operation of \C{Wd2}, and
-the latter does not hold because \C{Wd1} has not yet committed, and \C{Wd2}
-has already begun). Since \C{Wd2} also writes to \C{C}, this violates
-the $\psi_{SI}$ constraint which we assume to be an invariant. A proof
-that the post-condition holds now follows from the contradiction. It is
-informative to note that if the invariant is $\psi_{RC}$ instead of
-$\psi_{SI}$, we cannot derive a contradiction and we cannot rule out
-the interference, which (rightfully) causes  the proof to fail.
+\hboar \C{Wd2}$ holds ($\C{Wd2} \hboar \C{Wd1}$ does not hold because
+\C{Wd1}'s write to \C{C} clearly happened before \C{Wd2}'s commit, and
+$\C{Wd1} \hboar \C{Wd2}$ does not hold because \C{Wd1} has not yet
+committed, while \C{Wd2} has already begun). Since \C{Wd2} also writes
+to \C{C}, this violates the $\psi_{SI}$ constraint which we assume to
+be an invariant. A proof that the post-condition holds now follows
+from the contradiction. It is informative to note that if the
+invariant is $\psi_{RC}$ instead of $\psi_{SI}$, we cannot derive a
+contradiction and we cannot rule out the interference, which
+(rightfully) causes  the proof to fail.
 

### 2016-07-04 opsem.tex
typo fixes mostly
Final set of major edits.

@@
 in our model, composed of more fundamental \emph{session order}
-($\soZ$) and \emph{visibility} ($\visZ$) relations, which shall
-henceforth replace $\hbZ$ in the discussion.  A session order relation
-captures the sequential order of operations within a transaction. In
-particular, it relates two effects, $\eta_1$ and $\eta_2$, such that
-$\txn(\eta_1) = \txn(\eta_2)$ and $\id(\eta_1) < \id(\eta_2)$.  The
-semantics assigns monotonically increasing identifiers to effects, as
-defined by the $\id(\eta) > \maxId(\A)$ condition of
-\rulelabel{E-Aux}.  Evaluation contexts ($\ectx_i$) for
-transaction-bound terms are defined so as to enforce a deterministic
-sequential order of execution within a transaction, leading to a
-deterministic total order among effect ids, which defines the session
-order relation. Visibility ($\visZ$) on the other hand relates effects
-across concurrent transactions.  Intuitively, $\visZ$ relates $\eta_1$
-to $\eta_2$ if and only if $\eta_1$ was \emph{visible} to the
-operation that generated $\eta_2$ during its execution, thus effecting
-its return value ($\rval(\eta_2)$). For example, a read operation over
-\C{X} may pick the value ($\rval$) of the write effect with the highest id
-among its visible effects (this is made possible by appropriately
-defining $\interp{\cdot}$ in \rulelabel{E-Var}, as we show later).
-Thus, the value of a read depends on what write effects it can
-witness. An operation can only witness the effects of already
-concluded operations, which varies between executions due to the
+($\soZ$) and \emph{visibility} ($\visZ$) relations. In particular,
+$\hbZ \,=\, (\soZ \cup \visZ)^{+}$. Unlike $\hbZ$, $\visZ$ (defined
+below) is not a transitive relation, hence lets us capture finer
+properties than $\hbZ$. We therefore let $\visZ$ replace $\hbZ$ in our
+discussion henceforth. The session order relation captures the
+sequential order of operations within a transaction. In particular, it
+relates two effects, $\eta_1$ and $\eta_2$, such that $\txn(\eta_1) =
+\txn(\eta_2)$ and $\id(\eta_1) < \id(\eta_2)$.  The semantics assigns
+monotonically increasing identifiers to effects, as defined by the
+$\id(\eta) > \maxId(\A)$ condition of \rulelabel{E-Aux}. Evaluation
+contexts ($\ectx_i$) for transaction-bound terms are defined so as to
+enforce a deterministic sequential order of execution within a
+transaction, leading to a deterministic total order among effect ids,
+which defines the session order relation. Visibility ($\visZ$) on the
+other hand relates effects across concurrent transactions.
+Intuitively, $\visZ$ relates $\eta_1$ to $\eta_2$ if and only if
+$\eta_1$ was \emph{visible} to the operation that generated $\eta_2$
+during its execution, thus effecting its return value
+($\rval(\eta_2)$). For example, a read operation over \C{X} may pick
+the value ($\rval$) of the write effect with the highest id among its
+visible effects (this is made possible by appropriately defining
+$\interp{\cdot}$ in \rulelabel{E-Var}, as we show later). Thus, the
+value of a read depends on what write effects it can witness. An
+operation can only witness the effects of already concluded
+operations, which varies between executions due to the
 non-deterministic order of evaluating the parallel composition of

### 2016-07-04 ansi-isolation.tex
Final set of major edits.

@@
 containment relation is trivially lifted to the set of effects to
-define $\underE{S \subseteq T_i}$.  Visibility and session order
-relations are denoted by $\eta_1 \visar \eta_2$ and $\eta_1 \soar
-\eta_2$, respectively. A transaction $T_i$ is said to be visible to an
-effect $\eta$ if every effect $\eta_1$ of $T_i$ recorded by the trace
-is visible to $\eta$.  $T_i$ may be visible to $\eta$ but may not be
-visible to every other effect in the $\txn(\eta)$. For a transaction
-$T_i$ to be considered to be visible to a transaction $T_j$ in the
-context of a trace $\E$ (written $\underE{T_i \visar T_j}$), every
-effect ($\eta_1$) of $T_i$ present in $\E$ must be visible to every
-effect ($\eta_2$) of $T_j$ in $\E$.  Conversely, if none of the
-effects of $T_i$ present in $\E$ are visible to any effect of $T_j$,
-then $T_i$ is considered invisible to $T_j$ under $\E$ (written
-$\underE{T_i \invisar T_j}$). Transaction $T_i$ is said to have
-written to a variable $X$ under $\E$ (i.e., $\underE{T_i \wrstoar X}$)
-if there exists a writes-to-$X$ effect from $T_i$ in $\E$.
-\emph{Reads-from} ($\underE{T_i \rdsfmar X}$) is defined similarly.
-$T_i \usesar X$ holds if $T_i$ reads or writes $X$.
+define $\underE{S \subseteq T_i}$.  Visibility, session order, and
+happens-before relations are denoted by $\eta_1 \visar \eta_2$,
+$\eta_1 \soar \eta_2$, and $\eta_1 \hboar \eta_2$, respectively. A
+transaction $T_i$ is said to be visible to an effect $\eta$ if every
+effect $\eta_1$ of $T_i$ recorded by the trace is visible to $\eta$.
+$T_i$ may be visible to $\eta$ but may not be visible to every other
+effect in the $\txn(\eta)$. For a transaction $T_i$ to be considered
+to be visible to a transaction $T_j$ in the context of a trace $\E$
+(written $\underE{T_i \visar T_j}$), every effect ($\eta_1$) of $T_i$
+present in $\E$ must be visible to every effect ($\eta_2$) of $T_j$ in
+$\E$.  Conversely, if none of the effects of $T_i$ present in $\E$ are
+visible to any effect of $T_j$, then $T_i$ is considered invisible to
+$T_j$ under $\E$ (written $\underE{T_i \invisar T_j}$). Transaction
+$T_i$ is said to have written to a variable $X$ under $\E$ (i.e.,
+$\underE{T_i \wrstoar X}$) if there exists a $\C{WR}(X)$ effect from
+$T_i$ in $\E$.  \emph{Reads-from} ($\underE{T_i \rdsfmar X}$) is
+defined similarly. $T_i$ \emph{uses} $X$ ($T_i \usesar X$) if it reads
+or writes $X$.
 

### 2016-07-04 case-studies.tex

@@
 invariant. In particular, we were able to formally prove that to
-preserve the invariant, writers need to execute at \iso{Serializable}
-isolation, while readers can execute at \iso{Read Uncommitted}.
-Repeating the experiments with this configuration led to no
-violations, as expected. While executing all transactions at {\sc ser}
-also prevents violations, it led to an additional 113\% increase in
-the latency of writes on an average (32 concurrent writers; 5 rounds).
-We repeated the experiments with the same configuration on Postgres,
-and found no violations. However, Postgres's isolation
-levels~\cite{postgresiso} are stronger than ANSI SQL's in non-trivial
-ways. We therefore formalized Postgres' guarantees and repeated
-relevant parts of the proof to discover that \iso{Repeatable Read} is
-in fact sufficient for writers - an observation we confirmed through
-experiments. As usual, using {\sc ser} for all transactions also
-suffices, but on Postgres results in an additional 47\% increase in
-write latency.
+preserve the invariant, writers need to execute using
+\iso{Serializable} isolation, while readers can execute using
+\iso{Read Uncommitted}. Repeating the experiments with this
+configuration led to no violations, as expected. While executing all
+transactions at {\sc ser} also prevents violations, it led to an
+additional 113\% increase in the latency of writes averaged over 5
+rounds with 32 concurrent writers.  We repeated the experiments with
+the same configuration on Postgres, and found no violations. However,
+Postgres' isolation levels~\cite{postgresiso} are stronger than ANSI
+SQL's in non-trivial ways. We therefore formalized Postgres'
+guarantees and repeated relevant parts of the proof to discover that
+\iso{Repeatable Read} is in fact sufficient for writers - an
+observation we confirmed through experiments. As usual, using {\sc
+ser} for all transactions also suffices, but on Postgres results in an
+additional 47\% increase in write latency.
 
@@
 \vspace*{-3pt}
+
 \paragraph{Microblog.} Our second set of experiments focused on a
 Twitter-like microblogging application taken from~\cite{railsbook}.
-One of the invariants that the application chooses to enforce is
-referential integrity between microposts and authors. The application
-enforces this invariant entirely in user space by relying on
-database-backed transactions (this behavior is typical of Rails
+An important invariant that must be preserved is the referential
+integrity between microposts and authors. The application enforces
+this invariant entirely in user space by relying on database-backed
+transactions (this behavior is typical of Rails
 applications~\cite{bailisferal}).  For instance, deletion of a user is
 carried out in a transaction that also includes the deletion of
-microposts authored by the user.  However, concurrent observers can
-nonetheless witness violations of referential integrity. To measure
-the likelihood of such anomalies, we first populated a Postgres
-database with 1000 user accounts, each user with 50 microposts and 20
-followers, chosen uniformly at random. We then constructed experiments
-with 64 concurrent clients performing deletions and timeline reads of
-randomly chosen users in a 1:7 ratio, and witnessed 250 violations of
-referential integrity (the isolation level was left at the default
-{\sc rc} level for experiments). To determine appropriate isolation
-levels, we then instantiated our framework for an {\sc sc} store with
-various ANSI SQL isolation levels for user deletions and timeline
-reads, and obtained a proof that {\sc rc} is sufficient for deletions,
-whereas reads require {\sc rr}. The proof is immediately applicable
-for MySQL, which implements ANSI SQL isolation levels. Repeating our
-experiments for MySQL with this configuration resulted in no
-violations (as expectected), while yielding a 38\% reduction in
-latency than an {\sc ser} configuration.  Since Postgres's isolation
-levels are stronger than the corresponding ANSI levels,
-\rulelabel{RG-Conseq} rule (Fig.~\ref{fig:rg-rules}) lets us
+microposts authored by the user.  However, it is possible for
+concurrent observers to nonetheless witness invariant violations. To
+measure the likelihood of such anomalies, we first populated a
+Postgres database with 1000 user accounts, each user with 50
+microposts and 20 followers, chosen uniformly at random. We then
+constructed experiments with 64 concurrent clients performing
+deletions and timeline reads of randomly chosen users in a 1:7 ratio,
+and witnessed 250 violations of referential integrity (the isolation
+level was left at the default {\sc rc} level for experiments). To
+determine appropriate isolation levels, we then instantiated our
+framework for an {\sc sc} store with various ANSI SQL isolation levels
+for user deletions and timeline reads, and obtained a proof that {\sc
+rc} is sufficient for deletions, whereas reads require {\sc rr}. The
+proof is immediately applicable for MySQL, which implements ANSI SQL
+isolation levels. Repeating our experiments for MySQL with this
+configuration resulted in no violations (as expected), while yielding
+a 38\% reduction in latency than an {\sc ser} configuration.  Since
+Postgres's isolation levels are stronger than the corresponding ANSI
+levels, \rulelabel{RG-Conseq} rule (Fig.~\ref{fig:rg-rules}) lets us
 immediately deduce that the same choice of isolation levels also work

### 2016-07-04 motivation.tex

@@
 specifications described in \S\ref{sec:ansi-isolation} are more
-nuanced.} in Fig.~\ref{fig:interference-ex}.  \iso{Read Committed}
+nuanced.} in Fig.~\ref{fig:interference-ex}. The \iso{Read Committed}
 specification ($\psi_{RC}$) allows an operation $\eta_1$ of $T_1$ to
-happen before an operation $\eta_2$ of $T_2$ only if every operation
-in $T_1$ (including its commit) happens before $\eta_2$ (we let $T_1
+happen-before an operation $\eta_2$ of $T_2$ only if every operation
+in $T_1$ (including its commit) happens-before $\eta_2$ (we let $T_1
 \hboar \eta_2$ denote this). The specification does not require $T_1$
@@
 \begin{smathpar}
-\begin{array}{l}
-  \neg\committed(\C{Wd1}) \conj (\neg\committed(\C{Wd2}) \Rightarrow
-  \C{C = k-a1}) 
-                \\
-      \C{Wd1} \wrstoar \C{C} \conj (\committed(\C{Wd2})
+\begin{array}{c}
+ \neg\committed(\C{Wd1}) \conj \C{Wd1} \wrstoar \C{C}  \conj 
+    (\neg\committed(\C{Wd2}) \Rightarrow \C{C = k-a1}) \\
+    \conj (\committed(\C{Wd2})
                 \Rightarrow \C{C = k-a1-a2})

### 2016-07-04 reasoning.tex

@@
 the \rulelabel{E-Aux} reduction rule of Fig.~\ref{fig:txnimp}. Similar
-premises also appear in \rulelabel{RG-Asgn} and \rulelabel{RG-Txn}
+premises also appear in the \rulelabel{RG-Asgn} and \rulelabel{RG-Txn}
 rules, which define rely-guarantee judgments for assignments and
@@
 hold.  The soundness of strengthening $\I$ ensures that a program can
-be executed safely on any data store that offers isolation guarantees
+be safely executed on any system that offers isolation guarantees
 at least as strong as those assumed.
@@
 
-The interleaved step relation for transaction bound expressions
+\noindent The interleaved step relation for transaction bound expressions
 ($\txnbox{e}_i$) and commands ($\txnbox{c}_i$) is defined similarly.

### 2016-07-04 related-work.tex

@@
 consistency~\cite{bailisvldb, alvarocalm, gotsmanpopl16,redblueatc,
-  redblueosdi, ecinec}. All of them assume a system model that offers
-a choice between a \emph{coordination-free} weak consistency level
+redblueosdi, ecinec}. All of them assume a system model that offers a
+choice between a \emph{coordination-free} weak consistency level
 (\emph{e.g.}, eventual consistency~\cite{redblueosdi, redblueatc,
 ecinec, alvarocalm, bailisvldb}) or causal
-consistency~\cite{LBC16,gotsmanpopl16}). The key technical issue in
+consistency~\cite{lbc16,gotsmanpopl16}). The key technical issue in
 these efforts involves proving that atomic and fully isolated

### 2016-07-04 store-consistency.tex

@@
 describes {\sc sc} stores with on-demand weak isolation, such as
-off-the-shelf relational databases. Similar approaches can be employed
+provided by off-the-shelf relational databases. Similar approaches can be employed
 to capture the semantics of other kinds of stores, such as a causally
@@
 operational semantics\footnote{Details of the {\sc cc} instantiation
-are relegated to the supplementary in the interest of space.}.
+are given in the supplementary material.}.
 
@@
 Fig.~\ref{fig:wd1-decorated}.  \C{Wd2} is similar and not
-shown\footnote{Supplementary contains fully decorated program.}.
+shown\footnote{The supplementary material shows the fully decorated program.}.
 Temporary name \C{v3} has been introduced to capture the result of the

### 2016-07-04 case-studies.tex
edits

@@
 this invariant entirely in user space by relying on database-backed
-transactions (this behavior is typical of Rails
+transactions (this behaviour is typical of Rails
 applications~\cite{bailisferal}).  For instance, deletion of a user is

### 2016-07-04 case-studies.tex
edits to bib and wording changes in Sec. 3 and 6

@@
 In this section, we present two case studies that demonstrate the
-applicability of our reasoning approach in varied settings, and its
-consequent practical utility. Both our case studies target
-Ruby-on-Rails applications. To analyze these, and other similar
-open-source applications, we developed a symbolic execution engine
-called {\sc MagLev} that compiles a Rails program to an abstract
-program in an extended version of \txnimp\footnote{The details of {\sc
-MagLev} can be found in the supplementary. Its source code, along with
-that of our applications and experiments, is available at the
-following anonymized url: \url{...}}. The focus of our formal
-verification efforts are such \txnimp programs.
+applicability of our framework in varied settings, and its consequent
+practical utility. Both our case studies target Ruby-on-Rails
+applications. To analyze these, and other similar open-source
+applications, we have developed a symbolic execution engine that
+compiles a Rails program to an abstract program in an extended version
+of \txnimp, along with some proof automation support.\footnote{Details
+  ( can be found in the supplementary material.  Source code for our
+  implementation, along with applications and experiments, are
+  available at the following anonymized url: \url{...}} The focus of
+our formal verification efforts are such translated \txnimp programs.
 
@@
 read-increment-write sequence wrapped in a transaction).  The
-application is expected to preserve the \emph{monotonicity invariant},
+application is expected to preserve a \emph{monotonicity invariant},
 namely that the view count should never appear to be decreasing.
@@
 \iso{Repeatable Read} is in fact sufficient for writers - an
-observation we confirmed through experiments. As usual, using {\sc
-ser} for all transactions also suffices, but on Postgres results in an
-additional 47\% increase in write latency.
+observation further validated through experiments. As usual, using
+{\sc ser} for all transactions also suffices, but on Postgres results
+in an additional 47\% increase in write latency.
 
@@
 Twitter-like microblogging application taken from~\cite{railsbook}.
-An important invariant that must be preserved is the referential
-integrity between microposts and authors. The application enforces
-this invariant entirely in user space by relying on database-backed
+An important invariant that must be preserved is referential integrity
+between microposts (tweets) and authors. The application enforces this
+invariant entirely in user space by relying on database-backed
 transactions (this behaviour is typical of Rails
-applications~\cite{bailisferal}).  For instance, deletion of a user is
-carried out in a transaction that also includes the deletion of
-microposts authored by the user.  However, it is possible for
+applications~\cite{bailisferal}).  For instance, the deletion of a
+user is carried out in a transaction that also includes the deletion
+of microposts authored by the user.  However, it is possible for
 concurrent observers to nonetheless witness invariant violations. To
-measure the likelihood of such anomalies, we first populated a
-Postgres database with 1000 user accounts, each user with 50
-microposts and 20 followers, chosen uniformly at random. We then
-constructed experiments with 64 concurrent clients performing
-deletions and timeline reads of randomly chosen users in a 1:7 ratio,
-and witnessed 250 violations of referential integrity (the isolation
-level was left at the default {\sc rc} level for experiments). To
-determine appropriate isolation levels, we then instantiated our
-framework for an {\sc sc} store with various ANSI SQL isolation levels
-for user deletions and timeline reads, and obtained a proof that {\sc
-rc} is sufficient for deletions, whereas reads require {\sc rr}. The
-proof is immediately applicable for MySQL, which implements ANSI SQL
-isolation levels. Repeating our experiments for MySQL with this
-configuration resulted in no violations (as expected), while yielding
-a 38\% reduction in latency than an {\sc ser} configuration.  Since
-Postgres's isolation levels are stronger than the corresponding ANSI
-levels, \rulelabel{RG-Conseq} rule (Fig.~\ref{fig:rg-rules}) lets us
-immediately deduce that the same choice of isolation levels also work
-on Postgres.
+quantify the likelihood of such anomalies, we first populated a
+Postgres database with 1000 user accounts, in which each user is
+associated with 50 microposts and 20 followers, chosen uniformly at
+random. We then constructed experiments with 64 concurrent clients
+performing deletions and timeline reads of randomly chosen users in a
+1:7 ratio, and witnessed 250 violations of referential integrity (the
+isolation level was left at the default {\sc rc} level for these
+experiments). To determine appropriate (correct) isolation levels, we
+then instantiated our framework for an {\sc sc} store with various
+ANSI SQL isolation levels for user deletions and timeline reads, and
+obtained a proof that {\sc rc} is sufficient for deletions, whereas
+reads require {\sc rr}. The proof is immediately applicable for MySQL,
+which implements ANSI SQL isolation levels. Repeating our experiments
+for MySQL with this configuration resulted in no violations (as
+expected), while yielding a 38\% reduction in latency than an {\sc
+  ser} configuration.  Since Postgres' isolation levels are stronger
+than the corresponding ANSI levels, \rulelabel{RG-Conseq} rule
+(Fig.~\ref{fig:rg-rules}) lets us immediately deduce that the same
+choice of isolation levels also work on Postgres.
 
-Clearly, there exists a significant diversity among the weak
-consistency and isolation variants implemented on various real-world
-data stores. More such variants have been developed in recent years to
-meet the need for high availability (\emph{e.g.}, \emph{Parallel
-  Snapshot Isolation}~\cite{psi} (2011) or \emph{Non-Monotonic
-  Snapshot Isolation}~\cite{nmsi} (2013)).  As more commercial
+Clearly, there exists significant diversity among the weak consistency
+and isolation variants implemented on various real-world data
+stores. More such variants have been developed in recent years to meet
+the need for high availability (\emph{e.g.}, \emph{Parallel Snapshot
+  Isolation}~\cite{psi} (2011) or \emph{Non-Monotonic Snapshot
+  Isolation}~\cite{nmsi} (2013)).  As more commercial
 weakly-consistent stores adopt various kinds of transactional
 semantics, like Cassandra's recently introduced \emph{lightweight
-  transactions}, it is reasonable to expect new isolation
-definitions to continue to be proposed.  While it may be possible to
-carefully engineer a reasoning framework for each combination of
-consistency and isolation, such a strategy would lead to multiple
-semantic definitions and proof systems with no obvious way to compare
-and relate them. As illustrated here, having a parameterized semantics
-and a proof framework built on top of it allows us to support these
-variants as distinct, yet comparable, instantiations.
+  transactions}, it is reasonable to expect new isolation definitions
+to continue to be proposed.  While it may be possible to carefully
+engineer a reasoning framework for each combination of consistency and
+isolation, such a strategy would lead to multiple semantic definitions
+and proof systems with no obvious way to compare and relate them. As
+illustrated here, having a parameterized semantics and a proof
+framework built on top of it allows us to support these variants as
+distinct, yet comparable, instantiations.

### 2016-07-04 motivation.tex
edits to bib and wording changes in Sec. 3 and 6

@@
 \C{Wd1} also happening before $\eta_2$. $\psi_{SI}$ however disallows
-all dashed arrows, because dashed arrows establish $\hbZ$ between
-actions in \C{Wd2} and only a subset of the operations in \C{Wd1}.
-Since the result of an operation (\eg a read) depends on what happens
-before that operation, $\hbZ$ relation ultimately affects the result
-of the program. The solid arrow and the first (from left) dashed arrow
-in Fig.~\ref{fig:motiv-eg-1-hb} denote $\hbZ$ relationships that do
-not affect the value of \C{C} in a way that causes the program to
-violate its post-condition. The last two dashed arrows denote $\hbZ$
+all dashed arrows, because dashed arrows establish $\hbZ$ edges
+between actions in \C{Wd2} and only a subset of the operations in
+\C{Wd1}.  Since the result of an operation (\eg a read) depends on
+what happens before that operation, the structure of the $\hbZ$
+relation ultimately dictates the final result of the program. The
+solid arrow and the first (from left) dashed arrow in
+Fig.~\ref{fig:motiv-eg-1-hb} denote $\hbZ$ relationships that do not
+affect the value of \C{C} in a way that causes the program to violate
+its post-condition. The last two dashed arrows denote $\hbZ$
 relationships that lead to the violation of the post-condition. The

### 2016-07-04 case-studies.tex
edits

@@
 of \txnimp, along with some proof automation support.\footnote{Details
-  ( can be found in the supplementary material.  Source code for our
+   can be found in the supplementary material.  Source code for our
   implementation, along with applications and experiments, are

### 2016-07-05 introduction.tex
typo fixes

@@
 operation. Furthermore, weak isolation can exist independent of weak
-consistency, as evident from the presence of weakly isolated
+consistency, as evident from the presence of weakly-isolated
 transactions on sequentially consistent relational databases; such
@@
 build correctness proofs for transactional programs in the presence of
-a weakly isolated concurrency control mechanism.  A key novelty of our
+a weakly-isolated concurrency control mechanism.  A key novelty of our
 approach is that it is parametric over the \emph{isolation semantics}
@@
     well-formedness constraints on a program's execution.
-  \item We propose a compositional proof system for this language
+  \item We present a compositional proof system for this language
     capable of relating high-level application invariants to the structure
@@
 
-\noindent Our results provide the first (to the best of our
-knowledge) mechanism that precisely relates high-level program
+\noindent Our results provide the first (to the best of our knowledge)
+mechanism that precisely and uniformly relates high-level program
 invariants to low-level weak isolation and store consistency

### 2016-07-05 introduction.tex
edits

@@
 transactions preserves its invariants when executed on a store
-equipped with certain consistency guarantees (e.g., causal
+equipped with appropriate consistency guarantees  (\emph{e.g.}, causal
 visibility).
@@
 The paper makes the following contributions:
+
 \begin{itemize}

### 2016-07-05 abstract.tex
KC's comments.

@@
 
-  Serializability is a well-understood mechanism that simplifies
-  reasoning about the behavior of concurrent transactions by ensuring
-  transactions are \emph{isolated} from each other's effects once they
-  begin executing.  However, enforcing serializability comes at a
-  steep cost in performance because it necessarily restricts
-  opportunities to exploit concurrency even when such opportunities
-  would not violate application-specific invariants.
-
-  As a result, database systems (in practice) support, and often
-  encourage, developers to use weaker alternatives.  These
-  alternatives break the strong isolation guarantees offered by
-  serializable transactions to permit greater concurrency.
-  Unfortunately, the semantics of weak isolation is poorly understood,
-  and usually explained only informally in terms of low-level
-  implementation artifacts.  Consequently, verifying high-level
-  correctness properties in such environments is problematic.
+  Serializability is a well-understood correctness criterion that
+  simplifies reasoning about the behavior of concurrent transactions
+  by ensuring that transactions are \emph{isolated} from each other.
+  However, enforcing serializable isolation comes at a steep cost in
+  performance because it necessarily restricts opportunities to
+  exploit concurrency even when such opportunities would not violate
+  application-specific invariants. As a result, database systems in
+  practice support, and often encourage, developers to use weaker
+  alternatives. These alternatives break the strong isolation
+  guarantees offered by serializable transactions to permit greater
+  concurrency. Unfortunately, the semantics of weak isolation is
+  poorly understood, and usually explained only informally in terms of
+  low-level implementation artifacts. Consequently, verifying
+  high-level correctness properties in such environments remains a
+  challenging problem.
 
@@
   compositional reasoning about the behavior of concurrently executing
-  weakly-isolated transactions.  Notably, our development is
-  parametric over a transaction's specific isolation semantics, and
-  the consistency guarantees provided by the underlying data store,
-  allowing it to be applicable over a large range of concurrency
-  control mechanisms.  Case studies and experiments on real-world
-  applications demonstrate the utility of our approach, and provide
-  strong evidence that weakly-isolated transactions can be placed on
-  the same formal footing as their strongly-isolated serializable
-  counterparts.
+  weakly-isolated transactions. Notably, our development is parametric
+  over a transaction's specific isolation semantics, and the
+  consistency guarantees provided by the underlying data store,
+  allowing it to be applicable over a range of concurrency control
+  mechanisms.  Case studies and experiments on real-world applications
+  demonstrate the utility of our approach, and provide strong evidence
+  that weakly-isolated transactions can be placed on the same formal
+  footing as their strongly-isolated serializable counterparts.
 

### 2016-07-05 introduction.tex
KC's comments.

--- 2016-07-04 introduction.tex
@@
 Database transactions allow users to group operations on multiple
-objects into a coherent unit, equipped with a set of four key
+objects into a single logical unit, equipped with a set of four key
 properties - atomicity, consistency, isolation, and durability (ACID).
@@
 
-The guarantees provided by serializability, and strong isolation in
-particular, do not come for free, however - pessimistic concurrency
-control methods that enforce serializability, for example, require
-databases to provide expensive techniques such as two-phase locking
-that incur overhead to deal with deadlocks, rollbacks, and
-re-execution~\cite{twopl,ullmanbook}.  Similar criticisms apply to
-optimistic multi-version concurrency control methods that must deal
-with timestamp and version management~\cite{BG81}.  Regardless of the
-particular concurrency control method employed, enforcing
-serializability in a replicated data store additionally requires
-global coordination among geo-distributed replicas, which makes the
-store \emph{unavailable} in the presence of network
+The guarantees provided by serializability do not come for free,
+however - pessimistic concurrency control methods that enforce
+serializability, for example, require databases to provide expensive
+techniques such as two-phase locking that incur overhead to deal with
+deadlocks, rollbacks, and re-execution~\cite{twopl,ullmanbook}.
+Similar criticisms apply to optimistic multi-version concurrency
+control methods that must deal with timestamp and version
+management~\cite{BG81}.  Regardless of the particular concurrency
+control method employed, enforcing serializability in a replicated
+data store additionally requires global coordination among
+geo-distributed replicas, which makes the store \emph{unavailable} in
+the presence of network
 partitions~\cite{cap,sernotavlbl,bailishat,bernsigmod13}.
@@
 make clear that weakly-isolated transactions are quite prevalent in
-practice, and strongly-isolated (serializable) transactions are often
-eschewed altogether.
+practice, whereas serializable transactions are often eschewed.
 
@@
 redblueatc, ecinec, gotsmanpopl16} address broadly related concerns.
-At a high-level, reasoning about weak consistency involves reasoning
-about the concurrent execution of strongly atomic and fully isolated
-operations against different (potentially replicated), but consistent
-states of the database or data store, where each operation locally
-preserves application invariants. In contrast, reasoning about weak
-isolation involves reasoning about the concurrent execution of
-transactions, whose operations witness different states which,
-although consistent individually, may not be obviously reconciled into
-a consistent global view.  Intuitively, the latter problem reduces to
-the former if all transactions contain a single
-operation. Furthermore, weak isolation can exist independent of weak
-consistency, as evident from the presence of weakly isolated
-transactions on sequentially consistent relational databases; such
-systems existed long before the advent of weakly consistent NoSQL
-stores.  Thus, frameworks for reasoning about weak isolation will
-necessarily have to generalize the reasoning frameworks developed for
-weak consistency in new and important ways.
+Weak consistency is a phenomenon that usually manifests on replicated
+data stores, where operations are concurrently executed against
+different replicas, resulting in an order of execution inconsistent
+with their serial order. The operations, nonetheless, are atomic and
+fully isolated, and each operation is required to locally preserve
+application invariants. In contrast, weak isolation is a property of
+transactions, which are sets of atomic operations. Weak isolation
+manifests when successive atomic operations in a transaction witness
+different contemporary states (or different replicas) of the database
+which, although consistent individually, may not be obviously
+reconciled into a consistent global view. Furthermore, operations may
+temporarily break invariants as long as transaction as a whole
+preserves them.  Intuitively, the latter problem reduces to the former
+if all transactions contain a single operation. Furthermore, weak
+isolation can exist independent of weak consistency, as evident from
+the presence of weakly isolated transactions on sequentially
+consistent relational databases; such systems existed long before the
+advent of weakly consistent NoSQL stores.  Thus, frameworks for
+reasoning about weak isolation will necessarily have to generalize the
+reasoning frameworks developed for weak consistency in new and
+important ways.
 
@@
 semantics} of the underlying store. In concrete terms, this means
-that, unlike recent work focused on formalizing the semantics of weak
-consistency~\cite{gotsmanpopl16, redblueatc, ecinec}, our system model
-does not assume a minimum or predefined set of consistency or
+that, unlike recent work focused on reasoning about programs under
+weak consistency~\cite{gotsmanpopl16, redblueatc, ecinec}, our system
+model does not assume a minimum or predefined set of consistency or
 isolation levels. Instead, our operational semantics admits
 declarative specifications of transaction isolation and store
-consistency levels (\eg~\cite{pldi15,gotsmanconcur15}), and generates
-executions that are guaranteed to satisfy these specifications. The
-result is a flexible system model that is general enough to
-incorporate the semantics of a range of isolation levels on a variety
-of stores (\eg the sequentially consistent store of~\cite{adyaphd}, or
-the causally consistent store of~\cite{gotsmanpopl16}, etc).  Our key
-technical contribution is thus a set of proof rules that demonstrate
-that a program with a given selection of isolation levels for its
-transactions preserves its invariants when executed on a store
-equipped with certain consistency guarantees (e.g., causal
-visibility).
+consistency, and generates executions that are guaranteed to satisfy
+these specifications. The result is a flexible system model that is
+general enough to incorporate the semantics of a range of isolation
+levels on a variety of stores (\eg the sequentially consistent store
+of~\cite{adyaphd}, or the causally consistent store
+of~\cite{gotsmanpopl16}, etc).  Our key technical contribution is thus
+a set of proof rules that demonstrate that a program with a given
+selection of isolation levels for its transactions preserves its
+invariants when executed on a store equipped with certain consistency
+guarantees (\eg causal consistency).
 

### 2016-07-05 introduction.tex

@@
 Recent results on reasoning about application invariants in the
-presence of weak \emph{consistency}~\cite{burckhardt14, redblueosdi,
+presence of weak consistency~\cite{burckhardt14, redblueosdi,
 redblueatc, ecinec, gotsmanpopl16} address broadly related concerns.
@@
 isolation can exist independent of weak consistency, as evident from
-the presence of weakly isolated transactions on sequentially
+the presence of weakly-isolated transactions on sequentially
 consistent relational databases; such systems existed long before the

### 2016-07-05 opsem.tex

@@
 with \C{Wd1} and \C{Wd2}, respectively. The instantiated specification
-is the following term:
+is the following term:\vspace*{-8pt}
 
@@
 $\hbZ \,=\, (\soZ \cup \visZ)^{+}$. Unlike $\hbZ$, $\visZ$ (defined
-below) is not a transitive relation, hence lets us capture finer
-properties than $\hbZ$. We therefore let $\visZ$ replace $\hbZ$ in our
-discussion henceforth. The session order relation captures the
-sequential order of operations within a transaction. In particular, it
-relates two effects, $\eta_1$ and $\eta_2$, such that $\txn(\eta_1) =
+below) is not a transitive relation, and hence lets us capture
+finer-grained isolation properties than $\hbZ$, which we leverage in
+our development.  The session order relation captures the sequential
+order of operations within a transaction. In particular, it relates
+two effects, $\eta_1$ and $\eta_2$, such that $\txn(\eta_1) =
 \txn(\eta_2)$ and $\id(\eta_1) < \id(\eta_2)$.  The semantics assigns

### 2016-07-05 introduction.tex
C --> B. KC's suggestions incorporated.

@@
 manifests when successive atomic operations in a transaction witness
-different contemporary states (or different replicas) of the database
-which, although consistent individually, may not be obviously
-reconciled into a consistent global view. Furthermore, operations may
-temporarily break invariants as long as transaction as a whole
-preserves them.  Intuitively, the latter problem reduces to the former
-if all transactions contain a single operation. Furthermore, weak
-isolation can exist independent of weak consistency, as evident from
-the presence of weakly-isolated transactions on sequentially
-consistent relational databases; such systems existed long before the
-advent of weakly consistent NoSQL stores.  Thus, frameworks for
-reasoning about weak isolation will necessarily have to generalize the
-reasoning frameworks developed for weak consistency in new and
-important ways.
+different contemporary states of the database (or different replicas
+of the replicated store) which, although consistent individually, may
+not be obviously reconciled into a consistent global view.
+Intuitively, the latter problem reduces to the former if all
+transactions contain a single operation. Furthermore, weak isolation
+can exist independent of weak consistency, as evident from the
+presence of weakly-isolated transactions on conventional RDBS. With an
+added complexity of replication, richer mechanisms are needed to
+reason about weak isolation in tandem with weak consistency. Thus,
+frameworks for reasoning about weak isolation will necessarily have to
+generalize the reasoning frameworks developed for weak consistency in
+new and important ways.
 
@@
 mechanism that precisely and uniformly relates high-level program
-invariants to low-level weak isolation and store consistency
+invariants to low-level weak isolation and weak consistency
 guarantees, thereby allowing weakly-isolated transactions to enjoy the

### 2016-07-05 motivation.tex
C --> B. KC's suggestions incorporated.

@@
 \txnimp - a C-like imperative language equipped with a \C{txn} lexical
-block defining a transaction scope.  For exposition convenience, each
-\C{txn} block is associated with a single-quoted string in angle
+block defining a transaction scope.  For expositional convenience,
+each \C{txn} block is associated with a single-quoted string in angle
 braces that uniquely identifies the transaction. We use Hoare triple
@@
 \centering
-$\{\texttt{C}=\texttt{k} \conj \texttt{k}\ge\texttt{a1+a2}\}$
+$\{\texttt{B}=\texttt{k} \conj \texttt{k}\ge\texttt{a1+a2}\}$
 \begin{tabular}{l||l}
@@
   txn$\langle$'Wd1'$\rangle${
-    if (C $\ge$ a1) {
-      C := C - a1
+    if (B $\ge$ a1) {
+      B := B - a1
     }
@@
   txn$\langle$'Wd2'$\rangle${
-    if (C $\ge$ a2) {
-      C := C - a2
+    if (B $\ge$ a2) {
+      B := B - a2
     }
@@
 \end{tabular}
-$\{\texttt{C}=\texttt{k-a1-a2}\}$
+$\{\texttt{B}=\texttt{k-a1-a2}\}$
 \caption{\small Concurrent withdraw transactions}
@@
 Consider an implementation of a banking application that admits
-concurrent withdraw transactions on a checking account (\C{C}), as
+concurrent withdraw transactions on an account balance (\C{B}), as
 shown in Fig.~\ref{fig:motiv-eg-1}. If the initial balance (\C{k}) in
@@
 
-As an alternative, consider the execution of this transaction under
-a \emph{Read Committed} ({\sc rc}) isolation level, which is weaker than
+As an alternative, consider the execution of this transaction under a
+\emph{Read Committed} ({\sc rc}) isolation level, which is weaker than
 {\sc ser}.\footnote{{\sc rc} is in fact the default isolation level in
@@
 executions shown in Fig.~\ref{fig:rc-ex} on a strongly consistent
-({\sc sc}) store, such as a relational database.
+({\sc sc}) store, such as a conventional RDBMS.
 
@@
 operations. The subscript of an operation indicates the transaction
-that executes it. In the first execution, transaction \C{Wd1} reads
-the current balance (\C{k}) and writes the new balance (\C{k-a1}), but
+that executes it (Both transactions read \C{B} twice, hence two read
+operations). In the execution on left, transaction \C{Wd1} reads the
+current balance (\C{k}) and writes the new balance (\C{k-a1}), but
 before it commits, transaction \C{Wd2} executes and commits, writing
@@
 (the so-called \emph{lost update anomaly}~\cite{berenson}), resulting
-in an incorrect balance of \C{k-a1}. The second execution describes a
-similar scenario with \C{Wd1} and \C{Wd2} exchanging their roles.
+in an incorrect balance of \C{k-a1}. The execution on right describes
+a similar scenario with \C{Wd1} and \C{Wd2} exchanging their roles.
 Clearly, {\sc rc} is an excessively weak isolation level for this
@@
 transactions in the absence of write-write conflicts.} Since {\sc si},
-unlike {\sc ser}, does not rely on expensive lock-based concurrency
-control, it is also more efficient, making it appropriate for both
-\C{Wd} transactions.
+unlike {\sc ser}, does not need expensive mechanisms such as
+lock-based concurrency control or runtime monitoring, it is also more
+efficient, making it appropriate for both withdraw transactions.
 
@@
 \begin{array}{lcl}
-\psi_{RC} & \defeq & \forall T_1,T_2,\eta_1,\eta_2,.\; \txn(\eta_1) = T_1 
+\psi_{RC} & \defeq & \forall T_1,T_2,\eta_1,\eta_2.\; \txn(\eta_1) = T_1 
   \conj \txn(\eta_2) = T_2 \\
@@
 ($\psi_{SI}$) requires two transactions that write to the same
-variable to be related by happens-before.  This effectively prohibits
-interference due to actions in $T_1$ being interleaved in $T_2$ (or,
-vise versa).
+variable (denoted $\wrstoar$)to be related by happens-before.  This
+effectively prohibits interference due to actions in $T_1$ being
+interleaved in $T_2$ (or, vise versa).
 
@@
 Fig.~\ref{fig:motiv-eg-1-hb} denote $\hbZ$ relationships that do not
-affect the value of \C{C} in a way that causes the program to violate
+affect the value of \C{B} in a way that causes the program to violate
 its post-condition. The last two dashed arrows denote $\hbZ$
@@
 prohibited. For instance, consider the program point after the write
-to \C{C} in \C{Wd1}. The expected invariant ($\phi$) at that program
+to \C{B} in \C{Wd1}. The expected invariant ($\phi$) at that program
 point is shown below ($\committed$ stands for ``committed''):
@@
 \begin{array}{c}
- \neg\committed(\C{Wd1}) \conj \C{Wd1} \wrstoar \C{C}  \conj 
-    (\neg\committed(\C{Wd2}) \Rightarrow \C{C = k-a1}) \\
+ \neg\committed(\C{Wd1}) \conj \C{Wd1} \wrstoar \C{B}  \conj 
+    (\neg\committed(\C{Wd2}) \Rightarrow \C{B = k-a1}) \\
     \conj (\committed(\C{Wd2})
-                \Rightarrow \C{C = k-a1-a2})
+                \Rightarrow \C{B = k-a1-a2})
 \end{array}
@@
 \noindent $\phi$ asserts that \C{Wd1} is not yet committed, and that
-it wrote to \C{C}, and the value of \C{C} is either \C{k-a1-a2} or
+it wrote to \C{B}, and the value of \C{B} is either \C{k-a1-a2} or
 \C{k-a1} depending on whether or not \C{Wd2} is committed. If $\phi$
-remains invariant until \C{Wd1} commits, then the post-condition (\C{C
+remains invariant until \C{Wd1} commits, then the post-condition (\C{B
 = k-a1-a2}) can be established easily. However, an interference from
@@
 Fig.~\ref{fig:motiv-eg-1-hb}) may violate the invariant by writing
-$\C{k-a1}$ to \C{C} and committing \C{Wd2}, thus leading to
-$\committed(\C{Wd2}) \Rightarrow \C{C=k-a2}$. Fortunately,
+$\C{k-a1}$ to \C{B} and committing \C{Wd2}, thus leading to
+$\committed(\C{Wd2}) \Rightarrow \C{B=k-a2}$. Fortunately,
 \iso{Snapshot Isolation} prevents this interference, and this can be
@@
 execution state where neither $\C{Wd2} \hboar \C{Wd1}$ nor $\C{Wd1}
-\hboar \C{Wd2}$ holds ($\C{Wd2} \hboar \C{Wd1}$ does not hold because
-\C{Wd1}'s write to \C{C} clearly happened before \C{Wd2}'s commit, and
+\hboar \C{Wd2}$ holds; $\C{Wd2} \hboar \C{Wd1}$ does not hold because
+\C{Wd1}'s write to \C{B} clearly happened before \C{Wd2}'s commit, and
 $\C{Wd1} \hboar \C{Wd2}$ does not hold because \C{Wd1} has not yet
-committed, while \C{Wd2} has already begun). Since \C{Wd2} also writes
-to \C{C}, this violates the $\psi_{SI}$ constraint which we assume to
+committed, while \C{Wd2} has already begun. Since \C{Wd2} also writes
+to \C{B}, this violates the $\psi_{SI}$ constraint which we assume to
 be an invariant. A proof that the post-condition holds now follows
@@
 tailor-made for such stores. However, for the reasoning framework to
-be truly useful, it has to be parametric over different store
-consistency semantics, supporting stores that are weaker than {\sc sc}
-(e.g., causally or eventually consistent), and should be able to
-reconcile conflicts between consistency and isolation constraints.  We
+be truly useful, it has to be parametric over different consistency
+semantics, supporting stores that are weaker than {\sc sc} (e.g.,
+causally or eventually consistent), and should be able to reconcile
+conflicts between consistency and isolation constraints.  We
 demonstrate how our reasoning framework makes this possible in the

### 2016-07-05 reasoning.tex
C --> B. KC's suggestions incorporated.

@@
 similarly. For instance, consider the post-condition
-$\C{C\,=\,k-a1-a2}$ of the program in Fig.~\ref{fig:motiv-eg-1}. The
+$\C{B\,=\,k-a1-a2}$ of the program in Fig.~\ref{fig:motiv-eg-1}. The
 corresponding assertion on the post-state
-($\lambda\sigma.~\sigma(\C{C}) \,=\, \C{k-a1-a2}$) asserts that in all
-states resulting from executing the program, the value of \C{C} is
+($\lambda\sigma.~\sigma(\C{B}) \,=\, \C{k-a1-a2}$) asserts that in all
+states resulting from executing the program, the value of \C{B} is
 \C{k-a1-a2}. The equivalent execution-based assertion
-($\lambda\E.~\interp{\E}(\C{C})\,=\,\C{k-a1-a2}$) asserts that in all
-executions of the program, the value of \C{C} is \C{k-a1-a2}.
+($\lambda\E.~\interp{\E}(\C{B})\,=\,\C{k-a1-a2}$) asserts that in all
+executions of the program, the value of \C{B} is \C{k-a1-a2}.
 However, having access to an execution facilitates assertions that go
@@
   \lambda\E.~ \neg(\underE{\committed(\C{Wd1}}) \conj 
-      \underE{\C{Wd1} \wrstoar \C{C}}\\
+      \underE{\C{Wd1} \wrstoar \C{B}}\\
   \hspace*{0.2in}\conj(\neg(\underE{\committed(\C{Wd2})}) \Rightarrow 
-      \C{\interp{\E}(C) = k-a1}) \\
+      \C{\interp{\E}(B) = k-a1}) \\
    \hspace*{0.2in}\conj (\underE{\committed(\C{Wd2})} \Rightarrow 
-      \C{\interp{\E}(C) = k-a1-a2})
+      \C{\interp{\E}(B) = k-a1-a2})
 \end{array}
@@
 
-\noindent One of the conjuncts asserts that \C{Wd1} has written to \C{C} - a
-fact which cannot be deduced solely from the value of \C{C} (esp. in the
+\noindent One of the conjuncts asserts that \C{Wd1} has written to \C{B} - a
+fact which cannot be deduced solely from the value of \C{B} (esp. in the
 presence of interference), but can be expressed as a proposition over

### 2016-07-05 store-consistency.tex
C --> B. KC's suggestions incorporated.

@@
   \hspace*{0.3in}P : \{\neg\committed(\C{Wd1}) \conj \neg\committed(\C{Wd2}) \conj
-                          \C{C} = \C{k}\}
+                          \C{B} = \C{k}\}
 \end{decoration}$
@@
  P_1:\{ \neg\committed(\C{Wd1}) \conj
-        (\neg\committed(\C{Wd2}) \Rightarrow \C{C = k}) \conj\\
+        (\neg\committed(\C{Wd2}) \Rightarrow \C{B = k}) \conj\\
         \hspace*{0.3in}{\committed(\C{Wd2})} \Rightarrow 
-            \C{Wd2} \wrstoar \C{C} \wedge \C{C = k-a2} \}
+            \C{Wd2} \wrstoar \C{B} \wedge \C{B = k-a2} \}
  \end{decoration}$
@@
      $\begin{decoration}
-     \phi_4 : \{{\neg\committed(\C{Wd2})} \Rightarrow \C{C = k} \wedge \C{v3 = k-a1} \conj\\
+     \phi_4 : \{{\neg\committed(\C{Wd2})} \Rightarrow \C{B = k} \wedge \C{v3 = k-a1} \conj\\
         \hspace*{0.3in}({\committed(\C{Wd2})} \Rightarrow \C{Wd2}
-        \wrstoar \C{C}) \conj \neg\committed(\C{Wd1}) \conj \\
+        \wrstoar \C{B}) \conj \neg\committed(\C{Wd1}) \conj \\
         \hspace*{0.32in}{\committed(\C{Wd2})} \wedge
          {\C{Wd2} \visar \C{Wd1}} 
-        \Rightarrow \C{C = k-a2} \wedge \\
+        \Rightarrow \C{B = k-a2} \wedge \\
         \hspace*{1.9in}\C{v3 = k-a2-a1}\}
      \end{decoration}$ 
-     C := v3
+     B := v3
      $\begin{decoration}
-      \phi_5 : \{({\neg\committed(\C{Wd2})} \Rightarrow \C{C = k-a1}) 
+      \phi_5 : \{({\neg\committed(\C{Wd2})} \Rightarrow \C{B = k-a1}) 
           \conj \neg\committed(\C{Wd1}) \conj\\
          \hspace*{0.4in}{\committed(\C{Wd2})} 
-            \Rightarrow \C{Wd2} \wrstoar \C{C} \wedge \C{C = k-a1-a2}\}
+            \Rightarrow \C{Wd2} \wrstoar \C{B} \wedge \C{B = k-a1-a2}\}
       \end{decoration}$ 
@@
  $\begin{decoration}
-  Q_1 : \{{(\neg\committed(\C{Wd2})} \Rightarrow \C{C = k-a1})
+  Q_1 : \{{(\neg\committed(\C{Wd2})} \Rightarrow \C{B = k-a1})
             \conj \committed(\C{Wd1}) \conj\\
       \hspace*{0.4in}{\committed(\C{Wd2})} 
-          \Rightarrow \C{C = k-a1-a2} \}
+          \Rightarrow \C{B = k-a1-a2} \}
   \end{decoration}$
@@
   \hspace*{0.3in}Q : \{\committed(\C{Wd1}) \conj \committed(\C{Wd2}) \conj
-                          \C{C} = \C{k-a1-a2}\}
+                          \C{B} = \C{k-a1-a2}\}
 \end{decoration}$
@@
 $P_1$ of \C{Wd1} accounts for the possibility of \C{Wd2} committing
-before \C{Wd1}, writing \C{k-a2} to \C{C}. Since neither \C{Wd1} nor
+before \C{Wd1}, writing \C{k-a2} to \C{B}. Since neither \C{Wd1} nor
 \C{Wd2} are committed at the beginning, the pre-condition $P$ of the
@@
 \neg\committed(\C{Wd2})$, from which $P_1$ follows. The post-condition
-$Q_1$ of \C{Wd1} asserts different values for \C{C} depending on
+$Q_1$ of \C{Wd1} asserts different values for \C{B} depending on
 whether or not \C{Wd2} has already committed.  It also asserts that
@@
 the assertions that decorate \C{Wd1} are indeed valid. We consider the
-triple associated with the assignment to \C{C} for illustration. The
-pre-condition $\phi_4$ asserts different values for \C{C} and \C{v3}
+triple associated with the assignment to \C{B} for illustration. The
+pre-condition $\phi_4$ asserts different values for \C{B} and \C{v3}
 (temporary alias of RHS expression) depending on whether or not
@@
 this is sufficient to prove the post-condition $\phi_5$, which appears
-to make a stronger claim about the value of \C{C}. The proof follows
+to make a stronger claim about the value of \C{B}. The proof follows
 from the RG rule \rulelabel{RG-Asgn} for assignments, which allows us
@@
 \begin{array}{l}
-    {\neg\committed(\C{Wd1}) \conj \C{Wd1} \wrstoar \C{C} 
-    \conj (\neg\committed(\C{Wd2})} \Rightarrow \C{C = k-a1}) \conj\\
+    {\neg\committed(\C{Wd1}) \conj \C{Wd1} \wrstoar \C{B} 
+    \conj (\neg\committed(\C{Wd2})} \Rightarrow \C{B = k-a1}) \conj\\
     ({\committed(\C{Wd2})} \Rightarrow \C{Wd2} \wrstoar
-    \C{C}) \conj \\
+    \C{B}) \conj \\
     ({\committed(\C{Wd2})} \wedge {\C{Wd2} \visar \C{Wd1}} 
-    \Rightarrow \C{C = k-a2} \wedge \C{C = k-a2-a1})
+    \Rightarrow \C{B = k-a2} \wedge \C{B = k-a2-a1})
 \end{array}
@@
 invariant ($\I$), which asserts {\sc si} for both transactions.  Since
-both transactions now write to \C{C}, {\sc si} requires either
+both transactions now write to \C{B}, {\sc si} requires either
 $\C{Wd1} \visar \C{Wd2}$, or $\C{Wd2} \visar \C{Wd1}$. Since \C{Wd1}
@@
 deduce $\C{Wd2} \visar \C{Wd1}$. This lets us derive
-${\committed(\C{Wd2})}  \Rightarrow \C{C = k-a2} \wedge \C{C =
+${\committed(\C{Wd2})}  \Rightarrow \C{B = k-a2} \wedge \C{B =
 k-a2-a1}$, allowing us to prove the post-condition $\phi_5$.
@@
   \item  If \C{Wd2} commits, then it should have already written to
-  \C{C}. This fact is needed to show that \C{Wd2} conflicts with
+  \C{B}. This fact is needed to show that \C{Wd2} conflicts with
   \C{Wd1}, so that \C{SI(Wd1)} property can be exploited to deduce
@@
 
-  \item The value written to \C{C} is either $\C{k-a1-a2}$ or
+  \item The value written to \C{B} is either $\C{k-a1-a2}$ or
   $\C{k-a1}$ depending on whether or not \C{Wd1} has already
-  committed. This is clearly necessary to deduce the value of \C{C}
+  committed. This is clearly necessary to deduce the value of \C{B}
   before and after \C{Wd1} commits.
@@
 %   & & \hspace*{0.5in} \E' \Vdash \committed(\C{Wd2}) ~\Rightarrow~ \E'
-%       \Vdash \C{Wd2} \wrstoar \C{C} \conj \\ 
+%       \Vdash \C{Wd2} \wrstoar \C{B} \conj \\ 
 %   & & \hspace*{0.5in} \underE{\committed(\C{Wd1})} \conj \E' \Vdash
-%   \committed(\C{Wd2}) \Rightarrow \C{C=k-a1-a2} \\
+%   \committed(\C{Wd2}) \Rightarrow \C{B=k-a1-a2} \\
 %   & & \hspace*{0.5in} \neg\underE{\committed(\C{Wd1})} \conj
 %       (\E'-\E) \Vdash {\committed(\C{Wd2})}\\
-    & & \hspace*{0.8in}\Rightarrow \C{C=k-a2} \conj 
-        \E' \Vdash \C{Wd2} \wrstoar \C{C} \\
+    & & \hspace*{0.8in}\Rightarrow \C{B=k-a2} \conj 
+        \E' \Vdash \C{Wd2} \wrstoar \C{B} \\
     & & \hspace*{0.37in} \conj \underE{\committed(\C{Wd1})} \conj
@@
 %       \C{COMMIT(Wd2)} \in (\E'.\A-\E.\A) \\
-    & & \hspace*{0.6in}\Rightarrow \C{C=k-a1-a2} \conj
-        \E' \Vdash \C{Wd2} \wrstoar \C{C} \\
+    & & \hspace*{0.6in}\Rightarrow \C{B=k-a1-a2} \conj
+        \E' \Vdash \C{Wd2} \wrstoar \C{B} \\
     & & \hspace*{0.37in}\conj \forall(\eta \in \E'.\A - 
@@
 The stability for $\phi_4$ w.r.t. $R_1$ is trivial to establish: if
-$R_1$ does not commit \C{Wd2}, then the value of \C{C} is unaffected,
+$R_1$ does not commit \C{Wd2}, then the value of \C{B} is unaffected,
 and if it commits \C{Wd2}, then $\C{Wd2}\not\visar\C{Wd1}$, hence

### 2016-07-05 introduction.tex
tightened weak consistency discussion in intro

@@
 
+%%%SJ: tried to tighten this paragraph; the original now commented seemed overly
+%%complicated
 Recent results on reasoning about application invariants in the
@@
   redblueatc, ecinec, gotsmanpopl16} address broadly related concerns.
-Weak consistency is a phenomenon that usually manifests on replicated
-data stores, where operations are concurrently executed against
-different replicas, resulting in an order of execution inconsistent
-with their serial order. The operations, nonetheless, are atomic and
-fully isolated, and each operation is required to locally preserve
-application invariants. In contrast, weak isolation is a property of
-transactions, which are sets of atomic operations. Weak isolation
-manifests when successive atomic operations in a transaction witness
-different contemporary states of the database (or different replicas
-of the replicated store) which, although consistent individually, may
-not be obviously reconciled into a consistent global view.
-Intuitively, the latter problem reduces to the former if all
-transactions contain a single operation. Furthermore, weak isolation
-can exist independent of weak consistency, as evident from the
-presence of weakly-isolated transactions on conventional RDBS. With an
+Weak consistency is a phenomenon that manifests on replicated data
+stores, where operations are concurrently executed against different
+replicas, resulting in an execution order inconsistent with any serial
+(sequentially consistent) ordering.  In contrast, weak isolation is a
+property of atomic transactions whose effects interfere with one
+another in ways inconsistent with serializability.  Unlike weak
+consistency, weak isolation can manifest even in an unreplicated
+setting, as evident from the support for weakly-isolated transactions
+on conventional (unreplicated) databases as mentioned above.  With the
 added complexity of replication, richer mechanisms are needed to
@@
 new and important ways.
+
+%% Recent results on reasoning about application invariants in the
+%% presence of weak consistency~\cite{burckhardt14, redblueosdi,
+%% redblueatc, ecinec, gotsmanpopl16} address broadly related concerns.
+%% Weak consistency is a phenomenon that usually manifests on replicated
+%% data stores, where operations are concurrently executed against
+%% different replicas, resulting in an order of execution inconsistent
+%% with their serial order. The operations, nonetheless, are atomic and
+%% fully isolated, and each operation is required to locally preserve
+%% application invariants. In contrast, weak isolation is a property of
+%% transactions, which are sets of atomic operations. Weak isolation
+%% manifests when successive atomic operations in a transaction witness
+%% different contemporary states of the database (or different replicas
+%% of the replicated store) which, although consistent individually, may
+%% not be obviously reconciled into a consistent global view.
+%% Intuitively, the latter problem reduces to the former if all
+%% transactions contain a single operation. Furthermore, weak isolation
+%% can exist independent of weak consistency, as evident from the
+%% presence of weakly-isolated transactions on conventional RDBS. With an
+%% added complexity of replication, richer mechanisms are needed to
+%% reason about weak isolation in tandem with weak consistency. Thus,
+%% frameworks for reasoning about weak isolation will necessarily have to
+%% generalize the reasoning frameworks developed for weak consistency in
+%% new and important ways.
 

### 2016-07-05 store-consistency.tex
tightened weak consistency discussion in intro

@@
 pre-condition $\phi_4$ asserts different values for \C{B} and \C{v3}
-(temporary alias of RHS expression) depending on whether or not
-\C{Wd2} is committed. Inside \C{Wd1}, $\committed(\C{Wd2})$ may mean
-that either \C{Wd2} happened before \C{Wd1} (hence can be visible to
-all of \C{Wd1}), or that it committed concurrently with \C{Wd1} (hence
-cannot be visible to all of \C{Wd2}).  Nonetheless, $\phi_4$ only
-considers the case when $\C{Wd2} \visar \C{Wd1}$. As we show below,
-this is sufficient to prove the post-condition $\phi_5$, which appears
-to make a stronger claim about the value of \C{B}. The proof follows
-from the RG rule \rulelabel{RG-Asgn} for assignments, which allows us
-to easily conclude the following about the execution state after the
+depending on whether or not \C{Wd2} is committed. Inside \C{Wd1},
+$\committed(\C{Wd2})$ may mean that either \C{Wd2} happened before
+\C{Wd1} (hence can be visible to all of \C{Wd1}), or that it committed
+concurrently with \C{Wd1} (hence cannot be visible to all of \C{Wd2}).
+Nonetheless, $\phi_4$ only considers the case when $\C{Wd2} \visar
+\C{Wd1}$. As we show below, this is sufficient to prove the
+post-condition $\phi_5$, which appears to make a stronger claim about
+the value of \C{B}. The proof follows from the RG rule
+\rulelabel{RG-Asgn} for assignments, which allows us to easily
+conclude the following about the execution state after the
 assignment:\vspace*{-12pt}
@@
 The second part of the proof is to show that assertions are stable
-despite the interference from the the concurrent thread executing
-\C{Wd2}. We formalize the interference as a rely relation that
-describes the semantics of \C{Wd2} at just the right level of
+despite the interference from the concurrent thread executing
+\C{Wd2}. We formalize such interference as a rely relation that
+describes the semantics of \C{Wd2} at the right level of
 abstraction. In particular, we would like to capture two important

### 2016-07-05 introduction.tex
edits to abstract and intro

@@
 setting, as evident from the support for weakly-isolated transactions
-on conventional (unreplicated) databases as mentioned above.  With the
-added complexity of replication, richer mechanisms are needed to
-reason about weak isolation in tandem with weak consistency. Thus,
-frameworks for reasoning about weak isolation will necessarily have to
-generalize the reasoning frameworks developed for weak consistency in
-new and important ways.
+on conventional (unreplicated) databases as mentioned above.  In the
+presence of replication, however, the interaction between weak
+isolation and weak consistency can be subtle and non-trivial.
+Understanding weak isolation in these varied contexts thus requires
+new insights and substantial generalization of existing techniques.
 

### 2016-07-05 introduction.tex
Slight rewording of consistency-isolation para.

--- 2016-07-05 introduction.tex
@@
 Weak consistency is a phenomenon that manifests on replicated data
-stores, where operations are concurrently executed against different
-replicas, resulting in an execution order inconsistent with any serial
-(sequentially consistent) ordering.  In contrast, weak isolation is a
-property of atomic transactions whose effects interfere with one
-another in ways inconsistent with serializability.  Unlike weak
-consistency, weak isolation can manifest even in an unreplicated
-setting, as evident from the support for weakly-isolated transactions
-on conventional (unreplicated) databases as mentioned above.  With the
+stores, where atomic operations are concurrently executed against
+different replicas, resulting in an execution order inconsistent with
+%% We already say "inconsistent with.. ". 
+any sequential order. In contrast, weak isolation is a property of
+concurrent transactions interfering with one another resulting in an
+execution order that is not serializable. Unlike weak consistency,
+weak isolation can manifest even in an unreplicated setting, as
+evident from the support for weakly-isolated transactions on
+conventional (unreplicated) databases as mentioned above.  With the
 added complexity of replication, richer mechanisms are needed to
@@
 general enough to incorporate the semantics of a range of isolation
-levels on a variety of stores (\eg the sequentially consistent store
+levels on a variety of stores (\eg~the sequentially consistent store
 of~\cite{adyaphd}, or the causally consistent store
@@
 invariants when executed on a store equipped with certain consistency
-guarantees (\eg causal consistency).
+guarantees (\eg~causal consistency).
 

### 2016-07-05 abstract.tex

@@
   simplifies reasoning about the behavior of concurrent transactions
-  by ensuring that transactions are \emph{isolated} from each other.
-  However, enforcing serializable isolation comes at a steep cost in
-  performance because it necessarily restricts opportunities to
-  exploit concurrency even when such opportunities would not violate
-  application-specific invariants. As a result, database systems in
-  practice support, and often encourage, developers to use weaker
-  alternatives. These alternatives break the strong isolation
-  guarantees offered by serializable transactions to permit greater
-  concurrency. Unfortunately, the semantics of weak isolation is
-  poorly understood, and usually explained only informally in terms of
-  low-level implementation artifacts. Consequently, verifying
-  high-level correctness properties in such environments remains a
-  challenging problem.
+  by ensuring that transactions are \emph{isolated} from each other
+  while they execute.  However, enforcing serializable isolation comes
+  at a steep cost in performance because it necessarily restricts
+  opportunities to exploit concurrency even when such opportunities
+  would not violate application-specific invariants. As a result,
+  database systems in practice support, and often encourage,
+  developers to use weaker alternatives. These alternatives break the
+  strong isolation guarantees offered by serializable transactions to
+  permit greater concurrency. Unfortunately, the semantics of weak
+  isolation is poorly understood, and usually explained only
+  informally in terms of low-level implementation
+  artifacts. Consequently, verifying high-level correctness properties
+  in such environments remains a challenging problem.
 

### 2016-07-05 introduction.tex

@@
 evident from the support for weakly-isolated transactions on
-conventional (unreplicated) databases as mentioned above.  With the
-added complexity of replication, richer mechanisms are needed to
-reason about weak isolation in tandem with weak consistency. Thus,
-frameworks for reasoning about weak isolation will necessarily have to
-generalize the reasoning frameworks developed for weak consistency in
-new and important ways.
+conventional (unreplicated) databases as mentioned above. In the
+presence of replication, however, the interaction between weak
+isolation and weak consistency can be subtle and non-trivial.
+Understanding weak isolation in these varied contexts thus requires
+new insights and substantial generalization of existing techniques.
 

### 2016-07-05 introduction.tex
typos

@@
 
-Unfortunately, weak isolation admits behaviors that are difficult to
+Unfortunately, weak isolation admits behaviours that are difficult to
 comprehend~\cite{berenson}. To quantify weak isolation anomalies,
@@
 isolation semantics.  Adya~\cite{adyaphd} presents the first formal
-definitions of some of well-known isolation levels in the context of a
+definitions of some well-known isolation levels in the context of a
 sequentially consistent (SC) database.  However, there has been little
@@
 arguments. Consequently, reasoning about weak isolation remains an
-error prone endeavor, with major database vendors~\cite{postgresiso,
+error prone endeavour, with major database vendors~\cite{postgresiso,
   mysqliso, oracleiso} continuing to document their isolation levels
@@
 to the semantics. \S\ref{sec:store-consistency} generalizes the
-framework to integrate support for weakly consistent data stores. We
+framework to integrate support for weakly-consistent data stores. We
 describe the impact of our reasoning framework in the context of

### 2016-07-05 motivation.tex
Ids are not single quoted.
typos

@@
 style reasoning~\cite{rgjones}, a compositional proof technique that
-allows us to reason about the behavior of individual threads by
+allows us to reason about the behaviour of individual threads by
 abstracting away interferences induced by other threads (collectively
@@
 makes the effects of any operation immediately visible to all
-subsequent operations. The natural behavior of {\sc sc} to totally
+subsequent operations. The natural behaviour of {\sc sc} to totally
 order all operations w.r.t. the $\hbZ$ relation could be in conflict

### 2016-07-05 opsem.tex
typos

@@
 resulting trace satisfies the constraints imposed by $\I$. This
-behavior is captured by the auxiliary reduction rule
+behaviour is captured by the auxiliary reduction rule
 \rulelabel{E-Aux} that factors out the trace extension aspect of the
-evaluation by abstracting away the operation-specific behavior as a
+evaluation by abstracting away the operation-specific behaviour as a
 function that generates an appropriate effect. We let $\mathcal{F}$
@@
 As described previously, \rulelabel{E-Aux} abstracts away the
-operation-specific behavior of a machine step as a function ($\F$)
+operation-specific behaviour of a machine step as a function ($\F$)
 that accepts a set ($S$) of effects chosen by the machine to make

### 2016-07-05 store-consistency.tex
typos

@@
 to witness arbitrary subsets of the global state, effectively mimicking the
-behavior of an eventually consistent ({\sc ec}) data
+behaviour of an eventually consistent ({\sc ec}) data
 store\footnote{Eventual consistency guarantees that in the absence of further

### 2016-07-06 store-consistency.tex
edits

@@
 $\eta$ at replica $R_1$, which is then merged to $R_2$. If the
-\emph{store is {\sc cc}}, then $R_2$'s state at merge event is same or
-larger than $R_1$'s state at generation event (the difference is
+\emph{store is {\sc cc}}, then $R_2$'s state at the merge event is same or
+larger than $R_1$'s state at the generation event (the difference is
 highlighted). In our subset model, $\op$ witnesses $S_1 \subseteq
@@
 Fig.~\ref{fig:wd1-decorated}.  \C{Wd2} is similar and not
-shown\footnote{The supplementary material shows the fully decorated program.}.
+shown.\footnote{The supplementary material shows the fully decorated program.}
 Temporary name \C{v3} has been introduced to capture the result of the

### 2016-07-06 motivation.tex
Edits to the example.

--- 2016-07-05 motivation.tex
@@
 \txnimp - a C-like imperative language equipped with a \C{txn} lexical
-block defining a transaction scope.  For expositional convenience,
-each \C{txn} block is associated with a single-quoted string in angle
-braces that uniquely identifies the transaction. We use Hoare triple
-notation to annotate programs with pre- and post- conditions.
+block defining a transaction scope. Each \C{txn} block is associated
+with an indentifier in angle braces that uniquely identifies the
+transaction. We use Hoare triple notation to annotate programs with
+pre- and post- conditions.
 
@@
 Fig.~\ref{fig:motiv-eg-1-hb}) may violate the invariant by writing
-$\C{k-a1}$ to \C{B} and committing \C{Wd2}, thus leading to
+$\C{k-a2}$ to \C{B} and committing \C{Wd2}, thus leading to
 $\committed(\C{Wd2}) \Rightarrow \C{B=k-a2}$. Fortunately,

### 2016-07-06 opsem.tex
Edits to the example.

--- 2016-07-05 opsem.tex
@@
 We define a small-step operational semantics for this language in
-terms of an abstract machine that generates an execution trace
-($\E$). The first component of the trace is a set ($\A$) of
-\emph{effects}, where an effect ($\eta$) documents a read (\C{RD(X)}),
-a write (\C{WR(X)}), or a transaction commit (\C{COMMIT}) operation
-performed during the execution. Any value associated with the
-operation (\eg, value read or value written) is also documented, and
-can be accessed via $\rval$. Every effect has a unique identifier
-accessible via $\id$, and a transaction identifier accessible via
-$\txn$.  The latter identifies the transaction that generated the
-effect. In every step of the evaluation, the machine reduces a \txnimp\
-term by executing a read, write or commit operation, generating an
-effect, and extending the trace. Since effects include transaction
+terms of an abstract machine that generates an execution trace ($\E$).
+The first component of the trace is a set ($\A$) of \emph{effects},
+where an effect ($\eta$) is a tuple of (a). the identifier ($T_i$) of
+the transaction that generated this effect, (b). an indentifier ($j$)
+that uniquely identifies this effect, (c). the operation ($\op$)
+documented by this effect, which can be either a read (\C{RD(X)}), a
+write (\C{WR(X)}), or a transaction commit (\C{COMMIT}), and (d). the
+value, if any, associated with the operation (\eg, value read or value
+written). Functions $\txn$, $\id$, $\oper$, and $\rval$ project each
+of the four aforementioned individual components of the tuple.
+In every step of the evaluation, the machine reduces a \txnimp\ term
+by executing a read, write or commit operation, generating an effect,
+and extending the trace. Since effects include transaction
 identifiers, the semantics distinguishes between terms ($t$) of
@@
 monotonically increasing identifiers to effects, as defined by the
-$\id(\eta) > \maxId(\A)$ condition of \rulelabel{E-Aux}. Evaluation
+$\id(\eta) > \maxId(\A)$ condition of \rulelabel{E-Aux} ($\maxId(\A)$
+returns the maximum number identifying an effect in $\A$). Evaluation
 contexts ($\ectx_i$) for transaction-bound terms are defined so as to

### 2016-07-06 reasoning.tex
Edits to the example.

@@
 also facilitates state-based reasoning via the interpretation function
-($\interp{\cdot}$) introduced in \S\ref{sec:syntax}. For example, if a
-\emph{bi-state} rely relation relates every pair of states $\sigma$
-and $\sigma'$ such that $\sigma'(X) \ge \sigma(X) \ge 0$, the
-corresponding \emph{bi-execution} rely relation relates every pair of
-executions $\E$ and $\E'$ such that $\interp{\E}(X) \ge
-\interp{\E'}(X) \ge 0$.  Assertions on states are also written
-similarly. For instance, consider the post-condition
-$\C{B\,=\,k-a1-a2}$ of the program in Fig.~\ref{fig:motiv-eg-1}. The
-corresponding assertion on the post-state
-($\lambda\sigma.~\sigma(\C{B}) \,=\, \C{k-a1-a2}$) asserts that in all
-states resulting from executing the program, the value of \C{B} is
-\C{k-a1-a2}. The equivalent execution-based assertion
-($\lambda\E.~\interp{\E}(\C{B})\,=\,\C{k-a1-a2}$) asserts that in all
-executions of the program, the value of \C{B} is \C{k-a1-a2}.
+($\interp{\cdot}$) introduced in \S\ref{sec:syntax}, which interprets
+the given set of effects as a state. For example, if a \emph{bi-state}
+rely relation relates every pair of states $\sigma$ and $\sigma'$ such
+that $\sigma'(X) \ge \sigma(X) \ge 0$, the corresponding
+\emph{bi-execution} rely relation relates every pair of executions
+$\E$ and $\E'$ such that $\interp{\E.\A}(X) \ge \interp{\E'.\A}(X) \ge
+0$ . Assertions on states are also written similarly. For instance,
+consider the post-condition $\C{B\,=\,k-a1-a2}$ of the program in
+Fig.~\ref{fig:motiv-eg-1}. The corresponding assertion on the
+post-state ($\lambda\sigma.~\sigma(\C{B}) \,=\, \C{k-a1-a2}$) asserts
+that in all states resulting from executing the program, the value of
+\C{B} is \C{k-a1-a2}. The equivalent execution-based assertion
+($\lambda\E.~\interp{\E.\A}(\C{B})\,=\,\C{k-a1-a2}$) asserts that in
+all executions of the program, the value of \C{B} is \C{k-a1-a2}.
 However, having access to an execution facilitates assertions that go
@@
   \hspace*{0.2in}\conj(\neg(\underE{\committed(\C{Wd2})}) \Rightarrow 
-      \C{\interp{\E}(B) = k-a1}) \\
+      \C{\interp{\E.\A}(B) = k-a1}) \\
    \hspace*{0.2in}\conj (\underE{\committed(\C{Wd2})} \Rightarrow 
-      \C{\interp{\E}(B) = k-a1-a2})
+      \C{\interp{\E.\A}(B) = k-a1-a2})
 \end{array}
@@
 The rely-guarantee judgments defined by the rules in
-Fig.~\ref{fig:rg-rules} are sound with respect to
-the semantics of Definition~\ref{def:rg-semantics}.\footnote{Full proofs are provided in the supplementary material.}
+Fig.~\ref{fig:rg-rules} are sound with respect to the semantics of
+Definition~\ref{def:rg-semantics}.\footnote{Formal proof of soundness
+is provided in the supplementary material.}
 \end{theorem}

### 2016-07-06 store-consistency.tex
Edits to the example.

--- 2016-07-05 store-consistency.tex
@@
 \begin{definition}
+\label{def:maxvis}
 $\I_s' : \E \rightarrow \Prop$ is said to be a maximum visibility
@@
 We now consider the proof of the example in Fig.~\ref{fig:motiv-eg-1}
-in greater detail. We assume an {\sc sc} store, such as a relational
-database, whose store-specific trace invariant ($\I_s$) was shown
-previously. Both transactions are run at {\sc si} isolation, hence
-$\I_c$ is $\lambda\E.~\underE{\C{SI(Wd1)}} \conj
-\underE{\C{SI(Wd1)}}$. As usual, $\I$ is $\I_s \conj \I_c$.
+in greater detail. Both withdraw transactions are run at {\sc si}
+isolation, hence $\I_c$ is $\lambda\E.~\underE{\C{SI(Wd1)}} \conj
+\underE{\C{SI(Wd1)}}$. We assume an {\sc sc} store, such as a
+conventional RDBMS, whose store-specific trace invariant ($\I_s$) is
+the maximul visibility weakening (Def.~\ref{def:maxvis}) of {\sc sc}
+w.r.t. $\I_c$, as described in \S\ref{sec:store-consistency}. In
+concrete terms, this means that an effect added to $\E.\A$ by the
+current thread or a concurrent thread immediately becomes visible to
+the subsequent operations unless we produce a witness that such
+visibility violates $\I_c$. As usual, $\I$ is $\I_s \conj \I_c$. Note
+that the language of our specifications and (most of) the assertions
+on executions is a decidable subset of the first-order logic (in
+particular, our specifications and rely-guarantee rules are carefully
+crafted to avoid transitive closures and quantification over infinite
+domains). Consequently, the proof obligations generated by the
+application of our rules can often be discharged via an SMT solver. We
+nonetheless give full details in the following.
 
@@
 Decorated parts of \C{Wd1} relevant to the discussion are shown in
-Fig.~\ref{fig:wd1-decorated}.  \C{Wd2} is similar and not
-shown\footnote{The supplementary material shows the fully decorated program.}.
+Fig.~\ref{fig:wd1-decorated}.  \C{Wd2} is similar and not shown.
 Temporary name \C{v3} has been introduced to capture the result of the
-RHS expression \C{C-a1}.  Assertions implicitly refer to the current
+RHS expression \C{B-a1}.  Assertions implicitly refer to the current
 execution ($\E$), just as Hoare triples implicitly refer to the
-current state.  The context for propositions is also the implicit,
-i.e., we write $\psi$ instead of $\underE{\psi}$. The proposition \C{k
-$\ge$ a1+a2} remains an invariant, hence elided. The pre-condition
-$P_1$ of \C{Wd1} accounts for the possibility of \C{Wd2} committing
-before \C{Wd1}, writing \C{k-a2} to \C{B}. Since neither \C{Wd1} nor
-\C{Wd2} are committed at the beginning, the pre-condition $P$ of the
-program is extended with $\neg\committed(\C{Wd1}) \wedge
-\neg\committed(\C{Wd2})$, from which $P_1$ follows. The post-condition
-$Q_1$ of \C{Wd1} asserts different values for \C{B} depending on
-whether or not \C{Wd2} has already committed.  It also asserts that
-\C{Wd1} has committed. The post-condition $Q_2$ of \C{Wd2} (not shown)
-is similar, but with the roles of \C{Wd1} and \C{Wd2} exchanged. The
-post-condition $Q$ follows from $Q_1 \wedge Q_2$, hence valid. It
-remains to show that \C{Wd1} satisfies its specification (\C{Wd2} must
-also satisfy its spec, but the proof is similar).
+current state. The context for propositions the value computation of
+shared variables is also the implicit, i.e., we write $\psi$ instead
+of $\underE{\psi}$ and $\C{B = k}$ instead of $\C{\interp{\E.\A}(B) =
+k}$. The proposition \C{k $\ge$ a1+a2} remains an invariant, hence
+elided. The pre-condition $P_1$ of \C{Wd1} accounts for the
+possibility of \C{Wd2} committing before \C{Wd1}, writing \C{k-a2} to
+\C{B}. Since neither \C{Wd1} nor \C{Wd2} are committed at the
+beginning, the pre-condition $P$ of the program is extended with
+$\neg\committed(\C{Wd1}) \wedge \neg\committed(\C{Wd2})$, from which
+$P_1$ follows. The post-condition $Q_1$ of \C{Wd1} asserts different
+values for \C{B} depending on whether or not \C{Wd2} has already
+committed.  It also asserts that \C{Wd1} has committed. The
+post-condition $Q_2$ of \C{Wd2} (not shown) is similar, but with the
+roles of \C{Wd1} and \C{Wd2} exchanged. The post-condition $Q$ follows
+from $Q_1 \wedge Q_2$, hence valid. It remains to show that \C{Wd1}
+satisfies its specification (\C{Wd2} must also satisfy its spec, but
+the proof is similar).
 
@@
 
-\noindent The rule also lets us assume that the execution satisfies trace
-invariant ($\I$), which asserts {\sc si} for both transactions.  Since
-both transactions now write to \C{B}, {\sc si} requires either
+\noindent The rule also lets us assume that the execution satisfies
+trace invariant ($\I$), which asserts {\sc si} for both transactions.
+Since both transactions now write to \C{B}, {\sc si} requires either
 $\C{Wd1} \visar \C{Wd2}$, or $\C{Wd2} \visar \C{Wd1}$. Since \C{Wd1}
-is not yet committed ($\neg\committed(\C{Wd1})$), {\sc si} (which
-subsumes {\sc rc}) prohibits the former possibility, allowing us the
-deduce $\C{Wd2} \visar \C{Wd1}$. This lets us derive
-${\committed(\C{Wd2})}  \Rightarrow \C{B = k-a2} \wedge \C{B =
-k-a2-a1}$, allowing us to prove the post-condition $\phi_5$.
+is not yet committed ($\neg\committed(\C{Wd1})$), {\sc si}, which
+subsumes {\sc rc} and requires only committed transactions be visible,
+prohibits $\C{Wd1} \visar \C{Wd2}$, allowing us the deduce $\C{Wd2}
+\visar \C{Wd1}$. This lets us derive ${\committed(\C{Wd2})}
+\Rightarrow \C{B = k-a2} \wedge \C{B = k-a2-a1}$, allowing us to prove
+the post-condition $\phi_5$.
 
@@
 The stability for $\phi_4$ w.r.t. $R_1$ is trivial to establish: if
-$R_1$ does not commit \C{Wd2}, then the value of \C{B} is unaffected,
-and if it commits \C{Wd2}, then $\C{Wd2}\not\visar\C{Wd1}$, hence
-$\phi_4$ doesn't care. For $\phi_5$, {\sc si} condition should be used
-to show that any interference from \C{Wd2} is invalid. The proof
-proceeds on the same lines as the Hoare proof discussed above; it is
-also discussed in detail in \S\ref{sec:motivation}.
+$R_1$ does not commit \C{Wd2}, then the value of \C{B} is unaffected.
+If it commits \C{Wd2}, then since \C{Wd1} had already performed (read)
+operations that cannot witness \C{Wd2}, we have
+$\C{Wd2}\not\visar\C{Wd1}$, and since $\phi_4$ only makes assertions
+under the case $\C{Wd2}\visar\C{Wd1}$, it is stable. For $\phi_5$,
+{\sc si} condition should be used to show that any interference from
+\C{Wd2} is invalid. The proof proceeds on the same lines as the Hoare
+proof discussed above; it is also discussed in detail in
+\S\ref{sec:motivation}.
 

### 2016-07-06 abstract.tex

--- 2016-07-05 abstract.tex
@@
   Serializability is a well-understood correctness criterion that
-  simplifies reasoning about the behavior of concurrent transactions
-  by ensuring that transactions are \emph{isolated} from each other.
-  However, enforcing serializable isolation comes at a steep cost in
-  performance because it necessarily restricts opportunities to
-  exploit concurrency even when such opportunities would not violate
-  application-specific invariants. As a result, database systems in
-  practice support, and often encourage, developers to use weaker
-  alternatives. These alternatives break the strong isolation
+  simplifies reasoning about the behaviour of concurrent transactions
+  by ensuring they are \emph{isolated} from each other while they
+  execute.  However, enforcing serializable isolation comes at a steep
+  cost in performance because it necessarily restricts opportunities
+  to exploit concurrency even when such opportunities would not
+  violate application-specific invariants. As a result, database
+  systems in practice support, and often encourage, developers to use
+  weaker alternatives. These alternatives break the strong isolation
   guarantees offered by serializable transactions to permit greater
@@
   To address this issue, we present a program logic that enables
-  compositional reasoning about the behavior of concurrently executing
-  weakly-isolated transactions. Notably, our development is parametric
-  over a transaction's specific isolation semantics, and the
-  consistency guarantees provided by the underlying data store,
+  compositional reasoning about the behaviour of concurrently
+  executing weakly-isolated transactions. Notably, our development is
+  parametric over a transaction's specific isolation semantics, and
+  the consistency guarantees provided by the underlying data store,
   allowing it to be applicable over a range of concurrency control

### 2016-07-06 store-consistency.tex
He Zhu's qn addressed.
Edits.

@@
 isolation, hence $\I_c$ is $\lambda\E.~\underE{\C{SI(Wd1)}} \conj
-\underE{\C{SI(Wd1)}}$. We assume an {\sc sc} store, such as a
+\underE{\C{SI(Wd2)}}$. We assume an {\sc sc} store, such as a
 conventional RDBMS, whose store-specific trace invariant ($\I_s$) is
@@
 visibility violates $\I_c$. As usual, $\I$ is $\I_s \conj \I_c$. Note
-that the language of our specifications and (most of) the assertions
-on executions is a decidable subset of the first-order logic (in
-particular, our specifications and rely-guarantee rules are carefully
-crafted to avoid transitive closures and quantification over infinite
-domains). Consequently, the proof obligations generated by the
-application of our rules can often be discharged via an SMT solver. We
-nonetheless give full details in the following.
+that the language of our specifications is a decidable subset of the
+first-order logic. In particular, our specifications and
+rely-guarantee rules are carefully crafted to avoid transitive
+closures and quantification over infinite domains. The domain of
+assertions is often linear arithmetic, which is also decidable.
+Consequently, the proof obligations generated by the application of
+our rules can often be discharged via an SMT solver. We nonetheless
+give full details in the following.
 
@@
 prove that an interference from \C{Wd2} does not affect the local
-state of \C{Wd1}.
+state of \C{Wd1}. $R_1$ refers to labels (\C{Wd1} and \C{Wd2}) of both
+the transactions. This is allowed because labels are artifacts of the
+execution $\E$; they are not (local) variables.
 

### 2016-07-06 introduction.tex
Typos etc.

@@
 programmers to be prescient about the possible ways various
-undesirable phenomena might manifest in their applications, and
-determine, in each case, if the phenomenon can be allowed without
-violating application invariants. This is understandably hard,
-especially in the absence of any formal underpinning to define weak
-isolation semantics.  Adya~\cite{adyaphd} presents the first formal
-definitions of some well-known isolation levels in the context of a
-sequentially consistent (SC) database.  However, there has been little
-progress relating Adya's system model to a formal operational
-semantics or a proof system that can facilitate rigorous correctness
-arguments. Consequently, reasoning about weak isolation remains an
-error prone endeavour, with major database vendors~\cite{postgresiso,
-  mysqliso, oracleiso} continuing to document their isolation levels
-primarily in terms of the undesirable phenomena a particular isolation
-level may induce, placing the burden on the programmer to determine
-application correctness.
+undesirable phenomena might manifest in their applications, and in
+each case determine if the phenomenon can be allowed without violating
+application invariants. This is understandably hard, especially in the
+absence of any formal underpinning to define weak isolation semantics.
+Adya~\cite{adyaphd} presents the first formal definitions of some
+well-known isolation levels in the context of a sequentially
+consistent (SC) database.  However, there has been little progress
+relating Adya's system model to a formal operational semantics or a
+proof system that can facilitate rigorous correctness arguments.
+Consequently, reasoning about weak isolation remains an error prone
+endeavour, with major database vendors~\cite{postgresiso, mysqliso,
+oracleiso} continuing to document their isolation levels primarily in
+terms of the undesirable phenomena a particular isolation level may
+induce, placing the burden on the programmer to determine application
+correctness.
 

### 2016-07-06 motivation.tex
Typos etc.

@@
 framework should be able to use the specification to determine if an
-interference is valid or not, allowing the programmer to only
-focus on the former. Our second observation is that this is possible
+interference is valid or not, allowing the programmer to only focus on
+valid interferences. Our second observation is that this is possible
 if the reasoning framework adequately tracks $\hbZ$ at each program

### 2016-07-06 opsem.tex
Typos etc.

@@
 transaction, we use the number $i$ instead of $T_i$ for identification
-(\eg, in $\C{txn}\langle i \rangle$). For notational convenience, we
+(\eg in $\C{txn}\langle i \rangle$). For notational convenience, we
 let $t$ range over both expressions and commands.
@@
 ($\I$). $\I$ is a function from traces ($\E$) to first-order logical
-formula ($\Prop$) that define well-formedness
+formulas ($\Prop$) that define well-formedness
 constraints over traces.  The machine takes a step only if the
@@
 constraints to enforce isolation guarantees.
-\S\ref{sec:ansi-isolation} shows the specification of various ANSI SQL
-isolation levels stated as trace well-formedness constraints.
+\S\ref{sec:ansi-isolation} specifies various ANSI SQL isolation levels
+stated as trace well-formedness constraints.
 

### 2016-07-06 ansi-isolation.tex
typos
Kia's suggestions and typos.

@@
-\subsection{Isolation Specifications}
+\subsection{Specifications}
 \label{sec:ansi-isolation}
@@
 expressed as constraints over trace well-formedness. For brevity and
-convenience, we adopt few notations in this and subsequent sections.
-An execution trace is destructed as ($\A$,$\visZ$) whenever individual
-components of the pair are needed. Otherwise, it is written as $\E$.
-Sometimes, the dot notation (\eg~$\E.\A$) is also used. Since $\A$ and
-$\visZ$ are both sets, we lift the operations on sets to pairs of sets
-when updating $\E$. For example, $\E' = \E \cup
+convenience, we introduce some notations that are used in the following
+sections.  An execution trace is destructed as ($\A$,$\visZ$) whenever
+individual components of the pair are needed. Otherwise, it is written
+as $\E$.  Sometimes, the dot notation (\eg~$\E.\A$) is also
+used. Since $\A$ and $\visZ$ are both sets, we lift the operations on
+sets to pairs of sets when updating $\E$. For example, $\E' = \E \cup
 (\{\eta_2\},\{(\eta_1,\eta_2)\})$ expands to $\E' = (\E.\A \cup
@@
 and~\ref{fig:ansi-isolation}.
+
+\subsubsection{Trace Relations}
 
@@
 
-The ANSI SQL 92 standard requires \iso{Read Committed} isolation to
-avoid the dirty reads phenomenon, which is achieved by enforcing
-$\mathtt{AtomicVis}$ and $\mathtt{CommitVis}$ guarantees. The {\sc rc}
-specification (\C{RC})\footnote{In the following, we use small caps to abbreviate
-  isolation levels (\emph{e.g.,} {\sc rc} for \emph{Read Committed}), and
-  typewriter font for abbreviations of the specification of an isolation
-  level (\emph{e.g.,} \C{RC} denotes the specification for {\sc rc} as
-  given in Fig.~\ref{fig:ansi-isolation}).} is therefore a combination of these two
-guarantees. The specification also agrees with the description and
-implementation~\cite{bailishat,pldi15} of {\sc rc} for highly
-available replicated stores. On relational databases, however, {\sc
-rc} has also come to be associated with the $\mathtt{MonotonicVis}$
-guarantee.  Nonetheless, $\mathtt{AtomicVis}$ and $\mathtt{CommitVis}$
-are sufficient to reason about {\sc rc} isolation on relational stores
-too. The combination of these guarantees with the {\sc sc} property of
-relational stores (formalized in \S\ref{sec:store-consistency})
-automatically leads to the monotonicity guarantee, which explains why
-{\sc rc} comes with $\mathtt{MonotonicVis}$ on such stores regardless
-of the implementation. On weakly consistent stores however,
+\subsubsection{Isolation Levels}
+
+Fig.~\ref{fig:ansi-isolation} shows specifications of various
+well-known isolation guarantees.  The ANSI SQL 92 standard requires
+\iso{Read Committed} isolation to avoid the dirty reads phenomenon,
+which is achieved by enforcing $\mathtt{AtomicVis}$ and
+$\mathtt{CommitVis}$ guarantees. The {\sc rc} specification
+(\C{RC})\footnote{In the following, we use small caps to abbreviate
+isolation levels (\emph{e.g.,} {\sc rc} for \emph{Read Committed}),
+and typewriter font for abbreviations of the specification of an
+isolation level (\emph{e.g.,} \C{RC} denotes the specification for
+{\sc rc} as given in Fig.~\ref{fig:ansi-isolation}).} is therefore a
+combination of these two guarantees. The specification also agrees
+with the description and implementation~\cite{bailishat,pldi15} of
+{\sc rc} for highly available replicated stores. On relational
+databases, however, {\sc rc} has also come to be associated with the
+$\mathtt{MonotonicVis}$ guarantee.  Nonetheless, $\mathtt{AtomicVis}$
+and $\mathtt{CommitVis}$ are sufficient to reason about {\sc rc}
+isolation on relational stores too. The combination of these
+guarantees with the {\sc sc} property of relational stores (formalized
+in \S\ref{sec:store-consistency}) automatically leads to the
+monotonicity guarantee, which explains why {\sc rc} comes with
+$\mathtt{MonotonicVis}$ on such stores regardless of the
+implementation. On weakly consistent stores however,
 $\mathtt{AtomicVis}$ and $\mathtt{CommitVis}$ do not imply

### 2016-07-06 reasoning.tex
Kia's suggestions and typos.

@@
 traces ($\E$). In particular, our rely ($R$) and guarantee ($G$)
-relations relate executions (i.e., $R,\,Q\,\subseteq\,\E\times\E$),
+relations relate executions (i.e., $R,\,G\,\subseteq\,\E\times\E$),
 and our pre- ($P$) and post- ($Q$) conditions are assertions over
@@
 
-The \rulelabel{RG-Conseq} rule define ways to strengthen or
+The \rulelabel{RG-Conseq} rule defines ways to strengthen or
 weaken relations and assertions associated with the RG judgment of

### 2016-07-06 case-studies.tex
Reverted changes to ansi-isolation.

@@
 serializability requirement for increments, just by changing the
-definition of the $\interp{\E}(X)$ (\S\ref{sec:opsem}) to return the
-value of the largest write-to-$X$ instead of the latest write-to-$X$.
-In other words, if the reads chose the largest valued write effect,
-instead of the latest write effect they witness, then counter
-increments need not be serializable for reads to witness a
-monotonically increasing count.\footnote{The trade-off is that the view
-count is now no longer accurate, although it grows monotonically.}
-We implemented this version of the counter as an op-based RDT, which
-uses {\sc cc} reads and {\sc ec} increments, both un-isolated (i.e.,
-\emph{Read Uncommitted}), and obtained a 52\% improvement in overall
-latency compared to the version that uses serializable increments.
+definition of the $\interp{S}(X)$ (\S\ref{sec:opsem}) to return
+the value of the largest write-to-$X$ instead of the latest
+write-to-$X$ in $S$.  In other words, if the reads chose the largest
+valued write effect, instead of the latest write effect they witness,
+then counter increments need not be serializable for reads to witness
+a monotonically increasing count.\footnote{The trade-off is that the
+view count is now no longer accurate, although it grows
+monotonically.} We implemented this version of the counter as an
+op-based RDT, which uses {\sc cc} reads and {\sc ec} increments, both
+un-isolated (i.e., \emph{Read Uncommitted}), and obtained a 52\%
+improvement in overall latency compared to the version that uses
+serializable increments.
 \vspace*{-3pt}

### 2016-07-06 ansi-isolation.tex
Reverted changes to ansi-isolation.
Edits.

--- 2016-07-04 ansi-isolation.tex
@@
 Transaction $T_j$ is said to have experienced \emph{read-my-writes}
-visibility under $\E$ if every effect ($\eta_1$) of $T_j$ is visible
-to every subsequent effect ($\eta_2$) of the same transaction in $\E$.
-This lets $T_j$ to never lose its own updates. Monotonic visibility
-adds one more constraint to \emph{read-my-writes}; besides requiring $\eta_1$
-to be visible to $\eta_2$, it also requires every effect ($\eta$)
-visible to $\eta_1$ in $\E$ to be visible to $\eta_2$ as well. Thus,
-later operations in a transaction witness at least the same set of
-effects witnessed by the earlier operations, if not more (hence,
-``monotonic'' visibility). Atomic visibility allows an effect
-$\eta_2$ of $T_j$ to witness an effect $\eta_1$ of $T_i$ only if all
-effects of $T_i$ in $\E$ are also visible to $\eta_2$. Atomic
-visibility thus prevents a transaction from being partially visible.
-However, atomic visibility does not prevent an uncommitted transaction
-from being visible. This is addressed by \C{CommitVis}, which requires
-the commit effect of $T_i$ to be visible whenever any effect of $T_i$
-is visible.
+visibility (\C{RMWVis}) under $\E$ if every effect ($\eta_1$) of $T_j$
+is visible to every subsequent effect ($\eta_2$) of the same
+transaction in $\E$.  This lets $T_j$ to never lose its own updates.
+Monotonic visibility (\C{MonotonicVis}) adds one more constraint to
+\emph{read-my-writes}; besides requiring $\eta_1$ to be visible to
+$\eta_2$, it also requires every effect ($\eta$) visible to $\eta_1$
+in $\E$ to be visible to $\eta_2$ as well. Thus, later operations in a
+transaction witness at least the same set of effects witnessed by the
+earlier operations, if not more (hence, ``monotonic'' visibility).
+Atomic visibility (\C{AtomicVis}) allows an effect $\eta_2$ of $T_j$
+to witness an effect $\eta_1$ of $T_i$ only if all effects of $T_i$ in
+$\E$ are also visible to $\eta_2$. Atomic visibility thus prevents a
+transaction from being partially visible.  However, atomic visibility
+does not prevent an uncommitted transaction from being visible. This
+is addressed by \C{CommitVis}, which requires the commit effect of
+$T_i$ to be visible whenever any effect of $T_i$ is visible.
 
@@
 $\mathtt{AtomicVis}$ and $\mathtt{CommitVis}$ guarantees. The {\sc rc}
-specification (\C{RC})\footnote{In the following, we use small caps to abbreviate
-  isolation levels (\emph{e.g.,} {\sc rc} for \emph{Read Committed}), and
-  typewriter font for abbreviations of the specification of an isolation
-  level (\emph{e.g.,} \C{RC} denotes the specification for {\sc rc} as
-  given in Fig.~\ref{fig:ansi-isolation}).} is therefore a combination of these two
-guarantees. The specification also agrees with the description and
+specification (\C{RC})\footnote{In the following, we use small caps to
+abbreviate isolation levels (\emph{e.g.,} {\sc rc} for \emph{Read
+Committed}), and typewriter font for abbreviations of the
+specification of an isolation level (\emph{e.g.,} \C{RC} denotes the
+specification for {\sc rc} as given in
+Fig.~\ref{fig:ansi-isolation}).} is therefore a combination of these
+two guarantees. The specification also agrees with the description and
 implementation~\cite{bailishat,pldi15} of {\sc rc} for highly

### 2016-07-06 case-studies.tex
Github link added.

@@
   implementation, along with applications and experiments, are
-  available at the following anonymized url: \url{...}} The focus of
+  available at the following anonymized url:
+  \url{https://github.com/popl17/popl17-code}} The focus of
 our formal verification efforts are such translated \txnimp programs.

### 2016-07-06 case-studies.tex
edits to get to 12 pages

@@
+\vspace*{-4pt}
 \section{Case Studies}
@@
 
-In this section, we present two case studies that demonstrate the
-applicability of our framework in varied settings, and its consequent
-practical utility. Both our case studies target Ruby-on-Rails
-applications. To analyze these, and other similar open-source
-applications, we have developed a symbolic execution engine that
-compiles a Rails program to an abstract program in an extended version
-of \txnimp, along with some proof automation support.\footnote{Details
-   can be found in the supplementary material.  Source code for our
-  implementation, along with applications and experiments, are
-  available at the following anonymized url:
-  \url{https://github.com/popl17/popl17-code}} The focus of
-our formal verification efforts are such translated \txnimp programs.
+We present two case studies below that demonstrate the applicability
+of our framework and its practical utility. Both examples
+target Ruby-on-Rails applications, which are commonly used to
+interface applications to database systems. To analyze these, and
+other similar open-source applications, we have developed a symbolic
+execution engine that compiles a Rails program to an abstract program
+in an extended version of \txnimp, along with some proof automation
+support.\footnote{Details can be found in the supplementary material.
+  Source code for our implementation, along with applications and
+  experiments, are available at the following anonymized url:
+  \url{...}} The focus of our formal verification efforts are such
+translated \txnimp programs.
 
@@
 on an {\sc sc} store, but nonetheless observed monotonicity
-violations. The reason for such violations became evident when
-we instantiated our proof framework for an {\sc ec} store.
-Under this instantiation, it is impossible to derive the
+violations.  Under this instantiation, it is impossible to derive the
 \C{MonotonicVis} (\S\ref{sec:ansi-isolation}) property needed to show
 that successive reads in a session witness monotonically increasing
-state. However, assuming {\sc cc} store consistency for reads allowed
+state. However, assuming {\sc cc} for reads allowed
 us to complete the proof, and recover the monotonicity invariant. In
@@
 serializability requirement for increments, just by changing the
-definition of the $\interp{S}(X)$ (\S\ref{sec:opsem}) to return
-the value of the largest write-to-$X$ instead of the latest
-write-to-$X$ in $S$.  In other words, if the reads chose the largest
-valued write effect, instead of the latest write effect they witness,
-then counter increments need not be serializable for reads to witness
-a monotonically increasing count.\footnote{The trade-off is that the
+definition of the $\interp{S}(X)$ (\S\ref{sec:opsem}) to return the
+value of the largest write-to-$X$ instead of the latest write-to-$X$
+in $S$.  Thus, if the reads chose the largest-valued write
+effect, instead of the latest write effect they witness, then counter
+increments need not be serializable for reads to witness a
+monotonically increasing count.\footnote{The trade-off is that the
   view count is now no longer accurate, although it grows
@@
 un-isolated (i.e., \emph{Read Uncommitted}), and obtained a 52\%
-improvement in overall latency compared to the version that uses
-serializable increments.
+improvement in overall latency compared to the serializable original.
 \vspace*{-3pt}
@@
 for MySQL with this configuration resulted in no violations (as
-expected), while yielding a 38\% reduction in latency than an {\sc
-  ser} configuration.  Since Postgres' isolation levels are stronger
-than the corresponding ANSI levels, \rulelabel{RG-Conseq} rule
-(Fig.~\ref{fig:rg-rules}) lets us immediately deduce that the same
-choice of isolation levels also work on Postgres.
+expected), while yielding a 38\% reduction in latency compared to an
+{\sc ser} configuration.  Since Postgres' isolation levels are
+stronger than the corresponding ANSI levels, the \rulelabel{RG-Conseq}
+rule (Fig.~\ref{fig:rg-rules}) lets us immediately deduce that the
+same choice of isolation levels also work on Postgres.
 

### 2016-07-06 store-consistency.tex
edits to get to 12 pages

@@
 semantics and proof system are general enough to admit replication as
-a special case of our formulation. In this section, we explain how the
-standard artifacts of weakly-consistent replication manifest in our
-reasoning framework.
+a special case of our formulation, as we explain below.
 
@@
 conventional RDBMS, whose store-specific trace invariant ($\I_s$) is
-the maximul visibility weakening (Def.~\ref{def:maxvis}) of {\sc sc}
+the maximal visibility weakening (Def.~\ref{def:maxvis}) of {\sc sc}
 w.r.t. $\I_c$, as described in \S\ref{sec:store-consistency}. In
@@
 closures and quantification over infinite domains. The domain of
-assertions is often linear arithmetic, which is also decidable.
+assertions is drawn from linear arithmetic, which is also decidable.
 Consequently, the proof obligations generated by the application of
-our rules can often be discharged via an SMT solver. We nonetheless
-give full details in the following.
+our rules can often be discharged via an SMT solver.
 
@@
 is not committed in $\E$, but committed in $\E'$.):
-
+\vspace*{-1.5pt}
 \begin{smathpar}
@@
 \end{smathpar}
-For the sake of completeness, we also explicitly represent the fact
-that all the effects added by $R_1$ are of \C{Wd2}. This is needed to
+
+\noindent For the sake of completeness, we also explicitly represent the fact
+that all the effects added by $R_1$ are from \C{Wd2}. This is needed to
 prove that an interference from \C{Wd2} does not affect the local
-state of \C{Wd1}. $R_1$ refers to labels (\C{Wd1} and \C{Wd2}) of both
-the transactions. This is allowed because labels are artifacts of the
-execution $\E$; they are not (local) variables.
+state of \C{Wd1}.
 
@@
 $\C{Wd2}\not\visar\C{Wd1}$, and since $\phi_4$ only makes assertions
-under the case $\C{Wd2}\visar\C{Wd1}$, it is stable. For $\phi_5$,
-{\sc si} condition should be used to show that any interference from
-\C{Wd2} is invalid. The proof proceeds on the same lines as the Hoare
-proof discussed above; it is also discussed in detail in
-\S\ref{sec:motivation}.
+under the case $\C{Wd2}\visar\C{Wd1}$, it is stable. For $\phi_5$, the
+{\sc si} condition is used to show that any interference from \C{Wd2}
+is invalid. The proof proceeds on the same lines as the proof
+discussed above, and is also discussed in detail
+in\S\ref{sec:motivation}.
 
@@
 the rely relation $R_2$ for \C{Wd2}. Since the implementation of
-\C{Wd2} is same as \C{Wd1}, $R_2$ (hence $G_1$) is same as $R_1$ shown
-above, but with \C{Wd2} and \C{Wd1} interchanged. Continuing our focus
-on the assignment statement, the guarantee proof requires us to show
-that if executing the assignment takes the execution from $\E$ to
+\C{Wd2} is same as \C{Wd1}'s, $R_2$ (hence $G_1$) is same as $R_1$
+shown above, but with \C{Wd2} and \C{Wd1} interchanged. Continuing our
+focus on the assignment statement, the guarantee proof requires us to
+show that if executing the assignment takes the execution from $\E$ to
 $\E'$, such that $\E'$ satisfies $\I$, then $(\E,\E') \in G_1$.  The
@@
 that (a). the assignment statement does not commit \C{Wd2}, and (b).
-it adds a write effect $\eta$ such that $\txn(\eta)=\C{Wd1}$. This
-finishes the proof for the triple in focus. The proofs for other
-triples of \C{Wd1} can also be constructed similarly. The combination
-of correctness proofs for \C{Wd1} and \C{Wd2} yields a correctness
-proof for the program.
+it adds a write effect $\eta$ such that $\txn(\eta)=\C{Wd1}$.  The
+proofs for other triples of \C{Wd1} can also be constructed
+similarly. The combination of correctness proofs for \C{Wd1} and
+\C{Wd2} yields a correctness proof for the program.
 

### 2016-07-06 case-studies.tex
Latest comments by KC and Kia addressed.

--- 2016-07-06 case-studies.tex
@@
 guarantees and repeated relevant parts of the proof to discover that
-\iso{Repeatable Read} is in fact sufficient for writers - an
+Postgres' \iso{Repeatable Read} is in fact sufficient for writers - an
 observation further validated through experiments. As usual, using

### 2016-07-06 motivation.tex
Latest comments by KC and Kia addressed.

@@
 isolated from the writes of uncommitted transactions, and is therefore
-free from \emph{dirty reads}~\cite{berenson} that read uncommitted
-data. In the current example, {\sc rc} isolation admits the two
-executions shown in Fig.~\ref{fig:rc-ex} on a strongly consistent
-({\sc sc}) store, such as a conventional RDBMS.
+free from \emph{dirty reads}~\cite{berenson} of uncommitted data. In
+the current example, {\sc rc} isolation admits the two executions
+shown in Fig.~\ref{fig:rc-ex} on a strongly consistent ({\sc sc})
+store, such as a conventional RDBMS.
 
@@
 \end{smathpar}
-\caption{\small Visibility and interference properties for different weak isolation levels can be
-  captured by different instantiations of their happens-before relation.}
+\caption{\small Interference properties for different weak isolation
+levels can be captured as constraints over the happens-before
+relation.}
 \label{fig:interference-ex}
@@
 associated with a transaction. The first observation we make is that
-we can constrain interference by constraining the definition of the
+we can constrain interference by constraining the nature of the
 \emph{happens-before} ($\hbZ$) relation, which ultimately dictates
-what actions become visible to a transaction, or its constituents, and
-when. To do so, we axiomatize the $\hbZ$ relation to capture the
-interference characteristics of different isolation levels as their
-specifications. Two such specifications are shown\footnote{Note that
-specifications presented here are solely for illustration. The actual
-specifications described in \S\ref{sec:ansi-isolation} are more
-nuanced.} in Fig.~\ref{fig:interference-ex}. The \iso{Read Committed}
-specification ($\psi_{RC}$) allows an operation $\eta_1$ of $T_1$ to
-happen-before an operation $\eta_2$ of $T_2$ only if every operation
-in $T_1$ (including its commit) happens-before $\eta_2$ (we let $T_1
-\hboar \eta_2$ denote this). The specification does not require $T_1$
-to execute and commit before all operations of $T_2$, thus allowing it
-to interfere in $T_2$. The \iso{Snapshot Isolation} specification
-($\psi_{SI}$) requires two transactions that write to the same
-variable (denoted $\wrstoar$)to be related by happens-before.  This
-effectively prohibits interference due to actions in $T_1$ being
-interleaved in $T_2$ (or, vise versa).
+what other transactions become visible to a transaction, or its
+constituents, and when. To do so, we axiomatize the $\hbZ$ relation to
+capture the interference characteristics of different isolation levels
+as their specifications. Two such specifications are
+shown\footnote{Note that specifications presented here are solely for
+illustration. The actual specifications described in
+\S\ref{sec:ansi-isolation} are more nuanced.} in
+Fig.~\ref{fig:interference-ex}. The \iso{Read Committed} specification
+($\psi_{RC}$) allows an operation $\eta_1$ of a transaction $T_1$ to
+happen-before an operation $\eta_2$ of another transaction $T_2$ only
+if every operation in $T_1$ (including its commit) happens-before
+$\eta_2$ (we let $T_1 \hboar \eta_2$ denote this). The specification
+does not require $T_1$ to execute and commit before all operations of
+$T_2$, thus allowing it to interfere in $T_2$, just as \C{Wd2}
+interferes in \C{Wd1} in Fig.~\ref{fig:motiv-eg-1-a}. The
+\iso{Snapshot Isolation} specification ($\psi_{SI}$) requires two
+transactions that write to the same variable (denoted $\wrstoar$) to be
+related by happens-before.  This effectively prohibits interference
+due to actions in $T_1$ being interleaved in $T_2$ (or, vice versa).
 
@@
 
-We have thus far assumed a strongly consistent ({\sc sc}) store that,
-in the absence of transactions with special isolation requirements,
-makes the effects of any operation immediately visible to all
-subsequent operations. The natural behaviour of {\sc sc} to totally
-order all operations w.r.t. the $\hbZ$ relation could be in conflict
-with the constraints imposed by weak isolation. To meet the
-requirements of weaker isolation levels, say {\sc rc}, the store has
-to adapt itself to \emph{hide} the effects of concurrent transactions
-until they are committed. Once committed though, effects need to be
-immediately visible to subsequent operations. This semantics can be
-built into the reasoning framework, leading to a proof system
-tailor-made for such stores. However, for the reasoning framework to
-be truly useful, it has to be parametric over different consistency
-semantics, supporting stores that are weaker than {\sc sc} (e.g.,
-causally or eventually consistent), and should be able to reconcile
-conflicts between consistency and isolation constraints.  We
-demonstrate how our reasoning framework makes this possible in the
-following sections.
+Implicit in the above discussion is the assumption of a strongly
+consistent ({\sc sc}) store that guarantees the visibility of all
+previously committed transactions. {\sc sc} semantics can be built
+into the reasoning framework, leading to a proof system tailor-made
+for such stores. However, for the reasoning framework to be truly
+useful, it has to be parametric over different consistency semantics,
+supporting stores that are weaker than {\sc sc} (e.g., causally or
+eventually consistent), and should be able to reconcile conflicts
+between consistency and isolation constraints.  We demonstrate how our
+reasoning framework makes this possible in the following sections.

### 2016-07-06 store-consistency.tex
Latest comments by KC and Kia addressed.

--- 2016-07-06 store-consistency.tex
@@
 program-specific (or, client-specific) invariant, to capture
-consistency and isolation constraints, respectively.  $\I$ is now a
-conjunction: $\I \,=\, \lambda\E.~\I_s(\E) \wedge \I_c(\E)$ (often
-simplified to $\I \,=\, \I_s\wedge\I_c$).  While the program-specific
-trace invariant ($\I_c$) is independent of the underlying data store,
-the store specific invariant ($\I_s$) changes from store to store
-depending on the consistency level.  For an eventually consistent
-store, $\I_s$ is simply \emph{true}. Stronger stores, such as those
-that support strong consistency, have non-trivial definitions for
-$\I_s$.
+consistency and isolation constraints\footnote{$\I_c$ captures the
+combination of transaction-specific isolation constraints.},
+respectively.  $\I$ is now a conjunction: $\I \,=\,
+\lambda\E.~\I_s(\E) \wedge \I_c(\E)$ (often simplified to $\I \,=\,
+\I_s\wedge\I_c$). While the program-specific trace invariant ($\I_c$)
+is independent of the underlying data store, the store specific
+invariant ($\I_s$) changes from store to store depending on the
+consistency level.  For an eventually consistent store, $\I_s$ is
+simply \emph{true}. Stronger stores, such as those that support strong
+consistency, have non-trivial definitions for $\I_s$.
 
@@
 ($\E.\A$) that an operation witnesses. Constraints over replica states
-therefore manifest as constraints over a specific visibility relation.
-For example, instead of requiring the store to be causally consistent,
-an operation can witness a causally consistent subset of the state;
-such demands can be made via the trace invariant $\I$. For a
-pre-condition ($P$) of the operation to be useful, it has to be an
-assertion over every causally consistent subset of the global state.
-Since any replica that eventually executes the operation has to expose
-one such subset ($S$), the pre-condition is guaranteed to hold
-regardless of the replica. There is however one problem with this
-explanation - by considering subsets of just one global state, it
-ignores the fact that the global state (hence, the replica states)
-change during the execution of the operation. To account for such
-changes, we might choose to distinguish between an effect generation
-event at one replica $r_1$ and an effect merge event at replica $r_2$,
-requiring that \emph{non-conflicting} operations execute between these
-two events at $r_2$, and that they preserve certain
-invariants~\cite{gotsmanpopl16}.  Instead, our framework folds all such
-machinery into a stability condition predicated on $\I$
+therefore manifest as constraints over the visibility relation.  For
+example, instead of requiring the store to be causally consistent, an
+operation can witness a causally consistent subset of the state; such
+demands can be made via the trace invariant $\I$. For a pre-condition
+($P$) of the operation to be useful, it has to be an assertion over
+every causally consistent subset of the global state.  Since any
+replica that eventually executes the operation has to expose one such
+subset ($S$), the pre-condition is guaranteed to hold regardless of
+the replica. There is however one problem with this explanation - by
+considering subsets of just one global state, it ignores the fact that
+the global state (hence, the replica states) change during the
+execution of the operation. To account for such changes, we might
+choose to distinguish between an effect generation event at one
+replica $r_1$ and an effect merge event at replica $r_2$, requiring
+that \emph{non-conflicting} operations execute between these two
+events at $r_2$, and that they preserve certain
+invariants~\cite{gotsmanpopl16}.  Instead, our framework folds all
+such machinery into a stability condition predicated on $\I$
 (\S\ref{sec:rely-guarantee}).  Since any change to the global state
@@
 conventional RDBMS, whose store-specific trace invariant ($\I_s$) is
-the maximul visibility weakening (Def.~\ref{def:maxvis}) of {\sc sc}
-w.r.t. $\I_c$, as described in \S\ref{sec:store-consistency}. In
-concrete terms, this means that an effect added to $\E.\A$ by the
-current thread or a concurrent thread immediately becomes visible to
-the subsequent operations unless we produce a witness that such
-visibility violates $\I_c$. As usual, $\I$ is $\I_s \conj \I_c$. Note
-that the language of our specifications is a decidable subset of the
-first-order logic. In particular, our specifications and
-rely-guarantee rules are carefully crafted to avoid transitive
-closures and quantification over infinite domains. The domain of
-assertions is often linear arithmetic, which is also decidable.
-Consequently, the proof obligations generated by the application of
-our rules can often be discharged via an SMT solver. We nonetheless
-give full details in the following.
+the maximul visibility weakening of {\sc sc} w.r.t. $\I_c$, as
+described in \S\ref{sec:store-consistency}. In concrete terms, this
+means that an effect added to $\E.\A$ by the current thread or a
+concurrent thread immediately becomes visible to the subsequent
+operations unless we produce a witness that such visibility violates
+$\I_c$. As usual, $\I$ is $\I_s \conj \I_c$. Note that the language of
+our specifications is a decidable subset of the first-order logic. In
+particular, our specifications and rely-guarantee rules are carefully
+crafted to avoid generating proof obligations with transitive closures
+and quantification over infinite domains. The domain of assertions is
+often linear arithmetic, which is also decidable. Consequently, the
+generated proof obligations can often be discharged via an SMT solver.
+We nonetheless give full details in the following.
 
@@
 execution ($\E$), just as Hoare triples implicitly refer to the
-current state. The context for propositions the value computation of
-shared variables is also the implicit, i.e., we write $\psi$ instead
+current state. The context for propositions and the state for shared 
+variables is also the implicit, i.e., we write $\psi$ instead
 of $\underE{\psi}$ and $\C{B = k}$ instead of $\C{\interp{\E.\A}(B) =

### 2016-07-06 related-work.tex

@@
+\vspace*{-8pt}
 \section{Related Work}
@@
 Adya~\cite{adyaphd} specifies several weak isolation levels in terms
-of \emph{dependency graphs} between transactions, by defining the
-kinds of dependency cycles that are forbidden in each case. The
-operational nature of Adya's specifications make them suitable for
-runtime monitoring and anomaly
-detection~\cite{kemmevldb,feketesigmod08,pssi2011}, whereas the
-declarative nature of our specifications make them suitable for formal
-reasoning about program behaviour. Sivaramakrishnan \emph{et
+of \emph{dependency graphs} between transactions, and the kinds of
+dependencies that are forbidden in each case. The operational nature
+of Adya's specifications make them suitable for runtime monitoring and
+anomaly detection~\cite{kemmevldb,feketesigmod08,pssi2011}, whereas
+the declarative nature of our specifications make them suitable for
+formal reasoning about program behaviour. Sivaramakrishnan \emph{et
   al.}~\cite{pldi15} specify isolation levels declaratively as trace
 well-formedness conditions, but their specifications implicitly assume
-a complete trace with only committed transactions, and is thus not
-suitable to reason operationally about a program as it builds the
-trace; their framework also requires application invariants be
-expressed in terms of the visibility relations used to define
-isolation and consistency properties.  Cerone \emph{et
-  al.}~\cite{gotsmanconcur15} specify isolation levels with atomic
-visibility, but their specifications are also for complete traces, and
-assumes all transactions execute using either {\sc si} or {\sc ser},
-which is often not the case in practice. While all the aforementioned
-specification frameworks use the vocabulary introduced
-in~\cite{burckhardt14} to specify replicated data types, none of them
-are equipped with a reasoning framework that can use such
-specifications to verify programs under weak isolation.
+a complete trace with only committed transactions, making it difficult
+to reason about a program as it builds the trace; their framework also
+requires application invariants be expressed in terms of low-level
+visibility relations.  Cerone \emph{et al.}~\cite{gotsmanconcur15} specify
+isolation levels with atomic visibility, but their specifications are
+also for complete traces, and assumes all transactions execute using
+either {\sc si} or {\sc ser}, which is often not the case in
+practice. While all the aforementioned specification frameworks use
+the vocabulary introduced in~\cite{burckhardt14} to specify replicated
+data types, none of them are equipped with a reasoning framework that
+can use such specifications to verify programs under weak isolation.
 
+\vspace*{-4pt}
 \paragraph{Reasoning under weak isolation} In~\cite{feketessi}, Fekete
@@
 proposes an algorithm that allocates either {\sc si} or {\sc ser}
-isolation levels to transactions while guaranteeing the
+isolation levels to transactions while guaranteeing
 serializability. In~\cite{gotsmanpodc16}, Cerone \emph{et al.} improve
@@
 isolation levels (rooted in~\cite{berenson}) under a pre-defined store
-consistency model (usually {\sc sc}).
+consistency model.
 
+\vspace*{-4pt}
 \paragraph{Reasoning under weak consistency} There have been several
@@
 consistency~\cite{bailisvldb, alvarocalm, gotsmanpopl16,redblueatc,
-redblueosdi, ecinec}. All of them assume a system model that offers a
-choice between a \emph{coordination-free} weak consistency level
+  redblueosdi, ecinec}. All of them assume a system model that offers
+a choice between a \emph{coordination-free} weak consistency level
 (\emph{e.g.}, eventual consistency~\cite{redblueosdi, redblueatc,
   ecinec, alvarocalm, bailisvldb}) or causal
-consistency~\cite{lbc16,gotsmanpopl16}). The key technical issue in
-these efforts involves proving that atomic and fully isolated
-operations preserve application invariants when executed on these
-consistency levels.  In contrast, we admit weakly isolated
-transactions, and our system model accepts \emph{specifications} of
-consistency and isolation levels drawn from an expressive logic.
-Gotsman \emph{et al.}~\cite{gotsmanpopl16} adapts \iso{Parallel
-Snapshot Isolation} to the aforementioned setting by interpreting it
-as a consistency level that serializes writes to objects; a dedicated
-proof rule is developed to help prove prove program invariants hold
-under this model. By parameterizing our proof system over a gamut of
-weak isolation specifications, we avoid the need to define a separate
-proof rule for each new isolation level we may encounter.
+consistency~\cite{lbc16,gotsmanpopl16}). All these efforts involve
+proving that atomic and fully isolated operations preserve application
+invariants when executed under these consistency levels.  In contrast,
+we admit weakly-isolated transactions, and our system model accepts
+\emph{specifications} of consistency and isolation levels drawn from
+an expressive logic.  Gotsman \emph{et al.}~\cite{gotsmanpopl16}
+adapts \iso{Parallel Snapshot Isolation} to the aforementioned setting
+by interpreting it as a consistency level that serializes writes to
+objects; a dedicated proof rule is developed to help prove prove
+program invariants hold under this model. By parameterizing our proof
+system over a gamut of weak isolation specifications, we avoid the
+need to define a separate proof rule for each new isolation level we
+may encounter.
 
+\vspace*{-4pt}
 \paragraph{Reasoning under relaxed memory} The reasoning mechanisms
 used to describe and prove properties about weakly-isolated
-transactions bear resemblance to those used to formalize relaxed
+transactions bear some resemblance to those used to formalize relaxed
 memory behaviour~\cite{battycpp}.  Ridge~\cite{rgtso} generalizes
@@
 JMM in terms of memory reorderings, and an operational instantiation
-consistent with the TSO memory model, and demonstrate the equivalence
-of the two definitions.
+consistent with the TSO memory model.

### 2016-07-06 store-consistency.tex

@@
 conventional RDBMS, whose store-specific trace invariant ($\I_s$) is
-the maximul visibility weakening of {\sc sc} w.r.t. $\I_c$, as
+the maximum visibility weakening of {\sc sc} w.r.t. $\I_c$, as
 described in \S\ref{sec:store-consistency}. In concrete terms, this
@@
 generated proof obligations can often be discharged via an SMT solver.
-We nonetheless give full details in the following.
 

### 2016-07-06 related-work.tex
Edits to the related work.

@@
 a complete trace with only committed transactions, making it difficult
-to reason about a program as it builds the trace; their framework also
-requires application invariants be expressed in terms of low-level
-visibility relations.  Cerone \emph{et al.}~\cite{gotsmanconcur15} specify
-isolation levels with atomic visibility, but their specifications are
-also for complete traces, and assumes all transactions execute using
-either {\sc si} or {\sc ser}, which is often not the case in
-practice. While all the aforementioned specification frameworks use
-the vocabulary introduced in~\cite{burckhardt14} to specify replicated
-data types, none of them are equipped with a reasoning framework that
-can use such specifications to verify programs under weak isolation.
+to reason about a program as it builds the trace. Cerone \emph{et
+al.}~\cite{gotsmanconcur15} specify isolation levels with atomic
+visibility, but their specifications are also for complete traces.
+Like ours, all the aforementioned specification frameworks use the
+vocabulary introduced in~\cite{burckhardt14}. However, none of them
+are equipped with a reasoning framework that can use such
+specifications to verify programs under weak isolation.
 
@@
 an expressive logic.  Gotsman \emph{et al.}~\cite{gotsmanpopl16}
-adapts \iso{Parallel Snapshot Isolation} to the aforementioned setting
+adapt \iso{Parallel Snapshot Isolation} to a transaction-less setting
 by interpreting it as a consistency level that serializes writes to

### 2016-07-06 motivation.tex
edits

--- 2016-07-06 motivation.tex
@@
 
-We have thus far assumed a strongly consistent ({\sc sc}) store that,
-in the absence of transactions with special isolation requirements,
-makes the effects of any operation immediately visible to all
-subsequent operations. The natural behaviour of {\sc sc} to totally
-order all operations w.r.t. the $\hbZ$ relation could be in conflict
-with the constraints imposed by weak isolation. To meet the
-requirements of weaker isolation levels, say {\sc rc}, the store has
-to adapt itself to \emph{hide} the effects of concurrent transactions
-until they are committed. Once committed though, effects need to be
-immediately visible to subsequent operations. This semantics can be
-built into the reasoning framework, leading to a proof system
-tailor-made for such stores. However, for the reasoning framework to
-be truly useful, it has to be parametric over different consistency
+Implicit in the above discussion is the assumption of a strongly
+consistent ({\sc sc}) store that guarantees the visibility of all
+previously committed transactions. An {\sc sc} semantics can be built
+into the reasoning framework, leading to a proof system tailor-made
+for such stores.  However, for the reasoning framework to be truly
+useful, it should be capable of handling different consistency
 semantics, supporting stores that are weaker than {\sc sc} (e.g.,
@@
 following sections.
+
+%% We have thus far assumed a strongly consistent ({\sc sc}) store that,
+%% in the absence of transactions with special isolation requirements,
+%% makes the effects of any operation immediately visible to all
+%% subsequent operations. The natural behaviour of {\sc sc} to totally
+%% order all operations w.r.t. the $\hbZ$ relation could be in conflict
+%% with the constraints imposed by weak isolation. To meet the
+%% requirements of weaker isolation levels, say {\sc rc}, the store has
+%% to adapt itself to \emph{hide} the effects of concurrent transactions
+%% until they are committed. Once committed though, effects need to be
+%% immediately visible to subsequent operations. This semantics can be
+%% built into the reasoning framework, leading to a proof system
+%% tailor-made for such stores. However, for the reasoning framework to
+%% be truly useful, it has to be parametric over different consistency
+%% semantics, supporting stores that are weaker than {\sc sc} (e.g.,
+%% causally or eventually consistent), and should be able to reconcile
+%% conflicts between consistency and isolation constraints.  We
+%% demonstrate how our reasoning framework makes this possible in the
+%% following sections.

### 2016-07-06 store-consistency.tex
edits

--- 2016-07-06 store-consistency.tex
@@
 \end{itemize}
-The resultant rely relation ($R_1$) is shown below (note: we
-write $(\E'-\E) \Vdash {\committed(\C{Wd2})}$ to express that \C{Wd2}
-is not committed in $\E$, but committed in $\E'$.):
-\vspace*{-1.5pt}
+The resultant rely relation ($R_1$) is shown below\footnote{We
+  write $(\E'-\E) \Vdash {\committed(\C{Wd2})}$ to express the property
+  that \C{Wd2} is not committed in $\E$, but committed in $\E'$.}:
+
 \begin{smathpar}

### 2016-07-06 motivation.tex

--- 2016-07-06 motivation.tex
@@
 consistent ({\sc sc}) store that guarantees the visibility of all
-previously committed transactions. {\sc sc} semantics can be built
+previously committed transactions. An {\sc sc} semantics can be built
 into the reasoning framework, leading to a proof system tailor-made
 for such stores.  However, for the reasoning framework to be truly
-useful, it has to be parametric over different consistency semantics,
-supporting stores that are weaker than {\sc sc} (e.g., causally or
-eventually consistent), and should be able to reconcile conflicts
-between consistency and isolation constraints.  We demonstrate how our
-reasoning framework makes this possible in the following sections.
+useful, it should be capable of handling different consistency
+semantics, supporting stores that are weaker than {\sc sc} (e.g.,
+causally or eventually consistent), and should be able to reconcile
+conflicts between consistency and isolation constraints.  We
+demonstrate how our reasoning framework makes this possible in the
+following sections.
+
+%% We have thus far assumed a strongly consistent ({\sc sc}) store that,
+%% in the absence of transactions with special isolation requirements,
+%% makes the effects of any operation immediately visible to all
+%% subsequent operations. The natural behaviour of {\sc sc} to totally
+%% order all operations w.r.t. the $\hbZ$ relation could be in conflict
+%% with the constraints imposed by weak isolation. To meet the
+%% requirements of weaker isolation levels, say {\sc rc}, the store has
+%% to adapt itself to \emph{hide} the effects of concurrent transactions
+%% until they are committed. Once committed though, effects need to be
+%% immediately visible to subsequent operations. This semantics can be
+%% built into the reasoning framework, leading to a proof system
+%% tailor-made for such stores. However, for the reasoning framework to
+%% be truly useful, it has to be parametric over different consistency
+%% semantics, supporting stores that are weaker than {\sc sc} (e.g.,
+%% causally or eventually consistent), and should be able to reconcile
+%% conflicts between consistency and isolation constraints.  We
+%% demonstrate how our reasoning framework makes this possible in the
+%% following sections.

### 2016-07-06 introduction.tex
edits

@@
 selection of isolation levels for its transactions preserves its
-invariants when executed on a store equipped with certain consistency
-guarantees (\eg~causal consistency).
-
-The paper makes the following contributions:
-
-\begin{itemize}
+invariants when executed on a store equipped with appropriate consistency
+guarantees.  The paper makes the following contributions:
+\vspace*{-6pt}
+\begin{enumerate}
   \item We develop a semantics for a core language equipped with
@@
     the applicability and utility of our proof methodology.
-\end{itemize}
-
+\end{enumerate}
 \noindent Our results provide the first (to the best of our knowledge)

### 2016-07-06 motivation.tex
Fig. 4 redone.

--- 2016-07-06 motivation.tex
@@
 block defining a transaction scope. Each \C{txn} block is associated
-with an indentifier in angle braces that uniquely identifies the
+with an identifier in angle braces that uniquely identifies the
 transaction. We use Hoare triple notation to annotate programs with
@@
 are guaranteed to hold if both withdraw transactions are serialized,
-making \iso{Serializable} isolation ({\sc ser}) level a sufficent
-condition to preserve invaraints. But, is {\sc ser} necessary?
+making \iso{Serializable} isolation ({\sc ser}) level a sufficient
+condition to preserve invariants. But, is {\sc ser} necessary?
 
@@
 and is thus highly error-prone. Reasoning in terms of how weak
-isolation variants are implementated is no better since it requires
+isolation variants are implemented is no better since it requires
 programmers to understand low-level implementation details of the
@@
 \centering
-  \includegraphics[scale=0.4]{Figures/motiv-eg-1-hb}
+\subcaptionbox {
+  \label{fig:hb-eg-1}
+} [
+  0.25\columnwidth
+] {
+  \includegraphics[scale=0.6]{Figures/hb-eg-1}
+}
+\subcaptionbox {
+  \label{fig:hb-eg-2}
+} [
+  0.25\columnwidth
+] {
+  \includegraphics[scale=0.6]{Figures/hb-eg-2}
+}
+\subcaptionbox {
+  \label{fig:hb-eg-3}
+} [
+  0.25\columnwidth
+] {
+  \includegraphics[scale=0.6]{Figures/hb-eg-3}
+}
+\subcaptionbox {
+  \label{fig:hb-eg-4}
+}{
+  \includegraphics[scale=0.6]{Figures/hb-eg-4}
+}
 
-\caption{\small $\hbZ$ arrows allowed by {\sc si} (solid) and {\sc rc} (solid
-  and dashed) isolation levels for the example in
-  Fig.~\ref{fig:motiv-eg-1}. Each arrow represents an $\hbZ$ relationship
-  among all operations of \C{Wd2} and operations in \C{Wd1} that follow
-  the arrow head. Letters R, W and C stand for read, write and commit,
-  respectively.}
+\caption{\small $\hbZ$ arrows allowed by $\psi_{SI}$
+ and $\psi_{RC}$ specifications
+% (Figs.~\ref{fig:hb-eg-1},~\ref{fig:hb-eg-2},~\ref{fig:hb-eg-3},~\ref{fig:hb-eg-4})
+for the example in Fig.~\ref{fig:motiv-eg-1}. Each
+arrow represents an $\hbZ$ relationship between all operations of
+\C{Wd2} and an operation in \C{Wd1}. Letters R, W and C stand for
+read, write and commit, respectively.}
 \label{fig:motiv-eg-1-hb}
@@
 Fig.~\ref{fig:motiv-eg-1-hb} is a visualization of $\hbZ$ from \C{Wd2}
-to \C{Wd1} allowed by $\psi_{RC}$ and $\psi_{SI}$. All arrows are
-legal under $\psi_{RC}$ because in no case does an operation $\eta_1$
-from \C{Wd1} happen-before $\eta_2$ of \C{Wd2} without the commit of
-\C{Wd1} also happening before $\eta_2$. $\psi_{SI}$ however disallows
-all dashed arrows, because dashed arrows establish $\hbZ$ edges
-between actions in \C{Wd2} and only a subset of the operations in
+to \C{Wd1} allowed by $\psi_{RC}$ and $\psi_{SI}$. Arrows in all
+executions are legal under $\psi_{RC}$ because in no execution does an
+operation $\eta_1$ from \C{Wd1} happen-before $\eta_2$ of \C{Wd2}
+without the commit of \C{Wd1} also happening before $\eta_2$.
+$\psi_{SI}$ however disallows $\hbZ$ arrows (hence executions) in
+Figs.~\ref{fig:hb-eg-2},~\ref{fig:hb-eg-3},~\ref{fig:hb-eg-4} because
+such arrows establish $\hbZ$ edges
+between (operations in) \C{Wd2} and only a subset of the operations in
 \C{Wd1}. Since the result of an operation (\eg a read) depends on
@@
 relation ultimately dictates the final result of the program. The
-solid arrow and the first (from left) dashed arrow in
-Fig.~\ref{fig:motiv-eg-1-hb} denote $\hbZ$ relationships that do not
-affect the value of \C{B} in a way that causes the program to violate
-its post-condition. The last two dashed arrows denote $\hbZ$
+arrows in Figs.~\ref{fig:hb-eg-1} and~\ref{fig:hb-eg-2} denote
+$\hbZ$ relationships that do not affect the value of \C{B} in a way
+that causes the program to violate its post-condition. In contrast,
+arrows in Figs.~\ref{hb-eg-3} and~\ref{hb-eg-4} denote $\hbZ$
 relationships that lead to the violation of the post-condition. The

### 2016-07-06 motivation.tex
Missing fig refs fixed.
Typo.

@@
 $\psi_{SI}$ however disallows $\hbZ$ arrows (hence executions) in
-Figs.~\ref{fig:hb-eg-2},~\ref{fig:hb-eg-3},~\ref{fig:hb-eg-4} because
-such arrows establish $\hbZ$ edges
-between (operations in) \C{Wd2} and only a subset of the operations in
-\C{Wd1}. Since the result of an operation (\eg a read) depends on
-what happens before that operation, the structure of the $\hbZ$
-relation ultimately dictates the final result of the program. The
-arrows in Figs.~\ref{fig:hb-eg-1} and~\ref{fig:hb-eg-2} denote
-$\hbZ$ relationships that do not affect the value of \C{B} in a way
-that causes the program to violate its post-condition. In contrast,
-arrows in Figs.~\ref{hb-eg-3} and~\ref{hb-eg-4} denote $\hbZ$
+Figs.~\ref{fig:hb-eg-2},~\ref{fig:hb-eg-3}, and~\ref{fig:hb-eg-4}
+because such arrows establish $\hbZ$ edges between (operations in)
+\C{Wd2} and only a subset of the operations in \C{Wd1}. Since the
+result of an operation (\eg a read) depends on what happens before
+that operation, the structure of the $\hbZ$ relation ultimately
+dictates the final result of the program. The arrows in
+Figs.~\ref{fig:hb-eg-1} and~\ref{fig:hb-eg-2} denote $\hbZ$
+relationships that do not affect the value of \C{B} in a way that
+causes the program to violate its post-condition. In contrast, arrows
+in Figs.~\ref{fig:hb-eg-3} and~\ref{fig:hb-eg-4} denote $\hbZ$
 relationships that lead to the violation of the post-condition. The

### 2016-07-06 motivation.tex
Shrunk Fig. 4.
Minor edit.

@@
 ] {
-  \includegraphics[scale=0.6]{Figures/hb-eg-1}
+  \includegraphics[scale=0.5]{Figures/hb-eg-1}
 }
@@
 ] {
-  \includegraphics[scale=0.6]{Figures/hb-eg-2}
+  \includegraphics[scale=0.5]{Figures/hb-eg-2}
 }
@@
 ] {
-  \includegraphics[scale=0.6]{Figures/hb-eg-3}
+  \includegraphics[scale=0.5]{Figures/hb-eg-3}
 }
@@
 }{
-  \includegraphics[scale=0.6]{Figures/hb-eg-4}
+  \includegraphics[scale=0.5]{Figures/hb-eg-4}
 }
@@
 relationships that lead to the violation of the post-condition. The
-task of the reasoning framework is to determine if all $\hbZ$
+task of the reasoning framework is to determine if \emph{all} $\hbZ$
 relationships allowed by an isolation level lead to the satisfaction

### 2016-07-06 case-studies.tex
Source code url added.
edits
Reverted to non-anonymous code.

@@
-\vspace*{-4pt}
 \section{Case Studies}
@@
 support.\footnote{Details can be found in the supplementary material.
-  Source code for our implementation, along with applications and
-  experiments, are available at the following anonymized url:
-  \url{...}} The focus of our formal verification efforts are such
-translated \txnimp programs.
+Source code of our tools, applications and experiments, is also
+provided in the (non-anonymous) supplementary.} The focus of our
+formal verification efforts are such translated \txnimp programs.
 

### 2016-07-07 reasoning.tex
Footnote edit for KC.

@@
 important to keep in mind the distinction between high-level invariants
-(\eg~\C{X$\ge$0}) and trace invariants (\eg~\C{RC}). The former
+(\eg~\C{X$\ge$0}) and trace invariants (\eg~$\I =
+\lambda\E.\,\C{RC}(\C{Wd1})\wedge \C{SI}(\C{Wd2})$). The former
 constitute \emph{proof} \emph{obligations} for the programmer, whereas
-the latter captures \emph{assumptions} about the operational characteristics
-of transactions and the underlying data store. } 
+the latter is a combination of transaction-specific \emph{assumptions} about 
+operational characteristics of the underlying data store. } 
 Since the trace invariant captures transaction-isolation constraints,

### 2016-07-07 motivation.tex
Fixed Figs. 2 and 4.

@@
 \caption{\small A possible execution of the program shown in Fig.~\ref{fig:motiv-eg-1} under
-  \iso{Read Committed} isolation level.}
+  \iso{Read Committed} isolation level. Transaction \C{Wd1} is shown
+  against lighter green background, and transaction \C{Wd2} against
+  darker red background. Each transaction reads the balance (\C{B})
+  twice, hence two \C{Reads}.}
 \label{fig:rc-ex}
@@
 The figure depicts an execution as a series of read, write, and commit
-operations. The subscript of an operation indicates the transaction
-that executes it (Both transactions read \C{B} twice, hence two read
-operations). In the execution on left, transaction \C{Wd1} reads the
-current balance (\C{k}) and writes the new balance (\C{k-a1}), but
-before it commits, transaction \C{Wd2} executes and commits, writing
-the new balance (\C{k-a2}). {\sc rc} isolation prevent \C{Wd2} from
-witnessing the uncommitted writes of transaction \C{Wd1}.
-Subsequently committing \C{Wd1} leads to the loss of \C{Wd2}'s updates
-(the so-called \emph{lost update anomaly}~\cite{berenson}), resulting
-in an incorrect balance of \C{k-a1}. The execution on right describes
-a similar scenario with \C{Wd1} and \C{Wd2} exchanging their roles.
-Clearly, {\sc rc} is an excessively weak isolation level for this
-program because it loses the updates of one transaction, resulting in
-the violation of the post-condition.  A stronger isolation level that
-prevents lost updates is required. \iso{Snapshot Isolation} ({\sc
-si})~\cite{berenson} fits this requirement; {\sc si} effectively
-serializes transactions that update a shared data object by aborting
-and re-executing a transaction if write-write conflicts are detected
-during its commit.\footnote{{\sc si} however does not serialize
-transactions in the absence of write-write conflicts.} Since {\sc si},
-unlike {\sc ser}, does not need expensive mechanisms such as
-lock-based concurrency control or runtime monitoring, it is also more
-efficient, making it appropriate for both withdraw transactions.
+operations. In the execution on left, transaction \C{Wd1} (green)
+reads the current balance (\C{k}) and writes the new balance
+(\C{k-a1}), but before it commits, transaction \C{Wd2} (red) executes
+and commits, writing the new balance (\C{k-a2}). {\sc rc} isolation
+prevent \C{Wd2} from witnessing the uncommitted writes of transaction
+\C{Wd1}.  Subsequently committing \C{Wd1} leads to the loss of
+\C{Wd2}'s updates (the so-called \emph{lost update
+anomaly}~\cite{berenson}), resulting in an incorrect balance of
+\C{k-a1}. The execution on right describes a similar scenario with
+\C{Wd1} and \C{Wd2} exchanging their roles.  Clearly, {\sc rc} is an
+excessively weak isolation level for this program because it loses the
+updates of one transaction, resulting in the violation of the
+post-condition.  A stronger isolation level that prevents lost updates
+is required. \iso{Snapshot Isolation} ({\sc si})~\cite{berenson} fits
+this requirement; {\sc si} effectively serializes transactions that
+update a shared data object by aborting and re-executing a transaction
+if write-write conflicts are detected during its commit.\footnote{{\sc
+si} however does not serialize transactions in the absence of
+write-write conflicts.} Since {\sc si}, unlike {\sc ser}, does not
+need expensive mechanisms such as lock-based concurrency control or
+runtime monitoring, it is also more efficient, making it appropriate
+for both withdraw transactions.
 
@@
 
-\caption{\small $\hbZ$ arrows allowed by $\psi_{SI}$
- and $\psi_{RC}$ specifications
-% (Figs.~\ref{fig:hb-eg-1},~\ref{fig:hb-eg-2},~\ref{fig:hb-eg-3},~\ref{fig:hb-eg-4})
-for the example in Fig.~\ref{fig:motiv-eg-1}. Each
-arrow represents an $\hbZ$ relationship between all operations of
-\C{Wd2} and an operation in \C{Wd1}. Letters R, W and C stand for
-read, write and commit, respectively.}
+\caption{\small For the example in Fig.~\ref{fig:motiv-eg-1},
+$\psi_{SI}$ specification allows $\hbZ$ arrows (hence the execution)
+in Fig.~\ref{fig:hb-eg-1}. $\psi_{RC}$ specification allows
+Figs.~\ref{fig:hb-eg-1},~\ref{fig:hb-eg-2},~\ref{fig:hb-eg-3}, and
+~\ref{fig:hb-eg-4}. Each arrow represents an $\hbZ$ relationship
+between all operations of \C{Wd2} and an operation in \C{Wd1}. Letters
+R, W and C stand for read, write and commit, respectively.}
 \label{fig:motiv-eg-1-hb}

### 2016-07-07 motivation.tex
Edit.
Edits.

@@
 
-\caption{\small For the example in Fig.~\ref{fig:motiv-eg-1},
-$\psi_{SI}$ specification allows $\hbZ$ arrows (hence the execution)
-in Fig.~\ref{fig:hb-eg-1}. $\psi_{RC}$ specification allows
+\caption{\small Possible executions for the example in
+Fig.~\ref{fig:motiv-eg-1}. Letters R, W and C stand for read, write
+and commit, respectively. Each blue dashed arrow represents a $\hbZ$
+relationship between all operations of \C{Wd2} and an operation in
+\C{Wd1}. $\psi_{SI}$ specification allows $\hbZ$ arrows (hence the
+execution) in Fig.~\ref{fig:hb-eg-1}. $\psi_{RC}$ specification allows
+$\hbZ$ arrows in
 Figs.~\ref{fig:hb-eg-1},~\ref{fig:hb-eg-2},~\ref{fig:hb-eg-3}, and
-~\ref{fig:hb-eg-4}. Each arrow represents an $\hbZ$ relationship
-between all operations of \C{Wd2} and an operation in \C{Wd1}. Letters
-R, W and C stand for read, write and commit, respectively.}
+~\ref{fig:hb-eg-4}.  }
 \label{fig:motiv-eg-1-hb}

### 2016-07-07 motivation.tex
A simplification.

@@
 executions are legal under $\psi_{RC}$ because in no execution does an
-operation $\eta_1$ from \C{Wd1} happen-before $\eta_2$ of \C{Wd2}
-without the commit of \C{Wd1} also happening before $\eta_2$.
+operation from \C{Wd2} happen-before an operation of \C{Wd1} without
+the commit of \C{Wd2} also happening before the operation of \C{Wd1}.
 $\psi_{SI}$ however disallows $\hbZ$ arrows (hence executions) in

### 2017-06-28 motivation.tex
Fig. 4 shrunk to fit 12 pages.
Format changed to PACMPL.

@@
 In this section, we motivate our ideas via an example written in
-\txnimp - a C-like imperative language equipped with a \C{txn} lexical
-block defining a transaction scope. Each \C{txn} block is associated
-with an identifier in angle braces that uniquely identifies the
-transaction. We use Hoare triple notation to annotate programs with
-pre- and post- conditions.
+\txnimp - a functional language equipped with a \C{DB} monad to
+compose database computations. Besides the usual \C{bind} and
+\C{return} combinators, monad offers a \C{txn\_do} combinator that
+executes a database computation and returns the result.
 
@@
 \centering
-$\{\texttt{B}=\texttt{k} \conj \texttt{k}\ge\texttt{a1+a2}\}$
 \begin{tabular}{l||l}
-\begin{txnimpcode}
-  txn$\langle$'Wd1'$\rangle${
-    if (B $\ge$ a1) {
-      B := B - a1
-    }
-  }
-\end{txnimpcode}
+\begin{ocaml}
+let new_order_txn d_id c_d_id c_id
+      (ireqs: item_req list) = atomically do
+  dist <- SQL.select1 District
+               (fun d -> d.d_id = d_id);
+  let o_id = dist.d_next_o_id;
+  SQL.update District 
+    (fun d -> 
+      {d with d_next_o_id =d_next_o_id + 1})
+    (fun d -> d.d_id = d_id );
+  SQL.insert Order {o_id=o_id;  o_d_id=d_id; 
+                    o_c_id=c_id; o_ol_cnt=S.size ireqs; };
+  SQL.insert New_order {no_o_id=o_id; no_d_id=d_id};
+  foreach ireqs @@ fun ireq -> do
+    stk <- SQL.select1 Stock
+             (fun s -> s.s_i_id = ireq.ol_i_id);
+    let s_qty' = if stk.s_qty >= ireq.ol_qty + 10 
+                then stk.s_qty - ireq.ol_qty
+                else stk.s_qty - ireq.ol_qty + 91;
+    SQL.update Stock
+       (fun s -> {s with s_qty = s_qty'})
+       (fun s -> s.s_i_id = ireq.ol_i_id);
+    SQL.insert Order_line {ol_o_id=o_id; ol_d_id=d_id; 
+        ol_i_id=ireq.ol_i_id; ol_qty=ireq.ol_qty}
+
+\end{ocaml}
 &
-\begin{txnimpcode}
+\begin{ocaml}
   txn$\langle$'Wd2'$\rangle${
@@
   }
-\end{txnimpcode}
+\end{ocaml}
 \\
 \end{tabular}
-$\{\texttt{B}=\texttt{k-a1-a2}\}$
 \caption{\small Concurrent withdraw transactions}
@@
 ] {
-  \includegraphics[scale=0.5]{Figures/hb-eg-1}
+  \includegraphics[scale=0.45]{Figures/hb-eg-1}
 }
@@
 ] {
-  \includegraphics[scale=0.5]{Figures/hb-eg-2}
+  \includegraphics[scale=0.45]{Figures/hb-eg-2}
 }
@@
 ] {
-  \includegraphics[scale=0.5]{Figures/hb-eg-3}
+  \includegraphics[scale=0.45]{Figures/hb-eg-3}
 }
@@
 }{
-  \includegraphics[scale=0.5]{Figures/hb-eg-4}
+  \includegraphics[scale=0.45]{Figures/hb-eg-4}
 }

### 2017-06-28 motivation.tex
Example formatting.
Motivation Edited.

@@
 compose database computations. Besides the usual \C{bind} and
-\C{return} combinators, monad offers a \C{txn\_do} combinator that
+\C{return} combinators, monad offers an \C{atomically} combinator that
 executes a database computation and returns the result.
@@
 \centering
-\begin{tabular}{l||l}
 \begin{ocaml}
-let new_order_txn d_id c_d_id c_id
-      (ireqs: item_req list) = atomically do
-  dist <- SQL.select1 District
-               (fun d -> d.d_id = d_id);
+let new_order_ d_id c_id item_reqs = atomically do
+  dist <- SQL.select1 District (fun d -> d.d_id = d_id);
   let o_id = dist.d_next_o_id;
-  SQL.update District 
-    (fun d -> 
-      {d with d_next_o_id =d_next_o_id + 1})
+  SQL.update District (fun d -> {d with d_next_o_id =d_next_o_id + 1})
                       (fun d -> d.d_id = d_id );
   SQL.insert Order {o_id=o_id;  o_d_id=d_id; 
-                    o_c_id=c_id; o_ol_cnt=S.size ireqs; };
-  SQL.insert New_order {no_o_id=o_id; no_d_id=d_id};
-  foreach ireqs @@ fun ireq -> do
-    stk <- SQL.select1 Stock
-             (fun s -> s.s_i_id = ireq.ol_i_id);
-    let s_qty' = if stk.s_qty >= ireq.ol_qty + 10 
-                then stk.s_qty - ireq.ol_qty
-                else stk.s_qty - ireq.ol_qty + 91;
-    SQL.update Stock
-       (fun s -> {s with s_qty = s_qty'})
-       (fun s -> s.s_i_id = ireq.ol_i_id);
+                    o_c_id=c_id; o_ol_cnt=S.size item_reqs; };
+  foreach item_reqs @@ fun item_req -> do
+    stk <- SQL.select1 Stock (fun s -> s.s_i_id = item_req.ol_i_id);
+    let s_qty' = if stk.s_qty >= item_req.ol_qty + 10 
+                then stk.s_qty - item_req.ol_qty 
+                else stk.s_qty - item_req.ol_qty + 91;
+    SQL.update Stock (fun s -> {s with s_qty = s_qty'}) 
+                     (fun s -> s.s_i_id = item_req.ol_i_id);
     SQL.insert Order_line {ol_o_id=o_id; ol_d_id=d_id; 
-        ol_i_id=ireq.ol_i_id; ol_qty=ireq.ol_qty}
+                           ol_i_id=item_req.ol_i_id; ol_qty=item_req.ol_qty}
 
 \end{ocaml}
-&
-\begin{ocaml}
-  txn$\langle$'Wd2'$\rangle${
-    if (B $\ge$ a2) {
-      B := B - a2
-    }
-  }
-\end{ocaml}
-\\
-\end{tabular}
 \caption{\small Concurrent withdraw transactions}
-\label{fig:motiv-eg-1}
+\label{fig:new_order_code}
 \vspace*{-10pt}
@@
 
-Consider an implementation of a banking application that admits
+Fig.~\ref{fig:new_order_code} shows a simplified version of the TPC-C
+\C{new\_order} transaction written in \txnimp (the \C{do} syntax ala
+Haskell is used for convenience). TPC-C is an Online Transaction
+Processing (OLTP) benchmark that models an order-processing system for
+a wholesale parts supply business. The business logic is captured in 5
+database transactions that operate on 9 tables. \C{new\_order} is one
+such trasaction that operates on \C{District}, \C{Order},
+{New\_order}, \C{Stock}, and \C{Order\_line} tables. The transaction
+acts on the behalf of a customer, whose id is \C{c\_id}, to place a
+new order for the given set of items (\C{item\_reqs}), to be served by
+a warehouse under the district identified by \C{d\_id}. The
+transaction does so by invoking appropriate SQL functionality,
+captured in Fig.~\ref{fig:new_order_code} by various calls to
+functions under the \C{SQL} module. All \C{SQL} functions take the
+table name (a nullary constructor) as their first argument. The
+higher-order \C{SQL.select1} function accepts a boolean function that
+describes the selection criteria, and returns any record that meets
+the criteria (it models the SQL query \C{SELECT \ldots\xspace LIMIT
+1}). \C{SQL.update} also accepts a boolean function (3rd argument) to
+select the records to be updated. Its 2nd argument is a function that
+maps each selected record to a new (updated) record. \C{SQL.insert}
+inserts a given record into the specified table in the database.
+
+The \C{new\_order} transaction inserts a new \C{Order} record, whose
+id is the sequence number of the next order under the given district
+(\C{d\_id}). The sequence number is stored in the corresponding
+\C{District} record, and updated each time a new order added to the
+system. Since each order may request multiple items (\C{item\_reqs}),
+an \C{Order\_line} record is created for each requested item to relate
+the order with the item. Each item has a corresponding record in the
+\C{Stock} table, which keeps track of the quantity of the item left in
+stock (\C{s\_qty}). The quantity is updated by the transaction to
+reflect the processing of new order (if the stock quantity falls below
+10, it is automatically replenished by 91).
+
+TPC-C defines multiple invariants, called \emph{consistency
+conditions}, over the state of the application in the database. One
+such consistency condition is the requirement that for a given order
+\C{o}, the order-line-count field (\C{o.o\_ol\_cnt}) should reflect
+the number of order lines under the order, i.e., the number of
+\C{Order\_line} records whose \C{ol\_o\_id} is equal to \C{o.o\_id}.
+In a sequential exectution, it is more-or-less easy to see how this
+condition condition is preserved by the \C{new\_order} transaction by
+examining Fig.~\ref{fig:new_order_code}. A new \C{Order} record is
+added with its \C{o\_id} distinct from the existing order ids, and its
+\C{o\_ol\_cnt} set equal to the size of \C{item\_reqs} set. The
+\C{foreach} loop runs once for each \C{item\_req}, adding a new
+\C{Order\_line} record for each requested item, with its \C{ol\_o\_id}
+field set to \C{o\_id}. Thus, by the end of the loop, the number of
+\C{Order\_line} records in the database, whose \C{ol\_o\_id} is equal
+to \C{o\_id}, is equal to the size of the \C{item\_req} set, which
+inturn is equal to the \C{Order} record's \C{o\_ol\_cnt} field. Thus
+the consistency condition is preserved.
+
+While the aforementioned reasoning is simple to perform manually, the
+sequential verification of database programs, such as the one in
+Fig.~\ref{fig:new_order_code}, is difficult to automate owing to the
+presence of loops and collections. Thus verifying serializable
+transactions is at least as difficult as reasoning about sequential
+imperative programs. Nonetheless, program logics and automation
+techniques developed to reason about imperative programs can be used
+to reason about sequential transactions, perhaps with reasonable
+success.
+
 concurrent withdraw transactions on an account balance (\C{B}), as

### 2017-06-29 motivation.tex
Edits.

@@
 Fig.~\ref{fig:new_order_code}, is difficult to automate owing to the
-presence of loops and collections. Thus verifying serializable
-transactions is at least as difficult as reasoning about sequential
-imperative programs. Nonetheless, program logics and automation
-techniques developed to reason about imperative programs can be used
-to reason about sequential transactions, perhaps with reasonable
-success.
+presence of loops and collections. Thus verification of serializable
+transactions is comparable to the verification of sequential
+(imperative) programs in terms of difficulty. It is therefore not
+unreasonable to expect the reasoning machinery (i.e., program logics
+and inference techniques) developed for imperative programs to be
+equally effective on serializable transactions.
 
-concurrent withdraw transactions on an account balance (\C{B}), as
-shown in Fig.~\ref{fig:motiv-eg-1}. If the initial balance (\C{k}) in
-the account is enough to perform both withdraws, then the final
-balance, after both transactions commit, is expected to reflect the
-effects of both withdraws. The pre- and post-conditions in
-Fig.~\ref{fig:motiv-eg-1} reflect this expectation. Indeed, invariants
-are guaranteed to hold if both withdraw transactions are serialized,
-making \iso{Serializable} isolation ({\sc ser}) level a sufficient
-condition to preserve invariants. But, is {\sc ser} necessary?
+Unfortunately, transactions are rarely serializable in practice, and
+presence of weak isolation significantly complicates the reasoning
+about database programs. Weak isolation allows a transaction to
+witness the effects of concurrent transactions during its
+execution\footnote{Weak isolation doesn't violate atomicity as long as
+the witnessed effects are those of the committed transactions}.
+Various levels of weak isolation differ in what effects each allows
+the transaction to witness, and when. Admitting interference from
+concurrent transactions may violate the invariants that would
+otherwise hold under a fully isolated execution.
 
-As an alternative, consider the execution of this transaction under a
-\emph{Read Committed} ({\sc rc}) isolation level, which is weaker than
-{\sc ser}.\footnote{{\sc rc} is in fact the default isolation level in
-Postgres 9.5 and Oracle 11g databases.} An {\sc rc} transaction is
-isolated from the writes of uncommitted transactions, and is therefore
-free from \emph{dirty reads}~\cite{berenson} of uncommitted data. In
-the current example, {\sc rc} isolation admits the two executions
-shown in Fig.~\ref{fig:rc-ex} on a strongly consistent ({\sc sc})
-store, such as a conventional RDBMS.
+Let us reconsider the \C{new\_order} transaction, which has been
+previously shown to preserve TPC-C's consistency condition under a
+serializable execution. However, serializability is rarely the default
+isolation level on databases. For instance, the default on Postgres is
+Read Committed (RC), which is considerably weaker than
+serializability. An RC transaction is isolated from \emph{dirty
+writes}, i.e., writes of uncommitted transactions, but is allowed to
+witness the writes of concurrent transactions as soon as they are
+committed. Thus, with two concurrent instances of the \C{new\_order}
+transaction (call them $T_1$ and $T_2$) trying to place new orders for
+different customers under the same district (\C{d\_id}), executions
+shown in Fig.~\ref{fig:new_order_execs} are possible on Postgres.
 
@@
 The figure depicts an execution as a series of read, write, and commit
-operations. In the execution on left, transaction \C{Wd1} (green)
-reads the current balance (\C{k}) and writes the new balance
+operations. In the execution on left, the \C{new\_order} instance
+$T_1$ (green) reads the \C{d\_next\_o\_id} field of the district
+record for \C{d\_id}, but before it increments the field, another
+\C{new\_order} instance ($T_2$) begins its execution and commits. Note
+that $T_2$ reads the same \C{d\_next\_o\_id} value as $T_1$, and
+inserts new \C{Order} and \C{Order\_line} records with their \C{o\_id}
+and \C{ol\_o\_id} fields (resp.) equal to \C{d\_next\_o\_id}. $T_2$
+also increments the \C{d\_next\_o\_id} field, which $T_1$ has already
+acccessed. This is allowed because reads do not obtain a mutually
+exclusive lock on most databases, including Postgres. Post $T_2$'s
+commit, $T_1$ resumes execution and adds new \C{Order} and
+\C{Order\_line} fields with the same order id as $T_1$. Thus, by the
+end of the execution, \C{Order\_line} records inserted by $T_1$ and
+$T_2$ all bear the same order id. There are also two \C{Order} records
+with the same district id (\C{d\_id}) and order id, none of whose
+\C{o\_ol\_cnt} reflects the actual number of \C{Order\_line} records
+inserted with that order id. This clearly violates TPC-C's consistency
+condition.
+
+Note that, in the above example, there is no Write-Write conflict, or
+the \emph{lost update} anomaly that often characterizes concurrency
+bugs, such as illegal concurrent withdraws on a bank
+account~\cite{poloniexbug}. While $T_1$ and $T_2$ both increment the
+\C{d\_next\_o\_id} field of the district record, they do so atomically
+(Line 4 of Fig.~\ref{fig:new_order_code}), allowing both updates to be
+present in the final state. Likewise, four other well-known anomalies
+that characterize RC isolation~\cite{berenson}, namely, \emph{fuzzy
+reads}, \emph{phantom reads}, \emph{read skew}, and \emph{write skew},
+are also not exhibited by the example. Thus, any program analysis,
+static or dynamic, that aims to determine appropriate isolation by
+checking for possible manifestations of known anomalies fails to
+identify grounds for promoting the isolation level of \C{new\_order}.
+Yet, if we take the semantics of the application into account, it is
+clear that RC is not an appropriate isolation level for
+\C{new\_order}. Clearly, there is a need for a high-level verification
+framework that lets us reason about transaction semantics under weak
+isolation.
+
+\newpage
+
 (\C{k-a1}), but before it commits, transaction \C{Wd2} (red) executes

### 2017-06-29 motivation.tex
Edits.

@@
 \C{new\_order}. Clearly, there is a need for a high-level verification
-framework that lets us reason about transaction semantics under weak
-isolation.
+framework that lets us reason about transaction semantics in
+conjunction with the semantics of weak isolation.
+
+Facilitating high-level reasoning about weak isolation is however not
+easy owing to the low-level nature of its specification, either in
+terms of the implementation details~\cite{gray}, allowed
+anomalies~\cite{berenson}, or well-formedness condition on program
+traces~\cite{adyaphd,gotsmanconcur15}. For instance, does promoting
+the isolation level of \C{new\_order} to Repeatable Read (RR) help in
+preserving the consistency condition? To answer this question, one has
+to relate the low-level specification of RR given in terms of
+trace-based relations~\cite{adyaphd,gotsmanconcur15} to the semantics
+of TPC-C, which is difficult. Alternatively, one could reason about
+how RR is implemented, say on Postgres, and reason about the
+\C{new\_order} implementation in conjunction with Postgres's RR
+implementation. However, this is not any easier, considering that
+Postgres employs a host of techniques, including Multiversion, locks
+and rollbacks, to implement RR. The verification problem becomes even
+more difficult if the aim is to facilitate automation. Our experience
+suggests that a proof framework that requires programmers to relate
+trace-based isolation specifications to program semantics necessarily
+imposes heavy annotation burden, requires painstakingly long and
+verbose proofs, and does not lend itself to automation. Weak isolation
+aside, there are also the conventional challenges that face
+(automatic) verification of concurrent programs: facilitating modular
+proofs, inferring intermediary assertions and loop invariants that are
+stable w.r.t interference, etc. A proof system for weakly isolated
+transactions has to necessarily address these challenges, along with
+those that arise due to weak isolation.
+
+The proof system and the automation framework proposed in this paper
+overcomes or circumvents the aforementioned problems. The first
+observation that underlies our approach is that, while databases
+implement weak isolation employing a host of techniques to facilitate
+maximum concurrency despite the isolation, the operational semantics
+of such implementations can nonetheless be captured using a simple
+abstraction of the database state exposed to the application. For
+instance, consider how Postgres executes an RR transaction. A
+(conceptual) snapshot of the database state is created at the
+beginning of the transaction against which its reads are executed.
+When the transaction attempts to update a record, Postgres first
+checks if the current version of the record in the database is same as
+the snapshot version. If yes, then an exclusive lock is obtained on
+the record, and the updatation is carried out. If the record is
+already locked, Postgres waits (blocks the transaction) until the lock
+is released. Once obtained, an exclusive lock is held until the
+transaction commits or rolls back. If the aforementioned version check
+fails (i.e., database has a later version than the snapshot), then the
+transaction is rolled back and re-executed. Thus, the implementation
+of RR on Postgres is fairly involved. Nonetheless, its operational
+semantics can be described straightforwardly in terms of the legal
+transitions of the database state (denoted, $\Delta \longrightarrow
+\Delta'$) allowed while an RR transaction is in progress. Since
+transaction's reads are always served from a snapshot, no state
+changes are witnessed while the execution is in progress. Thus,
+insofar as an RR transaction is concerned, the database state doesn't
+change (i.e., $\Delta' = \Delta$) during the execution. Note that the
+uncommitted writes of the transaction are not part of $\Delta$;
+instead we imagine a transaction-local state ($\delta$) that is
+populated with only the uncommitted local writes. When the transaction
+commits, the local state ($\delta$) is flushed to the global state
+($\Delta$) to yeild a committed state. However, unlike reads, commit
+is performed against the current state of the database (call it
+$\Delta_c$); not the snapshot. Thus, after the transaction finishes
+execution, but before it commits, the transaction witnesses the
+effects of all concurrent transactions in a single database state
+transition $\Delta \longrightarrow \Delta_c$. Postgres's RR
+implementation effectively constrains this transition. Firstly, due to
+the version check, the current transaction cannot update a data item
+that was already updated by a concurrent transaction. Secondly, due to
+exclusive write locks, a data item updated by the current transaction
+cannot be overwritten by a concurrent transaction. Putting these two
+together, we know that the transition $\Delta \longrightarrow
+\Delta_c$ does not involve any concurrent transactions that write to
+the same data items as the current RR transaction. Since $\delta$
+records the writes of the current transaction, intuitively, forall
+$x\in\delta$, $\Delta_c(x) = \Delta(x)$. To summarize, the operational
+semantics of Postgres's RR implementation can be captured as
+constraints on the transitions of the database state ($\Delta
+\longrightarrow \Delta'$) during the lifetime an RR
+transaction ($T$):
+\begin{itemize}
+  \item While $T$ executes, $\Delta' = \Delta$.
+  \item After $T$ finishes execution, but before it commits its local
+    state $\delta$, $\forall(x\in\delta).~\Delta'(x) = \Delta(x)$.
+\end{itemize}
 

### 2017-06-29 motivation.tex
Edits.

@@
 
+Let us now see how the above simple characterization of Postgres's RR
+lets us verify \C{new\_order}. Firstly, since the database doesn't
+change ($\Delta' = \Delta$) during the execution of the transaction's
+body, we can reason about \C{new\_order} as if it is executing in
+complete isolation until it is ready to commit. This reasoning is same
+as the reasoning about sequential execution we performed earlier,
+except that it now also includes a local state ($\delta$) that stores
+local writes. Thus, when \C{new\_order} finishes execution and becomes
+ready to commit, immediately flushing its local writes ($\delta$) to
+the unchanged database state ($\Delta$) preserves the TPC-C invariant.
+However, we are required to consider the  interference from concurrent
+transitions at this point, which might change the database state from
+$\Delta$ to $\Delta'$. If this interference includes the effects of a
+concurrent \C{new\_order} transaction (with the same \C{d\_id}), then
+verification fails as described previously
+(Fig.~\ref{fig:new_order_execs}). Fortunately, we can show that this
+is impossible; RR pre-empts a concurrent \C{new\_order} transaction
+that modifies the same \C{District} record as the current transaction
+(concretely, since the record is already present in current
+transaction's $\delta$, transition from $\Delta$ to $\Delta'$ cannot
+change this record). Thus \C{new\_order} preserves the TPC-C
+invariant when executed under RR isolation on Postgres. Our proof
+framework generalizes this style of reasoning to various isolation
+levels on databases.
+
+The second observation that informs our approach is one that pertains
+to automation.
+
 \newpage

### 2017-06-29 motivation.tex
Edits.

@@
 The second observation that informs our approach is one that pertains
-to automation.
+to automation. Program verification, even when machine-aided, often
+entails significant annotation burden in the form of intermediary
+assertions and loop invariants required to prove a program correct.
+Even with concurrent program logics, such as Rely-Gurantee, which
+extend Hoare logic with additional artifacts, (stable) intermediary
+assertions and loop invariants remain a major source of annotation
+burden. Since database transactions inherently deal with the mutable
+state of the database, it is natural to formulate them as concurrent
+imperative programs. An advantage with this formulation is that it
+lets us immediately apply existing concurrent program logics to
+database transactions, albeit with necessary customizations to deal
+with weak isolation. The downside, however, is that there hardly is
+any automation support to apply such program logics on concurrent
+imperative programs. Attemps at automation face the problem of
+inferring stable and inductive loop invariants in its full generality,
+which, while being undecidable in theory, and hard in practice, also
+takes our focus away from weak isolation.
 
-\newpage
+In the light of aforementioned reservations, we refrain from
+formulating database transactions as concurrent imperative programs.
+Instead, we see value in viewing them as essentially functional
+programs that manage the database state via a monad akin to the
+\C{State} monad~\cite{statemonad}. We find it useful to reason about
+statements that mutate the database state, not in terms of a pre- and
+post-condition pair, but in terms of a state transformer that relates
+the pre- and post-states of a statement. Loops are substituted with
+higher-order combinators that have a well-defined control flow, and
+facilitate well-founded recursion. The static semantics of these
+combinators let us lift the state transformer of its higher-order
+argument (i.e., the loop body) to the state transformer of the
+combined expression (i.e., the loop). This is made possible due to the
+characteristic structure of database programs, which also gets
+reflected in the structure of the state transformer. We illustrate
+this point through a simple example that doesn't involve transactions,
+but is nonetheless relevant.
 
-(\C{k-a1}), but before it commits, transaction \C{Wd2} (red) executes
-and commits, writing the new balance (\C{k-a2}). {\sc rc} isolation
-prevent \C{Wd2} from witnessing the uncommitted writes of transaction
-\C{Wd1}.  Subsequently committing \C{Wd1} leads to the loss of
-\C{Wd2}'s updates (the so-called \emph{lost update
-anomaly}~\cite{berenson}), resulting in an incorrect balance of
-\C{k-a1}. The execution on right describes a similar scenario with
-\C{Wd1} and \C{Wd2} exchanging their roles.  Clearly, {\sc rc} is an
-excessively weak isolation level for this program because it loses the
-updates of one transaction, resulting in the violation of the
-post-condition.  A stronger isolation level that prevents lost updates
-is required. \iso{Snapshot Isolation} ({\sc si})~\cite{berenson} fits
-this requirement; {\sc si} effectively serializes transactions that
-update a shared data object by aborting and re-executing a transaction
-if write-write conflicts are detected during its commit.\footnote{{\sc
-si} however does not serialize transactions in the absence of
-write-write conflicts.} Since {\sc si}, unlike {\sc ser}, does not
-need expensive mechanisms such as lock-based concurrency control or
-runtime monitoring, it is also more efficient, making it appropriate
-for both withdraw transactions.
-
-Thinking in terms of anomalies, as described above, is how database
-programmers are often encouraged to reason about weak isolation.
-Unfortunately, such reasoning does not rest on any sound foundation,
-and is thus highly error-prone. Reasoning in terms of how weak
-isolation variants are implemented is no better since it requires
-programmers to understand low-level implementation details of the
-database that are far removed from the application semantics. An
-attractive  alternative in this context would be a principled
-reasoning approach that combines declarative reasoning about isolation
-guarantees with operational reasoning about programs. We demonstrate
-how our proof system makes this possible in the context of the current
-example.
-
-First, we note that the example in Fig.~\ref{fig:motiv-eg-1} is a
-concurrent program, and hence is amenable to \emph{rely-guarantee}
-style reasoning~\cite{rgjones}, a compositional proof technique that
-allows us to reason about the behaviour of individual threads by
-abstracting away interferences induced by other threads (collectively
-called \emph{the environment}) into a \emph{rely} relation. In an
-ordinary concurrent program, every environment step is a valid
-interference in the current thread. However, in the presence of
-transactions executing under various levels of weak isolation,
-determining what constitutes an interference is a non-trivial problem.
-For example, a \iso{Serializable} transaction admits no interference,
-whereas within a transaction executing under \iso{Read Uncommitted}
-isolation, all interferences are valid. Between these two extremes are
-various levels of isolation that admit some interferences while
-prohibiting others. For example, \iso{Read Committed} isolation admits
-interference of a committed transaction, but not those from
-uncommitted transactions. \iso{Snapshot Isolation} admits interference
-from committed non-conflicting transactions, and so on.
-
-\begin{figure}
-\begin{smathpar}
-\begin{array}{lcl}
-\psi_{RC} & \defeq & \forall T_1,T_2,\eta_1,\eta_2.\; \txn(\eta_1) = T_1 
-  \conj \txn(\eta_2) = T_2 \\
-  & & \hspace*{0.6in}\conj T_1 \neq T_2 \conj \eta_1 \hboar
-  \eta_2 \Rightarrow T_1 \hboar \eta_2 \\
-\psi_{SI} & \defeq & \forall T_1,T_2.\; T_1 \neq T_2 \conj
-  (\exists \C{X}.~{T_1 \wrstoar \C{X}} \conj 
-                      {T_2 \wrstoar \C{X}})\\
-  &  & \hspace*{0.6in}\Rightarrow{T_1 \hboar T_2} \disj {T_2 \hboar T_1} \\
-\end{array}
-\end{smathpar}
-\caption{\small Interference properties for different weak isolation
-levels can be captured as constraints over the happens-before
-relation.}
-\label{fig:interference-ex}
+% \begin{figure}[!t]
+% \centering
+% %
+% \begin{subfigure}[b]{0.46\textwidth}
+% \begin{ocaml}
+% Set s' = $\emptyset$;
+% foreach x in s {
+%   s'.add(f(x)); 
+% }
+% \end{ocaml}
+% \caption{}
+% \end{subfigure}
+% %
+% \begin{subfigure}[b]{0.5\textwidth}
+% \begin{ocaml}
+% let s' = ref [];
+% foreach s (fun x -> s' := x::(!s'));
+% \end{ocaml}
+% \caption{Lorem ipsum, lorem ipsum,Lorem ipsum, lorem ipsum,Lorem ipsum}
+% \end{subfigure}
+% \caption{Caption place holder}
+% %
+% \caption{New versions are created from existing versions either
+% through \C{push} or \C{merge}.}
+% \label{fig:syntactic-ancestors}
+% \end{figure}
+\begin{figure}[!h]
+\begin{ocaml}
+let l' = ref [];
+foreach l (fun x -> l' := x::(!l'));
+\end{ocaml}
+\caption{A program that uses mutable state to reverse a list of elements}
+\label{fig:mutable-rev}
 \end{figure}
 
-\begin{figure}
-\centering
-\subcaptionbox {
-  \label{fig:hb-eg-1}
-} [
-  0.25\columnwidth
-] {
-  \includegraphics[scale=0.45]{Figures/hb-eg-1}
-}
-\subcaptionbox {
-  \label{fig:hb-eg-2}
-} [
-  0.25\columnwidth
-] {
-  \includegraphics[scale=0.45]{Figures/hb-eg-2}
-}
-\subcaptionbox {
-  \label{fig:hb-eg-3}
-} [
-  0.25\columnwidth
-] {
-  \includegraphics[scale=0.45]{Figures/hb-eg-3}
-}
-\subcaptionbox {
-  \label{fig:hb-eg-4}
-}{
-  \includegraphics[scale=0.45]{Figures/hb-eg-4}
-}
-
-\caption{\small Possible executions for the example in
-Fig.~\ref{fig:motiv-eg-1}. Letters R, W and C stand for read, write
-and commit, respectively. Each blue dashed arrow represents a $\hbZ$
-relationship between all operations of \C{Wd2} and an operation in
-\C{Wd1}. $\psi_{SI}$ specification allows $\hbZ$ arrows (hence the
-execution) in Fig.~\ref{fig:hb-eg-1}. $\psi_{RC}$ specification allows
-$\hbZ$ arrows in
-Figs.~\ref{fig:hb-eg-1},~\ref{fig:hb-eg-2},~\ref{fig:hb-eg-3}, and
-~\ref{fig:hb-eg-4}.  }
-\label{fig:motiv-eg-1-hb}
-\vspace*{-8pt}
-\end{figure}
-
-For the rely-guarantee approach to be useful, it has to thus constrain
-allowed interference in accordance with the chosen level of isolation
-associated with a transaction. The first observation we make is that
-we can constrain interference by constraining the nature of the
-\emph{happens-before} ($\hbZ$) relation, which ultimately dictates
-what other transactions become visible to a transaction, or its
-constituents, and when. To do so, we axiomatize the $\hbZ$ relation to
-capture the interference characteristics of different isolation levels
-as their specifications. Two such specifications are
-shown\footnote{Note that specifications presented here are solely for
-illustration. The actual specifications described in
-\S\ref{sec:ansi-isolation} are more nuanced.} in
-Fig.~\ref{fig:interference-ex}. The \iso{Read Committed} specification
-($\psi_{RC}$) allows an operation $\eta_1$ of a transaction $T_1$ to
-happen-before an operation $\eta_2$ of another transaction $T_2$ only
-if every operation in $T_1$ (including its commit) happens-before
-$\eta_2$ (we let $T_1 \hboar \eta_2$ denote this). The specification
-does not require $T_1$ to execute and commit before all operations of
-$T_2$, thus allowing it to interfere in $T_2$, just as \C{Wd2}
-interferes in \C{Wd1} in Fig.~\ref{fig:motiv-eg-1-a}. The
-\iso{Snapshot Isolation} specification ($\psi_{SI}$) requires two
-transactions that write to the same variable (denoted $\wrstoar$) to be
-related by happens-before.  This effectively prohibits interference
-due to actions in $T_1$ being interleaved in $T_2$ (or, vice versa).
-
-Fig.~\ref{fig:motiv-eg-1-hb} is a visualization of $\hbZ$ from \C{Wd2}
-to \C{Wd1} allowed by $\psi_{RC}$ and $\psi_{SI}$. Arrows in all
-executions are legal under $\psi_{RC}$ because in no execution does an
-operation from \C{Wd2} happen-before an operation of \C{Wd1} without
-the commit of \C{Wd2} also happening before the operation of \C{Wd1}.
-$\psi_{SI}$ however disallows $\hbZ$ arrows (hence executions) in
-Figs.~\ref{fig:hb-eg-2},~\ref{fig:hb-eg-3}, and~\ref{fig:hb-eg-4}
-because such arrows establish $\hbZ$ edges between (operations in)
-\C{Wd2} and only a subset of the operations in \C{Wd1}. Since the
-result of an operation (\eg a read) depends on what happens before
-that operation, the structure of the $\hbZ$ relation ultimately
-dictates the final result of the program. The arrows in
-Figs.~\ref{fig:hb-eg-1} and~\ref{fig:hb-eg-2} denote $\hbZ$
-relationships that do not affect the value of \C{B} in a way that
-causes the program to violate its post-condition. In contrast, arrows
-in Figs.~\ref{fig:hb-eg-3} and~\ref{fig:hb-eg-4} denote $\hbZ$
-relationships that lead to the violation of the post-condition. The
-task of the reasoning framework is to determine if \emph{all} $\hbZ$
-relationships allowed by an isolation level lead to the satisfaction
-of the post-condition.
-
-% There is a $\hb$ edge (first dashed) that is allowed by RC but
-% prohibited by SI, which nonetheless does not lead to invariant
-% violation.
-
-The specification of an isolation level encodes its interference
-characteristics as constraints over the $\hbZ$ relation. However, for
-this to be useful in reasoning about programs, the rely-guarantee
-framework should be able to use the specification to determine if an
-interference is valid or not, allowing the programmer to only focus on
-valid interferences. Our second observation is that this is possible
-if the reasoning framework adequately tracks $\hbZ$ at each program
-point, while preserving $\hbZ$ constraints as invariants between
-program points. An interference that leads to the violation of $\hbZ$
-constraints (i.e., an invalid interference) is thus automatically
-prohibited. For instance, consider the program point after the write
-to \C{B} in \C{Wd1}. The expected invariant ($\phi$) at that program
-point is shown below ($\committed$ stands for ``committed''):
-
-\begin{smathpar}
-\begin{array}{c}
- \neg\committed(\C{Wd1}) \conj \C{Wd1} \wrstoar \C{B}  \conj 
-    (\neg\committed(\C{Wd2}) \Rightarrow \C{B = k-a1}) \\
-    \conj (\committed(\C{Wd2})
-                \Rightarrow \C{B = k-a1-a2})
-\end{array}
-\end{smathpar}
-
-\noindent $\phi$ asserts that \C{Wd1} is not yet committed, and that
-it wrote to \C{B}, and the value of \C{B} is either \C{k-a1-a2} or
-\C{k-a1} depending on whether or not \C{Wd2} is committed. If $\phi$
-remains invariant until \C{Wd1} commits, then the post-condition (\C{B
-= k-a1-a2}) can be established easily. However, an interference from
-\C{Wd2} at this stage (captured by the last dashed arrow in
-Fig.~\ref{fig:motiv-eg-1-hb}) may violate the invariant by writing
-$\C{k-a2}$ to \C{B} and committing \C{Wd2}, thus leading to
-$\committed(\C{Wd2}) \Rightarrow \C{B=k-a2}$. Fortunately,
-\iso{Snapshot Isolation} prevents this interference, and this can be
-shown by demonstrating that an interference from \C{Wd2} starting from
-an execution state that satisfies $\psi_{SI} \wedge \phi$ leads to an
-execution state where neither $\C{Wd2} \hboar \C{Wd1}$ nor $\C{Wd1}
-\hboar \C{Wd2}$ holds; $\C{Wd2} \hboar \C{Wd1}$ does not hold because
-\C{Wd1}'s write to \C{B} clearly happened before \C{Wd2}'s commit, and
-$\C{Wd1} \hboar \C{Wd2}$ does not hold because \C{Wd1} has not yet
-committed, while \C{Wd2} has already begun. Since \C{Wd2} also writes
-to \C{B}, this violates the $\psi_{SI}$ constraint which we assume to
-be an invariant. A proof that the post-condition holds now follows
-from the contradiction. It is informative to note that if the
-invariant is $\psi_{RC}$ instead of $\psi_{SI}$, we cannot derive a
-contradiction and we cannot rule out the interference, which
-(rightfully) causes  the proof to fail.
-
-Implicit in the above discussion is the assumption of a strongly
-consistent ({\sc sc}) store that guarantees the visibility of all
-previously committed transactions. An {\sc sc} semantics can be built
-into the reasoning framework, leading to a proof system tailor-made
-for such stores.  However, for the reasoning framework to be truly
-useful, it should be capable of handling different consistency
-semantics, supporting stores that are weaker than {\sc sc} (e.g.,
-causally or eventually consistent), and should be able to reconcile
-conflicts between consistency and isolation constraints.  We
-demonstrate how our reasoning framework makes this possible in the
-following sections.
-
-%% We have thus far assumed a strongly consistent ({\sc sc}) store that,
-%% in the absence of transactions with special isolation requirements,
-%% makes the effects of any operation immediately visible to all
-%% subsequent operations. The natural behaviour of {\sc sc} to totally
-%% order all operations w.r.t. the $\hbZ$ relation could be in conflict
-%% with the constraints imposed by weak isolation. To meet the
-%% requirements of weaker isolation levels, say {\sc rc}, the store has
-%% to adapt itself to \emph{hide} the effects of concurrent transactions
-%% until they are committed. Once committed though, effects need to be
-%% immediately visible to subsequent operations. This semantics can be
-%% built into the reasoning framework, leading to a proof system
-%% tailor-made for such stores. However, for the reasoning framework to
-%% be truly useful, it has to be parametric over different consistency
-%% semantics, supporting stores that are weaker than {\sc sc} (e.g.,
-%% causally or eventually consistent), and should be able to reconcile
-%% conflicts between consistency and isolation constraints.  We
-%% demonstrate how our reasoning framework makes this possible in the
-%% following sections.
+Consider the program in Fig.~\ref{fig:mutable-rev} that populates a
+mutable list (\C{l'}) with elements from other list (\C{l}), reversing
+their order. The aim is to show that, when the program ends, both
+\C{l'} and \C{l} contain the same set of elements. If we were to
+attempt Hoare-style verification, we are obligated to define an
+inductive invariant for the \C{foreach} loop. Unfortunately, none
+exists for the loop, as such. The reason is that there is no way in
+the program to refer to the elements in the list that have already
+been added to the list \C{l'}. This problem can be easily rectified by
+introducing an integer \C{i} that keeps track of the index of \C{x} in
+\C{l}.

### 2017-06-29 motivation.tex
motivation

--- 2017-06-29 motivation.tex
@@
 
-In this section, we motivate our ideas via an example written in
-\txnimp - a functional language equipped with a \C{DB} monad to
-compose database computations. Besides the usual \C{bind} and
+We have developed and implemented our ideas in an OCaml embedded DSL
+that supports a \C{DB} monad to define and compose database
+computations given in terms of SQL operations; the monadic \C{do}
+syntax is borrowed from Haskell.  Besides the usual \C{bind} and
 \C{return} combinators, monad offers an \C{atomically} combinator that
-executes a database computation and returns the result.
+executes a database computation as an atomic transaction and returns
+the result.
 
@@
 Fig.~\ref{fig:new_order_code} shows a simplified version of the TPC-C
-\C{new\_order} transaction written in \txnimp (the \C{do} syntax ala
-Haskell is used for convenience). TPC-C is an Online Transaction
+\C{new\_order} transaction.  TPC-C is a well-known Online Transaction
 Processing (OLTP) benchmark that models an order-processing system for
@@
 database transactions that operate on 9 tables. \C{new\_order} is one
-such trasaction that operates on \C{District}, \C{Order},
-{New\_order}, \C{Stock}, and \C{Order\_line} tables. The transaction
+such transaction that operates on \C{District}, \C{Order},
+\C{New\_order}, \C{Stock}, and \C{Order\_line} tables. The transaction
 acts on the behalf of a customer, whose id is \C{c\_id}, to place a
@@
 transaction does so by invoking appropriate SQL functionality,
-captured in Fig.~\ref{fig:new_order_code} by various calls to
-functions under the \C{SQL} module. All \C{SQL} functions take the
-table name (a nullary constructor) as their first argument. The
-higher-order \C{SQL.select1} function accepts a boolean function that
-describes the selection criteria, and returns any record that meets
-the criteria (it models the SQL query \C{SELECT \ldots\xspace LIMIT
-1}). \C{SQL.update} also accepts a boolean function (3rd argument) to
-select the records to be updated. Its 2nd argument is a function that
-maps each selected record to a new (updated) record. \C{SQL.insert}
-inserts a given record into the specified table in the database.
+captured by various calls to functions under the \C{SQL} module. All
+\C{SQL} functions take the table name (a nullary constructor) as their
+first argument. The higher-order \C{SQL.select1} function accepts a
+boolean function that describes the selection criteria, and returns
+any record that meets the criteria (it models the SQL query \C{SELECT
+  \ldots\xspace LIMIT 1}). \C{SQL.update} also accepts a Boolean
+function (3rd argument) to select the records to be updated. Its
+2$%nd$ argument is a function that maps each selected record to a new
+(updated) record. \C{SQL.insert} inserts a given record into the
+specified table in the database.
 
@@
 (\C{d\_id}). The sequence number is stored in the corresponding
-\C{District} record, and updated each time a new order added to the
+\C{District} record, and updated each time a new order is added to the
 system. Since each order may request multiple items (\C{item\_reqs}),
@@
 such consistency condition is the requirement that for a given order
-\C{o}, the order-line-count field (\C{o.o\_ol\_cnt}) should reflect
-the number of order lines under the order, i.e., the number of
+\C{o}, the \emph{order-line-count} field (\C{o.o\_ol\_cnt}) should
+reflect the number of order lines under the order, i.e., the number of
 \C{Order\_line} records whose \C{ol\_o\_id} is equal to \C{o.o\_id}.
-In a sequential exectution, it is more-or-less easy to see how this
-condition condition is preserved by the \C{new\_order} transaction by
-examining Fig.~\ref{fig:new_order_code}. A new \C{Order} record is
-added with its \C{o\_id} distinct from the existing order ids, and its
-\C{o\_ol\_cnt} set equal to the size of \C{item\_reqs} set. The
-\C{foreach} loop runs once for each \C{item\_req}, adding a new
-\C{Order\_line} record for each requested item, with its \C{ol\_o\_id}
-field set to \C{o\_id}. Thus, by the end of the loop, the number of
-\C{Order\_line} records in the database, whose \C{ol\_o\_id} is equal
-to \C{o\_id}, is equal to the size of the \C{item\_req} set, which
-inturn is equal to the \C{Order} record's \C{o\_ol\_cnt} field. Thus
-the consistency condition is preserved.
+In a sequential execution, it is easy to see how this condition is
+preserved.  A new \C{Order} record is added with its \C{o\_id}
+distinct from the existing order ids, and its \C{o\_ol\_cnt} is set to
+be equal to the size of \C{item\_reqs} set. The \C{foreach} loop runs
+once for each \C{item\_req}, adding a new \C{Order\_line} record for
+each requested item, with its \C{ol\_o\_id} field set to
+\C{o\_id}. Thus, at the end of the loop, the number of \C{Order\_line}
+records in the database, whose \C{ol\_o\_id} is equal to \C{o\_id}, is
+equal to the size of the \C{item\_req} set, which in turn is equal to
+the \C{Order} record's \C{o\_ol\_cnt} field, thus preserving the
+consistency condition.
 
-While the aforementioned reasoning is simple to perform manually, the
-sequential verification of database programs, such as the one in
-Fig.~\ref{fig:new_order_code}, is difficult to automate owing to the
-presence of loops and collections. Thus verification of serializable
-transactions is comparable to the verification of sequential
-(imperative) programs in terms of difficulty. It is therefore not
-unreasonable to expect the reasoning machinery (i.e., program logics
-and inference techniques) developed for imperative programs to be
-equally effective on serializable transactions.
+Because the aforementioned reasoning is reasonably simple to perform
+manually, it may be feasible to apply techniques for concurrent
+program verification~\cite{....} to automate much of this effort.
+Serializability aids the tractability of verification by preventing
+any interference among concurrently executing transactions.  Under
+weak isolation\footnote{Weak isolation doesn't violate atomicity as
+  long as the witnessed effects are those of committed transactions},
+interferences of various kinds are permitted.  Although various forms
+of weak isolation differ in the kinds of effects they allow and when
+those effects are made visible, admitting interference of any kind
+from concurrent atomic transactions makes verifying program invariants
+significantly harder than under a serializable execution.
 
-Unfortunately, transactions are rarely serializable in practice, and
-presence of weak isolation significantly complicates the reasoning
-about database programs. Weak isolation allows a transaction to
-witness the effects of concurrent transactions during its
-execution\footnote{Weak isolation doesn't violate atomicity as long as
-the witnessed effects are those of the committed transactions}.
-Various levels of weak isolation differ in what effects each allows
-the transaction to witness, and when. Admitting interference from
-concurrent transactions may violate the invariants that would
-otherwise hold under a fully isolated execution.
-
-Let us reconsider the \C{new\_order} transaction, which has been
-previously shown to preserve TPC-C's consistency condition under a
-serializable execution. However, serializability is rarely the default
-isolation level on databases. For instance, the default on Postgres is
-Read Committed (RC), which is considerably weaker than
-serializability. An RC transaction is isolated from \emph{dirty
-writes}, i.e., writes of uncommitted transactions, but is allowed to
-witness the writes of concurrent transactions as soon as they are
-committed. Thus, with two concurrent instances of the \C{new\_order}
-transaction (call them $T_1$ and $T_2$) trying to place new orders for
-different customers under the same district (\C{d\_id}), executions
-shown in Fig.~\ref{fig:new_order_execs} are possible on Postgres.
+To illustrate, reconsider the behavior of the \C{new\_order}
+transaction when executed with a \emph{Read Committed} (RC) isolation
+level, the default isolation level in PostgreSQL, a widely used
+open-source database system.  An RC transaction is isolated from
+\emph{dirty writes}, i.e., writes of uncommitted transactions, but is
+allowed to witness the writes of concurrent transactions as soon as
+they are committed. Thus, with two concurrent instances of the
+\C{new\_order} transaction (call them $T_1$ and $T_2$), both
+concurrently placing new orders for different customers under the same
+district (\C{d\_id}), PostgreSQL admits the two executions shown in
+Fig.~\ref{fig:new_order_execs}.
 
@@
 The figure depicts an execution as a series of read, write, and commit
-operations. In the execution on left, the \C{new\_order} instance
+operations. In the execution on the left, the \C{new\_order} instance
 $T_1$ (green) reads the \C{d\_next\_o\_id} field of the district
@@
 acccessed. This is allowed because reads do not obtain a mutually
-exclusive lock on most databases, including Postgres. Post $T_2$'s
+exclusive lock on most databases, including PostgreSQL. After $T_2$'s
 commit, $T_1$ resumes execution and adds new \C{Order} and
@@
 
-Note that, in the above example, there is no Write-Write conflict, or
-the \emph{lost update} anomaly that often characterizes concurrency
-bugs, such as illegal concurrent withdraws on a bank
-account~\cite{poloniexbug}. While $T_1$ and $T_2$ both increment the
-\C{d\_next\_o\_id} field of the district record, they do so atomically
-(Line 4 of Fig.~\ref{fig:new_order_code}), allowing both updates to be
-present in the final state. Likewise, four other well-known anomalies
-that characterize RC isolation~\cite{berenson}, namely, \emph{fuzzy
-reads}, \emph{phantom reads}, \emph{read skew}, and \emph{write skew},
-are also not exhibited by the example. Thus, any program analysis,
-static or dynamic, that aims to determine appropriate isolation by
-checking for possible manifestations of known anomalies fails to
-identify grounds for promoting the isolation level of \C{new\_order}.
+Notably, this example does not exhibit many common concurrency bugs
+that arise from incorrect use of weak isolation such as
+\emph{write-write} conflicts, or \emph{lost updates}.  While $T_1$ and
+$T_2$ both increment the \C{d\_next\_o\_id} field of the district
+record, they do so atomically (Line 4 of
+Fig.~\ref{fig:new_order_code}), allowing both updates to be present in
+the final state. Likewise, other well-known anomalies that
+characterize RC isolation~\cite{berenson} such as \emph{fuzzy reads},
+\emph{phantom reads}, \emph{read skew}, and \emph{write skew}, are
+also not exhibited by the example.  Thus, program analyses that aim to
+determine appropriate isolation by checking for possible
+manifestations of these anomalies would fail to identify grounds for
+promoting the isolation level of \C{new\_order} to something stronger.
 Yet, if we take the semantics of the application into account, it is
-clear that RC is not an appropriate isolation level for
-\C{new\_order}. Clearly, there is a need for a high-level verification
-framework that lets us reason about transaction semantics in
-conjunction with the semantics of weak isolation.
+quite clear that RC is not an appropriate isolation level for
+\C{new\_order}. Clearly, the demonstrates the need for a high-level
+verification framework that lets us reason about application semantics
+in conjunction with the semantics of weak isolation.
 
-Facilitating high-level reasoning about weak isolation is however not
-easy owing to the low-level nature of its specification, either in
-terms of the implementation details~\cite{gray}, allowed
-anomalies~\cite{berenson}, or well-formedness condition on program
-traces~\cite{adyaphd,gotsmanconcur15}. For instance, does promoting
-the isolation level of \C{new\_order} to Repeatable Read (RR) help in
-preserving the consistency condition? To answer this question, one has
-to relate the low-level specification of RR given in terms of
-trace-based relations~\cite{adyaphd,gotsmanconcur15} to the semantics
-of TPC-C, which is difficult. Alternatively, one could reason about
-how RR is implemented, say on Postgres, and reason about the
-\C{new\_order} implementation in conjunction with Postgres's RR
-implementation. However, this is not any easier, considering that
-Postgres employs a host of techniques, including Multiversion, locks
-and rollbacks, to implement RR. The verification problem becomes even
-more difficult if the aim is to facilitate automation. Our experience
-suggests that a proof framework that requires programmers to relate
-trace-based isolation specifications to program semantics necessarily
-imposes heavy annotation burden, requires painstakingly long and
-verbose proofs, and does not lend itself to automation. Weak isolation
-aside, there are also the conventional challenges that face
-(automatic) verification of concurrent programs: facilitating modular
-proofs, inferring intermediary assertions and loop invariants that are
-stable w.r.t interference, etc. A proof system for weakly isolated
-transactions has to necessarily address these challenges, along with
-those that arise due to weak isolation.
+While reasoning in terms of anomalies is cumbersome, and as the above
+example shows often inadequate, reasoning about weak isolation in
+terms of low-level operational details faces similar hurdles given the
+conceptual gap between properties defined by consistency conditions
+and data structures and state maintained by implementations.  To
+overcome these challenges, we develop a framework that axiomatizes the
+behavior of weak isolation implementations to enable modular reasoning
+using well-established techniques such as Rely-Guarantee~\cite{...}
+augmented to deal with weakly-isolated transactions.
 
-The proof system and the automation framework proposed in this paper
-overcomes or circumvents the aforementioned problems. The first
-observation that underlies our approach is that, while databases
-implement weak isolation employing a host of techniques to facilitate
-maximum concurrency despite the isolation, the operational semantics
-of such implementations can nonetheless be captured using a simple
-abstraction of the database state exposed to the application. For
-instance, consider how Postgres executes an RR transaction. A
-(conceptual) snapshot of the database state is created at the
-beginning of the transaction against which its reads are executed.
-When the transaction attempts to update a record, Postgres first
-checks if the current version of the record in the database is same as
-the snapshot version. If yes, then an exclusive lock is obtained on
-the record, and the updatation is carried out. If the record is
-already locked, Postgres waits (blocks the transaction) until the lock
-is released. Once obtained, an exclusive lock is held until the
-transaction commits or rolls back. If the aforementioned version check
-fails (i.e., database has a later version than the snapshot), then the
-transaction is rolled back and re-executed. Thus, the implementation
-of RR on Postgres is fairly involved. Nonetheless, its operational
-semantics can be described straightforwardly in terms of the legal
-transitions of the database state (denoted, $\Delta \longrightarrow
-\Delta'$) allowed while an RR transaction is in progress. Since
-transaction's reads are always served from a snapshot, no state
-changes are witnessed while the execution is in progress. Thus,
-insofar as an RR transaction is concerned, the database state doesn't
-change (i.e., $\Delta' = \Delta$) during the execution. Note that the
-uncommitted writes of the transaction are not part of $\Delta$;
-instead we imagine a transaction-local state ($\delta$) that is
-populated with only the uncommitted local writes. When the transaction
-commits, the local state ($\delta$) is flushed to the global state
-($\Delta$) to yeild a committed state. However, unlike reads, commit
-is performed against the current state of the database (call it
-$\Delta_c$); not the snapshot. Thus, after the transaction finishes
-execution, but before it commits, the transaction witnesses the
-effects of all concurrent transactions in a single database state
-transition $\Delta \longrightarrow \Delta_c$. Postgres's RR
-implementation effectively constrains this transition. Firstly, due to
-the version check, the current transaction cannot update a data item
-that was already updated by a concurrent transaction. Secondly, due to
-exclusive write locks, a data item updated by the current transaction
-cannot be overwritten by a concurrent transaction. Putting these two
-together, we know that the transition $\Delta \longrightarrow
-\Delta_c$ does not involve any concurrent transactions that write to
-the same data items as the current RR transaction. Since $\delta$
-records the writes of the current transaction, intuitively, forall
-$x\in\delta$, $\Delta_c(x) = \Delta(x)$. To summarize, the operational
-semantics of Postgres's RR implementation can be captured as
-constraints on the transitions of the database state ($\Delta
-\longrightarrow \Delta'$) during the lifetime an RR
-transaction ($T$):
+To illustrate, consider how we might verify that \C{new_order} is
+sound when executed under a \emph{Repeatable Read} (RR) isolation
+level.  PostgreSQL executes an RR transaction by taking a
+(conceptual) snapshot of the database state before the transaction
+begins that is used by reads performing the transaction.  When the
+transaction attempts to update a record, a check is made to determine
+if the current version of the record in the database is same as the
+snapshot version. If it is, an exclusive lock is obtained on the
+record, and the update is performed. If the record is already locked,
+the transaction blocks until the lock is released.  When the lock is
+obtained, is held exclusively until the transaction commits or rolls
+back. If the aforementioned version check fails (i.e., the database
+has a later version than the snapshot), the transaction is rolled back
+and re-executed.
+
+Although this implementation, comprising many thousands of lines code,
+is highly involved, its behavior in terms of state transitions on the
+database is fairly simple.  Since transaction's reads are always
+served from a snapshot, no state changes are witnessed while the
+execution is in progress. Thus, insofar as an RR transaction is
+concerned, the database state does not change during the execution.
+Uncommitted writes are recorded in a transaction-local state.  When
+the transaction commits, the local state is atomically written to the
+global state to yield a global state that reflects the transaction's
+updates.  However, unlike a strongly isolated serializable
+transaction, the commit operation is performed against the current
+state of the database, not the snapshot. Thus, after the transaction
+finishes execution, but before it commits, the transaction is able to
+witness the effects of all concurrent transactions.  The PostgreSQL RR
+implementation effectively constrains this transition.
+
+We can axiomatize this operational description by observing that (a)
+due to the version check, the current transaction cannot update a data
+item that was already updated by a concurrent transaction, and (b) due
+to the use of exclusive write locks, a data item updated by the
+current transaction cannot be overwritten by a concurrent transaction.
+If $\Delta$ is the state of the database when an RR transactions
+finishes, and $\Delta_c$ is the state visible to the transaction at
+the point of commit, we know the transition from $\Delta$ to
+$\Delta_c$ (written $\Delta \longrightarrow \Delta_c$) cannot exhibit
+effects from any concurrent transactions that write to the same data
+items as the current RR transaction.  Similarly, if $\delta$ denotes
+a local log that relates transaction variables being written with their updated values,
+then $\forall x\in\mathit{dom}(\delta)$, $\Delta_c(x) =
+\Delta(x)$. To summarize, the operational semantics of PostgresSQL's RR
+implementation can be captured as an axiomatization over transitions of
+the database state ($\Delta \longrightarrow \Delta'$) during the
+lifetime an RR transaction ($T$):
 \begin{itemize}
@@
 \end{itemize}
+
+This simple characterization of RR isolation allows us to verify the
+consistency condtions associated with the \C{new\_order}
+transaction. First, since the database doesn't change ($\Delta' =
+\Delta$) (where $\Delta'$ represents the state after the transaction
+commits), during the execution of the transaction's body, we can
+reason about \C{new\_order} as if it executed in complete isolation
+until its commit point, leading to a verification process similar to
+what would been applied when reasoning about serializability.  When
+\C{new\_order} finishes execution and becomes ready to commit, a
+transition that transfers the transaction'slocal writes ($\delta$) to
+the unchanged database state ($\Delta$).  However, we are required to
+consider the interference from concurrent transitions at this point,
+which might change the database state from $\Delta$ to $\Delta_c$. If
+this interference includes the effects of a concurrent \C{new\_order}
+transaction (with the same \C{d\_id}), then verification fails as
+described previously (Fig.~\ref{fig:new_order_execs}). Fortunately,
+sequential reasoning shows that this is impossible - RR prevents a
+concurrent \C{new\_order} transaction that modifies the same
+\C{District} record as the current transaction (concretely, since the
+record is already present in the current transaction's local log, any
+transition from $\Delta$ to $\Delta_c$ cannot change this record).
+Applying such axiomatic reasoning on \C{new\_order} allows us to prove
+that the TPC-C invariant holds when the transaction is executed under
+PostgreSQL's RR isolation.  Our proof framework generalizes this style
+of reasoning to various isolation levels on databases.
+
+The second observation that informs our approach is one that pertains
+to automation.
 
@@
 %% following sections.
+
+For example, \emph{Repeatable
+  Read} is a stronger isolation level that RC; we might ask whether
+executing the \C{new_order} transaction under this isolation level
+would preserve the consistency conditions of the transaction.  To
+answer this question, we must relate the low-level specification of RR
+given in terms of e.g., trace-based
+relations~\cite{adyaphd,gotsmanconcur15} to the semantics of TPC-C,
+which is difficult.   Alternatively, one could reason about how RR is
+implemented, say on PostgresSQL, and reason about the \C{new\_order}
+implementation in conjunction with Postgres's RR
+implementation. However, this is not any easier, considering that
+Postgres employs a host of techniques, including Multiversion, locks
+and rollbacks, to implement RR. The verification problem becomes even
+more difficult if the aim is to facilitate automation. Our experience
+suggests that a proof framework that requires programmers to relate
+trace-based isolation specifications to program semantics necessarily
+imposes heavy annotation burden, requires painstakingly long and
+verbose proofs, and does not lend itself to automation. Weak isolation
+aside, there are also the conventional challenges that face
+(automatic) verification of concurrent programs: facilitating modular
+proofs, inferring intermediary assertions and loop invariants that are
+stable w.r.t interference, etc. A proof system for weakly isolated
+transactions has to necessarily address these challenges, along with
+those that arise due to weak isolation.

### 2017-06-30 motiv-automation.tex
Motivating automation.

The second observation that informs our approach is one that pertains
to automation. Program verification, even when machine-aided, often
entails significant annotation burden in the form of intermediary
assertions and loop invariants required to prove a program correct.
Even with concurrent program logics, such as Rely-Gurantee, which
extend Hoare logic with additional artifacts, (stable) intermediary
assertions and loop invariants remain a major source of annotation
burden. Since database transactions inherently deal with the mutable
state of the database, it is natural to formulate them as concurrent
imperative programs. An advantage with this formulation is that it
lets us immediately apply existing concurrent program logics to
database transactions, albeit with necessary customizations to deal
with weak isolation. The downside, however, is that there hardly is
any automation support to apply such program logics on concurrent
imperative programs. Attemps at automation face the problem of
inferring stable and inductive loop invariants in its full generality,
which, while being undecidable in theory, and hard in practice, also
takes our focus away from weak isolation.

In the light of the above observations, we refrain from thinking of
database transactions as concurrent imperative programs insofar as
automation is concerned.  Instead, we see value in viewing them as
essentially functional programs that manage the database state via a
monad akin to the \C{State} monad~\cite{statemonad}. We find it useful
to reason about statements that mutate the database state, not in
terms of a pre- and post-condition pair, but in terms of a state
transformer that relates the pre- and post-states of a statement. The
state transformer semantics is defined algorithmically, just like the
predicate transformer semantics (e.g., strongest post-condition), but
the semantic domain doesn't admit arbitrary logical formulas. Instead,
it is the domain of sets equipped with a (set) monad bind ($\bind$)
operator.  Thus, a state transformer interprets a SQL statement in the
set domain, taking advantage of the fact that a a database is
essentially a set of tuples, and a SQL statement a set transformer.
The benefit of this approach is that loops can now be substituted with
higher-order combinators that automatically lift the state transformer
of its higher-order argument (i.e., the loop body) to the state
transformer of the combined expression (i.e., the loop).  Thus the
semantics of a \C{foreach} loop, for instance, can be captured as a
state transformer, where the state is a set, and the transformation
involves a bind operation. We illustrate this intuition on a simple
example.

% \begin{figure}[!t]
% \centering
% %
% \begin{subfigure}[b]{0.46\textwidth}
% \begin{ocaml}
% Set s' = $\emptyset$;
% foreach x in s {
%   s'.add(f(x)); 
% }
% \end{ocaml}
% \caption{}
% \end{subfigure}
% %
% \begin{subfigure}[b]{0.5\textwidth}
% \begin{ocaml}
% let s' = ref [];
% foreach s (fun x -> s' := x::(!s'));
% \end{ocaml}
% \caption{Lorem ipsum, lorem ipsum,Lorem ipsum, lorem ipsum,Lorem ipsum}
% \end{subfigure}
% \caption{Caption place holder}
% %
% \caption{New versions are created from existing versions either
% through \C{push} or \C{merge}.}
% \label{fig:syntactic-ancestors}
% \end{figure}
% let s' = ref (Set.empty) in
% foreach xs @@ fun x -> 
%   begin
%     s' := Set.union !s' @@ 
%             Set.map_selected s (fun y -> y<x)
%                                (fun y -> y+x);
%     s' := Set.add !s' x;
%   end
\begin{figure}[!h]
\begin{ocaml}
foreach item_reqs @@ fun item_req -> do
  SQL.update Stock (fun s -> {s with s_qty = k1}) 
                   (fun s -> s.s_i_id = item_req.ol_i_id);
  SQL.insert Order_line {ol_o_id=k2; ol_d_id=k3; 
                         ol_i_id=item_req.ol_i_id; ol_qty=item_req.ol_qty}
\end{ocaml}
\caption{Foreach loop from Fig.~\ref{fig:new_order_code}}
\label{fig:foreach_code}
\end{figure}

Fig.~\ref{fig:foreach_code} shows a (simplified) snippet of code taken
from Fig.~\ref{fig:new_order_code}. Few irrelevant expressions have
been replaced with constants (\C{k1} to \C{k3}). Body of the loop
executes a SQL update followed by an insert. Recall that a transaction
reads from the global database ($\Delta$), and writes to a
transaction-local database ($\delta$). An update
statement filters the tuples that match the search criteria from the
global database ($\Delta$), and computes the updated tuples to be
added to the local database. Thus, its state transformer (call it
$T_U$) is the following function on sets (Ocaml syntax is used when
convenient; $\bind$ has higher precedence than $\cup$):
\begin{ocaml}
  fun ($\Delta$,$\delta$) -> $\delta$ $\cup$ $\Delta$$\bind$(fun s -> if table(s) = stock && s.s_i_id = item_req.ol_i_id 
                                 then {{s with s_qty = k1}} (* a singleton set *)
                                 else $\emptyset$)
\end{ocaml}
% \begin{smathpar}
% \begin{array}{c}
%   \lambda\Delta.\lambda\delta.\, \delta \;\cup\; 
%       \Delta \,\bind\, (\lambda s.\, 
%           \C{if}\;{\C{stock}(s) \conj 
%                    \C{s.s\_i\_id} = \C{item\_req.ol\_i\_id}}\\
%            \hspace*{0.9in}
%            \C{then}\;{\{\{s \;\C{with}\; \C{s\_qty} = \C{k1}\}\}}\;
%            \C{else}\;{\emptyset})
% \end{array}
% \end{smathpar}
% \begin{verbatim}
%   Rmem(σ'(s')) = Rmem(σ(s')) U 
%                  Rmem[\x. Rmem[\y. if y<x then {y+x} else {}](σ(s)) U {x}](xs)
% \end{verbatim}
The transformer maps the current global ($\Delta$) and local
($\delta$) databases to the new local database. Observe that
$T_U(\Delta,\delta)$ is of the form $\delta \,\cup\,
F_U(\Delta)$. The transformer ($T_I(\Delta,\delta)$)
for the subsequent \C{insert} statement can be similarly calculated to
be of the form $\delta \cup F_I(\Delta)$.
Their composition gives the state transformer of the loop body:
\begin{ocaml}
  fun ($\Delta$,$\delta$) -> $\delta$ $\cup$ $F_U(\Delta) \,\cup\, F_I(\Delta)$
\end{ocaml}
Let $F_{body}(\Delta) = F_U(\Delta) \,\cup\, F_I(\Delta)$.  Observe
that the loop variable \C{item\_req} occurs free in $F_{body}$. The
transformer for the \C{foreach} loop can now be computed as following:
\begin{ocaml}
  fun ($\Delta$,$\delta$) -> $\delta$ $\cup$ item_reqs$\bind$(fun item_req -> $F_{body}(\Delta)$)
\end{ocaml}
Observe that the transformer captures the precise semantics of the
\C{foeach} loop as a formula in the domain of sets extended with a
bind function. The advantage of doing so is that we can now make use
of a  semantics-preserving translation from this domain to the
first-order logic, allowing us to leverage SMT solvers for automatic
proofs without having to infer a loop invariant.
Sec.~\ref{sec:automation} describes this translation. In the
exposition, we assumed the global database state ($\Delta$) to remain
invaraint, which is clearly not the case when we admit concurrency.
The concurrency extension of the state transformer semantics is also
covered in Sec.~\ref{sec:automation}. The immediate two sections,
however, focus on laying theoretical foundations for reasoning about
weak isolation.

### 2017-06-30 motivation.tex
Motivating automation.

--- 2017-06-29 motivation.tex
@@
 
-The second observation that informs our approach is one that pertains
-to automation. Program verification, even when machine-aided, often
-entails significant annotation burden in the form of intermediary
-assertions and loop invariants required to prove a program correct.
-Even with concurrent program logics, such as Rely-Gurantee, which
-extend Hoare logic with additional artifacts, (stable) intermediary
-assertions and loop invariants remain a major source of annotation
-burden. Since database transactions inherently deal with the mutable
-state of the database, it is natural to formulate them as concurrent
-imperative programs. An advantage with this formulation is that it
-lets us immediately apply existing concurrent program logics to
-database transactions, albeit with necessary customizations to deal
-with weak isolation. The downside, however, is that there hardly is
-any automation support to apply such program logics on concurrent
-imperative programs. Attemps at automation face the problem of
-inferring stable and inductive loop invariants in its full generality,
-which, while being undecidable in theory, and hard in practice, also
-takes our focus away from weak isolation.
-
-In the light of aforementioned reservations, we refrain from
-formulating database transactions as concurrent imperative programs.
-Instead, we see value in viewing them as essentially functional
-programs that manage the database state via a monad akin to the
-\C{State} monad~\cite{statemonad}. We find it useful to reason about
-statements that mutate the database state, not in terms of a pre- and
-post-condition pair, but in terms of a state transformer that relates
-the pre- and post-states of a statement. Loops are substituted with
-higher-order combinators that have a well-defined control flow, and
-facilitate well-founded recursion. The static semantics of these
-combinators let us lift the state transformer of its higher-order
-argument (i.e., the loop body) to the state transformer of the
-combined expression (i.e., the loop). This is made possible due to the
-characteristic structure of database programs, which also gets
-reflected in the structure of the state transformer. We illustrate
-this point through a simple example that doesn't involve transactions,
-but is nonetheless relevant.
-
-% \begin{figure}[!t]
-% \centering
-% %
-% \begin{subfigure}[b]{0.46\textwidth}
-% \begin{ocaml}
-% Set s' = $\emptyset$;
-% foreach x in s {
-%   s'.add(f(x)); 
-% }
-% \end{ocaml}
-% \caption{}
-% \end{subfigure}
-% %
-% \begin{subfigure}[b]{0.5\textwidth}
-% \begin{ocaml}
-% let s' = ref [];
-% foreach s (fun x -> s' := x::(!s'));
-% \end{ocaml}
-% \caption{Lorem ipsum, lorem ipsum,Lorem ipsum, lorem ipsum,Lorem ipsum}
-% \end{subfigure}
-% \caption{Caption place holder}
-% %
-% \caption{New versions are created from existing versions either
-% through \C{push} or \C{merge}.}
-% \label{fig:syntactic-ancestors}
-% \end{figure}
-\begin{figure}[!h]
-\begin{ocaml}
-let l' = ref [];
-foreach l (fun x -> l' := x::(!l'));
-\end{ocaml}
-\caption{A program that uses mutable state to reverse a list of elements}
-\label{fig:mutable-rev}
-\end{figure}
-
-Consider the program in Fig.~\ref{fig:mutable-rev} that populates a
-mutable list (\C{l'}) with elements from other list (\C{l}), reversing
-their order. The aim is to show that, when the program ends, both
-\C{l'} and \C{l} contain the same set of elements. If we were to
-attempt Hoare-style verification, we are obligated to define an
-inductive invariant for the \C{foreach} loop. Unfortunately, none
-exists for the loop, as such. The reason is that there is no way in
-the program to refer to the elements in the list that have already
-been added to the list \C{l'}. This problem can be easily rectified by
-introducing an integer \C{i} that keeps track of the index of \C{x} in
-\C{l}.
+\input{motiv-automation}

### 2017-06-30 abstract.tex

@@
   violate application-specific invariants. As a result, database
-  systems in practice support, and often encourage, developers to use
-  weaker alternatives. These alternatives break the strong isolation
-  guarantees offered by serializable transactions to permit greater
-  concurrency. Unfortunately, the semantics of weak isolation is
-  poorly understood, and usually explained only informally in terms of
-  low-level implementation artifacts. Consequently, verifying
-  high-level correctness properties in such environments remains a
-  challenging problem.
+  systems in practice support, and often encourage, developers to
+  implement transactions using weaker alternatives. These alternatives
+  break the strong isolation guarantees offered by serializable
+  transactions to permit greater concurrency. Unfortunately, the
+  semantics of weak isolation is poorly understood, and usually
+  explained only informally in terms of low-level implementation
+  artifacts. Consequently, verifying high-level correctness properties
+  in such environments remains a challenging problem.
 
-  To address this issue, we present a program logic that enables
-  compositional reasoning about the behaviour of concurrently
-  executing weakly-isolated transactions. Notably, our development is
-  parametric over a transaction's specific isolation semantics, and
-  the consistency guarantees provided by the underlying data store,
-  allowing it to be applicable over a range of concurrency control
-  mechanisms.  Case studies and experiments on real-world applications
-  demonstrate the utility of our approach, and provide strong evidence
-  that weakly-isolated transactions can be placed on the same formal
-  footing as their strongly-isolated serializable counterparts.
+  To address this issue, we present an automated verification
+  methodology based on a novel program logic that enables
+  compositional reasoning about the behavior of concurrently executing
+  weakly-isolated transactions. Notably, our development is parametric
+  over a transaction's specific isolation semantics, allowing it to be
+  applicable over a range of concurrency control mechanisms.  Given a
+  a collection of concurrently executing transactions, our analysis
+  verifies whether assertion ${\mathcal{A}}$ associated with
+  transaction $T$ in this collection, holds under a weak isolation
+  level ${\mathcal{I}}$.  Case studies and experiments on real-world
+  applications demonstrate the utility of our approach, and provide
+  strong evidence that verification of weakly-isolated transactions
+  can be placed on the same formal footing as their strongly-isolated
+  serializable counterparts.
 

### 2017-06-30 motivation.tex

--- 2017-06-29 motivation.tex
@@
 function (3rd argument) to select the records to be updated. Its
-2$%nd$ argument is a function that maps each selected record to a new
+2$^{nd}$ argument is a function that maps each selected record to a new
 (updated) record. \C{SQL.insert} inserts a given record into the
@@
 Because the aforementioned reasoning is reasonably simple to perform
-manually, it may be feasible to apply techniques for concurrent
-program verification~\cite{....} to automate much of this effort.
-Serializability aids the tractability of verification by preventing
-any interference among concurrently executing transactions.  Under
-weak isolation\footnote{Weak isolation doesn't violate atomicity as
-  long as the witnessed effects are those of committed transactions},
-interferences of various kinds are permitted.  Although various forms
-of weak isolation differ in the kinds of effects they allow and when
-those effects are made visible, admitting interference of any kind
-from concurrent atomic transactions makes verifying program invariants
-significantly harder than under a serializable execution.
+manually, verfiying that soundess of TPC-C's consistency conditions
+would appear to be feasible.  Serializability aids the tractability of
+verification by preventing any interference among concurrently
+executing transactions.  Under weak isolation\footnote{Weak isolation
+  doesn't violate atomicity as long as the witnessed effects are those
+  of committed transactions}, however, interferences of various kinds
+are permitted.  Although the verification problem for weakly isolated
+transactions would appear to be superficially similar to the
+verification of (racy) concurrent programs~\cite{concurrentGC...},
+weak isolation introduces new challenges arising from the use of
+transactions, and new opportunities arising from the fact that the
+store abstraction used by transaction is a relational database, not
+low-level memory.  Our proof framework addresses these challenges and
+exploits these opportunities to facilitiate fully automated
+correctness proofs.
 
-To illustrate, reconsider the behavior of the \C{new\_order}
-transaction when executed with a \emph{Read Committed} (RC) isolation
-level, the default isolation level in PostgreSQL, a widely used
-open-source database system.  An RC transaction is isolated from
-\emph{dirty writes}, i.e., writes of uncommitted transactions, but is
-allowed to witness the writes of concurrent transactions as soon as
-they are committed. Thus, with two concurrent instances of the
-\C{new\_order} transaction (call them $T_1$ and $T_2$), both
-concurrently placing new orders for different customers under the same
-district (\C{d\_id}), PostgreSQL admits the two executions shown in
+To illustrate, consider the behavior of the \C{new\_order} transaction
+when executed with a \emph{Read Committed} (RC) isolation level, the
+default isolation level in PostgreSQL, a widely used open-source
+database system.  An RC transaction is isolated from \emph{dirty
+  writes}, i.e., writes of uncommitted transactions, but is allowed to
+witness the writes of concurrent transactions as soon as they are
+committed. Thus, with two concurrent instances of the \C{new\_order}
+transaction (call them $T_1$ and $T_2$), both concurrently placing new
+orders for different customers under the same district (\C{d\_id}),
+PostgreSQL admits the two executions shown in
 Fig.~\ref{fig:new_order_execs}.
@@
 
-Notably, this example does not exhibit many common concurrency bugs
+Notably, this example does not exhibit any common concurrency bugs
 that arise from incorrect use of weak isolation such as
@@
 quite clear that RC is not an appropriate isolation level for
-\C{new\_order}. Clearly, the demonstrates the need for a high-level
-verification framework that lets us reason about application semantics
-in conjunction with the semantics of weak isolation.
+\C{new\_order}.
 
@@
 example shows often inadequate, reasoning about weak isolation in
-terms of low-level operational details faces similar hurdles given the
-conceptual gap between properties defined by consistency conditions
-and data structures and state maintained by implementations.  To
-overcome these challenges, we develop a framework that axiomatizes the
-behavior of weak isolation implementations to enable modular reasoning
-using well-established techniques such as Rely-Guarantee~\cite{...}
-augmented to deal with weakly-isolated transactions.
+terms of low-level traces~\cite{adyaphd,gotsmanconcur15} confounds
+high-level reasoning and automation.  Another alternative would
+interleave weak isolation implementation details within the program,
+yielding a (more-or-less) conventional concurrent program that can be
+then subject to classical concurrent verification methods.
+Considering the size and complexity of real-world transaction systems,
+this approach is unlikely to scale.
 
-To illustrate, consider how we might verify that \C{new_order} is
-sound when executed under a \emph{Repeatable Read} (RR) isolation
-level.  PostgreSQL executes an RR transaction by taking a
-(conceptual) snapshot of the database state before the transaction
+In this paper, we adopt a different approach that \emph{lifts}
+isolation semantics (\emph{not} their implementations) to the
+application layer, providing a unified framework to simultaneously
+reason about application invariants and isolation properties.  To
+illustrate this idea informally, consider how we might verify that
+$\C{new_order}$ is sound when executed under a \emph{Repeatable Read}
+(RR) isolation level.  PostgreSQL executes an RR transaction by taking
+a (conceptual) snapshot of the database state before the transaction
 begins that is used by reads performing the transaction.  When the
@@
 Although this implementation, comprising many thousands of lines code,
-is highly involved, its behavior in terms of state transitions on the
-database is fairly simple.  Since transaction's reads are always
-served from a snapshot, no state changes are witnessed while the
-execution is in progress. Thus, insofar as an RR transaction is
-concerned, the database state does not change during the execution.
-Uncommitted writes are recorded in a transaction-local state.  When
-the transaction commits, the local state is atomically written to the
-global state to yield a global state that reflects the transaction's
-updates.  However, unlike a strongly isolated serializable
-transaction, the commit operation is performed against the current
-state of the database, not the snapshot. Thus, after the transaction
-finishes execution, but before it commits, the transaction is able to
-witness the effects of all concurrent transactions.  The PostgreSQL RR
-implementation effectively constrains this transition.
+is highly involved, its semantic behavior in terms of how it effects
+transitions on the database state is fairly simple.  Since a
+transaction's reads are always served from a snapshot, no state
+changes are witnessed while the execution is in progress. Thus,
+insofar as an RR transaction is concerned, the database state does not
+change during the execution.  Uncommitted writes are recorded in a
+transaction-local state.  When the transaction commits, the local
+state is atomically written to the global state to yield a global
+state that reflects the transaction's updates.  However, unlike a
+strongly isolated serializable transaction, the commit operation is
+performed against the current state of the database, not the
+snapshot. Thus, after the transaction finishes execution, but before
+it commits, the transaction is able to witness the effects of all
+concurrent transactions.  The PostgreSQL RR implementation effectively
+constrains this transition.
 
@@
 The second observation that informs our approach is one that pertains
-to automation.
+to automation. Program verification, even when machine-aided, often
+entails significant annotation burden in the form of intermediary
+assertions and loop invariants required to prove a program correct.
+This is certainly true for concurrent program logics, such as
+Rely-Gurantee, which extend Hoare logic with additional artifacts and
+where (stable) intermediary assertions and loop invariants remain a
+major source of annotation burden.  However, a relational database is
+a significantly simpler abstraction than shared memory. There are no
+pointers, or linked data structures, or aliasing.  Although a database
+essentially abstracts a mutable state, the state is mutated through a
+well-defined fixed number of interfaces (SQL statements), each tagged
+with a logical formula describing what records are accessed and
+updated.
 
-\newpage
+This observation leads us away from thinking of database transactions
+as concurrent imperative programs.  Instead, we see value in viewing
+them as essentially functional computations that manage database state
+monadically~\cite{statemonad}. We find it useful to reason about
+statements that mutate the database state, not in terms of a pre- and
+post-condition pair, but in terms of a state transformer that relates
+the pre- and post-states of a statement. This state transformer
+semantics can be defined algorithmically, just like predicate
+transformer semantics (e.g., strongest post-condition).  Here, a state
+transformer interprets a SQL statement in the set domain, taking
+advantage of the fact that a a database is essentially a set of
+tuples, and a SQL statement is a transformer over these sets.  The
+benefit of this approach is that low-level loops can now be
+substituted with higher-order combinators that automatically lift the
+state transformer of its higher-order argument (i.e., the loop body)
+to the state transformer of the combined expression (i.e., the loop).
+Thus the semantics of a \C{foreach} loop, for instance, can be
+captured as a state transformer, where the state is a set, and the
+transformation defines a bind operation. We illustrate this intuition
+on a simple example.
 
-(\C{k-a1}), but before it commits, transaction \C{Wd2} (red) executes
-and commits, writing the new balance (\C{k-a2}). {\sc rc} isolation
-prevent \C{Wd2} from witnessing the uncommitted writes of transaction
-\C{Wd1}.  Subsequently committing \C{Wd1} leads to the loss of
-\C{Wd2}'s updates (the so-called \emph{lost update
-anomaly}~\cite{berenson}), resulting in an incorrect balance of
-\C{k-a1}. The execution on right describes a similar scenario with
-\C{Wd1} and \C{Wd2} exchanging their roles.  Clearly, {\sc rc} is an
-excessively weak isolation level for this program because it loses the
-updates of one transaction, resulting in the violation of the
-post-condition.  A stronger isolation level that prevents lost updates
-is required. \iso{Snapshot Isolation} ({\sc si})~\cite{berenson} fits
-this requirement; {\sc si} effectively serializes transactions that
-update a shared data object by aborting and re-executing a transaction
-if write-write conflicts are detected during its commit.\footnote{{\sc
-si} however does not serialize transactions in the absence of
-write-write conflicts.} Since {\sc si}, unlike {\sc ser}, does not
-need expensive mechanisms such as lock-based concurrency control or
-runtime monitoring, it is also more efficient, making it appropriate
-for both withdraw transactions.
-
-Thinking in terms of anomalies, as described above, is how database
-programmers are often encouraged to reason about weak isolation.
-Unfortunately, such reasoning does not rest on any sound foundation,
-and is thus highly error-prone. Reasoning in terms of how weak
-isolation variants are implemented is no better since it requires
-programmers to understand low-level implementation details of the
-database that are far removed from the application semantics. An
-attractive  alternative in this context would be a principled
-reasoning approach that combines declarative reasoning about isolation
-guarantees with operational reasoning about programs. We demonstrate
-how our proof system makes this possible in the context of the current
-example.
-
-First, we note that the example in Fig.~\ref{fig:motiv-eg-1} is a
-concurrent program, and hence is amenable to \emph{rely-guarantee}
-style reasoning~\cite{rgjones}, a compositional proof technique that
-allows us to reason about the behaviour of individual threads by
-abstracting away interferences induced by other threads (collectively
-called \emph{the environment}) into a \emph{rely} relation. In an
-ordinary concurrent program, every environment step is a valid
-interference in the current thread. However, in the presence of
-transactions executing under various levels of weak isolation,
-determining what constitutes an interference is a non-trivial problem.
-For example, a \iso{Serializable} transaction admits no interference,
-whereas within a transaction executing under \iso{Read Uncommitted}
-isolation, all interferences are valid. Between these two extremes are
-various levels of isolation that admit some interferences while
-prohibiting others. For example, \iso{Read Committed} isolation admits
-interference of a committed transaction, but not those from
-uncommitted transactions. \iso{Snapshot Isolation} admits interference
-from committed non-conflicting transactions, and so on.
-
-\begin{figure}
-\begin{smathpar}
-\begin{array}{lcl}
-\psi_{RC} & \defeq & \forall T_1,T_2,\eta_1,\eta_2.\; \txn(\eta_1) = T_1 
-  \conj \txn(\eta_2) = T_2 \\
-  & & \hspace*{0.6in}\conj T_1 \neq T_2 \conj \eta_1 \hboar
-  \eta_2 \Rightarrow T_1 \hboar \eta_2 \\
-\psi_{SI} & \defeq & \forall T_1,T_2.\; T_1 \neq T_2 \conj
-  (\exists \C{X}.~{T_1 \wrstoar \C{X}} \conj 
-                      {T_2 \wrstoar \C{X}})\\
-  &  & \hspace*{0.6in}\Rightarrow{T_1 \hboar T_2} \disj {T_2 \hboar T_1} \\
-\end{array}
-\end{smathpar}
-\caption{\small Interference properties for different weak isolation
-levels can be captured as constraints over the happens-before
-relation.}
-\label{fig:interference-ex}
+% \begin{figure}[!t]
+% \centering
+% %
+% \begin{subfigure}[b]{0.46\textwidth}
+% \begin{ocaml}
+% Set s' = $\emptyset$;
+% foreach x in s {
+%   s'.add(f(x)); 
+% }
+% \end{ocaml}
+% \caption{}
+% \end{subfigure}
+% %
+% \begin{subfigure}[b]{0.5\textwidth}
+% \begin{ocaml}
+% let s' = ref [];
+% foreach s (fun x -> s' := x::(!s'));
+% \end{ocaml}
+% \caption{Lorem ipsum, lorem ipsum,Lorem ipsum, lorem ipsum,Lorem ipsum}
+% \end{subfigure}
+% \caption{Caption place holder}
+% %
+% \caption{New versions are created from existing versions either
+% through \C{push} or \C{merge}.}
+% \label{fig:syntactic-ancestors}
+% \end{figure}
+% let s' = ref (Set.empty) in
+% foreach xs @@ fun x -> 
+%   begin
+%     s' := Set.union !s' @@ 
+%             Set.map_selected s (fun y -> y<x)
+%                                (fun y -> y+x);
+%     s' := Set.add !s' x;
+%   end
+\begin{figure}[!h]
+\begin{ocaml}
+foreach item_reqs @@ fun item_req -> do
+  SQL.update Stock (fun s -> {s with s_qty = k1}) 
+                   (fun s -> s.s_i_id = item_req.ol_i_id);
+  SQL.insert Order_line {ol_o_id=k2; ol_d_id=k3; 
+                         ol_i_id=item_req.ol_i_id; ol_qty=item_req.ol_qty}
+\end{ocaml}
+\caption{Foreach loop from Fig.~\ref{fig:new_order_code}}
+\label{fig:foreach_code}
 \end{figure}
 
-\begin{figure}
-\centering
-\subcaptionbox {
-  \label{fig:hb-eg-1}
-} [
-  0.25\columnwidth
-] {
-  \includegraphics[scale=0.45]{Figures/hb-eg-1}
-}
-\subcaptionbox {
-  \label{fig:hb-eg-2}
-} [
-  0.25\columnwidth
-] {
-  \includegraphics[scale=0.45]{Figures/hb-eg-2}
-}
-\subcaptionbox {
-  \label{fig:hb-eg-3}
-} [
-  0.25\columnwidth
-] {
-  \includegraphics[scale=0.45]{Figures/hb-eg-3}
-}
-\subcaptionbox {
-  \label{fig:hb-eg-4}
-}{
-  \includegraphics[scale=0.45]{Figures/hb-eg-4}
-}
-
-\caption{\small Possible executions for the example in
-Fig.~\ref{fig:motiv-eg-1}. Letters R, W and C stand for read, write
-and commit, respectively. Each blue dashed arrow represents a $\hbZ$
-relationship between all operations of \C{Wd2} and an operation in
-\C{Wd1}. $\psi_{SI}$ specification allows $\hbZ$ arrows (hence the
-execution) in Fig.~\ref{fig:hb-eg-1}. $\psi_{RC}$ specification allows
-$\hbZ$ arrows in
-Figs.~\ref{fig:hb-eg-1},~\ref{fig:hb-eg-2},~\ref{fig:hb-eg-3}, and
-~\ref{fig:hb-eg-4}.  }
-\label{fig:motiv-eg-1-hb}
-\vspace*{-8pt}
-\end{figure}
-
-For the rely-guarantee approach to be useful, it has to thus constrain
-allowed interference in accordance with the chosen level of isolation
-associated with a transaction. The first observation we make is that
-we can constrain interference by constraining the nature of the
-\emph{happens-before} ($\hbZ$) relation, which ultimately dictates
-what other transactions become visible to a transaction, or its
-constituents, and when. To do so, we axiomatize the $\hbZ$ relation to
-capture the interference characteristics of different isolation levels
-as their specifications. Two such specifications are
-shown\footnote{Note that specifications presented here are solely for
-illustration. The actual specifications described in
-\S\ref{sec:ansi-isolation} are more nuanced.} in
-Fig.~\ref{fig:interference-ex}. The \iso{Read Committed} specification
-($\psi_{RC}$) allows an operation $\eta_1$ of a transaction $T_1$ to
-happen-before an operation $\eta_2$ of another transaction $T_2$ only
-if every operation in $T_1$ (including its commit) happens-before
-$\eta_2$ (we let $T_1 \hboar \eta_2$ denote this). The specification
-does not require $T_1$ to execute and commit before all operations of
-$T_2$, thus allowing it to interfere in $T_2$, just as \C{Wd2}
-interferes in \C{Wd1} in Fig.~\ref{fig:motiv-eg-1-a}. The
-\iso{Snapshot Isolation} specification ($\psi_{SI}$) requires two
-transactions that write to the same variable (denoted $\wrstoar$) to be
-related by happens-before.  This effectively prohibits interference
-due to actions in $T_1$ being interleaved in $T_2$ (or, vice versa).
-
-Fig.~\ref{fig:motiv-eg-1-hb} is a visualization of $\hbZ$ from \C{Wd2}
-to \C{Wd1} allowed by $\psi_{RC}$ and $\psi_{SI}$. Arrows in all
-executions are legal under $\psi_{RC}$ because in no execution does an
-operation from \C{Wd2} happen-before an operation of \C{Wd1} without
-the commit of \C{Wd2} also happening before the operation of \C{Wd1}.
-$\psi_{SI}$ however disallows $\hbZ$ arrows (hence executions) in
-Figs.~\ref{fig:hb-eg-2},~\ref{fig:hb-eg-3}, and~\ref{fig:hb-eg-4}
-because such arrows establish $\hbZ$ edges between (operations in)
-\C{Wd2} and only a subset of the operations in \C{Wd1}. Since the
-result of an operation (\eg a read) depends on what happens before
-that operation, the structure of the $\hbZ$ relation ultimately
-dictates the final result of the program. The arrows in
-Figs.~\ref{fig:hb-eg-1} and~\ref{fig:hb-eg-2} denote $\hbZ$
-relationships that do not affect the value of \C{B} in a way that
-causes the program to violate its post-condition. In contrast, arrows
-in Figs.~\ref{fig:hb-eg-3} and~\ref{fig:hb-eg-4} denote $\hbZ$
-relationships that lead to the violation of the post-condition. The
-task of the reasoning framework is to determine if \emph{all} $\hbZ$
-relationships allowed by an isolation level lead to the satisfaction
-of the post-condition.
-
-% There is a $\hb$ edge (first dashed) that is allowed by RC but
-% prohibited by SI, which nonetheless does not lead to invariant
-% violation.
-
-The specification of an isolation level encodes its interference
-characteristics as constraints over the $\hbZ$ relation. However, for
-this to be useful in reasoning about programs, the rely-guarantee
-framework should be able to use the specification to determine if an
-interference is valid or not, allowing the programmer to only focus on
-valid interferences. Our second observation is that this is possible
-if the reasoning framework adequately tracks $\hbZ$ at each program
-point, while preserving $\hbZ$ constraints as invariants between
-program points. An interference that leads to the violation of $\hbZ$
-constraints (i.e., an invalid interference) is thus automatically
-prohibited. For instance, consider the program point after the write
-to \C{B} in \C{Wd1}. The expected invariant ($\phi$) at that program
-point is shown below ($\committed$ stands for ``committed''):
-
-\begin{smathpar}
-\begin{array}{c}
- \neg\committed(\C{Wd1}) \conj \C{Wd1} \wrstoar \C{B}  \conj 
-    (\neg\committed(\C{Wd2}) \Rightarrow \C{B = k-a1}) \\
-    \conj (\committed(\C{Wd2})
-                \Rightarrow \C{B = k-a1-a2})
-\end{array}
-\end{smathpar}
-
-\noindent $\phi$ asserts that \C{Wd1} is not yet committed, and that
-it wrote to \C{B}, and the value of \C{B} is either \C{k-a1-a2} or
-\C{k-a1} depending on whether or not \C{Wd2} is committed. If $\phi$
-remains invariant until \C{Wd1} commits, then the post-condition (\C{B
-= k-a1-a2}) can be established easily. However, an interference from
-\C{Wd2} at this stage (captured by the last dashed arrow in
-Fig.~\ref{fig:motiv-eg-1-hb}) may violate the invariant by writing
-$\C{k-a2}$ to \C{B} and committing \C{Wd2}, thus leading to
-$\committed(\C{Wd2}) \Rightarrow \C{B=k-a2}$. Fortunately,
-\iso{Snapshot Isolation} prevents this interference, and this can be
-shown by demonstrating that an interference from \C{Wd2} starting from
-an execution state that satisfies $\psi_{SI} \wedge \phi$ leads to an
-execution state where neither $\C{Wd2} \hboar \C{Wd1}$ nor $\C{Wd1}
-\hboar \C{Wd2}$ holds; $\C{Wd2} \hboar \C{Wd1}$ does not hold because
-\C{Wd1}'s write to \C{B} clearly happened before \C{Wd2}'s commit, and
-$\C{Wd1} \hboar \C{Wd2}$ does not hold because \C{Wd1} has not yet
-committed, while \C{Wd2} has already begun. Since \C{Wd2} also writes
-to \C{B}, this violates the $\psi_{SI}$ constraint which we assume to
-be an invariant. A proof that the post-condition holds now follows
-from the contradiction. It is informative to note that if the
-invariant is $\psi_{RC}$ instead of $\psi_{SI}$, we cannot derive a
-contradiction and we cannot rule out the interference, which
-(rightfully) causes  the proof to fail.
-
-Implicit in the above discussion is the assumption of a strongly
-consistent ({\sc sc}) store that guarantees the visibility of all
-previously committed transactions. An {\sc sc} semantics can be built
-into the reasoning framework, leading to a proof system tailor-made
-for such stores.  However, for the reasoning framework to be truly
-useful, it should be capable of handling different consistency
-semantics, supporting stores that are weaker than {\sc sc} (e.g.,
-causally or eventually consistent), and should be able to reconcile
-conflicts between consistency and isolation constraints.  We
-demonstrate how our reasoning framework makes this possible in the
-following sections.
-
-%% We have thus far assumed a strongly consistent ({\sc sc}) store that,
-%% in the absence of transactions with special isolation requirements,
-%% makes the effects of any operation immediately visible to all
-%% subsequent operations. The natural behaviour of {\sc sc} to totally
-%% order all operations w.r.t. the $\hbZ$ relation could be in conflict
-%% with the constraints imposed by weak isolation. To meet the
-%% requirements of weaker isolation levels, say {\sc rc}, the store has
-%% to adapt itself to \emph{hide} the effects of concurrent transactions
-%% until they are committed. Once committed though, effects need to be
-%% immediately visible to subsequent operations. This semantics can be
-%% built into the reasoning framework, leading to a proof system
-%% tailor-made for such stores. However, for the reasoning framework to
-%% be truly useful, it has to be parametric over different consistency
-%% semantics, supporting stores that are weaker than {\sc sc} (e.g.,
-%% causally or eventually consistent), and should be able to reconcile
-%% conflicts between consistency and isolation constraints.  We
-%% demonstrate how our reasoning framework makes this possible in the
-%% following sections.
-
-For example, \emph{Repeatable
-  Read} is a stronger isolation level that RC; we might ask whether
-executing the \C{new_order} transaction under this isolation level
-would preserve the consistency conditions of the transaction.  To
-answer this question, we must relate the low-level specification of RR
-given in terms of e.g., trace-based
-relations~\cite{adyaphd,gotsmanconcur15} to the semantics of TPC-C,
-which is difficult.   Alternatively, one could reason about how RR is
-implemented, say on PostgresSQL, and reason about the \C{new\_order}
-implementation in conjunction with Postgres's RR
-implementation. However, this is not any easier, considering that
-Postgres employs a host of techniques, including Multiversion, locks
-and rollbacks, to implement RR. The verification problem becomes even
-more difficult if the aim is to facilitate automation. Our experience
-suggests that a proof framework that requires programmers to relate
-trace-based isolation specifications to program semantics necessarily
-imposes heavy annotation burden, requires painstakingly long and
-verbose proofs, and does not lend itself to automation. Weak isolation
-aside, there are also the conventional challenges that face
-(automatic) verification of concurrent programs: facilitating modular
-proofs, inferring intermediary assertions and loop invariants that are
-stable w.r.t interference, etc. A proof system for weakly isolated
-transactions has to necessarily address these challenges, along with
-those that arise due to weak isolation.
+Fig.~\ref{fig:foreach_code} shows a (simplified) snippet of code taken
+from Fig.~\ref{fig:new_order_code}. Some irrelevant expressions have
+been replaced with constants (\C{k1} to \C{k3}).  The body of the loop
+executes a SQL update followed by an insert.  Recall that a transaction
+reads from the global database ($\Delta$), and writes to a
+transaction-local database ($\delta$) before committing these updates. An update
+statement filters the tuples that match the search criteria from $\Delta$
+and computes the updated tuples that are to be
+added to the local database. Thus, its state transformer (call it
+$T_U$) is the following function on sets\footnote{$\bind$ has higher precedence than $\cup$.}:
+\begin{ocaml}
+  fun ($\Delta$,$\delta$) -> $\delta$ $\cup$ $\Delta$$\bind$(fun s -> if table(s) = stock && s.s_i_id = item_req.ol_i_id 
+                                 then {{s with s_qty = k1}} (* a singleton set *)
+                                 else $\emptyset$)
+\end{ocaml}
+% \begin{smathpar}
+% \begin{array}{c}
+%   \lambda\Delta.\lambda\delta.\, \delta \;\cup\; 
+%       \Delta \,\bind\, (\lambda s.\, 
+%           \C{if}\;{\C{stock}(s) \conj 
+%                    \C{s.s\_i\_id} = \C{item\_req.ol\_i\_id}}\\
+%            \hspace*{0.9in}
+%            \C{then}\;{\{\{s \;\C{with}\; \C{s\_qty} = \C{k1}\}\}}\;
+%            \C{else}\;{\emptyset})
+% \end{array}
+% \end{smathpar}
+% \begin{verbatim}
+%   Rmem(σ'(s')) = Rmem(σ(s')) U 
+%                  Rmem[\x. Rmem[\y. if y<x then {y+x} else {}](σ(s)) U {x}](xs)
+% \end{verbatim}
+Observe that
+$T_U(\Delta,\delta)$ is of the form $\delta \,\cup\,
+F_U(\Delta)$. The transformer ($T_I(\Delta,\delta)$)
+for the subsequent \C{insert} statement can be similarly calculated to
+be of the form $\delta \cup F_I(\Delta)$.
+Their composition gives the state transformer of the loop body:
+\begin{ocaml}
+  fun ($\Delta$,$\delta$) -> $\delta$ $\cup$ $F_U(\Delta) \,\cup\, F_I(\Delta)$
+\end{ocaml}
+Let $F_{body}(\Delta) = F_U(\Delta) \,\cup\, F_I(\Delta)$.   The
+transformer for the \C{foreach} loop can now be computed as following:
+\begin{ocaml}
+  fun ($\Delta$,$\delta$) -> $\delta$ $\cup$ item_reqs$\bind$(fun item_req -> $F_{body}(\Delta)$)
+\end{ocaml}
+Observe that the transformer captures the precise semantics of the
+loop as a formula in the domain of sets extended with a bind
+function. The advantage in doing so is that we can now make use of a
+semantics-preserving translation from this domain to first-order
+logic, allowing us to leverage SMT solvers for automatic proofs
+without having to infer loop invariant, or provide intermediate
+assertions.  Sec.~\ref{sec:automation} describes this translation. In
+the exposition thus far, we assumed $\Delta$ remains invaraint, which
+is clearly not the case when we admit concurrency.  Necessary
+concurrency extensions of the state transformer semantics is also
+covered in Sec.~\ref{sec:automation}.  The immediate two sections,
+however, focus on laying theoretical foundations for reasoning about
+weak isolation.

### 2017-06-30 motivation.tex
Figures.

@@
 \begin{ocaml}
-let new_order_ d_id c_id item_reqs = atomically do
+let new_order d_id c_id item_reqs = atomically do
   dist <- SQL.select1 District (fun d -> d.d_id = d_id);
@@
   foreach item_reqs @@ fun item_req -> do
-    stk <- SQL.select1 Stock (fun s -> s.s_i_id = item_req.ol_i_id);
+    stk <- SQL.select1 Stock (fun s -> s.s_i_id = item_req.ol_i_id &&
+                                       s.s_d_id = d_id);
     let s_qty' = if stk.s_qty >= item_req.ol_qty + 10 
@@
 \end{ocaml}
-\caption{\small Concurrent withdraw transactions}
+\caption{TPC-C \C{new\_order} transaction}
 \label{fig:new_order_code}
@@
 specified table in the database.
+
+\begin{figure}[!t]
+\includegraphics[scale=0.5]{Figures/schema}
+\caption{Database schema of TPC-C's order management system.
+  Columns against highlighted background are primary keys. Arrows denote
+  foreign key relationships.}
+\label{fig:schema}
+\end{figure}
 
@@
   {\sc rc} Execution 1
-  \label{fig:motiv-eg-1-a}
+  \label{fig:motiv-eg-1-b}
 } [
@@
 ] {
-  \includegraphics[scale=0.6]{Figures/motiv-eg-1-a}
+  \includegraphics[scale=0.45]{Figures/motiv-eg-1-b}
 }
@@
   {\sc rc} Execution 2
-  \label{fig:motiv-eg-1-b}
+  \label{fig:motiv-eg-1-a}
 }{
-  \includegraphics[scale=0.6]{Figures/motiv-eg-1-b}
+  \includegraphics[scale=0.45]{Figures/motiv-eg-1-a}
 }
-\caption{\small A possible execution of the program shown in Fig.~\ref{fig:motiv-eg-1} under
-  \iso{Read Committed} isolation level. Transaction \C{Wd1} is shown
-  against lighter green background, and transaction \C{Wd2} against
-  darker red background. Each transaction reads the balance (\C{B})
-  twice, hence two \C{Reads}.}
+  \caption{\small RC executions involving two instances ($T_1$ and
+  $T_2$) of the \C{new\_order} transaction of
+  Fig.~\ref{fig:new_order_code}. 
+  Each instance reads the \C{d\_id} \C{District} record twice, second
+  time to (atomically) update the \C{d\_next\_o\_id} field.}
 \label{fig:rc-ex}

### 2017-07-01 motivation.tex
RR Figures added.

@@
 
+\begin{figure}[t]
+\includegraphics[scale=0.5]{Figures/RR-postgres}
+\caption{Database state transitions corresponding to an execution of
+  an RR transaction $T_1$ on Postgres. $T_2$ and $T_3$ are concurrent
+  transactions. State has three data items, $a$, $b$ and $c$. }
+\label{fig:rr-postgres}
+\end{figure}
+\begin{figure}[t]
+\includegraphics[scale=0.5]{Figures/RR-abstract}
+  \caption{An abstract execution equivalent to the one in
+  Fig.~\ref{fig:rr-postgres}. Database state remains invariant as long as
+  $T_1$ executes. The effects of concurrent transactions ($T_2$ and
+  $T_3$) are witnessed just before the commit.}
+\label{fig:rr-abstract}
+\end{figure}
+
 Although this implementation, comprising many thousands of lines code,

### 2017-07-01 motivation.tex
RR fig edited.

@@
 \includegraphics[scale=0.5]{Figures/RR-abstract}
-  \caption{An abstract execution equivalent to the one in
-  Fig.~\ref{fig:rr-postgres}. Database state remains invariant as long as
-  $T_1$ executes. The effects of concurrent transactions ($T_2$ and
-  $T_3$) are witnessed just before the commit.}
+  \caption{An abstract execution that includes the concrete execution
+  shown in Fig.~\ref{fig:rr-postgres}. It has no locks or snapshots,
+  admits fewer interleavings, yet results in the same post-state.  }
 \label{fig:rr-abstract}

### 2017-07-01 motivation.tex
RC figs edited.
edits to motivation

--- 2017-07-01 motivation.tex
@@
 
-We have developed and implemented our ideas in an OCaml embedded DSL
-that supports a \C{DB} monad to define and compose database
-computations given in terms of SQL operations; the monadic \C{do}
-syntax is borrowed from Haskell.  Besides the usual \C{bind} and
-\C{return} combinators, monad offers an \C{atomically} combinator that
-executes a database computation as an atomic transaction and returns
-the result.
+We have developed and implemented our ideas in an embedded DSL in
+OCaml that supports a \C{DB} monad to define and compose database
+computations given in terms of SQL operations.  Besides the usual
+\C{bind} and \C{return} combinators, the monad offers an
+\C{atomically} combinator that executes a database computation as an
+atomic transaction and returns the result.
 
@@
 Fig.~\ref{fig:new_order_code} shows a simplified version of the TPC-C
-\C{new\_order} transaction.  TPC-C is a well-known Online Transaction
-Processing (OLTP) benchmark that models an order-processing system for
-a wholesale parts supply business. The business logic is captured in 5
-database transactions that operate on 9 tables. \C{new\_order} is one
-such transaction that operates on \C{District}, \C{Order},
-\C{New\_order}, \C{Stock}, and \C{Order\_line} tables. The transaction
-acts on the behalf of a customer, whose id is \C{c\_id}, to place a
-new order for the given set of items (\C{item\_reqs}), to be served by
-a warehouse under the district identified by \C{d\_id}. The
-transaction does so by invoking appropriate SQL functionality,
-captured by various calls to functions under the \C{SQL} module. All
-\C{SQL} functions take the table name (a nullary constructor) as their
-first argument. The higher-order \C{SQL.select1} function accepts a
-boolean function that describes the selection criteria, and returns
-any record that meets the criteria (it models the SQL query \C{SELECT
-  \ldots\xspace LIMIT 1}). \C{SQL.update} also accepts a Boolean
-function (3rd argument) to select the records to be updated. Its
-2$^{nd}$ argument is a function that maps each selected record to a new
-(updated) record. \C{SQL.insert} inserts a given record into the
-specified table in the database.
+\C{new\_order} transaction written in this language.  TPC-C is a
+well-known Online Transaction Processing (OLTP) benchmark that models
+an order-processing system for a wholesale parts supply business. The
+business logic is captured in 5 database transactions that operate on
+9 tables; \C{new\_order} is one such transaction that operates on
+\C{District}, \C{Order}, \C{New\_order}, \C{Stock}, and
+\C{Order\_line} tables. The transaction acts on the behalf of a
+customer, whose id is \C{c\_id}, to place a new order for the given
+set of items (\C{item\_reqs}), to be served by a warehouse under the
+district identified by \C{d\_id}.  Fig.~\ef{fig:scheme} illustrates
+the relationship among these different tables.
+
+The transaction manages order placement by invoking appropriate SQL
+functionality, captured by various calls to functions defined by the
+\C{SQL} module. All \C{SQL} functions take the table name (a nullary
+constructor) as their first argument. The higher-order \C{SQL.select1}
+function accepts a boolean function that describes the selection
+criteria, and returns any record that meets the criteria (it models
+the SQL query \C{SELECT \ldots\xspace LIMIT 1}). \C{SQL.update} also
+accepts a Boolean function (its 3$^{rd}$ argument) to select the records to be
+updated. Its 2$^{nd}$ argument is a function that maps each selected
+record to a new (updated) record. \C{SQL.insert} inserts a given
+record into the specified table in the database.
+
+%%%SJ: Reviewers may not understand what primary and foreign keys are,
+%%%or why they are important.  There are also some fields (e.g., ol_i_id)
+%%%that are not described either in the caption or the text.
 
@@
 \C{o}, the \emph{order-line-count} field (\C{o.o\_ol\_cnt}) should
-reflect the number of order lines under the order, i.e., the number of
-\C{Order\_line} records whose \C{ol\_o\_id} is equal to \C{o.o\_id}.
-In a sequential execution, it is easy to see how this condition is
-preserved.  A new \C{Order} record is added with its \C{o\_id}
-distinct from the existing order ids, and its \C{o\_ol\_cnt} is set to
-be equal to the size of \C{item\_reqs} set. The \C{foreach} loop runs
-once for each \C{item\_req}, adding a new \C{Order\_line} record for
-each requested item, with its \C{ol\_o\_id} field set to
+reflect the number of order lines under the order; this is the number
+of \C{Order\_line} records whose \C{ol\_o\_id} field is the same as
+\C{o.o\_id}.  In a sequential execution, it is easy to see how this
+condition is preserved.  A new \C{Order} record is added with its
+\C{o\_id} distinct from existing order ids, and its \C{o\_ol\_cnt} is
+set to be equal to the size of the \C{item\_reqs} set. The \C{foreach}
+loop runs once for each \C{item\_req}, adding a new \C{Order\_line}
+record for each requested item, with its \C{ol\_o\_id} field set to
 \C{o\_id}. Thus, at the end of the loop, the number of \C{Order\_line}
-records in the database, whose \C{ol\_o\_id} is equal to \C{o\_id}, is
+records in the database, whose \C{ol\_o\_id} is equal to \C{o\_id} is
 equal to the size of the \C{item\_req} set, which in turn is equal to
@@
 Because the aforementioned reasoning is reasonably simple to perform
-manually, verfiying that soundess of TPC-C's consistency conditions
+manually, verfiying the soundess of TPC-C's consistency conditions
 would appear to be feasible.  Serializability aids the tractability of
 verification by preventing any interference among concurrently
-executing transactions.  Under weak isolation\footnote{Weak isolation
-  doesn't violate atomicity as long as the witnessed effects are those
-  of committed transactions}, however, interferences of various kinds
-are permitted.  Although the verification problem for weakly isolated
+executing transactions while the \C{new\_order} transaction executes.
+Under weak isolation\footnote{Weak isolation doesn't violate atomicity
+  as long as the witnessed effects are those of committed
+  transactions}, however, interferences of various kinds are
+permitted.  Although the verification problem for weakly isolated
 transactions would appear to be superficially similar to the
@@
 transactions, and new opportunities arising from the fact that the
-store abstraction used by transaction is a relational database, not
-low-level memory.  Our proof framework addresses these challenges and
-exploits these opportunities to facilitiate fully automated
-correctness proofs.
+store abstraction used by transactions is a relational database, not
+low-level memory.
 
-To illustrate, consider the behavior of the \C{new\_order} transaction
-when executed with a \emph{Read Committed} (RC) isolation level, the
-default isolation level in PostgreSQL, a widely used open-source
-database system.  An RC transaction is isolated from \emph{dirty
-  writes}, i.e., writes of uncommitted transactions, but is allowed to
-witness the writes of concurrent transactions as soon as they are
-committed. Thus, with two concurrent instances of the \C{new\_order}
-transaction (call them $T_1$ and $T_2$), both concurrently placing new
-orders for different customers under the same district (\C{d\_id}),
-PostgreSQL admits the two executions shown in
-Fig.~\ref{fig:new_order_execs}.
+To illustrate some of these challenges, consider the behavior of the
+\C{new\_order} transaction when executed with a \emph{Read Committed}
+(RC) isolation level, the default isolation level in PostgreSQL, a
+widely used open-source database system.  An RC transaction is
+isolated from \emph{dirty writes}, i.e., writes of uncommitted
+transactions, but is allowed to witness the writes of concurrent
+transactions as soon as they are committed. Thus, with two concurrent
+instances of the \C{new\_order} transaction (call them $T_1$ and
+$T_2$), both concurrently placing new orders for different customers
+under the same district (\C{d\_id}), PostgreSQL admits the two
+executions shown in Fig.~\ref{fig:new_order_execs}.
 
@@
   \caption{\small RC executions involving two instances ($T_1$ and
-  $T_2$) of the \C{new\_order} transaction of
+  $T_2$) of the \C{new\_order} transaction depicted in
   Fig.~\ref{fig:new_order_code}. 
-  Each instance reads the \C{d\_id} \C{District} record twice, second
+  Each instance reads the \C{d\_id} \C{District} record twice, the second
   time to (atomically) update the \C{d\_next\_o\_id} field.}
-\label{fig:rc-ex}
+\label{fig:new_order_execs}
 \end{figure}
 
-The figure depicts an execution as a series of read, write, and commit
+The figure depicts an execution as a series of transactional read, write, and commit
 operations. In the execution on the left, the \C{new\_order} instance
@@
 example shows often inadequate, reasoning about weak isolation in
-terms of low-level traces~\cite{adyaphd,gotsmanconcur15} confounds
-high-level reasoning and automation.  Another alternative would
-interleave weak isolation implementation details within the program,
-yielding a (more-or-less) conventional concurrent program that can be
-then subject to classical concurrent verification methods.
-Considering the size and complexity of real-world transaction systems,
-this approach is unlikely to scale.
+terms of low-level traces~\cite{adyaphd,gotsmanconcur15} complicates
+high-level reasoning.  A possible alternative would interleave weak
+isolation implementation details within the program, yielding a
+(more-or-less) conventional concurrent program that can be then
+subject to classical concurrent verification methods.  Considering the
+size and complexity of real-world transaction systems, this strategy
+is unlikely to scale.
 
@@
 isolation semantics (\emph{not} their implementations) to the
-application layer, providing a unified framework to simultaneously
+application layer, providing a principled framework to simultaneously
 reason about application invariants and isolation properties.  To
 illustrate this idea informally, consider how we might verify that
-$\C{new_order}$ is sound when executed under a \emph{Repeatable Read}
+\C{new\_order} is sound when executed under a \emph{Repeatable Read}
 (RR) isolation level.  PostgreSQL executes an RR transaction by taking
@@
 transaction attempts to update a record, a check is made to determine
-if the current version of the record in the database is same as the
+if the current version of the record in the database is the same as the
 snapshot version. If it is, an exclusive lock is obtained on the
@@
 \caption{Database state transitions corresponding to an execution of
-  an RR transaction $T_1$ on Postgres. $T_2$ and $T_3$ are concurrent
-  transactions. State has three data items, $a$, $b$ and $c$. }
+  an RR transaction $T_1$ on PostgresSQL. $T_2$ and $T_3$ are concurrent
+  transactions. The database state has three data items, $a$, $b$ and $c$. }
 \label{fig:rr-postgres}
@@
   \caption{An abstract execution equivalent to the one in
-  Fig.~\ref{fig:rr-postgres}. Database state remains invariant as long as
+  Fig.~\ref{fig:rr-postgres}. The database state remains invariant as long as
   $T_1$ executes. The effects of concurrent transactions ($T_2$ and
@@
 
-Although this implementation, comprising many thousands of lines code,
-is highly involved, its semantic behavior in terms of how it effects
-transitions on the database state is fairly simple.  Since a
+Although this implementation, comprising many thousands of lines of
+code, is highly involved, its semantic behavior in terms of how it
+effects transitions on the database state is fairly simple.  Since a
 transaction's reads are always served from a snapshot, no state
@@
 it commits, the transaction is able to witness the effects of all
-concurrent transactions.  The PostgreSQL RR implementation effectively
-constrains this transition.
+concurrent transactions that committed before it.  The PostgreSQL RR
+implementation effectively constrains this transition.
 
@@
 current transaction cannot be overwritten by a concurrent transaction.
-If $\Delta$ is the state of the database when an RR transactions
+If $\Delta$ is the state of the database when an RR transaction
 finishes, and $\Delta_c$ is the state visible to the transaction at
@@
 concurrency extensions of the state transformer semantics is also
-covered in Sec.~\ref{sec:automation}.  The immediate two sections,
+covered in Sec.~\ref{sec:automation}.  The following two sections,
 however, focus on laying theoretical foundations for reasoning about

### 2017-07-01 motivation.tex

@@
 \includegraphics[scale=0.5]{Figures/RR-abstract}
-  \caption{An abstract execution equivalent to the one in
-  Fig.~\ref{fig:rr-postgres}. The database state remains invariant as long as
-  $T_1$ executes. The effects of concurrent transactions ($T_2$ and
-  $T_3$) are witnessed just before the commit.}
+  \caption{An abstract execution that includes the concrete execution
+  shown in Fig.~\ref{fig:rr-postgres}. It has no locks or snapshots,
+  admits fewer interleavings, yet results in the same post-state.  }
 \label{fig:rr-abstract}

### 2017-07-02 opsem.tex
Latest comments by KC and Kia addressed.
Fig. 4 redone.
RG Rules added.

@@
 
-Fig.~\ref{fig:txnimp} shows the syntax and small-step semantics of
-\txnimp, a core language that we will use to formalize the intuitions
-presented in the previous section. Natural numbers, (shared) variables
-and arithmetic expressions constitute the syntactic class of
-expressions ($e$).  Commands ($c$) include $\cskip$, assignment
-statements, transaction (\C{txn}) lexical blocks, and their sequential
-and parallel composition. We let $T_i$ for $i \in \mathbb{N}$ range
-over transaction identifiers. When it is evident we are referring to a
-transaction, we use the number $i$ instead of $T_i$ for identification
-(\eg in $\C{txn}\langle i \rangle$). For notational convenience, we
-let $t$ range over both expressions and commands.
+% Fig.~\ref{fig:txnimp} shows the syntax and small-step semantics of
+% \txnimp, a core language that we will use to formalize the intuitions
+% presented in the previous section. Natural numbers, (shared) variables
+% and arithmetic expressions constitute the syntactic class of
+% expressions ($e$).  Commands ($c$) include $\cskip$, assignment
+% statements, transaction (\C{txn}) lexical blocks, and their sequential
+% and parallel composition. We let $T_i$ for $i \in \mathbb{N}$ range
+% over transaction identifiers. When it is evident we are referring to a
+% transaction, we use the number $i$ instead of $T_i$ for identification
+% (\eg in $\C{txn}\langle i \rangle$). Like variables, transaction
+% identifiers are globally accessible. For notational convenience, we
+% let $t$ range over both expressions and commands.
 
-We define a small-step operational semantics for this language in
-terms of an abstract machine that generates an execution trace ($\E$).
-The first component of the trace is a set ($\A$) of \emph{effects},
-where an effect ($\eta$) is a tuple of (a). the identifier ($T_i$) of
-the transaction that generated this effect, (b). an indentifier ($j$)
-that uniquely identifies this effect, (c). the operation ($\op$)
-documented by this effect, which can be either a read (\C{RD(X)}), a
-write (\C{WR(X)}), or a transaction commit (\C{COMMIT}), and (d). the
-value, if any, associated with the operation (\eg, value read or value
-written). Functions $\txn$, $\id$, $\oper$, and $\rval$ project each
-of the four aforementioned individual components of the tuple.
-In every step of the evaluation, the machine reduces a \txnimp\ term
-by executing a read, write or commit operation, generating an effect,
-and extending the trace. Since effects include transaction
-identifiers, the semantics distinguishes between terms ($t$) of
-different transactions. For example, $\txnbox{t}_i$ denotes a term $t$
-inside a transaction $T_i$.  Evaluation contexts are also
-appropriately marked. For example, $\ectx_i$ denotes the evaluation
-context for a term inside $T_i$. The other component of an execution
-trace is a visibility relation ($\visZ$) that establishes a visibility
-property between effects among different transactions. The intent and
-mechanics of $\visZ$ is described in the sequel.
+% We define a small-step operational semantics for this language in
+% terms of an abstract machine that generates an execution trace ($\E$).
+% The first component of the trace is a set ($\A$) of \emph{effects},
+% where an effect ($\eta$) is a tuple of (a). the identifier ($T_i$) of
+% the transaction that generated this effect, (b). the unique
+% identifier ($j$) of this effect, (c). the operation ($\op$)
+% documented by this effect, which can be either a read (\C{RD(X)}), a
+% write (\C{WR(X)}), or a transaction commit (\C{COMMIT}), and (d). the
+% value, if any, associated with the operation (\eg, value read or value
+% written). Functions $\txn$, $\id$, $\oper$, and $\rval$ project each
+% of the four aforementioned individual components of the tuple.  In
+% every step of the evaluation, the machine reduces a \txnimp\ term by
+% executing a read, write or commit operation, generating an effect, and
+% extending the trace. Since effects include transaction identifiers,
+% the semantics distinguishes between terms ($t$) of different
+% transactions. For example, $\txnbox{t}_i$ denotes a term $t$ inside a
+% transaction $T_i$.  Evaluation contexts are also appropriately marked.
+% For example, $\ectx_i$ denotes the evaluation context for a term
+% inside $T_i$. The other component of an execution trace is a
+% visibility relation ($\visZ$) that establishes a visibility property
+% between effects among different transactions. The intent and mechanics
+% of $\visZ$ is described in the sequel.
 
-Fundamental to our development is the notion of a trace invariant
-($\I$). $\I$ is a function from traces ($\E$) to first-order logical
-formulas ($\Prop$) that define well-formedness
-constraints over traces.  The machine takes a step only if the
-resulting trace satisfies the constraints imposed by $\I$. This
-behaviour is captured by the auxiliary reduction rule
-\rulelabel{E-Aux} that factors out the trace extension aspect of the
-evaluation by abstracting away the operation-specific behaviour as a
-function that generates an appropriate effect. We let $\mathcal{F}$
-denote this function.  \rulelabel{E-Aux} uses $\mathcal{F}$ to
-generate a new effect and extend the trace ($\E = (\A,\visZ)$)
-\emph{only if} the well-formedness constraints imposed by $\I$ on $\E$
-(i.e., $\I(\E')$) are satisfied. Otherwise, it gets stuck. In an
-execution that runs to completion, every small-step preserves the
-well-formedness of a trace, thus ensuring the invariance of $\I$.
-Note that the semantics makes no assumptions about $\I$ other than its
-type. As such, it can be instantiated with any trace-parametric
-proposition that expresses constraints over the given trace. For
-instance, consider the $\psi_{RC}$ specification from
-\S\ref{sec:motivation}, but with bounded $T_1$ and $T_2$ instantiated
-with \C{Wd1} and \C{Wd2}, respectively. The instantiated specification
-is the following term:\vspace*{-8pt}
+% Fundamental to our development is the notion of a trace invariant
+% ($\I$). $\I$ is a function from traces ($\E$) to first-order logical
+% formulas ($\Prop$) that define well-formedness
+% constraints over traces.  The machine takes a step only if the
+% resulting trace satisfies the constraints imposed by $\I$. This
+% behaviour is captured by the auxiliary reduction rule
+% \rulelabel{E-Aux} that factors out the trace extension aspect of the
+% evaluation by abstracting away the operation-specific behaviour as a
+% function that generates an appropriate effect. We let $\mathcal{F}$
+% denote this function.  \rulelabel{E-Aux} uses $\mathcal{F}$ to
+% generate a new effect and extend the trace ($\E = (\A,\visZ)$)
+% \emph{only if} the well-formedness constraints imposed by $\I$ on $\E$
+% (i.e., $\I(\E')$) are satisfied. Otherwise, it gets stuck. In an
+% execution that runs to completion, every small-step preserves the
+% well-formedness of a trace, thus ensuring the invariance of $\I$.
+% Note that the semantics makes no assumptions about $\I$ other than its
+% type. As such, it can be instantiated with any trace-parametric
+% proposition that expresses constraints over the given trace. For
+% instance, consider the $\psi_{RC}$ specification from
+% \S\ref{sec:motivation}, but with bounded $T_1$ and $T_2$ instantiated
+% with \C{Wd1} and \C{Wd2}, respectively. The instantiated specification
+% is the following term:\vspace*{-8pt}
 
-\begin{smathpar}
-\begin{array}{l}
-  \forall \eta_1,\eta_2,.\; \txn(\eta_1) = \C{Wd1}
-  \conj \txn(\eta_2) = \C{Wd2} \\
-  \hspace*{0.6in}\conj \C{Wd1} \neq \C{Wd2} \conj \eta_1 \hboar
-  \eta_2 \Rightarrow \C{Wd1} \hboar \eta_2 \\
-\end{array}
-\end{smathpar}
+% \begin{smathpar}
+% \begin{array}{l}
+%   \forall \eta_1,\eta_2.\; \txn(\eta_1) = \C{Wd1}
+%   \conj \txn(\eta_2) = \C{Wd2} \\
+%   \hspace*{0.6in}\conj \C{Wd1} \neq \C{Wd2} \conj \eta_1 \hboar
+%   \eta_2 \Rightarrow \C{Wd1} \hboar \eta_2 \\
+% \end{array}
+% \end{smathpar}
 
-\noindent It is easy to interpret the above specification in the
-context of a trace $\E$ that captures an execution of the program in
-Fig.~\ref{fig:motiv-eg-1}. Such a trace-parametric formula can be
-used to instantiate the trace invariant $\I$ in Fig.~\ref{fig:txnimp}.
-The resultant operational semantics describes an abstract machine that
-gets stuck if an operation of \C{Wd2} is executed in a state that
-incorporates some, but not all the effects (including the \C{COMMIT})
-of \C{Wd1}.
-% While the machine takes a step only if the constraints are
-% satisfied, it neither defines nor explicitly assumes an oracle to
-% check satisfaction.
+% \noindent It is easy to interpret the above specification in the
+% context of a trace $\E$ that captures an execution of the program in
+% Fig.~\ref{fig:motiv-eg-1}. Such a trace-parametric formula can be
+% used to instantiate the trace invariant $\I$ in Fig.~\ref{fig:txnimp}.
+% The resultant operational semantics describes an abstract machine that
+% gets stuck if an operation of \C{Wd2} is executed in a state that
+% incorporates some, but not all the effects (including the \C{COMMIT})
+% of \C{Wd1}.
+% % While the machine takes a step only if the constraints are
+% % satisfied, it neither defines nor explicitly assumes an oracle to
+% % check satisfaction.
 
-As described in \S\ref{sec:motivation}, the semantics of various
-isolation levels can be captured as constraints over the
-happens-before ($\hbZ$) relation. $\hbZ$ is however a derived relation
-in our model, composed of more fundamental \emph{session order}
-($\soZ$) and \emph{visibility} ($\visZ$) relations. In particular,
-$\hbZ \,=\, (\soZ \cup \visZ)^{+}$. Unlike $\hbZ$, $\visZ$ (defined
-below) is not a transitive relation, and hence lets us capture
-finer-grained isolation properties than $\hbZ$, which we leverage in
-our development.  The session order relation captures the sequential
-order of operations within a transaction. In particular, it relates
-two effects, $\eta_1$ and $\eta_2$, such that $\txn(\eta_1) =
-\txn(\eta_2)$ and $\id(\eta_1) < \id(\eta_2)$.  The semantics assigns
-monotonically increasing identifiers to effects, as defined by the
-$\id(\eta) > \maxId(\A)$ condition of \rulelabel{E-Aux} ($\maxId(\A)$
-returns the maximum number identifying an effect in $\A$). Evaluation
-contexts ($\ectx_i$) for transaction-bound terms are defined so as to
-enforce a deterministic sequential order of execution within a
-transaction, leading to a deterministic total order among effect ids,
-which defines the session order relation. Visibility ($\visZ$) on the
-other hand relates effects across concurrent transactions.
-Intuitively, $\visZ$ relates $\eta_1$ to $\eta_2$ if and only if
-$\eta_1$ was \emph{visible} to the operation that generated $\eta_2$
-during its execution, thus effecting its return value
-($\rval(\eta_2)$). For example, a read operation over \C{X} may pick
-the value ($\rval$) of the write effect with the highest id among its
-visible effects (this is made possible by appropriately defining
-$\interp{\cdot}$ in \rulelabel{E-Var}, as we show later). Thus, the
-value of a read depends on what write effects it can witness. An
-operation can only witness the effects of already concluded
-operations, which varies between executions due to the
-non-deterministic order of evaluating the parallel composition of
-transactions.
+% As described in \S\ref{sec:motivation}, the semantics of various
+% isolation levels can be captured as constraints over the
+% happens-before ($\hbZ$) relation. $\hbZ$ is however a derived relation
+% in our model, composed of more fundamental \emph{session order}
+% ($\soZ$) and \emph{visibility} ($\visZ$) relations. In particular,
+% $\hbZ \,=\, (\soZ \cup \visZ)^{+}$. Unlike $\hbZ$, $\visZ$ (defined
+% below) is not a transitive relation, and hence lets us capture
+% finer-grained isolation properties than $\hbZ$, which we leverage in
+% our development.  The session order relation captures the sequential
+% order of operations within a transaction. In particular, it relates
+% two effects, $\eta_1$ and $\eta_2$, such that $\txn(\eta_1) =
+% \txn(\eta_2)$ and $\id(\eta_1) < \id(\eta_2)$.  The semantics assigns
+% monotonically increasing identifiers to effects, as defined by the
+% $\id(\eta) > \maxId(\A)$ condition of \rulelabel{E-Aux} ($\maxId(\A)$
+% returns the maximum number identifying an effect in $\A$). Evaluation
+% contexts ($\ectx_i$) for transaction-bound terms are defined so as to
+% enforce a deterministic sequential order of execution within a
+% transaction, leading to a deterministic total order among effect ids,
+% which defines the session order relation. Visibility ($\visZ$) on the
+% other hand relates effects across concurrent transactions.
+% Intuitively, $\visZ$ relates $\eta_1$ to $\eta_2$ if and only if
+% $\eta_1$ was \emph{visible} to the operation that generated $\eta_2$
+% during its execution, thus effecting its return value
+% ($\rval(\eta_2)$). For example, a read operation over \C{X} may pick
+% the value ($\rval$) of the write effect with the highest id among its
+% visible effects (this is made possible by appropriately defining
+% $\interp{\cdot}$ in \rulelabel{E-Var}, as we show later). Thus, the
+% value of a read depends on what write effects it can witness. An
+% operation can only witness the effects of already concluded
+% operations, which varies between executions due to the
+% non-deterministic order of evaluating the parallel composition of
+% transactions.
 
-A more notable source of non-determinism, however, is the
-\rulelabel{E-Aux} rule, which allows the machine to expose an
-arbitrary subset ($S$) of existing effects ($\A$) to the incoming
-operation. In other words, the machine is not obligated to reveal the
-effects of all previous operations to an incoming operation. This
-relaxation allows the abstract machine to model the semantics of
-weakly-consistent data stores. For instance, operations issued to an
-eventually consistent ({\sc ec}) replicated store could be dispatched
-to different replicas whose states may not be in any well-defined
-relationship. By allowing operations to witness arbitrary subsets of
-the global state, the semantics models the weak visibility properties
-of such stores; we elaborate on the implication of this style of
-definition in \S\ref{sec:store-consistency}.  Stronger visibility
-properties can be expressed by imposing well-formedness constraints
-over $\visZ$ via the trace invariant ($\I$). Since the abstract
-machine is obligated to satisfy $\I$ at every step of the execution,
-operations are guaranteed to experience the level of isolation
-specified by $\I$.  Thus, in executions that run to completion, the
-abstract machine models a store that provides the required levels of
-isolation.  Notably, the machine achieves this without defining an
-operational semantics for isolation levels, instead solely relying on
-their declarative characterization as trace well-formedness
-constraints to enforce isolation guarantees.
-\S\ref{sec:ansi-isolation} specifies various ANSI SQL isolation levels
-stated as trace well-formedness constraints.
+% A more notable source of non-determinism, however, is the
+% \rulelabel{E-Aux} rule, which allows the machine to expose an
+% arbitrary subset ($S$) of existing effects ($\A$) to the incoming
+% operation. In other words, the machine is not obligated to reveal the
+% effects of all previous operations to an incoming operation. This
+% relaxation allows the abstract machine to model the semantics of
+% weakly-consistent data stores. For instance, operations issued to an
+% eventually consistent ({\sc ec}) replicated store could be dispatched
+% to different replicas whose states may not be in any well-defined
+% relationship. By allowing operations to witness arbitrary subsets of
+% the global state, the semantics models the weak visibility properties
+% of such stores; we elaborate on the implication of this style of
+% definition in \S\ref{sec:store-consistency}.  Stronger visibility
+% properties can be expressed by imposing well-formedness constraints
+% over $\visZ$ via the trace invariant ($\I$). Since the abstract
+% machine is obligated to satisfy $\I$ at every step of the execution,
+% operations are guaranteed to experience the level of isolation
+% specified by $\I$.  Thus, in executions that run to completion, the
+% abstract machine models a store that provides the required levels of
+% isolation.  Notably, the machine achieves this without defining an
+% operational semantics for isolation levels, instead solely relying on
+% their declarative characterization as trace well-formedness
+% constraints to enforce isolation guarantees.
+% \S\ref{sec:ansi-isolation} specifies various ANSI SQL isolation levels
+% stated as trace well-formedness constraints.
 
-% TRACE RELATIONS
-% ---------------
-\input{trace-relations}
+% % TRACE RELATIONS
+% % ---------------
+% \input{trace-relations}
 
-% However, a machine that lets operations witness arbitrary subset of
-% the global state offers no isolation whatsoever. For example, it may
-% allow a read operation to witness writes of an uncommitted
-% transaction, violating RC isolation. Fortunately, our ability to
-% express an isolation semantics as constraints over happens-before
-% order through $\visZ$ and $\soZ$ relations, and the property of the
-% abstract machine to be parametric over the trace invariant ($\I$),
-% lets us solve this problem.
-% In particular, we continue to define isolation semantics as
-% constraints over $\visZ$ and $\soZ$, but confine their domain of
-% interpretation to the given trace so that they now become trace
-% well-formedness constraints. Well-formedness constraints can be
-% combined into a trace invariant ($\I$).
+% % However, a machine that lets operations witness arbitrary subset of
+% % the global state offers no isolation whatsoever. For example, it may
+% % allow a read operation to witness writes of an uncommitted
+% % transaction, violating RC isolation. Fortunately, our ability to
+% % express an isolation semantics as constraints over happens-before
+% % order through $\visZ$ and $\soZ$ relations, and the property of the
+% % abstract machine to be parametric over the trace invariant ($\I$),
+% % lets us solve this problem.
+% % In particular, we continue to define isolation semantics as
+% % constraints over $\visZ$ and $\soZ$, but confine their domain of
+% % interpretation to the given trace so that they now become trace
+% % well-formedness constraints. Well-formedness constraints can be
+% % combined into a trace invariant ($\I$).
 
-As described previously, \rulelabel{E-Aux} abstracts away the
-operation-specific behaviour of a machine step as a function ($\F$)
-that accepts a set ($S$) of effects chosen by the machine to make
-visible to the operation, interprets the operation w.r.t. $S$, and
-returns an appropriate effect that encodes its return value. Rules
-\rulelabel{E-Var}, \rulelabel{E-Asgn} and \rulelabel{E-Commit} define
-such functions for read, write and commit operations, respectively.
-The effect returned by the function in each case includes its
-transaction id ($T_i$) along with an arbitrarily chosen effect id
-($j$) that is later verified to be unique in \rulelabel{E-Aux}. The
-$\rval$ for a write is the value being written, and for commit it is
-$\bot$. In case of a read, the value read depends on how the read
-operation chooses to interpret the given set ($S$) of visible effects.
-The interpretation may depend on application semantics. For example, a
-monotonically increasing counter application may choose to let a write
-with the largest value determine the value of a read. To accommodate
-multiple interpretations, the semantics is made parametric over an
-interpretation function ($\interp{\cdot}$) that accepts a set of
-effects and a variable name, and returns the value associated with the
-variable. A straightforward interpretation function that chooses the
-last write (i.e., write with largest id) is shown below:
+% As described previously, \rulelabel{E-Aux} abstracts away the
+% operation-specific behaviour of a machine step as a function ($\F$)
+% that accepts a set ($S$) of effects chosen by the machine to make
+% visible to the operation, interprets the operation w.r.t. $S$, and
+% returns an appropriate effect that encodes its return value. Rules
+% \rulelabel{E-Var}, \rulelabel{E-Asgn} and \rulelabel{E-Commit} define
+% such functions for read, write and commit operations, respectively.
+% The effect returned by the function in each case includes its
+% transaction id ($T_i$) along with an arbitrarily chosen effect id
+% ($j$) that is later verified to be unique in \rulelabel{E-Aux}. The
+% $\rval$ for a write is the value being written, and for commit it is
+% $\bot$. In case of a read, the value read depends on how the read
+% operation chooses to interpret the given set ($S$) of visible effects.
+% The interpretation may depend on application semantics. For example, a
+% monotonically increasing counter application may choose to let a write
+% with the largest value determine the value of a read. To accommodate
+% multiple interpretations, the semantics is made parametric over an
+% interpretation function ($\interp{\cdot}$) that accepts a set of
+% effects and a variable name, and returns the value associated with the
+% variable. A straightforward interpretation function that chooses the
+% last write (i.e., write with largest id) is shown below:
 
-\begin{smathpar}
-\begin{array}{lcl}
-  \isMax(S,\eta) & \Leftrightarrow &  \forall (\eta'\in S).  
-  \kind(\eta') = \kind(\eta) \\
-  & & \hspace*{0.4in}\Rightarrow \eta' = \eta \disj \id(\eta') < \id(\eta)\\
+% \begin{smathpar}
+% \begin{array}{lcl}
+%   \isMax(S,\eta) & \Leftrightarrow &  \forall (\eta'\in S).  
+%   \kind(\eta') = \kind(\eta) \\
+%   & & \hspace*{0.4in}\Rightarrow \eta' = \eta \disj \id(\eta') < \id(\eta)\\
 
-\interp{S}(X) & = & \C{if}\;(\exists (\eta \in S). \kind(\eta) = \C{WR}(X) 
-  \wedge \isMax(S,\eta)) \\
-  & & \C{then}\;\rval(\eta)\;\C{else}\;0\\
-\end{array}
-\end{smathpar}
+% \interp{S}(X) & = & \C{if}\;(\exists (\eta \in S). \kind(\eta) = \C{WR}(X) 
+%   \wedge \isMax(S,\eta)) \\
+%   & & \C{then}\;\rval(\eta)\;\C{else}\;0\\
+% \end{array}
+% \end{smathpar}
 
-\noindent Rules \rulelabel{E-Top-Ctx} and \rulelabel{E-Txn-Ctx} define
-congruence properties for top-level terms and transaction-bound terms,
-respectively. The rules and evaluation contexts ($\ectx$ and
-$\ectx_i$) are defined such that only certain kinds of terms are
-allowed at the top-level and inside a transaction. In particular, a
-\txnimp program at the top-level can either be a transaction, or a
-parallel composition of transactions. A command inside a \C{txn} block
-can either be an assignment, or a sequential composition of
-assignments.
+% \noindent Rules \rulelabel{E-Top-Ctx} and \rulelabel{E-Txn-Ctx} define
+% congruence properties for top-level terms and transaction-bound terms,
+% respectively. The rules and evaluation contexts ($\ectx$ and
+% $\ectx_i$) are defined such that only certain kinds of terms are
+% allowed at the top-level and inside a transaction. In particular, a
+% \txnimp program at the top-level can either be a transaction, or a
+% parallel composition of transactions. A command inside a \C{txn} block
+% can either be an assignment, or a sequential composition of
+% assignments.
 
-\input{ansi-isolation}
+% \input{ansi-isolation}

### 2017-07-02 reasoning.tex
RG Rules added.

--- 2016-07-06 reasoning.tex
@@
 
-We now describe a proof system that lets us demonstrate that a \txnimp
-program satisfies its high-level invariants in all small-step
-executions that satisfy a chosen trace invariant ($\I$).\footnote{It is
-important to keep in mind the distinction between high-level invariants
-(\eg~\C{X$\ge$0}) and trace invariants (\eg~\C{RC}). The former
-constitute \emph{proof} \emph{obligations} for the programmer, whereas
-the latter captures \emph{assumptions} about the operational characteristics
-of transactions and the underlying data store. } 
-Since the trace invariant captures transaction-isolation constraints,
-the demonstration is a proof that the given selection of isolation
-levels for transactions is sufficient to enforce a program's invariants.
+% We now describe a proof system that lets us demonstrate that a \txnimp
+% program satisfies its high-level invariants in all small-step
+% executions that satisfy a chosen trace invariant ($\I$).\footnote{It is
+% important to keep in mind the distinction between high-level invariants
+% (\eg~\C{X$\ge$0}) and trace invariants (\eg~$\I =
+% \lambda\E.\,\C{RC}(\C{Wd1})\wedge \C{SI}(\C{Wd2})$). The former
+% constitute \emph{proof} \emph{obligations} for the programmer, whereas
+% the latter is a combination of transaction-specific \emph{assumptions} about 
+% operational characteristics of the underlying data store. } 
+% Since the trace invariant captures transaction-isolation constraints,
+% the demonstration is a proof that the given selection of isolation
+% levels for transactions is sufficient to enforce a program's invariants.
 
-As mentioned earlier, our proof system is a rely-guarantee framework
-that admits compositional reasoning by abstracting away environmental
-interference as a rely relation. A conspicuous difference between a
-standard development of rely-guarantee and ours is that, while the
-former reasons in terms of program states (variable to value
-bindings), we reason in terms of executions as captured by their
-traces ($\E$). In particular, our rely ($R$) and guarantee ($G$)
-relations relate executions (i.e., $R,\,Q\,\subseteq\,\E\times\E$),
-and our pre- ($P$) and post- ($Q$) conditions are assertions over
-executions (i.e., $P,\,Q\, : \E \rightarrow \Prop$). Our development
-also facilitates state-based reasoning via the interpretation function
-($\interp{\cdot}$) introduced in \S\ref{sec:syntax}, which interprets
-the given set of effects as a state. For example, if a \emph{bi-state}
-rely relation relates every pair of states $\sigma$ and $\sigma'$ such
-that $\sigma'(X) \ge \sigma(X) \ge 0$, the corresponding
-\emph{bi-execution} rely relation relates every pair of executions
-$\E$ and $\E'$ such that $\interp{\E.\A}(X) \ge \interp{\E'.\A}(X) \ge
-0$ . Assertions on states are also written similarly. For instance,
-consider the post-condition $\C{B\,=\,k-a1-a2}$ of the program in
-Fig.~\ref{fig:motiv-eg-1}. The corresponding assertion on the
-post-state ($\lambda\sigma.~\sigma(\C{B}) \,=\, \C{k-a1-a2}$) asserts
-that in all states resulting from executing the program, the value of
-\C{B} is \C{k-a1-a2}. The equivalent execution-based assertion
-($\lambda\E.~\interp{\E.\A}(\C{B})\,=\,\C{k-a1-a2}$) asserts that in
-all executions of the program, the value of \C{B} is \C{k-a1-a2}.
-However, having access to an execution facilitates assertions that go
-beyond the state. An example is an invariant of \C{Wd1} described
-abstractly in \S\ref{sec:motivation}, and reified as an
-execution-based assertion below:
+% As mentioned earlier, our proof system is a rely-guarantee framework
+% that admits compositional reasoning by abstracting away environmental
+% interference as a rely relation. A conspicuous difference between a
+% standard development of rely-guarantee and ours is that, while the
+% former reasons in terms of program states (variable to value
+% bindings), we reason in terms of executions as captured by their
+% traces ($\E$). In particular, our rely ($R$) and guarantee ($G$)
+% relations relate executions (i.e., $R,\,G\,\subseteq\,\E\times\E$),
+% and our pre- ($P$) and post- ($Q$) conditions are assertions over
+% executions (i.e., $P,\,Q\, : \E \rightarrow \Prop$). Our development
+% also facilitates state-based reasoning via the interpretation function
+% ($\interp{\cdot}$) introduced in \S\ref{sec:syntax}, which interprets
+% the given set of effects as a state. For example, if a \emph{bi-state}
+% rely relation relates every pair of states $\sigma$ and $\sigma'$ such
+% that $\sigma'(X) \ge \sigma(X) \ge 0$, the corresponding
+% \emph{bi-execution} rely relation relates every pair of executions
+% $\E$ and $\E'$ such that $\interp{\E.\A}(X) \ge \interp{\E'.\A}(X) \ge
+% 0$ . Assertions on states are also written similarly. For instance,
+% consider the post-condition $\C{B\,=\,k-a1-a2}$ of the program in
+% Fig.~\ref{fig:motiv-eg-1}. The corresponding assertion on the
+% post-state ($\lambda\sigma.~\sigma(\C{B}) \,=\, \C{k-a1-a2}$) asserts
+% that in all states resulting from executing the program, the value of
+% \C{B} is \C{k-a1-a2}. The equivalent execution-based assertion
+% ($\lambda\E.~\interp{\E.\A}(\C{B})\,=\,\C{k-a1-a2}$) asserts that in
+% all executions of the program, the value of \C{B} is \C{k-a1-a2}.
+% However, having access to an execution facilitates assertions that go
+% beyond the state. An example is an invariant of \C{Wd1} described
+% abstractly in \S\ref{sec:motivation}, and reified as an
+% execution-based assertion below:
 
-\begin{smathpar}
-\begin{array}{l}
-  \lambda\E.~ \neg(\underE{\committed(\C{Wd1}}) \conj 
-      \underE{\C{Wd1} \wrstoar \C{B}}\\
-  \hspace*{0.2in}\conj(\neg(\underE{\committed(\C{Wd2})}) \Rightarrow 
-      \C{\interp{\E.\A}(B) = k-a1}) \\
-   \hspace*{0.2in}\conj (\underE{\committed(\C{Wd2})} \Rightarrow 
-      \C{\interp{\E.\A}(B) = k-a1-a2})
-\end{array}
-\end{smathpar}
-% Where, $\committed$ is defined as:
+% \begin{smathpar}
+% \begin{array}{l}
+%   \lambda\E.~ \neg(\underE{\committed(\C{Wd1}}) \conj 
+%       \underE{\C{Wd1} \wrstoar \C{B}}\\
+%   \hspace*{0.2in}\conj(\neg(\underE{\committed(\C{Wd2})}) \Rightarrow 
+%       \C{\interp{\E.\A}(B) = k-a1}) \\
+%    \hspace*{0.2in}\conj (\underE{\committed(\C{Wd2})} \Rightarrow 
+%       \C{\interp{\E.\A}(B) = k-a1-a2})
+% \end{array}
+% \end{smathpar}
+% % Where, $\committed$ is defined as:
+% % \begin{smathpar}
+% % \begin{array}{lcl}
+% % \underE{\committed(T_i)} & \Leftrightarrow &
+% %   \exists\eta.~\underE{\eta\in T_i} \conj \kind(\eta) = \C{COMMIT}\\
+% % \end{array}
+% % \end{smathpar}
+
+% \noindent One of the conjuncts asserts that \C{Wd1} has written to \C{B} - a
+% fact which cannot be deduced solely from the value of \C{B} (esp. in the
+% presence of interference), but can be expressed as a proposition over
+% the execution trace.
+
+% \subsection{The Rely-Guarantee Judgment}
+% \label{sec:rely-guarantee}
+
+\input{rg-rules}
+
+% The standard rely-guarantee judgment is a quintuple
+% $\rg{P,R}{c}{G,Q}$, which informally asserts that if a command $c$ is
+% executed in a state that satisfies its pre-condition $P$, provided
+% that every interference step during the execution is contained inside
+% the rely relation $R$, then the effect that each step of executing $c$
+% has on the state is captured by $G$, and the final state of execution
+% satisfies the post-condition $Q$. The terms ``execution'' and ``step''
+% are defined with respect to an operational semantics, which, in our
+% case, is parameterized on the trace invariant $\I$. Consequently, our
+% rely-guarantee judgment for commands is the sextuple
+% $\rg{\I,P,R}{c}{G,Q}$, whose semantics differs from the quintuple in
+% that it requires every step of the execution to additionally preserve
+% the trace invariant $\I$ for the post-condition to be valid.  As
+% usual, if $c$ is inside a transaction $T_i$, we write $\txnbox{c}_i$.
+% \txnimp expressions are side-effecting (they generate \C{RD} effects),
+% and admit interference during their evaluation.  Therefore, like
+% commands, expressions need to be reasoned about explicitly. But,
+% unlike commands, expressions evaluate to values, and the reasoning
+% framework should also admit assertions on such values.  We therefore
+% define a separate judgment for expressions - the septuple
+% $\rg{\I,P,R}{e}{G,C,Q}$. The new entrant $C : \mathbb{N} \rightarrow
+% \Prop$ is an assertion on the return value of $e$.
+
+% The rules that define rely-guarantee judgment are shown in
+% Fig.~\ref{fig:rg-rules}. Like standard rely-guarantee definitions,
+% these definitions also require a \emph{stability} condition, which
+% requires pre- and post- conditions to hold despite any interference
+% from concurrent threads (captured by $R$). Stability can be predicated
+% on the assumption that interference preserves the trace invariant
+% $\I$. Formally:\vspace*{-10pt}
+
 % \begin{smathpar}
 % \begin{array}{lcl}
-% \underE{\committed(T_i)} & \Leftrightarrow &
-%   \exists\eta.~\underE{\eta\in T_i} \conj \kind(\eta) = \C{COMMIT}\\
+%   \underI{\stable(R,P)} & \defeq & \forall \E, \E'.\, 
+%   \I(\E) \conj P(\E) \conj R(\E,\E') \\
+%   &   & \hspace*{1in}\conj \I(\E') \Rightarrow P(\E')\\
 % \end{array}
@@
 
-\noindent One of the conjuncts asserts that \C{Wd1} has written to \C{B} - a
-fact which cannot be deduced solely from the value of \C{B} (esp. in the
-presence of interference), but can be expressed as a proposition over
-the execution trace.
+% \noindent However, the assumption that interference preserves $\I$ (or,
+% dually, $\I$ withstands interference) needs to be justified
+% separately. We call this the stability requirement on $\I$:\vspace*{-10pt}
 
-\subsection{The Rely-Guarantee Judgment}
-\label{sec:rely-guarantee}
+% \begin{smathpar}
+% \begin{array}{lcl}
+% \stable(R,\I)& \defeq & \forall \E, \E'.\, 
+%   \I(\E) \conj R(\E,\E')\Rightarrow \I(\E')\\
+% \end{array}
+% \end{smathpar}
 
-\input{rg-rules}
+% \noindent The rule \rulelabel{RG-Var} defines the rely-guarantee judgment for
+% shared variable reads inside a transaction $T_i$. It requires $\I$ to
+% be stable, and pre- and post- conditions to be stable relative to
+% $\I$.  The quantified premise effectively requires a proof that if the
+% abstract machine of Fig.~\ref{fig:txnimp} takes a step starting from
+% an execution $\E$ that satisfies the pre-condition $P$, then the
+% resultant execution $\E'$ satisfies the post-condition $Q$, and that
+% the guarantee $G$ faithfully captures the transition from $\E$ to
+% $\E'$. It is informative to compare this premise with the premise of
+% the \rulelabel{E-Aux} reduction rule of Fig.~\ref{fig:txnimp}. Similar
+% premises also appear in the \rulelabel{RG-Asgn} and \rulelabel{RG-Txn}
+% rules, which define rely-guarantee judgments for assignments and
+% transactions, respectively. \rulelabel{RG-Var} however also requires
+% the return value ($\interp{S}(X)$) of the read to satisfy the
+% assertion $C$ meant for the value. \rulelabel{RG-Arith} defines the RG
+% judgment for an arithmetic expression $e_1\pm e_2$ in terms of the
+% corresponding judgments for the constituent expressions $e_1$ and
+% $e_2$. The quantified premise requires any value resulting from
+% evaluating $e_1 \pm e_2$ to satisfy the assertion $C$, provided that
+% $e_1$ and $e_2$ always evaluate to values that satisfy $C_1$ and
+% $C_2$, respectively. The rules for sequential and parallel composition
+% of commands are essentially the same as their counterparts in a
+% standard rely-guarantee formulation and hence elided.
 
-The standard rely-guarantee judgment is a quintuple
-$\rg{P,R}{c}{G,Q}$, which informally asserts that if a command $c$ is
-executed in a state that satisfies its pre-condition $P$, provided
-that every interference step during the execution is contained inside
-the rely relation $R$, then the effect that each step of executing $c$
-has on the state is captured by $G$, and the final state of execution
-satisfies the post-condition $Q$. The terms ``execution'' and ``step''
-are defined with respect to an operational semantics, which, in our
-case, is parameterized on the trace invariant $\I$. Consequently, our
-rely-guarantee judgment for commands is the sextuple
-$\rg{\I,P,R}{c}{G,Q}$, whose semantics differs from the quintuple in
-that it requires every step of the execution to additionally preserve
-the trace invariant $\I$ for the post-condition to be valid.  As
-usual, if $c$ is inside a transaction $T_i$, we write $\txnbox{c}_i$.
-\txnimp expressions are side-effecting (they generate \C{RD} effects),
-and admit interference during their evaluation.  Therefore, like
-commands, expressions need to be reasoned about explicitly. But,
-unlike commands, expressions evaluate to values, and the reasoning
-framework should also admit assertions on such values.  We therefore
-define a separate judgment for expressions - the septuple
-$\rg{\I,P,R}{e}{G,C,Q}$. The new entrant $C : \mathbb{N} \rightarrow
-\Prop$ is an assertion on the return value of $e$.
+% % RG judgment for the assignment $X:=e$ makes use of the corresponding
+% % judgment for the RHS expression $e$. The quantified premise asserts
+% % that evaluating the assigment after evaluating $e$ to a value $v$ and
+% % an execution $\E$, should result in an execution $\E'$ that satisfies
+% % the post-condition $Q$, while being related to $\E$ via the guarantee
+% % $G$. RG judgment for the \C{txn} lexical block is similar.  It uses
+% % the judgment for the transaction-bound command $c$ (i.e.,
+% % $\txnbox{c}_i$) to obtain an invariant $Q'$ for the execution $\E$
+% % before the commit, and verifies that committing the transaction under
+% % $\E$ results in an execution $\E'$ that satisfies transaction's
+% % post-condition $Q$. As usual, $\E$ and $\E'$ need to be related by $G$.
+% % The rules for sequential composition of transaction-bound comands
+% % (\rulelabel{RG-Seq}) and parallel composition of top-level commands
+% % (\rulelabel{RG-Par}) are straightforward, and more-or-less same as the
+% % corresponding rules in classical rely-guarantee.
 
-The rules that define rely-guarantee judgment are shown in
-Fig.~\ref{fig:rg-rules}. Like standard rely-guarantee definitions,
-these definitions also require a \emph{stability} condition, which
-requires pre- and post- conditions to hold despite any interference
-from concurrent threads (captured by $R$). Stability can be predicated
-on the assumption that interference preserves the trace invariant
-$\I$. Formally:\vspace*{-10pt}
+% The \rulelabel{RG-Conseq} rule defines ways to strengthen or
+% weaken relations and assertions associated with the RG judgment of
+% transaction-bound expressions. Similar rules exist for
+% transaction-bound and top-level commands, but are not
+% shown.\footnote{The supplementary materials provide the complete set
+%   of rules.} As is the case with a standard rely-guarantee
+% formulation,the rules allow the pre-condition $P$ and the rely relation
+% $R$ to be strengthened, and the post-condition $Q$ (also, $C$ in the
+% case of expressions) and the guarantee relation $G$ to be weakened.
+% The most notable aspect of the \rulelabel{RG-Conseq} rules is that
+% they allow the trace invariant $\I$ to be strengthened. Considering
+% that $\I$ captures isolation properties, this means that a program
+% proven correct under weaker isolation levels is also correct under
+% stronger ones.  Parametricity over the trace invariant $\I$, combined
+% with the ability to strengthen $\I$ as needed, allows our proof system
+% to support a highly flexible proof strategy to prove programs correct
+% over various isolation variants. For example, programmers can
+% \emph{define} isolation guarantees of their choice (by defining $\I$
+% appropriately) and then prove programs correct assuming the guarantees
+% hold.  The soundness of strengthening $\I$ ensures that a program can
+% be safely executed on any system that offers isolation guarantees
+% at least as strong as those assumed.
 
-\begin{smathpar}
-\begin{array}{lcl}
-  \underI{\stable(R,P)} & \defeq & \forall \E, \E'.\, 
-  \I(\E) \conj P(\E) \conj R(\E,\E') \\
-  &   & \hspace*{1in}\conj \I(\E') \Rightarrow P(\E')\\
-\end{array}
-\end{smathpar}
+% \subsection{Semantics and Soundness}
 
-\noindent However, the assumption that interference preserves $\I$ (or,
-dually, $\I$ withstands interference) needs to be justified
-separately. We call this the stability requirement on $\I$:\vspace*{-10pt}
+% \begin{definition}[\bfseries Step-indexed reflexive transitive closure]
+% For all $A:\text{Type}$, $R: A \rightarrow A \rightarrow \mathbb{P}$, and $n :
+% \mathbb{N}$, the step-indexed reflexive transitive closure $R^n$ of $R$ is
+% the smallest relation satisfying the following
+% properties:
+% \begin{itemize}
+% \item $\forall (x:A).\, R^0 (x,x)$
+% \item $\forall (n:\mathbb{N})(x,y,z : A).\, R(x,y) \conj R^n(y,z) \Rightarrow
+% R^{n+1}(x,z)$
+% \end{itemize}
+% \end{definition}
 
-\begin{smathpar}
-\begin{array}{lcl}
-\stable(R,\I)& \defeq & \forall \E, \E'.\, 
-  \I(\E) \conj R(\E,\E')\Rightarrow \I(\E')\\
-\end{array}
-\end{smathpar}
+% \begin{definition}[\bfseries Interleaved step and multi-step relations]
+% An interleaved step relation interleaves thread-local reductions with
+% interference from concurrent threads captured as the rely relation
+% ($R$).  It is defined thus:\vspace*{-10pt}
 
-\noindent The rule \rulelabel{RG-Var} defines the rely-guarantee judgment for
-shared variable reads inside a transaction $T_i$. It requires $\I$ to
-be stable, and pre- and post- conditions to be stable relative to
-$\I$.  The quantified premise effectively requires a proof that if the
-abstract machine of Fig.~\ref{fig:txnimp} takes a step starting from
-an execution $\E$ that satisfies the pre-condition $P$, then the
-resultant execution $\E'$ satisfies the post-condition $Q$, and that
-the guarantee $G$ faithfully captures the transition from $\E$ to
-$\E'$. It is informative to compare this premise with the premise of
-the \rulelabel{E-Aux} reduction rule of Fig.~\ref{fig:txnimp}. Similar
-premises also appear in the \rulelabel{RG-Asgn} and \rulelabel{RG-Txn}
-rules, which define rely-guarantee judgments for assignments and
-transactions, respectively. \rulelabel{RG-Var} however also requires
-the return value ($\interp{S}(X)$) of the read to satisfy the
-assertion $C$ meant for the value. \rulelabel{RG-Arith} defines the RG
-judgment for an arithmetic expression $e_1\pm e_2$ in terms of the
-corresponding judgments for the constituent expressions $e_1$ and
-$e_2$. The quantified premise requires any value resulting from
-evaluating $e_1 \pm e_2$ to satisfy the assertion $C$, provided that
-$e_1$ and $e_2$ always evaluate to values that satisfy $C_1$ and
-$C_2$, respectively. The rules for sequential and parallel composition
-of commands are essentially the same as their counterparts in a
-standard rely-guarantee formulation and hence elided.
+% \begin{smathpar}
+% \begin{array}{lcl}
+% \I \vdash (c,\E) \rstepsto (c',\E') & \defeq & \I \vdash 
+%   (c,\E) \stepsto (c',\E') \\
+%   &   & \disj (c' = c \conj R(\E, \E') \conj \I(\E'))\\
+% \end{array}
+% \end{smathpar}
 
-% RG judgment for the assignment $X:=e$ makes use of the corresponding
-% judgment for the RHS expression $e$. The quantified premise asserts
-% that evaluating the assigment after evaluating $e$ to a value $v$ and
-% an execution $\E$, should result in an execution $\E'$ that satisfies
-% the post-condition $Q$, while being related to $\E$ via the guarantee
-% $G$. RG judgment for the \C{txn} lexical block is similar.  It uses
-% the judgment for the transaction-bound command $c$ (i.e.,
-% $\txnbox{c}_i$) to obtain an invariant $Q'$ for the execution $\E$
-% before the commit, and verifies that committing the transaction under
-% $\E$ results in an execution $\E'$ that satisfies transaction's
-% post-condition $Q$. As usual, $\E$ and $\E'$ need to be related by $G$.
-% The rules for sequential composition of transaction-bound comands
-% (\rulelabel{RG-Seq}) and parallel composition of top-level commands
-% (\rulelabel{RG-Par}) are straightforward, and more-or-less same as the
-% corresponding rules in classical rely-guarantee.
+% \noindent The interleaved step relation for transaction bound expressions
+% ($\txnbox{e}_i$) and commands ($\txnbox{c}_i$) is defined similarly.
+% An interleaved multi-step relation ($\stepssto{n}$) is the step-indexed
+% reflexive transitive closure of the interleaved step relation.
+% \end{definition}
 
-The \rulelabel{RG-Conseq} rule define ways to strengthen or
-weaken relations and assertions associated with the RG judgment of
-transaction-bound expressions. Similar rules exist for
-transaction-bound and top-level commands, but are not
-shown.\footnote{The supplementary materials provide the complete set
-  of rules.} As is the case with a standard rely-guarantee
-formulation,the rules allow the pre-condition $P$ and the rely relation
-$R$ to be strengthened, and the post-condition $Q$ (also, $C$ in the
-case of expressions) and the guarantee relation $G$ to be weakened.
-The most notable aspect of the \rulelabel{RG-Conseq} rules is that
-they allow the trace invariant $\I$ to be strengthened. Considering
-that $\I$ captures isolation properties, this means that a program
-proven correct under weaker isolation levels is also correct under
-stronger ones.  Parametricity over the trace invariant $\I$, combined
-with the ability to strengthen $\I$ as needed, allows our proof system
-to support a highly flexible proof strategy to prove programs correct
-over various isolation variants. For example, programmers can
-\emph{define} isolation guarantees of their choice (by defining $\I$
-appropriately) and then prove programs correct assuming the guarantees
-hold.  The soundness of strengthening $\I$ ensures that a program can
-be safely executed on any system that offers isolation guarantees
-at least as strong as those assumed.
+% \begin{definition}[\bfseries Semantics of the RG judgment]
+% \label{def:rg-semantics}
+% The semantics of the RG sextuple $\rg{\mathbb{I},P,R}{c}{G,Q}$ is defined
+% in terms of the interleaved step relation thus:\vspace*{-10pt}
 
-\subsection{Semantics and Soundness}
+% \begin{smathpar}
+% \begin{array}{l}
+% \hspace*{-0.3in}
+% \rg{\mathbb{I},P,R}{c}{G,Q} \;\defeq\; \forall \E.\, P(\E)
+%   \wedge \mathbb{I}(\E) \\
+% \hspace*{0.4in}\Rightarrow (\forall n,\E'.\; \I \vdash (c,\E) 
+%     \rstepssto{n} (\cskip,\E') \Rightarrow Q(\E')) \\
+% \hspace*{0.5in}\conj \texttt{step-guaranteed}(\I,R,G,c,\E)\\
+% \end{array}
+% \end{smathpar}
 
-\begin{definition}[\bfseries Step-indexed reflexive transitive closure]
-For all $A:\text{Type}$, $R: A \rightarrow A \rightarrow \mathbb{P}$, and $n :
-\mathbb{N}$, the step-indexed reflexive transitive closure $R^n$ of $R$ is
-the smallest relation satisfying the following
-properties:
-\begin{itemize}
-\item $\forall (x:A).\, R^0 (x,x)$
-\item $\forall (n:\mathbb{N})(x,y,z : A).\, R(x,y) \conj R^n(y,z) \Rightarrow
-R^{n+1}(x,z)$
-\end{itemize}
-\end{definition}
+% \noindent The first conjunct in the consequent is called the \emph{Hoare
+% consequent} since it ascribes Hoare triple semantics to an RG sextuple.
+% The second conjunct, called the \emph{guarantee consequent}, uses the
+% $\texttt{step-guaranteed}$ predicate defined below:\vspace*{-10pt}
 
-\begin{definition}[\bfseries Interleaved step and multi-step relations]
-An interleaved step relation interleaves thread-local reductions with
-interference from concurrent threads captured as the rely relation
-($R$).  It is defined thus:\vspace*{-10pt}
+% \begin{smathpar}
+% \begin{array}{l}
+% \texttt{step-guaranteed}(\I,R,G,c,\E) \;\defeq\; \forall n,\E',c'',\E''.\\
+% \hspace*{0.2in}\I \vdash (c,\E) \rstepssto{n} (c',\E') \conj \I \vdash (c',\E') \stepsto
+%   (c'',\E'') \Rightarrow G(\E',\E'')\\
+% \end{array}
+% \end{smathpar}
 
-\begin{smathpar}
-\begin{array}{lcl}
-\I \vdash (c,\E) \rstepsto (c',\E') & \defeq & \I \vdash 
-  (c,\E) \stepsto (c',\E') \\
-  &   & \disj (c' = c \conj R(\E, \E') \conj \I(\E'))\\
-\end{array}
-\end{smathpar}
+% \noindent The guarantee consequent requires $G$ to capture the trace effect of
+% every small-step of $c$, where the reduction can be interleaved by the
+% interference ($R$) from concurrent threads. The semantics of the RG
+% sextuple for transaction-bound commands ($\txnbox{c}_i$) is defined
+% similarly. Expressions, unlike commands, evaluate to a value $v$, and
+% the semantics of their RG septuple ($\rg{\I,P,R}{\txnbox{e}_i}{G,C,Q}$) differs slightly in that its
+% Hoare consequent requires the value $v$ to satisfy the assertion $C$. 
+% \end{definition}
 
-\noindent The interleaved step relation for transaction bound expressions
-($\txnbox{e}_i$) and commands ($\txnbox{c}_i$) is defined similarly.
-An interleaved multi-step relation ($\stepssto{n}$) is the step-indexed
-reflexive transitive closure of the interleaved step relation.
-\end{definition}
+% Note that the semantics of all RG judgments, including the judgments
+% for transaction-bound terms, make similar demands of the guarantee
+% relation. Given that transactions are atomic (though not isolated), it
+% is not immediately apparent why a transaction's guarantee is required
+% to make explicit every step of its reduction. This requirement is
+% justified however because, in reality, a transaction's atomicity is
+% predicated on the isolation settings of the observer. A \iso{Read
+%   Uncommitted} transaction, for example, is permitted to observe the
+% internal state of a transaction $T$ even if $T$ is claimed to execute
+% atomically.  In the interest of modular verification, the transaction
+% must therefore make its internal state available via its guarantee
+% relation.
 
-\begin{definition}[\bfseries Semantics of the RG judgment]
-\label{def:rg-semantics}
-The semantics of the RG sextuple $\rg{\mathbb{I},P,R}{c}{G,Q}$ is defined
-in terms of the interleaved step relation thus:\vspace*{-10pt}
+% \begin{theorem}[\bfseries Soundness] 
+% The rely-guarantee judgments defined by the rules in
+% Fig.~\ref{fig:rg-rules} are sound with respect to the semantics of
+% Definition~\ref{def:rg-semantics}.\footnote{Formal proof of soundness
+% is provided in the supplementary material.}
+% \end{theorem}
 
-\begin{smathpar}
-\begin{array}{l}
-\hspace*{-0.3in}
-\rg{\mathbb{I},P,R}{c}{G,Q} \;\defeq\; \forall \E.\, P(\E)
-  \wedge \mathbb{I}(\E) \\
-\hspace*{0.4in}\Rightarrow (\forall n,\E'.\; \I \vdash (c,\E) 
-    \rstepssto{n} (\cskip,\E') \Rightarrow Q(\E')) \\
-\hspace*{0.5in}\conj \texttt{step-guaranteed}(\I,R,G,c,\E)\\
-\end{array}
-\end{smathpar}
-
-\noindent The first conjunct in the consequent is called the \emph{Hoare
-consequent} since it ascribes Hoare triple semantics to an RG sextuple.
-The second conjunct, called the \emph{guarantee consequent}, uses the
-$\texttt{step-guaranteed}$ predicate defined below:\vspace*{-10pt}
-
-\begin{smathpar}
-\begin{array}{l}
-\texttt{step-guaranteed}(\I,R,G,c,\E) \;\defeq\; \forall n,\E',c'',\E''.\\
-\hspace*{0.2in}\I \vdash (c,\E) \rstepssto{n} (c',\E') \conj \I \vdash (c',\E') \stepsto
-  (c'',\E'') \Rightarrow G(\E',\E'')\\
-\end{array}
-\end{smathpar}
-
-\noindent The guarantee consequent requires $G$ to capture the trace effect of
-every small-step of $c$, where the reduction can be interleaved by the
-interference ($R$) from concurrent threads. The semantics of the RG
-sextuple for transaction-bound commands ($\txnbox{c}_i$) is defined
-similarly. Expressions, unlike commands, evaluate to a value $v$, and
-the semantics of their RG septuple ($\rg{\I,P,R}{\txnbox{e}_i}{G,C,Q}$) differs slightly in that its
-Hoare consequent requires the value $v$ to satisfy the assertion $C$. 
-\end{definition}
-
-Note that the semantics of all RG judgments, including the judgments
-for transaction-bound terms, make similar demands of the guarantee
-relation. Given that transactions are atomic (though not isolated), it
-is not immediately apparent why a transaction's guarantee is required
-to make explicit every step of its reduction. This requirement is
-justified however because, in reality, a transaction's atomicity is
-predicated on the isolation settings of the observer. A \iso{Read
-  Uncommitted} transaction, for example, is permitted to observe the
-internal state of a transaction $T$ even if $T$ is claimed to execute
-atomically.  In the interest of modular verification, the transaction
-must therefore make its internal state available via its guarantee
-relation.
-
-\begin{theorem}[\bfseries Soundness] 
-The rely-guarantee judgments defined by the rules in
-Fig.~\ref{fig:rg-rules} are sound with respect to the semantics of
-Definition~\ref{def:rg-semantics}.\footnote{Formal proof of soundness
-is provided in the supplementary material.}
-\end{theorem}
-
-\noindent In particular, if $\rg{\I,P,R}{c}{G,Q}$ can be derived using
-the rules of Fig.~\ref{fig:rg-rules}, then (a) every interleaved
-multi-step reduction of $c$ starting from a trace that satisfies $P$
-and $\I$, results in a trace that satisfies $Q$, and (b) the effect
-that every small-step of $c$ has on the trace is contained in $G$.
-Soundness of the RG judgment for transaction-bound commands
-($\txnbox{c}_i$) is stated similarly.  For expressions, soundness of
-the judgment $\rg{\I,P,R}{\txnbox{e}_i}{G,C,Q}$ also proves that $e$
-is always evaluated to a value that satisfies $C$.
+% \noindent In particular, if $\rg{\I,P,R}{c}{G,Q}$ can be derived using
+% the rules of Fig.~\ref{fig:rg-rules}, then (a) every interleaved
+% multi-step reduction of $c$ starting from a trace that satisfies $P$
+% and $\I$, results in a trace that satisfies $Q$, and (b) the effect
+% that every small-step of $c$ has on the trace is contained in $G$.
+% Soundness of the RG judgment for transaction-bound commands
+% ($\txnbox{c}_i$) is stated similarly.  For expressions, soundness of
+% the judgment $\rg{\I,P,R}{\txnbox{e}_i}{G,C,Q}$ also proves that $e$
+% is always evaluated to a value that satisfies $C$.

### 2017-07-02 related-work.tex
edits

@@
-\vspace*{-8pt}
+\vspace*{-6pt}
 \section{Related Work}
@@
 \paragraph{Specifying weak isolation.}
+
 Adya~\cite{adyaphd} specifies several weak isolation levels in terms
@@
 specifications to verify programs under weak isolation.
+
+Recent work by Crooks \emph{et al.}~\cite{CPA+17} also explores the
+use of a state-based interpretation of isolation as we do, and share
+our position that enables specification of weak isolation that are not
+tied to implementation-specific artifacts.  However, they do not
+consider verification (manual or automated) of client programs, and it
+is not immediately apparent if their specification formalism is
+amenable for use within a verification toolchain.
 

### 2017-07-02 opsem.tex
Sec.3 text part 1.

@@
 
-% Fig.~\ref{fig:txnimp} shows the syntax and small-step semantics of
-% \txnimp, a core language that we will use to formalize the intuitions
-% presented in the previous section. Natural numbers, (shared) variables
-% and arithmetic expressions constitute the syntactic class of
-% expressions ($e$).  Commands ($c$) include $\cskip$, assignment
-% statements, transaction (\C{txn}) lexical blocks, and their sequential
-% and parallel composition. We let $T_i$ for $i \in \mathbb{N}$ range
-% over transaction identifiers. When it is evident we are referring to a
-% transaction, we use the number $i$ instead of $T_i$ for identification
-% (\eg in $\C{txn}\langle i \rangle$). Like variables, transaction
-% identifiers are globally accessible. For notational convenience, we
-% let $t$ range over both expressions and commands.
+Fig.~\ref{fig:txnimp} shows the syntax and small-step semantics of
+\txnimp, a core language that we use to formalize our intuitions about
+reasoning under weak isolation. Variables ($x$), Integer and boolean
+constants ($k$), records ($r$) of named constants, sets ($s$) of such
+records, arithmetic and boolean expressions ($e_1 \odot e_2$), and
+record expressions ($\{\bar{f}=\bar{e}\}$) constitute the syntactic
+class of expressions ($e$). Commands ($c$) include $\cskip$,
+conditional statements, \C{LET} constructs to bind names, \C{FOREACH}
+loops, SQL statements, their sequential composition ($c_1;c_2$), and
+transactions ($\ctxn{i}{\I}{c}$) and their parallel composition
+($c_1\,||\,c_2$). Certain terms that only appear at run-time are also
+present in $c$. These include a \C{TXN} block tagged with sets ($\stl$
+and $\stg$) of records that assume special meaning in operational
+semantics, and a \C{FOREACH} loop that keeps track of the set ($s_1$)
+of items already iterated, and the set ($s_2$) of items yet to be
+iterated. Note that even the surface-level \C{FOREACH} shown here is a
+little different from the one used in previous sections; its
+higher-order argument has two arguments, $y$ and $z$, which are
+invoked (during the reduction) with the set of already-iterated items,
+and the current item, respectively. This form of \C{FOREACH} lends
+itself to inductive reasoning, facilitating inductive proofs
+(Sec.~\ref{sec:reasoning}).
 
-% We define a small-step operational semantics for this language in
-% terms of an abstract machine that generates an execution trace ($\E$).
-% The first component of the trace is a set ($\A$) of \emph{effects},
-% where an effect ($\eta$) is a tuple of (a). the identifier ($T_i$) of
-% the transaction that generated this effect, (b). the unique
-% identifier ($j$) of this effect, (c). the operation ($\op$)
-% documented by this effect, which can be either a read (\C{RD(X)}), a
-% write (\C{WR(X)}), or a transaction commit (\C{COMMIT}), and (d). the
-% value, if any, associated with the operation (\eg, value read or value
-% written). Functions $\txn$, $\id$, $\oper$, and $\rval$ project each
-% of the four aforementioned individual components of the tuple.  In
-% every step of the evaluation, the machine reduces a \txnimp\ term by
-% executing a read, write or commit operation, generating an effect, and
-% extending the trace. Since effects include transaction identifiers,
-% the semantics distinguishes between terms ($t$) of different
-% transactions. For example, $\txnbox{t}_i$ denotes a term $t$ inside a
-% transaction $T_i$.  Evaluation contexts are also appropriately marked.
-% For example, $\ectx_i$ denotes the evaluation context for a term
-% inside $T_i$. The other component of an execution trace is a
-% visibility relation ($\visZ$) that establishes a visibility property
-% between effects among different transactions. The intent and mechanics
-% of $\visZ$ is described in the sequel.
+% also We let $T_i$ for $i \in \mathbb{N}$
+% range over transaction identifiers. When it is evident we are
+% referring to a transaction, we use the number $i$ instead of $T_i$ for
+% identification (\eg in $\C{txn}\langle i \rangle$). Like variables,
+% transaction identifiers are globally accessible. For notational
+% convenience, we let $t$ range over both expressions and commands.
+
+We define a small-step operational semantics for this language in
+terms of an abstract machine that executes a command, and updates
+either a transaction-local database ($\stl$), or the global database
+($\stg$). Database ($\stg$) is a modeled as a set of records of a
+pre-defined type, i.e., they all belong to a single table.
+Generalization to multiple tables is straightforward. Records in
+$\stg$ are uniquely identifiable through their $\idf$ field, which is
+auto-generated and does not belong to the surface language, i.e.,
+$\C{id}\notin f$. For a set $S$ of records, we define $\dom(S)$ as the
+set of unique ids of all the records in $S$. Thus $|\dom(\stg)| =
+|\stg|$. During its execution, a transaction may write to multiple
+records in $\stg$. Atomicity dictates that such writes shouldn't be
+available in $\stg$ until the transaction commits. We therefore
+associate each transaction with a local database ($\stl$) that stores
+\emph{only} the uncommitted records\footnote{While SQL's \C{UPDATE}
+admits writes at the granularity of record fields, databases, in
+reality, enforce record-level locking, allowing us to think of
+``uncommitted writes'' as ``uncommitted records''. }. Uncommitted
+records include deleted records, which are identified by a hidden
+$\delf$ field set to \C{true}. When the transaction commits, its local
+database is atomically \emph{flushed} to the global database,
+committing the uncommitted records. The flush operation ($\gg$) is defined as
+following:
+\begin{smathpar}
+\begin{array}{c}
+\forall r.~ r \in (\stl\gg\stg) ~\Leftrightarrow~ 
+  (r.\idf \notin \dom(\stl) \conj r \in \stg)
+\disj (r \in \stl \conj \neg r.\delf) 
+\end{array}
+\end{smathpar}
+Let $\stg' = \stl\gg\stg$. A record $r$ belongs to $\stg'$ iff it
+belongs to $\stg$ and not been updated in $\stl$, i.e., $r.\idf \notin
+\dom(\stl)$, or it belongs to $\stl$, i.e., it is either a new record,
+or an updated version of an old record, but the updatation is not a
+deletion ($\neg r.\delf$). Thus, flush defines the result of
+atomically applying transaction's local writes to the global database.
+Besides the commit, flush also helps a transaction read its own
+writes. Intuitively, the result of a read operation inside a
+transaction must be computed on the database resulting from flushing
+the current local state ($\stl$) to the global state ($\stg$). The
+abstract machine of Fig.~\ref{fig:txnimp}, however, doesn't let a
+transaction read its own writes. This reduces verbosity and simplifies
+semantics, without losing any generality; substituting $\stl\gg\stg$
+for $\stg$ at select places in reduction rules recovers the real
+
+Small-step semantics is stratified into a transaction-local reduction
+relation, and a top-level reduction relation. Transaction-local
+relation ($\stg \vdash (c,\stl) \stepsto (c',\stl')$) defines a
+small-step reduction for a command inside a transaction, when the
+database state is $\stg$; the command $c$ reduces to $c'$, while
+updating the transaction-local database $\stl$ to $\stl'$. The
+definition assumes a meta function $\eval$ that evaluates expressions
+with no free variables to values. The reduction relation for SQL
+statements is defined straightforwardly.  \C{INSERT} adds a new record
+to $\stl$ after adding a unique identifier (more discussion on
+uniqueness later). \C{DELETE} finds the records in $\stg$ that match
+the search criteria defined by its boolean function argument, and adds
+the records to $\stl$ after marking them for deletion. \C{SELECT}
+bounds the name introduced by \C{LET} to the set of records from
+$\stg$ that match the search criteria, and then executes the bound
+command $c$. \C{UPDATE} uses its first function argument to compute
+the updated version of the records that match the search criteria
+defined by its second function argument. Updated records are added to
+$\stl$.
+
+The reduction of \C{FOREACH} starts by first converting it to its
+run-time form that also keeps track of the iterated items ($s_1$),
+besides the yet-to-be-iterated items ($s_2$). Initially, $s_1$ is
+empty. As the elements are iterated, they are removed from $s_2$ and
+added to $s_1$ . Iteration involves invoking the higher-order function
+with $s_1$ and the current element $x$ (note: $\uplus$ in $\{x\}
+\uplus s_2$ denotes a disjoint union). The reduction ends when the
+$s_2$ becomes empty. The reduction rules for conditionals, \C{LET}
+binders, and sequences are straightforward, hence ommitted.
+
+The top-level reduction relation
 

### 2017-07-02 abstract.tex

@@
 
-  To address this issue, we present an automated verification
-  methodology based on a novel program logic that enables
+  To address this issue, we present a novel program logic that enables
   compositional reasoning about the behavior of concurrently executing
-  weakly-isolated transactions. Notably, our development is parametric
-  over a transaction's specific isolation semantics, allowing it to be
-  applicable over a range of concurrency control mechanisms.  Given a
-  a collection of concurrently executing transactions, our analysis
-  verifies whether assertion ${\mathcal{A}}$ associated with
-  transaction $T$ in this collection, holds under a weak isolation
-  level ${\mathcal{I}}$.  Case studies and experiments on real-world
-  applications demonstrate the utility of our approach, and provide
-  strong evidence that verification of weakly-isolated transactions
-  can be placed on the same formal footing as their strongly-isolated
-  serializable counterparts.
+  weakly-isolated transactions.  Recognizing that the proof burden
+  necessary to use this logic may dissuade application developers, we
+  also describe a fully automated verification tool based on this
+  foundation that determines the safety of high-level consistency
+  assertions associated with such transactions.  The key to effective
+  inference is the observation that weakly-isolated transactions can
+  be viewed as functional (monadic) computations over database state,
+  allowing us to treat their operations as state transformers over the
+  database.  This interpretation enables automated verification using
+  off-the-shelf SMT solvers.
+
+  Notably, our development is parametric over a transaction's specific
+  isolation semantics, allowing it to be applicable over a range of
+  concurrency control mechanisms.  Case studies and experiments on
+  real-world applications demonstrate the utility of our approach, and
+  provide strong evidence that automated verification of
+  weakly-isolated transactions can be placed on the same formal
+  footing as their strongly-isolated serializable counterparts.
 

### 2017-07-02 introduction.tex

@@
 The guarantees provided by serializability do not come for free,
-however - pessimistic concurrency control methods that enforce
-serializability, for example, require databases to provide expensive
-techniques such as two-phase locking that incur overhead to deal with
-deadlocks, rollbacks, and re-execution~\cite{twopl,ullmanbook}.
-Similar criticisms apply to optimistic multi-version concurrency
-control methods that must deal with timestamp and version
-management~\cite{BG81}.  Regardless of the particular concurrency
-control method employed, enforcing serializability in a replicated
-data store additionally requires global coordination among
-geo-distributed replicas, which makes the store \emph{unavailable} in
-the presence of network
-partitions~\cite{cap,sernotavlbl,bailishat,bernsigmod13}.
+however - pessimistic concurrency control methods require databases to
+use expensive mechanisms such as two-phase locking that incur overhead
+to deal with deadlocks, rollbacks, and
+re-execution~\cite{twopl,ullmanbook}.  Similar criticisms apply to
+optimistic multi-version concurrency control methods that must deal
+with timestamp and version management~\cite{BG81}.  These issues are
+further exacerbated when the database is replicated, requiring
+additional coordination
+mechanisms~\cite{cap,sernotavlbl,bailishat,bernsigmod13}.
 
 The tension between serializable transactions, which are easy to
-reason about but difficult to implement, and more pragmatic variants
-that are driven by performance and availability considerations, has
-motivated the development of weaker forms of transaction isolation,
-beginning as early as 1976~\cite{gray1976}. The ANSI SQL 92 standard
-defines three such weak isolation levels which are now implemented in
-many relational and NoSQL databases. Not surprisingly, weakly-isolated
-transactions have been found to significantly outperform serializable
-transactions on benchmark suites, both on single-node databases and
-multi-node replicated stores~\cite{dbtuningbook,bailishat,bailisvldb},
-leading to their overwhelming adoption. A 2013
-study~\cite{bailishotos} of 18 popular ACID and ``NewSQL'' databases
-found that only three of them offer serializability by default, and
-half, including Oracle 11g, do not offer it at all.  A 2015
-study~\cite{bailisferal} of a large corpus of database applications
-finds no evidence that applications manifestly change the default
-isolation level offered by the database. Taken together, these studies
-make clear that weakly-isolated transactions are quite prevalent in
-practice, whereas serializable transactions are often eschewed.
+reason about but difficult to implement efficiently, and more
+pragmatic variants that are driven by performance and availability
+considerations, has motivated the development of weaker forms of
+transaction isolation, beginning as early as 1976~\cite{gray1976}. The
+ANSI SQL 92 standard defines three such weak isolation levels which
+are now implemented in many relational and NoSQL databases. Not
+surprisingly, weakly-isolated transactions have been found to
+significantly outperform serializable transactions on benchmark
+suites, both on single-node databases and multi-node replicated
+stores~\cite{dbtuningbook,bailishat,bailisvldb}, leading to their
+overwhelming adoption. A 2013 study~\cite{bailishotos} of 18 popular
+ACID and ``NewSQL'' databases found that only three of them offer
+serializability by default, and half, including Oracle 11g, do not
+offer it at all.  A 2015 study~\cite{bailisferal} of a large corpus of
+database applications finds no evidence that applications manifestly
+change the default isolation level offered by the database. Taken
+together, these studies make clear that weakly-isolated transactions
+are quite prevalent in practice, whereas serializable transactions are
+often eschewed.
 
-Unfortunately, weak isolation admits behaviours that are difficult to
+Unfortunately, weak isolation admits behaviors that are difficult to
 comprehend~\cite{berenson}. To quantify weak isolation anomalies,
-Fekete \emph{et al.}~\cite{feketevldb09} devised and experimented
-with a microbenchmark suite that executes transactions under a
+Fekete \emph{et al.}~\cite{feketevldb09} devised and experimented with
+a microbenchmark suite that executes transactions under a
 weakly-isolated \emph{read committed} isolation level - the default
@@
 use weak isolation, and in what form, is therefore a prominent
-question facing all database programmers.
+question facing all database programmers.\footnote{This position has
+  been echoed by leading database researchers who lament the lack of
+  better understanding of this problem.  See: {\tt
+    http://www.bailis.org/blog/understanding-weak-isolation-is-a-serious-problem}.}
 
@@
 Consequently, reasoning about weak isolation remains an error prone
-endeavour, with major database vendors~\cite{postgresiso, mysqliso,
+endeavor, with major database vendors~\cite{postgresiso, mysqliso,
 oracleiso} continuing to document their isolation levels primarily in
@@
 
-%%%SJ: tried to tighten this paragraph; the original now commented seemed overly
-%%complicated
 Recent results on reasoning about application invariants in the
@@
 different replicas, resulting in an execution order inconsistent with
-%% We already say "inconsistent with.. ". 
 any sequential order. In contrast, weak isolation is a property of
@@
 evident from the support for weakly-isolated transactions on
-conventional (unreplicated) databases as mentioned above. In the
-presence of replication, however, the interaction between weak
-isolation and weak consistency can be subtle and non-trivial.
-Understanding weak isolation in these varied contexts thus requires
-new insights and substantial generalization of existing techniques.
+conventional (unreplicated) databases as mentioned above.
+
+%% In the
+%% presence of replication, however, the interaction between weak
+%% isolation and weak consistency can be subtle and non-trivial.
+%% Understanding weak isolation in these varied contexts thus requires
+%% new insights and substantial generalization of existing techniques.
 
@@
 In this paper, we propose a program logic for weakly-isolated
-transactions that realizes this goal.  In particular, we develop a set
-of syntax-directed compositional proof rules that allow developers to
-build correctness proofs for transactional programs in the presence of
-a weakly-isolated concurrency control mechanism.  A key novelty of our
-approach is that it is parametric over the \emph{isolation semantics}
-of transactions in the program, as well as the \emph{consistency
-semantics} of the underlying store. In concrete terms, this means
-that, unlike recent work focused on reasoning about programs under
-weak consistency~\cite{gotsmanpopl16, redblueatc, ecinec}, our system
-model does not assume a minimum or predefined set of consistency or
-isolation levels. Instead, our operational semantics admits
-declarative specifications of transaction isolation and store
-consistency, and generates executions that are guaranteed to satisfy
-these specifications. The result is a flexible system model that is
-general enough to incorporate the semantics of a range of isolation
-levels on a variety of stores (\eg~the sequentially consistent store
-of~\cite{adyaphd}, or the causally consistent store
-of~\cite{gotsmanpopl16}, etc).  Our key technical contribution is thus
-a set of proof rules that demonstrate that a program with a given
-selection of isolation levels for its transactions preserves its
-invariants when executed on a store equipped with appropriate consistency
-guarantees.  The paper makes the following contributions:
+transactions along with automated verification support that realizes
+this goal.  We develop a set of syntax-directed compositional proof
+rules that allow developers to build correctness proofs for
+transactional programs in the presence of a weakly-isolated
+concurrency control mechanism.  Moreover, we also present an inference
+algorithm based on a state transformer semantics that allows us to
+automatically verify the weakest isolation level for a transaction
+that ensures its consistency requirements are maintained.  The key to
+inference is our formulation of database state (represented as sets of
+tuples) as a monad, in which database computations are interpreted as
+transformers over these sets.  This interpretation leads to an
+encoding database operations amenable for verification by SMT solvers.
+
+We have realized these ideas as an embedded DSL within OCaml, with
+support for common SQL operations (updates, selects, inserts, deletes,
+joins, etc.).  Experimental results on real-world benchmarks
+demonstrate the feasibility of automated verification of weakly
+isolated transactions, an important advance in improving the safety of
+realistic database computations. The paper makes the following
+contributions:
 \vspace*{-6pt}
 \begin{enumerate}
-  \item We develop a semantics for a core language equipped with
-    weakly-isolated transactions, demonstrating that a general
-    parametric isolation semantics can be expressed as
-    well-formedness constraints on a program's execution.
-  \item We present a compositional proof system for this language
-    capable of relating high-level application invariants to the structure
-    of traces induced by the operational semantics, in which transactions
-    are associated with specific weak isolation levels.
-  \item We define a \emph{maximum visibility principle} to reconcile
-    store consistency guarantees with transaction isolation
-    requirements, and use it to instantiate our operational model and
-    the proof system for various kinds of (weakly-consistent) data
-    stores.
-  \item Case studies modeled after real-world scenarios demonstrate
-    the applicability and utility of our proof methodology.
+\item An operational semantics for an embedded DSL in OCaml that
+  expresses SQL-based relational database operations as monadic
+  computations on database state represented abstractly as sets of
+  tuples.
+ \item A compositional proof system for this language capable of
+   relating high-level application invariants to transformations on
+   database state, parameterized by a weak isolation semantics that
+   selectively exposes the visibility of these operations to other
+   transactions.
+ \item An inference algorithm capable of discovering the
+   weakest isolation level that is sound with respect to a
+   transaction's high-level consistency requirements.  The algorithm
+   exploits our monadic formulation, treating database computations as
+   predicate state transformers, that leads to a natural encoding
+   amenable for automated verification.
+ \item A realistic implementation along with results on real-world
+   database benchmarks that justify our approach and the utility of
+   our proof methodology.
 \end{enumerate}
 \noindent Our results provide the first (to the best of our knowledge)
-mechanism that precisely and uniformly relates high-level program
-invariants to low-level weak isolation and weak consistency
-guarantees, thereby allowing weakly-isolated transactions to enjoy the
-same rigorous reasoning capabilities as their strongly-isolated
-(serializable) counterparts.
+automated mechanism to verify the safety of high-level consistency
+conditions for weakly-isolated transactions, thereby allowing such
+transactions to enjoy the same rigorous reasoning capabilities as
+their strongly-isolated (serializable) counterparts.
 
@@
 invariants, and establishes the soundness of these rules with respect
-to the semantics. \S\ref{sec:store-consistency} generalizes the
-framework to integrate support for weakly-consistent data stores. We
-describe the impact of our reasoning framework in the context of
-several real-world case studies in \S\ref{sec:case-studies}.  Related
-work and conclusions are given in \S\ref{sec:relatedwork}.
+to the semantics. \S\ref{sec:inference} describes the inference
+algorithm, and the state transformer encoding.  We describe our
+implementation, benchmark results, and case studies in
+\S\ref{sec:case-studies}.  Related work and conclusions are given in
+\S\ref{sec:relatedwork}.

### 2017-07-02 motivation.tex

@@
 
-We have developed and implemented our ideas in an embedded DSL in
-OCaml that supports a \C{DB} monad to define and compose database
-computations given in terms of SQL operations.  Besides the usual
-\C{bind} and \C{return} combinators, the monad offers an
+We present our ideas in the context of an embedded DSL in OCaml that
+we have developed which supports a \C{DB} monad to define and compose
+database computations given in terms of SQL operations.  Besides the
+usual \C{bind} and \C{return} operators, the monad offers an
 \C{atomically} combinator that executes a database computation as an
@@
 business logic is captured in 5 database transactions that operate on
-9 tables; \C{new\_order} is one such transaction that operates on
+9 tables; \C{new\_order} is one such transaction that uses
 \C{District}, \C{Order}, \C{New\_order}, \C{Stock}, and
 \C{Order\_line} tables. The transaction acts on the behalf of a
-customer, whose id is \C{c\_id}, to place a new order for the given
+customer, whose id is \C{c\_id}, to place a new order for a given
 set of items (\C{item\_reqs}), to be served by a warehouse under the
-district identified by \C{d\_id}.  Fig.~\ef{fig:scheme} illustrates
+district identified by \C{d\_id}.  Fig.~\ref{fig:scheme} illustrates
 the relationship among these different tables.
@@
 executing transactions while the \C{new\_order} transaction executes.
-Under weak isolation\footnote{Weak isolation doesn't violate atomicity
-  as long as the witnessed effects are those of committed
+Under weak isolation\footnote{Weak isolation does not violate
+  atomicity as long as the witnessed effects are those of committed
   transactions}, however, interferences of various kinds are
@@
 transactions would appear to be superficially similar to the
-verification of (racy) concurrent programs~\cite{concurrentGC...},
-weak isolation introduces new challenges arising from the use of
-transactions, and new opportunities arising from the fact that the
-store abstraction used by transactions is a relational database, not
-low-level memory.
+verification of (racy) concurrent programs (e.g., garbage
+collectors~\cite{JLP+14,GHE15,HPQ+15}), weak isolation introduces new
+challenges arising from the use of transactions, and new opportunities
+arising from the fact that the store abstraction used by transactions
+is a relational database, not low-level memory.
 
@@
 them as essentially functional computations that manage database state
-monadically~\cite{statemonad}. We find it useful to reason about
+monadically. We find it useful to reason about
 statements that mutate the database state, not in terms of a pre- and

### 2017-07-02 opsem.tex
Edits.

@@
 transactions ($\ctxn{i}{\I}{c}$) and their parallel composition
-($c_1\,||\,c_2$). Certain terms that only appear at run-time are also
-present in $c$. These include a \C{TXN} block tagged with sets ($\stl$
-and $\stg$) of records that assume special meaning in operational
-semantics, and a \C{FOREACH} loop that keeps track of the set ($s_1$)
-of items already iterated, and the set ($s_2$) of items yet to be
-iterated. Note that even the surface-level \C{FOREACH} shown here is a
-little different from the one used in previous sections; its
-higher-order argument has two arguments, $y$ and $z$, which are
-invoked (during the reduction) with the set of already-iterated items,
-and the current item, respectively. This form of \C{FOREACH} lends
-itself to inductive reasoning, facilitating inductive proofs
-(Sec.~\ref{sec:reasoning}).
+($c_1\,||\,c_2$). The $\I$ in \C{TXN} block syntax is the
+transaction's isolation specification, which will be explained later.
+Certain terms that only appear at run-time are also present in $c$.
+These include a \C{TXN} block tagged with sets ($\stl$ and $\stg$) of
+records that assume special meaning in operational semantics, and a
+\C{FOREACH} loop that keeps track of the set ($s_1$) of items already
+iterated, and the set ($s_2$) of items yet to be iterated. Note that
+even the surface-level \C{FOREACH} shown here is a little different
+from the one used in previous sections; its higher-order argument has
+two arguments, $y$ and $z$, which are invoked (during the reduction)
+with the set of already-iterated items, and the current item,
+respectively. This form of \C{FOREACH} lends itself to inductive
+reasoning, facilitating inductive proofs (Sec.~\ref{sec:reasoning}).
 
@@
 
-The top-level reduction relation
+The top-level reduction relation defines small-step semantics of
+transactions, and their parallel composition. A transaction always
+comes tagged with an isolation specification $\I$ that dictates the
+timing and nature of iterferences that the transaction can witness.
+Formally, $\I$ is a function that maps a boolean ($\mathbb{B}$) and a
+triplet of database states (three sets of records:
+$\Pow(r)\times\Pow(r)\times\Pow(r)$) to a logical proposition
+($\Prop$). The abstract machine calls $\I$ while reducing
+$\ctxn{i}{\I}{c}$. The boolean argument identifies if the transaction
+is still executing (\C{true}), or if it has finished execution and
+ready to commit (\C{false}).  Among the triplet of sets (the second
+argument), first is the (current) transaction-local database state
+($\stl$), second is the state ($\stg$) of the global database when the
+transaction last took a step, and third is the current state ($\stg'$)
+of the global database.  Intuitively, $\stg'\neq\stg$ indicates an
+interference from a concurrent transaction, and the $\I$ decides if
+this interference is allowed or not, considering the local database
+state ($\stl$). For instance, as described in
+(Sec.~\ref{sec:motivation}), an RR transaction on Postgres defines
+$\I$ as following:
+\begin{smathpar}
+\begin{array}{lcl}
+\I\,\C{true}\,(\stl,\stg,\stg') & = & \stg' = \stg\\
+\I\,\C{false}\,(\stl,\stg,\stg') & = & \forall(r\in\stl)(r'\in\stg).~
+  r'.\idf = r.\idf \Rightarrow r'\in\stg'.
+\end{array}
+\end{smathpar}
 
+The reduction of a $\ctxn{i}{\I}{c}$ begins by first converting it to
+its run-time form $\ctxn{i}{\I,\stl,\stg}{c}$, where $\stl =
+\emptyset$, and $\stg$ the current (global) database. The top-level
+reduction rules in Fig.~\ref{fig:txnimp} assume this form. Rule
+\rulelabel{E-Txn} reduces $\ctxn{i}{\I,\stl,\stg}{c}$ under a database
+state ($\stg'$), only if the isolation specification ($\I$) allows the
+interference between $\stg$ and $\stg'$ during the execution. Rule
+\rulelabel{E-Commit} commits the transaction
+$\ctxn{i}{\I,\stl,\stg}{c}$ by flushing its uncommitted records to the
+database. This is done only if the interference between $\stg$ and
+$\stg'$ is allowed during the commit by the isolation specification.
+
+% For convenience, we define $\I_r = \I(\C{true})$ and $\I_c =
+% \I(\C{false})$.
 % Fundamental to our development is the notion of a trace invariant

### 2017-07-02 opsem.tex
Edits.

@@
 triplet of database states (three sets of records:
-$\Pow(r)\times\Pow(r)\times\Pow(r)$) to a logical proposition
+$\Pow{r}\times\Pow{r}\times\Pow{r}$) to a logical proposition
 ($\Prop$). The abstract machine calls $\I$ while reducing
@@
 
-% For convenience, we define $\I_r = \I(\C{true})$ and $\I_c =
-% \I(\C{false})$.
-% Fundamental to our development is the notion of a trace invariant
-% ($\I$). $\I$ is a function from traces ($\E$) to first-order logical
-% formulas ($\Prop$) that define well-formedness
-% constraints over traces.  The machine takes a step only if the
-% resulting trace satisfies the constraints imposed by $\I$. This
-% behaviour is captured by the auxiliary reduction rule
-% \rulelabel{E-Aux} that factors out the trace extension aspect of the
-% evaluation by abstracting away the operation-specific behaviour as a
-% function that generates an appropriate effect. We let $\mathcal{F}$
-% denote this function.  \rulelabel{E-Aux} uses $\mathcal{F}$ to
-% generate a new effect and extend the trace ($\E = (\A,\visZ)$)
-% \emph{only if} the well-formedness constraints imposed by $\I$ on $\E$
-% (i.e., $\I(\E')$) are satisfied. Otherwise, it gets stuck. In an
-% execution that runs to completion, every small-step preserves the
-% well-formedness of a trace, thus ensuring the invariance of $\I$.
-% Note that the semantics makes no assumptions about $\I$ other than its
-% type. As such, it can be instantiated with any trace-parametric
-% proposition that expresses constraints over the given trace. For
-% instance, consider the $\psi_{RC}$ specification from
-% \S\ref{sec:motivation}, but with bounded $T_1$ and $T_2$ instantiated
-% with \C{Wd1} and \C{Wd2}, respectively. The instantiated specification
-% is the following term:\vspace*{-8pt}
-
-% \begin{smathpar}
-% \begin{array}{l}
-%   \forall \eta_1,\eta_2.\; \txn(\eta_1) = \C{Wd1}
-%   \conj \txn(\eta_2) = \C{Wd2} \\
-%   \hspace*{0.6in}\conj \C{Wd1} \neq \C{Wd2} \conj \eta_1 \hboar
-%   \eta_2 \Rightarrow \C{Wd1} \hboar \eta_2 \\
-% \end{array}
-% \end{smathpar}
-
-% \noindent It is easy to interpret the above specification in the
-% context of a trace $\E$ that captures an execution of the program in
-% Fig.~\ref{fig:motiv-eg-1}. Such a trace-parametric formula can be
-% used to instantiate the trace invariant $\I$ in Fig.~\ref{fig:txnimp}.
-% The resultant operational semantics describes an abstract machine that
-% gets stuck if an operation of \C{Wd2} is executed in a state that
-% incorporates some, but not all the effects (including the \C{COMMIT})
-% of \C{Wd1}.
-% % While the machine takes a step only if the constraints are
-% % satisfied, it neither defines nor explicitly assumes an oracle to
-% % check satisfaction.
-
-% As described in \S\ref{sec:motivation}, the semantics of various
-% isolation levels can be captured as constraints over the
-% happens-before ($\hbZ$) relation. $\hbZ$ is however a derived relation
-% in our model, composed of more fundamental \emph{session order}
-% ($\soZ$) and \emph{visibility} ($\visZ$) relations. In particular,
-% $\hbZ \,=\, (\soZ \cup \visZ)^{+}$. Unlike $\hbZ$, $\visZ$ (defined
-% below) is not a transitive relation, and hence lets us capture
-% finer-grained isolation properties than $\hbZ$, which we leverage in
-% our development.  The session order relation captures the sequential
-% order of operations within a transaction. In particular, it relates
-% two effects, $\eta_1$ and $\eta_2$, such that $\txn(\eta_1) =
-% \txn(\eta_2)$ and $\id(\eta_1) < \id(\eta_2)$.  The semantics assigns
-% monotonically increasing identifiers to effects, as defined by the
-% $\id(\eta) > \maxId(\A)$ condition of \rulelabel{E-Aux} ($\maxId(\A)$
-% returns the maximum number identifying an effect in $\A$). Evaluation
-% contexts ($\ectx_i$) for transaction-bound terms are defined so as to
-% enforce a deterministic sequential order of execution within a
-% transaction, leading to a deterministic total order among effect ids,
-% which defines the session order relation. Visibility ($\visZ$) on the
-% other hand relates effects across concurrent transactions.
-% Intuitively, $\visZ$ relates $\eta_1$ to $\eta_2$ if and only if
-% $\eta_1$ was \emph{visible} to the operation that generated $\eta_2$
-% during its execution, thus effecting its return value
-% ($\rval(\eta_2)$). For example, a read operation over \C{X} may pick
-% the value ($\rval$) of the write effect with the highest id among its
-% visible effects (this is made possible by appropriately defining
-% $\interp{\cdot}$ in \rulelabel{E-Var}, as we show later). Thus, the
-% value of a read depends on what write effects it can witness. An
-% operation can only witness the effects of already concluded
-% operations, which varies between executions due to the
-% non-deterministic order of evaluating the parallel composition of
-% transactions.
-
-% A more notable source of non-determinism, however, is the
-% \rulelabel{E-Aux} rule, which allows the machine to expose an
-% arbitrary subset ($S$) of existing effects ($\A$) to the incoming
-% operation. In other words, the machine is not obligated to reveal the
-% effects of all previous operations to an incoming operation. This
-% relaxation allows the abstract machine to model the semantics of
-% weakly-consistent data stores. For instance, operations issued to an
-% eventually consistent ({\sc ec}) replicated store could be dispatched
-% to different replicas whose states may not be in any well-defined
-% relationship. By allowing operations to witness arbitrary subsets of
-% the global state, the semantics models the weak visibility properties
-% of such stores; we elaborate on the implication of this style of
-% definition in \S\ref{sec:store-consistency}.  Stronger visibility
-% properties can be expressed by imposing well-formedness constraints
-% over $\visZ$ via the trace invariant ($\I$). Since the abstract
-% machine is obligated to satisfy $\I$ at every step of the execution,
-% operations are guaranteed to experience the level of isolation
-% specified by $\I$.  Thus, in executions that run to completion, the
-% abstract machine models a store that provides the required levels of
-% isolation.  Notably, the machine achieves this without defining an
-% operational semantics for isolation levels, instead solely relying on
-% their declarative characterization as trace well-formedness
-% constraints to enforce isolation guarantees.
-% \S\ref{sec:ansi-isolation} specifies various ANSI SQL isolation levels
-% stated as trace well-formedness constraints.
-
-% % TRACE RELATIONS
-% % ---------------
-% \input{trace-relations}
-
-% % However, a machine that lets operations witness arbitrary subset of
-% % the global state offers no isolation whatsoever. For example, it may
-% % allow a read operation to witness writes of an uncommitted
-% % transaction, violating RC isolation. Fortunately, our ability to
-% % express an isolation semantics as constraints over happens-before
-% % order through $\visZ$ and $\soZ$ relations, and the property of the
-% % abstract machine to be parametric over the trace invariant ($\I$),
-% % lets us solve this problem.
-% % In particular, we continue to define isolation semantics as
-% % constraints over $\visZ$ and $\soZ$, but confine their domain of
-% % interpretation to the given trace so that they now become trace
-% % well-formedness constraints. Well-formedness constraints can be
-% % combined into a trace invariant ($\I$).
-
-% As described previously, \rulelabel{E-Aux} abstracts away the
-% operation-specific behaviour of a machine step as a function ($\F$)
-% that accepts a set ($S$) of effects chosen by the machine to make
-% visible to the operation, interprets the operation w.r.t. $S$, and
-% returns an appropriate effect that encodes its return value. Rules
-% \rulelabel{E-Var}, \rulelabel{E-Asgn} and \rulelabel{E-Commit} define
-% such functions for read, write and commit operations, respectively.
-% The effect returned by the function in each case includes its
-% transaction id ($T_i$) along with an arbitrarily chosen effect id
-% ($j$) that is later verified to be unique in \rulelabel{E-Aux}. The
-% $\rval$ for a write is the value being written, and for commit it is
-% $\bot$. In case of a read, the value read depends on how the read
-% operation chooses to interpret the given set ($S$) of visible effects.
-% The interpretation may depend on application semantics. For example, a
-% monotonically increasing counter application may choose to let a write
-% with the largest value determine the value of a read. To accommodate
-% multiple interpretations, the semantics is made parametric over an
-% interpretation function ($\interp{\cdot}$) that accepts a set of
-% effects and a variable name, and returns the value associated with the
-% variable. A straightforward interpretation function that chooses the
-% last write (i.e., write with largest id) is shown below:
-
-% \begin{smathpar}
-% \begin{array}{lcl}
-%   \isMax(S,\eta) & \Leftrightarrow &  \forall (\eta'\in S).  
-%   \kind(\eta') = \kind(\eta) \\
-%   & & \hspace*{0.4in}\Rightarrow \eta' = \eta \disj \id(\eta') < \id(\eta)\\
-
-% \interp{S}(X) & = & \C{if}\;(\exists (\eta \in S). \kind(\eta) = \C{WR}(X) 
-%   \wedge \isMax(S,\eta)) \\
-%   & & \C{then}\;\rval(\eta)\;\C{else}\;0\\
-% \end{array}
-% \end{smathpar}
-
-% \noindent Rules \rulelabel{E-Top-Ctx} and \rulelabel{E-Txn-Ctx} define
-% congruence properties for top-level terms and transaction-bound terms,
-% respectively. The rules and evaluation contexts ($\ectx$ and
-% $\ectx_i$) are defined such that only certain kinds of terms are
-% allowed at the top-level and inside a transaction. In particular, a
-% \txnimp program at the top-level can either be a transaction, or a
-% parallel composition of transactions. A command inside a \C{txn} block
-% can either be an assignment, or a sequential composition of
-% assignments.
-
-% \input{ansi-isolation}
+\input{isolation}

### 2017-07-02 motivation.tex
edits{

@@
 example shows often inadequate, reasoning about weak isolation in
-terms of low-level traces~\cite{adyaphd,gotsmanconcur15} complicates
-high-level reasoning.  A possible alternative would interleave weak
-isolation implementation details within the program, yielding a
-(more-or-less) conventional concurrent program that can be then
-subject to classical concurrent verification methods.  Considering the
-size and complexity of real-world transaction systems, this strategy
-is unlikely to scale.
+terms of low-level traces~\cite{adyaphd,gotsmanconcur15} on memory
+read and write actions complicates high-level reasoning.  A possible
+alternative would interleave weak isolation implementation details
+within the program, yielding a (more-or-less) conventional concurrent
+program that can be then subject to classical concurrent verification
+methods.  Considering the size and complexity of real-world
+transaction systems, this strategy is unlikely to scale.
 
@@
 This simple characterization of RR isolation allows us to verify the
-consistency condtions associated with the \C{new\_order}
+consistency conditions associated with the \C{new\_order}
 transaction. First, since the database doesn't change ($\Delta' =
@@
 commits), during the execution of the transaction's body, we can
-reason about \C{new\_order} as if it executed in complete isolation
+reason about \C{new\_order} as though it executed in complete isolation
 until its commit point, leading to a verification process similar to
@@
 \C{new\_order} finishes execution and becomes ready to commit, a
-transition that transfers the transaction'slocal writes ($\delta$) to
+transition that transfers the transaction's local writes ($\delta$) to
 the unchanged database state ($\Delta$).  However, we are required to
@@
 a significantly simpler abstraction than shared memory. There are no
-pointers, or linked data structures, or aliasing.  Although a database
+pointers, linked data structures, or aliasing.  Although a database
 essentially abstracts a mutable state, the state is mutated through a
@@
 logic, allowing us to leverage SMT solvers for automatic proofs
-without having to infer loop invariant, or provide intermediate
+without having to infer loop invariants, or provide intermediate
 assertions.  Sec.~\ref{sec:automation} describes this translation. In
@@
 covered in Sec.~\ref{sec:automation}.  The following two sections,
-however, focus on laying theoretical foundations for reasoning about
-weak isolation.
+however, first focus on laying the theoretical foundations for
+reasoning about weak isolation.

### 2017-07-02 isolation.tex
Isolation text.

\subsection{Isolation Specifications}
\label{sec:isolation}

Operational semantics of Fig.~\ref{fig:txnimp} does not employ
concurrency control constructs in any form. Any additional concurrency
control provided by the database implementation, beyond that which is
already incuded in the isolation specification, also needs to be
axiomatized in $\I$. We first describe such axiomatizations, and then
present the specifications of the standard isolation levels as
implemented by popular off-the-shelf databases.

\textbf{Unique Ids}. Most relational databases associate every record
in a table with a unique identifier. Often, auto-generated unique
identifiers are part of the schema itself (via SQL's \C{UNIQUE} and
\C{AUTO\_INCREMENT} keywords), but if not, database adds one (e.g.,
MySQL's InnoDB engine automatically adds a 6-byte identifier if none
exists). Enforcing global uniqueness of identifiers requires exclusive
locks, which are absent from Fig.~\ref{fig:txnimp}. The
\rulelabel{E-Insert} rule assigns an id to the newly added record
after checking that it is not already present in $\stg$ and $\stl$.
But the new record is only added to $\stl$, which is invisible outside
the transaction. Thus, the uniqueness check for the same id also
passes in a concurrent transaction, resulting in duplicate ids
after both commits. To prevent this, we need the following proposition
in $\I$ of every transaction, regardless of whether it is executing
or committing:
\begin{smathpar}
\begin{array}{lcl}
  \I_{id}(\stl,\stg,\stg') & = & \forall(r\in\stl).~
      r.\idf\notin \dom(\stg) \Rightarrow r.\idf\notin \dom(\stg').
\end{array}
\end{smathpar}
$\I_{id}$ makes sure that a globally unique id generated during a
transaction's execution remains globally unique until it commits. An
interference from a concurrent transaction that adds the same id to
$\stg$ is prohibited.

\textbf{Write Locks}. Transactions in MySQL and Postgres (effectively)
obtain exclusive locks on records they update, and do not release them
until they commit, thus preventing write-write (ww) conflicts. To
obtain this behavior on the abstract machine of Fig.~\ref{fig:txnimp},
the following proposition needs to be in $\I$ for every transaction:
\begin{smathpar}
\begin{array}{lcl}
  \I_{ww}(\stl,\stg,\stg') & = & \forall(r'\in\stl)(r \in \stg).~
      r.\idf = r'.\idf  \Rightarrow r\in\stg'.
\end{array}
\end{smathpar}
That is, given a record $r'\in\stl$, if there exists an $r\in\stg$
with the same id (i.e., $r'$ is an updated version of $r$), then $r$
must be present unmodified in $\stg'$. This prevents a concurrent
transaction from changing $r$, thus simulating the behavior of an
exclusive lock. There is however a caveat: if we assume extensional
equality over records, a write by a concurrent transaction that
doesn't change $r$'s contents is still allowed. While this in itself
is not a problem, the concurrent transaction may modify other records,
which then become visible in the current transaction. A write lock
prevents this behavior, whereas our axiomatization ($\I_{ww}$) allows
it. An easy fix for this is to add version timestamps to records that
effectively intensionalizes equality. Nonetheless, imprecise
axiomatization of write locks hasn't been a problem in practice.

\textbf{Read-Only Transactions}. Certain databases implement special
previleges for read-only transactions. Read-only behavior can be
enforced on a transaction by including the following proposition in
its $\I$:
\begin{smathpar}
\begin{array}{lcl}
  \I_{ro}(\stl,\stg,\stg') & = & \stl = \emptyset\\
\end{array}
\end{smathpar}
If a transaction declared as read-only performs a write, then its
$\stl\neq \emptyset$, and the transaction never commits.

### 2017-07-03 isolation.tex
Isolation.

@@
 
-Operational semantics of Fig.~\ref{fig:txnimp} does not employ
-concurrency control constructs in any form. Any additional concurrency
-control provided by the database implementation, beyond that which is
-already incuded in the isolation specification, also needs to be
-axiomatized in $\I$. We first describe such axiomatizations, and then
-present the specifications of the standard isolation levels as
-implemented by popular off-the-shelf databases.
+Databases implement isolation levels through a combination of various
+concurrency control mechanisms, which are absent from
+Fig.~\ref{fig:txnimp}. In this section, we first axiomatize the
+semantics of various concurrency control constructs, and then define
+the precise semantics of the isolation levels as they are implemented
+on Postgres and MySQL datbases.
+% Any additional concurrency control provided by the database
+% implementation, beyond that which is already incuded in the
+% isolation specification, also needs to be axiomatized in $\I$. We
+% first describe such axiomatizations, and then present the
+% specifications of the standard isolation levels as implemented by
+% popular off-the-shelf databases.
 
@@
 
+\textbf{Snapshots} Both MySQL and Postgres implement isolation levels
+that execute transactions against a static snapshot of the database.
+The semantics of a snapshot be axiomatized as following:
+\begin{smathpar}
+\begin{array}{lcl}
+  \I_{ss}(\stl,\stg,\stg') & = & \stg' = \stg.
+\end{array}
+\end{smathpar}
+% The snapshot semantics however may not be applicable throughout the
+% lifetime of the transaction, like in the case of Repeatable Read (RR).
+
+\textbf{Rollbacks} Rollbacks triggered by the post-facto detection of
+write-write conflicts can be axiomatized as following:
+\begin{smathpar}
+\begin{array}{lcl}
+  \I_{rb}(\stl,\stg,\stg') & = & \forall(r\in\stl)(r'\in\stg).~
+  r'.\idf = r.\idf \Rightarrow r'\in\stg'.
+\end{array}
+\end{smathpar}
+
 \textbf{Read-Only Transactions}. Certain databases implement special
@@
 $\stl\neq \emptyset$, and the transaction never commits.
+
+%\usepackage[table,xcdraw]{xcolor}
+\begin{table}[]
+\centering
+\begin{tabular}{l|l|l|l|l|l|l|l|l|l|l|l|l|}
+\cline{2-13}
+                                & \multicolumn{6}{c|}{{\color[HTML]{333333} Postgres}}                         & \multicolumn{6}{c|}{MySql}                                                   \\ \cline{2-13}
+                                & \multicolumn{2}{c|}{RC} & \multicolumn{2}{c|}{RR} & \multicolumn{2}{c|}{SER} & \multicolumn{2}{c|}{RC} & \multicolumn{2}{c|}{RR} & \multicolumn{2}{c|}{SER} \\ \cline{2-13}
+                                & T          & F          & T          & F          & T           & F          & T          & F          & T          & F          & T           & F          \\ \hline
+\multicolumn{1}{|l|}{$\I_{ss}$} &            &            & \checkmark &            & \checkmark  & \checkmark &            &            & \checkmark &            & \checkmark  & \checkmark \\ \hline
+\multicolumn{1}{|l|}{$\I_{rb}$} &            &            &            & \checkmark &             &            &            &            &            &            &             &            \\ \hline
+\multicolumn{1}{|l|}{$\I_{ro}$} &            &            &            &            &             &            &            &            & \checkmark & \checkmark &             &            \\ \hline
+\multicolumn{1}{|l|}{$\I_{id}$} & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark  & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark  & \checkmark \\ \hline
+\multicolumn{1}{|l|}{$\I_{ww}$} & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark  & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark  & \checkmark \\ \hline
+\end{tabular}
+
+\caption{The semantics of isolation levels on Postgres and MySQL}
+\label{fig:iso-table}
+\end{table}
+
+With help of the above axiomatizations, we can now define the precise
+semantics of various isolation levels on Postgres and MySQL. The table
+in Table~\ref{fig:iso-table} captures the summary.

### 2017-07-03 isolation.tex
new opsem

@@
 the precise semantics of the isolation levels as they are implemented
-on Postgres and MySQL datbases.
+on two real-world databases -  PostgreSQL and MySQL.
 % Any additional concurrency control provided by the database
@@
 But the new record is only added to $\stl$, which is invisible outside
-the transaction. Thus, the uniqueness check for the same id also
-passes in a concurrent transaction, resulting in duplicate ids
+the transaction. Thus, it is possible that the uniqueness check for the same id may
+also succeed in a concurrent transaction, resulting in duplicate ids
 after both commits. To prevent this, we need the following proposition
@@
 \end{smathpar}
-$\I_{id}$ makes sure that a globally unique id generated during a
-transaction's execution remains globally unique until it commits. An
-interference from a concurrent transaction that adds the same id to
-$\stg$ is prohibited.
+$\I_{id}$ ensures that a globally unique id generated during a
+transaction's execution remains globally unique until it commits, prohibiting
+interference from a concurrent transaction that adds the same id.
 
-\textbf{Write Locks}. Transactions in MySQL and Postgres (effectively)
+\textbf{Write Locks}. Transactions in MySQL and PostgreSQL (effectively)
 obtain exclusive locks on records they update, and do not release them
@@
 obtain this behavior on the abstract machine of Fig.~\ref{fig:txnimp},
-the following proposition needs to be in $\I$ for every transaction:
+the following proposition must hold for every transaction:
 \begin{smathpar}
@@
 
-\textbf{Snapshots} Both MySQL and Postgres implement isolation levels
+\textbf{Snapshots} Both MySQL and PostgresSQL implement isolation levels
 that execute transactions against a static snapshot of the database.
@@
 \cline{2-13}
-                                & \multicolumn{6}{c|}{{\color[HTML]{333333} Postgres}}                         & \multicolumn{6}{c|}{MySql}                                                   \\ \cline{2-13}
+                                & \multicolumn{6}{c|}{{\color[HTML]{333333} PostgreSQL}}                         & \multicolumn{6}{c|}{MySql}                                                   \\ \cline{2-13}
                                 & \multicolumn{2}{c|}{RC} & \multicolumn{2}{c|}{RR} & \multicolumn{2}{c|}{SER} & \multicolumn{2}{c|}{RC} & \multicolumn{2}{c|}{RR} & \multicolumn{2}{c|}{SER} \\ \cline{2-13}
@@
 
-\caption{The semantics of isolation levels on Postgres and MySQL}
+\caption{The semantics of isolation levels on PostgreSQL and MySQL}
 \label{fig:iso-table}
@@
 With help of the above axiomatizations, we can now define the precise
-semantics of various isolation levels on Postgres and MySQL. The table
+semantics of various isolation levels on PostgreSQL and MySQL. The table
 in Table~\ref{fig:iso-table} captures the summary.

### 2017-07-03 motivation.tex
new opsem

@@
 \caption{Database state transitions corresponding to an execution of
-  an RR transaction $T_1$ on PostgresSQL. $T_2$ and $T_3$ are concurrent
+  an RR transaction $T_1$ on PostgreSQL. $T_2$ and $T_3$ are concurrent
   transactions. The database state has three data items, $a$, $b$ and $c$. }
@@
 then $\forall x\in\mathit{dom}(\delta)$, $\Delta_c(x) =
-\Delta(x)$. To summarize, the operational semantics of PostgresSQL's RR
+\Delta(x)$. To summarize, the operational semantics of PostgreSQL's RR
 implementation can be captured as an axiomatization over transitions of

### 2017-07-03 opsem.tex
new opsem

@@
 transactions ($\ctxn{i}{\I}{c}$) and their parallel composition
-($c_1\,||\,c_2$). The $\I$ in \C{TXN} block syntax is the
+($c_1\,||\,c_2$). The $\I$ in the \C{TXN} block syntax is the
 transaction's isolation specification, which will be explained later.
 Certain terms that only appear at run-time are also present in $c$.
-These include a \C{TXN} block tagged with sets ($\stl$ and $\stg$) of
-records that assume special meaning in operational semantics, and a
-\C{FOREACH} loop that keeps track of the set ($s_1$) of items already
+These include a \R{TXN} block tagged with sets ($\stl$ and $\stg$) of
+records whose meaning is explained below, and a runtime \R{FOREACH}
+expression that keeps track of the set ($s_1$) of items already
 iterated, and the set ($s_2$) of items yet to be iterated. Note that
-even the surface-level \C{FOREACH} shown here is a little different
-from the one used in previous sections; its higher-order argument has
-two arguments, $y$ and $z$, which are invoked (during the reduction)
-with the set of already-iterated items, and the current item,
-respectively. This form of \C{FOREACH} lends itself to inductive
-reasoning, facilitating inductive proofs (Sec.~\ref{sec:reasoning}).
+even the surface-level syntax of the \C{FOREACH} command shown here is
+a little different from the one used in previous sections; its
+higher-order function has two arguments, $y$ and $z$, which are
+invoked (during the reduction) with the set of already-iterated items,
+and the current item, respectively. This form of \C{FOREACH} lends
+itself to inductive reasoning that will be useful for verification
+(Sec.~\ref{sec:reasoning}).
 
-% also We let $T_i$ for $i \in \mathbb{N}$
+% also We let $T_i$ for $i \in \matshbb{N}$
 % range over transaction identifiers. When it is evident we are
@@
 terms of an abstract machine that executes a command, and updates
-either a transaction-local database ($\stl$), or the global database
-($\stg$). Database ($\stg$) is a modeled as a set of records of a
-pre-defined type, i.e., they all belong to a single table.
-Generalization to multiple tables is straightforward. Records in
-$\stg$ are uniquely identifiable through their $\idf$ field, which is
-auto-generated and does not belong to the surface language, i.e.,
-$\C{id}\notin f$. For a set $S$ of records, we define $\dom(S)$ as the
-set of unique ids of all the records in $S$. Thus $|\dom(\stg)| =
-|\stg|$. During its execution, a transaction may write to multiple
-records in $\stg$. Atomicity dictates that such writes shouldn't be
-available in $\stg$ until the transaction commits. We therefore
-associate each transaction with a local database ($\stl$) that stores
-\emph{only} the uncommitted records\footnote{While SQL's \C{UPDATE}
-admits writes at the granularity of record fields, databases, in
-reality, enforce record-level locking, allowing us to think of
-``uncommitted writes'' as ``uncommitted records''. }. Uncommitted
-records include deleted records, which are identified by a hidden
-$\delf$ field set to \C{true}. When the transaction commits, its local
-database is atomically \emph{flushed} to the global database,
-committing the uncommitted records. The flush operation ($\gg$) is defined as
-following:
+either a transaction-local ($\stl$), or global ($\stg$) database, both
+of which are modeled as a set of records of a pre-defined type, i.e.,
+they all belong to a single table.  The generalization to multiple
+tables is straightforward.  Records in $\stg$ are uniquely identifiable
+through their $\idf$ field, which is auto-generated and does not
+belong to the surface language, i.e., $\C{id}\notin f$. For a set $S$
+of records, we define $\dom(S)$ as the set of unique ids of all 
+records in $S$. Thus $|\dom(\stg)| = |\stg|$. During its execution, a
+transaction may write to multiple records in $\stg$. Atomicity
+dictates that such writes should not be visible in $\stg$ until the
+transaction commits. We therefore associate each transaction with a
+local database ($\stl$) that stores such uncommitted
+records\footnote{While SQL's \C{UPDATE} admits writes at the
+  granularity of record fields, databases, in reality, enforce
+  record-level locking, allowing us to think of ``uncommitted writes''
+  as ``uncommitted records''. }. Uncommitted records include deleted
+records, which are identified by a hidden $\delf$ field set to
+\C{true}. When the transaction commits, its local database is
+atomically \emph{flushed} to the global database, committing these
+heretofore uncommitted records. The flush operation ($\gg$) is defined as
+follows:
 \begin{smathpar}
@@
 Let $\stg' = \stl\gg\stg$. A record $r$ belongs to $\stg'$ iff it
-belongs to $\stg$ and not been updated in $\stl$, i.e., $r.\idf \notin
-\dom(\stl)$, or it belongs to $\stl$, i.e., it is either a new record,
-or an updated version of an old record, but the updatation is not a
-deletion ($\neg r.\delf$). Thus, flush defines the result of
-atomically applying transaction's local writes to the global database.
-Besides the commit, flush also helps a transaction read its own
-writes. Intuitively, the result of a read operation inside a
+belongs to $\stg$ and has not been updated in $\stl$, i.e., $r.\idf
+\notin \dom(\stl)$, or it belongs to $\stl$, i.e., it is either a new
+record, or an updated version of an old record, but the update is not
+a deletion ($\neg r.\delf$). Thus, flush defines the result of
+atomically applying a transaction's local writes to the global
+database.  Besides the commit, flush also helps a transaction read its
+own writes. Intuitively, the result of a read operation inside a
 transaction must be computed on the database resulting from flushing
 the current local state ($\stl$) to the global state ($\stg$). The
-abstract machine of Fig.~\ref{fig:txnimp}, however, doesn't let a
-transaction read its own writes. This reduces verbosity and simplifies
-semantics, without losing any generality; substituting $\stl\gg\stg$
-for $\stg$ at select places in reduction rules recovers the real
+abstract machine of Fig.~\ref{fig:txnimp}, however, does not let a
+transaction read its own writes. This simplifies the semantics,
+without losing any generality, since substituting $\stl\gg\stg$ for
+$\stg$ at select places in the reduction rules effectively allows
+reads of uncommitted transaction writes to be realized.
 
-Small-step semantics is stratified into a transaction-local reduction
+The small-step semantics is stratified into a transaction-local reduction
 relation, and a top-level reduction relation. Transaction-local
@@
 statements is defined straightforwardly.  \C{INSERT} adds a new record
-to $\stl$ after adding a unique identifier (more discussion on
-uniqueness later). \C{DELETE} finds the records in $\stg$ that match
+to $\stl$ after adding a unique identifier. \C{DELETE} finds the records in $\stg$ that match
 the search criteria defined by its boolean function argument, and adds
@@
 The reduction of \C{FOREACH} starts by first converting it to its
-run-time form that also keeps track of the iterated items ($s_1$),
-besides the yet-to-be-iterated items ($s_2$). Initially, $s_1$ is
+run-time form to keep track of iterated items ($s_1$),
+as well as  yet-to-be-iterated items ($s_2$). Initially, $s_1$ is
 empty. As the elements are iterated, they are removed from $s_2$ and
-added to $s_1$ . Iteration involves invoking the higher-order function
+added to $s_1$ . Iteration involves invoking its function argument
 with $s_1$ and the current element $x$ (note: $\uplus$ in $\{x\}
-\uplus s_2$ denotes a disjoint union). The reduction ends when the
+\uplus s_2$ denotes a disjoint union). The reduction ends when 
 $s_2$ becomes empty. The reduction rules for conditionals, \C{LET}
-binders, and sequences are straightforward, hence ommitted.
+binders, and sequences are straightforward, and ommitted for brevity.
 
-The top-level reduction relation defines small-step semantics of
+The top-level reduction relation defines the small-step semantics of
 transactions, and their parallel composition. A transaction always
-comes tagged with an isolation specification $\I$ that dictates the
-timing and nature of iterferences that the transaction can witness.
-Formally, $\I$ is a function that maps a boolean ($\mathbb{B}$) and a
-triplet of database states (three sets of records:
-$\Pow{r}\times\Pow{r}\times\Pow{r}$) to a logical proposition
-($\Prop$). The abstract machine calls $\I$ while reducing
-$\ctxn{i}{\I}{c}$. The boolean argument identifies if the transaction
-is still executing (\C{true}), or if it has finished execution and
-ready to commit (\C{false}).  Among the triplet of sets (the second
-argument), first is the (current) transaction-local database state
-($\stl$), second is the state ($\stg$) of the global database when the
-transaction last took a step, and third is the current state ($\stg'$)
-of the global database.  Intuitively, $\stg'\neq\stg$ indicates an
-interference from a concurrent transaction, and the $\I$ decides if
-this interference is allowed or not, considering the local database
+comes tagged with an isolation specification $\I_e$ or $\I_c$ that
+dictates the timing and nature of interferences that a transaction can
+witness, both during its execution ($\I_e$), and when it is about to
+commit ($\I_c$).  Formally, $\I_e$ and $\I_c$ are functions that map
+the (current) transaction-local database state ($\stl$), the state
+($\stg$) of the global database when the transaction last took a step,
+and the current state ($\stg'$) of the global database.  Intuitively,
+$\stg'\neq\stg$ indicates an interference from another concurrent
+transaction, and the predicates $\I_e$ and $\I_c$ decide if this
+interference is allowed or not, taking into account the local database
 state ($\stl$). For instance, as described in
-(Sec.~\ref{sec:motivation}), an RR transaction on Postgres defines
+(Sec.~\ref{sec:motivation}), an RR transaction on PostgreSQL defines
 $\I$ as following:
@@
 \begin{array}{lcl}
-\I\,\C{true}\,(\stl,\stg,\stg') & = & \stg' = \stg\\
-\I\,\C{false}\,(\stl,\stg,\stg') & = & \forall(r\in\stl)(r'\in\stg).~
-  r'.\idf = r.\idf \Rightarrow r'\in\stg'.
+\I_e\,\,(\stl,\stg,\stg') & = & \stg' = \stg\\
+\I_c\,\,(\stl,\stg,\stg') & = & \forall(r\in\stl)(r'\in\stg).~ r'.\idf = r.\idf \Rightarrow r'\in\stg'.
 \end{array}
@@
 
-The reduction of a $\ctxn{i}{\I}{c}$ begins by first converting it to
-its run-time form $\ctxn{i}{\I,\stl,\stg}{c}$, where $\stl =
-\emptyset$, and $\stg$ the current (global) database. The top-level
-reduction rules in Fig.~\ref{fig:txnimp} assume this form. Rule
-\rulelabel{E-Txn} reduces $\ctxn{i}{\I,\stl,\stg}{c}$ under a database
-state ($\stg'$), only if the isolation specification ($\I$) allows the
-interference between $\stg$ and $\stg'$ during the execution. Rule
-\rulelabel{E-Commit} commits the transaction
-$\ctxn{i}{\I,\stl,\stg}{c}$ by flushing its uncommitted records to the
-database. This is done only if the interference between $\stg$ and
-$\stg'$ is allowed during the commit by the isolation specification.
+The reduction of a $\ctxn{i}{\I_e}{c}$ begins by first converting it
+to its run-time form $\ctxn{i}{\I_e,\stl,\stg}{c}$, where $\stl =
+\emptyset$, and $\stg$ the current (global) database.  Rule
+\rulelabel{E-Txn} reduces $\ctxn{i}{\I_e,\stl,\stg}{c}$ under a
+database state ($\stg'$), only if the isolation specification ($\I$)
+allows the interference between $\stg$ and $\stg'$ during that
+step. Rule \rulelabel{E-Commit} commits the transaction
+$\ctxn{i}{\I_c,\stl,\stg}{c}$ by flushing its uncommitted records to
+the database. This is done only if the interference between $\stg$ and
+$\stg'$ is allowed at the commit point by the isolation specification.
+The distinction between $\I_e$ and $\I_c$ allows us to model realistic
+database features such as snapshot isolation that isolates the effects
+a transaction can witness during its execution but exposes
+interferences from concurrent transactions at the commit point.
 

### 2017-07-03 reasoning.tex
Sec. 4 part 1.

@@
 
-% We now describe a proof system that lets us demonstrate that a \txnimp
-% program satisfies its high-level invariants in all small-step
-% executions that satisfy a chosen trace invariant ($\I$).\footnote{It is
-% important to keep in mind the distinction between high-level invariants
-% (\eg~\C{X$\ge$0}) and trace invariants (\eg~$\I =
-% \lambda\E.\,\C{RC}(\C{Wd1})\wedge \C{SI}(\C{Wd2})$). The former
-% constitute \emph{proof} \emph{obligations} for the programmer, whereas
-% the latter is a combination of transaction-specific \emph{assumptions} about 
-% operational characteristics of the underlying data store. } 
-% Since the trace invariant captures transaction-isolation constraints,
-% the demonstration is a proof that the given selection of isolation
-% levels for transactions is sufficient to enforce a program's invariants.
+We now describe a proof system that lets us prove the correctness of a \txnimp
+program $c$ w.r.t its high-level invariant $I$, on a machine that
+satisfies the isolation specifications ($\I$) of its
+transactions\footnote{Note the difference between $I$ and $\I$. The
+former constitute \emph{proof} \emph{obligations} for the programmer,
+whereas the latter describes a transaction's \emph{assumptions} about
+operational characteristics of the underlying machine.}.  Our proof
+system is essentially an adapatation of the rely-guarantee reasoning
+framework to the setting of weakly isolated database transactions. It
+deals with the additional complexity introduced by this setting, while
+also taking advantage of the opportunities it provides to simplify the
+reasoning. The primary challenge is the weak isolation; how do we
+relate a transaction's isolation specification ($\I$) to its rely
+relation ($R$) that describes the environment, so that the
+interference is considered only insofar as the isolation level allows
+it? Our proof system shows how to compute a rely relation ($R$) modulo
+the isolation specification ($\I$) of a transaction. Another
+characteristic of the transaction setting is the atomicity of a
+trasaction's aggregate writes; a transaction's writes are not visible
+to its peers until it commits. In the context of rely-guarantee, this
+means that the transactions guarantee ($G$) should capture the
+aggregate effect of a transaction, and not its individual writes.
+While \C{atomic} blocks are also present in the shared-memory
+programs, the fact that a transactions are weakly isolated introduces
+complexity.  Unlike an \C{atomic} block, the effect of a transaction
+is \emph{not} a sequential composition of the effects of its
+statements because each statement witnesses a potentially different
+version of the state. This is illustrated in
+Fig.~\ref{fig:atomic-vs-transaction}, where a series of operations on
+\C{x} can be composed to describe the effect of an \C{atomic} block,
+while same cannot be done in a transaction. Thus, weak isolation needs
+to be taken into account when defining and verifying guarantees, and
+our proof system shows how.
 
-% As mentioned earlier, our proof system is a rely-guarantee framework
-% that admits compositional reasoning by abstracting away environmental
-% interference as a rely relation. A conspicuous difference between a
-% standard development of rely-guarantee and ours is that, while the
-% former reasons in terms of program states (variable to value
-% bindings), we reason in terms of executions as captured by their
-% traces ($\E$). In particular, our rely ($R$) and guarantee ($G$)
-% relations relate executions (i.e., $R,\,G\,\subseteq\,\E\times\E$),
-% and our pre- ($P$) and post- ($Q$) conditions are assertions over
-% executions (i.e., $P,\,Q\, : \E \rightarrow \Prop$). Our development
-% also facilitates state-based reasoning via the interpretation function
-% ($\interp{\cdot}$) introduced in \S\ref{sec:syntax}, which interprets
-% the given set of effects as a state. For example, if a \emph{bi-state}
-% rely relation relates every pair of states $\sigma$ and $\sigma'$ such
-% that $\sigma'(X) \ge \sigma(X) \ge 0$, the corresponding
-% \emph{bi-execution} rely relation relates every pair of executions
-% $\E$ and $\E'$ such that $\interp{\E.\A}(X) \ge \interp{\E'.\A}(X) \ge
-% 0$ . Assertions on states are also written similarly. For instance,
-% consider the post-condition $\C{B\,=\,k-a1-a2}$ of the program in
-% Fig.~\ref{fig:motiv-eg-1}. The corresponding assertion on the
-% post-state ($\lambda\sigma.~\sigma(\C{B}) \,=\, \C{k-a1-a2}$) asserts
-% that in all states resulting from executing the program, the value of
-% \C{B} is \C{k-a1-a2}. The equivalent execution-based assertion
-% ($\lambda\E.~\interp{\E.\A}(\C{B})\,=\,\C{k-a1-a2}$) asserts that in
-% all executions of the program, the value of \C{B} is \C{k-a1-a2}.
-% However, having access to an execution facilitates assertions that go
-% beyond the state. An example is an invariant of \C{Wd1} described
-% abstractly in \S\ref{sec:motivation}, and reified as an
-% execution-based assertion below:
-
-% \begin{smathpar}
-% \begin{array}{l}
-%   \lambda\E.~ \neg(\underE{\committed(\C{Wd1}}) \conj 
-%       \underE{\C{Wd1} \wrstoar \C{B}}\\
-%   \hspace*{0.2in}\conj(\neg(\underE{\committed(\C{Wd2})}) \Rightarrow 
-%       \C{\interp{\E.\A}(B) = k-a1}) \\
-%    \hspace*{0.2in}\conj (\underE{\committed(\C{Wd2})} \Rightarrow 
-%       \C{\interp{\E.\A}(B) = k-a1-a2})
-% \end{array}
-% \end{smathpar}
-% % Where, $\committed$ is defined as:
-% % \begin{smathpar}
-% % \begin{array}{lcl}
-% % \underE{\committed(T_i)} & \Leftrightarrow &
-% %   \exists\eta.~\underE{\eta\in T_i} \conj \kind(\eta) = \C{COMMIT}\\
-% % \end{array}
-% % \end{smathpar}
-
-% \noindent One of the conjuncts asserts that \C{Wd1} has written to \C{B} - a
-% fact which cannot be deduced solely from the value of \C{B} (esp. in the
-% presence of interference), but can be expressed as a proposition over
-% the execution trace.
-
-% \subsection{The Rely-Guarantee Judgment}
-% \label{sec:rely-guarantee}
+\subsection{The Rely-Guarantee Judgment}
+\label{sec:rely-guarantee}
 
@@
 
-% The standard rely-guarantee judgment is a quintuple
-% $\rg{P,R}{c}{G,Q}$, which informally asserts that if a command $c$ is
-% executed in a state that satisfies its pre-condition $P$, provided
-% that every interference step during the execution is contained inside
-% the rely relation $R$, then the effect that each step of executing $c$
-% has on the state is captured by $G$, and the final state of execution
-% satisfies the post-condition $Q$. The terms ``execution'' and ``step''
-% are defined with respect to an operational semantics, which, in our
-% case, is parameterized on the trace invariant $\I$. Consequently, our
-% rely-guarantee judgment for commands is the sextuple
-% $\rg{\I,P,R}{c}{G,Q}$, whose semantics differs from the quintuple in
-% that it requires every step of the execution to additionally preserve
-% the trace invariant $\I$ for the post-condition to be valid.  As
-% usual, if $c$ is inside a transaction $T_i$, we write $\txnbox{c}_i$.
-% \txnimp expressions are side-effecting (they generate \C{RD} effects),
-% and admit interference during their evaluation.  Therefore, like
-% commands, expressions need to be reasoned about explicitly. But,
-% unlike commands, expressions evaluate to values, and the reasoning
-% framework should also admit assertions on such values.  We therefore
-% define a separate judgment for expressions - the septuple
-% $\rg{\I,P,R}{e}{G,C,Q}$. The new entrant $C : \mathbb{N} \rightarrow
-% \Prop$ is an assertion on the return value of $e$.
+Fig.~\ref{fig:rg-rules} shows an illustrative subset of the
+rely-guarantee (RG) reasoning rules for $\txnimp$. We define two RG
+judgments: top-level ($\rg{I,R}c{G,I}$), and transaction-local ($\R
+\vdash \hoare{P}c{Q}$).  Recall that the standard RG judgment is the
+quintuple $\rg{P,R}{c}{G,Q}$. Instead of a separate $P$ and $Q$, our
+top-level judgment has $I$ for both pre- and post-conditions, because
+our focus is on verifying that a \txnimp program \emph{preserves}
+database's consistency conditions\footnote{The terms \emph{consistency
+condition}, \emph{high-level invaraint}, and \emph{integrity
+constraint} all mean the same.}. Transaction-local RG judgment doesn't
+include a guarantee relation because transaction-local effects are not
+visible outside a transaction. Also, the rely relation ($\R$) of the
+transaction-local judgment is not same as the top-level rely relation
+($R$) as it takes into account the transaction's isolation
+specification ($\I$). Intuitively, $\R$ is $R$ modulo $\I$; formal
+definition will be given shortly. Recall that a transaction writes to
+its local database ($\stl$), which is then flushed when the
+transaction commits. Thus, the guarantee of a transaction depends on
+the state of its local database at the commit point. The pre- and
+post-condition assertions ($P$ and $Q$) in the local judgment
+facilitate tracking the changes to the transaction-local state, which
+eventually helps us prove the validity of the transaction's guarantee.
+Note that $P$ and $Q$ are bi-state assertions; they relate
+transaction-local database state ($\stl$) to the global database state
+($\stg$). Thus, the transaction-local judgment effectively tracks how
+transaction-local and global states change in relation to each other.
+
+The quintessential aspect of a rely-guarantee judgment is the
+stability condition, which, intuitively, requires the validity of an
+assertion $\phi$ to be unaffacted by the interference, i.e., the rely
+relation $R$. In conventional RG, stability is defined as following
+($\sigma$ denotes a state):
+\begin{smathpar}
+\begin{array}{lcl}
+\stable(R,\phi) & \Leftrightarrow & \forall \sigma,\sigma'.~
+\phi(\sigma) \conj R(\sigma,\sigma') \Rightarrow \phi(\sigma')\\
+\end{array}
+\end{smathpar}
+Due to the presence of local and global database states, and isolation
+specification, we use multiple definitions of stability in
+Fig.~\ref{fig:rg-rules}, but they all convey the same intuition as
+above. We will introduce them as they are encountered.
+
+\rulelabel{RG-Txn} is the top-level rule that lets us prove a
+transaction preserves the high-level invariant $I$ when executed under
+the required isolation as specified by $\I$. It relies on the
+transaction-local judgment to verify the the transaction body ($c$).
+The precondition $P$ of $c$ must follow from the fact that the
+transaction-local database ($\stl$) is initially empty, and the global
+database satisfies the high-level invariant $I$. The rely relation
+($\R_l$) for the local judgment is a ternary relation computed as $R$
+modulo $\I$ as following:
+\begin{smathpar}
+\begin{array}{lcl}
+\R_l(\stl,\stg,\stg') & \Leftrightarrow & R(\stg,\stg') \conj
+\I\,\C{true}\,(\stl,\stg,\stg')
+\end{array}
+\end{smathpar}
+Thus, $\R_l$ allows an interference only if it doesn't violate the
+execution-time isolation specification of the transaction. If a
+certain interference violates the isolation spec, i.e., $R(\stg,\stg')
+\Rightarrow \neg(\I\,\C{true}\,(\stl,\stg,\stg'))$, then
+$\R_l(\stl,\stg,\stg') \Leftrightarrow false$, and any assertion is
+trivially stable w.r.t that interference. This is sensible considering
+such interference is pre-empted in the operational semantics.
+
+Recall that $P$ and $Q$ of the transaction-local RG judgment are
+binary assertions; they relate local and global database states. The
+local judgment rules require one or both of them to be stable. The
+stability of a binary assertion $Q$ w.r.t a ternary rely relation $\R$
+is defined as following:
+\begin{smathpar}
+\begin{array}{c}
+\forall \stl,\stg,\stg'.~ Q(\stl,\stg) \conj \R(\stl,\stg,\stg')
+\Rightarrow Q(\stl,\stg')
+\end{array}
+\end{smathpar}
+That is, if $Q$ relates $\stl$ to $\stg$, and an interference allowed
+by the isolation specfication (which implicitly considers the local
+state $\stl$) takes $\stg$ to $\stg'$, the $Q$ must also relate $\stl$
+to $\stg'$.
+
+For the guarantee $G$ of the transaction to be valid, it must follow
+from the post-condition $Q$ of the body, provided that $Q$ is stable
+w.r.t the commit-time interference captured by $R_c$. $R_c$, like
+$R_l$, is computed as rely relation modulo isolation, except that
+commit-time isolation ($\I\,\C{false}$) is considered. The validity of
+$G$ is captured by the following implication:
+\begin{smathpar}
+\begin{array}{c}
+  \forall \stl,\stg.~ Q(\stl,\stg) \Rightarrow G(\stg, \stl \gg \stg)\spc
+\end{array}
+\end{smathpar}
+In other words, if $Q$ relates the transaction-local database state
+($\stl$) to the state of the global database ($\stg$) before the
+commit, then $G$ must relate the states of the global database before
+and after the commit. The act of commit captured by the flush
+operation ($\stl\gg\stg$). Once we establish the validity of $G$ as a
+faithful representative of the transaction, we can verify that the
+transaction preserves the high-level invariant $I$ by checking the
+stability of $I$ w.r.t $G$, i.e., $\forall \stg,\stg'.~I(\stg) \wedge
+G(\stg,\stg') \Rightarrow I(\stg')$.
+
+A characteristic of RG reasoning is that stability of an assertion is
+always proven w.r.t to $R$, and not $R^{*}$, although interference
+may include multiple environment steps, and $R$ only captures a single
+step. This is nonetheless sound due to the the inductive reasoning: if
+$Q$ is preserved by every step of $R$, then $Q$ is preserved by
+$R^{*}$, and vice-versa.  However, such reasoning does
+extend naturally to isolation-constrained interference ($\R$)
+because $R^{*}$ modulo $\I$ is not same as $\R^{*}$; the former is a
+transitive relation constrained by $\I$, whereas the latter is the
+transitive closure of a $\I$-constrained relation. We therefore
+introduce a side-condition on $\I$ that restores the equality. The
+condition requires $\I$ to allow an intereference $R^{*}(\stg,\stg'')$,
+for two database states $\stg$ and $\stg''$, only if it also allows
+interference for every prefix of $R^{*}(\stg,\stg'')$. In other words,
+if $\I$ disallows interference from $\stg$ to $\stg'$, then an
+$R$-step from $\stg'$ to $\stg''$ shouldn't make the interference from
+$\stg$ to $\stg''$ valid. We call this the stability condition on
+$\I$, defined as below:
+\begin{smathpar}
+\begin{array}{lcl}
+  \stable(R,\I) & \Leftrightarrow & \forall \stl,\stg,\stg',\stg''.~
+  \neg\I(\stl,\stg,\stg') \conj R(\stg',\stg'') \Rightarrow
+  \neg\I(\stl,\stg,\stg'')
+\end{array}
+\end{smathpar}
+It can be easily verified that the above stability condition is
+satisfied by the isolation axioms from Sec.~\ref{sec:isolation}. For
+instance, $\I_{ss}$, the snapshot axiom, is stable because if is
+invalid, then an interference has already modified a record, and no
+further interference will restore the original record. This is because
+records bear transaction ids, and the transaction that wrote the
+original record has long committed.
 

### 2017-07-03 opsem.tex
RG Rules and explanation.

@@
 Certain terms that only appear at run-time are also present in $c$.
-These include a \R{TXN} block tagged with sets ($\stl$ and $\stg$) of
-records whose meaning is explained below, and a runtime \R{FOREACH}
+These include a \textsc{TXN} block tagged with sets ($\stl$ and $\stg$) of
+records whose meaning is explained below, and a runtime \textsc{FOREACH}
 expression that keeps track of the set ($s_1$) of items already

### 2017-07-03 reasoning.tex
RG Rules and explanation.

@@
 original record has long committed.
+
+The \rulelabel{RG-Conseq} rule lets us safely strengthen the guarantee
+$G$, or weaken the rely $R$ of a transaction. Importantly, it also
+allows its isolation specification $\I$ to be strengthened. This means
+that a transaction proven correct under a weaker isolation level is
+also correct under a stronger level. Parametricity over the isolation
+specification $\I$, combined with the ability to strengthen $\I$ as
+needed, admits a flexible proof strategy to prove database programs
+correct. For example, programmers can declare isolation requirements
+of their choice through $\I$, and then prove programs correct assuming
+the guarantees hold. The soundness of strengthening $\I$ ensures that
+a program can be safely executed on any system that offers isolation
+guarantees at least as strong as those assumed.
+
+Two rules of the transaction-local RG judgment are shown in
+Fig.~\ref{fig:rg-rules}. The rule \rulelabel{RG-Update} is
+illustrative of the RG rules for SQL statements; they basically
+reflect the structure of the corresponding reduction rule from
+Fig.~\ref{fig:txnimp}, and require no further explanation. The rule
+\rulelabel{RG-Foreach} defines the RG judgment for a \C{FOREACH} loop.
+As is characteristic of loops, the reasoning is pivoted on a loop
+invariant $\I$ (not same as the high-level invariant), that needs to
+be stable w.r.t $\R$. $I$ must follow $P$, the pre-condition of
+\C{FOREACH}, when no elements have been iterated, i.e, when
+$y=\emptyset$. The body of the loop can assume the loop invariant, and
+the fact that $z$ is an element from the set $x$ being iterated, to
+prove its post-condition $Q_c$. Operational semantics ensures that $z$
+is added to $y$ at the end of the iteration, hence $Q_c \conj z\in y$
+is valid at the end of the iteration, from which $I$ must follow. When
+\C{FOREACH} has finished execution, $y$, the set of iterated items, is
+the entire set $x$. Thus $I \conj y=x$ must imply the post-condition
+$Q$, which also needs to be stable. The rules for conditions,
+sequencing etc., are more-or-less standard, hence elided.
+
+% \subsection{Semantics and Soundness}
 

### 2017-07-03 reasoning.tex
Edits.

@@
 $Q$ is preserved by every step of $R$, then $Q$ is preserved by
-$R^{*}$, and vice-versa.  However, such reasoning does
+$R^{*}$, and vice-versa.  However, such reasoning does not
 extend naturally to isolation-constrained interference ($\R$)
@@
 satisfied by the isolation axioms from Sec.~\ref{sec:isolation}. For
-instance, $\I_{ss}$, the snapshot axiom, is stable because if is
-invalid, then an interference has already modified a record, and no
-further interference will restore the original record. This is because
-records bear transaction ids, and the transaction that wrote the
-original record has long committed.
+instance, $\I_{ss}$, the snapshot axiom, is stable because if
+$\I_{ss}$ is invalid, then an interference has already modified a
+record, and no further interference will restore the original record,
+because original record bears the id of a transaction that has long
+committed. Thus, $\I_{ss}$ remains invalid.
 
@@
 
-% \subsection{Semantics and Soundness}
+\subsection{Semantics and Soundness}
 
-% \begin{definition}[\bfseries Step-indexed reflexive transitive closure]
-% For all $A:\text{Type}$, $R: A \rightarrow A \rightarrow \mathbb{P}$, and $n :
-% \mathbb{N}$, the step-indexed reflexive transitive closure $R^n$ of $R$ is
-% the smallest relation satisfying the following
-% properties:
-% \begin{itemize}
-% \item $\forall (x:A).\, R^0 (x,x)$
-% \item $\forall (n:\mathbb{N})(x,y,z : A).\, R(x,y) \conj R^n(y,z) \Rightarrow
-% R^{n+1}(x,z)$
-% \end{itemize}
-% \end{definition}
+\begin{definition}[\bfseries Step-indexed reflexive transitive closure]
+For all $A:\text{Type}$, $R: A \rightarrow A \rightarrow \mathbb{P}$, and $n :
+\mathbb{N}$, the step-indexed reflexive transitive closure $R^n$ of $R$ is
+the smallest relation satisfying the following
+properties:
+\begin{itemize}
+\item $\forall (x:A).\, R^0 (x,x)$
+\item $\forall (n:\mathbb{N})(x,y,z : A).\, R(x,y) \conj R^n(y,z) \Rightarrow
+R^{n+1}(x,z)$
+\end{itemize}
+\end{definition}
 
-% \begin{definition}[\bfseries Interleaved step and multi-step relations]
-% An interleaved step relation interleaves thread-local reductions with
-% interference from concurrent threads captured as the rely relation
-% ($R$).  It is defined thus:\vspace*{-10pt}
+\begin{definition}[\bfseries Interleaved step and multi-step relations]
+An interleaved step relation interleaves thread-local reductions with
+interference from concurrent threads captured as the rely relation
+($R$).  It is defined thus:\vspace*{-10pt}
 
-% \begin{smathpar}
-% \begin{array}{lcl}
-% \I \vdash (c,\E) \rstepsto (c',\E') & \defeq & \I \vdash 
-%   (c,\E) \stepsto (c',\E') \\
-%   &   & \disj (c' = c \conj R(\E, \E') \conj \I(\E'))\\
-% \end{array}
-% \end{smathpar}
+\begin{smathpar}
+\begin{array}{lcl}
+\I \vdash (c,\E) \rstepsto (c',\E') & \defeq & \I \vdash 
+  (c,\E) \stepsto (c',\E') \\
+  &   & \disj (c' = c \conj R(\E, \E') \conj \I(\E'))\\
+\end{array}
+\end{smathpar}
 
-% \noindent The interleaved step relation for transaction bound expressions
-% ($\txnbox{e}_i$) and commands ($\txnbox{c}_i$) is defined similarly.
-% An interleaved multi-step relation ($\stepssto{n}$) is the step-indexed
-% reflexive transitive closure of the interleaved step relation.
-% \end{definition}
+\noindent The interleaved step relation for transaction bound expressions
+($\txnbox{e}_i$) and commands ($\txnbox{c}_i$) is defined similarly.
+An interleaved multi-step relation ($\stepssto{n}$) is the step-indexed
+reflexive transitive closure of the interleaved step relation.
+\end{definition}
 
-% \begin{definition}[\bfseries Semantics of the RG judgment]
-% \label{def:rg-semantics}
-% The semantics of the RG sextuple $\rg{\mathbb{I},P,R}{c}{G,Q}$ is defined
-% in terms of the interleaved step relation thus:\vspace*{-10pt}
+\begin{definition}[\bfseries Semantics of the RG judgment]
+\label{def:rg-semantics}
+The semantics of the RG sextuple $\rg{\mathbb{I},P,R}{c}{G,Q}$ is defined
+in terms of the interleaved step relation thus:\vspace*{-10pt}
 
-% \begin{smathpar}
-% \begin{array}{l}
-% \hspace*{-0.3in}
-% \rg{\mathbb{I},P,R}{c}{G,Q} \;\defeq\; \forall \E.\, P(\E)
-%   \wedge \mathbb{I}(\E) \\
-% \hspace*{0.4in}\Rightarrow (\forall n,\E'.\; \I \vdash (c,\E) 
-%     \rstepssto{n} (\cskip,\E') \Rightarrow Q(\E')) \\
-% \hspace*{0.5in}\conj \texttt{step-guaranteed}(\I,R,G,c,\E)\\
-% \end{array}
-% \end{smathpar}
+\begin{smathpar}
+\begin{array}{l}
+\hspace*{-0.3in}
+\rg{\mathbb{I},P,R}{c}{G,Q} \;\defeq\; \forall \E.\, P(\E)
+  \wedge \mathbb{I}(\E) \\
+\hspace*{0.4in}\Rightarrow (\forall n,\E'.\; \I \vdash (c,\E) 
+    \rstepssto{n} (\cskip,\E') \Rightarrow Q(\E')) \\
+\hspace*{0.5in}\conj \texttt{step-guaranteed}(\I,R,G,c,\E)\\
+\end{array}
+\end{smathpar}
 
-% \noindent The first conjunct in the consequent is called the \emph{Hoare
-% consequent} since it ascribes Hoare triple semantics to an RG sextuple.
-% The second conjunct, called the \emph{guarantee consequent}, uses the
-% $\texttt{step-guaranteed}$ predicate defined below:\vspace*{-10pt}
+\noindent The first conjunct in the consequent is called the \emph{Hoare
+consequent} since it ascribes Hoare triple semantics to an RG sextuple.
+The second conjunct, called the \emph{guarantee consequent}, uses the
+$\texttt{step-guaranteed}$ predicate defined below:\vspace*{-10pt}
 
-% \begin{smathpar}
-% \begin{array}{l}
-% \texttt{step-guaranteed}(\I,R,G,c,\E) \;\defeq\; \forall n,\E',c'',\E''.\\
-% \hspace*{0.2in}\I \vdash (c,\E) \rstepssto{n} (c',\E') \conj \I \vdash (c',\E') \stepsto
-%   (c'',\E'') \Rightarrow G(\E',\E'')\\
-% \end{array}
-% \end{smathpar}
+\begin{smathpar}
+\begin{array}{l}
+\texttt{step-guaranteed}(\I,R,G,c,\E) \;\defeq\; \forall n,\E',c'',\E''.\\
+\hspace*{0.2in}\I \vdash (c,\E) \rstepssto{n} (c',\E') \conj \I \vdash (c',\E') \stepsto
+  (c'',\E'') \Rightarrow G(\E',\E'')\\
+\end{array}
+\end{smathpar}
 
-% \noindent The guarantee consequent requires $G$ to capture the trace effect of
-% every small-step of $c$, where the reduction can be interleaved by the
-% interference ($R$) from concurrent threads. The semantics of the RG
-% sextuple for transaction-bound commands ($\txnbox{c}_i$) is defined
-% similarly. Expressions, unlike commands, evaluate to a value $v$, and
-% the semantics of their RG septuple ($\rg{\I,P,R}{\txnbox{e}_i}{G,C,Q}$) differs slightly in that its
-% Hoare consequent requires the value $v$ to satisfy the assertion $C$. 
-% \end{definition}
+\noindent The guarantee consequent requires $G$ to capture the trace effect of
+every small-step of $c$, where the reduction can be interleaved by the
+interference ($R$) from concurrent threads. The semantics of the RG
+sextuple for transaction-bound commands ($\txnbox{c}_i$) is defined
+similarly. Expressions, unlike commands, evaluate to a value $v$, and
+the semantics of their RG septuple ($\rg{\I,P,R}{\txnbox{e}_i}{G,C,Q}$) differs slightly in that its
+Hoare consequent requires the value $v$ to satisfy the assertion $C$. 
+\end{definition}
 
-% Note that the semantics of all RG judgments, including the judgments
-% for transaction-bound terms, make similar demands of the guarantee
-% relation. Given that transactions are atomic (though not isolated), it
-% is not immediately apparent why a transaction's guarantee is required
-% to make explicit every step of its reduction. This requirement is
-% justified however because, in reality, a transaction's atomicity is
-% predicated on the isolation settings of the observer. A \iso{Read
-%   Uncommitted} transaction, for example, is permitted to observe the
-% internal state of a transaction $T$ even if $T$ is claimed to execute
-% atomically.  In the interest of modular verification, the transaction
-% must therefore make its internal state available via its guarantee
-% relation.
+Note that the semantics of all RG judgments, including the judgments
+for transaction-bound terms, make similar demands of the guarantee
+relation. Given that transactions are atomic (though not isolated), it
+is not immediately apparent why a transaction's guarantee is required
+to make explicit every step of its reduction. This requirement is
+justified however because, in reality, a transaction's atomicity is
+predicated on the isolation settings of the observer. A \iso{Read
+  Uncommitted} transaction, for example, is permitted to observe the
+internal state of a transaction $T$ even if $T$ is claimed to execute
+atomically.  In the interest of modular verification, the transaction
+must therefore make its internal state available via its guarantee
+relation.
 
-% \begin{theorem}[\bfseries Soundness] 
-% The rely-guarantee judgments defined by the rules in
-% Fig.~\ref{fig:rg-rules} are sound with respect to the semantics of
-% Definition~\ref{def:rg-semantics}.\footnote{Formal proof of soundness
-% is provided in the supplementary material.}
-% \end{theorem}
+\begin{theorem}[\bfseries Soundness] 
+The rely-guarantee judgments defined by the rules in
+Fig.~\ref{fig:rg-rules} are sound with respect to the semantics of
+Definition~\ref{def:rg-semantics}.\footnote{Formal proof of soundness
+is provided in the supplementary material.}
+\end{theorem}
 
-% \noindent In particular, if $\rg{\I,P,R}{c}{G,Q}$ can be derived using
-% the rules of Fig.~\ref{fig:rg-rules}, then (a) every interleaved
-% multi-step reduction of $c$ starting from a trace that satisfies $P$
-% and $\I$, results in a trace that satisfies $Q$, and (b) the effect
-% that every small-step of $c$ has on the trace is contained in $G$.
-% Soundness of the RG judgment for transaction-bound commands
-% ($\txnbox{c}_i$) is stated similarly.  For expressions, soundness of
-% the judgment $\rg{\I,P,R}{\txnbox{e}_i}{G,C,Q}$ also proves that $e$
-% is always evaluated to a value that satisfies $C$.
+\noindent In particular, if $\rg{\I,P,R}{c}{G,Q}$ can be derived using
+the rules of Fig.~\ref{fig:rg-rules}, then (a) every interleaved
+multi-step reduction of $c$ starting from a trace that satisfies $P$
+and $\I$, results in a trace that satisfies $Q$, and (b) the effect
+that every small-step of $c$ has on the trace is contained in $G$.
+Soundness of the RG judgment for transaction-bound commands
+($\txnbox{c}_i$) is stated similarly.  For expressions, soundness of
+the judgment $\rg{\I,P,R}{\txnbox{e}_i}{G,C,Q}$ also proves that $e$
+is always evaluated to a value that satisfies $C$.

### 2017-07-03 reasoning.tex
EDits.

@@
 A characteristic of RG reasoning is that stability of an assertion is
-always proven w.r.t to $R$, and not $R^{*}$, although interference
-may include multiple environment steps, and $R$ only captures a single
+always proven w.r.t to $R$, and not $R^{*}$, although interference may
+include multiple environment steps, and $R$ only captures a single
 step. This is nonetheless sound due to the the inductive reasoning: if
 $Q$ is preserved by every step of $R$, then $Q$ is preserved by
-$R^{*}$, and vice-versa.  However, such reasoning does not
-extend naturally to isolation-constrained interference ($\R$)
-because $R^{*}$ modulo $\I$ is not same as $\R^{*}$; the former is a
-transitive relation constrained by $\I$, whereas the latter is the
-transitive closure of a $\I$-constrained relation. We therefore
-introduce a side-condition on $\I$ that restores the equality. The
-condition requires $\I$ to allow an intereference $R^{*}(\stg,\stg'')$,
-for two database states $\stg$ and $\stg''$, only if it also allows
-interference for every prefix of $R^{*}(\stg,\stg'')$. In other words,
-if $\I$ disallows interference from $\stg$ to $\stg'$, then an
-$R$-step from $\stg'$ to $\stg''$ shouldn't make the interference from
-$\stg$ to $\stg''$ valid. We call this the stability condition on
-$\I$, defined as below:
+$R^{*}$, and vice-versa.  However, such reasoning does not extend
+naturally to isolation-constrained interference because $R^{*}$ modulo
+$\I$ is not same as $\R^{*}$; the former is a transitive relation
+constrained by $\I$, whereas the latter is the transitive closure of a
+$\I$-constrained relation. We therefore introduce a side-condition on
+$\I$ that restores the equality. The condition requires $\I$ to allow
+an intereference $R^{*}(\stg,\stg'')$, for two database states $\stg$
+and $\stg''$, only if it also allows interference for every prefix of
+$R^{*}(\stg,\stg'')$. In other words, if $\I$ disallows interference
+from $\stg$ to $\stg'$, then an $R$-step from $\stg'$ to $\stg''$
+shouldn't make the interference from $\stg$ to $\stg''$ valid. We call
+this the stability condition on $\I$, defined as below:
 \begin{smathpar}

### 2017-07-03 reasoning.tex
Edits.

@@
 the entire set $x$. Thus $I \conj y=x$ must imply the post-condition
-$Q$, which also needs to be stable. The rules for conditions,
+$Q$, which also needs to be stable. The rules for conditionals,
 sequencing etc., are more-or-less standard, hence elided.
-
-% \subsection{Semantics and Soundness}
-
-% The rules that define rely-guarantee judgment are shown in
-% Fig.~\ref{fig:rg-rules}. Like standard rely-guarantee definitions,
-% these definitions also require a \emph{stability} condition, which
-% requires pre- and post- conditions to hold despite any interference
-% from concurrent threads (captured by $R$). Stability can be predicated
-% on the assumption that interference preserves the trace invariant
-% $\I$. Formally:\vspace*{-10pt}
-
-% \begin{smathpar}
-% \begin{array}{lcl}
-%   \underI{\stable(R,P)} & \defeq & \forall \E, \E'.\, 
-%   \I(\E) \conj P(\E) \conj R(\E,\E') \\
-%   &   & \hspace*{1in}\conj \I(\E') \Rightarrow P(\E')\\
-% \end{array}
-% \end{smathpar}
-
-% \noindent However, the assumption that interference preserves $\I$ (or,
-% dually, $\I$ withstands interference) needs to be justified
-% separately. We call this the stability requirement on $\I$:\vspace*{-10pt}
-
-% \begin{smathpar}
-% \begin{array}{lcl}
-% \stable(R,\I)& \defeq & \forall \E, \E'.\, 
-%   \I(\E) \conj R(\E,\E')\Rightarrow \I(\E')\\
-% \end{array}
-% \end{smathpar}
-
-% \noindent The rule \rulelabel{RG-Var} defines the rely-guarantee judgment for
-% shared variable reads inside a transaction $T_i$. It requires $\I$ to
-% be stable, and pre- and post- conditions to be stable relative to
-% $\I$.  The quantified premise effectively requires a proof that if the
-% abstract machine of Fig.~\ref{fig:txnimp} takes a step starting from
-% an execution $\E$ that satisfies the pre-condition $P$, then the
-% resultant execution $\E'$ satisfies the post-condition $Q$, and that
-% the guarantee $G$ faithfully captures the transition from $\E$ to
-% $\E'$. It is informative to compare this premise with the premise of
-% the \rulelabel{E-Aux} reduction rule of Fig.~\ref{fig:txnimp}. Similar
-% premises also appear in the \rulelabel{RG-Asgn} and \rulelabel{RG-Txn}
-% rules, which define rely-guarantee judgments for assignments and
-% transactions, respectively. \rulelabel{RG-Var} however also requires
-% the return value ($\interp{S}(X)$) of the read to satisfy the
-% assertion $C$ meant for the value. \rulelabel{RG-Arith} defines the RG
-% judgment for an arithmetic expression $e_1\pm e_2$ in terms of the
-% corresponding judgments for the constituent expressions $e_1$ and
-% $e_2$. The quantified premise requires any value resulting from
-% evaluating $e_1 \pm e_2$ to satisfy the assertion $C$, provided that
-% $e_1$ and $e_2$ always evaluate to values that satisfy $C_1$ and
-% $C_2$, respectively. The rules for sequential and parallel composition
-% of commands are essentially the same as their counterparts in a
-% standard rely-guarantee formulation and hence elided.
-
-% % RG judgment for the assignment $X:=e$ makes use of the corresponding
-% % judgment for the RHS expression $e$. The quantified premise asserts
-% % that evaluating the assigment after evaluating $e$ to a value $v$ and
-% % an execution $\E$, should result in an execution $\E'$ that satisfies
-% % the post-condition $Q$, while being related to $\E$ via the guarantee
-% % $G$. RG judgment for the \C{txn} lexical block is similar.  It uses
-% % the judgment for the transaction-bound command $c$ (i.e.,
-% % $\txnbox{c}_i$) to obtain an invariant $Q'$ for the execution $\E$
-% % before the commit, and verifies that committing the transaction under
-% % $\E$ results in an execution $\E'$ that satisfies transaction's
-% % post-condition $Q$. As usual, $\E$ and $\E'$ need to be related by $G$.
-% % The rules for sequential composition of transaction-bound comands
-% % (\rulelabel{RG-Seq}) and parallel composition of top-level commands
-% % (\rulelabel{RG-Par}) are straightforward, and more-or-less same as the
-% % corresponding rules in classical rely-guarantee.
-
-% The \rulelabel{RG-Conseq} rule defines ways to strengthen or
-% weaken relations and assertions associated with the RG judgment of
-% transaction-bound expressions. Similar rules exist for
-% transaction-bound and top-level commands, but are not
-% shown.\footnote{The supplementary materials provide the complete set
-%   of rules.} As is the case with a standard rely-guarantee
-% formulation,the rules allow the pre-condition $P$ and the rely relation
-% $R$ to be strengthened, and the post-condition $Q$ (also, $C$ in the
-% case of expressions) and the guarantee relation $G$ to be weakened.
-% The most notable aspect of the \rulelabel{RG-Conseq} rules is that
-% they allow the trace invariant $\I$ to be strengthened. Considering
-% that $\I$ captures isolation properties, this means that a program
-% proven correct under weaker isolation levels is also correct under
-% stronger ones.  Parametricity over the trace invariant $\I$, combined
-% with the ability to strengthen $\I$ as needed, allows our proof system
-% to support a highly flexible proof strategy to prove programs correct
-% over various isolation variants. For example, programmers can
-% \emph{define} isolation guarantees of their choice (by defining $\I$
-% appropriately) and then prove programs correct assuming the guarantees
-% hold.  The soundness of strengthening $\I$ ensures that a program can
-% be safely executed on any system that offers isolation guarantees
-% at least as strong as those assumed.
 
@@
 
+% \begin{definition}[\bfseries Interleaved step and multi-step relations]
+% An interleaved step relation interleaves transaction-local reductions with
+% legal interference from concurrent transactions as defined by the
+% $\I$-constrained rely relation ($\R$).  It is defined thus:
+% \begin{smathpar}
+% \begin{array}{lcl}
+% (c,\stl,\stg) \rstepsto (c',\stl',\stg') & \defeq & \stg \vdash 
+%   (c,\stl) \stepsto (c',\stl') \conj \stg'=\stg \\
+%   &   & \disj (c' = c \conj \stl'=\stl \conj \R(\stl, \stg, \stg'))\\
+% \end{array}
+% \end{smathpar}
+
 \begin{definition}[\bfseries Interleaved step and multi-step relations]
-An interleaved step relation interleaves thread-local reductions with
-interference from concurrent threads captured as the rely relation
-($R$).  It is defined thus:\vspace*{-10pt}
-
+An interleaved step relation interleaves a transaction's reduction with
+interference from concurrent transactions as captured by the rely
+relation. It is defined thus:
 \begin{smathpar}
 \begin{array}{lcl}
-\I \vdash (c,\E) \rstepsto (c',\E') & \defeq & \I \vdash 
-  (c,\E) \stepsto (c',\E') \\
-  &   & \disj (c' = c \conj R(\E, \E') \conj \I(\E'))\\
+(c,\stg) \rstepsto (c',\stg') & \defeq &  
+  (c,\stg) \stepsto (c',\stg') \disj (c' = c \conj R(\stg, \stg'))\\
 \end{array}

### 2017-07-03 reasoning.tex
Soundness result added.

@@
 As is characteristic of loops, the reasoning is pivoted on a loop
-invariant $\I$ (not same as the high-level invariant), that needs to
+invariant $I$ (not same as the high-level invariant), that needs to
 be stable w.r.t $\R$. $I$ must follow $P$, the pre-condition of
@@
 
+We now formalize the semantics of the RG judgments defined in
+Fig.~\ref{fig:rg-rules}, and state their soundness guarantees.
+
 \begin{definition}[\bfseries Step-indexed reflexive transitive closure]
@@
 
-% \begin{definition}[\bfseries Interleaved step and multi-step relations]
-% An interleaved step relation interleaves transaction-local reductions with
-% legal interference from concurrent transactions as defined by the
-% $\I$-constrained rely relation ($\R$).  It is defined thus:
-% \begin{smathpar}
-% \begin{array}{lcl}
-% (c,\stl,\stg) \rstepsto (c',\stl',\stg') & \defeq & \stg \vdash 
-%   (c,\stl) \stepsto (c',\stl') \conj \stg'=\stg \\
-%   &   & \disj (c' = c \conj \stl'=\stl \conj \R(\stl, \stg, \stg'))\\
-% \end{array}
-% \end{smathpar}
-
 \begin{definition}[\bfseries Interleaved step and multi-step relations]
-An interleaved step relation interleaves a transaction's reduction with
-interference from concurrent transactions as captured by the rely
-relation. It is defined thus:
+Interleaved step relations interleave global and transaction-local
+reductions with interference as captured by the corresponding rely
+relations. They are defined thus:
 \begin{smathpar}
-\begin{array}{lcl}
+\begin{array}{lclr}
 (c,\stg) \rstepsto (c',\stg') & \defeq &  
-  (c,\stg) \stepsto (c',\stg') \disj (c' = c \conj R(\stg, \stg'))\\
+  (c,\stg) \stepsto (c',\stg') \disj (c' = c \conj R(\stg, \stg'))&
+  \texttt{global}\\
+(c,\stl,\stg) \rstepsto (c',\stl',\stg') & \defeq & \stg \vdash 
+  (c,\stl) \stepsto (c',\stl') \conj \stg'=\stg& \texttt{transaction-local}\\
+  &   & \disj (c' = c \conj \stl'=\stl \conj \R(\stl, \stg, \stg'))
 \end{array}
@@
 
-\noindent The interleaved step relation for transaction bound expressions
-($\txnbox{e}_i$) and commands ($\txnbox{c}_i$) is defined similarly.
-An interleaved multi-step relation ($\stepssto{n}$) is the step-indexed
-reflexive transitive closure of the interleaved step relation.
+\noindent An interleaved multi-step relation ($\stepssto{n}$) is the
+step-indexed reflexive transitive closure of the interleaved step
+relation.  
 \end{definition}
 
-\begin{definition}[\bfseries Semantics of the RG judgment]
+\begin{definition}[\bfseries Semantics of RG judgments]
 \label{def:rg-semantics}
-The semantics of the RG sextuple $\rg{\mathbb{I},P,R}{c}{G,Q}$ is defined
-in terms of the interleaved step relation thus:\vspace*{-10pt}
-
+The semantics of the global and transaction-local RG judgments are
+defined thus:
 \begin{smathpar}
-\begin{array}{l}
-\hspace*{-0.3in}
-\rg{\mathbb{I},P,R}{c}{G,Q} \;\defeq\; \forall \E.\, P(\E)
-  \wedge \mathbb{I}(\E) \\
-\hspace*{0.4in}\Rightarrow (\forall n,\E'.\; \I \vdash (c,\E) 
-    \rstepssto{n} (\cskip,\E') \Rightarrow Q(\E')) \\
-\hspace*{0.5in}\conj \texttt{step-guaranteed}(\I,R,G,c,\E)\\
+\begin{array}{lcl}
+\R \vdash \hoare{P}{c}{Q} & \defeq & \forall \stl,\stl',\stg,\stg'.~
+  P(\stl,\stg) \conj (c,\stl,\stg) \rstepssto{n} (\cskip, \stl',\stg')
+  \Rightarrow Q(\stl',\stg')\\
+\rg{I,R}{c}{G,I} & \defeq &  \forall \stg.\, I(\stg)
+  \Rightarrow (\forall n,\stg'.\; (c,\stg) 
+    \rstepssto{n} (\cskip,\stg') \Rightarrow I(\stg')) \\
+&&\hspace*{0.5in}\conj \texttt{step-guaranteed}(R,G,c,\stg)\\
 \end{array}
@@
 
-\noindent The first conjunct in the consequent is called the \emph{Hoare
-consequent} since it ascribes Hoare triple semantics to an RG sextuple.
-The second conjunct, called the \emph{guarantee consequent}, uses the
-$\texttt{step-guaranteed}$ predicate defined below:\vspace*{-10pt}
+\noindent The
+$\texttt{step-guaranteed}$ predicate used in the semantics of the
+global RG judgment is defined below:\vspace*{-10pt}
 
 \begin{smathpar}
-\begin{array}{l}
-\texttt{step-guaranteed}(\I,R,G,c,\E) \;\defeq\; \forall n,\E',c'',\E''.\\
-\hspace*{0.2in}\I \vdash (c,\E) \rstepssto{n} (c',\E') \conj \I \vdash (c',\E') \stepsto
-  (c'',\E'') \Rightarrow G(\E',\E'')\\
+\begin{array}{lcl}
+\texttt{step-guaranteed}(R,G,c,\stg) &\defeq& \forall n,\stg',c'',\stg''.
+(c,\stg) \rstepssto{n} (c',\stg') \conj (c',\stg') \stepsto
+  (c'',\stg'') \Rightarrow G(\stg',\stg'')\\
 \end{array}
 \end{smathpar}
-
-\noindent The guarantee consequent requires $G$ to capture the trace effect of
-every small-step of $c$, where the reduction can be interleaved by the
-interference ($R$) from concurrent threads. The semantics of the RG
-sextuple for transaction-bound commands ($\txnbox{c}_i$) is defined
-similarly. Expressions, unlike commands, evaluate to a value $v$, and
-the semantics of their RG septuple ($\rg{\I,P,R}{\txnbox{e}_i}{G,C,Q}$) differs slightly in that its
-Hoare consequent requires the value $v$ to satisfy the assertion $C$. 
 \end{definition}
 
-Note that the semantics of all RG judgments, including the judgments
-for transaction-bound terms, make similar demands of the guarantee
-relation. Given that transactions are atomic (though not isolated), it
-is not immediately apparent why a transaction's guarantee is required
-to make explicit every step of its reduction. This requirement is
-justified however because, in reality, a transaction's atomicity is
-predicated on the isolation settings of the observer. A \iso{Read
-  Uncommitted} transaction, for example, is permitted to observe the
-internal state of a transaction $T$ even if $T$ is claimed to execute
-atomically.  In the interest of modular verification, the transaction
-must therefore make its internal state available via its guarantee
-relation.
+We can now assert the soundness of the RG judgments in
+Fig.~\ref{fig:rg-rules} as following\footnote{Formal proof of soundness
+is provided in the supplementary material.}:
 
@@
 Fig.~\ref{fig:rg-rules} are sound with respect to the semantics of
-Definition~\ref{def:rg-semantics}.\footnote{Formal proof of soundness
-is provided in the supplementary material.}
+Definition~\ref{def:rg-semantics}.
 \end{theorem}
 
-\noindent In particular, if $\rg{\I,P,R}{c}{G,Q}$ can be derived using
-the rules of Fig.~\ref{fig:rg-rules}, then (a) every interleaved
-multi-step reduction of $c$ starting from a trace that satisfies $P$
-and $\I$, results in a trace that satisfies $Q$, and (b) the effect
-that every small-step of $c$ has on the trace is contained in $G$.
-Soundness of the RG judgment for transaction-bound commands
-($\txnbox{c}_i$) is stated similarly.  For expressions, soundness of
-the judgment $\rg{\I,P,R}{\txnbox{e}_i}{G,C,Q}$ also proves that $e$
-is always evaluated to a value that satisfies $C$.
+\noindent Thus, if $\rg{I,R}{c}{G,I}$ is a valid RG judgment, then (a)
+every interleaved multi-step reduction of $c$ preserves the database
+integrity constraint $I$, and (b) the effect that every transaction in
+$c$ has on the database state is captured by $G$.

### 2017-07-04 reasoning.tex
Weak isolation example added.

@@
 \label{sec:reasoning}
+
+\begin{figure}
+\centering
+\begin{tabular}{l|l}
+\begin{txnimpcode}
+$\begin{decoration}
+  \{X = k\}
+\end{decoration}$
+  atomic {
+    v1 := 0; v2 := 1;
+    while(v1<X) {
+      v2 := v2 * (X-v1);
+      v1 := v1+1;
+    }
+    X := v2; Y := v2;
+  }
+$\begin{decoration}
+  \{X = k! \conj Y = X\}
+\end{decoration}$
+\end{txnimpcode}
+&
+\begin{txnimpcode}
+$\begin{decoration}
+  \{X = k\}
+\end{decoration}$
+  transaction {
+    v1 := 0; v2 := 1;
+    while(v1<X) {
+      v2 := v2 * (X-v1);
+      v2 := v2 * v1;
+      v1 := v1+1;
+    }
+    X := v2; Y := v2;
+  }
+$\begin{decoration}
+  \{Y = X\}
+\end{decoration}$
+\end{txnimpcode}
+\end{tabular}
+  \caption{Factorial program written as an atomic block and a weakly
+  isolated transaction (shared variables have uppercase names).
+  Effect of multiple iteratons of the \C{atomic} loop can be
+  summarized as $k!$, while same cannot be done with the
+  \C{transaction} loop, where weak isolation allows reads in different
+  iterations to return different values for $X$.  Nonetheless,
+  post-state satisfies $X=Y$ in both cases due to atomicity of
+  writes.}
+\label{fig:atomic-vs-transaction}
+\end{figure}
 
@@
 Fig.~\ref{fig:atomic-vs-transaction}, where a series of operations on
-\C{x} can be composed to describe the effect of an \C{atomic} block,
+\C{X} can be composed to describe the effect of an \C{atomic} block,
 while same cannot be done in a transaction. Thus, weak isolation needs

### 2017-07-04 reasoning.tex
Edits.

@@
   isolated transaction (shared variables have uppercase names).
-  Effect of multiple iteratons of the \C{atomic} loop can be
+  Effect of multiple iterations of the \C{atomic} loop can be
   summarized as $k!$, while same cannot be done with the
@@
 
-We now describe a proof system that lets us prove the correctness of a \txnimp
-program $c$ w.r.t its high-level invariant $I$, on a machine that
-satisfies the isolation specifications ($\I$) of its
+We now describe a proof system that lets us prove the correctness of a
+\txnimp program $c$ w.r.t its high-level invariant $I$, on a machine
+that satisfies the isolation specifications ($\I$) of its
 transactions\footnote{Note the difference between $I$ and $\I$. The
@@
 operational characteristics of the underlying machine.}.  Our proof
-system is essentially an adapatation of the rely-guarantee reasoning
+system is essentially an adaptation of the rely-guarantee reasoning
 framework to the setting of weakly isolated database transactions. It
@@
 relate a transaction's isolation specification ($\I$) to its rely
-relation ($R$) that describes the environment, so that the
-interference is considered only insofar as the isolation level allows
-it? Our proof system shows how to compute a rely relation ($R$) modulo
-the isolation specification ($\I$) of a transaction. Another
-characteristic of the transaction setting is the atomicity of a
-trasaction's aggregate writes; a transaction's writes are not visible
-to its peers until it commits. In the context of rely-guarantee, this
-means that the transactions guarantee ($G$) should capture the
-aggregate effect of a transaction, and not its individual writes.
-While \C{atomic} blocks are also present in the shared-memory
-programs, the fact that a transactions are weakly isolated introduces
-complexity.  Unlike an \C{atomic} block, the effect of a transaction
-is \emph{not} a sequential composition of the effects of its
-statements because each statement witnesses a potentially different
-version of the state. This is illustrated in
+relation ($R$) that describes the transaction's environment, so that
+the interference is considered only insofar as the isolation level
+allows it? Our proof system shows how to compute a rely relation ($R$)
+modulo the isolation specification ($\I$) of a transaction. Another
+characteristic of the transaction setting is atomicity; a
+transaction's writes are not visible to its peers until it commits. In
+the context of rely-guarantee, this means that the transactions
+guarantee ($G$) should capture the aggregate effect of a transaction,
+and not its individual writes.  While shared memory \C{atomic} blocks
+also have the same characteristic, the fact that a transactions are
+weakly isolated introduces complexity.  Unlike an \C{atomic} block,
+the effect of a transaction is \emph{not} a sequential composition of
+the effects of its statements because each statement witnesses a
+potentially different version of the state. This is illustrated in
 Fig.~\ref{fig:atomic-vs-transaction}, where a series of operations on
-\C{X} can be composed to describe the effect of an \C{atomic} block,
-while same cannot be done in a transaction. Thus, weak isolation needs
-to be taken into account when defining and verifying guarantees, and
-our proof system shows how.
+a shared variable (\C{X}) can be composed to describe the effect of an
+\C{atomic} block, while same cannot be done in a transaction. Thus,
+weak isolation needs to be taken into account when defining and
+verifying guarantees, and our proof system shows how.
 
@@
 database's consistency conditions\footnote{The terms \emph{consistency
-condition}, \emph{high-level invaraint}, and \emph{integrity
+condition}, \emph{high-level invariant}, and \emph{integrity
 constraint} all mean the same.}. Transaction-local RG judgment doesn't
@@
 stability condition, which, intuitively, requires the validity of an
-assertion $\phi$ to be unaffacted by the interference, i.e., the rely
+assertion $\phi$ to be unaffected by the interference, i.e., the rely
 relation $R$. In conventional RG, stability is defined as following
@@
 That is, if $Q$ relates $\stl$ to $\stg$, and an interference allowed
-by the isolation specfication (which implicitly considers the local
+by the isolation specification (which implicitly considers the local
 state $\stl$) takes $\stg$ to $\stg'$, the $Q$ must also relate $\stl$
@@
 $\I$ that restores the equality. The condition requires $\I$ to allow
-an intereference $R^{*}(\stg,\stg'')$, for two database states $\stg$
+an interference $R^{*}(\stg,\stg'')$, for two database states $\stg$
 and $\stg''$, only if it also allows interference for every prefix of

### 2017-07-04 opsem.tex
Edits.

--- 2017-07-03 opsem.tex
@@
 Certain terms that only appear at run-time are also present in $c$.
-These include a \R{TXN} block tagged with sets ($\stl$ and $\stg$) of
-records whose meaning is explained below, and a runtime \R{FOREACH}
+These include a {\sf txn} block tagged with sets ($\stl$ and $\stg$) of
+records whose meaning is explained below, and a runtime {\sf foreach}
 expression that keeps track of the set ($s_1$) of items already
@@
 transactions, and their parallel composition. A transaction always
-comes tagged with an isolation specification $\I_e$ or $\I_c$ that
-dictates the timing and nature of interferences that a transaction can
-witness, both during its execution ($\I_e$), and when it is about to
-commit ($\I_c$).  Formally, $\I_e$ and $\I_c$ are functions that map
-the (current) transaction-local database state ($\stl$), the state
-($\stg$) of the global database when the transaction last took a step,
-and the current state ($\stg'$) of the global database.  Intuitively,
-$\stg'\neq\stg$ indicates an interference from another concurrent
-transaction, and the predicates $\I_e$ and $\I_c$ decide if this
-interference is allowed or not, taking into account the local database
-state ($\stl$). For instance, as described in
-(Sec.~\ref{sec:motivation}), an RR transaction on PostgreSQL defines
-$\I$ as following:
+comes tagged with an isolation specification $\I$, which has two
+components $\I_e$ and $\I_c$, the dictate the timing and nature of
+interferences that the transaction can witness, during its execution
+($\I_e$), and when it is about to commit ($\I_c$).  Formally, $\I_e$
+and $\I_c$ are predicates over by the (current) transaction-local
+database state ($\stl$), the state ($\stg$) of the global database
+when the transaction last took a step, and the current state ($\stg'$)
+of the global database.  Intuitively, $\stg'\neq\stg$ indicates an
+interference from another concurrent transaction, and the predicates
+$\I_e$ and $\I_c$ decide if this interference is allowed or not,
+taking into account the local database state ($\stl$). For instance,
+as described in (Sec.~\ref{sec:motivation}), an RR transaction on
+PostgreSQL defines $\I$ as following:
 \begin{smathpar}
@@
 \end{smathpar}
+For brevity in presentation, we use $\I$ instead of $\I_e$ and $\I_c$
+whenever the discussion applies to both $\I_e$ and $\I_c$.
 
-The reduction of a $\ctxn{i}{\I_e}{c}$ begins by first converting it
-to its run-time form $\ctxn{i}{\I_e,\stl,\stg}{c}$, where $\stl =
+The reduction of a $\ctxn{i}{\I}{c}$ begins by first converting it to
+its run-time form $\ctxnr{i}{\I,\stl,\stg}{c}$, where $\stl =
 \emptyset$, and $\stg$ the current (global) database.  Rule
-\rulelabel{E-Txn} reduces $\ctxn{i}{\I_e,\stl,\stg}{c}$ under a
-database state ($\stg'$), only if the isolation specification ($\I$)
-allows the interference between $\stg$ and $\stg'$ during that
-step. Rule \rulelabel{E-Commit} commits the transaction
-$\ctxn{i}{\I_c,\stl,\stg}{c}$ by flushing its uncommitted records to
+\rulelabel{E-Txn} reduces $\ctxnr{i}{\I,\stl,\stg}{c}$ under a
+database state ($\stg'$), only if the execution-time isolation
+specification ($\I_e$) allows the interference between $\stg$ and
+$\stg'$. Rule \rulelabel{E-Commit} commits the transaction
+$\ctxnr{i}{\I,\stl,\stg}{c}$ by flushing its uncommitted records to
 the database. This is done only if the interference between $\stg$ and
-$\stg'$ is allowed at the commit point by the isolation specification.
-The distinction between $\I_e$ and $\I_c$ allows us to model realistic
-database features such as snapshot isolation that isolates the effects
-a transaction can witness during its execution but exposes
-interferences from concurrent transactions at the commit point.
+$\stg'$ is allowed at the commit point by the isolation specification
+($\I_c$).  The distinction between $\I_e$ and $\I_c$ allows us to
+model snapshot semantics of realistic isolation levels that isolate a
+transaction from interference during its execution, but exposes
+interference at the commit point.
 

### 2017-07-04 reasoning.tex
Edits.

@@
 ($\R_l$) for the local judgment is a ternary relation computed as $R$
-modulo $\I$ as following:
+modulo $\I_e$ as following:
 \begin{smathpar}
@@
 \R_l(\stl,\stg,\stg') & \Leftrightarrow & R(\stg,\stg') \conj
-\I\,\C{true}\,(\stl,\stg,\stg')
+\I_e(\stl,\stg,\stg')
 \end{array}
@@
 Thus, $\R_l$ allows an interference only if it doesn't violate the
-execution-time isolation specification of the transaction. If a
+execution-time isolation specification ($\I_e$) of the transaction. If a
 certain interference violates the isolation spec, i.e., $R(\stg,\stg')
-\Rightarrow \neg(\I\,\C{true}\,(\stl,\stg,\stg'))$, then
-$\R_l(\stl,\stg,\stg') \Leftrightarrow false$, and any assertion is
-trivially stable w.r.t that interference. This is sensible considering
-such interference is pre-empted in the operational semantics.
+\Rightarrow \neg(\I_e(\stl,\stg,\stg'))$, then $\R_l(\stl,\stg,\stg')
+\Leftrightarrow false$, and any assertion is trivially stable w.r.t
+that interference. This is sensible considering such interference is
+pre-empted in the operational semantics.
 
@@
 $R_l$, is computed as rely relation modulo isolation, except that
-commit-time isolation ($\I\,\C{false}$) is considered. The validity of
+commit-time isolation ($\I_c$) is considered. The validity of
 $G$ is captured by the following implication:
@@
 
-The \rulelabel{RG-Conseq} rule lets us safely strengthen the guarantee
-$G$, or weaken the rely $R$ of a transaction. Importantly, it also
-allows its isolation specification $\I$ to be strengthened. This means
-that a transaction proven correct under a weaker isolation level is
-also correct under a stronger level. Parametricity over the isolation
-specification $\I$, combined with the ability to strengthen $\I$ as
-needed, admits a flexible proof strategy to prove database programs
-correct. For example, programmers can declare isolation requirements
-of their choice through $\I$, and then prove programs correct assuming
-the guarantees hold. The soundness of strengthening $\I$ ensures that
-a program can be safely executed on any system that offers isolation
-guarantees at least as strong as those assumed.
+The \rulelabel{RG-Conseq} rule lets us safely weaken the guarantee
+$G$, and strengthen the rely $R$ of a transaction. Importantly, it
+also allows its isolation specification $\I$ to be strengthened (both
+$\I_e$ and $\I_c$). This means that a transaction proven correct under
+a weaker isolation level is also correct under a stronger level.
+Parametricity over the isolation specification $\I$, combined with the
+ability to strengthen $\I$ as needed, admits a flexible proof strategy
+to prove database programs correct. For example, programmers can
+declare isolation requirements of their choice through $\I$, and then
+prove programs correct assuming the guarantees hold. The soundness of
+strengthening $\I$ ensures that a program can be safely executed on
+any system that offers isolation guarantees at least as strong as
+those assumed.
 
@@
 As is characteristic of loops, the reasoning is pivoted on a loop
-invariant $I$ (not same as the high-level invariant), that needs to
-be stable w.r.t $\R$. $I$ must follow $P$, the pre-condition of
+invariant $I$ (not same as the high-level invariant), that needs to be
+stable w.r.t $\R$. $I$ must follow $P$, the pre-condition of
 \C{FOREACH}, when no elements have been iterated, i.e, when

### 2017-07-04 reasoning.tex
Edits.

@@
 
-\begin{figure}
-\centering
-\begin{tabular}{l|l}
-\begin{txnimpcode}
-$\begin{decoration}
-  \{X = k\}
-\end{decoration}$
-  atomic {
-    v1 := 0; v2 := 1;
-    while(v1<X) {
-      v2 := v2 * (X-v1);
-      v1 := v1+1;
-    }
-    X := v2; Y := v2;
-  }
-$\begin{decoration}
-  \{X = k! \conj Y = X\}
-\end{decoration}$
-\end{txnimpcode}
-&
-\begin{txnimpcode}
-$\begin{decoration}
-  \{X = k\}
-\end{decoration}$
-  transaction {
-    v1 := 0; v2 := 1;
-    while(v1<X) {
-      v2 := v2 * (X-v1);
-      v2 := v2 * v1;
-      v1 := v1+1;
-    }
-    X := v2; Y := v2;
-  }
-$\begin{decoration}
-  \{Y = X\}
-\end{decoration}$
-\end{txnimpcode}
-\end{tabular}
-  \caption{Factorial program written as an atomic block and a weakly
-  isolated transaction (shared variables have uppercase names).
-  Effect of multiple iterations of the \C{atomic} loop can be
-  summarized as $k!$, while same cannot be done with the
-  \C{transaction} loop, where weak isolation allows reads in different
-  iterations to return different values for $X$.  Nonetheless,
-  post-state satisfies $X=Y$ in both cases due to atomicity of
-  writes.}
-\label{fig:atomic-vs-transaction}
-\end{figure}
+% \begin{figure}
+% \centering
+% \begin{tabular}{l|l}
+% \begin{txnimpcode}
+% $\begin{decoration}
+%   \{X = k\}
+% \end{decoration}$
+%   atomic {
+%     v1 := 0; v2 := 1;
+%     while(v1<X) {
+%       v2 := v2 * (X-v1);
+%       v1 := v1+1;
+%     }
+%     X := v2; Y := v2;
+%   }
+% $\begin{decoration}
+%   \{X = k! \conj Y = X\}
+% \end{decoration}$
+% \end{txnimpcode}
+% &
+% \begin{txnimpcode}
+% $\begin{decoration}
+%   \{X = k\}
+% \end{decoration}$
+%   transaction {
+%     v1 := 0; v2 := 1;
+%     while(v1<X) {
+%       v2 := v2 * (X-v1);
+%       v2 := v2 * v1;
+%       v1 := v1+1;
+%     }
+%     X := v2; Y := v2;
+%   }
+% $\begin{decoration}
+%   \{Y = X\}
+% \end{decoration}$
+% \end{txnimpcode}
+% \end{tabular}
+%   \caption{Factorial program written as an atomic block and a weakly
+%   isolated transaction (shared variables have uppercase names).
+%   Effect of multiple iterations of the \C{atomic} loop can be
+%   summarized as $k!$, while same cannot be done with the
+%   \C{transaction} loop, where weak isolation allows reads in different
+%   iterations to return different values for $X$.  Nonetheless,
+%   post-state satisfies $X=Y$ in both cases due to atomicity of
+%   writes.}
+% \label{fig:atomic-vs-transaction}
+% \end{figure}
 
@@
 the effects of its statements because each statement witnesses a
-potentially different version of the state. This is illustrated in
-Fig.~\ref{fig:atomic-vs-transaction}, where a series of operations on
-a shared variable (\C{X}) can be composed to describe the effect of an
-\C{atomic} block, while same cannot be done in a transaction. Thus,
-weak isolation needs to be taken into account when defining and
-verifying guarantees, and our proof system shows how.
+potentially different version of the state. For example,  a series of
+operations on a shared variable (\C{X}) can be composed to describe
+the effect of an \C{atomic} block, while same cannot be done in a
+transaction. Thus, weak isolation needs to be taken into account when
+defining and verifying guarantees, and our proof system shows how.
 

### 2017-07-04 motivation.tex
edits

@@
 set of items (\C{item\_reqs}), to be served by a warehouse under the
-district identified by \C{d\_id}.  Fig.~\ref{fig:scheme} illustrates
+district identified by \C{d\_id}.  Fig.~\ref{fig:schema} illustrates
 the relationship among these different tables.
@@
 stock (\C{s\_qty}). The quantity is updated by the transaction to
-reflect the processing of new order (if the stock quantity falls below
+reflect the processing of new orders (if the stock quantity falls below
 10, it is automatically replenished by 91).
@@
 Because the aforementioned reasoning is reasonably simple to perform
-manually, verfiying the soundess of TPC-C's consistency conditions
+manually, verifying the soundess of TPC-C's consistency conditions
 would appear to be feasible.  Serializability aids the tractability of
@@
 transaction. First, since the database doesn't change ($\Delta' =
-\Delta$) (where $\Delta'$ represents the state after the transaction
-commits), during the execution of the transaction's body, we can
-reason about \C{new\_order} as though it executed in complete isolation
-until its commit point, leading to a verification process similar to
-what would been applied when reasoning about serializability.  When
+\Delta$) during the execution of the transaction's body, we can reason
+about \C{new\_order} as though it executed in complete isolation until
+its commit point, leading to a verification process similar to what
+would been applied when reasoning about serializability.  When
 \C{new\_order} finishes execution and becomes ready to commit, a
-transition that transfers the transaction's local writes ($\delta$) to
-the unchanged database state ($\Delta$).  However, we are required to
+transition transfers the transaction's local writes ($\delta$) to the
+unchanged database state ($\Delta$).  However, we are required to
 consider the interference from concurrent transitions at this point,
@@
 transaction (with the same \C{d\_id}), then verification fails as
-described previously (Fig.~\ref{fig:new_order_execs}). Fortunately,
-sequential reasoning shows that this is impossible - RR prevents a
-concurrent \C{new\_order} transaction that modifies the same
-\C{District} record as the current transaction (concretely, since the
-record is already present in the current transaction's local log, any
-transition from $\Delta$ to $\Delta_c$ cannot change this record).
-Applying such axiomatic reasoning on \C{new\_order} allows us to prove
-that the TPC-C invariant holds when the transaction is executed under
-PostgreSQL's RR isolation.  Our proof framework generalizes this style
-of reasoning to various isolation levels on databases.
+described previously.  Fortunately, sequential reasoning shows that
+this is impossible - RR prevents a concurrent \C{new\_order}
+transaction that modifies the same \C{District} record as the current
+transaction (concretely, since the record is already present in the
+current transaction's local log, any transition from $\Delta$ to
+$\Delta_c$ cannot change this record).  Applying such axiomatic
+reasoning on \C{new\_order} allows us to prove that the TPC-C
+invariant holds when the transaction is executed under PostgreSQL's RR
+isolation.  Our proof framework generalizes this style of reasoning to
+various isolation levels on databases.
 
@@
 transformer interprets a SQL statement in the set domain, taking
-advantage of the fact that a a database is essentially a set of
+advantage of the fact that the database is essentially a set of
 tuples, and a SQL statement is a transformer over these sets.  The
@@
 Let $F_{body}(\Delta) = F_U(\Delta) \,\cup\, F_I(\Delta)$.   The
-transformer for the \C{foreach} loop can now be computed as following:
+transformer for the \C{foreach} loop can now be computed as follows:
 \begin{ocaml}

### 2017-07-04 isolation.tex
Edits.

@@
 
-Databases implement isolation levels through a combination of various
-concurrency control mechanisms, which are absent from
-Fig.~\ref{fig:txnimp}. In this section, we first axiomatize the
-semantics of various concurrency control constructs, and then define
-the precise semantics of the isolation levels as they are implemented
-on two real-world databases -  PostgreSQL and MySQL.
+A distinctive characteristic of our development is that it is
+parameterized on the isolation specification $\I$. $\I$ can be
+instantiated with the declarative characterization of an isolation
+guarantee or a concurrency control mechanism, regardless of its actual
+implementation. This allows us to model a range of isolation
+properties that are relevant to the theory and practice of database
+systems. We present some below:
+
+% Databases implement isolation levels through a combination of various
+% concurrency control mechanisms, which are absent from
+% Fig.~\ref{fig:txnimp}. In this section, we first axiomatize the
+% semantics of various concurrency control constructs, and then define
+% the precise semantics of the isolation levels as they are implemented
+% on two real-world databases -  PostgreSQL and MySQL.
 % Any additional concurrency control provided by the database
@@
 
-\textbf{Unique Ids}. Most relational databases associate every record
-in a table with a unique identifier. Often, auto-generated unique
-identifiers are part of the schema itself (via SQL's \C{UNIQUE} and
-\C{AUTO\_INCREMENT} keywords), but if not, database adds one (e.g.,
-MySQL's InnoDB engine automatically adds a 6-byte identifier if none
-exists). Enforcing global uniqueness of identifiers requires exclusive
-locks, which are absent from Fig.~\ref{fig:txnimp}. The
-\rulelabel{E-Insert} rule assigns an id to the newly added record
-after checking that it is not already present in $\stg$ and $\stl$.
-But the new record is only added to $\stl$, which is invisible outside
-the transaction. Thus, it is possible that the uniqueness check for the same id may
-also succeed in a concurrent transaction, resulting in duplicate ids
-after both commits. To prevent this, we need the following proposition
-in $\I$ of every transaction, regardless of whether it is executing
-or committing:
+\textbf{Unique Ids}. As the \C{new\_order} example
+(Sec.~\ref{sec:motivation}) demonstrates, enforcing global uniqueness
+of ordered identifiers requires stronger isolation. Alternatively,
+globally unique sequence numbers, regardless of the isolation level,
+can be from a requested from a relational databases via SQL's
+\C{UNIQUE} and \C{AUTO\_INCREMENT} keywords. Our development crucially
+relies on the uniqueness of record identifiers\footnote{This has
+parallels in real-world. For e.g., MySQL's InnoDB engine automatically
+adds a 6-byte unique identifier if none exists for a record.}, which
+are generated locally by the \rulelabel{E-Insert} rule. Global
+uniqueness of locally generated identifiers can be axiomatized as an
+isolation property as following:
 \begin{smathpar}
@@
 $\I_{id}$ ensures that a globally unique id generated during a
-transaction's execution remains globally unique until it commits, prohibiting
-interference from a concurrent transaction that adds the same id.
+transaction's execution remains globally unique until it commits,
+prohibiting interference from a concurrent transaction that adds the
+same id. The axiom thus simulates a global counter protected by an
+exclusive lock.
 
-\textbf{Write Locks}. Transactions in MySQL and PostgreSQL (effectively)
-obtain exclusive locks on records they update, and do not release them
-until they commit, thus preventing write-write (ww) conflicts. To
-obtain this behavior on the abstract machine of Fig.~\ref{fig:txnimp},
-the following proposition must hold for every transaction:
+\textbf{Write-Write Conflicts}. Databases often employ a combination of
+concurrency control methods, both optimistic (e.g., speculation and
+rollback) and pessimistic (e.g., exclusive locking), to eliminate
+write-write conflicts among concurrent transactions. The absence of
+write-write (ww) conflicts, regardless of how it is achieved, can be
+axiomatized as following:
 \begin{smathpar}
@@
 transaction from changing $r$, thus simulating the behavior of an
-exclusive lock. There is however a caveat: if we assume extensional
-equality over records, a write by a concurrent transaction that
-doesn't change $r$'s contents is still allowed. While this in itself
-is not a problem, the concurrent transaction may modify other records,
-which then become visible in the current transaction. A write lock
-prevents this behavior, whereas our axiomatization ($\I_{ww}$) allows
-it. An easy fix for this is to add version timestamps to records that
-effectively intensionalizes equality. Nonetheless, imprecise
-axiomatization of write locks hasn't been a problem in practice.
+exclusive lock or a speculative execution that succeeded (Note: a
+transaction writing to $r$ always changes $r$ because its $\txnf$
+field is updated). 
+% There is however a caveat: if we assume extensional
+% equality over records, a write by a concurrent transaction that
+% doesn't change $r$'s contents is still allowed. While this in itself
+% is not a problem, the concurrent transaction may modify other records,
+% which then become visible in the current transaction. A write lock
+% prevents this behavior, whereas our axiomatization ($\I_{ww}$) allows
+% it. An easy fix for this is to add version timestamps to records that
+% effectively intensionalizes equality. Nonetheless, imprecise
+% axiomatization of write locks hasn't been a problem in practice.
 
-\textbf{Snapshots} Both MySQL and PostgresSQL implement isolation levels
-that execute transactions against a static snapshot of the database.
-The semantics of a snapshot be axiomatized as following:
+\textbf{Snapshots} Almost all major relational databases implement
+isolation levels that execute transactions against a static snapshot
+of the database.  The semantics of a snapshot can be axiomatized as
+following:
 \begin{smathpar}
@@
 % lifetime of the transaction, like in the case of Repeatable Read (RR).
-
-\textbf{Rollbacks} Rollbacks triggered by the post-facto detection of
-write-write conflicts can be axiomatized as following:
-\begin{smathpar}
-\begin{array}{lcl}
-  \I_{rb}(\stl,\stg,\stg') & = & \forall(r\in\stl)(r'\in\stg).~
-  r'.\idf = r.\idf \Rightarrow r'\in\stg'.
-\end{array}
-\end{smathpar}
 
@@
 $\stl\neq \emptyset$, and the transaction never commits.
+
+We now define various isolation guarantees that are either found in
+the database and distributed systems literature, or implemented by a
+commercial vendor.
+
+\textbf{Read Committed (RC) and Monotonic Atomic View (MAV)} RC
+isolation allows a transaction to witness only the writes of the
+committed transactions. In particular, \emph{dirty writes} of
+uncommitted transactions are not made visible. Monotonic Atomic View
+(MAV)~\cite{hat,quelea} is an extension to RC that guarantees the
+continuous visibility of a committed transaction's writes if they once
+become visible in the current transaction. That is, an MAV transaction
+does not witness \emph{disappearing writes}, which can happen on a
+weakly consistent machine. Due to the SC nature of our abstract
+machine (there is always a single global database state $\stg$; not a
+vector of states indexed by vector clocks), and our choice to never
+violate atomicity of a transaction's writes, both RC and MAV are
+already guaranteed by our semantics. Thus $\I_e$ and $\I_c$ are both
+$true$ for RC and MAV.
+
+\textbf{Repeatable Read (RR)} By definition, a read in a Repeatable
+Read transaction is required to return the same value regardless of
+how many ever times it is repeated. RR is often implemented (for e.g.,
+in ~\cite{mysql,quelea,hat}) by executing the transaction against a
+(conceptual) snapshot of the database, but committing its writes to
+the actual database. This implementation of RR can be axiomatized as
+$\I_e = \I_{ss}$ and $\I_{c}=true$. However, this specification of RR
+is stronger than the ANSI SQL specification, which requires no more
+than the invariance of already read records. In particular, ANSI SQL
+RR allows \emph{phantom reads} phenomenon, where a \C{SELECT} query
+might return newer records that were not previously present. This
+specification is implemented, for e.g., in~\cite{mssql}, using
+record-level exclusive read locks, that prevent a record from being
+modified while it is read by an uncommitted transaction. ANSI SQL RR
+can be axiomatized in our framework, but it requires a minor extension
+to our operational semantics to track a transaction's reads. In
+particular, the records returned by \C{SELECT} should be added to the
+local database $\stl$, but without changing their transaction
+idenfitiers ($\txnf$ fields), and flush ($\gg$) should only flush the
+records that bear the current transaction's identifier. With this
+extension, ANSI SQL RR can be axiomatized thus ($i$ is the id of the
+current transaction):
+\begin{smathpar}
+\begin{array}{lcl}
+  \I_e(\stl,\stg,\stg') & \Leftrightarrow & \forall(r'\in\stl)(r \in \stg).~
+      r'.\idf \neq i \wedge r.\idf = r'.\idf  \Rightarrow r\in\stg'.\\
+  \I_c(\stl,\stg,\stg') & \Leftrightarrow & true\\
+\end{array}
+\end{smathpar}
+
+\textbf{Snapshot Isolation (SI)} The concept of executing a
+transaction against a consistent snapshot of the database was first
+proposed as Snapshot Isolation in~\cite{berenson}. SI doesn't admit
+write-write conflicts, and the original proposal, which is implemented
+in MS SQL Server~\cite{sqlserver}, required the
+database to roll-back an SI transaction if conflicts are detected
+during the commit. This behavior can be axiomatized as following:
+\begin{smathpar}
+\begin{array}{lcl}
+\I_e\,\,(\stl,\stg,\stg') & = & \stg' = \stg\\
+\I_c\,\,(\stl,\stg,\stg') & = & \forall(r\in\stl)(r'\in\stg).~ r'.\idf = r.\idf \Rightarrow r'\in\stg'.
+\end{array}
+\end{smathpar}
+Observe that the same axiomatization applies to Postgres's RR,
+although its implementation (described in Sec.~\ref{sec:motivation})
+differs considerably from the original proposal. Thus, reasoning done
+for an SI transaction on MS SQL server carries over Postgres's RR and
+vice-versa, demonstrating the benefits of reasoning axiomatically
+about isolation properties.
+
+\textbf{Serializability (SER)} Axiomatization of serializability is
+straightforward:
+\begin{smathpar}
+\begin{array}{lcl}
+\I_e\,\,(\stl,\stg,\stg') & = & \stg' = \stg\\
+\I_c\,\,(\stl,\stg,\stg') & = & \stg' = \stg.
+\end{array}
+\end{smathpar}
 

### 2017-07-04 abstract.tex

@@
   necessary to use this logic may dissuade application developers, we
-  also describe a fully automated verification tool based on this
-  foundation that determines the safety of high-level consistency
-  assertions associated with such transactions.  The key to effective
-  inference is the observation that weakly-isolated transactions can
-  be viewed as functional (monadic) computations over database state,
-  allowing us to treat their operations as state transformers over the
-  database.  This interpretation enables automated verification using
-  off-the-shelf SMT solvers.
+  also describe an inference procedure based on this foundation that
+  ascertains the weakest isolation level that still guarantees the
+  safety of high-level consistency assertions associated with such
+  transactions.  The key to effective inference is the observation
+  that weakly-isolated transactions can be viewed as functional
+  (monadic) computations over database state, allowing us to treat
+  their operations as state transformers over the database.  This
+  interpretation enables automated verification using off-the-shelf
+  SMT solvers.
 
@@
   concurrency control mechanisms.  Case studies and experiments on
-  real-world applications demonstrate the utility of our approach, and
-  provide strong evidence that automated verification of
-  weakly-isolated transactions can be placed on the same formal
-  footing as their strongly-isolated serializable counterparts.
+  real-world applications (written in an embedded DSL in OCaml)
+  demonstrate the utility of our approach, and provide strong evidence
+  that automated verification of weakly-isolated transactions can be
+  placed on the same formal footing as their strongly-isolated
+  serializable counterparts.
 

### 2017-07-04 introduction.tex

@@
 is sensible if transactions are serializable, but incorrect under the
-\emph{read committed} isolation level used in the experiments. The
-same study has found that 13\% of all invariants among 67 open source
-Ruby-on-Rails applications are at risk of being violated due to weak
+weak isolation level used in the experiments. The same study has found
+that 13\% of all invariants among 67 open source Ruby-on-Rails
+applications are at risk of being violated due to weak
 isolation. Indeed, incidents of safety violations due to executing
@@
 transactional programs in the presence of a weakly-isolated
-concurrency control mechanism.  Moreover, we also present an inference
-algorithm based on a state transformer semantics that allows us to
-automatically verify the weakest isolation level for a transaction
-that ensures its consistency requirements are maintained.  The key to
-inference is our formulation of database state (represented as sets of
-tuples) as a monad, in which database computations are interpreted as
-transformers over these sets.  This interpretation leads to an
-encoding database operations amenable for verification by SMT solvers.
+concurrency control mechanism.  Realizing that the proof burden to use
+these rules would likely discourage applications programmers, we
+present an inference procedure based on a state transformer semantics
+that allows us to automatically verify the weakest isolation level for
+a transaction that ensures its consistency requirements are
+maintained.  The key to inference is our formulation of database state
+(represented as sets of tuples) as a monad, and in which database
+computations are interpreted as transformers over these sets.  This
+interpretation leads to an encoding of database computations amenable
+for verification by off-the-shelf SMT solvers.
 
-We have realized these ideas as an embedded DSL within OCaml, with
-support for common SQL operations (updates, selects, inserts, deletes,
-joins, etc.).  Experimental results on real-world benchmarks
-demonstrate the feasibility of automated verification of weakly
-isolated transactions, an important advance in improving the safety of
-realistic database computations. The paper makes the following
-contributions:
+We have realized these ideas as an embedded DSL within OCaml, that
+supports common relational SQL operations (updates, selects, inserts,
+deletes, joins, etc.).  Experimental results on real-world benchmarks
+demonstrate the feasibility and value of automated verification for
+weakly isolated transactions, an important advance in improving the
+safety of realistic database computations. The paper makes the
+following contributions:
 \vspace*{-6pt}
 \begin{enumerate}
-\item An operational semantics for an embedded DSL in OCaml that
-  expresses SQL-based relational database operations as monadic
-  computations on database state represented abstractly as sets of
-  tuples.
- \item A compositional proof system for this language capable of
-   relating high-level application invariants to transformations on
-   database state, parameterized by a weak isolation semantics that
+\item An embedded DSL in OCaml that treats SQL-based relational
+  database operations as monadic computations over an abstract
+  database state.
+ \item An operational semantics and compositional proof system for
+   this language capable of relating high-level application invariants
+   to database state, parameterized by a weak isolation semantics that
    selectively exposes the visibility of these operations to other
    transactions.
- \item An inference algorithm capable of discovering the
-   weakest isolation level that is sound with respect to a
-   transaction's high-level consistency requirements.  The algorithm
-   exploits our monadic formulation, treating database computations as
-   predicate state transformers, that leads to a natural encoding
-   amenable for automated verification.
+ \item An inference algorithm capable of discovering the weakest
+   isolation level that is sound with respect to a transaction's
+   high-level consistency requirements.  The algorithm interprets
+   database operations as predicate state transformers leadig to a
+   natural encoding amenable for automated verification.
  \item A realistic implementation along with results on real-world
    database benchmarks that justify our approach and the utility of
-   our proof methodology.
+   our proof methodology and inference procedure.
 \end{enumerate}

### 2017-07-04 opsem.tex

@@
 The small-step semantics is stratified into a transaction-local reduction
-relation, and a top-level reduction relation. Transaction-local
+relation, and a top-level reduction relation. The transaction-local
 relation ($\stg \vdash (c,\stl) \stepsto (c',\stl')$) defines a
@@
 comes tagged with an isolation specification $\I$, which has two
-components $\I_e$ and $\I_c$, the dictate the timing and nature of
+components $\I_e$ and $\I_c$, that dictate the timing and nature of
 interferences that the transaction can witness, during its execution
 ($\I_e$), and when it is about to commit ($\I_c$).  Formally, $\I_e$
-and $\I_c$ are predicates over by the (current) transaction-local
+and $\I_c$ are predicates over the (current) transaction-local
 database state ($\stl$), the state ($\stg$) of the global database
@@
 For brevity in presentation, we use $\I$ instead of $\I_e$ and $\I_c$
-whenever the discussion applies to both $\I_e$ and $\I_c$.
+when its destructed form is not required.
 
@@
 ($\I_c$).  The distinction between $\I_e$ and $\I_c$ allows us to
-model snapshot semantics of realistic isolation levels that isolate a
-transaction from interference during its execution, but exposes
-interference at the commit point.
+model the snapshot semantics of realistic isolation levels that
+isolate a transaction from interference during its execution, but
+expose interferences at the commit point.
 

### 2017-07-04 reasoning.tex

@@
 We now describe a proof system that lets us prove the correctness of a
-\txnimp program $c$ w.r.t its high-level invariant $I$, on a machine
-that satisfies the isolation specifications ($\I$) of its
-transactions\footnote{Note the difference between $I$ and $\I$. The
-former constitute \emph{proof} \emph{obligations} for the programmer,
-whereas the latter describes a transaction's \emph{assumptions} about
-operational characteristics of the underlying machine.}.  Our proof
-system is essentially an adaptation of the rely-guarantee reasoning
-framework to the setting of weakly isolated database transactions. It
-deals with the additional complexity introduced by this setting, while
-also taking advantage of the opportunities it provides to simplify the
-reasoning. The primary challenge is the weak isolation; how do we
-relate a transaction's isolation specification ($\I$) to its rely
-relation ($R$) that describes the transaction's environment, so that
-the interference is considered only insofar as the isolation level
-allows it? Our proof system shows how to compute a rely relation ($R$)
-modulo the isolation specification ($\I$) of a transaction. Another
-characteristic of the transaction setting is atomicity; a
-transaction's writes are not visible to its peers until it commits. In
-the context of rely-guarantee, this means that the transactions
-guarantee ($G$) should capture the aggregate effect of a transaction,
-and not its individual writes.  While shared memory \C{atomic} blocks
-also have the same characteristic, the fact that a transactions are
-weakly isolated introduces complexity.  Unlike an \C{atomic} block,
-the effect of a transaction is \emph{not} a sequential composition of
-the effects of its statements because each statement witnesses a
-potentially different version of the state. For example,  a series of
-operations on a shared variable (\C{X}) can be composed to describe
-the effect of an \C{atomic} block, while same cannot be done in a
-transaction. Thus, weak isolation needs to be taken into account when
-defining and verifying guarantees, and our proof system shows how.
+\txnimp program $c$ w.r.t its high-level consistency conditions $I$,
+on an implementation that satisfies the isolation specifications
+($\I$) of its transactions\footnote{Note the difference between $I$
+  and $\I$. The former constitute \emph{proof} \emph{obligations} for
+  the programmer, whereas the latter describes a transaction's
+  \emph{assumptions} about the operational characteristics of the
+  underlying system.}.  Our proof system is essentially an adaptation
+of rely-guarantee reasoning framework~\cite{rgjones} to the setting of
+weakly isolated database transactions.  The primary challenge in the
+formulation deals with how we relate a transaction's isolation
+specification ($\I$) to its rely relation ($R$) that describes the
+transaction's environment, so that interference is considered only
+insofar as allowed by the isolation level.  Another characteristic of
+the transaction setting that affects the structure of the proof system
+is atomicity; a transaction's writes are not visible to its peers
+until it commits.  In the context of rely-guarantee, this means that
+the transaction's guarantee ($G$) should capture the aggregate effect
+of a transaction, and not its individual writes.  While shared memory
+\C{atomic} blocks also have the same characteristic, the fact that
+transactions are weakly isolated introduces complexity.  Unlike an
+\C{atomic} block, the effect of a transaction is \emph{not} a
+sequential composition of the effects of its statements because each
+statement witnesses a potentially different version of the state.
 
@@
 quintuple $\rg{P,R}{c}{G,Q}$. Instead of a separate $P$ and $Q$, our
-top-level judgment has $I$ for both pre- and post-conditions, because
-our focus is on verifying that a \txnimp program \emph{preserves}
+top-level judgment uses $I$ for both pre- and post-conditions, because
+our focus is on verifying that a \txnimp\ program \emph{preserves}
 database's consistency conditions\footnote{The terms \emph{consistency
     condition}, \emph{high-level invariant}, and \emph{integrity
-constraint} all mean the same.}. Transaction-local RG judgment doesn't
-include a guarantee relation because transaction-local effects are not
-visible outside a transaction. Also, the rely relation ($\R$) of the
-transaction-local judgment is not same as the top-level rely relation
-($R$) as it takes into account the transaction's isolation
-specification ($\I$). Intuitively, $\R$ is $R$ modulo $\I$; formal
-definition will be given shortly. Recall that a transaction writes to
-its local database ($\stl$), which is then flushed when the
-transaction commits. Thus, the guarantee of a transaction depends on
-the state of its local database at the commit point. The pre- and
-post-condition assertions ($P$ and $Q$) in the local judgment
-facilitate tracking the changes to the transaction-local state, which
-eventually helps us prove the validity of the transaction's guarantee.
-Note that $P$ and $Q$ are bi-state assertions; they relate
-transaction-local database state ($\stl$) to the global database state
-($\stg$). Thus, the transaction-local judgment effectively tracks how
-transaction-local and global states change in relation to each other.
+    constraint} all mean the same.}.  A transaction-local RG judgment
+does not include a guarantee relation because transaction-local
+effects are not visible outside a transaction. Also, the rely relation
+($\R$) of the transaction-local judgment is not the same as the
+top-level rely relation ($R$) because it must take into account the
+transaction's isolation specification ($\I$). Intuitively, $\R$ is $R$
+modulo $\I$.  Recall that a transaction writes to its local database
+($\stl$), which is then flushed when the transaction commits. Thus,
+the guarantee of a transaction depends on the state of its local
+database at the commit point. The pre- and post-condition assertions
+($P$ and $Q$) in the local judgment facilitate tracking the changes to
+the transaction-local state, which eventually helps us prove the
+validity of the transaction's guarantee.  Both $P$ and $Q$ are
+bi-state assertions; they relate transaction-local database state
+($\stl$) to the global database state ($\stg$). Thus, the
+transaction-local judgment effectively tracks how transaction-local
+and global states change in relation to each other.
 
-The quintessential aspect of a rely-guarantee judgment is the
-stability condition, which, intuitively, requires the validity of an
-assertion $\phi$ to be unaffected by the interference, i.e., the rely
-relation $R$. In conventional RG, stability is defined as following
-($\sigma$ denotes a state):
+A central feature of a rely-guarantee judgment is a stability
+condition that requires the validity of an assertion $\phi$ to be
+unaffected by the interference, i.e., the rely relation $R$. In
+conventional RG, stability is defined as follows, where $\sigma$ denotes a
+state:
 \begin{smathpar}
@@
 the required isolation as specified by $\I$. It relies on the
-transaction-local judgment to verify the the transaction body ($c$).
+transaction-local judgment to verify the transaction body ($c$).
 The precondition $P$ of $c$ must follow from the fact that the
@@
 ($\R_l$) for the local judgment is a ternary relation computed as $R$
-modulo $\I_e$ as following:
+modulo $\I_e$:
 \begin{smathpar}
@@
 \end{smathpar}
-Thus, $\R_l$ allows an interference only if it doesn't violate the
+Thus, $\R_l$ allows an interference only if it does not violate the
 execution-time isolation specification ($\I_e$) of the transaction. If a
@@
 that interference. This is sensible considering such interference is
-pre-empted in the operational semantics.
+prevented in the operational semantics.
 
@@
 stability of a binary assertion $Q$ w.r.t a ternary rely relation $\R$
-is defined as following:
+is defined as:
 \begin{smathpar}
@@
 by the isolation specification (which implicitly considers the local
-state $\stl$) takes $\stg$ to $\stg'$, the $Q$ must also relate $\stl$
+state $\stl$) takes $\stg$ to $\stg'$, then $Q$ must also relate $\stl$
 to $\stg'$.
 
-For the guarantee $G$ of the transaction to be valid, it must follow
+For the guarantee $G$ of a transaction to be valid, it must follow
 from the post-condition $Q$ of the body, provided that $Q$ is stable
 w.r.t the commit-time interference captured by $R_c$. $R_c$, like
-$R_l$, is computed as rely relation modulo isolation, except that
+$R_l$, is computed as a rely relation modulo isolation, except that
 commit-time isolation ($\I_c$) is considered. The validity of
@@
 In other words, if $Q$ relates the transaction-local database state
-($\stl$) to the state of the global database ($\stg$) before the
-commit, then $G$ must relate the states of the global database before
+($\stl$) to the state of the global database ($\stg$) before a transaction
+commits, then $G$ must relate the states of the global database before
 and after the commit. The act of commit captured by the flush
@@
 include multiple environment steps, and $R$ only captures a single
-step. This is nonetheless sound due to the the inductive reasoning: if
+step. This is nonetheless sound due to inductive reasoning: if
 $Q$ is preserved by every step of $R$, then $Q$ is preserved by
@@
 $\I$ is not same as $\R^{*}$; the former is a transitive relation
-constrained by $\I$, whereas the latter is the transitive closure of a
+constrained by $\I$, whereas the latter is the transitive closure of an
 $\I$-constrained relation. We therefore introduce a side-condition on
@@
 from $\stg$ to $\stg'$, then an $R$-step from $\stg'$ to $\stg''$
-shouldn't make the interference from $\stg$ to $\stg''$ valid. We call
-this the stability condition on $\I$, defined as below:
+should not make the interference from $\stg$ to $\stg''$ valid.  This
+stability condition on $\I$ is defined formally thus:
 \begin{smathpar}
@@
 record, and no further interference will restore the original record,
-because original record bears the id of a transaction that has long
-committed. Thus, $\I_{ss}$ remains invalid.
+because the original record bears the id of a transaction that has
+already committed. Thus, $\I_{ss}$ remains invalid.
 
@@
 
-Two rules of the transaction-local RG judgment are shown in
-Fig.~\ref{fig:rg-rules}. The rule \rulelabel{RG-Update} is
-illustrative of the RG rules for SQL statements; they basically
-reflect the structure of the corresponding reduction rule from
-Fig.~\ref{fig:txnimp}, and require no further explanation. The rule
+Salient rules of transaction-local RG judgment are shown in
+Fig.~\ref{fig:rg-rules}. These rules (\rulelabel{RG-Update},
+\rulelabel{RG-Select}, \rulelabel{RG-Delete}, and
+\rulelabel{RG-Insert}) reflect the structure of the corresponding
+reduction rule from Fig.~\ref{fig:txnimp}.  The rule
 \rulelabel{RG-Foreach} defines the RG judgment for a \C{FOREACH} loop.
 As is characteristic of loops, the reasoning is pivoted on a loop
-invariant $I$ (not same as the high-level invariant), that needs to be
-stable w.r.t $\R$. $I$ must follow $P$, the pre-condition of
-\C{FOREACH}, when no elements have been iterated, i.e, when
-$y=\emptyset$. The body of the loop can assume the loop invariant, and
-the fact that $z$ is an element from the set $x$ being iterated, to
-prove its post-condition $Q_c$. Operational semantics ensures that $z$
-is added to $y$ at the end of the iteration, hence $Q_c \conj z\in y$
-is valid at the end of the iteration, from which $I$ must follow. When
-\C{FOREACH} has finished execution, $y$, the set of iterated items, is
-the entire set $x$. Thus $I \conj y=x$ must imply the post-condition
-$Q$, which also needs to be stable. The rules for conditionals,
-sequencing etc., are more-or-less standard, hence elided.
+invariant $\psi$ that needs to be stable w.r.t $\R$. $\psi$ must be
+implied by $P$, the pre-condition of \C{FOREACH}, when no elements
+have been iterated, i.e, when $y=\emptyset$. The body of the loop can
+assume the loop invariant, and the fact that $z$ is an element from
+the set $x$ being iterated, to prove its post-condition $Q_c$. The
+operational semantics ensures that $z$ is added to $y$ at the end of
+the iteration, hence $Q_c \conj z\in y$ is valid at the end of the
+iteration, from which $\psi$ must follow. When the loop has finished
+execution, $y$, the set of iterated items, is the entire set $x$. Thus
+$\psi \conj y=x$ must imply the post-condition $Q$, which also needs
+to be stable. The rules for conditionals, sequencing etc., are
+standard, and hence elided.
 
@@
 We can now assert the soundness of the RG judgments in
-Fig.~\ref{fig:rg-rules} as following\footnote{Formal proof of soundness
-is provided in the supplementary material.}:
+Fig.~\ref{fig:rg-rules} as follows\footnote{Its proof is provided in
+  the supplementary material.}:
 
@@
 every interleaved multi-step reduction of $c$ preserves the database
-integrity constraint $I$, and (b) the effect that every transaction in
-$c$ has on the database state is captured by $G$.
+integrity constraint (consistency condition) $I$, and (b) the effect
+that every transaction in $c$ has on the database state is captured by
+$G$.

### 2017-07-05 isolation.tex
edits

@@
 implementation. This allows us to model a range of isolation
-properties that are relevant to the theory and practice of database
-systems. We present some below:
+properties that are relevant to the theory and practice of realistic
+transaction processing systems, some of the most well-known of which are
+given below:
 
@@
 (Sec.~\ref{sec:motivation}) demonstrates, enforcing global uniqueness
-of ordered identifiers requires stronger isolation. Alternatively,
-globally unique sequence numbers, regardless of the isolation level,
-can be from a requested from a relational databases via SQL's
-\C{UNIQUE} and \C{AUTO\_INCREMENT} keywords. Our development crucially
-relies on the uniqueness of record identifiers\footnote{This has
-parallels in real-world. For e.g., MySQL's InnoDB engine automatically
-adds a 6-byte unique identifier if none exists for a record.}, which
-are generated locally by the \rulelabel{E-Insert} rule. Global
+of ordered identifiers requires stronger isolation than possible using
+a weak read committed isolation level. Alternatively, globally unique
+sequence numbers, regardless of the isolation level, can be requested
+from a relational databases via SQL's \C{UNIQUE} and
+\C{AUTO\_INCREMENT} keywords. Our development crucially relies on the
+uniqueness of record identifiers\footnote{This has parallels in
+  real-world. For e.g., MySQL's InnoDB engine automatically adds a
+  6-byte unique identifier if none exists for a record.}, which are
+generated locally by the \rulelabel{E-Insert} rule.  The global
 uniqueness of locally generated identifiers can be axiomatized as an
@@
 same id. The axiom thus simulates a global counter protected by an
-exclusive lock.
+exclusive lock without explicitly appealing to specific
+implementation artifacts.
 
-\textbf{Write-Write Conflicts}. Databases often employ a combination of
-concurrency control methods, both optimistic (e.g., speculation and
-rollback) and pessimistic (e.g., exclusive locking), to eliminate
-write-write conflicts among concurrent transactions. The absence of
-write-write (ww) conflicts, regardless of how it is achieved, can be
-axiomatized as following:
+\textbf{Write-Write Conflicts}. Databases often employ a combination
+of concurrency control methods, both optimistic (e.g., speculation and
+rollback) and pessimistic (e.g., various degrees of locking), to
+eliminate write-write (ww) conflicts among concurrent
+transactions. We can specify the  absence of such conflicts using our
+tri-state formulation thus:
 \begin{smathpar}
@@
 isolation levels that execute transactions against a static snapshot
-of the database.  The semantics of a snapshot can be axiomatized as
-following:
+of the database that can be axiomatized thus:
 \begin{smathpar}
@@
 \textbf{Read-Only Transactions}. Certain databases implement special
-previleges for read-only transactions. Read-only behavior can be
-enforced on a transaction by including the following proposition in
-its $\I$:
+privileges for read-only transactions. Read-only behavior can be
+enforced on a transaction by including the following proposition as
+part of its isolation invariant:
 \begin{smathpar}
@@
 \end{smathpar}
-If a transaction declared as read-only performs a write, then its
-$\stl\neq \emptyset$, and the transaction never commits.
+%% If a transaction declared as read-only performs a write, then its
+%% $\stl\neq \emptyset$, and the transaction never commits.
 
-We now define various isolation guarantees that are either found in
-the database and distributed systems literature, or implemented by a
-commercial vendor.
+In addition to these properties, there are various isolation
+guarantees proposed in the database or distributed systems literature,
+or implemented by commercial vendors that can also be specified within
+this framework:
 
 \textbf{Read Committed (RC) and Monotonic Atomic View (MAV)} RC
-isolation allows a transaction to witness only the writes of the
-committed transactions. In particular, \emph{dirty writes} of
-uncommitted transactions are not made visible. Monotonic Atomic View
-(MAV)~\cite{hat,quelea} is an extension to RC that guarantees the
-continuous visibility of a committed transaction's writes if they once
-become visible in the current transaction. That is, an MAV transaction
+isolation allows a transaction to witness writes of committed
+transactions at any point during the transaction's execution.
+Although it offers only weak isolation guarantees, it nonetheless
+prevents witnessing \emph{dirty writes} (i.e., writes performed by
+uncommitted transactions).  Monotonic Atomic View
+(MAV)~\cite{bailishat} is an extension to RC that guarantees the
+continuous visibility of a committed transaction's writes once they
+become visible in the current transaction. That is, a MAV transaction
 does not witness \emph{disappearing writes}, which can happen on a
@@
 violate atomicity of a transaction's writes, both RC and MAV are
-already guaranteed by our semantics. Thus $\I_e$ and $\I_c$ are both
-$true$ for RC and MAV.
+already guaranteed by our semantics.  Thus, defining $\I_e$ and $\I_c$
+to \emph{true} ensures RC and MAV behavior under our semantics.
 
-\textbf{Repeatable Read (RR)} By definition, a read in a Repeatable
-Read transaction is required to return the same value regardless of
-how many ever times it is repeated. RR is often implemented (for e.g.,
-in ~\cite{mysql,quelea,hat}) by executing the transaction against a
+\textbf{Repeatable Read (RR)} By definition, a read to a transactional
+variable in a Repeatable Read transaction is required to return the
+same value whenever accessed.  RR is often implemented (for e.g., in
+~\cite{mysqliso,bailishat}) by executing the transaction against a
 (conceptual) snapshot of the database, but committing its writes to
@@
 than the invariance of already read records. In particular, ANSI SQL
-RR allows \emph{phantom reads} phenomenon, where a \C{SELECT} query
-might return newer records that were not previously present. This
-specification is implemented, for e.g., in~\cite{mssql}, using
+RR allows \emph{phantom reads}, a phenonmenon in which a \C{SELECT}
+query might return newer records, inserted into tables, that were not
+present at the point the RR transaction started.  This specification
+is implemented, for e.g., in Microsoft's SQL server, using
 record-level exclusive read locks, that prevent a record from being
-modified while it is read by an uncommitted transaction. ANSI SQL RR
-can be axiomatized in our framework, but it requires a minor extension
-to our operational semantics to track a transaction's reads. In
-particular, the records returned by \C{SELECT} should be added to the
-local database $\stl$, but without changing their transaction
-idenfitiers ($\txnf$ fields), and flush ($\gg$) should only flush the
-records that bear the current transaction's identifier. With this
-extension, ANSI SQL RR can be axiomatized thus ($i$ is the id of the
-current transaction):
+modified while it is read by an uncommitted transaction.  The ANSI SQL
+RR specification can be axiomatized in our framework, but it requires
+a minor extension to our operational semantics to track a
+transaction's reads. In particular, the records returned by \C{SELECT}
+should be added to the local database $\stl$, but without changing
+their transaction identifiers ($\txnf$ fields), and flush ($\gg$)
+should only flush the records that bear the current transaction's
+identifier.  With this small extension, ANSI SQL RR can be axiomatized
+thus ($i$ is the id of the current transaction):
 \begin{smathpar}
@@
 write-write conflicts, and the original proposal, which is implemented
-in MS SQL Server~\cite{sqlserver}, required the
-database to roll-back an SI transaction if conflicts are detected
-during the commit. This behavior can be axiomatized as following:
+in Microsoft SQL Server, required the database to roll-back an SI
+transaction if conflicts are detected during the commit. This behavior
+can be axiomatized thus:
 \begin{smathpar}
@@
 differs considerably from the original proposal. Thus, reasoning done
-for an SI transaction on MS SQL server carries over Postgres's RR and
-vice-versa, demonstrating the benefits of reasoning axiomatically
+for an SI transaction on MS SQL server carries over to Postgres's RR
+and vice-versa, demonstrating the benefits of reasoning axiomatically
 about isolation properties.
@@
 
-%\usepackage[table,xcdraw]{xcolor}
-\begin{table}[]
-\centering
-\begin{tabular}{l|l|l|l|l|l|l|l|l|l|l|l|l|}
-\cline{2-13}
-                                & \multicolumn{6}{c|}{{\color[HTML]{333333} PostgreSQL}}                         & \multicolumn{6}{c|}{MySql}                                                   \\ \cline{2-13}
-                                & \multicolumn{2}{c|}{RC} & \multicolumn{2}{c|}{RR} & \multicolumn{2}{c|}{SER} & \multicolumn{2}{c|}{RC} & \multicolumn{2}{c|}{RR} & \multicolumn{2}{c|}{SER} \\ \cline{2-13}
-                                & T          & F          & T          & F          & T           & F          & T          & F          & T          & F          & T           & F          \\ \hline
-\multicolumn{1}{|l|}{$\I_{ss}$} &            &            & \checkmark &            & \checkmark  & \checkmark &            &            & \checkmark &            & \checkmark  & \checkmark \\ \hline
-\multicolumn{1}{|l|}{$\I_{rb}$} &            &            &            & \checkmark &             &            &            &            &            &            &             &            \\ \hline
-\multicolumn{1}{|l|}{$\I_{ro}$} &            &            &            &            &             &            &            &            & \checkmark & \checkmark &             &            \\ \hline
-\multicolumn{1}{|l|}{$\I_{id}$} & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark  & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark  & \checkmark \\ \hline
-\multicolumn{1}{|l|}{$\I_{ww}$} & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark  & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark  & \checkmark \\ \hline
-\end{tabular}
+%% %\usepackage[table,xcdraw]{xcolor}
+%% \begin{table}[]
+%% \centering
+%% \begin{tabular}{l|l|l|l|l|l|l|l|l|l|l|l|l|}
+%% \cline{2-13}
+%%                                 & \multicolumn{6}{c|}{{\color[HTML]{333333} PostgreSQL}}                         & \multicolumn{6}{c|}{MySql}                                                   \\ \cline{2-13}
+%%                                 & \multicolumn{2}{c|}{RC} & \multicolumn{2}{c|}{RR} & \multicolumn{2}{c|}{SER} & \multicolumn{2}{c|}{RC} & \multicolumn{2}{c|}{RR} & \multicolumn{2}{c|}{SER} \\ \cline{2-13}
+%%                                 & T          & F          & T          & F          & T           & F          & T          & F          & T          & F          & T           & F          \\ \hline
+%% \multicolumn{1}{|l|}{$\I_{ss}$} &            &            & \checkmark &            & \checkmark  & \checkmark &            &            & \checkmark &            & \checkmark  & \checkmark \\ \hline
+%% \multicolumn{1}{|l|}{$\I_{rb}$} &            &            &            & \checkmark &             &            &            &            &            &            &             &            \\ \hline
+%% \multicolumn{1}{|l|}{$\I_{ro}$} &            &            &            &            &             &            &            &            & \checkmark & \checkmark &             &            \\ \hline
+%% \multicolumn{1}{|l|}{$\I_{id}$} & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark  & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark  & \checkmark \\ \hline
+%% \multicolumn{1}{|l|}{$\I_{ww}$} & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark  & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark  & \checkmark \\ \hline
+%% \end{tabular}
 
-\caption{The semantics of isolation levels on PostgreSQL and MySQL}
-\label{fig:iso-table}
-\end{table}
+%% \caption{The semantics of isolation levels on PostgreSQL and MySQL}
+%% \label{fig:iso-table}
+%% \end{table}
 
-With help of the above axiomatizations, we can now define the precise
-semantics of various isolation levels on PostgreSQL and MySQL. The table
-in Table~\ref{fig:iso-table} captures the summary.
+%% With help of the above axiomatizations, we can now define the precise
+%% semantics of various isolation levels on PostgreSQL and MySQL. The table
+%% in Table~\ref{fig:iso-table} captures the summary.

### 2017-07-05 isolation.tex
ANSI RR spec changed.

@@
 identifier.  With this small extension, ANSI SQL RR can be axiomatized
-thus ($i$ is the id of the current transaction):
+thus:
 \begin{smathpar}
 \begin{array}{lcl}
-  \I_e(\stl,\stg,\stg') & \Leftrightarrow & \forall(r'\in\stl)(r \in \stg).~
-      r'.\idf \neq i \wedge r.\idf = r'.\idf  \Rightarrow r\in\stg'.\\
+  \I_e(\stl,\stg,\stg') & \Leftrightarrow & \forall(r\in\stl).
+      r \in \Delta \Rightarrow r \in \Delta'\\
   \I_c(\stl,\stg,\stg') & \Leftrightarrow & true\\
@@
 \end{smathpar}
+If a record $r$ belongs to both $\stl$ and $\stg$, then it must be a
+record written by a different transaction and read by the current
+transaction. By requiring $r\in\stg'$, $\I_e$ guarantees the
+invariance of $r$, thus the repeatability of the read.
 

### 2017-07-05 inference.tex

\section{Inference}
\label{sec:inference}

\input{inference-rules}

### 2017-07-05 isolation.tex

@@
 than the invariance of already read records. In particular, ANSI SQL
-RR allows \emph{phantom reads}, a phenonmenon in which a \C{SELECT}
-query might return newer records, inserted into tables, that were not
-present at the point the RR transaction started.  This specification
-is implemented, for e.g., in Microsoft's SQL server, using
-record-level exclusive read locks, that prevent a record from being
-modified while it is read by an uncommitted transaction.  The ANSI SQL
-RR specification can be axiomatized in our framework, but it requires
-a minor extension to our operational semantics to track a
-transaction's reads. In particular, the records returned by \C{SELECT}
-should be added to the local database $\stl$, but without changing
-their transaction identifiers ($\txnf$ fields), and flush ($\gg$)
-should only flush the records that bear the current transaction's
-identifier.  With this small extension, ANSI SQL RR can be axiomatized
-thus:
+RR allows \emph{phantom reads}, a phenonmenon in which a repeated
+\C{SELECT} query might return newly inserted records that were not
+previously returned. This specification is implemented, for e.g., in
+Microsoft's SQL server, using record-level exclusive read locks, that
+prevent a record from being modified while it is read by an
+uncommitted transaction.  The ANSI SQL RR specification can be
+axiomatized in our framework, but it requires a minor extension to our
+operational semantics to track a transaction's reads. In particular,
+the records returned by \C{SELECT} should be added to the local
+database $\stl$, but without changing their transaction identifiers
+($\txnf$ fields), and flush ($\gg$) should only flush the records that
+bear the current transaction's identifier. With this extension, ANSI
+SQL RR can be axiomatized thus:
 \begin{smathpar}

### 2017-07-06 inference.tex
Translation to MSFOL.
Sec. 4 Part 1.

@@
 
+Rely-Guarantee framework presented in the previous section facilitates
+modular proofs for weakly-isolated transactions, but imposes a
+non-trivial annotation burden on the programmer. In particular, it
+requires each statement ($c$) of the transaction to be annotated with
+a stable pre- ($P$) and post-condition ($Q$), and loops to be
+annotated with stable inductive invariants ($\psi$). While weakest
+pre-condition style predicate transformers can help in inferring $P$'s
+and $Q$'s for regular statements, inferring inductive loop invariants
+is still a challenge, even for the simpler form of loops we admit.
+Thus, predicate transformer-based approach to proof automation is
+unlikely to be effective in this case. We therefore present an
+alternative approach to RG proof automation in this section that is
+based on state transformers instead of predicate transformers. The
+idea is to infer the effect that each statement has on the
+transaction-local database state $\stl$ (i.e., how it transforms
+$\stl$), and compose multiple such effects together to describe the
+effect of the transaction as a whole. Importantly, this approach
+generalizes to loops, where the effect of a loop can be inferred as a
+well-defined function of the effect of its body, thanks to certain
+pleasant properties enjoyed by the database programs modeled by
+$\txnimp$.
+
+\input{logic-syntax}
+
+At the core of our approach is a simple language ($\SL$) to express
+set transformations shown in Fig.~\ref{fig:logic-syntax}. The language
+admits only set expressions that include variables ($x$), literals of
+the form $\{x \,|\, \phi^0\}$ where $\phi^0$ is a propositional
+(quantifier-free) formula on $x$, a restricted form of existential
+quantification that binds a set $x$ satisfying proposition $\phi^0$ in
+a set expression $s$, a monadic composition of two set expressions
+($s_1$ and $s_2$) using bind ($\bind$), a conditional set expression
+where the condition is a propositional formula, and a union of two set
+expressions. Symbols $\stl$ and $\stg$ are also variables in $\SL$,
+but are used to denote local and database states, respectively.  The
+language is carefully chosen to be expressive enough to capture the
+semantics of $\txnimp$ statements, yet simple enough to have a
+semantics-preserving translation to the first-order logic.
+
 \input{inference-rules}
+
+Fig.~\ref{fig:inference-rules} shows the syntax-directed inference
+rules for $\txnimp$ commands inside a transaction $\C{TXN}_i$.
+Inference depends on the $\I$-constrained rely relation $\R$, and the
+high-level invariant $I$ for the reasons that will become clear
+shortly.  Inference rules compute, for each command $c$, a (meta)
+function $\F$ that returns a set as an expression in $\SL$, given a
+pair of sets ($\stl$ and $\stg$) that describe local and a global
+database states, respectively. The expression returned by
+$\F(\stl,\stg)$ (abstractly) describes the set of records that get
+added to $\stl$ as a result of executing $c$ under $\stl$ and $\stg$.
+Thus, $\F$ captures the \emph{effect} part of the state transformer of
+$c$, which is the function $\lambda(\stl,\stg).~\stl \cup
+\F(\stl,\stg)$. For $\F$ to be useful in RG verification, it needs to
+be stable w.r.t the rely relation $\R$. The stability condition on
+effects is defined below:
+\begin{smathpar}
+\begin{array}{lcl}
+  \stable(\R,\F) & \Leftrightarrow & \forall \stl,\stg,\stg'.~
+  \R(\stl,\stg,\stg') \Rightarrow \F(\stl,\stg) \equiv \F(\stl,\stg')
+\end{array}
+\end{smathpar}
+Intuitively, the stability condition requires an effect to describe
+the same set of records before and after the interference. The
+equivalence in $\SL$ translates to the equality in first-order logic,
+as will be explained shortly. In the inference rules, stability is
+enforced constructively by a meta function $\stabilize{\cdot}$, which
+accepts an effect and returns a new effect that is guaranteed to be
+stable under $\R$. $\stabilize{\cdot}$ achieves the stability
+guarantee by abstracting away the bound $\stg$ in an unstable $\F$ to
+an existentially bound $\stg'$ as described below:
+\begin{smathpar}
+\begin{array}{lcll}
+  \stabilize{\F} & = & \F & \texttt{if } \stable(\R,\F)\\
+  & = & \lambda (\stl,\stg).~\existsl(\stg',I(\stg'),\F(\stl,\stg')) & \texttt{otherwise}\\
+\end{array}
+\end{smathpar}
+Observe that when $\F$ is not stable, then $\stabilize{\F}$ returns an
+$\F'$ that simply ignores its $\stg$ argument in favor of a generic
+$\stg'$, making $\F'$ trivially stable. The existentially bound
+$\stg'$ satisfies $I$ since commands always executed against a valid
+database state.
+
+The inference rules of SQL statements, like the RG rules, closely
+follow the corresponding reduction rules in Fig.~\ref{fig:txnimp},
+except that their language of expression is $\SL$. For instance, while
+the reduction rule for \C{UPDATE} declaratively specifies the set of
+updated records, the inference rule uses $\SL$'s bind operation to
+\emph{compute} the set. Other SQL rules do likewise. The inference
+rules for \C{LET} binders, conditionals, and sequences compose the
+effects inferred for their subcommands. For instance, the effect of a
+sequence of commands $c_1;c_2$ is the union of effects $F_1$ and $F_2$
+of $c_1$ and $c_2$, respectively, except that $F_2$ is applied to a
+state ($\stl$) to which $F_1$ has already been applied, reflecting
+their order of reduction. The inference rule for \C{FOREACH} takes
+advantage of the $\SL$'s bind operator to lift the effect inferred for
+the loop body to the level of the loop. Since records added to $\stl$
+in each iteration of \C{FOREACH} are independent of the previous
+iteration, sequential composition of the effects of different
+iterations is same as their parallel composition. Since the loop body
+is executed once per each $z\in x$, the effect of the the loop is the
+union of effects ($\F$) for all $z\in x$, all applied to the same
+state ($\stl$ and $\stg$). That is, $\F_{loop}(\stl,\stg) =
+\bigcup_{z\in x}\F_{body}(\stl,\stg)$. From the definition of
+set monad's bind, $\F_{loop}(\stl,\stg) = x \bind (\lambda
+z.~F_{body}(\stl,\stg))$, which is what the rule infers.
+
+\input{logic}
+
+\begin{theorem}
+if $c \elabsto \F$ and $\semof{\F(\stl,\stg)} = (\phi^1,g)$ and
+$P(\stl,\stg) = true$ and $Q(\stl',\stg) = \exists\stl.~ \phi^1 \wedge
+  \forall a. a\in\stl' \Leftrightarrow a\in\stl \vee g(a)$, then
+  $\R \vdash \hoare{P}{c}{Q}$
+\end{theorem}

### 2017-07-06 inference.tex

@@
 \begin{array}{lcll}
-  \stabilize{\F} & = & \F & \texttt{if } \stable(\R,\F)\\
-  & = & \lambda (\stl,\stg).~\existsl(\stg',I(\stg'),\F(\stl,\stg')) & \texttt{otherwise}\\
+  \stabilize{\F} & = & \F & \texttt{if } \stable(\R,\F).\\
+  & = & \lambda (\stl,\stg).~\existsl(\stg',I(\stg'),\F(\stl,\stg')) 
+      & \texttt{otherwise. }\stg'\texttt{ is a fresh name.}\\
 \end{array}
@@
 $\stg'$, making $\F'$ trivially stable. The existentially bound
-$\stg'$ satisfies $I$ since commands always executed against a valid
-database state.
+$\stg'$ satisfies $I$ since transactions are always executed against a
+valid database state.
 
@@
 
+% \begin{figure}[h]
+% \begin{smathpar}
+% \begin{array}{l}
+%   \foreache{xs}{(\lambda y.\lambda x.~ 
+%         \updatee{(\lambda z.~ \{\idf=z.\idf;\,\C{name}=z.\C{name};\,
+%                                \C{s\_id}=x.\idf\})}
+%                 {(\lambda z.~ z.\idf = x.\C{s\_id})};\\
+%         \hspace*{1.27in} \inserte{x})}
+% \end{array}
+% \end{smathpar}
+% \caption{A $\txnimp$ program that populates a database
+%   of couples.}
+% \label{fig:people_code}
+% \end{figure}
+
+\begin{example}
+Consider a database of couples that tracks the spouse
+relationship. Each record has an $\idf$ field, a \C{name} field, and a
+spouse id (\C{s\_id}) field. A $\txnimp$ transaction bearing id $i$
+that populates the database with a given set ($xs$) of couples will
+include the following code snippet:
+\begin{smathpar}
+\begin{array}{l}
+  \foreache{xs}{(\lambda y.\lambda x.~ 
+        \updatee{(\lambda z.~ \{\idf=z.\idf;\,\C{name}=z.\C{name};\,
+                               \C{s\_id}=x.\idf\})}
+                {(\lambda z.~ z.\idf = x.\C{s\_id})};\\
+        \hspace*{1.27in} \inserte{x})}
+\end{array}
+\end{smathpar}
+The effect inferred for the program, assuming no interference, is as
+following:
+\begin{smathpar}
+\begin{array}{l}
+  \lambda(\stl,\stg).~xs \bind 
+    (\lambda x. \stg \bind 
+      (\lambda z. \itel{z.\idf = x.\C{s\_id}}
+                       {\{\{\idf=z.\idf;\,\C{name}=z.\C{name};\,
+                               \C{s\_id}=x.\idf\}\}}
+                       {\emptyset}) \\
+      \hspace*{1cm} ~\cup~ \{y \,|\, y = \{x \with \delf=false; \}\})
+\end{array}
+\end{smathpar}
+Under the possibility of an interference affecting the stability, the
+following stable effect is inferred ($I$ is the database invariant):
+\begin{smathpar}
+\begin{array}{l}
+  \lambda(\stl,\stg).~xs \bind 
+    (\lambda x.~\existsl(\stg',I,\stg' \bind 
+      (\lambda z. \itel{z.\idf = x.\C{s\_id}}
+                       {\{\{\idf=z.\idf;\,\C{name}=z.\C{name};\,
+                               \C{s\_id}=x.\idf\}\}}
+                       {\emptyset})) \\
+      \hspace*{1cm} ~\cup~ \{y \,|\, y = \{x \with \delf=false; \}\})
+\end{array}
+\end{smathpar}
+\end{example}
+
 \input{logic}
 
+\subsection{Soundness of Inference}
+
+The correspondence between the inference rules and the RG judgment of
+Sec.~\ref{sec:reasoning} is stated thus\footnote{Proof can be found in
+the supplementary material}:
 \begin{theorem}
-if $c \elabsto \F$ and $\semof{\F(\stl,\stg)} = (\phi^1,g)$ and
-$P(\stl,\stg) = true$ and $Q(\stl',\stg) = \exists\stl.~ \phi^1 \wedge
-  \forall a. a\in\stl' \Leftrightarrow a\in\stl \vee g(a)$, then
-  $\R \vdash \hoare{P}{c}{Q}$
+\label{thm:inference-sound}
+  Forall i,$R$,$I$,$c$,$\F$, if $\stable(\R,I)$ and $c \elabsto \F$,
+  then $\R \vdash \hoare{{\sf true}}{c}{\F}$
 \end{theorem}
+Where, ${\sf true} = \lambda(\stl,\stg).~true$ and $\stable(\R,I)
+\Leftrightarrow \forall\stl,\stg,\stg'. I(\stg) \conj
+\R(\stl,\stg,\stg') \Rightarrow \I(\stg')$.
+
+\subsection{From $\SL$ to the first-order logic}
+
+Theorem~\ref{thm:inference-sound} lets us replace the local judgment
+of the \rulelabel{RG-Txn} rule (Fig.~\ref{fig:rg-rules}) by the
+inference judgment. The soundness of a transaction's guarantee can now
+be established w.r.t the effect $\F$ of the body. The
+\rulelabel{RG-Txn} rule updated to use the inference judgment is shown
+below:
+\begin{smathpar}
+\begin{array}{c}
+\RULE
+{
+  \stable(R,\I)\spc
+  \stable(R,I)\\
+  \R(\stl,\stg,\stg') \Leftrightarrow \I(\C{true},\stl,\stg,\stg') 
+        \wedge R(\stg,\stg') \spc
+  c \elabsto \F\\
+  \R_c(\stl,\stg,\stg') \Leftrightarrow \I(\C{false},\stl,\stg,\stg') 
+        \wedge R(\stg,\stg') \spc
+  \stable(\R_c,\F)\\
+  \forall \stl,\stg.~ \F(\stl,\stg) \Rightarrow 
+    G(\stg, \stl \gg \stg)\spc
+  \forall \stg,\stg'.~I(\stg) \wedge G(\stg,\stg') \Rightarrow I(\stg')\\
+}
+{
+  \rg{I,R}{\ctxn{i}{\I}{c}}{G,I}
+}
+\end{array}
+\end{smathpar}
+Since the inference judgment $c\elabsto \F$ is fully automatic,
+automating the application of the \rulelabel{RG-Txn} rule for a
+transaction now requires automating the multiple implication checks in
+the premise. While $R$, $\R$, $\R_c$, $\I$ and $\I$ are formulas in
+the first-order logic (FOL) with a relatively simple structure, $\F$
+is an expression in the set language $\SL$
+(Fig.~\ref{fig:logic-syntax}) with a possibly complex structure.
+Fortunately, however, there exists a semantics-preserving translation
+from $\SL$ to a restricted subset of first-order logic (FOL) that
+lends itself to automatic reasoning.
+
+The algorithm ($\mssemof{\cdot}$) shown in Fig.~\ref{fig:logic}
+encodes an $\SL$ expression in a FOL. Given a set expression $s$ of
+$\SL$, $\mssemof{s}$ is a pair $(\phi,f)$, where $\phi$ encodes $s$ in
+FOL, and $f$ is a boolean function that names the set defined by $s$.
+Naming provides an easy handle to $s$, simplifying $\mssemof{\cdot}$.
+Set variables ($x$, $\stl$, $\stg$, $\dots$) are encoded as
+unconstrained boolean functions ($g_x$, $g_{\stl}$, $g_{\stg}$,
+$\ldots$) that are named after the variable.  Encoding of a set
+literal $\{a\,|\,\varphi\}$ introduces a new boolean function $g$ that
+is $true$ for only those $a$ that satisfy $\varphi$.  Thus $g$
+faithfully encodes the literal. The encoding of the $\existsl{}$
+expression first eliminates the second-order quantification over sets.
+Existentially bound set variables ($a$) is skolemized, i.e.,
+instantiated with a new unbound variable $a'$, before encoding the
+bound expression $s$ and the first-order constraint $\phi$. The
+algorithm for encoding $\phi$, which is already a first-order formula,
+is straightforward, hence not shown. It basically involves replacing
+set variables with boolean functions named after the variables. The
+first-order encoding for a bind expression describes the semantics of
+set monad's $\bind$ in FOL. Let $s_1$ be a set, and let $f$ be a
+function that maps each variable in $s_1$ to a new set. Then, $s_2 =
+s_1 \bind f$ if and only if for all $y\in s_2$, there exists an $x \in
+s_1$ such that $y = f(x)$, and forall $x\in s_1$, $f(x)\in s_2$. The
+encoding adds new constraints to this effect. Conditional set
+expressions and set union are encoded straightforwardly.

### 2017-07-06 reasoning.tex

@@
 \begin{array}{lcl}
-\R_l(\stl,\stg,\stg') & \Leftrightarrow & R(\stg,\stg') \conj
-\I_e(\stl,\stg,\stg')
+%\R_l(\stl,\stg,\stg') & \Leftrightarrow & R(\stg,\stg') \conj
+%\I_e(\stl,\stg,\stg')
+  \R_l(\stl,\stg,\stg') & \Leftrightarrow & \exists \stg_1. \I_e(\stl, \stg_1, \stg) \wedge R(\stg, \stg') \wedge \I_e(\stl, \stg_1, \stg')
 \end{array}
 \end{smathpar}
-Thus, $\R_l$ allows an interference only if it does not violate the
-execution-time isolation specification ($\I_e$) of the transaction. If a
-certain interference violates the isolation spec, i.e., $R(\stg,\stg')
-\Rightarrow \neg(\I_e(\stl,\stg,\stg'))$, then $\R_l(\stl,\stg,\stg')
-\Leftrightarrow false$, and any assertion is trivially stable w.r.t
-that interference. This is sensible considering such interference is
-prevented in the operational semantics.
+In words, if there exists $\stg_1$ which was the global database state during the last step by a transaction and if the transaction can take its next step at the current global database state $\stg$, then even after the interference, the transaction should still be able to take its next step.
+If there is no such state $\stg_1$, then a transaction will never be able to take its next step after the interference, and hence such interference can be safely discarded.
+
+%\begin{smathpar}
+%\exists \stg_1. \I_e(\stl, \stg_1, \stg) \wedge R(\stg, \stg') \wedge \I_e(\stl, \stg_1, \stg') \Rightarrow \forall \stg_2 
+%\end{smathpar}
+%Thus, $\R_l$ allows an interference only if it does not violate the
+%execution-time isolation specification ($\I_e$) of the transaction. If a
+%certain interference violates the isolation spec, i.e., $R(\stg,\stg')
+%\Rightarrow \neg(\I_e(\stl,\stg,\stg'))$, then $\R_l(\stl,\stg,\stg')
+%\Leftrightarrow false$, and any assertion is trivially stable w.r.t
+%that interference. This is sensible considering such interference is
+%prevented in the operational semantics.
 

### 2017-07-06 inference.tex
Sec. 5 full.

@@
   Forall i,$R$,$I$,$c$,$\F$, if $\stable(\R,I)$ and $c \elabsto \F$,
-  then $\R \vdash \hoare{{\sf true}}{c}{\F}$
+  then:\\\vspace*{-0.2cm}
+  \begin{smathpar}
+  \begin{array}{c}
+  \R \vdash \hoare{\lambda(\stl,\stg).~\stl=\emptyset \conj
+  I(\stg)}{~c~}{\lambda(\stl,\stg).\stl = \F(\emptyset,\stg)}
+  \end{array}
+  \end{smathpar}
 \end{theorem}
-Where, ${\sf true} = \lambda(\stl,\stg).~true$ and $\stable(\R,I)
-\Leftrightarrow \forall\stl,\stg,\stg'. I(\stg) \conj
-\R(\stl,\stg,\stg') \Rightarrow \I(\stg')$.
+Where the set expression $\F(\emptyset,\stg)$ has a natural
+interpretation, on the lines discussed in the next sub-section.
+% Where, ${\sf true} = \lambda(\stl,\stg).~true$ and $\stable(\R,I)
+% \Leftrightarrow \forall\stl,\stg,\stg'. I(\stg) \conj
+% \R(\stl,\stg,\stg') \Rightarrow \I(\stg')$.
 
@@
   \stable(\R_c,\F)\\
-  \forall \stl,\stg.~ \F(\stl,\stg) \Rightarrow 
-    G(\stg, \stl \gg \stg)\spc
+  \forall \stg.~ G(\stg, F(\emptyset,\stg) \gg \stg)\spc
   \forall \stg,\stg'.~I(\stg) \wedge G(\stg,\stg') \Rightarrow I(\stg')\\
@@
 expression first eliminates the second-order quantification over sets.
-Existentially bound set variables ($a$) is skolemized, i.e.,
+Existentially bound set variable ($a$) is skolemized, i.e.,
 instantiated with a new unbound variable $a'$, before encoding the
@@
 algorithm for encoding $\phi$, which is already a first-order formula,
-is straightforward, hence not shown. It basically involves replacing
-set variables with boolean functions named after the variables. The
-first-order encoding for a bind expression describes the semantics of
-set monad's $\bind$ in FOL. Let $s_1$ be a set, and let $f$ be a
-function that maps each variable in $s_1$ to a new set. Then, $s_2 =
-s_1 \bind f$ if and only if for all $y\in s_2$, there exists an $x \in
-s_1$ such that $y = f(x)$, and forall $x\in s_1$, $f(x)\in s_2$. The
-encoding adds new constraints to this effect. Conditional set
-expressions and set union are encoded straightforwardly.
+is straightforward. It basically involves replacing set variables with
+boolean functions named after the variables. The first-order encoding
+for a bind expression describes the semantics of set monad's $\bind$
+in FOL. Let $s_1$ be a set, and let $f$ be a function that maps each
+variable in $s_1$ to a new set. Then, $s_2 = s_1 \bind f$ if and only
+if for all $y\in s_2$, there exists an $x \in s_1$ such that $y =
+f(x)$, and forall $x\in s_1$, $f(x)\in s_2$. The encoding adds new
+constraints to this effect. Conditional set expressions and set union
+are encoded straightforwardly.
+
+Observe that the language of encoding shown in Fig.~\ref{fig:logic} is
+a subset of logic that satisfies the following syntactic properties:
+\begin{itemize}
+  \item All quantification is first-order; second-order objects, such
+    as sets and functions, are never quantified.
+  \item Quantifiers appear only at the prenex position, i.e., at the
+    beginning of a quantified formula.  
+  \item All function symbols, modulo those that might appear in
+    $\varphi\in\Prop^0$, are uninterpreted and boolean.
+\end{itemize}
+The fragment of FOL that satisfies the above syntactic conditions is
+called Effectively Propositional first-order logic (denoted {\sf
+EPR}), or Bernay-Sch{\"o}nfinkel-Ramsey (BSR) logic, which is known to
+be decidable~\cite{z3epr}. The language of encoding, however, is a
+combination of {\sf EPR} with (a). $\Prop^0$, the theory from which
+quantifier-free propositions ($\varphi$) that encode object language
+expressions are drawn, and (b).  $\Prop^1$, the theory from which
+invariants ($I$) are drawn. We write $\SL[\Prop^0,\Prop^1]$ to
+highlight the parameterization of $\SL$ on $\Prop^0$ and $\Prop^1$,
+and state the following theorem:
+\begin{theorem}
+  $\SL[\Prop^0,\Prop^1]$ is decidable if ${\sf EPR}+\Prop^0+\Prop^1$
+  is decidable.
+\end{theorem}
+A useful instantiation of $\SL$ is $\SL[{\sf SLA},{\sf EPR}+{\sf
+SLA}]$, where ${\sf SLA}$ is the theory of simple linear arithmetic.
+{\sf EPR}+{\sf SLA} is known to be decidable~\cite{eprsla}, hence:
+\begin{theorem}
+  $\SL[{\sf SLA},{\sf EPR}]$ is decidable.
+\end{theorem}
+The $\SL[{\sf SLA},{\sf EPR}+{\sf SLA}]$ instantiation requires $I$ to
+be drawn from {\sf EPR}+{\sf SLA}, which is expressive enough to
+describe common database integrity constraints, such as referential
+integrity, non-negativeness of all integer values in a column etc.
+Isolation specifications presented in Sec.~\ref{sec:isolation} are
+already simple first-order formulas that can be encoded in {\sf EPR}.
+Furthermore, it is also reasonable to expect the guarantee ($G$) of a
+transaction to be expressible in the same logic as its inferred $\F$,
+since $\F$ (without the stability check) is essentially a complete
+characterization of the transaction, while $G$ is only an abstraction.
+Thus, with $\SL[{\sf SLA},{\sf EPR}+{\sf SLA}]$ as the language of
+inference, the verification problem for weakly isolated transactions
+is decidable. Moreover, off-the-shelf SMT solvers (e.g., Z3) are
+equipped with efficient decision procedures for ${\sf EPR}+{\sf SLA}$
+fragment, making the verification a practical exercise.

### 2017-07-06 introduction.tex
edits

@@
 question facing all database programmers.\footnote{This position has
-  been echoed by leading database researchers who lament the lack of
+  been echoed by leading database researchers who lament the lack of a
   better understanding of this problem.  See: {\tt
@@
 A major problem with weak isolation is that its semantics in the
-context of user programs is not well-understood. The original
+context of user programs is not precisely understood. The original
 proposal~\cite{gray1976} defines multiple ``degrees'' of weak
@@
 endeavor, with major database vendors~\cite{postgresiso, mysqliso,
-oracleiso} continuing to document their isolation levels primarily in
-terms of the undesirable phenomena a particular isolation level may
+  oracleiso} continuing to document their isolation levels primarily
+in terms of the undesirable phenomena a particular isolation level may
 induce, placing the burden on the programmer to determine application
@@
 transactional programs in the presence of a weakly-isolated
-concurrency control mechanism.  Realizing that the proof burden to use
-these rules would likely discourage applications programmers, we
-present an inference procedure based on a state transformer semantics
-that allows us to automatically verify the weakest isolation level for
-a transaction that ensures its consistency requirements are
-maintained.  The key to inference is our formulation of database state
-(represented as sets of tuples) as a monad, and in which database
-computations are interpreted as transformers over these sets.  This
-interpretation leads to an encoding of database computations amenable
-for verification by off-the-shelf SMT solvers.
+concurrency control mechanism.  Realizing that the proof burden
+imposed by these rules would likely discourage applications
+programmers from using them, we present an inference procedure based
+on a state transformer semantics that automatically verifies the
+weakest isolation level for a transaction that ensures its consistency
+requirements are maintained.  The key to inference is a novel formulation
+of database state (represented as sets of tuples) as a monad, and in
+which database computations are interpreted as transformers over these
+sets.  This interpretation leads to an encoding of database
+computations amenable for verification by off-the-shelf SMT solvers.
 
@@
    high-level consistency requirements.  The algorithm interprets
-   database operations as predicate state transformers leadig to a
-   natural encoding amenable for automated verification.
+   database operations as state transformers expressed in a language
+   amenable for translation into a decidable fragment of first-order
+   logic, and is thus suitable for automated verification using
+   off-the-shelf SMT solvers.
  \item A realistic implementation along with results on real-world
@@
 \noindent Our results provide the first (to the best of our knowledge)
-automated mechanism to verify the safety of high-level consistency
-conditions for weakly-isolated transactions, thereby allowing such
-transactions to enjoy the same rigorous reasoning capabilities as
-their strongly-isolated (serializable) counterparts.
+automated mechanism capable of verifying the safety of high-level
+consistency conditions for weakly-isolated database transactions,
+thereby allowing such transactions to enjoy the same rigorous
+reasoning capabilities as their strongly-isolated (serializable)
+counterparts.
 

### 2017-07-06 related-work.tex

@@
 \paragraph{Specifying weak isolation.}
-
 Adya~\cite{adyaphd} specifies several weak isolation levels in terms
@@
 \vspace*{-4pt}
-\paragraph{Reasoning under weak isolation} In~\cite{feketessi}, Fekete
+\paragraph{Reasoning under weak isolation.} In~\cite{feketessi}, Fekete
 \emph{et al.} propose a theory to characterize non-serializable
@@
 \vspace*{-4pt}
-\paragraph{Reasoning under weak consistency} There have been several
+\paragraph{Reasoning under weak consistency.} There have been several
 recent proposals to reason about programs executing under weak
 consistency~\cite{bailisvldb, alvarocalm, gotsmanpopl16,redblueatc,
-  redblueosdi, ecinec}. All of them assume a system model that offers
-a choice between a \emph{coordination-free} weak consistency level
+redblueosdi, ecinec}. All of them assume a system model that offers a
+choice between a \emph{coordination-free} weak consistency level
 (\emph{e.g.}, eventual consistency~\cite{redblueosdi, redblueatc,
@@
 invariants when executed under these consistency levels.  In contrast,
-we admit weakly-isolated transactions, and our system model accepts
-\emph{specifications} of consistency and isolation levels drawn from
-an expressive logic.  Gotsman \emph{et al.}~\cite{gotsmanpopl16}
-adapt \iso{Parallel Snapshot Isolation} to a transaction-less setting
-by interpreting it as a consistency level that serializes writes to
-objects; a dedicated proof rule is developed to help prove prove
-program invariants hold under this model. By parameterizing our proof
-system over a gamut of weak isolation specifications, we avoid the
-need to define a separate proof rule for each new isolation level we
-may encounter.
+our focus in on reasoning in the presence of weakly-isolated
+transactions under a strongly consistent store.  Gotsman \emph{et
+al.}~\cite{gotsmanpopl16} adapt \iso{Parallel Snapshot Isolation} to a
+transaction-less setting by interpreting it as a consistency level
+that serializes writes to objects; a dedicated proof rule is developed
+to help prove prove program invariants hold under this model. By
+parameterizing our proof system over a gamut of weak isolation
+specifications, we avoid the need to define a separate proof rule for
+each new isolation level we may encounter.
+
+%% \vspace*{-4pt}
+%% \paragraph{Reasoning under relaxed memory.} The reasoning mechanisms
+%% used to describe and prove properties about weakly-isolated
+%% transactions bear some resemblance to those used to formalize relaxed
+%% memory behaviour~\cite{battycpp}.  Ridge~\cite{rgtso} generalizes
+%% rely-guarantee reasoning to the x86-TSO memory model.  Likewise,
+%% Vafeiadis \emph{et al.}~\cite{rsl13} generalize concurrent separation
+%% logic (CSL)~\cite{csl} to the C11 relaxed memory model.  Ferreira
+%% \emph{et al.}~\cite{ferreira10} propose a parameterized operational
+%% semantics for relaxed memory models, but the parameterization is over
+%% a relation between relaxed memory programs and related {\sc sc}
+%% programs. Demange \emph{et al.}~\cite{DLZ+13} present a \emph{buffered
+%%   memory model} for Java that defines an axiomatic definition for the
+%% JMM in terms of memory reorderings, and an operational instantiation
+%% consistent with the TSO memory model.
 
 \vspace*{-4pt}
-\paragraph{Reasoning under relaxed memory} The reasoning mechanisms
-used to describe and prove properties about weakly-isolated
-transactions bear some resemblance to those used to formalize relaxed
-memory behaviour~\cite{battycpp}.  Ridge~\cite{rgtso} generalizes
-rely-guarantee reasoning to the x86-TSO memory model.  Likewise,
-Vafeiadis \emph{et al.}~\cite{rsl13} generalize concurrent separation
-logic (CSL)~\cite{csl} to the C11 relaxed memory model.  Ferreira
-\emph{et al.}~\cite{ferreira10} propose a parameterized operational
-semantics for relaxed memory models, but the parameterization is over
-a relation between relaxed memory programs and related {\sc sc}
-programs. Demange \emph{et al.}~\cite{DLZ+13} present a \emph{buffered
-  memory model} for Java that defines an axiomatic definition for the
-JMM in terms of memory reorderings, and an operational instantiation
-consistent with the TSO memory model.
+\paragraph{Inference.}  Prior work by Vafeiadis~\cite{Vaf10,Vaf10a}
+describe \emph{action inference}, an inference procedure for computing
+rely and guarantee relations in the conetxt of RGSep~\cite{VP07}, an
+integration of rely-guarantee and separation logic~\cite{Rey02} that
+allows one to precisely reason about local and shared state of a
+concurrent program. The ideas underlying action inference have been
+used to prove memory safety, linearizability, shape invariant
+inference, etc.  of fine-grained concurrent data structures.  While
+our motivation is similar to theirs (inference of intermediate
+assertions and invariants), the context of study (transactions
+vs. shared-memory concurrency), the objects being analyzed (relational
+database tables vs. concurrent data structures), and the analysis
+technique used to drive inference (state transformers vs. abstract
+interpretation) are quite different.

### 2017-07-06 inference.tex
Sec. 5 weakening example.

@@
 \begin{array}{lcl}
-  \stable(\R,\F) & \Leftrightarrow & \forall \stl,\stg,\stg'.~
+  \stable(\R,\F) & \Leftrightarrow & \forall \stl,\stg,\stg',\bar{v}.~
   \R(\stl,\stg,\stg') \Rightarrow \F(\stl,\stg) \equiv \F(\stl,\stg')
@@
 \end{smathpar}
-Intuitively, the stability condition requires an effect to describe
-the same set of records before and after the interference. The
-equivalence in $\SL$ translates to the equality in first-order logic,
-as will be explained shortly. In the inference rules, stability is
-enforced constructively by a meta function $\stabilize{\cdot}$, which
-accepts an effect and returns a new effect that is guaranteed to be
-stable under $\R$. $\stabilize{\cdot}$ achieves the stability
-guarantee by abstracting away the bound $\stg$ in an unstable $\F$ to
-an existentially bound $\stg'$ as described below:
+where, $\bar{v}$ are the variables that occur free in $\F$ (possible
+because of how the inference rules are structured).  Intuitively, the
+stability condition requires an effect to describe the same set of
+records before and after the interference. The equivalence in $\SL$
+translates to the equality in first-order logic, as will become clear
+shortly. In the inference rules, stability is enforced constructively
+by a meta function $\stabilize{\cdot}$, which accepts an effect and
+returns a new effect that is guaranteed to be stable under $\R$.
+$\stabilize{\cdot}$ achieves the stability guarantee by abstracting
+away the bound $\stg$ in an unstable $\F$ to an existentially bound
+$\stg'$ as described below:
 \begin{smathpar}
@@
 $\F'$ that simply ignores its $\stg$ argument in favor of a generic
-$\stg'$, making $\F'$ trivially stable. The existentially bound
-$\stg'$ satisfies $I$ since transactions are always executed against a
-valid database state.
+$\stg'$, making $\F'$ trivially stable. It is safe to assume
+$I(\stg')$ because all verified transactions preserve the invariant,
+hence only valid database states will ever be witnessed. From the
+perspective of RG reasoning, $\stabilize{\cdot}$ effectively weakens
+the post-condition of a statement, as done by the
+\rulelabel{RG-Conseq} rule for transaction-bound commands (cf.  the
+supplementary material). The weakening semantics chosen by
+$\stabilize{\cdot}$, while being simple, is nonetheless useful because
+of the $I(\stg')$ assumption on the existentially bound $\stg'$. The
+example in Fig.~\ref{fig:weakening-example} demonstrates. 
+\begin{figure}[h]
+\begin{ocaml}
+let add_interest acc_id pc = atomically @@ do
+  let a = SQL.select1 BankAccount (fun acc -> acc.id = acc_id);
+  let y = a.bal + pc*a.bal;
+  SQL.update BankAccount (fun acc -> {acc with bal = acc.bal + y})
+                         (fun acc -> acc.id = acc_id)
+\end{ocaml}
+\caption{A transaction that deposits an interest to a bank account}
+\label{fig:weakening-example}
+\end{figure}
+Here, an \C{add\_interest} transaction adds a positive interest to the
+balance of a bank account, which is required to be positive ($I(\stg)
+\Leftrightarrow \forall(r\in\stg).~r.\C{bal}\ge 0$). The transaction
+starts by issuing a \C{select1} query, whose effect implicitly asserts
+that there exists a record in the bank account database that is equal
+to $a$ ($\exists(r\in\stg).~r=a$). However, this assertion is unstable
+because a concurrent \C{withdraw} or \C{deposit} transactions might
+update the account balance, and there will no longer exist an
+$r\in\stg$ that is equal to $a$. Fortunately, the weakening operator
+($\stabilize{\cdot}$) comes to the rescue here by weakening the
+assertion to $\exists \stg'.I(\stg') \conj \exists(r\in\stg').~a=r$,
+which is enough to prove that $\C{a.bal + pc*a.bal}\ge 0$, and verify
+that \C{add\_interest} preserves the positive balance invariant.
 

### 2017-07-07 abstract.tex
edits to inference

@@
   Serializability is a well-understood correctness criterion that
-  simplifies reasoning about the behaviour of concurrent transactions
+  simplifies reasoning about the behavior of concurrent transactions
   by ensuring they are \emph{isolated} from each other while they
@@
   that weakly-isolated transactions can be viewed as functional
-  (monadic) computations over database state, allowing us to treat
-  their operations as state transformers over the database.  This
-  interpretation enables automated verification using off-the-shelf
-  SMT solvers.
+  (monadic) computations over an abstract database state, allowing us
+  to treat their operations as state transformers over the database.
+  This interpretation enables automated verification using
+  off-the-shelf SMT solvers.
 

### 2017-07-07 inference.tex
Edit.
edits to inference

@@
 
-Rely-Guarantee framework presented in the previous section facilitates
-modular proofs for weakly-isolated transactions, but imposes a
-non-trivial annotation burden on the programmer. In particular, it
-requires each statement ($c$) of the transaction to be annotated with
-a stable pre- ($P$) and post-condition ($Q$), and loops to be
-annotated with stable inductive invariants ($\psi$). While weakest
-pre-condition style predicate transformers can help in inferring $P$'s
-and $Q$'s for regular statements, inferring inductive loop invariants
-is still a challenge, even for the simpler form of loops we admit.
-Thus, predicate transformer-based approach to proof automation is
-unlikely to be effective in this case. We therefore present an
-alternative approach to RG proof automation in this section that is
-based on state transformers instead of predicate transformers. The
-idea is to infer the effect that each statement has on the
-transaction-local database state $\stl$ (i.e., how it transforms
-$\stl$), and compose multiple such effects together to describe the
-effect of the transaction as a whole. Importantly, this approach
-generalizes to loops, where the effect of a loop can be inferred as a
-well-defined function of the effect of its body, thanks to certain
-pleasant properties enjoyed by the database programs modeled by
-$\txnimp$.
+The Rely-Guarantee framework presented in the previous section
+facilitates modular proofs for weakly-isolated transactions, but
+imposes a non-trivial annotation burden.  In particular, it requires
+each statement ($c$) of the transaction to be annotated with a stable
+pre- ($P$) and post-condition ($Q$), and loops to be annotated with
+stable inductive invariants ($\psi$). While weakest pre-condition
+style predicate transformers can help in inferring intermediate
+assertions for regular statements, loop invariant inference remains
+challenging, even for the simple form of loops considered here.  As an
+alternative, we present an inference algorithm based on state
+transformers.  The idea is to infer the logical effect that each
+statement has on the transaction-local database state $\stl$ (i.e.,
+how it transforms $\stl$), and compose multiple such effects together
+to describe the effect of the transaction as a whole.  Importantly,
+this approach generalizes to loops, where the effect of a loop can be
+inferred as a well-defined function of the effect of its body, thanks
+to certain pleasant properties enjoyed by the database programs
+modeled by $\txnimp$.  Interpreting database semantics as functional
+transformations on sets (described in terms of their logical effects)
+enables an inference mechanism that can leverage off-the-shelf SMT
+solvers for automated verification.
 
@@
 At the core of our approach is a simple language ($\SL$) to express
-set transformations shown in Fig.~\ref{fig:logic-syntax}. The language
-admits only set expressions that include variables ($x$), literals of
-the form $\{x \,|\, \phi^0\}$ where $\phi^0$ is a propositional
+set transformations (see \ref{fig:logic-syntax}). The language admits
+only set expressions that include variables ($x$), literals of the
+form $\{x \,|\, \varphi\}$ where $\varphi$ is a propositional
 (quantifier-free) formula on $x$, a restricted form of existential
-quantification that binds a set $x$ satisfying proposition $\phi^0$ in
+quantification that binds a set $x$ satisfying proposition $\phi$ in
 a set expression $s$, a monadic composition of two set expressions
-($s_1$ and $s_2$) using bind ($\bind$), a conditional set expression
-where the condition is a propositional formula, and a union of two set
-expressions. Symbols $\stl$ and $\stg$ are also variables in $\SL$,
-but are used to denote local and database states, respectively.  The
-language is carefully chosen to be expressive enough to capture the
-semantics of $\txnimp$ statements, yet simple enough to have a
-semantics-preserving translation to the first-order logic.
+($s_1$ and $s_2$) composed using a bind ($\bind$) operation, a
+conditional set expression where the condition is a propositional
+formula, and a union of two set expressions. Symbols $\stl$ and $\stg$
+are also variables in $\SL$, but are used to denote local and database
+states, respectively.  The language is carefully chosen to be
+expressive enough to capture the semantics of $\txnimp$ statements
+(and is representative of SQL semantics), yet simple enough to have a
+semantics-preserving translation to a decidable fragment of
+first-order logic.
 
@@
 
-Fig.~\ref{fig:inference-rules} shows the syntax-directed inference
-rules for $\txnimp$ commands inside a transaction $\C{TXN}_i$.
-Inference depends on the $\I$-constrained rely relation $\R$, and the
-high-level invariant $I$ for the reasons that will become clear
-shortly.  Inference rules compute, for each command $c$, a (meta)
-function $\F$ that returns a set as an expression in $\SL$, given a
-pair of sets ($\stl$ and $\stg$) that describe local and a global
-database states, respectively. The expression returned by
-$\F(\stl,\stg)$ (abstractly) describes the set of records that get
-added to $\stl$ as a result of executing $c$ under $\stl$ and $\stg$.
-Thus, $\F$ captures the \emph{effect} part of the state transformer of
-$c$, which is the function $\lambda(\stl,\stg).~\stl \cup
-\F(\stl,\stg)$. For $\F$ to be useful in RG verification, it needs to
-be stable w.r.t the rely relation $\R$. The stability condition on
-effects is defined below:
+Fig.~\ref{fig:inference-rules} shows the syntax-directed state
+transformer inference rules for $\txnimp$ commands inside a
+transaction $\C{TXN}_i$.  Inference depends on the $\I$-constrained
+rely relation $\R$, and the high-level invariant $I$ for the reasons
+that will become clear shortly.  Inference rules compute, for each
+command $c$, a (meta) function $\F$ that returns a set as an
+expression in $\SL$, given a pair of sets ($\stl$ and $\stg$) that
+describe local and a global database states, respectively. The
+expression returned by $\F(\stl,\stg)$ (abstractly) describes the set
+of records that get added to $\stl$ as a result of executing $c$ under
+$\stl$ and $\stg$.  Thus, $\F$ captures the \emph{effect} part of the
+state transformer of $c$, which is the function
+$\lambda(\stl,\stg).~\stl \cup \F(\stl,\stg)$\footnote{Recall that the
+  operational semantics treats deletion of records as the addition of
+  the deleted record with its \C{del} field set to true in the local
+  store.}. For $\F$ to be useful in RG verification, it needs to be
+stable w.r.t the rely relation $\R$. The stability condition on
+effects can be defined thus:
 \begin{smathpar}
@@
 \end{smathpar}
-where, $\bar{v}$ are the variables that occur free in $\F$ (possible
-because of how the inference rules are structured).  Intuitively, the
-stability condition requires an effect to describe the same set of
-records before and after the interference. The equivalence in $\SL$
-translates to the equality in first-order logic, as will become clear
-shortly. In the inference rules, stability is enforced constructively
-by a meta function $\stabilize{\cdot}$, which accepts an effect and
-returns a new effect that is guaranteed to be stable under $\R$.
-$\stabilize{\cdot}$ achieves the stability guarantee by abstracting
-away the bound $\stg$ in an unstable $\F$ to an existentially bound
-$\stg'$ as described below:
+where, $\bar{v}$ are the variables that occur free in $\F$; this is
+possible because of how the inference rules are structured.
+Intuitively, the stability condition requires an effect to describe
+the same set of records before and after the interference. The
+equivalence in $\SL$ translates to the equality in first-order logic,
+as we describe below. In the inference rules, stability is
+enforced constructively by a meta function $\stabilize{\cdot}$, which
+accepts an effect and returns a new effect that is guaranteed to be
+stable under $\R$.  $\stabilize{\cdot}$ achieves the stability
+guarantee by abstracting away the bound $\stg$ in an unstable $\F$ to
+an existentially bound $\stg'$ as described below:
 \begin{smathpar}
@@
 \end{smathpar}
-Observe that when $\F$ is not stable, then $\stabilize{\F}$ returns an
-$\F'$ that simply ignores its $\stg$ argument in favor of a generic
+Observe that when $\F$ is not stable, then $\stabilize{\F}$ returns a
+transformer $\F'$ that simply ignores its $\stg$ argument in favor of a generic
 $\stg'$, making $\F'$ trivially stable. It is safe to assume
 $I(\stg')$ because all verified transactions preserve the invariant,
-hence only valid database states will ever be witnessed. From the
+and hence only valid database states will ever be witnessed. From the
 perspective of RG reasoning, $\stabilize{\cdot}$ effectively weakens
 the post-condition of a statement, as done by the
-\rulelabel{RG-Conseq} rule for transaction-bound commands (cf.  the
-supplementary material). The weakening semantics chosen by
+\rulelabel{RG-Conseq} rule for transaction-bound commands.  The weakening semantics chosen by
 $\stabilize{\cdot}$, while being simple, is nonetheless useful because
@@
 \end{ocaml}
-\caption{A transaction that deposits an interest to a bank account}
+\caption{A transaction that deposits an interest to a bank account.}
 \label{fig:weakening-example}
 \end{figure}
-Here, an \C{add\_interest} transaction adds a positive interest to the
-balance of a bank account, which is required to be positive ($I(\stg)
-\Leftrightarrow \forall(r\in\stg).~r.\C{bal}\ge 0$). The transaction
-starts by issuing a \C{select1} query, whose effect implicitly asserts
-that there exists a record in the bank account database that is equal
-to $a$ ($\exists(r\in\stg).~r=a$). However, this assertion is unstable
+Here, an \C{add\_interest} transaction adds a positive interest
+(\C{pc}) to the balance of a bank account, which is required to be
+positive ($I(\stg) \Leftrightarrow \forall(r\in\stg).~r.\C{bal}\ge
+0$). The transaction starts by issuing a \C{select1} query, whose
+effect implicitly asserts that there exists a record in the bank
+account database that is equal to $a$
+($\exists(r\in\stg).~r=a$). However, this assertion is unstable
 because a concurrent \C{withdraw} or \C{deposit} transactions might
-update the account balance, and there will no longer exist an
-$r\in\stg$ that is equal to $a$. Fortunately, the weakening operator
-($\stabilize{\cdot}$) comes to the rescue here by weakening the
-assertion to $\exists \stg'.I(\stg') \conj \exists(r\in\stg').~a=r$,
-which is enough to prove that $\C{a.bal + pc*a.bal}\ge 0$, and verify
-that \C{add\_interest} preserves the positive balance invariant.
+update the account balance, so that such a record no longer exists in
+the database.  Fortunately, the weakening operator
+($\stabilize{\cdot}$) allows us to weaken the assertion to $\exists
+\stg'.I(\stg') \conj \exists(r\in\stg').~a=r$ (Fig.~\ref{fig:logic}
+formalizes the encoding of {\sf exists} to logic), which is enough to
+prove that $\C{a.bal + pc*a.bal}\ge 0$, and verify that
+\C{add\_interest} preserves the positive balance invariant.
 
-The inference rules of SQL statements, like the RG rules, closely
-follow the corresponding reduction rules in Fig.~\ref{fig:txnimp},
-except that their language of expression is $\SL$. For instance, while
-the reduction rule for \C{UPDATE} declaratively specifies the set of
-updated records, the inference rule uses $\SL$'s bind operation to
-\emph{compute} the set. Other SQL rules do likewise. The inference
+The state transformer rules, like the earlier RG rules, closely follow
+the corresponding reduction rules in Fig.~\ref{fig:txnimp}, except
+that their language of expression is $\SL$. For instance, while the
+reduction rule for \C{UPDATE} declaratively specifies the set of
+updated records, the state transformer rule uses $\SL$'s bind
+operation to \emph{compute} the set. Other SQL rules do likewise. The
 rules for \C{LET} binders, conditionals, and sequences compose the
@@
 state ($\stl$ and $\stg$). That is, $\F_{loop}(\stl,\stg) =
-\bigcup_{z\in x}\F_{body}(\stl,\stg)$. From the definition of
-set monad's bind, $\F_{loop}(\stl,\stg) = x \bind (\lambda
-z.~F_{body}(\stl,\stg))$, which is what the rule infers.
+\bigcup_{z\in x}\F_{body}(\stl,\stg)$. From the definition of the set
+monad's bind operator, $\F_{loop}(\stl,\stg) = x \bind (\lambda
+z.~F_{body}(\stl,\stg))$, which mirrors the definition of the rule.
 
@@
 
-\begin{example}
-Consider a database of couples that tracks the spouse
-relationship. Each record has an $\idf$ field, a \C{name} field, and a
-spouse id (\C{s\_id}) field. A $\txnimp$ transaction bearing id $i$
-that populates the database with a given set ($xs$) of couples will
-include the following code snippet:
-\begin{smathpar}
-\begin{array}{l}
-  \foreache{xs}{(\lambda y.\lambda x.~ 
-        \updatee{(\lambda z.~ \{\idf=z.\idf;\,\C{name}=z.\C{name};\,
-                               \C{s\_id}=x.\idf\})}
-                {(\lambda z.~ z.\idf = x.\C{s\_id})};\\
-        \hspace*{1.27in} \inserte{x})}
-\end{array}
-\end{smathpar}
-The effect inferred for the program, assuming no interference, is as
-following:
-\begin{smathpar}
-\begin{array}{l}
-  \lambda(\stl,\stg).~xs \bind 
-    (\lambda x. \stg \bind 
-      (\lambda z. \itel{z.\idf = x.\C{s\_id}}
-                       {\{\{\idf=z.\idf;\,\C{name}=z.\C{name};\,
-                               \C{s\_id}=x.\idf\}\}}
-                       {\emptyset}) \\
-      \hspace*{1cm} ~\cup~ \{y \,|\, y = \{x \with \delf=false; \}\})
-\end{array}
-\end{smathpar}
-Under the possibility of an interference affecting the stability, the
-following stable effect is inferred ($I$ is the database invariant):
-\begin{smathpar}
-\begin{array}{l}
-  \lambda(\stl,\stg).~xs \bind 
-    (\lambda x.~\existsl(\stg',I,\stg' \bind 
-      (\lambda z. \itel{z.\idf = x.\C{s\_id}}
-                       {\{\{\idf=z.\idf;\,\C{name}=z.\C{name};\,
-                               \C{s\_id}=x.\idf\}\}}
-                       {\emptyset})) \\
-      \hspace*{1cm} ~\cup~ \{y \,|\, y = \{x \with \delf=false; \}\})
-\end{array}
-\end{smathpar}
-\end{example}
+%% \begin{example}
+%% Consider a database of couples that tracks the spouse
+%% relationship. Each record has an $\idf$ field, a \C{name} field, and a
+%% spouse id (\C{s\_id}) field. A $\txnimp$ transaction bearing id $i$
+%% that populates the database with a given set ($xs$) of couples will
+%% include the following code snippet:
+%% \begin{smathpar}
+%% \begin{array}{l}
+%%   \foreache{xs}{(\lambda y.\lambda x.~ 
+%%         \updatee{(\lambda z.~ \{\idf=z.\idf;\,\C{name}=z.\C{name};\,
+%%                                \C{s\_id}=x.\idf\})}
+%%                 {(\lambda z.~ z.\idf = x.\C{s\_id})};\\
+%%         \hspace*{1.27in} \inserte{x})}
+%% \end{array}
+%% \end{smathpar}
+%% The effect captured by a state transformer inferred for the program, assuming no interference would be:
+%% \begin{smathpar}
+%% \begin{array}{l}
+%%   \lambda(\stl,\stg).~xs \bind 
+%%     (\lambda x. \stg \bind 
+%%       (\lambda z. \itel{z.\idf = x.\C{s\_id}}
+%%                        {\{\{\idf=z.\idf;\,\C{name}=z.\C{name};\,
+%%                                \C{s\_id}=x.\idf\}\}}
+%%                        {\emptyset}) \\
+%%       \hspace*{1cm} ~\cup~ \{y \,|\, y = \{x \with \delf=\mathit{false}; \}\})
+%% \end{array}
+%% \end{smathpar}
+%% Under the possibility of an interference affecting the stability, the
+%% following stable effect is inferred ($I$ is the database invariant):
+%% \begin{smathpar}
+%% \begin{array}{l}
+%%   \lambda(\stl,\stg).~xs \bind 
+%%     (\lambda x.~\existsl(\stg',I,\stg' \bind 
+%%       (\lambda z. \itel{z.\idf = x.\C{s\_id}}
+%%                        {\{\{\idf=z.\idf;\,\C{name}=z.\C{name};\,
+%%                                \C{s\_id}=x.\idf\}\}}
+%%                        {\emptyset})) \\
+%%       \hspace*{1cm} ~\cup~ \{y \,|\, y = \{x \with \delf=false; \}\})
+%% \end{array}
+%% \end{smathpar}
+%% \end{example}
 
@@
 The correspondence between the inference rules and the RG judgment of
-Sec.~\ref{sec:reasoning} is stated thus\footnote{Proof can be found in
-the supplementary material}:
+Sec.~\ref{sec:reasoning} is stated thus\footnote{The proof can be found in
+the supplementary material.}:
 \begin{theorem}
 \label{thm:inference-sound}
-  Forall i,$R$,$I$,$c$,$\F$, if $\stable(\R,I)$ and $c \elabsto \F$,
+  \emph{Forall} i,$R$,$I$,$c$,$\F$, if $\stable(\R,I)$ and $c \elabsto \F$,
   then:\\\vspace*{-0.2cm}
@@
 \end{theorem}
-Where the set expression $\F(\emptyset,\stg)$ has a natural
-interpretation, on the lines discussed in the next sub-section.
+\noindent where the set expression $\F(\emptyset,\stg)$ has a natural
+interpretation as discussed below.
 % Where, ${\sf true} = \lambda(\stl,\stg).~true$ and $\stable(\R,I)
@@
 Theorem~\ref{thm:inference-sound} lets us replace the local judgment
-of the \rulelabel{RG-Txn} rule (Fig.~\ref{fig:rg-rules}) by the
-inference judgment. The soundness of a transaction's guarantee can now
+of the \rulelabel{RG-Txn} rule (Fig.~\ref{fig:rg-rules}) by a state
+transformer inference judgment. The soundness of a transaction's guarantee can now
 be established w.r.t the effect $\F$ of the body. The
-\rulelabel{RG-Txn} rule updated to use the inference judgment is shown
+\rulelabel{RG-Txn} rule so updated is shown
 below:
@@
 \end{smathpar}
-Since the inference judgment $c\elabsto \F$ is fully automatic,
-automating the application of the \rulelabel{RG-Txn} rule for a
-transaction now requires automating the multiple implication checks in
+Automating the application of the \rulelabel{RG-Txn} rule for a
+transaction requires automating the multiple implication checks in
 the premise. While $R$, $\R$, $\R_c$, $\I$ and $\I$ are formulas in
-the first-order logic (FOL) with a relatively simple structure, $\F$
+first-order logic (FOL) with a relatively simple structure, $\F$
 is an expression in the set language $\SL$
@@
 The algorithm ($\mssemof{\cdot}$) shown in Fig.~\ref{fig:logic}
-encodes an $\SL$ expression in a FOL. Given a set expression $s$ of
-$\SL$, $\mssemof{s}$ is a pair $(\phi,f)$, where $\phi$ encodes $s$ in
-FOL, and $f$ is a boolean function that names the set defined by $s$.
-Naming provides an easy handle to $s$, simplifying $\mssemof{\cdot}$.
-Set variables ($x$, $\stl$, $\stg$, $\dots$) are encoded as
-unconstrained boolean functions ($g_x$, $g_{\stl}$, $g_{\stg}$,
-$\ldots$) that are named after the variable.  Encoding of a set
-literal $\{a\,|\,\varphi\}$ introduces a new boolean function $g$ that
-is $true$ for only those $a$ that satisfy $\varphi$.  Thus $g$
-faithfully encodes the literal. The encoding of the $\existsl{}$
-expression first eliminates the second-order quantification over sets.
-Existentially bound set variable ($a$) is skolemized, i.e.,
-instantiated with a new unbound variable $a'$, before encoding the
-bound expression $s$ and the first-order constraint $\phi$. The
-algorithm for encoding $\phi$, which is already a first-order formula,
-is straightforward. It basically involves replacing set variables with
-boolean functions named after the variables. The first-order encoding
-for a bind expression describes the semantics of set monad's $\bind$
-in FOL. Let $s_1$ be a set, and let $f$ be a function that maps each
-variable in $s_1$ to a new set. Then, $s_2 = s_1 \bind f$ if and only
-if for all $y\in s_2$, there exists an $x \in s_1$ such that $y =
-f(x)$, and forall $x\in s_1$, $f(x)\in s_2$. The encoding adds new
-constraints to this effect. Conditional set expressions and set union
-are encoded straightforwardly.
+encodes an $\SL$ expression in a decidable fragment of FOL. Given a
+set expression $s$ of $\SL$, $\mssemof{s}$ is a pair $(\phi,f)$, where
+$\phi$ encodes $s$ in FOL, and $f$ is a Boolean function that names
+the set defined by $s$.  Naming provides an easy handle to $s$,
+simplifying $\mssemof{\cdot}$.  Set variables ($x$, $\stl$, $\stg$,
+$\dots$) are encoded as unconstrained Boolean functions ($g_x$,
+$g_{\stl}$, $g_{\stg}$, $\ldots$) that are named after the variable.
+Encoding of a set literal $\{a\,|\,\varphi\}$ introduces a new Boolean
+function $g$ that is \emph{true} for only those $a$ that satisfy $\varphi$.
+Thus $g$ faithfully encodes the literal. The encoding of the
+$\existsl{}$ expression first eliminates the second-order
+quantification over sets.  Existentially bound set variable ($a$) is
+skolemized, i.e., instantiated with a new unbound variable $a'$,
+before encoding the bound expression $s$ and the first-order
+constraint $\phi$. The algorithm for encoding $\phi$, which is already
+a first-order formula, is straightforward. It basically involves
+replacing set variables with Boolean functions named after the
+variables. The first-order encoding for a bind expression describes
+the semantics of set monad's bind operator in FOL. Let $s_1$ be a set, and
+let $f$ be a function that maps each variable in $s_1$ to a new
+set. Then, $s_2 = s_1 \bind f$ if and only if for all $y\in s_2$,
+there exists an $x \in s_1$ such that $y = f(x)$, and forall $x\in
+s_1$, $f(x)\in s_2$. The encoding adds new constraints to this
+effect. Conditional set expressions and set union are encoded
+straightforwardly.
 
-Observe that the language of encoding shown in Fig.~\ref{fig:logic} is
+Observe that the encoding shown in Fig.~\ref{fig:logic} maps to
 a subset of logic that satisfies the following syntactic properties:
@@
   \item All function symbols, modulo those that might appear in
-    $\varphi\in\Prop^0$, are uninterpreted and boolean.
+    $\varphi\in\Prop^0$, are uninterpreted and Boolean.
 \end{itemize}
-The fragment of FOL that satisfies the above syntactic conditions is
-called Effectively Propositional first-order logic (denoted {\sf
-EPR}), or Bernay-Sch{\"o}nfinkel-Ramsey (BSR) logic, which is known to
-be decidable~\cite{z3epr}. The language of encoding, however, is a
+This fragment of FOL known as EPR (Effectively Propositional logic)
+is known to be decidable~\cite{z3epr}. The language of encoding, however, is a
 combination of {\sf EPR} with (a). $\Prop^0$, the theory from which
@@
 SLA}]$, where ${\sf SLA}$ is the theory of simple linear arithmetic.
-{\sf EPR}+{\sf SLA} is known to be decidable~\cite{eprsla}, hence:
+Since {\sf EPR}+{\sf SLA} is known to be decidable~\cite{eprsla}, hence:
 \begin{theorem}
-  $\SL[{\sf SLA},{\sf EPR}]$ is decidable.
+  $\SL[{\sf SLA},{\sf EPR}+{\sf SLA}]$ is decidable.
 \end{theorem}
@@
 integrity, non-negativeness of all integer values in a column etc.
-Isolation specifications presented in Sec.~\ref{sec:isolation} are
+The isolation specifications presented in Sec.~\ref{sec:isolation} are
 already simple first-order formulas that can be encoded in {\sf EPR}.
@@
 is decidable. Moreover, off-the-shelf SMT solvers (e.g., Z3) are
-equipped with efficient decision procedures for ${\sf EPR}+{\sf SLA}$
-fragment, making the verification a practical exercise.
+equipped with efficient decision procedures for ${\sf EPR}+{\sf SLA}$,
+making automated verification a practical exercise.

### 2017-07-07 related-work.tex
edits to inference

@@
 describe \emph{action inference}, an inference procedure for computing
-rely and guarantee relations in the conetxt of RGSep~\cite{VP07}, an
+rely and guarantee relations in the context of RGSep~\cite{VP07}, an
 integration of rely-guarantee and separation logic~\cite{Rey02} that
@@
 inference, etc.  of fine-grained concurrent data structures.  While
-our motivation is similar to theirs (inference of intermediate
-assertions and invariants), the context of study (transactions
+our motivation is similar (automated inference of intermediate
+assertions and local invariants), the context of study (transactions
 vs. shared-memory concurrency), the objects being analyzed (relational
-database tables vs. concurrent data structures), and the analysis
-technique used to drive inference (state transformers vs. abstract
-interpretation) are quite different.
+database tables vs. concurrent data structures), the properties being
+verified (integrity constraints over relational tables vs. memory
+safety, or linearizability of concurrent data strucgture operations)
+and the analysis technique used to drive inference (state transformers
+vs. abstract interpretation) are quite different.

### 2017-07-07 opsem.tex
Opsem txnid added to local reduction.
Edits.

@@
 transactions ($\ctxn{i}{\I}{c}$) and their parallel composition
-($c_1\,||\,c_2$). The $\I$ in the \C{TXN} block syntax is the
+($c_1\,||\,c_2$). Each transaction is assumed to have a unique
+identifier $i$. The $\I$ in the \C{TXN} block syntax is the
 transaction's isolation specification, which will be explained later.
@@
 they all belong to a single table.  The generalization to multiple
-tables is straightforward.  Records in $\stg$ are uniquely identifiable
-through their $\idf$ field, which is auto-generated and does not
-belong to the surface language, i.e., $\C{id}\notin f$. For a set $S$
+tables is straightforward. The semantics assumes that the records in
+$\stg$ are uniquely identifiable through their $\idf$ field, and
+enforces this invariant wherever necessary. Certain hidden fields are
+treated specially by the operational semantics, and they are hidden
+from the surface language. These include a $\txnf$ field that tracks
+the identifier of the transaction that last updated the record, and a
+$\delf$ field that flags the deleted records in $\stl$.  For a set $S$
 of records, we define $\dom(S)$ as the set of unique ids of all
@@
 records\footnote{While SQL's \C{UPDATE} admits writes at the
-  granularity of record fields, databases, in reality, enforce
+  granularity of record fields, most popular databases enforce
   record-level locking, allowing us to think of ``uncommitted writes''
   as ``uncommitted records''. }. Uncommitted records include deleted
-records, which are identified by a hidden $\delf$ field set to
-\C{true}. When the transaction commits, its local database is
-atomically \emph{flushed} to the global database, committing these
-heretofore uncommitted records. The flush operation ($\gg$) is defined as
-follows:
+records, whose $\delf$ field is set to \C{true}. When the transaction
+commits, its local database is atomically \emph{flushed} to the global
+database, committing these heretofore uncommitted records. The flush
+operation ($\gg$) is defined as follows:
 \begin{smathpar}
@@
 statements is defined straightforwardly. \C{INSERT} adds a new record
-to $\stl$ after adding a unique identifier. \C{DELETE} finds the records in $\stg$ that match
-the search criteria defined by its boolean function argument, and adds
-the records to $\stl$ after marking them for deletion. \C{SELECT}
-bounds the name introduced by \C{LET} to the set of records from
-$\stg$ that match the search criteria, and then executes the bound
-command $c$. \C{UPDATE} uses its first function argument to compute
-the updated version of the records that match the search criteria
-defined by its second function argument. Updated records are added to
-$\stl$.
+to $\stl$ after checking the uniqueness of its id. \C{DELETE} finds
+the records in $\stg$ that match the search criteria defined by its
+boolean function argument, and adds the records to $\stl$ after
+marking them for deletion. \C{SELECT} bounds the name introduced by
+\C{LET} to the set of records from $\stg$ that match the search
+criteria, and then executes the bound command $c$. \C{UPDATE} uses its
+first function argument to compute the updated version of the records
+that match the search criteria defined by its second function
+argument. Updated records are added to $\stl$.
 

### 2017-07-07 inference.tex

@@
 At the core of our approach is a simple language ($\SL$) to express
-set transformations (see \ref{fig:logic-syntax}). The language admits
-only set expressions that include variables ($x$), literals of the
-form $\{x \,|\, \varphi\}$ where $\varphi$ is a propositional
+set transformations (see Fig.~\ref{fig:logic-syntax}). The language
+admits only set expressions that include variables ($x$), literals of
+the form $\{x \,|\, \varphi\}$ where $\varphi$ is a propositional
 (quantifier-free) formula on $x$, a restricted form of existential
-quantification that binds a set $x$ satisfying proposition $\phi$ in
-a set expression $s$, a monadic composition of two set expressions
+quantification that binds a set $x$ satisfying proposition $\phi$ in a
+set expression $s$, a monadic composition of two set expressions
 ($s_1$ and $s_2$) composed using a bind ($\bind$) operation, a
@@
 states, respectively.  The language is carefully chosen to be
-expressive enough to capture the semantics of $\txnimp$ statements
-(and is representative of SQL semantics), yet simple enough to have a
+expressive enough to capture the semantics of $\txnimp$ statements (as
+well as SQL operations more generally), yet simple enough to have a
 semantics-preserving translation to a decidable fragment of

### 2017-07-07 isolation.tex
Edits.

@@
 implementation. This allows us to model a range of isolation
-properties that are relevant to the theory and practice of realistic
-transaction processing systems, some of the most well-known of which are
-given below:
+properties that are relevant to the theory and practice of transaction
+processing systems. Few such well-known properties are discussed
+below:
 
@@
 (Sec.~\ref{sec:motivation}) demonstrates, enforcing global uniqueness
-of ordered identifiers requires stronger isolation than possible using
-a weak read committed isolation level. Alternatively, globally unique
+of ordered identifiers requires stronger isolation levels than ones
+that are default on most databases. Alternatively, globally unique
 sequence numbers, regardless of the isolation level, can be requested
@@
   6-byte unique identifier if none exists for a record.}, which are
-generated locally by the \rulelabel{E-Insert} rule.  The global
-uniqueness of locally generated identifiers can be axiomatized as an
-isolation property as following:
+checked locally for uniqueness by the \rulelabel{E-Insert} rule. The
+global uniqueness of locally unique identifiers can be axiomatized as
+an isolation property as following:
 \begin{smathpar}
@@
 \end{smathpar}
-$\I_{id}$ ensures that a globally unique id generated during a
-transaction's execution remains globally unique until it commits,
-prohibiting interference from a concurrent transaction that adds the
-same id. The axiom thus simulates a global counter protected by an
-exclusive lock without explicitly appealing to specific
-implementation artifacts.
+$\I_{id}$ ensures that if the id of a record is globally unique when
+it is added to a transaction's $\stl$, it remains globally unique
+until the transaction commits. This is achieved by prohibiting the
+interference from a concurrent transaction that adds the same id. The
+axiom thus simulates a global counter protected by an exclusive lock
+without explicitly appealing to an implementation artifact.
 
@@
 \end{smathpar}
-% The snapshot semantics however may not be applicable throughout the
-% lifetime of the transaction, like in the case of Repeatable Read (RR).
 
@@
 \end{smathpar}
-%% If a transaction declared as read-only performs a write, then its
-%% $\stl\neq \emptyset$, and the transaction never commits.
 
-In addition to these properties, there are various isolation
-guarantees proposed in the database or distributed systems literature,
-or implemented by commercial vendors that can also be specified within
-this framework:
+In addition to these properties, various specific isolation levels
+proposed in the database or distributed systems literature, or
+implemented by commercial vendors can also be specified within this
+framework:
 
@@
 
-\textbf{Repeatable Read (RR)} By definition, a read to a transactional
-variable in a Repeatable Read transaction is required to return the
-same value whenever accessed.  RR is often implemented (for e.g., in
+\textbf{Repeatable Read (RR)} By definition, multiple reads to a
+transactional variable in a Repeatable Read transaction are required
+to return the same value.  RR is often implemented (for e.g., in
 ~\cite{mysqliso,bailishat}) by executing the transaction against a
@@
 record written by a different transaction and read by the current
-transaction. By requiring $r\in\stg'$, $\I_e$ guarantees the
+transaction (since the current transaction's records are not yet
+present in $\stg$). By requiring $r\in\stg'$, $\I_e$ guarantees the
 invariance of $r$, thus the repeatability of the read.
@@
 transaction if conflicts are detected during the commit. This behavior
-can be axiomatized thus:
-\begin{smathpar}
-\begin{array}{lcl}
-\I_e\,\,(\stl,\stg,\stg') & = & \stg' = \stg\\
-\I_c\,\,(\stl,\stg,\stg') & = & \forall(r\in\stl)(r'\in\stg).~ r'.\idf = r.\idf \Rightarrow r'\in\stg'.
-\end{array}
-\end{smathpar}
-Observe that the same axiomatization applies to Postgres's RR,
+can be axiomatized as $\I_e = \I_{ss}$ (execution against a snapshot),
+and $\I_c = \I_{ww}$ (avoiding write-write conflicts during the
+commit).
+% \begin{smathpar}
+% \begin{array}{lcl}
+% \I_e\,\,(\stl,\stg,\stg') & = & \stg' = \stg\\
+% \I_c\,\,(\stl,\stg,\stg') & = & \forall(r\in\stl)(r'\in\stg).~ r'.\idf = r.\idf \Rightarrow r'\in\stg'.
+% \end{array}
+% \end{smathpar}
+Note that the same axiomatization applies to Postgres's RR,
 although its implementation (described in Sec.~\ref{sec:motivation})

### 2017-07-07 opsem.tex
Edits.

@@
 
+\textbf{Local Context Independence} As mentioned previously, our
+operational semantics does not let a transaction read its own writes.
+It also does not let a transaction overwrite its own writes, due to
+the premise $\dom(\stl)\cap\dom(s) = \emptyset$ on
+\rulelabel{E-Delete} and \rulelabel{E-Update} rules. We refer to this
+restriction as \emph{local context independence}. As such, this
+restriction is easy to relax in operational semantics and the
+reasoning framework presented in the next section, however, the
+automation framework introduced in Sec.~\ref{sec:inference} has a
+non-trivial dependence on this assumption. Nonetheless, our experience
+suggests that local context independence is not a severe restriction
+in practice. For instance, all 17 transactions in the TPC-C and TPC-E
+benchmarks satisfy this assumption.
+
 \input{isolation}

### 2017-07-07 inference.tex
Edits.

@@
   \R \vdash \hoare{\lambda(\stl,\stg).~\stl=\emptyset \conj
-  I(\stg)}{~c~}{\lambda(\stl,\stg).\stl = \F(\emptyset,\stg)}
+  I(\stg)}{c}{\lambda(\stl,\stg).\stl = \F(\emptyset,\stg)}
   \end{array}
@@
 transaction requires automating the multiple implication checks in
-the premise. While $R$, $\R$, $\R_c$, $\I$ and $\I$ are formulas in
+the premise. While $R$, $\R$, $\R_c$, $\I$ and $I$ are formulas in
 first-order logic (FOL) with a relatively simple structure, $\F$

### 2017-07-07 reasoning.tex
Edits.

@@
 the required isolation as specified by $\I$. It relies on the
-transaction-local judgment to verify the transaction body ($c$).
-The precondition $P$ of $c$ must follow from the fact that the
+transaction-local judgment to verify the body ($c$) of a transaction
+$i$. The precondition $P$ of $c$ must follow from the fact that the
 transaction-local database ($\stl$) is initially empty, and the global
@@
 instance, $\I_{ss}$, the snapshot axiom, is stable because if
-$\I_{ss}$ is invalid, then an interference has already modified a
-record, and no further interference will restore the original record,
-because the original record bears the id of a transaction that has
-already committed. Thus, $\I_{ss}$ remains invalid.
+$\I_{ss}$ is invalid ($\neg\I_{ss}$), then an interference has already
+modified a record, and no further interference will restore the
+original record, because the original record bears the id of a
+transaction that has already committed. Thus, $\I_{ss}$ remains
+invalid.
 
@@
 operational semantics ensures that $z$ is added to $y$ at the end of
-the iteration, hence $Q_c \conj z\in y$ is valid at the end of the
-iteration, from which $\psi$ must follow. When the loop has finished
-execution, $y$, the set of iterated items, is the entire set $x$. Thus
-$\psi \conj y=x$ must imply the post-condition $Q$, which also needs
-to be stable. The rules for conditionals, sequencing etc., are
+the iteration, hence $Q_c$ must imply $[y\cup\{z\}/y]\psi$. When the
+loop has finished execution, $y$, the set of iterated items, is the
+entire set $x$. Thus $[x/y]\psi$ is true at the end of the loop, from
+which the post-condition $Q$ must follow. As with the other rules, $Q$
+needs to be stable. The rules for conditionals, sequencing etc., are
 standard, and hence elided.
@@
 Fig.~\ref{fig:rg-rules}, and state their soundness guarantees.
-
-\begin{definition}[\bfseries Step-indexed reflexive transitive closure]
-For all $A:\text{Type}$, $R: A \rightarrow A \rightarrow \mathbb{P}$, and $n :
-\mathbb{N}$, the step-indexed reflexive transitive closure $R^n$ of $R$ is
-the smallest relation satisfying the following
-properties:
-\begin{itemize}
-\item $\forall (x:A).\, R^0 (x,x)$
-\item $\forall (n:\mathbb{N})(x,y,z : A).\, R(x,y) \conj R^n(y,z) \Rightarrow
-R^{n+1}(x,z)$
-\end{itemize}
-\end{definition}
 
@@
   \texttt{global}\\
-(c,\stl,\stg) \rstepsto (c',\stl',\stg') & \defeq & \stg \vdash 
-  (c,\stl) \stepsto (c',\stl') \conj \stg'=\stg& \texttt{transaction-local}\\
+(\tbox{c}_i,\stl,\stg) \rstepsto (\tbox{c'}_i,\stl',\stg') & \defeq & \stg \vdash 
+  (\tbox{c}_i,\stl) \stepsto (\tbox{c'}_i,\stl') \conj \stg'=\stg& \texttt{transaction-local}\\
   &   & \disj (c' = c \conj \stl'=\stl \conj \R(\stl, \stg, \stg'))
@@
 \noindent An interleaved multi-step relation ($\stepssto{n}$) is the
-step-indexed reflexive transitive closure of the interleaved step
-relation.  
+reflexive transitive closure of the interleaved step relation.  
 \end{definition}
@@
 \R \vdash \hoare{P}{c}{Q} & \defeq & \forall \stl,\stl',\stg,\stg'.~
-  P(\stl,\stg) \conj (c,\stl,\stg) \rstepssto{n} (\cskip, \stl',\stg')
+  P(\stl,\stg) \conj (\tbox{c}_i,\stl,\stg) \rstepssto{n}
+  (\tbox{\cskip}_i, \stl',\stg')
   \Rightarrow Q(\stl',\stg')\\
@@
 \begin{array}{lcl}
-\texttt{step-guaranteed}(R,G,c,\stg) &\defeq& \forall n,\stg',c'',\stg''.
+\texttt{step-guaranteed}(R,G,c,\stg) &\defeq& \forall \stg',c'',\stg''.
 (c,\stg) \rstepssto{n} (c',\stg') \conj (c',\stg') \stepsto

### 2017-07-07 motivation.tex
Edits to Sec. 2.

@@
 
-We present our ideas in the context of an embedded DSL in OCaml that
-we have developed which supports a \C{DB} monad to define and compose
-database computations given in terms of SQL operations.  Besides the
-usual \C{bind} and \C{return} operators, the monad offers an
-\C{atomically} combinator that executes a database computation as an
-atomic transaction and returns the result.
+We present our ideas in the context of our DSL embedded in OCaml,
+which offers a \C{DB} monad to define and compose database
+computations given in terms of SQL operations.  Besides the usual
+\C{bind} and \C{return} operators, the monad offers an \C{atomically}
+combinator that executes a database computation as an atomic
+transaction and returns the result.
 
@@
 \C{new\_order} transaction written in this language. TPC-C is a
-well-known Online Transaction Processing (OLTP) benchmark that models
+widely-used Online Transaction Processing (OLTP) benchmark that models
 an order-processing system for a wholesale parts supply business. The
@@
 
-\begin{figure}[!t]
-\includegraphics[scale=0.5]{Figures/schema}
+% \begin{figure}[!t]
+% \includegraphics[scale=0.5]{Figures/schema}
+% \caption{Database schema of TPC-C's order management system.
+%   Columns against highlighted background are primary keys. Arrows denote
+%   foreign key relationships.}
+% \label{fig:schema}
+% \end{figure}
+
+\begin{figure}[t]
+  \centering
+	\begin{subfigure}{0.48\textwidth}
+		\includegraphics[width=0.9\textwidth]{Figures/schema1}
+    \caption{A valid TPC-C database. The only existing order belongs
+      to the district with \C{d\_id}=11. It's id (\C{o\_id}) is one
+      less than the district's \C{d\_next\_o\_id}, and it's order
+      count (\C{o\_ol\_cnt}) is equal to the number of order line
+      records whose \C{ol\_o\_id} is equal to the order's id.  }
+		\label{fig:tpcc_db1}
+	\end{subfigure}
+	\begin{subfigure}{0.48\textwidth}
+		\includegraphics[width=0.9\textwidth]{Figures/schema2}
+    \caption{The database in Fig.~\ref{fig:tpcc_db1} after correctly
+      executing a \C{new\_order} transaction. A new order record is
+      added whose \C{o\_id} is equal to the \C{d\_next\_o\_id} from
+      Fig.~\ref{fig:tpcc_db1}. The district's \C{d\_next\_o\_id} is
+      incremented. The order's \C{o\_ol\_cnt} is 2, reflecting the
+      actual number of order line records whose \C{ol\_o\_id} is equal
+      to the order's id (2).}
+		\label{fig:tpcc_db2}
+	\end{subfigure}
+
 \caption{Database schema of TPC-C's order management system.
-  Columns against highlighted background are primary keys. Arrows denote
-  foreign key relationships.}
+  The naming
+  convention indicates primary keys and foreign keys. For e.g.,
+  \C{ol\_id} is the primary key column of the order line table,
+  whereas \C{ol\_o\_id} is a foreign key that refers to the \C{o\_id}
+  column of the order table.}
 \label{fig:schema}

### 2017-07-07 motivation.tex
EDits to Sec. 2.

@@
 \C{new\_order} transaction when executed with a \emph{Read Committed}
-(RC) isolation level, the default isolation level in PostgreSQL, a
-widely used open-source database system.  An RC transaction is
+(RC) isolation level, the default isolation level in 8 of the 18
+databases studied in~\ref{bailishotos}.  An RC transaction is
 isolated from \emph{dirty writes}, i.e., writes of uncommitted
@@
 $T_2$), both concurrently placing new orders for different customers
-under the same district (\C{d\_id}), PostgreSQL admits the two
-executions shown in Fig.~\ref{fig:new_order_execs}.
+under the same district (\C{d\_id}), RC isolation allows the
+execution shown in Fig.~\ref{fig:new_order_execs}.
 
-\begin{figure}[!h]
-\centering
-\subcaptionbox {
-  {\sc rc} Execution 1
-  \label{fig:motiv-eg-1-b}
-} [
-  0.55\columnwidth
-] {
+\begin{figure}[!t]
 \includegraphics[scale=0.45]{Figures/motiv-eg-1-b}
-}
-%\hspace*{0.5in}
-\subcaptionbox {
-  {\sc rc} Execution 2
-  \label{fig:motiv-eg-1-a}
-}{
-  \includegraphics[scale=0.45]{Figures/motiv-eg-1-a}
-}
-  \caption{\small RC executions involving two instances ($T_1$ and
+\caption{\small An RC execution involving two instances ($T_1$ and
   $T_2$) of the \C{new\_order} transaction depicted in
   Fig.~\ref{fig:new_order_code}. 
-  Each instance reads the \C{d\_id} \C{District} record twice, the second
-  time to (atomically) update the \C{d\_next\_o\_id} field.}
-\label{fig:new_order_execs}
+  Both instances read the \C{d\_id} \C{District} record concurrently,
+  because neither transaction is committed when the reads are
+  executed.  The subsequent operations are effectively sequentialized,
+  since $T_2$ commits. Nonetheless, reading the same value for
+  \C{d\_next\_o\_id} prompts both instances to add \C{Order} records
+  with same ids, which inturn triggers the the violation of TPC-C's
+  consistency condition.}
+\label{fig:new_order_exec}
 \end{figure}
 
-The figure depicts an execution as a series of transactional read, write, and commit
-operations. In the execution on the left, the \C{new\_order} instance
+% \begin{figure}[!h]
+% \centering
+% \subcaptionbox {
+%   {\sc rc} Execution 1
+%   \label{fig:motiv-eg-1-b}
+% } [
+%   0.55\columnwidth
+% ] {
+%   \includegraphics[scale=0.45]{Figures/motiv-eg-1-b}
+% }
+% %\hspace*{0.5in}
+% \subcaptionbox {
+%   {\sc rc} Execution 2
+%   \label{fig:motiv-eg-1-a}
+% }{
+%   \includegraphics[scale=0.45]{Figures/motiv-eg-1-a}
+% }
+%   \caption{\small RC executions involving two instances ($T_1$ and
+%   $T_2$) of the \C{new\_order} transaction depicted in
+%   Fig.~\ref{fig:new_order_code}. 
+%   Each instance reads the \C{d\_id} \C{District} record twice, the second
+%   time to (atomically) update the \C{d\_next\_o\_id} field.}
+% \label{fig:new_order_execs}
+% \end{figure}
+
+The figure depicts an execution as a series of SQL 
+operations. In the execution, the \C{new\_order} instance
 $T_1$ (green) reads the \C{d\_next\_o\_id} field of the district
@@
 acccessed. This is allowed because reads do not obtain a mutually
-exclusive lock on most databases, including PostgreSQL. After $T_2$'s
-commit, $T_1$ resumes execution and adds new \C{Order} and
-\C{Order\_line} fields with the same order id as $T_1$. Thus, by the
-end of the execution, \C{Order\_line} records inserted by $T_1$ and
-$T_2$ all bear the same order id. There are also two \C{Order} records
-with the same district id (\C{d\_id}) and order id, none of whose
-\C{o\_ol\_cnt} reflects the actual number of \C{Order\_line} records
-inserted with that order id.  This clearly violates TPC-C's consistency
-condition.
+exclusive lock on most databases. After $T_2$'s commit, $T_1$ resumes
+execution and adds new \C{Order} and \C{Order\_line} fields with the
+same order id as $T_1$. Thus, by the end of the execution,
+\C{Order\_line} records inserted by $T_1$ and $T_2$ all bear the same
+order id. There are also two \C{Order} records with the same district
+id (\C{d\_id}) and order id, none of whose \C{o\_ol\_cnt} reflects the
+actual number of \C{Order\_line} records inserted with that order id.
+This clearly violates TPC-C's consistency condition.
 
-Notably, this example does not exhibit any common concurrency bugs
-that arise from incorrect use of weak isolation such as
-\emph{write-write} conflicts, or \emph{lost updates}.  While $T_1$ and
-$T_2$ both increment the \C{d\_next\_o\_id} field of the district
-record, they do so atomically (Line 4 of
-Fig.~\ref{fig:new_order_code}), allowing both updates to be present in
-the final state. Likewise, other well-known anomalies that
-characterize RC isolation~\cite{berenson} such as \emph{fuzzy reads},
-\emph{phantom reads}, \emph{read skew}, and \emph{write skew}, are
-also not exhibited by the example.  Thus, program analyses that aim to
+Notably, this example does not exhibit any of the anomalies that
+characterize RC isolation~\cite{berenson}. For instance, there are no
+\emph{lost writes}; both concurrent transactions' writes are present
+in the final state of the database. Thus, program analyses that aim to
 determine appropriate isolation by checking for possible
@@
 
-While reasoning in terms of anomalies is cumbersome, and as the above
-example shows often inadequate, reasoning about weak isolation in
-terms of low-level traces~\cite{adyaphd,gotsmanconcur15} on memory
-read and write actions complicates high-level reasoning.  A possible
-alternative would interleave weak isolation implementation details
-within the program, yielding a (more-or-less) conventional concurrent
-program that can be then subject to classical concurrent verification
-methods.  Considering the size and complexity of real-world
-transaction systems, this strategy is unlikely to scale.
+While reasoning in terms of anomalies is cumbersome and inadequate,
+reasoning about weak isolation in terms of low-level
+traces~\cite{adyaphd,gotsmanconcur15} on memory read and write actions
+complicates high-level reasoning.  A possible alternative would
+interleave weak isolation implementation details within the program,
+yielding a (more-or-less) conventional concurrent program that can be
+then subject to classical concurrent verification methods.
+Considering the size and complexity of real-world transaction systems,
+this strategy is unlikely to scale.
 
@@
 illustrate this idea informally, consider how we might verify that
-\C{new\_order} is sound when executed under a \emph{Repeatable Read}
-(RR) isolation level.  PostgreSQL executes an RR transaction by taking
-a (conceptual) snapshot of the database state before the transaction
-begins that is used by reads performing the transaction.  When the
-transaction attempts to update a record, a check is made to determine
-if the current version of the record in the database is the same as the
-snapshot version. If it is, an exclusive lock is obtained on the
-record, and the update is performed. If the record is already locked,
-the transaction blocks until the lock is released.  When the lock is
-obtained, is held exclusively until the transaction commits or rolls
-back. If the aforementioned version check fails (i.e., the database
-has a later version than the snapshot), the transaction is rolled back
-and re-executed.
+\C{new\_order} is sound when executed under \emph{Snapshot Isolation}
+(SI), and isolation level stronger than RC. Snapshot isolation
+allows transactions to be executed against their respective private
+snapshots of the database, thus admitting concurrency, but it also
+requires that there not be any write-writeconflicts among
+concurrent transactions. Write-write conflicts can be eliminated in various
+ways; either through detection followed by a rollback, or through
+exclusive locks, or a combination of both, leading to different ways
+in which SI can be realized. For instance, one possible implementation
+of SI, close to the one use by PostgreSQL~\cite{postgres-ssi},
+executes a transaction against its private snapshot of the database,
+but obtains exclusive locks on the actual records in the database
+before performing writes. A write is performed only of the record
+hasn't already been updated by a concurrent transaction, i.e,  when it
+doesn't result in a write-write conflict.  Otherwise, the transaction is
+rolledback. This implementation is demonstrated for a simple database
+with three records, $a$, $b$, and $c$ in Fig.~\ref{fig:RR-postgres}.
 
@@
 \caption{Database state transitions corresponding to an execution of
-  an RR transaction $T_1$ on PostgreSQL. $T_2$ and $T_3$ are concurrent
-  transactions. The database state has three data items, $a$, $b$ and $c$. }
+  an SI transaction $T_1$ on a PostgreSQL-like store. $T_2$ and $T_3$
+  are concurrent transactions. }
 \label{fig:rr-postgres}
@@
 
-Although this implementation, comprising many thousands of lines of
-code, is highly involved, its semantic behavior in terms of how it
-effects transitions on the database state is fairly simple.  Since a
-transaction's reads are always served from a snapshot, no state
-changes are witnessed while the execution is in progress. Thus,
-insofar as an RR transaction is concerned, the database state does not
-change during the execution.  Uncommitted writes are recorded in a
-transaction-local state.  When the transaction commits, the local
-state is atomically written to the global state to yield a global
-state that reflects the transaction's updates.  However, unlike a
-strongly isolated serializable transaction, the commit operation is
-performed against the current state of the database, not the
-snapshot. Thus, after the transaction finishes execution, but before
-it commits, the transaction is able to witness the effects of all
-concurrent transactions that committed before it.  The PostgreSQL RR
-implementation effectively constrains this transition.
-
-We can axiomatize this operational description by observing that (a)
-due to the version check, the current transaction cannot update a data
-item that was already updated by a concurrent transaction, and (b) due
-to the use of exclusive write locks, a data item updated by the
-current transaction cannot be overwritten by a concurrent transaction.
-If $\Delta$ is the state of the database when an RR transaction
-finishes, and $\Delta_c$ is the state visible to the transaction at
-the point of commit, we know the transition from $\Delta$ to
-$\Delta_c$ (written $\Delta \longrightarrow \Delta_c$) cannot exhibit
-effects from any concurrent transactions that write to the same data
-items as the current RR transaction.  Similarly, if $\delta$ denotes
-a local log that relates transaction variables being written with their updated values,
-then $\forall x\in\mathit{dom}(\delta)$, $\Delta_c(x) =
-\Delta(x)$. To summarize, the operational semantics of PostgreSQL's RR
-implementation can be captured as an axiomatization over transitions of
-the database state ($\Delta \longrightarrow \Delta'$) during the
-lifetime an RR transaction ($T$):
+Implementations of SI on real databases (e.g., PostgreSQL) are
+complicated, often running into thousands of lines of code.
+Nonetheless, the semantics of Snapshot Isolation, in terms of how it
+effects the transitions on the database state, can be captured in a
+fairly simple model. Firstly, although SI admits concurrent
+transactions, their effects are  not witnessed in the current
+transaction (call it $T$), during $T$'s execution. Thus, insofar as
+$T$ is concerned, as long as it executes, the global state doesn't
+change. More formally, for every consecutive pair of global states
+$(\stg,\stg')$ witnessed by $T$ during its execution, $\stg'=\stg$.
+After $T$ finishes execution, it commits to the actual database, which
+incorporates the effects of concurrent transactions. In executions
+where $T$ successfully commits, concurrent transactions are guaranteed
+to not be in write-write conflict with $T$. Thus, if $\stg$ is the
+global state that $T$ witnessed when it finished execution (which
+would be same as the snapshot state), and $\stg'$ is the state to
+which $T$ commits, then the diff between $\stg$ and $\stg'$ shouldn't
+be in a write-write conflict with $T$. To concretize this notion, let
+the database state be a map from transaction variables to values, and
+let $\stl$ denote a transaction-local log that maps the variables
+being written to their updated values. Then the absence of write-write
+conflicts between $T$ and the diff between $\stg$ and $\stg'$ can be
+defined as: $\forall x\in\mathit{dom}(\delta)$, $\stg'(x) = \stg(x)$.
+To summarize, the semantics of SI can be captured as an axiomatization
+over transitions of the database state ($\Delta \longrightarrow
+\Delta'$) during the lifetime an SI transaction ($T$):
 \begin{itemize}
@@
 
-This simple characterization of RR isolation allows us to verify the
-consistency conditions associated with the \C{new\_order}
-transaction. First, since the database doesn't change ($\Delta' =
-\Delta$) during the execution of the transaction's body, we can reason
-about \C{new\_order} as though it executed in complete isolation until
-its commit point, leading to a verification process similar to what
-would been applied when reasoning about serializability.  When
-\C{new\_order} finishes execution and becomes ready to commit, a
-transition transfers the transaction's local writes ($\delta$) to the
-unchanged database state ($\Delta$).  However, we are required to
-consider the interference from concurrent transitions at this point,
-which might change the database state from $\Delta$ to $\Delta_c$. If
-this interference includes the effects of a concurrent \C{new\_order}
-transaction (with the same \C{d\_id}), then verification fails as
-described previously.  Fortunately, sequential reasoning shows that
-this is impossible - RR prevents a concurrent \C{new\_order}
-transaction that modifies the same \C{District} record as the current
-transaction (concretely, since the record is already present in the
-current transaction's local log, any transition from $\Delta$ to
-$\Delta_c$ cannot change this record).  Applying such axiomatic
-reasoning on \C{new\_order} allows us to prove that the TPC-C
-invariant holds when the transaction is executed under PostgreSQL's RR
-isolation.  Our proof framework generalizes this style of reasoning to
-various isolation levels on databases.
+% nor $T$'s effects are
+% witnessed by concurrent transactions, during $T$'s execution. Thus,
+% $T$'s execution can be thought of as a series of transitions of a
+% state local to $T$ (call it $\stl$), that has no bearing outside $T$. Conversely, any
+% transitions performed by the concurrent transactions on the global
+% state (call it $\stg$) has no
+% bearing on $T$ since $T$ reads off a static snapshot established at
+% the beginning of the transaction. Consequently, any
+% state transitions performed by the concurrent transactions can be
+% \emph{moved past} $T$'s local state transitions. $T$'s commit,
+% however, is a global state transition that is enabled only if
+% there are no WW conflicts between $T$ and concurrent transactions.
+% Thus, concurrent transactions' state transitions cannot be moved past
+% $T$'s commit, and $T$ can be thought of witnessing the effects of
+% concurrent transactions just before its commit. 
+% therefore treat Since a
+% transaction's reads are always served from a snapshot, no state
+% changes are witnessed while the execution is in progress. Thus,
+% insofar as an RR transaction is concerned, the database state does not
+% change during the execution.  Uncommitted writes are recorded in a
+% transaction-local state.  When the transaction commits, the local
+% state is atomically written to the global state to yield a global
+% state that reflects the transaction's updates.  However, unlike a
+% strongly isolated serializable transaction, the commit operation is
+% performed against the current state of the database, not the
+% snapshot. Thus, after the transaction finishes execution, but before
+% it commits, the transaction is able to witness the effects of all
+% concurrent transactions that committed before it.  The PostgreSQL RR
+% implementation effectively constrains this transition.
+
+% We can axiomatize this operational description by observing that (a)
+% due to the version check, the current transaction cannot update a data
+% item that was already updated by a concurrent transaction, and (b) due
+% to the use of exclusive write locks, a data item updated by the
+% current transaction cannot be overwritten by a concurrent transaction.
+% If $\Delta$ is the state of the database when an RR transaction
+% finishes, and $\Delta_c$ is the state visible to the transaction at
+% the point of commit, we know the transition from $\Delta$ to
+% $\Delta_c$ (written $\Delta \longrightarrow \Delta_c$) cannot exhibit
+% effects from any concurrent transactions that write to the same data
+% items as the current RR transaction.  Similarly, if $\delta$ denotes
+% a local log that relates transaction variables being written with their updated values,
+% then $\forall x\in\mathit{dom}(\delta)$, $\Delta_c(x) =
+% \Delta(x)$. To summarize, the operational semantics of PostgreSQL's RR
+% implementation can be captured as an axiomatization over transitions of
+% the database state ($\Delta \longrightarrow \Delta'$) during the
+% lifetime an RR transaction ($T$):
+% \begin{itemize}
+%   \item While $T$ executes, $\Delta' = \Delta$.
+%   \item After $T$ finishes execution, but before it commits its local
+%     state $\delta$, $\forall(x\in\delta).~\Delta'(x) = \Delta(x)$.
+% \end{itemize}
+
+This simple characterization of SI isolation allows us to verify the
+consistency conditions associated with the \C{new\_order} transaction.
+First, since the database doesn't change ($\Delta' = \Delta$) during
+the execution of the transaction's body, we can reason about
+\C{new\_order} as though it executed in complete isolation until its
+commit point, leading to a verification process similar to what would
+been applied when reasoning about serializability.  When
+\C{new\_order} finishes execution, but before it commits, the SI
+axiomatization shown above requires us to consider global state
+transitions $\stg \longrightarrow \stg'$ that do not include changes
+to the records ($\stl$) written by \C{new\_order}, i.e.,
+$\forall(x\in\delta).~\Delta'(x) = \Delta(x)$. Since concurrent
+\C{new\_order} transactions that write to the same C{District} record
+modify the record by incrementing its \C{d\_next\_o\_id} field,
+their transitions do not satisfy the necessary condition imposed by
+the SI axiomatization, allowing us to rightfully ignore their
+interference. This leaves the interference due to concurrent
+\C{new\_order} transactions that write to a different district record. 
+Fortunately, such interference does not prompt \C{new\_order} to
+violate the invariant, which can be established by reasoning on the
+similar lines as concurrent program logics, such as the Rely-Guarantee
+logic.
+
+% and becomes ready to commit, a
+% transition transfers the transaction's local writes ($\delta$) to the
+% unchanged database state ($\Delta$).  However, we are required to
+% consider the interference from concurrent transitions at this point,
+% which might change the database state from $\Delta$ to $\Delta_c$. If
+% this interference includes the effects of a concurrent \C{new\_order}
+% transaction (with the same \C{d\_id}), then verification fails as
+% described previously.  Fortunately, sequential reasoning shows that
+% this is impossible - RR prevents a concurrent \C{new\_order}
+% transaction that modifies the same \C{District} record as the current
+% transaction (concretely, since the record is already present in the
+% current transaction's local log, any transition from $\Delta$ to
+% $\Delta_c$ cannot change this record).  Applying such axiomatic
+% reasoning on \C{new\_order} allows us to prove that the TPC-C
+% invariant holds when the transaction is executed under PostgreSQL's RR
+% isolation.  Our proof framework generalizes this style of reasoning to
+% various isolation levels on databases.
 

### 2017-07-07 motivation.tex
Edits.

@@
 $\forall(x\in\delta).~\Delta'(x) = \Delta(x)$. Since concurrent
-\C{new\_order} transactions that write to the same C{District} record
-modify the record by incrementing its \C{d\_next\_o\_id} field,
-their transitions do not satisfy the necessary condition imposed by
-the SI axiomatization, allowing us to rightfully ignore their
-interference. This leaves the interference due to concurrent
-\C{new\_order} transactions that write to a different district record. 
-Fortunately, such interference does not prompt \C{new\_order} to
-violate the invariant, which can be established by reasoning on the
-similar lines as concurrent program logics, such as the Rely-Guarantee
-logic.
+\C{new\_order} transactions that write to the same \C{District} record
+modify the record by incrementing its \C{d\_next\_o\_id} field, their
+transitions do not satisfy the necessary condition imposed by the SI
+axiomatization, allowing us to rightfully ignore their interference.
+This leaves the interference due to concurrent \C{new\_order}
+transactions that write to a different district record.  Fortunately,
+such interference does not prompt \C{new\_order} to violate the
+invariant, which can be established by adopting a reasoning similar to
+the one used on concurrent program logics, such as the Rely-Guarantee
+logic, customized for the database programs.
 

### 2017-07-07 motivation.tex
Sec. 2 edits.

@@
 but obtains exclusive locks on the actual records in the database
-before performing writes. A write is performed only of the record
+before performing writes. A write is performed only if the record
 hasn't already been updated by a concurrent transaction, i.e,  when it
@@
 transaction (call it $T$), during $T$'s execution. Thus, insofar as
-$T$ is concerned, as long as it executes, the global state doesn't
-change. More formally, for every consecutive pair of global states
-$(\stg,\stg')$ witnessed by $T$ during its execution, $\stg'=\stg$.
-After $T$ finishes execution, it commits to the actual database, which
-incorporates the effects of concurrent transactions. In executions
-where $T$ successfully commits, concurrent transactions are guaranteed
-to not be in write-write conflict with $T$. Thus, if $\stg$ is the
-global state that $T$ witnessed when it finished execution (which
-would be same as the snapshot state), and $\stg'$ is the state to
-which $T$ commits, then the diff between $\stg$ and $\stg'$ shouldn't
-be in a write-write conflict with $T$. To concretize this notion, let
-the database state be a map from transaction variables to values, and
-let $\stl$ denote a transaction-local log that maps the variables
-being written to their updated values. Then the absence of write-write
-conflicts between $T$ and the diff between $\stg$ and $\stg'$ can be
-defined as: $\forall x\in\mathit{dom}(\delta)$, $\stg'(x) = \stg(x)$.
-To summarize, the semantics of SI can be captured as an axiomatization
-over transitions of the database state ($\Delta \longrightarrow
-\Delta'$) during the lifetime an SI transaction ($T$):
+$T$ is concerned, the global state doesn't change as long as it
+executes. More formally, for every consecutive pair of global states
+$(\stg,\stg')$ witnessed by $T$ during its execution, SI requires
+$\stg'=\stg$.  After $T$ finishes execution, it commits to the actual
+database, which incorporates the effects of concurrent transactions.
+In executions where $T$ successfully commits, concurrent transactions
+are guaranteed to not be in write-write conflict with $T$. Thus, if
+$\stg$ is the global state that $T$ witnessed when it finished
+execution (which would be same as the snapshot state), and $\stg'$ is
+the state to which $T$ commits, then the diff between $\stg$ and
+$\stg'$ shouldn't be in a write-write conflict with $T$. To concretize
+this notion, let the database state be a map from transaction
+variables to values, and let $\stl$ denote a transaction-local log
+that maps the variables being written to their updated values. Then
+the absence of write-write conflicts between $T$ and the diff between
+$\stg$ and $\stg'$ can be defined as: $\forall
+x\in\mathit{dom}(\delta)$, $\stg'(x) = \stg(x)$.  To summarize, the
+semantics of SI can be captured as an axiomatization over transitions
+of the database state ($\Delta \longrightarrow \Delta'$) during the
+lifetime an SI transaction ($T$):
 \begin{itemize}
@@
 the one used on concurrent program logics, such as the Rely-Guarantee
-logic, customized for the database programs.
+logic, customized for the database programs. Our approach generalizes
+this style of reasoning to a range of isolation levels, concurrency
+control techniques, and their combinations used frequently in practice.
 
@@
 added to the local database. Thus, its state transformer (call it
-$T_U$) is the following function on sets\footnote{$\bind$ has higher precedence than $\cup$.}:
-\begin{ocaml}
-  fun ($\Delta$,$\delta$) -> $\delta$ $\cup$ $\Delta$$\bind$(fun s -> if table(s) = stock && s.s_i_id = item_req.ol_i_id 
-                                 then {{s with s_qty = k1}} (* a singleton set *)
-                                 else $\emptyset$)
-\end{ocaml}
+$\T_U$) is the following function on sets\footnote{$\bind$ has higher
+precedence than $\cup$.}:
+\begin{smathpar}
+\begin{array}{l}
+  \lambda(\stl,\stg).~ \stl \cup \stg \bind(\lambda s. 
+      \itel{{\sf table}(s) = \C{Stock} \conj 
+                s.\C{s\_i\_id} = \C{item\_req.ol\_i\_id}\\\hspace*{1.3in}} 
+           {\{\{s \with \C{s\_qty} = \C{k1}\}\}\\\hspace*{1.3in}}
+           {\emptyset}
+\end{array}
+\end{smathpar}
 % \begin{smathpar}
@@
 % \end{verbatim}
-Observe that
-$T_U(\Delta,\delta)$ is of the form $\delta \,\cup\,
-F_U(\Delta)$. The transformer ($T_I(\Delta,\delta)$)
-for the subsequent \C{insert} statement can be similarly calculated to
-be of the form $\delta \cup F_I(\Delta)$.
-Their composition gives the state transformer of the loop body:
-\begin{ocaml}
-  fun ($\Delta$,$\delta$) -> $\delta$ $\cup$ $F_U(\Delta) \,\cup\, F_I(\Delta)$
-\end{ocaml}
-Let $F_{body}(\Delta) = F_U(\Delta) \,\cup\, F_I(\Delta)$.   The
-transformer for the \C{foreach} loop can now be computed as follows:
-\begin{ocaml}
-  fun ($\Delta$,$\delta$) -> $\delta$ $\cup$ item_reqs$\bind$(fun item_req -> $F_{body}(\Delta)$)
-\end{ocaml}
-Observe that the transformer captures the precise semantics of the
-loop as a formula in the domain of sets extended with a bind
-function. The advantage in doing so is that we can now make use of a
-semantics-preserving translation from this domain to first-order
-logic, allowing us to leverage SMT solvers for automatic proofs
-without having to infer loop invariants, or provide intermediate
-assertions.  Sec.~\ref{sec:automation} describes this translation. In
-the exposition thus far, we assumed $\Delta$ remains invaraint, which
-is clearly not the case when we admit concurrency.  Necessary
-concurrency extensions of the state transformer semantics is also
-covered in Sec.~\ref{sec:automation}.  The following two sections,
-however, first focus on laying the theoretical foundations for
-reasoning about weak isolation.
+% Observe that
+% $\T_U(\Delta,\delta)$ is of the form $\delta \,\cup\,
+% F_U(\Delta)$. 
+The transformer ($\T_I(\Delta,\delta)$) for the subsequent \C{insert}
+statement can be similarly calculated as the following:
+\begin{smathpar}
+\begin{array}{l}
+  \lambda(\stl,\stg).~ \stl \cup \{\C{ol\_o\_id}=\C{k2};\, 
+        \C{ol\_d\_id}=\C{k3};\, \C{ol\_i\_id}=\C{item\_req.ol\_i\_id};\, 
+        \C{ol\_qty}=\C{item\_req.ol\_qty}\}
+\end{array}
+\end{smathpar}
+Observe that both transformers are of the form $T(\Delta,\delta) =
+\stl \cup F(\stg)$, where $F$ is a function that returns the set of
+records added by the transformer to the transaction-local database
+($\stl$). Let $\F_U$ and $\F_I$ be the corresponding functions for $\T_U$
+and $\T_I$ shown above. The state transformed of the loop body in
+Fig.~\ref{fig:new} can be expressed as the following composition of
+$\F_U$ and $\F_I$:
+\begin{smathpar}
+\begin{array}{l}
+  \lambda(\stl,\stg).~ \stl \cup \F_U(\stg) \cup \F_I(\stg)
+\end{array}
+\end{smathpar}
+The transformer for the loop itself can now be computed as following:
+\begin{smathpar}
+\begin{array}{l}
+  \lambda(\stl,\stg).~ \stl \cup \C{item\_reqs}\bind
+      (\lambda\C{item\_req}.~ \F_U(\stg) \cup \F_I(\stg))
+\end{array}
+\end{smathpar}
+Observe that the structure of the transformer mirrors the structure of
+the program itself. In particular, SQL statements become set
+operations, and the \C{DB} monad's $\bind$ becoms set monadic $\bind$.
+Thus, state transformer inference is completely mechanical. The
+advantage of inferring such transformers is that we can now make use
+of a semantics-preserving translation from the domain of sets equipped
+with $\bind$ to the first-order logic, allowing us to leverage SMT
+solvers for automatic proofs without having to infer loop invariants,
+or provide intermediate assertions.  Sec.~\ref{sec:inference}
+describes this translation. In the exposition thus far, we assumed
+$\Delta$ remains invaraint, which is clearly not the case when we
+admit concurrency.  Necessary concurrency extensions of the state
+transformer semantics is also covered in Sec.~\ref{sec:inference}.
+The following two sections, however, first focus on laying the
+theoretical foundations for reasoning about weak isolation.

### 2017-07-07 motivation.tex
Edits.

@@
   let o_id = dist.d_next_o_id;
-  SQL.update District (fun d -> {d with d_next_o_id =d_next_o_id + 1})
-                      (fun d -> d.d_id = d_id );
-  SQL.insert Order {o_id=o_id;  o_d_id=d_id; 
-                    o_c_id=c_id; o_ol_cnt=S.size item_reqs; };
+  SQL.update(* UPDATE *) District 
+            (* SET *)(fun d -> {d with d_next_o_id =d_next_o_id + 1})
+            (* WHERE *)(fun d -> d.d_id = d_id );
+  SQL.insert(* INSERT INTO *) Order (* VALUES *){o_id=o_id;  
+            o_d_id=d_id; o_c_id=c_id; o_ol_cnt=S.size item_reqs; };
   foreach item_reqs @@ fun item_req -> do
-    stk <- SQL.select1 Stock (fun s -> s.s_i_id = item_req.ol_i_id &&
-                                       s.s_d_id = d_id);
+    stk <- SQL.select1(* SELECT * FROM *) Stock 
+              (* WHERE *)(fun s -> s.s_i_id = item_req.ol_i_id 
+                                  && s.s_d_id = d_id)(* LIMIT 1 *); 
     let s_qty' = if stk.s_qty >= item_req.ol_qty + 10 

### 2017-07-07 motivation.tex
edits

--- 2017-07-07 motivation.tex
@@
 
-We present our ideas in the context of our DSL embedded in OCaml,
-which offers a \C{DB} monad to define and compose database
-computations given in terms of SQL operations.  Besides the usual
-\C{bind} and \C{return} operators, the monad offers an \C{atomically}
-combinator that executes a database computation as an atomic
-transaction and returns the result.
+We present our ideas in the context of an embedded DSL in OCaml, which
+offers a \C{DB} monad to define and compose database computations
+given in terms of SQL operations.  Besides the usual \C{bind} and
+\C{return} operators, the monad includes an \C{atomically} combinator
+that executes a database computation as an atomic transaction and
+returns the result.
 
@@
 \C{new\_order} transaction written in this language. TPC-C is a
-widely-used Online Transaction Processing (OLTP) benchmark that models
-an order-processing system for a wholesale parts supply business. The
-business logic is captured in 5 database transactions that operate on
-9 tables; \C{new\_order} is one such transaction that uses
-\C{District}, \C{Order}, \C{New\_order}, \C{Stock}, and
-\C{Order\_line} tables. The transaction acts on the behalf of a
-customer, whose id is \C{c\_id}, to place a new order for a given
-set of items (\C{item\_reqs}), to be served by a warehouse under the
-district identified by \C{d\_id}.  Fig.~\ref{fig:schema} illustrates
-the relationship among these different tables.
+widely-used and well-studied Online Transaction Processing (OLTP)
+benchmark that models an order-processing system for a wholesale parts
+supply business. The business logic is captured in 5 database
+transactions that operate on 9 tables; \C{new\_order} is one such
+transaction that uses \C{District}, \C{Order}, \C{New\_order},
+\C{Stock}, and \C{Order\_line} tables. The transaction acts on the
+behalf of a customer, whose id is \C{c\_id}, to place a new order for
+a given set of items (\C{item\_reqs}), to be served by a warehouse
+under the district identified by \C{d\_id}.  Fig.~\ref{fig:schema}
+illustrates the relationship among these different tables.
 
@@
 functionality, captured by various calls to functions defined by the
-\C{SQL} module. All \C{SQL} functions take the table name (a nullary
-constructor) as their first argument. The higher-order \C{SQL.select1}
-function accepts a boolean function that describes the selection
-criteria, and returns any record that meets the criteria (it models
-the SQL query \C{SELECT \ldots\xspace LIMIT 1}). \C{SQL.update} also
-accepts a Boolean function (its 3$^{rd}$ argument) to select the records to be
-updated. Its 2$^{nd}$ argument is a function that maps each selected
-record to a new (updated) record. \C{SQL.insert} inserts a given
-record into the specified table in the database.
+\C{SQL} module. All \C{SQL} operators supported by the module take a
+table name (a nullary constructor) as their first argument. The
+higher-order \C{SQL.select1} function accepts a boolean function that
+describes the selection criteria, and returns any record that meets
+the criteria (it models the SQL query \C{SELECT \ldots\xspace LIMIT
+  1}). \C{SQL.update} also accepts a Boolean function (its 3$^{rd}$
+argument) to select the records to be updated. Its 2$^{nd}$ argument
+is a function that maps each selected record to a new (updated)
+record. \C{SQL.insert} inserts a given record into the specified table
+in the database.
 
@@
   \centering
-	\begin{subfigure}{0.48\textwidth}
+	\begin{subfigure}{0.4\textwidth}
 		\includegraphics[width=0.9\textwidth]{Figures/schema1}
     \caption{A valid TPC-C database. The only existing order belongs
-      to the district with \C{d\_id}=11. It's id (\C{o\_id}) is one
-      less than the district's \C{d\_next\_o\_id}, and it's order
+      to the district with \C{d\_id}=11. Its id (\C{o\_id}) is one
+      less than the district's \C{d\_next\_o\_id}, and its order
       count (\C{o\_ol\_cnt}) is equal to the number of order line
@@
 	\end{subfigure}
-	\begin{subfigure}{0.48\textwidth}
+	\begin{subfigure}{0.52\textwidth}
 		\includegraphics[width=0.9\textwidth]{Figures/schema2}
@@
   transactions}, however, interferences of various kinds are
-permitted.  Although the verification problem for weakly isolated
-transactions would appear to be superficially similar to the
-verification of (racy) concurrent programs (e.g., garbage
-collectors~\cite{JLP+14,GHE15,HPQ+15}), weak isolation introduces new
-challenges arising from the use of transactions, and new opportunities
-arising from the fact that the store abstraction used by transactions
-is a relational database, not low-level memory.
-
-To illustrate some of these challenges, consider the behavior of the
-\C{new\_order} transaction when executed with a \emph{Read Committed}
-(RC) isolation level, the default isolation level in 8 of the 18
-databases studied in~\ref{bailishotos}.  An RC transaction is
-isolated from \emph{dirty writes}, i.e., writes of uncommitted
-transactions, but is allowed to witness the writes of concurrent
-transactions as soon as they are committed. Thus, with two concurrent
-instances of the \C{new\_order} transaction (call them $T_1$ and
-$T_2$), both concurrently placing new orders for different customers
-under the same district (\C{d\_id}), RC isolation allows the
-execution shown in Fig.~\ref{fig:new_order_execs}.
-
-\begin{figure}[!t]
+permitted, leading to executions superficially similar to executions
+permitted by concurrent (racy) programs~\cite{GHE15,HPQ+15}.
+\begin{wrapfigure}{l}{.4\textwidth}
 \includegraphics[scale=0.45]{Figures/motiv-eg-1-b}
@@
   \C{d\_next\_o\_id} prompts both instances to add \C{Order} records
-  with same ids, which inturn triggers the the violation of TPC-C's
+  with same ids, which in turn triggers a violation of TPC-C's
   consistency condition.}
 \label{fig:new_order_exec}
-\end{figure}
+\end{wrapfigure}
+To illustrate, consider the behavior of the \C{new\_order} transaction
+when executed with a \emph{Read Committed} (RC) isolation level, the
+default isolation level in 8 of the 18 databases studied
+in~\ref{bailishat}.  An RC transaction is isolated from \emph{dirty
+  writes}, i.e., writes of uncommitted transactions, but is allowed to
+witness the writes of concurrent transactions as soon as they are
+committed. Thus, with two concurrent instances of the \C{new\_order}
+transaction (call them $T_1$ and $T_2$), both concurrently placing new
+orders for different customers under the same district (\C{d\_id}), RC
+isolation allows the execution shown in
+Fig.~\ref{fig:new_order_execs}.
 
@@
 \emph{lost writes}; both concurrent transactions' writes are present
-in the final state of the database. Thus, program analyses that aim to
-determine appropriate isolation by checking for possible
-manifestations of these anomalies would fail to identify grounds for
-promoting the isolation level of \C{new\_order} to something stronger.
-Yet, if we take the semantics of the application into account, it is
-quite clear that RC is not an appropriate isolation level for
-\C{new\_order}.
+in the final state of the database. Indeed, program analyses that aim
+to determine appropriate isolation by checking for possible
+manifestations of RC-induced anomalies would fail to identify grounds
+for promoting the isolation level of \C{new\_order} to something
+stronger.  Yet, if we take the semantics of the application into
+account, it is quite clear that RC is not an appropriate isolation
+level for \C{new\_order}.
 
@@
 traces~\cite{adyaphd,gotsmanconcur15} on memory read and write actions
-complicates high-level reasoning.  A possible alternative would
+can complicate high-level reasoning.  A possible alternative would
 interleave weak isolation implementation details within the program,
@@
 \C{new\_order} is sound when executed under \emph{Snapshot Isolation}
-(SI), and isolation level stronger than RC. Snapshot isolation
-allows transactions to be executed against their respective private
-snapshots of the database, thus admitting concurrency, but it also
-requires that there not be any write-writeconflicts among
-concurrent transactions. Write-write conflicts can be eliminated in various
-ways; either through detection followed by a rollback, or through
-exclusive locks, or a combination of both, leading to different ways
-in which SI can be realized. For instance, one possible implementation
-of SI, close to the one use by PostgreSQL~\cite{postgres-ssi},
-executes a transaction against its private snapshot of the database,
-but obtains exclusive locks on the actual records in the database
-before performing writes. A write is performed only of the record
-hasn't already been updated by a concurrent transaction, i.e,  when it
-doesn't result in a write-write conflict.  Otherwise, the transaction is
-rolledback. This implementation is demonstrated for a simple database
-with three records, $a$, $b$, and $c$ in Fig.~\ref{fig:RR-postgres}.
+(SI), a stronger isolation level than RC. Snapshot isolation allows
+transactions to be executed against a private snapshot of the
+database, thus admitting concurrency, but it also requires that there
+not be any write-write conflicts among concurrent transactions when
+they commit. Write-write conflicts can be eliminated in various ways,
+e.g., through conflict detection followed by a rollback, or through
+exclusive locks, or a combination of both.  For instance, one possible
+implementation of SI, close to the one used by
+PostgreSQL~\cite{postgres-ssi}, executes a transaction against its
+private snapshot of the database, but obtains exclusive locks on the
+actual records in the database before performing writes. A write is
+performed only if the record that is to be written has not already
+been updated by a concurrent transaction. Conflicts are resolved by
+abort and roll back.
 
-\begin{figure}[t]
-\includegraphics[scale=0.5]{Figures/RR-postgres}
-\caption{Database state transitions corresponding to an execution of
-  an SI transaction $T_1$ on a PostgreSQL-like store. $T_2$ and $T_3$
-  are concurrent transactions. }
-\label{fig:rr-postgres}
-\end{figure}
-\begin{figure}[t]
-\includegraphics[scale=0.5]{Figures/RR-abstract}
-  \caption{An abstract execution that includes the concrete execution
-  shown in Fig.~\ref{fig:rr-postgres}. It has no locks or snapshots,
-  admits fewer interleavings, yet results in the same post-state.  }
-\label{fig:rr-abstract}
-\end{figure}
+%% This implementation is demonstrated for a simple
+%% database with three records, $a$, $b$, and $c$ in
+%% Fig.~\ref{fig:RR-postgres}.
+
+%% \begin{figure}[t]
+%% \includegraphics[scale=0.5]{Figures/RR-postgres}
+%% \caption{Database state transitions corresponding to an execution of
+%%   an SI transaction $T_1$ on a PostgreSQL-like store. $T_2$ and $T_3$
+%%   are concurrent transactions. }
+%% \label{fig:rr-postgres}
+%% \end{figure}
+%% \begin{figure}[t]
+%% \includegraphics[scale=0.5]{Figures/RR-abstract}
+%%   \caption{An abstract execution that includes the concrete execution
+%%   shown in Fig.~\ref{fig:rr-postgres}. It has no locks or snapshots,
+%%   admits fewer interleavings, yet results in the same post-state.  }
+%% \label{fig:rr-abstract}
+%% \end{figure}
 
@@
 complicated, often running into thousands of lines of code.
-Nonetheless, the semantics of Snapshot Isolation, in terms of how it
-effects the transitions on the database state, can be captured in a
-fairly simple model. Firstly, although SI admits concurrent
-transactions, their effects are  not witnessed in the current
-transaction (call it $T$), during $T$'s execution. Thus, insofar as
-$T$ is concerned, as long as it executes, the global state doesn't
-change. More formally, for every consecutive pair of global states
-$(\stg,\stg')$ witnessed by $T$ during its execution, $\stg'=\stg$.
-After $T$ finishes execution, it commits to the actual database, which
-incorporates the effects of concurrent transactions. In executions
-where $T$ successfully commits, concurrent transactions are guaranteed
-to not be in write-write conflict with $T$. Thus, if $\stg$ is the
-global state that $T$ witnessed when it finished execution (which
-would be same as the snapshot state), and $\stg'$ is the state to
-which $T$ commits, then the diff between $\stg$ and $\stg'$ shouldn't
-be in a write-write conflict with $T$. To concretize this notion, let
-the database state be a map from transaction variables to values, and
-let $\stl$ denote a transaction-local log that maps the variables
-being written to their updated values. Then the absence of write-write
-conflicts between $T$ and the diff between $\stg$ and $\stg'$ can be
-defined as: $\forall x\in\mathit{dom}(\delta)$, $\stg'(x) = \stg(x)$.
-To summarize, the semantics of SI can be captured as an axiomatization
-over transitions of the database state ($\Delta \longrightarrow
-\Delta'$) during the lifetime an SI transaction ($T$):
+Nonetheless, the semantics of SI, in terms of how it effects
+transitions on the database state, can be captured in a fairly simple
+model. First, effects induced by one transaction are not visible to
+another concurrently executing one (call it $T$), during $T$'s
+execution. Thus, from $T$'s perspective, the global state does not
+change during its execution.  More formally, for every operation
+performed by $T$, the global state $T$ witness before ($\stg$) and
+after ($\stg')$ executing the operation is the same ($\stg'=\stg$).
+After $T$ finishes execution, it commits its changes to the actual
+database, which may have already incorporated the effects of concurrent
+transactions. In executions where $T$ successfully commits, concurrent
+transactions are guaranteed to not be in write-write conflict with
+$T$. Thus, if $\stg$ is the global state that $T$ witnessed when it
+finished execution (the snapshot state), and
+$\stg'$ is the state to which $T$ commits, then the difference between
+$\stg$ and $\stg'$ should not result in a write-write conflict with $T$. To
+concretize this notion, let the database state be a map from
+transaction variables to values, and let $\stl$ denote a
+transaction-local log that maps the variables being written to their
+updated values.  The absence of write-write conflicts between $T$
+and the diff between $\stg$ and $\stg'$ can be expressed as: $\forall
+x\in\mathit{dom}(\delta)$, $\stg'(x) = \stg(x)$.  In other words, the
+semantics of SI can be captured as an axiomatization over transitions
+of the database state ($\Delta \longrightarrow \Delta'$) during the
+lifetime an SI transaction ($T$):
 \begin{itemize}
@@
 
-This simple characterization of SI isolation allows us to verify the
-consistency conditions associated with the \C{new\_order} transaction.
-First, since the database doesn't change ($\Delta' = \Delta$) during
-the execution of the transaction's body, we can reason about
-\C{new\_order} as though it executed in complete isolation until its
-commit point, leading to a verification process similar to what would
-been applied when reasoning about serializability.  When
+\noindent This simple characterization of SI isolation allows us to
+verify the consistency conditions associated with the \C{new\_order}
+transaction.  First, since the database doesn't change ($\Delta' =
+\Delta$) during the execution of the transaction's body, we can reason
+about \C{new\_order} as though it executed in complete isolation until
+its commit point, leading to a verification process similar to what
+would been applied when reasoning about serializability.  When
 \C{new\_order} finishes execution, but before it commits, the SI
@@
 to the records ($\stl$) written by \C{new\_order}, i.e.,
-$\forall(x\in\delta).~\Delta'(x) = \Delta(x)$. Since concurrent
-\C{new\_order} transactions that write to the same \C{District} record
-modify the record by incrementing its \C{d\_next\_o\_id} field, their
-transitions do not satisfy the necessary condition imposed by the SI
-axiomatization, allowing us to rightfully ignore their interference.
-This leaves the interference due to concurrent \C{new\_order}
-transactions that write to a different district record.  Fortunately,
-such interference does not prompt \C{new\_order} to violate the
-invariant, which can be established by adopting a reasoning similar to
-the one used on concurrent program logics, such as the Rely-Guarantee
-logic, customized for the database programs.
+$\forall(x\in\delta).~\Delta'(x) = \Delta(x)$.  The axiomatization
+precludes any execution in which there may concurrent updates to
+shared table fields (e.g., \C{d\_next\_o\_id} on the same \C{District}
+table), but does not prohibit interferences that write to different
+tables, or different records in the same table.  We need to reason
+about the safety of such interferences with respect to the
+transaction's consistency invariants.
+
+%% Since concurrent
+%% \C{new\_order} transactions that write to the same \C{District} record
+%% modify the record by incrementing its \C{d\_next\_o\_id} field, their
+%% transitions do not satisfy the necessary condition imposed by the SI
+%% axiomatization, allowing us to rightfully ignore their interference.
+%% This leaves the interference due to concurrent \C{new\_order}
+%% transactions that write to a different district record.  Fortunately,
+%% such interference does not prompt \C{new\_order} to violate the
+%% invariant, which can be established by adopting a reasoning similar to
+%% the one used on concurrent program logics, such as the Rely-Guarantee
+%% logic, customized for the database programs.
 
@@
 
-The second observation that informs our approach is one that pertains
-to automation. Program verification, even when machine-aided, often
-entails significant annotation burden in the form of intermediary
-assertions and loop invariants required to prove a program correct.
-This is certainly true for concurrent program logics, such as
-Rely-Gurantee, which extend Hoare logic with additional artifacts and
-where (stable) intermediary assertions and loop invariants remain a
-major source of annotation burden.  However, a relational database is
-a significantly simpler abstraction than shared memory. There are no
-pointers, linked data structures, or aliasing.  Although a database
-essentially abstracts a mutable state, the state is mutated through a
+%% The second observation that informs our approach is one that pertains
+%% to automation. Program verification, even when machine-aided, often
+%% entails significant annotation burden in the form of intermediary
+%% assertions and loop invariants required to prove a program correct.
+%% This is certainly true for concurrent program logics, such as
+%% Rely-Gurantee, which extend Hoare logic with additional artifacts and
+%% where (stable) intermediary assertions and loop invariants remain a
+%% major source of annotation burden.
+
+To do so, we observe that a relational database is a significantly
+simpler abstraction than shared memory. There are no pointers, linked
+data structures, or aliasing.  Although a database essentially
+abstracts a mutable state, the state is affected through a
 well-defined fixed number of interfaces (SQL statements), each tagged
@@
 them as essentially functional computations that manage database state
-monadically. We find it useful to reason about
-statements that mutate the database state, not in terms of a pre- and
-post-condition pair, but in terms of a state transformer that relates
-the pre- and post-states of a statement. This state transformer
-semantics can be defined algorithmically, just like predicate
-transformer semantics (e.g., strongest post-condition).  Here, a state
-transformer interprets a SQL statement in the set domain, taking
-advantage of the fact that the database is essentially a set of
-tuples, and a SQL statement is a transformer over these sets.  The
-benefit of this approach is that low-level loops can now be
-substituted with higher-order combinators that automatically lift the
-state transformer of its higher-order argument (i.e., the loop body)
-to the state transformer of the combined expression (i.e., the loop).
-Thus the semantics of a \C{foreach} loop, for instance, can be
-captured as a state transformer, where the state is a set, and the
-transformation defines a bind operation. We illustrate this intuition
-on a simple example.
+monadically.  By doing so, we can formulate the semantics of database
+operations as state transformers that relate the operation's pre- and
+post-states.  A state transformer semantics can be defined
+algorithmically, just like classical predicate transformer semantics
+(e.g., strongest post-condition).  Here, a transformer interprets a
+SQL statement in the set domain, modeling the database as a set of
+tuples, and a SQL statement a function over this set.  Among other
+things, one benefit of this approach is that low-level loops can now
+be substituted with higher-order combinators that automatically lift
+the state transformer of its higher-order argument (i.e., the loop
+body) to the state transformer of the combined expression (i.e., the
+loop).  We illustrate this intuition on a simple example.
+
+%% Thus the semantics of a \C{foreach} loop, for instance, can be
+%% captured as a state transformer, where the state is a set, and the
+%% transformation defines a bind operation.
 

### 2017-07-07 implementation.tex

\section{Implementation}
\label{sec:implementation}

\input{evaluation}

### 2017-07-07 motivation.tex

@@
 % \end{verbatim}
-Observe that
-$T_U(\Delta,\delta)$ is of the form $\delta \,\cup\,
-F_U(\Delta)$. The transformer ($T_I(\Delta,\delta)$)
-for the subsequent \C{insert} statement can be similarly calculated to
-be of the form $\delta \cup F_I(\Delta)$.
-Their composition gives the state transformer of the loop body:
-\begin{ocaml}
-  fun ($\Delta$,$\delta$) -> $\delta$ $\cup$ $F_U(\Delta) \,\cup\, F_I(\Delta)$
-\end{ocaml}
-Let $F_{body}(\Delta) = F_U(\Delta) \,\cup\, F_I(\Delta)$.   The
-transformer for the \C{foreach} loop can now be computed as follows:
-\begin{ocaml}
-  fun ($\Delta$,$\delta$) -> $\delta$ $\cup$ item_reqs$\bind$(fun item_req -> $F_{body}(\Delta)$)
-\end{ocaml}
-Observe that the transformer captures the precise semantics of the
-loop as a formula in the domain of sets extended with a bind
-function. The advantage in doing so is that we can now make use of a
-semantics-preserving translation from this domain to first-order
-logic, allowing us to leverage SMT solvers for automatic proofs
-without having to infer loop invariants, or provide intermediate
-assertions.  Sec.~\ref{sec:automation} describes this translation. In
-the exposition thus far, we assumed $\Delta$ remains invaraint, which
-is clearly not the case when we admit concurrency.  Necessary
-concurrency extensions of the state transformer semantics is also
-covered in Sec.~\ref{sec:automation}.  The following two sections,
-however, first focus on laying the theoretical foundations for
-reasoning about weak isolation.
+% Observe that
+% $\T_U(\Delta,\delta)$ is of the form $\delta \,\cup\,
+% F_U(\Delta)$. 
+The transformer ($\T_I(\Delta,\delta)$) for the subsequent \C{insert}
+statement can be similarly constructed:
+\begin{smathpar}
+\begin{array}{l}
+  \lambda(\stl,\stg).~ \stl \cup \{\C{ol\_o\_id}=\C{k2};\, 
+        \C{ol\_d\_id}=\C{k3};\, \C{ol\_i\_id}=\C{item\_req.ol\_i\_id};\, 
+        \C{ol\_qty}=\C{item\_req.ol\_qty}\}
+\end{array}
+\end{smathpar}
+Observe that both transformers are of the form $T(\Delta,\delta) =
+\stl \cup F(\stg)$, where $F$ is a function that returns the set of
+records added by the transformer to the transaction-local database
+($\stl$). Let $\F_U$ and $\F_I$ be the corresponding functions for $\T_U$
+and $\T_I$ shown above. The state transformed of the loop body in
+Fig.~\ref{fig:new} can be expressed as the following composition of
+$\F_U$ and $\F_I$:
+\begin{smathpar}
+\begin{array}{l}
+  \lambda(\stl,\stg).~ \stl \cup \F_U(\stg) \cup \F_I(\stg)
+\end{array}
+\end{smathpar}
+The transformer for the loop itself can now be computed to be:
+\begin{smathpar}
+\begin{array}{l}
+  \lambda(\stl,\stg).~ \stl \cup \C{item\_reqs}\bind
+      (\lambda\C{item\_req}.~ \F_U(\stg) \cup \F_I(\stg))
+\end{array}
+\end{smathpar}
+Observe that the structure of the transformer mirrors the structure of
+the program itself. In particular, SQL statements become set
+operations, and the \C{DB} monad's $\bind$ becoms set monadic $\bind$.
+Thus, state transformer inference is completely mechanical. The
+advantage of inferring such transformers is that we can now make use
+of a semantics-preserving translation from the domain of sets equipped
+with $\bind$ to first-order logic, allowing us to leverage SMT solvers
+for automatic proofs without having to infer loop invariants, or
+provide intermediate assertions.  Sec.~\ref{sec:inference} describes
+this translation. In the exposition thus far, we assumed $\Delta$
+remains invaraint, which is clearly not the case when we admit
+concurrency.  Necessary concurrency extensions of the state
+transformer semantics to deal with interference is also covered in
+Sec.~\ref{sec:inference}.  The following two sections, however, first
+focus on laying the theoretical foundations for reasoning about weak
+isolation.

### 2017-07-07 motivation.tex
Edit.

--- 2017-07-07 motivation.tex
@@
 \begin{ocaml}
-let new_order d_id c_id item_reqs = atomically do
+let new_order (d_id, c_id, item_reqs) = atomically do
   dist <- SQL.select1 District (fun d -> d.d_id = d_id);

### 2017-07-07 motivation.tex
edits

--- 2017-07-07 motivation.tex
@@
 	\end{subfigure}
+  \hspace*{.4in}
 	\begin{subfigure}{0.52\textwidth}
@@
    	\end{subfigure}
-
 \caption{Database schema of TPC-C's order management system.
@@
 permitted by concurrent (racy) programs~\cite{GHE15,HPQ+15}.
+To illustrate, consider the behavior of the \C{new\_order} transaction
 \begin{wrapfigure}{l}{.4\textwidth}
@@
 \end{wrapfigure}
-To illustrate, consider the behavior of the \C{new\_order} transaction
 when executed with a \emph{Read Committed} (RC) isolation level, the
 default isolation level in 8 of the 18 databases studied
-in~\ref{bailishat}.  An RC transaction is isolated from \emph{dirty
+in~\cite{bailishotos}.  An RC transaction is isolated from \emph{dirty
   writes}, i.e., writes of uncommitted transactions, but is allowed to
@@
 orders for different customers under the same district (\C{d\_id}), RC
-isolation allows the execution shown in
-Fig.~\ref{fig:new_order_execs}.
+isolation allows the execution shown in Fig.~\ref{fig:new_order_exec}.
 
@@
 implementation of SI, close to the one used by
-PostgreSQL~\cite{postgres-ssi}, executes a transaction against its
+PostgreSQL~\cite{postgresiso}, executes a transaction against its
 private snapshot of the database, but obtains exclusive locks on the
@@
 
-Implementations of SI on real databases (e.g., PostgreSQL) are
-complicated, often running into thousands of lines of code.
-Nonetheless, the semantics of SI, in terms of how it effects
-transitions on the database state, can be captured in a fairly simple
-model. First, effects induced by one transaction are not visible to
-another concurrently executing one (call it $T$), during $T$'s
-execution. Thus, from $T$'s perspective, the global state does not
-change during its execution.  More formally, for every operation
-performed by $T$, the global state $T$ witness before ($\stg$) and
-after ($\stg')$ executing the operation is the same ($\stg'=\stg$).
-After $T$ finishes execution, it commits its changes to the actual
-database, which may have already incorporated the effects of concurrent
-transactions. In executions where $T$ successfully commits, concurrent
-transactions are guaranteed to not be in write-write conflict with
-$T$. Thus, if $\stg$ is the global state that $T$ witnessed when it
-finished execution (the snapshot state), and
-$\stg'$ is the state to which $T$ commits, then the difference between
-$\stg$ and $\stg'$ should not result in a write-write conflict with $T$. To
-concretize this notion, let the database state be a map from
-transaction variables to values, and let $\stl$ denote a
+As this discussion indicates, implementations of SI on real databases
+such as PostgresSQL are highly complicated, often running into
+thousands of lines of code.  Nonetheless, the semantics of SI, in
+terms of how it effects transitions on the database state, can be
+captured in a fairly simple model. First, effects induced by one
+transaction are not visible to another concurrently executing one
+(call it $T$), during $T$'s execution. Thus, from $T$'s perspective,
+the global state does not change during its execution.  More formally,
+for every operation performed by $T$, the global state $T$ witness
+before ($\stg$) and after ($\stg')$ executing the operation is the
+same ($\stg'=\stg$).  After $T$ finishes execution, it commits its
+changes to the actual database, which may have already incorporated
+the effects of concurrent transactions. In executions where $T$
+successfully commits, concurrent transactions are guaranteed to not be
+in write-write conflict with $T$. Thus, if $\stg$ is the global state
+that $T$ witnessed when it finished execution (the snapshot state),
+and $\stg'$ is the state to which $T$ commits, then the difference
+between $\stg$ and $\stg'$ should not result in a write-write conflict
+with $T$. To concretize this notion, let the database state be a map
+from transaction variables to values, and let $\stl$ denote a
 transaction-local log that maps the variables being written to their
-updated values.  The absence of write-write conflicts between $T$
-and the diff between $\stg$ and $\stg'$ can be expressed as: $\forall
+updated values.  The absence of write-write conflicts between $T$ and
+the diff between $\stg$ and $\stg'$ can be expressed as: $\forall
 x\in\mathit{dom}(\delta)$, $\stg'(x) = \stg(x)$.  In other words, the
@@
 records added by the transformer to the transaction-local database
-($\stl$). Let $\F_U$ and $\F_I$ be the corresponding functions for $\T_U$
-and $\T_I$ shown above. The state transformed of the loop body in
-Fig.~\ref{fig:new} can be expressed as the following composition of
-$\F_U$ and $\F_I$:
+($\stl$). Let $\F_U$ and $\F_I$ be the corresponding functions for
+$\T_U$ and $\T_I$ shown above. The state transformed of the loop body
+in Fig.~\ref{fig:new_order_code} can be expressed as the following
+composition of $\F_U$ and $\F_I$:
 \begin{smathpar}
@@
 the program itself. In particular, SQL statements become set
-operations, and the \C{DB} monad's $\bind$ becoms set monadic $\bind$.
-Thus, state transformer inference is completely mechanical. The
-advantage of inferring such transformers is that we can now make use
-of a semantics-preserving translation from the domain of sets equipped
-with $\bind$ to first-order logic, allowing us to leverage SMT solvers
-for automatic proofs without having to infer loop invariants, or
-provide intermediate assertions.  Sec.~\ref{sec:inference} describes
-this translation. In the exposition thus far, we assumed $\Delta$
-remains invaraint, which is clearly not the case when we admit
-concurrency.  Necessary concurrency extensions of the state
-transformer semantics to deal with interference is also covered in
-Sec.~\ref{sec:inference}.  The following two sections, however, first
-focus on laying the theoretical foundations for reasoning about weak
-isolation.
+operations, and the \C{DB} monad's $\bind$ becoms set monadic $\bind$,
+leading to a completely mechanical inference procedure.  The advantage
+of inferring such transformers is that we can now make use of a
+semantics-preserving translation from the domain of sets equipped with
+$\bind$ to first-order logic, allowing us to leverage SMT solvers for
+automated proofs without having to infer potentially complex
+thread-local invariants or intermediate assertions.
+Sec.~\ref{sec:inference} describes this translation. In the exposition
+thus far, we assumed $\Delta$ remains invariant, which is clearly not
+the case when we admit concurrency.  Necessary concurrency extensions
+of the state transformer semantics to deal with interference is also
+covered in Sec.~\ref{sec:inference}.  The following two sections,
+however, first focus on laying the theoretical foundations for
+reasoning about weak isolation.

### 2017-07-07 opsem.tex
Edits.
edits

@@
 ($c_1\,||\,c_2$). Each transaction is assumed to have a unique
-identifier $i$. The $\I$ in the \C{TXN} block syntax is the
-transaction's isolation specification, which will be explained later.
-Certain terms that only appear at run-time are also present in $c$.
-These include a {\sf txn} block tagged with sets ($\stl$ and $\stg$) of
-records whose meaning is explained below, and a runtime {\sf foreach}
-expression that keeps track of the set ($s_1$) of items already
-iterated, and the set ($s_2$) of items yet to be iterated. Note that
-even the surface-level syntax of the \C{FOREACH} command shown here is
-a little different from the one used in previous sections; its
-higher-order function has two arguments, $y$ and $z$, which are
-invoked (during the reduction) with the set of already-iterated items,
-and the current item, respectively. This form of \C{FOREACH} lends
-itself to inductive reasoning that will be useful for verification
-(Sec.~\ref{sec:reasoning}).
+identifier $i$, and executes at the top-level; our semantics does not
+supported nested transactions. The $\I$ in the \C{TXN} block syntax is
+the transaction's isolation specification, which will be explained
+later.  Certain terms that only appear at run-time are also present in
+$c$.  These include a {\sf txn} block tagged with sets ($\stl$ and
+$\stg$) of records whose meaning is explained below, and a runtime
+{\sf foreach} expression that keeps track of the set ($s_1$) of items
+already iterated, and the set ($s_2$) of items yet to be
+iterated. Note that even the surface-level syntax of the \C{FOREACH}
+command shown here is a little different from the one used in previous
+sections; its higher-order function has two arguments, $y$ and $z$,
+which are invoked (during the reduction) with the set of
+already-iterated items, and the current item, respectively. This form
+of \C{FOREACH} lends itself to inductive reasoning that will be useful
+for verification (Sec.~\ref{sec:reasoning}).
 
@@
 the identifier of the transaction that last updated the record, and a
-$\delf$ field that flags the deleted records in $\stl$.  For a set $S$
-of records, we define $\dom(S)$ as the set of unique ids of all
-records in $S$. Thus $|\dom(\stg)| = |\stg|$. During its execution, a
+$\delf$ field that flags deleted records in $\stl$.  For a set $S$ of
+records, we define $\dom(S)$ as the set of unique ids of all records
+in $S$. Thus $|\dom(\stg)| = |\stg|$. During its execution, a
 transaction may write to multiple records in $\stg$. Atomicity
@@
 automation framework introduced in Sec.~\ref{sec:inference} has a
-non-trivial dependence on this assumption. Nonetheless, our experience
-suggests that local context independence is not a severe restriction
-in practice. For instance, all 17 transactions in the TPC-C and TPC-E
+non-trivial dependence on this assumption. Nonetheless, we haven't
+found local context independence to be a severe restriction in
+practice. For instance, all 17 transactions in the TPC-C and TPC-E
 benchmarks satisfy this assumption.

### 2017-07-07 implementation.tex
Implementation sec.

@@
 
+\begin{figure}
+\begin{ocaml}
+type table_name =  District | Order | Order_line | Stock
+
+type district = {d_id: int; d_next_o_id: int}
+type order = {o_id: int; o_d_id: int; o_c_id: int; o_ol_cnt: int}
+type order_line = {ol_o_id: int; ol_d_id: int; ol_i_id: int; ol_qty: int}
+type stock = {s_i_id: int; s_d_id:int; s_qty: int}
+\end{ocaml}
+\caption{OCaml type definitions corresponding to the TPC-C schema from
+Fig.~\ref{fig:schema1}}
+\label{fig:ocaml-schema}
+\end{figure}
+
+We have implemented our DSL to define transactions as monadic
+computations in OCaml (modulo the syntactic sugar), and our automatic
+reasoning framework as an extra frontend pass (called \tool) in ocamlc
+4.03 compiler\footnote{The source code is available at available at
+\emph{\ldots}}. The input to \tool is a program in our DSL that
+describes the schema of the database as a collection of OCaml type
+definitions, and transactions as OCaml functions, whose top-level
+expression is an application of the \C{atomically\_do} combinator to a
+function of type $a \rightarrow \C{DB}\;b$, for some (base) types $a$
+and $b$. For instance, TPC-C's schema from Fig.~\ref{fig:schema1} can
+be described via the OCaml type definitions shown in
+Fig.~\ref{fig:ocaml-schema}.  \tool also requires a specification of
+the program in the form of a collection of guarantees ($G$), one per
+transaction, and an invariant $I$ that is a conjunction of the
+integrity constraints on the database. An auxiliary DSL that includes
+the first-order logic (FOL) combinators has been implemented for this
+purpose. \tool's verification pass follows OCaml's type checking pass,
+hence the concrete artifact of verification is OCaml's typed AST. The
+tool is already equipped with  an axiomatization of PostgresSQL and
+MySQL's isolation levels expressed in our FOL DSL. Other data stores
+can be similarly axiomatized. The concrete result of verification is
+an assignment of an isolation level of the selected data store to each
+transaction in the program.
+
+At the top-level, \tool runs a loop that pics an unverified
+transaction and progressively strengthens its isolation level until it
+passes the verification. If the selected data store provides
+serializable isolation level, and if the program is sequentially
+corred, then the verification is guaranteed to succeed. Within the
+loop, the \tool first computes the various rely relations needed for
+verification ($R$, $\R_l$, and $\R_c$). It then traverses the AST of a
+transaction, applying the inference rules to construct a state
+tansformer, check its stability, and weaken it ($\stabilize{\cdot}$)
+if it is not stable. The result of traversing the transaction's AST is
+therefore a stable state transformer ($\F$), which is then checked
+against the transaction's stated guarantee ($G$). If the check passes,
+then the guarantee is verified to check if it preserves the invariant
+$I$. The successful result from both the checks results in the
+transaction being certified correct under the current choice of its
+isolation level. Succesful verification of all transactions concludes
+the top-level execution, returning the inferred isolation levels as
+its output.
+
+\tool uses Z3 SMT solver as its underlying reasoning engine. Each
+implication check described above is first encoded in the FOL,
+applying the translation described in Sec.~\ref{sec:inference}
+wherever necessary, and the validity of the resultant formula ($\phi$)
+is decided by asking Z3 to check the satisfiability of its negation;
+unsatisfiability of $\neg\phi$ proves the validity.
+
+\subsection{Pragmatics}
+
+\textbf{Real-World Isolation Levels} The axiomatization of the
+isolation levels presented in Sec.~\ref{sec:isolation} leave out
+certain nuances of their implementations on real data stores, which
+need to be taken into account for the verification to be effective in
+practice.  We consider these into account while linking \tool with
+store-specific semantics (isolation specifications etc). As an
+example, consider how PostgresSQL implements the \C{UPDATE} operation.
+\C{UPDATE} first selects the records that meet the search criteria
+from the snapshot against which it is executing (the snapshot is
+established at the beginning of the transaction if the isolation level
+is RR, or at the beginning of the \C{UPDATE} statement if the
+isolation level is RC). The selected records are then visited in the
+actual database (if it still exists), write locks are obtained, and
+the updatation is carried out, provided that the record still meets
+\C{UPDATE}'s search criteria. If the record in no longer meets the
+search criteria (due to a concurrent update) the record is excluded
+from the update, and the write lock is immediately released.
+Otherwise, the record remains locked till the transaction commits.
+
+Clearly, the \C{UPDATE} operation on PostgreSQL is not atomic, as the
+our formal model assumes it to be. It admits interference between the
+updatation of individual records that meet the search criteria.
+Nonetheless, through a series of relatively straightforward
+deductions, we can show that PostgreSQL's \C{UPDATE} is infact
+equivalent (in semantics) to a sequential composition of two atomic
+operations $c_1;c_2$, where $c_1$ is effectively a \C{SELECT}
+operation with the same search criteria as \C{UPDATE}, and $c_2$ is
+the a slight variation of the original \C{UPDATE} that updates a
+record only if a record with the same id is present in the records
+returned by the \C{SELECT}. This transformation is summarized below:
+\begin{smathpar}
+\begin{array}{lcl}
+\updatee{(\lambda x. ~e_1)}{(\lambda x.~e_2)}
+&
+\longrightarrow
+&
+\lete{y}{\selecte{(\lambda x.~e_1})}
+     {\updatee{(\lambda x.~e_1 \wedge x.\idf\in\dom(y))}
+              {(\lambda x.~e_2})}\\
+\end{array}
+\end{smathpar}
+The intuition behind this translation is the observation that all
+interferences possible during the execution of the \C{UPDATE} can be
+accommodated between the time the records are selected from the
+snapshot, and the time they are actually updated.  \tool performs this
+translation if the selected store is PostgreSQL, allowing it to reason
+about \C{UPDATE} operations in a way that is faithful to its semantics
+on PostgreSQL. \tool also admits similar compensatory logic for
+certain combinations of isolation levels and operations on MySQL.
+
+\textbf{Set functions} SQL's \C{SELECT} query admits projections of
+record fields, and also application of auxiliary functions such as
+\C{MAX} and \C{MIN}. For example, \C{SELECT MAX(ol\_o\_id) FROM
+Order\_line WHERE $\ldots$}. We admit such extensions as set functions
+in our DSL (e.g., \C{project}, \C{max}, \C{min}), and axiomatize their
+behavior. For instance:
+\begin{smathpar}
+\begin{array}{lcl}
+  s_2 \;=\;\C{project}\,s_1\,(\lambda z.~e) & \Leftrightarrow &
+  \forall y.~y\in s_2 \Leftrightarrow  \exists(x \in s_1).~y = [x/z]e\\
+  x \;=\; \C{max}\,s & \Leftrightarrow & x \in s \conj \forall(y \in
+  s).~y\le x\\
+\end{array}
+\end{smathpar}
+There are however certain set functions that
+
 \input{evaluation}

### 2017-07-07 opsem.tex
edits

@@
 \txnimp, a core language that we use to formalize our intuitions about
-reasoning under weak isolation. Variables ($x$), Integer and boolean
+reasoning under weak isolation. Variables ($x$), integer and Boolean
 constants ($k$), records ($r$) of named constants, sets ($s$) of such
@@
 
-The small-step semantics is stratified into a transaction-local reduction
-relation, and a top-level reduction relation. The transaction-local
-relation ($\stg \vdash (c,\stl) \stepsto (c',\stl')$) defines a
-small-step reduction for a command inside a transaction, when the
-database state is $\stg$; the command $c$ reduces to $c'$, while
-updating the transaction-local database $\stl$ to $\stl'$. The
-definition assumes a meta function $\eval$ that evaluates expressions
-with no free variables to values. The reduction relation for SQL
-statements is defined straightforwardly. \C{INSERT} adds a new record
-to $\stl$ after checking the uniqueness of its id. \C{DELETE} finds
-the records in $\stg$ that match the search criteria defined by its
-boolean function argument, and adds the records to $\stl$ after
-marking them for deletion. \C{SELECT} bounds the name introduced by
-\C{LET} to the set of records from $\stg$ that match the search
-criteria, and then executes the bound command $c$. \C{UPDATE} uses its
-first function argument to compute the updated version of the records
-that match the search criteria defined by its second function
-argument. Updated records are added to $\stl$.
+The small-step semantics is stratified into a transaction-local
+reduction relation, and a top-level reduction relation. The
+transaction-local relation ($\stg \vdash (c,\stl) \stepsto
+(c',\stl')$) defines a small-step reduction for a command inside a
+transaction, when the database state is $\stg$; the command $c$
+reduces to $c'$, while updating the transaction-local database $\stl$
+to $\stl'$. The definition assumes a meta function $\eval$ that
+evaluates expressions with no free variables to values. The reduction
+relation for SQL statements is defined straightforwardly. \C{INSERT}
+adds a new record to $\stl$ after checking the uniqueness of its
+id. \C{DELETE} finds the records in $\stg$ that match the search
+criteria defined by its boolean function argument, and adds the
+records to $\stl$ after marking them for deletion. \C{SELECT} bounds
+the name introduced by \C{LET} to the set of records from $\stg$ that
+match the search criteria, and then executes the bound command
+$c$. \C{UPDATE} uses its first function argument to compute the
+updated version of the records that match the search criteria defined
+by its second function argument. Updated records are added to $\stl$.
 

### 2017-07-07 evaluation.tex

\subsection{Evaluation}

To explore the space of correct isolation levels for different transactions, we  also  ran a number  of experiments using  a real transaction OLTP workload   based on TPC-C benchmark.  
 The  workload  involves  a mix of five TPC-C transaction types: 
\C{New\_Order},  \C{Payment},  \C{Delivery}, \C{Stock Level} and  \C{Order Status}, 
operating on top of  a $10$-warehouse database.  
Each warehouse has  $10$ districts, each serving $3000$ customers. 
TPC-C  specification model  requires  maintaining multiple   consistency conditions.  We focus only on   four conditions, named $I_1$, $I_2$, $I_3$, and $I_4$.   
Consistency condition $I_1$  requires that  sales of each  warehouse  equals to the  sum of    its  district's sales.  Conditions $I_2$ and $I_3$  enforce  
 sequentially unique IDs assigned to  orders placed in the same   warehouse and district. Condition $I_4$ requires 
 the  number of  item orders   in each district 
  reflexes  the district's roll-up. 
Given the consistency conditions,  Table  \ref{tab:tpcc-eval} gives the isolation level  for the \C{New\_Order},  \C{Payment},  \C{Delivery} transactions using the  MySql and Postrgres databases   to maintain   each consistency condition. 
Recall that the \C{New\_Order} transaction places a new order for
 items from a particular district in a warehouse.  
The \C{Payment} transaction  processes  a payment for a particular customer and updates the customer, district, and warehouse tables. 
 The  \C{Delivery}  transaction processes  orders for a particular   district at  a warehouse. 
 Our experimental results  shows that  the concurrent executions of   \C{New\_Order}   
    and  \C{Delivery} transactions under a weak isolation level (i.e.,  $RC$) is anomalous, i.e., it may violate the consistency conditions $I_2$ and $I_3$. Depending on the isolation guarantees provided  by  underlying database,   the stronger  isolation level is required. For instance, the isolation level $RR$ in Postgres  is sufficient to maintain the condition while   MySql database  requires  $SER$.

 	\begin{table*}[t]\small
	\centering
\begin{tabular}{|c|c|c|c|}
		\hline
	\textbf{Transaction}   & \textbf{Invariant} 
	& \textbf{MYSQL-Isolation} & \textbf{Postgres-Isolation} \\ 
	\hline
	\multirow{4}{*}{new\_order }  & $I_1$ 
	& RC &  RC\\ 
	&  $I_2$ &SER & RR \\
	&  $I_3$ & SER  &  RR  \\
	& $I_4$ & RC & RC   \\
	\hline
	\multirow{4}{*}{Payment}  & $I_1 $ 
	& RC &  RC\\ 
	&  $I_2$  &RC & RC \\
	&  $I_3 $ & RC  &  RC  \\
	& $I_4$  & RC & RC   \\
	\hline
	\multirow{4}{*}{Delivery}  & $I_1$  
	& RC &  RC \\ 
	&  $I_2$ &SER & RR \\
	&  $I_3$ & SER  &  RR \\
	& $I_4$  & RC & RC   \\
	\hline
\end{tabular}
	\caption{ Isolation level required by TPC-C transactions. }
	\label{tab:tpcc-eval}
\end{table*}

### 2017-07-07 evaluation.tex
Update evaluation.tex

@@
 Each warehouse has  $10$ districts, each serving $3000$ customers. 
-TPC-C  specification model  requires  maintaining multiple   consistency conditions.  We focus only on   four conditions, named $I_1$, $I_2$, $I_3$, and $I_4$.   
+TPC-C  specification model  requires  maintaining multiple   consistency conditions.  We focus  on   four conditions, named $I_1$, $I_2$, $I_3$, and $I_4$.   
 Consistency condition $I_1$  requires that  sales of each  warehouse  equals to the  sum of    its  district's sales.  Conditions $I_2$ and $I_3$  enforce  
@@
   reflexes  the district's roll-up. 
-Given the consistency conditions,  Table  \ref{tab:tpcc-eval} gives the isolation level  for the \C{New\_Order},  \C{Payment},  \C{Delivery} transactions using the  MySql and Postrgres databases   to maintain   each consistency condition.
+Given the consistency conditions,  Table  \ref{tab:tpcc-eval} gives the isolation level  for the \C{New\_Order},  \C{Payment},  \C{Delivery} transactions using the  MySQL and PostgreSQL databases   to maintain   each consistency condition.
+Note,   \C{Stock Level} and  \C{Order Status}   are read-only transactions can be executed under RC.
 Recall that the \C{New\_Order} transaction places a new order for
@@
  The  \C{Delivery}  transaction processes  orders for a particular   district at  a warehouse. 
- Our experimental results  shows that  the concurrent executions of   \C{New\_Order}   
-    and  \C{Delivery} transactions under a weak isolation level (i.e.,  $RC$) is anomalous, i.e., it may violate the consistency conditions $I_2$ and $I_3$.
+ Our experimental results  show that  the concurrent executions of   \C{New\_Order}   
+    and  \C{Delivery} transactions under a weak isolation level (i.e.,  RC) is anomalous, i.e., it may violate the consistency conditions $I_2$ and $I_3$.
 Depending on the isolation guarantees provided  by  underlying database,   the stronger  isolation level is required.
-For instance, the isolation level $RR$ in Postgres  is sufficient to maintain the condition while   MySql database  requires  $SER$.
+For instance, the isolation level RR in PostgreSQL  is sufficient to preserve the consistency  conditions while  operating on MySQL database  requires  these transaction to be serialized.
 
@@
 	\textbf{Transaction}   & \textbf{Invariant} 
-	& \textbf{MYSQL-Isolation} & \textbf{Postgres-Isolation} \\ 
+	& \textbf{MySQL-Isolation} & \textbf{PostgreSQL-Isolation} \\ 
 	\hline
-	\multirow{4}{*}{new\_order }  & $I_1$ 
+	\multirow{4}{*}{\C{New\_Order} }  & $I_1$ 
 	& RC &  RC\\ 
@@
 	\hline
-	\multirow{4}{*}{Payment}  & $I_1 $ 
+	\multirow{4}{*}{\C{Payment}}  & $I_1 $ 
 	& RC &  RC\\ 
@@
 	\hline
-	\multirow{4}{*}{Delivery}  & $I_1$  
+	\multirow{4}{*}{\C{Delivery}}  & $I_1$  
 	& RC &  RC \\ 

### 2017-07-07 implementation.tex
Impl.

@@
 if it is not stable. The result of traversing the transaction's AST is
-therefore a stable state transformer ($\F$), which is then checked
-against the transaction's stated guarantee ($G$). If the check passes,
-then the guarantee is verified to check if it preserves the invariant
-$I$. The successful result from both the checks results in the
-transaction being certified correct under the current choice of its
-isolation level. Succesful verification of all transactions concludes
-the top-level execution, returning the inferred isolation levels as
-its output.
+therefore a state transformer ($\F$) that is stable w.r.t $R_l$, which
+is also stabilized against $\R_c$ (using $\stabilize{\cdot}$), and
+then checked against the transaction's stated guarantee ($G$). If the
+check passes, then the guarantee is verified to check if it preserves
+the invariant $I$. The successful result from both the checks results
+in the transaction being certified correct under the current choice of
+its isolation level. Succesful verification of all transactions
+concludes the top-level execution, returning the inferred isolation
+levels as its output.
 
@@
 \end{smathpar}
-There are however certain set functions that
+There are however certain set functions whose behavior cannot be
+completely axiomatized in FOL. These include \C{sum}, \C{count} etc.
+For them we admit imprecise axiomatization in form certain lemmas on
+these functions.
+
+\textbf{Annotation Burden} \tool significantly reduces the annotation
+burden in verifying a weakly isolated transactions by eliminating the
+need to annotate intermediary assertions and loop invariants.
+Guarantees ($G$) and invariants ($I$), however, still need to be
+provided. Alternatively, a weakly isolated transaction $T$ can be
+verified against the generic serializability condition, to eliminate
+the need for any annotation. In this mode, \tool first infers the
+transformer $\F_{SER}$ of $T$ without considering any interference,
+which then becomes its guarantee ($G$). Doing likewise for every
+transaction results in a rely relation ($R$) that includes $\F_{SER}$
+of every transaction. The verification can now proceed as usual. The
+result of this verification is an assignment of (possibly weak)
+isolation levels to transactions which guarantees serializability of
+all executions.
 

### 2017-07-07 evaluation.tex

@@
   reflexes  the district's roll-up. 
-Given the consistency conditions,  Table  \ref{tab:tpcc-eval} gives the isolation level  for the \C{New\_Order},  \C{Payment},  \C{Delivery} transactions using the  MySQL and PostgreSQL databases   to maintain   each consistency condition.
+Given the consistency conditions,  Table  \ref{tab:tpcc-eval} gives the isolation level  for the \C{New\_Order},  \C{Payment},  and \C{Delivery} transactions using the  MySQL and PostgreSQL databases   to maintain   each consistency condition.
 Note,   \C{Stock Level} and  \C{Order Status}   are read-only transactions can be executed under RC.

### 2017-07-07 implementation.tex
Edits.

@@
 transaction results in a rely relation ($R$) that includes $\F_{SER}$
-of every transaction. The verification can now proceed as usual. The
-result of this verification is an assignment of (possibly weak)
-isolation levels to transactions which guarantees serializability of
-all executions.
+of every transaction. Verification now proceeds by taking interference
+into account, and verifying that each transaction still yeilds the
+same $F$ as its $F_{SER}$. The result of this verification is an
+assignment of (possibly weak) isolation levels to transactions which
+nonetheless guarantees serializability of all executions.
 

### 2017-07-07 introduction.tex

@@
 considerations, has motivated the development of weaker forms of
-transaction isolation, beginning as early as 1976~\cite{gray1976}. The
-ANSI SQL 92 standard defines three such weak isolation levels which
-are now implemented in many relational and NoSQL databases. Not
-surprisingly, weakly-isolated transactions have been found to
-significantly outperform serializable transactions on benchmark
-suites, both on single-node databases and multi-node replicated
-stores~\cite{dbtuningbook,bailishat,bailisvldb}, leading to their
-overwhelming adoption. A 2013 study~\cite{bailishotos} of 18 popular
-ACID and ``NewSQL'' databases found that only three of them offer
-serializability by default, and half, including Oracle 11g, do not
-offer it at all.  A 2015 study~\cite{bailisferal} of a large corpus of
-database applications finds no evidence that applications manifestly
-change the default isolation level offered by the database. Taken
-together, these studies make clear that weakly-isolated transactions
-are quite prevalent in practice, whereas serializable transactions are
-often eschewed.
+transaction isolation, beginning as early as 1976~\cite{gray1976};
+these instantiations permit a transaction to witness various effects
+of newly committed, or even concurrently running, transactions while
+it executes.  The ANSI SQL 92 standard defines three such weak
+isolation levels which are now implemented in many relational and
+NoSQL databases. Not surprisingly, weakly-isolated transactions have
+been found to significantly outperform serializable transactions on
+benchmark suites, both on single-node databases and multi-node
+replicated stores~\cite{dbtuningbook,bailishat,bailisvldb}, leading to
+their overwhelming adoption. A 2013 study~\cite{bailishotos} of 18
+popular ACID and ``NewSQL'' databases found that only three of them
+offer serializability by default, and half, including Oracle 11g, do
+not offer it at all.  A 2015 study~\cite{bailisferal} of a large
+corpus of database applications finds no evidence that applications
+manifestly change the default isolation level offered by the
+database. Taken together, these studies make clear that
+weakly-isolated transactions are quite prevalent in practice, whereas
+serializable transactions are often eschewed.
 
@@
 computations amenable for verification by off-the-shelf SMT solvers.
-
-We have realized these ideas as an embedded DSL within OCaml, that
-supports common relational SQL operations (updates, selects, inserts,
-deletes, joins, etc.).  Experimental results on real-world benchmarks
-demonstrate the feasibility and value of automated verification for
-weakly isolated transactions, an important advance in improving the
-safety of realistic database computations. The paper makes the
-following contributions:
-\vspace*{-6pt}
+The paper makes the following contributions:
+%% We have realized these ideas as an embedded DSL within OCaml, that
+%% supports common relational SQL operations (updates, selects, inserts,
+%% deletes, joins, etc.).  Experimental results on real-world benchmarks
+%% demonstrate the feasibility and value of automated verification for
+%% weakly isolated transactions, an important advance in improving the
+%% safety of realistic database computations.
 \begin{enumerate}
-\item An embedded DSL in OCaml that treats SQL-based relational
-  database operations as monadic computations over an abstract
-  database state.
- \item An operational semantics and compositional proof system for
-   this language capable of relating high-level application invariants
-   to database state, parameterized by a weak isolation semantics that
-   selectively exposes the visibility of these operations to other
-   transactions.
- \item An inference algorithm capable of discovering the weakest
-   isolation level that is sound with respect to a transaction's
-   high-level consistency requirements.  The algorithm interprets
-   database operations as state transformers expressed in a language
-   amenable for translation into a decidable fragment of first-order
-   logic, and is thus suitable for automated verification using
-   off-the-shelf SMT solvers.
- \item A realistic implementation along with results on real-world
-   database benchmarks that justify our approach and the utility of
-   our proof methodology and inference procedure.
+\item We analyze properties of weak isolation in the context of an
+  embedded DSL for OCaml that treats SQL-based relational database
+  operations (e.g., inserts, selects, deletes, joins, updates, etc.)
+  as monadic computations over an abstract database state.
+ \item We develop an operational semantics and compositional proof
+   system for this language capable of relating high-level application
+   invariants to database state, parameterized by a weak isolation
+   semantics that selectively exposes the visibility of these
+   operations to other transactions.
+ \item Exploiting the monadic definition of database state, we devise
+   an inference algorithm capable of discovering the weakest isolation
+   level that is sound with respect to a transaction's high-level
+   consistency requirements.  The algorithm interprets database
+   operations as state transformers expressed in a language amenable
+   for translation into a decidable fragment of first-order logic, and
+   is thus suitable for automated verification using off-the-shelf SMT
+   solvers.
+ \item We present details of a realistic implementation along with an
+   evaluation study on real-world database benchmarks that justify our
+   approach and the utility of our proof methodology and inference
+   mechanism..
 \end{enumerate}
 \noindent Our results provide the first (to the best of our knowledge)
-automated mechanism capable of verifying the safety of high-level
-consistency conditions for weakly-isolated database transactions,
-thereby allowing such transactions to enjoy the same rigorous
-reasoning capabilities as their strongly-isolated (serializable)
-counterparts.
+formalization of weakly-isolated transactions, along with an
+expressive and compositional proof automation framework capable of
+verifying the safety of high-level consistency conditions attached to
+these transactions.  Collectively, these contributions allow
+weakly-isolated transactions to enjoy the same rigorous reasoning
+capabilities as their strongly-isolated (serializable) counterparts.
 

### 2017-07-07 motivation.tex

@@
 
-We present our ideas in the context of an embedded DSL in OCaml, which
-offers a \C{DB} monad to define and compose database computations
-given in terms of SQL operations.  Besides the usual \C{bind} and
-\C{return} operators, the monad includes an \C{atomically} combinator
-that executes a database computation as an atomic transaction and
-returns the result.
+We present our ideas in the context of an embedded DSL in OCaml that
+we have developed, which offers a \C{DB} monad to define and compose
+database computations given in terms of SQL operations.  Besides the
+usual \C{bind} and \C{return} operators, the monad includes an
+\C{atomically} combinator that executes a database computation as an
+atomic transaction and returns the result.
 
@@
 \C{o\_id}. Thus, at the end of the loop, the number of \C{Order\_line}
-records in the database, whose \C{ol\_o\_id} is equal to \C{o\_id} is
-equal to the size of the \C{item\_req} set, which in turn is equal to
-the \C{Order} record's \C{o\_ol\_cnt} field, thus preserving the
-consistency condition.
+records in the database (i.e., the number of records whose
+\C{ol\_o\_id} field is equal to \C{o\_id}) is guaranteed to be equal
+to the size of the \C{item\_req} set, which in turn is equal to the
+\C{Order} record's \C{o\_ol\_cnt} field; these constraints ensure that
+the transaction's consistency condition is preserved.
 
@@
 verification by preventing any interference among concurrently
-executing transactions while the \C{new\_order} transaction executes.
-Under weak isolation\footnote{Weak isolation does not violate
-  atomicity as long as the witnessed effects are those of committed
-  transactions}, however, interferences of various kinds are
-permitted, leading to executions superficially similar to executions
-permitted by concurrent (racy) programs~\cite{GHE15,HPQ+15}.
-To illustrate, consider the behavior of the \C{new\_order} transaction
+executing transactions while the \C{new\_order} transaction executes,
+essentially yielding serial behaviors.  Under weak
+isolation\footnote{Weak isolation does not violate atomicity as long
+  as the witnessed effects are those of committed transactions},
+however, interferences of various kinds are permitted, leading to
+executions superficially similar to executions permitted by concurrent
+(racy) programs~\cite{GHE15,HPQ+15}.  To illustrate, consider the
+behavior of the \C{new\_order} transaction
 \begin{wrapfigure}{l}{.4\textwidth}
@@
 
-The figure depicts an execution as a series of SQL 
-operations. In the execution, the \C{new\_order} instance
-$T_1$ (green) reads the \C{d\_next\_o\_id} field of the district
-record for \C{d\_id}, but before it increments the field, another
-\C{new\_order} instance ($T_2$) begins its execution and commits. Note
-that $T_2$ reads the same \C{d\_next\_o\_id} value as $T_1$, and
-inserts new \C{Order} and \C{Order\_line} records with their \C{o\_id}
-and \C{ol\_o\_id} fields (resp.) equal to \C{d\_next\_o\_id}. $T_2$
-also increments the \C{d\_next\_o\_id} field, which $T_1$ has already
-acccessed. This is allowed because reads do not obtain a mutually
-exclusive lock on most databases. After $T_2$'s commit, $T_1$ resumes
-execution and adds new \C{Order} and \C{Order\_line} fields with the
-same order id as $T_1$. Thus, by the end of the execution,
-\C{Order\_line} records inserted by $T_1$ and $T_2$ all bear the same
-order id. There are also two \C{Order} records with the same district
-id (\C{d\_id}) and order id, none of whose \C{o\_ol\_cnt} reflects the
-actual number of \C{Order\_line} records inserted with that order id.
-This clearly violates TPC-C's consistency condition.
+The figure depicts an execution as a series of SQL operations. In the
+execution, the \C{new\_order} instance $T_1$ (green) reads the
+\C{d\_next\_o\_id} field of the district record for \C{d\_id}, but
+before it increments the field, another \C{new\_order} instance
+($T_2$) begins its execution and commits. Note that $T_2$ reads the
+same \C{d\_next\_o\_id} value as $T_1$, and inserts new \C{Order} and
+\C{Order\_line} records with their \C{o\_id} and \C{ol\_o\_id} fields
+(resp.) equal to \C{d\_next\_o\_id}. $T_2$ also increments the
+\C{d\_next\_o\_id} field, which $T_1$ has already acccessed. This is
+allowed because reads typically do not obtain a mutually exclusive
+lock on most databases. After $T_2$'s commit, $T_1$ resumes execution
+and adds new \C{Order} and \C{Order\_line} fields with the same order
+id as $T_1$. Thus, by the end of the execution, \C{Order\_line}
+records inserted by $T_1$ and $T_2$ all bear the same order id. There
+are also two \C{Order} records with the same district id (\C{d\_id})
+and order id, none of whose \C{o\_ol\_cnt} reflects the actual number
+of \C{Order\_line} records inserted with that order id.  This clearly
+violates TPC-C's consistency condition.
 
@@
 characterize RC isolation~\cite{berenson}. For instance, there are no
-\emph{lost writes}; both concurrent transactions' writes are present
+\emph{lost writes} - both concurrent transactions' writes are present
 in the final state of the database. Indeed, program analyses that aim
@@
 While reasoning in terms of anomalies is cumbersome and inadequate,
-reasoning about weak isolation in terms of low-level
+reasoning about weak isolation in terms of
 traces~\cite{adyaphd,gotsmanconcur15} on memory read and write actions
@@
 database, thus admitting concurrency, but it also requires that there
-not be any write-write conflicts among concurrent transactions when
-they commit. Write-write conflicts can be eliminated in various ways,
-e.g., through conflict detection followed by a rollback, or through
+not be any write-write conflicts (i.e,, both transactions modifying
+the same record) among concurrent transactions when they
+commit. Write-write conflicts can be eliminated in various ways, e.g.,
+through conflict detection followed by a rollback, or through
 exclusive locks, or a combination of both.  For instance, one possible
@@
 semantics of SI can be captured as an axiomatization over transitions
-of the database state ($\Delta \longrightarrow \Delta'$) during the
-lifetime an SI transaction ($T$):
+of the database state ($\Delta \longrightarrow \Delta'$) during a
+transaction's ($T$) lifetime:
 \begin{itemize}
@@
 verify the consistency conditions associated with the \C{new\_order}
-transaction.  First, since the database doesn't change ($\Delta' =
-\Delta$) during the execution of the transaction's body, we can reason
+transaction.  First, since the database does not change ($\Delta' =
+\Delta$) during execution of the transaction's body, we can reason
 about \C{new\_order} as though it executed in complete isolation until
 its commit point, leading to a verification process similar to what
-would been applied when reasoning about serializability.  When
-\C{new\_order} finishes execution, but before it commits, the SI
+would been applied when reasoning sequentially.  When \C{new\_order}
+finishes execution, however, but before it commits, the SI
 axiomatization shown above requires us to consider global state
@@
 $\forall(x\in\delta).~\Delta'(x) = \Delta(x)$.  The axiomatization
-precludes any execution in which there may concurrent updates to
+precludes any execution in which there are concurrent updates to
 shared table fields (e.g., \C{d\_next\_o\_id} on the same \C{District}
 table), but does not prohibit interferences that write to different
-tables, or different records in the same table.  We need to reason
-about the safety of such interferences with respect to the
-transaction's consistency invariants.
+tables, or write to different records in the same table.  We need to
+reason about the safety of such interferences with respect to
+\C{new\_order}'s consistency invariants.
 
@@
 
-To do so, we observe that a relational database is a significantly
-simpler abstraction than shared memory. There are no pointers, linked
-data structures, or aliasing.  Although a database essentially
-abstracts a mutable state, the state is affected through a
-well-defined fixed number of interfaces (SQL statements), each tagged
-with a logical formula describing what records are accessed and
-updated.
+We approach the verification problem by first observing that a
+relational database is a significantly simpler abstraction than shared
+memory. There are no pointers, linked data structures, or aliasing.
+Although a database essentially abstracts a mutable state, this state
+is affected through a well-defined fixed number of interfaces (SQL
+statements), each tagged with a logical formula describing what
+records are accessed and updated.
 
-This observation leads us away from thinking of database transactions
-as concurrent imperative programs.  Instead, we see value in viewing
-them as essentially functional computations that manage database state
-monadically.  By doing so, we can formulate the semantics of database
-operations as state transformers that relate the operation's pre- and
-post-states.  A state transformer semantics can be defined
-algorithmically, just like classical predicate transformer semantics
-(e.g., strongest post-condition).  Here, a transformer interprets a
+This observation leads us away from thinking of a collection of
+database transactions as a simple variant of a concurrent imperative
+program.  Instead, we see value in viewing them as essentially
+functional computations that manage database state monadically,
+mirroring the structure of our DSL.  By doing so, we can formulate the
+semantics of database operations as state transformers that explicitly
+relate an operation's pre- and post-states, defining the semantics of
+the corresponding transformer algorithmically, just like classical
+predicate transformer semantics (e.g., weakest pre-condition or
+strongest post-condition).  In our case, a transformer interprets a
 SQL statement in the set domain, modeling the database as a set of
-tuples, and a SQL statement a function over this set.  Among other
+tuples, and a SQL statement as a function over this set.  Among other
 things, one benefit of this approach is that low-level loops can now
@@
 been replaced with constants (\C{k1} to \C{k3}).  The body of the loop
-executes a SQL update followed by an insert.  Recall that a transaction
-reads from the global database ($\Delta$), and writes to a
-transaction-local database ($\delta$) before committing these updates. An update
-statement filters the tuples that match the search criteria from $\Delta$
-and computes the updated tuples that are to be
-added to the local database. Thus, its state transformer (call it
-$\T_U$) is the following function on sets\footnote{$\bind$ has higher
-precedence than $\cup$.}:
+executes a SQL update followed by an insert.  Recall that a
+transaction reads from the global database ($\Delta$), and writes to a
+transaction-local database ($\delta$) before committing these
+updates. An update statement filters the tuples that match the search
+criteria from $\Delta$ and computes the updated tuples that are to be
+added to the local database. Thus, the state transformer for the update
+statement (call it $\T_U$) is the following function on
+sets\footnote{$\bind$ has higher precedence than $\cup$.}:
 \begin{smathpar}
@@
 ($\stl$). Let $\F_U$ and $\F_I$ be the corresponding functions for
-$\T_U$ and $\T_I$ shown above. The state transformed of the loop body
-in Fig.~\ref{fig:new_order_code} can be expressed as the following
-composition of $\F_U$ and $\F_I$:
+$\T_U$ and $\T_I$ shown above. The state transformation induced by the
+loop body in Fig.~\ref{fig:new_order_code} can be expressed as the
+following composition of $\F_U$ and $\F_I$:
 \begin{smathpar}
@@
 the program itself. In particular, SQL statements become set
-operations, and the \C{DB} monad's $\bind$ becoms set monadic $\bind$,
-leading to a completely mechanical inference procedure.  The advantage
-of inferring such transformers is that we can now make use of a
-semantics-preserving translation from the domain of sets equipped with
-$\bind$ to first-order logic, allowing us to leverage SMT solvers for
-automated proofs without having to infer potentially complex
-thread-local invariants or intermediate assertions.
-Sec.~\ref{sec:inference} describes this translation. In the exposition
-thus far, we assumed $\Delta$ remains invariant, which is clearly not
-the case when we admit concurrency.  Necessary concurrency extensions
-of the state transformer semantics to deal with interference is also
-covered in Sec.~\ref{sec:inference}.  The following two sections,
-however, first focus on laying the theoretical foundations for
-reasoning about weak isolation.
+operations, and the \C{DB} monad's $\bind$ becoms set monadic $\bind$.
+The advantage of inferring such transformers is that we can now make
+use of a semantics-preserving translation from the domain of sets
+equipped with $\bind$ to a decidable fragment of first-order logic,
+allowing us to leverage SMT solvers for automated proofs without
+having to infer potentially complex thread-local invariants or
+intermediate assertions.  Sec.~\ref{sec:inference} describes this
+translation. In the exposition thus far, we assumed $\Delta$ remains
+invariant, which is clearly not the case when we admit concurrency.
+Necessary concurrency extensions of the state transformer semantics to
+deal with interference is also covered in Sec.~\ref{sec:inference}.
+The following two sections, however, first focus on laying the
+theoretical foundations for reasoning about weak isolation.

### 2017-07-07 full-rg-rules.tex
Proofs

\section{Operational Semantics}

%\renewcommand{\ctxn}[3]{\C{TXN}_{#1}\langle #2 \rangle\{#3\}}
\begin{figure*}[!ht]
\raggedright
%
\textbf{Syntax}\\
%
\begin{smathpar}
\renewcommand{\arraystretch}{1.2}
\begin{array}{lclcl}
\multicolumn{5}{c} {
  {x,y} \in \mathtt{Variables}\qquad
  {f} \in \mathtt{Field\;Names} \qquad
% {\tau} \in \mathtt{Table\;Names}
  {i,j} \in \mathbb{N} \qquad
  {\odot} \in \{+,-,\le,\ge,=\}\qquad
  {k} \in \mathbb{Z}\cup\mathbb{B} \qquad
  {\rec} \in \{\bar{f}=\bar{k}\}\
}\\
{\stl,\stg,s} & \in & \mathtt{State} & \coloneqq &  \Pow{\{\bar{f}=\bar{k}\}} \\
{\I_e, \I_c }  & \in & \mathtt{Isolation Spec} & \coloneqq & (\stl,\stg,\stg') \rightarrow \Prop\\
v & \in & \mathtt{Values} & \coloneqq & k \ALT \rec \ALT s\\
e & \in & \mathtt{Expressions} & \coloneqq & k \ALT x \ALT x.f 
    \ALT \{\bar{f}=\bar{e}\} \ALT e_1 \odot e_2\\ 
c & \in & \mathtt{Commands} & \coloneqq & \cskip \ALT \lete{x}{e}{c}
    \ALT \ite{e}{c_1}{c_2}\ALT c_1;c_2 \ALT \inserte{x}  \\
&&&&\ALT \deletee{\lambda x.e}
    \ALT \lete{x}{\selecte{\lambda x.e}}{c}
    \ALT \updatee{\lambda x.e_1}{\lambda x.e_2}\\
&&&&\ALT \foreache{x}{\lambda y.\lambda z. c} 
    \ALT \foreachr{s_1}{s_2}{\lambda x.\lambda y. e}\\
&&&&\ALT \txn{i}{\I_e,\I_c}{ c } \ALT \ctxn{i}{\I_e,\I_c,\stl,\stg}{c} \ALT c1 || c2\\
% t & \in & \mathtt{Terms} & \coloneqq & e \ALT c\\
\ectx & \in & \mathtt{Eval\;Ctx} & ::= & \bullet \ALT  
  \bullet || c_2 \ALT c_1 || \bullet \ALT \bullet;\,c_2 
  \ALT \ctxn{i}{\I_e,\I_c,\stl,\stg}{\bullet} \\
\end{array}
\end{smathpar}
%
\bigskip

\renewcommand{\arraystretch}{1.2}

%
\textbf{Local Reduction} \quad 
\fbox {\(\stg \vdash (c,\stl) \stepsto (c',\stl')\)}\\
%
\begin{minipage}{2.8in}
\rulelabel{E-Insert}
\begin{smathpar}
\begin{array}{c}
\RULE
{
  i \not\in \dom(\stl \cup \stg)\\
  r = \{\bar{f}=\bar{k};\,\idf=i;\,\delf=\C{false}\}
}
{
  \stg \vdash (\inserte{\{\bar{f}=\bar{k}\}},\stl) \stepsto
  (\cskip,\stl \cup \{r\})
}
\end{array}
\end{smathpar}
\end{minipage}
%
%
\begin{minipage}{2.8in}
\rulelabel{E-Delete}
\begin{smathpar}
\begin{array}{c}
\RULE
{
  s = \{r' \,|\, \exists(r\in\Delta).~ \eval([r/x]e)=\C{true} \\
        \hspace*{0.7in}\conj r'=\{\bar{f}=r.\bar{f}; \idf=r.\idf;
        \delf=\C{true}\}\}\\
% \dom(s) \cap \dom(\delta) = \emptyset
}
{
  \stg \vdash (\deletee{\lambda x.e},\stl) \stepsto (\cskip,\stl \cup s)
}
\end{array}
\end{smathpar}
\end{minipage}
%
\bigskip

%
\begin{minipage}{2.8in}
\rulelabel{E-Select}
\begin{smathpar}
\begin{array}{c}
\RULE
{
  s = \{r\in\Delta \,|\, \eval([r/x]e)=\C{true}\}\spc
  c' = [s/y]c
}
{
  \stg \vdash (\lete{y}{\selecte{\lambda x.e}}{c}, \stl) \stepsto 
              (c',\stl)
}
\end{array}
\end{smathpar}
\end{minipage}
%
%
\begin{minipage}{2.8in}
\rulelabel{E-Update}
\begin{smathpar}
\begin{array}{c}
\RULE
{
  s = \{r' \,|\, \exists(r\in\Delta).~ \eval([r/x]e_2)=\C{true} \conj r'=[r/x]e_1\}\\
}
{
  \stg \vdash (\updatee{\lambda x.e_1}{\lambda x.e_2},\stl) \stepsto 
              (\cskip,\stl \cup s)
}
\end{array}
\end{smathpar}
\end{minipage}
%

\begin{minipage}{2.8in}
\rulelabel{E-Seq1}
\begin{smathpar}
\begin{array}{c}
\RULE
{
 \stg \vdash (c1, \stl) \stepsto (c1', \stl') \spc c1 \neq \cskip
}
{
  \stg \vdash (c1;c2, \stl) \stepsto 
              (c1';c2,\stl')
}
\end{array}
\end{smathpar}
\end{minipage}

\begin{minipage}{2.8in}
\rulelabel{E-Seq2}
\begin{smathpar}
\begin{array}{c}
\RULE
{
 \stg \vdash (c1, \stl) \stepsto (\cskip, \stl')
}
{
  \stg \vdash (c1;c2, \stl) \stepsto 
              (c2,\stl')
}
\end{array}
\end{smathpar}
\end{minipage}

\begin{minipage}{2.8in}
\rulelabel{E-IfTrue}
\begin{smathpar}
\begin{array}{c}
\RULE
{
 \eval(e) = \C{true}
}
{
  \stg \vdash (\ite{e}{c_1}{c_2}, \stl) \stepsto 
              (c1,\stl)
}
\end{array}
\end{smathpar}
\end{minipage}

\begin{minipage}{2.8in}
\rulelabel{E-IfFalse}
\begin{smathpar}
\begin{array}{c}
\RULE
{
 \eval(e) = \C{false}
}
{
  \stg \vdash (\ite{e}{c_1}{c_2}, \stl) \stepsto 
              (c2,\stl)
}
\end{array}
\end{smathpar}
\end{minipage}

\begin{smathpar}
\begin{array}{ll}
  \rulelabel{E-Foreach1} & \stg \vdash (\foreache{s}{\lambda y.\lambda
  z.c},\stl) \stepsto (\foreachr{\emptyset}{s}{\lambda y.\lambda z. c})\\
  \rulelabel{E-Foreach2} & \stg \vdash (\foreachr{s_1}{\{r\} \uplus s_2}{\lambda y.\lambda
  z.c},\stl) \stepsto ([r/z][s_1/y]c;\,\foreachr{s_1 \cup \{r\}}{s_2}{\lambda y.\lambda z. c})\\
  \rulelabel{E-Foreach3} & \stg \vdash (\foreachr{s}{\emptyset}{\lambda y.\lambda
  z.c},\stl) \stepsto (\cskip,\stl)\\
\end{array}
\end{smathpar}
%
\bigskip

%
\textbf{Top-Level Reduction} \quad 
\fbox {\((c,\stg) \stepsto (c',\stg')\)}\\
%
\begin{minipage}{3in}
  \rulelabel{E-Txn-Start}
  \begin{smathpar}
  \begin{array}{c}
    \RULE{}
         {(\txn{i}{\I_e,\I_c}{c},\stg) \stepsto (\ctxn{i}{\I_e,\I_c,\emptyset,\stg}{c},\stg)}
  \end{array}
  \end{smathpar}
\end{minipage}%
\hfill
\begin{minipage}{3in}
\rulelabel{E-Txn}
\begin{smathpar}
\begin{array}{c}
\RULE
{
  \I_e\,\,(\stl,\stg,\stg')\spc
  \stg \vdash (c,\stl) \stepsto (c',\stl')
}
{
  (\ctxn{i}{\I_e,\I_c,\stl,\stg}{c},\stg') \stepsto
  (\ctxn{i}{\I_e,\I_c,\stl',\stg'}{c'},\stg')
}
\end{array}
\end{smathpar}
\end{minipage}\\

\begin{center}
\begin{minipage}{3in}
\rulelabel{E-Commit}
\begin{smathpar}
\begin{array}{c}
\RULE
{
  \I_c\,\,(\stl,\stg,\stg')
}
{
  (\ctxn{i}{\I_e,\I_c,\stl,\stg}{\cskip},\stg') \stepsto (\cskip,\stl\gg\stg')
}
\end{array}
\end{smathpar}
\end{minipage}
\end{center}
\hfill
%
\caption{\small \txnimp: Syntax and Small-step semantics}
\label{fig:txnimp}
\end{figure*}

\begin{figure}[t]
\raggedright
%
\textbf{Txn-Local Reasoning} \quad 
  \fbox {\( R \vdash \hoare{P}{c}{Q} \)} \\
%\fbox{\( \rg{\mathbb{I},P,R}{\txnbox{c}_i}{G,Q}\)} \quad
%\fbox{\( \rg{\mathbb{I},P,R}{c}{G,Q}\)} \quad\\
%
\begin{minipage}{3.2in}
\rulelabel{RG-Insert}
\begin{smathpar}
\begin{array}{c}
\RULE
{
  \stable(R,P)\\
  \hspace*{-1.1in}\forall\stl,\stl',\stg,i.~P(\stl,\stg) \conj i \not\in
  \dom(\stl\cup\stg) \\
  \conj \stl'=\stl \cup 
  \{\{\bar{f}=x.\bar{f};\,\idf=i;\,\delf=\C{false}\} \Rightarrow 
  Q(\stl',\stg)
}
{
  R \vdash \hoare{P}{\inserte{x}}{Q}
}
\end{array}
\end{smathpar}
\end{minipage}
%

%
\begin{minipage}{3.2in}
\rulelabel{RG-Delete}
\begin{smathpar}
\begin{array}{c}
\RULE
{
  \stable(R,P)\\
  \hspace*{-0.2in}\forall\stl,\stl',\stg.~P(\stl,\stg) \conj 
  \stl' = \stl \cup \{r' \,|\, \exists(r\in\Delta).~ [r/x]e=\C{true} \\
        \hspace*{0.7in}\conj r'=\{\bar{f}=r.\bar{f}; \idf=r.\idf;
        \delf=\C{true}\}\}
  \Rightarrow 
  Q(\stl',\stg)
}
{
  R \vdash \hoare{P}{\deletee{\lambda x.e}}{Q}
}
\end{array}
\end{smathpar}
\end{minipage}
%

%
\begin{minipage}{3in}
\rulelabel{RG-Update}
\begin{smathpar}
\begin{array}{c}
\RULE
{
  \stable(R,P)\\
  \hspace*{-0.2in}\forall\stl,\stl',\stg.~P(\stl,\stg) \conj 
  \stl' = \stl \cup \{r' \,|\, \exists(r\in\Delta).~ [r/x]e_2 =\C{true} \\
        \hspace*{1.4in}\conj r'=[r/x]e_1\}
  \Rightarrow 
  Q(\stl',\stg)
}
{
  R \vdash \hoare{P}{\updatee{\lambda x.e_1}{\lambda x.e_2}}{Q}
}
\end{array}
\end{smathpar}
\end{minipage}
%

%
\begin{minipage}{3in}
\rulelabel{RG-Select}
\begin{smathpar}
\begin{array}{c}
\RULE
{
  R \vdash \hoare{P'}{c}{Q}\spc
  \stable(R,P)\\
  \hspace*{-0.2in}P'(\stl,\stg) \Leftrightarrow P(\stl,\stg) \wedge
  x = \{r' \,|\, \exists(r\in\Delta).~ [r/x]e_2 =\C{true}\} \\
}
{
  R \vdash \hoare{P}{\lete{y}{\selecte{\lambda x.e}}{c}}{Q}
}
\end{array}
\end{smathpar}
\end{minipage}
%

%
\begin{minipage}{3in}
\rulelabel{RG-Foreach}
\begin{smathpar}
\begin{array}{c}
\RULE
{
  \stable(\R,Q)\spc
  \stable(\R,\psi)\\
  P \Rightarrow [y/\phi]\psi\spc
  \R \vdash \hoare{\psi \wedge z\in x}{c}{Q_c}\\
  Q_c  \Rightarrow [y \cup \{z\} / y]\psi \spc
  [x / y]\psi \Rightarrow Q
}
{
  \R \vdash \hoare{P}{\foreache{x}{\lambda y.\lambda z.c}}{Q}
}
\end{array}
\end{smathpar}
\end{minipage}
%

\begin{minipage}{3in}
\rulelabel{RG-Seq}
\begin{smathpar}
\begin{array}{c}
\RULE
{
\hoare{P} {c1} {Q^{'}} \spc \hoare{Q^{'}}{c2}{Q}\\
\stable(R, Q^{'})
}
{
  R \vdash \hoare{P}{c1;c2}{Q}
}
\end{array}
\end{smathpar}
\end{minipage}

\begin{minipage}{3in}
\rulelabel{RG-If}
\begin{smathpar}
\begin{array}{c}
\RULE
{
\hoare{P \wedge e} {c1} {Q} \spc \hoare{P \wedge \neg e}{c2}{Q}\\
\stable(R, P)
}
{
  R \vdash \hoare{P}{\ite{e}{c_1}{c_2}}{Q}
}
\end{array}
\end{smathpar}
\end{minipage}
%
\begin{minipage}{3in}
\rulelabel{RG-Conseq}
\begin{smathpar}
\begin{array}{c}
\RULE
{
  R \vdash \hoare{P}{c}{Q}\\
  P' \Rightarrow P \spc
  Q \Rightarrow Q' \spc
  \stable(R,P')\spc
  \stable(R,Q')\spc
}
{
  R \vdash \hoare{P'}{c}{Q'}
}
\end{array}
\end{smathpar}
\end{minipage}
%
\bigskip

%
\textbf{Top-Level Reasoning} \quad \fbox {\(
\rg{I,R}{c}{G,I} \)}\\
%
%
\begin{minipage}{3.5in}
\rulelabel{RG-Txn}
\begin{smathpar}
\begin{array}{c}
\RULE
{
 \stable(R, I) \spc
  \stable(R,\I)\spc
  P(\stl,\stg) \Leftrightarrow \stl=\emptyset \wedge I(\stg)\\
  R_r(\stl,\stg,\stg') \Leftrightarrow \exists \stg_1. \I_e(\stl, \stg_1, \stg) \wedge R(\stg, \stg') \wedge \I_e(\stl, \stg_1, \stg') \spc
  R_r \vdash \rg{P}{c}{Q}\\
  R_c(\stl,\stg,\stg') \Leftrightarrow \exists \stg_1. \I_c(\stl, \stg_1, \stg) \wedge R(\stg, \stg') \wedge \I_c(\stl, \stg_1, \stg')  \spc
  \stable(R_c,Q)\\
  \forall \stl,\stg.~ Q(\stl,\stg) \Rightarrow 
    G(\stg, \stl \gg \stg)\spc
  \forall \stg,\stg'.~I(\stg) \wedge G(\stg,\stg') \Rightarrow I(\stg')\\
}
{
  \rg{I,R}{\ctxn{i}{\I}{c}}{G \cup ID,I}
}
\end{array}
\end{smathpar}
\end{minipage}
%

%
\begin{minipage}{3.6in}
\rulelabel{RG-Par}
\begin{smathpar}
\begin{array}{c}
\RULE
{
 \rg{I,R \cup G_2 \cup ID}{t_1}{G_1 \cup ID,I} \\
 \rg{I,R \cup G_1 \cup ID}{t_2}{G_2 \cup ID,I}
}
{\rg{I,R}{t1||t2}{G_1\cup G_2 \cup ID,I}}
\end{array}
\end{smathpar}
\end{minipage}
%
%
\begin{minipage}{2in}
\rulelabel{RG-Conseq2}
\begin{smathpar}
\begin{array}{c}
\RULE
{
  \rg{I,R}{\ctxn{i}{\I}{c}}{G,I}\\
  \I' \Rightarrow \I \spc 
  R' \subseteq R \spc 
  \stable(R',\I')\\
  G \subseteq G' \spc
  \forall \stg,\stg'.~I(\stg) \wedge G'(\stg,\stg') \Rightarrow I(\stg')\\
}
{
  \rg{I,R'}{\ctxn{i}{\I'}{c}}{G',I}
}
\end{array}
\end{smathpar}
\end{minipage}
%

\caption{\small \txnimp: Rely-Guarantee Rules}
\label{fig:rg-rules}
\vspace*{-12pt}
\end{figure}

\begin{definition}[ Step-indexed reflexive transitive closure]
For all $A:\text{Type}$, $R: A \rightarrow A \rightarrow \mathbb{P}$, and $n :
\mathbb{N}$, the step-indexed reflexive transitive closure $R^n$ of $R$ is
the smallest relation satisfying the following
properties:
\begin{itemize}
\item $\forall (x:A).\, R^0 (x,x)$
\item $\forall (x,y,z : A).\, R(x,y) \conj R^{n-1}(y,z) \Rightarrow
R^{n}(x,z)$
\end{itemize}
\end{definition}

\begin{definition} [Interleaved step relation]
The interleaved step relation (denoted as $\rightarrow_R$) interleaves transaction local reduction with interference from concurrent transactions captured as the Rely relation (R). It is defined as follows:
\begin{mathpar}
\begin{array}{lcl}
(t, \stg) \rightarrow_{R}  (t^{'}, \stg^{'}) & \defeq & (t = t^{'} \wedge R(\sigma, \sigma^{'}))  \vee ((t, \stg) \rightarrow (t^{'}, \sigma^{'}))% \wedge\\
%& & (t.c = \texttt{SKIP} \Rightarrow \mathbb{I}_c(t.\sigma_l, t.\sigma_g, \sigma)) \wedge (\neg(t.c = \texttt{SKIP}) \Rightarrow \mathbb{I}_r(t.\sigma_l, t.\sigma_g, \sigma)))
\end{array}
\end{mathpar}

The interleaved multistep relation (denoted as $\rightarrow_R^n$) is the step-indexed reflexive transitive closure of $\rightarrow_R$.
\end{definition}

Given a transaction $t = \texttt{txn}\langle \mathbb{I}, \stl, \stg \rangle \{c\}$, we use the notation $t.\stl, t.\stg, t.\mathbb{I}$ and $t.c$ to denote the various components of $t$. Below, we provide a more precise definition of the transaction-local RG judgement:

\begin{mathpar}
\begin{array}{lcl}
\R \vdash \hoare{P}{c}{Q} & \defeq & \forall t, \stg,\stg',\bar{v}.
  P(t.\stl,\stg) \conj t.c = c \wedge (t,\stg) \rightarrow_{\R}^{n} (t_2, \stg^{'}) \rightarrow (t^{'}, \stg^{'}) \wedge t^{'}.c = \texttt{SKIP} \wedge Q(t^{'}.\stl, \stg^{'})
\end{array}
\end{mathpar}

Here, we have explicitly stated that the last step in the reduction sequence is taken by the transaction (and not by the environment),  finishing in the state satisfying the assertion $Q$. The nature of interference before and after the last step of the transaction are different (after the last step and before the commit step, the interference is controlled by $\mathbb{I}_c$, while before the last step, the interference is controlled by $\mathbb{I}_e$). Also, $\bar{v}$ denote valuations to all free variables in $c$.

\begin{lemma}
If $\stable(R, Q)$, then $\forall \stl, \stg, \stg', k. Q(\stl, \stg) \wedge R^{k}(\stg, \stg') \Rightarrow Q(\stl, \stg')$
\end{lemma}

\begin{proof}
We use induction on $k$.

\textbf{Base Case}: For $k = 0$, $\stg = \stg'$ and hence $Q(\stl, \stg')$.

\textbf{Inductive Case}: For the inductive case, assume that for $k'$, $\forall \stl, \stg, \stg'. Q(\stl, \stg) \wedge R^{k'}(\stg, \stg') \Rightarrow Q(\stl, \stg')$.
Given $\stl, \stg, \stg_1$ such that $Q(\stl, \stg)$, $R^{k'+1}(\stl, \stg_1)$, we have to show $Q(\stl, \stg_1)$. There exists $\stg'$ such that $R^{k'}(\stg, \stg')$ and $R(\stg', \stg_1)$. By the inductive hypothesis, $Q(\stl, \stg')$. $\stable(R,Q)$ is defined as follows:
$$
\stable(R,Q) = \forall \stl, \stg, \stg'. Q(\stl, \stg) \wedge R(\stg, \stg') \Rightarrow Q(\stl, \stg')
$$
Instantiating the above statement with $\stl, \stg', \stg_1$, we get $Q(\stl, \stg_1)$
\end{proof}

\begin{theorem}
RG-Txn is sound.
\end{theorem}
\begin{proof}
\begin{mathpar}
\begin{array}{lc}
  \stable(R, I) & HI\\
  \stable(R,\I) & H\I \\
  P(\stl,\stg) \Leftrightarrow \stl=\emptyset \wedge I(\stg) & HP\\ 
  \R_e(\stl,\stg,\stg') \Leftrightarrow \exists \stg_1. \I_e(\stl, \stg_1, \stg) \wedge R(\stg, \stg') \wedge \I_e(\stl, \stg_1, \stg') & H\R_l \\
  \R_e\vdash \rg{P}{c}{Q} & Hc\\
  \R_c(\stl,\stg,\stg') \Leftrightarrow \exists \stg_1. \I_c(\stl, \stg_1, \stg) \wedge R(\stg, \stg') \wedge \I_c(\stl, \stg_1, \stg') & H\R_c \\
  \stable(\R_c,Q) & HQ \\
  \forall \stl,\stg.~ Q(\stl,\stg) \Rightarrow 
    G(\stg, \stl \gg \stg)\spc  & HQG \\
  \forall \stg,\stg'.~I(\stg) \wedge G(\stg,\stg') \Rightarrow I(\stg') & HG\\
  \forall \stg. G(\stg, \stg) & HID
\end{array}
\end{mathpar}

Let $t_s = \ctxn{i}{\I}{c}$. Consider $\stg$ such that $I(\stg)$, and let $(t_s, \stg) \rightarrow_{R}^{n} (\texttt{SKIP}, \stg^{'})$. We have to show (1) $I(\stg^{'})$ and (2) $\texttt{step-guaranteed}(R, G, t_s, \stg)$. We break down the sequence of reductions into four parts :

\begin{itemize}
\item $\pi_1 = (t_s, \stg) \rightarrow_{R}^{n_1} (t_s, \stg_1) \rightarrow (t, \stg_1)$, where initially only the environment takes steps and the last step in the sequence is the start of the transaction using the rule E-Txn-Start. 
\item $(t, \stg_1) \rightarrow_{R}^{n_2} (t^{'}, \stg_2)$, which begins from $t$ taking its first step at state $\stg_1$ and ends at the first configuration where $t^{'}.c = \texttt{SKIP}$. We denote this sub-sequence by $\pi$.
\item $(t^{'}, \stg_2) \rightarrow_{R}^{n_3} (\texttt{SKIP}, \stg_3)$ which ends at the step where $t$ commits.
\item $(\texttt{SKIP}, \stg_3) \rightarrow_{R}^{n_4} (\texttt{SKIP}, \stg^{'})$ where only the environment takes a step.
\end{itemize}

In the sequence $\pi_1$, $R^{n_1}(\stg, \stg_1)$. By $I(\stg)$, $HI$ and Lemma 6.3, $I(\stg_1)$. By the rule E-Txn-Start, $t.\stl = \phi, t.\stg = \stg_1$ and $t.c = c$. Hence $P(t.\stl, \stg_1)$.

Expanding the definition of the assertion $Hc$ and instantiating it with $\stg = \stg_1$ and $\stg^{'} = \stg_2$, we would get $Q(t^{'}.\stl, \stg_2)$. However, the environment steps in run $\pi$ are in $R$, while the environment steps in assertion $Hc$ are in $\R_e$. Hence, we will now show that only environment steps in $\R_e$ can actually happen in the run $\pi$. We will show this in two steps. In the first step, we will prove that for all configurations $(t_p, \stg_p)$ in the run $\pi$ except possibly the last configuration, $\I_e(t_p.\stl, t_p.\stg, \stg_p)$.

We will prove this by contradiction. Assume that there is a configuration $(t_1, \stg_b)$ such that $\neg \I_e( t_1.\stl, t_1.\stg, \stg_b)$. Let $(t_1, \stg_{b}^{'}) \rightarrow (t_{1}^{'}, \stg_{b}^{'})$ be the next step in $\pi$ taken by the transaction. We know that this step always exists because the last step in $\pi$ is taken by the transaction.  Then $\mathbb{I}_e(t_1.\stl, t_1.\stg, \stg_{b}^{'})$. All steps between $(t_1, \stg_b)$ and $(t_1, \stg_{b}^{'})$ are taken by the environment, i.e. $R^{k}(\stg_b, \stg_{b}^{'})$ for some $k$. However,  $\neg \I_e(t_1.\stl, t_1.\stg, \stg_b)$, the assertion $H\I$ and a simple induction on $k$ shows that $\neg \I_e(t_1.\stl, t_1.\stg, \stg_{b}^{'})$. This is a contradiction. Hence, $\I_e(t_1.\stl, t_1.\stg, \stg_b)$.

Now, we will show that every environment step in $\pi$ is in $\R_e$. Assume that $(t_1, \stg_a) \rightarrow_R (t_1, \stg_b)$ is an environment step such that $R(\stg_a, \stg_b)$. Then, we know that $\I_e(t_1.\stl, t_1.\stg, \stg_a)$ and $\I_e(t_1.\stl, t_1.\stg, \stg_b)$. Hence, $t_1.\stg$ provides the existence of $\stg_1$ in the definition of $\R_e$. Thus, $\R_e(t_1.\stl, \stg_a, \stg_b)$. We can use $Hc$ and make the assertion $Q(t^{'}.\stl, \stg_2)$.

Note that $t^{'}.\stg = \stg_2$. Also, since all the changes in the global database state have so far been made by the environment, $I(\stg_2)$.

$(t^{'}, \stg_2) \rightarrow_{R}^{n_3 - 1} (t^{'}, \stg_{2}^{'}) \rightarrow (\texttt{SKIP}, \stg_3)$, where the first $n_3-1$ steps are only performed by the environment. Since the transaction commits at state $\stg_{2}^{'}$, by the E-Commit rule, $\I_c(t^{'}.\stl, \stg_2, \stg_{2}^{'})$. We will now show that all environment steps in the above run must be in $\R_c$. Again, we will show this in two steps. Let $m = n_3 - 1$ and $(t^{'}, \stg_2) \rightarrow_R (t^{'}, \stg_{21}) \rightarrow_R (t^{'}, \stg_{22}) \ldots \rightarrow_R (t^{'}, \stg_{2m}) \rightarrow (\texttt{SKIP}, \stg_3)$. We will show that $\I_c(t^{'}.\stl, \stg_2, \stg_{2k})$ for all $k, 1 \leq k \leq m$.

We will prove this by contradiction. Suppose for some $i$, $\neg \I_c(t^{'}.\stl, \stg_2, \stg_{2i})$. Clearly, $R^{j}(\stg_{2i}, \stg^{'})$. Then, by $H\I$ and a simple induction on $j$, we can show that $\neg \I_c(t^{'}.\stl, \stg_2, \stg_{2}^{'})$. However, this is a contradiction. Hence, $\forall k$, $\I_c(t^{'}.\stl, \stg_2, \stg_{2k})$.

Now, we will show that every environment step is in $\R_c$. Consider the step $(t^{'}, \stg_{2k}) \rightarrow_R (t^{'}, \stg_{2(k+1)})$. We have $\I_c(t^{'}.\stl, \stg_2, \stg_{2k})$ and $\I_c(t^{'}.\stl, \stg_2, \stg_{2(k+1)})$. Hence, $\stg_2$ provides the existence of $\stg_1$ in the definition of $\R_c$. Thus, $\R_c(t^{'}.\stl, \stg_{2k}, \stg_{2(k+1)})$.

By $HQ$, $Q(t^{'}.\stl, \stg_2)$ and Lemma 6.3 we have $Q(t^{'}.\stl, \stg_{2}^{'})$. Since all state changes so far have been made by the environment, $I(\stg_{2}^{'})$. By $HQG$, $G(\stg_{2}^{'}, t^{'}.\stl \gg \stg_{2}^{'})$. By the E-Commit rule, $\stg_3 = (t^{'}.\stl \gg \stg_{2}^{'})$. Hence, $G(\stg_{2}^{'}, \stg_3)$. All the steps of the transaction except the commit step do not change the global database state and by $HID$ belong to $G$. The commit step satisfies $G$. This proves the \texttt{step-guaranteed} assertion. Finally, by $HG$, $I(\stg_3)$.

%{\color{red}Note : Since E-Commit is applied at $(t^{'}, \sigma_{2}^{'})$ and $t^{'}.\sigma_g = \sigma_2$, $\mathbb{I}_c(t^{'}.\sigma_l, \sigma_2, \sigma_2^{'})$. $\sigma_{2}^{'}$ can be reached from $\sigma_2$ by any number of environment steps, each of which may not obey $\mathbb{I}_c$. That is, there may exist $\sigma_3$ such that $R(\sigma_2, \sigma_3)$ and $\neg \mathbb{I}_c(t^{'}.\sigma_l, \sigma_2, \sigma_3)$. Such states will be overlooked by the $\texttt{stable}_c$ function}.

All the steps in $(\texttt{SKIP}, \stg_3) \rightarrow_{R}^{n_4} (\texttt{SKIP}, \stg^{'})$ are performed by the environment. Since $I(\stg_3)$, by $HI$ and Lemma 6.3, $I(\stg^{'})$.
\end{proof}

%\RULE
%{
%  R \vdash \hoare{P'}{c}{Q}\spc
%  \stable(R,P')\\
%  \hspace*{-0.2in}P'(\stl,\stg) \Leftrightarrow P(\stl,\stg) \wedge
%  x = \{r' \,|\, \exists(r\in\Delta).~ [r/x]e_2 =\C{true}\} \\
%}
%{
%  R \vdash \hoare{P}{\selecte{\lambda x.e}}{Q}
%}

% s = \{r\in\Delta \,|\, \eval([r/x]e)=\C{true}\}\spc
%  c' = [s/y]c
\begin{theorem}
RG-Select is sound
\end{theorem}

\begin{proof}
Given the premise of RG-Select, $t, \stg$ such that $t.c = \lete{x}{\selecte{\lambda y.e}}{c}$, $(t, \stg) \rightarrow_{R}^{m} (t_2, \stg^{'}) \rightarrow (t^{'}, \stg^{'})$, $P(t.\stl, \stg)$ and $t^{'}.c = \texttt{SKIP}$, we have to show that $Q(t^{'}.\stl, \stg^{'})$. The reduction sequence can be broken down into following parts:

\begin{itemize}
\item $\pi_1 = (t, \stg) \rightarrow_{R}^{n_1} (t, \stg_1) \rightarrow (t_1, \stg_1)$ where initially only the environment takes steps, and ends with the application of the E-Select rule.
\item $\pi_2 = (t_1, \stg_1) \rightarrow_{R}^{n_2} (t_2, \stg^{'}) \rightarrow (t^{'}, \stg^{'}) $ which corresponds to the execution of $\C{c}$
\end{itemize}

In $\pi_1$, $R^{n_1}(\stg, \stg_1)$. By $P(t.\stl, \stg)$ and $\stable(R, P)$, we get $P(t.\stl, \stg_1)$. By applying the E-Select rule, $t_1.\stl = t.\stl$, $t_1.c = [s/x]c$, where $s = \{r\in\Delta_1 \,|\, \eval([r/y]e)=\C{true}\}$. By definition of $P'$, $P'(t_1.\stl, \stg_1)$. The following property holds trivially:

$$
R \vdash \hoare{P \wedge x = s} {c} {Q} \Leftrightarrow R \vdash \hoare{P} {[s/x]c} {Q}
$$

Since $R \vdash \hoare{P'}{c}{Q}$, by the above property, $R \vdash \hoare{P}{[s/x]c}{Q}$. Since $P(t_1.\stl, \stg_1)$, by definition of $R \vdash \hoare{P}{[s/x]c}{Q}$, we get $Q(t^{'}.\stl, \stg')$.

%Suppose the transaction next takes a step at state $\stg_1$, i.e. for some $k$ $R^{k}(\stg, \stg_1)$ and the rest of reduction is $(t_1, \stg_1) \rightarrow (t_{1}^{'}, \stg_1) \rightarrow_{R}^{m^{'}}  (t_2, \stg^{'}) \rightarrow (t^{'}, \sigma^{'})$. Now, $P'(t_1.\stl, \stg)$ and by $\stable(R, P')$, $P'(t_1.\stl, \stg_1)$. Since $P'(t_1.\stl, \stg)$ binds $y$ to $s$ and $t_1.c = [s/y]c$, by the assertion $R \vdash \hoare{P'}{c}{Q}$, $Q(t^{'}.\stl, \stg^{'})$.

\end{proof}

\begin{theorem}
RG-Update is sound
\end{theorem}

\begin{proof}
Given the premise of RG-Update, $t, \stg$ such that $t.c = \updatee{\lambda x.e_1}{\lambda x.e_2}$, $(t, \stg) \rightarrow_{R}^{m} (t_2, \stg^{'}) \rightarrow (t^{'}, \stg^{'})$, $P(t.\stl, \stg)$ and $t^{'}.c = \texttt{SKIP}$, we have to show that $Q(t^{'}.\stl, \stg^{'})$.

Since only a single step needs to be taken by the transaction (by applying the E-Update rule), $t_2.c = t.c$, $t_2.\stl = t.\stl$ and $R^m(\stg, \stg')$. By $\stable(R, P)$, $P(t_2.\stl, \stg')$. According to E-Update,  $t^{'}.\stl = t_2.\stl \cup \{r' \,|\, \exists(r\in \stg').~ \eval([r/x]e_2)=\C{true} \conj r'=[r/x]e_1\}$. From the premise of RG-Update, we know that

$$
\forall\stl,\stl',\stg.~P(\stl,\stg) \conj 
  \stl' = \stl \cup \{r' \,|\, \exists(r \in \stg).~ [r/x]e_2 =\C{true} \conj r'=[r/x]e_1\} \Rightarrow 
  Q(\stl',\stg)
$$

Instantiating the above statement with $\stl = t_2.\stl$ and $\stg = \stg'$, we get $Q(\stl',\stg')$. However, $\stl^{'} = t^{'}.\stl$. Hence, $Q(t^{'}.\stl,\stg')$.
\end{proof}

\begin{theorem}
RG-Insert is sound
\end{theorem}

\begin{proof}
Given the premise of RG-Insert, $t, \stg$ such that $t.c = \inserte{x}$, $(t, \stg) \rightarrow_{R}^{m} (t_2, \stg^{'}) \rightarrow (t^{'}, \stg^{'})$, $P(t.\stl, \stg)$ and $t^{'}.c = \texttt{SKIP}$, we have to show that $Q(t^{'}.\stl, \stg^{'})$.

Since only a single step needs to be taken by the transaction (by applying the E-Insert rule), $t_2.c = t.c$, $t_2.\stl = t.\stl$ and $R^m(\stg, \stg')$. By $\stable(R, P)$, $P(t_2.\stl, \stg')$. According to E-Insert, $t^{'}.\stl = t_2.\stl \cup \{\bar{f}=\bar{k};\,\idf=i;\,\delf=\C{false}\}$ and $i \not\in \dom(t_2.\stl \cup \stg')$. From the premise of RG-Insert, we know that

$$
\forall\stl,\stl',\stg,i.~P(\stl,\stg) \conj i \not\in
  \dom(\stl\cup\stg) \\
  \conj \stl'=\stl \cup 
  \{\{\bar{f}=x.\bar{f};\,\idf=i;\,\delf=\C{false}\} \Rightarrow 
  Q(\stl',\stg)
$$

Instantiating the above statement with $\stl = t_2.\stl$ and $\stg = stg'$, we get $Q(\stl',\stg')$. However, $\stl^{'} = t^{'}.\stl$. Hence, $Q(t^{'}.\stl,\stg)$.
\end{proof}

\begin{theorem}
RG-Delete is sound
\end{theorem}

\begin{proof}
Given the premise of RG-Delete $t, \stg$ such that $t.c = \deletee{\lambda x.e}$, $(t, \stg) \rightarrow_{R}^{m} (t_2, \stg^{'}) \rightarrow (t^{'}, \sigma^{'})$, $P(t.\stl, \stg)$ and $t^{'}.c = \texttt{SKIP}$, we have to show that $Q(t^{'}.\stl, \stg^{'})$.

Since only a single step needs to be taken by the transaction (by applying the E-Delete rule), $t_2.c = t.c$, $t_2.\stl = t.\stl$ and $R^m(\stg, \stg')$. By $\stable(R, P)$, $P(t_2.\stl, \stg')$. According to E-Delete,  $t^{'}.\stl = t_2.\stl \cup \{r' \,|\, \exists(r\in\stg').~ \eval([r/x]e)=\C{true} \conj r'=\{\bar{f}=r.\bar{f}; \idf=r.\idf;\delf=\C{true}\}\}$. From the premise of RG-Delete, we know that

$$
\forall\stl,\stl',\stg.~P(\stl,\stg) \conj 
  \stl' = \stl \cup \{r' \,|\, \exists(r\in\Delta).~ [r/x]e=\C{true}
        \conj r'=\{\bar{f}=r.\bar{f}; \idf=r.\idf;
        \delf=\C{true}\}\}  \Rightarrow 
  Q(\stl',\stg)
$$

Instantiating the above statement with $\stl = t_2.\stl$ and $\stg = \stg'$, we get $Q(\stl',\stg')$. However, $\stl^{'} = t^{'}.\stl$. Hence, $Q(t^{'}.\stl,\stg')$.
\end{proof}

%\RULE
%{
%  \stable(R,Q)\spc
%  \stable(R,I)\\
%  P \wedge y=\emptyset \Rightarrow I\spc
%  R \vdash \hoare{I \wedge z\in x}{c}{Q_c}\\
%  Q_c \wedge z\in y \Rightarrow I\spc
%  I \wedge y=x \Rightarrow Q
%}
%{
%  R \vdash \hoare{P}{\foreache{x}{\lambda y.\lambda z.c}}{Q}
%}

\begin{theorem}
RG-Foreach is sound
\end{theorem}

\begin{proof}
\begin{mathpar}
\begin{array}{lc}
\stable(R,Q) & HQ\\
  \stable(R,\psi) & HI\\
  \stable(R,P) & HP\\
  P  \Rightarrow [\phi / y]\psi & H1\\
  R \vdash \hoare{\psi \wedge z\in x}{c}{Q_c} & Hc\\
  Q_c \Rightarrow [y \cup \{z\} / y]\psi & H2\\
  %[x / y] \psi \Rightarrow Q & HF \\
 \end{array}
  \end{mathpar}
Given $t, \stg$ such that $t.c = \foreache{x}{\lambda y.\lambda z.c}$, $(t, \stg) \rightarrow_{R}^{n} (t_2, \stg^{'}) \rightarrow (t^{'}, \stg^{'})$, $P(t.\stl, \stg)$ and $t^{'}.c = \texttt{SKIP}$, we have to show that $Q(t^{'}.\stl, \stg^{'})$.

The operational semantics of \texttt{foreach} (E-Foreach1, E-Foreach2, E-Foreach3) essentially execute the command $\texttt{c}$ for a number of iterations, where in each iteration, $\texttt{z}$ is bound to a record $r \in \texttt{x}$, while $\C{y}$ is bound to a set containing records bound to $\C{z}$ in previous iterations. $\C{z}$ is bound to a different record in each iteration, and the loop stops when all records in $\C{x}$ are iterated over.

Assuming that $|x| = s$, the reduction sequence for \texttt{foreach} will have the following structure :

$$
(t, \stg) \rightarrow_{R}^{m} (t_{1}, \stg_{1}) \rightarrow_{R}^{n_1} (t_{2}^{'}, \stg_{2}^{'}) \rightarrow_{R}^{n_{1}^{'}} (t_{2}, \stg_{2}) \rightarrow_{R}^{n_2} (t_{3}^{'},\stg_{3}^{'}) \rightarrow_{R}^{n_{2}^{'}} (t_{3}, \stg_{3}) \ldots (t_{s}, \stg_{s}) \rightarrow_{R}^{n_s} (t_{s+1}^{'}, \stg_{s+1}^{'}) \rightarrow_{R}^{l} (t^{'}, \stg^{'})
$$

The reduction sequence $\pi_i = (t_{i}, \stg_{i}) \rightarrow_{R}^{n_i} (t_{i+1}^{'}, \stg_{i+1}^{'})$ corresponds to the execution of the command $\C{c}$ in the $i$th iteration, such that the first and last steps in $\pi_i$ are not environment steps. The sequence $\pi_0 = (t, \stg) \rightarrow_{R}^{m} (t_{1}, \stg_{1})$ corresponds to the steps E-Foreach1 and E-Foreach2 along with environment steps. Similarly, the sequence $\pi_{i}^{'} = (t_{i+1}^{'}, \stg_{i+1}^{'}) \rightarrow_{R}^{n_{1}^{'}} (t_{i+1}, \stg_{2})$ corresponds to the execution of the E-Foreach2 step required to prepare the $(i+1)$th iteration along with environment steps.

Let $x = \{r_1, \ldots, r_s\}$, and assume that the records are picked in the increasing order. Then at the start of the $i$th iteration, $\C{z}$ is bound to $r_i$, while $\C{y}$ is bound to $\{r_1, \ldots, r_{i-1}\}$. We will show that $[\{r_1,\ldots, r_i\} / y]\psi$ holds at the end of iteration $i$, for all $1 \leq i \leq s$. More precisely, we will show $[\{r_1,\ldots, r_i\} / y]\psi(t_{i+1}^{'}.\stl, \stg_{i+1}^{'})$. We will use induction on $i$.

\textbf{Base Case}: The steps E-Foreach1 and E-Foreach2 do not change $\stl$. Also, $P(t.\stl, \stg)$ and $\stable(R, P)$. Hence, at the end of the sequence $\pi_0$, $P(t_{1}.\stl, \stg_{1})$. By $H1$, this implies $[\phi / y] \psi(t_1.\stl, \stg_1)$. The sequence $\pi_1 = (t_{1}, \stg_{1}) \rightarrow_{R}^{n_1} (t_{2}^{'}, \stg_{2}^{'})$ corresponds the execution of $\C{c}$ in the first iteration with $\C{z}$ bound to $r_1$ and $\C{y}$ bound to $\phi$. Clearly, $\psi(t_1.\stl, \stg_1) \wedge z \in x$ holds. Hence, by $Hc$, $Q_c(t_{2}^{'}.\stl, \stg_{2}^{'})$. By H2, this implies $[\{r_1\} / y] \psi(t_2^{'}.\stl, \stg_{2}^{'})$.

\textbf{Inductive Case}: Assume that $[\{r_1,\ldots, r_{k-1}\} / y]\psi(t_{k}^{'}.\stl, \stg_{k}^{'})$. The next sequence of reductions $(t_{k}^{'}, \stg_{k}^{'}) \rightarrow_{R}^{n_{k}^{'}} (t_k, \stg_k)$ only corresponds to the execution of the E-Foreach2 step for the $k$th iteration and environment steps. E-Foreach2 does not change $\stl$, and since $\stable(R, \psi)$, we get $[\{r_1,\ldots, r_k\} / y]\psi(t_{k}.\stl, \stg_{k})$. At the start of the next iteration, $\C{z}$ is bound to $r_{k}$, and $\C{y}$ is bound to $\{r_1, \ldots, r_{k-1}\}$. Hence, $\psi(t_k.\stl, \stg_k) \wedge z \in x$. By Hc, this implies $Q_c(t_{k+1}^{'}.\stl, \stg_{k+1}^{'})$. By $H2$, this implies $[y \cup z / y] \psi(t_{k+1}^{'}.\stl, \stg_{k+1}^{'}) = [\{r_1, \ldots, r_k\} / y]\psi(t_{k+1}^{'}.\stl, \stg_{k+1}^{'})$. This proves the inductive step.

Hence, at the end of the $s$th iteration, $[x / y]\psi(t_{s+1}^{'}.\stl, \stg_{s+1}^{'})$. This implies $Q(t_{s+1}^{'}.\stl, \stg_{s+1}^{'})$. Finally, the last part of the reduction, $(t_{s+1}^{'}, \stg_{s+1}^{'}) \rightarrow_{R}^{l} (t^{'}, \stg^{'})$ corresponds environment steps and E-Foreach3 (as the last step). Since $\stable(R, Q)$ and E-Foreach3 does not change $\stl$, we have $Q(t^{'}.\stl, \stg)$.

\end{proof}

\begin{theorem}
RG-Seq is sound
\end{theorem}

\begin{proof}
\begin{mathpar}
\begin{array}{lc}
\hoare{P} {c1} {Q^{'}} & H1\\
 \hoare{Q^{'}}{c2}{Q} & H2\\
\stable(R, Q^{'}) & H3
\end{array}
\end{mathpar}
Given $t, \stg$ such that $t.c = c1;c2$, $(t, \stg) \rightarrow_{R}^{m} (t_2, \stg^{'}) \rightarrow (t^{'}, \stg^{'})$, $P(t.\stl, \stg)$ and $t^{'}.c = \texttt{SKIP}$, we have to show that $Q(t^{'}.\stl, \stg^{'})$. We can divide the reduction sequence into three parts :

\begin{itemize}
\item $(t,\stg) \rightarrow_{R}^{m_1} (t_{m}^{'}, \stg_1) \rightarrow (t_{m}, \stg_1)$, where $t_m.c = c2$. We denote this sequence as $\pi_1$.
\item $(t_m, \stg_1) \rightarrow_{R}^{m_2} (t_m, \stg_{1}^{'}) $ where all steps are taken by the environment. This sequence is denoted as $\pi_2$.
\item $(t_m, \stg_{1}^{'}) \rightarrow_{R}^{m_3} (t_{2}, \stg^{'}) \rightarrow (t^{'}, \stg^{'})$. This run is denoted as $\pi_3$.
\end{itemize}

By the premise of the E-Seq1 and E-Seq2 rules, all the reductions in the sequence $\pi_1$ are also applicable to $c1$. Hence, consider transaction $s$ such that $s.c = c1$, $s.\stl=t.\stl$. Then, there exists the sequence $(s, \stg)  \rightarrow_{R}^{m_1} (s_2, \stg_1) \rightarrow (s^{'}, \stg_1)$ with $s^{'}.c = \texttt{SKIP}$, $s^{'}.\stl = t_{m}.\stl$. Since $P(s.\stl, \stg)$, by H1, $Q^{'}(s^{'}.\stl, \stg_1)$. This implies $Q^{'}(t_m.\stl, \stg_1)$.

In the sequence $\pi_2$, all steps are taken by the environment. By H3, $Q^{'}(t_m.\stl, \stg_{1}^{'})$.

Since $t_m.c = c2$, by H3, $Q(t^{'}.\stl, \stg^{'})$.

%There exists $s$ such that $s.c = c1$, $s.\sigma_l=t.\sigma_l$, $s.\sigma_v = t.\sigma_v$ and the run $(s, \sigma) \rightarrow (s_1, \sigma) \rightarrow_{R}^{m_1} (s_2, \sigma_1) \rightarrow (s^{'}, \sigma_1)$ with $s^{'}.c = \texttt{SKIP}$. Similarly, there exists $u$ such that $u.c = c2$, $u.\sigma_l = s^{'}.\sigma_l$, $u.\sigma_v = s^{'}.\sigma_v$ and the run $(u, \sigma_{1}^{'}) \rightarrow (u_1, \sigma_{1}^{'}) \rightarrow_{R}^{m_2} (u_2, \sigma^{'}) \rightarrow (u^{'}, \sigma^{'})$ with $R^{m_3}(\sigma_1, \sigma_{1}^{'})$ and $u^{'}.c = \texttt{SKIP}$. Also, $u^{'}.\sigma_l = t^{'}.\sigma_l$ and $u^{'}.\sigma_v = t^{'}.\sigma_v$.

%By H1, $Q^{'}(s^{'}.\sigma_l, \sigma_3, s^{'}.\sigma_v)$. By H2, $Q^{'}(s^{'}.\sigma_l, \sigma_{3}^{'}, s^{'}.\sigma_v)$. Also, $Q^{'}(u.\sigma_l, \sigma_{3}^{'}, u.\sigma_v)$. By H3, $Q(u^{'}.\sigma_l, \sigma^{'}, u^{'}.\sigma_v)$. Hence $Q(t^{'}.\sigma_l, \sigma^{'}, t^{'}.\sigma_v)$.

\end{proof}

\begin{theorem}
RG-If is sound
\end{theorem}
\begin{proof}
\begin{mathpar}
\begin{array}{lc}
\hoare{P \wedge e} {c1} {Q} & H1 \\
\hoare{P \wedge \neg e}{c2}{Q} & H2\\
\stable(R, P) & H3
\end{array}
\end{mathpar}

Given $t, \stg$ such that $t.c = \ite{e}{c_1}{c_2}$, $(t, \stg) \rightarrow_{R}^{m} (t_2, \stg^{'}) \rightarrow (t^{'}, \stg^{'})$, $P(t.\stl, \stg)$ and $t^{'}.c = \texttt{SKIP}$, we have to show that $Q(t^{'}.\stl, \stg^{'})$. Assume that $\texttt{eval}(e) = \texttt{true}$. We divide the sequence of steps into two parts:

\begin{itemize}
\item $\pi_1 = (t, \stg) \rightarrow_{R}^{n_1} (t, \stg_1) \rightarrow (t_1, \stg_1)$ where initially only the environment takes steps, and the last step is taken by the transaction using E-IfTrue.
\item $\pi_2 = (t_1, \stg_1) \rightarrow_{R}^{n_2} (t_2, \stg^{'}) \rightarrow (t^{'}, \stg^{'})$
\end{itemize}

Since $P(t.\stl, \stg)$ and $R^{n_1}(\stg, \stg_1)$, by H3, we have $P(t.\stl, \stg_1)$. By applying the rule E-IfTrue, we have $t_1.\stl = t.\stl$, $t_1.c = c1$. Hence, $P(t_1.\stl, \stg_1)$. By the definition of $H1$, $Q(t^{'}, \stg')$. A similar proof follows for the case $\texttt{eval}(e) = \texttt{false}$
%In the first step by $t$, only the rules E-IfTrue and E-IfFalse are applicable. Assume that $t.\sigma_v(e)$. Then, $t_1.c = c1$, $t_1.\sigma_v = t.\sigma_v$, $t_1.\sigma_l = t.\sigma_l$ and $t_1.\sigma_g= \sigma$. Assume that the next step is taken by the transaction at global database state $\sigma_1$. That is, the environment takes $m_1$ steps in $(t_1, \sigma) \rightarrow_{R}^{m_1} (t_1, \sigma_1)$. Then, since the transaction takes a step, $\mathbb{I}_r(t_1.\sigma_l, \sigma, \sigma_1)$. Due to $\texttt{stable}_{\mathbb{I}}(R, \mathbb{I}_r)$, all the $m_1$ must also follow $\mathbb{I}_r$, and hence by $H3$, $P(t_1.\sigma_l, \sigma_1, t_1.\sigma_1)$. By $H1$, $Q(t^{'}.\sigma_l, \sigma^{'}, t^{'}.\sigma_v)$.

%A similar proof follows for the case $\neg t.\sigma_v(e)$.
\end{proof}

%\begin{lemma}
%If $\{\mathbb{I}, I, R\} t \{G, I\}$, then $\{\mathbb{I}, I, R \cup ID\} t \{G, I\}$, where $ID$ is the identity function
%\end{lemma}
%\begin{proof}
%Given $\sigma$ such that $I(\sigma)$ and the run $\pi_1$, $(t,\sigma) \rightarrow_{R \cup ID}^{n_1} (\texttt{SKIP}, \sigma^{'})$, we have to show (1) $I(\sigma^{'})$ and (2)$\texttt{commit-guarantee}(\mathbb{I}, R \cup ID, G, t, \sigma)$.
%
%We will transform the run $\pi_1$ to run $\pi_2$ of the form $(t,\sigma) \rightarrow_{R}^{n_2} (\texttt{SKIP}, \sigma^{'})$. Consider any sequence of two steps in the run $\pi_1$ of the form $(t_1, \sigma_1) \rightarrow_{R \cup ID} (t_1, \sigma_1) \rightarrow_{R \cup ID} (t_2, \sigma_2)$, i.e. the first step is taken by the environment in the form of the $ID$ function. Then we remove the first step and obtain the valid step $(t_1, \sigma_1) \rightarrow_{R \cup ID} (t_2, \sigma_2)$. We perform this transformation from left to right (i.e. increasing values of the step index) whenever the environment takes a step in the form of the $ID$ function. It is clear that the resulting run will only have environment steps in the form of $R$. Since the run $\pi_2$ also ends in $\sigma^{'}$, by the hypothesis $I(\sigma^{'})$. Also, the commit step in the run $\pi_2$ will involve the same states as the commit step in $\pi_1$. Hence, $\texttt{commit-guarantee}(\mathbb{I}, R \cup ID, G, t, \sigma)$. 
%\end{proof}

\begin{lemma}
If $\stable(R, Q)$ and $R' \subseteq R$, then $\stable(R', Q)$
\end{lemma}
\begin{proof}
Given $\stl, \stg, \stg'$ such that $Q(\stl, \stg)$ and $R'(\stg, \stg')$, we have to show that $Q(\stl, \stg')$. Since $R' \subseteq R$, $R(\stg, \stg')$. Hence, by $\stable(R,Q)$, $Q(\stl, \stg')$.
\end{proof}

\begin{lemma}
If $\rg{I,R}{\ctxn{i}{\I}{c}}{G \cup ID,I}$ and $R^{'} \subseteq R$, then $\rg{I,R'}{\ctxn{i}{\I}{c}}{G \cup ID,I}$
\end{lemma}
\begin{proof}
Let $t = \ctxn{i}{\I}{c}$. Then, given $\stg$ such that $I(\stg)$ and $(t, \stg) \rightarrow_{R'}^{n} (\C{SKIP}, \stg^{'})$, we have to show (1) $I(\stg^{'})$ and (2) $\texttt{step-guaranteed}(R', G \cup ID, t, \stg)$. Since $R' \subseteq R$, every environment step in the above reduction sequence is in $R$. Thus,  $(t, \stg) \rightarrow_{R'}^{n} (\C{SKIP}, \stg^{'})$, which by definition of $\rg{I,R}{\ctxn{i}{\I}{c}}{G \cup ID,I}$ implies $I(\stg')$. The same argument holds for $\texttt{step-guaranteed}(R', G \cup ID, t, \stg)$.
\end{proof}

\begin{theorem}
RG-Par is sound
\end{theorem}

\begin{proof}
\begin{mathpar}
\begin{array}{lc}
 \rg{I,R \cup G_2 \cup ID}{t_1}{G_1 \cup ID,I} & H1\\
 \rg{I,R \cup G_1 \cup ID}{t_2}{G_2 \cup ID,I} & H2
\end{array}
\end{mathpar}

Consider $\stg$ such that $I(\stg)$, and let $(t_1 || t_2, \stg) \rightarrow_{R}^{n} (\texttt{SKIP}, \stg^{'})$. We have to show (1) $I(\stg^{'})$ and (2) $\texttt{step-guaranteed}( R, G_1 \cup G_2 \cup ID, t1 || t_2, \stg)$.

Suppose that $t_1$ commits before $t_2$ in the execution sequence. Consider the sequence upto (and including) the commit step of $t_1$, i.e. $(t_1 || t_2, \stg) \rightarrow_{R}^{n_1} (t_{1}^{'} || t_{2}^{'}, \stg_1) \rightarrow (t_{2}^{'}, \stg_{1}^{'})$. In this sequence, all steps apart from the steps taken by $t_1$ belong to $R \cup ID$, since any step taken by $t_2$ cannot change the global database state. Hence, there exists the sequence $(t_1, \stg) \rightarrow_{R \cup ID}^{n_1} (t_{1}^{'}, \stg_1) \rightarrow (\texttt{SKIP}, \stg_{1}^{'})$. Since $R \cup ID \subseteq R \cup G_2 \cup ID$, by H1 and Lemma, $I(\stg_{1}^{'})$ and $G_1(\stg_1, \stg_{1}^{'})$. Now, consider the entire run from the perspective of $t_2$. All steps taken by $t_1$ except the commit step do not change the global database state, and the change during the commit step belongs to $G_1$. Hence, all steps in the run apart from the steps taken by $t_2$ belong to $R \cup G_1 \cup ID$. Hence, there exists a run $(t_2, \stg) \rightarrow_{R \cup G_1 \cup ID}^{n} (\texttt{SKIP}, \stg^{'})$. By H2 $I(\stg^{'})$.

Finally, the commit step of $t_1$ belongs to $G_1$, while the commit step of $t_2$ belongs to $G_2$, and every other step of either transaction does not change the global database state. Hence,  $\texttt{step-guaranteed}(R, G_1 \cup G_2 \cup ID, t1 || t2, \stg)$. The proof for the case where $t_2$ commits before $t_1$ would be similar.
\end{proof}

\begin{theorem}
RG-Conseq is sound
\end{theorem}

\begin{proof}
\begin{mathpar}
\begin{array}{lc}
 R \vdash \hoare{P}{t}{Q} & H1\\
  P' \Rightarrow P & H2\\
  Q \Rightarrow Q' & H3\\
\end{array}
\end{mathpar}
Given $t, \stg$ such that $(t, \stg) \rightarrow_{R}^{m} (t_2, \stg^{'}) \rightarrow (t^{'}, \stg^{'})$, $P'(t.\stl, \stg)$ and $t^{'}.c = \texttt{SKIP}$, we have to show that $Q'(t^{'}.\stl, \stg^{'})$. By H2, $P(t.\stl, \stg)$. Then, expanding the definition in H1, we get $Q(t^{'}.\stl, \stg')$. By H3, $Q'(t^{'}.\stl, \stg^{'})$.
\end{proof}

\begin{theorem}
RG-Conseq2 is sound
\end{theorem}

\begin{proof}
\begin{mathpar}
\begin{array}{lc}
\rg{I,R}{\ctxn{i}{\I}{c}}{G,I} & H1\\
  \I' \Rightarrow \I \spc & H2\\
  R' \subseteq R \spc & H2\\
  \stable(R',\I') & H3\\
  G \subseteq G' & H4\\
  \forall \stg,\stg'.~I(\stg) \wedge G'(\stg,\stg') \Rightarrow I(\stg')& H5
  \end{array}
  \end{mathpar}
Let $t = \ctxn{i}{\I'}{c}$. Given $\stg$ such that $I(\stg)$ and reduction sequence $\pi = (t, \stg) \rightarrow_{R'}^{n} (\C{SKIP}, \stg')$, we have to show that $I(\stg')$ and $\texttt{step-guaranteed}(R', G', t, \stg)$. First, we will show that the above reduction sequence is valid even if the isolation level of $t$ is changed to $\mathbb{I}$. Assume that the transaction performs $m$ steps in $\pi$. We will use induction on $m$ to show that every step of the transaction is valid for isolation level $\mathbb{I}$.

For the base case, the first step is always valid irrespective of any isolation level. For the inductive case, assume that all steps upto the $k$th step of the transaction in $t$ are valid with isolation level $\mathbb{I}$. Let the $(k+1)$th step of the transaction be $(t_1, \stg_1) \rightarrow (t_2, \stg_1)$. Then $\mathbb{I}'(t_1.\stl, t_1.\stg, \stg_1)$. By H2, $\mathbb{I}(t_1.\stl, t_1.\stg, \stg_1)$. Hence, the $k+1$th step is also valid for isolation level $\mathbb{I}$. This shows that the entire reduction sequence is valid even if the isolation level of $t$ is changed to $\mathbb{I}$. Let $t' = \ctxn{i}{\I'}{c}$. Since $R' \subseteq R$, it follows that the reduction sequence $\pi' = (t', \stg) \rightarrow_{R}^{n} (\C{SKIP}, \stg')$ comprising of the same steps as $\pi$ is valid. By H1, $I(\stg')$. Finally, by $\texttt{step-guaranteed}(R, G, t', \stg)$, all global database state changes caused by $t'$ in $\pi'$ are in $G$. But these are the same global database stage changes in $\pi$. Since $G \subseteq G'$, these state changes are also in $G'$.
\end{proof}

### 2017-07-08 supplement.tex
Supplement with proofs

--- 2017-07-07 full-rg-rules.tex
@@
+%-----------------------------------------------------------------------------
+%
+%               Template for sigplanconf LaTeX Class
+%
+% Name:         sigplanconf-template.tex
+%
+% Purpose:      A template for sigplanconf.cls, which is a LaTeX 2e class
+%               file for SIGPLAN conference proceedings.
+%
+% Guide:        Refer to "Author's Guide to the ACM SIGPLAN Class,"
+%               sigplanconf-guide.pdf
+%
+% Author:       Paul C. Anagnostopoulos
+%               Windfall Software
+%               978 371-2316
+%               paul@windfall.com
+%
+% Created:      15 February 2005
+%
+%-----------------------------------------------------------------------------
+
+%\documentclass[preprint,numbers]{sigplanconf}
+\documentclass[acmlarge,anonymous]{acmart}\settopmatter{printfolios=true}
+
+% The following \documentclass options may be useful:
+
+% preprint      Remove this option only once the paper is in final form.
+% 10pt          To set in 10-point type instead of 9-point.
+% 11pt          To set in 11-point type instead of 9-point.
+% numbers       To obtain numeric citation style instead of author/year.
+
+\usepackage{amssymb}
+\usepackage{amsmath}
+\usepackage{amsfonts}
+\usepackage{caption}
+\usepackage{subcaption}
+\usepackage{xspace}
+\usepackage{mathtools}
+\usepackage{mathpartir}
+\usepackage{ifpdf}
+\usepackage{graphicx}
+%\usepackage[usenames,dvipsnames]{color}
+\usepackage{stmaryrd}
+%\usepackage[numbers]{natbib}
+\usepackage{amsthm}
+\usepackage{listings}          % format code
+\usepackage{wrapfig}
+\usepackage{textcomp}
+\usepackage{tabularx}
+\usepackage{color}
+\usepackage{url}
+\usepackage{tikz}
+\usepackage{multirow,array}
+%\usepackage{float}
+
+\input{mydefs.tex}
+
+\begin{document}
+
+\special{papersize=8.5in,11in}
+\setlength{\pdfpageheight}{\paperheight}
+\setlength{\pdfpagewidth}{\paperwidth}
+
+\makeatletter\if@ACM@journal\makeatother
+%% Journal information (used by PACMPL format)
+%% Supplied to authors by publisher for camera-ready submission
+\acmJournal{PACMPL}
+\acmVolume{1}
+\acmNumber{1}
+\acmArticle{1}
+\acmYear{2017}
+\acmMonth{1}
+\acmDOI{10.1145/nnnnnnn.nnnnnnn}
+\startPage{1}
+\else\makeatother
+%% Conference information (used by SIGPLAN proceedings format)
+%% Supplied to authors by publisher for camera-ready submission
+\acmConference[PL'17]{ACM SIGPLAN Conference on Programming Languages}{January 01--03, 2017}{New York, NY, USA}
+\acmYear{2017}
+\acmISBN{978-x-xxxx-xxxx-x/YY/MM}
+\acmDOI{10.1145/nnnnnnn.nnnnnnn}
+\startPage{1}
+\fi
+
+%% Copyright information
+%% Supplied to authors (based on authors' rights management selection;
+%% see authors.acm.org) by publisher for camera-ready submission
+\setcopyright{none}             %% For review submission
+%\setcopyright{acmcopyright}
+%\setcopyright{acmlicensed}
+%\setcopyright{rightsretained}
+%\copyrightyear{2017}           %% If different from \acmYear
+
+%% Bibliography style
+\bibliographystyle{ACM-Reference-Format}
+% Uncomment the publication rights you want to use.
+%\publicationrights{transferred}
+%\publicationrights{licensed}     % this is the default
+%\publicationrights{author-pays}
+%% Citation style
+%% Note: author/year citations are required for papers published as an
+%% issue of PACMPL.
+\citestyle{acmauthoryear}   %% For author/year citations
+
+\title{Alone Together:
+  Compositional Reasoning and Inference for Weak Isolation}
+
+%% Author information
+%% Contents and number of authors suppressed with 'anonymous'.
+%% Each author should be introduced by \author, followed by
+%% \authornote (optional), \orcid (optional), \affiliation, and
+%% \email.
+%% An author may have multiple affiliations and/or emails; repeat the
+%% appropriate command.
+%% Many elements are not rendered, but should be provided for metadata
+%% extraction tools.
+
+%% Author with single affiliation.
+\author{First1 Last1}
+\authornote{with author1 note}          %% \authornote is optional;
+                                        %% can be repeated if necessary
+\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
+\affiliation{
+  \position{Position1}
+  \department{Department1}              %% \department is recommended
+  \institution{Institution1}            %% \institution is required
+  \streetaddress{Street1 Address1}
+  \city{City1}
+  \state{State1}
+  \postcode{Post-Code1}
+  \country{Country1}
+}
+\email{first1.last1@inst1.edu}          %% \email is recommended
+
+%% Author with two affiliations and emails.
+\author{First2 Last2}
+\authornote{with author2 note}          %% \authornote is optional;
+                                        %% can be repeated if necessary
+\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
+\affiliation{
+  \position{Position2a}
+  \department{Department2a}             %% \department is recommended
+  \institution{Institution2a}           %% \institution is required
+  \streetaddress{Street2a Address2a}
+  \city{City2a}
+  \state{State2a}
+  \postcode{Post-Code2a}
+  \country{Country2a}
+}
+\email{first2.last2@inst2a.com}         %% \email is recommended
+\affiliation{
+  \position{Position2b}
+  \department{Department2b}             %% \department is recommended
+  \institution{Institution2b}           %% \institution is required
+  \streetaddress{Street3b Address2b}
+  \city{City2b}
+  \state{State2b}
+  \postcode{Post-Code2b}
+  \country{Country2b}
+}
+\email{first2.last2@inst2b.org}         %% \email is recommended
+
+\maketitle
+%% Abstract
+%% Note: \begin{abstract}...\end{abstract} environment must come
+%% before \maketitle command
 \section{Operational Semantics}
@@
 %\renewcommand{\ctxn}[3]{\C{TXN}_{#1}\langle #2 \rangle\{#3\}}
-\begin{figure*}[!ht]
-\raggedright
+%\begin{figure*}[H]
+%\raggedright
 %
@@
     \ALT \foreachr{s_1}{s_2}{\lambda x.\lambda y. e}\\
-&&&&\ALT \txn{i}{\I_e,\I_c}{ c } \ALT \ctxn{i}{\I_e,\I_c,\stl,\stg}{c} \ALT c1 || c2\\
+&&&&\ALT \ctxn{i}{\I_e,\I_c}{ c } \ALT \ctxn{i}{\I_e,\I_c,\stl,\stg}{c} \ALT c1 || c2\\
 % t & \in & \mathtt{Terms} & \coloneqq & e \ALT c\\
@@
 \end{minipage}
-
 \begin{minipage}{2.8in}
@@
 %
-\caption{\small \txnimp: Syntax and Small-step semantics}
-\label{fig:txnimp}
-\end{figure*}
-
-\begin{figure}[t]
-\raggedright
+%\caption{\small \txnimp: Syntax and Small-step semantics}
+%\label{fig:txnimp}
+%\end{figure*}
+
+\section{Rely-Guarantee Reasoning}
+
+%\begin{figure}[t]
+%\raggedright
 %
@@
 %
-\begin{minipage}{3.2in}
+\begin{minipage}{0.5\textwidth}
 \rulelabel{RG-Insert}
@@
 \end{minipage}
-%
-
-%
-\begin{minipage}{3.2in}
+\begin{minipage}{0.5\textwidth}
 \rulelabel{RG-Delete}
@@
 \end{minipage}
-%
-
-%
 \begin{minipage}{3in}
@@
 \end{minipage}
-%
-
 \begin{minipage}{3in}
@@
 
-\caption{\small \txnimp: Rely-Guarantee Rules}
-\label{fig:rg-rules}
-\vspace*{-12pt}
-\end{figure}
-
+%\caption{\small \txnimp: Rely-Guarantee Rules}
+%\label{fig:rg-rules}
+%\vspace*{-12pt}
+%\end{figure}
+
+\section{Soundness of RG-Reasoning}
 \begin{definition}[ Step-indexed reflexive transitive closure]
@@
 \forall\stl,\stl',\stg,i.~P(\stl,\stg) \conj i \not\in
-  \dom(\stl\cup\stg) \\
+  \dom(\stl\cup\stg) 
   \conj \stl'=\stl \cup 
@@
 \end{proof}
+
+% We recommend abbrvnat bibliography style.
+
+%\bibliographystyle{plainnat}
+\small
+\bibliography{all}
+
+\end{document}

### 2017-07-08 document.tex

While a developer can examine the isolation level for each transaction, it is 
an error-prone and difficult task to  determine  whether   the 
combination of  transaction's isolations  preserve the application correctness properties.     
Given  the three possible  isolation levels RC, RR, and SER for each five  TPC-C transaction,  
we  run  the TPC-C benchmark  under all 243  possible   isolation combinations.  Then, 
 for each combination, we  check whether  the four consistency conditions are maintained. 
Out of $243$ runs, $171$ cases fails, i.e., they could not preserve the consistency conditions.  
We have observed that \C{Delivery} and \C{New_order} transactions are  problematic when 
there are executing under a weak isolation level. 
While the   \C{New_order} transaction triggers an  invariant violation  under  weak isolation levels  (RC and RR),  
out of $160$ runs in which   the  \C{Delivery}  transaction is 
executed under  RC and RR, only $11$ runs 
fails.

### 2017-07-08 supplement.tex
Edits

@@
 
+%\begin{minipage}
 \begin{smathpar}
@@
 %
-\bigskip
+%\bigskip
 
@@
   P(\stl,\stg) \Leftrightarrow \stl=\emptyset \wedge I(\stg)\\
-  R_r(\stl,\stg,\stg') \Leftrightarrow \exists \stg_1. \I_e(\stl, \stg_1, \stg) \wedge R(\stg, \stg') \wedge \I_e(\stl, \stg_1, \stg') \spc
-  R_r \vdash \rg{P}{c}{Q}\\
-  R_c(\stl,\stg,\stg') \Leftrightarrow \exists \stg_1. \I_c(\stl, \stg_1, \stg) \wedge R(\stg, \stg') \wedge \I_c(\stl, \stg_1, \stg')  \spc
-  \stable(R_c,Q)\\
+  \R_e(\stl,\stg,\stg') \Leftrightarrow \exists \stg_1. \I_e(\stl, \stg_1, \stg) \wedge R(\stg, \stg') \wedge \I_e(\stl, \stg_1, \stg') \spc
+  \R_e \vdash \rg{P}{c}{Q}\\
+  \R_c(\stl,\stg,\stg') \Leftrightarrow \exists \stg_1. \I_c(\stl, \stg_1, \stg) \wedge R(\stg, \stg') \wedge \I_c(\stl, \stg_1, \stg')  \spc
+  \stable(\R_c,Q)\\
   \forall \stl,\stg.~ Q(\stl,\stg) \Rightarrow 
@@
 }
-{\rg{I,R}{t1||t2}{G_1\cup G_2 \cup ID,I}}
+{\rg{I,R}{t_1||t_2}{G_1\cup G_2 \cup ID,I}}
 \end{array}
@@
 
-In the sequence $\pi_1$, $R^{n_1}(\stg, \stg_1)$. By $I(\stg)$, $HI$ and Lemma 6.3, $I(\stg_1)$. By the rule E-Txn-Start, $t.\stl = \phi, t.\stg = \stg_1$ and $t.c = c$. Hence $P(t.\stl, \stg_1)$.
+In the sequence $\pi_1$, $R^{n_1}(\stg, \stg_1)$. By $I(\stg)$, $HI$ and Lemma 3.3, $I(\stg_1)$. By the rule E-Txn-Start, $t.\stl = \phi, t.\stg = \stg_1$ and $t.c = c$. Hence $P(t.\stl, \stg_1)$.
 
 Expanding the definition of the assertion $Hc$ and instantiating it with $\stg = \stg_1$ and $\stg^{'} = \stg_2$, we would get $Q(t^{'}.\stl, \stg_2)$.
-However, the environment steps in run $\pi$ are in $R$, while the environment steps in assertion $Hc$ are in $\R_e$.
-Hence, we will now show that only environment steps in $\R_e$ can actually happen in the run $\pi$.
+However, the environment steps in sequence $\pi$ are in $R$, while the environment steps in assertion $Hc$ are in $\R_e$.
+Hence, we will now show that only environment steps in $\R_e$ can actually happen in the sequence $\pi$.
 We will show this in two steps.
-In the first step, we will prove that for all configurations $(t_p, \stg_p)$ in the run $\pi$ except possibly the last configuration, $\I_e(t_p.\stl, t_p.\stg, \stg_p)$.
+In the first step, we will prove that for all configurations $(t_p, \stg_p)$ in the sequence $\pi$ except possibly the last configuration, $\I_e(t_p.\stl, t_p.\stg, \stg_p)$.
 
@@
 Since the transaction commits at state $\stg_{2}^{'}$, by the E-Commit rule, $\I_c(t^{'}.\stl, \stg_2, \stg_{2}^{'})$.
-We will now show that all environment steps in the above run must be in $\R_c$.
+We will now show that all environment steps in the above sequence must be in $\R_c$.
 Again, we will show this in two steps.
@@
 
-By $HQ$, $Q(t^{'}.\stl, \stg_2)$ and Lemma 6.3 we have $Q(t^{'}.\stl, \stg_{2}^{'})$.
+By $HQ$, $Q(t^{'}.\stl, \stg_2)$ and Lemma 3.3 we have $Q(t^{'}.\stl, \stg_{2}^{'})$.
 Since all state changes so far have been made by the environment, $I(\stg_{2}^{'})$.
@@
 
-All the steps in $(\texttt{SKIP}, \stg_3) \rightarrow_{R}^{n_4} (\texttt{SKIP}, \stg^{'})$ are performed by the environment. Since $I(\stg_3)$, by $HI$ and Lemma 6.3, $I(\stg^{'})$.
+All the steps in $(\texttt{SKIP}, \stg_3) \rightarrow_{R}^{n_4} (\texttt{SKIP}, \stg^{'})$ are performed by the environment. Since $I(\stg_3)$, by $HI$ and Lemma 3.3, $I(\stg^{'})$.
 \end{proof}
@@
 \item $(t_m, \stg_1) \rightarrow_{R}^{m_2} (t_m, \stg_{1}^{'}) $ where all steps are taken by the environment. This sequence is denoted as $\pi_2$.
-\item $(t_m, \stg_{1}^{'}) \rightarrow_{R}^{m_3} (t_{2}, \stg^{'}) \rightarrow (t^{'}, \stg^{'})$. This run is denoted as $\pi_3$.
+\item $(t_m, \stg_{1}^{'}) \rightarrow_{R}^{m_3} (t_{2}, \stg^{'}) \rightarrow (t^{'}, \stg^{'})$. This sequence is denoted as $\pi_3$.
 \end{itemize}
@@
 
-%There exists $s$ such that $s.c = c1$, $s.\sigma_l=t.\sigma_l$, $s.\sigma_v = t.\sigma_v$ and the run $(s, \sigma) \rightarrow (s_1, \sigma) \rightarrow_{R}^{m_1} (s_2, \sigma_1) \rightarrow (s^{'}, \sigma_1)$ with $s^{'}.c = \texttt{SKIP}$.
-Similarly, there exists $u$ such that $u.c = c2$, $u.\sigma_l = s^{'}.\sigma_l$, $u.\sigma_v = s^{'}.\sigma_v$ and the run $(u, \sigma_{1}^{'}) \rightarrow (u_1, \sigma_{1}^{'}) \rightarrow_{R}^{m_2} (u_2, \sigma^{'}) \rightarrow (u^{'}, \sigma^{'})$ with $R^{m_3}(\sigma_1, \sigma_{1}^{'})$ and $u^{'}.c = \texttt{SKIP}$.
+%There exists $s$ such that $s.c = c1$, $s.\sigma_l=t.\sigma_l$, $s.\sigma_v = t.\sigma_v$ and the sequence $(s, \sigma) \rightarrow (s_1, \sigma) \rightarrow_{R}^{m_1} (s_2, \sigma_1) \rightarrow (s^{'}, \sigma_1)$ with $s^{'}.c = \texttt{SKIP}$.
+Similarly, there exists $u$ such that $u.c = c2$, $u.\sigma_l = s^{'}.\sigma_l$, $u.\sigma_v = s^{'}.\sigma_v$ and the sequence $(u, \sigma_{1}^{'}) \rightarrow (u_1, \sigma_{1}^{'}) \rightarrow_{R}^{m_2} (u_2, \sigma^{'}) \rightarrow (u^{'}, \sigma^{'})$ with $R^{m_3}(\sigma_1, \sigma_{1}^{'})$ and $u^{'}.c = \texttt{SKIP}$.
 Also, $u^{'}.\sigma_l = t^{'}.\sigma_l$ and $u^{'}.\sigma_v = t^{'}.\sigma_v$.
@@
 %\begin{proof}
-%Given $\sigma$ such that $I(\sigma)$ and the run $\pi_1$, $(t,\sigma) \rightarrow_{R \cup ID}^{n_1} (\texttt{SKIP}, \sigma^{'})$, we have to show (1) $I(\sigma^{'})$ and (2)$\texttt{commit-guarantee}(\mathbb{I}, R \cup ID, G, t, \sigma)$.
-%
-%We will transform the run $\pi_1$ to run $\pi_2$ of the form $(t,\sigma) \rightarrow_{R}^{n_2} (\texttt{SKIP}, \sigma^{'})$.
-Consider any sequence of two steps in the run $\pi_1$ of the form $(t_1, \sigma_1) \rightarrow_{R \cup ID} (t_1, \sigma_1) \rightarrow_{R \cup ID} (t_2, \sigma_2)$, i.e.
+%Given $\sigma$ such that $I(\sigma)$ and the sequence $\pi_1$, $(t,\sigma) \rightarrow_{R \cup ID}^{n_1} (\texttt{SKIP}, \sigma^{'})$, we have to show (1) $I(\sigma^{'})$ and (2)$\texttt{commit-guarantee}(\mathbb{I}, R \cup ID, G, t, \sigma)$.
+%
+%We will transform the sequence $\pi_1$ to sequence $\pi_2$ of the form $(t,\sigma) \rightarrow_{R}^{n_2} (\texttt{SKIP}, \sigma^{'})$.
+Consider any sequence of two steps in the sequence $\pi_1$ of the form $(t_1, \sigma_1) \rightarrow_{R \cup ID} (t_1, \sigma_1) \rightarrow_{R \cup ID} (t_2, \sigma_2)$, i.e.
 the first step is taken by the environment in the form of the $ID$ function.
@@
 increasing values of the step index) whenever the environment takes a step in the form of the $ID$ function.
-It is clear that the resulting run will only have environment steps in the form of $R$.
-Since the run $\pi_2$ also ends in $\sigma^{'}$, by the hypothesis $I(\sigma^{'})$.
-Also, the commit step in the run $\pi_2$ will involve the same states as the commit step in $\pi_1$.
+It is clear that the resulting sequence will only have environment steps in the form of $R$.
+Since the sequence $\pi_2$ also ends in $\sigma^{'}$, by the hypothesis $I(\sigma^{'})$.
+Also, the commit step in the sequence $\pi_2$ will involve the same states as the commit step in $\pi_1$.
 Hence, $\texttt{commit-guarantee}(\mathbb{I}, R \cup ID, G, t, \sigma)$.
@@
 Consider $\stg$ such that $I(\stg)$, and let $(t_1 || t_2, \stg) \rightarrow_{R}^{n} (\texttt{SKIP}, \stg^{'})$.
-We have to show (1) $I(\stg^{'})$ and (2) $\texttt{step-guaranteed}( R, G_1 \cup G_2 \cup ID, t1 || t_2, \stg)$.
+We have to show (1) $I(\stg^{'})$ and (2) $\texttt{step-guaranteed}( R, G_1 \cup G_2 \cup ID, t_1 || t_2, \stg)$.
 
@@
 Hence, there exists the sequence $(t_1, \stg) \rightarrow_{R \cup ID}^{n_1} (t_{1}^{'}, \stg_1) \rightarrow (\texttt{SKIP}, \stg_{1}^{'})$.
-Since $R \cup ID \subseteq R \cup G_2 \cup ID$, by H1 and Lemma, $I(\stg_{1}^{'})$ and $G_1(\stg_1, \stg_{1}^{'})$.
-Now, consider the entire run from the perspective of $t_2$.
+Since $R \cup ID \subseteq R \cup G_2 \cup ID$, by H1 and Lemma 3.13, $I(\stg_{1}^{'})$ and $G_1(\stg_1, \stg_{1}^{'})$.
+Now, consider the entire sequence from the perspective of $t_2$.
 All steps taken by $t_1$ except the commit step do not change the global database state, and the change during the commit step belongs to $G_1$.
-Hence, all steps in the run apart from the steps taken by $t_2$ belong to $R \cup G_1 \cup ID$.
-Hence, there exists a run $(t_2, \stg) \rightarrow_{R \cup G_1 \cup ID}^{n} (\texttt{SKIP}, \stg^{'})$.
+Hence, all steps in the sequence apart from the steps taken by $t_2$ belong to $R \cup G_1 \cup ID$.
+Hence, there exists a sequence $(t_2, \stg) \rightarrow_{R \cup G_1 \cup ID}^{n} (\texttt{SKIP}, \stg^{'})$.
 By H2 $I(\stg^{'})$.

### 2017-07-08 evaluation.tex
Edits.

--- 2017-07-07 evaluation.tex
@@
 
-To explore the space of correct isolation levels for different transactions, we  also  ran a number  of experiments using  a real transaction OLTP workload   based on TPC-C benchmark.  
- The  workload  involves  a mix of five TPC-C transaction types: 
-\C{New\_Order},  \C{Payment},  \C{Delivery}, \C{Stock Level} and  \C{Order Status}, 
-operating on top of  a $10$-warehouse database.  
-Each warehouse has  $10$ districts, each serving $3000$ customers. 
-TPC-C  specification model  requires  maintaining multiple   consistency conditions.  We focus  on   four conditions, named $I_1$, $I_2$, $I_3$, and $I_4$.   
-Consistency condition $I_1$  requires that  sales of each  warehouse  equals to the  sum of    its  district's sales.  Conditions $I_2$ and $I_3$  enforce  
- sequentially unique IDs assigned to  orders placed in the same   warehouse and district. Condition $I_4$ requires 
- the  number of  item orders   in each district 
-  reflexes  the district's roll-up. 
-Given the consistency conditions,  Table  \ref{tab:tpcc-eval} gives the isolation level  for the \C{New\_Order},  \C{Payment},  \C{Delivery} transactions using the  MySQL and PostgreSQL databases   to maintain   each consistency condition.
-Note,   \C{Stock Level} and  \C{Order Status}   are read-only transactions can be executed under RC.
-Recall that the \C{New\_Order} transaction places a new order for
- items from a particular district in a warehouse.  
-The \C{Payment} transaction  processes  a payment for a particular customer and updates the customer, district, and warehouse tables. 
- The  \C{Delivery}  transaction processes  orders for a particular   district at  a warehouse. 
- Our experimental results  show that  the concurrent executions of   \C{New\_Order}   
-    and  \C{Delivery} transactions under a weak isolation level (i.e.,  RC) is anomalous, i.e., it may violate the consistency conditions $I_2$ and $I_3$.
-Depending on the isolation guarantees provided  by  underlying database,   the stronger  isolation level is required.
-For instance, the isolation level RR in PostgreSQL  is sufficient to preserve the consistency  conditions while  operating on MySQL database  requires  these transaction to be serialized.
+We evaluate \tool on two different applications: 1. Courseware: a
+course registration system taken from~\cite{gotsmanpopl16}, and
+% 2. RUBiS: an eBay-like auction site~\cite{rubis} that allows users to
+% bid for items on sale, and pay for items from a wallet modeled after a
+% bank account, and 
+2. The TPC-C benchmark.
+
+\textbf{Courseware} Courseware application allows new courses to be
+added (\C{add\_course} transaction), and new students to be
+registered (\C{register\_student}) into the system. A registered
+student can enroll (\C{enroll}) in one of the existing courses,
+provided that the enrollments haven't already exceeded the course
+capacity (\C{c\_capacity}). A course with no enrollments can be
+canceled (\C{cancel\_course}). Likewise, a student who is not enrolled
+in any course can be deregistered (\C{deregister\_student}). Besides
+\C{Student} and \C{Course} tables, there is also an \C{Enrollment}
+table tracks the many-to-many enrollment relationship between courses
+and students. The application is required to preserve the following
+invariants on the database: 1.  $I_1$: An enrollment record should
+always refer to an existing student and an existing course, and 2.
+$I_2$: the capacity (\C{c\_capacity}) of a course should always be a
+non-negative quantity. Both $I_1$ and $I_2$ can be violated under weak
+isolation. $I_1$ can be violated, for example, when
+\C{deregister\_student} runs concurrently with \C{enroll}, both at RC
+isolation. While the former transaction removes the student record
+after checking that no enrollments for that student exist, the latter
+transaction concurrently adds an enrollment record after checking the
+student exist. Both can succeed concurrently, resulting in an invalid
+state. Invariant $I_2$ can be violated by two \C{enroll}s, both
+reading the \C{c\_capacity}=1, and both (atomically) decrementing it,
+resulting in \C{c\_capacity}=-1.  We ran \tool on the courseware
+application written in our DSL, annotated with guarantees for all
+transactions, and provided the $I = I_1 \wedge I_2$ as the correctness
+condition. \tool was able to automatically discover anomalous
+executions described above, and was subsequently able to infer that
+the anomalies can be preempted by promoting the isolation level of
+\C{enroll} and \C{deregister\_student} to SER (on both MySQL and
+PostgreSQL), leaving the isolation levels of remaining transactions at
+RC.
+
+\begin{table}[]
+\centering
+\begin{tabular}{l|c|c|c|c|c|}
+\cline{2-6}
+                                 & \multicolumn{1}{l|}{\C{new\_order}} & \multicolumn{1}{l|}{\C{delivery}} & \multicolumn{1}{l|}{\C{payment}} & \multicolumn{1}{l|}{\C{order\_status}} & \multicolumn{1}{l|}{\C{stock\_level}} \\ \hline
+\multicolumn{1}{|l|}{MySQL}      & SER                                   & SER                                 & RC                                 & RC                                       & RC                                      \\ \hline
+\multicolumn{1}{|l|}{PostgreSQL} & RR                                    & RR                                  & RC                                 & RC                                       & RC                                      \\ \hline
+\end{tabular}
+\caption{The discovered isolation levels for TPC-C transactions}
+\label{tab:tpcc}
+\end{table}
+
+\textbf{TPC-C} The simplified schema of the TPC-C benchmark has been
+described in Sec.~\ref{sec:motivation}. In addition to the tables
+shown in Fig.~\ref{fig:schema}, the TPC-C schema also has
+\C{Warehouse} and \C{New\_order} tables that are relevant for
+verification. To verify TPC-C, we consider four of the twelve
+consistency conditions specified by the standard, which we name $I_1$
+to $I_4$. Consistency condition $I_1$  requires that sales bottom line
+of each warehouse equals to the sum of the sales bottom lines of all
+district served by the warehouse.  Conditions $I_2$ and $I_3$
+effectively enforce uniqueness of ids assigned to \C{Order} and
+\C{New\_order} records, respectively, under a district. Condition
+$I_4$ requires that the number of order lines under a district must
+match the sum of order line counts of all orders under the district.
+% There are five transactions in TPC-C, out of which two are read-only.
+% The three remaining transactions are \C{new\_order}, \C{delivery}, and
+% \C{payment}.  
+Similar to the example discussed in Sec.~\ref{sec:motivation}, there
+are a number of ways TPC-C's transactions violate the aforementioned
+invariants under weak isolation. \tool was able to discover all such
+violations when verifying the benchmark against $I =
+\bigwedge_{i}I_i$, with guarantees of all three transactions
+provided. The isolation levels were subsequently strengthened  as
+shown in Table.~\ref{tab:tpcc}.
 
@@
 \end{tabular}
-	\caption{ Isolation level required by TPC-C transactions. }
+\caption{Various invariant violations witnessed for the TPC-C
+  benchmark on MySQL and PostgreSQL}
 \label{tab:tpcc-eval}
 \end{table*}
+
+To sanity-check the results of \tool, we conducted experiments with
+the high-contention OLTP workload  on TPC-C aiming to explore the
+space of correct isolation levels for different transactions. The
+workload involves  a mix of all five TPC-C transactions executing
+against a TPC-C database with 10 warehouses. Each warehouse has 10
+districts, and each district serves 3000 customers. There are a total
+of 5 transactions in TPC-C, and given that MySQL and PostgreSQL
+support 3 isolation levels each, there are a total of $3^5 = 243$
+different configurations of isolation levels for TPC-C transactions on
+MySQL and PostgreSQL. We executed OLTP benchmark with all 243
+configurations, and found 171 of them violating at least one invariant.
+Table~\ref{fig:tpcc-eval} shows the summary of these results. As
+expected, the isolation levels that \tool infers for the TPC-C
+transactions do not result in invariant violations, either on MySQL or
+on PostgreSQL.

### 2017-07-08 implementation.tex
Edits.

@@
 hence the concrete artifact of verification is OCaml's typed AST. The
-tool is already equipped with  an axiomatization of PostgresSQL and
+tool is already equipped with  an axiomatization of PostgreSQL and
 MySQL's isolation levels expressed in our FOL DSL. Other data stores
@@
 serializable isolation level, and if the program is sequentially
-corred, then the verification is guaranteed to succeed. Within the
+correct, then the verification is guaranteed to succeed. Within the
 loop, the \tool first computes the various rely relations needed for
@@
 transaction, applying the inference rules to construct a state
-tansformer, check its stability, and weaken it ($\stabilize{\cdot}$)
+transformer, check its stability, and weaken it ($\stabilize{\cdot}$)
 if it is not stable. The result of traversing the transaction's AST is
@@
 in the transaction being certified correct under the current choice of
-its isolation level. Succesful verification of all transactions
+its isolation level. Successful verification of all transactions
 concludes the top-level execution, returning the inferred isolation
@@
 store-specific semantics (isolation specifications etc). As an
-example, consider how PostgresSQL implements the \C{UPDATE} operation.
+example, consider how PostgreSQL implements the \C{UPDATE} operation.
 \C{UPDATE} first selects the records that meet the search criteria
@@
 Nonetheless, through a series of relatively straightforward
-deductions, we can show that PostgreSQL's \C{UPDATE} is infact
+deductions, we can show that PostgreSQL's \C{UPDATE} is in fact
 equivalent (in semantics) to a sequential composition of two atomic
@@
 of every transaction. Verification now proceeds by taking interference
-into account, and verifying that each transaction still yeilds the
+into account, and verifying that each transaction still yields the
 same $F$ as its $F_{SER}$. The result of this verification is an

### 2017-07-08 reasoning.tex
Edits to Sec. 2.

@@
 If there is no such state $\stg_1$, then a transaction will never be able to take its next step after the interference, and hence such interference can be safely discarded.
+Often, this significantly simplifies the form of $\R_l$ irrespective of $R$.
+For example, when a transaction is executed against a snapshot of the database (i.e.
+$\mathbb{I}_{ss}$), $\R_l$ will be the identity function, since if $\stg \neq \stg'$, there does not exist any $\stg_1$ which satisfies the constraint for $\R_l(\stl,\stg,\stg')$.
 
@@
 \begin{smathpar}
-\begin{array}{lcl}
-\R \vdash \hoare{P}{c}{Q} & \defeq & \forall \stl,\stl',\stg,\stg'.~
-  P(\stl,\stg) \conj (\tbox{c}_i,\stl,\stg) \rstepssto{n}
+\begin{array}{lclr}
+\R \vdash \hoare{P}{c}{Q} & \defeq & \forall \stl,\stl',\stg,\stg',
+  \bar{v}.~ P(\stl,\stg) \conj (\tbox{c}_i,\stl,\stg) \rstepssto{n}
   (\tbox{\cskip}_i, \stl',\stg')
-  \Rightarrow Q(\stl',\stg')\\
+  \Rightarrow Q(\stl',\stg') & \bar{v} = {\sf freevars}(c)\\
 \rg{I,R}{c}{G,I} & \defeq &  \forall \stg.\, I(\stg)
-  \Rightarrow (\forall n,\stg'.\; (c,\stg) 
+  \Rightarrow (\forall \stg'.\; (c,\stg) 
     \rstepssto{n} (\cskip,\stg') \Rightarrow I(\stg')) \\

### 2017-07-08 reasoning.tex
Edits

@@
 %\I_e(\stl,\stg,\stg')
-  \R_l(\stl,\stg,\stg') & \Leftrightarrow & \exists \stg_1. \I_e(\stl, \stg_1, \stg) \wedge R(\stg, \stg') \wedge \I_e(\stl, \stg_1, \stg')
+  \R_e(\stl,\stg,\stg') & \Leftrightarrow & \exists \stg_1.  R(\stg, \stg') \wedge \I_e(\stl, \stg_1, \stg')
 \end{array}
 \end{smathpar}
-In words, if there exists $\stg_1$ which was the global database state during the last step by a transaction and if the transaction can take its next step at the current global database state $\stg$, then even after the interference, the transaction should still be able to take its next step.
+
+While $R$ describes changes in the global database state, $\I_e$ will constrain the changes that will be made visible to the transaction.
+However, $\I_e$ uses the global database state that was seen by the transaction during its last step.
+We would like to find the interference steps that will be made visible to the transaction (since we only need to verify against that interference), but finding whether an interference step is allowed depends on the last seen global database state which could be anything.
+Hence, we say that if there exists some database state that was last seen by the transaction such that the transaction can continue its execution even after the interference, then such an interference could be allowed by $\I_e$ and hence is included in $\R_e$.
 If there is no such state $\stg_1$, then a transaction will never be able to take its next step after the interference, and hence such interference can be safely discarded.
-Often, this significantly simplifies the form of $\R_l$ irrespective of $R$.
+%Often, this significantly simplifies the form of $\R_l$ irrespective of $R$.
 For example, when a transaction is executed against a snapshot of the database (i.e.
 $\mathbb{I}_{ss}$), $\R_l$ will be the identity function, since if $\stg \neq \stg'$, there does not exist any $\stg_1$ which satisfies the constraint for $\R_l(\stl,\stg,\stg')$.
+
+%In words, if there exists $\stg_1$ which was the global database state during the last step by a transaction and if the transaction can take its next step at the current global database state $\stg$, then even after the interference, the transaction should still be able to take its next step.
 

### 2017-07-08 isolation.tex
edits

@@
 A distinctive characteristic of our development is that it is
-parameterized on the isolation specification $\I$. $\I$ can be
+parameterized on a weak isolation specification $\I$. $\I$ can be
 instantiated with the declarative characterization of an isolation
-guarantee or a concurrency control mechanism, regardless of its actual
-implementation. This allows us to model a range of isolation
-properties that are relevant to the theory and practice of transaction
-processing systems. Few such well-known properties are discussed
-below:
+guarantee or a concurrency control mechanism, regardless of the actual
+implementation used to realize it. This allows us to model a range of
+isolation properties that are relevant to the theory and practice of
+transaction processing systems without appealing to specific
+implementation artifacts like locks, versions, logs, speculation, etc.
+A few well-known properties are discussed below:
 
@@
 (Sec.~\ref{sec:motivation}) demonstrates, enforcing global uniqueness
-of ordered identifiers requires stronger isolation levels than ones
-that are default on most databases. Alternatively, globally unique
-sequence numbers, regardless of the isolation level, can be requested
-from a relational databases via SQL's \C{UNIQUE} and
-\C{AUTO\_INCREMENT} keywords. Our development crucially relies on the
-uniqueness of record identifiers\footnote{This has parallels in
-  real-world. For e.g., MySQL's InnoDB engine automatically adds a
-  6-byte unique identifier if none exists for a record.}, which are
-checked locally for uniqueness by the \rulelabel{E-Insert} rule. The
-global uniqueness of locally unique identifiers can be axiomatized as
-an isolation property as following:
+of ordered identifiers requires stronger isolation levels than the
+ones that are default on most databases (e.g., Read
+Committed). Alternatively, globally unique sequence numbers,
+regardless of the isolation level, can be requested from a relational
+database via SQL's \C{UNIQUE} and \C{AUTO\_INCREMENT} keywords. Our
+development crucially relies on the uniqueness of record
+identifiers\footnote{The importance of unique ids is recognized in
+  real-world implementations.  For example, MySQL's InnoDB engine
+  automatically adds a 6-byte unique identifier if none exists for a
+  record.}, which are checked locally for uniqueness by the
+\rulelabel{E-Insert} rule.  The global uniqueness of locally unique
+identifiers can be captured as an isolation property thus:
 \begin{smathpar}
@@
 it is added to a transaction's $\stl$, it remains globally unique
-until the transaction commits. This is achieved by prohibiting the
-interference from a concurrent transaction that adds the same id. The
-axiom thus simulates a global counter protected by an exclusive lock
-without explicitly appealing to an implementation artifact.
+until the transaction commits. This would be achieved within our
+semantic framework by prohibiting the interference from a concurrent
+transaction that adds the same id. The axiom thus simulates a global
+counter protected by an exclusive lock without explicitly appealing to
+an implementation artifact.
 
@@
 rollback) and pessimistic (e.g., various degrees of locking), to
-eliminate write-write (ww) conflicts among concurrent
+eliminate write-write (\emph{ww}) conflicts among concurrent
 transactions. We can specify the absence of such conflicts using our
@@
 prevent a record from being modified while it is read by an
-uncommitted transaction. The ANSI SQL RR specification can be
-axiomatized in our framework, but it requires a minor extension to our
-operational semantics to track a transaction's reads. In particular,
-the records returned by \C{SELECT} should be added to the local
-database $\stl$, but without changing their transaction identifiers
-($\txnf$ fields), and flush ($\gg$) should only flush the records that
-bear the current transaction's identifier. With this extension, ANSI
-SQL RR can be axiomatized thus:
+uncommitted transaction, but not prohibiting insertion of new
+records. The ANSI SQL RR specification can be axiomatized in our
+framework, but it requires a minor extension to our operational
+semantics to track a transaction's reads. In particular, the records
+returned by \C{SELECT} should be added to the local database $\stl$,
+but without changing their transaction identifiers ($\txnf$ fields),
+and flush ($\gg$) should only flush the records that bear the current
+transaction's identifier. With this extension, ANSI SQL RR can be
+axiomatized thus:
 \begin{smathpar}
@@
 
-\textbf{Serializability (SER)} Axiomatization of serializability is
+\textbf{Serializability (SER)} The specification of serializability is
 straightforward:

### 2017-07-08 opsem.tex
edits

@@
 supported nested transactions. The $\I$ in the \C{TXN} block syntax is
-the transaction's isolation specification, which will be explained
-later.  Certain terms that only appear at run-time are also present in
+the transaction's isolation specification, whose purpose is explained
+below.  Certain terms that only appear at run-time are also present in
 $c$.  These include a {\sf txn} block tagged with sets ($\stl$ and
-$\stg$) of records whose meaning is explained below, and a runtime
-{\sf foreach} expression that keeps track of the set ($s_1$) of items
-already iterated, and the set ($s_2$) of items yet to be
-iterated. Note that even the surface-level syntax of the \C{FOREACH}
-command shown here is a little different from the one used in previous
+$\stg$) of records representing local and global database state, and a
+runtime {\sf foreach} expression that keeps track of the set ($s_1$)
+of items already iterated, and the set ($s_2$) of items yet to be
+iterated. Note that the surface-level syntax of the \C{FOREACH}
+command shown here is slightly different from the one used in previous
 sections; its higher-order function has two arguments, $y$ and $z$,
@@
 they all belong to a single table.  The generalization to multiple
-tables is straightforward. The semantics assumes that the records in
-$\stg$ are uniquely identifiable through their $\idf$ field, and
-enforces this invariant wherever necessary. Certain hidden fields are
+tables is straightforward. The semantics assumes that  records in
+$\stg$ can be uniquely identified via their $\idf$ field, and
+enforces this property wherever necessary. Certain hidden fields are
 treated specially by the operational semantics, and they are hidden
@@
 \notin \dom(\stl)$, or it belongs to $\stl$, i.e., it is either a new
-record, or an updated version of an old record, but the update is not
-a deletion ($\neg r.\delf$). Thus, flush defines the result of
-atomically applying a transaction's local writes to the global
-database.  Besides the commit, flush also helps a transaction read its
-own writes. Intuitively, the result of a read operation inside a
-transaction must be computed on the database resulting from flushing
-the current local state ($\stl$) to the global state ($\stg$). The
-abstract machine of Fig.~\ref{fig:txnimp}, however, does not let a
-transaction read its own writes. This simplifies the semantics,
-without losing any generality, since substituting $\stl\gg\stg$ for
-$\stg$ at select places in the reduction rules effectively allows
-reads of uncommitted transaction writes to be realized.
+record, or an updated version of an old record, provided the update is
+not a deletion ($\neg r.\delf$).  Besides the commit, flush also helps
+a transaction read its own writes. Intuitively, the result of a read
+operation inside a transaction must be computed on the database
+resulting from flushing the current local state ($\stl$) to the global
+state ($\stg$). The abstract machine of Fig.~\ref{fig:txnimp},
+however, does not let a transaction read its own writes. This
+simplifies the semantics, without losing any generality, since
+substituting $\stl\gg\stg$ for $\stg$ at select places in the
+reduction rules effectively allows reads of uncommitted transaction
+writes to be realized.
 
@@
 reduces to $c'$, while updating the transaction-local database $\stl$
-to $\stl'$. The definition assumes a meta function $\eval$ that
-evaluates expressions with no free variables to values. The reduction
-relation for SQL statements is defined straightforwardly. \C{INSERT}
-adds a new record to $\stl$ after checking the uniqueness of its
-id. \C{DELETE} finds the records in $\stg$ that match the search
-criteria defined by its boolean function argument, and adds the
-records to $\stl$ after marking them for deletion. \C{SELECT} bounds
-the name introduced by \C{LET} to the set of records from $\stg$ that
-match the search criteria, and then executes the bound command
-$c$. \C{UPDATE} uses its first function argument to compute the
-updated version of the records that match the search criteria defined
-by its second function argument. Updated records are added to $\stl$.
+to $\stl'$. The definition assumes a meta-function $\eval$ that
+evaluates closed terms to values. The reduction relation for SQL
+statements is defined straightforwardly. \C{INSERT} adds a new record
+to $\stl$ after checking the uniqueness of its id. \C{DELETE} finds
+the records in $\stg$ that match the search criteria defined by its
+Boolean function argument, and adds the records to $\stl$ after
+marking them for deletion. \C{SELECT} bounds the name introduced by
+\C{LET} to the set of records from $\stg$ that match the search
+criteria, and then executes the bound command $c$. \C{UPDATE} uses its
+first function argument to compute the updated version of the records
+that match the search criteria defined by its second function
+argument. Updated records are added to $\stl$.
 
 The reduction of \C{FOREACH} starts by first converting it to its
-run-time form to keep track of iterated items ($s_1$),
-as well as  yet-to-be-iterated items ($s_2$). Initially, $s_1$ is
-empty. As the elements are iterated, they are removed from $s_2$ and
-added to $s_1$ . Iteration involves invoking its function argument
-with $s_1$ and the current element $x$ (note: $\uplus$ in $\{x\}
-\uplus s_2$ denotes a disjoint union). The reduction ends when 
-$s_2$ becomes empty. The reduction rules for conditionals, \C{LET}
-binders, and sequences are straightforward, and ommitted for brevity.
+run-time form to keep track of iterated items ($s_1$), as well as
+yet-to-be-iterated items ($s_2$).  Iteration involves invoking its
+function argument with $s_1$ and the current element $x$ (note:
+$\uplus$ in $\{x\} \uplus s_2$ denotes a disjoint union). The
+reduction ends when $s_2$ becomes empty. The reduction rules for
+conditionals, \C{LET} binders, and sequences are standard, and
+ommitted for brevity.
 
 The top-level reduction relation defines the small-step semantics of
-transactions, and their parallel composition. A transaction always
-comes tagged with an isolation specification $\I$, which has two
-components $\I_e$ and $\I_c$, that dictate the timing and nature of
-interferences that the transaction can witness, during its execution
-($\I_e$), and when it is about to commit ($\I_c$).  Formally, $\I_e$
-and $\I_c$ are predicates over the (current) transaction-local
-database state ($\stl$), the state ($\stg$) of the global database
-when the transaction last took a step, and the current state ($\stg'$)
-of the global database.  Intuitively, $\stg'\neq\stg$ indicates an
+transactions, and their parallel composition. A transaction comes
+tagged with an isolation specification $\I$, which has two components
+$\I_e$ and $\I_c$, that dictate the timing and nature of interferences
+that the transaction can witness, during its execution ($\I_e$), and
+when it is about to commit ($\I_c$).  Formally, $\I_e$ and $\I_c$ are
+predicates over the (current) transaction-local database state
+($\stl$), the state ($\stg$) of the global database when the
+transaction last took a step, and the current state ($\stg'$) of the
+global database.  Intuitively, $\stg'\neq\stg$ indicates an
 interference from another concurrent transaction, and the predicates
@@
 taking into account the local database state ($\stl$). For instance,
-as described in (Sec.~\ref{sec:motivation}), an RR transaction on
-PostgreSQL defines $\I$ as following:
+as described in (Sec.~\ref{sec:motivation}), an SI transaction on
+PostgreSQL defines $\I$ as follows:
 \begin{smathpar}
@@
 \end{smathpar}
-For brevity in presentation, we use $\I$ instead of $\I_e$ and $\I_c$
+To simplify the presentation, we use $\I$ instead of $\I_e$ and $\I_c$
 when its destructed form is not required.
@@
 its run-time form $\ctxnr{i}{\I,\stl,\stg}{c}$, where $\stl =
-\emptyset$, and $\stg$ the current (global) database.  Rule
+\emptyset$, and $\stg$ is the current (global) database.  Rule
 \rulelabel{E-Txn} reduces $\ctxnr{i}{\I,\stl,\stg}{c}$ under a
-database state ($\stg'$), only if the execution-time isolation
+database state ($\stg'$), only if the transaction-body isolation
 specification ($\I_e$) allows the interference between $\stg$ and
@@
 It also does not let a transaction overwrite its own writes, due to
-the premise $\dom(\stl)\cap\dom(s) = \emptyset$ on
+the premise $\dom(\stl)\cap\dom(s) = \emptyset$ on the
 \rulelabel{E-Delete} and \rulelabel{E-Update} rules. We refer to this
-restriction as \emph{local context independence}. As such, this
-restriction is easy to relax in operational semantics and the
-reasoning framework presented in the next section, however, the
-automation framework introduced in Sec.~\ref{sec:inference} has a
-non-trivial dependence on this assumption. Nonetheless, we haven't
-found local context independence to be a severe restriction in
-practice. For instance, all 17 transactions in the TPC-C and TPC-E
+restriction as \emph{local context independence}.  This restriction is
+easy to relax in the operational semantics and the reasoning framework
+presented in the next section; the automation framework introduced in
+Sec.~\ref{sec:inference}, however, has a non-trivial dependence on
+this assumption. Nonetheless, we have encountered few instances in
+practice where enforcing local context independence turns out to be a
+severe restriction. Indeed, all 17 transactions in the TPC-C and TPC-E
 benchmarks satisfy this assumption.

### 2017-07-08 reasoning.tex
edits

--- 2017-07-08 reasoning.tex
@@
 top-level judgment uses $I$ for both pre- and post-conditions, because
-our focus is on verifying that a \txnimp\ program \emph{preserves}
-database's consistency conditions\footnote{The terms \emph{consistency
+our focus is on verifying that a \txnimp\ program \emph{preserves} a
+databases' consistency conditions\footnote{The terms \emph{consistency
     condition}, \emph{high-level invariant}, and \emph{integrity
-    constraint} all mean the same.}.  A transaction-local RG judgment
-does not include a guarantee relation because transaction-local
-effects are not visible outside a transaction. Also, the rely relation
-($\R$) of the transaction-local judgment is not the same as the
-top-level rely relation ($R$) because it must take into account the
-transaction's isolation specification ($\I$). Intuitively, $\R$ is $R$
-modulo $\I$.  Recall that a transaction writes to its local database
-($\stl$), which is then flushed when the transaction commits. Thus,
-the guarantee of a transaction depends on the state of its local
-database at the commit point. The pre- and post-condition assertions
-($P$ and $Q$) in the local judgment facilitate tracking the changes to
-the transaction-local state, which eventually helps us prove the
-validity of the transaction's guarantee.  Both $P$ and $Q$ are
-bi-state assertions; they relate transaction-local database state
-($\stl$) to the global database state ($\stg$). Thus, the
-transaction-local judgment effectively tracks how transaction-local
-and global states change in relation to each other.
+    constraint} are used interchangeably throughout the paper.}.  A
+transaction-local RG judgment does not include a guarantee relation
+because transaction-local effects are not visible outside a
+transaction. Also, the rely relation ($\R$) of the transaction-local
+judgment is not the same as the top-level rely relation ($R$) because
+it must take into account the transaction's isolation specification
+($\I$). Intuitively, $\R$ is $R$ modulo $\I$.  Recall that a
+transaction writes to its local database ($\stl$), which is then
+flushed when the transaction commits. Thus, the guarantee of a
+transaction depends on the state of its local database at the commit
+point. The pre- and post-condition assertions ($P$ and $Q$) in the
+local judgment facilitate tracking the changes to the
+transaction-local state, which eventually helps us prove the validity
+of the transaction's guarantee.  Both $P$ and $Q$ are bi-state
+assertions; they relate transaction-local database state ($\stl$) to
+the global database state ($\stg$). Thus, the transaction-local
+judgment effectively tracks how transaction-local and global states
+change in relation to each other.
 
@@
 \end{smathpar}
-Due to the presence of local and global database states, and isolation
-specification, we use multiple definitions of stability in
-Fig.~\ref{fig:rg-rules}, but they all convey the same intuition as
-above. We will introduce them as they are encountered.
+Due to the presence of local and global database states, and the
+availability of an isolation specification, we use multiple
+definitions of stability in Fig.~\ref{fig:rg-rules}, but they all
+convey the same intuition as above.
 
@@
 \end{smathpar}
-In words, if there exists $\stg_1$ which was the global database state during the last step by a transaction and if the transaction can take its next step at the current global database state $\stg$, then even after the interference, the transaction should still be able to take its next step.
-If there is no such state $\stg_1$, then a transaction will never be able to take its next step after the interference, and hence such interference can be safely discarded.
-Often, this significantly simplifies the form of $\R_l$ irrespective of $R$.
-For example, when a transaction is executed against a snapshot of the database (i.e.
-$\mathbb{I}_{ss}$), $\R_l$ will be the identity function, since if $\stg \neq \stg'$, there does not exist any $\stg_1$ which satisfies the constraint for $\R_l(\stl,\stg,\stg')$.
+In words, if there exists $\stg_1$ which was the global database state
+during the last step by a transaction and if the transaction can take
+its next step at the current global database state $\stg$, then even
+after the interference, the transaction should still be able to take
+its next step. If there is no such state $\stg_1$, then the
+transaction will be stuck, and the interference can be safely
+discarded. Often, this significantly simplifies the form of $\R_l$
+irrespective of $R$. For example, when a transaction is executed
+against a snapshot of the database (i.e. $\mathbb{I}_{ss}$), $\R_l$
+will be the identity function, since if $\stg \neq \stg'$, there does
+not exist any $\stg_1$ which satisfies the constraint for
+$\R_l(\stl,\stg,\stg')$.
 
@@
 from the post-condition $Q$ of the body, provided that $Q$ is stable
-w.r.t the commit-time interference captured by $R_c$. $R_c$, like
-$R_l$, is computed as a rely relation modulo isolation, except that
+w.r.t the commit-time interference captured by $\R_c$. $\R_c$, like
+$\R_l$, is computed as a rely relation modulo isolation, except that
 commit-time isolation ($\I_c$) is considered. The validity of
@@
 commits, then $G$ must relate the states of the global database before
-and after the commit. The act of commit captured by the flush
+and after the commit. The act of commit is captured by the flush
 operation ($\stl\gg\stg$). Once we establish the validity of $G$ as a
@@
 original record, because the original record bears the id of a
-transaction that has already committed. Thus, $\I_{ss}$ remains
-invalid.
+transaction that has already committed.
 
@@
 
-We can now assert the soundness of the RG judgments in
+\noindent We can now assert the soundness of the RG judgments in
 Fig.~\ref{fig:rg-rules} as follows\footnote{Its proof is provided in

### 2017-07-08 evaluation.tex
Edits. URL added.

@@
 \hline
-\multirow{4}{*}{\C{Payment}}  & $I_1 $ 
-& RC &  RC\\ 
-&  $I_2$  &RC & RC \\
-&  $I_3 $ & RC  &  RC  \\
-& $I_4$  & RC & RC   \\
-\hline
+% \multirow{4}{*}{\C{Payment}}  & $I_1 $ 
+% & RC &  RC\\ 
+% &  $I_2$  &RC & RC \\
+% &  $I_3 $ & RC  &  RC  \\
+% & $I_4$  & RC & RC   \\
+% \hline
 \multirow{4}{*}{\C{Delivery}}  & $I_1$  
@@
 different configurations of isolation levels for TPC-C transactions on
-MySQL and PostgreSQL. We executed OLTP benchmark with all 243
-configurations, and found 171 of them violating at least one invariant.
-Table~\ref{fig:tpcc-eval} shows the summary of these results. As
-expected, the isolation levels that \tool infers for the TPC-C
-transactions do not result in invariant violations, either on MySQL or
-on PostgreSQL.
+MySQL and PostgreSQL. We executed the OLTP workload on the TPC-C
+benchmark with all 243 configurations, and found 171 of them violating
+at least one invariant.  Table~\ref{fig:tpcc-eval} shows the summary
+of these results. As expected, the isolation levels that \tool infers
+for the TPC-C transactions do not result in invariant violations,
+either on MySQL or on PostgreSQL.

### 2017-07-08 implementation.tex
Edits. URL added.

@@
 \caption{OCaml type definitions corresponding to the TPC-C schema from
-Fig.~\ref{fig:schema1}}
+Fig.~\ref{fig:schema}}
 \label{fig:ocaml-schema}
@@
 reasoning framework as an extra frontend pass (called \tool) in ocamlc
-4.03 compiler\footnote{The source code is available at available at
-\emph{\ldots}}. The input to \tool is a program in our DSL that
+4.03 compiler\footnote{The source code is available at
+\emph{https://goo.gl/nmUXZK}}. The input to \tool is a program in our DSL that
 describes the schema of the database as a collection of OCaml type
@@
 function of type $a \rightarrow \C{DB}\;b$, for some (base) types $a$
-and $b$. For instance, TPC-C's schema from Fig.~\ref{fig:schema1} can
+and $b$. For instance, TPC-C's schema from Fig.~\ref{fig:schema} can
 be described via the OCaml type definitions shown in

### 2017-07-08 evaluation.tex

--- 2017-07-08 evaluation.tex
@@
 \subsection{Evaluation}
+\label{sec:case-studies}
 
-We evaluate \tool on two different applications: 1. Courseware: a
-course registration system taken from~\cite{gotsmanpopl16}, and
+In this section, we present our experience in running \tool on two different applications: \emph{Courseware}: a
+course registration system described by~\cite{gotsmanpopl16}, and
 % 2. RUBiS: an eBay-like auction site~\cite{rubis} that allows users to
@@
 % bank account, and 
-2. The TPC-C benchmark.
+the \emph{TPC-C} benchmark.
 
-\textbf{Courseware} Courseware application allows new courses to be
-added (\C{add\_course} transaction), and new students to be
-registered (\C{register\_student}) into the system. A registered
-student can enroll (\C{enroll}) in one of the existing courses,
-provided that the enrollments haven't already exceeded the course
+\textbf{Courseware} The Courseware application allows new courses to be
+added (via an \C{add\_course} transaction), and new students to be
+registered (via a \C{register\_student} transaction) into a database. A registered
+student can enroll (\C{enroll}) in an existing course,
+provided that enrollment has not already exceeded the course
 capacity (\C{c\_capacity}). A course with no enrollments can be
@@
 \C{Student} and \C{Course} tables, there is also an \C{Enrollment}
-table tracks the many-to-many enrollment relationship between courses
+table to track the many-to-many enrollment relationship between courses
 and students. The application is required to preserve the following
-invariants on the database: 1.  $I_1$: An enrollment record should
-always refer to an existing student and an existing course, and 2.
-$I_2$: the capacity (\C{c\_capacity}) of a course should always be a
-non-negative quantity. Both $I_1$ and $I_2$ can be violated under weak
+invariants on the database:
+
+\begin{enumerate}
+\item  $I_1$: An enrollment record should always refer to an existing student and an existing course.
+\item  $I_2$: The capacity (\C{c\_capacity}) of a course should always be a
+  non-negative quantity.
+\end{enumerate}
+\noindent Both $I_1$ and $I_2$ can be violated under weak
 isolation. $I_1$ can be violated, for example, when
@@
 isolation. While the former transaction removes the student record
-after checking that no enrollments for that student exist, the latter
+after checking that no enrollments for that student exists, the latter
 transaction concurrently adds an enrollment record after checking the
-student exist. Both can succeed concurrently, resulting in an invalid
-state. Invariant $I_2$ can be violated by two \C{enroll}s, both
-reading the \C{c\_capacity}=1, and both (atomically) decrementing it,
+student exists.  Both can succeed concurrently, resulting in an
+invalid state. Invariant $I_2$ can be violated by two \C{enroll}s,
+both reading \C{c\_capacity}=1, and both (atomically) decrementing it,
 resulting in \C{c\_capacity}=-1.  We ran \tool on the courseware
 application written in our DSL, annotated with guarantees for all
-transactions, and provided the $I = I_1 \wedge I_2$ as the correctness
-condition. \tool was able to automatically discover anomalous
-executions described above, and was subsequently able to infer that
-the anomalies can be preempted by promoting the isolation level of
-\C{enroll} and \C{deregister\_student} to SER (on both MySQL and
-PostgreSQL), leaving the isolation levels of remaining transactions at
-RC.
+transactions taken from ~\cite{gotsmanpopl16}, and asserted $I = I_1
+\wedge I_2$ as the correctness condition. \tool was able to
+automatically discover the anomalous executions described above, and
+was subsequently able to infer that the anomalies can be preempted by
+promoting the isolation level of \C{enroll} and
+\C{deregister\_student} to SER (on both MySQL and PostgreSQL), leaving
+the isolation levels of remaining transactions at RC.  The total time
+for inference and verification took less than a minute running on a
+conventional laptop.
 
@@
 \C{Warehouse} and \C{New\_order} tables that are relevant for
-verification. To verify TPC-C, we consider four of the twelve
+verification.  To verify TPC-C, we examined four of the twelve
 consistency conditions specified by the standard, which we name $I_1$
-to $I_4$. Consistency condition $I_1$  requires that sales bottom line
-of each warehouse equals to the sum of the sales bottom lines of all
-district served by the warehouse.  Conditions $I_2$ and $I_3$
-effectively enforce uniqueness of ids assigned to \C{Order} and
-\C{New\_order} records, respectively, under a district. Condition
-$I_4$ requires that the number of order lines under a district must
-match the sum of order line counts of all orders under the district.
+to $I_4$:
+
+\begin{enumerate}
+\item Consistency condition $I_1$  requires that the sales bottom line
+of each warehouse equals the sum of the sales bottom lines of all
+districts served by the warehouse.
+
+\item Conditions $I_2$ and $I_3$ effectively enforce uniqueness of ids assigned
+  to \C{Order} and \C{New\_order} records, respectively, under a district.
+
+\item  Condition $I_4$ requires that the number of order lines under a district
+  must match the sum of order line counts of all orders under the district.
 % There are five transactions in TPC-C, out of which two are read-only.
@@
   % \C{payment}.
+\end{enumerate}
+
 Similar to the example discussed in Sec.~\ref{sec:motivation}, there
@@
 provided. The isolation levels were subsequently strengthened  as
-shown in Table.~\ref{tab:tpcc}.
+shown in Table.~\ref{tab:tpcc}.  As before, inference and verification
+took less than a minute.
 
-\begin{table*}[t]\small
-\centering
-\begin{tabular}{|c|c|c|c|}
-  \hline
-\textbf{Transaction}   & \textbf{Invariant} 
-& \textbf{MySQL-Isolation} & \textbf{PostgreSQL-Isolation} \\ 
-\hline
-\multirow{4}{*}{\C{New\_Order} }  & $I_1$ 
-& RC &  RC\\ 
-&  $I_2$ &SER & RR \\
-&  $I_3$ & SER  &  RR  \\
-& $I_4$ & RC & RC   \\
-\hline
-\multirow{4}{*}{\C{Payment}}  & $I_1 $ 
-& RC &  RC\\ 
-&  $I_2$  &RC & RC \\
-&  $I_3 $ & RC  &  RC  \\
-& $I_4$  & RC & RC   \\
-\hline
-\multirow{4}{*}{\C{Delivery}}  & $I_1$  
-& RC &  RC \\ 
-&  $I_2$ &SER & RR \\
-&  $I_3$ & SER  &  RR \\
-& $I_4$  & RC & RC   \\
-\hline
-\end{tabular}
-\caption{Various invariant violations witnessed for the TPC-C
-  benchmark on MySQL and PostgreSQL}
-\label{tab:tpcc-eval}
-\end{table*}
+%% \begin{table*}[t]\small
+%% \centering
+%% \begin{tabular}{|c|c|c|c|}
+%%   \hline
+%% \textbf{Transaction}   & \textbf{Invariant} 
+%% & \textbf{MySQL-Isolation} & \textbf{PostgreSQL-Isolation} \\ 
+%% \hline
+%% \multirow{4}{*}{\C{New\_Order} }  & $I_1$ 
+%% & RC &  RC\\ 
+%% &  $I_2$ &SER & RR \\
+%% &  $I_3$ & SER  &  RR  \\
+%% & $I_4$ & RC & RC   \\
+%% \hline
+%% \multirow{4}{*}{\C{Payment}}  & $I_1 $ 
+%% & RC &  RC\\ 
+%% &  $I_2$  &RC & RC \\
+%% &  $I_3 $ & RC  &  RC  \\
+%% & $I_4$  & RC & RC   \\
+%% \hline
+%% \multirow{4}{*}{\C{Delivery}}  & $I_1$  
+%% & RC &  RC \\ 
+%% &  $I_2$ &SER & RR \\
+%% &  $I_3$ & SER  &  RR \\
+%% & $I_4$  & RC & RC   \\
+%% \hline
+%% \end{tabular}
+%% \caption{Various invariant violations witnessed for the TPC-C
+%%   benchmark on MySQL and PostgreSQL}
+%% \label{tab:tpcc-eval}
+%% \end{table*}
 
-To sanity-check the results of \tool, we conducted experiments with
-the high-contention OLTP workload  on TPC-C aiming to explore the
-space of correct isolation levels for different transactions. The
-workload involves  a mix of all five TPC-C transactions executing
-against a TPC-C database with 10 warehouses. Each warehouse has 10
-districts, and each district serves 3000 customers. There are a total
-of 5 transactions in TPC-C, and given that MySQL and PostgreSQL
-support 3 isolation levels each, there are a total of $3^5 = 243$
-different configurations of isolation levels for TPC-C transactions on
-MySQL and PostgreSQL. We executed OLTP benchmark with all 243
-configurations, and found 171 of them violating at least one invariant.
-Table~\ref{fig:tpcc-eval} shows the summary of these results. As
-expected, the isolation levels that \tool infers for the TPC-C
-transactions do not result in invariant violations, either on MySQL or
-on PostgreSQL.
+To sanity-check the results of \tool, we conducted experiments with a
+high-contention OLTP workload on TPC-C aiming to explore the space of
+correct isolation levels for different transactions. The workload
+involves a mix of all five TPC-C transactions executing against a
+TPC-C database with 10 warehouses. Each warehouse has 10 districts,
+and each district serves 3000 customers. There are a total of 5
+transactions in TPC-C, and given that MySQL and PostgreSQL support 3
+isolation levels each, there are a total of $3^5 = 243$ different
+configurations of isolation levels for TPC-C transactions on MySQL and
+PostgreSQL. We executed the benchmark with all 243 configurations, and
+found 171 of them violated at least one of the four invariants we
+considered.  Table~\ref{fig:tpcc-eval} shows the summary of these
+results. As expected, the isolation levels that \tool infers for the
+TPC-C transactions do not result in invariant violations, either on
+MySQL or on PostgreSQL, and were determined to be the weakest safe
+assignments possible.

### 2017-07-08 implementation.tex

@@
 \caption{OCaml type definitions corresponding to the TPC-C schema from
-Fig.~\ref{fig:schema}}
+Fig.~\ref{fig:schema1}}
 \label{fig:ocaml-schema}
@@
 We have implemented our DSL to define transactions as monadic
-computations in OCaml (modulo the syntactic sugar), and our automatic
-reasoning framework as an extra frontend pass (called \tool) in ocamlc
-4.03 compiler\footnote{The source code is available at
-\emph{https://goo.gl/nmUXZK}}. The input to \tool is a program in our DSL that
+computations in OCaml (modulo some syntactic sugar), and our automatic
+reasoning framework as an extra frontend pass (called \tool) in the ocamlc
+4.03 compiler\footnote{The source code is available at available at
+\emph{\ldots}}. The input to \tool is a program in our DSL that
 describes the schema of the database as a collection of OCaml type
@@
 
-At the top-level, \tool runs a loop that pics an unverified
+At the top-level, \tool runs a loop that picks an unverified
 transaction and progressively strengthens its isolation level until it
-passes the verification. If the selected data store provides
+passes verification. If the selected data store provides a
 serializable isolation level, and if the program is sequentially
@@
 
-\tool uses Z3 SMT solver as its underlying reasoning engine. Each
-implication check described above is first encoded in the FOL,
-applying the translation described in Sec.~\ref{sec:inference}
-wherever necessary, and the validity of the resultant formula ($\phi$)
-is decided by asking Z3 to check the satisfiability of its negation;
-unsatisfiability of $\neg\phi$ proves the validity.
+\tool uses the Z3 SMT solver as its underlying reasoning engine. Each
+implication check described above is first encoded in FOL, applying
+the translation described in Sec.~\ref{sec:inference} wherever
+necessary.
 
@@
 certain nuances of their implementations on real data stores, which
-need to be taken into account for the verification to be effective in
+need to be taken into account for verification to be effective in
 practice.  We consider these into account while linking \tool with
 store-specific semantics (isolation specifications etc). As an
-example, consider how PostgreSQL implements the \C{UPDATE} operation.
+example, consider how PostgreSQL implements an \C{UPDATE} operation.
 \C{UPDATE} first selects the records that meet the search criteria
@@
 established at the beginning of the transaction if the isolation level
-is RR, or at the beginning of the \C{UPDATE} statement if the
+is SI, or at the beginning of the \C{UPDATE} statement if the
 isolation level is RC). The selected records are then visited in the
-actual database (if it still exists), write locks are obtained, and
-the updatation is carried out, provided that the record still meets
-\C{UPDATE}'s search criteria. If the record in no longer meets the
+actual database (if they still exist), write locks are obtained, and
+the update is performed, provided that the record still meets
+\C{UPDATE}'s search criteria. If the record no longer meets the
 search criteria (due to a concurrent update) the record is excluded
 from the update, and the write lock is immediately released.
-Otherwise, the record remains locked till the transaction commits.
+Otherwise, the record remains locked until the transaction commits.
 
-Clearly, the \C{UPDATE} operation on PostgreSQL is not atomic, as the
-our formal model assumes it to be. It admits interference between the
-updatation of individual records that meet the search criteria.
-Nonetheless, through a series of relatively straightforward
+Clearly, this sequence of events is not atomic, unlike the assumption
+made by our formal model.  The implementation admits interference
+between the updates of individual records that meet the search
+criteria.  Nonetheless, through a series of relatively straightforward
 deductions, we can show that PostgreSQL's \C{UPDATE} is in fact
-equivalent (in semantics) to a sequential composition of two atomic
+equivalent (in behavior) to a sequential composition of two atomic
 operations $c_1;c_2$, where $c_1$ is effectively a \C{SELECT}
 operation with the same search criteria as \C{UPDATE}, and $c_2$ is
-the a slight variation of the original \C{UPDATE} that updates a
-record only if a record with the same id is present in the records
+a slight variation of the original \C{UPDATE} that updates a
+record only if a record with the same id is present in the set of records
 returned by the \C{SELECT}. This transformation is summarized below:
@@
 record fields, and also application of auxiliary functions such as
-\C{MAX} and \C{MIN}. For example, \C{SELECT MAX(ol\_o\_id) FROM
+\C{MAX} and \C{MIN}, e.g., \C{SELECT MAX(ol\_o\_id) FROM
 Order\_line WHERE $\ldots$}. We admit such extensions as set functions
@@
 completely axiomatized in FOL. These include \C{sum}, \C{count} etc.
-For them we admit imprecise axiomatization in form certain lemmas on
-these functions.
+For these, we admit imprecise axiomatizations expressed as lemmas
+over these functions.
 
@@
 burden in verifying a weakly isolated transactions by eliminating the
-need to annotate intermediary assertions and loop invariants.
-Guarantees ($G$) and invariants ($I$), however, still need to be
+need to annotate intermediate assertions and loop invariants.
+Guarantees ($G$) and global invariants ($I$), however, still need to be
 provided. Alternatively, a weakly isolated transaction $T$ can be
-verified against the generic serializability condition, to eliminate
+verified against a generic serializability condition,  eliminating
 the need for any annotation. In this mode, \tool first infers the
@@
 assignment of (possibly weak) isolation levels to transactions which
-nonetheless guarantees serializability of all executions.
+nonetheless guarantees behavior equivalent to a  serializable execution.
 

### 2017-07-08 inference.tex
Edits to Sec. 2.
Edits
changes to R_l
Edits. URL added.

@@
 
-The Rely-Guarantee framework presented in the previous section
+The rely-guarantee framework presented in the previous section
 facilitates modular proofs for weakly-isolated transactions, but
@@
 first-order logic.
-
 \input{inference-rules}
-
 Fig.~\ref{fig:inference-rules} shows the syntax-directed state
@@
 \end{smathpar}
-where, $\bar{v}$ are the variables that occur free in $\F$; this is
+where $\bar{v}$ are the variables that occur free in $\F$; this is
 possible because of how the inference rules are structured.
@@
 the same set of records before and after the interference. The
-equivalence in $\SL$ translates to the equality in first-order logic,
+equivalence in $\SL$ translates to equality in first-order logic,
 as we describe below. In the inference rules, stability is
@@
 (\C{pc}) to the balance of a bank account, which is required to be
-positive ($I(\stg) \Leftrightarrow \forall(r\in\stg).~r.\C{bal}\ge
+non-negatives ($I(\stg) \Leftrightarrow \forall(r\in\stg).~r.\C{bal}\ge
 0$). The transaction starts by issuing a \C{select1} query, whose
@@
 rules for \C{LET} binders, conditionals, and sequences compose the
-effects inferred for their subcommands. For instance, the effect of a
-sequence of commands $c_1;c_2$ is the union of effects $F_1$ and $F_2$
-of $c_1$ and $c_2$, respectively, except that $F_2$ is applied to a
-state ($\stl$) to which $F_1$ has already been applied, reflecting
-their order of reduction. The inference rule for \C{FOREACH} takes
-advantage of the $\SL$'s bind operator to lift the effect inferred for
-the loop body to the level of the loop. Since records added to $\stl$
-in each iteration of \C{FOREACH} are independent of the previous
-iteration, sequential composition of the effects of different
-iterations is same as their parallel composition. Since the loop body
-is executed once per each $z\in x$, the effect of the the loop is the
-union of effects ($\F$) for all $z\in x$, all applied to the same
-state ($\stl$ and $\stg$). That is, $\F_{loop}(\stl,\stg) =
-\bigcup_{z\in x}\F_{body}(\stl,\stg)$. From the definition of the set
-monad's bind operator, $\F_{loop}(\stl,\stg) = x \bind (\lambda
+effects inferred for their subcommands. Thus, the effect of a sequence
+of commands $c_1;c_2$ is the union of effects $F_1$ and $F_2$ of $c_1$
+and $c_2$, respectively, except that $F_2$ is applied to a state
+($\stl$) to which $F_1$ has already been applied, reflecting their
+order of reduction. The inference rule for \C{FOREACH} takes advantage
+of the $\SL$'s bind operator to lift the effect inferred for the loop
+body to the level of the loop. Since records added to $\stl$ in each
+iteration of \C{FOREACH} are independent of the previous iteration,
+sequential composition of the effects of different iterations is same
+as their parallel composition. Since the loop body is executed once
+per each $z\in x$, the effect of the the loop is the union of effects
+($\F$) for all $z\in x$, all applied to the same state ($\stl$ and
+$\stg$). That is, $\F_{loop}(\stl,\stg) = \bigcup_{z\in
+  x}\F_{body}(\stl,\stg)$. From the definition of the set monad's bind
+operator, $\F_{loop}(\stl,\stg) = x \bind (\lambda
 z.~F_{body}(\stl,\stg))$, which mirrors the definition of the rule.
+This development mirrors ideas found in type checkers for certain
+dependent type systems~\cite{KJ14}.
 
@@
   \stable(R,I)\\
-  \R(\stl,\stg,\stg') \Leftrightarrow \I(\C{true},\stl,\stg,\stg') 
-        \wedge R(\stg,\stg') \spc
-  c \elabsto \F\\
-  \R_c(\stl,\stg,\stg') \Leftrightarrow \I(\C{false},\stl,\stg,\stg') 
-        \wedge R(\stg,\stg') \spc
-  \stable(\R_c,\F)\\
+   \R_e(\stl,\stg,\stg') \Leftrightarrow \exists \stg_1. \I_e(\stl, \stg_1, \stg) \wedge R(\stg, \stg') \wedge \I_e(\stl, \stg_1, \stg') \spc  c \elabsto \F\\
+  \R_c(\stl,\stg,\stg') \Leftrightarrow \exists \stg_1. \I_c(\stl, \stg_1, \stg) \wedge R(\stg, \stg') \wedge \I_c(\stl, \stg_1, \stg') \spc \stable(\R_c,\F)\\
   \forall \stg.~ G(\stg, F(\emptyset,\stg) \gg \stg)\spc
@@
 SLA}]$, where ${\sf SLA}$ is the theory of simple linear arithmetic.
-Since {\sf EPR}+{\sf SLA} is known to be decidable~\cite{eprsla}, hence:
+Since {\sf EPR}+{\sf SLA} is known to be decidable~\cite{eprsla}:
 \begin{theorem}

### 2017-07-08 reasoning.tex

@@
 database satisfies the high-level invariant $I$. The rely relation
-($\R_l$) for the local judgment is a ternary relation computed as $R$
+($\R_e$) for the local judgment is a ternary relation computed as $R$
 modulo $\I_e$:
@@
 \begin{array}{lcl}
-%\R_l(\stl,\stg,\stg') & \Leftrightarrow & R(\stg,\stg') \conj
+%\R_e(\stl,\stg,\stg') & \Leftrightarrow & R(\stg,\stg') \conj
 %\I_e(\stl,\stg,\stg')
-  \R_l(\stl,\stg,\stg') & \Leftrightarrow & \exists \stg_1. \I_e(\stl, \stg_1, \stg) \wedge R(\stg, \stg') \wedge \I_e(\stl, \stg_1, \stg')
+  \R_e(\stl,\stg,\stg') & \Leftrightarrow & \exists \stg_1.  R(\stg, \stg') \wedge \I_e(\stl, \stg_1, \stg')
 \end{array}
 \end{smathpar}
-In words, if there exists $\stg_1$ which was the global database state
-during the last step by a transaction and if the transaction can take
-its next step at the current global database state $\stg$, then even
-after the interference, the transaction should still be able to take
-its next step. If there is no such state $\stg_1$, then the
-transaction will be stuck, and the interference can be safely
-discarded. Often, this significantly simplifies the form of $\R_l$
-irrespective of $R$. For example, when a transaction is executed
-against a snapshot of the database (i.e. $\mathbb{I}_{ss}$), $\R_l$
-will be the identity function, since if $\stg \neq \stg'$, there does
-not exist any $\stg_1$ which satisfies the constraint for
-$\R_l(\stl,\stg,\stg')$.
+
+While $R$ describes changes in the global database state, $\I_e$
+constrain the global effects that will visible to the transaction.
+Recall, however that $\I_e$ uses the global state in effect at the
+point the transaction last took a step ($\stg_1$) to determine whether
+the transaction can proceed from the current global state ($\stg$).
+Moreover, the global rely relation $R$ defines the permissible
+interferences from $\Delta$ to a future global state ($\Delta'$).  The
+local rely relation $\R_e$ determines allowed interferences as
+specified by $\I_e$ assuming $\stg_1$.  In effect, the existential on
+$\stg_1$ defines this reley relation to be an over-approximation on
+the interferences in the presence of which the transaction can
+proceed, simplifying verification burden.
+
+%Often, this significantly simplifies the form of $\R_l$ irrespective
+%of $R$. For example, when a transaction is executed against a
+%snapshot of the database (i.e. $\mathbb{I}_{ss}$), $\R_l$ will be the
+%identity function, since if $\stg \neq \stg'$, there does not exist
+%any $\stg_1$ which satisfies the constraint for
+%$\R_l(\stl,\stg,\stg')$.
+
+%In words, if there exists $\stg_1$ which was the global database state during the last step by a transaction and if the transaction can take its next step at the current global database state $\stg$, then even after the interference, the transaction should still be able to take its next step.
 
@@
 w.r.t the commit-time interference captured by $\R_c$. $\R_c$, like
-$\R_l$, is computed as a rely relation modulo isolation, except that
+$\R_e$, is computed as a rely relation modulo isolation, except that
 commit-time isolation ($\I_c$) is considered. The validity of

### 2017-07-08 related-work.tex

@@
 vs. abstract interpretation) are quite different.
+
+\section{Conclusions}
+\label{sec:conclusions}
+
+To improve performance, modern database systems employ techniques that
+weaken the strong isolation guarantees provided by serializable
+transactions in favor of alternatives that allow a transaction to
+witness the effects of other concurrently executing transactions that
+happen commit during its execution.  Typically, it is the
+responsibility of the database programmer to determine if an available
+weak isolation level would violate a transaction's consistency
+constraints.  Although this has proven to be a difficult and
+error-prone process, there has heretofore been no attempt to formalize
+notions of weak isolation with respect to application semantics, or
+consider how we might verify the correctness of database programs that
+use weakly-isolated transactions.  In this paper, we provide such a
+formalization.  We develop a rely-guarantee proof framework cognizant
+of weak isolation semantics, and build on this foundation to devise an
+inference procedure that facilitates automated verification of
+weakly-isolated transactions, and have applied our ideas on
+widely-used database systems to justify their utility.  Our solution
+enables database applications to leverage the performance advantages
+offered by weak isolation, without compromising safety and
+correctness.

### 2017-07-08 inference.tex

@@
 alternative, we present an inference algorithm based on state
-transformers.  The idea is to infer the logical effect that each
-statement has on the transaction-local database state $\stl$ (i.e.,
-how it transforms $\stl$), and compose multiple such effects together
-to describe the effect of the transaction as a whole.  Importantly,
-this approach generalizes to loops, where the effect of a loop can be
-inferred as a well-defined function of the effect of its body, thanks
-to certain pleasant properties enjoyed by the database programs
-modeled by $\txnimp$.  Interpreting database semantics as functional
-transformations on sets (described in terms of their logical effects)
-enables an inference mechanism that can leverage off-the-shelf SMT
-solvers for automated verification.
+transformers that alleviates this burden.  The idea is to infer the
+logical effect that each statement has on the transaction-local
+database state $\stl$ (i.e., how it transforms $\stl$), and compose
+multiple such effects together to describe the effect of the
+transaction as a whole.  Importantly, this approach generalizes to
+loops, where the effect of a loop can be inferred as a well-defined
+function of the effect of its body, thanks to certain pleasant
+properties enjoyed by the database programs modeled by $\txnimp$.
+Interpreting database semantics as functional transformations on sets
+(described in terms of their logical effects) enables an inference
+mechanism that can leverage off-the-shelf SMT solvers for automated
+verification.
 
@@
 The correspondence between the inference rules and the RG judgment of
-Sec.~\ref{sec:reasoning} is stated thus\footnote{The proof can be found in
+\S~\ref{sec:reasoning} is stated thus\footnote{The proof can be found in
 the supplementary material.}:
@@
   \R \vdash \hoare{\lambda(\stl,\stg).~\stl=\emptyset \conj
-  I(\stg)}{c}{\lambda(\stl,\stg).\stl = \F(\emptyset,\stg)}
+  I(\stg)}{c}{\lambda(\stl,\stg).\stl \subseteq \F(\emptyset,\stg)}
   \end{array}
@@
   \stable(R,I)\\
-   \R_e(\stl,\stg,\stg') \Leftrightarrow \exists \stg_1. \I_e(\stl, \stg_1, \stg) \wedge R(\stg, \stg') \wedge \I_e(\stl, \stg_1, \stg') \spc  c \elabsto \F\\
-  \R_c(\stl,\stg,\stg') \Leftrightarrow \exists \stg_1. \I_c(\stl, \stg_1, \stg) \wedge R(\stg, \stg') \wedge \I_c(\stl, \stg_1, \stg') \spc \stable(\R_c,\F)\\
+   \R_e(\stl,\stg,\stg') \Leftrightarrow \exists \stg_1.  R(\stg, \stg') \wedge \I_e(\stl, \stg_1, \stg') \spc  c \elabsto \F\\
+  \R_c(\stl,\stg,\stg') \Leftrightarrow \exists \stg_1.  R(\stg, \stg') \wedge \I_c(\stl, \stg_1, \stg') \spc \stable(\R_c,\F)\\
   \forall \stg.~ G(\stg, F(\emptyset,\stg) \gg \stg)\spc
@@
 integrity, non-negativeness of all integer values in a column etc.
-The isolation specifications presented in Sec.~\ref{sec:isolation} are
+The isolation specifications presented in \S\ref{sec:isolation} are
 already simple first-order formulas that can be encoded in {\sf EPR}.

### 2017-07-08 isolation.tex

@@
 A distinctive characteristic of our development is that it is
-parameterized on a weak isolation specification $\I$. $\I$ can be
+parameterized on a weak isolation specification $\I$ that can be
 instantiated with the declarative characterization of an isolation
@@
 \textbf{Unique Ids}. As the \C{new\_order} example
-(Sec.~\ref{sec:motivation}) demonstrates, enforcing global uniqueness
+(\S\ref{sec:motivation}) demonstrates, enforcing global uniqueness
 of ordered identifiers requires stronger isolation levels than the
@@
 
-\textbf{Read Committed (RC) and Monotonic Atomic View (MAV)} RC
+\textbf{Read Committed (RC) and Monotonic Atomic View (MAV).} RC
 isolation allows a transaction to witness writes of committed
@@
 prevent a record from being modified while it is read by an
-uncommitted transaction, but not prohibiting insertion of new
+uncommitted transaction, but which does not prohibit insertion of new
 records. The ANSI SQL RR specification can be axiomatized in our

### 2017-07-08 reasoning.tex

@@
   underlying system.}.  Our proof system is essentially an adaptation
-of rely-guarantee reasoning framework~\cite{rgjones} to the setting of
+of a rely-guarantee reasoning framework~\cite{rgjones} to the setting of
 weakly isolated database transactions.  The primary challenge in the
@@
 \C{atomic} blocks also have the same characteristic, the fact that
-transactions are weakly isolated introduces complexity.  Unlike an
+transactions are weakly isolated introduces surprising complexity.  Unlike an
 \C{atomic} block, the effect of a transaction is \emph{not} a
 sequential composition of the effects of its statements because each
-statement witnesses a potentially different version of the state.
+statement can witness a potentially different version of the state.
 
@@
 transaction preserves the high-level invariant $I$ when executed under
-the required isolation as specified by $\I$. It relies on the
-transaction-local judgment to verify the body ($c$) of a transaction
-$i$. The precondition $P$ of $c$ must follow from the fact that the
+the required isolation as specified by $\I$. It depends on a
+transaction-local judgment to verify the body ($c$) of a transaction with
+id $i$. The precondition $P$ of $c$ must follow from the fact that the
 transaction-local database ($\stl$) is initially empty, and the global
@@
 \end{smathpar}
-
-While $R$ describes changes in the global database state, $\I_e$
-constrain the global effects that will visible to the transaction.
-Recall, however that $\I_e$ uses the global state in effect at the
+\noindent While $R$ describes changes in the global database state, $\I_e$
+constrains the global effects visible to the transaction.
+Recall however that $\I_e$ uses the global state in effect at the
 point the transaction last took a step ($\stg_1$) to determine whether
@@
 
-Salient rules of transaction-local RG judgment are shown in
+Salient rules of transaction-local RG judgments are shown in
 Fig.~\ref{fig:rg-rules}. These rules (\rulelabel{RG-Update},

### 2017-07-08 implementation.tex
Fixing typos.

--- 2017-07-08 implementation.tex
@@
 
-At the top-level, \tool runs a loop that pics an unverified
+At the top-level, \tool runs a loop that picks an unverified
 transaction and progressively strengthens its isolation level until it

### 2017-07-08 motivation.tex

@@
 behavior of the \C{new\_order} transaction
+when executed with a \emph{Read Committed} (RC) isolation level, the
 \begin{wrapfigure}{l}{.4\textwidth}
@@
   executed.  The subsequent operations are effectively sequentialized,
-  since $T_2$ commits. Nonetheless, reading the same value for
-  \C{d\_next\_o\_id} prompts both instances to add \C{Order} records
-  with same ids, which in turn triggers a violation of TPC-C's
+  since $T_2$ commits. Nonetheless, both transactions read the same value for
+  \C{d\_next\_o\_id} resulting in them  adding \C{Order} records
+  with the same ids, which in turn triggers a violation of TPC-C's
   consistency condition.}
@@
 \end{wrapfigure}
-when executed with a \emph{Read Committed} (RC) isolation level, the
 default isolation level in 8 of the 18 databases studied
@@
 database, thus admitting concurrency, but it also requires that there
-not be any write-write conflicts (i.e,, both transactions modifying
-the same record) among concurrent transactions when they
-commit. Write-write conflicts can be eliminated in various ways, e.g.,
-through conflict detection followed by a rollback, or through
-exclusive locks, or a combination of both.  For instance, one possible
-implementation of SI, close to the one used by
-PostgreSQL~\cite{postgresiso}, executes a transaction against its
+not be any write-write conflicts (i.e,, such a conflict occurs if
+concurrentlyh executing transactions modify the same record) among
+concurrent transactions when they commit. Write-write conflicts can be
+eliminated in various ways, e.g., through conflict detection followed
+by a rollback, or through exclusive locks, or a combination of both.
+For instance, one possible implementation of SI, close to the one used
+by PostgreSQL~\cite{postgresiso}, executes a transaction against its
 private snapshot of the database, but obtains exclusive locks on the

### 2017-07-08 opsem.tex

@@
 taking into account the local database state ($\stl$). For instance,
-as described in (Sec.~\ref{sec:motivation}), an SI transaction on
+as described in \S\ref{sec:motivation}, an SI transaction on
 PostgreSQL defines $\I$ as follows:
@@
 easy to relax in the operational semantics and the reasoning framework
-presented in the next section; the automation framework introduced in
-Sec.~\ref{sec:inference}, however, has a non-trivial dependence on
+presented in the next section; our inference procedure described in
+\S\ref{sec:inference}, however, has a non-trivial dependence on
 this assumption. Nonetheless, we have encountered few instances in

### 2017-09-13 popl-rebuttal.md
Rebuttal.

Firstly, we thank the reviewers for their valuable feedback.

1. Read Uncommitted (RU), while a valid ANSI SQL isolation level, is
   often not implemented by the databases. For eg., Oracle supports
   only RC and SER, while Postgres maps RU to RC. Even when RU is
   implemented, it is never the default isolation level (Bailis et al,
   HotOS'13), and applications almost never demote the isolation level
   in practice (Bailis et al, SIGMOD'15). Thus, RC can be considered
   as a baseline isolation level that meets the performance demands of
   real-world applications. Lastly, RU violates atomicity, the basic
   tenet underlying transactions, and accommodating it would
   significantly complicate the formal framework. Weighing such pros
   and cons, we decided to elide RU.

2. While our language only includes the four basic SQL operations,
   they are nonetheless enough to implement database community's
   standard benchmarks, such TPC-C and TPC-E. Moreover, MySQL's
   (developer) documentation seems to indicate that its core calculus
   for SQL is indeed composed of these four basic operations, making
   our technique likely applicable on MySQL's core language, if not
   its surface language. We haven't analyzed other databases, but if
   the need arises, we can accommodate database-specific semantics as
   elaboration rules on the lines described in Sec. 6.1 para 1.

3. About our framework not admitting all SI executions, it seems that
   reviewer might have misunderstood our specification for SI. The
   spec allows concurrent transactions that do not conflict with the
   current SI transaction to commit concurrently, since SI's I_c spec
   (p.11) allows Δ to change as long as it doesn't conflict with δ.

4. We have ignored heap memory to focus on the database state in this
   paper. N Swamy et al (PLDI'13) have proposed Dijkstra monad to
   reason about heap memory monadically, and it would be interesting
   to see if Diskstra and DB monads can be cleanly composed to reason
   about the combination of heap and database.

### 2017-09-13 rebuttal.txt
edits

We thank the reviewers for their valuable feedback.

Reviewer (1):
-------------

- No support for Read Uncommitted (RU) isolation.  While RU is a valid
  ANSI SQL isolation level, it is typically not implemented by modern
  databases. For example, Oracle supports only RC and SER, while
  Postgres maps RU to RC. Even when RU is supported, it is never the
  default isolation level (Bailis et al, HotOS'13)., and applications
  rarely demote isolation levels to RU in practice (Bailis et al,
  SIGMOD'15). Thus, RC can be considered as a baseline isolation level
  that meets the performance demands of real-world applications.
  Furthermore, RU violates atomicity, the basic tenet underlying
  transactions.  Given its lack of acceptance in the database
  community, and the complications it would introduce to our
  formalization, we opted to elide it from the development.

- Limited set of operations in the DSL.  While it is true our language
  only includes four basic SQL operations, they have proven sufficient
  to implement widely accewpted standard benchmarks, such TPC-C and
  TPC-E.  We also note that MySQL's developer documentation (see:
  ....) indicates that its core calculus for SQL is indeed
  composed of these four basic operations, making our technique likely
  applicable on MySQL's core language, if not its surface language. We
  expect similar conclusions to hold on other databases.

Reviewer (2):
-------------

- Not admitting valid snapshot isolation (SI) executions.  We believe
  the reviewer has misunderstood our specification for SI as $I_{ss}$,
  which only describes the semantics of "executing against a
  snapshot". SI indeed executes against a snapshot (thus its $I_e =
  I_{ss}$), but it *commits* against any non-write-write-conflicting
  database state (thus its $I_c = I_{ww}$). This is explained on p.13
  ("Snapshot Isolation").  The presented specification indeed allows
  non-conflicting concurrent transactions to commit in parallel.

- Decidability.  (a) The RG-TXN (Sec. 5.2, p. 20) rule does not lead
  to second-order quantification or inversion of quantifier positions.
  First, recall that each premise of the rule generates an independent
  verification condition (VC) dispatched to the solver. The VC
  corresponding to $\mathbf{F}\Rightarrow G$ first introduces $\delta$
  and $\Delta$ into the context as some sets (uninterpreted
  functions).  Next, it asserts the logical encoding of $F$ computed
  using the rules in Fig.10. Finally it asserts the negation of $G$,
  and checks for satisfiability. Note that in this process: (i). sets
  or functions are never quantified (thus no second-order
  quantification), and (ii). $\mathbf{F}$ is never negated (thus no
  inversion of quantifiers in $\mathbf{F}$). Other premises are
  similarly encoded.

  The reviewer is correct in observing that our encoding is not in the
  EPR fragment of FOL since we admit formulas of the form $\forall
  x.\exists y.\phi$.  However, our encoding is still decidable as it
  belongs to a closely related variant of EPR, Godel-Kalmar-Schutte,
  that admits sentences of the form: $\exists^{}\forall^{2}\exists^{}$
  (see: Borger et al, The Classical Decision Problem, p.261; Gurevich,
  On the Classical Decision Problem, p.7); formulas generated by our
  encoding clearly belong to this fragment.  We will clarify this
  issue in the final version.

Reviewer (3);
-------------

- Dealing with heap state: It is indeed true that we have ignored heap
  memory to focus exclusively on database state in our
  formalization. Swamy et al. (PLDI'13), however, have proposed the
  Dijkstra monad to reason about heap memory monadically.  As part of
  future work, we intend to explore the composition of the DB and
  Dijkstra monads to enable reasoning over both heap and database
  state.

### 2017-09-13 popl-rebuttal.md

@@
    tenet underlying transactions, and accommodating it would
-   significantly complicate the formal framework. Weighing such pros
-   and cons, we decided to elide RU.
+   significantly complicate the formal framework, jeopardizing
+   automation. Weighing such pros and cons, we decided to elide RU.
 
@@
    standard benchmarks, such TPC-C and TPC-E. Moreover, MySQL's
-   (developer) documentation seems to indicate that its core calculus
-   for SQL is indeed composed of these four basic operations, making
-   our technique likely applicable on MySQL's core language, if not
-   its surface language. We haven't analyzed other databases, but if
-   the need arises, we can accommodate database-specific semantics as
-   elaboration rules on the lines described in Sec. 6.1 para 1.
+   [developer documentation](https://dev.mysql.com/doc/internals/en/)
+   seems to indicate that its core calculus for SQL is indeed composed
+   of these four basic operations, making our technique likely
+   applicable on MySQL's core language, if not its surface language.
+   We haven't analyzed other databases, but if the need arises, we can
+   accommodate database-specific semantics as elaboration rules on the
+   lines described in Sec. 6.1 para 1.
 
-3. About our framework not admitting all SI executions, it seems that
-   reviewer might have misunderstood our specification for SI. The
-   spec allows concurrent transactions that do not conflict with the
-   current SI transaction to commit concurrently, since SI's I_c spec
-   (p.11) allows Δ to change as long as it doesn't conflict with δ.
+3. About our framework not admitting all SI executions, the reviewer has 
+   possibly misunderstood our specification for SI as
+   $I_{ss}$, which only describes the semantics of "executing
+   against a snapshot". SI indeed executes against a snapshot (thus its $I_e
+   = I_{ss}$), but it commits against any non-write-write-conflicting
+   database state (thus its $I_c = I_{ww}$). This is explained under
+   "Snapshot Isolation" heading on p.13. Observe that the spec allows
+   non-conflicting concurrent transactions to commit in parallel.
 
-4. We have ignored heap memory to focus on the database state in this
+4. The RG-TXN rule does not lead to second-order quantification or
+   inversion of quantifier positions for the reasons described below.
+   Firstly, lets recall that each premise of the rule generates an
+   independent verification condition (VC) dispatched to the solver.
+   The VC corresponding to $F\Rightarrow G$ first introduces $\delta$
+   and $\Delta$ into the context as *some* sets (uninterpreted
+   functions, to be precise). Next, it asserts the logical encoding of
+   $F$ computed using the rules in Fig.10. Finally it asserts the
+   negation of $G$, and checks for satisfiability. Note that in this
+   process: (a). Sets or functions are never quantified (thus no
+   second-order quantification), and (b). $F$ is never negated (thus
+   no inversion of quantifiers in $F$). Other premises are similarly
+   encoded.
+
+5. We have been a bit sloppy about describing the fragment of FOL that
+   our logic belongs to. The reviewer is indeed right that $\forall
+   x.\exists y.\phi$ doesn't belong to EPR. The appropriate fragment
+   in our case is a closely related fragment called
+   Godel-Kalmar-Schutte, which admits sentences of form
+   $\exists^{*}\forall^{2}\exists^{*}$, while remaining decidable
+   (Borger et al, The Classical Decision Problem, p.261; Gurevich, On
+   the Classical Decision Problem, p.7). We apologize for not being
+   precise, and will promptly correct the mistake.
+
+6. We have ignored heap memory to focus on the database state in this
    paper. N Swamy et al (PLDI'13) have proposed Dijkstra monad to
    reason about heap memory monadically, and it would be interesting
-   to see if Diskstra and DB monads can be cleanly composed to reason
+   to see if Dijkstra and DB monads can be cleanly composed to reason
    about the combination of heap and database.

### 2017-09-13 evaluation.tex

@@
-\subsection{Evaluation}
+\section{Evaluation}
 \label{sec:case-studies}
 
-In this section, we present our experience in running \tool on two different applications: \emph{Courseware}: a
-course registration system described by~\cite{gotsmanpopl16}, and
-% 2. RUBiS: an eBay-like auction site~\cite{rubis} that allows users to
-% bid for items on sale, and pay for items from a wallet modeled after a
-% bank account, and 
-the \emph{TPC-C} benchmark.
+In this section, we present our experience in running \tool on two
+different applications: \emph{Courseware}: a course registration
+system described by~\cite{gotsmanpopl16}, and the \emph{TPC-C}
+benchmark.
 
@@
 \multicolumn{1}{|l|}{MySQL}      & SER                                   & SER                                 & RC                                 & RC                                       & RC                                      \\ \hline
-\multicolumn{1}{|l|}{PostgreSQL} & RR                                    & RR                                  & RC                                 & RC                                       & RC                                      \\ \hline
+\multicolumn{1}{|l|}{PostgreSQL} & SI                                    & SI                                  & RC                                 & RC                                       & RC                                      \\ \hline
 \end{tabular}
@@
 found 171 of them violated at least one of the four invariants we
-considered.  Table~\ref{fig:tpcc-eval} shows the summary of these
-results. As expected, the isolation levels that \tool infers for the
+considered.  As expected, the isolation levels that \tool infers for the
 TPC-C transactions do not result in invariant violations, either on

### 2017-09-13 implementation.tex

--- 2017-07-08 implementation.tex
@@
 4.03 compiler\footnote{The source code is available at available at
-\emph{\ldots}}. The input to \tool is a program in our DSL that
+\emph{https://goo.gl/nmUXZK}}. The input to \tool is a program in our DSL that
 describes the schema of the database as a collection of OCaml type
@@
 implication check described above is first encoded in FOL, applying
-the translation described in Sec.~\ref{sec:inference} wherever
+the translation described in \S\ref{sec:inference} wherever
 necessary.
@@
 \textbf{Real-World Isolation Levels} The axiomatization of the
-isolation levels presented in Sec.~\ref{sec:isolation} leave out
+isolation levels presented in \S\ref{sec:isolation} leave out
 certain nuances of their implementations on real data stores, which

### 2017-09-13 introduction.tex

@@
 programmers from using them, we present an inference procedure based
-on a state transformer semantics that automatically verifies the
-weakest isolation level for a transaction that ensures its consistency
-requirements are maintained.  The key to inference is a novel formulation
-of database state (represented as sets of tuples) as a monad, and in
-which database computations are interpreted as transformers over these
-sets.  This interpretation leads to an encoding of database
-computations amenable for verification by off-the-shelf SMT solvers.
-The paper makes the following contributions:
+that automatically verifies the weakest isolation level for a
+transaction that ensures its consistency requirements are maintained.
+The key to inference is a novel formulation of database state
+(represented as sets of tuples) as a monad, and in which database
+computations are interpreted as state transformers over these sets.
+This interpretation leads to an encoding of database computations
+amenable for verification by off-the-shelf SMT solvers.  The paper
+makes the following contributions:
 %% We have realized these ideas as an embedded DSL within OCaml, that
@@
 algorithm, and the state transformer encoding.  We describe our
-implementation, benchmark results, and case studies in
-\S\ref{sec:case-studies}.  Related work and conclusions are given in
-\S\ref{sec:relatedwork}.
+implementation in \S\ref{sec:implementation}, and provide case studies
+and benchmark results in \S\ref{sec:case-studies}.  Related work is
+given in \S\ref{sec:relatedwork}, and \S\ref{sec:conclusions} presents
+conclusions.

### 2017-10-03 popl-rebuttal.md

--- 2017-09-13 popl-rebuttal.md
@@
-Firstly, we thank the reviewers for their valuable feedback.
+We thank the reviewers for their valuable feedback.
 
-1. Read Uncommitted (RU), while a valid ANSI SQL isolation level, is
-   often not implemented by the databases. For eg., Oracle supports
-   only RC and SER, while Postgres maps RU to RC. Even when RU is
-   implemented, it is never the default isolation level (Bailis et al,
-   HotOS'13), and applications almost never demote the isolation level
-   in practice (Bailis et al, SIGMOD'15). Thus, RC can be considered
-   as a baseline isolation level that meets the performance demands of
-   real-world applications. Lastly, RU violates atomicity, the basic
-   tenet underlying transactions, and accommodating it would
-   significantly complicate the formal framework. Weighing such pros
-   and cons, we decided to elide RU.
+Reviewer A
+-----------
 
-2. While our language only includes the four basic SQL operations,
-   they are nonetheless enough to implement database community's
-   standard benchmarks, such TPC-C and TPC-E. Moreover, MySQL's
-   (developer) documentation seems to indicate that its core calculus
-   for SQL is indeed composed of these four basic operations, making
-   our technique likely applicable on MySQL's core language, if not
-   its surface language. We haven't analyzed other databases, but if
-   the need arises, we can accommodate database-specific semantics as
-   elaboration rules on the lines described in Sec. 6.1 para 1.
+- Read Uncommitted (RU) isolation: While RU is a valid ANSI SQL
+  isolation level, it is typically not implemented by modern
+  databases. For example, Oracle supports only RC and SER, while
+  Postgres maps RU to RC. Even when RU is supported, it is never the
+  default isolation level (Bailis et al, HotOS'13)., and applications
+  rarely demote isolation levels to RU in practice (Bailis et al,
+  SIGMOD'15). Furthermore, RU violates atomicity, the basic tenet
+  underlying transactions.  Given its lack of acceptance in the
+  database community, and the complications it would introduce to our
+  formalization, we opted to use RC as the baseline isolation level.
 
-3. About our framework not admitting all SI executions, it seems that
-   reviewer might have misunderstood our specification for SI. The
-   spec allows concurrent transactions that do not conflict with the
-   current SI transaction to commit concurrently, since SI's I_c spec
-   (p.11) allows Δ to change as long as it doesn't conflict with δ.
+- Limited set of operations in the DSL: While it is true our language
+  only includes four basic SQL operations, they have proven sufficient
+  to implement widely accepted benchmarks, such TPC-C and TPC-E. We
+  also note that MySQL's [developer
+  documentation](https://dev.mysql.com/doc/internals/en/) (also see: S
+  Pachev, Understanding MySQL Internals, 1st Edition, Chapter 9)
+  indicates that its core calculus for SQL is indeed composed of these
+  four basic operations, making our technique likely applicable on
+  MySQL's core language, if not its surface language. We expect
+  similar conclusions to hold on other databases.
 
-4. We have ignored heap memory to focus on the database state in this
-   paper. N Swamy et al (PLDI'13) have proposed Dijkstra monad to
-   reason about heap memory monadically, and it would be interesting
-   to see if Diskstra and DB monads can be cleanly composed to reason
-   about the combination of heap and database.
+Reviewer B
+-------------
+
+- Isolation levels weaker than RC: please see above.
+
+- Not admitting valid snapshot isolation (SI) executions:  We believe
+  the reviewer has misunderstood our specification for SI as $I_{ss}$,
+  which only describes the semantics of "executing against a
+  snapshot". SI indeed executes against a snapshot (thus its $I_e =
+  I_{ss}$), but it *commits* against any non-write-write-conflicting
+  database state (thus its $I_c = I_{ww}$). This is explained on p.13
+  ("Snapshot Isolation"). The presented specification indeed allows
+  non-conflicting concurrent transactions to commit in parallel, and hence
+  admits all SI executions.
+
+- Decidability: The RG-TXN (Sec. 5.2, p. 20) rule does not lead to
+  second-order quantification or quantified formulas with negation at
+  prenex position. Sets are encoded as uninterpreted functions (Fig.
+  10), and quantified set variables are simply replaced with new
+  function symbols that occur free in the encoded formula.  Consider
+  the ${\sf F} \Rightarrow G$ premise of RG-TXN. The quantified set
+  variables $\delta$ and $\Delta$ are instantiated with fresh function
+  variables $f_{\delta}$ and $f_{\Delta}$, respectively, resulting in
+  a quantifier-free implication with free function symbols. The
+  antecedent of this implication (${\sf F}$) is asserted as such, but
+  its consequent ($G$) is negated before checking the satisfiability
+  of the entire formula. This is a standard approach often employed to
+  discharge verification conditions of the form $\forall a. \phi_1(a)
+  \Rightarrow \phi_2(a)$.  In this process: (i). sets or functions are
+  never quantified (thus no second-order quantification), and
+  (ii). ${\sf F}$ is never negated (thus no inversion of quantifiers
+  in ${\sf F}$). Other premises are similarly encoded.
+
+  The reviewer is correct in observing that our encoding is not in the
+  EPR fragment of FOL.  However, our encoding is still decidable as it
+  belongs to a closely related variant of EPR, Godel-Kalmar-Schutte,
+  that admits sentences of the form:
+  $\exists^{*}\forall^{2}\exists^{*}$ (see: Borger et al, The
+  Classical Decision Problem, 1st Edition, p.261, and Gurevich, On the
+  Classical Decision Problem, p.7); formulas generated by our encoding
+  clearly belong to this fragment.  We will clarify this issue in the
+  final version.
+
+- Local Context Independence: This assumption is used by the inference
+  rule for `FOREACH`. The rule computes the transformer of a loop by
+  lifting the transformer (${\sf F}$) of its body. In the process, it
+  assumes that the effect of each loop iteration depends only on
+  the loop variable $z$, but not on local ($\delta$) and global
+  ($\Delta$) states (observe that ${\sf F}$ is applied to the same
+  $\delta$ and $\Delta$ forall $z$). While independence on $\Delta$ is
+  gauranteed by the stability condition (checked separately), the
+  independence on the local state $\delta$ is simply assumed.
+
+Reviewer C
+-------------
+
+- Dealing with heap state: It is indeed true that we have ignored heap
+  memory to focus exclusively on database state in our
+  formalization. Swamy et al. (PLDI'13), however, have proposed the
+  Dijkstra monad to reason about heap memory monadically.  As part of
+  future work, we intend to explore the composition of the DB and
+  Dijkstra monads to enable reasoning over both heap and database
+  state.

### 2017-10-03 popl-rebuttal.txt

--- 2017-10-03 popl-rebuttal.md
@@
   rarely demote isolation levels to RU in practice (Bailis et al,
-  SIGMOD'15). Furthermore, RU violates atomicity, the basic tenet
-  underlying transactions.  Given its lack of acceptance in the
-  database community, and the complications it would introduce to our
-  formalization, we opted to use RC as the baseline isolation level.
+  SIGMOD'15). Thus, RC can be considered as a baseline isolation level
+  that meets the performance demands of real-world applications.
+  Furthermore, RU violates atomicity, the basic tenet underlying
+  transactions.  Given its lack of acceptance in the database
+  community, and the complications it would introduce to our
+  formalization, we opted to elide it from the development.
 
@@
   only includes four basic SQL operations, they have proven sufficient
-  to implement widely accepted benchmarks, such TPC-C and TPC-E. We
-  also note that MySQL's [developer
+  to implement widely accewpted standard benchmarks, such TPC-C and
+  TPC-E. We also note that MySQL's [developer
   documentation](https://dev.mysql.com/doc/internals/en/) (also see: S
@@
 
-- Isolation levels weaker than RC: please see above.
+- Isolation levels weaker than RC: please refer to our comments on RU
+  isolation above.
 
@@
   ("Snapshot Isolation"). The presented specification indeed allows
-  non-conflicting concurrent transactions to commit in parallel, and hence
+  non-conflicting concurrent transactions to commit in parallel, hence
   admits all SI executions.
@@
   10), and quantified set variables are simply replaced with new
-  function symbols that occur free in the encoded formula.  Consider
-  the ${\sf F} \Rightarrow G$ premise of RG-TXN. The quantified set
-  variables $\delta$ and $\Delta$ are instantiated with fresh function
-  variables $f_{\delta}$ and $f_{\Delta}$, respectively, resulting in
-  a quantifier-free implication with free function symbols. The
-  antecedent of this implication (${\sf F}$) is asserted as such, but
-  its consequent ($G$) is negated before checking the satisfiability
-  of the entire formula. This is a standard approach often employed to
-  discharge verification conditions of the form $\forall a. \phi_1(a)
-  \Rightarrow \phi_2(a)$.  In this process: (i). sets or functions are
-  never quantified (thus no second-order quantification), and
-  (ii). ${\sf F}$ is never negated (thus no inversion of quantifiers
-  in ${\sf F}$). Other premises are similarly encoded.
+  function symbols that occur free in the encoded formula. For
+  instance consider the ${\sf F} \Rightarrow G$ premise of RG-TXN. The
+  quantified set variables $\delta$ and $\Delta$ are instantiated with
+  fresh function variables $f_{\delta}$ and $f_{\Delta}$,
+  respectively, resulting in a quantifier-free implication with free
+  function symbols. The antecedent of this implication (${\sf F}$) is
+  asserted as such, but its consequent ($G$) is negated before
+  checking the satisfiability of the entire formula. This is a
+  standard approach often employed to discharge the verification
+  conditions of form $\forall a. \phi_1(a) \Rightarrow \phi_2(a)$.
+  Note that in this process: (i). sets or functions are never
+  quantified (thus no second-order quantification), and (ii). ${\sf
+  F}$ is never negated (thus no inversion of quantifiers in ${\sf
+  F}$). Other premises are similarly encoded.
 
   The reviewer is correct in observing that our encoding is not in the
-  EPR fragment of FOL.  However, our encoding is still decidable as it
+  EPR fragment of FOL since we admit formulas of the form $\forall
+  x.\exists y.\phi$.  However, our encoding is still decidable as it
   belongs to a closely related variant of EPR, Godel-Kalmar-Schutte,

### 2017-10-10 motivation.tex

@@
 Because the aforementioned reasoning is reasonably simple to perform
-manually, verifying the soundess of TPC-C's consistency conditions
+manually, verifying the soundness of TPC-C's consistency conditions
 would appear to be feasible.  Serializability aids the tractability of
@@
 not be any write-write conflicts (i.e,, such a conflict occurs if
-concurrentlyh executing transactions modify the same record) among
+concurrently executing transactions modify the same record) among
 concurrent transactions when they commit. Write-write conflicts can be
@@
 As this discussion indicates, implementations of SI on real databases
-such as PostgresSQL are highly complicated, often running into
+such as PostgreSQL are highly complicated, often running into
 thousands of lines of code.  Nonetheless, the semantics of SI, in
@@
 the program itself. In particular, SQL statements become set
-operations, and the \C{DB} monad's $\bind$ becoms set monadic $\bind$.
+operations, and the \C{DB} monad's $\bind$ becomes set monadic $\bind$.
 The advantage of inferring such transformers is that we can now make

### 2017-10-10 reasoning.tex

@@
 specified by $\I_e$ assuming $\stg_1$.  In effect, the existential on
-$\stg_1$ defines this reley relation to be an over-approximation on
+$\stg_1$ defines this rely relation to be an over-approximation on
 the interferences in the presence of which the transaction can

### 2017-10-10 related-work.tex

@@
 the declarative nature of our specifications make them suitable for
-formal reasoning about program behaviour. Sivaramakrishnan \emph{et
+formal reasoning about program behavior. Sivaramakrishnan \emph{et
 al.}~\cite{pldi15} specify isolation levels declaratively as trace
@@
 verified (integrity constraints over relational tables vs. memory
-safety, or linearizability of concurrent data strucgture operations)
+safety, or linearizability of concurrent data structure operations)
 and the analysis technique used to drive inference (state transformers

### 2017-10-18 inference.tex
Fixing typos.
Minor Changes in RG-rules

@@
   \R \vdash \hoare{\lambda(\stl,\stg).~\stl=\emptyset \conj
-  I(\stg)}{c}{\lambda(\stl,\stg).\stl \subseteq \F(\emptyset,\stg)}
+  I(\stg)}{c}{\lambda(\stl,\stg).\stl = \F(\emptyset,\stg)}
   \end{array}
@@
 interpretation as discussed below.
+
+\begin{proof}
+We only provide a brief proof sketch here. We prove a slightly stronger statement :
+
+\emph{Forall} i,$R$,$I$,$c$,$\F$,$s$ if $\stable(\R,I)$ and $c \elabsto \F$,
+  then:\\
+
+\begin{smathpar}
+  \begin{array}{c}
+  \R \vdash \hoare{\lambda(\stl,\stg).~\stl=s \conj
+  I(\stg)}{c}{\lambda(\stl,\stg).\stl = \F(s,\stg)}
+  \end{array}
+  \end{smathpar}
+
+We use structural induction on $c$ to prove the statement.
+Assume that $P \Leftrightarrow \lambda(\stl,\stg).~\stl=s \conj I(\stg)$ and $Q \Leftrightarrow \lambda(\stl,\stg).\stl = \F(s,\stg)$.
+For each type of $c$, our strategy is to show the premises of the corresponding RG-rule.
+The base cases correspond to the INSERT, UPDATE and DELETE statements.
+In all these cases, the form of $F$ directly corresponds to the condition linking the pre-condition $P$ and post-condition $Q$ in the premises of the RG-rules.
+If $\stable(\R, F)$, then $\stabilize{F} = F$ and hence the proof directly follows.
+Otherwise, $\stabilize{F} = \lambda (\stl, \stg).\existsl(\stg', I(\stg'), F(\stl, \stg'))$.
+Since the pre-condition $P(\stl, \stg)$ asserts $I(\stg)$, any state which obeys $P$ provides the existential $\stg'$ in the above set expression, and hence $F$ applied on that state matches the condition linking $P$ and $Q$.
+The stability of $P$ with respect to $\R$ follows directly from the stability of $I$.
+
+The proofs for SELECT, SEQ and IF-THE-ELSE directly use the inductive hypothesis to infer the RG-judgements present in the premises of their corresponding RG-rules.
+The interesting case is the FOREACH statement, for which we use the loop invariant $\psi(\stl, \stg) \Leftrightarrow \stl = s \cup y \bind (\lambda z.~\F(s, \stg))$, (where assuming that $c$ is the body of the loop, $c \elabsto F$) to prove all the premises of \rulelabel{RG-Foreach}.
+Using the same notation as the rule $\rulelabel{RG-Foreach}$, $y$ refers to the records already processed in previous iterations of the loop, while $z$ refers to the record being processed in the current iteration.
+At the beginning of the loop $[\phi/y]\psi(\stl, \stg)$ just reduces to $\stl = s$ which is implied by the pre-condition $P$.
+From the inductive hypothesis, we can infer that each iteration corresponds to the application of $F$.
+Since all iterations are assumed to be independent of each other, and $z$ is bound to a record in $x$ for each iteration, we can infer that at the end of every iteration the loop invariant $[y \cup \{z\}/y]\psi$ will be satisfied.
+\end{proof}
 % Where, ${\sf true} = \lambda(\stl,\stg).~true$ and $\stable(\R,I)
@@
   \stable(R,\I)\spc
-  \stable(R,I)\\
+  \stable(R,I)\spc
    \R_e(\stl,\stg,\stg') \Leftrightarrow \exists \stg_1.  R(\stg, \stg') \wedge \I_e(\stl, \stg_1, \stg') \spc  c \elabsto \F\\
   \R_c(\stl,\stg,\stg') \Leftrightarrow \exists \stg_1.  R(\stg, \stg') \wedge \I_c(\stl, \stg_1, \stg') \spc \stable(\R_c,\F)\\
-  \forall \stg.~ G(\stg, F(\emptyset,\stg) \gg \stg)\spc
+  \forall \stl,\stg.~ \stl \in F(\emptyset,\stg) \Rightarrow 
+    G(\stg, \stl\gg\stg)\spc
   \forall \stg,\stg'.~I(\stg) \wedge G(\stg,\stg') \Rightarrow I(\stg')\\

### 2017-10-18 reasoning.tex
Minor Changes in RG-rules

@@
 convey the same intuition as above.
+In our setting, we only need to prove the stability of an assertion ($Q$) against those environment steps which lead to a global database state on which the transaction itself can take its next step according to its isolation specification ($\I$).
+
+\begin{smathpar}
+\begin{array}{lcl}
+\stable(R, Q) & \Leftrightarrow & \forall \stl, \stg, \stg'. Q(\stl, \stg) \wedge R^{*}(\stg, \stg') \wedge \I(\stl, \stg, \stg') \Rightarrow Q(\stl, \stg')
+\end{array}
+\end{smathpar}
+
+The above condition uses the transitive closure of $R$, while the standard RG stability assertion only uses $R$ and then relies on an inductive argument to prove stability for the transitive closure.
+We cannot directly replace $R^{*}$ by $R$ in the above condition because the inductive argument would not hold, as $R^{*}$ modulo $\I$ is not the same as ($R$ modulo $\I$$)^{*}$.To obtain an equivalent form in our setting, we require an additional condition on the isolation specification, which we call the stability condition on $\I$.
+The condition requires $\I$ to allow
+an interference $R^{*}(\stg,\stg'')$, for two database states $\stg$
+and $\stg''$, only if it also allows interference for every prefix of
+$R^{*}(\stg,\stg'')$. In other words, if $\I$ disallows interference
+from $\stg$ to $\stg'$, then an $R$-step from $\stg'$ to $\stg''$
+should not make the interference from $\stg$ to $\stg''$ valid.  This
+stability condition on $\I$ is defined formally thus:
+\begin{smathpar}
+\begin{array}{lcl}
+  \stable(R,\I) & \Leftrightarrow & \forall \stl,\stg,\stg',\stg''.~
+  \neg\I(\stl,\stg,\stg') \conj R(\stg',\stg'') \Rightarrow
+  \neg\I(\stl,\stg,\stg'')
+\end{array}
+\end{smathpar}
+It can be easily verified that the above stability condition is
+satisfied by the isolation axioms from Sec.~\ref{sec:isolation}. For
+instance, $\I_{ss}$, the snapshot axiom, is stable because if
+$\I_{ss}$ is invalid ($\neg\I_{ss}$), then an interference has already
+modified a record, and no further interference will restore the
+original record, because the original record bears the id of a
+transaction that has already committed.
+We can now generalize the stability condition for any arbitrary assertion $Q$ by only considering those environment steps which in addition to preserving $Q$, also preserve the condition that the transaction can continue to take its next step, i.e.
+the condition $\I$ holds on both the beginning and the end states.
+
+ \begin{smathpar}
+\forall \stl, \stg, \stg', \stg_1. Q(\stl, \stg) \wedge \I(\stl, \stg_1, \stg) \wedge R^{*}(\stg, \stg') \wedge \I(\stl, \stg_1, \stg') \Rightarrow Q(\stl, \stg')
+\end{smathpar}
+
+In the above condition, we use a separate notation ($\stg_1$) for the global database state that the transaction witnessed during its last step.
+$Q$ does not constrain this global database state, but the stability of $\I$ ensures that any database state constrained by $Q$ must also obey $\I$.
+We can now replace the transitive closure in the above condition with the single step relation $R$, backed by the inductive argument for stability.
+Pushing the universal quantifier for $\stg_1$ inside the antecedent and separating the predicates involving $\stg_1$, we can rewrite the condition in the form below, by defining a new ternary relation $\R$, which can be thought of as $R$ modulo $\I$:%stating that an assertion $Q$ on a database state $\stg$ on which the transaction can take its next step (hence the condition $\I(\stl, \stg_1, \stg)$) is stable w.r.t.
+environment steps to $\stg'$
+
+\begin{smathpar}
+\begin{array}{lcl}
+\stable((R \backslash \I), Q) & \Leftrightarrow & \forall \stl,\stg,\stg'. Q(\stl,\stg) \wedge (R \backslash \I)(\stl,\stg,\stg') \Rightarrow Q(\stl,\stg')\\
+(R \backslash \I)(\stl, \stg, \stg') & \Leftrightarrow & \exists \stg_1. R(\stg, \stg') \wedge \I(\stl, \stg_1, \stg) \wedge \I(\stl, \stg_1, \stg')
+\end{array}
+\end{smathpar}
+
+The above condition often significantly simplifies the form of $\R$ irrespective
+of $R$. For example, when a transaction is executed against a
+snapshot of the database (i.e. $\mathbb{I}_{ss}$), $\R$ will be the
+identity function, since if $\stg \neq \stg'$, there does not exist
+any $\stg_1$ which satisfies the constraint for
+$\R_l(\stl,\stg,\stg')$.
 
@@
 database satisfies the high-level invariant $I$. The rely relation
-($\R_e$) for the local judgment is a ternary relation computed as $R$
-modulo $\I_e$:
-\begin{smathpar}
-\begin{array}{lcl}
+($\R_e$) is obtained from the global rely relation $R$ and the isolation specification $\I_e$ as explained above.
+Note that $\I_e$ constrains the global effects visible to the transaction while it is executing but has not yet committed.
+
+%for the local judgment is a ternary relation computed as $R$
+%modulo $\I_e$:
+%\begin{smathpar}
+%\begin{array}{lcl}
 %\R_e(\stl,\stg,\stg') & \Leftrightarrow & R(\stg,\stg') \conj
 %\I_e(\stl,\stg,\stg')
-  \R_e(\stl,\stg,\stg') & \Leftrightarrow & \exists \stg_1.  R(\stg, \stg') \wedge \I_e(\stl, \stg_1, \stg')
-\end{array}
-\end{smathpar}
-\noindent While $R$ describes changes in the global database state, $\I_e$
-constrains the global effects visible to the transaction.
-Recall however that $\I_e$ uses the global state in effect at the
-point the transaction last took a step ($\stg_1$) to determine whether
-the transaction can proceed from the current global state ($\stg$).
-Moreover, the global rely relation $R$ defines the permissible
-interferences from $\Delta$ to a future global state ($\Delta'$).  The
-local rely relation $\R_e$ determines allowed interferences as
-specified by $\I_e$ assuming $\stg_1$.  In effect, the existential on
-$\stg_1$ defines this rely relation to be an over-approximation on
-the interferences in the presence of which the transaction can
-proceed, simplifying verification burden.
-
-%Often, this significantly simplifies the form of $\R_l$ irrespective
-%of $R$. For example, when a transaction is executed against a
-%snapshot of the database (i.e. $\mathbb{I}_{ss}$), $\R_l$ will be the
-%identity function, since if $\stg \neq \stg'$, there does not exist
-%any $\stg_1$ which satisfies the constraint for
-%$\R_l(\stl,\stg,\stg')$.
+%  \R_e(\stl,\stg,\stg') & \Leftrightarrow & \exists \stg_1.  R(\stg, \stg') \wedge \I_e(\stl, \stg_1, \stg) \wedge \I_e(\stl, \stg_1, \stg')
+%\end{array}
+%\end{smathpar}
+%\noindent While $R$ describes changes in the global database state, $\I_e$
+%constrains the global effects visible to the transaction.
+%Recall however that $\I_e$ uses the global state in effect at the
+%point the transaction last took a step ($\stg_1$) to determine whether
+%the transaction can proceed from the current global state ($\stg$).
+%Moreover, the global rely relation $R$ defines the permissible
+%interferences from $\Delta$ to a future global state ($\Delta'$).  The
+%local rely relation $\R_e$ determines allowed interferences as
+%specified by $\I_e$ assuming $\stg_1$.  In effect, the existential on
+%$\stg_1$ defines this rely relation to be an over-approximation on
+%the interferences in the presence of which the transaction can
+%proceed, simplifying verification burden.
 
@@
 binary assertions; they relate local and global database states. The
-local judgment rules require one or both of them to be stable. The
-stability of a binary assertion $Q$ w.r.t a ternary rely relation $\R$
-is defined as:
-\begin{smathpar}
-\begin{array}{c}
-\forall \stl,\stg,\stg'.~ Q(\stl,\stg) \conj \R(\stl,\stg,\stg')
-\Rightarrow Q(\stl,\stg')
-\end{array}
-\end{smathpar}
-That is, if $Q$ relates $\stl$ to $\stg$, and an interference allowed
-by the isolation specification (which implicitly considers the local
-state $\stl$) takes $\stg$ to $\stg'$, then $Q$ must also relate $\stl$
-to $\stg'$.
+local judgment rules require one or both of them to be stable with respect to the constrained rely relation $\R_e$.
+
+%The
+%stability of a binary assertion $Q$ w.r.t a ternary rely relation $\R$
+%is defined as:
+%\begin{smathpar}
+%\begin{array}{c}
+%\forall \stl,\stg,\stg'.~ Q(\stl,\stg) \conj \R(\stl,\stg,\stg')
+%\Rightarrow Q(\stl,\stg')
+%\end{array}
+%\end{smathpar}
+%That is, if $Q$ relates $\stl$ to $\stg$, and an interference allowed
+%by the isolation specification (which implicitly considers the local
+%state $\stl$) takes $\stg$ to $\stg'$, then $Q$ must also relate $\stl$
+%to $\stg'$.
 
@@
 
-A characteristic of RG reasoning is that stability of an assertion is
-always proven w.r.t to $R$, and not $R^{*}$, although interference may
-include multiple environment steps, and $R$ only captures a single
-step. This is nonetheless sound due to inductive reasoning: if
-$Q$ is preserved by every step of $R$, then $Q$ is preserved by
-$R^{*}$, and vice-versa.  However, such reasoning does not extend
-naturally to isolation-constrained interference because $R^{*}$ modulo
-$\I$ is not same as $\R^{*}$; the former is a transitive relation
-constrained by $\I$, whereas the latter is the transitive closure of an
-$\I$-constrained relation. We therefore introduce a side-condition on
-$\I$ that restores the equality. The condition requires $\I$ to allow
-an interference $R^{*}(\stg,\stg'')$, for two database states $\stg$
-and $\stg''$, only if it also allows interference for every prefix of
-$R^{*}(\stg,\stg'')$. In other words, if $\I$ disallows interference
-from $\stg$ to $\stg'$, then an $R$-step from $\stg'$ to $\stg''$
-should not make the interference from $\stg$ to $\stg''$ valid.  This
-stability condition on $\I$ is defined formally thus:
-\begin{smathpar}
-\begin{array}{lcl}
-  \stable(R,\I) & \Leftrightarrow & \forall \stl,\stg,\stg',\stg''.~
-  \neg\I(\stl,\stg,\stg') \conj R(\stg',\stg'') \Rightarrow
-  \neg\I(\stl,\stg,\stg'')
-\end{array}
-\end{smathpar}
-It can be easily verified that the above stability condition is
-satisfied by the isolation axioms from Sec.~\ref{sec:isolation}. For
-instance, $\I_{ss}$, the snapshot axiom, is stable because if
-$\I_{ss}$ is invalid ($\neg\I_{ss}$), then an interference has already
-modified a record, and no further interference will restore the
-original record, because the original record bears the id of a
-transaction that has already committed.
+%A characteristic of RG reasoning is that stability of an assertion is
+%always proven w.r.t to $R$, and not $R^{*}$, although interference may
+%include multiple environment steps, and $R$ only captures a single
+%step. This is nonetheless sound due to inductive reasoning: if
+%$Q$ is preserved by every step of $R$, then $Q$ is preserved by
+%$R^{*}$, and vice-versa.  However, such reasoning does not extend
+%naturally to isolation-constrained interference because $R^{*}$ modulo
+%$\I$ is not same as $\R^{*}$; the former is a transitive relation
+%constrained by $\I$, whereas the latter is the transitive closure of an
+%$\I$-constrained relation. We therefore introduce a side-condition on
+%$\I$ that restores the equality. The condition requires $\I$ to allow
+%an interference $R^{*}(\stg,\stg'')$, for two database states $\stg$
+%and $\stg''$, only if it also allows interference for every prefix of
+%$R^{*}(\stg,\stg'')$. In other words, if $\I$ disallows interference
+%from $\stg$ to $\stg'$, then an $R$-step from $\stg'$ to $\stg''$
+%should not make the interference from $\stg$ to $\stg''$ valid.  This
+%stability condition on $\I$ is defined formally thus:
+%\begin{smathpar}
+%\begin{array}{lcl}
+%  \stable(R,\I) & \Leftrightarrow & \forall \stl,\stg,\stg',\stg''.~
+%  \neg\I(\stl,\stg,\stg') \conj R(\stg',\stg'') \Rightarrow
+%  \neg\I(\stl,\stg,\stg'')
+%\end{array}
+%\end{smathpar}
+%It can be easily verified that the above stability condition is
+%satisfied by the isolation axioms from Sec.~\ref{sec:isolation}. For
+%instance, $\I_{ss}$, the snapshot axiom, is stable because if
+%$\I_{ss}$ is invalid ($\neg\I_{ss}$), then an interference has already
+%modified a record, and no further interference will restore the
+%original record, because the original record bears the id of a
+%transaction that has already committed.
 
@@
 
+\begin{proof}
+Here, we provide only a brief proof sketch.
+The most important rule is the top-level rule \rulelabel{RG-Txn}, which proves that a transaction $c$ which begins its execution in global database state satisfying $I$ and suffering interference $R$ while executing under isolation specification $\I$ finishes its execution in a database state also satisfying $I$, and also guarantees that its commit step satisfies $G$.
+The rule uses the transaction-local RG judgement $\R_e \vdash \rg{P}{c}{Q}$.
+By \rulelabel{E-Txn-Start}, the local and global database states at the start of a transaction satisfy $P$, and the only challenge is that environment steps in an execution covered by $\R_e \vdash \rg{P}{c}{Q}$ are in $\R_e$, while the top-level judgement requires environment steps in $R$.
+We show that it is enough to consider only those enviroment steps in $\R_e$.
+First, we use an inductive argument and stability of $\I_e$ ($\stable(R, \I_e)$) to show that any execution in which the transaction does not get stuck and completes all its steps must always preserve the isolation specification $\I_e$ after every environment step.
+Intuitively, this is because once $\I_e$ gets broken after some environment step, it will continue to remain broken and the transaction would get stuck.
+Then, we show that $\R$ contains exactly those environment steps which preserve $\I$.
+Hence, the local-level RG judgement can be soundly used, which guarantees that after the transaction finishes its execution, its local state $\stl$ and global state $\stg$ will satisfy the invariant $Q$.
+Environment steps between the last step of the transaction and its commit step can modify the global state, and hence we also require $Q$ to be stable againt $R$.
+Again, we use an inductive argument, the stability of $\I_c$, and the fact that the transaction does not get stuck and must execute its commit step to show that all environment steps must preserve $\I_c$, and hence it is enough to require $\stable(\R_c, Q)$.
+$Q$ guarantees that the commit step is in $G$, and $G$ in turn guarantees that after execution, the global database state will obey the invariant $I$.
+
+The local RG-judgement rules are mostly straightforward and directly follow from the operational semantics.
+At this point, the transaction can only change its local database state by adding new records to it, and hence all the rules have the form : $P(\stl, \stg)$ and $\stl'$ is the new state should satisfy $Q(\stl', \stg)$.
+An important thing to note is that the rules for Update, Select, Delete and Insert only require the pre-condition $P$ to be stable against interference.
+This is because the local RG-judgement rules only make assertions about the database state just after the transaction takes its last step.
+For the aforementioned commands, the last step would be the only step taken by the transaction, and hence there would be no environment steps which could affect $Q$ (the environment steps between the last execution step of the transaction and the commit step would be controlled by $\I_c$ and would be taken care of by $\stable(\R_c, Q)$ at the top level).
+
+The rule $\rulelabel{RG-Foreach}$ makes use of an inductive loop invariant $\psi$ which assumes that the variable $y$ is bound to records which have been processed in previous iterations and $z$ is bound to the record processed in the current iteration.
+Since the operational semantics ensure that every iteration processes a unique record, the proof is straightforward.
+\end{proof}
+
 \noindent Thus, if $\rg{I,R}{c}{G,I}$ is a valid RG judgment, then (a)

### 2017-10-18 supplement.tex
Edits
Minor Changes in RG-rules

@@
   P(\stl,\stg) \Leftrightarrow \stl=\emptyset \wedge I(\stg)\\
-  \R_e(\stl,\stg,\stg') \Leftrightarrow \exists \stg_1. \I_e(\stl, \stg_1, \stg) \wedge R(\stg, \stg') \wedge \I_e(\stl, \stg_1, \stg') \spc
+  \R_e(\stl,\stg,\stg') \Leftrightarrow \exists \stg_1. R(\stg, \stg') \wedge \I_e(\stl, \stg_1, \stg) \wedge \I_e(\stl, \stg_1, \stg') \spc
   \R_e \vdash \rg{P}{c}{Q}\\
-  \R_c(\stl,\stg,\stg') \Leftrightarrow \exists \stg_1. \I_c(\stl, \stg_1, \stg) \wedge R(\stg, \stg') \wedge \I_c(\stl, \stg_1, \stg')  \spc
+  \R_c(\stl,\stg,\stg') \Leftrightarrow \exists \stg_1. R(\stg, \stg') \wedge \I_c(\stl, \stg_1, \stg) \wedge \I_c(\stl, \stg_1, \stg')  \spc
   \stable(\R_c,Q)\\
@@
   P(\stl,\stg) \Leftrightarrow \stl=\emptyset \wedge I(\stg) & HP\\ 
-  \R_e(\stl,\stg,\stg') \Leftrightarrow \exists \stg_1. \I_e(\stl, \stg_1, \stg) \wedge R(\stg, \stg') \wedge \I_e(\stl, \stg_1, \stg') & H\R_l \\
+  \R_e(\stl,\stg,\stg') \Leftrightarrow \exists \stg_1. R(\stg, \stg') \wedge \I_e(\stl, \stg_1, \stg) \wedge  \I_e(\stl, \stg_1, \stg') & H\R_l \\
   \R_e\vdash \rg{P}{c}{Q} & Hc\\
-  \R_c(\stl,\stg,\stg') \Leftrightarrow \exists \stg_1. \I_c(\stl, \stg_1, \stg) \wedge R(\stg, \stg') \wedge \I_c(\stl, \stg_1, \stg') & H\R_c \\
+  \R_c(\stl,\stg,\stg') \Leftrightarrow \exists \stg_1. R(\stg, \stg') \wedge \I_c(\stl, \stg_1, \stg) \wedge  \I_c(\stl, \stg_1, \stg') & H\R_c \\
   \stable(\R_c,Q) & HQ \\
@@
 Assume that $(t_1, \stg_a) \rightarrow_R (t_1, \stg_b)$ is an environment step such that $R(\stg_a, \stg_b)$.
-Then, we know that $\I_e(t_1.\stl, t_1.\stg, \stg_a)$ and $\I_e(t_1.\stl, t_1.\stg, \stg_b)$.
+Then, we know that $\I_e(t_1.\stl, t_1.\stg, \stg_b)$ and $\I_e(t_1.\stl, t_1.\stg, \stg_a)$.
 Hence, $t_1.\stg$ provides the existence of $\stg_1$ in the definition of $\R_e$.
@@
 Consider the step $(t^{'}, \stg_{2k}) \rightarrow_R (t^{'}, \stg_{2(k+1)})$.
-We have $\I_c(t^{'}.\stl, \stg_2, \stg_{2k})$ and $\I_c(t^{'}.\stl, \stg_2, \stg_{2(k+1)})$.
+We have $\I_c(t^{'}.\stl, \stg_2, \stg_{2(k+1)})$ and $\I_c(t^{'}.\stl, \stg_2, \stg_{2k})$.
 Hence, $\stg_2$ provides the existence of $\stg_1$ in the definition of $\R_c$.
@@
 
+\begin{theorem}
+If 
+$$
+\forall \stl, \stg, \stg'. \R_e(\stl,\stg,\stg')  \Leftrightarrow  \exists \stg_1.  R(\stg, \stg') \wedge \I_e(\stl, \stg_1, \stg')
+$$
+and
+$$
+  \forall \stl,\stg,\stg',\stg''.~
+  \neg\I_e(\stl,\stg,\stg') \conj R(\stg',\stg'') \Rightarrow
+  \neg\I_e(\stl,\stg,\stg'')
+$$
+then $\forall \stl, \stg, \stg'.\exists \stg_1. R^{*}(\stg, \stg') \wedge \I_e(\stl, \stg_1, \stg') \Rightarrow \R_e^{*}(\stl, \stg, \stg')$
+\end{theorem}
+\begin{proof}
+Consider $\stl, \stg, \stg', \stg_1$ and $n$ such that $R^{n-1}(\stg, \stg')$ and $\I_e(\stl, \stg_1, \stg')$.
+Then $R(\stg'_1, \stg'_2)$, $R(\stg'_2, \stg'_3), \ldots, R(\stg'_{n-1}, \stg'_{n})$, where $\stg'_1 = \stg$ and $\stg'_n = \stg'$.
+We will show that $\R_e^{n-1}(\stl, \stg, \stg')$.
+Clearly, $\R_e(\stl, \stg'_{n-1}, \stg'_{n})$.
+Also, $\I_e(\stl, \stg_1, \stg'_{n-1})$.
+Because otherwise, if $\neg \I_e(\stl, \stg_1, \stg'_{n-1})$, then with $R(\stg'_{n-1}, \stg'_n)$ would imply $\neg \I_e(\stl, \stg_1, \stg'_{n})$ which contradicts our assumption that $\I_e(\stl, \stg_1, \stg')$.
+Hence, by definition $\R_e(\stl, \stg'_{n-2}, \stg'_{n-1})$.
+In this manner, we can show that $\forall i, 1 \leq i \leq n-1.
+\R_e(\stl, \stg'_{i}, \stg'_{i+1})$, because $\I_e(\stl, \stg_1, \stg'_{i+1})$.
+Hence, $\R_e^{n-1}(\stl, \stg, \stg')$.
+\end{proof}
+
+\begin{theorem}
+If $\I_e = \I_{ss}$, then $\forall \stl, \stg, \stg'. \R_e(\stl, \stg, \stg') \Rightarrow \stg = \stg'$ 
+\end{theorem}
+\begin{proof}
+First, we show that 
+$$
+  \forall \stl,\stg,\stg',\stg''.~
+  \neg\I_e(\stl,\stg,\stg') \conj R(\stg',\stg'') \Rightarrow
+  \neg\I_e(\stl,\stg,\stg'')
+$$
+Given $\stl, \stg, \stg'$ such that $\neg \I_e(\stl, \stg, \stg')$, $\stg \neq \stg'$.
+Since $R(\stg', \stg'')$ corresponds to the commit of a transaction, either the transaction is read-only, in which case $\stg' = \stg''$ and hence $\stg \neq \stg''$ which implies $\neg \I_e(\stl, \stg, \stg'')$, or the transaction modifies/inserts a record, in which case it will also add its own unique transaction id to the record, so that $\stg \neq \stg''$, which again implies the result.
+
+Now, consider $\stl, \stg, \stg'$ such that $\R_e(\stl, \stg, \stg')$.By definition of $\R_e$, there exists $\stg_1$ such that $\I_e(\stl, \stg_1, \stg')$.
+Hence, $\stg_1 = \stg'$.
+Now, $\I_e(\stl, \stg_1, \stg)$, because otherwise, if $\neg \I_e(\stl, \stg_1, \stg)$, then by the earlier result, $\neg \I_e((\stl, \stg_1, \stg')$ which is a contradiction.
+Hence, $\stg_1 = \stg$.
+This implies that $\stg = \stg'$.
+\end{proof}
+
+\begin{theorem}
+If $\I_e = \I_{ww}$, then $\forall \stl, \stg, \stg'. \R_e(\stl, \stg, \stg') \Rightarrow \stg = \stg'$ 
+\end{theorem}
+
+\input{inference-rules}
+\newpage
+\input{inference-proof}
+
 % We recommend abbrvnat bibliography style.

### 2017-10-18 introduction.tex
Added missing file
edits upto Sec. 5

@@
 
-The tension between serializable transactions, which are easy to
-reason about but difficult to implement efficiently, and more
-pragmatic variants that are driven by performance and availability
-considerations, has motivated the development of weaker forms of
-transaction isolation, beginning as early as 1976~\cite{gray1976};
-these instantiations permit a transaction to witness various effects
-of newly committed, or even concurrently running, transactions while
-it executes.  The ANSI SQL 92 standard defines three such weak
+Because serializable transactions favor correctness over performance,
+there has been long-standing interest~\cite{gray1976} in the database
+community to consider weaker variants that try to recover performance,
+even at the expense of simplicity and ease of reasoning.  These
+instantiations permit a transaction to witness various effects of
+newly committed, or even concurrently running, transactions while it
+executes, thus weakening serializability's strong isolation
+guarantees.  The ANSI SQL 92 standard defines three such weak
 isolation levels which are now implemented in many relational and
@@
 database. Taken together, these studies make clear that
-weakly-isolated transactions are quite prevalent in practice, whereas
+weakly-isolated transactions are quite prevalent in practice, and
 serializable transactions are often eschewed.
@@
 been echoed by leading database researchers who lament the lack of a
-  better understanding of this problem.  See: {\tt
+better understanding of this problem; see e.g., {\tt
   http://www.bailis.org/blog/understanding-weak-isolation-is-a-serious-problem}.}
 
-A major problem with weak isolation is that its semantics in the
-context of user programs is not precisely understood. The original
-proposal~\cite{gray1976} defines multiple ``degrees'' of weak
-isolation in terms of implementation details such as the nature and
-duration of locks held in each case. The ANSI SQL 92 standard defines
-four levels of isolation (including serializability) in terms of
-various undesirable \emph{phenomena} (\eg \emph{dirty reads}) each is
-required to prevent. While this is an improvement, it requires
-programmers to be prescient about the possible ways various
-undesirable phenomena might manifest in their applications, and in
-each case determine if the phenomenon can be allowed without violating
-application invariants. This is understandably hard, especially in the
-absence of any formal underpinning to define weak isolation semantics.
-Adya~\cite{adyaphd} presents the first formal definitions of some
-well-known isolation levels in the context of a sequentially
-consistent (SC) database.  However, there has been little progress
-relating Adya's system model to a formal operational semantics or a
-proof system that can facilitate rigorous correctness arguments.
-Consequently, reasoning about weak isolation remains an error prone
-endeavor, with major database vendors~\cite{postgresiso, mysqliso,
-  oracleiso} continuing to document their isolation levels primarily
-in terms of the undesirable phenomena a particular isolation level may
-induce, placing the burden on the programmer to determine application
-correctness.
+A major problem with weak isolation as currently specified is that its
+semantics in the context of user programs is not easily
+understood. The original proposal~\cite{gray1976} defines multiple
+``degrees'' of weak isolation in terms of implementation details such
+as the nature and duration of locks held in each case. The ANSI SQL 92
+standard defines four levels of isolation (including serializability)
+in terms of various undesirable \emph{phenomena} (\eg \emph{dirty
+  reads} - reading data written by an uncommitted transaction) each is
+required to prevent. While this is an improvement, this style of
+definition still requires programmers to be prescient about the
+possible ways various undesirable phenomena might manifest in their
+applications, and in each case determine if the phenomenon can be
+allowed without violating application invariants. This is
+understandably hard, especially in the absence of any formal
+underpinning to define weak isolation semantics.  Adya~\cite{adyaphd}
+presents the first formal definitions of some well-known isolation
+levels in the context of a sequentially consistent (SC) database.
+However, there has been little progress relating Adya's system model
+to a formal operational semantics or a proof system that can
+facilitate rigorous correctness arguments.  Consequently, reasoning
+about weak isolation remains an error prone endeavor, with major
+database vendors~\cite{postgresiso, mysqliso, oracleiso} continuing to
+document their isolation levels primarily in terms of the undesirable
+phenomena a particular isolation level may induce, placing the burden
+on the programmer to determine application correctness.
 
@@
 any sequential order. In contrast, weak isolation is a property of
-concurrent transactions interfering with one another resulting in an
+concurrent transactions interfering with one another, resulting in an
 execution order that is not serializable. Unlike weak consistency,
@@
 In this paper, we propose a program logic for weakly-isolated
-transactions along with automated verification support that realizes
-this goal.  We develop a set of syntax-directed compositional proof
-rules that allow developers to build correctness proofs for
+transactions along with automated verification support to allow
+developers to verify the soundness of their applications, without
+having to resort to low-level operational reasoning as they are forced
+to do currently.  We develop a set of syntax-directed compositional
+proof rules that enable the construction of correctness proofs for
 transactional programs in the presence of a weakly-isolated
 concurrency control mechanism.  Realizing that the proof burden
-imposed by these rules would likely discourage applications
-programmers from using them, we present an inference procedure based
-that automatically verifies the weakest isolation level for a
-transaction that ensures its consistency requirements are maintained.
-The key to inference is a novel formulation of database state
-(represented as sets of tuples) as a monad, and in which database
-computations are interpreted as state transformers over these sets.
-This interpretation leads to an encoding of database computations
-amenable for verification by off-the-shelf SMT solvers.  The paper
-makes the following contributions:
+imposed by these rules may discourage applications programmers from
+using them, we also present an inference procedure based on these
+rules that automatically verifies the weakest isolation level for a
+transaction that still ensures its invariants are maintained.  The key
+to inference is a novel formulation of database state (represented as
+sets of tuples) as a monad, and in which database computations are
+interpreted as state transformers over these sets.  This
+interpretation leads to an encoding of database computations amenable
+for verification by off-the-shelf SMT solvers.  The paper makes the
+following contributions:
 %% We have realized these ideas as an embedded DSL within OCaml, that
@@
     as monadic computations over an abstract database state.
- \item We develop an operational semantics and compositional proof
-   system for this language capable of relating high-level application
-   invariants to database state, parameterized by a weak isolation
-   semantics that selectively exposes the visibility of these
-   operations to other transactions.
+ \item We develop an operational semantics and a compositional
+   rely-guarantee style proof system for this language capable of
+   relating high-level application invariants to database state,
+   parameterized by a weak isolation semantics that selectively
+   exposes the visibility of these operations to other transactions.
  \item Exploiting the monadic definition of database state, we devise
@@
    approach and the utility of our proof methodology and inference
-   mechanism..
+   mechanism.
 \end{enumerate}

### 2017-10-18 isolation.tex
edits upto Sec. 5

@@
 than the invariance of already read records. In particular, ANSI SQL
-RR allows \emph{phantom reads}, a phenonmenon in which a repeated
+RR allows \emph{phantom reads}, a phenomenon in which a repeated
 \C{SELECT} query might return newly inserted records that were not
@@
 but without changing their transaction identifiers ($\txnf$ fields),
-and flush ($\gg$) should only flush the records that bear the current
+and flush ($\rhd$) should only flush the records that bear the current
 transaction's identifier. With this extension, ANSI SQL RR can be

### 2017-10-18 reasoning.tex
edits upto Sec. 5

@@
   underlying system.}.  Our proof system is essentially an adaptation
-of a rely-guarantee reasoning framework~\cite{rgjones} to the setting of
-weakly isolated database transactions.  The primary challenge in the
-formulation deals with how we relate a transaction's isolation
+of a rely-guarantee reasoning framework~\cite{rgjones} to the setting
+of weakly isolated database transactions.  The primary challenge in
+the formulation deals with how we relate a transaction's isolation
 specification ($\I$) to its rely relation ($R$) that describes the
@@
 the transaction setting that affects the structure of the proof system
-is atomicity; a transaction's writes are not visible to its peers
+is atomicity; we do not permit a transaction's writes to be visible
 until it commits.  In the context of rely-guarantee, this means that
@@
 \C{atomic} blocks also have the same characteristic, the fact that
-transactions are weakly isolated introduces surprising complexity.  Unlike an
-\C{atomic} block, the effect of a transaction is \emph{not} a
-sequential composition of the effects of its statements because each
+transactions are weakly-isolated introduces surprising complexity.
+Unlike an \C{atomic} block, the effect of a transaction is \emph{not}
+a sequential composition of the effects of its statements because each
 statement can witness a potentially different version of the state.
@@
 \vdash \hoare{P}c{Q}$).  Recall that the standard RG judgment is the
-quintuple $\rg{P,R}{c}{G,Q}$. Instead of a separate $P$ and $Q$, our
-top-level judgment uses $I$ for both pre- and post-conditions, because
-our focus is on verifying that a \txnimp\ program \emph{preserves} a
-databases' consistency conditions\footnote{The terms \emph{consistency
-    condition}, \emph{high-level invariant}, and \emph{integrity
-    constraint} are used interchangeably throughout the paper.}.  A
-transaction-local RG judgment does not include a guarantee relation
-because transaction-local effects are not visible outside a
-transaction. Also, the rely relation ($\R$) of the transaction-local
-judgment is not the same as the top-level rely relation ($R$) because
-it must take into account the transaction's isolation specification
-($\I$). Intuitively, $\R$ is $R$ modulo $\I$.  Recall that a
-transaction writes to its local database ($\stl$), which is then
-flushed when the transaction commits. Thus, the guarantee of a
-transaction depends on the state of its local database at the commit
-point. The pre- and post-condition assertions ($P$ and $Q$) in the
-local judgment facilitate tracking the changes to the
-transaction-local state, which eventually helps us prove the validity
-of the transaction's guarantee.  Both $P$ and $Q$ are bi-state
-assertions; they relate transaction-local database state ($\stl$) to
-the global database state ($\stg$). Thus, the transaction-local
-judgment effectively tracks how transaction-local and global states
-change in relation to each other.
+quintuple $\rg{P,R}{c}{G,Q}$. Instead of separate $P$ and $Q$
+assertions, our top-level judgment uses $I$ as both a pre- and
+post-condition, because our focus is on verifying that a
+\txnimp\ program \emph{preserves} a databases' consistency
+conditions\footnote{The terms \emph{consistency condition},
+  \emph{high-level invariant}, and \emph{integrity constraint} are
+  used interchangeably throughout the paper.}.  A transaction-local RG
+judgment does not include a guarantee relation because
+transaction-local effects are not visible outside a transaction. Also,
+the rely relation ($\R$) of the transaction-local judgment is not the
+same as the top-level rely relation ($R$) because it must take into
+account the transaction's isolation specification ($\I$). Intuitively,
+$\R$ is $R$ modulo $\I$.  Recall that a transaction writes to its
+local database ($\stl$), which is then flushed when the transaction
+commits. Thus, the guarantee of a transaction depends on the state of
+its local database at the commit point. The pre- and post-condition
+assertions ($P$ and $Q$) in the local judgment facilitate tracking the
+changes to the transaction-local state, which eventually helps us
+prove the validity of the transaction's guarantee.  Both $P$ and $Q$
+are bi-state assertions; they relate transaction-local database state
+($\stl$) to the global database state ($\stg$). Thus, the
+transaction-local judgment effectively tracks how transaction-local
+and global states change in relation to each other.
+
+\subsubsection{Stability}
 
@@
 condition that requires the validity of an assertion $\phi$ to be
-unaffected by the interference, i.e., the rely relation $R$. In
-conventional RG, stability is defined as follows, where $\sigma$ denotes a
-state:
+unaffected by interference from other concurrently executing
+transactions, i.e., the rely relation $R$. In conventional RG,
+stability is defined as follows, where $\sigma$ and $\sigma'$ denote
+states:
 \begin{smathpar}
@@
 convey the same intuition as above.
-In our setting, we only need to prove the stability of an assertion ($Q$) against those environment steps which lead to a global database state on which the transaction itself can take its next step according to its isolation specification ($\I$).
+In our setting, we only need to prove the stability of an assertion ($\phi$) against those environment steps which lead to a global database state on which the transaction itself can take its next step according to its isolation specification ($\I$).
 
@@
 \begin{array}{lcl}
-\stable(R, Q) & \Leftrightarrow & \forall \stl, \stg, \stg'. Q(\stl, \stg) \wedge R^{*}(\stg, \stg') \wedge \I(\stl, \stg, \stg') \Rightarrow Q(\stl, \stg')
+\stable(R, \phi) & \Leftrightarrow & \forall \stl, \stg, \stg'. \phi(\stl, \stg) \wedge R^{*}(\stg, \stg') \wedge \I(\stl, \stg, \stg') \Rightarrow \phi(\stl, \stg')
 \end{array}
@@
 
-The above condition uses the transitive closure of $R$, while the standard RG stability assertion only uses $R$ and then relies on an inductive argument to prove stability for the transitive closure.
-We cannot directly replace $R^{*}$ by $R$ in the above condition because the inductive argument would not hold, as $R^{*}$ modulo $\I$ is not the same as ($R$ modulo $\I$$)^{*}$.To obtain an equivalent form in our setting, we require an additional condition on the isolation specification, which we call the stability condition on $\I$.
-The condition requires $\I$ to allow
-an interference $R^{*}(\stg,\stg'')$, for two database states $\stg$
-and $\stg''$, only if it also allows interference for every prefix of
-$R^{*}(\stg,\stg'')$. In other words, if $\I$ disallows interference
-from $\stg$ to $\stg'$, then an $R$-step from $\stg'$ to $\stg''$
-should not make the interference from $\stg$ to $\stg''$ valid.  This
-stability condition on $\I$ is defined formally thus:
+A characteristic of RG reasoning is that stability of an assertion is
+always proven w.r.t to $R$, and not $R^{*}$, although interference may
+include multiple environment steps, and $R$ only captures a single
+step. This is nonetheless sound due to inductive reasoning: if $\phi$
+is preserved by every step of $R$, then $\phi$ is preserved by
+$R^{*}$, and vice-versa.  However, such reasoning does not extend
+naturally to isolation-constrained interference because $R^{*}$ modulo
+$\I$ is not same as $\R^{*}$; the former is a transitive relation
+constrained by $\I$, whereas the latter is the transitive closure of
+an This means, unfortunately, that we cannot directly replace $R^{*}$
+by $R$ in the above condition because of our need to consider the weak
+isolation specification $\I$ that selectively exposes interferences.
+
+To obtain an equivalent form in our setting, we require an additional
+condition on the isolation specification, which we call the
+\emph{stability condition on $\I$}.  The condition requires $\I$ to
+allow interferences $R^{*}(\stg,\stg'')$, for two database states
+$\stg$ and $\stg''$, only if it also allows interference for every
+prefix of $R^{*}(\stg,\stg'')$. In other words, if $\I$ disallows
+interference from $\stg$ to $\stg'$, then an $R$-step from $\stg'$ to
+$\stg''$ should not make the interference from $\stg$ to $\stg''$
+valid.  This stability condition on $\I$ is defined formally thus:
 \begin{smathpar}
@@
 original record, because the original record bears the id of a
-transaction that has already committed.
-We can now generalize the stability condition for any arbitrary assertion $Q$ by only considering those environment steps which in addition to preserving $Q$, also preserve the condition that the transaction can continue to take its next step, i.e.
-the condition $\I$ holds on both the beginning and the end states.
+transaction that has already committed. We can now generalize the
+stability condition for any arbitrary assertion $\phi$ by only
+considering those environment steps which in addition to preserving
+$\phi$, also preserve the condition that the transaction can continue
+to take its next step, i.e. the condition $\I$ holds on both the
+beginning and the end states.
 
@@
 
-In the above condition, we use a separate notation ($\stg_1$) for the global database state that the transaction witnessed during its last step.
-$Q$ does not constrain this global database state, but the stability of $\I$ ensures that any database state constrained by $Q$ must also obey $\I$.
-We can now replace the transitive closure in the above condition with the single step relation $R$, backed by the inductive argument for stability.
-Pushing the universal quantifier for $\stg_1$ inside the antecedent and separating the predicates involving $\stg_1$, we can rewrite the condition in the form below, by defining a new ternary relation $\R$, which can be thought of as $R$ modulo $\I$:%stating that an assertion $Q$ on a database state $\stg$ on which the transaction can take its next step (hence the condition $\I(\stl, \stg_1, \stg)$) is stable w.r.t.
-environment steps to $\stg'$
+In the above condition, we use a separate notation ($\stg_1$) for the
+global database state that the transaction witnessed during its last
+step. $\phi$ does not constrain this global database state, but the
+stability of $\I$ ensures that any database state constrained by
+$\phi$ must also obey $\I$. We can now replace the transitive closure
+in the above condition with the single step relation $R$, backed by
+the inductive argument for stability. Pushing the universal quantifier
+for $\stg_1$ inside the antecedent and separating the predicates
+involving $\stg_1$, we can rewrite the above condition into the form
+shown below, by defining a new ternary relation $\R$, which can be
+thought of as $R$ modulo $\I$:
 
@@
 \begin{array}{lcl}
-\stable((R \backslash \I), Q) & \Leftrightarrow & \forall \stl,\stg,\stg'. Q(\stl,\stg) \wedge (R \backslash \I)(\stl,\stg,\stg') \Rightarrow Q(\stl,\stg')\\
-(R \backslash \I)(\stl, \stg, \stg') & \Leftrightarrow & \exists \stg_1. R(\stg, \stg') \wedge \I(\stl, \stg_1, \stg) \wedge \I(\stl, \stg_1, \stg')
+\stable(\R, Q) & \Leftrightarrow & \forall \stl,\stg,\stg'. Q(\stl,\stg) \wedge \R(\stl,\stg,\stg') \Rightarrow Q(\stl,\stg')\\
+\R(\stl, \stg, \stg') & \Leftrightarrow & \exists \stg_1. R(\stg, \stg') \wedge \I(\stl, \stg_1, \stg) \wedge \I(\stl, \stg_1, \stg')
 \end{array}
@@
 
-The above condition often significantly simplifies the form of $\R$ irrespective
-of $R$. For example, when a transaction is executed against a
-snapshot of the database (i.e. $\mathbb{I}_{ss}$), $\R$ will be the
-identity function, since if $\stg \neq \stg'$, there does not exist
-any $\stg_1$ which satisfies the constraint for
-$\R_l(\stl,\stg,\stg')$.
+The above condition often significantly simplifies the form of $\R$
+irrespective of $R$. For example, when a transaction is executed
+against a snapshot of the database (i.e. $\mathbb{I}_{ss}$), $\R$ will
+be the identity function, since if $\stg \neq \stg'$, there does not
+exist any $\stg_1$ which satisfies the constraint for
+$\R_e(\stl,\stg,\stg')$.
+
+\subsubsection{Rules}
 
@@
 the required isolation as specified by $\I$. It depends on a
-transaction-local judgment to verify the body ($c$) of a transaction with
-id $i$. The precondition $P$ of $c$ must follow from the fact that the
-transaction-local database ($\stl$) is initially empty, and the global
-database satisfies the high-level invariant $I$. The rely relation
-($\R_e$) is obtained from the global rely relation $R$ and the isolation specification $\I_e$ as explained above.
-Note that $\I_e$ constrains the global effects visible to the transaction while it is executing but has not yet committed.
+transaction-local judgment to verify the body ($c$) of a transaction
+with id $i$. The precondition $P$ of $c$ must follow from the fact
+that the transaction-local database ($\stl$) is initially empty, and
+the global database satisfies the high-level invariant $I$. The rely
+relation ($\R_e$) is obtained from the global rely relation $R$ and the
+isolation specification $\I_e$ as explained above. Recall that $\I_e$
+constrains the global effects visible to the transaction while it is
+executing but has not yet committed.
 
@@
 binary assertions; they relate local and global database states. The
-local judgment rules require one or both of them to be stable with respect to the constrained rely relation $\R_e$.
+local judgment rules require one or both of them to be stable with
+respect to the constrained rely relation $\R_e$.
 
@@
 \begin{array}{c}
-  \forall \stl,\stg.~ Q(\stl,\stg) \Rightarrow G(\stg, \stl \gg \stg)\spc
+  \forall \stl,\stg.~ Q(\stl,\stg) \Rightarrow G(\stg, \stl \rhd \stg)\spc
 \end{array}
@@
 and after the commit. The act of commit is captured by the flush
-operation ($\stl\gg\stg$). Once we establish the validity of $G$ as a
+operation ($\stl\rhd\stg$). Once we establish the validity of $G$ as a
 faithful representative of the transaction, we can verify that the
@@
 \begin{array}{lclr}
-\R \vdash \hoare{P}{c}{Q} & \defeq & \forall \stl,\stl',\stg,\stg',
-  \bar{v}.~ P(\stl,\stg) \conj (\tbox{c}_i,\stl,\stg) \rstepssto{n}
+\R \vdash \hoare{P}{c}{Q} & \defeq & \forall \stl,\stl',\stg,\stg'.~ P(\stl,\stg) \conj (\tbox{c}_i,\stl,\stg) \rstepssto{n}
   (\tbox{\cskip}_i, \stl',\stg')
-  \Rightarrow Q(\stl',\stg') & \bar{v} = {\sf freevars}(c)\\
+  \Rightarrow Q(\stl',\stg') \\
 \rg{I,R}{c}{G,I} & \defeq &  \forall \stg.\, I(\stg)
@@
 \begin{array}{lcl}
-\texttt{step-guaranteed}(R,G,c,\stg) &\defeq& \forall \stg',c'',\stg''.
+\texttt{step-guaranteed}(R,G,c,\stg) &\defeq& \forall c',c''\stg',\stg''.
 (c,\stg) \rstepssto{n} (c',\stg') \conj (c',\stg') \stepsto
@@
 
-\begin{proof}
-Here, we provide only a brief proof sketch.
-The most important rule is the top-level rule \rulelabel{RG-Txn}, which proves that a transaction $c$ which begins its execution in global database state satisfying $I$ and suffering interference $R$ while executing under isolation specification $\I$ finishes its execution in a database state also satisfying $I$, and also guarantees that its commit step satisfies $G$.
-The rule uses the transaction-local RG judgement $\R_e \vdash \rg{P}{c}{Q}$.
-By \rulelabel{E-Txn-Start}, the local and global database states at the start of a transaction satisfy $P$, and the only challenge is that environment steps in an execution covered by $\R_e \vdash \rg{P}{c}{Q}$ are in $\R_e$, while the top-level judgement requires environment steps in $R$.
-We show that it is enough to consider only those enviroment steps in $\R_e$.
-First, we use an inductive argument and stability of $\I_e$ ($\stable(R, \I_e)$) to show that any execution in which the transaction does not get stuck and completes all its steps must always preserve the isolation specification $\I_e$ after every environment step.
-Intuitively, this is because once $\I_e$ gets broken after some environment step, it will continue to remain broken and the transaction would get stuck.
-Then, we show that $\R$ contains exactly those environment steps which preserve $\I$.
-Hence, the local-level RG judgement can be soundly used, which guarantees that after the transaction finishes its execution, its local state $\stl$ and global state $\stg$ will satisfy the invariant $Q$.
-Environment steps between the last step of the transaction and its commit step can modify the global state, and hence we also require $Q$ to be stable againt $R$.
-Again, we use an inductive argument, the stability of $\I_c$, and the fact that the transaction does not get stuck and must execute its commit step to show that all environment steps must preserve $\I_c$, and hence it is enough to require $\stable(\R_c, Q)$.
-$Q$ guarantees that the commit step is in $G$, and $G$ in turn guarantees that after execution, the global database state will obey the invariant $I$.
-
-The local RG-judgement rules are mostly straightforward and directly follow from the operational semantics.
-At this point, the transaction can only change its local database state by adding new records to it, and hence all the rules have the form : $P(\stl, \stg)$ and $\stl'$ is the new state should satisfy $Q(\stl', \stg)$.
-An important thing to note is that the rules for Update, Select, Delete and Insert only require the pre-condition $P$ to be stable against interference.
-This is because the local RG-judgement rules only make assertions about the database state just after the transaction takes its last step.
-For the aforementioned commands, the last step would be the only step taken by the transaction, and hence there would be no environment steps which could affect $Q$ (the environment steps between the last execution step of the transaction and the commit step would be controlled by $\I_c$ and would be taken care of by $\stable(\R_c, Q)$ at the top level).
-
-The rule $\rulelabel{RG-Foreach}$ makes use of an inductive loop invariant $\psi$ which assumes that the variable $y$ is bound to records which have been processed in previous iterations and $z$ is bound to the record processed in the current iteration.
-Since the operational semantics ensure that every iteration processes a unique record, the proof is straightforward.
-\end{proof}
-
 \noindent Thus, if $\rg{I,R}{c}{G,I}$ is a valid RG judgment, then (a)

### 2017-10-18 abstract.tex
edits upto Sec. 5

@@
   implement transactions using weaker alternatives. These alternatives
-  break the strong isolation guarantees offered by serializable
-  transactions to permit greater concurrency. Unfortunately, the
-  semantics of weak isolation is poorly understood, and usually
-  explained only informally in terms of low-level implementation
+  break the strong isolation guarantees offered by serializablity to
+  permit greater concurrency. Unfortunately, the semantics of weak
+  isolation is poorly understood, and usually explained only
+  informally in terms of low-level implementation
   artifacts. Consequently, verifying high-level correctness properties
@@
   ascertains the weakest isolation level that still guarantees the
-  safety of high-level consistency assertions associated with such
+  safety of high-level consistency invariants associated with such
   transactions.  The key to effective inference is the observation
@@
 
-  Notably, our development is parametric over a transaction's specific
+  Our development is parametric over a transaction's specific
   isolation semantics, allowing it to be applicable over a range of
-  concurrency control mechanisms.  Case studies and experiments on
+  weak isolation mechanisms.  Case studies and experiments on
   real-world applications (written in an embedded DSL in OCaml)

### 2017-10-18 motivation.tex
edits upto Sec. 5

--- 2017-07-08 motivation.tex
@@
   dist <- SQL.select1 District (fun d -> d.d_id = d_id);
-  let o_id = dist.d_next_o_id;
+  let o_id = dist.d_next_o_id in
       SQL.update(* UPDATE *) District 
@@
                     then stk.s_qty - item_req.ol_qty 
-                else stk.s_qty - item_req.ol_qty + 91;
+                    else stk.s_qty - item_req.ol_qty + 91 in
             SQL.update Stock (fun s -> {s with s_qty = s_qty'}) 
@@
 \C{ol\_o\_id} field is equal to \C{o\_id}) is guaranteed to be equal
-to the size of the \C{item\_req} set, which in turn is equal to the
+to the size of the \C{item\_reqs} set, which in turn is equal to the
 \C{Order} record's \C{o\_ol\_cnt} field; these constraints ensure that
@@
   executed.  The subsequent operations are effectively sequentialized,
-  since $T_2$ commits. Nonetheless, both transactions read the same value for
+  since $T_2$ commits before $T_1$. Nonetheless, both transactions read the same value for
   \C{d\_next\_o\_id} resulting in them  adding \C{Order} records
@@
 default isolation level in 8 of the 18 databases studied
-in~\cite{bailishotos}.  An RC transaction is isolated from \emph{dirty
-  writes}, i.e., writes of uncommitted transactions, but is allowed to
-witness the writes of concurrent transactions as soon as they are
-committed. Thus, with two concurrent instances of the \C{new\_order}
-transaction (call them $T_1$ and $T_2$), both concurrently placing new
-orders for different customers under the same district (\C{d\_id}), RC
-isolation allows the execution shown in Fig.~\ref{fig:new_order_exec}.
+in~\cite{bailishotos}.  An executing RC transaction is isolated from
+\emph{dirty writes}, i.e., writes of uncommitted transactions, but is
+allowed to witness the writes of concurrent transactions as soon as
+they are committed. Thus, with two concurrent instances of the
+\C{new\_order} transaction (call them $T_1$ and $T_2$), both
+concurrently placing new orders for different customers under the same
+district (\C{d\_id}), RC isolation allows the execution shown in
+Fig.~\ref{fig:new_order_exec}.
 
@@
 
-Notably, this example does not exhibit any of the anomalies that
+Clearly, this example does not exhibit any of the anomalies that
 characterize RC isolation~\cite{berenson}. For instance, there are no
 \emph{lost writes} - both concurrent transactions' writes are present
-in the final state of the database. Indeed, program analyses that aim
-to determine appropriate isolation by checking for possible
+in the final state of the database.  Program analyses that aim to
+determine appropriate isolation by checking for possible
 manifestations of RC-induced anomalies would fail to identify grounds
@@
 traces~\cite{adyaphd,gotsmanconcur15} on memory read and write actions
-can complicate high-level reasoning.  A possible alternative would
-interleave weak isolation implementation details within the program,
-yielding a (more-or-less) conventional concurrent program that can be
-then subject to classical concurrent verification methods.
-Considering the size and complexity of real-world transaction systems,
-this strategy is unlikely to scale.
+can complicate high-level reasoning.  A possible alternative would be
+to utilize concurrent program verfication methods where the
+implementation details. of weak isolation are interleaved within the
+program, yielding a (more-or-less) conventional concurrent program;
+but, considering the size and complexity of real-world transaction
+systems, this strategy is unlikely to scale.
 
@@
 database, thus admitting concurrency, but it also requires that there
-not be any write-write conflicts (i.e,, such a conflict occurs if
-concurrentlyh executing transactions modify the same record) among
+not be any write-write conflicts (i.e., such a conflict occurs if
+concurrently executing transactions modify the same record) among
 concurrent transactions when they commit. Write-write conflicts can be
@@
 
-As this discussion indicates, implementations of SI on real databases
+As this discussion hints, implementations of SI on real databases
 such as PostgresSQL are highly complicated, often running into
@@
 captured in a fairly simple model. First, effects induced by one
-transaction are not visible to another concurrently executing one
-(call it $T$), during $T$'s execution. Thus, from $T$'s perspective,
+transaction (call it $T$) are not visible to another concurrently executing one
+during $T$'s execution. Thus, from $T$'s perspective,
 the global state does not change during its execution.  More formally,
-for every operation performed by $T$, the global state $T$ witness
+for every operation performed by $T$, the global state $T$ witnesses
 before ($\stg$) and after ($\stg')$ executing the operation is the
@@
 with $T$. To concretize this notion, let the database state be a map
-from transaction variables to values, and let $\stl$ denote a
-transaction-local log that maps the variables being written to their
+from database locations to values, and let $\stl$ denote a
+transaction-local log that maps the locations being written to their
 updated values.  The absence of write-write conflicts between $T$ and
@@
   \item After $T$ finishes execution, but before it commits its local
-    state $\delta$, $\forall(x\in\delta).~\Delta'(x) = \Delta(x)$.
+    state $\delta$, $\forall(x\in\mathit{dom}(delta)).~\Delta'(x) = \Delta(x)$.
 \end{itemize}
@@
 to the records ($\stl$) written by \C{new\_order}, i.e.,
-$\forall(x\in\delta).~\Delta'(x) = \Delta(x)$.  The axiomatization
+$\forall(x\in\mathit{dom}(\delta)).~\Delta'(x) = \Delta(x)$.  The axiomatization
 precludes any execution in which there are concurrent updates to
@@
 
-We approach the verification problem by first observing that a
+We approach the verification problem by first observing that a typical
 relational database is a significantly simpler abstraction than shared
-memory. There are no pointers, linked data structures, or aliasing.
-Although a database essentially abstracts a mutable state, this state
-is affected through a well-defined fixed number of interfaces (SQL
-statements), each tagged with a logical formula describing what
-records are accessed and updated.
+memory.  Its primary data structure is a table, with no primitive
+support for pointers, linked data structures, or aliasing.  Although a
+database essentially abstracts a mutable state, this state is managed
+through a well-defined fixed number of interfaces (SQL statements),
+each tagged with a logical formula describing what records are
+accessed and updated.
 
@@
 SQL statement in the set domain, modeling the database as a set of
-tuples, and a SQL statement as a function over this set.  Among other
+records, and a SQL statement as a function over this set.  Among other
 things, one benefit of this approach is that low-level loops can now
@@
 from Fig.~\ref{fig:new_order_code}. Some irrelevant expressions have
-been replaced with constants (\C{k1} to \C{k3}).  The body of the loop
-executes a SQL update followed by an insert.  Recall that a
+been replaced with constants (\C{k1},\C{k2}, and \C{k3}).  The body of
+the loop executes a SQL update followed by an insert.  Recall that a
 transaction reads from the global database ($\Delta$), and writes to a
 transaction-local database ($\delta$) before committing these
-updates. An update statement filters the tuples that match the search
-criteria from $\Delta$ and computes the updated tuples that are to be
-added to the local database. Thus, the state transformer for the update
-statement (call it $\T_U$) is the following function on
+updates. An update statement filters the records that match the search
+criteria from $\Delta$ and computes the updated records that are to be
+added to the local database. Thus, the state transformer for the
+update statement (call it $\T_U$) is the following function on
 sets\footnote{$\bind$ has higher precedence than $\cup$.}:
@@
         s.\C{s\_i\_id} = \C{item\_req.ol\_i\_id}\\\hspace*{1.3in}}
-           {\{\{s \with \C{s\_qty} = \C{k1}\}\}\\\hspace*{1.3in}}
-           {\emptyset}
+           {\{ \langle s\ {\sf with}\  s_\C{qty} = \C{k1}\rangle \}\\\hspace*{1.3in}}
+           {\emptyset})
 \end{array}
@@
 % F_U(\Delta)$.
-The transformer ($\T_I(\Delta,\delta)$) for the subsequent \C{insert}
+Here, the set bind operator extracts record elements ($\C{s}$) from
+the database, checks the precondition of the update action, and if
+satisfied, constructs a new set containing a single record that is
+identical to $\C{s}$ except that it binds field $\C{s_{qty}}$ to value
+$\C{k_1}$.  This new set is added (via set union) to the existing
+local database state $\delta$.\footnote{For now, assume that the
+  record being added to the local database has not previously been
+  updated.}
+
+The transformer ($\T_I(\delta,\Delta)$) for the subsequent \C{insert}
 statement can be similarly constructed:
@@
 \begin{array}{l}
-  \lambda(\stl,\stg).~ \stl \cup \{\C{ol\_o\_id}=\C{k2};\, 
-        \C{ol\_d\_id}=\C{k3};\, \C{ol\_i\_id}=\C{item\_req.ol\_i\_id};\, 
-        \C{ol\_qty}=\C{item\_req.ol\_qty}\}
+  \lambda(\stl,\stg).~ \stl \cup
+       \{\langle\C{ol\_o\_id}=\C{k2},
+                 \C{ol\_d\_id}=\C{k3},\, \C{ol\_i\_id}=\C{item\_req.ol\_i\_id},\, 
+                 \C{ol\_qty}=\C{item\_req.ol\_qty}\rangle\}
+
 \end{array}
 \end{smathpar}
-Observe that both transformers are of the form $T(\Delta,\delta) =
-\stl \cup F(\stg)$, where $F$ is a function that returns the set of
-records added by the transformer to the transaction-local database
+Observe that both transformers are of the form $T(\delta,\Delta) =
+\stl \cup F(\stl,\stg)$, where $F$ is a function that returns the set
+of records added  to the transaction-local database
 ($\stl$). Let $\F_U$ and $\F_I$ be the corresponding functions for
@@
 \begin{array}{l}
-  \lambda(\stl,\stg).~ \stl \cup \F_U(\stg) \cup \F_I(\stg)
+  \lambda(\stl,\stg).~ \stl \cup \F_U(\stl,\stg) \cup \F_I(\stl,\stg)
 \end{array}
@@
   \lambda(\stl,\stg).~ \stl \cup \C{item\_reqs}\bind
-      (\lambda\C{item\_req}.~ \F_U(\stg) \cup \F_I(\stg))
+      (\lambda\C{item\_req}.~ \F_U(\stl,\stg) \cup \F_I(\stl,\stg))
 \end{array}
@@
 operations, and the \C{DB} monad's $\bind$ becoms set monadic $\bind$.
-The advantage of inferring such transformers is that we can now make
-use of a semantics-preserving translation from the domain of sets
-equipped with $\bind$ to a decidable fragment of first-order logic,
-allowing us to leverage SMT solvers for automated proofs without
-having to infer potentially complex thread-local invariants or
-intermediate assertions.  Sec.~\ref{sec:inference} describes this
-translation. In the exposition thus far, we assumed $\Delta$ remains
-invariant, which is clearly not the case when we admit concurrency.
-Necessary concurrency extensions of the state transformer semantics to
-deal with interference is also covered in Sec.~\ref{sec:inference}.
-The following two sections, however, first focus on laying the
-theoretical foundations for reasoning about weak isolation.
+As we demonstrate, the advantage of inferring such transformers is
+that we can now make use of a semantics-preserving translation from
+the domain of sets equipped with $\bind$ to a decidable fragment of
+first-order logic, allowing us to leverage SMT solvers for automated
+proofs without having to infer potentially complex thread-local
+invariants or intermediate assertions.  Sec.~\ref{sec:inference}
+describes this translation. In the exposition thus far, we assumed
+$\Delta$ remains invariant, which is clearly not the case when we
+admit concurrency.  Necessary concurrency extensions of the state
+transformer semantics to deal with interference is also covered in
+Sec.~\ref{sec:inference}.  Before presenting the transformer
+semantics, we first focus our attention in the following two sections
+on the theoretical foundations for weak isolation, upon which this
+semantics is based.

### 2017-10-18 opsem.tex
edits upto Sec. 5

@@
 records, arithmetic and boolean expressions ($e_1 \odot e_2$), and
-record expressions ($\{\bar{f}=\bar{e}\}$) constitute the syntactic
+record expressions ($[\bar{f}=\bar{e}]$) constitute the syntactic
 class of expressions ($e$). Commands ($c$) include $\cskip$,
 conditional statements, \C{LET} constructs to bind names, \C{FOREACH}
-loops, SQL statements, their sequential composition ($c_1;c_2$), and
+loops, SQL statements, their sequential composition ($c_1;c_2$), 
 transactions ($\ctxn{i}{\I}{c}$) and their parallel composition
@@
 identifier $i$, and executes at the top-level; our semantics does not
-supported nested transactions. The $\I$ in the \C{TXN} block syntax is
+support nested transactions. The $\I$ in the \C{TXN} block syntax is
 the transaction's isolation specification, whose purpose is explained
@@
 they all belong to a single table.  The generalization to multiple
-tables is straightforward. The semantics assumes that  records in
-$\stg$ can be uniquely identified via their $\idf$ field, and
+tables is straightforward, e.g., by having the machine manipulate a
+set of sets, one for each table.  The semantics assumes that records
+in $\stg$ can be uniquely identified via their $\idf$ field, and
 enforces this property wherever necessary. Certain hidden fields are
-treated specially by the operational semantics, and they are hidden
+treated specially by the operational semantics, and are hidden
 from the surface language. These include a $\txnf$ field that tracks
@@
 database, committing these heretofore uncommitted records. The flush
-operation ($\gg$) is defined as follows:
+operation ($\rhd$) is defined as follows:
 \begin{smathpar}
 \begin{array}{c}
-\forall r.~ r \in (\stl\gg\stg) ~\Leftrightarrow~ 
+\forall r.~ r \in (\stl\rhd\stg) ~\Leftrightarrow~ 
   (r.\idf \notin \dom(\stl) \conj r \in \stg)
@@
 \end{smathpar}
-Let $\stg' = \stl\gg\stg$. A record $r$ belongs to $\stg'$ iff it
+Let $\stg' = \stl\rhd\stg$. A record $r$ belongs to $\stg'$ iff it
 belongs to $\stg$ and has not been updated in $\stl$, i.e., $r.\idf
@@
 simplifies the semantics, without losing any generality, since
-substituting $\stl\gg\stg$ for $\stg$ at select places in the
+substituting $\stl\rhd\stg$ for $\stg$ at select places in the
 reduction rules effectively allows reads of uncommitted transaction
@@
 transactions, and their parallel composition. A transaction comes
-tagged with an isolation specification $\I$, which has two components
+tagged with an \emph{isolation specification} $\I$, which has two components
 $\I_e$ and $\I_c$, that dictate the timing and nature of interferences
@@
 \end{smathpar}
+This definition dictates that no change to the global database state
+can be visible to an SI transaction while it executes ($\I_e$), and
+there should be no concurrent updates to records written by the
+transaction by other concurrently executing ones ($\I_c$).
 To simplify the presentation, we use $\I$ instead of $\I_e$ and $\I_c$
@@
 specification ($\I_e$) allows the interference between $\stg$ and
-$\stg'$. Rule \rulelabel{E-Commit} commits the transaction
-$\ctxnr{i}{\I,\stl,\stg}{c}$ by flushing its uncommitted records to
-the database. This is done only if the interference between $\stg$ and
-$\stg'$ is allowed at the commit point by the isolation specification
-($\I_c$).  The distinction between $\I_e$ and $\I_c$ allows us to
-model the snapshot semantics of realistic isolation levels that
-isolate a transaction from interference during its execution, but
-expose interferences at the commit point.
+$\stg'$.   Rule \rulelabel{E-Commit} commits the
+transaction $\ctxnr{i}{\I,\stl,\stg}{c}$ by flushing its uncommitted
+records to the database. This is done only if the interference between
+$\stg$ and $\stg'$ is allowed at the commit point by the isolation
+specification ($\I_c$).  The distinction between $\I_e$ and $\I_c$
+allows us to model the snapshot semantics of realistic isolation
+levels that isolate a transaction from interference during its
+execution, but expose interferences at the commit point.
 

### 2017-10-18 reasoning.tex
edits upto Sec. 5

@@
  \begin{smathpar}
-\forall \stl, \stg, \stg', \stg_1. Q(\stl, \stg) \wedge \I(\stl, \stg_1, \stg) \wedge R^{*}(\stg, \stg') \wedge \I(\stl, \stg_1, \stg') \Rightarrow Q(\stl, \stg')
+   \forall \stl, \stg, \stg', \stg_1. \phi(\stl, \stg) \wedge \I(\stl, \stg_1, \stg) \wedge R^{*}(\stg, \stg') \wedge \I(\stl, \stg_1, \stg')
+   \Rightarrow \phi(\stl, \stg')
 \end{smathpar}
@@
 step. $\phi$ does not constrain this global database state, but the
-stability of $\I$ ensures that any database state constrained by
-$\phi$ must also obey $\I$. We can now replace the transitive closure
-in the above condition with the single step relation $R$, backed by
-the inductive argument for stability. Pushing the universal quantifier
-for $\stg_1$ inside the antecedent and separating the predicates
-involving $\stg_1$, we can rewrite the above condition into the form
-shown below, by defining a new ternary relation $\R$, which can be
-thought of as $R$ modulo $\I$:
+stability of $\I$ ensures that any database state constrained by $\phi$
+must also obey $\I$. We can now replace the transitive closure in the
+above condition with the single step relation $R$, backed by the
+inductive argument for stability. Pushing the universal quantifier for
+$\stg_1$ inside the antecedent and separating the predicates involving
+$\stg_1$, we can rewrite the condition in the form below, by defining
+a new ternary relation $R \backslash \I$, which can be thought of as
+$R$ modulo $\I$:
 
@@
 \begin{array}{lcl}
-\stable(\R, Q) & \Leftrightarrow & \forall \stl,\stg,\stg'. Q(\stl,\stg) \wedge \R(\stl,\stg,\stg') \Rightarrow Q(\stl,\stg')\\
-\R(\stl, \stg, \stg') & \Leftrightarrow & \exists \stg_1. R(\stg, \stg') \wedge \I(\stl, \stg_1, \stg) \wedge \I(\stl, \stg_1, \stg')
+\stable((R \backslash \I), \phi) & \Leftrightarrow & \forall \stl,\stg,\stg'. \phi(\stl,\stg) \wedge (R \backslash \I)(\stl,\stg,\stg') \Rightarrow \phi(\stl,\stg')\\
+(R \backslash \I)(\stl, \stg, \stg') & \Leftrightarrow & \exists \stg_1. R(\stg, \stg') \wedge \I(\stl, \stg_1, \stg) \wedge \I(\stl, \stg_1, \stg')
 \end{array}
@@
 
-The above condition often significantly simplifies the form of $\R$
-irrespective of $R$. For example, when a transaction is executed
-against a snapshot of the database (i.e. $\mathbb{I}_{ss}$), $\R$ will
-be the identity function, since if $\stg \neq \stg'$, there does not
-exist any $\stg_1$ which satisfies the constraint for
-$\R_e(\stl,\stg,\stg')$.
+The above condition often significantly simplifies the form of $R
+\backslash \I$ irrespective of $R$. For example, when a transaction is
+executed against a snapshot of the database (i.e. $\mathbb{I}_{ss}$),
+$R \backslash \I_{ss}$ will be the identity function, since if $\stg
+\neq \stg'$, there does not exist any $\stg_1$ which satisfies the
+constraint for $(R \backslash \I_{ss})(\stl,\stg,\stg')$.
 

### 2017-10-18 inference.tex
Encoding algorithm.

--- 2017-07-08 inference.tex
@@
 (\C{pc}) to the balance of a bank account, which is required to be
-non-negatives ($I(\stg) \Leftrightarrow \forall(r\in\stg).~r.\C{bal}\ge
+non-negative ($I(\stg) \Leftrightarrow \forall(r\in\stg).~r.\C{bal}\ge
 0$). The transaction starts by issuing a \C{select1} query, whose
 effect implicitly asserts that there exists a record in the bank
-account database that is equal to $a$
-($\exists(r\in\stg).~r=a$). However, this assertion is unstable
-because a concurrent \C{withdraw} or \C{deposit} transactions might
-update the account balance, so that such a record no longer exists in
-the database.  Fortunately, the weakening operator
-($\stabilize{\cdot}$) allows us to weaken the assertion to $\exists
-\stg'.I(\stg') \conj \exists(r\in\stg').~a=r$ (Fig.~\ref{fig:logic}
-formalizes the encoding of {\sf exists} to logic), which is enough to
-prove that $\C{a.bal + pc*a.bal}\ge 0$, and verify that
-\C{add\_interest} preserves the positive balance invariant.
+account database that is equal to $a$ ($\exists(r\in\stg).~r=a$).
+However, this assertion is unstable because a concurrent \C{withdraw}
+or \C{deposit} transactions might update the account balance, so that
+such a record no longer exists in the database.  Fortunately, the
+weakening operator ($\stabilize{\cdot}$) allows us to weaken the
+assertion to $\exists \stg'.I(\stg') \conj \exists(r\in\stg').~a=r$
+({\sf exists} translates to $\exists$ in first-order logic, as
+formalized in Fig.~\ref{fig:logic}), which is enough to prove that
+$\C{a.bal + pc*a.bal}\ge 0$, and verify that \C{add\_interest}
+preserves the positive balance invariant.
 
@@
 \noindent where the set expression $\F(\emptyset,\stg)$ has a natural
-interpretation as discussed below.
+interpretation in the theory of sets.
 % Where, ${\sf true} = \lambda(\stl,\stg).~true$ and $\stable(\R,I)
@@
 of the \rulelabel{RG-Txn} rule (Fig.~\ref{fig:rg-rules}) by a state
-transformer inference judgment. The soundness of a transaction's guarantee can now
-be established w.r.t the effect $\F$ of the body. The
-\rulelabel{RG-Txn} rule so updated is shown
-below:
+transformer inference judgment. The soundness of a transaction's
+guarantee can now be established w.r.t the effect $\F$ of the body.
+The \rulelabel{RG-Txn} rule so updated is shown below:
 \begin{smathpar}
@@
   \stable(R,\I)\spc
-  \stable(R,I)\\
-   \R_e(\stl,\stg,\stg') \Leftrightarrow \exists \stg_1.  R(\stg, \stg') \wedge \I_e(\stl, \stg_1, \stg') \spc  c \elabsto \F\\
+  \stable(R,I)\spc
+  \R_e(\stl,\stg,\stg') \Leftrightarrow \exists \stg_1.  R(\stg, \stg') \wedge \I_e(\stl, \stg_1, \stg') \spc  c \Longrightarrow_{\langle i,\R_e,I \rangle}\F\\
   \R_c(\stl,\stg,\stg') \Leftrightarrow \exists \stg_1.  R(\stg, \stg') \wedge \I_c(\stl, \stg_1, \stg') \spc \stable(\R_c,\F)\\
-  \forall \stg.~ G(\stg, F(\emptyset,\stg) \gg \stg)\spc
+  \forall \stl,\stg.~ \stl \in F(\emptyset,\stg) \Rightarrow 
+    G(\stg, \stl\gg\stg)\spc
   \forall \stg,\stg'.~I(\stg) \wedge G(\stg,\stg') \Rightarrow I(\stg')\\

### 2017-10-18 inference.tex

@@
 
-\input{logic}
-
 \subsection{Soundness of Inference}
@@
 interpretation in the theory of sets.
-% Where, ${\sf true} = \lambda(\stl,\stg).~true$ and $\stable(\R,I)
-% \Leftrightarrow \forall\stl,\stg,\stg'. I(\stg) \conj
-% \R(\stl,\stg,\stg') \Rightarrow \I(\stg')$.
+
+\begin{proof}
+  We only provide a brief proof sketch here. We prove a slightly
+  stronger statement :
+
+  \emph{Forall} i,$R$,$I$,$c$,$\F$,$s$ if $\stable(\R,I)$ and $c
+  \elabsto \F$, then:\\
+
+  \begin{smathpar}
+    \begin{array}{c}
+    \R \vdash \hoare{\lambda(\stl,\stg).~\stl=s \conj
+    I(\stg)}{c}{\lambda(\stl,\stg).\stl = \F(s,\stg)}
+    \end{array}
+  \end{smathpar}
+
+  We use structural induction on $c$ to prove the statement. Assume
+  that $P \Leftrightarrow \lambda(\stl,\stg).~\stl=s \conj I(\stg)$
+  and $Q \Leftrightarrow \lambda(\stl,\stg).\stl = \F(s,\stg)$. For
+  each type of $c$, our strategy is to show the premises of the
+  corresponding RG-rule. The base cases correspond to the INSERT,
+  UPDATE and DELETE statements. In all these cases, the form of $F$
+  directly corresponds to the condition linking the pre-condition $P$
+  and post-condition $Q$ in the premises of the RG-rules. If
+  $\stable(\R, F)$, then $\stabilize{F} = F$ and hence the proof
+  directly follows. Otherwise, $\stabilize{F} = \lambda (\stl,
+  \stg).\existsl(\stg', I(\stg'), F(\stl, \stg'))$. Since the
+  pre-condition $P(\stl, \stg)$ asserts $I(\stg)$, any state which
+  obeys $P$ provides the existential $\stg'$ in the above set
+  expression, and hence $F$ applied on that state matches the
+  condition linking $P$ and $Q$. The stability of $P$ with respect to
+  $\R$ follows directly from the stability of $I$.
+
+  The proofs for SELECT, SEQ and IF-THE-ELSE directly use the inductive
+  hypothesis to infer the RG-judgements present in the premises of
+  their corresponding RG-rules. The interesting case is the FOREACH
+  statement, for which we use the loop invariant $\psi(\stl, \stg)
+  \Leftrightarrow \stl = s \cup y \bind (\lambda z.~\F(s, \stg))$,
+  (where assuming that $c$ is the body of the loop, $c \elabsto F$) to
+  prove all the premises of \rulelabel{RG-Foreach}. Using the same
+  notation as the rule $\rulelabel{RG-Foreach}$, $y$ refers to the
+  records already processed in previous iterations of the loop, while
+  $z$ refers to the record being processed in the current iteration.
+  At the beginning of the loop $[\phi/y]\psi(\stl, \stg)$ just reduces
+  to $\stl = s$ which is implied by the pre-condition $P$. From the
+  inductive hypothesis, we can infer that each iteration corresponds
+  to the application of $F$. Since all iterations are assumed to be
+  independent of each other, and $z$ is bound to a record in $x$ for
+  each iteration, we can infer that at the end of every iteration the
+  loop invariant $[y \cup \{z\}/y]\psi$ will be satisfied. 
+\end{proof}
 
 \subsection{From $\SL$ to the first-order logic}
+
+\input{logic}
 

### 2017-10-19 inference.tex
Full encoding algorithm

@@
 
-The algorithm ($\mssemof{\cdot}$) shown in Fig.~\ref{fig:logic}
+The algorithm ($\mssemof{\nubar}{\cdot}$) shown in Fig.~\ref{fig:logic}
 encodes an $\SL$ expression in a decidable fragment of FOL. Given a
-set expression $s$ of $\SL$, $\mssemof{s}$ is a pair $(\phi,f)$, where
+set expression $s$ of $\SL$, $\mssemof{\nubar}{s}$ is a pair $(\phi,f)$, where
 $\phi$ encodes $s$ in FOL, and $f$ is a Boolean function that names
 the set defined by $s$.  Naming provides an easy handle to $s$,
-simplifying $\mssemof{\cdot}$.  Set variables ($x$, $\stl$, $\stg$,
+simplifying $\mssemof{\nubar}{\cdot}$.  Set variables ($x$, $\stl$, $\stg$,
 $\dots$) are encoded as unconstrained Boolean functions ($g_x$,

### 2017-10-19 inference.tex
Notational changes
Also added proof sketch of theorems in Section 4 and 5

@@
 The correspondence between the inference rules and the RG judgment of
-\S~\ref{sec:reasoning} is stated thus\footnote{The proof can be found in
-the supplementary material.}:
+\S~\ref{sec:reasoning} is stated thus:
 \begin{theorem}
@@
 \begin{proof}
-  We only provide a brief proof sketch here. We prove a slightly
+  We only provide a brief proof sketch here\footnote{The complete proof can be found in
+the supplementary material.}. We prove a slightly
   stronger statement :
@@
   \emph{Forall} i,$R$,$I$,$c$,$\F$,$s$ if $\stable(\R,I)$ and $c
-  \elabsto \F$, then:\\
+  \elabsto \F$, then:
 
@@
 
-  We use structural induction on $c$ to prove the statement. Assume
+  We use structural induction on $c$. Assume
   that $P \Leftrightarrow \lambda(\stl,\stg).~\stl=s \conj I(\stg)$
@@
 
-  The proofs for SELECT, SEQ and IF-THE-ELSE directly use the inductive
+  The proofs for SELECT, SEQ and IF-THEN-ELSE directly use the inductive
   hypothesis to infer the RG-judgements present in the premises of
@@
   statement, for which we use the loop invariant $\psi(\stl, \stg)
-  \Leftrightarrow \stl = s \cup y \bind (\lambda z.~\F(s, \stg))$,
+  \Leftrightarrow \stl = s \cup (y \bind (\lambda z.~\F(s, \stg)))$,
   (where assuming that $c$ is the body of the loop, $c \elabsto F$) to
@@
   \stable(R,I)\spc
-  \R_e(\stl,\stg,\stg') \Leftrightarrow \exists \stg_1.  R(\stg, \stg') \wedge \I_e(\stl, \stg_1, \stg') \spc  c \Longrightarrow_{\langle i,\R_e,I \rangle}\F\\
-  \R_c(\stl,\stg,\stg') \Leftrightarrow \exists \stg_1.  R(\stg, \stg') \wedge \I_c(\stl, \stg_1, \stg') \spc \stable(\R_c,\F)\\
-  \forall \stl,\stg.~ \stl \in F(\emptyset,\stg) \Rightarrow 
-    G(\stg, \stl\gg\stg)\spc
+  P(\stl,\stg) \Leftrightarrow \stl=\emptyset \wedge I(\stg)\\
+  \R_e = R \backslash \I_e \spc \R_c = R \backslash \I_c \spc 
+   \R_e \vdash \rg{P}{c}{Q} \spc \stable(\R_c,Q) \\ 
+  \forall \stl,\stg.~ Q(\stl,\stg) \Rightarrow 
+    G(\stg, \stl \rhd \stg)\spc
   \forall \stg,\stg'.~I(\stg) \wedge G(\stg,\stg') \Rightarrow I(\stg')\\

### 2017-10-19 reasoning.tex
Notational changes
Also added proof sketch of theorems in Section 4 and 5

@@
 constrained by $\I$, whereas the latter is the transitive closure of
-an This means, unfortunately, that we cannot directly replace $R^{*}$
+a constrained relation. This means, unfortunately, that we cannot directly replace $R^{*}$
 by $R$ in the above condition because of our need to consider the weak
@@
 
+\noindent Thus, if $\rg{I,R}{c}{G,I}$ is a valid RG judgment, then (a)
+every interleaved multi-step reduction of $c$ preserves the database
+integrity constraint (consistency condition) $I$, and (b) the effect
+that every transaction in $c$ has on the database state is captured by
+$G$.
 \noindent We can now assert the soundness of the RG judgments in
-Fig.~\ref{fig:rg-rules} as follows\footnote{Its proof is provided in
-  the supplementary material.}:
+Fig.~\ref{fig:rg-rules} as follows:
 
@@
 
-\noindent Thus, if $\rg{I,R}{c}{G,I}$ is a valid RG judgment, then (a)
-every interleaved multi-step reduction of $c$ preserves the database
-integrity constraint (consistency condition) $I$, and (b) the effect
-that every transaction in $c$ has on the database state is captured by
-$G$.
+\begin{proof}
+Here, we provide only a brief proof sketch\footnote{The complete proof is provided in the supplementary material.}.
+The most important rule is the top-level rule \rulelabel{RG-Txn}, which proves that a transaction $c$ which begins its execution in global database state satisfying $I$ and suffering interference $R$ while executing under isolation specification $\I$ finishes its execution in a database state also satisfying $I$, and also guarantees that its commit step satisfies $G$.
+The rule uses the transaction-local RG judgement $\R_e \vdash \rg{P}{c}{Q}$.
+By \rulelabel{E-Txn-Start}, the local and global database states at the start of a transaction satisfy $P$, and the only challenge is that environment steps in an execution covered by $\R_e \vdash \rg{P}{c}{Q}$ are in $\R_e$, while the top-level judgement requires environment steps in $R$.
+We show that it is enough to consider only those enviroment steps in $\R_e$.
+First, we use an inductive argument and stability of $\I_e$ ($\stable(R, \I_e)$) to show that any execution in which the transaction completes all its steps must always preserve the isolation specification $\I_e$ after every environment step.
+Intuitively, this is because once $\I_e$ gets broken after some environment step, it will continue to remain broken and the transaction would not be able to proceed (according to \rulelabel{E-Txn}).
+Since $\R_e$ contains exactly those environment steps which preserve $\I_e$, the local-level RG judgement can be soundly used, which guarantees that after the transaction finishes its execution, its local state $\stl$ and global state $\stg$ will satisfy the assertion $Q$.
+Environment steps between the last step of the transaction and its commit step can modify the global state, and hence we also require $Q$ to be stable againt $R$.
+Again, we use an inductive argument, the stability of $\I_c$, and the fact that the transaction must execute its commit step to show that all environment steps must preserve $\I_c$, and hence it is enough to require $\stable(\R_c, Q)$.
+$Q$ guarantees that the commit step is in $G$, and $G$ in turn guarantees that after execution, the global database state will obey the invariant $I$.
+
+The local RG-judgement rules are mostly straightforward and directly follow from the operational semantics.
+At this point, the transaction can only change its local database state by adding new records to it, and hence all the rules have the form : $P(\stl, \stg)$ and $\stl'$ is the new state should satisfy $Q(\stl', \stg)$.
+An important thing to note is that the rules for Update, Select, Delete and Insert only require the pre-condition $P$ to be stable against interference.
+This is because the local RG-judgement rules only make assertions about the database state just after the transaction takes its last step.
+For the aforementioned commands, the last step would be the only step taken by the transaction, and hence $Q$ would not be affected by any environment steps, but instead directly reflect the changes made by the command (the environment steps between the last execution step of the transaction and the commit step would be controlled by $\I_c$ and would be taken care of by $\stable(\R_c, Q)$ at the top level).
+
+The rule $\rulelabel{RG-Foreach}$ makes use of an inductive loop invariant $\psi$ which assumes that the variable $y$ is bound to records which have been processed in previous iterations and $z$ is bound to the record processed in the current iteration.
+Since the operational semantics ensure that every iteration processes a unique record, the proof is straightforward.
+\end{proof}

### 2017-10-19 inference.tex
Encoding explanation part 1.

--- 2017-10-18 inference.tex
@@
 transaction requires automating the multiple implication checks in
-the premise. While $R$, $\R$, $\R_c$, $\I$ and $I$ are formulas in
+the premise. While $R$, $\R$, $\I$ and $I$ are formulas in
 first-order logic (FOL) with a relatively simple structure, $\F$
@@
 
-The algorithm ($\mssemof{\cdot}$) shown in Fig.~\ref{fig:logic}
-encodes an $\SL$ expression in a decidable fragment of FOL. Given a
-set expression $s$ of $\SL$, $\mssemof{s}$ is a pair $(\phi,f)$, where
-$\phi$ encodes $s$ in FOL, and $f$ is a Boolean function that names
-the set defined by $s$.  Naming provides an easy handle to $s$,
-simplifying $\mssemof{\cdot}$.  Set variables ($x$, $\stl$, $\stg$,
-$\dots$) are encoded as unconstrained Boolean functions ($g_x$,
-$g_{\stl}$, $g_{\stg}$, $\ldots$) that are named after the variable.
-Encoding of a set literal $\{a\,|\,\varphi\}$ introduces a new Boolean
-function $g$ that is \emph{true} for only those $a$ that satisfy $\varphi$.
-Thus $g$ faithfully encodes the literal. The encoding of the
-$\existsl{}$ expression first eliminates the second-order
-quantification over sets.  Existentially bound set variable ($a$) is
-skolemized, i.e., instantiated with a new unbound variable $a'$,
-before encoding the bound expression $s$ and the first-order
-constraint $\phi$. The algorithm for encoding $\phi$, which is already
-a first-order formula, is straightforward. It basically involves
-replacing set variables with Boolean functions named after the
-variables. The first-order encoding for a bind expression describes
-the semantics of set monad's bind operator in FOL. Let $s_1$ be a set, and
-let $f$ be a function that maps each variable in $s_1$ to a new
-set. Then, $s_2 = s_1 \bind f$ if and only if for all $y\in s_2$,
-there exists an $x \in s_1$ such that $y = f(x)$, and forall $x\in
-s_1$, $f(x)\in s_2$. The encoding adds new constraints to this
-effect. Conditional set expressions and set union are encoded
-straightforwardly.
+The algorithm ($\mssemof{\cdot}{\cdot}$) shown in Fig.~\ref{fig:logic}
+translates an $\SL$ expression ($s$) to FOL. The translation is based
+on encoding a set of element type $T$ as a unary predicate on $T$.
+The predicate is represented as a meta function that accepts a
+an $x:T$ and returns a quantifier-free proposition that evaluates to
+true ($\top$) if and only if $x$ is present in the set. Alternatively,
+translation may also encode the set as a function in the logic itself,
+in which case a quantified proposition constraining the function is
+also generated. For instance, consider the set $\{1,2\}$. The
+predicate describing the set can be encoded as the function $\lambda
+\v. \v=1 \vee \v=2$, with no further constraints (i.e., with $\top$
+constraint), or it can be encoded as the function $\lambda \v. g(\v)$
+with an associated constraint ($\phi\in\Prop^1$) defining the
+uninterpreted boolean functon $g$:
+\begin{smathpar}
+  \begin{array}{lcl}
+    \phi & = & \forall \nu.~g(\nu) \Leftrightarrow \nu=1 \vee \nu=2
+  \end{array}
+\end{smathpar}
+The translation adopts one or the other approach, depending on the
+need. For uniformity, we consider the encoding of a set as pair
+($\phi$,$\G$), where $\G$ is a meta function, and $\phi$ is a FOL
+formula constraining any uninterpreted functions used in $G$.
+
+Due to the presence of bind ($\bind$) in $\SL$, a set expression $s$
+may contain free variables introduced by an enclosing binder. For
+instance, consider the $\SL$ expression $s_1\bind(\lambda x. \{ y\,|\,
+y=x+1\})$, where $s_1$ is an integer set (expression). The
+subexpression $\{ y\,|\, y=x+1\}$ (call it $s_2$) contains $x$ as a
+free variable. In such cases, the predicate associated with the
+subexpression should also be indexed by its free variables, so that a
+unique set exists for each instantiation of the free variables. For
+instance, the predicate ($\G$) associated with subexpression from the
+above example should be $\lambda\v_1.\lambda\v_2.~\v_2 = \v_1 + 1$, so
+that the set $\G\; x_1$ is different from the set $\G\; x_2$ for
+distinct $x_1,x_2 \in s_1$. Intuitively, the bind expression
+$s_1\bind(\lambda x. \{ y\,|\, y=x+1\})$ denotes the set
+$\bigcup\limits_{x\in s_1}\G\;x$.
+
+The translation algorithm (Fig.~\ref{fig:logic}) takes free variables
+into account. Given a set expression $s\in\SL$, whose (possible) free
+variables are $\nubar$ in the order of their introduction (top-most
+binder first), $\mssemof{\nubar}{s}$ returns the encoding of $s$ as
+$(\phi,\G)$.  The meta function $\G$ is a predicate indexed by the
+(possible) free variables of $s$, thus, its arity is $|\nubar| + 1$.
+Note that $\nubar$ is only a sequence of variables introduced by the
+enclosing binders of $s$, and not all may actually occur free in $s$.
+Nonetheless, its predicate $\G$ is always indexed by $|\nubar|$ free
+variables for uniformity. The translation encodes database state as an
+uninterpreted sort. Considering that the state is actually a set of
+records, we define an uninterpreted relation $\in$ to relate records
+and states. Thus, a variable set expression $\stg$ denoting a database
+state is encoded as the predicate $\lambda (\vbar,r).~r\in\stg$, where
+$|\vbar|=|\nubar|$ (predicates are uncurried for simplicity; $\vbar$
+is a comma-separated sequence; $r\not\in\SL$ is a special variable).
+The constraints associated with the encoding of a state are trivial
+($\top$). The set literal expression $\{x\,|\, \varphi\}$ is encoded
+straightforwardly. The conditional set expression is encoded as an
+if-then-else predicate in FOL, where the predicates on true and false
+branches are computed from the set subexpressions $s_1$ and $s_2$,
+respectively. The conjunction of constraints $\phi_1$ and $\phi_2$,
+from $\mssemof{\nubar}{s_1}$ and $\mssemof{\nubar}{s_2}$ (resp.), is
+propagated upwards as the constraint of the conditional expression.
+Set union expression is encoded similar.
+
+The first-order encoding of a bind expression describes the semantics
+of set monad's bind operator in FOL. Let $s_1$ be a set, and let $f =
+\lambda x.~s_2$ be a function that maps each variable in $s_1$ to a
+new set. Then, $s = s_1 \bind f$ if and only if forall $x$, $y$, if
+$x\in s_1$ and $y \in f(x)$, then  $y \in s$, and vice versa. The
+translation first encodes $s_1$ and $s_2$ to obtain predicates
+$(\phi_1,\G_1)$ and $(\phi_2,\G_2)$, respectively. Since $s_2$ is
+under a new binder that binds $x$, free variable sequence of $s_2$ is
+$\nubar,x$. For technical reasons, we substitute a fresh $a$ for $x$,
+making the sequence $\nubar,a$. The set $s$ is encoded as a a new
+uninterpreted predicate $g$ indexed by $s$'s free variables
+($\nubar$). Since the set denoted by $g$ is the result of the bind
+$s_1 \bind \lambda x.s_2$, a first-order constraint defining the bind
+operation (as described above) is generated. The constraint relates the
+predicates $\G_1$ and $\G_2$, representing $s_1$ and $s_2$ (resp.), to
+the uninterpreted predicate $g$ that represents $s$.
 

### 2017-10-19 inference.tex
Explanation added.

@@
 propagated upwards as the constraint of the conditional expression.
-Set union expression is encoded similar.
+Set union expression is encoded similarly.
 
@@
 $s_1 \bind \lambda x.s_2$, a first-order constraint defining the bind
-operation (as described above) is generated. The constraint relates the
-predicates $\G_1$ and $\G_2$, representing $s_1$ and $s_2$ (resp.), to
-the uninterpreted predicate $g$ that represents $s$.
+operation (as described above) is generated. The constraint relates
+the predicates $\G_1$ and $\G_2$, representing $s_1$ and $s_2$
+(resp.), to the uninterpreted predicate $g$ that represents $s$. The
+constraint is assigned a name ($\pi$) so that it can be enabled only
+for the choice of free variables ($\nubar$) for which $g$ needs to be
+defined; observe that the lambda returned by
+$\mssemof{\nubar}{s_1 \bind \lambda x.s_2}$ asserts $\pi(\vbar)$ only
+in conjunction with $g(\vbar,r)$, for some $r$.
+
+The first-order encoding of the $\existsl$ expression eliminates the
+existential by Skolemizing it. Skolemizing is the process of
+substituting an existentially bound $x$ in $\phi\in\Prop^1$ with
+$f(\nubar)$, where $f$ is a fresh uninterpreted function, and $\nubar$
+are the free variables in $\phi$ bound by enclosing universal
+quantifiers. Due to the decidability restrictions that will become
+clear shortly, the only uninterpreted functions we admit in our logic
+are boolean (i.e., predicates/relations). Consequently, Skolemizing an
+existentially bound $x:T$ with $f(\nubar):T$, where $T$ is a
+non-boolean type, is not really an option. Instead, we define $f$ as
+an uninterpreted relation that behaves as a partial function:
+\begin{smathpar}
+  \begin{array}{c}
+    \forall \nubar.\forall a.\forall b.~f(\nubar,a) \wedge f(\nubar,b)
+    \Rightarrow a = b
+  \end{array}
+\end{smathpar}
+We then replace the existentially bound $x$ with a new universally
+bound $a$ in $\phi$, under the condition that $f(\nubar,a)$ holds.
+Due to the partial function property of $f$, there can be utmost one
+instantiation for $a$, as desired. There can however be no
+instantiations, i.e., no $a$ that satisfies $f(\nubar,a)$, thus
+violating the exists property. Fortunately, this problem can be solved
+in our case by observing that the existential always occurs in the
+context (antecedent) in the verification conditions (implications)
+generated by the \rulelabel{RG-Txn} rule. Intuitively, we prove that a
+set of records is equal to, or contained in, other set of records
+(e.g., $\F_1 = \F_2$ or $\F \subseteq G$) \emph{assuming} that there
+exists a state $\stg$ that satisfies $I$. If there does not exist such
+a state, then our premise itself is false, and the goal must be
+vacuously provable. We make use of this observation, and let the
+context derive $\bot$ when there does not exist a desired $a$.
 

### 2017-10-20 inference.tex
Sec. 5

--- 2017-10-19 inference.tex
@@
 assertion to $\exists \stg'.I(\stg') \conj \exists(r\in\stg').~a=r$
-({\sf exists} translates to $\exists$ in first-order logic, as
-formalized in Fig.~\ref{fig:logic}), which is enough to prove that
-$\C{a.bal + pc*a.bal}\ge 0$, and verify that \C{add\_interest}
-preserves the positive balance invariant.
+({\sf exists} has the same semantics as $\exists$, as formalized in
+Fig.~\ref{fig:logic}), which is enough to prove that $\C{a.bal +
+pc*a.bal}\ge 0$, and verify that \C{add\_interest} preserves the
+positive balance invariant.
 
@@
 $x\in s_1$ and $y \in f(x)$, then  $y \in s$, and vice versa. The
-translation first encodes $s_1$ and $s_2$ to obtain predicates
-$(\phi_1,\G_1)$ and $(\phi_2,\G_2)$, respectively. Since $s_2$ is
-under a new binder that binds $x$, free variable sequence of $s_2$ is
-$\nubar,x$. For technical reasons, we substitute a fresh $a$ for $x$,
-making the sequence $\nubar,a$. The set $s$ is encoded as a a new
-uninterpreted predicate $g$ indexed by $s$'s free variables
-($\nubar$). Since the set denoted by $g$ is the result of the bind
-$s_1 \bind \lambda x.s_2$, a first-order constraint defining the bind
-operation (as described above) is generated. The constraint relates the
-predicates $\G_1$ and $\G_2$, representing $s_1$ and $s_2$ (resp.), to
-the uninterpreted predicate $g$ that represents $s$.
+translation first encodes $s_1$ and $s_2$ to obtain $(\phi_1,\G_1)$
+and $(\phi_2,\G_2)$, respectively. Since $s_2$ is under a new binder
+that binds $x$, free variable sequence of $s_2$ is $\nubar,x$. In the
+interest of hygiene, we substitute a fresh $a$ for $x$, making the
+sequence $\nubar,a$. The set $s$ is encoded as a a new uninterpreted
+predicate $g$ indexed by $s$'s free variables ($\nubar$). Since the
+set denoted by $g$ is the result of the bind $s_1 \bind \lambda
+x.s_2$, a first-order constraint defining the bind operation (as
+described above) is generated. The constraint relates the predicates
+$\G_1$ and $\G_2$, representing $s_1$ and $s_2$ (resp.), to the
+uninterpreted predicate $g$ that represents $s$. The constraint is
+assigned a name ($\pi$) so that it can be enabled only for the choice
+of free variables ($\nubar$) for which $g$ needs to be defined;
+observe that the lambda returned by $\mssemof{\nubar}{s_1 \bind
+\lambda x.s_2}$ asserts $\pi(\vbar)$ only in conjunction with
+$g(\vbar,r)$, for some $r$.
+
+The first-order encoding of the $\existsl(\stg,\phi,s)$ expression
+eliminates the existential by Skolemizing it. Skolemizing is the
+process of substituting an existentially bound $x$ in
+$\phi_x\in\Prop^1$ with $f(\nubar)$, where $f$ is a fresh
+uninterpreted function, and $\nubar$ are the free variables in
+$\phi_x$ bound by enclosing universal quantifiers. Due to the
+decidability restrictions (Sec.~\ref{sec:decidability}), the only
+uninterpreted functions we admit in our logic are boolean (i.e.,
+predicates/relations). Consequently, Skolemizing an existentially
+bound $\stg$ with $f(\nubar)$ is not really an option in our case.
+Instead, we define $f$ as an uninterpreted relation that behaves as a
+partial function:
+\begin{smathpar}
+  \begin{array}{c}
+    \forall \nubar.\forall a.\forall b.~f(\nubar,a) \wedge f(\nubar,b)
+    \Rightarrow a = b
+  \end{array}
+\end{smathpar}
+We then replace the existentially bound $\stg$ with a new universally
+bound $a$ in $\phi$ and $s$, and define the set $[a/\stg]s$ under the
+condition that $f(\nubar,a) \conj [a/\stg]\phi$ holds. Since the
+existence of a $\stg$ that satisfies $\phi$ is axiomatized tp
+define $s$, if the aforementioned condition does not hold, i.e., if
+there does exist such a $\stg$, we let the context derive $\bot$.
+
+\noindent {\bf Example} Let us reconsider the  TPC-C \C{new\_order}
+transaction from Sec.~\ref{sec:motivation}. Recall that the state
+transformer ($\T$) for the foreach loop shown in
+Fig.~\ref{fig:foreach_code} is (\C{k1}, \C{k2}, and \C{k3} are
+constants):
+\begin{smathpar}
+\begin{array}{l}
+  \lambda(\stl,\stg).~ \stl \cup \C{item\_reqs}\bind
+      (\lambda\C{item\_req}.~ \F_U(\stl,\stg) \cup \F_I(\stl,\stg))
+\end{array}
+\end{smathpar}
+Where:
+\begin{smathpar}
+  \begin{array}{lcl}
+    \F_U & = & \lambda(\stl,\stg).~ \stg \bind(\lambda s. 
+                \itel{{\sf table}(s) = \C{Stock} \conj 
+                  s.\C{s\_i\_id} = \C{item\_req.ol\_i\_id}\\
+         &   & \hspace*{0.9in}}
+                     {\{ \langle \C{s\_i\_id}=s.\C{s\_i\_id};\, 
+                                 \C{s\_d\_id}=s.\C{s\_d\_id};\,
+                                 \C{s\_qty} = \C{k1}\rangle \}\\
+         &   & \hspace*{0.9in}} {\emptyset}) \\
+    \F_I & = & \lambda(\stl,\stg).~ \{\langle\C{ol\_o\_id}=\C{k2},\,
+                 \C{ol\_d\_id}=\C{k3},\, \C{ol\_i\_id}=\C{item\_req.ol\_i\_id},\, 
+                 \C{ol\_qty}=\C{item\_req.ol\_qty}\rangle\}\\
+  \end{array}
+\end{smathpar}
+For any $\stl$ and $\stg$, $\F_U(\stl,\stg)$ and $\F_I(\stl,\stg)$ are
+expressions in $\SL$, so can be translated to FOL by the encoding
+algorithm in Fig.~\ref{fig:logic}. Since the iteration variable
+\C{item\_req} occurs free in these expressions, the appropriate
+application of the encoding algorithm is
+$\mssemof{\C{item\_req}}{\F_U(\stl,\stg)}$ and
+$\mssemof{\C{item\_req}}{\F_I(\stl,\stg)}$, which result in
+$(\phi_U,\G_U)$ and $(\phi_I,\G_I)$, respectively, where $\phi_U$,
+$\phi_I$, $\G_U$, $\G_I$ are as shown below:
+\begin{smathpar}
+  \begin{array}{lcl}
+    \phi_U & = & \forall \C{item\_req}.\forall s. \forall s'.~  \\
+    & & \hspace*{0.1in}\pi(\C{item\_req}) \Leftrightarrow(s \in \stg) \conj 
+        (\itec{{\sf table}(s)=\C{Stock} \wedge 
+              s.\C{s\_i\_id} = \C{item\_req.ol\_i\_id}\\
+    & & \hspace*{1.45in}}
+              {s' = \langle \C{s\_i\_id}=s.\C{s\_i\_id};\, 
+                            \C{s\_d\_id}=s.\C{s\_d\_id};\,
+                            \C{s\_qty} = \C{k1}\rangle\\
+    & & \hspace*{1.45in}} {\bot}) \\
+    & & \hspace*{1.2in} \Leftrightarrow g(\C{item\_req},s')\\
+    \G_U & = & \lambda (\C{item\_req},r).~ \pi(\C{item\_req}) 
+        \wedge g(\C{item\_req},r)\\
+    \phi_I & = & \top \\ 
+    \G_U & = & \lambda (\C{item\_req},r).~ 
+        r = \langle \C{ol\_o\_id}=\C{k2},\,
+                    \C{ol\_d\_id}=\C{k3},\, 
+                    \C{ol\_i\_id}=\C{item\_req.ol\_i\_id},\, 
+                    \C{ol\_qty}=\C{item\_req.ol\_qty}
+             \rangle\\
+  \end{array}
+\end{smathpar}
+
+\subsection{Decidability}
+\label{sec:decidability} %this label is referenced earlier.
 
@@
     as sets and functions, are never quantified.
-  \item Quantifiers appear only at the prenex position, i.e., at the
-    beginning of a quantified formula.  
+  \item All quantifiers are universal, and appear only at the prenex
+    position, i.e., at the beginning of a quantified formula.  
   \item All function symbols, modulo those that might appear in
@@
 \end{itemize}
-This fragment of FOL known as EPR (Effectively Propositional logic)
-is known to be decidable~\cite{z3epr}. The language of encoding, however, is a
-combination of {\sf EPR} with (a). $\Prop^0$, the theory from which
-quantifier-free propositions ($\varphi$) that encode object language
-expressions are drawn, and (b).  $\Prop^1$, the theory from which
-invariants ($I$) are drawn. We write $\SL[\Prop^0,\Prop^1]$ to
+This fragment of FOL known as EPR (Effectively Propositional logic) is
+known to be decidable~\cite{z3epr}. The language of encoding, however,
+is a combination of {\sf EPR} with (a). $\Prop^0$, the theory from
+which quantifier-free propositions ($\varphi$) that encode object
+language expressions are drawn, and (b).  $\Prop^1$, the theory from
+which invariants ($I$) are drawn. We write $\SL[\Prop^0,\Prop^1]$ to
 highlight the parameterization of $\SL$ on $\Prop^0$ and $\Prop^1$,

### 2017-10-20 motivation.tex
motivation

@@
   \item After $T$ finishes execution, but before it commits its local
-    state $\delta$, $\forall(x\in\mathit{dom}(delta)).~\Delta'(x) = \Delta(x)$.
+    state $\delta$, $\forall(x\in\mathit{dom}(\delta)).~\Delta'(x) = \Delta(x)$.
 \end{itemize}
@@
 
-We approach the verification problem by first observing that a typical
+We approach the verification problem by first observing that a
 relational database is a significantly simpler abstraction than shared
@@
 update statement (call it $\T_U$) is the following function on
-sets\footnote{$\bind$ has higher precedence than $\cup$.}:
+sets\footnote{$\bind$ has higher precedence than $\cup$; angle braces
+($\langle \ldots \rangle$) are used to denote records in math}:
 \begin{smathpar}
@@
         s.\C{s\_i\_id} = \C{item\_req.ol\_i\_id}\\\hspace*{1.3in}}
-           {\{ \langle s\ {\sf with}\  s_\C{qty} = \C{k1}\rangle \}\\\hspace*{1.3in}}
+           {\{ \langle \C{s\_i\_id}=s.\C{s\_i\_id};\, 
+                       \C{s\_d\_id}=s.\C{s\_d\_id};\,
+                       \C{s\_qty} = \C{k1}\rangle \}\\\hspace*{1.3in}}
            {\emptyset})
@@
   \lambda(\stl,\stg).~ \stl \cup
-       \{\langle\C{ol\_o\_id}=\C{k2},
+       \{\langle\C{ol\_o\_id}=\C{k2},\,
                  \C{ol\_d\_id}=\C{k3},\, \C{ol\_i\_id}=\C{item\_req.ol\_i\_id},\, 
@@
 \end{smathpar}
-Observe that both transformers are of the form $T(\delta,\Delta) =
-\stl \cup F(\stl,\stg)$, where $F$ is a function that returns the set
-of records added  to the transaction-local database
-($\stl$). Let $\F_U$ and $\F_I$ be the corresponding functions for
-$\T_U$ and $\T_I$ shown above. The state transformation induced by the
-loop body in Fig.~\ref{fig:new_order_code} can be expressed as the
-following composition of $\F_U$ and $\F_I$:
+Observe that both transformers are of the form $\T(\delta,\Delta) =
+\stl \cup \F(\stl,\stg)$, where $\F$ is a function that returns the
+set of records added  to the transaction-local database ($\stl$). Let
+$\F_U$ and $\F_I$ be the corresponding functions for $\T_U$ and $\T_I$
+shown above. The state transformation induced by the loop body in
+Fig.~\ref{fig:new_order_code} can be expressed as the following
+composition of $\F_U$ and $\F_I$:
 \begin{smathpar}

### 2017-10-20 popl-rebuttal.md
Shepard rebuttal.

@@
   state.
+
+Shepard
+----------
+- The primary concern raised by reviewer B is that the fragment of FOL
+  in which the translation algorithm (Fig. 10, Sec.5) encodes set
+  expressions is not EPR, hence its decidability is doubtful. In
+  particular, the concern was that an existential quantifier follows a
+  universal quantifier in the logical encoding of bind (`>>=`), and
+  this pattern is not admitted by EPR. In the rebuttal phase we
+  addressed the concern by pointing out that the fragment, while not
+  EPR as claimed, is nonetheless decidable since it is subsumed by the
+  GKS fragment, which is known to be decidable. However, as the
+  reviewer rightly points out, there are no known results on the
+  combination of GKS and Simple Linear Arithmetic (SLA) to support
+  theorem 5.3, so SLA invariants, which are common in databases, might
+  make the encoding undecidable. While this concern can be addressed
+  by considering machine (finite-bit) arithmetic instead of SLA, we
+  decided to take a long, hard look at the encoding algorithm and see
+  if it can be simplified. In this we were successful as we were able
+  to modify our encoding to eliminate the existential altogether. Our
+  modifications are based on the observation that the semantics of
+  bind, which were previously written using two implications one of
+  which involves an existential, can be equivalently written using a
+  single bi-implication that only involves universal quantifiers. A
+  rigorous proof to this effect can be found in the techreport, while
+  an automatic Z3-assisted proof can be found
+  [here](https://rise4fun.com/Z3/G1a). As a consequence of this
+  change, the fragment of encoding is back to being EPR, hence the
+  Theorems 5.2 and 5.3 continue to hold.

### 2017-10-20 inference.tex
edits

@@
 \begin{figure}[h]
+\begin{center}
 \begin{ocaml}
@@
 \end{ocaml}
+\end{center}
 \caption{A transaction that deposits an interest to a bank account.}
@@
 
-The correspondence between the inference rules and the RG judgment of
-\S~\ref{sec:reasoning} is stated thus:
+We can now establish a correspondence between the inference rules given
+above and the RG judgment of \S\ref{sec:reasoning}:
 \begin{theorem}
   \label{thm:inference-sound}
-  \emph{Forall} i,$R$,$I$,$c$,$\F$, if $\stable(\R,I)$ and $c \elabsto \F$,
+  Let the set expression $\F(\emptyset,\stg)$ have a natural interpretation
+  in the theory of sets.  Then, for all $i$,$R$,$I$,$c$,$\F$,
+  if $\stable(\R,I)$ and $c \elabsto \F$,
   then:\\\vspace*{-0.2cm}
@@
 \end{theorem}
-\noindent where the set expression $\F(\emptyset,\stg)$ has a natural
-interpretation in the theory of sets.
 
-\begin{proof}
-  We only provide a brief proof sketch here\footnote{The complete proof can be found in
-the supplementary material.}. We prove a slightly
-  stronger statement :
-
-  \emph{Forall} i,$R$,$I$,$c$,$\F$,$s$ if $\stable(\R,I)$ and $c
-  \elabsto \F$, then:
-
+\paragraph{{\sc Proof Sketch.}} We prove a slightly stronger statement :
+For all  $i$,$R$,$I$,$c$,$\F$,$s$ if $\stable(\R,I)$ and $c \elabsto \F$, then:
   \begin{smathpar}
@@
   \end{smathpar}
+The proof follows by structural induction on $c$. Assume that $P
+\Leftrightarrow \lambda(\stl,\stg).~\stl=s \conj I(\stg)$ and $Q
+\Leftrightarrow \lambda(\stl,\stg).\stl = \F(s,\stg)$. For each type
+of $c$, our strategy is to show the premises of the corresponding
+RG-rule. The base cases correspond to \C{INSERT}, \C{UPDATE} and \C{DELETE}
+statements. In all these cases, the form of $\F$ directly corresponds
+to the condition linking the pre-condition $P$ and post-condition $Q$
+in the premises of the RG-rules. If $\stable(\R, \F)$, then
+$\stabilize{\F} = \F$ and hence the proof directly follows. Otherwise,
+$\stabilize{\F} = \lambda (\stl, \stg).\existsl(\stg', I(\stg'),
+\F(\stl, \stg'))$. Since the pre-condition $P(\stl, \stg)$ asserts
+$I(\stg)$, any state which obeys $P$ provides the existential $\stg'$
+in the above set expression, and hence $\F$ applied on that state
+matches the condition linking $P$ and $Q$. The stability of $P$ with
+respect to $\R$ follows directly from the stability of $I$.
 
-  We use structural induction on $c$. Assume
-  that $P \Leftrightarrow \lambda(\stl,\stg).~\stl=s \conj I(\stg)$
-  and $Q \Leftrightarrow \lambda(\stl,\stg).\stl = \F(s,\stg)$. For
-  each type of $c$, our strategy is to show the premises of the
-  corresponding RG-rule. The base cases correspond to the INSERT,
-  UPDATE and DELETE statements. In all these cases, the form of $F$
-  directly corresponds to the condition linking the pre-condition $P$
-  and post-condition $Q$ in the premises of the RG-rules. If
-  $\stable(\R, F)$, then $\stabilize{F} = F$ and hence the proof
-  directly follows. Otherwise, $\stabilize{F} = \lambda (\stl,
-  \stg).\existsl(\stg', I(\stg'), F(\stl, \stg'))$. Since the
-  pre-condition $P(\stl, \stg)$ asserts $I(\stg)$, any state which
-  obeys $P$ provides the existential $\stg'$ in the above set
-  expression, and hence $F$ applied on that state matches the
-  condition linking $P$ and $Q$. The stability of $P$ with respect to
-  $\R$ follows directly from the stability of $I$.
-
-  The proofs for SELECT, SEQ and IF-THEN-ELSE directly use the inductive
-  hypothesis to infer the RG-judgements present in the premises of
-  their corresponding RG-rules. The interesting case is the FOREACH
-  statement, for which we use the loop invariant $\psi(\stl, \stg)
-  \Leftrightarrow \stl = s \cup (y \bind (\lambda z.~\F(s, \stg)))$,
-  (where assuming that $c$ is the body of the loop, $c \elabsto F$) to
-  prove all the premises of \rulelabel{RG-Foreach}. Using the same
-  notation as the rule $\rulelabel{RG-Foreach}$, $y$ refers to the
-  records already processed in previous iterations of the loop, while
-  $z$ refers to the record being processed in the current iteration.
-  At the beginning of the loop $[\phi/y]\psi(\stl, \stg)$ just reduces
-  to $\stl = s$ which is implied by the pre-condition $P$. From the
-  inductive hypothesis, we can infer that each iteration corresponds
-  to the application of $F$. Since all iterations are assumed to be
-  independent of each other, and $z$ is bound to a record in $x$ for
-  each iteration, we can infer that at the end of every iteration the
-  loop invariant $[y \cup \{z\}/y]\psi$ will be satisfied. 
-\end{proof}
+The proofs for \C{SELECT}, sequencing, and conditionals use use the
+inductive hypothesis to infer the RG-judgements present in the
+premises of their corresponding RG-rules. The interesting case is the
+\C{FOREACH} statement, for which we use the loop invariant
+$\psi(\stl, \stg) \Leftrightarrow \stl = s \cup (y \bind (\lambda
+z.~\F(s, \stg)))$, (where assuming that $c$ is the body of the loop,
+$c \elabsto \F$) to prove all the premises of
+\rulelabel{RG-Foreach}. Using the same notation as the rule
+$\rulelabel{RG-Foreach}$, $y$ refers to the records already processed
+in previous iterations of the loop, while $z$ refers to the record
+being processed in the current iteration.  At the beginning of the
+loop $[\phi/y]\psi(\stl, \stg)$ just reduces to $\stl = s$ which is
+implied by the pre-condition $P$. From the inductive hypothesis, we
+can infer that each iteration corresponds to the application of
+$F$. Since all iterations are assumed to be independent of each other,
+and $z$ is bound to a record in $x$ for each iteration, we conclude
+that at the end of every iteration, the loop invariant $[y \cup
+  \{z\}/y]\psi$ will be satisfied.
 
@@
 on encoding a set of element type $T$ as a unary predicate on $T$.
-The predicate is represented as a meta function that accepts a
-an $x:T$ and returns a quantifier-free proposition that evaluates to
+The predicate is represented as a meta function that accepts an
+$x:T$ and returns a quantifier-free proposition that evaluates to
 true ($\top$) if and only if $x$ is present in the set. Alternatively,
@@
 predicate describing the set can be encoded as the function $\lambda
-\v. \v=1 \vee \v=2$, with no further constraints (i.e., with $\top$
-constraint), or it can be encoded as the function $\lambda \v. g(\v)$
+\v. \v=1 \vee \v=2$, with no further constraints, or it can be encoded as the function $\lambda \v. g(\v)$
 with an associated constraint ($\phi\in\Prop^1$) defining the
@@
 subexpression should also be indexed by its free variables, so that a
-unique set exists for each instantiation of the free variables. For
-instance, the predicate ($\G$) associated with subexpression from the
+unique set exists for each instantiation of the free variables. Thus,
+the predicate ($\G$) associated with the subexpression from the
 above example should be $\lambda\v_1.\lambda\v_2.~\v_2 = \v_1 + 1$, so
@@
 The constraints associated with the encoding of a state are trivial
-($\top$). The set literal expression $\{x\,|\, \varphi\}$ is encoded
+(denoted $\top$). The set literal expression $\{x\,|\, \varphi\}$ is encoded
 straightforwardly. The conditional set expression is encoded as an
@@
 The first-order encoding of a bind expression describes the semantics
-of set monad's bind operator in FOL. Let $s_1$ be a set, and let $f =
+of the set monad's bind operator in FOL. Let $s_1$ be a set, and let $f =
 \lambda x.~s_2$ be a function that maps each variable in $s_1$ to a
@@
 and $(\phi_2,\G_2)$, respectively. Since $s_2$ is under a new binder
-that binds $x$, free variable sequence of $s_2$ is $\nubar,x$. In the
+that binds $x$, the free variable sequence of $s_2$ is $\nubar,x$. In the
 interest of hygiene, we substitute a fresh $a$ for $x$, making the
-sequence $\nubar,a$. The set $s$ is encoded as a a new uninterpreted
+sequence $\nubar,a$. The set $s$ is encoded as a new uninterpreted
 predicate $g$ indexed by $s$'s free variables ($\nubar$). Since the
@@
 of free variables ($\nubar$) for which $g$ needs to be defined;
-observe that the lambda returned by $\mssemof{\nubar}{s_1 \bind
+observe that the lambda expression returned by $\mssemof{\nubar}{s_1 \bind
 \lambda x.s_2}$ asserts $\pi(\vbar)$ only in conjunction with
@@
 condition that $f(\nubar,a) \conj [a/\stg]\phi$ holds. Since the
-existence of a $\stg$ that satisfies $\phi$ is axiomatized tp
+existence of a $\stg$ that satisfies $\phi$ is axiomatized to
 define $s$, if the aforementioned condition does not hold, i.e., if
@@
 transaction from Sec.~\ref{sec:motivation}. Recall that the state
-transformer ($\T$) for the foreach loop shown in
+transformer for the \C{foreach} loop shown in
 Fig.~\ref{fig:foreach_code} is (\C{k1}, \C{k2}, and \C{k3} are
@@
 \end{smathpar}
-Where:
+where:
 \begin{smathpar}
@@
 $\mssemof{\C{item\_req}}{\F_U(\stl,\stg)}$ and
-$\mssemof{\C{item\_req}}{\F_I(\stl,\stg)}$, which result in
+$\mssemof{\C{item\_req}}{\F_I(\stl,\stg)}$, which results in
 $(\phi_U,\G_U)$ and $(\phi_I,\G_I)$, respectively, where $\phi_U$,

### 2017-10-20 opsem.tex
edits

@@
 of \C{FOREACH} lends itself to inductive reasoning that will be useful
-for verification (Sec.~\ref{sec:reasoning}).
+for verification (Sec.~\ref{sec:reasoning}).  Our language ensures
+that all effectful actions are encapsulated within database commands,
+and that all shared state among processes are only manipulated via
+transactions and its supported operations.  In particular, we do not
+consider programs in which objects resident on e.g., the OCaml heap
+are concurrently manipulated by OCaml expressions as well as database
+actions.
 

### 2017-10-20 reasoning.tex
edits

@@
 In our setting, we only need to prove the stability of an assertion ($\phi$) against those environment steps which lead to a global database state on which the transaction itself can take its next step according to its isolation specification ($\I$).
-
 \begin{smathpar}
@@
 \end{smathpar}
-
-A characteristic of RG reasoning is that stability of an assertion is
+\noindent A characteristic of RG reasoning is that stability of an assertion is
 always proven w.r.t to $R$, and not $R^{*}$, although interference may
@@
 beginning and the end states.
-
  \begin{smathpar}
@@
 \end{smathpar}
-
-In the above condition, we use a separate notation ($\stg_1$) for the
-global database state that the transaction witnessed during its last
-step. $\phi$ does not constrain this global database state, but the
-stability of $\I$ ensures that any database state constrained by $\phi$
-must also obey $\I$. We can now replace the transitive closure in the
-above condition with the single step relation $R$, backed by the
-inductive argument for stability. Pushing the universal quantifier for
-$\stg_1$ inside the antecedent and separating the predicates involving
-$\stg_1$, we can rewrite the condition in the form below, by defining
-a new ternary relation $R \backslash \I$, which can be thought of as
-$R$ modulo $\I$:
-
+Here, we use a separate notation ($\stg_1$) for the global database
+state that the transaction witnessed during its last step. $\phi$ does
+not constrain this global database state, but the stability of $\I$
+ensures that any database state constrained by $\phi$ must also obey
+$\I$. We can now replace the transitive closure in the rule with the
+single step relation $R$, backed by the inductive argument for
+stability. Pushing the universal quantifier for $\stg_1$ inside the
+antecedent and separating the predicates involving $\stg_1$, we can
+rewrite the condition in the form below, by defining a new ternary
+relation $R \backslash \I$, which can be thought of as $R$ modulo
+$\I$:
 \begin{smathpar}
@@
 \end{smathpar}
-
-The above condition often significantly simplifies the form of $R
+This condition often significantly simplifies the form of $R
 \backslash \I$ irrespective of $R$. For example, when a transaction is
@@
 the global database satisfies the high-level invariant $I$. The rely
-relation ($\R_e$) is obtained from the global rely relation $R$ and the
-isolation specification $\I_e$ as explained above. Recall that $\I_e$
-constrains the global effects visible to the transaction while it is
-executing but has not yet committed.
+relation ($\R_e$) is obtained from the global rely relation $R$ and
+the isolation specification $\I_e$ as explained above. Recall that
+$\I_e$ constrains the global effects visible to the transaction while
+it is executing but has not yet committed, and $P$ and $Q$ of the
+transaction-local RG judgment are binary assertions; they relate local
+and global database states. The local judgment rules require one or
+both of them to be stable with respect to the constrained rely
+relation $\R_e$.
 
@@
 
-Recall that $P$ and $Q$ of the transaction-local RG judgment are
-binary assertions; they relate local and global database states. The
-local judgment rules require one or both of them to be stable with
-respect to the constrained rely relation $\R_e$.
-
 %The
@@
 and after the commit. The act of commit is captured by the flush
-operation ($\stl\rhd\stg$). Once we establish the validity of $G$ as a
+action ($\stl\rhd\stg$). Once we establish the validity of $G$ as a
 faithful representative of the transaction, we can verify that the
@@
 
-\begin{proof}
-Here, we provide only a brief proof sketch\footnote{The complete proof is provided in the supplementary material.}.
-The most important rule is the top-level rule \rulelabel{RG-Txn}, which proves that a transaction $c$ which begins its execution in global database state satisfying $I$ and suffering interference $R$ while executing under isolation specification $\I$ finishes its execution in a database state also satisfying $I$, and also guarantees that its commit step satisfies $G$.
-The rule uses the transaction-local RG judgement $\R_e \vdash \rg{P}{c}{Q}$.
-By \rulelabel{E-Txn-Start}, the local and global database states at the start of a transaction satisfy $P$, and the only challenge is that environment steps in an execution covered by $\R_e \vdash \rg{P}{c}{Q}$ are in $\R_e$, while the top-level judgement requires environment steps in $R$.
-We show that it is enough to consider only those enviroment steps in $\R_e$.
-First, we use an inductive argument and stability of $\I_e$ ($\stable(R, \I_e)$) to show that any execution in which the transaction completes all its steps must always preserve the isolation specification $\I_e$ after every environment step.
-Intuitively, this is because once $\I_e$ gets broken after some environment step, it will continue to remain broken and the transaction would not be able to proceed (according to \rulelabel{E-Txn}).
-Since $\R_e$ contains exactly those environment steps which preserve $\I_e$, the local-level RG judgement can be soundly used, which guarantees that after the transaction finishes its execution, its local state $\stl$ and global state $\stg$ will satisfy the assertion $Q$.
-Environment steps between the last step of the transaction and its commit step can modify the global state, and hence we also require $Q$ to be stable againt $R$.
-Again, we use an inductive argument, the stability of $\I_c$, and the fact that the transaction must execute its commit step to show that all environment steps must preserve $\I_c$, and hence it is enough to require $\stable(\R_c, Q)$.
-$Q$ guarantees that the commit step is in $G$, and $G$ in turn guarantees that after execution, the global database state will obey the invariant $I$.
+\paragraph{{\sc Proof Sketch.}}  The most important rule is the top-level
+rule \rulelabel{RG-Txn}, which proves that a transaction $c$ which
+begins its execution in global database state satisfying $I$ and
+encountering interference $R$ while executing under isolation
+specification $\I$ finishes its execution in a database state also
+satisfying $I$, and also guarantees that its commit step satisfies
+$G$. The rule uses the transaction-local RG judgement $\R_e \vdash
+\rg{P}{c}{Q}$. By \rulelabel{E-Txn-Start}, the local and global
+database states at the start of a transaction satisfy $P$, and the
+only challenge is that environment steps in an execution covered by
+$\R_e \vdash \rg{P}{c}{Q}$ are in $\R_e$, while the top-level
+judgement requires environment steps in $R$. We show that it is enough
+to consider only those enviroment steps in $\R_e$. First, we use an
+inductive argument and stability of $\I_e$ ($\stable(R, \I_e)$) to
+show that any execution in which the transaction completes all its
+steps must always preserve the isolation specification $\I_e$ after
+every environment step. Intuitively, this is because once $\I_e$ gets
+broken after some environment step, it will continue to remain broken
+and the transaction would not be able to proceed (according to
+\rulelabel{E-Txn}). Since $\R_e$ contains exactly those environment
+steps which preserve $\I_e$, the local-level RG judgement can be
+soundly used, which guarantees that after the transaction finishes its
+execution, its local state $\stl$ and global state $\stg$ will satisfy
+the assertion $Q$. Environment steps between the last step of the
+transaction and its commit step can modify the global state, and hence
+we also require $Q$ to be stable againt $R$. Again, we use an
+inductive argument, the stability of $\I_c$, and the fact that the
+transaction must execute its commit step to show that all environment
+steps must preserve $\I_c$, and hence it is enough to require
+$\stable(\R_c, Q)$. $Q$ guarantees that the commit step is in $G$, and
+$G$ in turn guarantees that after execution, the global database state
+will obey the invariant $I$.
 
-The local RG-judgement rules are mostly straightforward and directly follow from the operational semantics.
-At this point, the transaction can only change its local database state by adding new records to it, and hence all the rules have the form : $P(\stl, \stg)$ and $\stl'$ is the new state should satisfy $Q(\stl', \stg)$.
-An important thing to note is that the rules for Update, Select, Delete and Insert only require the pre-condition $P$ to be stable against interference.
-This is because the local RG-judgement rules only make assertions about the database state just after the transaction takes its last step.
-For the aforementioned commands, the last step would be the only step taken by the transaction, and hence $Q$ would not be affected by any environment steps, but instead directly reflect the changes made by the command (the environment steps between the last execution step of the transaction and the commit step would be controlled by $\I_c$ and would be taken care of by $\stable(\R_c, Q)$ at the top level).
+%% The local RG-judgement rules are mostly straightforward and directly
+%% follow from the operational semantics. At this point, the transaction
+%% can only change its local database state by adding new records to it,
+%% and hence all the rules have the form : $P(\stl, \stg)$ and $\stl'$ is
+%% the new state should satisfy $Q(\stl', \stg)$. Note that note is that
+%% the {\sc RG-Update}, {\sf RG-Select}, {\sc RG-Delete} and {\sc
+%%   RG-Insert} rules only require the pre-condition $P$ to be stable
+%% against interference. This is because the these rules only make
+%% assertions about the database state just after the transaction takes
+%% its last step. For the aforementioned commands, the last step would be
+%% the only step taken by the transaction, and hence $Q$ would not be
+%% affected by any environment steps, but instead directly reflect the
+%% changes made by the command (the environment steps between the last
+%% execution step of the transaction and the commit step would be
+%% controlled by $\I_c$ and would be taken care of by $\stable(\R_c, Q)$
+%% at the top level).
 
-The rule $\rulelabel{RG-Foreach}$ makes use of an inductive loop invariant $\psi$ which assumes that the variable $y$ is bound to records which have been processed in previous iterations and $z$ is bound to the record processed in the current iteration.
-Since the operational semantics ensure that every iteration processes a unique record, the proof is straightforward.
-\end{proof}
+%% The rule $\rulelabel{RG-Foreach}$ makes use of an inductive loop
+%% invariant $\psi$ which assumes that the variable $y$ is bound to
+%% records which have been processed in previous iterations and $z$ is
+%% bound to the record processed in the current iteration. Since the
+%% operational semantics ensure that every iteration processes a unique
+%% record, the proof is straightforward.

### 2017-10-20 introduction.tex
DSL syntax made same as OCaml syntax.

@@
 \begin{enumerate}
-\item We analyze properties of weak isolation in the context of an
-    embedded DSL for OCaml that treats SQL-based relational database
-    operations (e.g., inserts, selects, deletes, joins, updates, etc.)
-    as monadic computations over an abstract database state.
+  \item We analyze properties of weak isolation in the context of a
+    DSL embedded in OCaml that treats SQL-based relational database
+    operations (e.g., inserts, selects, deletes, updates, etc.) as
+    computations over an abstract database state.
   \item We develop an operational semantics and a compositional
@@
     exposes the visibility of these operations to other transactions.
- \item Exploiting the monadic definition of database state, we devise
-   an inference algorithm capable of discovering the weakest isolation
-   level that is sound with respect to a transaction's high-level
-   consistency requirements.  The algorithm interprets database
-   operations as state transformers expressed in a language amenable
-   for translation into a decidable fragment of first-order logic, and
-   is thus suitable for automated verification using off-the-shelf SMT
-   solvers.
- \item We present details of a realistic implementation along with an
-   evaluation study on real-world database benchmarks that justify our
-   approach and the utility of our proof methodology and inference
-   mechanism.
+  \item We devise an inference algorithm capable of discovering the
+    weakest isolation level that is sound with respect to a
+    transaction's high-level consistency requirements. The algorithm
+    interprets database operations as state transformers expressed in
+    a language amenable for translation into a decidable fragment of
+    first-order logic, and is thus suitable for automated verification
+    using off-the-shelf SMT solvers.
+  \item We present details of an implementation along with an
+    evaluation study on real database benchmarks that justify our
+    approach, and demonstrate the utility of our inference mechanism.
 \end{enumerate}

### 2017-10-20 motivation.tex
DSL syntax made same as OCaml syntax.

@@
 
-We present our ideas in the context of an embedded DSL in OCaml that
-we have developed, which offers a \C{DB} monad to define and compose
-database computations given in terms of SQL operations.  Besides the
-usual \C{bind} and \C{return} operators, the monad includes an
-\C{atomically} combinator that executes a database computation as an
-atomic transaction and returns the result.
+We present our ideas in the context of a DSL embedded in OCaml that
+manages an abstract database state, and allows it be manipulated via a
+well-defined \C{SQL} interface. Arbitrary database computations can be
+built around this interface, which can then be run as transactions
+using the \C{atomically\_do} combinator provided by the DSL.
 
@@
 \begin{ocaml}
-let new_order (d_id, c_id, item_reqs) = atomically do
-  dist <- SQL.select1 District (fun d -> d.d_id = d_id);
+let new_order (d_id, c_id, item_reqs) = atomically_do @@ fun () ->
+  let dist = SQL.select1 District (fun d -> d.d_id = d_id) in
   let o_id = dist.d_next_o_id in
@@
             o_d_id=d_id; o_c_id=c_id; o_ol_cnt=S.size item_reqs; };
-      foreach item_reqs @@ fun item_req -> do
-        stk <- SQL.select1(* SELECT * FROM *) Stock 
+  foreach item_reqs @@ fun item_req ->
+    let stk = SQL.select1(* SELECT * FROM *) Stock 
               (* WHERE *)(fun s -> s.s_i_id = item_req.ol_i_id 
@@
 (racy) programs~\cite{GHE15,HPQ+15}.  To illustrate, consider the
-behavior of the \C{new\_order} transaction
-when executed with a \emph{Read Committed} (RC) isolation level, the
+behavior of the \C{new\_order} transaction when executed under
+\emph{Read Committed} (RC) isolation level, the default isolation
+level in 8 of the 18 databases studied in~\cite{bailishotos}.
 \begin{wrapfigure}{l}{.4\textwidth}
@@
 \end{wrapfigure}
-default isolation level in 8 of the 18 databases studied
-in~\cite{bailishotos}.  An executing RC transaction is isolated from
-\emph{dirty writes}, i.e., writes of uncommitted transactions, but is
-allowed to witness the writes of concurrent transactions as soon as
-they are committed. Thus, with two concurrent instances of the
-\C{new\_order} transaction (call them $T_1$ and $T_2$), both
-concurrently placing new orders for different customers under the same
-district (\C{d\_id}), RC isolation allows the execution shown in
-Fig.~\ref{fig:new_order_exec}.
+An executing RC transaction is isolated from \emph{dirty writes},
+i.e., writes of uncommitted transactions, but is allowed to witness
+the writes of concurrent transactions as soon as they are committed.
+Thus, with two concurrent instances of the \C{new\_order} transaction
+(call them $T_1$ and $T_2$), both concurrently placing new orders for
+different customers under the same district (\C{d\_id}), RC isolation
+allows the execution shown in Fig.~\ref{fig:new_order_exec}.
 
@@
 Clearly, this example does not exhibit any of the anomalies that
-characterize RC isolation~\cite{berenson}. For instance, there are no
-\emph{lost writes} - both concurrent transactions' writes are present
-in the final state of the database.  Program analyses that aim to
-determine appropriate isolation by checking for possible
+\emph{characterize} RC isolation~\cite{berenson}\footnote{Berenson et
+al~\cite{berenson} characterize isolation levels in terms of the
+anomalies they \emph{admit}. For e.g., RC is characterized by
+\emph{lost writes} because it admits the anomaly.}. For instance,
+there are no \emph{lost writes} - both concurrent transactions' writes
+are present in the final state of the database.  Program analyses that
+aim to determine appropriate isolation by checking for possible
 manifestations of RC-induced anomalies would fail to identify grounds
@@
 program.  Instead, we see value in viewing them as essentially
-functional computations that manage database state monadically,
+functional computations that manage database state abstractly,
 mirroring the structure of our DSL.  By doing so, we can formulate the
@@
 \begin{ocaml}
-foreach item_reqs @@ fun item_req -> do
+foreach item_reqs @@ fun item_req ->
   SQL.update Stock (fun s -> {s with s_qty = k1}) 
@@
 \begin{array}{l}
-  \lambda(\stl,\stg).~ \stl \cup \stg \bind(\lambda s. 
-      \itel{{\sf table}(s) = \C{Stock} \conj 
-        s.\C{s\_i\_id} = \C{item\_req.ol\_i\_id}\\\hspace*{1.3in}}
-           {\{ \langle \C{s\_i\_id}=s.\C{s\_i\_id};\, 
-                       \C{s\_d\_id}=s.\C{s\_d\_id};\,
-                       \C{s\_qty} = \C{k1}\rangle \}\\\hspace*{1.3in}}
+  \lambda(\stl,\stg).~ \stl \cup \stg \bind(\lambda \C{s}. 
+      \itel{{\sf table}(\C{s}) = \C{Stock} \conj 
+        \C{s}.\C{s\_i\_id} = \C{item\_req.ol\_i\_id}\\\hspace*{1.15in}}
+           {\{ \langle \C{s\_i\_id}=\C{s.s\_i\_id};\, 
+                       \C{s\_d\_id}=\C{s.s\_d\_id};\,
+                       \C{s\_qty} = \C{k1}\rangle \}\\\hspace*{1.15in}}
            {\emptyset})
@@
 local database state $\delta$.\footnote{For now, assume that the
-  record being added to the local database has not previously been
-  updated.}
+  record being added to $\stl$ is not already present in $\stl$.}
 
@@
 the program itself. In particular, SQL statements become set
-operations, and the \C{DB} monad's $\bind$ becoms set monadic $\bind$.
-As we demonstrate, the advantage of inferring such transformers is
-that we can now make use of a semantics-preserving translation from
-the domain of sets equipped with $\bind$ to a decidable fragment of
-first-order logic, allowing us to leverage SMT solvers for automated
-proofs without having to infer potentially complex thread-local
-invariants or intermediate assertions.  Sec.~\ref{sec:inference}
-describes this translation. In the exposition thus far, we assumed
-$\Delta$ remains invariant, which is clearly not the case when we
-admit concurrency.  Necessary concurrency extensions of the state
-transformer semantics to deal with interference is also covered in
-Sec.~\ref{sec:inference}.  Before presenting the transformer
-semantics, we first focus our attention in the following two sections
-on the theoretical foundations for weak isolation, upon which this
-semantics is based.
+operations, and the \C{foreach} combinator becoms set monad's bind
+($\bind$) combinator.  As we demonstrate, the advantage of inferring
+such transformers is that we can now make use of a
+semantics-preserving translation from the domain of sets equipped with
+$\bind$ to a decidable fragment of first-order logic, allowing us to
+leverage SMT solvers for automated proofs without having to infer
+potentially complex thread-local invariants or intermediate
+assertions.  Sec.~\ref{sec:inference} describes this translation. In
+the exposition thus far, we assumed $\Delta$ remains invariant, which
+is clearly not the case when we admit concurrency.  Necessary
+concurrency extensions of the state transformer semantics to deal with
+interference is also covered in Sec.~\ref{sec:inference}.  Before
+presenting the transformer semantics, we first focus our attention in
+the following two sections on the theoretical foundations for weak
+isolation, upon which this semantics is based.

### 2017-10-20 related-work.tex

@@
 is not immediately apparent if their specification formalism is
-amenable for use within a verification toolchain.
+amenable for use within a verification toolchain.  ~\cite{WB17}
+presents a dynamic analysis for weak isolation that attempts to
+discover weak isolation anomalies from SQL log files.  Their solution,
+while capable of identifying database attacks due to the use of weaker
+isolation levels than necessary, does not consider how to verify
+application correctness, infer proper isolation levels, or formally
+reason about the relationship between weak-isolation levels and
+application invariants.
 

### 2017-10-21 inference.tex
DSL syntax made same as OCaml syntax.
add_interest example better explained.

@@
 are also variables in $\SL$, but are used to denote local and database
-states, respectively.  The language is carefully chosen to be
+states, respectively. Constant sets can be written using the set
+literatl expressions. For e.g., the set $\{1,2\}$ can be written as
+$\{x \,|\, x=1 \vee x=2\}$. The language is carefully chosen to be
 expressive enough to capture the semantics of $\txnimp$ statements (as
@@
 Observe that when $\F$ is not stable, then $\stabilize{\F}$ returns a
-transformer $\F'$ that simply ignores its $\stg$ argument in favor of a generic
-$\stg'$, making $\F'$ trivially stable. It is safe to assume
+transformer $\F'$ that simply ignores its $\stg$ argument in favor of
+a generic $\stg'$, making $\F'$ trivially stable. It is safe to assume
 $I(\stg')$ because all verified transactions preserve the invariant,
@@
 the post-condition of a statement, as done by the
-\rulelabel{RG-Conseq} rule for transaction-bound commands.  The weakening semantics chosen by
-$\stabilize{\cdot}$, while being simple, is nonetheless useful because
-of the $I(\stg')$ assumption on the existentially bound $\stg'$. The
-example in Fig.~\ref{fig:weakening-example} demonstrates. 
+\rulelabel{RG-Conseq} rule for transaction-bound commands.  The
+weakening semantics chosen by $\stabilize{\cdot}$, while being simple,
+is nonetheless useful because of the $I(\stg')$ assumption on the
+existentially bound $\stg'$. The example in
+Fig.~\ref{fig:weakening-example} demonstrates. 
 \begin{figure}[h]
@@
 \begin{ocaml}
-let add_interest acc_id pc = atomically @@ do
-  let a = SQL.select1 BankAccount (fun acc -> acc.id = acc_id);
-  let y = a.bal + pc*a.bal;
+let add_interest acc_id pc = atomically_do @@ fun () ->
+  let a = SQL.select1 BankAccount (fun acc -> acc.id = acc_id) in
+  let y = a.bal + pc*a.bal in
   SQL.update BankAccount (fun acc -> {acc with bal = acc.bal + y})
@@
 Here, an \C{add\_interest} transaction adds a positive interest
-(\C{pc}) to the balance of a bank account, which is required to be
-non-negative ($I(\stg) \Leftrightarrow \forall(r\in\stg).~r.\C{bal}\ge
-0$). The transaction starts by issuing a \C{select1} query, whose
-effect implicitly asserts that there exists a record in the bank
-account database that is equal to $a$ ($\exists(r\in\stg).~r=a$).
-However, this assertion is unstable because a concurrent \C{withdraw}
-or \C{deposit} transactions might update the account balance, so that
-such a record no longer exists in the database.  Fortunately, the
-weakening operator ($\stabilize{\cdot}$) allows us to weaken the
-assertion to $\exists \stg'.I(\stg') \conj \exists(r\in\stg').~a=r$
-({\sf exists} has the same semantics as $\exists$, as formalized in
-Fig.~\ref{fig:logic}), which is enough to prove that $\C{a.bal +
-pc*a.bal}\ge 0$, and verify that \C{add\_interest} preserves the
-positive balance invariant.
+(determined by \C{pc}) to the balance of a bank account, which is
+required to be non-negative ($I(\stg) \Leftrightarrow
+\forall(r\in\stg).~r.\C{bal}\ge 0$). The transaction starts by issuing
+a \C{select1} query, whose effect ($\F$) is essentially a singleton
+set containing a record $r$ whose id is \C{acc\_id} (i.e.,
+$\F(\stl,\stg) = \{r \,|\, r\in\stg \wedge r.\C{id} = \C{acc\_id}\}$).
+However, this effect is unstable because $\F(\stl,\stg')$ may not be
+the same set as $\F(\stl,\stg)$ when $\stg'\neq\stg$. A record
+$r\in\stg$ whose $\C{id}=\C{acc\_id}$ may have its balance updated by
+a concurrent \C{withdraw} or \C{deposit} transaction in $\stg'$,
+making the record in $\stg'$ different from the record in $\stg$.
+Hence the stability check fails.  Fortunately, the weakening operator
+($\stabilize{\cdot}$) allows us to weaken the effect to
+$\existsl(\stg, I(\stg), \{ r \,|\, r\in\stg \wedge
+r.\C{id}=\C{acc\_id}\})$, which effectively asserts that the
+\C{select1} query returns a record with $\C{id}=\C{acc\_id}$ from
+\emph{some} database state that satisfies the non-negative balance
+invariant $I$.  This weakened assertion is nonetheless enough to
+deduce that $\C{a.bal}\ge0$, and subsequently prove that $\C{a.bal +
+pc*a.bal}\ge 0$, allowing us to verify the \C{add\_interest}
+transaction.
 

### 2017-10-21 implementation.tex
Edits.

@@
 computations in OCaml (modulo some syntactic sugar), and our automatic
-reasoning framework as an extra frontend pass (called \tool) in the ocamlc
-4.03 compiler\footnote{The source code is available at available at
-\emph{https://goo.gl/nmUXZK}}. The input to \tool is a program in our DSL that
-describes the schema of the database as a collection of OCaml type
-definitions, and transactions as OCaml functions, whose top-level
-expression is an application of the \C{atomically\_do} combinator to a
-function of type $a \rightarrow \C{DB}\;b$, for some (base) types $a$
-and $b$. For instance, TPC-C's schema from Fig.~\ref{fig:schema} can
-be described via the OCaml type definitions shown in
-Fig.~\ref{fig:ocaml-schema}.  \tool also requires a specification of
-the program in the form of a collection of guarantees ($G$), one per
-transaction, and an invariant $I$ that is a conjunction of the
-integrity constraints on the database. An auxiliary DSL that includes
-the first-order logic (FOL) combinators has been implemented for this
-purpose. \tool's verification pass follows OCaml's type checking pass,
-hence the concrete artifact of verification is OCaml's typed AST. The
-tool is already equipped with  an axiomatization of PostgreSQL and
-MySQL's isolation levels expressed in our FOL DSL. Other data stores
-can be similarly axiomatized. The concrete result of verification is
-an assignment of an isolation level of the selected data store to each
-transaction in the program.
+reasoning framework as an extra frontend pass (called \tool) in the
+ocamlc 4.03 compiler\footnote{The source code is available at
+available at \emph{https://goo.gl/nmUXZK}}. The input to \tool is a
+program in our DSL that describes the schema of the database as a
+collection of OCaml type definitions, and transactions as OCaml
+functions, whose top-level expression is an application of the
+\C{atomically\_do} combinator. For instance, TPC-C's schema from
+Fig.~\ref{fig:schema} can be described via the OCaml type definitions
+shown in Fig.~\ref{fig:ocaml-schema}.  \tool also requires a
+specification of the program in the form of a collection of guarantees
+($G$), one per transaction, and an invariant $I$ that is a conjunction
+of the integrity constraints on the database. An auxiliary DSL that
+includes the first-order logic (FOL) combinators has been implemented
+for this purpose. \tool's verification pass follows OCaml's type
+checking pass, hence the concrete artifact of verification is OCaml's
+typed AST. The tool is already equipped with  an axiomatization of
+PostgreSQL and MySQL's isolation levels expressed in our FOL DSL.
+Other data stores can be similarly axiomatized. The concrete result of
+verification is an assignment of an isolation level of the selected
+data store to each transaction in the program.
 
@@
 if it is not stable. The result of traversing the transaction's AST is
-therefore a state transformer ($\F$) that is stable w.r.t $R_l$, which
+therefore a state transformer ($\F$) that is stable w.r.t $\R_l$, which
 is also stabilized against $\R_c$ (using $\stabilize{\cdot}$), and

### 2017-10-21 inference.tex
Edits.

@@
 states, respectively. Constant sets can be written using the set
-literatl expressions. For e.g., the set $\{1,2\}$ can be written as
+literal expressions. For e.g., the set $\{1,2\}$ can be written as
 $\{x \,|\, x=1 \vee x=2\}$. The language is carefully chosen to be
@@
 \begin{array}{lcl}
-  \stable(\R,\F) & \Leftrightarrow & \forall \stl,\stg,\stg',\bar{v}.~
+  \stable(\R,\F) & \Leftrightarrow & \forall \stl,\stg,\stg',\nubar.~
   \R(\stl,\stg,\stg') \Rightarrow \F(\stl,\stg) \equiv \F(\stl,\stg')
@@
 \end{smathpar}
-where $\bar{v}$ are the variables that occur free in $\F$; this is
+where $\nubar$ are the variables that occur free in $\F$; this is
 possible because of how the inference rules are structured.
@@
 The proofs for \C{SELECT}, sequencing, and conditionals use use the
-inductive hypothesis to infer the RG-judgements present in the
+inductive hypothesis to infer the RG-judgments present in the
 premises of their corresponding RG-rules. The interesting case is the
@@
 on encoding a set of element type $T$ as a unary predicate on $T$.
-The predicate is represented as a meta function that accepts an
-$x:T$ and returns a quantifier-free proposition that evaluates to
-true ($\top$) if and only if $x$ is present in the set. Alternatively,
-translation may also encode the set as a function in the logic itself,
-in which case a quantified proposition constraining the function is
-also generated. For instance, consider the set $\{1,2\}$. The
-predicate describing the set can be encoded as the function $\lambda
-\v. \v=1 \vee \v=2$, with no further constraints, or it can be encoded as the function $\lambda \v. g(\v)$
-with an associated constraint ($\phi\in\Prop^1$) defining the
-uninterpreted boolean functon $g$:
+The predicate is represented as a meta function that accepts an $x:T$
+and returns a quantifier-free proposition that evaluates to true
+($\top$) if and only if $x$ is present in the set. Alternatively,
+translation may also encode the set as a predicate in the logic
+itself, in which case a quantified proposition constraining the
+predicate is also generated. For instance, consider the set $\{1,2\}$.
+The predicate describing the set can be encoded as the function
+$\lambda \v. \v=1 \vee \v=2$, with no further constraints, or it can
+be encoded as the function $\lambda \v. g(\v)$ with an associated
+constraint ($\phi\in\Prop^1$) defining the uninterpreted boolean
+function (i.e., uninterpreted predicate) $g$:
 \begin{smathpar}
@@
 ($\phi$,$\G$), where $\G$ is a meta function, and $\phi$ is a FOL
-formula constraining any uninterpreted functions used in $G$.
+formula constraining any uninterpreted predicates used in $G$.
 
@@
 uninterpreted sort. Considering that the state is actually a set of
-records, we define an uninterpreted relation $\in$ to relate records
+records, we define an uninterpreted relation ``$\in$'' to relate records
 and states. Thus, a variable set expression $\stg$ denoting a database

### 2017-10-21 opsem.tex
Edits.

@@
 records, arithmetic and boolean expressions ($e_1 \odot e_2$), and
-record expressions ($[\bar{f}=\bar{e}]$) constitute the syntactic
-class of expressions ($e$). Commands ($c$) include $\cskip$,
+record expressions ($\langle \bar{f}=\bar{e} \rangle$) constitute the
+syntactic class of expressions ($e$). Commands ($c$) include $\cskip$,
 conditional statements, \C{LET} constructs to bind names, \C{FOREACH}

### 2017-10-21 reasoning.tex
Edits.

@@
 \C{atomic} blocks also have the same characteristic, the fact that
-transactions are weakly-isolated introduces surprising complexity.
+transactions are weakly-isolated introduces non-trivial complexity.
 Unlike an \C{atomic} block, the effect of a transaction is \emph{not}
@@
 definitions of stability in Fig.~\ref{fig:rg-rules}, but they all
-convey the same intuition as above.
-In our setting, we only need to prove the stability of an assertion ($\phi$) against those environment steps which lead to a global database state on which the transaction itself can take its next step according to its isolation specification ($\I$).
+convey the same intuition as above. In our setting, we only need to
+prove the stability of an assertion ($\phi$) against those environment
+steps which lead to a global database state on which the transaction
+itself can take its next step according to its isolation specification
+($\I$). 
 \begin{smathpar}
@@
 modified a record, and no further interference will restore the
-original record, because the original record bears the id of a
-transaction that has already committed. We can now generalize the
-stability condition for any arbitrary assertion $\phi$ by only
-considering those environment steps which in addition to preserving
-$\phi$, also preserve the condition that the transaction can continue
-to take its next step, i.e. the condition $\I$ holds on both the
-beginning and the end states.
+original record, because new commits bear new transaction ids
+different from the transaction id (\C{txn} field) of the original
+record. We can now generalize the stability condition for any
+arbitrary assertion $\phi$ by only considering those environment steps
+which in addition to preserving $\phi$, also preserve the condition
+that the transaction can continue to take its next step, i.e. the
+condition $\I$ holds on both the beginning and the end states.
  \begin{smathpar}
@@
 \begin{array}{lclr}
-\R \vdash \hoare{P}{c}{Q} & \defeq & \forall \stl,\stl',\stg,\stg'.~ P(\stl,\stg) \conj (\tbox{c}_i,\stl,\stg) \rstepssto{n}
+\R \vdash \hoare{P}{c}{Q} & \defeq & \forall
+  \stl,\stl',\stg,\stg',\nubar.~ P(\stl,\stg) \conj (\tbox{c}_i,\stl,\stg) \rstepssto{n}
   (\tbox{\cskip}_i, \stl',\stg')
-  \Rightarrow Q(\stl',\stg') \\
+  \Rightarrow Q(\stl',\stg') & \nubar = \texttt{freevars}(c)\\
 \rg{I,R}{c}{G,I} & \defeq &  \forall \stg.\, I(\stg)
@@
     \rstepssto{n} (\cskip,\stg') \Rightarrow I(\stg')) \\
-&&\hspace*{0.5in}\conj \texttt{step-guaranteed}(R,G,c,\stg)\\
+&&\hspace*{0.6in}\conj \texttt{step-guaranteed}(R,G,c,\stg)\\
 \end{array}

### 2017-10-21 inference.tex
Edits.

@@
    \stable(R,\I)\spc
-  \stable(R,I)\spc
-  P(\stl,\stg) \Leftrightarrow \stl=\emptyset \wedge I(\stg)\\
+   \stable(R,I)\\
    \R_e = R \backslash \I_e \spc \R_c = R \backslash \I_c \spc  
-   \R_e \vdash \rg{P}{c}{Q} \spc \stable(\R_c,Q) \\ 
-  \forall \stl,\stg.~ Q(\stl,\stg) \Rightarrow 
-    G(\stg, \stl \rhd \stg)\spc
+   c \elabsto \F \spc \stable(\R_c,\F)\\
+   \forall \stl,\stg.~ \stl = F(\emptyset,\stg) \Rightarrow 
+   G(\stg, \stl\gg\stg)\spc
    \forall \stg,\stg'.~I(\stg) \wedge G(\stg,\stg') \Rightarrow I(\stg')\\
@@
 \end{smathpar}
+% \begin{smathpar}
+% \begin{array}{c}
+% \RULE
+% {
+%   \stable(R,\I)\spc
+%   \stable(R,I)\spc
+%   P(\stl,\stg) \Leftrightarrow \stl=\emptyset \wedge I(\stg)\\
+%   \R_e = R \backslash \I_e \spc \R_c = R \backslash \I_c \spc 
+%    \R_e \vdash \rg{P}{c}{Q} \spc \stable(\R_c,Q) \\ 
+%   \forall \stl,\stg.~ Q(\stl,\stg) \Rightarrow 
+%     G(\stg, \stl \rhd \stg)\spc
+%   \forall \stg,\stg'.~I(\stg) \wedge G(\stg,\stg') \Rightarrow I(\stg')\\
+% }
+% {
+%   \rg{I,R}{\ctxn{i}{\I}{c}}{G,I}
+% }
+% \end{array}
+% \end{smathpar}
 Automating the application of the \rulelabel{RG-Txn} rule for a
 transaction requires automating the multiple implication checks in
-the premise. While $R$, $\R$, $\I$ and $I$ are formulas in
+the premise. While $R$, $G$, $\I$ and $I$ are formulas in
 first-order logic (FOL) with a relatively simple structure, $\F$

### 2017-10-22 implementation.tex
Stability definition changed.

@@
 \textbf{Real-World Isolation Levels} The axiomatization of the
-isolation levels presented in \S\ref{sec:isolation} leave out
+isolation levels presented in \S\ref{sec:isolation} leaves out
 certain nuances of their implementations on real data stores, which

### 2017-10-22 motivation.tex
Stability definition changed.

@@
 As this discussion hints, implementations of SI on real databases
-such as PostgresSQL are highly complicated, often running into
+such as PostgreSQL are highly complicated, often running into
 thousands of lines of code.  Nonetheless, the semantics of SI, in
@@
 its commit point, leading to a verification process similar to what
-would been applied when reasoning sequentially.  When \C{new\_order}
-finishes execution, however, but before it commits, the SI
-axiomatization shown above requires us to consider global state
+would have been applied when reasoning sequentially.  When
+\C{new\_order} finishes execution, however, but before it commits, the
+SI axiomatization shown above requires us to consider global state
 transitions $\stg \longrightarrow \stg'$ that do not include changes
 to the records ($\stl$) written by \C{new\_order}, i.e.,
-$\forall(x\in\mathit{dom}(\delta)).~\Delta'(x) = \Delta(x)$.  The axiomatization
-precludes any execution in which there are concurrent updates to
-shared table fields (e.g., \C{d\_next\_o\_id} on the same \C{District}
-table), but does not prohibit interferences that write to different
-tables, or write to different records in the same table.  We need to
-reason about the safety of such interferences with respect to
+$\forall(x\in\mathit{dom}(\delta)).~\Delta'(x) = \Delta(x)$.  The
+axiomatization precludes any execution in which there are concurrent
+updates to shared table fields (e.g., \C{d\_next\_o\_id} on the same
+\C{District} table), but does not prohibit interferences that write to
+different tables, or write to different records in the same table.  We
+need to reason about the safety of such interferences with respect to
 \C{new\_order}'s consistency invariants.

### 2017-10-22 reasoning.tex
Stability definition changed.

@@
 \end{smathpar}
-\noindent A characteristic of RG reasoning is that stability of an assertion is
-always proven w.r.t to $R$, and not $R^{*}$, although interference may
-include multiple environment steps, and $R$ only captures a single
-step. This is nonetheless sound due to inductive reasoning: if $\phi$
-is preserved by every step of $R$, then $\phi$ is preserved by
-$R^{*}$, and vice-versa.  However, such reasoning does not extend
-naturally to isolation-constrained interference because $R^{*}$ modulo
-$\I$ is not same as $\R^{*}$; the former is a transitive relation
-constrained by $\I$, whereas the latter is the transitive closure of
-a constrained relation. This means, unfortunately, that we cannot directly replace $R^{*}$
-by $R$ in the above condition because of our need to consider the weak
-isolation specification $\I$ that selectively exposes interferences.
+\noindent A characteristic of RG reasoning is that stability of an
+assertion is always proven w.r.t to $R$, and not $R^{*}$, although
+interference may include multiple environment steps, and $R$ only
+captures a single step. This is nonetheless sound due to inductive
+reasoning: if $\phi$ is preserved by every step of $R$, then $\phi$ is
+preserved by $R^{*}$, and vice-versa.  However, such reasoning does
+not extend naturally to isolation-constrained interference because
+$R^{*}$ modulo $\I$ is not same as $\R^{*}$; the former is a
+transitive relation constrained by $\I$, whereas the latter is the
+transitive closure of a relation constrained by $\I$. This means,
+unfortunately, that we cannot directly replace $R^{*}$ by $R$ in the
+above condition because of our need to consider the weak isolation
+specification $\I$ that selectively exposes interferences.
 
@@
 \emph{stability condition on $\I$}.  The condition requires $\I$ to
-allow interferences $R^{*}(\stg,\stg'')$, for two database states
-$\stg$ and $\stg''$, only if it also allows interference for every
-prefix of $R^{*}(\stg,\stg'')$. In other words, if $\I$ disallows
-interference from $\stg$ to $\stg'$, then an $R$-step from $\stg'$ to
-$\stg''$ should not make the interference from $\stg$ to $\stg''$
-valid.  This stability condition on $\I$ is defined formally thus:
+admit the interference of multiple $R$ steps (i.e.,
+$R^{*}(\stg,\stg'')$, for two database states $\stg$ and $\stg''$),
+only if it also admits interference of each $R$ step along the way.
+Formally:
 \begin{smathpar}
@@
   \stable(R,\I) & \Leftrightarrow & \forall \stl,\stg,\stg',\stg''.~
-  \neg\I(\stl,\stg,\stg') \conj R(\stg',\stg'') \Rightarrow
-  \neg\I(\stl,\stg,\stg'')
+  \I(\stl,\stg,\stg'') \conj R(\stg',\stg'') \Rightarrow
+  \I(\stl,\stg,\stg') \conj \I(\stl,\stg',\stg'')
 \end{array}
@@
 satisfied by the isolation axioms from Sec.~\ref{sec:isolation}. For
-instance, $\I_{ss}$, the snapshot axiom, is stable because if
-$\I_{ss}$ is invalid ($\neg\I_{ss}$), then an interference has already
-modified a record, and no further interference will restore the
-original record, because new commits bear new transaction ids
-different from the transaction id (\C{txn} field) of the original
-record. We can now generalize the stability condition for any
-arbitrary assertion $\phi$ by only considering those environment steps
-which in addition to preserving $\phi$, also preserve the condition
-that the transaction can continue to take its next step, i.e. the
-condition $\I$ holds on both the beginning and the end states.
- \begin{smathpar}
-   \forall \stl, \stg, \stg', \stg_1. \phi(\stl, \stg) \wedge \I(\stl, \stg_1, \stg) \wedge R^{*}(\stg, \stg') \wedge \I(\stl, \stg_1, \stg')
-   \Rightarrow \phi(\stl, \stg')
-\end{smathpar}
-Here, we use a separate notation ($\stg_1$) for the global database
-state that the transaction witnessed during its last step. $\phi$ does
-not constrain this global database state, but the stability of $\I$
-ensures that any database state constrained by $\phi$ must also obey
-$\I$. We can now replace the transitive closure in the rule with the
-single step relation $R$, backed by the inductive argument for
-stability. Pushing the universal quantifier for $\stg_1$ inside the
-antecedent and separating the predicates involving $\stg_1$, we can
-rewrite the condition in the form below, by defining a new ternary
-relation $R \backslash \I$, which can be thought of as $R$ modulo
-$\I$:
+instance, $\I_{ss}$, the snapshot axiom, is stable because if a the
+state is unmodified between $\stg$ and $\stg''$, then it is clearly
+unmodified between $\stg$ and $\stg'$, and also between $\stg'$ and
+$\stg''$, where $\stg'$ is an intermediary state.  Modifying and
+restoring the state $\stg$ is not possible because each new commit
+bears a new transaction id different from the transaction ids (\C{txn}
+fields) present in $\stg$.
+
+The stability condition on $\I$ guarantees that an interference from
+$R^*$ is admissible only if the interference due to each individual
+$R$ step is admissible. In other words, it makes isolation-constrained
+$R^*$ relation equal to the transitive closure of the
+isolation-constrained $R$ relation. We call $R$ constrained by
+isolation $\I$ as $R$ modulo $\I$ ($R\backslash \I$; written shortly
+as $\R$), which is the following ternary relation:
 \begin{smathpar}
   \begin{array}{lcl}
-\stable((R \backslash \I), \phi) & \Leftrightarrow & \forall \stl,\stg,\stg'. \phi(\stl,\stg) \wedge (R \backslash \I)(\stl,\stg,\stg') \Rightarrow \phi(\stl,\stg')\\
-(R \backslash \I)(\stl, \stg, \stg') & \Leftrightarrow & \exists \stg_1. R(\stg, \stg') \wedge \I(\stl, \stg_1, \stg) \wedge \I(\stl, \stg_1, \stg')
+    (R \backslash \I)(\stl, \stg, \stg') & \Leftrightarrow & R(\stg,
+      \stg') \wedge \I(\stl, \stg, \stg')\\
+  \end{array}
+\end{smathpar}
+It is now enough to prove the stability of an RG assertion $\phi$
+w.r.t $R\backslash \I$:
+\begin{smathpar}
+  \begin{array}{lcl}
+    \stable((R \backslash \I), \phi) & \Leftrightarrow & \forall
+      \stl,\stg,\stg'.~ \phi(\stl,\stg) \wedge (R \backslash \I)(\stl,\stg,\stg') \Rightarrow \phi(\stl,\stg')\\
   \end{array}
@@
 executed against a snapshot of the database (i.e. $\mathbb{I}_{ss}$),
-$R \backslash \I_{ss}$ will be the identity function, since if $\stg
-\neq \stg'$, there does not exist any $\stg_1$ which satisfies the
-constraint for $(R \backslash \I_{ss})(\stl,\stg,\stg')$.
+$R \backslash \I_{ss}$ will be the identity function, since any
+non-trivial interference will violate the $\stg' = \stg$ condition
+imposed by $\I_{ss}$.
 
@@
 satisfying $I$, and also guarantees that its commit step satisfies
-$G$. The rule uses the transaction-local RG judgement $\R_e \vdash
+$G$. The rule uses the transaction-local RG judgment $\R_e \vdash
 \rg{P}{c}{Q}$. By \rulelabel{E-Txn-Start}, the local and global
@@
 $\R_e \vdash \rg{P}{c}{Q}$ are in $\R_e$, while the top-level
-judgement requires environment steps in $R$. We show that it is enough
-to consider only those enviroment steps in $\R_e$. First, we use an
+judgment requires environment steps in $R$. We show that it is enough
+to consider only those environment steps in $\R_e$. First, we use an
 inductive argument and stability of $\I_e$ ($\stable(R, \I_e)$) to
@@
 \rulelabel{E-Txn}). Since $\R_e$ contains exactly those environment
-steps which preserve $\I_e$, the local-level RG judgement can be
+steps which preserve $\I_e$, the local-level RG judgment can be
 soundly used, which guarantees that after the transaction finishes its
@@
 transaction and its commit step can modify the global state, and hence
-we also require $Q$ to be stable againt $R$. Again, we use an
+we also require $Q$ to be stable against $R$. Again, we use an
 inductive argument, the stability of $\I_c$, and the fact that the

### 2017-10-22 motivation.tex
Typos and cosmetic changes.

@@
 can complicate high-level reasoning.  A possible alternative would be
-to utilize concurrent program verfication methods where the
+to utilize concurrent program verification methods where the
 implementation details. of weak isolation are interleaved within the
@@
 the program itself. In particular, SQL statements become set
-operations, and the \C{foreach} combinator becoms set monad's bind
+operations, and the \C{foreach} combinator becomes set monad's bind
 ($\bind$) combinator.  As we demonstrate, the advantage of inferring

### 2017-10-22 reasoning.tex
Edit.

@@
 unfortunately, that we cannot directly replace $R^{*}$ by $R$ in the
-above condition because of our need to consider the weak isolation
-specification $\I$ that selectively exposes interferences.
+above condition.
 

### 2017-10-22 isolation.tex

@@
 % \end{smathpar}
-Note that the same axiomatization applies to Postgres's RR,
+Note that the same axiomatization applies to PostgreSQL's RR,
 although its implementation (described in Sec.~\ref{sec:motivation})
 differs considerably from the original proposal. Thus, reasoning done
-for an SI transaction on MS SQL server carries over to Postgres's RR
+for an SI transaction on MS SQL server carries over to PostgreSQL's RR
 and vice-versa, demonstrating the benefits of reasoning axiomatically

### 2017-10-22 popl-rebuttal.md
Rebuttal updated with proof.

@@
   $\delta$ and $\Delta$ forall $z$). While independence on $\Delta$ is
-  gauranteed by the stability condition (checked separately), the
+  guaranteed by the stability condition (checked separately), the
   independence on the local state $\delta$ is simply assumed.
@@
   if it can be simplified. In this we were successful as we were able
-  to modify our encoding to eliminate the existential altogether. Our
+  to modify our bind encoding to eliminate the existential altogether. Our
   modifications are based on the observation that the semantics of
@@
   single bi-implication that only involves universal quantifiers. A
-  rigorous proof to this effect can be found in the techreport, while
-  an automatic Z3-assisted proof can be found
-  [here](https://rise4fun.com/Z3/G1a). As a consequence of this
-  change, the fragment of encoding is back to being EPR, hence the
-  Theorems 5.2 and 5.3 continue to hold.
+  formal proof to this effect is given below, while an automatic
+  Z3-assisted proof can be found [here](https://rise4fun.com/Z3/G1a).
+  As a consequence of this change, the fragment of encoding is back to
+  being EPR, hence the Theorems 5.2 and 5.3 continue to hold. We have
+  also tightened our treatment of free variables in the encoding
+  algorithm, and added an example to Sec. 5.2 to demonstrate how
+  exactly the encoding works.
+
+#### Bind Theorem and Proof ####
+
+Theorem:
+  Let S be a set containing elements of type T, and let R be a binary
+  relation over T. Let S1 be a T set defined using the following two
+  implications (note: (x,y: T) is a shorthand for (x:T, y:T)):
+
+           ∀(x,y: T). x∈S ∧ R(x,y) ⇒ y∈S1			(Call it H0)
+           ∀(y: T). y∈S1 ⇒ ∃(x:T). x∈S ∧ R(x,y)		(H1)
+
+  Likewise, let S2 be a T set defined via a single bi-implication as
+  following:
+
+          ∀(x,y: T). x∈S ∧ R(x,y) ⇔ y∈S2			(H2)
+
+  Then, S1 = S2. That is:
+
+          ∀(y: T). y∈S1 ⇔ y∈S2        (G)
+
+Proof:
+  We will show G be proving the following two implications (goals)
+  separately:
+
+        ∀(y: T). y∈S1 ⇒ y∈S2					(G0)
+        ∀(y: T). y∈S2 ⇒ y∈S1					(G1)
+
+  Let us first note that H2 is a conjunction of the following two
+  hypothesis:
+
+          ∀(x,y: T). x∈S ∧ R(x,y) ⇒ y∈S2			(H3)
+          ∀(x,y: T). y∈S2 ⇒ x∈S ∧ R(x,y)			(H4)
+
+  Proof of G0:
+  Intros on the goal G0 introduces a new y0:T, gives us the following
+  hypothesis:
+
+          y0∈S1							(H5)
+
+  And reduces the goal to proving that y0∈S2.  Instantiating H1 with
+  (y:=y0), and then applying it on H5 gives:
+
+          ∃(x:T). x∈S ∧ R(x,y0)				(H6)
+
+  Destructing the existential in H6 introduces a new x0:T, and gives the
+  following:
+
+          x0∈S ∧ R(x0,y0)				(H7)
+
+  Now, instantiating H3 with (x:=x0)(y:=y0), and applying it on H7 gives
+  y0∈S2, which is what needs to be proven.
+
+  Proof of G1:
+  Intros on the goal G1 introduces a new y0:T, gives us the following
+  hypothesis:
+
+          y0∈S2					(H8)
+
+  And reduces the goal to proving that y0∈S1. At this point we prove a
+  simple lemma that asserts the non-emptiness of type T:
+
+          ∃(x:T)			(Lemma1)
+
+  The proof for Lemma1 requires an evidence for a value of type T, which
+  is readily available in the form of y0:T.  Destructing the existential
+  in Lemma1 introduces a new x0:T into the context (note that x0 may or
+  may not be equal to y0). Now instantiating H4 with (x:=x0)(y:=y0), and
+  then applying it on H8 gives:
+
+          x0∈S ∧ R(x0,y0)				(H9)
+
+  Finally, instantiating H0 with (x:=x0)(y:=y0), and applying it on H9
+  gives y0∈S1, which is what needs to be proven.
+Qed.

### 2017-10-22 popl-rebuttal.md
Edits.

@@
 
-  And reduces the goal to proving that y0∈S1. At this point we prove a
-  simple lemma that asserts the non-emptiness of type T:
+  And reduces the goal to proving that y0∈S1. Instantiating H4 with
+  (x:=y0)(y:=y0), and then applying it on H8 gives:
 
-          ∃(x:T)			(Lemma1)
+          y0∈S ∧ R(y0,y0)				(H9)
 
-  The proof for Lemma1 requires an evidence for a value of type T, which
-  is readily available in the form of y0:T.  Destructing the existential
-  in Lemma1 introduces a new x0:T into the context (note that x0 may or
-  may not be equal to y0). Now instantiating H4 with (x:=x0)(y:=y0), and
-  then applying it on H8 gives:
-
-          x0∈S ∧ R(x0,y0)				(H9)
-
-  Finally, instantiating H0 with (x:=x0)(y:=y0), and applying it on H9
+  Finally, instantiating H0 with (x:=y0)(y:=y0), and applying it on H9
   gives y0∈S1, which is what needs to be proven.

### 2017-10-22 popl-rebuttal.md
edits

--- 2017-10-20 popl-rebuttal.md
@@
 
-Shepard
-----------
-- The primary concern raised by reviewer B is that the fragment of FOL
+Shepherd ----------
+
+Dear Peter -
+
+Please find attached a revised near-final version of our POPL
+submission.  As per your guidance, while we have addressed the
+comments from reviewers (A) and (C), we have taken special care
+to consider the issues raised by reviewer (B).
+
+The primary concern raised by the reviewer is that the fragment of FOL
 in which the translation algorithm (Fig. 10, Sec.5) encodes set
-  expressions is not EPR, hence its decidability is doubtful. In
-  particular, the concern was that an existential quantifier follows a
-  universal quantifier in the logical encoding of bind (`>>=`), and
-  this pattern is not admitted by EPR. In the rebuttal phase we
-  addressed the concern by pointing out that the fragment, while not
-  EPR as claimed, is nonetheless decidable since it is subsumed by the
-  GKS fragment, which is known to be decidable. However, as the
-  reviewer rightly points out, there are no known results on the
-  combination of GKS and Simple Linear Arithmetic (SLA) to support
-  theorem 5.3, so SLA invariants, which are common in databases, might
-  make the encoding undecidable. While this concern can be addressed
-  by considering machine (finite-bit) arithmetic instead of SLA, we
-  decided to take a long, hard look at the encoding algorithm and see
-  if it can be simplified. In this we were successful as we were able
-  to modify our encoding to eliminate the existential altogether. Our
-  modifications are based on the observation that the semantics of
-  bind, which were previously written using two implications one of
-  which involves an existential, can be equivalently written using a
-  single bi-implication that only involves universal quantifiers. A
-  rigorous proof to this effect can be found in the techreport, while
-  an automatic Z3-assisted proof can be found
-  [here](https://rise4fun.com/Z3/G1a). As a consequence of this
-  change, the fragment of encoding is back to being EPR, hence the
-  Theorems 5.2 and 5.3 continue to hold.
+expressions is not EPR, and hence raises doubts about its
+decidability.  In particular, the concern was that an existential
+quantifier follows a universal quantifier in the logical encoding of
+bind (`>>=`), and this pattern is not admitted by EPR. In our
+rebuttal, we addressed the concern by pointing out that the fragment,
+while not EPR as claimed, is nonetheless decidable since it is
+subsumed by GKS (Godel-Kalmar-Schutte), which is known to be
+decidable. However, as the reviewer correctly points out, there are no
+known results on the combination of GKS and Simple Linear Arithmetic
+(SLA) to support theorem 5.3, so SLA invariants, which are common in
+databases, might still make the encoding undecidable.  While this
+concern can be addressed by considering machine (finite-bit)
+arithmetic instead of SLA, we decided a better approach would be
+revisit our encoding algorithm to see if it could be simplified.  We
+were successful in this effort and were able to modify our encoding to
+eliminate the existential altogether. Our modifications are based on
+the observation that the semantics of bind, which were previously
+written using two implications one of which involves an existential,
+can be equivalently written using a single bi-implication that only
+involves universal quantifiers. A proof of the equivalence between the
+two encodings is given below; an automatic Z3-assisted proof, can be
+found [here](https://rise4fun.com/Z3/G1a). As a consequence of this
+change, the fragment of encoding is now unambiguously in EPR, and
+hence Theorems 5.2 and 5.3 hold without the need for any modification
+or qualification.  We believe this new development addresses the main
+concerns of reviewer B, and trust that you will consider the revision
+suitable for acceptance.
+
+Sincerely,
+
+Gowtham (for Kartik, Mahsa, and Suresh)
+
+---------------------------------------

### 2017-10-22 popl-rebuttal.md
Added a note to address reviewer C.

--- 2017-10-22 popl-rebuttal.md
@@
 
+- We added the following note to the end of the first paragraph of
+  Sec. 3 to address reviewer C's comment about heap manipulating
+  programs: "Our language ensures that all effectful actions are
+  encapsulated within database commands, and that all shared state
+  among processes are only manipulated via transactions and its
+  supported operations.  In particular, we do not consider programs in
+  which objects resident on e.g., the OCaml heap are concurrently
+  manipulated by OCaml expressions as well as database actions."
+
 #### Bind Theorem and Proof ####

### 2017-10-22 popl-rebuttal.md

--- 2017-10-22 popl-rebuttal.md
@@
 
-Shepherd ----------
+Shepherd
+----------
 
@@
 Please find attached a revised near-final version of our POPL
-submission.  As per your guidance, while we have addressed the
-comments from reviewers (A) and (C), we have taken special care
-to consider the issues raised by reviewer (B).
+submission.  Besides numerous minor changes throughout in response to
+general themes raised in the reviews (e.g., the issue of of heap state
+raised by reviewer (C) is addressed in Sec. 3 (last two sentences of
+para. 1)), the most significant change from the original submission
+deal with issues raised specifically by reviewer (B), which is
+consistent with your guidance as well.
 
@@
 ---------------------------------------
+
+Theorem:
+  Let S be a set containing elements of type T, and let R be a binary
+  relation over T. Let S1 be a T set defined using the following two
+  implications (note: (x,y: T) is a shorthand for (x:T, y:T)):
+
+           ∀(x,y: T). x∈S ∧ R(x,y) ⇒ y∈S1			(Call it H0)
+           ∀(y: T). y∈S1 ⇒ ∃(x:T). x∈S ∧ R(x,y)		(H1)
+
+  Likewise, let S2 be a T set defined via a single bi-implication as
+  following:
+
+          ∀(x,y: T). x∈S ∧ R(x,y) ⇔ y∈S2			(H2)
+
+  Then, S1 = S2. That is:
+
+          ∀(y: T). y∈S1 ⇔ y∈S2        (G)
+
+Proof:
+  We will show G by proving the following two implications (goals)
+  separately:
+
+        ∀(y: T). y∈S1 ⇒ y∈S2					(G0)
+        ∀(y: T). y∈S2 ⇒ y∈S1					(G1)
+
+  Let us first note that H2 is a conjunction of the following two
+  hypothesis:
+
+          ∀(x,y: T). x∈S ∧ R(x,y) ⇒ y∈S2			(H3)
+          ∀(x,y: T). y∈S2 ⇒ x∈S ∧ R(x,y)			(H4)
+
+  Proof of G0:
+  Intros on the goal G0 introduces a new y0:T, gives us the following
+  hypothesis:
+
+          y0∈S1							(H5)
+
+  And reduces the goal to proving that y0∈S2.  Instantiating H1 with
+  (y:=y0), and then applying it on H5 gives:
+
+          ∃(x:T). x∈S ∧ R(x,y0)				(H6)
+
+  Destructing the existential in H6 introduces a new x0:T, and gives the
+  following:
+
+          x0∈S ∧ R(x0,y0)				(H7)
+
+  Now, instantiating H3 with (x:=x0)(y:=y0), and applying it on H7 gives
+  y0∈S2, which is what needs to be proven.
+
+  Proof of G1:
+  Intros on the goal G1 introduces a new y0:T, gives us the following
+  hypothesis:
+
+          y0∈S2					(H8)
+
+  And reduces the goal to proving that y0∈S1. Instantiating H4 with
+  (x:=y0)(y:=y0), and then applying it on H8 gives:
+
+          y0∈S ∧ R(y0,y0)				(H9)
+
+  Finally, instantiating H0 with (x:=y0)(y:=y0), and applying it on H9
+  gives y0∈S1, which is what needs to be proven.
+Qed.

### 2017-10-22 popl-rebuttal.md

@@
 submission.  Besides numerous minor changes throughout in response to
-general themes raised in the reviews (e.g., the issue of of heap state
+general themes raised in the reviews (e.g., the issue of heap state
 raised by reviewer (C) is addressed in Sec. 3 (last two sentences of
-para. 1)), the most significant change from the original submission
+para. 1)), the most significant changes from the original submission
 deal with issues raised specifically by reviewer (B), which is
@@
 in which the translation algorithm (Fig. 10, Sec.5) encodes set
-expressions is not EPR, and hence raises doubts about its
+expressions is not EPR, and hence this raises doubts about its
 decidability.  In particular, the concern was that an existential
@@
 arithmetic instead of SLA, we decided a better approach would be
-revisit our encoding algorithm to see if it could be simplified.  We
-were successful in this effort and were able to modify our encoding to
-eliminate the existential altogether. Our modifications are based on
-the observation that the semantics of bind, which were previously
-written using two implications one of which involves an existential,
-can be equivalently written using a single bi-implication that only
-involves universal quantifiers. A proof of the equivalence between the
-two encodings is given below; an automatic Z3-assisted proof, can be
-found [here](https://rise4fun.com/Z3/G1a). As a consequence of this
-change, the fragment of encoding is now unambiguously in EPR, and
-hence Theorems 5.2 and 5.3 hold without the need for any modification
-or qualification.  We believe this new development addresses the main
-concerns of reviewer B, and trust that you will consider the revision
-suitable for acceptance.
+revisit our encoding algorithm to see if it could be simplified.
+
+We are pleased to report that we were successful in this effort and
+were able to modify our encoding to eliminate the existential
+altogether. Our modifications are based on the observation that the
+semantics of bind, which were previously written using two
+implications one of which involves an existential, can be equivalently
+written using a single bi-implication that only involves universal
+quantifiers. A proof of the equivalence between the two encodings is
+given below; a mechanized Z3-assisted proof can be found at:
+https://rise4fun.com/Z3/G1a. As a consequence of this change, the
+fragment of encoding is now unambiguously in EPR, and hence Theorems
+5.2 and 5.3 hold without the need for any modification or
+qualification.  We believe this new development addresses the main
+concerns of reviewer B, and consequently trust that you will consider
+the revision suitable for acceptance.
 

### 2017-10-22 inference.tex

@@
 body to the level of the loop. Since records added to $\stl$ in each
-iteration of \C{FOREACH} are independent of the previous iteration,
-sequential composition of the effects of different iterations is same
+iteration of \C{FOREACH} are independent of the previous iteration (the
+rule assumes local context independence as discussed in \S\ref{sec:opsem}),
+the sequential composition of the effects of different iterations is the same
 as their parallel composition. Since the loop body is executed once
@@
 z.~F_{body}(\stl,\stg))$, which mirrors the definition of the rule.
-This development mirrors ideas found in type checkers for certain
-dependent type systems~\cite{KJ14}.
+%This development mirrors ideas found in type checkers for certain
+%dependent type systems~\cite{KJ14}.
 

### 2017-10-22 introduction.tex

@@
 question facing all database programmers.\footnote{This position has
-been echoed by leading database researchers who lament the lack of a
+been echoed by database researchers who lament the lack of a
 better understanding of this problem; see e.g., {\tt

### 2017-10-22 opsem.tex
Typos and cosmetic changes.

@@
 reduction rules effectively allows reads of uncommitted transaction
-writes to be realized.
+writes to be realized, if so desired.
 
@@
 conditionals, \C{LET} binders, and sequences are standard, and
-ommitted for brevity.
+omitted for brevity.
 
@@
 presented in the next section; our inference procedure described in
-\S\ref{sec:inference}, however, has a non-trivial dependence on
-this assumption. Nonetheless, we have encountered few instances in
+\S\ref{sec:inference}, however, has a non-trivial dependence on this
+assumption.  Nonetheless, we have encountered few instances in
 practice where enforcing local context independence turns out to be a
-severe restriction. Indeed, all 17 transactions in the TPC-C and TPC-E
-benchmarks satisfy this assumption.
+severe restriction. Indeed, all of the transactions we have considered
+in our benchmarks (e.g., TPC-C) satisfy this assumption.
 

### 2017-10-22 inference.tex
Minor edits.

--- 2017-10-21 inference.tex
@@
 (quantifier-free) formula on $x$, a restricted form of existential
-quantification that binds a set $x$ satisfying proposition $\phi$ in a
+quantification that binds a set $\stg$ satisfying proposition $\phi$ in a
 set expression $s$, a monadic composition of two set expressions
@@
 body to the level of the loop. Since records added to $\stl$ in each
-iteration of \C{FOREACH} are independent of the previous iteration,
-sequential composition of the effects of different iterations is same
-as their parallel composition. Since the loop body is executed once
-per each $z\in x$, the effect of the the loop is the union of effects
-($\F$) for all $z\in x$, all applied to the same state ($\stl$ and
-$\stg$). That is, $\F_{loop}(\stl,\stg) = \bigcup_{z\in
+iteration of \C{FOREACH} are independent of the previous iteration
+(recall local context independence; Sec. ~\ref{sec:opsem}), sequential
+composition of the effects of different iterations is same as their
+parallel composition. Since the loop body is executed once per each
+$z\in x$, the effect of the the loop is the union of effects ($\F$)
+for all $z\in x$, all applied to the same state ($\stl$ and $\stg$).
+That is, $\F_{loop}(\stl,\stg) = \bigcup_{z\in
 x}\F_{body}(\stl,\stg)$. From the definition of the set monad's bind

### 2017-10-22 introduction.tex
Minor edits.

--- 2017-10-20 introduction.tex
@@
 transactions along with automated verification support to allow
-developers to verify the soundness of their applications, without
+developers verify the soundness of their applications, without
 having to resort to low-level operational reasoning as they are forced
@@
 rules that automatically verifies the weakest isolation level for a
-transaction that still ensures its invariants are maintained.  The key
-to inference is a novel formulation of database state (represented as
+transaction while ensuring its invariants are maintained.  The key to
+inference is a novel formulation of database state (represented as
 sets of tuples) as a monad, and in which database computations are

### 2017-10-22 motivation.tex
Minor edits.

@@
 need to reason about the safety of such interferences with respect to
-\C{new\_order}'s consistency invariants.
+\C{new\_order}'s consistency invariants to verify \C{new\_order}.
 
@@
 satisfied, constructs a new set containing a single record that is
-identical to $\C{s}$ except that it binds field $\C{s_{qty}}$ to value
+identical to $\C{s}$ except that it binds field $\C{s\_qty}$ to value
 $\C{k_1}$.  This new set is added (via set union) to the existing

### 2017-10-22 inference.tex

@@
 effects inferred for their subcommands. Thus, the effect of a sequence
-of commands $c_1;c_2$ is the union of effects $F_1$ and $F_2$ of $c_1$
-and $c_2$, respectively, except that $F_2$ is applied to a state
-($\stl$) to which $F_1$ has already been applied, reflecting their
+of commands $c_1;c_2$ is the union of effects $\F_1$ and $\F_2$ of $c_1$
+and $c_2$, respectively, except that $\F_2$ is applied to a state
+($\stl$) to which $\F_1$ has already been applied, reflecting their
 order of reduction. The inference rule for \C{FOREACH} takes advantage
@@
 can infer that each iteration corresponds to the application of
-$F$. Since all iterations are assumed to be independent of each other,
+$\F$. Since all iterations are assumed to be independent of each other,
 and $z$ is bound to a record in $x$ for each iteration, we conclude
@@
    \R_e = R \backslash \I_e \spc \R_c = R \backslash \I_c \spc  
-   c \elabsto \F \spc \stable(\R_c,\F)\\
+   c \Longrightarrow_{\langle i,\R_e,I \rangle}\F \spc \stable(\R_c,\F)\\
    \forall \stl,\stg.~ \stl = F(\emptyset,\stg) \Rightarrow 

### 2017-10-22 reasoning.tex
Minor edits.

@@
 isolation-constrained $R$ relation. We call $R$ constrained by
-isolation $\I$ as $R$ modulo $\I$ ($R\backslash \I$; written shortly
+isolation $\I$ as $R$ modulo $\I$ ($R\backslash \I$; written equivalently
 as $\R$), which is the following ternary relation:
@@
 \noindent We can now assert the soundness of the RG judgments in
-Fig.~\ref{fig:rg-rules} as follows:
+Fig.~\ref{fig:rg-rules} as follows\footnote{Full proofs for theorems
+in this paper can be found in the accompanying tech report}:
 

### 2017-10-26 inference.tex
Encoding to GKS.

@@
 transaction from Sec.~\ref{sec:motivation}. Recall that the state
-transformer for the \C{foreach} loop shown in
+transformer ($\T$) for the \C{foreach} loop shown in
 Fig.~\ref{fig:foreach_code} is (\C{k1}, \C{k2}, and \C{k3} are
@@
   \begin{array}{lcl}
-    \phi_U & = & \forall \C{item\_req}.\forall s. \forall s'.~  \\
-    & & \hspace*{0.1in}\pi(\C{item\_req}) \Leftrightarrow(s \in \stg) \conj 
+    \phi_U & = & \hspace*{0.2in}\forall \C{item\_req}.\forall s. \forall s'.~ 
+        \pi_1(\C{item\_req}) \Leftrightarrow \\
+    & & \hspace*{0.3in}(s \in \stg) \conj 
         (\itec{{\sf table}(s)=\C{Stock} \wedge 
@@
                             \C{s\_qty} = \C{k1}\rangle\\
-    & & \hspace*{1.45in}} {\bot}) \\
-    & & \hspace*{1.2in} \Leftrightarrow g(\C{item\_req},s')\\
-    \G_U & = & \lambda (\C{item\_req},r).~ \pi(\C{item\_req}) 
-        \wedge g(\C{item\_req},r)\\
+    & & \hspace*{1.45in}} {\bot}) \Rightarrow g_0(\C{item\_req},s')\\
+    & & \conj \hspace*{0.08in}\forall \C{item\_req}.\forall s'. \exists s.~ 
+        \pi_2(\C{item\_req}) \Leftrightarrow \\
+    & & \hspace*{0.28in} g_0(\C{item\_req},s') \Rightarrow (s \in \stg) \conj
+        (\itec{{\sf table}(s)=\C{Stock} \wedge 
+              s.\C{s\_i\_id} = \C{item\_req.ol\_i\_id}\\
+    & & \hspace*{1.8in}}
+              {s' = \langle \C{s\_i\_id}=s.\C{s\_i\_id};\, 
+                            \C{s\_d\_id}=s.\C{s\_d\_id};\,
+                            \C{s\_qty} = \C{k1}\rangle\\
+    & & \hspace*{1.8in}} {\bot}) \\
+    \G_U & = & \lambda (\C{item\_req},r).~ \pi_1(\C{item\_req}) \wedge 
+        \pi_2(\C{item\_req}) \wedge g_0(\C{item\_req},r)\\
         \phi_I & = & \top \\ 
-    \G_U & = & \lambda (\C{item\_req},r).~ 
+    \G_I & = & \lambda (\C{item\_req},r).~ 
         r = \langle \C{ol\_o\_id}=\C{k2},\,
@@
              \rangle\\
+  \end{array}
+\end{smathpar}
+Since the transformer ($\T$) of the \C{foreach} loop is not nested
+does not contain any free iteration variables, the appropriate
+application of the encoding algorithm  is
+$\mssemof{\emptyset}{\T(\stl,\stg)}$, which results in the
+$(\phi,\G)$, as defined below:
+\begin{smathpar}
+  \begin{array}{lcl}
+    \phi & = & \phi_I \conj \phi_U \conj \\
+    & & \hspace*{0.2in}\forall \C{ireq}. \forall s.~ 
+        \pi_3 \Leftrightarrow (\C{ireq} \in \C{item\_reqs} \conj 
+        (\G_U(\C{ireq},s') \vee \G_I(\C{ireq},s'))
+        \Rightarrow g_1(s))\\
+    & & \conj \hspace*{0.05in}\forall s.\exists \C{ireq}. ~ 
+        \pi_4 \Leftrightarrow (g_1(s) \Rightarrow (\C{ireq} \in \C{item\_reqs} 
+        \conj (\G_U(\C{ireq},s') \vee \G_I(\C{ireq},s')))\\
+    \G & = & \lambda(r).~\pi_3 \wedge \pi_4 \wedge g_1(r)\\
   \end{array}

### 2017-10-26 inference.tex
edits

@@
 are also variables in $\SL$, but are used to denote local and database
-states, respectively. Constant sets can be written using the set
-literal expressions. For e.g., the set $\{1,2\}$ can be written as
+states, respectively. Constant sets can be written using set
+literal expressions. For example, the set $\{1,2\}$ can be written as
 $\{x \,|\, x=1 \vee x=2\}$. The language is carefully chosen to be
@@
 well as SQL operations more generally), yet simple enough to have a
-semantics-preserving translation to a decidable fragment of
-first-order logic.
+semantics-preserving translation amenable for automated verification.
 \input{inference-rules}
@@
 transaction $\C{TXN}_i$.  Inference depends on the $\I$-constrained
-rely relation $\R$, and the high-level invariant $I$ for the reasons
+rely relation $\R$, and the high-level invariant $I$ for reasons
 that will become clear shortly.  Inference rules compute, for each
@@
 Intuitively, the stability condition requires an effect to describe
-the same set of records before and after the interference. The
-equivalence in $\SL$ translates to equality in first-order logic,
-as we describe below. In the inference rules, stability is
-enforced constructively by a meta function $\stabilize{\cdot}$, which
-accepts an effect and returns a new effect that is guaranteed to be
-stable under $\R$.  $\stabilize{\cdot}$ achieves the stability
-guarantee by abstracting away the bound $\stg$ in an unstable $\F$ to
-an existentially bound $\stg'$ as described below:
+the same set of records before and after an interference. The
+equivalence in $\SL$ translates to equality in first-order logic, as
+we describe below. In the inference rules, stability is enforced
+constructively by a meta function $\stabilize{\cdot}$, which accepts
+an effect and returns a new effect that is guaranteed to be stable
+under $\R$.  $\stabilize{\cdot}$ achieves the stability guarantee by
+abstracting away the bound global state ($\stg$) in an unstable $\F$
+to an existentially bound $\stg'$ as described below:
 \begin{smathpar}
@@
 \end{smathpar}
-Observe that when $\F$ is not stable, then $\stabilize{\F}$ returns a
+Observe that when $\F$ is not stable, $\stabilize{\F}$ returns a
 transformer $\F'$ that simply ignores its $\stg$ argument in favor of
 a generic $\stg'$, making $\F'$ trivially stable. It is safe to assume
-$I(\stg')$ because all verified transactions preserve the invariant,
-and hence only valid database states will ever be witnessed. From the
-perspective of RG reasoning, $\stabilize{\cdot}$ effectively weakens
-the post-condition of a statement, as done by the
+$I(\stg')$ because all verified transactions must preserve the
+invariant, and hence only valid database states will ever be
+witnessed. From the perspective of RG reasoning, $\stabilize{\cdot}$
+effectively weakens the post-condition of a statement, as done by the
 \rulelabel{RG-Conseq} rule for transaction-bound commands.  The
 weakening semantics chosen by $\stabilize{\cdot}$, while being simple,
-is nonetheless useful because of the $I(\stg')$ assumption on the
-existentially bound $\stg'$. The example in
+is nonetheless useful because of the $I(\stg')$ assumption imposed on
+the existentially bound $\stg'$. The example in
 Fig.~\ref{fig:weakening-example} demonstrates.
@@
 effects inferred for their subcommands. Thus, the effect of a sequence
-of commands $c_1;c_2$ is the union of effects $\F_1$ and $\F_2$ of $c_1$
-and $c_2$, respectively, except that $\F_2$ is applied to a state
-($\stl$) to which $\F_1$ has already been applied, reflecting their
-order of reduction. The inference rule for \C{FOREACH} takes advantage
-of the $\SL$'s bind operator to lift the effect inferred for the loop
-body to the level of the loop. Since records added to $\stl$ in each
-iteration of \C{FOREACH} are independent of the previous iteration
-(recall local context independence; Sec. ~\ref{sec:opsem}), sequential
-composition of the effects of different iterations is same as their
+of commands $c_1;c_2$ is the union of effects $\F_1$ and $\F_2$ of
+$c_1$ and $c_2$, respectively, except that $\F_2$ is applied to a
+state ($\stl$) to which $\F_1$ has already been applied, reflecting
+their order of reduction. The inference rule for \C{FOREACH} takes
+advantage of the $\SL$'s bind operator to lift the effect inferred for
+the loop body to the level of the loop. Since records added to $\stl$
+in each iteration of \C{FOREACH} are independent of the previous
+iteration (recall that we make a local context independence assumption
+about database programs; Sec. ~\ref{sec:opsem}), sequential
+composition of the effects of different iterations is the same as their
 parallel composition. Since the loop body is executed once per each
-$z\in x$, the effect of the the loop is the union of effects ($\F$)
+$z\in x$, the effect of the the loop is a union of effects ($\F$)
 for all $z\in x$, all applied to the same state ($\stl$ and $\stg$).
@@
 ($\top$) if and only if $x$ is present in the set. Alternatively,
-translation may also encode the set as a predicate in the logic
+the translation may also encode the set as a predicate in the logic
 itself, in which case a quantified proposition constraining the
@@
 $(\phi,\G)$.  The meta function $\G$ is a predicate indexed by the
-(possible) free variables of $s$, thus, its arity is $|\nubar| + 1$.
+(possible) free variables of $s$, and thus its arity is $|\nubar| + 1$.
 Note that $\nubar$ is only a sequence of variables introduced by the
@@
 propagated upwards as the constraint of the conditional expression.
-Set union expression is encoded similarly.
+A set union expression is encoded similarly.
 

### 2017-10-26 inference.tex
GKS decidability.

--- 2017-10-26 inference.tex
@@
               s.\C{s\_i\_id} = \C{item\_req.ol\_i\_id}\\
-    & & \hspace*{1.45in}}
+    & & \hspace*{0.85in}}
               {s' = \langle \C{s\_i\_id}=s.\C{s\_i\_id};\, 
@@
                             \C{s\_qty} = \C{k1}\rangle\\
-    & & \hspace*{1.45in}} {\bot}) \Rightarrow g_0(\C{item\_req},s')\\
+    & & \hspace*{0.85in}} {\bot}) \Rightarrow g_0(\C{item\_req},s')\\
     & & \conj \hspace*{0.08in}\forall \C{item\_req}.\forall s'. \exists s.~ 
@@
     \phi & = & \phi_I \conj \phi_U \conj \\
-    & & \hspace*{0.2in}\forall \C{ireq}. \forall s.~ 
-        \pi_3 \Leftrightarrow (\C{ireq} \in \C{item\_reqs} \conj 
-        (\G_U(\C{ireq},s') \vee \G_I(\C{ireq},s'))
+    & & \hspace*{0.2in}\forall \C{item\_req}. \forall s.~ 
+        \pi_3 \Leftrightarrow (\C{item\_req} \in \C{item\_reqs} \conj 
+        (\G_U(\C{item\_req},s') \vee \G_I(\C{item\_req},s'))
         \Rightarrow g_1(s))\\
-    & & \conj \hspace*{0.05in}\forall s.\exists \C{ireq}. ~ 
-        \pi_4 \Leftrightarrow (g_1(s) \Rightarrow (\C{ireq} \in \C{item\_reqs} 
-        \conj (\G_U(\C{ireq},s') \vee \G_I(\C{ireq},s')))\\
+    & & \conj \hspace*{0.05in}\forall s.\exists \C{item\_req}. ~ 
+        \pi_4 \Leftrightarrow (g_1(s) \Rightarrow (\C{item\_req} \in \C{item\_reqs} 
+        \conj (\G_U(\C{item\_req},s') \vee \G_I(\C{item\_req},s')))\\
     \G & = & \lambda(r).~\pi_3 \wedge \pi_4 \wedge g_1(r)\\
@@
 Observe that the encoding shown in Fig.~\ref{fig:logic} maps to
-a subset of logic that satisfies the following syntactic properties:
+a fragment of FOL that satisfies the following syntactic properties:
 \begin{itemize}
+  \item All function symbols, modulo those that might appear in
+    $\varphi\in\Prop^0$, are uninterpreted and Boolean.
   \item All quantification is first-order; second-order objects, such
     as sets and functions, are never quantified.
-  \item All quantifiers are universal, and appear only at the prenex
-    position, i.e., at the beginning of a quantified formula.  
-  \item All function symbols, modulo those that might appear in
-    $\varphi\in\Prop^0$, are uninterpreted and Boolean.
+  \item Quantifiers appear only at the prenex position, i.e., at the
+    beginning of a quantified formula.
 \end{itemize}
-This fragment of FOL known as EPR (Effectively Propositional logic) is
-known to be decidable~\cite{z3epr}. The language of encoding, however,
-is a combination of {\sf EPR} with (a). $\Prop^0$, the theory from
-which quantifier-free propositions ($\varphi$) that encode object
-language expressions are drawn, and (b).  $\Prop^1$, the theory from
-which invariants ($I$) are drawn. We write $\SL[\Prop^0,\Prop^1]$ to
-highlight the parameterization of $\SL$ on $\Prop^0$ and $\Prop^1$,
-and state the following theorem:
+The simple syntactic structure of the fragment already makes is
+amenable for automatic reasoning via an off-the-shelf SMT solver, such
+as Z3. The decidability of this fragment, however, is more subtle and
+discussed below.
+
+Consider a set expression $s$ with no free variables (i.e., $\nubar =
+\emptyset$, like $\T(\stl,\stg)$ from the above example). Let
+$(\phi,G) = \mssemof{\emptyset}{s}$. Note that $\phi$ is a conjunction
+of (a).  $\phi_i$'s, where each $\phi_i$ results from encoding a
+subexpression $s_i$ of $s$, and (b). a $\phi_s$, resulting from
+encoding $s$ itself (i.e., its top-level expression). From Fig.~\ref{fig:logic},
+it is clear that $\phi_s$ is either $\top$ (for the first four cases),
+or it is a prenex-quantified formula, where quantification is either
+$\forall^2$, or $\exists$, or $\forall\exists$. Generalizing this
+observation, for a set expression $s$ with $|\nubar|$ free variables,
+$\phi_s$, if quantified, is a prenex-quantified formula, where
+quantification assumes one among the forms of $\forall^{|\nubar|+2}$,
+or $\forall^{|\nubar|}\exists$, or $\forall^{|\nubar|+1}\exists$.  In
+other words, the number of $\forall$ quantifiers preceding an
+$\exists$ quantifier is utmost two more than the number of free
+variables ($\nubar$) in $s$. For the convenience of this discussion,
+let us call $\forall^{|\nubar|+1}\exists$ as the prenex signature of
+$\phi_s$.
+
+Next, in Fig.~\ref{fig:logic}, observe that the (ordered) set $\nubar$
+is extended only in the encoding rules for $\bind$ and $\existsl$
+expression. Since $\existsl$ is a result of a particular choice of the
+weakening semantics (Sec.~\ref{sec:inference}), let us ignore it to
+focus on the core set expression language. This leaves $\bind$ as the
+only case where $\nubar$ is extended.  Since an occurance of $\bind$
+adds one to $|\nubar|$, if $s$ is a subexpression of a top-level
+$\bind$ expression (like $\F_U(\stl,\stg)$ from the above example),
+then the prenex signature of $\phi_s$ is $\forall^2\exists$.
+Furthermore, if the subexpressions of $s$ are not $\bind$ expressions,
+then none of the $\phi_i$'s are quantified, and the prenex signature
+of $\phi = \bigwedge_i\phi_i \wedge \phi_s$ remains
+$\forall^2\exists$.  The fragment of FOL that admits formulas with
+prenex signatures of form $\forall^2\exists^*$ is called
+Godel-Kalmar-Schutte ({\sf GKS}) fragment, which is known to be
+decidable. The language of encoding, however, is a combination of {\sf
+GKS} with (a). $\Prop^0$, the theory from which quantifier-free
+propositions ($\varphi$) that encode object language expressions are
+drawn, and (b).  $\Prop^1$, the theory from which invariants ($I$) are
+drawn. Thus the encoding is decidable if the combination of ${\sf GKS}
++ \Prop^0 + \Prop^1$ is decidable.
+
+To formally state the above property, let $\SL[n,\Prop^0,\Prop^1]$
+denote the core set expression language that admits $n$ nested $\bind$
+expressions, and allows program's (boolean) expressions ($e$), and
+invariants ($\I$) to be drawn from theories $\Prop^0$ and $\Prop^1$,
+respectively. Then, we have the following result:
 \begin{theorem}
-  $\SL[\Prop^0,\Prop^1]$ is decidable if ${\sf EPR}+\Prop^0+\Prop^1$
+  $\SL[0,\Prop^0,\Prop^1]$ is decidable if ${\sf GKS}+\Prop^0+\Prop^1$
   is decidable.
 \end{theorem}
-A useful instantiation of $\SL$ is $\SL[{\sf SLA},{\sf EPR}+{\sf
-SLA}]$, where ${\sf SLA}$ is the theory of simple linear arithmetic.
-Since {\sf EPR}+{\sf SLA} is known to be decidable~\cite{eprsla}:
-\begin{theorem}
-  $\SL[{\sf SLA},{\sf EPR}+{\sf SLA}]$ is decidable.
-\end{theorem}
-The $\SL[{\sf SLA},{\sf EPR}+{\sf SLA}]$ instantiation requires $I$ to
-be drawn from {\sf EPR}+{\sf SLA}, which is expressive enough to
-describe common database integrity constraints, such as referential
-integrity, non-negativeness of all integer values in a column etc.
-The isolation specifications presented in \S\ref{sec:isolation} are
-already simple first-order formulas that can be encoded in {\sf EPR}.
-Furthermore, it is also reasonable to expect the guarantee ($G$) of a
-transaction to be expressible in the same logic as its inferred $\F$,
-since $\F$ (without the stability check) is essentially a complete
-characterization of the transaction, while $G$ is only an abstraction.
-Thus, with $\SL[{\sf SLA},{\sf EPR}+{\sf SLA}]$ as the language of
-inference, the verification problem for weakly isolated transactions
-is decidable. Moreover, off-the-shelf SMT solvers (e.g., Z3) are
-equipped with efficient decision procedures for ${\sf EPR}+{\sf SLA}$,
-making automated verification a practical exercise.
+The $n=0$ case denotes the lack of nested bind expressions. Since
+nested \C{foreach} statements are the only program constructs in
+$\txnimp$ that lead to nested bind expressions, the $n=0$ case
+essentially denotes the encoding language for $\txnimp$ programs
+without nested \C{foreach} loops. Observe that the \C{new\_order}
+transaction (Fig.~\ref{fig:new_order_code}) satisfies this constraint.
+Indeed, the restricted language without nested \C{foreach} loops is
+nonetheless expressive enough to cover all TPC-C transactions.
+
+% This
+% fragment of FOL known as EPR (Effectively Propositional
+% logic) is
+% known to be decidable~\cite{z3epr}. The language of encoding, however,
+% is a combination of {\sf EPR} with (a). $\Prop^0$, the theory from
+% which quantifier-free propositions ($\varphi$) that encode object
+% language expressions are drawn, and (b).  $\Prop^1$, the theory from
+% which invariants ($I$) are drawn. We write $\SL[\Prop^0,\Prop^1]$ to
+% highlight the parameterization of $\SL$ on $\Prop^0$ and $\Prop^1$,
+% and state the following theorem:
+% \begin{theorem}
+%   $\SL[\Prop^0,\Prop^1]$ is decidable if ${\sf EPR}+\Prop^0+\Prop^1$
+%   is decidable.
+% \end{theorem}
+% A useful instantiation of $\SL$ is $\SL[{\sf SLA},{\sf EPR}+{\sf
+% SLA}]$, where ${\sf SLA}$ is the theory of simple linear arithmetic.
+% Since {\sf EPR}+{\sf SLA} is known to be decidable~\cite{eprsla}:
+% \begin{theorem}
+%   $\SL[{\sf SLA},{\sf EPR}+{\sf SLA}]$ is decidable.
+% \end{theorem}
+% The $\SL[{\sf SLA},{\sf EPR}+{\sf SLA}]$ instantiation requires $I$ to
+% be drawn from {\sf EPR}+{\sf SLA}, which is expressive enough to
+% describe common database integrity constraints, such as referential
+% integrity, non-negativeness of all integer values in a column etc.
+% The isolation specifications presented in \S\ref{sec:isolation} are
+% already simple first-order formulas that can be encoded in {\sf EPR}.
+% Furthermore, it is also reasonable to expect the guarantee ($G$) of a
+% transaction to be expressible in the same logic as its inferred $\F$,
+% since $\F$ (without the stability check) is essentially a complete
+% characterization of the transaction, while $G$ is only an abstraction.
+% Thus, with $\SL[{\sf SLA},{\sf EPR}+{\sf SLA}]$ as the language of
+% inference, the verification problem for weakly isolated transactions
+% is decidable. Moreover, off-the-shelf SMT solvers (e.g., Z3) are
+% equipped with efficient decision procedures for ${\sf EPR}+{\sf SLA}$,
+% making automated verification a practical exercise.

### 2017-10-26 related-work.tex

--- 2017-10-10 related-work.tex
@@
 visibility, but their specifications are also for complete traces.
-Like ours, all the aforementioned specification frameworks use the
-vocabulary introduced in~\cite{burckhardt14}. However, none of them
-are equipped with a reasoning framework that can use such
-specifications to verify programs under weak isolation.
+Both the aforementioned specification frameworks use the vocabulary
+introduced in~\cite{burckhardt14}. However, none of them are equipped
+with a reasoning framework that can use such specifications to verify
+programs under weak isolation.
 
@@
 is not immediately apparent if their specification formalism is
-amenable for use within a verification toolchain.
+amenable for use within a verification toolchain.  ~\cite{WB17}
+presents a dynamic analysis for weak isolation that attempts to
+discover weak isolation anomalies from SQL log files.  Their solution,
+while capable of identifying database attacks due to the use of
+incorrect (excessively weak) isolation levels, does not consider how
+to verify application correctness, infer proper isolation levels, or
+formally reason about the relationship between weak-isolation levels
+and application invariants.
 

### 2017-10-26 inference.tex
edits

@@
 we describe below. In the inference rules, stability is enforced
-constructively by a meta function $\stabilize{\cdot}$, which accepts
+constructively by a meta-function $\stabilize{\cdot}$, which accepts
 an effect and returns a new effect that is guaranteed to be stable
@@
 free variable. In such cases, the predicate associated with the
-subexpression should also be indexed by its free variables, so that a
+subexpression should also be indexed by its free variables so that a
 unique set exists for each instantiation of the free variables. Thus,
@@
 binder first), $\mssemof{\nubar}{s}$ returns the encoding of $s$ as
-$(\phi,\G)$.  The meta function $\G$ is a predicate indexed by the
+$(\phi,\G)$.  The meta-function $\G$ is a predicate indexed by the
 (possible) free variables of $s$, and thus its arity is $|\nubar| + 1$.
@@
 is extended only in the encoding rules for $\bind$ and $\existsl$
-expression. Since $\existsl$ is a result of a particular choice of the
-weakening semantics (Sec.~\ref{sec:inference}), let us ignore it to
+expressions. Since $\existsl$ is a result of a particular choice of the
+weakening semantics (Sec.~\ref{sec:inference}), let us ignore it for now to
 focus on the core set expression language. This leaves $\bind$ as the
-only case where $\nubar$ is extended.  Since an occurance of $\bind$
+only case where $\nubar$ is extended.  Since an occurrence of $\bind$
 adds one to $|\nubar|$, if $s$ is a subexpression of a top-level
-$\bind$ expression (like $\F_U(\stl,\stg)$ from the above example),
+{\sf bind} expression (like $\F_U(\stl,\stg)$ from the above example),
 then the prenex signature of $\phi_s$ is $\forall^2\exists$.
-Furthermore, if the subexpressions of $s$ are not $\bind$ expressions,
+Furthermore, if the subexpressions of $s$ are not {\sf bind} expressions,
 then none of the $\phi_i$'s are quantified, and the prenex signature
@@
 $\forall^2\exists$.  The fragment of FOL that admits formulas with
-prenex signatures of form $\forall^2\exists^*$ is called
-Godel-Kalmar-Schutte ({\sf GKS}) fragment, which is known to be
+prenex signatures of form $\forall^2\exists^*$ is called the
+Godel-Kalmar-Schutte ({\sf GKS}) fragment~\cite{...}, which is known to be
 decidable. The language of encoding, however, is a combination of {\sf
@@
 To formally state the above property, let $\SL[n,\Prop^0,\Prop^1]$
-denote the core set expression language that admits $n$ nested $\bind$
-expressions, and allows program's (boolean) expressions ($e$), and
+denote the core set expression language that admits $n$ nested {\sf bind}
+expressions, and allows program (Boolean) expressions ($e$), and
 invariants ($\I$) to be drawn from theories $\Prop^0$ and $\Prop^1$,
-respectively. Then, we have the following result:
+respectively. Then:
 \begin{theorem}
@@
 \end{theorem}
-The $n=0$ case denotes the lack of nested bind expressions. Since
-nested \C{foreach} statements are the only program constructs in
-$\txnimp$ that lead to nested bind expressions, the $n=0$ case
-essentially denotes the encoding language for $\txnimp$ programs
+The $n=0$ case denotes the lack of nested {\sf bind}
+expressions. Since nested \C{foreach} statements are the only program
+constructs in $\txnimp$ that lead to nested bind expressions, this
+instantiation denotes the encoding language for $\txnimp$ programs
 without nested \C{foreach} loops. Observe that the \C{new\_order}
 transaction (Fig.~\ref{fig:new_order_code}) satisfies this constraint.
-Indeed, the restricted language without nested \C{foreach} loops is
-nonetheless expressive enough to cover all TPC-C transactions.
+Indeed, this restricted language is nonetheless expressive enough to
+cover all of the benchmarks we have considered.
 

### 2017-10-26 inference.tex
Edits.

--- 2017-10-26 inference.tex
@@
 
+A useful instantiation of $\SL$ is $\SL[{\sf BV},{\sf GKS}+{\sf BV}]$,
+where ${\sf BV}$ is the theory of bit-vector arithmetic, which is
+generally used to encode the finite-bit arithmetic of real programs.
+Bit-vector arithmetic has a finite quantifier-free
+axiomatization, 
+Since {\sf EPR}+{\sf SLA} is known to be decidable~\cite{eprsla}:
+\begin{theorem}
+  $\SL[{\sf SLA},{\sf EPR}+{\sf SLA}]$ is decidable.
+\end{theorem}
+The $\SL[{\sf SLA},{\sf EPR}+{\sf SLA}]$ instantiation requires $I$ to
+be drawn from {\sf EPR}+{\sf SLA}, which is expressive enough to
+describe common database integrity constraints, such as referential
+integrity, non-negativeness of all integer values in a column etc.
+The isolation specifications presented in \S\ref{sec:isolation} are
+already simple first-order formulas that can be encoded in {\sf EPR}.
+Furthermore, it is also reasonable to expect the guarantee ($G$) of a
+transaction to be expressible in the same logic as its inferred $\F$,
+since $\F$ (without the stability check) is essentially a complete
+characterization of the transaction, while $G$ is only an abstraction.
+Thus, with $\SL[{\sf SLA},{\sf EPR}+{\sf SLA}]$ as the language of
+inference, the verification problem for weakly isolated transactions
+is decidable. Moreover, off-the-shelf SMT solvers (e.g., Z3) are
+equipped with efficient decision procedures for ${\sf EPR}+{\sf SLA}$,
+making automated verification a practical exercise.
+
 % This

### 2017-10-26 inference.tex
GKS-related edits.

--- 2017-10-26 inference.tex
@@
 The first-order encoding of a bind expression describes the semantics
-of the set monad's bind operator in FOL. Let $s_1$ be a set, and let $f =
-\lambda x.~s_2$ be a function that maps each variable in $s_1$ to a
-new set. Then, $s = s_1 \bind f$ if and only if forall $x$, $y$, if
-$x\in s_1$ and $y \in f(x)$, then  $y \in s$, and vice versa. The
-translation first encodes $s_1$ and $s_2$ to obtain $(\phi_1,\G_1)$
-and $(\phi_2,\G_2)$, respectively. Since $s_2$ is under a new binder
-that binds $x$, the free variable sequence of $s_2$ is $\nubar,x$. In the
-interest of hygiene, we substitute a fresh $a$ for $x$, making the
-sequence $\nubar,a$. The set $s$ is encoded as a new uninterpreted
-predicate $g$ indexed by $s$'s free variables ($\nubar$). Since the
-set denoted by $g$ is the result of the bind $s_1 \bind \lambda
-x.s_2$, a first-order constraint defining the bind operation (as
-described above) is generated. The constraint relates the predicates
-$\G_1$ and $\G_2$, representing $s_1$ and $s_2$ (resp.), to the
-uninterpreted predicate $g$ that represents $s$. The constraint is
-assigned a name ($\pi$) so that it can be enabled only for the choice
-of free variables ($\nubar$) for which $g$ needs to be defined;
-observe that the lambda expression returned by $\mssemof{\nubar}{s_1 \bind
-\lambda x.s_2}$ asserts $\pi(\vbar)$ only in conjunction with
-$g(\vbar,r)$, for some $r$.
+of the set monad's bind operator in FOL. Let $s_1$ be a set, and let
+$f$ be a function that maps each variable in $s_1$ to a new set. Then,
+$s_2 = s_1 \bind f$ if and only if for all $y\in s_2$, there exists an
+$x \in s_1$ such that $y = f(x)$, and forall $x\in s_1$, $f(x)\in
+s_2$. The encoding essentially adds new constraints to this effect.
+The translation first encodes $s_1$ and $s_2$ to obtain
+$(\phi_1,\G_1)$ and $(\phi_2,\G_2)$, respectively. Since $s_2$ is
+under a new binder that binds $x$, the free variable sequence of $s_2$
+is $\nubar,x$. In the interest of hygiene, we substitute a fresh $a$
+for $x$, making the sequence $\nubar,a$. The set $s$ is encoded as a
+new uninterpreted predicate $g$ indexed by $s$'s free variables
+($\nubar$). Since the set denoted by $g$ is the result of the bind
+$s_1 \bind \lambda x.s_2$, first-order constraints defining the bind
+operation (as described above) are generated. The constraints relate
+the predicates $\G_1$ and $\G_2$, representing $s_1$ and $s_2$
+(resp.), to the uninterpreted predicate $g$ that represents $s$. The
+constraints are assigned names ($\pi_1$ and $\pi_2$) to give them an
+easy handle.
 
 The first-order encoding of the $\existsl(\stg,\phi,s)$ expression
-eliminates the existential by Skolemizing it. Skolemizing is the
+essentially Skolemizes the existential. Skolemizing is the
 process of substituting an existentially bound $x$ in
 $\phi_x\in\Prop^1$ with $f(\nubar)$, where $f$ is a fresh
-uninterpreted function, and $\nubar$ are the free variables in
-$\phi_x$ bound by enclosing universal quantifiers. Due to the
+uninterpreted function (called the Skolem function), and $\nubar$ are
+the free variables in $\phi_x$ bound by enclosing universal
+quantifiers. Due to the
 decidability restrictions (Sec.~\ref{sec:decidability}), the only
 uninterpreted functions we admit in our logic are boolean (i.e.,
-predicates/relations). Consequently, Skolemizing an existentially
-bound $\stg$ with $f(\nubar)$ is not really an option in our case.
-Instead, we define $f$ as an uninterpreted relation that behaves as a
-partial function:
+predicates/relations). Consequently, we cannot define the Skolem
+function $f$ directly. Instead, we define it via an uninterpreted
+relation, by explicitly asserting the function property:
 \begin{smathpar}
   \begin{array}{c}
-    \forall \nubar.\forall a.\forall b.~f(\nubar,a) \wedge f(\nubar,b)
-    \Rightarrow a = b
+    (\forall \nubar.\forall a.\forall b.~f(\nubar,a) \wedge f(\nubar,b)
+    \Rightarrow a = b)
+    ~~\conj~~ (\forall \nubar.\exists a. f(\nubar,a))
   \end{array}
@@
 We then replace the existentially bound $\stg$ with a new universally
-bound $a$ in $\phi$ and $s$, and define the set $[a/\stg]s$ under the
-condition that $f(\nubar,a) \conj [a/\stg]\phi$ holds. Since the
-existence of a $\stg$ that satisfies $\phi$ is axiomatized to
-define $s$, if the aforementioned condition does not hold, i.e., if
-there does exist such a $\stg$, we let the context derive $\bot$.
+bound $a$ in $\phi$ and $s$, such that $f(\nubar,a) $ holds, before
+encoding the existentially bound $s$.
 
@@
 \begin{itemize}
-  \item All function symbols, modulo those that might appear in
-    $\varphi\in\Prop^0$, are uninterpreted and Boolean.
+  \item All function symbols, modulo those that are drawn from
+    $\Prop^0$ and $\Prop^1$, are uninterpreted and Boolean.
   \item All quantification is first-order; second-order objects, such
@@
 other words, the number of $\forall$ quantifiers preceding an
-$\exists$ quantifier is utmost two more than the number of free
+$\exists$ quantifier is utmost one more than the number of free
 variables ($\nubar$) in $s$. For the convenience of this discussion,
@@
 Next, in Fig.~\ref{fig:logic}, observe that the (ordered) set $\nubar$
-is extended only in the encoding rules for $\bind$ and $\existsl$
-expressions. Since $\existsl$ is a result of a particular choice of the
-weakening semantics (Sec.~\ref{sec:inference}), let us ignore it for now to
-focus on the core set expression language. This leaves $\bind$ as the
-only case where $\nubar$ is extended.  Since an occurrence of $\bind$
-adds one to $|\nubar|$, if $s$ is a subexpression of a top-level
-{\sf bind} expression (like $\F_U(\stl,\stg)$ from the above example),
-then the prenex signature of $\phi_s$ is $\forall^2\exists$.
-Furthermore, if the subexpressions of $s$ are not {\sf bind} expressions,
-then none of the $\phi_i$'s are quantified, and the prenex signature
-of $\phi = \bigwedge_i\phi_i \wedge \phi_s$ remains
-$\forall^2\exists$.  The fragment of FOL that admits formulas with
-prenex signatures of form $\forall^2\exists^*$ is called the
-Godel-Kalmar-Schutte ({\sf GKS}) fragment~\cite{...}, which is known to be
-decidable. The language of encoding, however, is a combination of {\sf
-GKS} with (a). $\Prop^0$, the theory from which quantifier-free
-propositions ($\varphi$) that encode object language expressions are
-drawn, and (b).  $\Prop^1$, the theory from which invariants ($I$) are
-drawn. Thus the encoding is decidable if the combination of ${\sf GKS}
-+ \Prop^0 + \Prop^1$ is decidable.
+is extended only in the encoding rule for $\bind$. Since an occurrence
+of $\bind$ adds one to $|\nubar|$, if $s$ is a bind expression nested
+inside a top-level bind expression (like $\F_U(\stl,\stg)$ from
+the above example), then the prenex signature of $\phi_s$ is
+$\forall^2\exists$.  Furthermore, if the subexpressions of $s$ are
+neither {bind} nor $\existsl$ expressions, then none of the
+$\phi_i$'s are quantified, and the prenex signature of $\phi =
+\bigwedge_i\phi_i \wedge \phi_s$ remains $\forall^2\exists$. Similar
+observation holds when $s$ is an $\existsl$ expression nested inside a
+top-level {bind} expression.  Since $\existsl$ is generated as a
+result of stabilizing a SQL command transformer, which is always a
+non-nested bind expression, the subexpression ($s'$) of $\existsl$ is
+a non-nested bind expression. $s'$ is however nested inside the
+top-level bind expression, hence its prenex signature is
+$\forall^2\exists$.  Since $\existsl$ doesn't extend $\nubar$, the
+prenex signature of $s$ remains $\forall^2\exists$. When $s$ is an
+expression other than $\bind$ or $\existsl$, then $\phi_s$ is not a
+quantified formula, and its prenex signature is trivially subsumed by
+$\forall^2\exists$. Thus, for the subset of $\SL$, where bind
+expressions are restricted to one level of nesting, the FOL formulas
+generated by the encoding have the prenex signature as
+$\forall^2\exists$.
 
-To formally state the above property, let $\SL[n,\Prop^0,\Prop^1]$
-denote the core set expression language that admits $n$ nested {\sf bind}
-expressions, and allows program (Boolean) expressions ($e$), and
-invariants ($\I$) to be drawn from theories $\Prop^0$ and $\Prop^1$,
-respectively. Then:
+The fragment of FOL that admits formulas with prenex signatures of
+form $\forall^2\exists^*$ is called the G\"odel-K\'almar-Sch\"utte ({\sf
+GKS}) fragment~\cite{gks}, which is known to be decidable. The
+language of encoding, however, is a combination of {\sf GKS} with (a).
+$\Prop^0$, the theory from which quantifier-free propositions
+($\varphi$) that encode object language expressions are drawn, and
+(b). $\Prop^1$, the theory from which invariants ($I$) are drawn. Thus
+the encoding of the subset of $\SL$ described above is decidable if the
+combination of ${\sf GKS} + \Prop^0 + \Prop^1$ is decidable. We write
+$\SL[\Prop^0,\Prop^1]$ to highlight the parameterization of $\SL$ on
+$\Prop^0$ and $\Prop^1$.  The discussion in the previous paragraph
+points to the existence of subsets in $\SL[\Prop^0,\Prop^1]$ that are
+decidable:
 \begin{theorem}
-  $\SL[0,\Prop^0,\Prop^1]$ is decidable if ${\sf GKS}+\Prop^0+\Prop^1$
+  There exists an $\SL'[\Prop^0,\Prop^1] \subset \SL[\Prop^0,\Prop^1]$
+  such that $\SL'$ is decidable if ${\sf GKS}+\Prop^0+\Prop^1$ is
+  decidable.
+\end{theorem}
+An example of such an $\SL'$ is the subset described above: $\SL$ with
+bind expressions confined to one level of nesting. We denote this
+subset as $\SL^1[\Prop^0,\Prop^1]$, for which we assert decidability:
+\begin{theorem}
+  $\SL^1[\Prop^0,\Prop^1]$ is decidable if ${\sf GKS}+\Prop^0+\Prop^1$ is
+  decidable.
+\end{theorem}
+$\SL^1$ is a useful subset of $\SL$, for it corresponds to $\txnimp$
+programs without nested \C{foreach} loops. Observe that the
+\C{new\_order} transaction (Fig.~\ref{fig:new_order_code}) belongs to
+this subset.  Indeed, $\SL^1$, while being a restricted version of
+$\SL$, is nonetheless expressive enough to cover all the benchmarks we
+considered in Sec.~\ref{sec:case-studies}.
+
+A useful instantiation of $\SL^1$ is $\SL[{\sf BV},{\sf GKS}+{\sf
+BV}]$, where ${\sf BV}$ is the theory of bit-vector arithmetic, which
+is often used to encode the finite-bit integer arithmetic of real
+programs. Finite-bit integer arithmetic has a finite axiomatization in
+{\sf GKS}. For instance, 32-bit integers can be encoded as $2^{32}$
+distinct constants of an uninterpreted sort $T$, while integer
+operations like addition and multiplication can be encoded as
+uninterpreted functions whose properties are enumerated for the entire
+domain of $T$. Thus {\sf BV} is subsumed by {\sf GKS}. Since the
+latter is decidable, the combination is decidable:
+\begin{theorem}
+  $\SL^1[{\sf GKS},{\sf GKS}+{\sf BV}]$ is decidable.
+\end{theorem}
+The $\SL^1[{\sf BV},{\sf GKS}+{\sf BV}]$ instantiation requires $I$ to
+be drawn from {\sf GKS}+{\sf BV}, which is expressive enough to
+describe common database integrity constraints, such as referential
+integrity, non-negativeness of all integer values in a column etc.
+The isolation specifications presented in \S\ref{sec:isolation} are
+already simple first-order formulas that can be encoded in {\sf GKS}.
+Furthermore, it is also reasonable to expect the guarantee ($G$) of a
+transaction to be expressible in the same logic as its inferred $\F$,
+since $\F$ (without the stability check) is essentially a complete
+characterization of the transaction, while $G$ is only an abstraction.
+Thus, with $\SL^1[{\sf GKS},{\sf GKS}+{\sf BV}]$ as the language of
+inference, the verification problem for weakly isolated transactions
 is decidable.
-\end{theorem}
-The $n=0$ case denotes the lack of nested {\sf bind}
-expressions. Since nested \C{foreach} statements are the only program
-constructs in $\txnimp$ that lead to nested bind expressions, this
-instantiation denotes the encoding language for $\txnimp$ programs
-without nested \C{foreach} loops. Observe that the \C{new\_order}
-transaction (Fig.~\ref{fig:new_order_code}) satisfies this constraint.
-Indeed, this restricted language is nonetheless expressive enough to
-cover all of the benchmarks we have considered.
 

### 2017-11-04 inference.tex

@@
 is extended only in the encoding rule for $\bind$. Since an occurrence
-of $\bind$ adds one to $|\nubar|$, if $s$ is a bind expression nested
-inside a top-level bind expression (like $\F_U(\stl,\stg)$ from
-the above example), then the prenex signature of $\phi_s$ is
-$\forall^2\exists$.  Furthermore, if the subexpressions of $s$ are
-neither {bind} nor $\existsl$ expressions, then none of the
-$\phi_i$'s are quantified, and the prenex signature of $\phi =
-\bigwedge_i\phi_i \wedge \phi_s$ remains $\forall^2\exists$. Similar
+of $\bind$ adds a quantifier to $|\nubar|$, if $s$ is a bind
+expression nested inside a top-level bind expression (like
+$\F_U(\stl,\stg)$ from the above example), then the prenex signature
+of $\phi_s$ is $\forall^2\exists$.  Furthermore, if the subexpressions
+of $s$ are neither {bind} nor $\existsl$ expressions, then none of the
+$\phi_i$'s are quantified, and the prenex signature of $\phi = \bigwedge_i\phi_i \wedge \phi_s$
+remains $\forall^2\exists$. A similar
 observation holds when $s$ is an $\existsl$ expression nested inside a
@@
 non-nested bind expression, the subexpression ($s'$) of $\existsl$ is
-a non-nested bind expression. $s'$ is however nested inside the
+a non-nested bind expression. $s'$ is however nested inside a
 top-level bind expression, hence its prenex signature is
@@
 
-The fragment of FOL that admits formulas with prenex signatures of
+The fragment of FOL that admits formulas with prenex signatures of the
 form $\forall^2\exists^*$ is called the G\"odel-K\'almar-Sch\"utte ({\sf
@@
 ($\varphi$) that encode object language expressions are drawn, and
-(b). $\Prop^1$, the theory from which invariants ($I$) are drawn. Thus
+(b). $\Prop^1$, the theory from which invariants ($I$) are drawn. Thus,
 the encoding of the subset of $\SL$ described above is decidable if the
@@
 $\Prop^0$ and $\Prop^1$.  The discussion in the previous paragraph
-points to the existence of subsets in $\SL[\Prop^0,\Prop^1]$ that are
+points to the existence of non-trivial subsets in $\SL[\Prop^0,\Prop^1]$ that are
 decidable:
@@
 \end{theorem}
-An example of such an $\SL'$ is the subset described above: $\SL$ with
+One interesting example of such an $\SL'$ is the subset described above: $\SL$ with
 bind expressions confined to one level of nesting. We denote this
 subset as $\SL^1[\Prop^0,\Prop^1]$, for which we assert decidability:
-\begin{theorem}
+\begin{corollary}
   $\SL^1[\Prop^0,\Prop^1]$ is decidable if ${\sf GKS}+\Prop^0+\Prop^1$ is
   decidable.
-\end{theorem}
+\end{corollary}
 $\SL^1$ is a useful subset of $\SL$, for it corresponds to $\txnimp$
@@
 \begin{theorem}
-  $\SL^1[{\sf GKS},{\sf GKS}+{\sf BV}]$ is decidable.
+  $\SL^1[{\sf BV},{\sf GKS}+{\sf BV}]$ is decidable.
 \end{theorem}
-The $\SL^1[{\sf BV},{\sf GKS}+{\sf BV}]$ instantiation requires $I$ to
-be drawn from {\sf GKS}+{\sf BV}, which is expressive enough to
-describe common database integrity constraints, such as referential
-integrity, non-negativeness of all integer values in a column etc.
-The isolation specifications presented in \S\ref{sec:isolation} are
-already simple first-order formulas that can be encoded in {\sf GKS}.
-Furthermore, it is also reasonable to expect the guarantee ($G$) of a
-transaction to be expressible in the same logic as its inferred $\F$,
-since $\F$ (without the stability check) is essentially a complete
+This instantiation requires $I$ to be drawn from {\sf GKS}+{\sf BV},
+which is expressive enough to describe common database integrity
+constraints, such as referential integrity, non-negativeness of all
+integer values in a column etc.  The isolation specifications
+presented in \S\ref{sec:isolation} are already simple first-order
+formulas that can be encoded in {\sf GKS}.  Furthermore, it is also
+reasonable to expect the guarantee ($G$) of a transaction to be
+expressible in the same logic as its inferred $\F$, since $\F$
+(without the stability check) is essentially a complete
 characterization of the transaction, while $G$ is only an abstraction.
-Thus, with $\SL^1[{\sf GKS},{\sf GKS}+{\sf BV}]$ as the language of
+Thus, with $\SL^1[{\sf BV},{\sf GKS}+{\sf BV}]$ as the language of
 inference, the verification problem for weakly isolated transactions

### 2017-11-05 inference.tex
Inference rules changed to include F_{ctxt}

@@
 semantics-preserving translation amenable for automated verification.
+
 \input{inference-rules}
+
 Fig.~\ref{fig:inference-rules} shows the syntax-directed state
 transformer inference rules for $\txnimp$ commands inside a
-transaction $\C{TXN}_i$.  Inference depends on the $\I$-constrained
-rely relation $\R$, and the high-level invariant $I$ for reasons
-that will become clear shortly.  Inference rules compute, for each
-command $c$, a (meta) function $\F$ that returns a set as an
-expression in $\SL$, given a pair of sets ($\stl$ and $\stg$) that
-describe local and a global database states, respectively. The
-expression returned by $\F(\stl,\stg)$ (abstractly) describes the set
-of records that get added to $\stl$ as a result of executing $c$ under
-$\stl$ and $\stg$.  Thus, $\F$ captures the \emph{effect} part of the
-state transformer of $c$, which is the function
-$\lambda(\stl,\stg).~\stl \cup \F(\stl,\stg)$\footnote{Recall that the
-  operational semantics treats deletion of records as the addition of
-  the deleted record with its \C{del} field set to true in the local
-  store.}. For $\F$ to be useful in RG verification, it needs to be
-stable w.r.t the rely relation $\R$. The stability condition on
-effects can be defined thus:
+transaction $\C{TXN}_i$. The rules compute, for each command $c$, a
+(meta) function $\F$ that returns a set of records as an expression in
+$\SL$, given a global database $\stg$. Intuitively, $\F(\stg)$
+abstracts the set of records added to the local database $\stl$ as a
+result of executing $c$ under $\stg$ (i.e., $\stg \vdash
+(\tbox{c}_i,\stl) \stepsto^{*}_{R} (\tbox{\cskip}_i, \stl \cup
+\F(\stg))$)\footnote{Recall that the operational semantics treats 
+  deletion of records as the addition of the deleted record with its
+  \C{del} field set to true in the local store.}. Note that $\F$
+actually captures the \emph{effect} part of the state transformer,
+but we nonetheless refer to it as a transformer. Since the execution
+is subject to isolation-constrained interference, the inference
+judgment depends on the isolation-constrained rely relation $\R$,
+which is used to enforce the stability of the state transformer $\F$.
+Recall that $\R$ is a tri-state rely relation over $\stl$, $\stg$ and
+$\stg'$, that admits an interference from $\stg$ and $\stg'$ depending
+on the local database state $\stl$. Thus, the stability of the state
+transformer $\F$ of $c$ with respect to $\R$ needs to take into
+account the (possible) prior state of the local database $\stl$, which
+depends on the context (sequence of previous commands) of $c$, and
+computed by the corresponding state transformer $\Fx$. Thus, the
+semantics of the state transformer can be understood in terms of the
+RG judgment as following (formalized as
+Theorem~\ref{thm:inference-sound} in Sec.~\ref{sec:inference-sound}):
+\begin{smathpar}
+  \begin{array}{c}
+    \R \vdash \hoare{\lambda(\stl,\stg).~ \stl = \Fx(\stg)}{c}
+    {\lambda(\stl,\stg).~ \stl = \Fx(\stg) \cup \F(\stg)}
+  \end{array}
+\end{smathpar}
+In the above RG judgment, let us call the pre-condition $P$, and
+post-condition $Q$. The stability condition on the state transformer
+$\F$ can be derived from the stability condition on $Q$. Observe that
+for $Q$ to be stable, $\Fx(\stg') \cup \F(\stg')$ must be equal to
+$\Fx(\stg) \cup \F(\stg)$, where $\stg$ and $\stg')$ are related by
+$R$ (ignore $\I$ for the moment). Assuming that $P$ is stable,
+$\Fx(\stg')=\Fx(\stg)$ is already given, leaving $\F(\stg')=\F(\stg)$
+to be enforced. Thus, the stability of $\F$ in in the context of $\Fx$
+(written $\inctxt{\Fx}{\F}$) is defined as following:
 \begin{smathpar}
 \begin{array}{lcl}
-  \stable(\R,\F) & \Leftrightarrow & \forall \stl,\stg,\stg',\nubar.~
-  \R(\stl,\stg,\stg') \Rightarrow \F(\stl,\stg) \equiv \F(\stl,\stg')
+  \stable(\R,\inctxt{\Fx}{\F}) & \Leftrightarrow & \forall \stg,\stg',\nubar.~
+  \R(\Fx(\stg) \cup \F(\stg),\stg,\stg') \Rightarrow \F(\stg) = \F(\stg')
 \end{array}
@@
 where $\nubar$ are the variables that occur free in $\F$; this is
-possible because of how the inference rules are structured.
-Intuitively, the stability condition requires an effect to describe
-the same set of records before and after an interference. The
-equivalence in $\SL$ translates to equality in first-order logic, as
-we describe below. In the inference rules, stability is enforced
+possible because of how the inference rules are structured. The
+equality in $\SL$ translates to equivalence in first-order logic, as
+we describe later. In the inference rules, stability is enforced
 constructively by a meta-function $\stabilize{\cdot}$, which accepts
-an effect and returns a new effect that is guaranteed to be stable
-under $\R$.  $\stabilize{\cdot}$ achieves the stability guarantee by
-abstracting away the bound global state ($\stg$) in an unstable $\F$
-to an existentially bound $\stg'$ as described below:
+a transformer $\F$ (in its context $\Fx$) and returns a new
+transformer that is guaranteed to be stable under $\R$.
+$\stabilize{\cdot}$ achieves the stability guarantee by abstracting
+away the bound global state ($\stg$) in an unstable $\F$ to an
+existentially bound $\stg'$ as described below:
 \begin{smathpar}
 \begin{array}{lcll}
-  \stabilize{\F} & = & \F & \texttt{if } \stable(\R,\F).\\
-  & = & \lambda (\stl,\stg).~\existsl(\stg',I(\stg'),\F(\stl,\stg')) 
+  \stabilize{\inctxt{\Fx}{\F}} & = & \F & \texttt{if }
+  \stable(\R,\inctxt{\Fx}{\F}).\\
+  & = & \lambda (\stg).~\existsl(\stg',I(\stg'),\F(\stg')) 
       & \texttt{otherwise. }\stg'\texttt{ is a fresh name.}\\
@@
 is nonetheless useful because of the $I(\stg')$ assumption imposed on
-the existentially bound $\stg'$. The example in
+an existentially bound $\stg'$. The example in
 Fig.~\ref{fig:weakening-example} demonstrates.
@@
 \forall(r\in\stg).~r.\C{bal}\ge 0$). The transaction starts by issuing
-a \C{select1} query, whose effect ($\F$) is essentially a singleton
+a \C{select1} query, whose transformer $\F$ is essentially a singleton
 set containing a record $r$ whose id is \C{acc\_id} (i.e.,
-$\F(\stl,\stg) = \{r \,|\, r\in\stg \wedge r.\C{id} = \C{acc\_id}\}$).
-However, this effect is unstable because $\F(\stl,\stg')$ may not be
-the same set as $\F(\stl,\stg)$ when $\stg'\neq\stg$. A record
+$\F(\stg) = \{r \,|\, r\in\stg \wedge r.\C{id} = \C{acc\_id}\}$).
+However, $\F$ is unstable because $\F(\stg')$ may not be
+the same set as $\F(\stg)$ when $\stg'\neq\stg$. A record
 $r\in\stg$ whose $\C{id}=\C{acc\_id}$ may have its balance updated by
@@
 of commands $c_1;c_2$ is the union of effects $\F_1$ and $\F_2$ of
-$c_1$ and $c_2$, respectively, except that $\F_2$ is applied to a
-state ($\stl$) to which $\F_1$ has already been applied, reflecting
-their order of reduction. The inference rule for \C{FOREACH} takes
+$c_1$ and $c_2$, respectively, except that $\F_2$ is computed in a
+context that includes $\F_1$. The inference rule for \C{FOREACH} takes
 advantage of the $\SL$'s bind operator to lift the effect inferred for
@@
 about database programs; Sec. ~\ref{sec:opsem}), sequential
-composition of the effects of different iterations is the same as their
-parallel composition. Since the loop body is executed once per each
-$z\in x$, the effect of the the loop is a union of effects ($\F$)
+composition of the effects of different iterations is the same as
+their parallel composition. Since the loop body is executed once per
+each $z\in x$, the effect of the the loop is a union of effects ($\F$)
 for all $z\in x$, all applied to the same state ($\stl$ and $\stg$).
-That is, $\F_{loop}(\stl,\stg) = \bigcup_{z\in
-  x}\F_{body}(\stl,\stg)$. From the definition of the set monad's bind
-operator, $\F_{loop}(\stl,\stg) = x \bind (\lambda
-z.~F_{body}(\stl,\stg))$, which mirrors the definition of the rule.
+That is, $\F_{loop}(\stg) = \bigcup_{z\in
+x}\F_{body}(\stg)$. From the definition of the set monad's bind
+operator, $\F_{loop}(\stg) = x \bind (\lambda
+z.~F_{body}(\stg))$, which mirrors the definition of the rule.
 %This development mirrors ideas found in type checkers for certain
@@
 \subsection{Soundness of Inference}
+\label{sec:inference-sound}
 
-We can now establish a correspondence between the inference rules given
-above and the RG judgment of \S\ref{sec:reasoning}:
+We now formally state the correspondence between the inference rules
+given above and the RG judgment of \S\ref{sec:reasoning}:
 \begin{theorem}
   \label{thm:inference-sound}
-  Let the set expression $\F(\emptyset,\stg)$ have a natural interpretation
-  in the theory of sets.  Then, for all $i$,$R$,$I$,$c$,$\F$,
-  if $\stable(\R,I)$ and $c \elabsto \F$,
-  then:\\\vspace*{-0.2cm}
+  For all $i$,$R$,$I$,$c$,$\Fx$, $\F$, if $\stable(\R,I)$ and $\Fx
+  \vdash c \elabsto \F$, then:\\\vspace*{-0.2cm}
   \begin{smathpar}
   \begin{array}{c}
-  \R \vdash \hoare{\lambda(\stl,\stg).~\stl=\emptyset \conj
-  I(\stg)}{c}{\lambda(\stl,\stg).\stl = \F(\emptyset,\stg)}
+    \R \vdash \hoare{\lambda(\stl,\stg).~\stl=\Fx(\stg) \conj
+    I(\stg)}{c}{\lambda(\stl,\stg).~\stl = \Fx(\stg) \cup \F(\stg)
+    \conj I(\stg)}
   \end{array}
@@
 
-\paragraph{{\sc Proof Sketch.}} We prove a slightly stronger statement :
-For all  $i$,$R$,$I$,$c$,$\F$,$s$ if $\stable(\R,I)$ and $c \elabsto \F$, then:
-  \begin{smathpar}
-    \begin{array}{c}
-    \R \vdash \hoare{\lambda(\stl,\stg).~\stl=s \conj
-    I(\stg)}{c}{\lambda(\stl,\stg).\stl = \F(s,\stg)}
-    \end{array}
-  \end{smathpar}
+\paragraph{{\sc Proof Sketch.}} 
+% We prove a slightly stronger statement :
+% For all  $i$,$R$,$I$,$c$,$\F$,$s$ if $\stable(\R,I)$ and $c \elabsto \F$, then:
+%   \begin{smathpar}
+%     \begin{array}{c}
+%     \R \vdash \hoare{\lambda(\stl,\stg).~\stl=s \conj
+%     I(\stg)}{c}{\lambda(\stl,\stg).\stl = \F(s,\stg)}
+%     \end{array}
+%   \end{smathpar}
 The proof follows by structural induction on $c$. Assume that $P
@@
 $\stabilize{\F} = \F$ and hence the proof directly follows. Otherwise,
-$\stabilize{\F} = \lambda (\stl, \stg).\existsl(\stg', I(\stg'),
-\F(\stl, \stg'))$. Since the pre-condition $P(\stl, \stg)$ asserts
+$\stabilize{\F} = \lambda (\stg).\existsl(\stg', I(\stg'),
+\F( \stg'))$. Since the pre-condition $P(\stl, \stg)$ asserts
 $I(\stg)$, any state which obeys $P$ provides the existential $\stg'$
@@
 guarantee can now be established w.r.t the effect $\F$ of the body.
-The \rulelabel{RG-Txn} rule so updated is shown below:
+The \rulelabel{RG-Txn} rule so updated is shown below ($\Fempty =
+\lambda(\stg).~\emptyset$ denotes an empty context):
 \begin{smathpar}
@@
    \R_e = R \backslash \I_e \spc \R_c = R \backslash \I_c \spc
-   c \Longrightarrow_{\langle i,\R_e,I \rangle}\F \spc \stable(\R_c,\F)\\
-   \forall \stl,\stg.~ \stl = F(\emptyset,\stg) \Rightarrow 
-   G(\stg, \stl\gg\stg)\spc
+   \Fempty \vdash c \Longrightarrow_{\langle i,\R_e,I \rangle}\F \\
+   \stable(\R_c,\inctxt{\Fempty}{\F}) \spc
+   \forall \stg.~ G(\stg,\F(\stg)) \spc
    \forall \stg,\stg'.~I(\stg) \wedge G(\stg,\stg') \Rightarrow I(\stg')\\

### 2017-11-06 inference.tex

@@
 $c_1$ and $c_2$, respectively, except that $\F_2$ is computed in a
-context that includes $\F_1$. The inference rule for \C{FOREACH} takes
-advantage of the $\SL$'s bind operator to lift the effect inferred for
-the loop body to the level of the loop. Since records added to $\stl$
-in each iteration of \C{FOREACH} are independent of the previous
-iteration (recall that we make a local context independence assumption
-about database programs; Sec. ~\ref{sec:opsem}), sequential
-composition of the effects of different iterations is the same as
-their parallel composition. Since the loop body is executed once per
-each $z\in x$, the effect of the the loop is a union of effects ($\F$)
-for all $z\in x$, all applied to the same state ($\stl$ and $\stg$).
-That is, $\F_{loop}(\stg) = \bigcup_{z\in
+context that includes $\F_1$ (we write $\F_1 \cup \F_2$ as a shorthand
+for $\lambda(\stg).~\F_1(\stg) \cup \F_2(\stg)$). The inference rule
+for \C{FOREACH} takes advantage of the $\SL$'s bind operator to lift
+the effect inferred for the loop body to the level of the loop. Since
+records added to $\stl$ in each iteration of \C{FOREACH} are
+independent of the previous iteration (recall that we make a local
+context independence assumption about database programs; Sec.
+~\ref{sec:opsem}), sequential composition of the effects of different
+iterations is the same as their parallel composition. Since the loop
+body is executed once per each $z\in x$, the effect of the the loop is
+a union of effects ($\F$) for all $z\in x$, all applied to the same
+state ($\stl$ and $\stg$).  That is, $\F_{loop}(\stg) = \bigcup_{z\in
 x}\F_{body}(\stg)$. From the definition of the set monad's bind
-operator, $\F_{loop}(\stg) = x \bind (\lambda
-z.~F_{body}(\stg))$, which mirrors the definition of the rule.
+operator, $\F_{loop}(\stg) = x \bind (\lambda z.~F_{body}(\stg))$,
+which mirrors the definition of the rule.
 %This development mirrors ideas found in type checkers for certain

### 2017-11-06 inference.tex
edits

@@
 function of the effect of its body, thanks to certain pleasant
-properties enjoyed by the database programs modeled by $\txnimp$.
-Interpreting database semantics as functional transformations on sets
-(described in terms of their logical effects) enables an inference
-mechanism that can leverage off-the-shelf SMT solvers for automated
-verification.
+properties enjoyed by the database programs modeled by our core
+language.  Interpreting database semantics as functional
+transformations on sets (described in terms of their logical effects)
+enables an inference mechanism that can leverage off-the-shelf SMT
+solvers for automated verification.
 
@@
 set transformations (see Fig.~\ref{fig:logic-syntax}). The language
-admits only set expressions that include variables ($x$), literals of
-the form $\{x \,|\, \varphi\}$ where $\varphi$ is a propositional
+admits set expressions that include variables ($x$), literals of the
+form $\{x \,|\, \varphi\}$ where $\varphi$ is a propositional
 (quantifier-free) formula on $x$, a restricted form of existential
-quantification that binds a set $\stg$ satisfying proposition $\phi$ in a
-set expression $s$, a monadic composition of two set expressions
+quantification that binds a set $\stg$ satisfying proposition $\phi$
+in a set expression $s$, a monadic composition of two set expressions
 ($s_1$ and $s_2$) composed using a bind ($\bind$) operation, a
@@
 are also variables in $\SL$, but are used to denote local and database
-states, respectively. Constant sets can be written using set
-literal expressions. For example, the set $\{1,2\}$ can be written as
-$\{x \,|\, x=1 \vee x=2\}$. The language is carefully chosen to be
-expressive enough to capture the semantics of $\txnimp$ statements (as
-well as SQL operations more generally), yet simple enough to have a
-semantics-preserving translation amenable for automated verification.
+states (also represented as sets), respectively. Constant sets can be
+written using set literal expressions. For example, the set $\{1,2\}$
+can be written as $\{x \,|\, x=1 \vee x=2\}$. The language is
+carefully chosen to be expressive enough to capture the semantics of
+$\txnimp$ statements (as well as SQL operations more generally), yet
+simple enough to have a semantics-preserving translation amenable for
+automated verification.
 
@@
 \end{smathpar}
-In the above RG judgment, let us call the pre-condition $P$, and
-post-condition $Q$. The stability condition on the state transformer
+In the above RG judgment, let us call the pre-condition
+($\hoare{\lambda(\stl,\stg).~ \stl = \Fx(\stg)}$) $P$, and
+post-condition (${\lambda(\stl,\stg).~ \stl = \Fx(\stg) \cup
+  \F(\stg)}$), $Q$. The stability condition on the state transformer
 $\F$ can be derived from the stability condition on $Q$. Observe that
 for $Q$ to be stable, $\Fx(\stg') \cup \F(\stg')$ must be equal to
-$\Fx(\stg) \cup \F(\stg)$, where $\stg$ and $\stg')$ are related by
+$\Fx(\stg) \cup \F(\stg)$, where $\stg$ and $\stg'$ are related by
 $R$ (ignore $\I$ for the moment). Assuming that $P$ is stable,
@@
 
-The proofs for \C{SELECT}, sequencing, and conditionals use use the
+The proofs for \C{SELECT}, sequencing, and conditionals use the
 inductive hypothesis to infer the RG-judgments present in the

### 2017-11-06 evaluation.tex
Courseware code and guarantees.

@@
 
-In this section, we present our experience in running \tool on two
+In this section, we present our experience in running \thetool on two
 different applications: \emph{Courseware}: a course registration
@@
 benchmark.
+
+\begin{figure}[!t]
+\begin{ocaml}
+  type table_name = Student | Course | Enrollment
+  type student = {s_id: id; s_name: string}
+  type course = {c_id: id; c_name: string; c_capacity: int}
+  type enrollment = {e_id: id; e_s_id: id; e_c_id: id}
+
+  let enroll_txn sid cid = 
+    let crse = SQL.select1 [Course] (fun c -> c.c_id = cid) in
+    let s_c_enrs = SQL.select [Enrollment] (fun e -> e.e_s_id = sid && 
+                                                     e.e_c_id = cid) in
+    if crse.c_capacity > 0 && Set.is_empty s_c_enrs then
+      (SQL.insert Enrollment {e_id=new_id (); e_s_id=sid; e_c_id=cid};
+       SQL.update Course 
+         (fun c -> {c with c_capacity = c.c_capacity - 1})
+         (fun c -> c.c_id = cid))
+    else ()
+
+  let deregister_student_txn sid = 
+    let s_enrs = SQL.select [Enrollment] (fun e -> e.e_s_id = sid) in
+    if Set.is_empty s_enrs then
+      SQL.delete Student (fun s -> s.s_id = sid)
+    else ()
+\end{ocaml}
+\caption{Courseware Application}
+\label{fig:courseware_code}
+\end{figure}
 
@@
 \C{Student} and \C{Course} tables, there is also an \C{Enrollment}
-table to track the many-to-many enrollment relationship between courses
-and students. The application is required to preserve the following
-invariants on the database:
+table to track the many-to-many enrollment relationship between
+courses and students. The simplified code for the Courseware
+application with only \C{enroll} and \C{deregister\_student}
+transactions is shown in Fig.~\ref{fig:courseware_code}. The
+application is required to preserve the following invariants on the
+database:
 
@@
 \end{enumerate}
-\noindent Both $I_1$ and $I_2$ can be violated under weak
-isolation. $I_1$ can be violated, for example, when
-\C{deregister\_student} runs concurrently with \C{enroll}, both at RC
-isolation. While the former transaction removes the student record
-after checking that no enrollments for that student exists, the latter
-transaction concurrently adds an enrollment record after checking the
-student exists.  Both can succeed concurrently, resulting in an
-invalid state. Invariant $I_2$ can be violated by two \C{enroll}s,
-both reading \C{c\_capacity}=1, and both (atomically) decrementing it,
-resulting in \C{c\_capacity}=-1.  We ran \tool on the courseware
-application written in our DSL, annotated with guarantees for all
-transactions taken from ~\cite{gotsmanpopl16}, and asserted $I = I_1
-\wedge I_2$ as the correctness condition. \tool was able to
-automatically discover the anomalous executions described above, and
-was subsequently able to infer that the anomalies can be preempted by
-promoting the isolation level of \C{enroll} and
+\noindent Both $I_1$ and $I_2$ can be violated under weak isolation.
+$I_1$ can be violated, for example, when \C{deregister\_student} runs
+concurrently with \C{enroll}, both at RC isolation. While the former
+transaction removes the student record after checking that no
+enrollments for that student exists, the latter transaction
+concurrently adds an enrollment record after checking the student
+exists.  Both can succeed concurrently, resulting in an invalid state.
+Invariant $I_2$ can be violated by two \C{enroll}s, both reading
+\C{c\_capacity}=1, and both (atomically) decrementing it, resulting in
+\C{c\_capacity}=-1.  We ran \thetool on the courseware application
+(Fig.~\ref{fig:courseware_code}) after annotating transactions with
+their respective guarantees, and asserting $I = I_1 \wedge I_2$ as the
+correctness condition. The guarantees $G_e$ and $G_d$ for \C{enroll}
+and \C{deregister\_student} transactions, respectively, are shown below:
+\begin{smathpar}
+  \begin{array}{lcl}
+    G_e(\stg,\stg') & \Leftrightarrow & \stg_s'=\stg_s
+      \conj \exists\C{cid}.\exists\C{sid}.\\
+    & & \hspace*{0.3in} \stg_c' = \stg_c \bind \lambda c.~
+        \itel{c.\C{c\_id}=\C{cid}\\
+    & & \hspace*{1.15in}}
+          {\existsl(c',~c'.\C{id}=c.\C{id} \wedge
+              c'.\C{c\_name}=c.\C{c\_name} \wedge c'.\C{c\_capacity}\ge0,
+              ~\{c'\})\\
+    & & \hspace*{1.15in}}
+          {\{c\}}\\
+    & & \hspace*{0.15in}\conj \stg_e = \stg_e' \bind \lambda e.~ 
+        \itel{e.\C{e\_c\_id}=\C{cid} \wedge e.\C{e\_s\_id}=\C{sid}}
+          {\emptyset}{\{e\}}\\
+%   & & \hspace*{1.15in}} 
+%         {\emptyset} {\{e\}}\\
+    G_d(\stg,\stg') & \Leftrightarrow & \stg_c' = \stg_c \conj
+      \stg_e' = \stg_e \conj \exists \C{sid}.~
+      \itel{\forall(e \in \stg_e).~e.\C{e\_s\_id}\neq\C{sid}\\
+%   & & \hspace*{0.3in} \itel{\forall(e \in
+%           \stg_e).~e.\C{e\_s\_id}\neq\C{sid}\\
+    & & \hspace*{1.55in}}
+          {\stg_s' = \stg_s \bind \lambda s.~
+           \itel{s.\C{id}=\C{sid}}{\emptyset}{\{s\}}\\
+    & & \hspace*{1.55in}}
+          {\stg_s' = \stg_s}\\
+  \end{array}
+\end{smathpar}
+For the sake of this presentation we split $\stg$ into three disjoint
+sets of records, $\stg_s$, $\stg_c$, and $\stg_e$, standing for
+\C{Student}, \C{Course}, and \C{Enrollment} tables, respectively.
+Observing that the set language $\SL$ (Sec.~\ref{sec:inference}),
+besides being useful for automatic verification, also facilitates
+succinct expression of transaction semantics, we define $G_e$ and
+$G_d$ in a combination of FOL and $\SL$. $G_e$ essentially says that
+the \C{enroll} transaction leaves the \C{Student} table unchanged,
+while it may update the \C{c\_capacity} field of a \C{Course} record
+to a non-negative value (even when it doesn't update, it is the case
+that $c'.\C{c\_capacity}\ge0$, because $c'=c$, and $c\in\stg_c$, and
+we know that $I_2(\stg_c)$). $G_e$ also conveys that \C{enroll} might
+insert a new \C{Enrollment} record by stating that $\stg_e$, the
+\C{Enrollment} table in pre-state, contains all records $e$ from
+$\stg_e'$, the table in post-state, except when $e.\C{e\_c\_id}$ and
+$e.\C{e\_s\_id}$ match \C{cid} and \C{sid}, respectively. The
+guarantee $G_d$ of \C{deregister\_student} asserts that the
+transaction doesn't write to \C{Course} and \C{Enrollment} tables. The
+transaction might however delete a \C{Student} record bearing an
+\C{id}=\C{sid} (formally, $\stg_s' = \stg_s \bind \lambda s.~
+\itel{s.\C{id}=\C{sid}}{\emptyset}{\{s\}}$), for some \C{sid} for
+which no corresponding \C{Enrollment} records are present in the
+pre-state (in other words, $\forall(e \in
+\stg_e).~e.\C{e\_s\_id}\neq\C{sid}$).
+
+With help of the guarantees, such as those described above, \thetool
+was able to automatically discover the aforementioned anomalous
+executions, and was subsequently able to infer that the anomalies can
+be preempted by promoting the isolation level of \C{enroll} and
 \C{deregister\_student} to SER (on both MySQL and PostgreSQL), leaving
@@
 are a number of ways TPC-C's transactions violate the aforementioned
-invariants under weak isolation. \tool was able to discover all such
+invariants under weak isolation. \thetool was able to discover all such
 violations when verifying the benchmark against $I =
@@
 
-To sanity-check the results of \tool, we conducted experiments with a
+To sanity-check the results of \thetool, we conducted experiments with a
 high-contention OLTP workload on TPC-C aiming to explore the space of
@@
 found 171 of them violated at least one of the four invariants we
-considered.  As expected, the isolation levels that \tool infers for the
+considered.  As expected, the isolation levels that \thetool infers for the
 TPC-C transactions do not result in invariant violations, either on

### 2017-11-06 implementation.tex
Courseware code and guarantees.

@@
 computations in OCaml (modulo some syntactic sugar), and our automatic
-reasoning framework as an extra frontend pass (called \tool) in the
+reasoning framework as an extra frontend pass (called \thetool) in the
 ocamlc 4.03 compiler\footnote{The source code is available at
-available at \emph{https://goo.gl/nmUXZK}}. The input to \tool is a
+available at \emph{https://goo.gl/nmUXZK}}. The input to \thetool is a
 program in our DSL that describes the schema of the database as a
@@
 Fig.~\ref{fig:schema} can be described via the OCaml type definitions
-shown in Fig.~\ref{fig:ocaml-schema}.  \tool also requires a
+shown in Fig.~\ref{fig:ocaml-schema}.  \thetool also requires a
 specification of the program in the form of a collection of guarantees
@@
 includes the first-order logic (FOL) combinators has been implemented
-for this purpose. \tool's verification pass follows OCaml's type
+for this purpose. \thetool's verification pass follows OCaml's type
 checking pass, hence the concrete artifact of verification is OCaml's
@@
 
-At the top-level, \tool runs a loop that picks an unverified
+At the top-level, \thetool runs a loop that picks an unverified
 transaction and progressively strengthens its isolation level until it
@@
 correct, then the verification is guaranteed to succeed. Within the
-loop, the \tool first computes the various rely relations needed for
+loop, the \thetool first computes the various rely relations needed for
 verification ($R$, $\R_l$, and $\R_c$). It then traverses the AST of a
@@
 
-\tool uses the Z3 SMT solver as its underlying reasoning engine. Each
+\thetool uses the Z3 SMT solver as its underlying reasoning engine. Each
 implication check described above is first encoded in FOL, applying
@@
 need to be taken into account for verification to be effective in
-practice.  We consider these into account while linking \tool with
+practice.  We consider these into account while linking \thetool with
 store-specific semantics (isolation specifications etc). As an
@@
 accommodated between the time the records are selected from the
-snapshot, and the time they are actually updated.  \tool performs this
+snapshot, and the time they are actually updated.  \thetool performs this
 translation if the selected store is PostgreSQL, allowing it to reason
 about \C{UPDATE} operations in a way that is faithful to its semantics
-on PostgreSQL. \tool also admits similar compensatory logic for
+on PostgreSQL. \thetool also admits similar compensatory logic for
 certain combinations of isolation levels and operations on MySQL.
@@
 
-\textbf{Annotation Burden} \tool significantly reduces the annotation
+\textbf{Annotation Burden} \thetool significantly reduces the annotation
 burden in verifying a weakly isolated transactions by eliminating the
@@
 verified against a generic serializability condition,  eliminating
-the need for any annotation. In this mode, \tool first infers the
+the need for any annotation. In this mode, \thetool first infers the
 transformer $\F_{SER}$ of $T$ without considering any interference,

### 2017-11-06 inference.tex
Courseware code and guarantees.

@@
 \end{smathpar}
-In the above RG judgment, let us call the pre-condition
-($\hoare{\lambda(\stl,\stg).~ \stl = \Fx(\stg)}$) $P$, and
-post-condition (${\lambda(\stl,\stg).~ \stl = \Fx(\stg) \cup
-  \F(\stg)}$), $Q$. The stability condition on the state transformer
-$\F$ can be derived from the stability condition on $Q$. Observe that
-for $Q$ to be stable, $\Fx(\stg') \cup \F(\stg')$ must be equal to
-$\Fx(\stg) \cup \F(\stg)$, where $\stg$ and $\stg'$ are related by
-$R$ (ignore $\I$ for the moment). Assuming that $P$ is stable,
-$\Fx(\stg')=\Fx(\stg)$ is already given, leaving $\F(\stg')=\F(\stg)$
-to be enforced. Thus, the stability of $\F$ in in the context of $\Fx$
-(written $\inctxt{\Fx}{\F}$) is defined as following:
+In the above RG judgment, let $P$ denote the pre-condition
+$\lambda(\stl,\stg).~ \stl = \Fx(\stg)$, and let $Q$ denote the
+post-condition $\lambda(\stl,\stg).~ \stl = \Fx(\stg) \cup \F(\stg)$.
+The stability condition on the state transformer $\F$ can be derived
+from the stability condition on $Q$. Observe that for $Q$ to be
+stable, $\Fx(\stg') \cup \F(\stg')$ must be equal to $\Fx(\stg) \cup
+\F(\stg)$, where $\stg$ and $\stg'$ are related by $R$ (ignore $\I$
+for the moment). Assuming that $P$ is stable, $\Fx(\stg')=\Fx(\stg)$
+is already given, leaving $\F(\stg')=\F(\stg)$ to be enforced. Thus,
+the stability of $\F$ in in the context of $\Fx$ (written
+$\inctxt{\Fx}{\F}$) is defined as following:
 \begin{smathpar}

### 2017-11-06 evaluation.tex
Typos.

@@
 different applications: \emph{Courseware}: a course registration
-system described by~\cite{gotsmanpopl16}, and the \emph{TPC-C}
-benchmark.
+system described by~\cite{gotsmanpopl16}, and the TPC-C benchmark.
 
@@
       (SQL.insert Enrollment {e_id=new_id (); e_s_id=sid; e_c_id=cid};
-       SQL.update Course 
-         (fun c -> {c with c_capacity = c.c_capacity - 1})
+       SQL.update Course (fun c -> {c with c_capacity = c.c_capacity - 1})
                          (fun c -> c.c_id = cid))
@@
 \C{c\_capacity}=1, and both (atomically) decrementing it, resulting in
-\C{c\_capacity}=-1.  We ran \thetool on the courseware application
+\C{c\_capacity}=-1.  We ran \thetool on the Courseware application
 (Fig.~\ref{fig:courseware_code}) after annotating transactions with

### 2017-11-06 motivation.tex
Typos.

@@
   let o_id = dist.d_next_o_id in
+  begin
     SQL.update(* UPDATE *) District 
-            (* SET *)(fun d -> {d with d_next_o_id =d_next_o_id + 1})
+              (* SET *)(fun d -> {d with d_next_o_id = d.d_next_o_id + 1})
               (* WHERE *)(fun d -> d.d_id = d_id );
@@
       let stk = SQL.select1(* SELECT * FROM *) Stock 
-              (* WHERE *)(fun s -> s.s_i_id = item_req.ol_i_id 
-                                  && s.s_d_id = d_id)(* LIMIT 1 *); 
+                (* WHERE *)(fun s -> s.s_i_id = item_req.ol_i_id &&
+                                     s.s_d_id = d_id)(* LIMIT 1 *) in
       let s_qty' = if stk.s_qty >= item_req.ol_qty + 10 
@@
                              ol_i_id=item_req.ol_i_id; ol_qty=item_req.ol_qty}
+  end
 

### 2017-11-06 implementation.tex
Typos etc.

@@
 
-\textbf{Annotation Burden} \thetool significantly reduces the annotation
-burden in verifying a weakly isolated transactions by eliminating the
-need to annotate intermediate assertions and loop invariants.
-Guarantees ($G$) and global invariants ($I$), however, still need to be
-provided. Alternatively, a weakly isolated transaction $T$ can be
-verified against a generic serializability condition,  eliminating
-the need for any annotation. In this mode, \thetool first infers the
-transformer $\F_{SER}$ of $T$ without considering any interference,
-which then becomes its guarantee ($G$). Doing likewise for every
-transaction results in a rely relation ($R$) that includes $\F_{SER}$
-of every transaction. Verification now proceeds by taking interference
-into account, and verifying that each transaction still yields the
-same $F$ as its $F_{SER}$. The result of this verification is an
-assignment of (possibly weak) isolation levels to transactions which
-nonetheless guarantees behavior equivalent to a  serializable execution.
+\textbf{Annotation Burden} \thetool significantly reduces the
+annotation burden in verifying a weakly isolated transactions by
+eliminating the need to annotate intermediate assertions and loop
+invariants.  Guarantees ($G$) and global invariants ($I$), however,
+still need to be provided. Alternatively, a weakly isolated
+transaction $T$ can be verified against a generic serializability
+condition,  eliminating the need for guarantee annotations. In this
+mode, \thetool first infers the transformer $\F_{SER}$ of $T$ without
+considering any interference, which then becomes its guarantee ($G$).
+Doing likewise for every transaction results in a rely relation ($R$)
+that includes $\F_{SER}$ of every transaction. Verification now
+proceeds by taking interference into account, and verifying that each
+transaction still yields the same $F$ as its $F_{SER}$. The result of
+this verification is an assignment of (possibly weak) isolation levels
+to transactions which nonetheless guarantees behavior equivalent to a
+serializable execution.
 

### 2017-11-06 inference.tex
Typos etc.

@@
   deletion of records as the addition of the deleted record with its
-  \C{del} field set to true in the local store.}. Note that $\F$
-actually captures the \emph{effect} part of the state transformer,
-but we nonetheless refer to it as a transformer. Since the execution
-is subject to isolation-constrained interference, the inference
-judgment depends on the isolation-constrained rely relation $\R$,
-which is used to enforce the stability of the state transformer $\F$.
-Recall that $\R$ is a tri-state rely relation over $\stl$, $\stg$ and
-$\stg'$, that admits an interference from $\stg$ and $\stg'$ depending
-on the local database state $\stl$. Thus, the stability of the state
-transformer $\F$ of $c$ with respect to $\R$ needs to take into
-account the (possible) prior state of the local database $\stl$, which
-depends on the context (sequence of previous commands) of $c$, and
-computed by the corresponding state transformer $\Fx$. Thus, the
-semantics of the state transformer can be understood in terms of the
-RG judgment as following (formalized as
-Theorem~\ref{thm:inference-sound} in Sec.~\ref{sec:inference-sound}):
+  \C{del} field set to true in the local store.}. Note that the
+function $\F$ we call state transformer here is actually the
+\emph{effect} part of the state transformer introduced in
+Sec.~\ref{sec:motivation}, which is a function $\T$ of form
+$\lambda(\stl,\stg).~\stl \cup \F(\stg)$. Nonetheless, we keep
+referring to $\F$ as state transformer. Since the execution is subject
+to isolation-constrained interference, the inference judgment depends
+on the isolation-constrained rely relation $\R$, which is used to
+enforce the stability of the state transformer $\F$.  Recall that $\R$
+is a tri-state rely relation over $\stl$, $\stg$ and $\stg'$, that
+admits an interference from $\stg$ and $\stg'$ depending on the local
+database state $\stl$. Thus, the stability of the state transformer
+$\F$ of $c$ with respect to $\R$ needs to take into account the
+(possible) prior state of the local database $\stl$, which depends on
+the context (sequence of previous commands) of $c$, and computed by
+the corresponding state transformer $\Fx$. Thus, the semantics of the
+state transformer can be understood in terms of the RG judgment as
+following (formalized as Theorem~\ref{thm:inference-sound} in
+Sec.~\ref{sec:inference-sound}):
 \begin{smathpar}
@@
 a union of effects ($\F$) for all $z\in x$, all applied to the same
-state ($\stl$ and $\stg$).  That is, $\F_{loop}(\stg) = \bigcup_{z\in
+state ($\stg$).  That is, $\F_{loop}(\stg) = \bigcup_{z\in
 x}\F_{body}(\stg)$. From the definition of the set monad's bind
@@
   \lambda(\stl,\stg).~ \stl \cup \C{item\_reqs}\bind
-      (\lambda\C{item\_req}.~ \F_U(\stl,\stg) \cup \F_I(\stl,\stg))
+      (\lambda\C{item\_req}.~ \F_U(\stg) \cup \F_I(\stg))
 \end{array}
@@
   \begin{array}{lcl}
-    \F_U & = & \lambda(\stl,\stg).~ \stg \bind(\lambda s. 
+    \F_U & = & \lambda(\stg).~ \stg \bind(\lambda s. 
                 \itel{{\sf table}(s) = \C{Stock} \conj 
                   s.\C{s\_i\_id} = \C{item\_req.ol\_i\_id}\\
-         &   & \hspace*{0.9in}}
+         &   & \hspace*{0.8in}}
                      {\{ \langle \C{s\_i\_id}=s.\C{s\_i\_id};\, 
@@
                                  \C{s\_qty} = \C{k1}\rangle \}\\
-         &   & \hspace*{0.9in}} {\emptyset}) \\
-    \F_I & = & \lambda(\stl,\stg).~ \{\langle\C{ol\_o\_id}=\C{k2},\,
+         &   & \hspace*{0.8in}} {\emptyset}) \\
+    \F_I & = & \lambda(\stg).~ \{\langle\C{ol\_o\_id}=\C{k2},\,
                  \C{ol\_d\_id}=\C{k3},\, \C{ol\_i\_id}=\C{item\_req.ol\_i\_id},\, 
@@
 \end{smathpar}
-For any $\stl$ and $\stg$, $\F_U(\stl,\stg)$ and $\F_I(\stl,\stg)$ are
-expressions in $\SL$, so can be translated to FOL by the encoding
-algorithm in Fig.~\ref{fig:logic}. Since the iteration variable
-\C{item\_req} occurs free in these expressions, the appropriate
-application of the encoding algorithm is
-$\mssemof{\C{item\_req}}{\F_U(\stl,\stg)}$ and
-$\mssemof{\C{item\_req}}{\F_I(\stl,\stg)}$, which results in
+For any $\stg$, $\F_U(\stg)$ and $\F_I(\stg)$ are expressions in
+$\SL$, so can be translated to FOL by the encoding algorithm in
+Fig.~\ref{fig:logic}. Since the iteration variable \C{item\_req}
+occurs free in these expressions, the appropriate application of the
+encoding algorithm is $\mssemof{\C{item\_req}}{\F_U(\stg)}$ and
+$\mssemof{\C{item\_req}}{\F_I(\stg)}$, which results in
 $(\phi_U,\G_U)$ and $(\phi_I,\G_I)$, respectively, where $\phi_U$,
@@
 expression nested inside a top-level bind expression (like
-$\F_U(\stl,\stg)$ from the above example), then the prenex signature
+$\F_U(\stg)$ from the above example), then the prenex signature
 of $\phi_s$ is $\forall^2\exists$.  Furthermore, if the subexpressions

### 2017-11-06 motivation.tex
Typos etc.

@@
 update statement (call it $\T_U$) is the following function on
-sets\footnote{$\bind$ has higher precedence than $\cup$; angle braces
-($\langle \ldots \rangle$) are used to denote records in math}:
+sets\footnote{Bind ($\bind$) has higher precedence than union
+($\cup$). Angle braces ($\langle \ldots \rangle$) are used to denote
+records in math}:
 \begin{smathpar}
@@
 Observe that both transformers are of the form $\T(\delta,\Delta) =
-\stl \cup \F(\stl,\stg)$, where $\F$ is a function that returns the
+\stl \cup \F(\stg)$, where $\F$ is a function that returns the
 set of records added  to the transaction-local database ($\stl$). Let
@@
 \begin{array}{l}
-  \lambda(\stl,\stg).~ \stl \cup \F_U(\stl,\stg) \cup \F_I(\stl,\stg)
+  \lambda(\stl,\stg).~ \stl \cup \F_U(\stg) \cup \F_I(\stg)
 \end{array}

### 2017-11-06 evaluation.tex
edits

@@
 insert a new \C{Enrollment} record by stating that $\stg_e$, the
-\C{Enrollment} table in pre-state, contains all records $e$ from
-$\stg_e'$, the table in post-state, except when $e.\C{e\_c\_id}$ and
+\C{Enrollment} table in the pre-state, contains all records $e$ from
+$\stg_e'$, the table in the post-state, except when $e.\C{e\_c\_id}$ and
 $e.\C{e\_s\_id}$ match \C{cid} and \C{sid}, respectively. The

### 2017-11-06 implementation.tex
edits

@@
 correct, then the verification is guaranteed to succeed. Within the
-loop, the \thetool first computes the various rely relations needed for
+loop, \thetool first computes the various rely relations needed for
 verification ($R$, $\R_l$, and $\R_c$). It then traverses the AST of a
 transaction, applying the inference rules to construct a state
-transformer, check its stability, and weaken it ($\stabilize{\cdot}$)
+transformer, checks its stability, and weakens it ($\stabilize{\cdot}$)
 if it is not stable. The result of traversing the transaction's AST is
@@
 check passes, then the guarantee is verified to check if it preserves
-the invariant $I$. The successful result from both the checks results
+the invariant $I$. The successful result from both checks results
 in the transaction being certified correct under the current choice of
@@
 need to be taken into account for verification to be effective in
-practice.  We consider these into account while linking \thetool with
-store-specific semantics (isolation specifications etc). As an
+practice.  We take these into account while linking \thetool with
+store-specific semantics (isolation specifications, etc.). As an
 example, consider how PostgreSQL implements an \C{UPDATE} operation.
@@
 actual database (if they still exist), write locks are obtained, and
-the update is performed, provided that the record still meets
-\C{UPDATE}'s search criteria. If the record no longer meets the
-search criteria (due to a concurrent update) the record is excluded
+the update is performed, provided that each matched record still meets
+\C{UPDATE}'s search criteria. If a record no longer meets the
+search criteria (due to a concurrent update), it is excluded
 from the update, and the write lock is immediately released.
@@
 Clearly, this sequence of events is not atomic, unlike the assumption
-made by our formal model.  The implementation admits interference
+made by our formal model because  the implementation admits interference
 between the updates of individual records that meet the search
@@
 \C{MAX} and \C{MIN}, e.g., \C{SELECT MAX(ol\_o\_id) FROM
-Order\_line WHERE $\ldots$}. We admit such extensions as set functions
+Order\_line WHERE $\ldots$}, etc. We admit such extensions as set functions
 in our DSL (e.g., \C{project}, \C{max}, \C{min}), and axiomatize their

### 2017-11-06 inference.tex
edits

@@
 \begin{theorem}
-  There exists an $\SL'[\Prop^0,\Prop^1] \subset \SL[\Prop^0,\Prop^1]$
+  There exist $\SL'[\Prop^0,\Prop^1] \subset \SL[\Prop^0,\Prop^1]$
   such that $\SL'$ is decidable if ${\sf GKS}+\Prop^0+\Prop^1$ is

### 2017-11-07 introduction.tex
edits

--- 2017-10-22 introduction.tex
@@
 imposed by these rules may discourage applications programmers from
-using them, we also present an inference procedure based on these
-rules that automatically verifies the weakest isolation level for a
-transaction that still ensures its invariants are maintained.  The key
-to inference is a novel formulation of database state (represented as
-sets of tuples) as a monad, and in which database computations are
-interpreted as state transformers over these sets.  This
-interpretation leads to an encoding of database computations amenable
-for verification by off-the-shelf SMT solvers.  The paper makes the
-following contributions:
+using them, we also present an inference procedure that automatically
+verifies the weakest isolation level for a transaction while ensuring
+its invariants are maintained.  The key to inference is a novel
+formulation of database state (represented as sets of tuples) as a
+monad, and in which database computations are interpreted as state
+transformers over these sets.  This interpretation leads to an
+encoding of database computations amenable for verification by
+off-the-shelf SMT solvers.  The paper makes the following
+contributions:
 %% We have realized these ideas as an embedded DSL within OCaml, that
@@
 \end{enumerate}
-\noindent Our results provide the first (to the best of our knowledge)
-formalization of weakly-isolated transactions, along with an
-expressive and compositional proof automation framework capable of
-verifying the safety of high-level consistency conditions attached to
-these transactions.  Collectively, these contributions allow
-weakly-isolated transactions to enjoy the same rigorous reasoning
-capabilities as their strongly-isolated (serializable) counterparts.
+\noindent Our results provide the first formalization of
+weakly-isolated transactions, along with an expressive and
+compositional proof automation framework capable of verifying the
+safety of high-level consistency conditions attached to these
+transactions.  Collectively, these contributions allow weakly-isolated
+transactions to enjoy the same rigorous reasoning capabilities as
+their strongly-isolated (serializable) counterparts.
 
@@
 and benchmark results in \S\ref{sec:case-studies}.  Related work is
-given in \S\ref{sec:relatedwork}, and \S\ref{sec:conclusions} presents
-conclusions.
+given in \S\ref{sec:relatedwork}, and \S\ref{sec:conclusions} concludes.

### 2017-11-07 motivation.tex
edits

@@
 We present our ideas in the context of a DSL embedded in OCaml that
-manages an abstract database state, and allows it be manipulated via a
+manages an abstract database state that can be manipulated via a
 well-defined \C{SQL} interface. Arbitrary database computations can be
@@
 (racy) programs~\cite{GHE15,HPQ+15}.  To illustrate, consider the
-behavior of the \C{new\_order} transaction when executed under
+behavior of the \C{new\_order} transaction when executed under a
 \emph{Read Committed} (RC) isolation level, the default isolation
@@
 before it increments the field, another \C{new\_order} instance
-($T_2$) begins its execution and commits. Note that $T_2$ reads the
+$T_2$ (red) begins its execution and commits. Note that $T_2$ reads the
 same \C{d\_next\_o\_id} value as $T_1$, and inserts new \C{Order} and
@@
 and adds new \C{Order} and \C{Order\_line} fields with the same order
-id as $T_1$. Thus, by the end of the execution, \C{Order\_line}
+id as $T_1$. Thus, at the end of the execution, \C{Order\_line}
 records inserted by $T_1$ and $T_2$ all bear the same order id. There
@@
 
-Clearly, this example does not exhibit any of the anomalies that
+This example does not exhibit any of the anomalies that
 \emph{characterize} RC isolation~\cite{berenson}\footnote{Berenson et
 al~\cite{berenson} characterize isolation levels in terms of the
-anomalies they \emph{admit}. For e.g., RC is characterized by
+anomalies they \emph{admit}. For example, RC is characterized by
 \emph{lost writes} because it admits the anomaly.}. For instance,
-there are no \emph{lost writes} - both concurrent transactions' writes
+there are no \emph{lost writes} since both concurrent transactions' writes
 are present in the final state of the database.  Program analyses that

### 2017-11-07 reasoning.tex
edits

@@
 \noindent We can now assert the soundness of the RG judgments in
-Fig.~\ref{fig:rg-rules} as follows\footnote{Full proofs for theorems
-in this paper can be found in the accompanying tech report}:
+Fig.~\ref{fig:rg-rules} as follows\footnote{Full proofs for the major
+  theorems and lemmas defined in this paper are available from
+  ~\cite{KN+18_arxiv}.}:
 

### 2017-11-07 related-work.tex
edits

@@
-\vspace*{-6pt}
 \section{Related Work}
@@
 the declarative nature of our specifications make them suitable for
-formal reasoning about program behavior. Sivaramakrishnan \emph{et
-al.}~\cite{pldi15} specify isolation levels declaratively as trace
-well-formedness conditions, but their specifications implicitly assume
-a complete trace with only committed transactions, making it difficult
-to reason about a program as it builds the trace. Cerone \emph{et
-al.}~\cite{gotsmanconcur15} specify isolation levels with atomic
-visibility, but their specifications are also for complete traces.
-Both the aforementioned specification frameworks use the vocabulary
-introduced in~\cite{burckhardt14}. However, none of them are equipped
-with a reasoning framework that can use such specifications to verify
-programs under weak isolation.
+formal reasoning about program behavior. ~\cite{pldi15} specify
+isolation levels declaratively as trace well-formedness conditions,
+but their specifications implicitly assume a complete trace with only
+committed transactions, making it difficult to reason about a program
+as it builds the trace. ~\cite{gotsmanconcur15} specify isolation
+levels with atomic visibility, but their specifications are also for
+complete traces.  Both the aforementioned specification frameworks use
+the vocabulary introduced in~\cite{burckhardt14}. However, none of
+them are equipped with a reasoning framework that can use such
+specifications to verify programs under weak isolation.
 
-Recent work by Crooks \emph{et al.}~\cite{CPA+17} also explores the
-use of a state-based interpretation of isolation as we do, and share
-our position that enables specification of weak isolation that are not
+Recent work described in~\cite{CPA+17} also explores the
+use of a state-based interpretation of isolation as we do, and 
+like our approach, develops specifications of weak isolation that are not
 tied to implementation-specific artifacts.  However, they do not
@@
 
-\vspace*{-4pt}
-\paragraph{Reasoning under weak isolation.} In~\cite{feketessi}, Fekete
-\emph{et al.} propose a theory to characterize non-serializable
-executions that arise under {\sc si}. Fekete~\cite{fekete2005} also
-proposes an algorithm that allocates either {\sc si} or {\sc ser}
-isolation levels to transactions while guaranteeing
-serializability. In~\cite{gotsmanpodc16}, Cerone \emph{et al.} improve
+\paragraph{Reasoning under weak isolation.} ~\cite{feketessi} propose
+a theory to characterize non-serializable executions that arise under
+{\sc si}. They also propose an algorithm that
+allocates either {\sc si} or {\sc ser} isolation levels to
+transactions while guaranteeing
+serializability. ~\cite{gotsmanpodc16} improve
 on Adya's {\sc si} specification and use it to derive a static
@@
 techniques are tailor-made for a finite set of well-understood
-isolation levels (rooted in~\cite{berenson}) under a pre-defined store
-consistency model.
+isolation levels (rooted in~\cite{berenson}).
 
-\vspace*{-4pt}
 \paragraph{Reasoning under weak consistency.} There have been several
@@
 our focus in on reasoning in the presence of weakly-isolated
-transactions under a strongly consistent store.  Gotsman \emph{et
-al.}~\cite{gotsmanpopl16} adapt \iso{Parallel Snapshot Isolation} to a
-transaction-less setting by interpreting it as a consistency level
-that serializes writes to objects; a dedicated proof rule is developed
-to help prove prove program invariants hold under this model. By
-parameterizing our proof system over a gamut of weak isolation
-specifications, we avoid the need to define a separate proof rule for
-each new isolation level we may encounter.
+transactions under a strongly consistent store.  ~\cite{gotsmanpopl16}
+adapt \iso{Parallel Snapshot Isolation} to a transaction-less setting
+by interpreting it as a consistency level that serializes writes to
+objects; a dedicated proof rule is developed to help prove prove
+program invariants hold under this model. By parameterizing our proof
+system over a gamut of weak isolation specifications, we avoid the
+need to define a separate proof rule for each new isolation level we
+may encounter.
 
@@
 
-\vspace*{-4pt}
-\paragraph{Inference.}  Prior work by Vafeiadis~\cite{Vaf10,Vaf10a}
+\paragraph{Inference.}  ~\cite{Vaf10,Vaf10a}
 describe \emph{action inference}, an inference procedure for computing

### 2017-11-08 evaluation.tex
Formatting changed as per PACM rules.

@@
 
-  let deregister_student_txn sid = 
+  let deregister_txn sid = 
     let s_enrs = SQL.select [Enrollment] (fun e -> e.e_s_id = sid) in
@@
 added (via an \C{add\_course} transaction), and new students to be
-registered (via a \C{register\_student} transaction) into a database. A registered
+registered (via a \C{register} transaction) into a database. A registered
 student can enroll (\C{enroll}) in an existing course,
@@
 canceled (\C{cancel\_course}). Likewise, a student who is not enrolled
-in any course can be deregistered (\C{deregister\_student}). Besides
+in any course can be deregistered (\C{deregister}). Besides
 \C{Student} and \C{Course} tables, there is also an \C{Enrollment}
@@
 courses and students. The simplified code for the Courseware
-application with only \C{enroll} and \C{deregister\_student}
+application with only \C{enroll} and \C{deregister}
 transactions is shown in Fig.~\ref{fig:courseware_code}. The
@@
 \noindent Both $I_1$ and $I_2$ can be violated under weak isolation.
-$I_1$ can be violated, for example, when \C{deregister\_student} runs
+$I_1$ can be violated, for example, when \C{deregister} runs
 concurrently with \C{enroll}, both at RC isolation. While the former
@@
 correctness condition. The guarantees $G_e$ and $G_d$ for \C{enroll}
-and \C{deregister\_student} transactions, respectively, are shown below:
+and \C{deregister} transactions, respectively, are shown below:
 \begin{smathpar}
@@
           {\existsl(c',~c'.\C{id}=c.\C{id} \wedge
-              c'.\C{c\_name}=c.\C{c\_name} \wedge c'.\C{c\_capacity}\ge0,
-              ~\{c'\})\\
+              c'.\C{c\_name}=c.\C{c\_name} \\
+    & & \hspace*{2.6in}\wedge~ c'.\C{c\_capacity}\ge0, ~\{c'\})\\
     & & \hspace*{1.15in}}
@@
 $e.\C{e\_s\_id}$ match \C{cid} and \C{sid}, respectively. The
-guarantee $G_d$ of \C{deregister\_student} asserts that the
+guarantee $G_d$ of \C{deregister} asserts that the
 transaction doesn't write to \C{Course} and \C{Enrollment} tables. The
@@
 be preempted by promoting the isolation level of \C{enroll} and
-\C{deregister\_student} to SER (on both MySQL and PostgreSQL), leaving
+\C{deregister} to SER (on both MySQL and PostgreSQL), leaving
 the isolation levels of remaining transactions at RC. The total time

### 2017-11-08 implementation.tex
Formatting changed as per PACM rules.

@@
 ocamlc 4.03 compiler\footnote{The source code is available at
-available at \emph{https://goo.gl/nmUXZK}}. The input to \thetool is a
-program in our DSL that describes the schema of the database as a
-collection of OCaml type definitions, and transactions as OCaml
-functions, whose top-level expression is an application of the
-\C{atomically\_do} combinator. For instance, TPC-C's schema from
+available at \url{https://github.com/gowthamk/acidifier}}. The input
+to \thetool is a program in our DSL that describes the schema of the
+database as a collection of OCaml type definitions, and transactions
+as OCaml functions, whose top-level expression is an application of
+the \C{atomically\_do} combinator. For instance, TPC-C's schema from
 Fig.~\ref{fig:schema} can be described via the OCaml type definitions

### 2017-11-08 inference.tex
edits
Formatting changed as per PACM rules.

@@
 \emph{effect} part of the state transformer introduced in
-Sec.~\ref{sec:motivation}, which is a function $\T$ of form
-$\lambda(\stl,\stg).~\stl \cup \F(\stg)$. Nonetheless, we keep
-referring to $\F$ as state transformer. Since the execution is subject
+Sec.~\ref{sec:motivation}, which is a function $\T$ of the form
+$\lambda(\stl,\stg).~\stl \cup \F(\stg)$. Nonetheless, for simplicity,
+we will continue to refer to $\F$ as state transformer. Since the execution is subject
 to isolation-constrained interference, the inference judgment depends
@@
          &   & \hspace*{0.8in}} {\emptyset}) \\
-    \F_I & = & \lambda(\stg).~ \{\langle\C{ol\_o\_id}=\C{k2},\,
-                 \C{ol\_d\_id}=\C{k3},\, \C{ol\_i\_id}=\C{item\_req.ol\_i\_id},\, 
-                 \C{ol\_qty}=\C{item\_req.ol\_qty}\rangle\}\\
+    \F_I & = & \lambda(\stg).~ \{\langle\C{ol\_o\_id}=\C{k2};\,
+                 \C{ol\_d\_id}=\C{k3};\,
+                 \C{ol\_i\_id}=\C{item\_req.ol\_i\_id};\,\\
+         &   & \hspace*{2.3in}\C{ol\_qty}=\C{item\_req.ol\_qty}\rangle\}\\
   \end{array}
@@
   \begin{array}{lcl}
-    \phi_U & = & \hspace*{0.2in}\forall \C{item\_req}.\forall s. \forall s'.~ 
+    \phi_U & = & \hspace*{0.1in}\forall \C{item\_req}.\forall s. \forall s'.~ 
         \pi_1(\C{item\_req}) \Leftrightarrow \\
@@
     & & \hspace*{0.85in}} {\bot}) \Rightarrow g_0(\C{item\_req},s')\\
-    & & \conj \hspace*{0.08in}\forall \C{item\_req}.\forall s'. \exists s.~ 
+    & & \hspace*{-0.05in}\conj \forall \C{item\_req}.\forall s'. \exists s.~ 
         \pi_2(\C{item\_req}) \Leftrightarrow \\
-    & & \hspace*{0.28in} g_0(\C{item\_req},s') \Rightarrow (s \in \stg) \conj
-        (\itec{{\sf table}(s)=\C{Stock} \wedge 
+    & & \hspace*{0.15in} g_0(\C{item\_req},s') \Rightarrow s \in \stg \conj
+        \itec{{\sf table}(s)=\C{Stock} \wedge 
               s.\C{s\_i\_id} = \C{item\_req.ol\_i\_id}\\
-    & & \hspace*{1.8in}}
+    & & \hspace*{1.6in}}
               {s' = \langle \C{s\_i\_id}=s.\C{s\_i\_id};\, 
@@
                             \C{s\_qty} = \C{k1}\rangle\\
-    & & \hspace*{1.8in}} {\bot}) \\
+    & & \hspace*{1.6in}} {\bot} \\
     \G_U & = & \lambda (\C{item\_req},r).~ \pi_1(\C{item\_req}) \wedge 
@@
     \G_I & = & \lambda (\C{item\_req},r).~ 
-        r = \langle \C{ol\_o\_id}=\C{k2},\,
-                    \C{ol\_d\_id}=\C{k3},\, 
-                    \C{ol\_i\_id}=\C{item\_req.ol\_i\_id},\, 
-                    \C{ol\_qty}=\C{item\_req.ol\_qty}
-             \rangle\\
+        r = \langle \C{ol\_o\_id}=\C{k2};\,
+                    \C{ol\_d\_id}=\C{k3};\, 
+                    \C{ol\_i\_id}=\C{item\_req.ol\_i\_id};\,\\
+    & & \hspace*{3in}\C{ol\_qty}=\C{item\_req.ol\_qty} \rangle\\
   \end{array}
@@
 $\mssemof{\emptyset}{\T(\stl,\stg)}$, which results in the
-$(\phi,\G)$, as defined below:
+$(\phi_I \wedge \phi_U \wedge \phi_1 \wedge \phi_2,\G)$, where
+$\phi_1$, $\phi_2$, and $\G$ are as defined below:
 \begin{smathpar}
   \begin{array}{lcl}
-    \phi & = & \phi_I \conj \phi_U \conj \\
-    & & \hspace*{0.2in}\forall \C{item\_req}. \forall s.~ 
-        \pi_3 \Leftrightarrow (\C{item\_req} \in \C{item\_reqs} \conj 
-        (\G_U(\C{item\_req},s') \vee \G_I(\C{item\_req},s'))
-        \Rightarrow g_1(s))\\
-    & & \conj \hspace*{0.05in}\forall s.\exists \C{item\_req}. ~ 
-        \pi_4 \Leftrightarrow (g_1(s) \Rightarrow (\C{item\_req} \in \C{item\_reqs} 
-        \conj (\G_U(\C{item\_req},s') \vee \G_I(\C{item\_req},s')))\\
+    \phi_1 & = & \forall \C{item\_req}. \forall s.~ 
+        \pi_3 \Leftrightarrow \C{item\_req} \in \C{item\_reqs} \wedge
+        \G_U(\C{item\_req},s') \vee \G_I(\C{item\_req},s')
+        \Rightarrow g_1(s)\\
+    \phi_2 & = & \forall s.\exists \C{item\_req}. ~ 
+        \pi_4 \Leftrightarrow g_1(s) \Rightarrow \C{item\_req} \in \C{item\_reqs} 
+        \wedge \G_U(\C{item\_req},s') \vee \G_I(\C{item\_req},s')\\
     \G & = & \lambda(r).~\pi_3 \wedge \pi_4 \wedge g_1(r)\\

### 2017-11-08 motivation.tex
Formatting changed as per PACM rules.

@@
 	\end{subfigure}
-  \hspace*{.4in}
+  \hspace*{.2in}
 	\begin{subfigure}{0.52\textwidth}
@@
 
+\begin{wrapfigure}{l}{.4\textwidth}
+\includegraphics[scale=0.45]{Figures/motiv-eg-1-b}
+\caption{\small An RC execution involving two instances ($T_1$ and
+  $T_2$) of the \C{new\_order} transaction depicted in
+  Fig.~\ref{fig:new_order_code}. 
+  Both instances read the \C{d\_id} \C{District} record concurrently,
+  because neither transaction is committed when the reads are
+  executed.  The subsequent operations are effectively sequentialized,
+  since $T_2$ commits before $T_1$. Nonetheless, both transactions read the same value for
+  \C{d\_next\_o\_id} resulting in them  adding \C{Order} records
+  with the same ids, which in turn triggers a violation of TPC-C's
+  consistency condition.}
+\label{fig:new_order_exec}
+\end{wrapfigure}
 Because the aforementioned reasoning is reasonably simple to perform
@@
 level in 8 of the 18 databases studied in~\cite{bailishotos}.
-\begin{wrapfigure}{l}{.4\textwidth}
-\includegraphics[scale=0.45]{Figures/motiv-eg-1-b}
-\caption{\small An RC execution involving two instances ($T_1$ and
-  $T_2$) of the \C{new\_order} transaction depicted in
-  Fig.~\ref{fig:new_order_code}. 
-  Both instances read the \C{d\_id} \C{District} record concurrently,
-  because neither transaction is committed when the reads are
-  executed.  The subsequent operations are effectively sequentialized,
-  since $T_2$ commits before $T_1$. Nonetheless, both transactions read the same value for
-  \C{d\_next\_o\_id} resulting in them  adding \C{Order} records
-  with the same ids, which in turn triggers a violation of TPC-C's
-  consistency condition.}
-\label{fig:new_order_exec}
-\end{wrapfigure}
 An executing RC transaction is isolated from \emph{dirty writes},
@@
   \lambda(\stl,\stg).~ \stl \cup
-       \{\langle\C{ol\_o\_id}=\C{k2},\,
-                 \C{ol\_d\_id}=\C{k3},\, \C{ol\_i\_id}=\C{item\_req.ol\_i\_id},\, 
+       \{\langle\C{ol\_o\_id}=\C{k2};\,
+                 \C{ol\_d\_id}=\C{k3};\, \C{ol\_i\_id}=\C{item\_req.ol\_i\_id};\, 
                  \C{ol\_qty}=\C{item\_req.ol\_qty}\rangle\}

### 2017-11-08 inference.tex
Updated proof sketch in Section 5

--- 2017-11-06 inference.tex
@@
   \label{thm:inference-sound}
-  For all $i$,$R$,$I$,$c$,$\Fx$, $\F$, if $\stable(\R,I)$ and $\Fx
+  For all $i$,$R$,$I$,$c$,$\Fx$, $\F$, if $\stable(\R,I)$, $\stable(\R, \Fx)$ and $\Fx
   \vdash c \elabsto \F$, then:\\\vspace*{-0.2cm}
@@
     \R \vdash \hoare{\lambda(\stl,\stg).~\stl=\Fx(\stg) \conj
-    I(\stg)}{c}{\lambda(\stl,\stg).~\stl = \Fx(\stg) \cup \F(\stg)
-    \conj I(\stg)}
+    I(\stg)}{c}{\lambda(\stl,\stg).~\stl = \Fx(\stg) \cup \F(\stg)}
   \end{array}
@@
 The proof follows by structural induction on $c$. Assume that $P
-\Leftrightarrow \lambda(\stl,\stg).~\stl=s \conj I(\stg)$ and $Q
-\Leftrightarrow \lambda(\stl,\stg).\stl = \F(s,\stg)$. For each type
+\Leftrightarrow \lambda(\stl,\stg).~\stl=\Fx(\stg) \conj I(\stg)$ and $Q
+\Leftrightarrow \lambda(\stl,\stg).\stl = \Fx(\stg) \cup \F(\stg)$. For each type
 of $c$, our strategy is to show the premises of the corresponding
@@
 to the condition linking the pre-condition $P$ and post-condition $Q$
-in the premises of the RG-rules. If $\stable(\R, \F)$, then
-$\stabilize{\F} = \F$ and hence the proof directly follows. Otherwise,
-$\stabilize{\F} = \lambda (\stg).\existsl(\stg', I(\stg'),
+in the premises of the RG-rules. If $\stable(\R, \Fx[\F])$, then
+$\stabilize{\Fx[\F]} = \F$ and hence the proof directly follows. Otherwise,
+$\stabilize{\Fx[\F]} =$ \\%\hspace{2cm}
+ $\lambda (\stg).\existsl(\stg', I(\stg'),
 \F( \stg'))$. Since the pre-condition $P(\stl, \stg)$ asserts
@@
 matches the condition linking $P$ and $Q$. The stability of $P$ with
-respect to $\R$ follows directly from the stability of $I$.
+respect to $\R$ follows directly from $\stable(\R, I)$ and $\stable(\R, \Fx)$.
 
@@
 \C{FOREACH} statement, for which we use the loop invariant
-$\psi(\stl, \stg) \Leftrightarrow \stl = s \cup (y \bind (\lambda
-z.~\F(s, \stg)))$, (where assuming that $c$ is the body of the loop,
+$\psi(\stl, \stg) \Leftrightarrow \stl = \Fx(\stg) \cup (y \bind (\lambda
+z.~\F(\stg)))$, (where assuming that $c$ is the body of the loop,
 $c \elabsto \F$) to prove all the premises of
@@
 being processed in the current iteration.  At the beginning of the
-loop $[\phi/y]\psi(\stl, \stg)$ just reduces to $\stl = s$ which is
+loop $[\phi/y]\psi(\stl, \stg)$ just reduces to $\stl = \Fx(\stg)$ which is
 implied by the pre-condition $P$. From the inductive hypothesis, we

### 2017-11-08 evaluation.tex
edits

@@
 different applications: \emph{Courseware}: a course registration
-system described by~\cite{gotsmanpopl16}, and the TPC-C benchmark.
+system described by~\cite{gotsmanpopl16}, and TPC-C.
 
@@
        SQL.update Course (fun c -> {c with c_capacity = c.c_capacity - 1})
-                         (fun c -> c.c_id = cid))
-    else ()
+                         (fun c -> c.c_id = cid)) else ()
 
@@
     if Set.is_empty s_enrs then
-      SQL.delete Student (fun s -> s.s_id = sid)
-    else ()
+      SQL.delete Student (fun s -> s.s_id = sid) else ()
 \end{ocaml}
@@
 \label{tab:tpcc}
+\vspace*{-10pt}
 \end{table}

### 2017-11-08 implementation.tex
edits

@@
 concludes the top-level execution, returning the inferred isolation
-levels as its output.
-
-\thetool uses the Z3 SMT solver as its underlying reasoning engine. Each
+levels as its output.  \thetool uses the Z3 SMT solver as its underlying reasoning engine. Each
 implication check described above is first encoded in FOL, applying
@@
 record only if a record with the same id is present in the set of records
-returned by the \C{SELECT}. This transformation is summarized below:
+returned by \C{SELECT}: % This transformation is summarized below:
 \begin{smathpar}
@@
 completely axiomatized in FOL. These include \C{sum}, \C{count} etc.
-For these, we admit imprecise axiomatizations expressed as lemmas
-over these functions.
+For these, we admit imprecise axiomatizations. % expressed as lemmas over these functions.
 

### 2017-11-08 introduction.tex
edits

@@
 with timestamp and version management~\cite{BG81}.  These issues are
-further exacerbated when the database is replicated, requiring
+exacerbated when the database is replicated, requiring
 additional coordination

### 2017-11-08 related-work.tex
edits

@@
 while capable of identifying database attacks due to the use of
-incorrect (excessively weak) isolation levels, does not consider how
+incorrect isolation levels, does not consider how
 to verify application correctness, infer proper isolation levels, or
@@
 and application invariants.
-
 \paragraph{Reasoning under weak isolation.} ~\cite{feketessi} propose
@@
 isolation levels (rooted in~\cite{berenson}).
-
+\vspace*{-6pt}
 \paragraph{Reasoning under weak consistency.} There have been several
@@
 may encounter.
-
 %% \vspace*{-4pt}
@@
 %% consistent with the TSO memory model.
-
 \paragraph{Inference.}  ~\cite{Vaf10,Vaf10a}
@@
 vs. abstract interpretation) are quite different.
-
+\vspace*{-7pt}
 \section{Conclusions}
@@
 enables database applications to leverage the performance advantages
-offered by weak isolation, without compromising safety and
-correctness.
+offered by weak isolation, without compromising correctness.

### 2017-11-08 inference.tex
Formatting fixed.

@@
   Sec.~\ref{sec:motivation}, which is a function $\T$ of form
-$\lambda(\stl,\stg).~\stl \cup \F(\stg)$. Nonetheless, we keep
-referring to $\F$ as state transformer. Since the execution is subject
-to isolation-constrained interference, the inference judgment depends
-on the isolation-constrained rely relation $\R$, which is used to
-enforce the stability of the state transformer $\F$.  Recall that $\R$
-is a tri-state rely relation over $\stl$, $\stg$ and $\stg'$, that
-admits an interference from $\stg$ and $\stg'$ depending on the local
+  $\lambda(\stl,\stg).~\stl \cup \F(\stg)$. Nonetheless, for
+  simplicity, we will continue to refer to $\F$ as state transformer.
+  Since the execution is subject to isolation-constrained
+  interference, the inference judgment depends on the
+  isolation-constrained rely relation $\R$, which is used to enforce
+  the stability of the state transformer $\F$.  Recall that $\R$ is a
+  tri-state rely relation over $\stl$, $\stg$ and $\stg'$, that admits
+  an interference from $\stg$ and $\stg'$ depending on the local
   database state $\stl$. Thus, the stability of the state transformer
   $\F$ of $c$ with respect to $\R$ needs to take into account the
-(possible) prior state of the local database $\stl$, which depends on
-the context (sequence of previous commands) of $c$, and computed by
-the corresponding state transformer $\Fx$. Thus, the semantics of the
-state transformer can be understood in terms of the RG judgment as
-following (formalized as Theorem~\ref{thm:inference-sound} in
+  (possible) prior state of the local database $\stl$, which depends
+  on the context (sequence of previous commands) of $c$, and computed
+  by the corresponding state transformer $\Fx$. Thus, the semantics of
+  the state transformer can be understood in terms of the RG judgment
+  as following (formalized as Theorem~\ref{thm:inference-sound} in
   Sec.~\ref{sec:inference-sound}):
@@
     & & \hspace*{0.85in}} {\bot}) \Rightarrow g_0(\C{item\_req},s')\\
-    & & \conj \hspace*{0.08in}\forall \C{item\_req}.\forall s'. \exists s.~ 
+    & & \hspace*{-0.05in}\conj \forall \C{item\_req}.\forall s'. \exists s.~ 
         \pi_2(\C{item\_req}) \Leftrightarrow \\
-    & & \hspace*{0.28in} g_0(\C{item\_req},s') \Rightarrow (s \in \stg) \conj
-        (\itec{{\sf table}(s)=\C{Stock} \wedge 
+    & & \hspace*{0.15in} g_0(\C{item\_req},s') \Rightarrow s \in \stg \conj
+        \itec{{\sf table}(s)=\C{Stock} \wedge 
               s.\C{s\_i\_id} = \C{item\_req.ol\_i\_id}\\
-    & & \hspace*{1.8in}}
+    & & \hspace*{1.5in}}
               {s' = \langle \C{s\_i\_id}=s.\C{s\_i\_id};\, 
@@
                             \C{s\_qty} = \C{k1}\rangle\\
-    & & \hspace*{1.8in}} {\bot}) \\
+    & & \hspace*{1.5in}} {\bot} \\
     \G_U & = & \lambda (\C{item\_req},r).~ \pi_1(\C{item\_req}) \wedge 

### 2017-11-08 introduction.tex
Formatting fixed.

@@
 been echoed by database researchers who lament the lack of a
-better understanding of this problem; see e.g., {\tt
+better understanding of this problem; see e.g., \url{
   http://www.bailis.org/blog/understanding-weak-isolation-is-a-serious-problem}.}

### 2017-11-08 motivation.tex
Formatting fixed.

@@
 
-We present our ideas in the context of a DSL embedded in OCaml that
-manages an abstract database state that can be manipulated via a
-well-defined \C{SQL} interface. Arbitrary database computations can be
-built around this interface, which can then be run as transactions
-using the \C{atomically\_do} combinator provided by the DSL.
-
-\begin{figure}
+\begin{figure}[!t]
 \centering
@@
 \end{figure}
+
+We present our ideas in the context of a DSL embedded in OCaml that
+manages an abstract database state that can be manipulated via a
+well-defined \C{SQL} interface. Arbitrary database computations can be
+built around this interface, which can then be run as transactions
+using the \C{atomically\_do} combinator provided by the DSL.
 
@@
 \emph{characterize} RC isolation~\cite{berenson}\footnote{Berenson et
-al~\cite{berenson} characterize isolation levels in terms of the
+al characterize isolation levels in terms of the
 anomalies they \emph{admit}. For example, RC is characterized by

### 2017-11-08 inference.tex
Edits to fit to 31 pages.

@@
 %   \end{smathpar}
-The proof follows by structural induction on $c$. Assume that $P
-\Leftrightarrow \lambda(\stl,\stg).~\stl=\Fx(\stg) \conj I(\stg)$ and $Q
-\Leftrightarrow \lambda(\stl,\stg).\stl = \Fx(\stg) \cup \F(\stg)$. For each type
-of $c$, our strategy is to show the premises of the corresponding
-RG-rule. The base cases correspond to \C{INSERT}, \C{UPDATE} and \C{DELETE}
-statements. In all these cases, the form of $\F$ directly corresponds
-to the condition linking the pre-condition $P$ and post-condition $Q$
-in the premises of the RG-rules. If $\stable(\R, \Fx[\F])$, then
-$\stabilize{\Fx[\F]} = \F$ and hence the proof directly follows. Otherwise,
-$\stabilize{\Fx[\F]} =$ \\%\hspace{2cm}
- $\lambda (\stg).\existsl(\stg', I(\stg'),
-\F( \stg'))$. Since the pre-condition $P(\stl, \stg)$ asserts
-$I(\stg)$, any state which obeys $P$ provides the existential $\stg'$
-in the above set expression, and hence $\F$ applied on that state
-matches the condition linking $P$ and $Q$. The stability of $P$ with
-respect to $\R$ follows directly from $\stable(\R, I)$ and $\stable(\R, \Fx)$.
-
-The proofs for \C{SELECT}, sequencing, and conditionals use the
-inductive hypothesis to infer the RG-judgments present in the
-premises of their corresponding RG-rules. The interesting case is the
-\C{FOREACH} statement, for which we use the loop invariant
-$\psi(\stl, \stg) \Leftrightarrow \stl = \Fx(\stg) \cup (y \bind (\lambda
-z.~\F(\stg)))$, (where assuming that $c$ is the body of the loop,
-$c \elabsto \F$) to prove all the premises of
+The proof follows by structural induction on $c$. Let $P
+=\lambda(\stl,\stg).~\stl=\Fx(\stg) \conj I(\stg)$ and $Q
+=\lambda(\stl,\stg).\stl = \Fx(\stg) \cup \F(\stg)$. The base cases
+correspond to \C{INSERT}, \C{UPDATE} and \C{DELETE} statements, where
+the proof is straightforward. The proofs for \C{SELECT}, sequencing,
+and conditionals use the inductive hypothesis to infer the
+RG-judgments present in the premises of their corresponding RG-rules.
+The interesting case is the \C{FOREACH} statement, for which we use
+the loop invariant $\psi(\stl, \stg) \Leftrightarrow \stl = \Fx(\stg)
+\cup (y \bind (\lambda z.~\F(\stg)))$, (where assuming that $c$ is the
+body of the loop, $c \elabsto \F$) to prove all the premises of
 \rulelabel{RG-Foreach}. Using the same notation as the rule
@@
 being processed in the current iteration.  At the beginning of the
-loop $[\phi/y]\psi(\stl, \stg)$ just reduces to $\stl = \Fx(\stg)$ which is
-implied by the pre-condition $P$. From the inductive hypothesis, we
-can infer that each iteration corresponds to the application of
-$\F$. Since all iterations are assumed to be independent of each other,
-and $z$ is bound to a record in $x$ for each iteration, we conclude
-that at the end of every iteration, the loop invariant $[y \cup
-  \{z\}/y]\psi$ will be satisfied.
+loop $[\phi/y]\psi(\stl, \stg)$ just reduces to $\stl = \Fx(\stg)$
+which is implied by the pre-condition $P$. From the inductive
+hypothesis, we can infer that each iteration corresponds to the
+application of $\F$. Since all iterations are assumed to be
+independent of each other, and $z$ is bound to a record in $x$ for
+each iteration, we conclude that at the end of every iteration, the
+loop invariant $[y \cup \{z\}/y]\psi$ will be satisfied.
 
@@
    \stable(R,\I)\spc
-   \stable(R,I)\\
-   \R_e = R \backslash \I_e \spc \R_c = R \backslash \I_c \spc
+   \stable(R,I)\spc
+   \R_e = R \backslash \I_e \spc 
+   \R_c = R \backslash \I_c \spc
    \Fempty \vdash c \Longrightarrow_{\langle i,\R_e,I \rangle}\F \\
@@
 be encoded as the function $\lambda \v. g(\v)$ with an associated
-constraint ($\phi\in\Prop^1$) defining the uninterpreted boolean
-function (i.e., uninterpreted predicate) $g$:
-\begin{smathpar}
-  \begin{array}{lcl}
-    \phi & = & \forall \nu.~g(\nu) \Leftrightarrow \nu=1 \vee \nu=2
-  \end{array}
-\end{smathpar}
+constraint, $\phi\in\Prop^1 = \forall \nu.~g(\nu) \Leftrightarrow
+\nu=1 \vee \nu=2$, defining the uninterpreted predicate $g$.
 The translation adopts one or the other approach, depending on the

### 2017-11-08 motivation.tex
Edits to fit to 31 pages.

@@
 \label{sec:motivation}
+
+We present our ideas in the context of a DSL embedded in OCaml that
+manages an abstract database state that can be manipulated via a
+well-defined \C{SQL} interface. Arbitrary database computations can be
+built around this interface, which can then be run as transactions
+using the \C{atomically\_do} combinator provided by the DSL.
 
@@
 
-We present our ideas in the context of a DSL embedded in OCaml that
-manages an abstract database state that can be manipulated via a
-well-defined \C{SQL} interface. Arbitrary database computations can be
-built around this interface, which can then be run as transactions
-using the \C{atomically\_do} combinator provided by the DSL.
-
 Fig.~\ref{fig:new_order_code} shows a simplified version of the TPC-C
@@
 
-%%%SJ: Reviewers may not understand what primary and foreign keys are,
-%%%or why they are important.  There are also some fields (e.g., ol_i_id)
-%%%that are not described either in the caption or the text.
+The \C{new\_order} transaction inserts a new \C{Order} record, whose
+id is the sequence number of the next order under the given district
+(\C{d\_id}). The sequence number is stored in the corresponding
+\C{District} record, and updated each time a new order is added to the
+system. Since each order may request multiple items (\C{item\_reqs}),
+an \C{Order\_line} record is created for each requested item to relate
+the order with the item. Each item has a corresponding record in the
+\C{Stock} table, which keeps track of the quantity of the item left in
+stock (\C{s\_qty}). The quantity is updated by the transaction to
+reflect the processing of new orders (if the stock quantity falls below
+10, it is automatically replenished by 91).
 
-% \begin{figure}[!t]
-% \includegraphics[scale=0.5]{Figures/schema}
-% \caption{Database schema of TPC-C's order management system.
-%   Columns against highlighted background are primary keys. Arrows denote
-%   foreign key relationships.}
-% \label{fig:schema}
-% \end{figure}
-
-\begin{figure}[t]
+\begin{figure}[!t]
   \centering
@@
 
-The \C{new\_order} transaction inserts a new \C{Order} record, whose
-id is the sequence number of the next order under the given district
-(\C{d\_id}). The sequence number is stored in the corresponding
-\C{District} record, and updated each time a new order is added to the
-system. Since each order may request multiple items (\C{item\_reqs}),
-an \C{Order\_line} record is created for each requested item to relate
-the order with the item. Each item has a corresponding record in the
-\C{Stock} table, which keeps track of the quantity of the item left in
-stock (\C{s\_qty}). The quantity is updated by the transaction to
-reflect the processing of new orders (if the stock quantity falls below
-10, it is automatically replenished by 91).
-
 TPC-C defines multiple invariants, called \emph{consistency
@@
 
+Because the aforementioned reasoning is reasonably simple to perform
+manually, verifying the soundness of TPC-C's consistency conditions
+would appear to be feasible.  Serializability aids the tractability of
+verification by preventing any interference among concurrently
+executing transactions while the \C{new\_order} transaction executes,
+essentially yielding serial behaviors.  Under weak
+isolation\footnote{Weak isolation does not violate atomicity as long
+  as the witnessed effects are those of committed transactions},
+however, interferences of various kinds are permitted, leading to
+executions superficially similar to executions permitted by concurrent
+(racy) programs~\cite{GHE15,HPQ+15}. 
 \begin{wrapfigure}{l}{.4\textwidth}
@@
 \end{wrapfigure}
-Because the aforementioned reasoning is reasonably simple to perform
-manually, verifying the soundness of TPC-C's consistency conditions
-would appear to be feasible.  Serializability aids the tractability of
-verification by preventing any interference among concurrently
-executing transactions while the \C{new\_order} transaction executes,
-essentially yielding serial behaviors.  Under weak
-isolation\footnote{Weak isolation does not violate atomicity as long
-  as the witnessed effects are those of committed transactions},
-however, interferences of various kinds are permitted, leading to
-executions superficially similar to executions permitted by concurrent
-(racy) programs~\cite{GHE15,HPQ+15}.  To illustrate, consider the
+To illustrate, consider the
 behavior of the \C{new\_order} transaction when executed under a

### 2017-11-08 motivation.tex
CCS and other things.

@@
 executing transactions while the \C{new\_order} transaction executes,
-essentially yielding serial behaviors.  Under weak
-isolation\footnote{Weak isolation does not violate atomicity as long
-  as the witnessed effects are those of committed transactions},
-however, interferences of various kinds are permitted, leading to
-executions superficially similar to executions permitted by concurrent
-(racy) programs~\cite{GHE15,HPQ+15}. 
+essentially yielding serial behaviors.  
 \begin{wrapfigure}{l}{.4\textwidth}
@@
 \label{fig:new_order_exec}
+\vspace*{-10pt}
 \end{wrapfigure}
-To illustrate, consider the
-behavior of the \C{new\_order} transaction when executed under a
-\emph{Read Committed} (RC) isolation level, the default isolation
-level in 8 of the 18 databases studied in~\cite{bailishotos}.
-An executing RC transaction is isolated from \emph{dirty writes},
-i.e., writes of uncommitted transactions, but is allowed to witness
-the writes of concurrent transactions as soon as they are committed.
-Thus, with two concurrent instances of the \C{new\_order} transaction
-(call them $T_1$ and $T_2$), both concurrently placing new orders for
-different customers under the same district (\C{d\_id}), RC isolation
-allows the execution shown in Fig.~\ref{fig:new_order_exec}.
+Under weak isolation\footnote{Weak isolation does not violate
+  atomicity as long as the witnessed effects are those of committed 
+transactions}, however, interferences of various kinds are permitted,
+leading to executions superficially similar to executions permitted by
+concurrent (racy) programs~\cite{GHE15,HPQ+15}.  To illustrate,
+consider the behavior of the \C{new\_order} transaction when executed
+under a \emph{Read Committed} (RC) isolation level, the default
+isolation level in 8 of the 18 databases studied
+in~\cite{bailishotos}.  An executing RC transaction is isolated from
+\emph{dirty writes}, i.e., writes of uncommitted transactions, but is
+allowed to witness the writes of concurrent transactions as soon as
+they are committed.  Thus, with two concurrent instances of the
+\C{new\_order} transaction (call them $T_1$ and $T_2$), both
+concurrently placing new orders for different customers under the same
+district (\C{d\_id}), RC isolation allows the execution shown in
+Fig.~\ref{fig:new_order_exec}.
 

### 2017-11-08 motivation.tex
minor

@@
 ($\cup$). Angle braces ($\langle \ldots \rangle$) are used to denote
-records in math}:
+records.}:
 \begin{smathpar}
@@
 local database state $\delta$.\footnote{For now, assume that the
-  record being added to $\stl$ is not already present in $\stl$.}
+  record being added is not already present in $\stl$.}
 

### 2017-11-08 introduction.tex
Mahsa's suggestions.

@@
 Fekete \emph{et al.}~\cite{feketevldb09} devised and experimented with
-a microbenchmark suite that executes transactions under a
-weakly-isolated \emph{read committed} isolation level - the default
-level for 8 of the 18 databases studied in~\cite{bailishotos}, and
-found that 25 out of every 1000 rows in the database violate at least
-one integrity constraint. Bailis \emph{et al.}~\cite{bailisferal} rely
-on Rails' \emph{uniqueness validation} to maintain uniqueness of
-records while serving Linkbench's~\cite{linkbench} insertion workload
-(6400 records distributed over 1000 keys; 64 concurrent clients), and
-report discovering more than 10 duplicate records.  Rails relies on
-database transactions to validate uniqueness during insertions, which
-is sensible if transactions are serializable, but incorrect under the
+a microbenchmark suite that executes transactions under \emph{Read
+Committed} weak isolation level - default level for 8 of the 18
+databases studied in~\cite{bailishotos}, and found that 25 out of
+every 1000 rows in the database violate at least one integrity
+constraint. Bailis \emph{et al.}~\cite{bailisferal} rely on Rails'
+\emph{uniqueness validation} to maintain uniqueness of records while
+serving Linkbench's~\cite{linkbench} insertion workload (6400 records
+distributed over 1000 keys; 64 concurrent clients), and report
+discovering more than 10 duplicate records.  Rails relies on database
+transactions to validate uniqueness during insertions, which is
+sensible if transactions are serializable, but incorrect under the
 weak isolation level used in the experiments. The same study has found
 that 13\% of all invariants among 67 open source Ruby-on-Rails
-applications are at risk of being violated due to weak
-isolation. Indeed, incidents of safety violations due to executing
-applications in a weakly-isolated environment have been reported on
-web services in production~\cite{starbucksbug, scimedbug}, including
-in safety-critical applications such as bitcoin
+applications are at risk of being violated due to weak isolation.
+Indeed, incidents of safety violations due to executing applications
+in a weakly-isolated environment have been reported on web services in
+production~\cite{starbucksbug, scimedbug}, including in
+safety-critical applications such as bitcoin
 exchanges~\cite{poloniexbug, bitcoinbug}. While enforcing
@@
 question facing all database programmers.\footnote{This position has
-been echoed by database researchers who lament the lack of a
-better understanding of this problem; see e.g., \url{
+been echoed by database researchers who lament the lack of a better
+understanding of this problem; see e.g., \url{
   http://www.bailis.org/blog/understanding-weak-isolation-is-a-serious-problem}.}

### 2017-11-08 motivation.tex
Mahsa's suggestions.

--- 2017-11-08 motivation.tex
@@
 This example does not exhibit any of the anomalies that
-\emph{characterize} RC isolation~\cite{berenson}\footnote{Berenson et
-al characterize isolation levels in terms of the
-anomalies they \emph{admit}. For example, RC is characterized by
-\emph{lost writes} because it admits the anomaly.}. For instance,
-there are no \emph{lost writes} since both concurrent transactions' writes
-are present in the final state of the database.  Program analyses that
-aim to determine appropriate isolation by checking for possible
+\emph{characterize} RC isolation~\cite{berenson}\footnote{Berenson
+\emph{et al.} characterize isolation levels in terms of the anomalies
+they \emph{admit}. For example, RC is characterized by \emph{lost
+writes} because it admits the anomaly.}. For instance, there are no
+\emph{lost writes} since both concurrent transactions' writes are
+present in the final state of the database.  Program analyses that aim
+to determine appropriate isolation by checking for possible
 manifestations of RC-induced anomalies would fail to identify grounds
@@
 to utilize concurrent program verification methods where the
-implementation details. of weak isolation are interleaved within the
-program, yielding a (more-or-less) conventional concurrent program;
-but, considering the size and complexity of real-world transaction
+implementation details of weak isolation are interleaved within the
+program, yielding a (more-or-less) conventional concurrent program.
+But, considering the size and complexity of real-world transaction
 systems, this strategy is unlikely to scale.
@@
 be substituted with higher-order combinators that automatically lift
-the state transformer of its higher-order argument (i.e., the loop
-body) to the state transformer of the combined expression (i.e., the
-loop).  We illustrate this intuition on a simple example.
+the state transformer of its higher-order argument, i.e., the loop
+body, to the state transformer of the combined expression, i.e., the
+loop.  We illustrate this intuition on a simple example.
 
@@
 from Fig.~\ref{fig:new_order_code}. Some irrelevant expressions have
-been replaced with constants (\C{k1},\C{k2}, and \C{k3}).  The body of
+been replaced with constants (\C{k1}, \C{k2}, and \C{k3}).  The body of
 the loop executes a SQL update followed by an insert.  Recall that a

### 2017-11-08 opsem.tex
Minor edits.
Mahsa's suggestions.

@@
 \txnimp, a core language that we use to formalize our intuitions about
-reasoning under weak isolation. Variables ($x$), integer and Boolean
+reasoning under weak isolation. Variables ($x$), Integer and Boolean
 constants ($k$), records ($r$) of named constants, sets ($s$) of such

### 2017-11-08 inference.tex
Typos.

@@
   \item All function symbols, modulo those that are drawn from
-    $\Prop^0$ and $\Prop^1$, are uninterpreted and Boolean.
+    $\Prop^0$ and $\Prop^1$, are uninterpreted and boolean.
   \item All quantification is first-order; second-order objects, such
@@
 top-level bind expression, hence its prenex signature is
-$\forall^2\exists$.  Since $\existsl$ doesn't extend $\nubar$, the
+$\forall^2\exists$.  Since $\existsl$ does not extend $\nubar$, the
 prenex signature of $s$ remains $\forall^2\exists$. When $s$ is an

### 2017-11-08 opsem.tex
Typos.

@@
 \txnimp, a core language that we use to formalize our intuitions about
-reasoning under weak isolation. Variables ($x$), Integer and Boolean
+reasoning under weak isolation. Variables ($x$), integer and boolean
 constants ($k$), records ($r$) of named constants, sets ($s$) of such
@@
 the records in $\stg$ that match the search criteria defined by its
-Boolean function argument, and adds the records to $\stl$ after
+boolean function argument, and adds the records to $\stl$ after
 marking them for deletion. \C{SELECT} bounds the name introduced by
@@
 \I_e\,\,(\stl,\stg,\stg') & = & \stg' = \stg\\
-\I_c\,\,(\stl,\stg,\stg') & = & \forall(r\in\stl)(r'\in\stg).~ r'.\idf = r.\idf \Rightarrow r'\in\stg'.
+\I_c\,\,(\stl,\stg,\stg') & = & \forall(r\in\stl)(r'\in\stg).~ r'.\idf = r.\idf \Rightarrow r'\in\stg'
 \end{array}

### 2017-11-08 reasoning.tex
Formatting fixed.
Typos.

@@
 \begin{array}{lcl}
-\stable(R, \phi) & \Leftrightarrow & \forall \stl, \stg, \stg'. \phi(\stl, \stg) \wedge R^{*}(\stg, \stg') \wedge \I(\stl, \stg, \stg') \Rightarrow \phi(\stl, \stg')
+\stable(R, \phi) & \Leftrightarrow & \forall \stl, \stg, \stg'. \phi(\stl, \stg) 
+  \wedge R^{*}(\stg, \stg') \wedge \I(\stl, \stg, \stg') \Rightarrow \phi(\stl, \stg')
 \end{array}
@@
 %to $\stg'$.
-
+\raggedbottom
 For the guarantee $G$ of a transaction to be valid, it must follow
@@
 \R \vdash \hoare{P}{c}{Q} & \defeq & \forall
-  \stl,\stl',\stg,\stg',\nubar.~ P(\stl,\stg) \conj (\tbox{c}_i,\stl,\stg) \rstepssto{n}
+  \stl,\stl',\stg,\stg'.~ P(\stl,\stg) \conj (\tbox{c}_i,\stl,\stg) \rstepssto{n}
   (\tbox{\cskip}_i, \stl',\stg')
-  \Rightarrow Q(\stl',\stg') & \nubar = \texttt{freevars}(c)\\
+  \Rightarrow Q(\stl',\stg') &\\
 \rg{I,R}{c}{G,I} & \defeq &  \forall \stg.\, I(\stg)
@@
     \rstepssto{n} (\cskip,\stg') \Rightarrow I(\stg')) \\
-&&\hspace*{0.6in}\conj \texttt{step-guaranteed}(R,G,c,\stg)\\
+&&\hspace*{0.6in}\conj \texttt{txn-guaranteed}(R,G,c,\stg)\\
 \end{array}
@@
 \noindent The
-$\texttt{step-guaranteed}$ predicate used in the semantics of the
+$\texttt{txn-guaranteed}$ predicate used in the semantics of the
 global RG judgment is defined below:\vspace*{-10pt}
@@
 \begin{array}{lcl}
-\texttt{step-guaranteed}(R,G,c,\stg) &\defeq& \forall c',c''\stg',\stg''.
+\texttt{txn-guaranteed}(R,G,c,\stg) &\defeq& \forall c',c''\stg',\stg''.
 (c,\stg) \rstepssto{n} (c',\stg') \conj (c',\stg') \stepsto
